### 2015-07-17 manuscript/manuscript.Rnw
add start of manuscript draft

\documentclass[english,12pt]{article}
\usepackage{fontspec}
\setmainfont{FreeSerif}
\setsansfont{FreeSans}
\setmonofont{FreeMono}
\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
\usepackage{url}
\usepackage{fancyhdr}
\usepackage{graphicx}
\usepackage[bf]{caption}
\usepackage{rotating}
\usepackage{wrapfig}
\usepackage{booktabs}
\usepackage{float}
\usepackage{multirow}
\usepackage{setspace}
%\usepackage{txfonts}
%\usepackage{fancyref}
\usepackage{acronym}
\usepackage{array}
\usepackage{dcolumn}
\usepackage{adjustbox}
\usepackage{longtable}
\usepackage{pdflscape}
\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
\addbibresource{references.bib}
\usepackage[noblocks,auth-sc]{authblk}
\usepackage[hyperfigures,bookmarks,colorlinks]{hyperref}
%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
% Floats at end for journals
% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
\usepackage[nomargin,inline,draft]{fixme}
\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
\usepackage[x11names,svgnames]{xcolor}
\usepackage{texshade}
\usepackage{tikz}
\usepackage{nameref}
\usepackage{zref-xr,zref-user}
\newenvironment{narrow}[2]{%
  \begin{list}{}{%
      \setlength{\topsep}{0pt}%
      \setlength{\leftmargin}{#1}%
      \setlength{\rightmargin}{#2}%
      \setlength{\listparindent}{\parindent}%
      \setlength{\itemindent}{\parindent}%
      \setlength{\parsep}{\parskip}}%
  \item[]}{\end{list}}
\def\newblock{\hskip}
\newenvironment{paperquote}{%
  \begin{quote}%
     \it
  }%
  {\end{quote}}
\renewcommand{\textfraction}{0.15}
\renewcommand{\topfraction}{0.85}
\renewcommand{\bottomfraction}{0.65}
\renewcommand{\floatpagefraction}{0.60}
%\renewcommand{\baselinestretch}{1.8}
\newenvironment{enumerate*}%
  {\begin{enumerate}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{enumerate}}
\newenvironment{itemize*}%
  {\begin{itemize}%
    \setlength{\itemsep}{0pt}%
    \setlength{\parskip}{0pt}}%
  {\end{itemize}}
\oddsidemargin 0.0in 
\textwidth 6.5in
\raggedbottom
\clubpenalty = 10000
\widowpenalty = 10000
% make the afill smaller
\renewcommand\Affilfont{\itshape\small}
\pagestyle{fancy}

\begin{document}
\author[1,2]{Michael R. McGowen}
\author[1]{Don Armstrong}
\author[1]{Amy Weckle}
\author[1]{Jason Caravas}
\author[3]{Dalen Agnew}
\author[4]{Kurt Benirschke}
\author[5]{Sue Savage-Rumbaugh}
\author[6,7]{Chong Jai Kim}
\author[8]{Günter P. Wagner}
\author[6]{Roberto Romero}
\author[1,6,9,*]{Derek E. Wildman}
\title{The evolution of gene expression in the term placenta of viviparous mammals}
% These are all wrong, but we'll change them later.
\affil[1]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
\affil[2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
\affil[3]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
\affil[4]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
\affil[5]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
\affil[6]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
\affil[7]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
\affil[8]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
\affil[9]{Department of Obstetrics and Gynecology, Wayne State University School of Medicine, Detroit, MI, USA}
\affil[*]{To whom correspondence should be addressed}
\date{\today}

% \twocolumn
\maketitle
\onehalfspacing

<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
opts_chunk$set(cache=TRUE, autodep=TRUE)
options(device = function(file, width = 8, height = 7, ...) {
  cairo_pdf(tempfile(), width = width, height = height, ...)
})
library("data.table")
library("ggplot2")
library("reshape2")
@

<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R")))>>=
for (a in paste0("../common_r_code/",c("to_latex.R"))) {
    source(a)
}
@

\section{Abstract}

The placenta is arguably the most anatomically variable organ in
mammals. Understanding gene expression patterns among placentas from
multiple species can illuminate the molecular basis of placenta
function as well as potentially enrich our knowledge of the biology of
obstetrical syndromes. We profiled the term placenta by using Illumina
RNA-Seq technology to sequence the transcriptomes of 6 species of
therian (i.e. marsupials and “placental”) mammals. We identified
significant shifts in gene expression along specific evolutionary
lineages using a model of character evolution. Genes that showed
significant elevation in expression along the human lineage
($\mathrm{FDR}\le 0.01$) were enriched in GO categories such as
angiogenesis, tube morphogenesis, cell migration, and regulation of
inflammatory response. Three of these genes (KISS1, ADAM12, GPC3)
showed exceptionally high expression in humans; these genes are
implicated in preeclampsia and may indicate that the risk of
preeclampsia recently emerged on the human lineage.

Keywords: mammal, gene expression, evolution, RNA-seq, placenta

\section{Introduction}

The main role of the placenta in viviparous mammals is the transfer of
nutrients, oxygen, and waste between mother and offspring. Although
these functions are conserved among mammals, the placenta is perhaps
the most variable organ in terms of structure and form (Mossman,
1937). Species differ in the shape of attachment (i.e., discoid,
diffuse, cotyledonary, zonary), degree of invasiveness of fetal
tissues into maternal tissues (i.e. epitheliochorial,
endotheliochorial, hemochorial), and degree of interdigitation of
maternal-fetal tissue (i.e., villous, folded, trabecular, lamellar,
labrynthine) (Mossman, 1987). In addition, therian mammals differ
drastically in the number and size of their offspring, as well as the
length of gestation, traits often associated with specific aspects of
placental morphology (Capellini et al., 2011).

The placenta arose at some point before the origin of therian mammals,
a group that includes both marsupials and eutherian or “placental”
mammals. The term “placental mammals” is misleading, as marsupials
also possess a placenta, although its direct apposition to maternal
tissues lasts for a proportionally shorter time during gestation
(Renfree, 2010). Marsupials also differ from eutherian mammals in
that most possess only a choriovitelline placenta, produced from the
fusion of the chorion and yolk sac, during most of gestation. Although
some eutherian mammals may possess a choriovitelline placenta at some
point in development, the major site of nutrient exchange in
placentals occurs via the chorioallantoic placenta, produced from the
fusion of the chorion and allantois (Mossman, 1987). It is unclear
whether the first placenta was more like that of marsupials or
eutherians; however, using phylogenetic methods, the ancestral
eutherian was reconstructed as possessing a relatively invasive
hemochorial placenta, in which the maternal and fetal blood supply is
separated by one tissue layer, as in humans and mouse (Elliot
and Crespi, 2009; Wildman et al., 2006).

Although the morphology of the mammalian placenta has been well
characterized in a number of species (Mossman, 1987), the diversity of
the molecular environment is only beginning to be understood from a
genetic perspective (Carter, 2012). Studies of certain model species
(i.e., humans, mice, cattle) have revealed a stunning divergence in
the genetics of the placenta, especially in later stages of pregnancy
(Carter, 2012; Hou et al., 2012; Knox and Baker, 2008; Su et al.,
2002). For example, each lineage has evolved expansive
placenta-specific gene families with expression patterns that can vary
in space and time (Green et al., 2000; Simmons et al., 2008). These
include independent expansions of the prolactins in mouse and cow,
pregnancy-specific glycoproteins (PSGs) in humans and mouse,
pregnancy-associated glycoproteins (PAGs) in cow, and the expansion of
the growth hormone and chorionic gonadotropin beta genes in hominoid
primates (Haig, 2008; Rawn and Cross, 2008). Expression patterns of
specific genes also can vary widely between species. For example, the
expression of orthologous genes in the placenta have been found to
vary substantially between humans and mice (Su et al., 2002), and the
recruitment of SIGLEC6 to placental expression in humans differs from
both chimpanzees and mice (Brinkman-Van der Linden et al., 2007).

One explanation for the divergent paths of the placental interface is
the distinct evolutionary interplay between maternal and fetal
strategies during pregnancy (Haig, 1993). The maternal strategy is to
requisition an adequate degree of nutrients to the fetus without
risking the health of the mother, while the fetal strategy via
paternal genes is to extract as many resources from the mother as
possible. Therefore, an interplay between these extremes can set up an
intricate “arms race”, involving the imprinting of parental alleles,
production of new genes, and enlistment of existing genes to
expression at the maternal-fetal interface (Haig, 1993).

One symptom
of this maternal-fetal conflict is its potential affect on human
health (Hollegaard et al., 2013). In addition to having an invasive
hemochorial placenta, humans possess a deeply invasive trophoblast
layer that can penetrate the myometrium of the uterus (Carter and
Pijnenborg, 2011). Defects in placental growth are known to be
associated with the devastating condition of preeclampsia, which can
cause hypertension, proteinuria, and has the potential to lead to
death of the mother and/or infant (citation). Although cases have been
documented in chimpanzees and gorillas (citation), preeclampsia has a
far higher incidence in human pregnancies with as much as 4\% of
pregnancies being affected in the United States (citation).
Chimpanzees and gorillas also have deeply invasive trophoblasts,
although the progression of invasion in great apes may be slower than
humans (Carter and Pijnenborg, 2011); however, large-scale expression
studies of great ape placentas have not been conducted. The increasing
affordability of next-generation sequencing technology has enabled the
large-scale study of specific tissues from a wide array of species to
address broader phylogenetic questions in the evolution of gene
expression (Brawand et al., 2011). Here we use Illumina RNA-Seq
technology to sequence and quantify the mRNA environment of six
mammalian placentas at term, including that of the human, bonobo (the
sister species to the chimpanzee), dog, cow, African elephant, and
opossum, representing marsupials as well as three major branches of
eutherian mammals (Euarchontoglires, Laurasiatheria, Afrotheria).
These placentas also represent a wide array of placental morphologies
as well as traits such as number and size of offspring and length of
gestation (Table 1). We use this data, along with existing RNA-Seq
data from mouse (Wang et al., 2011), to address the major differences
between expression in marsupial and eutherian mammals at term,
identify the evolution of expression in important placental genes such
as hormones and galectins, identify novel lineage-specific expression
patterns, and quantify human specific differences in gene expression
that may be related to obstetrical disorders such as preeclampsia.

\section{Methods}

\subsection{Collection of Placental Tissue and RNA Extraction}

Fetal tissue was collected from the placenta of six mammalian species
(Homo sapiens [human], Pan paniscus [bonobo], Canis familiaris
[domestic dog], Bos taurus [domestic cow], Loxodonta africana [African
savanna elephant], and Monodelphis domestica [gray short-tailed
opossum]). Information on specimen ID, place and time of collection,
and collector is shown for each sample in Supplemental Table 1. In the
case of Pan, Bos, and Loxodonta, the placenta was collected at term
upon expulsion from the mother following delivery. Placental tissue
from Homo sapiens was collected upon delivery via a caesarian section.
Near-term placental tissue from Canis familiaris was collected upon
the death of the mother from a car accident. Placental tissue from
Monodelphis domestica was sampled at fetal stage 32, roughly one day
prior to birth. Only the fetal portion of the placenta was sampled at
the maternal-fetal interface (i.e. villous portion for Homo, Pan,
Canis, and Loxodonta; cotyledon for Bos; diffuse placenta for
Monodelphis). All tissues were collected and stored in RNAlater (Life
Technologies, Carlsbad, CA) until RNA isolation could be performed.
Tissues were homogenized in Trizol, and total RNA was isolated
following the manufacturer’s instructions. RNA was further purified
using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction with the
RNase-Free DNase Set (Qiagen, Valencia, CA), following the
manufacturer’s protocol to remove any contaminating DNA from the
sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
Clara, CA) to confirm concentration and quality of the isolation. RNA
quality was assessed using A260/A280 and A260/A230 absorbance values.

\subsection{Sequencing}

All RNA-Seq mRNA isolation, library construction, and Illumina
sequencing was performed by the Applied Genomics Technology Center
(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
total RNA and standard Illumina protocol. Paired-end sequencing was
performed using the Illumina Genome Analyzer II with an insert size of
230 base pairs (bp). Bos, Loxodonta, and Monodelphis were sequenced
with read lengths of 75 bp; Canis, Homo, and Pan were sequenced with
read lengths of 150 bp.

\subsection{Transcriptome Assembly and Mapping}

The quality of reads was evaluated using FASTQC v. 0.9.2
(http://www.bioinformatics.babraham.ac.uk/projects/fastqc/). Following
this, reads were trimmed based on quality in Galaxy using the FASTQ
Trimmer v. 1.0.0 (Blankenberg et al., 2010). All transcriptomes were
then aligned to reference genomes and assembled using the Tuxedo suite
of next-generation sequencing analysis tools. Annotated reference
genomes were downloaded from Ensembl v. 64 (Flicek et al., 2012),
including Homo sapiens GRCh37, Pan troglodytes CHIMP2.1 (6X coverage),
Canis familiaris BROADD2 (7.6X), Bos taurus UMD3.1 (9.5X), Loxodonta
africana loxAfr3 (7X), and Monodelphis domestica BROADO5 (6.8X). The
genome of the closely-related Pan troglodytes was used to map the
transcriptome of Pan paniscus; the separation between the two
recognized Pan species is generally perceived to be ~1 Mya (Prufer et
al., 2012). Reads were aligned to reference genomes using Bowtie v.
0.12.7 (Langmead et al., 2009) and TopHat v. 1.3.3 (Trapnell et al.,
2009). Cufflinks v. 1.2.1 was then used to assemble transcripts and
quantify expression values using fragments per kilobase of exon per
million mapped reads (FPKM) (Trapnell et al., 2010). RNA-Seq metrics
were computed using Picard (http://picard.sourceforge.net/). For
comparison, we also downloaded the placenta Illumina RNA-Seq
transcriptome of a mouse from (Wang et al., 2011). Raw reads were
mapped as above using the Ensembl v. 64 annotation of the Mus musculus
(house mouse) genome (NCBI m37).

\subsection{Exploration of Individual Transcriptome Data}

To determine enrichment levels of functional categories of genes in
highly-expressed transcripts in each transcriptome, we performed
functional annotation cluster analyses per transcriptome using DAVID
(Huang et al., 2009; Sherman et al., 2007). DAVID ranks functional
annotation clusters according to enrichment score (ES); functional
annotation clusters are considered significantly enriched if ES>1.3,
as this corresponds to p<0.05 (Huang et al., 2009). We executed
analyses with two gene lists per transcriptome: 1) annotated genes
with expression levels of the top 5\% of the frequency distribution of
FPKM and 2) the top 100 genes with the highest FPKM. Due to the
paucity of annotated gene ontology terms for some species, gene lists
were compared using gene ontology terms derived from the human genome.

\subsection{Comparative Analysis of Gene Expression}

All 1:1 orthologues present among human, bonobo, mouse, dog, cow,
elephant, and opossum were determined using Ensembl v.64 gene trees
via the Biomart online data management tool (Kasprzyk, 2011). A total
of 6882 1:1 orthologous genes were expressed in at least one species.
For further analysis, we eliminated all genes that did not have at
least one species with an FPKM>10. A list of these 4259 1:1
orthologous genes was retained for future analyses of gene expression
evolution described below.

We used the software OUCH (Ornstein–Uhlenbeck for Comparative
Hypotheses) in the R statistical environment to investigate
significant shifts in gene expression levels over time (Butler and
King, 2004). OUCH institutes an Ornstein-Uhlenbeck (OU) model of
evolution, which allows for different optima for quantitative trait
values along user-specified branches of a phylogenetic tree. The OU
model explicitly accounts for change in quantitative trait values by
introducing two parameters into the model of stochastic Brownian
motion of change in trait value along a phylogenetic tree through
time. These parameters include, ?, or an optimum trait value, and ,
or the strength of selection (interpreted as “pull” toward the
optimum). In addition, the OU model can also integrate evolutionary
change by assuming that each user-specified set of braches has its own
separate OU parameters (see Butler and King, 2004).

Various OU branch models were tested against a model in which no
variation in optima among branches was allowed. The following branches
were specified in separate models to have a different optimum than all
other branches: human, bonobo, mouse, dog, cow, elephant, opossum,
human+bonobo, Euarchontoglires, and Boreoeutheria. Likelihood values
were computed for each branch model and compared to an OU model with
one optimum using likelihood ratio tests with df=1. We filtered out
genes with universally low expression levels, and included only genes
in which expression in at least one species had an FPKM≥10 as
described above. We corrected for multiple testing using the false
discovery rate (FDR) and calculated adjusted p values (q-values) using
the Benjamini-Hochstein correction. We used the time-calibrated
phylogenetic tree of (Meredith et al., 2011) with all species not used
in this analyses pruned.

To determine enrichment of functional categories in lists of genes
that display a significant shift in expression on a specific branch,
we performed functional annotation cluster analyses on each branch
using DAVID (Huang et al., 2009; Sherman et al., 2007). DAVID ranks
functional annotation clusters according to enrichment score (ES);
functional annotation clusters are considered significantly enriched
if ES>1.3, as this corresponds to p<0.05 (Huang et al., 2009). As
above, due to the paucity of annotated gene ontology terms for some
species, gene lists were compared using gene ontology terms derived
from the human genome.

\subsection{Correlation of Tissue Specific Index with Lineage Specific Genes}

To evaluate the extent of tissue specificity of each gene, we
calculated the tissue specificity index based on human microarray data
from (Su et al., 2004) available via BioGPS. The tissue specificity
index (TSI) measures the extent to which a specific gene has
predominant expression in only one tissue (Yanai et al., 2005). A TSI
of 1.0 would signify that a gene was solely expressed in only one
tissue, while a TSI of 0.0 would signify that a gene had equal
expression in all tissues evaluated. We used the formula available in
Yanai et al. (2005) to calculate the TSI, and identified the top four
tissues which expressed each probe set from (Su et al., 2004).

\subsection{Validation of Primate Expression using Spider Monkey Placenta Transcriptome}

To verify expression levels in human and bonobo, we also generated
RNA-Seq data for the placenta transcriptome of the black-headed spider
monkey (Ateles fusciceps) using similar protocols discussed above.
Ateles is outside the clade, which includes hominids and can be used
to polarize expression values of highly expressed genes in Homo and
Pan. We sampled placental tissue from an Ateles placenta expelled from
the mother shortly after birth (see Supplemental Table X). We used
similar procedures discussed above to extract and sequence this
sample. [DEREK—I’m unsure how to proceed on describing this so as to
not step on Natalie’s toes]

\section{Results}

\subsection{Transcriptome Statistics}

For transcriptomes sequenced with 75 bp, (opossum, elephant, and cow),
the number of reads ranged from 3.61x107 to 3.70x107 and number of
bases ranged from 1.43x109 to 1.64x109 with at least 96\% of reads
mapping to the respective reference genome (Table 1). For
transcriptomes sequenced with 150 bp (human, bonobo, and dog), number
of reads ranged from 6.47x107 to 9.27x107 and number of bases ranged
from 5.20x109 to 6.25x109 with at least 99\% of reads mapping to the
respective reference genome. All reads were submitted to the NCBI
Sequence Read Archive (SRA) (Accession Number XXX-YYY).

The percentage of bases that mapped to Ensembl annotated mRNA ranged
from ~35\% in elephant to ~85\% in human, with percentages of bases
mapping to annotated protein-coding segments of genes ranging from
~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
ranged from ~43\% in elephant to 80\% in bonobo. The number of
annotated transcripts to which reads mapped in each respective genomes
ranged from 4119 (bonobo) to 6506 (human) for transcripts with FPKM>10
and 11689 (dog) to 30733 (human) for transcripts with FPKM>1. The
greater degree of mapping to annotated protein-coding sequences in
humans corresponds to the greater number of Ensembl annotated
sequences in the human genome compared to the other five reference
genomes (Flicek, 2012).

The top 50 expressed genes in each genome analyzed are shown in Tables
2-7, and all genes for each transcriptome are listed in Supplementary
Tables X-XX [These are large Excel files to be included with the
Supplemental Material]. In addition, specific genes including
placental hormones and hormone-related genes, galectins, and
insulin-like growth factor related genes are shown with FPKM values
side to side for all transcriptomes including the reanalysis of mouse
from Wang et al. (2011) in Table 8. Below we describe each
transcriptome separately, specifically concentrating on
highly-expressed genes.

\subsection{Human Placenta Transcriptome}

The top 50 expressed genes as measured by FPKM in the term placenta of
the human are shown in Table 2 with a list of all genes in
Supplemental Table X. The gene with by far the highest FPKM value
(101962) in humans is chorionic somatomammotropin hormone 1 (CSH1),
followed by chorionic somatomammotropin hormone 2 (CSH2), and
chorionic somatomammotropin hormone-like 1 (CSHL1), all three of which
are members of a distinct expansion of the growth hormone gene family
within hominoids. Another member of the growth hormone gene family,
GH2 (FPKM=729.639), is also expressed to a high degree in the human
placenta (Table 2). Ten pregnancy-specific beta-1-glycoprotein genes
are highly-expressed (FPKM>100) in the term placenta of humans, with
PSG4 showing the greatest degree of expression (FPKM= 5201.31; Table
2). These genes are members of a hominoid-specific expansion of the
carcinoembryonic antigen (CEA) family and known to be expressed by the
syncytiotrophoblast (Lei et al., 1992; Rand and Cross, 2008). Genes
involved in the production of estrogen (CYP19A1) and progesterone
(HSD3B1) show exceptionally high expression in the human placenta
(FPKM = 4817.15 and 829.846 respectively), as well as CGA (FPKM=
4793.08), which generates the alpha subnit of multiple human chorionic
gonadotropin hormones (Table 2). The genes producing the beta subunits
of the chorionic gonadotropins (CGB, CGB1, CGB2, CGB5, CGB7, CGB8) are
either not expressed (CGB8) or expressed modestly (FPKM≤20.826; Table
8). Other highly expressed genes in the top 50 include members of the
hemoglobin family (e.g. HBA1, HBB, HBG2, HBA2), the maternally
imprinted long noncoding RNA gene H19, genes coding for proteins
required for growth and/or cell proliferation (ADAM12, KISS1, IGF2,
GPC3, CSF3R), anticoagulation (TFPI2), and glucose transport (SLC2A1)
among others (Table 2, Supplemental Table X)

\subsection{Bonobo Placenta Transcriptome}

The top 50 expressed genes as measured by FPKM in the term placenta of
the bonobo are shown in Table 3 with a list of all genes in
Supplemental Table X. Expression in the bonobo placenta differs
somewhat from that of the human, as evidenced by the makeup of the top
50 genes (Table 3). The highest expression (FPKM=10149.1) in bonobo is
proteoglycan 2, bone marrow (PRG2), which is involved in insulin
growth factor signaling and angiogenesis during pregnancy (Weyer and
Glerup, 2011). PRG2 is also expressed at relatively high levels in the
human placenta (FPKM=250.675; Supplemental Table X), although at
comparatively much lower levels than the bonobo. A bonobo gene closely
related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
expressed at high levels (FPKM=1232.56; Table 3). Two other Pan genes
included within this family (ENSPTRG00000030805 which shows similarity
to GH2 and ENSPTRG00000034560) show expression levels of FPKM>89. Six
pregnancy-specific beta-1-glycoprotein genes were expressed in the
bonobo placenta with FPKM ranging from 20.885-167.961 (Supplemental
Table X). Placental expression of the genes involved in estrogen and
progesterone production (CYP19A1, HSD3B1 respectively) show lower
expression in the bonobo as compared to humans (CYP19A1 FPKM= 48.6877;
HSD3B1 FPKM= 54.5301; Table 8). In addition, CGA is expressed in the
bonobo placenta (FPKM=202.629), but at much lower levels than humans
as well (Table 8). Two Ensembl-annotated beta subunits of the
chorionic gonadotropin genes are expressed in bonobo, one related to
CGB and another related to CGB7; both are not highly expressed (Table
8). Among the other top 50 expressed genes, there is high expression
of ribosomal proteins (e.g., RPL18, RPL23), keratins (e.g., KRT7,
KRT19), heat-shock proteins (e.g., HSPA5, HSPA1A), serpins (SERPINE1,
SERPINE2), insulin growth factor 2 (IGF2) and the hormone
adrenomedullin (ADM).

\subsection{Dog Placenta Transcriptome}

The top 50 expressed genes as measured by FPKM in the term placenta of
the dog are shown in Table 4 with a list of all genes in Supplemental
Table X. The gene with the highest expression (FPKM=48399.9) in the
dog placenta is the hormone relaxin (RLN1). Other highly expressed
genes among the top 50 (Table 4) include genes involved in the
cytoskeleton (TMSB10, TMSB4X), anticoagulation (TFPI2, PLA2G7), innate
immunity (LCN2, GM2A), lipid metabolism (APOE), inhibition of matrix
metalloproteinases (TIMP2, TIMP3), and growth (PGF, EGFR, HTRA1). The
gene LGALS1, which shows evidence of immunosupression during pregnancy
(Than et al., 2008), shows high expression in dog (FPKM= 1405.02). In
addition, the gene SSLP1, a gene expressed in the placentas of cow and
sheep but not present in primates (Ushizawa et al., 2009), is also
expressed at high levels (FPKM= 1853.98). There is no placental
expression of the estrogen synthase gene (CYP19A1) in dogs, however
expression of HSD3B1 is relatively high (FPKM=157.622).

\subsection{Cow Placenta Transcriptome}

The top 50 expressed genes as measured by FPKM in the term placenta of
the dog are shown in Table 5 with a list of all genes in Supplemental
Table X. The gene with the highest expression (FPKM=12488.8) was the
miRNA gene bta-mir-2887-1. This gene occurs in several copies in the
cow genome, and it was unclear as to how many or to which of 12
bta-mir-2887 genes this mapped (Supplemental Table X). A previous
study indicated that this miRNA gene was expressed in the bovine ovary
(Hossain et al., 2009). Several members of the ruminant-specific
pregnancy-associated glycoprotein family show especially high
expression in the cow placenta (Table 4, Supplemental Table X)
including PAG2 (FPKM=7255.34), PAG10 (2076.42), BT.83780 (1793.72),
PAG11 (962.19), PAG12 (469.751), BT.88769 (408.584), PAG1 (301.61),
PAG4 (186.496), PAG6 (157.348), PAG8 (153.225), and PAG16 (103.573).
In addition, 14 other PAGs were expressed with FPKM<40 (Supplemental
Table X). Other genes with high expression include genes involved with
regulation of growth (IGF2, PHLDA2), imprinted genes (GNAS, DLK1,
PHLDA2), and CSH1, a gene that is a member of a ruminant specific
expansion of placental prolactins and not directly related to the gene
of the same name in hominids. Other ruminant-specific placental
prolactins also show high expression in the cow placenta including
PRP1 (632.766), PRP5 (396.832), PRP6 (345.296), PRP2 (195.771), in
addition to 7 other placental prolactins with expression of FPKM<51
(Table 5; Supplemental Table X). SSLP1 (also called SOLD1), a gene
expressed in the cotyledonary villi of the ruminiant placenta and
involved in cell proliferation (Ushizawa et al., 2009, 2010), also
shows evidence of high expression (FPKM=1001.45). There is some
placental expression of the estrogen synthase gene (CYP19A1) in cows
(75.1346), and comparatively greater expression of HSD3B1 (FPKM=
157.622).

\subsection{Elephant Placenta Transcriptome}

The top 50 expressed genes as measured by FPKM in the term placenta of
the elephant are shown in Table 6 with a list of all genes in
Supplemental Table X. The gene with the highest expression
(FPKM=33439.5) was the natriuretic peptide precursor B (NPPB). The
gene with the second highest expression (FPKM=20233) was the microRNA
miR-675, which is imbedded in the first exon of the maternally
imprinted long non-coding RNA gene H19 (Keniry et al., 2012), a gene
also highly expressed in the elephant placenta (FPKM=7164.3). Five
uncharacterized genes are expressed to a high degree in the elephant
placenta (Table 6), three of which are a member of an
elephant-specific expansion of 11 genes closely related to the
thymus-specific serine protease gene PRSS16. The other 8 PRSS16-like
genes are also expressed to a lesser degree (FPKM<284.014). Two other
novel genes with high expression are related to the gene SSLP1 (Table
6), which is also expressed in the cow and dog placentas (Tables 4 and
5). Placental expression of the hormones relaxin (RLN1) and prolactin
(PRL) are exceptionally high in the elephant when compared to other
mammals (Tables 6 and 8). There is little to no placental expression
of the estrogen synthase gene, CYP19A1, (FPKM<1) and the enzyme that
produces progesterone HSD3B1 (FPKM=0). Other genes with high
expression in the elephant placenta include genes involved in
oxidative stress (TXN), growth (IGF2), and PAEP, a gene usually
expressed in the endometrium of the uterus in humans (Rutanen et al.,
1987).

\subsection{Opossum Placenta Transcriptome}

The top 50 expressed genes as measured by FPKM in the term placenta of
the opossum are shown in Table 7 with a list of all genes in
Supplemental Table X. The gene with the highest expression
(FPKM=41740.6) was hemoglobin, epsilon (HBE), which is expressed at
the embryonic stage of development in marsupials, as in other mammals
(Holland and Gooley, 1997). HBE subunits link up with protein products
of two hemoglobin, zeta (HBZ) genes, which are also expressed to a
high degree in the opossum placenta (FPKM=2452.19 and FPKM=1139.25).
In addition, there is high expression (FPKM=5827.37) of the gene
hemoglobin, omega (HBO), a hemoglobin found only in marsupials and
monotremes (Opazo et al., 2008). The hemoglobin gene HBA is also
expressed in the placenta (FPKM=456.978). The gene with the second
highest expression (FPKM=22854.8) is superoxide dismutase 3,
extracellular (SOD3), which functions as an antioxidant. Four cystatin
genes (CSTB, CST3, two CSTA-like genes) are expressed to a high
degree, along with two cathepsin-L related genes (Table 7), which are
known to interact with cystatins (Turk et al., 2001). The
uncharacterized gene XM\_001364916.1 also shows high expression in the
opossum placenta (FPKM=13572.8). The gene WFDC2, a member of the whey
acidic protein (WAP) family is also expressed to a high degree in the
opossum placenta. Other genes with high expression include genes
involved in retinol binding (RBP4), growth (IGFBP3), lipid metabolism
(APOA1, APOA2), and immunosupression (LGALS1). In addition, there was
little production of CYP19A1 and HSD3B1 (FPKM<2), and moderate
expression of relaxin RLN1 (FPKM= 41.3949).

\subsection{DAVID Analyses of Individual Transcriptomes}

In order to identify major functional categories of genes expressed in
the individual transcriptomes, we undertook functional annotation
clustering analyses using DAVID (see Methods). Concentrating on genes
with relatively high expression values, we conducted DAVID analyses
using a list of genes with FPKM values included within the top 5\%
distribution of expressed genes. Functional annotation clusters from
all six placenta transcriptomes show enrichment in similar clusters of
genes (Supplemental Table X). Significant enrichment in genes (ES>1.3,
p>0.05) in the top 5\% of FPKM values in at least five transcriptomes
included clusters such as ribosome, vesicle, cytoskeleton,
glycoprotein, lysosome, annexins, hormone stimulus or response,
growth, calcium binding, cell migration, basement membrane,
antioxidant activity or oxidative stress, cell adhesion, and
mitochondrion among others (Supplemental Table X).

We also investigated enrichment in functional annotation clusters in
the top 100 genes of each transcriptome (Table 9). As in the analyses
of the top 5\% of genes, the top 100 genes show significant enrichment
(ES>1.3) in all six transcriptomes in genes associated with ribosomes.
Human shows enrichment in clusters such as collagen-binding, hormone
activity, female pregnancy, coagulation, and the defense response.
Bonobo shows enrichment in clusters related to apoptosis,
cytoskeleton-related genes, and the stress response. Cow shows only
four significantly enriched clusters; however many genes highly
expressed in the cow are ruminant-specific placenta genes and are not
yet annotated to specific GO categories. Dog shows enrichment in genes
associated with RNA, as well as homeostasis, glycoproteins, lyosome,
and calcium binding. Elephant shows enrichment in genes associated
with RNA, coagulation and calcium binding. Opossum shows enrichment in
genes involved in apoptosis, oxidative stress, mitochondrion, hormone
metabolism, and lipids.

\subsection{Comparative Analysis of Gene Expression}

We performed a comparative analysis of gene expression by
reconstructing the evolution of expression values and detecting
significant shifts in expression along specific lineages. Likelihood
values for each model for each of the X genes reconstructed, as well
as likelihood ratio tests, P- and Q-values (after correcting for
multiple testing) are shown in Supplemental Table X. We identified
between 4 (Boreoeutheria) and 550 (Mouse) significant shifts (Q <
0.01) in expression on specific lineages (Table 10). The top 10 genes
with significant shifts (if applicable) per lineage are shown in Table
10 and all genes for each branch are shown in Supplemental Tables
X-XX. Genes that showed shifts on the human lineage were
overwhelmingly identified as placenta-specific by the
tissue-specificity index (TSI), with 56 out of 279 genes (20\%) with
both TSI>0.75 and placenta as one the top two tissues with the highest
expression. Unsurpisingly, this was not the case with other lineages,
since we used expression values from a human-derived microarray to
generate TSI values for each gene. However, this does underscore the
dynamic nature of expression patterns in the placenta and indeed all
tissues across mammals.

We then performed DAVID functional annotation clustering analyses on
each list of genes from each lineage. The top three clusters for each
lineage are shown in Table 10 and a full list of clusters in
Supplemental Table X. The top enrichment cluster in the human lineage
is blood vessel development, followed by extracellular matrix and
metal ion binding (Table 10). Other significant clusters include those
related to cell communication/signaling, cytokines, defense response,
tryptophan metabolism, apoptosis, and skeletal system development
(Supplemental Table X). The bonobo lineage was enriched in only three
clusters: apoptosis, Toll-like receptor signalling and apoptosis. The
mouse displayed enrichment in clusters involved in epigenetic
modifications such as methyltransferase activity and histone
modification, as well embryonic morphogeneisis, lipid biosynthesis,
among others (Supplemental Table X). The top enriched cluster in the
cow lineage was the response to vitamin A, as well as many clusters
related to the immune system. The dog showed enrichment in skeletal
system development, as well as lipid metabolic processes and blood
circulation. The elephant lineage had only one enriched cluster:
carbohydrate binding. Genes that showed enrichment on the opossum
lineage (essentially representing the difference between placental
mammals and marsupials) were clustered in many categories not seen in
the other lineages, including clusters related to the mitochondria,
urea cycle, peroxisome, glycolysis, and transport across the cell
membrane (Table 10 and Supplemental Table X).

\section{Discussion}

\subsection{Evolution of Placental Hormones}

Reproductive hormones are central to the regulation of pregnancy and
their source, timing, and quantity can vary widely from species to
species (Carter, 2012). Expression and synthesis of hormones is
perhaps the most well-documented physiological process in the placenta
in many species. Here we quantified the extreme disparity in genes
with hormone products or enzymatic products that directly produce
hormones in multiple mammalian species (Table 8).

In both ruminants and primates, the placenta is responsible for the
production of estrogen and progesterone. High relative expression of
CYP19A1, the enzyme that synthesizes estrogen, and HSD3B1, the enzyme
that synthesizes progesterone, in the human, bonobo, and cow,
generally matches the high levels of placental production of estrogen
and progesterone in humans, other primates, and ruminants in late
pregnancy. In addition, HSD3B1 shows high levels in dog, which matches
the high levels of placental production of progesterone at term in
that species.

The elephant is strikingly the only mammal analyzed that shows high
expression of the primarily pituitary hormone prolactin (PRL) in the
placenta. All afrotherians yet analyzed contain only one prolactin
gene, and prolactin has undergone adaptive evolution on the elephant
lineage (Wallis, 2009). (Yamamoto et al., 2011) noted the presence of
a prolactin in the trophoblast of the elephant placenta using
immunohistological staining and human derived antibodies. A previous
analysis of the elephant placenta also detected the expression of
prolactin (Hou et al., 2012). Here we confirm that the sole prolactin
in the elephant is not only expressed but highly expressed in the
placenta, and likely functions both in the maintenance of the corpus
luteum and as a placental lactogen (Yamamoto et al., 2011). PRL is not
expressed in the opossum, even though there is evidence that it is
expressed in the placenta of another marsupial, the tammar wallaby
(Menzies et al., 2011). This suggests that placental prolactin
expression in marsupials may be as diverse as in eutherian mammals.
Although, PRL is not expressed to any great degree in the placenta
other mammals, we see very high expression of genes which resulted
from duplications of PRL in both the cow (Table X), as well as mouse
(Supplemental Table X). Describe these

The growth hormone (GH) family has expanded in anthropoid primates,
resulting in approximately 5 genes in humans (GH1, GH2, CSH1, CSH2,
CSHL1). All but GH1 show substantial expression in the placenta (Table
8). There is some evidence that increased expression of these genes
may be correlated with overall birth weight of offspring in humans
(Mannik et al., 2010), and it is possible that the extreme expression
levels of these genes may be due to high birth weight. Ensembl
annotations show 4 GH genes present in the chimpanzee genome, and
three of these genes show high expression in bonobo, although to a
lesser degree than in human (Table 8). Human placental GH genes are
involved in increasing blood flow, replacing maternal pituitary
expression of GH, as well as trophoblast invasion (Alsat et al., 1998;
Lacroix et al., 2005; Schiessl et al., 2007), and it is likely that
they have the same function in the bonobo.

The human gene CGA codes for the alpha subunit of multiple hormones
including pituitary hormones (luteinizing hormone,
follicle-stimulating hormone, thyroid-stimulating hormone), as well as
numerous chorionic gonadotropins with variable beta subunits (Maston
and Ruvolo, 2002). The function of the chorionic gonadotropins is to
prevent the regression of the corpus luteum (Hanson et al., 1971).
Among the six placentas we analyzed here, CGA is only expressed in the
placentas of humans and bonobos (Table 8), which corresponds with the
emergence of the chorionic gonadotropins in anthropoid primates
(Maston and Ruvolo, 2002). However, oddly, CGA is expressed at much
higher levels in both species than the beta subunits with which its
product dimerizes (Table 8). CGA is one of several genes noted by
(Uddin et al., 2008) to have undergone adaptive evolution of the
human-chimpanzee stem lineage.

Relaxin is a hormone that has multiple effects during pregnancy
including increasing maternal blood flow, widening the pubic bone and
relaxing uterine musculature in preparation for labor. In both dogs
and elephants, there is high expression of relaxin in the term
placenta, with little evidence in other species examined here (Table
8). Relaxin is known to be produced in the placenta of the pregnant
dog and is used in the diagnosis of canine pregnancy (Steinetz et al.,
1989; Tsutsui and Stewart, 1991); however, the source of relaxin
produced by pregnant African elephants remains unclear (Meyer et al.,
2004). The high rise in relaxin in elephants has been suggested to be
a major signal for the initiation of labor (Niemuller et al., 1998).
Here we show that the placenta of elephants show high expression of
relaxin and is likely the predominant source of relaxin at term (Table
8).

Two genes, NPPB and NPPC, code for peptide hormones that are important
in the inducement of vasodilation, and may be important in increasing
maternal blood flow to the fetus (Cameron et al., 1996). There is some
indication that these proteins are present in the decidua of mice
(Cameron et al., 1996), but we find very little expression in the
fetal portion of the mouse placenta (Table 8). Here we find high
expression of at least one of these genes in dog, cow, and elephant.
The lack of expression in the hemochorial placentas of
Euarchontoglires may mean that high expression of these genes are
essential in increasing maternal blood flow in epitheliochorial and
endotheliochorial placentas

\subsection{The Evolution of Placental Expression of Galectins}

Galectins are recognized as major regulators of pregnancy in both
fetal and maternal tissues (Than et al., 2012). Indeed, the evolution
of new lineage-specific galectins has occurred in both primates and
ruminants, with the genes LGALS13, LGALS14, and LGALS16 recognized as
being placenta-specific in humans and regulating immune responses by
inducing apoptosis of maternal T-cells (Than et al., 2009). Indeed,
LGALS14 has the highest expression of any galectin in the human
placenta (Table 8), and other primate galectins such as LGALS13 and
LGALS16 are also expressed in human placenta. However, in the closely
related bonobo, LGALS14 is not found, and LGALS1 is the most abundant
galectin. LGALS1 shows generally high expression in all placentas
analyzed. LGALS1 is also involved in mediating maternal-fetal immune
tolerance, and is expressed in both maternal and fetal tissue in
primates (Than et al., 2008). LGALS1 placental expression was thought
to be confined to eutherian mammals, as there is evidence of adaptive
evolution at this node, as well as the appearance of estrogen response
elements in eutherian cis-regulatory regions (Than et al., 2008). It
was unclear as to whether galectins such as LGALS1 were important in
the regulation of pregnancy in marsupials, as well as eutherians. Here
we show substantial expression of galectins in the opossum, especially
LGALS1 (Table 8). In addition, LGALS2, LGALS3, and LGALS8 are also
expressed. The presence of galectins at the maternal-fetal interface
in the opossum implies that it was also present in the ancestral
placenta of therian mammals. Indeed, adaptive evolution of LGALS1 on
the eutherian stem is likely due to the increased invasive nature of
eutehrian placentation rather than the origination of maternal-fetal
tolerance. Although the function of LGALS1 in the marsupial placenta
needs to be confirmed, it is likely that the mediation of
maternal-fetal immune tolerance by galectins is an ancient feature of
mammalian viviparity.

\subsection{The Opossum Placenta and the Evolution of Viviparity}

The placenta of the opossum Monodelphis domestica is distinctly
different from that of the term placenta of eutherian mammals in that
it is derived from both the chorion and yolk sac. Indeed other
differences between the placentas of marsupials and eutherians include
lack of invasive extravillous trophoblasts in marsupials, the
retention of a permeable shell coat for 2/3 of marsupial gestation,
and the greater degree of imprinting in placenta-expressed genes in
eutherians (Renfree, 2010; Renfree et al., 2013). In addition, the
placenta of Monodelphis is also distinct from other marsupials, in
that it is considered one of the least specialized, and thought be
closer to the original marsupial condition (Renfree, 2010). In
Monodelphis, the duration of direct connection between maternal and
fetal tissues is as short as two days in a gestational period of 13-15
days. During this period, the placenta is directly engaged in the
exchange of urea with maternal tissues, owing to the lack of a
vesicular allantois (Freyer et al., 2007). This differs drastically
with the situation in another well-studied marsupial, the tammar
wallaby (Freyer et al., 2007). Therefore, we must interpret
differences between the opossum and eutherian placentas with caution,
as highly expressed genes and functional annotation clusters may
represent didelphid or Monodelphis specific adaptations.

Genes that show a significant shift on the opossum lineage include
functional annotation clusters that correspond to active generation of
energy, transport, and lipid breakdown via the peroxisome (Table 10;
Supplemental Table X). The choriovitelline placenta of both didelphids
and the tammar wallaby are known to be involved in glycolysis and
protein synethsis (Freyer et al., 2003). This implies that the
choriovitelline placenta in marsupials is an active site of metabolism
and molecular synthesis, as well as the transport of nutrients from
the maternal tissues. This is a distinct difference that is not seen
in significant shifts on other lineages, and may be due to the
formation of the placenta in the opossum from the yolk sac. In
addition, the urea cycle is a highly enriched functional annotation
cluster which directly correlates with the high levels of urea in the
Monodelphis placenta (Freyer et al., 2007). This is potentially a
Monodelphis-specific characteristic as discussed above.

The opossum placenta shows expression of genes that either do not
exist in eutherians or are not found in the term eutherian placenta
(Table X). The existence of embryonic forms of hemoglobin in fetal
tissues at term, as evidenced by the expression of HBE, HBZ, and HBO
is a feature of marsupials that is generally due to their very short
gestation periods. Embryonic hemoglobins in marsupials have a lower
affinity for oxygen and allow for a lower rate of oxygen uptake from
maternal tissues (Holland and Gooley, 1997). One novel uncharacterized
gene that shows high expression (XM\_001364916.1) is not found in
eutherians; Ensembl gene trees reveal that homologs of this gene exist
in other amniotes such as birds and turtles, but are not present in
eutherian mammals. The homolog of this gene in chicken is found to
expressed in reproductive tract (Froman et al., 2006), but its
function is unknown. WFDC2 is a member of the whey acidic protein
(WAP) family and involved in innate immunity. It is expressed in the
mammary gland of the tammar wallaby during pregnancy and early
lactation (Sharp et al., 2007), and here in abundance in the opossum
placenta.

\subsection{Novel Gene Expansions in the Elephant Placenta}

We found evidence of high expression in two novel expansions of genes
in the elephant genome. One family includes 11 genes that are related
to the gene PRSS16 gene, a thymus-specific serine protease that may be
involved regulating the presentation of self-peptides in CD4+
T-lymphocytes (Fornari et al., 2012). All 11 of these genes are
expressed in the elephant placenta, three of which have very abundant
mRNAs (Table 6). It is possible that this elephant-specific gene
family may be involved in immunity in the elephant placenta and the
mediation of maternal-fetal immune tolerance; however, future work
will need to test these hypotheses. Three other genes abundantly
expressed in the elephant placenta are homologues of SSLP1, a gene
with predominant expression in the seminal vesicle in mouse (Li et
al., 2006). Homologues of SSLP1 have been identified as being
expressed in the placental villi of cow, sheep, and goat. SSLP1 also
contains an Alu retrotransposon and may be involved in the development
of villi via cell proliferation (Ushizawa et al., 2009, 2010). We
verified expression in cow, however its abundant expression was also
identified in dog. This gene is not found in primates. Expansion of
these genes in elephants may indicate an elaboration of this system.

\subsection{Human Specific Expression Shifts and the Evolution of Preeclampsia}

The human placenta is distinctive from many other primates in that it
possesses a deeply invasive trophoblast that penetrates the inner
myometrium of the uterus (Carter and Pijnenborg, 2011). It is
hypothesized that this feature is related to the increased demand for
resources by the developing fetus to feed the expanded size of the
human brain (Carter and Pijnenborg, 2011; Robillard et al., 2002).
Humans have the largest brain in terms of relative size and this
differs markedly from closely related great apes (Boddy et al., 2012).
Deep penetration of the myometrium occurs in chimpanzees and gorillas,
but this penetration may be accelerated in humans (Carter and
Pijnenborg, 2011).

The near exclusivity of deep trophoblast invasion in humans has also
led many to speculate on the origin of various disorders, which are
almost exclusively symptomatic of human pregnancy. For example,
preeclampsia is a human disease that occurs in 4\% of pregnancies in
the United States. Preeclampsia can be attributed to the lack of
adequate invasion of maternal tissue, and is noted by the onset of
hypertension and proteinuria (citation). Some cases of preeclampsia
can develop into eclampsia, a more severe condition resulting in
seizures during pregnancy and death. Lack of invasion in preeclampsia
may be attributed to the inefficiency of growth regulators as well as
the inability to sufficiently suppress the maternal immune response.
Although there are speculative cases of eclampsia in great apes
(Baird, 1981; Stout and Lemmon, 1969; Thornton and Onwude, 1992), it
is exceedingly rare in these species (Carter and Pijnenborg, 2011). It
is possible that a comparative evolutionary approach to gene
expression can identify genes that have divergent expression patterns
from other mammals as well as association with preeclampsia.

Functional annotation clusters significantly enriched on the human
lineage (blood vessel development, extracellular matrix, tube
morphogenesis) seem to suggest involvement in remodeling of tissues
and increased blood flow, both of which may correlate with the
invasion of the trophoblast (Table 8; Supplemental Table X). In
addition, clusters such as cytokines and defense response
(Supplemental Table X) also indicate an a significant change in
expression of the immune response that may be involved in
immunosupression at the maternal-fetal boundary to further allow
invasion of fetal tissues.

Many genes that both show expression shifts on the human lineage and
have very high expression (FPKM>200) are directly related to the
highly invasive nature of the human placenta. For example, GPC3,
glypican 3, is an imprinted gene and cell growth regulator, associated
with Simpson-Golabi-Behmel fetal overgrowth syndrome, and involved in
intrauterine growth restriction (Monk and Moore, 2004; Pilia et al.,
1996). GPC3 has the highest expression in humans (Table 2), but is
also expressed in other mammals to a lesser degree, especially those
with invasive hemochorial placentas like bonobos (FPKM= 199.924),
spider monkeys (FPKM= 96.24), and the mouse (FPKM= 80.7545)
(Supplemental Tables X-XX). ADAM12 increases the availability of
insulin-like growth factors that lead to the promotion of cell growth
and placental invasion (Laigaard et al., 2005). (Hou et al., 2009)
found evidence of adaptive evolution in ADAM12 on the primate lineage,
and like GPC3 it is expressed in the bonobo (FPKM= 46.1553), spider
monkey (FPKM=30.41) and mouse (FPKM=2.25391) as well, although to a
lesser degree (Supplemental Tables X-XX). RHOBTB3 is imprinted and
associated with trophoblast invasion as well (Bilban et al., 2010;
Yuen et al., 2011). In addition, one other gene, KISS1, shows
exceptional abundance in the human placenta (Table 2); however due to
its absence in elephant Ensmebl gene annotations, it was not part of
the 1:1 orthologues examined in the comparative study. Here, it is
only expressed in the term placentas of humans (Table 2) and bonobos
(FPKM=17.6394; Supplemental Table X). KISS1 is thought to be involved
in the regulation of trophoblast invasion, and shows highest
expression in the first trimester of pregnancy (Bilban et al., 2004;
Reynolds et al., 2009).

In addition, several other genes that show expression shifts on the
human lineage also have major effects on placental function. Genes
such as TXK and EBI3 are intimately involved in the immune system in
the placenta. TXK is involved in natural killer cell differentiation
(Bezman et al., 2012). EBI3 is implicated in natural killer cell
function an maternal-fetal immunomodulation in the syncytiotrophoblast
(Devergne et al., 2001). KYNU and KMO are interacting genes that are
involved in tryptophan metabolism, which may be involved in
immunosupression and the prevention of fetal rejection (Takikawa,
2005). THSD7A is a thrombospondin, which inhibits endothelial cell
migration and tube morphogenesis (Wang et al., 2010).

Many genes that show evidence of a shift on the human lineage show
evidence of implication in preeclampsia, including ADAM12 (Gack et
al., 2005; Laigaard et al., 2005) MAOA (Carrasco et al., 2000;
Bottalico et al., 2004), BCL2 (Ishihara et al., 2002), GPC3 (Gui et
al., 2012; Ogawa, 2007), TMEM8A, CLIC3, EBI3, FLT1, BCL6, SYDE1, and
MAN1C1 (Enquobahrie et al., 2008; Winn et al., 2009; Nishizawa, 2011).
In addition KISS1 is also involved in preeclampsia as well (Sado et
al., 2011; Qiao et al, 2012). Due to the association of many genes
that show human-specific shifts as well as high expression, we
conclude that preeclampsia is possibly a symptom of recent evolution
of gene expression on the human lineage. Although many genes such as
GPC3 and ADAM12 show distinctive relatively high expression in bonobo,
this expression has intensified on the human lineage. Sequencing of
more placentas of nonhuman primates, including other individuals of
the genus Pan will help to further characterize this transition.

\section{Conclusion}

Here we sequenced six placentas spanning the phylogenetic tree of
mammals. This revealed a dynamic expression pattern of expression of
lineage-specific expansions of gene families and differential gene
expression that agreed with the scenario of an “arms race” between
maternal and fetal tissues that have shaped such disparate gene
expression profiles. In addition, we highlighted the broad spectrum of
hormone gene expression in multiple species that suggested that there
was a lack of conservation in hormone signaling during mammalian
pregnancy. Major differences between expression in marsupial and
eutherian mammals at term seemed to be correlated with the differences
between choriovitelline and chorioamniotic placenta, with the
marsupial placenta more enriched in genes the contributed to the
active generation of energy. We also identified that galectin
expression, especially that of LGALS1, is a common feature of all
mammalian placentas and was most likely present in the ancestral
therian. Also, we identified two new expansions of genes that are
highly expressed in the elephant placenta hat may be involved in
trophoplast differentiation and immune system function, mirroring
expansions in other groups such as primates, rodents, and ruminants.
We also quantified significant lineage-specific shifts in gene
expression, concentrating on the identification of genes that showed
gene expression changes on the human lineage. We cautiously propose
that many of these genes may be related to the invasive nature of the
trophoblast in humans, although we note that in many of these genes,
there is in some cases substantial expression in the bonobo as well.
These genes have the potential to spur further research and may be at
the root of obstetrical disorders such as preeclampsia. We propose
that more placentas from diverse lineages, especially those of
marsupials and primates, to further identify genes and expression
levels that are specific to marsupials and eutehrians, as well as
confirm the identity of human-specific expression shifts.

\printbibliography

\end{document}

### 2015-08-19 manuscript/manuscript.Rnw
site mossman and carter

@@
 these functions are conserved among mammals, the placenta is perhaps
-the most variable organ in terms of structure and form (Mossman,
-1937). Species differ in the shape of attachment (i.e., discoid,
-diffuse, cotyledonary, zonary), degree of invasiveness of fetal
-tissues into maternal tissues (i.e. epitheliochorial,
-endotheliochorial, hemochorial), and degree of interdigitation of
-maternal-fetal tissue (i.e., villous, folded, trabecular, lamellar,
-labrynthine) (Mossman, 1987). In addition, therian mammals differ
-drastically in the number and size of their offspring, as well as the
-length of gestation, traits often associated with specific aspects of
-placental morphology (Capellini et al., 2011).
+the most variable organ in terms of structure and form
+\cite{Mossman.1937:comparativemorphogenesisofthe}. Species differ in
+the shape of attachment (i.e., discoid, diffuse, cotyledonary,
+zonary), degree of invasiveness of fetal tissues into maternal tissues
+(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
+interdigitation of maternal-fetal tissue (i.e., villous, folded,
+trabecular, lamellar, labrynthine)
+\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
+mammals differ drastically in the number and size of their offspring,
+as well as the length of gestation, traits often associated with
+specific aspects of placental morphology (Capellini et al., 2011).
 
@@
 placentals occurs via the chorioallantoic placenta, produced from the
-fusion of the chorion and allantois (Mossman, 1987). It is unclear
+fusion of the chorion and allantois \cite{Mossman.1987:vertebratefetalmembranes}. It is unclear
 whether the first placenta was more like that of marsupials or
@@
 Although the morphology of the mammalian placenta has been well
-characterized in a number of species (Mossman, 1987), the diversity of
+characterized in a number of species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
 the molecular environment is only beginning to be understood from a
-genetic perspective (Carter, 2012). Studies of certain model species
+genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}. Studies of certain model species
 (i.e., humans, mice, cattle) have revealed a stunning divergence in

### 2015-08-24 manuscript/manuscript.Rnw
add more missing references

@@
 as well as the length of gestation, traits often associated with
-specific aspects of placental morphology (Capellini et al., 2011).
+specific aspects of placental morphology
+\cite{Capellini.ea2011:Placentationandmaternalinvestment}.
 
@@
 tissues lasts for a proportionally shorter time during gestation
-(Renfree, 2010). Marsupials also differ from eutherian mammals in
+\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. Marsupials also differ from eutherian mammals in
 that most possess only a choriovitelline placenta, produced from the
@@
 hemochorial placenta, in which the maternal and fetal blood supply is
-separated by one tissue layer, as in humans and mouse (Elliot
-and Crespi, 2009; Wildman et al., 2006).
+separated by one tissue layer, as in humans and mouse
+\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
 
@@
 the genetics of the placenta, especially in later stages of pregnancy
-(Carter, 2012; Hou et al., 2012; Knox and Baker, 2008; Su et al.,
-2002). For example, each lineage has evolved expansive
+\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
+For example, each lineage has evolved expansive
 placenta-specific gene families with expression patterns that can vary
-in space and time (Green et al., 2000; Simmons et al., 2008). These
+in space and time \cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}. These
 include independent expansions of the prolactins in mouse and cow,
@@
 the growth hormone and chorionic gonadotropin beta genes in hominoid
-primates (Haig, 2008; Rawn and Cross, 2008). Expression patterns of
+primates \cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}. Expression patterns of
 specific genes also can vary widely between species. For example, the
 expression of orthologous genes in the placenta have been found to
-vary substantially between humans and mice (Su et al., 2002), and the
+vary substantially between humans and mice \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
 recruitment of SIGLEC6 to placental expression in humans differs from
-both chimpanzees and mice (Brinkman-Van der Linden et al., 2007).
+both chimpanzees and mice \cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
 
@@
 the distinct evolutionary interplay between maternal and fetal
-strategies during pregnancy (Haig, 1993). The maternal strategy is to
+strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}. The maternal strategy is to
 requisition an adequate degree of nutrients to the fetus without
@@
 production of new genes, and enlistment of existing genes to
-expression at the maternal-fetal interface (Haig, 1993).
+expression at the maternal-fetal interface \cite{Haig1993:Geneticconflictsinhuman}.
 

### 2015-08-28 manuscript/manuscript.Rnw
add more missing references to the manuscript

@@
 of this maternal-fetal conflict is its potential affect on human
-health (Hollegaard et al., 2013). In addition to having an invasive
+health \cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. 
+In addition to having an invasive
 hemochorial placenta, humans possess a deeply invasive trophoblast
-layer that can penetrate the myometrium of the uterus (Carter and
-Pijnenborg, 2011). Defects in placental growth are known to be
+layer that can penetrate the myometrium of the uterus 
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
+ Defects in placental growth are known to be
 associated with the devastating condition of preeclampsia, which can
 cause hypertension, proteinuria, and has the potential to lead to
-death of the mother and/or infant (citation). Although cases have been
-documented in chimpanzees and gorillas (citation), preeclampsia has a
+death of the mother and/or infant \DLA{(citation)}. Although cases have been
+documented in chimpanzees and gorillas \DLA{(citation)}, preeclampsia has a
 far higher incidence in human pregnancies with as much as 4\% of
-pregnancies being affected in the United States (citation).
+pregnancies being affected in the United States \DLA{(citation)}.
 Chimpanzees and gorillas also have deeply invasive trophoblasts,
 although the progression of invasion in great apes may be slower than
-humans (Carter and Pijnenborg, 2011); however, large-scale expression
+humans \cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however, large-scale expression
 studies of great ape placentas have not been conducted. The increasing
@@
 address broader phylogenetic questions in the evolution of gene
-expression (Brawand et al., 2011). Here we use Illumina RNA-Seq
+expression \cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use Illumina RNA-Seq
 technology to sequence and quantify the mRNA environment of six
@@
 as well as traits such as number and size of offspring and length of
-gestation (Table 1). We use this data, along with existing RNA-Seq
-data from mouse (Wang et al., 2011), to address the major differences
+gestation \DLA{(Table 1)}. We use this data, along with existing RNA-Seq
+data from mouse \cite{Wang.ea2011:surveyfornovelimprinted}, to address the major differences
 between expression in marsupial and eutherian mammals at term,
@@
 Fetal tissue was collected from the placenta of six mammalian species
-(Homo sapiens [human], Pan paniscus [bonobo], Canis familiaris
-[domestic dog], Bos taurus [domestic cow], Loxodonta africana [African
-savanna elephant], and Monodelphis domestica [gray short-tailed
-opossum]). Information on specimen ID, place and time of collection,
-and collector is shown for each sample in Supplemental Table 1. In the
-case of Pan, Bos, and Loxodonta, the placenta was collected at term
+(\textit{Homo sapiens} [human], \textit{Pan paniscus} [bonobo], \textit{Canis familiaris}
+[domestic dog], \textit{Bos taurus} [domestic cow], \textit{Loxodonta africana} [African
+savanna elephant], \textit{Monodelphis domestica} [gray short-tailed
+opossum], \textit{Nannospalax gallili} [?], \textit{Spalax carmelli} [?]).
+Information on specimen ID, place and time of collection,
+and collector is shown for each sample in \DLA{Supplemental Table 1}. In the
+case of \textit{Pan}, \textit{Bos}, and \textit{Loxodonta}, the placenta was collected at term
 upon expulsion from the mother following delivery. Placental tissue
-from Homo sapiens was collected upon delivery via a caesarian section.
-Near-term placental tissue from Canis familiaris was collected upon
+from \textit{Homo sapiens} was collected upon delivery via a caesarian section.
+Near-term placental tissue from \textit{Canis familiaris} was collected upon
 the death of the mother from a car accident. Placental tissue from
-Monodelphis domestica was sampled at fetal stage 32, roughly one day
+\textit{Monodelphis domestica} was sampled at fetal stage 32, roughly one day
 prior to birth. Only the fetal portion of the placenta was sampled at
-the maternal-fetal interface (i.e. villous portion for Homo, Pan,
-Canis, and Loxodonta; cotyledon for Bos; diffuse placenta for
-Monodelphis). All tissues were collected and stored in RNAlater (Life
+the maternal-fetal interface (i.e. villous portion for \textit{Homo}, \textit{Pan},
+\textit{Canis}, and \textit{Loxodonta}; cotyledon for \textit{Bos}; diffuse placenta for
+\textit{Monodelphis}). All tissues were collected and stored in RNAlater (Life
 Technologies, Carlsbad, CA) until RNA isolation could be performed.
@@
 
+\DLA{Add details for new sequences and fix numbers}
+
 \subsection{Sequencing}
@@
 performed using the Illumina Genome Analyzer II with an insert size of
-230 base pairs (bp). Bos, Loxodonta, and Monodelphis were sequenced
-with read lengths of 75 bp; Canis, Homo, and Pan were sequenced with
+230 base pairs (bp). \textit{Bos}, \textit{Loxodonta}, and
+\textit{Monodelphis} were sequenced with read lengths of 75 bp;
+\textit{Canis}, \textit{Homo}, and \textit{Pan} were sequenced with
 read lengths of 150 bp.
@@
 
+\DLA{This isn't how these were analyzed anymore; fix this.}
+
 The quality of reads was evaluated using FASTQC v. 0.9.2
-(http://www.bioinformatics.babraham.ac.uk/projects/fastqc/). Following
+(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}). Following
 this, reads were trimmed based on quality in Galaxy using the FASTQ
@@
 possesses a deeply invasive trophoblast that penetrates the inner
-myometrium of the uterus (Carter and Pijnenborg, 2011). It is
+myometrium of the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
 hypothesized that this feature is related to the increased demand for
@@
 (Baird, 1981; Stout and Lemmon, 1969; Thornton and Onwude, 1992), it
-is exceedingly rare in these species (Carter and Pijnenborg, 2011). It
+is exceedingly rare in these species \cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It
 is possible that a comparative evolutionary approach to gene

### 2015-08-28 manuscript/manuscript.Rnw
add more missing references to the manuscript

@@
 The quality of reads was evaluated using FASTQC v. 0.9.2
-(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}). Following
-this, reads were trimmed based on quality in Galaxy using the FASTQ
-Trimmer v. 1.0.0 (Blankenberg et al., 2010). All transcriptomes were
-then aligned to reference genomes and assembled using the Tuxedo suite
-of next-generation sequencing analysis tools. Annotated reference
-genomes were downloaded from Ensembl v. 64 (Flicek et al., 2012),
-including Homo sapiens GRCh37, Pan troglodytes CHIMP2.1 (6X coverage),
-Canis familiaris BROADD2 (7.6X), Bos taurus UMD3.1 (9.5X), Loxodonta
-africana loxAfr3 (7X), and Monodelphis domestica BROADO5 (6.8X). The
-genome of the closely-related Pan troglodytes was used to map the
-transcriptome of Pan paniscus; the separation between the two
-recognized Pan species is generally perceived to be ~1 Mya (Prufer et
-al., 2012). Reads were aligned to reference genomes using Bowtie v.
-0.12.7 (Langmead et al., 2009) and TopHat v. 1.3.3 (Trapnell et al.,
-2009). Cufflinks v. 1.2.1 was then used to assemble transcripts and
-quantify expression values using fragments per kilobase of exon per
-million mapped reads (FPKM) (Trapnell et al., 2010). RNA-Seq metrics
-were computed using Picard (http://picard.sourceforge.net/). For
-comparison, we also downloaded the placenta Illumina RNA-Seq
-transcriptome of a mouse from (Wang et al., 2011). Raw reads were
-mapped as above using the Ensembl v. 64 annotation of the Mus musculus
-(house mouse) genome (NCBI m37).
+(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
+Following this, reads were trimmed based on quality in Galaxy using
+the FASTQ Trimmer v. 1.0.0
+\cite{Blankenberg.ea2010:ManipulationofFASTQdata}. All transcriptomes
+were then aligned to reference genomes and assembled using the Tuxedo
+suite of next-generation sequencing analysis tools. Annotated
+reference genomes were downloaded from Ensembl v. 64
+\cite{Flicek.ea2012:Ensembl2012}, including \textit{Homo sapiens}
+GRCh37, \textit{Pan troglodytes} CHIMP2.1 (6X coverage), \textit{Canis
+  familiaris} BROADD2 (7.6X), \textit{Bos taurus} UMD3.1 (9.5X),
+\textit{Loxodonta africana} loxAfr3 (7X), and \textit{Monodelphis
+  domestica} BROADO5 (6.8X). The genome of the closely-related Pan
+troglodytes was used to map the transcriptome of Pan paniscus; the
+separation between the two recognized Pan species is generally
+perceived to be ~1 Mya (Prufer et al., 2012). Reads were aligned to
+reference genomes using Bowtie v. 0.12.7\cite{Langmead2009} and TopHat
+v. 1.3.3 \cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
+Cufflinks v. 1.2.1 was then used to assemble transcripts and quantify
+expression values using fragments per kilobase of exon per million
+mapped reads (FPKM)
+\cite{Trapnell.ea2010:Transcriptassemblyandquantification}. RNA-Seq
+metrics were computed using Picard
+(\url{http://picard.sourceforge.net/}). For comparison, we also
+downloaded the placenta Illumina RNA-Seq transcriptome of a mouse from
+\cite{Wang.ea2011:surveyfornovelimprinted}. Raw reads were mapped as
+above using the Ensembl v. 64 annotation of the \textit{Mus musculus} (house
+mouse) genome (NCBI m37).
 
@@
 functional annotation cluster analyses per transcriptome using DAVID
-(Huang et al., 2009; Sherman et al., 2007). DAVID ranks functional
-annotation clusters according to enrichment score (ES); functional
-annotation clusters are considered significantly enriched if ES>1.3,
-as this corresponds to p<0.05 (Huang et al., 2009). We executed
+\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
+DAVID ranks functional annotation clusters according to enrichment
+score (ES); functional annotation clusters are considered
+significantly enriched if ES>1.3, as this corresponds to p<0.05
+\cite{Huang.ea2009:Systematicandintegrativeanalysis}. We executed
 analyses with two gene lists per transcriptome: 1) annotated genes
@@
 elephant, and opossum were determined using Ensembl v.64 gene trees
-via the Biomart online data management tool (Kasprzyk, 2011). A total
-of 6882 1:1 orthologous genes were expressed in at least one species.
-For further analysis, we eliminated all genes that did not have at
-least one species with an FPKM>10. A list of these 4259 1:1
-orthologous genes was retained for future analyses of gene expression
-evolution described below.
+via the Biomart online data management
+tool\cite{Kasprzyk2011:BioMartdrivingparadigmchange}. A total of 6882
+1:1 orthologous genes were expressed in at least one species. For
+further analysis, we eliminated all genes that did not have at least
+one species with an $\mathrm{FPKM}\ge10$. A list of these 4259 1:1 orthologous
+genes was retained for future analyses of gene expression evolution
+described below.
 
@@
 Hypotheses) in the R statistical environment to investigate
-significant shifts in gene expression levels over time (Butler and
-King, 2004). OUCH institutes an Ornstein-Uhlenbeck (OU) model of
-evolution, which allows for different optima for quantitative trait
-values along user-specified branches of a phylogenetic tree. The OU
-model explicitly accounts for change in quantitative trait values by
-introducing two parameters into the model of stochastic Brownian
-motion of change in trait value along a phylogenetic tree through
-time. These parameters include, ?, or an optimum trait value, and ,
-or the strength of selection (interpreted as “pull” toward the
-optimum). In addition, the OU model can also integrate evolutionary
-change by assuming that each user-specified set of braches has its own
-separate OU parameters (see Butler and King, 2004).
+significant shifts in gene expression levels over
+time\cite{Butler.2004:phlogeneticcomparativeanalysis}. OUCH institutes
+an Ornstein-Uhlenbeck (OU) model of evolution, which allows for
+different optima for quantitative trait values along user-specified
+branches of a phylogenetic tree. The OU model explicitly accounts for
+change in quantitative trait values by introducing two parameters into
+the model of stochastic Brownian motion of change in trait value along
+a phylogenetic tree through time. These parameters include, $\theta$,
+or an optimum trait value, and $\alpha$, or the strength of selection
+(interpreted as “pull” toward the optimum). In addition, the OU model
+can also integrate evolutionary change by assuming that each
+user-specified set of braches has its own separate OU parameters
+\cite{Butler.2004:phlogeneticcomparativeanalysis}.
 
@@
 genes with universally low expression levels, and included only genes
-in which expression in at least one species had an FPKM≥10 as
-described above. We corrected for multiple testing using the false
-discovery rate (FDR) and calculated adjusted p values (q-values) using
-the Benjamini-Hochstein correction. We used the time-calibrated
-phylogenetic tree of (Meredith et al., 2011) with all species not used
-in this analyses pruned.
+in which expression in at least one species had an
+$\mathrm{FPKM}\ge10$ as described above. We corrected for multiple
+testing using the false discovery rate (FDR) and calculated adjusted
+$p$ values (q-values) using the Benjamini-Hochberg
+correction\cite{Benjamini1995:bh_procedure}. We used the
+time-calibrated phylogenetic tree of
+\cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
+not used in this analyses pruned.
 
@@
 we performed functional annotation cluster analyses on each branch
-using DAVID (Huang et al., 2009; Sherman et al., 2007). DAVID ranks
-functional annotation clusters according to enrichment score (ES);
-functional annotation clusters are considered significantly enriched
-if ES>1.3, as this corresponds to p<0.05 (Huang et al., 2009). As
+using DAVID
+\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
+DAVID ranks functional annotation clusters according to enrichment
+score (ES); functional annotation clusters are considered
+significantly enriched if $\mathrm{ES}\ge1.3$, as this corresponds to
+$p\le0.05$ \cite{Huang.ea2009:Systematicandintegrativeanalysis}. As
 above, due to the paucity of annotated gene ontology terms for some

### 2015-08-28 manuscript/manuscript.Rnw
add more missing references to the manuscript

@@
 
+\DLA{Complete discussion of the ateles placenta}
+
 \section{Results}
@@
 Sequence Read Archive (SRA) (Accession Number XXX-YYY).
+
+\DLA{Need to do an embargoed submission to SRA}
 
@@
 
-The top 50 expressed genes in each genome analyzed are shown in Tables
-2-7, and all genes for each transcriptome are listed in Supplementary
-Tables X-XX [These are large Excel files to be included with the
-Supplemental Material]. In addition, specific genes including
-placental hormones and hormone-related genes, galectins, and
+\DLA{Summarize all of this in a table}
+
+The top 50 expressed genes in each genome analyzed are shown in
+\DLA{Tables 2-7}, and all genes for each transcriptome are listed in
+\DLA{Supplementary Tables X-XX} [These are large Excel files to be
+included with the Supplemental Material]. In addition, specific genes
+including placental hormones and hormone-related genes, galectins, and
 insulin-like growth factor related genes are shown with FPKM values
 side to side for all transcriptomes including the reanalysis of mouse
-from Wang et al. (2011) in Table 8. Below we describe each
-transcriptome separately, specifically concentrating on
-highly-expressed genes.
+from \cite{Wang.ea2011:surveyfornovelimprinted} in \DLA{Table 8}.
+Below we describe each transcriptome separately, specifically
+concentrating on highly-expressed genes.
 
@@
 their source, timing, and quantity can vary widely from species to
-species (Carter, 2012). Expression and synthesis of hormones is
+species \cite{Carter2012:Evolutionofplacentalfunction}. Expression and synthesis of hormones is
 perhaps the most well-documented physiological process in the placenta
@@
 with hormone products or enzymatic products that directly produce
-hormones in multiple mammalian species (Table 8).
+hormones in multiple mammalian species (\DLA{Table 8}).
 
@@
 the high levels of placental production of progesterone at term in
-that species.
+that species. \DLA{I think we need citations for this}
 
@@
 gene, and prolactin has undergone adaptive evolution on the elephant
-lineage (Wallis, 2009). (Yamamoto et al., 2011) noted the presence of
-a prolactin in the trophoblast of the elephant placenta using
-immunohistological staining and human derived antibodies. A previous
-analysis of the elephant placenta also detected the expression of
-prolactin (Hou et al., 2012). Here we confirm that the sole prolactin
-in the elephant is not only expressed but highly expressed in the
-placenta, and likely functions both in the maintenance of the corpus
-luteum and as a placental lactogen (Yamamoto et al., 2011). PRL is not
-expressed in the opossum, even though there is evidence that it is
-expressed in the placenta of another marsupial, the tammar wallaby
-(Menzies et al., 2011). This suggests that placental prolactin
-expression in marsupials may be as diverse as in eutherian mammals.
-Although, PRL is not expressed to any great degree in the placenta
-other mammals, we see very high expression of genes which resulted
-from duplications of PRL in both the cow (Table X), as well as mouse
-(Supplemental Table X). Describe these
+lineage \cite{Wallis2009:ProlactininAfrotheriacharacterization}.
+\cite{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
+presence of a prolactin in the trophoblast of the elephant placenta
+using immunohistological staining and human derived antibodies. A
+previous analysis of the elephant placenta also detected the
+expression of prolactin (Hou et al., 2012). Here we confirm that the
+sole prolactin in the elephant is not only expressed but highly
+expressed in the placenta, and likely functions both in the
+maintenance of the corpus luteum and as a placental lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}. PRL is not expressed in the opossum, even though there
+is evidence that it is expressed in the placenta of another marsupial,
+the tammar wallaby \cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests that
+placental prolactin expression in marsupials may be as diverse as in
+eutherian mammals. Although, PRL is not expressed to any great degree
+in the placenta other mammals, we see very high expression of genes
+which resulted from duplications of PRL in both the cow \DLA{(Table X)}, as
+well as mouse \DLA{(Supplemental Table X)}. \DLA{Describe these}
 
@@
 may be correlated with overall birth weight of offspring in humans
-(Mannik et al., 2010), and it is possible that the extreme expression
+\DLA{(Mannik et al., 2010)}, and it is possible that the extreme expression
 levels of these genes may be due to high birth weight. Ensembl
@@
 involved in increasing blood flow, replacing maternal pituitary
-expression of GH, as well as trophoblast invasion (Alsat et al., 1998;
-Lacroix et al., 2005; Schiessl et al., 2007), and it is likely that
+expression of GH, as well as trophoblast invasion \cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth}, and it is likely that
 they have the same function in the bonobo.
@@
 follicle-stimulating hormone, thyroid-stimulating hormone), as well as
-numerous chorionic gonadotropins with variable beta subunits (Maston
-and Ruvolo, 2002). The function of the chorionic gonadotropins is to
-prevent the regression of the corpus luteum (Hanson et al., 1971).
+numerous chorionic gonadotropins with variable beta subunits \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of the chorionic gonadotropins is to
+prevent the regression of the corpus luteum \cite{Hanson.ea1971:EffectsofHCGand}.
 Among the six placentas we analyzed here, CGA is only expressed in the
-placentas of humans and bonobos (Table 8), which corresponds with the
+placentas of humans and bonobos (\DLA{Table 8}), which corresponds with the
 emergence of the chorionic gonadotropins in anthropoid primates
-(Maston and Ruvolo, 2002). However, oddly, CGA is expressed at much
+\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly, CGA is expressed at much
 higher levels in both species than the beta subunits with which its
-product dimerizes (Table 8). CGA is one of several genes noted by
-(Uddin et al., 2008) to have undergone adaptive evolution of the
+product dimerizes (\DLA{Table 8}). CGA is one of several genes noted by
+\cite{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone adaptive evolution of the
 human-chimpanzee stem lineage.
@@
 8). Relaxin is known to be produced in the placenta of the pregnant
-dog and is used in the diagnosis of canine pregnancy (Steinetz et al.,
-1989; Tsutsui and Stewart, 1991); however, the source of relaxin
-produced by pregnant African elephants remains unclear (Meyer et al.,
-2004). The high rise in relaxin in elephants has been suggested to be
-a major signal for the initiation of labor (Niemuller et al., 1998).
-Here we show that the placenta of elephants show high expression of
-relaxin and is likely the predominant source of relaxin at term (Table
-8).
+dog and is used in the diagnosis of canine pregnancy
+\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof};
+however, the source of relaxin produced by pregnant African elephants
+remains unclear \cite{Meyer.ea2004:Speciesandfetalgender}. The high
+rise in relaxin in elephants has been suggested to be a major signal
+for the initiation of labor
+\cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive} Here we
+show that the placenta of elephants show high expression of relaxin
+and is likely the predominant source of relaxin at term \DLA{(Table 8)}.
 

### 2015-09-24 manuscript/manuscript.Rnw
start working on loading results
fix transcriptome assembly and mapping section

@@
 \usepackage{setspace}
-%\usepackage{txfonts}
-%\usepackage{fancyref}
+\usepackage{wrapfig}
+% \usepackage{txfonts}
+\usepackage[plain]{fancyref}
+\renewcommand*{\fref}[1]{\Fref{#1}}
 \usepackage{acronym}
@@
 
+% Acronyms
+\acrodef{EV}[EV]{Extracellular Vesicle}
+\acrodef{SVM}{Support Vector Machine}
+\acrodef{ROC}{Receiver Operating Characteristic}
+\acrodef{KNN}{K Nearest Neighbors}
+\acrodef{SRA}{Sequence Read Archive}
+\acrodef{FPKM}{Fragments Per Kilobase of exon per Million reads \cite{Trapnell.ea2010:Transcriptassemblyandquantification}}
+
 % \twocolumn
@@
 library("reshape2")
+library("xtable")
 @
@@
     source(a)
+}
+@
+
+<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm")))>>=
+for (a in paste0("../data/",c("combined_fpkm"))) {
+    load(a)
 }
@@
 
+\DLA{Fix how the sequencing was done}
+
 All RNA-Seq mRNA isolation, library construction, and Illumina
@@
 
+For comparison, additional sequences of placenta in \textit{Mus
+  musculus} \cite{Wang.ea2011:surveyfornovelimprinted} and
+\textit{Homo sapiens}
+\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} were obtained
+from the \ac{SRA}.
+
 \subsection{Transcriptome Assembly and Mapping}
 
-\DLA{This isn't how these were analyzed anymore; fix this.}
-
-The quality of reads was evaluated using FASTQC v. 0.9.2
+The quality of reads was evaluated using FASTQC v0.11.2
 (\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
-Following this, reads were trimmed based on quality in Galaxy using
-the FASTQ Trimmer v. 1.0.0
-\cite{Blankenberg.ea2010:ManipulationofFASTQdata}. All transcriptomes
-were then aligned to reference genomes and assembled using the Tuxedo
-suite of next-generation sequencing analysis tools. Annotated
-reference genomes were downloaded from Ensembl v. 64
-\cite{Flicek.ea2012:Ensembl2012}, including \textit{Homo sapiens}
-GRCh37, \textit{Pan troglodytes} CHIMP2.1 (6X coverage), \textit{Canis
-  familiaris} BROADD2 (7.6X), \textit{Bos taurus} UMD3.1 (9.5X),
-\textit{Loxodonta africana} loxAfr3 (7X), and \textit{Monodelphis
-  domestica} BROADO5 (6.8X). The genome of the closely-related Pan
-troglodytes was used to map the transcriptome of Pan paniscus; the
-separation between the two recognized Pan species is generally
-perceived to be ~1 Mya (Prufer et al., 2012). Reads were aligned to
-reference genomes using Bowtie v. 0.12.7\cite{Langmead2009} and TopHat
-v. 1.3.3 \cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
-Cufflinks v. 1.2.1 was then used to assemble transcripts and quantify
-expression values using fragments per kilobase of exon per million
-mapped reads (FPKM)
-\cite{Trapnell.ea2010:Transcriptassemblyandquantification}. RNA-Seq
-metrics were computed using Picard
-(\url{http://picard.sourceforge.net/}). For comparison, we also
-downloaded the placenta Illumina RNA-Seq transcriptome of a mouse from
-\cite{Wang.ea2011:surveyfornovelimprinted}. Raw reads were mapped as
-above using the Ensembl v. 64 annotation of the \textit{Mus musculus} (house
-mouse) genome (NCBI m37).
+Transcriptomes without a suitable reference genome (Reference of ---
+in \fref{tab:ref_genome_aligned_species}) were assembled into a
+reference using Trinity v2014-07-17
+\cite{Haas.ea2013:Denovotranscriptsequence} and annotated using
+Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against either
+the \textit{Homo sapiens} or \textit{Mus musculus} transcriptome
+(\fref{tab:ref_genome_aligned_species}).
+% 
+All transcriptomes were then aligned to the appropriate reference
+genome using STAR v2.4.2a
+\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
+values in \ac{FPKM} were quantified using Cufflinks v2.2.1
+\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
+%
+The genome of the closely-related \textit{Pan troglodytes} was used as
+a reference for \textit{Pan paniscus}. The separation between the two
+recognized Pan species is generally perceived to be $\approx 1$ Mya
+\cite{Prüfer.ea2012:bonobogenomecomparedwith}.
+
+\begin{table}
+  \begin{tabular}{c c c c}
+    \toprule
+    Species                       & Common Name        & Reference & Annotation \\
+    \midrule
+    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & Ensembl 80 \\
+    \textit{Ateles fusciceps}     & Spider monkey      & ---        & \textit{Homo sapiens} Ensembl 80 \\
+    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & Ensembl 80 \\
+    \textit{Bos taurus}           & Cow                & UMD3.1     & Ensembl 80 \\
+    \textit{Canis familiaris}     & Dog                & CanFam3.1  & Ensembl 80 \\
+    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & Ensembl 80 \\
+    \textit{Homo sapiens}         & Human              & GRCh38     & Ensembl 80 \\
+    \textit{Mus musculus}         & Mouse              & GRCm38     & Ensembl 80 \\
+    \textit{Nannospalax galili}   & Blind mole rat     & ----       & \textit{Mus musculus} Ensembl 80 \\
+    \textit{Monodelphis domestica} & Opossum           & BROAD05    & Ensembl 80 \\
+    \textit{Spalax carmeli}       & Blind mole rat     & ----       & \textit{Mus musculus} Ensembl 80 \\
+    \bottomrule
+  \end{tabular}
+  \caption{Reference genomes for aligned species}
+  \label{tab:ref_genome_aligned_species}
+\end{table}
 
@@
 with expression levels of the top 5\% of the frequency distribution of
-FPKM and 2) the top 100 genes with the highest FPKM. Due to the
+\ac{FPKM} and 2) the top 100 genes with the highest \ac{FPKM}. Due to the
 paucity of annotated gene ontology terms for some species, gene lists
@@
 
+<<transcriptome.statistics>>=
+@
+
 For transcriptomes sequenced with 75 bp, (opossum, elephant, and cow),
@@
 annotated transcripts to which reads mapped in each respective genomes
-ranged from 4119 (bonobo) to 6506 (human) for transcripts with FPKM>10
-and 11689 (dog) to 30733 (human) for transcripts with FPKM>1. The
+ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
+and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
 greater degree of mapping to annotated protein-coding sequences in
@@
 including placental hormones and hormone-related genes, galectins, and
-insulin-like growth factor related genes are shown with FPKM values
+insulin-like growth factor related genes are shown with \ac{FPKM} values
 side to side for all transcriptomes including the reanalysis of mouse
@@
 
+<<top.genes.function,echo=FALSE,results="hide">>=
+top.genes.table <- function(species.name,n=50) {
+    temp <- combined.fpkm[species==species.name &
+                      !grepl("^MT-?",gene_short_name)][order(-mean_fpkm)][!duplicated(gene_short_name)][1:n,list(gene_short_name,human_name,mean_fpkm)]
+    setnames(temp,c("Gene name","Human Name","FPKM"))
+    if (species.name=="homo sapiens") {
+        temp <- temp[,-2,with=FALSE]
+    }
+    return(temp)
+}
+@
+
 \subsection{Human Placenta Transcriptome}
 
-The top 50 expressed genes as measured by FPKM in the term placenta of
-the human are shown in Table 2 with a list of all genes in
-Supplemental Table X. The gene with by far the highest FPKM value
+\begin{wraptable}{I}{0.3\textwidth}
+  \tiny
+<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
+top.human.genes <- top.genes.table(species.name="homo sapiens")
+print(xtable(top.human.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.human.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+  \textit{Homo sapiens}}
+\label{tab:top50_human_placenta}
+\end{wraptable}
+
+The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
+the human are shown in \fref{tab:top50_human_placenta} with a list of all genes in
+Supplemental Table X. The gene with by far the highest \ac{FPKM} value
 (101962) in humans is chorionic somatomammotropin hormone 1 (CSH1),
@@
 within hominoids. Another member of the growth hormone gene family,
-GH2 (FPKM=729.639), is also expressed to a high degree in the human
+GH2 (\ac{FPKM}=729.639), is also expressed to a high degree in the human
 placenta (Table 2). Ten pregnancy-specific beta-1-glycoprotein genes
 are highly-expressed (FPKM>100) in the term placenta of humans, with
-PSG4 showing the greatest degree of expression (FPKM= 5201.31; Table
+PSG4 showing the greatest degree of expression (\ac{FPKM}= 5201.31; Table
 2). These genes are members of a hominoid-specific expansion of the
@@
 (HSD3B1) show exceptionally high expression in the human placenta
-(FPKM = 4817.15 and 829.846 respectively), as well as CGA (FPKM=
+(\ac{FPKM} = 4817.15 and 829.846 respectively), as well as CGA (\ac{FPKM}=
 4793.08), which generates the alpha subnit of multiple human chorionic
@@
 
-The top 50 expressed genes as measured by FPKM in the term placenta of
+\begin{wraptable}{O}{0.3\textwidth}
+  \tiny
+<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
+top.human.genes <- top.genes.table(species.name="pan paniscus")
+print(xtable(top.human.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.human.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+  \textit{Pan paniscus}}
+\label{tab:top50_bonobo_placenta}
+\end{wraptable}
+
+The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
 the bonobo are shown in Table 3 with a list of all genes in
@@
 somewhat from that of the human, as evidenced by the makeup of the top
-50 genes (Table 3). The highest expression (FPKM=10149.1) in bonobo is
+50 genes (Table 3). The highest expression (\ac{FPKM}=10149.1) in bonobo is
 proteoglycan 2, bone marrow (PRG2), which is involved in insulin
@@
 Glerup, 2011). PRG2 is also expressed at relatively high levels in the
-human placenta (FPKM=250.675; Supplemental Table X), although at
+human placenta (\ac{FPKM}=250.675; Supplemental Table X), although at
 comparatively much lower levels than the bonobo. A bonobo gene closely
 related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
-expressed at high levels (FPKM=1232.56; Table 3). Two other Pan genes
+expressed at high levels (\ac{FPKM}=1232.56; Table 3). Two other Pan genes
 included within this family (ENSPTRG00000030805 which shows similarity
-to GH2 and ENSPTRG00000034560) show expression levels of FPKM>89. Six
+to GH2 and ENSPTRG00000034560) show expression levels of \ac{FPKM}>89. Six
 pregnancy-specific beta-1-glycoprotein genes were expressed in the
-bonobo placenta with FPKM ranging from 20.885-167.961 (Supplemental
+bonobo placenta with \ac{FPKM} ranging from 20.885-167.961 (Supplemental
 Table X). Placental expression of the genes involved in estrogen and
 progesterone production (CYP19A1, HSD3B1 respectively) show lower
-expression in the bonobo as compared to humans (CYP19A1 FPKM= 48.6877;
-HSD3B1 FPKM= 54.5301; Table 8). In addition, CGA is expressed in the
-bonobo placenta (FPKM=202.629), but at much lower levels than humans
+expression in the bonobo as compared to humans (CYP19A1 \ac{FPKM}= 48.6877;
+HSD3B1 \ac{FPKM}= 54.5301; Table 8). In addition, CGA is expressed in the
+bonobo placenta (\ac{FPKM}=202.629), but at much lower levels than humans
 as well (Table 8). Two Ensembl-annotated beta subunits of the
@@
 
-The top 50 expressed genes as measured by FPKM in the term placenta of
+The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
 the dog are shown in Table 4 with a list of all genes in Supplemental
-Table X. The gene with the highest expression (FPKM=48399.9) in the
+Table X. The gene with the highest expression (\ac{FPKM}=48399.9) in the
 dog placenta is the hormone relaxin (RLN1). Other highly expressed
@@
 gene LGALS1, which shows evidence of immunosupression during pregnancy
-(Than et al., 2008), shows high expression in dog (FPKM= 1405.02). In
+(Than et al., 2008), shows high expression in dog (\ac{FPKM}= 1405.02). In
 addition, the gene SSLP1, a gene expressed in the placentas of cow and
 sheep but not present in primates (Ushizawa et al., 2009), is also
-expressed at high levels (FPKM= 1853.98). There is no placental
+expressed at high levels (\ac{FPKM}= 1853.98). There is no placental
 expression of the estrogen synthase gene (CYP19A1) in dogs, however
-expression of HSD3B1 is relatively high (FPKM=157.622).
+expression of HSD3B1 is relatively high (\ac{FPKM}=157.622).
 
@@
 
-The top 50 expressed genes as measured by FPKM in the term placenta of
+The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
 the dog are shown in Table 5 with a list of all genes in Supplemental
-Table X. The gene with the highest expression (FPKM=12488.8) was the
+Table X. The gene with the highest expression (\ac{FPKM}=12488.8) was the
 miRNA gene bta-mir-2887-1. This gene occurs in several copies in the
@@
 expression in the cow placenta (Table 4, Supplemental Table X)
-including PAG2 (FPKM=7255.34), PAG10 (2076.42), BT.83780 (1793.72),
+including PAG2 (\ac{FPKM}=7255.34), PAG10 (2076.42), BT.83780 (1793.72),
 PAG11 (962.19), PAG12 (469.751), BT.88769 (408.584), PAG1 (301.61),
 PAG4 (186.496), PAG6 (157.348), PAG8 (153.225), and PAG16 (103.573).
-In addition, 14 other PAGs were expressed with FPKM<40 (Supplemental
+In addition, 14 other PAGs were expressed with \ac{FPKM}<40 (Supplemental
 Table X). Other genes with high expression include genes involved with
@@
 PRP1 (632.766), PRP5 (396.832), PRP6 (345.296), PRP2 (195.771), in
-addition to 7 other placental prolactins with expression of FPKM<51
+addition to 7 other placental prolactins with expression of \ac{FPKM}<51
 (Table 5; Supplemental Table X). SSLP1 (also called SOLD1), a gene
@@
 involved in cell proliferation (Ushizawa et al., 2009, 2010), also
-shows evidence of high expression (FPKM=1001.45). There is some
+shows evidence of high expression (\ac{FPKM}=1001.45). There is some
 placental expression of the estrogen synthase gene (CYP19A1) in cows
-(75.1346), and comparatively greater expression of HSD3B1 (FPKM=
+(75.1346), and comparatively greater expression of HSD3B1 (\ac{FPKM}=
 157.622).
@@
 
-The top 50 expressed genes as measured by FPKM in the term placenta of
+The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
 the elephant are shown in Table 6 with a list of all genes in
 Supplemental Table X. The gene with the highest expression
-(FPKM=33439.5) was the natriuretic peptide precursor B (NPPB). The
-gene with the second highest expression (FPKM=20233) was the microRNA
+(\ac{FPKM}=33439.5) was the natriuretic peptide precursor B (NPPB). The
+gene with the second highest expression (\ac{FPKM}=20233) was the microRNA
 miR-675, which is imbedded in the first exon of the maternally
 imprinted long non-coding RNA gene H19 (Keniry et al., 2012), a gene
-also highly expressed in the elephant placenta (FPKM=7164.3). Five
+also highly expressed in the elephant placenta (\ac{FPKM}=7164.3). Five
 uncharacterized genes are expressed to a high degree in the elephant
@@
 of the estrogen synthase gene, CYP19A1, (FPKM<1) and the enzyme that
-produces progesterone HSD3B1 (FPKM=0). Other genes with high
+produces progesterone HSD3B1 (\ac{FPKM}=0). Other genes with high
 expression in the elephant placenta include genes involved in
@@
 
-The top 50 expressed genes as measured by FPKM in the term placenta of
+The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
 the opossum are shown in Table 7 with a list of all genes in
 Supplemental Table X. The gene with the highest expression
-(FPKM=41740.6) was hemoglobin, epsilon (HBE), which is expressed at
+(\ac{FPKM}=41740.6) was hemoglobin, epsilon (HBE), which is expressed at
 the embryonic stage of development in marsupials, as in other mammals
@@
 of two hemoglobin, zeta (HBZ) genes, which are also expressed to a
-high degree in the opossum placenta (FPKM=2452.19 and FPKM=1139.25).
-In addition, there is high expression (FPKM=5827.37) of the gene
+high degree in the opossum placenta (\ac{FPKM}=2452.19 and \ac{FPKM}=1139.25).
+In addition, there is high expression (\ac{FPKM}=5827.37) of the gene
 hemoglobin, omega (HBO), a hemoglobin found only in marsupials and
 monotremes (Opazo et al., 2008). The hemoglobin gene HBA is also
-expressed in the placenta (FPKM=456.978). The gene with the second
-highest expression (FPKM=22854.8) is superoxide dismutase 3,
+expressed in the placenta (\ac{FPKM}=456.978). The gene with the second
+highest expression (\ac{FPKM}=22854.8) is superoxide dismutase 3,
 extracellular (SOD3), which functions as an antioxidant. Four cystatin
@@
 uncharacterized gene XM\_001364916.1 also shows high expression in the
-opossum placenta (FPKM=13572.8). The gene WFDC2, a member of the whey
+opossum placenta (\ac{FPKM}=13572.8). The gene WFDC2, a member of the whey
 acidic protein (WAP) family is also expressed to a high degree in the
@@
 little production of CYP19A1 and HSD3B1 (FPKM<2), and moderate
-expression of relaxin RLN1 (FPKM= 41.3949).
+expression of relaxin RLN1 (\ac{FPKM}= 41.3949).
 
@@
 with relatively high expression values, we conducted DAVID analyses
-using a list of genes with FPKM values included within the top 5\%
+using a list of genes with \ac{FPKM} values included within the top 5\%
 distribution of expressed genes. Functional annotation clusters from
@@
 genes (Supplemental Table X). Significant enrichment in genes (ES>1.3,
-p>0.05) in the top 5\% of FPKM values in at least five transcriptomes
+p>0.05) in the top 5\% of \ac{FPKM} values in at least five transcriptomes
 included clusters such as ribosome, vesicle, cytoskeleton,
@@
 also expressed in other mammals to a lesser degree, especially those
-with invasive hemochorial placentas like bonobos (FPKM= 199.924),
-spider monkeys (FPKM= 96.24), and the mouse (FPKM= 80.7545)
+with invasive hemochorial placentas like bonobos (\ac{FPKM}= 199.924),
+spider monkeys (\ac{FPKM}= 96.24), and the mouse (\ac{FPKM}= 80.7545)
 (Supplemental Tables X-XX). ADAM12 increases the availability of
@@
 found evidence of adaptive evolution in ADAM12 on the primate lineage,
-and like GPC3 it is expressed in the bonobo (FPKM= 46.1553), spider
-monkey (FPKM=30.41) and mouse (FPKM=2.25391) as well, although to a
+and like GPC3 it is expressed in the bonobo (\ac{FPKM}= 46.1553), spider
+monkey (\ac{FPKM}=30.41) and mouse (\ac{FPKM}=2.25391) as well, although to a
 lesser degree (Supplemental Tables X-XX). RHOBTB3 is imprinted and
@@
 only expressed in the term placentas of humans (Table 2) and bonobos
-(FPKM=17.6394; Supplemental Table X). KISS1 is thought to be involved
+(\ac{FPKM}=17.6394; Supplemental Table X). KISS1 is thought to be involved
 in the regulation of trophoblast invasion, and shows highest

### 2015-10-01 data/fastq_fpkm/combine_fpkms.R
include star logfiles

@@ -129,7 +129,7 @@ for (file in star.log.files) {
     star.log <- read.table(file,sep="|",fill=TRUE,stringsAsFactors=FALSE)
     colnames(star.log) <- c("field","value")
     star.log$value <- gsub("\\t","",star.log$value)
-    star.log$field <- gsub("^\\s+","",star.log$field)
+    star.log$field <- gsub("(^\\s+|\\s+$)","",star.log$field)
     star.logs[[file]] <- data.table(star.log)[!grepl(":",field),]
     star.logs[[file]][,species := gsub("_"," ",
                                    gsub("_log_star.txt","",

### 2015-10-01 manuscript/manuscript.Rnw
include top dog and cow genes

@@
 \renewcommand\Affilfont{\itshape\small}
+\renewcommand\cite[1]{\autocite{#1}}
 \pagestyle{fancy}
@@
 \acrodef{SRA}{Sequence Read Archive}
-\acrodef{FPKM}{Fragments Per Kilobase of exon per Million reads \cite{Trapnell.ea2010:Transcriptassemblyandquantification}}
+\acrodef{FPKM}{Fragments Per Kilobase of exon per Million reads}
 
@@
 
-\subsection{Transcriptome Assembly and Mapping}
+\subsection{Transcriptome Assembly, Alignment, and Quantification}
 
@@
 \cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
-values in \ac{FPKM} were quantified using Cufflinks v2.2.1
+values in \ac{FPKM}
+\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
+quantified using Cufflinks v2.2.1
 \cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
@@
 
+\begin{figure}
+<<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
+print(ggplot(star.logs[grepl("%",field),],
+             aes(x=species,y=as.numeric(gsub("%","",value)),
+                 color=field))
+      + geom_boxplot()
+      + facet_grid(field~.,scales="free")
+      + ylab("Percent of Reads")
+      + theme(legend.position="top",
+              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1))
+      )
+@ 
+\caption{Transcript statistics for each species}
+\label{fig:transcriptome_statistics}
+\end{figure}
+
 <<transcriptome.statistics>>=
@@
     temp <- combined.fpkm[species==species.name &
-                      !grepl("^MT-?",gene_short_name)][order(-mean_fpkm)][!duplicated(gene_short_name)][1:n,list(gene_short_name,human_name,mean_fpkm)]
+                      !grepl("^MT-?",gene_short_name) & 
+                          !grepl("(RNA\\d|MT)-",human_name)
+                          ][order(-mean_fpkm)][!duplicated(gene_short_name)][1:n,list(gene_short_name,human_name,mean_fpkm)]
     setnames(temp,c("Gene name","Human Name","FPKM"))
@@
 
-\begin{wraptable}{I}{0.3\textwidth}
+\begin{wraptable}{R}{0.3\textwidth}
   \tiny
@@
 
-\begin{wraptable}{O}{0.3\textwidth}
+\begin{wraptable}{L}{0.3\textwidth}
   \tiny
 <<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
-top.human.genes <- top.genes.table(species.name="pan paniscus")
-print(xtable(top.human.genes),
+top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
+print(xtable(top.bonobo.genes),
       booktabs=TRUE,
@@
       )
-setkey(top.human.genes,"Gene name")
+setkey(top.bonobo.genes,"Gene name")
 @ 
@@
 
+\begin{wraptable}{R}{0.3\textwidth}
+  \tiny
+<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
+top.dog.genes <- top.genes.table(species.name="canis familiaris")
+print(xtable(top.dog.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.dog.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+  \textit{Canis familiaris}}
+\label{tab:top50_dog_placenta}
+\end{wraptable}
+
 The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
@@
 \subsection{Cow Placenta Transcriptome}
+
+\begin{wraptable}{L}{0.3\textwidth}
+  \tiny
+<<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
+top.cow.genes <- top.genes.table(species.name="bos taurus")
+print(xtable(top.cow.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.cow.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+  \textit{Bos taurus}}
+\label{tab:top50_cow_placenta}
+\end{wraptable}
 

### 2015-10-01 data/fastq_fpkm/combine_fpkms.R
fix grep of star logs to only exclude the separator lines, not all lines containing :

@@ -130,7 +130,7 @@ for (file in star.log.files) {
     colnames(star.log) <- c("field","value")
     star.log$value <- gsub("\\t","",star.log$value)
     star.log$field <- gsub("(^\\s+|\\s+$)","",star.log$field)
-    star.logs[[file]] <- data.table(star.log)[!grepl(":",field),]
+    star.logs[[file]] <- data.table(star.log)[!grepl(":$",field),]
     star.logs[[file]][,species := gsub("_"," ",
                                    gsub("_log_star.txt","",
                                         gsub("SRR\\d+_","",file)))]

### 2015-10-02 data/combine_fpkm.R
also save star logs

@@ -92,5 +92,5 @@ gene.fpkms[is.na(human_name),
 ### data.table code
 combined.fpkm <- data.table(data.frame(gene.fpkms))
 
-save(file=output.file,combined.fpkm)
+save(file=output.file,combined.fpkm,star.logs)
 

### 2015-10-02 manuscript/manuscript.Rnw
add table of sequencing results and figure of read efficiency

@@
 library("xtable")
+library("grid")
+
+capwords <- function(s, strict = FALSE) {
+    cap <- function(s) paste(toupper(substring(s, 1, 1)),
+                             {s <- substring(s, 2); if(strict) tolower(s) else s},
+                             sep = "", collapse = " " )
+    sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
+}
+
+capfirst <- function(s, strict = FALSE) {
+    paste(toupper(substring(s, 1, 1)),
+          {s <- substring(s, 2); if(strict) tolower(s) else s},
+          sep = "")
+}
+
 @
@@
 }
+star.logs[,value:=gsub("%$","",value)]
 @
@@
 obstetrical syndromes. We profiled the term placenta by using Illumina
-RNA-Seq technology to sequence the transcriptomes of 6 species of
-therian (i.e. marsupials and “placental”) mammals. We identified
-significant shifts in gene expression along specific evolutionary
-lineages using a model of character evolution. Genes that showed
-significant elevation in expression along the human lineage
-($\mathrm{FDR}\le 0.01$) were enriched in GO categories such as
-angiogenesis, tube morphogenesis, cell migration, and regulation of
-inflammatory response. Three of these genes (KISS1, ADAM12, GPC3)
-showed exceptionally high expression in humans; these genes are
-implicated in preeclampsia and may indicate that the risk of
-preeclampsia recently emerged on the human lineage.
-
-Keywords: mammal, gene expression, evolution, RNA-seq, placenta
+RNA-Seq technology to sequence the transcriptomes of
+\Sexpr{star.logs[,length(unique(species))]} species of therian (i.e.
+marsupials and “placental”) mammals. We identified significant shifts
+in gene expression along specific evolutionary lineages using a model
+of character evolution. Genes that showed significant elevation in
+expression along the human lineage ($\mathrm{FDR}\le 0.01$) were
+enriched in GO categories such as angiogenesis, tube morphogenesis,
+cell migration, and regulation of inflammatory response. Three of
+these genes (KISS1, ADAM12, GPC3) showed exceptionally high expression
+in humans; these genes are implicated in preeclampsia and may indicate
+that the risk of preeclampsia recently emerged on the human lineage.
+
+\textbf{Keywords:} mammal, gene expression, evolution, RNA-seq, placenta
 
@@
     \toprule
-    Species                       & Common Name        & Reference & Annotation \\
+    Species                       & Common Name        & Reference & Read Length  \\
     \midrule
-    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & Ensembl 80 \\
-    \textit{Ateles fusciceps}     & Spider monkey      & ---        & \textit{Homo sapiens} Ensembl 80 \\
-    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & Ensembl 80 \\
-    \textit{Bos taurus}           & Cow                & UMD3.1     & Ensembl 80 \\
-    \textit{Canis familiaris}     & Dog                & CanFam3.1  & Ensembl 80 \\
-    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & Ensembl 80 \\
-    \textit{Homo sapiens}         & Human              & GRCh38     & Ensembl 80 \\
-    \textit{Mus musculus}         & Mouse              & GRCm38     & Ensembl 80 \\
-    \textit{Nannospalax galili}   & Blind mole rat     & ----       & \textit{Mus musculus} Ensembl 80 \\
-    \textit{Monodelphis domestica} & Opossum           & BROAD05    & Ensembl 80 \\
-    \textit{Spalax carmeli}       & Blind mole rat     & ----       & \textit{Mus musculus} Ensembl 80 \\
+    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
+    \textit{Bos taurus}           & Cow                & UMD3.1     & 76 \\
+    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 151 \\
+    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 101 \\
+    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 76 \\
+    \textit{Homo sapiens}         & Human              & GRCh38     & 76, 151 \\
+    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 76 \\
+    \textit{Mus musculus}         & Mouse              & GRCm38     & 36,44,151 \\
+    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
+    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 151 \\
+    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
     \bottomrule
@@
 <<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
-print(ggplot(star.logs[grepl("%",field),],
-             aes(x=species,y=as.numeric(gsub("%","",value)),
-                 color=field))
+star.logs.unmapped <- 
+    star.logs[grepl("%",field) & 
+                  grepl("unmapped",field) & 
+                      !grepl("too many mismatches",field),]
+star.logs.unmapped <- 
+    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
+pushViewport(viewport(layout=grid.layout(ncol=1,nrow=2,heights=c(2,1))))
+print(ggplot(star.logs.unmapped,
+             aes(x=capfirst(species),y=as.numeric(gsub("%","",value))))
       + geom_boxplot()
@@
       + ylab("Percent of Reads")
+      + xlab("Species")
       + theme(legend.position="top",
-              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1))
+              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
+      vp=viewport(layout.pos.col=1,layout.pos.row=1)
       )
+pushViewport(viewport(layout.pos.col=1,layout.pos.row=1))
+grid.text("A",
+          y=unit(1,"npc")-unit(0.1,"cm"),
+          x=unit(0,"npc")+unit(0.1,"cm"),
+          gp=gpar(fontsize=24),
+          just=c("left","top"))
+popViewport()          
+print(ggplot(star.logs[grepl("input reads$",field),],
+             aes(x=capfirst(species),y=as.numeric(value)))
+      + geom_boxplot()
+      + scale_y_log10()
+      + ylab("Reads")
+      + xlab("Species")
+      + theme(legend.position="top",
+              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
+      vp=viewport(layout.pos.col=1,layout.pos.row=2))
+pushViewport(viewport(layout.pos.col=1,layout.pos.row=2))
+grid.text("B",
+          y=unit(1,"npc")-unit(0.1,"cm"),
+          x=unit(0,"npc")+unit(0.1,"cm"),
+          gp=gpar(fontsize=24),
+          just=c("left","top"))
+popViewport(2)
 @ 
-\caption{Transcript statistics for each species}
+\caption{RNAseq read and alignment statistics for all species. A)
+  Boxplot of percentages of unmapped reads in each species; 0\% of
+  reads in all species were unmapped because of mismatches. B) Boxplot
+  of number of RNAseq reads in each species.}
 \label{fig:transcriptome_statistics}
 \end{figure}
-
-<<transcriptome.statistics>>=
-@
 

### 2015-10-05 manuscript/manuscript.Rnw
add more plots for transcriptome stats

@@
 })
+options(scipen = -2, digits = 3)
 library("data.table")
@@
 library("grid")
+library("gridExtra")
 
@@
 
+species <- function(s,join=TRUE) {
+    temp <- paste0("\\textit{",capfirst(s),"}")
+    if (join) {
+        return(english.join(temp))
+    } else {
+        return(temp)
+    }
+}
+
+english.join <- function(s){
+    ret <- c()
+    ## everything but the last gets a , appended if there are more than two
+    if (length(s) > 2) {
+        s[-length(s)] <- 
+            paste0(s[-length(s)],",")
+    }
+    ## the second to last gets an and appended if there are more than one
+    if (length(s) > 1) {
+        s[length(s)-1] <- 
+            paste0(s[length(s)-1]," and")
+    }
+    return(paste(sep="",collapse=" ",s))
+}
+
 @
@@
 star.logs[,value:=gsub("%$","",value)]
+star.logs.wide <- 
+    data.table(dcast(star.logs,species+file~field))
+setnames(star.logs.wide,
+         "Average input read length",
+         "input.read.length")
+setnames(star.logs.wide,
+         "Number of input reads",
+         "input.reads")
+setnames(star.logs.wide,
+         "Uniquely mapped reads number",
+         "unique.map.reads")
 @
@@
       + ylab("Percent of Reads")
-      + xlab("Species")
+      + xlab(NULL)
+      + theme_bw()
       + theme(legend.position="top",
-              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
+              plot.margin=unit(c(0,0,0,1.1),"strwidth",data=list("","","","e8")),
+              axis.title.x=element_blank(),
+              axis.text.x = element_blank()),#element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
       vp=viewport(layout.pos.col=1,layout.pos.row=1)
@@
       + xlab("Species")
+      + theme_bw()
       + theme(legend.position="top",
+              plot.margin=unit(c(0,1.6,0,0),"strheight",data=list("","TO","","")),
               axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
@@
 
-For transcriptomes sequenced with 75 bp, (opossum, elephant, and cow),
-the number of reads ranged from 3.61x107 to 3.70x107 and number of
-bases ranged from 1.43x109 to 1.64x109 with at least 96\% of reads
-mapping to the respective reference genome (Table 1). For
-transcriptomes sequenced with 150 bp (human, bonobo, and dog), number
-of reads ranged from 6.47x107 to 9.27x107 and number of bases ranged
-from 5.20x109 to 6.25x109 with at least 99\% of reads mapping to the
-respective reference genome. All reads were submitted to the NCBI
-Sequence Read Archive (SRA) (Accession Number XXX-YYY).
+\begin{figure}
+<<reads.vs.read.length,fig.width=8,fig.height=8,echo=FALSE>>=
+print(ggplot(star.logs.wide,
+             aes(x=as.numeric(input.read.length),
+                 y=as.numeric(input.reads),
+                 color=capfirst(species),
+                 shape=capfirst(species)
+                 ))
+      + scale_shape_manual(values=1:nlevels(as.factor(star.logs.wide[,species])),
+                           guide=guide_legend(title="Species")
+                           )
+      + scale_color_discrete(guide=guide_legend(title="Species")
+                             )
+      + geom_point(size=5)
+      + xlab("Trimmed Read Length")
+      + ylab("Read number")
+      + scale_y_log10()
+      + theme(legend.text=element_text(face="italic")
+              )
+      )
+@ 
+\caption{RNA Read number as a function of trimmed read length}
+\label{fig:transcriptome_reads_vs_read_length}
+\end{figure}
+
+For transcriptomes sequenced with 75 bp, %
+(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
+%
+,
+the number of trimmed reads ranged from %
+\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
+%
+to %
+\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
+with at least 96\% of reads \DLA{This is wrong}
+mapping to the respective reference genome (\fref{tab:ref_genome_aligned_species}).
+
+For transcriptomes sequenced with 150 bp, %
+% The mismatch here is because we're reporting trimmed reads here, not
+% full length reads
+(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
+%
+,
+the number of trimmed reads ranged from %
+\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
+%
+to %
+\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
+%
+with at least 99\% of reads mapping to the
+respective reference genome \DLA{This is also wrong}.
+%
+All reads were submitted to the NCBI Sequence Read Archive (SRA)
+(Accession Number XXX-YYY).
+
+For transcriptomes sequenced with 35 bp, %
+% The mismatch here is because we're reporting trimmed reads here, not
+% full length reads
+(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
+%
+,
+the number of trimmed reads ranged from %
+\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
+%
+to %
+\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
+%
+with at least 99\% of reads mapping to the
+respective reference genome \DLA{This is also wrong}.
+%
+All reads were submitted to the NCBI Sequence Read Archive (SRA)
+(Accession Number XXX-YYY).
 

### 2015-10-06 common_r_code/plot_tree_matrix.R
add plot tree matrix code

@@ -1,0 +1,45 @@
+plot.tree.matrix <- function(gene.tree,gene.tree.table,expression) {
+    gene.tree$tip.label <- gsub("^(ENS.+)\\1$","\\1",gene.tree$tip.label)
+    gene.tree.table[,species:=capfirst(gsub("_"," ",species))]
+    gene.tree.table[,symbol_or_id:=ifelse(symbol=="NULL",gene_id,paste0(symbol," ",species))]
+    gene.tree.expression <-
+        combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+    setkey(gene.tree.table,"protein_id")
+    gene.tree$tip.label <-
+        gene.tree.table[gene.tree$tip.label,gene_id]
+    gene.tree.subset <-
+        drop.tip(gene.tree,gene.tree$tip.label[!(gene.tree$tip.label %in%
+                                                     gene.tree.expression[,gene_id])])
+
+    gene.tree.expression.matrix <-
+        dcast(gene.tree.expression,
+              gene_id~species,value.var="mean_fpkm")
+    rownames(gene.tree.expression.matrix) <-
+        gene.tree.expression.matrix[,1]
+    gene.tree.expression.matrix <-
+        gene.tree.expression.matrix[,-1]
+    colnames(gene.tree.expression.matrix) <-
+        capfirst(colnames(gene.tree.expression.matrix))
+    colnames(gene.tree.expression.matrix) <-
+        gsub("(^)(.)[^ ]*","\\2.",colnames(gene.tree.expression.matrix))
+    colnames(gene.tree.expression.matrix) <-
+        gsub("( )(..)[^ ]*","\\2.",colnames(gene.tree.expression.matrix))
+
+
+    setkey(gene.tree.table,"gene_id")
+    rownames(gene.tree.expression.matrix) <-
+        gene.tree.table[rownames(gene.tree.expression.matrix),symbol_or_id]
+        
+    
+    gene.tree.subset$tip.label <-
+        gene.tree.table[gene.tree.subset$tip.label,symbol_or_id]
+    p <- ggtree(gene.tree.subset)# %>% add_legend(x=2008, y=5)
+    p <- p + geom_tiplab(size=3,align=TRUE)
+    
+    
+    print(gheatmap(p,data=gene.tree.expression.matrix,offset=0.3,colnames=TRUE) +
+              theme(legend.title=element_text())
+          + scale_fill_gradient(low="white",high="green",na.value="gray50",
+                                name="FPKM")
+          )    
+}

### 2015-10-06 data/fastq_fpkm/combine_fpkms.R
do not sum un-annotated transcripts

@@ -68,7 +68,6 @@ for (file in gene.files) {
         gsub("_"," ",
              gsub("_genes.fpkm_tracking","",
                   gsub("SRR\\d+_","",file)))
-    gene.fpkms[[file]][,species := species.name]
     if (any(species.name %in% trinity.diamond.species)) {
         ## need to merge in the gene names here and then merge down
         ## the calls on the genes to only a single gene
@@ -82,18 +81,19 @@ for (file in gene.files) {
             trinity.diamond[[diamond.file]][gene.fpkms[[file]]]
         ## sum the FPKM in order to determine the total expression for
         ## this gene
-        gene.fpkms[[file]][,
+        gene.fpkms[[file]][!is.na(gene_id),
                            FPKM:=sum(FPKM),gene_id]
-        gene.fpkms[[file]][,
+        gene.fpkms[[file]][!is.na(gene_id),
                            FPKM_conf_lo:=sum(FPKM_conf_lo),gene_id]
-        gene.fpkms[[file]][,
+        gene.fpkms[[file]][!is.na(gene_id),
                            FPKM_conf_hi:=sum(FPKM_conf_hi),gene_id]
         gene.fpkms[[file]] <-
-            gene.fpkms[[file]][!duplicated(gene_id),]
+            gene.fpkms[[file]][is.na(gene_id) | !duplicated(gene_id),]
         gene.fpkms[[file]][,i.gene_id:=NULL]
         gene.fpkms[[file]][,i.gene_short_id:=NULL]
         gene.fpkms[[file]][,i.species:=NULL]
     }
+    gene.fpkms[[file]][,species := species.name]
     gene.fpkms[[file]][,file:=file]
     i <- i+1
     setTxtProgressBar(pb,i)

### 2015-10-06 data/combine_fpkm.R
include all of the oma groups, build long orthologs file

@@ -92,5 +92,17 @@ gene.fpkms[is.na(human_name),
 ### data.table code
 combined.fpkm <- data.table(data.frame(gene.fpkms))
 
+setkey(oma.groups,"ensembl")
+setkey(combined.fpkm,"tracking_id")
+combined.fpkm <- oma.groups[,list(group_num,fingerprint,ensembl)][combined.fpkm]
+combined.fpkm[,ensembl:=NULL]
+setnames(combined.fpkm,"group_num","oma_group_num")
+
+combined.fpkm[,oma_group_name:=NA]
+combined.fpkm[!is.na(oma_group_num),
+              oma_group_name:=paste(collapse=",",
+                  grep("^-$",unique(gene_short_name),value=TRUE,invert=TRUE)),
+              by=as.numeric(oma_group_num)]
+
 save(file=output.file,combined.fpkm,star.logs)
 

### 2015-10-06 manuscript/manuscript.Rnw
add start of tree matrix plots

@@
 library("gridExtra")
+library("ggtree")
 
@@
 
-<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R")))>>=
-for (a in paste0("../common_r_code/",c("to_latex.R"))) {
+<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R")))>>=
+for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R"))) {
     source(a)
@@
 
-<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm")))>>=
-for (a in paste0("../data/",c("combined_fpkm"))) {
+<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees")))>>=
+for (a in paste0("../data/",c("combined_fpkm","gene_trees"))) {
     load(a)
 }
+setkey(genes.to.tree,"gene_id")
 star.logs[,value:=gsub("%$","",value)]
@@
 RNA-Seq data for the placenta transcriptome of the black-headed spider
-monkey (Ateles fusciceps) using similar protocols discussed above.
+monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
 Ateles is outside the clade, which includes hominids and can be used
@@
 mapping to the respective reference genome (\fref{tab:ref_genome_aligned_species}). 
-
+%
 For transcriptomes sequenced with 150 bp, %
@@
 %
-All reads were submitted to the NCBI Sequence Read Archive (SRA)
-(Accession Number XXX-YYY).
-
+%
 For transcriptomes sequenced with 35 bp, %
@@
 
+\subsection{Spider Monkey Transcriptome}
+
+\begin{wraptable}{L}{0.3\textwidth}
+  \tiny
+<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
+top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
+print(xtable(top.spider.monkey.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.spider.monkey.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+  \textit{Ateles fusciceps}}
+\label{tab:top50_spider_monkey_placenta}
+\end{wraptable}
+
 \subsection{Dog Placenta Transcriptome}
@@
 
+\begin{figure}
+<<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of prolactin}
+\label{fig:prolactin_evolution_tree}
+\end{figure}
+
 The elephant is strikingly the only mammal analyzed that shows high
@@
 \subsection{The Evolution of Placental Expression of Galectins}
+
+\begin{figure}
+<<galectins.evolution.tree,echo=FALSE,fig.width=8,fig.height=8>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of galectins}
+\label{fig:galectins_evolution_tree}
+\end{figure}
 

### 2015-10-08 manuscript/acronyms.tex
move acronyms to their own file

\acrodef{EV}[EV]{Extracellular Vesicle}
\acrodef{SVM}{Support Vector Machine}
\acrodef{ROC}{Receiver Operating Characteristic}
\acrodef{KNN}{K Nearest Neighbors}
\acrodef{SRA}{Sequence Read Archive}
\acrodef{PRG2}[\textit{PRG2}]{Proteoglycan 2, bone marrow}
\acrodef{FPKM}{Fragments Per Kilobase of exon per Million reads}
\acrodef{CSH1}[\textit{CSH1}]{chorionic somatomammotropin hormone 1}
\acrodef{CSH2}[\textit{CSH2}]{chorionic somatomammotropin hormone 2}
\acrodef{CSHL1}[\textit{CSHL1}]{chorionic somatomammotropin hormone-like 1}
\acrodef{GH2}[\textit{GH2}]{growth hormone 2}

%%% Local Variables:
%%% mode: latex
%%% TeX-master: "manuscript.Rnw"
%%% End:

### 2015-10-08 manuscript/manuscript.Rnw
move acronyms to their own file
calculate core placenta transcriptome

@@
 % Acronyms
-\acrodef{EV}[EV]{Extracellular Vesicle}
-\acrodef{SVM}{Support Vector Machine}
-\acrodef{ROC}{Receiver Operating Characteristic}
-\acrodef{KNN}{K Nearest Neighbors}
-\acrodef{SRA}{Sequence Read Archive}
-\acrodef{FPKM}{Fragments Per Kilobase of exon per Million reads}
-
+\input{acronyms}
 % \twocolumn
@@
 
-<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees")))>>=
-for (a in paste0("../data/",c("combined_fpkm","gene_trees"))) {
+<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset")))>>=
+for (a in paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset"))) {
     load(a)
@@
 
+\subsection{Core Placenta Transcriptome}
+
+\begin{figure}
+<<core.placenta.transcriptome,fig.width=8,fig.height=12,echo=FALSE>>=
+combined.fpkm.wide <- 
+    data.table(dcast(combined.fpkm,
+                     human_name~species,
+                     fun.aggregate=sum,
+                     value.var="mean_fpkm"))
+temp <- 
+    combined.fpkm.wide[,median_expression := apply(combined.fpkm.wide[,-1,with=FALSE],1,median)]
+temp <- 
+    combined.fpkm.wide[,min_expression := apply(combined.fpkm.wide[,-1,with=FALSE],1,min)]
+setkey(combined.fpkm.wide,"human_name")
+core.placenta.transcriptome.long <- 
+    data.table(melt(combined.fpkm.wide[min_expression >= 10,
+                                       ][order(-median_expression)
+                                         ][!(human_name %in% 
+                                                 housekeeping.genes.superset[,gene_short_name])],
+                    id.vars="human_name",
+                    variable.name="Species",
+                    value.name="FPKM"
+                    ))
+temp <- core.placenta.transcriptome.long[,human_name:=
+                                     reorder(human_name,
+                                             combined.fpkm.wide[core.placenta.transcriptome.long[,human_name],median_expression])]
+rm(temp)
+
+print(ggplot(core.placenta.transcriptome.long[!Species %in% c("median_expression","min_expression"),],
+             aes(y=human_name,x=capfirst(Species)))
+      + geom_tile(aes(fill=log10(FPKM)))
+      + scale_x_discrete(expand=c(0,0))
+      + scale_y_discrete(expand=c(0,0))
+      + scale_fill_gradient(low="white",
+                            high="blue",
+                            limits=c(0,4),
+                            guide=guide_colorbar(title=expression(log[10](FPKM)))
+                            )
+      + ylab("Human Ortholog Name")
+      + xlab("Species")
+      + theme_grey()
+      + theme(legend.position="top",
+              axis.ticks = element_blank(), 
+              axis.text.x = element_text(angle = 340, hjust = 0.3, vjust=0.8,face="italic"))
+      )
+
+# 
+# print(xtable(combined.fpkm.wide[min_expression > 10,
+#                                 ][order(-median_expression)
+#                                   ][!(human_name %in% 
+#                                           housekeeping.genes.superset[,gene_short_name])][1:75,],
+#              digits=0
+#              ),
+#       include.rownames=FALSE,
+#       booktabs=TRUE,
+#       floating=FALSE
+#       )
+@ 
+\caption{Core Placenta Transcriptome.
+  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
+  genes expressed (\ac{FPKM}$\ge 10$) in the placenta of all organisms
+  studied with housekeeping
+  genes\cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
+  removed.}
+\label{fig:core_placenta_transcriptome}
+\end{figure}
+
+Annexins?
+
+TIMP3 -- maybe in pre-eclampsia (23172037)
+SERPINH1? with TIMP3 
+% http://www.proteinatlas.org/ENSG00000149257-SERPINH1/tissue/placenta
+
+Why are RPL17 and RPS26 here? They are ribosomal proteins. These
+should be missing now.
+
 \subsection{DAVID Analyses of Individual Transcriptomes}

### 2015-10-08 manuscript/manuscript.Rnw
add expression tree figure

@@
 })
-options(scipen = -2, digits = 3)
+options(scipen = -2, digits = 2)
 library("data.table")
@@
 library("gridExtra")
+library("ape")
 library("ggtree")
@@
 top.genes.table <- function(species.name,n=50) {
-    temp <- combined.fpkm[species==species.name &
-                      !grepl("^MT-?",gene_short_name) & 
+    temp <- 
+        combined.fpkm[species==species.name &
+                          !(human_name %in% housekeeping.genes.superset[,gene_short_name]) &
                               !grepl("(RNA\\d|MT)-",human_name)
-                          ][order(-mean_fpkm)][!duplicated(gene_short_name)][1:n,list(gene_short_name,human_name,mean_fpkm)]
+                      ][order(-mean_fpkm)][!duplicated(gene_short_name)][1:n,list(gene_short_name,
+                                                                                  human_name,mean_fpkm)]
     setnames(temp,c("Gene name","Human Name","FPKM"))
@@
 }
+gene.expression <- function(species.name,gene) {
+    combined.fpkm[species==species.name &
+                      gene_short_name==gene,mean_fpkm]
+}
+
 @
@@
 
-\begin{wraptable}{L}{0.3\textwidth}
+\begin{wraptable}{R}{0.3\textwidth}
   \tiny
@@
 \begin{figure}
-<<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8>>=
+<<prolactin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
 
@@
 \begin{figure}
-<<galectins.evolution.tree,echo=FALSE,fig.width=8,fig.height=8>>=
+<<galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
 

### 2015-10-08 manuscript/manuscript.Rnw
update the bonobo description and homo sapiens description

@@
 
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the human are shown in \fref{tab:top50_human_placenta} with a list of all genes in
+The top 50 expressed genes as measured by \ac{FPKM} in the term
+placenta of \textit{Homo sapiens} are shown in
+\fref{tab:top50_human_placenta} with a list of all genes in
 Supplemental Table X. The gene with by far the highest \ac{FPKM} value
-(101962) in humans is chorionic somatomammotropin hormone 1 (CSH1),
-followed by chorionic somatomammotropin hormone 2 (CSH2), and
-chorionic somatomammotropin hormone-like 1 (CSHL1), all three of which
-are members of a distinct expansion of the growth hormone gene family
-within hominoids. Another member of the growth hormone gene family,
-GH2 (\ac{FPKM}=729.639), is also expressed to a high degree in the human
-placenta (Table 2). Ten pregnancy-specific beta-1-glycoprotein genes
-are highly-expressed (FPKM>100) in the term placenta of humans, with
-PSG4 showing the greatest degree of expression (\ac{FPKM}= 5201.31; Table
-2). These genes are members of a hominoid-specific expansion of the
+(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
+in humans is \ac{CSH1}, followed by \ac{CSH2}, and \ac{CSHL1}, all
+three of which are members of a distinct expansion of the growth
+hormone gene family within hominoids. Another member of the growth
+hormone gene family, \ac{GH2} (\ac{FPKM}=%
+\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
+is also expressed to a high degree in the human placenta
+(\fref{tab:top50_human_placenta}).
+% 
+Ten pregnancy-specific beta-1-glycoprotein genes are highly-expressed
+(\ac{FPKM}>100) in the term placenta of humans, with \ac{PSG4} showing the
+greatest degree of expression %
+(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG4")}). % 
+These
+genes are members of a hominoid-specific expansion of the
 carcinoembryonic antigen (CEA) family and known to be expressed by the
@@
 (HSD3B1) show exceptionally high expression in the human placenta
-(\ac{FPKM} = 4817.15 and 829.846 respectively), as well as CGA (\ac{FPKM}=
-4793.08), which generates the alpha subnit of multiple human chorionic
-gonadotropin hormones (Table 2). The genes producing the beta subunits
-of the chorionic gonadotropins (CGB, CGB1, CGB2, CGB5, CGB7, CGB8) are
-either not expressed (CGB8) or expressed modestly (FPKM≤20.826; Table
-8). Other highly expressed genes in the top 50 include members of the
-hemoglobin family (e.g. HBA1, HBB, HBG2, HBA2), the maternally
-imprinted long noncoding RNA gene H19, genes coding for proteins
-required for growth and/or cell proliferation (ADAM12, KISS1, IGF2,
-GPC3, CSF3R), anticoagulation (TFPI2), and glucose transport (SLC2A1)
-among others (Table 2, Supplemental Table X)
+(\ac{FPKM} = 4817.15 and 829.846 respectively), as well as CGA
+(\ac{FPKM}= 4793.08), which generates the alpha subnit of multiple
+human chorionic gonadotropin hormones (Table 2). The genes producing
+the beta subunits of the chorionic gonadotropins (CGB, CGB1, CGB2,
+CGB5, CGB7, CGB8) are either not expressed (CGB8) or expressed
+modestly (FPKM≤20.826; Table 8). Other highly expressed genes in the
+top 50 include members of the hemoglobin family (e.g. HBA1, HBB, HBG2,
+HBA2), the maternally imprinted long noncoding RNA gene H19, genes
+coding for proteins required for growth and/or cell proliferation
+(ADAM12, KISS1, IGF2, GPC3, CSF3R), anticoagulation (TFPI2), and
+glucose transport (SLC2A1) among others (Table 2, Supplemental Table
+X)
 
@@
 The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the bonobo are shown in Table 3 with a list of all genes in
+the bonobo are shown in \fref{tab:top50_bonobo_placenta} with a list of all genes in
 Supplemental Table X. Expression in the bonobo placenta differs
 somewhat from that of the human, as evidenced by the makeup of the top
-50 genes (Table 3). The highest expression (\ac{FPKM}=10149.1) in bonobo is
-proteoglycan 2, bone marrow (PRG2), which is involved in insulin
-growth factor signaling and angiogenesis during pregnancy (Weyer and
-Glerup, 2011). PRG2 is also expressed at relatively high levels in the
-human placenta (\ac{FPKM}=250.675; Supplemental Table X), although at
+50 genes (Table 3). The highest expression (\ac{FPKM}=%
+\Sexpr{gene.expression("pan paniscus","PRG2")}%
+%
+) in bonobo is \ac{PRG2}, which is involved in insulin
+growth factor signaling and angiogenesis during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}. \ac{PRG2} is also expressed at relatively high levels in the
+human placenta (\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")}; Supplemental Table X), although at
 comparatively much lower levels than the bonobo. A bonobo gene closely
 related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
-expressed at high levels (\ac{FPKM}=1232.56; Table 3). Two other Pan genes
+expressed at high levels (\ac{FPKM}=1232.56; \fref{tab:top50_bonobo_placenta}). Two other Pan genes
 included within this family (ENSPTRG00000030805 which shows similarity

### 2015-10-08 common_r_code/plot_tree_matrix.R
plot the log of FPKM

@@ -11,9 +11,10 @@ plot.tree.matrix <- function(gene.tree,gene.tree.table,expression) {
         drop.tip(gene.tree,gene.tree$tip.label[!(gene.tree$tip.label %in%
                                                      gene.tree.expression[,gene_id])])
 
+    gene.tree.expression[,mean_fpkm_log:=log10(mean_fpkm+1)]
     gene.tree.expression.matrix <-
         dcast(gene.tree.expression,
-              gene_id~species,value.var="mean_fpkm")
+              gene_id~species,value.var="mean_fpkm_log")
     rownames(gene.tree.expression.matrix) <-
         gene.tree.expression.matrix[,1]
     gene.tree.expression.matrix <-
@@ -40,6 +41,6 @@ plot.tree.matrix <- function(gene.tree,gene.tree.table,expression) {
     print(gheatmap(p,data=gene.tree.expression.matrix,offset=0.3,colnames=TRUE) +
               theme(legend.title=element_text())
           + scale_fill_gradient(low="white",high="green",na.value="gray50",
-                                name="FPKM")
-          )    
+                                name=expression(log[10](FPKM)))
+          )
 }

### 2015-10-13 manuscript/supplemental_material.Rnw
add supplemental material

--- 2015-10-08 manuscript/manuscript.Rnw
@@
 \begin{document}
-\author[1,2]{Michael R. McGowen}
-\author[1]{Don Armstrong}
-\author[1]{Amy Weckle}
-\author[1]{Jason Caravas}
-\author[3]{Dalen Agnew}
-\author[4]{Kurt Benirschke}
-\author[5]{Sue Savage-Rumbaugh}
-\author[6,7]{Chong Jai Kim}
-\author[8]{Günter P. Wagner}
-\author[6]{Roberto Romero}
-\author[1,6,9,*]{Derek E. Wildman}
-\title{The evolution of gene expression in the term placenta of viviparous mammals}
-% These are all wrong, but we'll change them later.
-\affil[1]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
-\affil[2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
-\affil[3]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
-\affil[4]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
-\affil[5]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
-\affil[6]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
-\affil[7]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
-\affil[8]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
-\affil[9]{Department of Obstetrics and Gynecology, Wayne State University School of Medicine, Detroit, MI, USA}
-\affil[*]{To whom correspondence should be addressed}
-\date{\today}
 
-% Acronyms
-\input{acronyms}
-% \twocolumn
-\maketitle
-\onehalfspacing
+\makeatletter 
+\renewcommand{\thefigure}{S\@arabic\c@figure} 
+\makeatother
+
+\makeatletter 
+\renewcommand{\thetable}{S\@arabic\c@table} 
+\makeatother
+
+\title{Supplemental Material for: The evolution of gene expression in the term placenta of viviparous mammals}
+
+% \maketitle
 
@@
 library("gridExtra")
+library("ape")
 library("ggtree")
+library("topGO")
+library("org.Hs.eg.db")
 
@@
 
-<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset")))>>=
-for (a in paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset"))) {
+<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","palcenta_go_results_mf")))>>=
+for (a in paste0("../data/",c("combined_fpkm","gene_trees",
+                              "housekeeping_genes_superset",
+                              "placenta_core_transcriptome",
+                              "placenta_go_results_bp",
+                              "placenta_go_results_mf"
+                              ))) {
     load(a)
@@
 @
-
-\section{Abstract}
-
-The placenta is arguably the most anatomically variable organ in
-mammals. Understanding gene expression patterns among placentas from
-multiple species can illuminate the molecular basis of placenta
-function as well as potentially enrich our knowledge of the biology of
-obstetrical syndromes. We profiled the term placenta by using Illumina
-RNA-Seq technology to sequence the transcriptomes of
-\Sexpr{star.logs[,length(unique(species))]} species of therian (i.e.
-marsupials and “placental”) mammals. We identified significant shifts
-in gene expression along specific evolutionary lineages using a model
-of character evolution. Genes that showed significant elevation in
-expression along the human lineage ($\mathrm{FDR}\le 0.01$) were
-enriched in GO categories such as angiogenesis, tube morphogenesis,
-cell migration, and regulation of inflammatory response. Three of
-these genes (KISS1, ADAM12, GPC3) showed exceptionally high expression
-in humans; these genes are implicated in preeclampsia and may indicate
-that the risk of preeclampsia recently emerged on the human lineage.
-
-\textbf{Keywords:} mammal, gene expression, evolution, RNA-seq, placenta
-
-\section{Introduction}
-
-The main role of the placenta in viviparous mammals is the transfer of
-nutrients, oxygen, and waste between mother and offspring. Although
-these functions are conserved among mammals, the placenta is perhaps
-the most variable organ in terms of structure and form
-\cite{Mossman.1937:comparativemorphogenesisofthe}. Species differ in
-the shape of attachment (i.e., discoid, diffuse, cotyledonary,
-zonary), degree of invasiveness of fetal tissues into maternal tissues
-(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
-interdigitation of maternal-fetal tissue (i.e., villous, folded,
-trabecular, lamellar, labrynthine)
-\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
-mammals differ drastically in the number and size of their offspring,
-as well as the length of gestation, traits often associated with
-specific aspects of placental morphology
-\cite{Capellini.ea2011:Placentationandmaternalinvestment}.
-
-The placenta arose at some point before the origin of therian mammals,
-a group that includes both marsupials and eutherian or “placental”
-mammals. The term “placental mammals” is misleading, as marsupials
-also possess a placenta, although its direct apposition to maternal
-tissues lasts for a proportionally shorter time during gestation
-\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. Marsupials also differ from eutherian mammals in
-that most possess only a choriovitelline placenta, produced from the
-fusion of the chorion and yolk sac, during most of gestation. Although
-some eutherian mammals may possess a choriovitelline placenta at some
-point in development, the major site of nutrient exchange in
-placentals occurs via the chorioallantoic placenta, produced from the
-fusion of the chorion and allantois \cite{Mossman.1987:vertebratefetalmembranes}. It is unclear
-whether the first placenta was more like that of marsupials or
-eutherians; however, using phylogenetic methods, the ancestral
-eutherian was reconstructed as possessing a relatively invasive
-hemochorial placenta, in which the maternal and fetal blood supply is
-separated by one tissue layer, as in humans and mouse
-\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
-
-Although the morphology of the mammalian placenta has been well
-characterized in a number of species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
-the molecular environment is only beginning to be understood from a
-genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}. Studies of certain model species
-(i.e., humans, mice, cattle) have revealed a stunning divergence in
-the genetics of the placenta, especially in later stages of pregnancy
-\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
-For example, each lineage has evolved expansive
-placenta-specific gene families with expression patterns that can vary
-in space and time \cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}. These
-include independent expansions of the prolactins in mouse and cow,
-pregnancy-specific glycoproteins (PSGs) in humans and mouse,
-pregnancy-associated glycoproteins (PAGs) in cow, and the expansion of
-the growth hormone and chorionic gonadotropin beta genes in hominoid
-primates \cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}. Expression patterns of
-specific genes also can vary widely between species. For example, the
-expression of orthologous genes in the placenta have been found to
-vary substantially between humans and mice \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
-recruitment of SIGLEC6 to placental expression in humans differs from
-both chimpanzees and mice \cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
-
-One explanation for the divergent paths of the placental interface is
-the distinct evolutionary interplay between maternal and fetal
-strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}. The maternal strategy is to
-requisition an adequate degree of nutrients to the fetus without
-risking the health of the mother, while the fetal strategy via
-paternal genes is to extract as many resources from the mother as
-possible. Therefore, an interplay between these extremes can set up an
-intricate “arms race”, involving the imprinting of parental alleles,
-production of new genes, and enlistment of existing genes to
-expression at the maternal-fetal interface \cite{Haig1993:Geneticconflictsinhuman}.
-
-One symptom
-of this maternal-fetal conflict is its potential affect on human
-health \cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. 
-In addition to having an invasive
-hemochorial placenta, humans possess a deeply invasive trophoblast
-layer that can penetrate the myometrium of the uterus 
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
- Defects in placental growth are known to be
-associated with the devastating condition of preeclampsia, which can
-cause hypertension, proteinuria, and has the potential to lead to
-death of the mother and/or infant \DLA{(citation)}. Although cases have been
-documented in chimpanzees and gorillas \DLA{(citation)}, preeclampsia has a
-far higher incidence in human pregnancies with as much as 4\% of
-pregnancies being affected in the United States \DLA{(citation)}.
-Chimpanzees and gorillas also have deeply invasive trophoblasts,
-although the progression of invasion in great apes may be slower than
-humans \cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however, large-scale expression
-studies of great ape placentas have not been conducted. The increasing
-affordability of next-generation sequencing technology has enabled the
-large-scale study of specific tissues from a wide array of species to
-address broader phylogenetic questions in the evolution of gene
-expression \cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use Illumina RNA-Seq
-technology to sequence and quantify the mRNA environment of six
-mammalian placentas at term, including that of the human, bonobo (the
-sister species to the chimpanzee), dog, cow, African elephant, and
-opossum, representing marsupials as well as three major branches of
-eutherian mammals (Euarchontoglires, Laurasiatheria, Afrotheria).
-These placentas also represent a wide array of placental morphologies
-as well as traits such as number and size of offspring and length of
-gestation \DLA{(Table 1)}. We use this data, along with existing RNA-Seq
-data from mouse \cite{Wang.ea2011:surveyfornovelimprinted}, to address the major differences
-between expression in marsupial and eutherian mammals at term,
-identify the evolution of expression in important placental genes such
-as hormones and galectins, identify novel lineage-specific expression
-patterns, and quantify human specific differences in gene expression
-that may be related to obstetrical disorders such as preeclampsia.
-
-\section{Methods}
-
-\subsection{Collection of Placental Tissue and RNA Extraction}
-
-Fetal tissue was collected from the placenta of six mammalian species
-(\textit{Homo sapiens} [human], \textit{Pan paniscus} [bonobo], \textit{Canis familiaris}
-[domestic dog], \textit{Bos taurus} [domestic cow], \textit{Loxodonta africana} [African
-savanna elephant], \textit{Monodelphis domestica} [gray short-tailed
-opossum], \textit{Nannospalax gallili} [?], \textit{Spalax carmelli} [?]).
-Information on specimen ID, place and time of collection,
-and collector is shown for each sample in \DLA{Supplemental Table 1}. In the
-case of \textit{Pan}, \textit{Bos}, and \textit{Loxodonta}, the placenta was collected at term
-upon expulsion from the mother following delivery. Placental tissue
-from \textit{Homo sapiens} was collected upon delivery via a caesarian section.
-Near-term placental tissue from \textit{Canis familiaris} was collected upon
-the death of the mother from a car accident. Placental tissue from
-\textit{Monodelphis domestica} was sampled at fetal stage 32, roughly one day
-prior to birth. Only the fetal portion of the placenta was sampled at
-the maternal-fetal interface (i.e. villous portion for \textit{Homo}, \textit{Pan},
-\textit{Canis}, and \textit{Loxodonta}; cotyledon for \textit{Bos}; diffuse placenta for
-\textit{Monodelphis}). All tissues were collected and stored in RNAlater (Life
-Technologies, Carlsbad, CA) until RNA isolation could be performed.
-Tissues were homogenized in Trizol, and total RNA was isolated
-following the manufacturer’s instructions. RNA was further purified
-using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction with the
-RNase-Free DNase Set (Qiagen, Valencia, CA), following the
-manufacturer’s protocol to remove any contaminating DNA from the
-sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
-Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
-Clara, CA) to confirm concentration and quality of the isolation. RNA
-quality was assessed using A260/A280 and A260/A230 absorbance values.
-
-\DLA{Add details for new sequences and fix numbers}
-
-\subsection{Sequencing}
-
-\DLA{Fix how the sequencing was done}
-
-All RNA-Seq mRNA isolation, library construction, and Illumina
-sequencing was performed by the Applied Genomics Technology Center
-(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
-total RNA and standard Illumina protocol. Paired-end sequencing was
-performed using the Illumina Genome Analyzer II with an insert size of
-230 base pairs (bp). \textit{Bos}, \textit{Loxodonta}, and
-\textit{Monodelphis} were sequenced with read lengths of 75 bp;
-\textit{Canis}, \textit{Homo}, and \textit{Pan} were sequenced with
-read lengths of 150 bp.
-
-For comparison, additional sequences of placenta in \textit{Mus
-  musculus} \cite{Wang.ea2011:surveyfornovelimprinted} and
-\textit{Homo sapiens}
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} were obtained
-from the \ac{SRA}.
-
-\subsection{Transcriptome Assembly, Alignment, and Quantification}
-
-The quality of reads was evaluated using FASTQC v0.11.2
-(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
-Transcriptomes without a suitable reference genome (Reference of ---
-in \fref{tab:ref_genome_aligned_species}) were assembled into a
-reference using Trinity v2014-07-17
-\cite{Haas.ea2013:Denovotranscriptsequence} and annotated using
-Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against either
-the \textit{Homo sapiens} or \textit{Mus musculus} transcriptome
-(\fref{tab:ref_genome_aligned_species}).
-% 
-All transcriptomes were then aligned to the appropriate reference
-genome using STAR v2.4.2a
-\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
-values in \ac{FPKM}
-\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
-quantified using Cufflinks v2.2.1
-\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
-%
-The genome of the closely-related \textit{Pan troglodytes} was used as
-a reference for \textit{Pan paniscus}. The separation between the two
-recognized Pan species is generally perceived to be $\approx 1$ Mya
-\cite{Prüfer.ea2012:bonobogenomecomparedwith}.
-
-\begin{table}
-  \begin{tabular}{c c c c}
-    \toprule
-    Species                       & Common Name        & Reference & Read Length  \\
-    \midrule
-    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
-    \textit{Bos taurus}           & Cow                & UMD3.1     & 76 \\
-    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 151 \\
-    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 101 \\
-    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 76 \\
-    \textit{Homo sapiens}         & Human              & GRCh38     & 76, 151 \\
-    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 76 \\
-    \textit{Mus musculus}         & Mouse              & GRCm38     & 36,44,151 \\
-    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
-    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 151 \\
-    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
-    \bottomrule
-  \end{tabular}
-  \caption{Reference genomes for aligned species}
-  \label{tab:ref_genome_aligned_species}
-\end{table}
-
-\subsection{Exploration of Individual Transcriptome Data}
-
-To determine enrichment levels of functional categories of genes in
-highly-expressed transcripts in each transcriptome, we performed
-functional annotation cluster analyses per transcriptome using DAVID
-\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
-DAVID ranks functional annotation clusters according to enrichment
-score (ES); functional annotation clusters are considered
-significantly enriched if ES>1.3, as this corresponds to p<0.05
-\cite{Huang.ea2009:Systematicandintegrativeanalysis}. We executed
-analyses with two gene lists per transcriptome: 1) annotated genes
-with expression levels of the top 5\% of the frequency distribution of
-\ac{FPKM} and 2) the top 100 genes with the highest \ac{FPKM}. Due to the
-paucity of annotated gene ontology terms for some species, gene lists
-were compared using gene ontology terms derived from the human genome.
-
-\subsection{Comparative Analysis of Gene Expression}
-
-All 1:1 orthologues present among human, bonobo, mouse, dog, cow,
-elephant, and opossum were determined using Ensembl v.64 gene trees
-via the Biomart online data management
-tool\cite{Kasprzyk2011:BioMartdrivingparadigmchange}. A total of 6882
-1:1 orthologous genes were expressed in at least one species. For
-further analysis, we eliminated all genes that did not have at least
-one species with an $\mathrm{FPKM}\ge10$. A list of these 4259 1:1 orthologous
-genes was retained for future analyses of gene expression evolution
-described below.
-
-We used the software OUCH (Ornstein–Uhlenbeck for Comparative
-Hypotheses) in the R statistical environment to investigate
-significant shifts in gene expression levels over
-time\cite{Butler.2004:phlogeneticcomparativeanalysis}. OUCH institutes
-an Ornstein-Uhlenbeck (OU) model of evolution, which allows for
-different optima for quantitative trait values along user-specified
-branches of a phylogenetic tree. The OU model explicitly accounts for
-change in quantitative trait values by introducing two parameters into
-the model of stochastic Brownian motion of change in trait value along
-a phylogenetic tree through time. These parameters include, $\theta$,
-or an optimum trait value, and $\alpha$, or the strength of selection
-(interpreted as “pull” toward the optimum). In addition, the OU model
-can also integrate evolutionary change by assuming that each
-user-specified set of braches has its own separate OU parameters
-\cite{Butler.2004:phlogeneticcomparativeanalysis}.
-
-Various OU branch models were tested against a model in which no
-variation in optima among branches was allowed. The following branches
-were specified in separate models to have a different optimum than all
-other branches: human, bonobo, mouse, dog, cow, elephant, opossum,
-human+bonobo, Euarchontoglires, and Boreoeutheria. Likelihood values
-were computed for each branch model and compared to an OU model with
-one optimum using likelihood ratio tests with df=1. We filtered out
-genes with universally low expression levels, and included only genes
-in which expression in at least one species had an
-$\mathrm{FPKM}\ge10$ as described above. We corrected for multiple
-testing using the false discovery rate (FDR) and calculated adjusted
-$p$ values (q-values) using the Benjamini-Hochberg
-correction\cite{Benjamini1995:bh_procedure}. We used the
-time-calibrated phylogenetic tree of
-\cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
-not used in this analyses pruned.
-
-To determine enrichment of functional categories in lists of genes
-that display a significant shift in expression on a specific branch,
-we performed functional annotation cluster analyses on each branch
-using DAVID
-\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
-DAVID ranks functional annotation clusters according to enrichment
-score (ES); functional annotation clusters are considered
-significantly enriched if $\mathrm{ES}\ge1.3$, as this corresponds to
-$p\le0.05$ \cite{Huang.ea2009:Systematicandintegrativeanalysis}. As
-above, due to the paucity of annotated gene ontology terms for some
-species, gene lists were compared using gene ontology terms derived
-from the human genome.
-
-\subsection{Correlation of Tissue Specific Index with Lineage Specific Genes}
-
-To evaluate the extent of tissue specificity of each gene, we
-calculated the tissue specificity index based on human microarray data
-from (Su et al., 2004) available via BioGPS. The tissue specificity
-index (TSI) measures the extent to which a specific gene has
-predominant expression in only one tissue (Yanai et al., 2005). A TSI
-of 1.0 would signify that a gene was solely expressed in only one
-tissue, while a TSI of 0.0 would signify that a gene had equal
-expression in all tissues evaluated. We used the formula available in
-Yanai et al. (2005) to calculate the TSI, and identified the top four
-tissues which expressed each probe set from (Su et al., 2004).
-
-\subsection{Validation of Primate Expression using Spider Monkey Placenta Transcriptome}
-
-To verify expression levels in human and bonobo, we also generated
-RNA-Seq data for the placenta transcriptome of the black-headed spider
-monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
-Ateles is outside the clade, which includes hominids and can be used
-to polarize expression values of highly expressed genes in Homo and
-Pan. We sampled placental tissue from an Ateles placenta expelled from
-the mother shortly after birth (see Supplemental Table X). We used
-similar procedures discussed above to extract and sequence this
-sample. [DEREK—I’m unsure how to proceed on describing this so as to
-not step on Natalie’s toes]
-
-\DLA{Complete discussion of the ateles placenta}
-
-\section{Results}
-
-\subsection{Transcriptome Statistics}
 
@@
 
-For transcriptomes sequenced with 75 bp, %
-(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
-%
-,
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
-with at least 96\% of reads \DLA{This is wrong}
-mapping to the respective reference genome (\fref{tab:ref_genome_aligned_species}). 
-%
-For transcriptomes sequenced with 150 bp, %
-% The mismatch here is because we're reporting trimmed reads here, not
-% full length reads
-(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
-%
-,
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
-%
-with at least 99\% of reads mapping to the
-respective reference genome \DLA{This is also wrong}.
-%
-%
-For transcriptomes sequenced with 35 bp, %
-% The mismatch here is because we're reporting trimmed reads here, not
-% full length reads
-(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
-%
-,
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
-%
-with at least 99\% of reads mapping to the
-respective reference genome \DLA{This is also wrong}.
-%
-All reads were submitted to the NCBI Sequence Read Archive (SRA)
-(Accession Number XXX-YYY).
-
-\DLA{Need to do an embargoed submission to SRA}
-
-The percentage of bases that mapped to Ensembl annotated mRNA ranged
-from ~35\% in elephant to ~85\% in human, with percentages of bases
-mapping to annotated protein-coding segments of genes ranging from
-~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
-from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
-from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
-ranged from ~43\% in elephant to 80\% in bonobo. The number of
-annotated transcripts to which reads mapped in each respective genomes
-ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
-and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
-greater degree of mapping to annotated protein-coding sequences in
-humans corresponds to the greater number of Ensembl annotated
-sequences in the human genome compared to the other five reference
-genomes (Flicek, 2012).
-
-\DLA{Summarize all of this in a table}
-
-The top 50 expressed genes in each genome analyzed are shown in
-\DLA{Tables 2-7}, and all genes for each transcriptome are listed in
-\DLA{Supplementary Tables X-XX} [These are large Excel files to be
-included with the Supplemental Material]. In addition, specific genes
-including placental hormones and hormone-related genes, galectins, and
-insulin-like growth factor related genes are shown with \ac{FPKM} values
-side to side for all transcriptomes including the reanalysis of mouse
-from \cite{Wang.ea2011:surveyfornovelimprinted} in \DLA{Table 8}.
-Below we describe each transcriptome separately, specifically
-concentrating on highly-expressed genes.
-
-<<top.genes.function,echo=FALSE,results="hide">>=
-top.genes.table <- function(species.name,n=50) {
-    temp <- combined.fpkm[species==species.name &
-                      !grepl("^MT-?",gene_short_name) & 
-                          !grepl("(RNA\\d|MT)-",human_name)
-                          ][order(-mean_fpkm)][!duplicated(gene_short_name)][1:n,list(gene_short_name,human_name,mean_fpkm)]
+<<genes.expression,echo=FALSE,results="hide">>=
+genes.expression.table <- function(species.name,min.fpkm=50) {
+    temp <- 
+        combined.fpkm[species==species.name][order(-mean_fpkm)][,list(name_or_id,
+                          human_name,mean_fpkm)]
+    temp <- temp[!duplicated(name_or_id)][mean_fpkm>=min.fpkm]
     setnames(temp,c("Gene name","Human Name","FPKM"))
@@
 
-\subsection{Human Placenta Transcriptome}
-
-\begin{wraptable}{R}{0.3\textwidth}
-  \tiny
-<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
-top.human.genes <- top.genes.table(species.name="homo sapiens")
-print(xtable(top.human.genes),
+<<human.expressed.genes,echo=FALSE,results="asis">>=
+print(xtable(genes.expression.table(species.name="homo sapiens"),
+             caption="Expression of all genes with FPKM $\\ge 50$ in \\textit{Homo sapiens}",
+             label="tab:all_human_placenta"
+             ),
       booktabs=TRUE,
-      floating=FALSE
+      tabular.environment="longtable",
+      floating=FALSE,
+      caption.placement="top",
+      add.to.row = list(pos = list(0),command = "\\midrule \\endhead "),
+      size="\\tiny"
       )
-setkey(top.human.genes,"Gene name")
 @
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
-  \textit{Homo sapiens}}
-\label{tab:top50_human_placenta}
-\end{wraptable}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the human are shown in \fref{tab:top50_human_placenta} with a list of all genes in
-Supplemental Table X. The gene with by far the highest \ac{FPKM} value
-(101962) in humans is chorionic somatomammotropin hormone 1 (CSH1),
-followed by chorionic somatomammotropin hormone 2 (CSH2), and
-chorionic somatomammotropin hormone-like 1 (CSHL1), all three of which
-are members of a distinct expansion of the growth hormone gene family
-within hominoids. Another member of the growth hormone gene family,
-GH2 (\ac{FPKM}=729.639), is also expressed to a high degree in the human
-placenta (Table 2). Ten pregnancy-specific beta-1-glycoprotein genes
-are highly-expressed (FPKM>100) in the term placenta of humans, with
-PSG4 showing the greatest degree of expression (\ac{FPKM}= 5201.31; Table
-2). These genes are members of a hominoid-specific expansion of the
-carcinoembryonic antigen (CEA) family and known to be expressed by the
-syncytiotrophoblast (Lei et al., 1992; Rand and Cross, 2008). Genes
-involved in the production of estrogen (CYP19A1) and progesterone
-(HSD3B1) show exceptionally high expression in the human placenta
-(\ac{FPKM} = 4817.15 and 829.846 respectively), as well as CGA (\ac{FPKM}=
-4793.08), which generates the alpha subnit of multiple human chorionic
-gonadotropin hormones (Table 2). The genes producing the beta subunits
-of the chorionic gonadotropins (CGB, CGB1, CGB2, CGB5, CGB7, CGB8) are
-either not expressed (CGB8) or expressed modestly (FPKM≤20.826; Table
-8). Other highly expressed genes in the top 50 include members of the
-hemoglobin family (e.g. HBA1, HBB, HBG2, HBA2), the maternally
-imprinted long noncoding RNA gene H19, genes coding for proteins
-required for growth and/or cell proliferation (ADAM12, KISS1, IGF2,
-GPC3, CSF3R), anticoagulation (TFPI2), and glucose transport (SLC2A1)
-among others (Table 2, Supplemental Table X)
-
-\subsection{Bonobo Placenta Transcriptome}
-
-\begin{wraptable}{L}{0.3\textwidth}
-  \tiny
-<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
-top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
-print(xtable(top.bonobo.genes),
-      booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.bonobo.genes,"Gene name")
-@ 
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
-  \textit{Pan paniscus}}
-\label{tab:top50_bonobo_placenta}
-\end{wraptable}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the bonobo are shown in Table 3 with a list of all genes in
-Supplemental Table X. Expression in the bonobo placenta differs
-somewhat from that of the human, as evidenced by the makeup of the top
-50 genes (Table 3). The highest expression (\ac{FPKM}=10149.1) in bonobo is
-proteoglycan 2, bone marrow (PRG2), which is involved in insulin
-growth factor signaling and angiogenesis during pregnancy (Weyer and
-Glerup, 2011). PRG2 is also expressed at relatively high levels in the
-human placenta (\ac{FPKM}=250.675; Supplemental Table X), although at
-comparatively much lower levels than the bonobo. A bonobo gene closely
-related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
-expressed at high levels (\ac{FPKM}=1232.56; Table 3). Two other Pan genes
-included within this family (ENSPTRG00000030805 which shows similarity
-to GH2 and ENSPTRG00000034560) show expression levels of \ac{FPKM}>89. Six
-pregnancy-specific beta-1-glycoprotein genes were expressed in the
-bonobo placenta with \ac{FPKM} ranging from 20.885-167.961 (Supplemental
-Table X). Placental expression of the genes involved in estrogen and
-progesterone production (CYP19A1, HSD3B1 respectively) show lower
-expression in the bonobo as compared to humans (CYP19A1 \ac{FPKM}= 48.6877;
-HSD3B1 \ac{FPKM}= 54.5301; Table 8). In addition, CGA is expressed in the
-bonobo placenta (\ac{FPKM}=202.629), but at much lower levels than humans
-as well (Table 8). Two Ensembl-annotated beta subunits of the
-chorionic gonadotropin genes are expressed in bonobo, one related to
-CGB and another related to CGB7; both are not highly expressed (Table
-8). Among the other top 50 expressed genes, there is high expression
-of ribosomal proteins (e.g., RPL18, RPL23), keratins (e.g., KRT7,
-KRT19), heat-shock proteins (e.g., HSPA5, HSPA1A), serpins (SERPINE1,
-SERPINE2), insulin growth factor 2 (IGF2) and the hormone
-adrenomedullin (ADM).
-
-\subsection{Spider Monkey Transcriptome}
-
-\begin{wraptable}{L}{0.3\textwidth}
-  \tiny
-<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
-top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
-print(xtable(top.spider.monkey.genes),
-      booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.spider.monkey.genes,"Gene name")
-@ 
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
-  \textit{Ateles fusciceps}}
-\label{tab:top50_spider_monkey_placenta}
-\end{wraptable}
-
-\subsection{Dog Placenta Transcriptome}
-
-\begin{wraptable}{R}{0.3\textwidth}
-  \tiny
-<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
-top.dog.genes <- top.genes.table(species.name="canis familiaris")
-print(xtable(top.dog.genes),
-      booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.dog.genes,"Gene name")
-@ 
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
-  \textit{Canis familiaris}}
-\label{tab:top50_dog_placenta}
-\end{wraptable}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the dog are shown in Table 4 with a list of all genes in Supplemental
-Table X. The gene with the highest expression (\ac{FPKM}=48399.9) in the
-dog placenta is the hormone relaxin (RLN1). Other highly expressed
-genes among the top 50 (Table 4) include genes involved in the
-cytoskeleton (TMSB10, TMSB4X), anticoagulation (TFPI2, PLA2G7), innate
-immunity (LCN2, GM2A), lipid metabolism (APOE), inhibition of matrix
-metalloproteinases (TIMP2, TIMP3), and growth (PGF, EGFR, HTRA1). The
-gene LGALS1, which shows evidence of immunosupression during pregnancy
-(Than et al., 2008), shows high expression in dog (\ac{FPKM}= 1405.02). In
-addition, the gene SSLP1, a gene expressed in the placentas of cow and
-sheep but not present in primates (Ushizawa et al., 2009), is also
-expressed at high levels (\ac{FPKM}= 1853.98). There is no placental
-expression of the estrogen synthase gene (CYP19A1) in dogs, however
-expression of HSD3B1 is relatively high (\ac{FPKM}=157.622).
-
-\subsection{Cow Placenta Transcriptome}
-
-\begin{wraptable}{L}{0.3\textwidth}
-  \tiny
-<<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
-top.cow.genes <- top.genes.table(species.name="bos taurus")
-print(xtable(top.cow.genes),
-      booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.cow.genes,"Gene name")
-@ 
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
-  \textit{Bos taurus}}
-\label{tab:top50_cow_placenta}
-\end{wraptable}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the dog are shown in Table 5 with a list of all genes in Supplemental
-Table X. The gene with the highest expression (\ac{FPKM}=12488.8) was the
-miRNA gene bta-mir-2887-1. This gene occurs in several copies in the
-cow genome, and it was unclear as to how many or to which of 12
-bta-mir-2887 genes this mapped (Supplemental Table X). A previous
-study indicated that this miRNA gene was expressed in the bovine ovary
-(Hossain et al., 2009). Several members of the ruminant-specific
-pregnancy-associated glycoprotein family show especially high
-expression in the cow placenta (Table 4, Supplemental Table X)
-including PAG2 (\ac{FPKM}=7255.34), PAG10 (2076.42), BT.83780 (1793.72),
-PAG11 (962.19), PAG12 (469.751), BT.88769 (408.584), PAG1 (301.61),
-PAG4 (186.496), PAG6 (157.348), PAG8 (153.225), and PAG16 (103.573).
-In addition, 14 other PAGs were expressed with \ac{FPKM}<40 (Supplemental
-Table X). Other genes with high expression include genes involved with
-regulation of growth (IGF2, PHLDA2), imprinted genes (GNAS, DLK1,
-PHLDA2), and CSH1, a gene that is a member of a ruminant specific
-expansion of placental prolactins and not directly related to the gene
-of the same name in hominids. Other ruminant-specific placental
-prolactins also show high expression in the cow placenta including
-PRP1 (632.766), PRP5 (396.832), PRP6 (345.296), PRP2 (195.771), in
-addition to 7 other placental prolactins with expression of \ac{FPKM}<51
-(Table 5; Supplemental Table X). SSLP1 (also called SOLD1), a gene
-expressed in the cotyledonary villi of the ruminiant placenta and
-involved in cell proliferation (Ushizawa et al., 2009, 2010), also
-shows evidence of high expression (\ac{FPKM}=1001.45). There is some
-placental expression of the estrogen synthase gene (CYP19A1) in cows
-(75.1346), and comparatively greater expression of HSD3B1 (\ac{FPKM}=
-157.622).
-
-\subsection{Elephant Placenta Transcriptome}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the elephant are shown in Table 6 with a list of all genes in
-Supplemental Table X. The gene with the highest expression
-(\ac{FPKM}=33439.5) was the natriuretic peptide precursor B (NPPB). The
-gene with the second highest expression (\ac{FPKM}=20233) was the microRNA
-miR-675, which is imbedded in the first exon of the maternally
-imprinted long non-coding RNA gene H19 (Keniry et al., 2012), a gene
-also highly expressed in the elephant placenta (\ac{FPKM}=7164.3). Five
-uncharacterized genes are expressed to a high degree in the elephant
-placenta (Table 6), three of which are a member of an
-elephant-specific expansion of 11 genes closely related to the
-thymus-specific serine protease gene PRSS16. The other 8 PRSS16-like
-genes are also expressed to a lesser degree (FPKM<284.014). Two other
-novel genes with high expression are related to the gene SSLP1 (Table
-6), which is also expressed in the cow and dog placentas (Tables 4 and
-5). Placental expression of the hormones relaxin (RLN1) and prolactin
-(PRL) are exceptionally high in the elephant when compared to other
-mammals (Tables 6 and 8). There is little to no placental expression
-of the estrogen synthase gene, CYP19A1, (FPKM<1) and the enzyme that
-produces progesterone HSD3B1 (\ac{FPKM}=0). Other genes with high
-expression in the elephant placenta include genes involved in
-oxidative stress (TXN), growth (IGF2), and PAEP, a gene usually
-expressed in the endometrium of the uterus in humans (Rutanen et al.,
-1987).
-
-\subsection{Opossum Placenta Transcriptome}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the opossum are shown in Table 7 with a list of all genes in
-Supplemental Table X. The gene with the highest expression
-(\ac{FPKM}=41740.6) was hemoglobin, epsilon (HBE), which is expressed at
-the embryonic stage of development in marsupials, as in other mammals
-(Holland and Gooley, 1997). HBE subunits link up with protein products
-of two hemoglobin, zeta (HBZ) genes, which are also expressed to a
-high degree in the opossum placenta (\ac{FPKM}=2452.19 and \ac{FPKM}=1139.25).
-In addition, there is high expression (\ac{FPKM}=5827.37) of the gene
-hemoglobin, omega (HBO), a hemoglobin found only in marsupials and
-monotremes (Opazo et al., 2008). The hemoglobin gene HBA is also
-expressed in the placenta (\ac{FPKM}=456.978). The gene with the second
-highest expression (\ac{FPKM}=22854.8) is superoxide dismutase 3,
-extracellular (SOD3), which functions as an antioxidant. Four cystatin
-genes (CSTB, CST3, two CSTA-like genes) are expressed to a high
-degree, along with two cathepsin-L related genes (Table 7), which are
-known to interact with cystatins (Turk et al., 2001). The
-uncharacterized gene XM\_001364916.1 also shows high expression in the
-opossum placenta (\ac{FPKM}=13572.8). The gene WFDC2, a member of the whey
-acidic protein (WAP) family is also expressed to a high degree in the
-opossum placenta. Other genes with high expression include genes
-involved in retinol binding (RBP4), growth (IGFBP3), lipid metabolism
-(APOA1, APOA2), and immunosupression (LGALS1). In addition, there was
-little production of CYP19A1 and HSD3B1 (FPKM<2), and moderate
-expression of relaxin RLN1 (\ac{FPKM}= 41.3949).
-
-\subsection{Core Placenta Transcriptome}
-
-\begin{figure}
-<<core.placenta.transcriptome,fig.width=8,fig.height=12,echo=FALSE>>=
-combined.fpkm.wide <- 
-    data.table(dcast(combined.fpkm,
-                     human_name~species,
-                     fun.aggregate=sum,
-                     value.var="mean_fpkm"))
-temp <- 
-    combined.fpkm.wide[,median_expression := apply(combined.fpkm.wide[,-1,with=FALSE],1,median)]
-temp <- 
-    combined.fpkm.wide[,min_expression := apply(combined.fpkm.wide[,-1,with=FALSE],1,min)]
-setkey(combined.fpkm.wide,"human_name")
-core.placenta.transcriptome.long <- 
-    data.table(melt(combined.fpkm.wide[min_expression >= 10,
-                                       ][order(-median_expression)
-                                         ][!(human_name %in% 
-                                                 housekeeping.genes.superset[,gene_short_name])],
-                    id.vars="human_name",
-                    variable.name="Species",
-                    value.name="FPKM"
-                    ))
-temp <- core.placenta.transcriptome.long[,human_name:=
-                                     reorder(human_name,
-                                             combined.fpkm.wide[core.placenta.transcriptome.long[,human_name],median_expression])]
-rm(temp)
-
-print(ggplot(core.placenta.transcriptome.long[!Species %in% c("median_expression","min_expression"),],
-             aes(y=human_name,x=capfirst(Species)))
-      + geom_tile(aes(fill=log10(FPKM)))
-      + scale_x_discrete(expand=c(0,0))
-      + scale_y_discrete(expand=c(0,0))
-      + scale_fill_gradient(low="white",
-                            high="blue",
-                            limits=c(0,4),
-                            guide=guide_colorbar(title=expression(log[10](FPKM)))
-                            )
-      + ylab("Human Ortholog Name")
-      + xlab("Species")
-      + theme_grey()
-      + theme(legend.position="top",
-              axis.ticks = element_blank(), 
-              axis.text.x = element_text(angle = 340, hjust = 0.3, vjust=0.8,face="italic"))
-      )
-
-# 
-# print(xtable(combined.fpkm.wide[min_expression > 10,
-#                                 ][order(-median_expression)
-#                                   ][!(human_name %in% 
-#                                           housekeeping.genes.superset[,gene_short_name])][1:75,],
-#              digits=0
-#              ),
-#       include.rownames=FALSE,
-#       booktabs=TRUE,
-#       floating=FALSE
-#       )
-@ 
-\caption{Core Placenta Transcriptome.
-  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-  genes expressed (\ac{FPKM}$\ge 10$) in the placenta of all organisms
-  studied with housekeeping
-  genes\cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
-  removed.}
-\label{fig:core_placenta_transcriptome}
-\end{figure}
-
-Annexins?
-
-TIMP3 -- maybe in pre-eclampsia (23172037)
-SERPINH1? with TIMP3 
-% http://www.proteinatlas.org/ENSG00000149257-SERPINH1/tissue/placenta
-
-Why are RPL17 and RPS26 here? They are ribosomal proteins. These
-should be missing now.
-
-\subsection{DAVID Analyses of Individual Transcriptomes}
-
-In order to identify major functional categories of genes expressed in
-the individual transcriptomes, we undertook functional annotation
-clustering analyses using DAVID (see Methods). Concentrating on genes
-with relatively high expression values, we conducted DAVID analyses
-using a list of genes with \ac{FPKM} values included within the top 5\%
-distribution of expressed genes. Functional annotation clusters from
-all six placenta transcriptomes show enrichment in similar clusters of
-genes (Supplemental Table X). Significant enrichment in genes (ES>1.3,
-p>0.05) in the top 5\% of \ac{FPKM} values in at least five transcriptomes
-included clusters such as ribosome, vesicle, cytoskeleton,
-glycoprotein, lysosome, annexins, hormone stimulus or response,
-growth, calcium binding, cell migration, basement membrane,
-antioxidant activity or oxidative stress, cell adhesion, and
-mitochondrion among others (Supplemental Table X).
-
-We also investigated enrichment in functional annotation clusters in
-the top 100 genes of each transcriptome (Table 9). As in the analyses
-of the top 5\% of genes, the top 100 genes show significant enrichment
-(ES>1.3) in all six transcriptomes in genes associated with ribosomes.
-Human shows enrichment in clusters such as collagen-binding, hormone
-activity, female pregnancy, coagulation, and the defense response.
-Bonobo shows enrichment in clusters related to apoptosis,
-cytoskeleton-related genes, and the stress response. Cow shows only
-four significantly enriched clusters; however many genes highly
-expressed in the cow are ruminant-specific placenta genes and are not
-yet annotated to specific GO categories. Dog shows enrichment in genes
-associated with RNA, as well as homeostasis, glycoproteins, lyosome,
-and calcium binding. Elephant shows enrichment in genes associated
-with RNA, coagulation and calcium binding. Opossum shows enrichment in
-genes involved in apoptosis, oxidative stress, mitochondrion, hormone
-metabolism, and lipids.
-
-\subsection{Comparative Analysis of Gene Expression}
-
-We performed a comparative analysis of gene expression by
-reconstructing the evolution of expression values and detecting
-significant shifts in expression along specific lineages. Likelihood
-values for each model for each of the X genes reconstructed, as well
-as likelihood ratio tests, P- and Q-values (after correcting for
-multiple testing) are shown in Supplemental Table X. We identified
-between 4 (Boreoeutheria) and 550 (Mouse) significant shifts (Q <
-0.01) in expression on specific lineages (Table 10). The top 10 genes
-with significant shifts (if applicable) per lineage are shown in Table
-10 and all genes for each branch are shown in Supplemental Tables
-X-XX. Genes that showed shifts on the human lineage were
-overwhelmingly identified as placenta-specific by the
-tissue-specificity index (TSI), with 56 out of 279 genes (20\%) with
-both TSI>0.75 and placenta as one the top two tissues with the highest
-expression. Unsurpisingly, this was not the case with other lineages,
-since we used expression values from a human-derived microarray to
-generate TSI values for each gene. However, this does underscore the
-dynamic nature of expression patterns in the placenta and indeed all
-tissues across mammals.
-
-We then performed DAVID functional annotation clustering analyses on
-each list of genes from each lineage. The top three clusters for each
-lineage are shown in Table 10 and a full list of clusters in
-Supplemental Table X. The top enrichment cluster in the human lineage
-is blood vessel development, followed by extracellular matrix and
-metal ion binding (Table 10). Other significant clusters include those
-related to cell communication/signaling, cytokines, defense response,
-tryptophan metabolism, apoptosis, and skeletal system development
-(Supplemental Table X). The bonobo lineage was enriched in only three
-clusters: apoptosis, Toll-like receptor signalling and apoptosis. The
-mouse displayed enrichment in clusters involved in epigenetic
-modifications such as methyltransferase activity and histone
-modification, as well embryonic morphogeneisis, lipid biosynthesis,
-among others (Supplemental Table X). The top enriched cluster in the
-cow lineage was the response to vitamin A, as well as many clusters
-related to the immune system. The dog showed enrichment in skeletal
-system development, as well as lipid metabolic processes and blood
-circulation. The elephant lineage had only one enriched cluster:
-carbohydrate binding. Genes that showed enrichment on the opossum
-lineage (essentially representing the difference between placental
-mammals and marsupials) were clustered in many categories not seen in
-the other lineages, including clusters related to the mitochondria,
-urea cycle, peroxisome, glycolysis, and transport across the cell
-membrane (Table 10 and Supplemental Table X).
-
-\section{Discussion}
-
-\subsection{Evolution of Placental Hormones}
-
-Reproductive hormones are central to the regulation of pregnancy and
-their source, timing, and quantity can vary widely from species to
-species \cite{Carter2012:Evolutionofplacentalfunction}. Expression and synthesis of hormones is
-perhaps the most well-documented physiological process in the placenta
-in many species. Here we quantified the extreme disparity in genes
-with hormone products or enzymatic products that directly produce
-hormones in multiple mammalian species (\DLA{Table 8}).
-
-In both ruminants and primates, the placenta is responsible for the
-production of estrogen and progesterone. High relative expression of
-CYP19A1, the enzyme that synthesizes estrogen, and HSD3B1, the enzyme
-that synthesizes progesterone, in the human, bonobo, and cow,
-generally matches the high levels of placental production of estrogen
-and progesterone in humans, other primates, and ruminants in late
-pregnancy. In addition, HSD3B1 shows high levels in dog, which matches
-the high levels of placental production of progesterone at term in
-that species. \DLA{I think we need citations for this}
-
-\begin{figure}
-<<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{Expression of prolactin}
-\label{fig:prolactin_evolution_tree}
-\end{figure}
-
-The elephant is strikingly the only mammal analyzed that shows high
-expression of the primarily pituitary hormone prolactin (PRL) in the
-placenta. All afrotherians yet analyzed contain only one prolactin
-gene, and prolactin has undergone adaptive evolution on the elephant
-lineage \cite{Wallis2009:ProlactininAfrotheriacharacterization}.
-\cite{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
-presence of a prolactin in the trophoblast of the elephant placenta
-using immunohistological staining and human derived antibodies. A
-previous analysis of the elephant placenta also detected the
-expression of prolactin (Hou et al., 2012). Here we confirm that the
-sole prolactin in the elephant is not only expressed but highly
-expressed in the placenta, and likely functions both in the
-maintenance of the corpus luteum and as a placental lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}. PRL is not expressed in the opossum, even though there
-is evidence that it is expressed in the placenta of another marsupial,
-the tammar wallaby \cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests that
-placental prolactin expression in marsupials may be as diverse as in
-eutherian mammals. Although, PRL is not expressed to any great degree
-in the placenta other mammals, we see very high expression of genes
-which resulted from duplications of PRL in both the cow \DLA{(Table X)}, as
-well as mouse \DLA{(Supplemental Table X)}. \DLA{Describe these}
-
-The growth hormone (GH) family has expanded in anthropoid primates,
-resulting in approximately 5 genes in humans (GH1, GH2, CSH1, CSH2,
-CSHL1). All but GH1 show substantial expression in the placenta (Table
-8). There is some evidence that increased expression of these genes
-may be correlated with overall birth weight of offspring in humans
-\DLA{(Mannik et al., 2010)}, and it is possible that the extreme expression
-levels of these genes may be due to high birth weight. Ensembl
-annotations show 4 GH genes present in the chimpanzee genome, and
-three of these genes show high expression in bonobo, although to a
-lesser degree than in human (Table 8). Human placental GH genes are
-involved in increasing blood flow, replacing maternal pituitary
-expression of GH, as well as trophoblast invasion \cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth}, and it is likely that
-they have the same function in the bonobo.
-
-The human gene CGA codes for the alpha subunit of multiple hormones
-including pituitary hormones (luteinizing hormone,
-follicle-stimulating hormone, thyroid-stimulating hormone), as well as
-numerous chorionic gonadotropins with variable beta subunits \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of the chorionic gonadotropins is to
-prevent the regression of the corpus luteum \cite{Hanson.ea1971:EffectsofHCGand}.
-Among the six placentas we analyzed here, CGA is only expressed in the
-placentas of humans and bonobos (\DLA{Table 8}), which corresponds with the
-emergence of the chorionic gonadotropins in anthropoid primates
-\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly, CGA is expressed at much
-higher levels in both species than the beta subunits with which its
-product dimerizes (\DLA{Table 8}). CGA is one of several genes noted by
-\cite{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone adaptive evolution of the
-human-chimpanzee stem lineage.
-
-Relaxin is a hormone that has multiple effects during pregnancy
-including increasing maternal blood flow, widening the pubic bone and
-relaxing uterine musculature in preparation for labor. In both dogs
-and elephants, there is high expression of relaxin in the term
-placenta, with little evidence in other species examined here (Table
-8). Relaxin is known to be produced in the placenta of the pregnant
-dog and is used in the diagnosis of canine pregnancy
-\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof};
-however, the source of relaxin produced by pregnant African elephants
-remains unclear \cite{Meyer.ea2004:Speciesandfetalgender}. The high
-rise in relaxin in elephants has been suggested to be a major signal
-for the initiation of labor
-\cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive} Here we
-show that the placenta of elephants show high expression of relaxin
-and is likely the predominant source of relaxin at term \DLA{(Table 8)}.
-
-Two genes, NPPB and NPPC, code for peptide hormones that are important
-in the inducement of vasodilation, and may be important in increasing
-maternal blood flow to the fetus (Cameron et al., 1996). There is some
-indication that these proteins are present in the decidua of mice
-(Cameron et al., 1996), but we find very little expression in the
-fetal portion of the mouse placenta (Table 8). Here we find high
-expression of at least one of these genes in dog, cow, and elephant.
-The lack of expression in the hemochorial placentas of
-Euarchontoglires may mean that high expression of these genes are
-essential in increasing maternal blood flow in epitheliochorial and
-endotheliochorial placentas
-
-\subsection{The Evolution of Placental Expression of Galectins}
-
-\begin{figure}
-<<galectins.evolution.tree,echo=FALSE,fig.width=8,fig.height=8>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{Expression of galectins}
-\label{fig:galectins_evolution_tree}
-\end{figure}
-
-Galectins are recognized as major regulators of pregnancy in both
-fetal and maternal tissues (Than et al., 2012). Indeed, the evolution
-of new lineage-specific galectins has occurred in both primates and
-ruminants, with the genes LGALS13, LGALS14, and LGALS16 recognized as
-being placenta-specific in humans and regulating immune responses by
-inducing apoptosis of maternal T-cells (Than et al., 2009). Indeed,
-LGALS14 has the highest expression of any galectin in the human
-placenta (Table 8), and other primate galectins such as LGALS13 and
-LGALS16 are also expressed in human placenta. However, in the closely
-related bonobo, LGALS14 is not found, and LGALS1 is the most abundant
-galectin. LGALS1 shows generally high expression in all placentas
-analyzed. LGALS1 is also involved in mediating maternal-fetal immune
-tolerance, and is expressed in both maternal and fetal tissue in
-primates (Than et al., 2008). LGALS1 placental expression was thought
-to be confined to eutherian mammals, as there is evidence of adaptive
-evolution at this node, as well as the appearance of estrogen response
-elements in eutherian cis-regulatory regions (Than et al., 2008). It
-was unclear as to whether galectins such as LGALS1 were important in
-the regulation of pregnancy in marsupials, as well as eutherians. Here
-we show substantial expression of galectins in the opossum, especially
-LGALS1 (Table 8). In addition, LGALS2, LGALS3, and LGALS8 are also
-expressed. The presence of galectins at the maternal-fetal interface
-in the opossum implies that it was also present in the ancestral
-placenta of therian mammals. Indeed, adaptive evolution of LGALS1 on
-the eutherian stem is likely due to the increased invasive nature of
-eutehrian placentation rather than the origination of maternal-fetal
-tolerance. Although the function of LGALS1 in the marsupial placenta
-needs to be confirmed, it is likely that the mediation of
-maternal-fetal immune tolerance by galectins is an ancient feature of
-mammalian viviparity.
-
-\subsection{The Opossum Placenta and the Evolution of Viviparity}
-
-The placenta of the opossum Monodelphis domestica is distinctly
-different from that of the term placenta of eutherian mammals in that
-it is derived from both the chorion and yolk sac. Indeed other
-differences between the placentas of marsupials and eutherians include
-lack of invasive extravillous trophoblasts in marsupials, the
-retention of a permeable shell coat for 2/3 of marsupial gestation,
-and the greater degree of imprinting in placenta-expressed genes in
-eutherians (Renfree, 2010; Renfree et al., 2013). In addition, the
-placenta of Monodelphis is also distinct from other marsupials, in
-that it is considered one of the least specialized, and thought be
-closer to the original marsupial condition (Renfree, 2010). In
-Monodelphis, the duration of direct connection between maternal and
-fetal tissues is as short as two days in a gestational period of 13-15
-days. During this period, the placenta is directly engaged in the
-exchange of urea with maternal tissues, owing to the lack of a
-vesicular allantois (Freyer et al., 2007). This differs drastically
-with the situation in another well-studied marsupial, the tammar
-wallaby (Freyer et al., 2007). Therefore, we must interpret
-differences between the opossum and eutherian placentas with caution,
-as highly expressed genes and functional annotation clusters may
-represent didelphid or Monodelphis specific adaptations.
-
-Genes that show a significant shift on the opossum lineage include
-functional annotation clusters that correspond to active generation of
-energy, transport, and lipid breakdown via the peroxisome (Table 10;
-Supplemental Table X). The choriovitelline placenta of both didelphids
-and the tammar wallaby are known to be involved in glycolysis and
-protein synethsis (Freyer et al., 2003). This implies that the
-choriovitelline placenta in marsupials is an active site of metabolism
-and molecular synthesis, as well as the transport of nutrients from
-the maternal tissues. This is a distinct difference that is not seen
-in significant shifts on other lineages, and may be due to the
-formation of the placenta in the opossum from the yolk sac. In
-addition, the urea cycle is a highly enriched functional annotation
-cluster which directly correlates with the high levels of urea in the
-Monodelphis placenta (Freyer et al., 2007). This is potentially a
-Monodelphis-specific characteristic as discussed above.
-
-The opossum placenta shows expression of genes that either do not
-exist in eutherians or are not found in the term eutherian placenta
-(Table X). The existence of embryonic forms of hemoglobin in fetal
-tissues at term, as evidenced by the expression of HBE, HBZ, and HBO
-is a feature of marsupials that is generally due to their very short
-gestation periods. Embryonic hemoglobins in marsupials have a lower
-affinity for oxygen and allow for a lower rate of oxygen uptake from
-maternal tissues (Holland and Gooley, 1997). One novel uncharacterized
-gene that shows high expression (XM\_001364916.1) is not found in
-eutherians; Ensembl gene trees reveal that homologs of this gene exist
-in other amniotes such as birds and turtles, but are not present in
-eutherian mammals. The homolog of this gene in chicken is found to
-expressed in reproductive tract (Froman et al., 2006), but its
-function is unknown. WFDC2 is a member of the whey acidic protein
-(WAP) family and involved in innate immunity. It is expressed in the
-mammary gland of the tammar wallaby during pregnancy and early
-lactation (Sharp et al., 2007), and here in abundance in the opossum
-placenta.
-
-\subsection{Novel Gene Expansions in the Elephant Placenta}
-
-We found evidence of high expression in two novel expansions of genes
-in the elephant genome. One family includes 11 genes that are related
-to the gene PRSS16 gene, a thymus-specific serine protease that may be
-involved regulating the presentation of self-peptides in CD4+
-T-lymphocytes (Fornari et al., 2012). All 11 of these genes are
-expressed in the elephant placenta, three of which have very abundant
-mRNAs (Table 6). It is possible that this elephant-specific gene
-family may be involved in immunity in the elephant placenta and the
-mediation of maternal-fetal immune tolerance; however, future work
-will need to test these hypotheses. Three other genes abundantly
-expressed in the elephant placenta are homologues of SSLP1, a gene
-with predominant expression in the seminal vesicle in mouse (Li et
-al., 2006). Homologues of SSLP1 have been identified as being
-expressed in the placental villi of cow, sheep, and goat. SSLP1 also
-contains an Alu retrotransposon and may be involved in the development
-of villi via cell proliferation (Ushizawa et al., 2009, 2010). We
-verified expression in cow, however its abundant expression was also
-identified in dog. This gene is not found in primates. Expansion of
-these genes in elephants may indicate an elaboration of this system.
-
-\subsection{Human Specific Expression Shifts and the Evolution of Preeclampsia}
-
-The human placenta is distinctive from many other primates in that it
-possesses a deeply invasive trophoblast that penetrates the inner
-myometrium of the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
-hypothesized that this feature is related to the increased demand for
-resources by the developing fetus to feed the expanded size of the
-human brain (Carter and Pijnenborg, 2011; Robillard et al., 2002).
-Humans have the largest brain in terms of relative size and this
-differs markedly from closely related great apes (Boddy et al., 2012).
-Deep penetration of the myometrium occurs in chimpanzees and gorillas,
-but this penetration may be accelerated in humans (Carter and
-Pijnenborg, 2011).
-
-The near exclusivity of deep trophoblast invasion in humans has also
-led many to speculate on the origin of various disorders, which are
-almost exclusively symptomatic of human pregnancy. For example,
-preeclampsia is a human disease that occurs in 4\% of pregnancies in
-the United States. Preeclampsia can be attributed to the lack of
-adequate invasion of maternal tissue, and is noted by the onset of
-hypertension and proteinuria (citation). Some cases of preeclampsia
-can develop into eclampsia, a more severe condition resulting in
-seizures during pregnancy and death. Lack of invasion in preeclampsia
-may be attributed to the inefficiency of growth regulators as well as
-the inability to sufficiently suppress the maternal immune response.
-Although there are speculative cases of eclampsia in great apes
-(Baird, 1981; Stout and Lemmon, 1969; Thornton and Onwude, 1992), it
-is exceedingly rare in these species \cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It
-is possible that a comparative evolutionary approach to gene
-expression can identify genes that have divergent expression patterns
-from other mammals as well as association with preeclampsia.
-
-Functional annotation clusters significantly enriched on the human
-lineage (blood vessel development, extracellular matrix, tube
-morphogenesis) seem to suggest involvement in remodeling of tissues
-and increased blood flow, both of which may correlate with the
-invasion of the trophoblast (Table 8; Supplemental Table X). In
-addition, clusters such as cytokines and defense response
-(Supplemental Table X) also indicate an a significant change in
-expression of the immune response that may be involved in
-immunosupression at the maternal-fetal boundary to further allow
-invasion of fetal tissues.
-
-Many genes that both show expression shifts on the human lineage and
-have very high expression (FPKM>200) are directly related to the
-highly invasive nature of the human placenta. For example, GPC3,
-glypican 3, is an imprinted gene and cell growth regulator, associated
-with Simpson-Golabi-Behmel fetal overgrowth syndrome, and involved in
-intrauterine growth restriction (Monk and Moore, 2004; Pilia et al.,
-1996). GPC3 has the highest expression in humans (Table 2), but is
-also expressed in other mammals to a lesser degree, especially those
-with invasive hemochorial placentas like bonobos (\ac{FPKM}= 199.924),
-spider monkeys (\ac{FPKM}= 96.24), and the mouse (\ac{FPKM}= 80.7545)
-(Supplemental Tables X-XX). ADAM12 increases the availability of
-insulin-like growth factors that lead to the promotion of cell growth
-and placental invasion (Laigaard et al., 2005). (Hou et al., 2009)
-found evidence of adaptive evolution in ADAM12 on the primate lineage,
-and like GPC3 it is expressed in the bonobo (\ac{FPKM}= 46.1553), spider
-monkey (\ac{FPKM}=30.41) and mouse (\ac{FPKM}=2.25391) as well, although to a
-lesser degree (Supplemental Tables X-XX). RHOBTB3 is imprinted and
-associated with trophoblast invasion as well (Bilban et al., 2010;
-Yuen et al., 2011). In addition, one other gene, KISS1, shows
-exceptional abundance in the human placenta (Table 2); however due to
-its absence in elephant Ensmebl gene annotations, it was not part of
-the 1:1 orthologues examined in the comparative study. Here, it is
-only expressed in the term placentas of humans (Table 2) and bonobos
-(\ac{FPKM}=17.6394; Supplemental Table X). KISS1 is thought to be involved
-in the regulation of trophoblast invasion, and shows highest
-expression in the first trimester of pregnancy (Bilban et al., 2004;
-Reynolds et al., 2009).
-
-In addition, several other genes that show expression shifts on the
-human lineage also have major effects on placental function. Genes
-such as TXK and EBI3 are intimately involved in the immune system in
-the placenta. TXK is involved in natural killer cell differentiation
-(Bezman et al., 2012). EBI3 is implicated in natural killer cell
-function an maternal-fetal immunomodulation in the syncytiotrophoblast
-(Devergne et al., 2001). KYNU and KMO are interacting genes that are
-involved in tryptophan metabolism, which may be involved in
-immunosupression and the prevention of fetal rejection (Takikawa,
-2005). THSD7A is a thrombospondin, which inhibits endothelial cell
-migration and tube morphogenesis (Wang et al., 2010).
-
-Many genes that show evidence of a shift on the human lineage show
-evidence of implication in preeclampsia, including ADAM12 (Gack et
-al., 2005; Laigaard et al., 2005) MAOA (Carrasco et al., 2000;
-Bottalico et al., 2004), BCL2 (Ishihara et al., 2002), GPC3 (Gui et
-al., 2012; Ogawa, 2007), TMEM8A, CLIC3, EBI3, FLT1, BCL6, SYDE1, and
-MAN1C1 (Enquobahrie et al., 2008; Winn et al., 2009; Nishizawa, 2011).
-In addition KISS1 is also involved in preeclampsia as well (Sado et
-al., 2011; Qiao et al, 2012). Due to the association of many genes
-that show human-specific shifts as well as high expression, we
-conclude that preeclampsia is possibly a symptom of recent evolution
-of gene expression on the human lineage. Although many genes such as
-GPC3 and ADAM12 show distinctive relatively high expression in bonobo,
-this expression has intensified on the human lineage. Sequencing of
-more placentas of nonhuman primates, including other individuals of
-the genus Pan will help to further characterize this transition.
-
-\section{Conclusion}
-
-Here we sequenced six placentas spanning the phylogenetic tree of
-mammals. This revealed a dynamic expression pattern of expression of
-lineage-specific expansions of gene families and differential gene
-expression that agreed with the scenario of an “arms race” between
-maternal and fetal tissues that have shaped such disparate gene
-expression profiles. In addition, we highlighted the broad spectrum of
-hormone gene expression in multiple species that suggested that there
-was a lack of conservation in hormone signaling during mammalian
-pregnancy. Major differences between expression in marsupial and
-eutherian mammals at term seemed to be correlated with the differences
-between choriovitelline and chorioamniotic placenta, with the
-marsupial placenta more enriched in genes the contributed to the
-active generation of energy. We also identified that galectin
-expression, especially that of LGALS1, is a common feature of all
-mammalian placentas and was most likely present in the ancestral
-therian. Also, we identified two new expansions of genes that are
-highly expressed in the elephant placenta hat may be involved in
-trophoplast differentiation and immune system function, mirroring
-expansions in other groups such as primates, rodents, and ruminants.
-We also quantified significant lineage-specific shifts in gene
-expression, concentrating on the identification of genes that showed
-gene expression changes on the human lineage. We cautiously propose
-that many of these genes may be related to the invasive nature of the
-trophoblast in humans, although we note that in many of these genes,
-there is in some cases substantial expression in the bonobo as well.
-These genes have the potential to spur further research and may be at
-the root of obstetrical disorders such as preeclampsia. We propose
-that more placentas from diverse lineages, especially those of
-marsupials and primates, to further identify genes and expression
-levels that are specific to marsupials and eutehrians, as well as
-confirm the identity of human-specific expression shifts.
 

### 2015-10-13 manuscript/manuscript.Rnw
include supplemental material external references
add go analysis of core placenta transcriptome

@@
 \usepackage{zref-xr,zref-user}
+\zxrsetup{toltxlabel}
+\zexternaldocument*{supplemental_material}
 \newenvironment{narrow}[2]{%
@@
 })
-options(scipen = -2, digits = 2)
+options(scipen = -1, digits = 2)
 library("data.table")
@@
 library("ggtree")
+library("topGO")
+library("org.Hs.eg.db")
 
@@
 \begin{figure}
-<<core.placenta.transcriptome,fig.width=8,fig.height=12,echo=FALSE>>=
-combined.fpkm.wide <- 
-    data.table(dcast(combined.fpkm,
-                     human_name~species,
-                     fun.aggregate=sum,
-                     value.var="mean_fpkm"))
-temp <- 
-    combined.fpkm.wide[,median_expression := apply(combined.fpkm.wide[,-1,with=FALSE],1,median)]
-temp <- 
-    combined.fpkm.wide[,min_expression := apply(combined.fpkm.wide[,-1,with=FALSE],1,min)]
-setkey(combined.fpkm.wide,"human_name")
-core.placenta.transcriptome.long <- 
-    data.table(melt(combined.fpkm.wide[min_expression >= 10,
-                                       ][order(-median_expression)
-                                         ][!(human_name %in% 
-                                                 housekeeping.genes.superset[,gene_short_name])],
-                    id.vars="human_name",
-                    variable.name="Species",
-                    value.name="FPKM"
-                    ))
-temp <- core.placenta.transcriptome.long[,human_name:=
-                                     reorder(human_name,
-                                             combined.fpkm.wide[core.placenta.transcriptome.long[,human_name],median_expression])]
-rm(temp)
-
+<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
 print(ggplot(core.placenta.transcriptome.long[!Species %in% c("median_expression","min_expression"),],
@@
       )
-
-# 
-# print(xtable(combined.fpkm.wide[min_expression > 10,
-#                                 ][order(-median_expression)
-#                                   ][!(human_name %in% 
-#                                           housekeeping.genes.superset[,gene_short_name])][1:75,],
-#              digits=0
-#              ),
-#       include.rownames=FALSE,
-#       booktabs=TRUE,
-#       floating=FALSE
-#       )
 @ 
@@
   \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-  genes expressed (\ac{FPKM}$\ge 10$) in the placenta of all organisms
+  genes expressed (\ac{FPKM}$ \ge 10$) in the placenta of all organisms
   studied with housekeeping
-  genes\cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
+  genes \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
   removed.}
@@
 \end{figure}
+
+<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=
+go.results <- rbindlist(list(placenta_go_results_mf[1:10,],
+                             placenta_go_results_bp[1:10,]))
+go.results <- go.results[,Annotated:=NULL]
+setnames(go.results,"classic","p value")
+setnames(go.results,"classic.fdr","FDR")
+setnames(go.results,"GO.ID","GO ID")
+setnames(go.results,"Term","GO Term")
+print(xtable(go.results,
+             digits=2,
+             display=c("d","s","s","d","f","e","e","s"),
+             caption="Top go MF results",
+             label="tab:core_placenta_go_terms"),
+      include.rownames=FALSE,
+      floating=TRUE,
+      booktabs=TRUE,
+      size="\\tiny"
+      )
+@
+
+The
+\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
+genes comprising the core placenta transcriptome are shown in
+\fref{fig:core_placenta_transcriptome} ordered by their median
+transcription level in \ac{FPKM}. We considered a gene to be core to
+the transcriptome if an ortholog of that gene was expressed with
+$\mathrm{FPKM} \ge 10$ in all species studied and was not a
+housekeeping gene
+\cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
+\textit{ANXA1} is involved in the modulation of the inflammatory
+response \DLA{It does more stuff; read the huthcinson2011 review and
+  fill this in}
+%
+Another annexin, \textit{ANXA2} 
+%
+\textit{TIMP3}, which may be down-regulated in pre-eclampsia
+\cite{Ma.ea2014:Down-regulationofTIMP3leads} (though another study
+\cite{Xiang.ea2013:PromoterhypomethylationofTIMP3} claims
+up-regulation)
 

### 2015-10-13 manuscript/manuscript.Rnw
add a more complete expression function to make printing expression easier

@@
         combined.fpkm[species==species.name &
-                          !(human_name %in% housekeeping.genes.superset[,gene_short_name]) &
-                              !grepl("(RNA\\d|MT)-",human_name)
-                      ][order(-mean_fpkm)][!duplicated(gene_short_name)][1:n,list(gene_short_name,
+                           !(!is.na(human_name) & 
+                                 human_name %in% housekeeping.genes.superset[,gene_short_name]) &
+                               !grepl("((RNA\\d|MT)-|^RPS\\d+$)",human_name) &
+                                   !grepl("(^MT-|^RP[LS]\\d+[ABCD]?(?:-.+)?$|rRNA$)",gene_short_name)
+                      ][order(-mean_fpkm)][,list(name_or_id,
                                                  human_name,mean_fpkm)]
+    temp <- temp[!duplicated(name_or_id)][1:n]
     setnames(temp,c("Gene name","Human Name","FPKM"))
@@
 }
-gene.expression <- function(species.name,gene) {
-    combined.fpkm[species==species.name &
-                      gene_short_name==gene,mean_fpkm]
+gene.expression <- function(species.name,gene,complete=FALSE) {
+    fpkm <- combined.fpkm[species==species.name &
+                              gene_short_name==gene,
+                          mean_fpkm]
+
+    if (complete) {
+        return(paste0("\\textit{",gene,"}=",
+                      formatC(fpkm,digits=2,format="f"),
+                      " \\ac{FPKM}"
+                      ))
+    } else {
+        return(fpkm)
+    }
 }

### 2015-10-13 manuscript/manuscript.Rnw
discuss more completely the transcription profile in homo sapiens

@@
 
+<<cgb.min.expression,results="hide",echo=FALSE,warnings=FALSE>>=
+cgb.min.expression <- 
+    min(sapply(paste0("CGB",c("","1","2","5","7","8")),
+               function(x){gene.expression(species.name="homo sapiens",gene=x)}))
+cgb.sum.expression <- 
+    sum(sapply(paste0("CGB",c("","1","2","5","7","8")),
+               function(x){gene.expression(species.name="homo sapiens",gene=x)}))
+@
+
 The top 50 expressed genes as measured by \ac{FPKM} in the term
 placenta of \textit{Homo sapiens} are shown in
-\fref{tab:top50_human_placenta} with a list of all genes in
-Supplemental Table X. The gene with by far the highest \ac{FPKM} value
+\fref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
+in Supplemental \fref{tab:all_human_placenta}. The gene with by far
+the highest \ac{FPKM} value
+%
 (\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
-in humans is \ac{CSH1}, followed by \ac{CSH2}, and \ac{CSHL1}, all
+%
+in humans is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
 three of which are members of a distinct expansion of the growth
-hormone gene family within hominoids. Another member of the growth
+hormone gene family within hominoids (~\ref{fig:galectins_evolution_tree}). Another member of the growth
 hormone gene family, \ac{GH2} (\ac{FPKM}=%
@@
 % 
-Ten pregnancy-specific beta-1-glycoprotein genes are highly-expressed
-(\ac{FPKM}>100) in the term placenta of humans, with \ac{PSG4} showing the
+Eleven pregnancy-specific beta-1-glycoprotein genes are highly-expressed
+(\ac{FPKM}$\ge 100$) in the term placenta of humans, with \textit{PSG1} showing the
 greatest degree of expression %
-(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG4")}). % 
+(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
+and the pseudogene \textit{PSG10P} showing the least %
+(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \fref{fig:psg_evolution_tree}) . %
 These
-genes are members of a hominoid-specific expansion of the
-carcinoembryonic antigen (CEA) family and known to be expressed by the
-syncytiotrophoblast (Lei et al., 1992; Rand and Cross, 2008). Genes
-involved in the production of estrogen (CYP19A1) and progesterone
-(HSD3B1) show exceptionally high expression in the human placenta
-(\ac{FPKM} = 4817.15 and 829.846 respectively), as well as CGA
-(\ac{FPKM}= 4793.08), which generates the alpha subnit of multiple
-human chorionic gonadotropin hormones (Table 2). The genes producing
-the beta subunits of the chorionic gonadotropins (CGB, CGB1, CGB2,
-CGB5, CGB7, CGB8) are either not expressed (CGB8) or expressed
-modestly (FPKM≤20.826; Table 8). Other highly expressed genes in the
-top 50 include members of the hemoglobin family (e.g. HBA1, HBB, HBG2,
-HBA2), the maternally imprinted long noncoding RNA gene H19, genes
-coding for proteins required for growth and/or cell proliferation
-(ADAM12, KISS1, IGF2, GPC3, CSF3R), anticoagulation (TFPI2), and
-glucose transport (SLC2A1) among others (Table 2, Supplemental Table
-X)
+genes are members of a hominoid-specific expansion of the \ac{CEA} family
+and known to be expressed by the syncytiotrophoblast
+\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
+%
+Genes involved in the production of estrogens (%
+\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
+% 
+\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
+% 
+\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
+% 
+) and progesterone (%
+\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
+\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
+\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
+%
+show exceptionally high expression in the human placenta, as well as %
+%
+\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
+%
+which generates the alpha subnit of multiple human chorionic
+gonadotropin hormones (\fref{tab:top50_human_placenta}). The genes
+producing the beta subunits of the chorionic gonadotropins
+(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
+\textit{CGB7}, \textit{CGB8}) are either not expressed (%
+%
+\Sexpr{gene.expression(species.name="homo sapiens",gene="CGB8")}
+%
+) have reasonable expression ($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$),
+and the sum of their expression is on the order of the expression of
+expression of \textit{CGA}
+($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
+%
+%
+Other highly expressed genes in the top 50 include members of the
+hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
+\textit{HBA2}), the maternally imprinted long noncoding RNA gene
+\textit{H19}, genes coding for proteins required for growth and/or
+cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
+\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
+glucose transport (\textit{SLC2A1}) among others
+(\fref{tab:top50_human_placenta}, Supplemental
+\fref{tab:all_human_placenta}).
+
+\begin{figure}
+<<psg.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=5)
+@ 
+\caption{Expression of pregnancy-specific glycoproteins}
+\label{fig:psg_evolution_tree}
+\end{figure}
 

### 2015-10-13 manuscript/manuscript.Rnw
add estrogen, progesterone, and cgh gene trees

@@
 \begin{figure}
+<<estrogen.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CYP19A1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of estrogen sythesis genes}
+\label{fig:estrogen_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<progesterone.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="HSD3B1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of progesterone sythesis genes}
+\label{fig:progesterone_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<chorionic.gonadotropin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CGB1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of chorionic gonadotropin}
+\label{fig:cgb_evolution_tree}
+\end{figure}
+
+\begin{figure}
 <<prolactin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

### 2015-10-13 manuscript/manuscript.Rnw
add opossum and elephant transcriptome tables

@@
 
-\subsection{Elephant Placenta Transcriptome}
+\subsection{African Elephant Placenta Transcriptome}
+
+\begin{wraptable}{L}{0.3\textwidth}
+  \tiny
+<<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
+top.elephant.genes <- top.genes.table(species.name="loxodonta africana")
+print(xtable(top.elephant.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.elephant.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+  \textit{Loxodonta africana}}
+\label{tab:top50_elphant_placenta}
+\end{wraptable}
 
@@
 \subsection{Opossum Placenta Transcriptome}
+
+\begin{wraptable}{L}{0.3\textwidth}
+  \tiny
+<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
+top.opossum.genes <- top.genes.table(species.name="monodelphis domestica")
+print(xtable(top.opossum.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.opossum.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+  \textit{Monodelphis domestica}}
+\label{tab:top50_opossum_placenta}
+\end{wraptable}
 

### 2015-10-13 data/combine_fpkm.R
add name_or_id and remove useless columns

@@ -104,5 +104,12 @@ combined.fpkm[!is.na(oma_group_num),
                   grep("^-$",unique(gene_short_name),value=TRUE,invert=TRUE)),
               by=as.numeric(oma_group_num)]
 
-save(file=output.file,combined.fpkm,star.logs)
+combined.fpkm[,c("class_code","nearest_ref_id",
+                 "i.gene_short_name",
+                 "tss_id","locus","length","coverage"):=
+                     list(NULL,NULL,NULL,NULL,NULL,NULL)]
+combined.fpkm[,name_or_id:=gene_short_name]
+combined.fpkm[is.null(name_or_id),name_or_id:=gene_id]
+combined.fpkm[name_or_id=="-" | name_or_id=="",name_or_id:=gene_id]
 
+save(file=output.file,combined.fpkm,star.logs)

### 2015-10-13 common_r_code/plot_tree_matrix.R
allow specifying a minimum fpkm

@@ -1,9 +1,13 @@
-plot.tree.matrix <- function(gene.tree,gene.tree.table,expression) {
+plot.tree.matrix <- function(gene.tree,gene.tree.table,expression,min.fpkm=NA) {
     gene.tree$tip.label <- gsub("^(ENS.+)\\1$","\\1",gene.tree$tip.label)
     gene.tree.table[,species:=capfirst(gsub("_"," ",species))]
     gene.tree.table[,symbol_or_id:=ifelse(symbol=="NULL",gene_id,paste0(symbol," ",species))]
     gene.tree.expression <-
         combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+    if (!is.na(min.fpkm)) {
+        gene.tree.expression <-
+            gene.tree.expression[mean_fpkm >= min.fpkm ,]
+    }
     setkey(gene.tree.table,"protein_id")
     gene.tree$tip.label <-
         gene.tree.table[gene.tree$tip.label,gene_id]

### 2015-10-14 manuscript/manuscript.Rnw
add the tables of top genes for all of the organisms we have placenta data for

@@
     \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
-    \textit{Bos taurus}           & Cow                & UMD3.1     & 76 \\
-    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 151 \\
-    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 101 \\
-    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 76 \\
-    \textit{Homo sapiens}         & Human              & GRCh38     & 76, 151 \\
-    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 76 \\
-    \textit{Mus musculus}         & Mouse              & GRCm38     & 36,44,151 \\
+    \textit{Bos taurus}           & Cow                & UMD3.1     & 75 \\
+    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 150 \\
+    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 100 \\
+    \textit{Equus caballus}       & Horse              & EquCab2    & 100 \\ 
+    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 75 \\
+    \textit{Homo sapiens}         & Human              & GRCh38     & 75, 150 \\
+    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 75 \\
+    \textit{Mus musculus}         & Mouse              & GRCm38     & 35,44,150 \\
     \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
-    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 151 \\
+    \textit{Ovis aries}           & Sheep              & Oar_v3.1   & 151 \\
+    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 150 \\
     \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
+    \textit{Sus scrofa}           & Wild Boar          & Sscrofa10.2 & 51 \\
     \bottomrule
@@
                                  human_name %in% housekeeping.genes.superset[,gene_short_name]) &
-                               !grepl("((RNA\\d|MT)-|^RPS\\d+$)",human_name) &
-                                   !grepl("(^MT-|^RP[LS]\\d+[ABCD]?(?:-.+)?$|rRNA$)",gene_short_name)
+                               !grepl("((RNA\\d|MT)-|^RP[SL]\\d+$)",human_name) &
+                                   !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[l]?(?:-.+)?$|rRNA$)",gene_short_name)
                       ][order(-mean_fpkm)][,list(name_or_id,
@@
     temp <- temp[!duplicated(name_or_id)][1:n]
+    ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
+    temp <- temp[!(is.na(name_or_id)|name_or_id==""),]
     setnames(temp,c("Gene name","Human Name","FPKM"))
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
-  \textit{Homo sapiens}}
+\caption{Expression in \ac{FPKM} of top 50 genes which are not
+  ribosomal, mitochondrial, or a housekeeping gene in \textit{Homo
+    sapiens}}
 \label{tab:top50_human_placenta}
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
   \textit{Pan paniscus}}
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
   \textit{Ateles fusciceps}}
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
   \textit{Canis familiaris}}
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
   \textit{Bos taurus}}
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
   \textit{Loxodonta africana}}
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 non-mitochondrial genes in
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
   \textit{Monodelphis domestica}}
@@
 
+\subsection{Mouse Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
+top.mouse.genes <- top.genes.table(species.name="mus musculus")
+print(xtable(top.mouse.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.mouse.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
+  \textit{Mus musculus}}
+\label{tab:top50_mouse_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \textit{Mus musculus} are shown in
+\fref{tab:top50_mouse_placenta} with a list of all highly expressed
+genes in Supplemental \fref{tab:all_mouse_placenta}.
+
+% No idea what mir-6236 is doing
+
+% AC099934.1 probably overlaps with Rn7sl (part of SRP?)
+
+% AC157822.1 overlaps with Rn7s2
+
+% AC122546.1 = no idea
+
+% Gm3788 = psuedogene (Thymosin beta 10)
+
+\DLA{Why are there so many psuedogenes in the mouse? Maybe because of
+  the short read length?}
+
+\subsection{\textit{Nannospalax galili} (Blind Mole Rat) Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
+top.galili.genes <- top.genes.table(species.name="nannospalax galili")
+print(xtable(top.galili.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.galili.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
+  \textit{Nannospalax galili}}
+\label{tab:top50_n_galili_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \textit{Nannospalax galili} are shown in
+\fref{tab:top50_n_galili_placenta} with a list of all highly expressed
+genes in Supplemental \fref{tab:all_n_galili_placenta}.
+
+\subsection{\textit{Spalax carmeli} (Blind Mole Rat) Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
+top.carmeli.genes <- top.genes.table(species.name="spalax carmeli")
+print(xtable(top.carmeli.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.carmeli.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
+  \textit{Spalax carmeli}}
+\label{tab:top50_s_carmeli_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \textit{Spalax carmeli} are shown in
+\fref{tab:top50_s_carmeli_placenta} with a list of all highly expressed
+genes in Supplemental \fref{tab:all_s_carmeli_placenta}.
+
+\subsection{Wild Boar Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
+top.pig.genes <- top.genes.table(species.name="sus scrofa")
+print(xtable(top.pig.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.pig.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
+  \textit{Sus scrofa}}
+\label{tab:top50_pig_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \textit{Sus scrofa} are shown in
+\fref{tab:top50_pig_placenta} with a list of all highly expressed
+genes in Supplemental \fref{tab:all_pig_placenta}.
+
+\subsection{Horse Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
+top.horse.genes <- top.genes.table(species.name="equus caballus")
+print(xtable(top.horse.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.horse.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
+  \textit{Equus caballus}}
+\label{tab:top50_horse_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \textit{Equus Caballus} are shown in
+\fref{tab:top50_horse_placenta} with a list of all highly expressed
+genes in Supplemental \fref{tab:all_horse_placenta}.
+
+\subsection{Sheep Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
+top.sheep.genes <- top.genes.table(species.name="ovis aries")
+print(xtable(top.sheep.genes),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+setkey(top.sheep.genes,"Gene name")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
+  \textit{Ovis aries}}
+\label{tab:top50_sheep_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \textit{Ovis aries} are shown in
+\fref{tab:top50_sheep_placenta} with a list of all highly expressed
+genes in Supplemental \fref{tab:all_sheep_placenta}.
+
 \subsection{Core Placenta Transcriptome}
@@
 
-Annexins?
-
-TIMP3 -- maybe in pre-eclampsia (23172037)
 SERPINH1? with TIMP3 
 % http://www.proteinatlas.org/ENSG00000149257-SERPINH1/tissue/placenta
-
-Why are RPL17 and RPS26 here? They are ribosomal proteins. These
-should be missing now.
 

### 2015-10-16 manuscript/manuscript.Rnw
show the shape of the core placenta transcriptome

@@
 
+\begin{figure}
+<<core.placenta.transcriptome.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
+cpt.distribution.shape <- 
+    data.table(melt(
+        core.placenta.transcriptome.genes.shape[fifth_min_expression > 0,
+                                                list(human_name,
+                                                     min_expression,
+                                                     second_min_expression,
+                                                     third_min_expression,
+                                                     fourth_min_expression,
+                                                     fifth_min_expression)],
+               id.vars="human_name",
+               variable.name="Lowest Expression",
+               value.name="Expression (FPKM)"
+               ))
+cpt.distribution.shape <- 
+    cpt.distribution.shape[,`Lowest Expression`:=
+                               gsub("^min_expression$","First",
+                                    `Lowest Expression`)]
+cpt.distribution.shape <- 
+    cpt.distribution.shape[,`Lowest Expression`:=
+                               factor(gsub("^(.)(.+)_min_expression$","\\U\\1\\L\\2",
+                                           `Lowest Expression`,perl=TRUE),
+                                      levels=c("First","Second","Third","Fourth","Fifth"))
+                           ]
+print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0,],
+             aes(x=`Expression (FPKM)`+1,color=`Lowest Expression`))
+      + geom_histogram(aes(y=..count..+1))
+      + xlab(expression(FPKM+1))
+      + ylab("Genes with N-th Lowest Expression + 1")
+      + scale_y_log10()
+      + scale_x_log10())
+@ 
+\caption{Distribution shape of placenta core genes. A histogram and
+  density plot of the first through fifth lowest expression in each of
+  \Sexpr{star.logs[,length(unique(species))]} species for each of
+  \Sexpr{nrow(core.placenta.transcriptome.genes.shape)} human ortholog
+  genes.}
+\label{fig:core_placenta_transcriptome_shape}
+\end{figure}
+
 <<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=

### 2015-10-16 data/placenta_core_transcriptome.R

@@ -15,11 +15,39 @@ combined.fpkm.wide <-
                      fun.aggregate=sum,
                      value.var="mean_fpkm"))
 
+min.na.zero <- function(x){
+    x[is.na(x)] <- 0
+    return(min(x))
+}
+
+max.na.zero <- function(x){
+    return(max(x,na.rm=TRUE))
+}
+
+min.but.one <- function(x,n=1){
+    x[is.na(x)] <- 0
+    while (n > 0) {
+        x <- x[-which.min(x)]
+        n <- n - 1
+    }
+    return(min(x))
+}
+
 combined.fpkm.wide[,median_expression :=
                        apply(combined.fpkm.wide[,-1,with=FALSE],
                              1,median)]
+combined.fpkm.wide[,max_expression :=
+                       apply(combined.fpkm.wide[,-1,with=FALSE],1,max.na.zero)]
+combined.fpkm.wide[,fifth_min_expression :=
+                       apply(combined.fpkm.wide[,-1,with=FALSE],1,function(x){min.but.one(x,4)})]
+combined.fpkm.wide[,fourth_min_expression :=
+                       apply(combined.fpkm.wide[,-1,with=FALSE],1,function(x){min.but.one(x,3)})]
+combined.fpkm.wide[,third_min_expression :=
+                       apply(combined.fpkm.wide[,-1,with=FALSE],1,function(x){min.but.one(x,2)})]
+combined.fpkm.wide[,second_min_expression :=
+                       apply(combined.fpkm.wide[,-1,with=FALSE],1,min.but.one)]
 combined.fpkm.wide[,min_expression :=
-                       apply(combined.fpkm.wide[,-1,with=FALSE],1,min)]
+                       apply(combined.fpkm.wide[,-1,with=FALSE],1,min.na.zero)]
 setkey(combined.fpkm.wide,"human_name")
 
 ## all genes which are not housekeeping genes which have minimum 
@@ -37,10 +65,19 @@ core.placenta.transcriptome.long[,human_name:=
                                              combined.fpkm.wide[core.placenta.transcriptome.long[,human_name],median_expression])]
 
 
+core.placenta.transcriptome.genes.shape <-
+    combined.fpkm.wide[,list(human_name,
+                             median_expression,
+                             min_expression,
+                             second_min_expression,
+                             third_min_expression,
+                             fourth_min_expression,
+                             fifth_min_expression,
[…]

### 2015-10-19 go_annotation/go_tissue_analysis.R
run go enrichment in each tissue subset

@@ -1,0 +1,107 @@
+library("data.table")
+library(doMC)
+ncore = multicore:::detectCores()
+registerDoMC(cores = ncore)
+library("topGO")
+library("org.Hs.eg.db")
+
+args <- c("combined_fpkm","housekeeping_genes_superset","go_bos_taurus_mf")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+load(args[1])
+load(args[2])
+
+
+
+do.go.analysis <- function(species.name,go.ontology,top.100=TRUE,exclude.housekeeping=FALSE) {
+
+    all.genes <- combined.fpkm[species==species.name & !is.na(human_name),]
+    all.genes[,egid:=as.vector(unlist(sapply(mget(all.genes[,human_name],
+                   org.Hs.egSYMBOL2EG,ifnotfound=NA),function(x){x[1]})))]
+    all.genes <- all.genes[!is.na(egid),]
+
+    analysis.type <- ""
+    if (exclude.housekeeping) {
+        all.genes <- all.genes[!(human_name %in% housekeeping.genes.superset[,gene_short_name]),]
+        analysis.type <- " No Housekeeping"
+    }
+    
+    all.genes[,rank:=rank(-mean_fpkm)]
+
+    if (top.100) {
+        analysis.type <- paste0("Top 100",analysis.type)
+        gene.list <- all.genes[,factor(as.integer(rank <=100))]
+    } else {
+        analysis.type <- paste0("Top 5%",analysis.type)
+        gene.list <- all.genes[,factor(as.integer(rank <= (nrow(all.genes)*0.05)))]
+    }
+    names(gene.list) <- all.genes[,egid]
+    
+    godata <- new("topGOdata",
+                  ontology=go.ontology,
+                  allGenes=gene.list,
+                  annot=annFUN.org,
+                  mapping="org.Hs.eg.db")
+    test.stat <- new("classicCount", testStatistic = GOFisherTest, name = "Fisher test")
+    result.fisher <- getSigGroups(godata,test.stat)
+    res.table <- 
+        data.table(GenTable(godata,
+                            classic=result.fisher,
+                            orderBy="classic",
+                            topNodes=length(result.fisher@score)))
+    res.table[,classic:=as.numeric(gsub("< ","",classic))]
+    setnames(res.table,"classic","p value")
+    res.table[,FDR:=p.adjust(method="BH",`p value`)]
+    res.table[,Ontology:=go.ontology]
[…]

### 2015-10-19 data/placenta_core_transcriptome.R
only take one of the egid hits in the placenta core transcriptome

@@ -82,7 +82,8 @@ placenta.transcriptome.genes <-
 pts <- data.table(all.genes=all.genes)
 pts <- pts[,placenta.ts:=all.genes %in% placenta.transcriptome.genes]
 pts <- pts[,housekeeping:=all.genes %in% housekeeping.genes.superset[,gene_short_name]]
-pts <- pts[,egid:=as.vector(unlist(mget(all.genes,org.Hs.egSYMBOL2EG,ifnotfound=NA)))]
+pts <- pts[,egid:=as.vector(unlist(sapply(mget(all.genes,org.Hs.egSYMBOL2EG,ifnotfound=NA),
+                                          function(x){x[1]})))]
 ## this list contains all possible genes with a 0 if it is not in the
 ## core placenta transcriptome, or a 1 if it is in the core placenta
 ## transcriptome

### 2015-10-19 manuscript/manuscript.Rnw
remove first_min etc. from placenta core heatmap fix axes labels

@@
 <<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
-print(ggplot(core.placenta.transcriptome.long[!Species %in% c("median_expression","min_expression"),],
+print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median)_expression",Species),],
              aes(y=human_name,x=capfirst(Species)))
@@
               axis.ticks = element_blank(), 
-              axis.text.x = element_text(angle = 340, hjust = 0.3, vjust=0.8,face="italic"))
+              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
       )

### 2015-10-19 manuscript/manuscript.Rnw
change the distribution plot to be a line graph instead of a histogram

@@
 <<core.placenta.transcriptome.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
+calculate.distribution.shape <- 
+    function(min.vector) {
+        min.vector <- sort(min.vector,decreasing=TRUE)
+        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
+        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
+        return(temp)
+    }
+
+cptg.nh <- core.placenta.transcriptome.genes.shape[!(human_name %in% housekeeping.genes.superset[,gene_short_name])]
+
 cpt.distribution.shape <- 
-    data.table(melt(
-        core.placenta.transcriptome.genes.shape[fifth_min_expression > 0,
-                                                list(human_name,
-                                                     min_expression,
-                                                     second_min_expression,
-                                                     third_min_expression,
-                                                     fourth_min_expression,
-                                                     fifth_min_expression)],
-               id.vars="human_name",
-               variable.name="Lowest Expression",
-               value.name="Expression (FPKM)"
-               ))
+    lapply(c("min_expression",
+             "second_min_expression",
+             "third_min_expression",
+             "fourth_min_expression",
+             "fifth_min_expression"),
+           function(x){
+               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Lowest Expression`:=x]
+           })
+cpt.distribution.shape <- 
+    rbindlist(cpt.distribution.shape)
+
 cpt.distribution.shape <- 
@@
                            ]
-print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0,],
-             aes(x=`Expression (FPKM)`+1,color=`Lowest Expression`))
-      + geom_histogram(aes(y=..count..+1))
-      + xlab(expression(FPKM+1))
-      + ylab("Genes with N-th Lowest Expression + 1")
+setnames(cpt.distribution.shape,
+         "FPKM",
+         "Expression (FPKM)")
+
+print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 1,],
+             aes(x=`Expression (FPKM)`,color=`Lowest Expression`,y=num_genes))
+      + geom_line(lwd=1)
+      + xlab(expression(FPKM))
+      + ylab("Genes with N-th Lowest Expression")
       + scale_y_log10()
-      + scale_x_log10())
-@ 
-\caption{Distribution shape of placenta core genes. A histogram and
-  density plot of the first through fifth lowest expression in each of
-  \Sexpr{star.logs[,length(unique(species))]} species for each of
-  \Sexpr{nrow(core.placenta.transcriptome.genes.shape)} human ortholog
-  genes.}
+      + scale_x_log10()
+      + geom_vline(xintercept=10)
+      + geom_segment(mapping=aes(y=V1,
+                         yend=V1,
+                         xend=10,
+                         x=100,
+                         color=`Lowest Expression`),
+                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
+                                              max(num_genes),`Lowest Expression`],
+                     arrow=arrow(length=unit(0.5,"cm"))
+                   )
+      + geom_text(mapping=aes(x=100,
+                      y=V1,
+                      label=paste0(V1," genes"),override=TRUE),
+                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
+                      max(num_genes),`Lowest Expression`],
+                  show_guide=FALSE,
+                  hjust=0
+                  )
+      + guides(color=guide_legend(title="Lowest\nExpression"))
+      )
+@ 
+\caption{Distribution shape of placenta core genes. The Y axis shows
+  the number of orthologous gene sets included in the core placenta
+  transcriptome if the least expression cutoff (or second-least, etc.)
+  was set to the X axis. For example, if the cutoff is set to 10
+  (vertical black line), 37 genes would be included if the FPKM of all
+  species must be at least 10, 134 genes would be included if all but
+  one species must have FPKM of at least 10. 655 genes would be
+  included if all but five species must have FPKM of at least 10.}
 \label{fig:core_placenta_transcriptome_shape}

### 2015-10-19 manuscript/supplemental_material.Rnw
output all expression tables in supplemental materials output all go tables in supplemental materials

@@
 
+<<load.go.results,echo=FALSE,results="hide",cache.extra=tools::md5sum("../go_annotation/go_tissue_analysis")>>=
+load("../go_annotation/go_tissue_analysis")
+
+@
+
 \begin{figure}
@@
 
+<<go.analysis.all.species,echo=FALSE,results="asis">>=
+for (this.species in go.results[grepl("100 No Housekeeping",type) & FDR <= 0.0005,unique(species)]) {
+    print(xtable(go.results[grepl("100 No Housekeeping",type) & FDR <= 0.0005 & species == this.species,
+                            list(GO.ID,Term,FDR,Ontology,type)
+                            ],
+                 caption=paste0("Signficantly enriched ($\\mathrm{FDR}\\le0.0005$) GO terms in ",
+                     "\\textit{",capfirst(this.species),"}"),
+                 label=paste0("tab:go_terms_enriched_",gsub(" ","_",this.species)),
+                 display=c("s","s","s","e","s","s")
+                 ),
+          size="\\small",
+          booktabs=TRUE,
+          floating=TRUE)
+}
+@
+
 <<genes.expression,echo=FALSE,results="hide">>=
@@
 
-<<human.expressed.genes,echo=FALSE,results="asis">>=
-print(xtable(genes.expression.table(species.name="homo sapiens"),
-             caption="Expression of all genes with FPKM $\\ge 50$ in \\textit{Homo sapiens}",
-             label="tab:all_human_placenta"
+<<top.expressed.genes,echo=FALSE,results="asis">>=
+for (this.species in combined.fpkm[,sort(unique(species))]) {
+    print(xtable(genes.expression.table(species.name=this.species),
+                 caption=paste0("Expression of all genes with FPKM $\\ge 50$ in \\textit{",
+                     capfirst(this.species),"}"),
+                 label=paste0("tab:all_",gsub(" ","_",this.species),"_placenta")
                  ),
@@
           )
+}
 @

### 2015-10-19 manuscript/manuscript.Rnw
include all tissue go analysis

@@
 
-<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","palcenta_go_results_mf")))>>=
+<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","placenta_go_results_mf")))>>=
 for (a in paste0("../data/",c("combined_fpkm","gene_trees",
@@
          "unique.map.reads")
+@
+
+<<load.go.results,echo=FALSE,results="hide",cache.extra=tools::md5sum("../go_annotation/go_tissue_analysis")>>=
+load("../go_annotation/go_tissue_analysis")
+
 @
@@
 <<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
-print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median)_expression",Species),],
+print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
              aes(y=human_name,x=capfirst(Species)))
@@
 <<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=
-go.results <- rbindlist(list(placenta_go_results_mf[1:10,],
+placenta.go.results <- rbindlist(list(placenta_go_results_mf[1:10,],
                              placenta_go_results_bp[1:10,]))
-go.results <- go.results[,Annotated:=NULL]
-setnames(go.results,"classic","p value")
-setnames(go.results,"classic.fdr","FDR")
-setnames(go.results,"GO.ID","GO ID")
-setnames(go.results,"Term","GO Term")
-print(xtable(go.results,
+placenta.go.results <- placenta.go.results[,Annotated:=NULL]
+setnames(placenta.go.results,"classic","p value")
+setnames(placenta.go.results,"classic.fdr","FDR")
+setnames(placenta.go.results,"GO.ID","GO ID")
+setnames(placenta.go.results,"Term","GO Term")
+print(xtable(placenta.go.results,
              digits=2,
@@
 metabolism, and lipids.
+
+\begin{table}
+<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
+go.analysis.wide <- 
+    data.table(dcast(data=go.results,
+                     formula=GO.ID+type+Term+Ontology~species,
+                     value.var="FDR"))
+go.analysis.wide <- 
+    go.analysis.wide[,`max(FDR)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],
+                                    1,max,na.rm=TRUE)]
+go.analysis.wide.table <- 
+    go.analysis.wide[`max(FDR)` <= 0.1 & grepl("No Housekeeping",type),
+                     list(GO.ID,Term,Ontology,`max(FDR)`)]
+go.analysis.wide.table <- 
+    go.analysis.wide.table[order(`max(FDR)`)]
+print(xtable(go.analysis.wide.table,
+             display=c("d","s","s","s","e")
+             ),
+      booktabs=TRUE,
+      floating=FALSE
+      )
+@ 
+\caption{GO terms with enrichment in all species analyzed with
+  housekeeping genes removed}
+\label{tab:go_terms_all_species}
+\end{table}
 

### 2015-10-21 common_r_code/plot_tree_matrix.R
handle duplicate row names more elegantly

@@ -2,6 +2,7 @@ plot.tree.matrix <- function(gene.tree,gene.tree.table,expression,min.fpkm=NA) {
     gene.tree$tip.label <- gsub("^(ENS.+)\\1$","\\1",gene.tree$tip.label)
     gene.tree.table[,species:=capfirst(gsub("_"," ",species))]
     gene.tree.table[,symbol_or_id:=ifelse(symbol=="NULL",gene_id,paste0(symbol," ",species))]
+    gene.tree.table[duplicated(symbol_or_id),symbol_or_id:=paste0(symbol_or_id," dup")]
     gene.tree.expression <-
         combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
     if (!is.na(min.fpkm)) {

### 2015-10-21 manuscript/supplemental_material.Rnw
include more go results

@@
 <<go.analysis.all.species,echo=FALSE,results="asis">>=
-for (this.species in go.results[grepl("100 No Housekeeping",type) & FDR <= 0.0005,unique(species)]) {
-    print(xtable(go.results[grepl("100 No Housekeeping",type) & FDR <= 0.0005 & species == this.species,
+for (this.species in go.results[grepl("100 No Housekeeping",type) & FDR <= 0.05,unique(species)]) {
+    print(xtable(go.results[grepl("100 No Housekeeping",type) & FDR <= 0.05 & species == this.species,
                             list(GO.ID,Term,FDR,Ontology,type)
-                            ],
-                 caption=paste0("Signficantly enriched ($\\mathrm{FDR}\\le0.0005$) GO terms in ",
+                            ][order(FDR)],
+                 caption=paste0("Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                      "\\textit{",capfirst(this.species),"}"),
@@
           booktabs=TRUE,
-          floating=TRUE)
+          tabular.environment="longtable",
+          caption.placement="top",
+          add.to.row = list(pos = list(0),command = "\\midrule \\endhead "),
+          floating=FALSE)
 }

### 2015-10-21 manuscript/manuscript.Rnw
order species more usefully

@@
 
+data.matrix.dt <- function(x){
+    temp <- data.matrix(as.data.frame(x)[,-1])
+    rownames(temp) <- as.data.frame(x)[,1]
+    temp
+}
+
 capwords <- function(s, strict = FALSE) {
@@
 }
+
+species.names <- 
+    list(
+        "ateles fusciceps"=
+            list(sort.order=1,
+                 common.name="Spider monkey",
+                 scientific.name="Ateles fuciceps"),
+        "homo sapiens"=
+            list(sort.order=2,
+                 common.name="Human",
+                 scientific.name="Homo sapiens"),
+        "pan paniscus"=
+            list(sort.order=3,
+                 common.name="Bonobo",
+                 scientific.name="Pan paniscus"),
+        "mus musculus"=
+            list(sort.order=4,
+                 common.name="mouse",
+                 scientific.name="Mus musculus"),
+        "nannospalax galili"=
+            list(sort.order=5,
+                 name="nannospalax galili",
+                 scientific.name="Nannospalax galili"),
+        "spalax carmeli"=
+            list(sort.order=6,
+                 name="spalax carmeli",
+                 scientific.name="Spalax carmeli"),
+        "bos taurus"=
+            list(sort.order=7,
+                 name="Cow",
+                 scientific.name="Bos taurus"),
+        "ovis aries"=
+            list(sort.order=8,
+                 name="Sheep",
+                 scientific.name="Ovis aries"),
+        "sus scrofa"=
+            list(sort.order=9,
+                 name="Boar",
+                 scientific.name="Sus scrofa"),
+        "equus caballus"=
+            list(sort.order=10,
+                 name="Horse",
+                 scientific.name="Equus caballus"),
+        "canis familiaris"=
+            list(sort.order=11,
+                 name="Dog",
+                 scientific.name="Canis familiaris"),
+        "loxodonta africana"=
+            list(sort.order=12,
+                 name="Elephant",
+                 scientific.name="Loxodonta africana"),
+        "dasypus novemcinctus"=
+            list(sort.order=13,
+                 name="Nine banded armadillo",
+                 scientific.name="Dasypus novemcinctus"),
+        "monodelphis domestica"=
+            list(sort.order=14,
+                 name="Opposum",
+                 scientific.name="Monodelphis domestica"))
 
@@
 }
+combined.fpkm[,species:=factor(species,levels=names(species.names))]
 setkey(genes.to.tree,"gene_id")

### 2015-10-21 manuscript/manuscript.Rnw
don't output decimals for FPKM
switch to table from wraptable for now
add PRSS16 gene tree

@@
         return(paste0("\\textit{",gene,"}=",
-                      formatC(fpkm,digits=2,format="f"),
+                      formatC(fpkm,digits=0,format="f"),
                       " \\ac{FPKM}"
@@
 
-\begin{wraptable}{R}{0.3\textwidth}
+\begin{table}
   \tiny
@@
 \label{tab:top50_human_placenta}
-\end{wraptable}
+\end{table}
 
@@
 
-\begin{wraptable}{L}{0.3\textwidth}
+\begin{table}
   \tiny
@@
 \label{tab:top50_bonobo_placenta}
-\end{wraptable}
+\end{table}
 
@@
 \label{tab:top50_spider_monkey_placenta}
-\end{wraptable}
+\end{table}
 
@@
 \label{tab:top50_dog_placenta}
-\end{wraptable}
+\end{table}
 
@@
 \label{tab:top50_cow_placenta}
-\end{wraptable}
+\end{table}
 
@@
 
-\begin{wraptable}{L}{0.3\textwidth}
+\begin{table}
 \tiny
@@
 \label{tab:top50_elphant_placenta}
-\end{wraptable}
+\end{table}
 
@@
 \label{tab:top50_opossum_placenta}
-\end{wraptable}
+\end{table}
 
@@
 these genes in elephants may indicate an elaboration of this system.
+
+\begin{figure}
+<<prss16.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of \textit{PRSS16} and related genes}
+\label{fig:prss16_evolution_tree}
+\end{figure}
 

### 2015-10-21 manuscript/manuscript.Rnw
add lgals13 gene tree

@@
 
+\begin{figure}
+<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of galectins}
+\label{fig:galectins_lgals13_evolution_tree}
+\end{figure}
+
 Galectins are recognized as major regulators of pregnancy in both

### 2015-10-21 manuscript/manuscript.Rnw
add csh1 gene tree

@@
 \begin{figure}
+<<chorionic.somatomammotropin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CSH1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of chorionic somatomammotropin}
+\label{fig:csh1_evolution_tree}
+\end{figure}
+
+\begin{figure}
 <<prolactin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

### 2015-10-21 manuscript/manuscript.Rnw
add serpine tree

@@
 
+\begin{figure}
+<<serpine.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="SERPINE2" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1)
+@ 
+\caption{Expression of \textit{SERPINE2} genes with $\mathrm{FPKM} \ge 1$}
+\label{fig:serpine_evolution_tree}
+\end{figure}
+
 \subsection{Spider Monkey Transcriptome}

### 2015-10-21 manuscript/manuscript.Rnw
fix supplemental material reference to human table
clean up homo sapiens results section

@@
 \fref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
-in Supplemental \fref{tab:all_human_placenta}. The gene with by far
+in Supplemental \fref{tab:all_homo_sapiens_placenta}. The gene with by far
 the highest \ac{FPKM} value
@@
 three of which are members of a distinct expansion of the growth
-hormone gene family within hominoids (~\ref{fig:galectins_evolution_tree}). Another member of the growth
+hormone gene family within hominoids (\fref{fig:galectins_evolution_tree}). Another member of the growth
 hormone gene family, \ac{GH2} (\ac{FPKM}=%
@@
 (\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
-\textit{CGB7}, \textit{CGB8}) are either not expressed (%
-%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="CGB8")}
-%
-) have reasonable expression ($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$),
-and the sum of their expression is on the order of the expression of
-expression of \textit{CGA}
-($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
+\textit{CGB7}, \textit{CGB8}) have reasonable expression
+($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
+expression is on the order of the expression of expression of
+\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
 %
@@
 (\fref{tab:top50_human_placenta}, Supplemental
-\fref{tab:all_human_placenta}).
+\fref{tab:all_homo_sapiens_placenta}).
 

### 2015-10-21 manuscript/manuscript.Rnw
clean up bonobo results section some more

@@
 
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the bonobo are shown in \fref{tab:top50_bonobo_placenta} with a list of all genes in
-Supplemental Table X. Expression in the bonobo placenta differs
-somewhat from that of the human, as evidenced by the makeup of the top
-50 genes (Table 3). The highest expression (\ac{FPKM}=%
+The top 50 expressed genes as measured by \ac{FPKM} in the term
+placenta of the bonobo are shown in \fref{tab:top50_bonobo_placenta}
+with a list of all genes in Supplemental
+\fref{tab:all_pan_paniscus_placenta}. Expression in the bonobo
+placenta differs somewhat from that of the human, as evidenced by the
+makeup of the top 50 genes (\fref{tab:top50_bonobo_placenta}). The highest expression
+(\ac{FPKM}=%
 \Sexpr{gene.expression("pan paniscus","PRG2")}%
 %
-) in bonobo is \ac{PRG2}, which is involved in insulin
-growth factor signaling and angiogenesis during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}. \ac{PRG2} is also expressed at relatively high levels in the
-human placenta (\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")}; Supplemental Table X), although at
+) in bonobo is \textit{PRG2}, which is involved in insulin
+growth factor signaling and angiogenesis during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}. \textit{PRG2} is also expressed at relatively high levels in the
+human placenta (\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")}; Supplemental \fref{tab:all_homo_sapiens_placenta}), although at
 comparatively much lower levels than the bonobo. A bonobo gene closely

### 2015-10-21 manuscript/manuscript.Rnw
include prolactin receptor gene tree

@@
 
+\begin{figure}
+<<prolactin.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRLR" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of prolactin receptor}
+\label{fig:prolactin_receptor_evolution_tree}
+\end{figure}
+
 The elephant is strikingly the only mammal analyzed that shows high

### 2015-10-29 manuscript/manuscript.Rnw
discussion of prolactin needs work
describe the sequence and code deposition

@@
 
+\subsection{Sequence Deposition and Code}
+
+All of the code used to produce every analysis presented in this paper
+is available at
+\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
+sequences used are deposited in SRA, and can be obtained using
+\DLA{add GSE label for the experiment}.
+
 \section{Results}
@@
 expression of the primarily pituitary hormone prolactin (PRL) in the
-placenta. All afrotherians yet analyzed contain only one prolactin
+placenta. %
+\DLA{This is totally not true. \textit{A. fusciceps} expresses PRL,
+  and so does \textit{S. carmeli}, and \textit{B. taurus}, \textit{M.
+    musculus}, all have expansions of prolatacin.}
+%
+All afrotherians yet analyzed contain only one prolactin
 gene, and prolactin has undergone adaptive evolution on the elephant

### 2015-11-03 manuscript/manuscript.Rnw
add a species tree

@@
 
+\begin{figure}
+<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
+species.tree <- 
+    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodona_africana),((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus)))),(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus),((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)))));")
+plot(species.tree,x.lim=18)
+edgelabels("Eutheria",2,frame="none",bg="white",font=1,adj=c(0.75,0),cex=0.9)
+edgelabels("Euarchontoglires",16,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
+edgelabels("Laurasiatheria",7,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
+edgelabels("Atlantogenata",3,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
+edgelabels("Boreoeutheria",6,frame="none",bg="white",font=1,adj=c(0.7,0),cex=0.9)
+@ 
+\caption{Species tree for organisms studied}
+\label{fig:species_tree}
+\end{figure}
+
 \subsection{Exploration of Individual Transcriptome Data}

### 2015-11-03 manuscript/manuscript.Rnw
fix _ in ovis
switch to cleveref from fref

@@
 % \usepackage{txfonts}
-\usepackage[plain]{fancyref}
-\renewcommand*{\fref}[1]{\Fref{#1}}
+%\usepackage[plain]{fancyref}
 \usepackage{acronym}
@@
 \zexternaldocument*{supplemental_material}
+\usepackage{cleveref}
 \newenvironment{narrow}[2]{%
@@
 Transcriptomes without a suitable reference genome (Reference of ---
-in \fref{tab:ref_genome_aligned_species}) were assembled into a
+in \cref{tab:ref_genome_aligned_species}) were assembled into a
 reference using Trinity v2014-07-17
@@
 the \textit{Homo sapiens} or \textit{Mus musculus} transcriptome
-(\fref{tab:ref_genome_aligned_species}).
+(\cref{tab:ref_genome_aligned_species}).
 % 
@@
     \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
-    \textit{Ovis aries}           & Sheep              & Oar_v3.1   & 151 \\
+    \textit{Ovis aries}           & Sheep              & Oar\_v3.1   & 151 \\
     \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 150 \\
@@
 with at least 96\% of reads \DLA{This is wrong}
-mapping to the respective reference genome (\fref{tab:ref_genome_aligned_species}). 
+mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
 %
@@
 placenta of \textit{Homo sapiens} are shown in
-\fref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
-in Supplemental \fref{tab:all_homo_sapiens_placenta}. The gene with by far
+\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
+in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
 the highest \ac{FPKM} value
@@
 three of which are members of a distinct expansion of the growth
-hormone gene family within hominoids (\fref{fig:galectins_evolution_tree}). Another member of the growth
+hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
 hormone gene family, \ac{GH2} (\ac{FPKM}=%
@@
 is also expressed to a high degree in the human placenta
-(\fref{tab:top50_human_placenta}).
+(\cref{tab:top50_human_placenta}).
 % 
@@
 and the pseudogene \textit{PSG10P} showing the least %
-(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \fref{fig:psg_evolution_tree}) . %
+(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:psg_evolution_tree}) . %
 These
@@
 which generates the alpha subnit of multiple human chorionic
-gonadotropin hormones (\fref{tab:top50_human_placenta}). The genes
+gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
 producing the beta subunits of the chorionic gonadotropins
@@
 glucose transport (\textit{SLC2A1}) among others
-(\fref{tab:top50_human_placenta}, Supplemental
-\fref{tab:all_homo_sapiens_placenta}).
+(\cref{tab:top50_human_placenta}, Supplemental
+\cref{tab:all_homo_sapiens_placenta}).
 
@@
 The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the bonobo are shown in \fref{tab:top50_bonobo_placenta}
+placenta of the bonobo are shown in \cref{tab:top50_bonobo_placenta}
 with a list of all genes in Supplemental
-\fref{tab:all_pan_paniscus_placenta}. Expression in the bonobo
+\cref{tab:all_pan_paniscus_placenta}. Expression in the bonobo
 placenta differs somewhat from that of the human, as evidenced by the
-makeup of the top 50 genes (\fref{tab:top50_bonobo_placenta}). The highest expression
+makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
 (\ac{FPKM}=%
@@
 growth factor signaling and angiogenesis during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}. \textit{PRG2} is also expressed at relatively high levels in the
-human placenta (\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")}; Supplemental \fref{tab:all_homo_sapiens_placenta}), although at
+human placenta (\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")}; Supplemental \cref{tab:all_homo_sapiens_placenta}), although at
 comparatively much lower levels than the bonobo. A bonobo gene closely
 related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
-expressed at high levels (\ac{FPKM}=1232.56; \fref{tab:top50_bonobo_placenta}). Two other Pan genes
+expressed at high levels (\ac{FPKM}=1232.56; \cref{tab:top50_bonobo_placenta}). Two other Pan genes
 included within this family (ENSPTRG00000030805 which shows similarity
@@
 in the term placenta of \textit{Mus musculus} are shown in
-\fref{tab:top50_mouse_placenta} with a list of all highly expressed
-genes in Supplemental \fref{tab:all_mouse_placenta}.
+\cref{tab:top50_mouse_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_mouse_placenta}.
 
@@
 in the term placenta of \textit{Nannospalax galili} are shown in
-\fref{tab:top50_n_galili_placenta} with a list of all highly expressed
-genes in Supplemental \fref{tab:all_n_galili_placenta}.
+\cref{tab:top50_n_galili_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_n_galili_placenta}.
 
@@
 in the term placenta of \textit{Spalax carmeli} are shown in
-\fref{tab:top50_s_carmeli_placenta} with a list of all highly expressed
-genes in Supplemental \fref{tab:all_s_carmeli_placenta}.
+\cref{tab:top50_s_carmeli_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_s_carmeli_placenta}.
 
@@
 in the term placenta of \textit{Sus scrofa} are shown in
-\fref{tab:top50_pig_placenta} with a list of all highly expressed
-genes in Supplemental \fref{tab:all_pig_placenta}.
+\cref{tab:top50_pig_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_pig_placenta}.
 
@@
 in the term placenta of \textit{Equus Caballus} are shown in
-\fref{tab:top50_horse_placenta} with a list of all highly expressed
-genes in Supplemental \fref{tab:all_horse_placenta}.
+\cref{tab:top50_horse_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_horse_placenta}.
 
@@
 in the term placenta of \textit{Ovis aries} are shown in
-\fref{tab:top50_sheep_placenta} with a list of all highly expressed
-genes in Supplemental \fref{tab:all_sheep_placenta}.
+\cref{tab:top50_sheep_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_sheep_placenta}.
 
@@
 genes comprising the core placenta transcriptome are shown in
-\fref{fig:core_placenta_transcriptome} ordered by their median
+\cref{fig:core_placenta_transcriptome} ordered by their median
 transcription level in \ac{FPKM}. We considered a gene to be core to
@@
 \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
-\textit{ANXA1} is involved in the modulation of the inflammatory
-response \DLA{It does more stuff; read the huthcinson2011 review and
-  fill this in}
+\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
+which would be included in the core placenta set if the cutoff was set
+differently.
+%
+\textit{ANXA1} is
+involved in the modulation of the inflammatory response \DLA{It does
+  more stuff; read the huthcinson2011 review and fill this in}
 %

### 2015-11-03 manuscript/manuscript.Rnw
discuss go enrichement more

@@
 
-\subsection{DAVID Analyses of Individual Transcriptomes}
+\subsection{topGO Analyses of Individual Transcriptomes}
 
@@
 the individual transcriptomes, we undertook functional annotation
-clustering analyses using DAVID (see Methods). Concentrating on genes
-with relatively high expression values, we conducted DAVID analyses
-using a list of genes with \ac{FPKM} values included within the top 5\%
-distribution of expressed genes. Functional annotation clusters from
-all six placenta transcriptomes show enrichment in similar clusters of
-genes (Supplemental Table X). Significant enrichment in genes (ES>1.3,
-p>0.05) in the top 5\% of \ac{FPKM} values in at least five transcriptomes
-included clusters such as ribosome, vesicle, cytoskeleton,
-glycoprotein, lysosome, annexins, hormone stimulus or response,
-growth, calcium binding, cell migration, basement membrane,
-antioxidant activity or oxidative stress, cell adhesion, and
-mitochondrion among others (Supplemental Table X).
-
-We also investigated enrichment in functional annotation clusters in
-the top 100 genes of each transcriptome (Table 9). As in the analyses
-of the top 5\% of genes, the top 100 genes show significant enrichment
-(ES>1.3) in all six transcriptomes in genes associated with ribosomes.
+clustering analyses using topGO. We examined \ac{GSE} in 1:1 human
+orthologs which had \ac{FPKM} values in the top 5\% of expressed genes
+%
+Significant enrichment in genes ($\mathrm{FDR}\le0.05$) in the top 5\%
+of \ac{FPKM} values in at least 12 of the
+\Sexpr{length(species.names)} species included clusters such as
+membrane organization , locomotion, tissue development, extracellular
+matrix organization, regulation of peptidase activity, and various
+cell death related pathways among others
+(\cref{tab:go_terms_all_species}).
+
+(Supplemental %
+\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}.) %
+
 Human shows enrichment in clusters such as collagen-binding, hormone
@@
 \begin{table}
+  \footnotesize
 <<go.analysis.all.tissue,echo=FALSE,results="asis">>=
@@
 go.analysis.wide <- 
-    go.analysis.wide[,`max(FDR)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],
-                                    1,max,na.rm=TRUE)]
+    go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
+                                         function(x){sum(x <= 0.05)})]
+go.analysis.wide <- 
+    go.analysis.wide[,`max(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
+                                           function(x){if (!any(is.finite(x) & x<=0.05)) {
+                                                           return(NA)
+                                                       } else {
+                                                           max(x[is.finite(x) & x<=0.05],na.rm=TRUE)
+                                                       }})]
 go.analysis.wide.table <- 
-    go.analysis.wide[`max(FDR)` <= 0.1 & grepl("No Housekeeping",type),
-                     list(GO.ID,Term,Ontology,`max(FDR)`)]
+    go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
+                         grepl("Top 5% No Housekeeping",type),
+                     list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
 go.analysis.wide.table <- 
-    go.analysis.wide.table[order(`max(FDR)`)]
+    go.analysis.wide.table[order(Ontology,-`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
+setnames(go.analysis.wide.table,"GO.ID","GO ID")
+
 print(xtable(go.analysis.wide.table,
-             display=c("d","s","s","s","e")
+             display=c("d","s","s","s","d","e"),
+             align=  c("c","c","c","c","c","c")
              ),
+      math.style.exponents=TRUE,
+      include.rownames=FALSE,
       booktabs=TRUE,
@@
       )
-@ 
-\caption{GO terms with enrichment in all species analyzed with
-  housekeeping genes removed}
+
+@ 
+\caption{GO terms with significant enrichment ($\mathrm{FDR}\le0.05$)
+  in 12 or more of the species analyzed from genes in the top 5\% of
+  expression. $n(\mathrm{FDR}\le0.05)$ is the number of species with
+  significant enrichment ($\mathrm{FDR}\le 0.05$) in the GO term
+  listed. $max(\mathrm{FDR}\le0.05)$ is the maximum FDR value in all
+  species with an FDR of less than $0.05$.}
 \label{tab:go_terms_all_species}

### 2015-11-03 manuscript/manuscript.Rnw
include all go results FDR≤ 0.05

@@
 <<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=
-placenta.go.results <- rbindlist(list(placenta_go_results_mf[1:10,],
-                             placenta_go_results_bp[1:10,]))
+placenta.go.results <- rbindlist(list(placenta_go_results_mf[classic.fdr < 0.05,],
+                             placenta_go_results_bp[classic.fdr < 0.05,]))
 placenta.go.results <- placenta.go.results[,Annotated:=NULL]

### 2015-11-03 manuscript/manuscript.Rnw
use horizontal and vertical axis

@@
 @ 
-\caption{Distribution shape of placenta core genes. The Y axis shows
+\caption{Distribution shape of placenta core genes. The vertical axis shows
   the number of orthologous gene sets included in the core placenta
   transcriptome if the least expression cutoff (or second-least, etc.)
-  was set to the X axis. For example, if the cutoff is set to 10
+  was set to the horizontal axis. For example, if the cutoff is set to 10
   (vertical black line), 37 genes would be included if the FPKM of all

### 2015-11-03 go_annotation/go_tissue_analysis.R
use only expressed genes

@@ -34,7 +34,7 @@ do.go.analysis <- function(species.name,go.ontology,top.100=TRUE,exclude.houseke
         gene.list <- all.genes[,factor(as.integer(rank <=100))]
     } else {
         analysis.type <- paste0("Top 5%",analysis.type)
-        gene.list <- all.genes[,factor(as.integer(rank <= (nrow(all.genes)*0.05)))]
+        gene.list <- all.genes[,factor(as.integer(rank <= (sum(mean_fpkm>0)*0.05)))]
     }
     names(gene.list) <- all.genes[,egid]
     

### 2015-11-20 data/combine_fpkm.R
only remove duplicated fpkms if we're not doing per-sample

@@ -112,4 +112,9 @@ combined.fpkm[,name_or_id:=gene_short_name]
 combined.fpkm[is.null(name_or_id),name_or_id:=gene_id]
 combined.fpkm[name_or_id=="-" | name_or_id=="",name_or_id:=gene_id]
 
+if (!all(grepl("_per_sample",output.file))) {
+    combined.fpkm <-
+        combined.fpkm[!duplicated(tracking_id)]
+}
+
 save(file=output.file,combined.fpkm,star.logs)

### 2015-11-20 data/fastq_fpkm/combine_fpkms.R

@@ -104,7 +104,7 @@ gene.fpkms <- rbindlist(gene.fpkms,fill=TRUE)
 
 gene.fpkms[,mean_fpkm := mean(FPKM),tracking_id]
 gene.fpkms[,sd_fpkm := sd(FPKM),tracking_id]
-gene.fpkms <- gene.fpkms[!duplicated(tracking_id),]
+# gene.fpkms <- gene.fpkms[!duplicated(tracking_id),]
 
 isoform.fpkms <- list();
 for (file in isoform.files) {

### 2015-11-20 go_annotation/go_tissue_analysis.R
only include significant go results in the output

@@ -102,6 +102,8 @@ print(results)
 
 go.results <- rbindlist(results)
 
+go.results <- go.results[`p value` <= 0.05,]
+
 save(file=args[length(args)],
      go.results)
                

### 2015-11-20 manuscript/manuscript.Rnw
add notes on core placenta genes to discuss
abstract out utility functions

@@
 
-data.matrix.dt <- function(x){
-    temp <- data.matrix(as.data.frame(x)[,-1])
-    rownames(temp) <- as.data.frame(x)[,1]
-    temp
-}
-
-capwords <- function(s, strict = FALSE) {
-    cap <- function(s) paste(toupper(substring(s, 1, 1)),
-                             {s <- substring(s, 2); if(strict) tolower(s) else s},
-                             sep = "", collapse = " " )
-    sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
-}
-
-capfirst <- function(s, strict = FALSE) {
-    paste(toupper(substring(s, 1, 1)),
-          {s <- substring(s, 2); if(strict) tolower(s) else s},
-          sep = "")
-}
-
-species.names <- 
-    list(
-        "ateles fusciceps"=
-            list(sort.order=1,
-                 common.name="Spider monkey",
-                 scientific.name="Ateles fuciceps"),
-        "homo sapiens"=
-            list(sort.order=2,
-                 common.name="Human",
-                 scientific.name="Homo sapiens"),
-        "pan paniscus"=
-            list(sort.order=3,
-                 common.name="Bonobo",
-                 scientific.name="Pan paniscus"),
-        "mus musculus"=
-            list(sort.order=4,
-                 common.name="mouse",
-                 scientific.name="Mus musculus"),
-        "nannospalax galili"=
-            list(sort.order=5,
-                 name="nannospalax galili",
-                 scientific.name="Nannospalax galili"),
-        "spalax carmeli"=
-            list(sort.order=6,
-                 name="spalax carmeli",
-                 scientific.name="Spalax carmeli"),
-        "bos taurus"=
-            list(sort.order=7,
-                 name="Cow",
-                 scientific.name="Bos taurus"),
-        "ovis aries"=
-            list(sort.order=8,
-                 name="Sheep",
-                 scientific.name="Ovis aries"),
-        "sus scrofa"=
-            list(sort.order=9,
-                 name="Boar",
-                 scientific.name="Sus scrofa"),
-        "equus caballus"=
-            list(sort.order=10,
-                 name="Horse",
-                 scientific.name="Equus caballus"),
-        "canis familiaris"=
-            list(sort.order=11,
-                 name="Dog",
-                 scientific.name="Canis familiaris"),
-        "loxodonta africana"=
-            list(sort.order=12,
-                 name="Elephant",
-                 scientific.name="Loxodonta africana"),
-        "dasypus novemcinctus"=
-            list(sort.order=13,
-                 name="Nine banded armadillo",
-                 scientific.name="Dasypus novemcinctus"),
-        "monodelphis domestica"=
-            list(sort.order=14,
-                 name="Opposum",
-                 scientific.name="Monodelphis domestica"))
-
-species <- function(s,join=TRUE) {
-    temp <- paste0("\\textit{",capfirst(s),"}")
-    if (join) {
-        return(english.join(temp))
-    } else {
-        return(temp)
-    }
-}
-
-english.join <- function(s){
-    ret <- c()
-    ## everything but the last gets a , appended if there are more than two
-    if (length(s) > 2) {
-        s[-length(s)] <- 
-            paste0(s[-length(s)],",")
-    }
-    ## the second to last gets an and appended if there are more than one
-    if (length(s) > 1) {
-        s[length(s)-1] <- 
-            paste0(s[length(s)-1]," and")
-    }
-    return(paste(sep="",collapse=" ",s))
-}
-
-@
-
-<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R")))>>=
-for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R"))) {
+@
+
+<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
+for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R"))) {
     source(a)
@@
 
+% SPINT2 -- Serine Peptidase inhibitor
+
+% SLC3A2 -- this is probably actually a housekeeping gene; it's
+% involved in amino acid transport
+
+% SQSTM1 -- NFκB signaling (immune supression?)
+
+% KRTCAP2  N-glycosylation (probably housekeeping)
+
+% NRDG1 -- interacts with HSP90
+
+% SDC4 -- proteoglycan
+
 SERPINH1? with TIMP3 

### 2015-11-20 common_r_code/utility_functions.R

@@ -1,0 +1,106 @@
+data.matrix.dt <- function(x){
+    temp <- data.matrix(as.data.frame(x)[,-1])
+    rownames(temp) <- as.data.frame(x)[,1]
+    temp
+}
+
+capwords <- function(s, strict = FALSE) {
+    cap <- function(s) paste(toupper(substring(s, 1, 1)),
+                             {s <- substring(s, 2); if(strict) tolower(s) else s},
+                             sep = "", collapse = " " )
+    sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
+}
+
+capfirst <- function(s, strict = FALSE) {
+    paste(toupper(substring(s, 1, 1)),
+          {s <- substring(s, 2); if(strict) tolower(s) else s},
+          sep = "")
+}
+
+species.names <- 
+    list(
+        "ateles fusciceps"=
+            list(sort.order=1,
+                 common.name="Spider monkey",
+                 scientific.name="Ateles fuciceps"),
+        "homo sapiens"=
+            list(sort.order=2,
+                 common.name="Human",
+                 scientific.name="Homo sapiens"),
+        "pan paniscus"=
+            list(sort.order=3,
+                 common.name="Bonobo",
+                 scientific.name="Pan paniscus"),
+        "mus musculus"=
+            list(sort.order=4,
+                 common.name="mouse",
+                 scientific.name="Mus musculus"),
+        "nannospalax galili"=
+            list(sort.order=5,
+                 name="nannospalax galili",
+                 scientific.name="Nannospalax galili"),
+        "spalax carmeli"=
+            list(sort.order=6,
+                 name="spalax carmeli",
+                 scientific.name="Spalax carmeli"),
+        "bos taurus"=
+            list(sort.order=7,
+                 name="Cow",
+                 scientific.name="Bos taurus"),
+        "ovis aries"=
+            list(sort.order=8,
+                 name="Sheep",
+                 scientific.name="Ovis aries"),
+        "sus scrofa"=
+            list(sort.order=9,
+                 name="Boar",
+                 scientific.name="Sus scrofa"),
+        "equus caballus"=
+            list(sort.order=10,
+                 name="Horse",
+                 scientific.name="Equus caballus"),
+        "canis familiaris"=
+            list(sort.order=11,
+                 name="Dog",
+                 scientific.name="Canis familiaris"),
+        "loxodonta africana"=
[…]

### 2015-11-20 data/placenta_lineage_specific.R
calculate lineage specificity

@@ -1,0 +1,79 @@
+library("data.table")
+library("reshape2")
+library("org.Hs.eg.db")
+
+args <- c("combined_fpkm","housekeeping_genes_superset","placenta_core_transcriptome")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+load(args[1])
+load(args[2])
+
+combined.fpkm.wide <- 
+    data.table(dcast(combined.fpkm,
+                     human_name~species,
+                     fun.aggregate=sum,
+                     value.var="mean_fpkm"))
+
+### ignore genes which aren't expressed at least 10 FPKM in some
+### tissue
+evolution.of.expression <-
+    combined.fpkm.wide[apply(combined.fpkm.wide[,-1,with=FALSE],1,max) >= 100,]
+
+species <- c("ateles fusciceps", "homo sapiens", "pan paniscus", "mus musculus", 
+             "nannospalax galili", "spalax carmeli", "bos taurus", "ovis aries", 
+             "sus scrofa", "equus caballus", "canis familiaris", "loxodonta africana", 
+             "dasypus novemcinctus", "monodelphis domestica")
+
+### infraclass
+eutharians <- species[1:13]
+metatheria <- species[14]
+
+### magnorder
+boreoeutharians <- species[1:11]
+atlantogenata <- species[12:13]
+
+### superorder
+euarchontoglires <- species[1:6]
+laurasiatheria <- species[7:11]
+
+### order
+primates <- species[1:3]
+rodentia <- species[4:6]
+
+
+calculate.t.test <- function(x,g1,g2){
+    p.value <- NA;
+    try(p.value <- t.test(as.numeric(x[g1]),
+                          as.numeric(x[g2]))$p.value,
+        silent=TRUE);
+    return(p.value)
+}
+
+calculate.z.test <- function(x,g1,g2){
+    p.value <- NA;
+    try(p.value <- pnorm(q=as.numeric(x[g1]),
+                         mean=mean(as.numeric(x[g2]),na.rm=TRUE),
+                         sd=sd(as.numeric(x[g2]),na.rm=TRUE)),
+        silent=TRUE);
+    return(p.value)
+}
+
+evolution.of.expression[,bor.atl.p:=apply(evolution.of.expression,1,
+                             calculate.t.test,boreoeutharians,atlantogenata)]
+evolution.of.expression[,bor.atl.fdr:=p.adjust(bor.atl.p,method="BH")]
+
[…]

### 2015-11-20 data/ensembl_orthologs_long.R
get all of the ensembl orthologs

@@ -1,0 +1,40 @@
+library(biomaRt)
+library(data.table)
+
+args <- commandArgs(trailingOnly=TRUE)
+
+human <- useMart("ensembl",dataset="hsapiens_gene_ensembl")
+
+ensembl.orthologs.long <- list()
+
+for (species in c("dnovemcinctus",
+                  "ptroglodytes",
+                  "btaurus",
+                  "cfamiliaris",
+                  "lafricana",
+                  "mmusculus",
+                  "mdomestica")) {
+    ensembl.orthologs.long[[species]] <-
+        data.table(getBM(c("ensembl_gene_id",
+                           paste0(species,"_",
+                                  "homolog_ensembl_gene")),
+                         values =TRUE, mart = human, bmHeader=FALSE))
+    setnames(ensembl.orthologs.long[[species]],
+             paste0(species,"_",
+                    "homolog_ensembl_gene"),
+             "homolog_ensembl_gene")
+    ensembl.orthologs.long[[species]][,species:=species]
+    ensembl.orthologs.long[[species]] <-
+        ensembl.orthologs.long[[species]][homolog_ensembl_gene!="",]
+}
+
+hgnc.symbols <-
+    data.table(getBM(c("ensembl_gene_id",
+                       "hgnc_symbol"),
+                     mart=human))
+
+
+ensembl.orthologs.long <- rbindlist(ensembl.orthologs.long)
+
+save(ensembl.orthologs.long,
+     file=args[length(args)])

### 2015-11-20 manuscript/manuscript.Rnw
work on GSE analyses

@@
 
-% SDC4 -- proteoglycan
+% SDC4 -- proteoglycan (syndecan) cell adhesion
+
+% LAMC1 -- laminin
+
+% LUM -- proteoglycan
+
+% VDAC1 -- mitochondrial gene?
+
+% F11R -- tight junctions
 
@@
 
-\subsection{topGO Analyses of Individual Transcriptomes}
+\subsection{\ac{GSE} Analyses of Individual Transcriptomes}
 
 In order to identify major functional categories of genes expressed in
-the individual transcriptomes, we undertook functional annotation
-clustering analyses using topGO. We examined \ac{GSE} in 1:1 human
+the individual transcriptomes, we examined \ac{GSE} in 1:1 human
 orthologs which had \ac{FPKM} values in the top 5\% of expressed genes
+(Supplemental %
+\cref{tab:go_terms_enriched_top_five_dasypus_novemcinctus,tab:go_terms_enriched_top_five_equus_caballus,tab:go_terms_enriched_top_five_homo_sapiens,tab:go_terms_enriched_top_five_loxodonta_africana,tab:go_terms_enriched_top_five_monodelphis_domestica,tab:go_terms_enriched_top_five_mus_musculus,tab:go_terms_enriched_top_five_ovis_aries,tab:go_terms_enriched_top_five_pan_paniscus,tab:go_terms_enriched_top_five_sus_scrofa,tab:go_terms_enriched_top_five_ateles_fusciceps,tab:go_terms_enriched_top_five_nannospalax_galili,tab:go_terms_enriched_top_five_spalax_carmeli,tab:go_terms_enriched_top_five_bos_taurus,tab:go_terms_enriched_top_five_canis_familiaris}) using topGO \cite{topGO:2.20.0}.
+%
 %
@@
 cell death related pathways among others
-(\cref{tab:go_terms_all_species}).
-
-(Supplemental %
-\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}.) %
-
-Human shows enrichment in clusters such as collagen-binding, hormone
-activity, female pregnancy, coagulation, and the defense response.
-Bonobo shows enrichment in clusters related to apoptosis,
-cytoskeleton-related genes, and the stress response. Cow shows only
-four significantly enriched clusters; however many genes highly
-expressed in the cow are ruminant-specific placenta genes and are not
-yet annotated to specific GO categories. Dog shows enrichment in genes
-associated with RNA, as well as homeostasis, glycoproteins, lyosome,
-and calcium binding. Elephant shows enrichment in genes associated
-with RNA, coagulation and calcium binding. Opossum shows enrichment in
-genes involved in apoptosis, oxidative stress, mitochondrion, hormone
-metabolism, and lipids.
+(\cref{tab:go_terms_all_species}). If we instead examine the top 100
+genes by \ac{FPKM} in each of the \Sexpr{length(species.names)}
+species (Supplemental %
+\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris})
+notable enrichments include extracellular matrix, female pregnancy,
+integrin and collagen-related processes, and various cell death
+related processes.
+
+% \textit{Ateles fuciceps} shows enrichment in extracellular
+% matrix and structural molecular activity (Supplemental
+% \cref{tab:go_terms_enriched_ateles_fusciceps}). \textit{Homo sapiens}
+% placentas are enriched in female pregnancy, estrogen biosynthesis,
+% reproductive process, hormone aactivity, oxygen transporter activity,
+% and haptoglobin binding (Supplemental
+% \cref{tab:go_terms_enriched_homo_sapiens}). \textit{Pan paniscus}
+% placentas are enriched in clusters related to stress, apoptosis, and
+% response to abiotic stimuli. \textit{Mus musculus} placentas are
+% enriched in cell junction, vesicle transport, in utero embryonic
+% development, and many cytoskeletal processes. \textit{Nannospalax
+%   galili} placentas are enriched in extracellular matrix organization,
+% integrin binding, and cell death.
+% 
+% Cow shows only four
+% significantly enriched clusters; however many genes highly expressed
+% in the cow are ruminant-specific placenta genes and are not yet
+% annotated to specific GO categories. Dog shows enrichment in genes
+% associated with RNA, as well as homeostasis, glycoproteins, lyosome,
+% and calcium binding. Elephant shows enrichment in genes associated
+% with RNA, coagulation and calcium binding. Opossum shows enrichment in
+% genes involved in apoptosis, oxidative stress, mitochondrion, hormone
+% metabolism, and lipids.
 

### 2015-11-20 manuscript/supplemental_material.Rnw
include list of tables and figures in supplemental material
switch to cleveref
abstract out utility functions and table code

--- 2015-10-13 manuscript/supplemental_material.Rnw
@@
 % \usepackage{txfonts}
-\usepackage[plain]{fancyref}
-\renewcommand*{\fref}[1]{\Fref{#1}}
 \usepackage{acronym}
@@
 \usepackage{zref-xr,zref-user}
+\usepackage{cleveref}
 \newenvironment{narrow}[2]{%
@@
 
+\listoftables
+
+\listoffigures
+
 <<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
@@
 
-capwords <- function(s, strict = FALSE) {
-    cap <- function(s) paste(toupper(substring(s, 1, 1)),
-                             {s <- substring(s, 2); if(strict) tolower(s) else s},
-                             sep = "", collapse = " " )
-    sapply(strsplit(s, split = " "), cap, USE.NAMES = !is.null(names(s)))
-}
-
-capfirst <- function(s, strict = FALSE) {
-    paste(toupper(substring(s, 1, 1)),
-          {s <- substring(s, 2); if(strict) tolower(s) else s},
-          sep = "")
-}
-
-species <- function(s,join=TRUE) {
-    temp <- paste0("\\textit{",capfirst(s),"}")
-    if (join) {
-        return(english.join(temp))
-    } else {
-        return(temp)
-    }
-}
-
-english.join <- function(s){
-    ret <- c()
-    ## everything but the last gets a , appended if there are more than two
-    if (length(s) > 2) {
-        s[-length(s)] <- 
-            paste0(s[-length(s)],",")
-    }
-    ## the second to last gets an and appended if there are more than one
-    if (length(s) > 1) {
-        s[length(s)-1] <- 
-            paste0(s[length(s)-1]," and")
-    }
-    return(paste(sep="",collapse=" ",s))
-}
-
 @
 
-<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R")))>>=
-for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R"))) {
+<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
+for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R"))) {
     source(a)
@@
 @
+
+<<load.go.results,echo=FALSE,results="hide",cache.extra=tools::md5sum("../go_annotation/go_tissue_analysis")>>=
+load("../go_annotation/go_tissue_analysis")
+
+@
+
+\section{Transcriptome statistics}
 
@@
 
+\section{Gene Ontology Analyses}
+
+\subsection{Top 100 Transcripts Enrichment}
+
+<<go.analysis.all.species,echo=FALSE,results="asis">>=
+for (this.species in names(species.names)) {
+    cat(paste0("\\subsubsection{Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
+                           " the top 100 \\textit{",
+                           capfirst(this.species),"} expressed transcripts}\n\n"))
+    this.caption <- paste0("Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
+                           " the top 100 \\textit{",capfirst(this.species),"} expressed transcripts")
+    this.label <- paste0("tab:go_terms_enriched_",gsub(" ","_",this.species))
+    print(xtable(go.results[grepl("100 No Housekeeping",type) & FDR <= 0.05 & species == this.species,
+                            list(GO.ID,Term,FDR,Ontology)
+                            ][order(FDR)],
+                 caption=this.caption,
+                 label=this.label,
+                 display=c("s","s","s","e","s")
+                 ),
+          size="\\small",
+          booktabs=TRUE,
+          tabular.environment="longtable",
+          caption.placement="top",
+          floating=FALSE)
+}
+@
+
+\subsection{Top 5\% Transcripts Enrichment}
+
+<<go.analysis.top.5p.all.species,echo=FALSE,results="asis">>=
+for (this.species in names(species.names)) {
+    cat(paste0("\\subsubsection{Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
+                           " the top 5\\% of \\textit{",
+                           capfirst(this.species),"} expressed transcripts}\n\n"))
+    this.caption <- paste0("Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
+                           " the top 5\\% of \\textit{",
+                           capfirst(this.species),"} expressed transcripts.")
+    print(xtable(go.results[grepl("Top 5% No Housekeeping",type) & FDR <= 0.05 & species == this.species,
+                            list(GO.ID,Term,FDR,Ontology)
+                            ][order(FDR)],
+                 caption=this.caption,
+                 label=paste0("tab:go_terms_enriched_top_five_",gsub(" ","_",this.species)),
+                 display=c("s","s","s","e","s")
+                 ),
+          size="\\small",
+          booktabs=TRUE, 
+          tabular.environment="longtable",
+          caption.placement="top",
+          floating=FALSE)
+}
+@
+
 <<genes.expression,echo=FALSE,results="hide">>=
@@
 
-<<human.expressed.genes,echo=FALSE,results="asis">>=
-print(xtable(genes.expression.table(species.name="homo sapiens"),
-             caption="Expression of all genes with FPKM $\\ge 50$ in \\textit{Homo sapiens}",
-             label="tab:all_human_placenta"
+\section{Gene Expression}
+<<top.expressed.genes,echo=FALSE,results="asis">>=
+for (this.species in names(species.names)) {
+    this.caption <- paste0("Expression of all genes with FPKM $\\ge 50$ in \\textit{",
+                           capfirst(this.species),"}")
+    cat(paste0("\\subsection{",this.caption,"}\n\n"))
+    print(xtable(genes.expression.table(species.name=this.species),
+                 caption=this.caption,
+                 label=paste0("tab:all_",gsub(" ","_",this.species),"_placenta")
                  ),
@@
           caption.placement="top",
-      add.to.row = list(pos = list(0),command = "\\midrule \\endhead "),
           size="\\tiny"
           )
+}
 @

### 2015-11-21 manuscript/manuscript.Rnw
discuss start of OUCH analysis

@@
 
-We used the software OUCH (Ornstein–Uhlenbeck for Comparative
-Hypotheses) in the R statistical environment to investigate
-significant shifts in gene expression levels over
-time\cite{Butler.2004:phlogeneticcomparativeanalysis}. OUCH institutes
-an Ornstein-Uhlenbeck (OU) model of evolution, which allows for
-different optima for quantitative trait values along user-specified
-branches of a phylogenetic tree. The OU model explicitly accounts for
-change in quantitative trait values by introducing two parameters into
-the model of stochastic Brownian motion of change in trait value along
-a phylogenetic tree through time. These parameters include, $\theta$,
-or an optimum trait value, and $\alpha$, or the strength of selection
-(interpreted as “pull” toward the optimum). In addition, the OU model
-can also integrate evolutionary change by assuming that each
-user-specified set of braches has its own separate OU parameters
+We used the software \ac{OUCH} in the R statistical environment to
+investigate significant shifts in gene expression levels over
+time\cite{Butler.2004:phlogeneticcomparativeanalysis}. \ac{OUCH}
+institutes an \ac{OU} model of evolution, which allows for different
+optima for quantitative trait values along user-specified branches of
+a phylogenetic tree. The \ac{OU} model explicitly accounts for change
+in quantitative trait values by introducing two parameters into the
+model of stochastic Brownian motion of change in trait value along a
+phylogenetic tree through time. These parameters are $\theta$, an
+optimum trait value, and $\alpha$, the strength of selection
+(interpreted as “pull” toward the optimum). In addition, the \ac{OU}
+model can also integrate evolutionary change by assuming that each
+user-specified set of braches has its own separate \ac{OU} parameters
 \cite{Butler.2004:phlogeneticcomparativeanalysis}.

### 2015-11-21 manuscript/manuscript.Rnw
add discussion of tissue specificity index

@@
 To evaluate the extent of tissue specificity of each gene, we
-calculated the tissue specificity index based on human microarray data
-from (Su et al., 2004) available via BioGPS. The tissue specificity
-index (TSI) measures the extent to which a specific gene has
-predominant expression in only one tissue (Yanai et al., 2005). A TSI
-of 1.0 would signify that a gene was solely expressed in only one
-tissue, while a TSI of 0.0 would signify that a gene had equal
-expression in all tissues evaluated. We used the formula available in
-Yanai et al. (2005) to calculate the TSI, and identified the top four
-tissues which expressed each probe set from (Su et al., 2004).
+calculated the \ac{TSI} based on human RNASeq data in 3267 samples in
+126 tissues the Roadmap Epigenomics
+\cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping} and GTEx projects
+\cite{Gtexconsortium.2015:Thegenotypetissueexpression} as well
+additional placenta
+\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} and uterus
+\cite{Chan.ea2014:Assessmentofmyometrialtranscriptome} samples from
+SRA. The \ac{TSI} measures the extent to which a specific gene has
+predominant expression in only one tissue
+\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}. A gene
+expressed in only one tissue would have a \ac{TSI} of 1, whereas a
+gene expressed in all tissues would have a \ac{TSI} of 0.
 

### 2015-11-21 manuscript/manuscript.Rnw
get rid of duplication of go analysis in the ouch bit

@@
 
-To determine enrichment of functional categories in lists of genes
-that display a significant shift in expression on a specific branch,
-we performed functional annotation cluster analyses on each branch
-using DAVID
-\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
-DAVID ranks functional annotation clusters according to enrichment
-score (ES); functional annotation clusters are considered
-significantly enriched if $\mathrm{ES}\ge1.3$, as this corresponds to
-$p\le0.05$ \cite{Huang.ea2009:Systematicandintegrativeanalysis}. As
-above, due to the paucity of annotated gene ontology terms for some
-species, gene lists were compared using gene ontology terms derived
-from the human genome.
+We also calculated the enrichment of functional categories in the
+lists of genes which displayed a significant shift in expression on
+each branch using the same methodology as in the per-species gene
+lists \DLA{reference the tables once this analysis is complete}.
 

### 2015-11-21 manuscript/manuscript.Rnw
clean up some more of the OU analysis

@@
 
-Various OU branch models were tested against a model in which no
-variation in optima among branches was allowed. The following branches
-were specified in separate models to have a different optimum than all
-other branches: human, bonobo, mouse, dog, cow, elephant, opossum,
-human+bonobo, Euarchontoglires, and Boreoeutheria. Likelihood values
-were computed for each branch model and compared to an OU model with
-one optimum using likelihood ratio tests with df=1. We filtered out
-genes with universally low expression levels, and included only genes
-in which expression in at least one species had an
-$\mathrm{FPKM}\ge10$ as described above. We corrected for multiple
-testing using the false discovery rate (FDR) and calculated adjusted
-$p$ values (q-values) using the Benjamini-Hochberg
-correction\cite{Benjamini1995:bh_procedure}. We used the
-time-calibrated phylogenetic tree of
+\DLA{The description of the OU process is terrible; it needs to be
+  completely redone}
+
+Various \ac{OU} branch models were tested against a model in which no
+variation in optima ($\theta$) among branches was allowed. The
+following branches were specified in separate models to have a
+different optimum than all other branches: human, bonobo, mouse, dog,
+cow, elephant, opossum, human+bonobo, Euarchontoglires, and
+Boreoeutheria. Likelihood values were computed for each branch model
+and compared to an \ac{OU} model with one optimum using likelihood
+ratio tests with df=1. We filtered out genes with universally low
+expression levels, and included only genes in which expression in at
+least one species had an $\mathrm{FPKM}\ge10$ as described above. We
+corrected for multiple testing using the \ac{FDR} and using the
+Benjamini-Hochberg correction\cite{Benjamini1995:bh_procedure}. We
+used the time-calibrated phylogenetic tree of
 \cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species

### 2015-11-21 manuscript/manuscript.Rnw
discuss how go analysis was done properly

@@
 highly-expressed transcripts in each transcriptome, we performed
-functional annotation cluster analyses per transcriptome using DAVID
-\cite{Huang.ea2009:Systematicandintegrativeanalysis,Sherman.ea2007:DAVIDKnowledgebasegene-centereddatabase}.
-DAVID ranks functional annotation clusters according to enrichment
-score (ES); functional annotation clusters are considered
-significantly enriched if ES>1.3, as this corresponds to p<0.05
-\cite{Huang.ea2009:Systematicandintegrativeanalysis}. We executed
-analyses with two gene lists per transcriptome: 1) annotated genes
-with expression levels of the top 5\% of the frequency distribution of
-\ac{FPKM} and 2) the top 100 genes with the highest \ac{FPKM}. Due to the
-paucity of annotated gene ontology terms for some species, gene lists
-were compared using gene ontology terms derived from the human genome.
+functional annotation cluster analyses per transcriptome using topGO
+\cite{topGO:2.20.0}. Gene Ontology clusters were considered
+significantly enriched if the \ac{FDR} as estimated by the Benjamini
+and Hochberg procedure\cite{Benjamini1995:bh_procedure} on from the
+$p$ values obtained from the Fisher Exact Test were less than or equal
+to 0.05. Enrichment tests were considered with two gene lists per
+transcriptome: annotated genes with expression levels of the top
+5\% of the frequency distribution of \ac{FPKM} (Supplemental
+\cref{tab:go_terms_enriched_top_five_dasypus_novemcinctus,tab:go_terms_enriched_top_five_equus_caballus,tab:go_terms_enriched_top_five_homo_sapiens,tab:go_terms_enriched_top_five_loxodonta_africana,tab:go_terms_enriched_top_five_monodelphis_domestica,tab:go_terms_enriched_top_five_mus_musculus,tab:go_terms_enriched_top_five_ovis_aries,tab:go_terms_enriched_top_five_pan_paniscus,tab:go_terms_enriched_top_five_sus_scrofa,tab:go_terms_enriched_top_five_ateles_fusciceps,tab:go_terms_enriched_top_five_nannospalax_galili,tab:go_terms_enriched_top_five_spalax_carmeli,tab:go_terms_enriched_top_five_bos_taurus,tab:go_terms_enriched_top_five_canis_familiaris});
+the top 100 genes with the highest \ac{FPKM} (Supplemental
+\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
+Due to the paucity of annotated gene ontology terms for some species,
+gene lists were compared using the \textit{Homo sapiens} Gene Ontology
+using the 1:1 \textit{Homo sapiens} ortholog.
 

### 2015-11-21 manuscript/manuscript.Rnw
make ortholog discussion accurate

@@
 
-All 1:1 orthologues present among human, bonobo, mouse, dog, cow,
-elephant, and opossum were determined using Ensembl v.64 gene trees
-via the Biomart online data management
-tool\cite{Kasprzyk2011:BioMartdrivingparadigmchange}. A total of 6882
-1:1 orthologous genes were expressed in at least one species. For
-further analysis, we eliminated all genes that did not have at least
-one species with an $\mathrm{FPKM}\ge10$. A list of these 4259 1:1 orthologous
-genes was retained for future analyses of gene expression evolution
-described below.
+All 1:1 orthologues present among \textit{D. novemcinctus}, \textit{P.
+  troglodytes}, \textit{B. taurus}, \textit{C. familiaris}, \textit{L.
+  africana}, \textit{M. musculus}, \textit{H. sapiens}, and \textit{M.
+  domestica} were determined using Ensembl v.80 gene trees via
+Biomart\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the
+biomaRt R package \DLA{Cite the biomaRt package}. A total of
+\Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
+  1,length(unique(human_short_name))]} 1:1 orthologous genes were
+expressed in at least one species. Of these,
+\Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
+  10,length(unique(human_short_name))]} 1:1 orthologous genes with
+\ac{FPKM} $\ge 10$ were retained for future analyses of gene
+expression evolution described below.
 

### 2015-11-30 manuscript/manuscript.Rnw
fix up the sequencing and the human length error

@@
 
-\DLA{Fix how the sequencing was done}
-
 All RNA-Seq mRNA isolation, library construction, and Illumina
@@
 performed using the Illumina Genome Analyzer II with an insert size of
-230 base pairs (bp). \textit{Bos}, \textit{Loxodonta}, and
-\textit{Monodelphis} were sequenced with read lengths of 75 bp;
-\textit{Canis}, \textit{Homo}, and \textit{Pan} were sequenced with
-read lengths of 150 bp.
-
-For comparison, additional sequences of placenta in \textit{Mus
-  musculus} \cite{Wang.ea2011:surveyfornovelimprinted} and
+230 base pairs (bp). Read lengths for species are shown in
+\cref{tab:ref_genome_aligned_species}.
+
+\subsubsection{Additional Sequences}
+
+For more complete comparisons additional sequences of placenta in
+\textit{Mus musculus} \cite{Wang.ea2011:surveyfornovelimprinted},
 \textit{Homo sapiens}
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} were obtained
-from the \ac{SRA}.
+\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \textit{Equus
+  caballus} \DLA{citation}, \textit{Ovis aries} \DLA{citation}, and
+\textit{Sus scrofa} \DLA{citation} were obtained from the \ac{SRA}
+(Supplemental \cref{tab:srr_numbers}).
 
@@
 \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
-  1,length(unique(human_short_name))]} 1:1 orthologous genes were
+  1,length(unique(human_name))]} 1:1 orthologous genes were
 expressed in at least one species. Of these,
 \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
-  10,length(unique(human_short_name))]} 1:1 orthologous genes with
+  10,length(unique(human_name))]} 1:1 orthologous genes with
 \ac{FPKM} $\ge 10$ were retained for future analyses of gene

### 2015-11-30 manuscript/supplemental_material.Rnw
fix up the sequencing and the human length error

@@
 @
+
+\section{Sequence Read Archive}
+
+\DLA{Add in SRR and GSE numbers for everything here once I have them.}
+
+\begin{table}
+  \begin{tabular}{c c p{3in}}
+    \toprule
+    Species                       & Experiment Number & Sample Number(s) \\
+    \midrule
+    \textit{Ateles fusciceps}     &  & \\
+    \textit{Bos taurus}           &  & \\
+    \textit{Canis familiaris}     &  & \\
+    \textit{Dasypus novemcinctus} &  & \\
+    \textit{Equus caballus}       &  & SRR850299,  SRR850300 \\
+    \textit{Loxodonta africana}   &  & \\
+    \textit{Homo sapiens}         & GSE66622 & SRR1850945,
+                                               SRR1850946, 
+                                               SRR1850951, 
+                                               SRR1850952, 
+                                               SRR1850953, 
+                                               SRR1850954, 
+                                               SRR1850957, 
+                                               SRR1850958, 
+                                               SRR1850965, 
+                                               SRR1850966, 
+                                               SRR1850975, 
+                                               SRR1850976, 
+                                               SRR1850983, 
+                                               SRR1850984, 
+                                               SRR1850987, 
+                                               SRR1850988, 
+                                               SRR1850991, 
+                                               SRR1850992, 
+                                               SRR1850993, 
+                                               SRR1850994, 
+                                               SRR1850995, 
+                                               SRR1850996, 
+                                               SRR1851003, 
+                                               SRR1851004, 
+                                               SRR1851011, 
+                                               SRR1851012 \\
+    \textit{Monodelphis domestica} & & \\
+    \textit{Mus musculus}         &  & SRR243594,
+                                       SRR243595,
+                                       SRR243596,
+                                       SRR243597,
+                                       SRR243598,
+                                       SRR243599,
+                                       SRR243600,
+                                       SRR243601,
+                                       SRR243602,
+                                       SRR243586,
+                                       SRR243587,
+                                       SRR243588,
+                                       SRR243589,
+                                       SRR243590,
+                                       SRR243591,
+                                       SRR243592,
+                                       SRR243593\\
+    \textit{Nannospalax galili}   &  & \\
+    \textit{Ovis aries}           &  & ERR489144, ERR489145\\
+    \textit{Pan paniscus}         &  & \\
+    \textit{Spalax carmeli}       &  & \\
+    \textit{Sus scrofa}           &           & SRR543893 \\
+    \bottomrule
+  \end{tabular}
+  \caption{Species, experiment and species numbers used in the analyses presented in this paper}
+  \label{tab:srr_numbers}
+\end{table}
+
+\begin{table}
+  \begin{tabular}
+  \end{tabular}
+\end{table}
 

### 2015-12-11 data/hugo_nomenclature.R
include hugo nomenclature for full gene names

@@ -1,0 +1,11 @@
+library("data.table")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+hugo.nomenclature <- fread(args[1])
+setnames(hugo.nomenclature,"Ensembl ID(supplied by Ensembl)","gene_id")
+hugo.nomenclature <- hugo.nomenclature[gene_id!="",]
+setkey(hugo.nomenclature,"gene_id")
+
+save(hugo.nomenclature,
+     file=args[length(args)])

### 2015-12-11 data/combine_fpkm.R
fix issue with tracking_id disappearing
also remove NULL assignment error

@@ -95,7 +95,7 @@ combined.fpkm <- data.table(data.frame(gene.fpkms))
 setkey(oma.groups,"ensembl")
 setkey(combined.fpkm,"tracking_id")
 combined.fpkm <- oma.groups[,list(group_num,fingerprint,ensembl)][combined.fpkm]
-combined.fpkm[,ensembl:=NULL]
+setnames(combined.fpkm,"ensembl","tracking_id")
 setnames(combined.fpkm,"group_num","oma_group_num")
 
 combined.fpkm[,oma_group_name:=as.character(NA)]
@@ -107,7 +107,7 @@ combined.fpkm[!is.na(oma_group_num),
 combined.fpkm[,c("class_code","nearest_ref_id",
                  "i.gene_short_name",
                  "tss_id","locus","length","coverage"):=
-                     list(NULL,NULL,NULL,NULL,NULL,NULL)]
+                     list(NULL,NULL,NULL,NULL,NULL,NULL,NULL)]
 combined.fpkm[,name_or_id:=gene_short_name]
 combined.fpkm[is.null(name_or_id),name_or_id:=gene_id]
 combined.fpkm[name_or_id=="-" | name_or_id=="",name_or_id:=gene_id]

### 2015-12-11 manuscript/manuscript.Rnw
load hugo nomenclature files

@@
 
-<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","placenta_go_results_mf")))>>=
+<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","placenta_go_results_mf","hugo_nomenclature")))>>=
 for (a in paste0("../data/",c("combined_fpkm","gene_trees",
@@
                               "placenta_go_results_bp",
-                              "placenta_go_results_mf"
+                              "placenta_go_results_mf",
+                              "hugo_nomenclature"
                               ))) {

### 2015-12-11 manuscript/manuscript.Rnw
make links black

--- 2015-11-30 manuscript/manuscript.Rnw
@@
 \usepackage[noblocks,auth-sc]{authblk}
-\usepackage[hyperfigures,bookmarks,colorlinks]{hyperref}
+\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
 %\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
@@
          "unique.map.reads")
+setkey(hugo.nomenclature,"Approved Symbol")
 @
@@
                                                  human_name,mean_fpkm)]
-    temp <- temp[!duplicated(name_or_id)][1:n]
     ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
     temp <- temp[!(is.na(name_or_id)|name_or_id==""),]
-    setnames(temp,c("Gene name","Human Name","FPKM"))
+    temp <- temp[!duplicated(name_or_id)][1:n]
+    setkey(temp,"human_name")
+    temp <- 
+        hugo.nomenclature[temp][order(-mean_fpkm),
+                                list(name_or_id,`Approved Symbol`,`Approved Name`,mean_fpkm)]
+    temp[!is.na(`Approved Name`),
+         `Approved Name`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Name`),"}")]
+    temp[!is.na(`Approved Symbol`),
+         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Symbol`),"}")]
+    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM"))
     if (species.name=="homo sapiens") {
@@
     return(temp)
+}
+genes.table.sanitize <- function(str){
+    return(str)
+}
+xtable.sanitize.text <- function(str){
+    result <- str
+    result <- gsub("\\\\", "SANITIZE.BACKSLASH", result)
+    result <- gsub("$", "\\$", result, fixed = TRUE)
+    result <- gsub(">", "$>$", result, fixed = TRUE)
+    result <- gsub("<", "$<$", result, fixed = TRUE)
+    result <- gsub("|", "$|$", result, fixed = TRUE)
+    result <- gsub("{", "\\{", result, fixed = TRUE)
+    result <- gsub("}", "\\}", result, fixed = TRUE)
+    result <- gsub("%", "\\%", result, fixed = TRUE)
+    result <- gsub("&", "\\&", result, fixed = TRUE)
+    result <- gsub("_", "\\_", result, fixed = TRUE)
+    result <- gsub("#", "\\#", result, fixed = TRUE)
+    result <- gsub("^", "\\verb|^|", result, fixed = TRUE)
+    result <- gsub("~", "\\~{}", result, fixed = TRUE)
+    result <- gsub("SANITIZE.BACKSLASH", "$\\backslash$",
+                   result, fixed = TRUE)
+    return(result)
 }
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.human.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.human.genes,"Gene symbol")
 @ 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.bonobo.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.bonobo.genes,"Gene symbol")
 @ 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.spider.monkey.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.spider.monkey.genes,"Gene symbol")
 @ 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.dog.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.dog.genes,"Gene symbol")
 @ 
@@
 
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the dog are shown in Table 4 with a list of all genes in Supplemental
-Table X. The gene with the highest expression (\ac{FPKM}=48399.9) in the
-dog placenta is the hormone relaxin (RLN1). Other highly expressed
-genes among the top 50 (Table 4) include genes involved in the
-cytoskeleton (TMSB10, TMSB4X), anticoagulation (TFPI2, PLA2G7), innate
-immunity (LCN2, GM2A), lipid metabolism (APOE), inhibition of matrix
-metalloproteinases (TIMP2, TIMP3), and growth (PGF, EGFR, HTRA1). The
-gene LGALS1, which shows evidence of immunosupression during pregnancy
-(Than et al., 2008), shows high expression in dog (\ac{FPKM}= 1405.02). In
-addition, the gene SSLP1, a gene expressed in the placentas of cow and
-sheep but not present in primates (Ushizawa et al., 2009), is also
-expressed at high levels (\ac{FPKM}= 1853.98). There is no placental
-expression of the estrogen synthase gene (CYP19A1) in dogs, however
-expression of HSD3B1 is relatively high (\ac{FPKM}=157.622).
+The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
+in the term placenta of \textit{Canis familiaris} are shown in
+\cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
+\cref{tab:all_canis_familiaris_placenta}. The gene with the highest
+expression %
+(\ac{FPKM}=\Sexpr{gene.expression("canis
+  familiaris","ENSCAFG00000002115")}) %
+is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
+among the top 50 (\cref{tab:top50_dog_placenta}) include genes
+involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
+anticoagulation (\textit{TFPI2}, \textit{PLA2G7}), innate immunity
+(\textit{LCN2}, \textit{GM2A}), lipid metabolism (\textit{APOE}),
+inhibition of matrix metalloproteinases (\textit{TIMP2},
+\textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
+HTRA1). The gene \textit{LGALS1}, which shows evidence of
+immunosupression during pregnancy
+\cite{Than.ea2008:Emergenceofhormonaland}, shows high expression in
+dog %
+(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
+In addition, the gene SSLP1, a gene expressed in the placentas of cow
+and sheep but not present in primates (Ushizawa et al., 2009), is also
+expressed at high levels
+%
+(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","SSLP1")}). %
+There is no placental expression of the estrogen synthase gene
+(CYP19A1) in dogs, however expression of HSD3B1 is relatively high
+(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","HSD3B1")}).
 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.cow.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.cow.genes,"Gene symbol")
 @ 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.elephant.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.elephant.genes,"Gene symbol")
 @ 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.opossum.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.opossum.genes,"Gene symbol")
 @ 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.mouse.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.mouse.genes,"Gene symbol")
 @ 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.galili.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.galili.genes,"Gene symbol")
 @ 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.carmeli.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.carmeli.genes,"Gene symbol")
 @ 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.pig.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.pig.genes,"Gene symbol")
 @ 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.horse.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.horse.genes,"Gene symbol")
 @ 
@@
       booktabs=TRUE,
-      floating=FALSE
-      )
-setkey(top.sheep.genes,"Gene name")
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.sheep.genes,"Gene symbol")
 @ 

### 2015-12-11 manuscript/supplemental_material.Rnw
load hugo nomenclature

@@
 
-<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","palcenta_go_results_mf")))>>=
+<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","palcenta_go_results_mf","hugo_nomenclature")))>>=
 for (a in paste0("../data/",c("combined_fpkm","gene_trees",

### 2015-12-14 data/ensembl_gene_annotation.R
build ensembl gene annotations

@@ -1,0 +1,27 @@
+library("biomaRt")
+library("data.table")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+
+
+gene.annotation <- list()
+for (species in c("hsapiens","btaurus","cfamiliaris","dnovemcinctus","ecaballus",
+                  "lafricana","mdomestica","mmusculus","oaries","ptroglodytes",
+                  "sscrofa")) {
+    mart <- useMart("ensembl",dataset=paste0(species,"_gene_ensembl"))
+
+    gene.annotation[[species]] <-
+        data.table(getBM(c("ensembl_gene_id",
+                           "external_gene_name",
+                           "gene_biotype",
+                           "chromosome_name",
+                           "start_position"
+                           ),
+                         value=TRUE,mart=mart,bmHeader=FALSE))
+    
+    gene.annotation[[species]][,species:=species]
+}
+gene.annotation <- rbindlist(gene.annotation)
+save(gene.annotation,
+     file=args[length(args)])

### 2015-12-15 manuscript/manuscript.Rnw
add species name aliases
load gene annotation

@@
 
+\newcommand{\smonkey}[0]{\textit{A. fusciceps}}
+\newcommand{\cow}[0]{\textit{B. taurus}}
+\newcommand{\dog}[0]{\textit{C. familiaris}}
+\newcommand{\armadillo}[0]{\textit{D. novemcinctus}}
+\newcommand{\horse}[0]{\textit{E. galili}}
+\newcommand{\elephant}[0]{\textit{L. africana}}
+\newcommand{\human}[0]{\textit{H. sapiens}}
+\newcommand{\opossum}[0]{\textit{M. domestica}}
+\newcommand{\mouse}[0]{\textit{M. musculus}}
+\newcommand{\galili}[0]{\textit{N. galili}}
+\newcommand{\sheep}[0]{\textit{O. aries}}
+\newcommand{\bonobo}[0]{\textit{P. paniscus}}
+\newcommand{\carmeli}[0]{\textit{S. carmeli}}
+\newcommand{\boar}[0]{\textit{S. scrofa}}
+
 \begin{document}
@@
 
-<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","placenta_go_results_mf","hugo_nomenclature")))>>=
+<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","placenta_go_results_mf","hugo_nomenclature","gene_annotation")))>>=
 for (a in paste0("../data/",c("combined_fpkm","gene_trees",
@@
                               "placenta_go_results_mf",
-                              "hugo_nomenclature"
+                              "hugo_nomenclature",
+                              "gene_annotation"
                               ))) {

### 2015-12-15 manuscript/manuscript.Rnw
use species aliases

@@
 technology to sequence and quantify the mRNA environment of six
-mammalian placentas at term, including that of the human, bonobo (the
-sister species to the chimpanzee), dog, cow, African elephant, and
-opossum, representing marsupials as well as three major branches of
+mammalian placentas at term, including that of the \human, \bonobo (the
+sister species to the chimpanzee), \dog, \cow, \elephant, and
+\opossum, representing marsupials as well as three major branches of
 eutherian mammals (Euarchontoglires, Laurasiatheria, Afrotheria).
@@
 Fetal tissue was collected from the placenta of six mammalian species
-(\textit{Homo sapiens} [human], \textit{Pan paniscus} [bonobo], \textit{Canis familiaris}
-[domestic dog], \textit{Bos taurus} [domestic cow], \textit{Loxodonta africana} [African
-savanna elephant], \textit{Monodelphis domestica} [gray short-tailed
-opossum], \textit{Nannospalax gallili} [?], \textit{Spalax carmelli} [?]).
-Information on specimen ID, place and time of collection,
-and collector is shown for each sample in \DLA{Supplemental Table 1}. In the
-case of \textit{Pan}, \textit{Bos}, and \textit{Loxodonta}, the placenta was collected at term
-upon expulsion from the mother following delivery. Placental tissue
-from \textit{Homo sapiens} was collected upon delivery via a caesarian section.
-Near-term placental tissue from \textit{Canis familiaris} was collected upon
-the death of the mother from a car accident. Placental tissue from
-\textit{Monodelphis domestica} was sampled at fetal stage 32, roughly one day
-prior to birth. Only the fetal portion of the placenta was sampled at
-the maternal-fetal interface (i.e. villous portion for \textit{Homo}, \textit{Pan},
-\textit{Canis}, and \textit{Loxodonta}; cotyledon for \textit{Bos}; diffuse placenta for
-\textit{Monodelphis}). All tissues were collected and stored in RNAlater (Life
-Technologies, Carlsbad, CA) until RNA isolation could be performed.
-Tissues were homogenized in Trizol, and total RNA was isolated
-following the manufacturer’s instructions. RNA was further purified
-using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction with the
-RNase-Free DNase Set (Qiagen, Valencia, CA), following the
-manufacturer’s protocol to remove any contaminating DNA from the
-sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
-Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
-Clara, CA) to confirm concentration and quality of the isolation. RNA
-quality was assessed using A260/A280 and A260/A230 absorbance values.
+(\human [human], \bonobo [bonobo], \dog [domestic dog], \cow [domestic
+cow], \elephant [African savanna elephant], \opossum [gray
+short-tailed opossum], \gallili [?], \carmelli [?]). Information on
+specimen ID, place and time of collection, and collector is shown for
+each sample in \DLA{Supplemental Table 1}. In the case of \bonobo,
+\cow, and \elephant, the placenta was collected at term upon expulsion
+from the mother following delivery. Placental tissue from \human was
+collected upon delivery via a caesarian section. Near-term placental
+tissue from \dog was collected upon the death of the mother from a car
+accident. Placental tissue from \opossum was sampled at fetal stage
+32, roughly one day prior to birth. Only the fetal portion of the
+placenta was sampled at the maternal-fetal interface (i.e. villous
+portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
+diffuse placenta for \opossum). All tissues were collected and stored
+in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
+could be performed. Tissues were homogenized in Trizol, and total RNA
+was isolated following the manufacturer’s instructions. RNA was
+further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
+conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
+following the manufacturer’s protocol to remove any contaminating DNA
+from the sample. Total RNA was then evaluated on a Nanodrop 1000
+(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
+(Santa Clara, CA) to confirm concentration and quality of the
+isolation. RNA quality was assessed using A260/A280 and A260/A230
+absorbance values.
 
@@
 For more complete comparisons additional sequences of placenta in
-\textit{Mus musculus} \cite{Wang.ea2011:surveyfornovelimprinted},
-\textit{Homo sapiens}
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \textit{Equus
-  caballus} \DLA{citation}, \textit{Ovis aries} \DLA{citation}, and
-\textit{Sus scrofa} \DLA{citation} were obtained from the \ac{SRA}
-(Supplemental \cref{tab:srr_numbers}).
+\mouse \cite{Wang.ea2011:surveyfornovelimprinted}, \human
+\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
+\DLA{citation}, \sheep \DLA{citation}, and \boar \DLA{citation} were
+obtained from the \ac{SRA} (Supplemental \cref{tab:srr_numbers}).
 
@@
 Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against either
-the \textit{Homo sapiens} or \textit{Mus musculus} transcriptome
+the \human or \mouse transcriptome
 (\cref{tab:ref_genome_aligned_species}).
@@
 The genome of the closely-related \textit{Pan troglodytes} was used as
-a reference for \textit{Pan paniscus}. The separation between the two
+a reference for \bonobo. The separation between the two
 recognized Pan species is generally perceived to be $\approx 1$ Mya
@@
 Due to the paucity of annotated gene ontology terms for some species,
-gene lists were compared using the \textit{Homo sapiens} Gene Ontology
-using the 1:1 \textit{Homo sapiens} ortholog.
+gene lists were compared using the \human Gene Ontology
+using the 1:1 \human ortholog obtained
 
@@
 
-All 1:1 orthologues present among \textit{D. novemcinctus}, \textit{P.
-  troglodytes}, \textit{B. taurus}, \textit{C. familiaris}, \textit{L.
-  africana}, \textit{M. musculus}, \textit{H. sapiens}, and \textit{M.
-  domestica} were determined using Ensembl v.80 gene trees via
+All 1:1 orthologues present among \opossum, \textit{Pan troglodytes},
+\cow, \dog, \elephant, \mouse, and \human were determined using
+Ensembl v.80 gene trees via
 Biomart\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the

### 2015-12-15 manuscript/manuscript.Rnw
comment out OU and spider monkey sections

@@
 \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
-  1,length(unique(human_name))]} 1:1 orthologous genes were
-expressed in at least one species. Of these,
-\Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
-  10,length(unique(human_name))]} 1:1 orthologous genes with
-\ac{FPKM} $\ge 10$ were retained for future analyses of gene
-expression evolution described below.
-
-We used the software \ac{OUCH} in the R statistical environment to
-investigate significant shifts in gene expression levels over
-time\cite{Butler.2004:phlogeneticcomparativeanalysis}. \ac{OUCH}
-institutes an \ac{OU} model of evolution, which allows for different
-optima for quantitative trait values along user-specified branches of
-a phylogenetic tree. The \ac{OU} model explicitly accounts for change
-in quantitative trait values by introducing two parameters into the
-model of stochastic Brownian motion of change in trait value along a
-phylogenetic tree through time. These parameters are $\theta$, an
-optimum trait value, and $\alpha$, the strength of selection
-(interpreted as “pull” toward the optimum). In addition, the \ac{OU}
-model can also integrate evolutionary change by assuming that each
-user-specified set of braches has its own separate \ac{OU} parameters
-\cite{Butler.2004:phlogeneticcomparativeanalysis}.
-
-\DLA{The description of the OU process is terrible; it needs to be
-  completely redone}
-
-Various \ac{OU} branch models were tested against a model in which no
-variation in optima ($\theta$) among branches was allowed. The
-following branches were specified in separate models to have a
-different optimum than all other branches: human, bonobo, mouse, dog,
-cow, elephant, opossum, human+bonobo, Euarchontoglires, and
-Boreoeutheria. Likelihood values were computed for each branch model
-and compared to an \ac{OU} model with one optimum using likelihood
-ratio tests with df=1. We filtered out genes with universally low
-expression levels, and included only genes in which expression in at
-least one species had an $\mathrm{FPKM}\ge10$ as described above. We
-corrected for multiple testing using the \ac{FDR} and using the
-Benjamini-Hochberg correction\cite{Benjamini1995:bh_procedure}. We
-used the time-calibrated phylogenetic tree of
-\cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
-not used in this analyses pruned.
-
-We also calculated the enrichment of functional categories in the
-lists of genes which displayed a significant shift in expression on
-each branch using the same methodology as in the per-species gene
-lists \DLA{reference the tables once this analysis is complete}.
+  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
+in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
+  sapiens" & mean_fpkm >= 10,length(unique(human_name))]} 1:1
+orthologous genes with \ac{FPKM} $\ge 10$ were retained for future
+analyses of gene expression evolution described below.
+
+% We used the software \ac{OUCH} in the R statistical environment to
+% investigate significant shifts in gene expression levels over
+% time\cite{Butler.2004:phlogeneticcomparativeanalysis}. \ac{OUCH}
+% institutes an \ac{OU} model of evolution, which allows for different
+% optima for quantitative trait values along user-specified branches of
+% a phylogenetic tree. The \ac{OU} model explicitly accounts for change
+% in quantitative trait values by introducing two parameters into the
+% model of stochastic Brownian motion of change in trait value along a
+% phylogenetic tree through time. These parameters are $\theta$, an
+% optimum trait value, and $\alpha$, the strength of selection
+% (interpreted as “pull” toward the optimum). In addition, the \ac{OU}
+% model can also integrate evolutionary change by assuming that each
+% user-specified set of braches has its own separate \ac{OU} parameters
+% \cite{Butler.2004:phlogeneticcomparativeanalysis}.
+% 
+% \DLA{The description of the OU process is terrible; it needs to be
+%   completely redone}
+% 
+% Various \ac{OU} branch models were tested against a model in which no
+% variation in optima ($\theta$) among branches was allowed. The
+% following branches were specified in separate models to have a
+% different optimum than all other branches: human, bonobo, mouse, dog,
+% cow, elephant, opossum, human+bonobo, Euarchontoglires, and
+% Boreoeutheria. Likelihood values were computed for each branch model
+% and compared to an \ac{OU} model with one optimum using likelihood
+% ratio tests with df=1. We filtered out genes with universally low
+% expression levels, and included only genes in which expression in at
+% least one species had an $\mathrm{FPKM}\ge10$ as described above. We
+% corrected for multiple testing using the \ac{FDR} and using the
+% Benjamini-Hochberg correction\cite{Benjamini1995:bh_procedure}. We
+% used the time-calibrated phylogenetic tree of
+% \cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
+% not used in this analyses pruned.
+% 
+% We also calculated the enrichment of functional categories in the
+% lists of genes which displayed a significant shift in expression on
+% each branch using the same methodology as in the per-species gene
+% lists \DLA{reference the tables once this analysis is complete}.
 
@@
 
-\subsection{Validation of Primate Expression using Spider Monkey Placenta Transcriptome}
-
-To verify expression levels in human and bonobo, we also generated
-RNA-Seq data for the placenta transcriptome of the black-headed spider
-monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
-Ateles is outside the clade, which includes hominids and can be used
-to polarize expression values of highly expressed genes in Homo and
-Pan. We sampled placental tissue from an Ateles placenta expelled from
-the mother shortly after birth (see Supplemental Table X). We used
-similar procedures discussed above to extract and sequence this
-sample. [DEREK—I’m unsure how to proceed on describing this so as to
-not step on Natalie’s toes]
-
-\DLA{Complete discussion of the ateles placenta}
+% \subsection{Validation of Primate Expression using Spider Monkey Placenta Transcriptome}
+% 
+% To verify expression levels in human and bonobo, we also generated
+% RNA-Seq data for the placenta transcriptome of the black-headed spider
+% monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
+% Ateles is outside the clade, which includes hominids and can be used
+% to polarize expression values of highly expressed genes in Homo and
+% Pan. We sampled placental tissue from an Ateles placenta expelled from
+% the mother shortly after birth (see Supplemental Table X). We used
+% similar procedures discussed above to extract and sequence this
+% sample. [DEREK—I’m unsure how to proceed on describing this so as to
+% not step on Natalie’s toes]
+% 
+% \DLA{Complete discussion of the ateles placenta}
 

### 2015-12-15 manuscript/manuscript.Rnw
discuss SSLP1

@@
 expression %
-(\ac{FPKM}=\Sexpr{gene.expression("canis
-  familiaris","ENSCAFG00000002115")}) %
+(\ac{FPKM}=%
+\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
 is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
@@
 \textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
-HTRA1). The gene \textit{LGALS1}, which shows evidence of
+\textit{HTRA1}). The gene \textit{LGALS1}, which shows evidence of
 immunosupression during pregnancy
-\cite{Than.ea2008:Emergenceofhormonaland}, shows high expression in
-dog %
+\cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
 (\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
-In addition, the gene SSLP1, a gene expressed in the placentas of cow
-and sheep but not present in primates (Ushizawa et al., 2009), is also
+In addition, the gene \textit{SSLP1}, a gene expressed in the
+placentas of \textit{Bos tarus} (%
+\Sexpr{gene.expression("bos tarus","SSLP1")}, %
+\cref{tab:top50_cow_placenta}) and \textit{Ovis Aries}
+(\Sexpr{gene.expression("ovis aries","SSLP1")},
+\cref{tab:top50_sheep_placenta}) but is not present in primates
+\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, is also
 expressed at high levels
@@
     genes.to.tree[combined.fpkm[gene_short_name=="PRLR" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of prolactin receptor}
+\label{fig:prolactin_receptor_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<sslp1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="SSLP1" & species=="homo sapiens",gene_id],tree]
 gene.tree.table <- trees[[tree.id]]$table

### 2015-12-15 manuscript/manuscript.Rnw
clean up text in gene tables

@@
          `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Symbol`),"}")]
+    if (species.name=="homo sapiens") {
+        temp[,name_or_id:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",name_or_id,"}{",xtable.sanitize.text(name_or_id),"}")]
+    } else {
+        temp[,name_or_id:=xtable.sanitize.text(name_or_id)]
+    }
     setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM"))

### 2015-12-15 manuscript/manuscript.Rnw
use xspace
work on discussion of SSLP1

@@
 
-\newcommand{\smonkey}[0]{\textit{A. fusciceps}}
-\newcommand{\cow}[0]{\textit{B. taurus}}
-\newcommand{\dog}[0]{\textit{C. familiaris}}
-\newcommand{\armadillo}[0]{\textit{D. novemcinctus}}
-\newcommand{\horse}[0]{\textit{E. galili}}
-\newcommand{\elephant}[0]{\textit{L. africana}}
-\newcommand{\human}[0]{\textit{H. sapiens}}
-\newcommand{\opossum}[0]{\textit{M. domestica}}
-\newcommand{\mouse}[0]{\textit{M. musculus}}
-\newcommand{\galili}[0]{\textit{N. galili}}
-\newcommand{\sheep}[0]{\textit{O. aries}}
-\newcommand{\bonobo}[0]{\textit{P. paniscus}}
-\newcommand{\carmeli}[0]{\textit{S. carmeli}}
-\newcommand{\boar}[0]{\textit{S. scrofa}}
+\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
+\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
+\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
+\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
+\newcommand{\horse}[0]{\textit{E. galili}\xspace}
+\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
+\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
+\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
+\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
+\newcommand{\galili}[0]{\textit{N. galili}\xspace}
+\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
+\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
+\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
+\newcommand{\boar}[0]{\textit{S. scrofa}\xspace}
 
@@
 
-\subsection{Dog Placenta Transcriptome}
+\subsection{\dog Placenta Transcriptome}
 
@@
 The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
-in the term placenta of \textit{Canis familiaris} are shown in
+in the term placenta of \dog are shown in
 \cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
@@
 (\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
-In addition, the gene \textit{SSLP1}, a gene expressed in the
-placentas of \textit{Bos tarus} (%
+In addition, an ortholog of SSLP1 (ENSCAFG00000010638) a gene
+expressed in the placentas of \cow (%
 \Sexpr{gene.expression("bos tarus","SSLP1")}, %
-\cref{tab:top50_cow_placenta}) and \textit{Ovis Aries}
+\cref{tab:top50_cow_placenta}) and \sheep
 (\Sexpr{gene.expression("ovis aries","SSLP1")},
 \cref{tab:top50_sheep_placenta}) but is not present in primates
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, is also
-expressed at high levels
-%
-(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","SSLP1")}). %
+\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}
+(\cref{fig:sslp1_evolution_tree}), is also expressed at high levels
+%
+(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
 There is no placental expression of the estrogen synthase gene
-(CYP19A1) in dogs, however expression of HSD3B1 is relatively high
+(\textit{CYP19A1}) in \dog (~\ref{fig:estrogen_evolution_tree}), 
+however expression of \textit{HSD3B1} is relatively high
 (\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","HSD3B1")}).
@@
 tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="SSLP1" & species=="homo sapiens",gene_id],tree]
+    genes.to.tree[combined.fpkm[gene_id=="ENSCAFG00000010638" & species=="canis familiaris",gene_id],tree]
 gene.tree.table <- trees[[tree.id]]$table
@@
 @ 
-\caption{Expression of prolactin receptor}
-\label{fig:prolactin_receptor_evolution_tree}
+\caption{Expression of SSLP1}
+\label{fig:sslp1_evolution_tree}
 \end{figure}

### 2015-12-18 manuscript/manuscript.Rnw
load xspace properly
use capitalized crefs
fix ref usage
The pig is actually Duroc, not Wild Boar.
(The testes came from wild boar, and the pig placenta came from a duroc
x duroc cross.)
include bad reads

@@
 \usepackage{nameref}
+\usepackage{xspace}
 \usepackage{zref-xr,zref-user}
@@
 \zexternaldocument*{supplemental_material}
-\usepackage{cleveref}
+\usepackage[capitalize]{cleveref}
 \newenvironment{narrow}[2]{%
@@
 \newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
-\newcommand{\boar}[0]{\textit{S. scrofa}\xspace}
+\newcommand{\pig}[0]{\textit{S. scrofa}\xspace}
 
@@
          "unique.map.reads")
+star.logs.wide[,`% bad reads`:=(as.numeric(`% of reads mapped to too many loci`)+
+                            as.numeric(`% of reads unmapped: other`)+
+                            as.numeric(`% of reads unmapped: too many mismatches`)+
+                            as.numeric(`% of reads unmapped: too short`))]
+star.logs.wide[,`% good reads`:=100-`% bad reads`]
 setkey(hugo.nomenclature,"Approved Symbol")
@@
     \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
-    \textit{Sus scrofa}           & Wild Boar          & Sscrofa10.2 & 51 \\
+    \textit{Sus scrofa}           & Duroc (pig)        & Sscrofa10.2 & 51 \\
     \bottomrule
@@
 
-For transcriptomes sequenced with 75 bp, %
+\Cref{fig:transcriptome_statistics} shows the RNAseq read and
+alignment statistics for all species. For transcriptomes sequenced
+with 75 bp %
+%
 (\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
@@
 \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
-with at least 96\% of reads \DLA{This is wrong}
+with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
 mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
@@
 %
-with at least 99\% of reads mapping to the
-respective reference genome \DLA{This is also wrong}.
+with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
+of reads mapping to the
+respective reference genome (cref{tab:ref_genome_aligned_species}).
 %
@@
 There is no placental expression of the estrogen synthase gene
-(\textit{CYP19A1}) in \dog (~\ref{fig:estrogen_evolution_tree}), 
+(\textit{CYP19A1}) in \dog (\cref{fig:estrogen_evolution_tree}), 
 however expression of \textit{HSD3B1} is relatively high

### 2015-12-18 manuscript/manuscript.Rnw
include tmbs1 tree

@@
 \begin{figure}
+<<tmsb1.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="TMSB10" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of \textit{TMSB1}}
+\label{fig:tmsb1_evolution_tree}
+\end{figure}
+
+\begin{figure}
 <<sslp1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

### 2015-12-18 manuscript/manuscript.Rnw
add discussion of primate vs non-primate changes

@@
 edgelabels("Boreoeutheria",6,frame="none",bg="white",font=1,adj=c(0.7,0),cex=0.9)
+edgelabels("Primates",22,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
+edgelabels("Rodentia",17,frame="none",bg="white",font=1,adj=c(0.7,0),cex=0.9)
 @ 
@@
 
-\subsection{Human Placenta Transcriptome}
+<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("placenta_lineage_specific")))>>=
+load("../data/placenta_lineage_specific")
+source("/home/don/projects/xtable/pkg/R/print.xtable.R")
+
+placenta.lineage.specific.table <- function(analysis,output=TRUE) {
+    temp.table <- 
+        head(placenta.lineage.specific.analysis[analysis.name==analysis &
+                                                p.value <= 0.05][order(p.value),
+                                                                 list(human_name,p.value,fdr,fc),],
+             n=5)
+    temp.table <- data.table(temp.table)
+    setkey(temp.table,"human_name")
+    temp.table <- 
+        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
+                                            `Approved Name`,p.value,fdr,fc)]
+    setkey(temp.table,"Approved Symbol")
+    setnames(temp.table,
+              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
+              )
+    if (output) {
+        print(xtable(temp.table[order(p.value),],
+                     display=c("s","s","s","e","g","f"),
+                     ),
+              booktabs=TRUE,
+              floating=FALSE,
+              math.style.exponents=TRUE,
+              sanitize.text.function=genes.table.sanitize,
+              include.rownames=FALSE)
+
+    }
+    return(invisible(temp.table[order(p.value),]))
+}
+@
+
+\subsection{Primate Placenta Transcriptomes}
+
+\begin{table}
+  \tiny
+<<primate.lineage.specific,echo=FALSE,results="asis">>=
+pri.all.lin.spec <- NULL
+pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
+@ 
+\caption{Genes whose expression changed significantly in the primate
+  order as compared to all other analyzed species.}
+\label{tab:primate_lineage_specific}
+\end{table}
+
+We examined the expression of three primates, \human, \bonobo, and
+\smonkey. The genes whose expressions are different on the primate
+lineage in comparison to all other analyzed species are listed in
+\cref{tab:primate_lineage_specific}. \textit{FSTL1} has significantly
+higher expression in the primate lineage
+(\cref{fig:fstl1_evolution_tree},
+$\mathrm{FDR}=\Sexpr{pri.all.lin.spec["FSTL1","FDR"]}$), is implicated
+in BMP signaling, and has previously been implicated in inflammation
+and invasiveness
+\cite{Mouillet.ea2015:expressionandpost-transcriptionalregulation}. Of
+non-primates, \horse shows the highest expression
+%
+\Sexpr{gene.expression(species.name="equus caballus",gene="FSTL1",complete=TRUE)}, 
+%
+which is still lower than the lowest primate, \smonkey
+%
+(\Sexpr{gene.expression(species.name="ateles fusciceps",gene="FSTL1",complete=TRUE)}. 
+%
+No other genes were significantly differentially expressed on this
+lineage by t-test.
+
+\begin{figure}
+<<fstl1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="FSTL1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=5)
+@ 
+\caption{Expression of follistatin-like 1 \textit{FSTLl1} and related
+  genes.}
+\label{fig:fstl1_evolution_tree}
+\end{figure}
+
+\subsubsection{\human Placenta Transcriptome}
+\label{sec:human_placenta}
 

### 2015-12-18 manuscript/manuscript.Rnw
clean up sequencing discussion

@@
 could be performed. Tissues were homogenized in Trizol, and total RNA
-was isolated following the manufacturer’s instructions. RNA was
+was isolated following the manufacturer's instructions. RNA was
 further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
@@
 (AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
-total RNA and standard Illumina protocol. Paired-end sequencing was
-performed using the Illumina Genome Analyzer II with an insert size of
-230 base pairs (bp). Read lengths for species are shown in
+total RNA and the standard Illumina protocol. Paired-end sequencing
+was performed using the Illumina Genome Analyzer II with an insert
+size of 230 base pairs (bp). Read lengths for species are shown in
 \cref{tab:ref_genome_aligned_species}.
@@
 
-For more complete comparisons additional sequences of placenta in
-\mouse \cite{Wang.ea2011:surveyfornovelimprinted}, \human
+For more complete evolutionary comparisons additional sequences of
+placenta in \mouse \cite{Wang.ea2011:surveyfornovelimprinted}, \human
 \cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
-\DLA{citation}, \sheep \DLA{citation}, and \boar \DLA{citation} were
-obtained from the \ac{SRA} (Supplemental \cref{tab:srr_numbers}).
+\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
+\DLA{citation; e-mailed Alec Archibald from the ISGC about what the
+  right one should be}, and \pig
+\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
+\ac{SRA} (Supplemental \cref{tab:srr_numbers}).
 
@@
 %
-with at least 99\% of reads mapping to the
-respective reference genome \DLA{This is also wrong}.
-%
-All reads were submitted to the NCBI Sequence Read Archive (SRA)
-(Accession Number XXX-YYY).
-
-\DLA{Need to do an embargoed submission to SRA}
-
-The percentage of bases that mapped to Ensembl annotated mRNA ranged
-from ~35\% in elephant to ~85\% in human, with percentages of bases
-mapping to annotated protein-coding segments of genes ranging from
-~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
-from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
-from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
-ranged from ~43\% in elephant to 80\% in bonobo. The number of
-annotated transcripts to which reads mapped in each respective genomes
-ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
-and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
-greater degree of mapping to annotated protein-coding sequences in
-humans corresponds to the greater number of Ensembl annotated
-sequences in the human genome compared to the other five reference
-genomes (Flicek, 2012).
-
-\DLA{Summarize all of this in a table}
-
+with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
+of reads mapping to the
+respective reference genome (cref{tab:ref_genome_aligned_species}).
 The top 50 expressed genes in each genome analyzed are shown in
-\DLA{Tables 2-7}, and all genes for each transcriptome are listed in
-\DLA{Supplementary Tables X-XX} [These are large Excel files to be
-included with the Supplemental Material]. In addition, specific genes
-including placental hormones and hormone-related genes, galectins, and
-insulin-like growth factor related genes are shown with \ac{FPKM} values
-side to side for all transcriptomes including the reanalysis of mouse
-from \cite{Wang.ea2011:surveyfornovelimprinted} in \DLA{Table 8}.
-Below we describe each transcriptome separately, specifically
-concentrating on highly-expressed genes.
+\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elphant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
+and all genes in Supplemental
+\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_nannospalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.
+
+% The percentage of bases that mapped to Ensembl annotated mRNA ranged
+% from ~35\% in elephant to ~85\% in human, with percentages of bases
+% mapping to annotated protein-coding segments of genes ranging from
+% ~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
+% from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
+% from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
+% ranged from ~43\% in elephant to 80\% in bonobo. The number of
+% annotated transcripts to which reads mapped in each respective genomes
+% ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
+% and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
+% greater degree of mapping to annotated protein-coding sequences in
+% humans corresponds to the greater number of Ensembl annotated
+% sequences in the human genome compared to the other five reference
+% genomes (Flicek, 2012).
 

### 2015-12-18 manuscript/manuscript.Rnw
use macros for species names

@@
 The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \textit{Homo sapiens} are shown in
+placenta of \human are shown in
 \cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
@@
 Eleven pregnancy-specific beta-1-glycoprotein genes are highly-expressed
-(\ac{FPKM}$\ge 100$) in the term placenta of humans, with \textit{PSG1} showing the
+(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
 greatest degree of expression %
@@
 
-\subsection{Bonobo Placenta Transcriptome}
+\subsubsection{\bonobo Placenta Transcriptome}
 
@@
 
-\subsection{Spider Monkey Transcriptome}
+\subsubsection{\smonkey Transcriptome}
 
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Ateles fusciceps}}
+\caption{Expression in \ac{FPKM} of top 50 genes in \smonkey which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene}
 \label{tab:top50_spider_monkey_placenta}
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Canis familiaris}}
+\caption{Expression in \ac{FPKM} of the top 50 genes in \dog which are
+  not ribosomal, mitochondrial, or whose human ortholog is a
+  housekeeping gene}
 \label{tab:top50_dog_placenta}
@@
 expression %
+(\ac{FPKM}=%
+\Sexpr{gene.expression("canis familiaris","TMSB10")}) is the thymosin 
+is 
 (\ac{FPKM}=%
@@
 The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Mus musculus} are shown in
+in the term placenta of \mouse are shown in
 \cref{tab:top50_mouse_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_mouse_placenta}.
+genes in Supplemental \cref{tab:all_mus_musculus_placenta}.
 
@@
 \cref{tab:top50_s_carmeli_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_s_carmeli_placenta}.
+genes in Supplemental \cref{tab:all_spalax_carmeli_placenta}.
 
@@
 
-\subsection{Horse Placenta Transcriptome}
+\subsection{\horse Placenta Transcriptome}
 
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Equus caballus}}
+\caption{Expression in \ac{FPKM} of top 50 genes in \horse which are
+  not ribosomal, mitochondrial, or whose human ortholog is a
+  housekeeping gene}
 \label{tab:top50_horse_placenta}
@@
 \cref{tab:top50_horse_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_horse_placenta}.
-
-\subsection{Sheep Placenta Transcriptome}
+genes in Supplemental \cref{tab:all_equus_caballus_placenta}.
+
+\subsection{\sheep Placenta Transcriptome}
 
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Ovis aries}}
+\caption{Expression in \ac{FPKM} of top 50 genes in \sheep which are
+  not ribosomal, mitochondrial, or whose human ortholog is a
+  housekeeping gene}
 \label{tab:top50_sheep_placenta}
@@
 The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Ovis aries} are shown in
+in the term placenta of \sheep are shown in
 \cref{tab:top50_sheep_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_sheep_placenta}.
+genes in Supplemental \cref{tab:all_ovis_aries_placenta}.
 

### 2015-12-20 manuscript/manuscript.Rnw
fix up bonobo discussion

@@
 The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the bonobo are shown in \cref{tab:top50_bonobo_placenta}
+placenta of the \bonobo are shown in \cref{tab:top50_bonobo_placenta}
 with a list of all genes in Supplemental
-\cref{tab:all_pan_paniscus_placenta}. Expression in the bonobo
-placenta differs somewhat from that of the human, as evidenced by the
+\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
+placenta differs somewhat from that of the \human, as evidenced by the
 makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
-(\ac{FPKM}=%
-\Sexpr{gene.expression("pan paniscus","PRG2")}%
-%
-) in bonobo is \textit{PRG2}, which is involved in insulin
+ in \bonobo is %
+\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
+%
+ which is involved in insulin
 growth factor signaling and angiogenesis during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}. \textit{PRG2} is also expressed at relatively high levels in the
 human placenta (\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")}; Supplemental \cref{tab:all_homo_sapiens_placenta}), although at
-comparatively much lower levels than the bonobo. A bonobo gene closely
+comparatively much lower levels than the \bonobo. A \bonobo gene closely
 related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
@@
 pregnancy-specific beta-1-glycoprotein genes were expressed in the
-bonobo placenta with \ac{FPKM} ranging from 20.885-167.961 (Supplemental
-Table X). Placental expression of the genes involved in estrogen and
+\bonobo placenta with \ac{FPKM} ranging from 20.885-167.961 (Supplemental
+\cref{tab:all_pan_paniscus_placenta}). Placental expression of the genes involved in estrogen and
 progesterone production (CYP19A1, HSD3B1 respectively) show lower

### 2016-01-04 common_r_code/plot_tree_matrix.R
fix tree
fix spacing of gene names
vertical species names
ggplot default gradient

@@ -1,21 +1,25 @@
-plot.tree.matrix <- function(gene.tree,gene.tree.table,expression,min.fpkm=NA) {
+library("magrittr")
+library("tidyr")
+plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offset.ratio=0.5,min.fpkm=NA) {
     gene.tree$tip.label <- gsub("^(ENS.+)\\1$","\\1",gene.tree$tip.label)
     gene.tree.table[,species:=capfirst(gsub("_"," ",species))]
-    gene.tree.table[,symbol_or_id:=ifelse(symbol=="NULL",gene_id,paste0(symbol," ",species))]
+    gene.tree.table[,short.species:=gsub("^(.)[^ ]* +([^ ]*)","\\1. \\2",species)]
+    gene.tree.table[,symbol_or_id:=ifelse(symbol=="NULL",gene_id,paste0(symbol," ",short.species))]
     gene.tree.table[duplicated(symbol_or_id),symbol_or_id:=paste0(symbol_or_id," dup")]
-    gene.tree.expression <-
-        combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+    gene.tree.expression.subset <- gene.tree.expression
     if (!is.na(min.fpkm)) {
-        gene.tree.expression <-
+        gene.tree.expression.subset <-
             gene.tree.expression[mean_fpkm >= min.fpkm ,]
     }
     setkey(gene.tree.table,"protein_id")
     gene.tree$tip.label <-
         gene.tree.table[gene.tree$tip.label,gene_id]
     gene.tree.subset <-
-        drop.tip(gene.tree,gene.tree$tip.label[!(gene.tree$tip.label %in%
-                                                     gene.tree.expression[,gene_id])])
-
+        drop.tip(gene.tree,
+                 gene.tree$tip.label[!(gene.tree$tip.label %in%
+                                       gene.tree.expression.subset[,gene_id])])
+    gene.tree.expression <-
+        gene.tree.expression[gene_id %in% gene.tree$tip.label,]
     gene.tree.expression[,mean_fpkm_log:=log10(mean_fpkm+1)]
     gene.tree.expression.matrix <-
         dcast(gene.tree.expression,
@@ -29,7 +33,7 @@ plot.tree.matrix <- function(gene.tree,gene.tree.table,expression,min.fpkm=NA) {
     colnames(gene.tree.expression.matrix) <-
[…]

### 2016-01-04 manuscript/manuscript.Rnw
fix caballus name
fix references to reference genome table

@@
 \newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
-\newcommand{\horse}[0]{\textit{E. galili}\xspace}
+\newcommand{\horse}[0]{\textit{E. caballus}\xspace}
 \newcommand{\elephant}[0]{\textit{L. africana}\xspace}
@@
 of reads mapping to the
-respective reference genome (cref{tab:ref_genome_aligned_species}).
+respective reference genome (\cref{tab:ref_genome_aligned_species}).
 %
@@
 of reads mapping to the
-respective reference genome (cref{tab:ref_genome_aligned_species}).
+respective reference genome (\cref{tab:ref_genome_aligned_species}).
 The top 50 expressed genes in each genome analyzed are shown in

### 2016-01-04 manuscript/manuscript.Rnw
fix top primate changes table

@@
                                             `Approved Name`,p.value,fdr,fc)]
-    setkey(temp.table,"Approved Symbol")
     setnames(temp.table,
@@
               )
+    setkey(temp.table,"Human Symbol")
     if (output) {
-        print(xtable(temp.table[order(p.value),],
+        print(xtable(temp.table[order(`p (t-test)`),],
                      display=c("s","s","s","e","g","f"),
@@
     }
-    return(invisible(temp.table[order(p.value),]))
+    return(invisible(temp.table[order(`p (t-test)`),]))
 }
@@
 <<primate.lineage.specific,echo=FALSE,results="asis">>=
-pri.all.lin.spec <- NULL
 pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
+setkey(pri.all.lin.spec,"Human Symbol")
 @ 

### 2016-01-04 manuscript/manuscript.Rnw
include hormone/igf/galectin heatmap figure

@@
 library("org.Hs.eg.db")
+library("scales")
 
@@
 \begin{figure}
+<<hormone.genes.plot,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+
+gene.expression.nozero <- function(genes) {
+    temp <-
+        data.matrix.dt(dcast(combined.fpkm[human_name %in% genes,],
+                             human_name~species,
+                             value.var="FPKM",
+                             fun.aggregate=function(x){x[1]}))
+    temp <-
+        temp[apply(temp,1,function(x){sum(x,na.rm=TRUE)})>5,]
+    return(temp)
+}
+
+### hormone genes
+hormone.genes <- c("CGA","CGB","CGB1","CGB2","CGB5","CGB8","CSH1","CSH2","CSHL1","CYP19A1",
+                   "GH2","GHRH","HSD3B1","NPPB","NPPC","PRL","RLN1")
+hormone.expression.nozero <-
+    gene.expression.nozero(hormone.genes)
+
+hor.exp.long <- melt(log10(hormone.expression.nozero+1)[rev(1:nrow(hormone.expression.nozero)),])
+colnames(hor.exp.long)[1:3] <- c("gene","species","fpkm")
+hor.exp.long <- hor.exp.long[hor.exp.long$gene!="IGF2",]
+
+### galectin genes
+galectin.genes <- unique(grep("^LGALS",combined.fpkm[,human_name],value=TRUE))
+
+galectin.expression.nozero <-
+    gene.expression.nozero(galectin.genes)
+
+gal.exp.long <- melt(log10(galectin.expression.nozero+1)[rev(1:nrow(galectin.expression.nozero)),])
+colnames(gal.exp.long)[1:3] <- c("gene","species","fpkm")
+
+### IGF related genes
+
+igfrelated.genes <- unique(grep("^IGF",combined.fpkm[,human_name],value=TRUE))
+
+igfrelated.expression.nozero <-
+    gene.expression.nozero(igfrelated.genes)
+
+igf.exp.long <- melt(log10(igfrelated.expression.nozero+1)[rev(1:nrow(igfrelated.expression.nozero)),])
+colnames(igf.exp.long)[1:3] <- c("gene","species","fpkm")
+
+igf.exp.long$type <- "IGF Related"
+gal.exp.long$type <- "Galectin"
+hor.exp.long$type <- "Hormone"
+
+### scalerange <- range(c(igf.exp.long$fpkm,gal.exp.long$fpkm,hor.exp.long$fpkm),na.rm=TRUE)
+### gradientends <- scalerange + rep(c(0,100,200), each=2)
+### colorends <- c("white", "red", "white", "green", "white", "blue")
+### 
+### gal.exp.long$fpkm <- gal.exp.long$fpkm+100
+### hor.exp.long$fpkm <- hor.exp.long$fpkm+200
+
+combined.long <-
+    rbind(igf.exp.long,
+          gal.exp.long,
+          hor.exp.long)
+combined.long$species <-
+    factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(combined.long$species)),
+           levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(names(species.names))))
+
+common.plot.options <- 
+    list(geom_tile(aes(fill = fpkm), colour = "white"),
+##         scale_fill_gradientn(colours = colorends, values = rescale(gradientends)),
+         scale_fill_gradient(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = "grey50",
+                             guide=guide_colorbar(title=expression(log[10](FPKM)))),
+         scale_x_discrete("", expand = c(0, 0)),
+         scale_y_discrete("", expand = c(0, 0)),
+         theme_grey(base_size = 9),
+         theme(legend.position = "top",
+               axis.ticks = element_blank(), 
+               axis.text.x = element_text(angle = 300, hjust = 0, vjust=1,face="italic")),
+         facet_wrap(~type,scale="free_y",ncol=1)
+         )
+print(ggplot(combined.long, aes(y=gene, x=species))
+      + common.plot.options
+      )
+@ 
+\caption{Expression of Hormone Genes, Galectins, and IGF related genes}
+\label{fig:hormone_genes}
+\end{figure}
+
+\begin{figure}
 <<estrogen.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

### 2016-01-05 manuscript/manuscript.Rnw
show lower levels in PSG tree plot

@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=5)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1)
 @ 

### 2016-01-05 manuscript/manuscript.Rnw
work on bonobo section

@@
 \caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Pan paniscus}}
+mitochondrial, or whose \human ortholog is a housekeeping gene in \bonobo }
 \label{tab:top50_bonobo_placenta}
@@
 The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the \bonobo are shown in \cref{tab:top50_bonobo_placenta}
+placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
 with a list of all genes in Supplemental
 \cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
-placenta differs somewhat from that of the \human, as evidenced by the
+placenta differs somewhat from that of \human, as evidenced by the
 makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
@@
 comparatively much lower levels than the \bonobo. A \bonobo gene closely
-related to the chorionic somatomammotropins (Q95MK7\_PANTR) is also
-expressed at high levels (\ac{FPKM}=1232.56; \cref{tab:top50_bonobo_placenta}). Two other Pan genes
-included within this family (ENSPTRG00000030805 which shows similarity
-to GH2 and ENSPTRG00000034560) show expression levels of \ac{FPKM}>89. Six
+related to the chorionic somatomammotropins (ENSPTRG00000034560) is also
+expressed at high levels %
+%
+(\ac{FPKM}=\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560")}); %
+%
+\cref{tab:top50_bonobo_placenta}). Two other Pan genes included within
+this family (ENSPTRG00000030805 which shows similarity to GH2 and
+ENSPTRG00000034560) show expression levels of \ac{FPKM}>89. Six
 pregnancy-specific beta-1-glycoprotein genes were expressed in the
-\bonobo placenta with \ac{FPKM} ranging from 20.885-167.961 (Supplemental
-\cref{tab:all_pan_paniscus_placenta}). Placental expression of the genes involved in estrogen and
-progesterone production (CYP19A1, HSD3B1 respectively) show lower
-expression in the bonobo as compared to humans (CYP19A1 \ac{FPKM}= 48.6877;
-HSD3B1 \ac{FPKM}= 54.5301; Table 8). In addition, CGA is expressed in the
-bonobo placenta (\ac{FPKM}=202.629), but at much lower levels than humans
-as well (Table 8). Two Ensembl-annotated beta subunits of the
-chorionic gonadotropin genes are expressed in bonobo, one related to
-CGB and another related to CGB7; both are not highly expressed (Table
-8). Among the other top 50 expressed genes, there is high expression
-of ribosomal proteins (e.g., RPL18, RPL23), keratins (e.g., KRT7,
-KRT19), heat-shock proteins (e.g., HSPA5, HSPA1A), serpins (SERPINE1,
-SERPINE2), insulin growth factor 2 (IGF2) and the hormone
-adrenomedullin (ADM).
+\bonobo placenta with \ac{FPKM} ranging from 20.885-167.961
+(Supplemental \cref{tab:all_pan_paniscus_placenta}). Placental
+expression of the genes involved in estrogen and progesterone
+production (CYP19A1, HSD3B1 respectively) show lower expression in the
+bonobo as compared to humans %
+(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
+\cref{fig:estrogen_evolution_tree};
+\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
+\cref{fig:progesterone_evolution_tree} and
+\cref{tab:top50_human_placenta}) although the expression is fairly similar to that of 
+. In addition, \textit{CGA} is
+expressed in the bonobo placenta
+(\ac{FPKM}=\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
+lower levels than \human as well (\cref{tab:top50_human_placenta}). 
+%
+Two Ensembl-annotated beta subunits of the chorionic gonadotropin
+genes are expressed in \bonobo, one related to CGB and another related
+to CGB7; both are not highly expressed (Table 8). Among the other top
+50 expressed genes, there is high expression of ribosomal proteins
+(e.g., RPL18, RPL23), keratins (e.g., KRT7, KRT19), heat-shock
+proteins (e.g., HSPA5, HSPA1A), serpins (SERPINE1, SERPINE2), insulin
+growth factor 2 (IGF2) and the hormone adrenomedullin (ADM).
 
@@
       + scale_y_discrete(expand=c(0,0))
-      + scale_fill_gradient(low="white",
-                            high="blue",
+      + scale_fill_gradient(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = "grey50",
                             limits=c(0,4),

### 2016-01-05 manuscript/manuscript.Rnw
add missing parenthesis
use \human, not just human

@@
 %
-(\Sexpr{gene.expression(species.name="ateles fusciceps",gene="FSTL1",complete=TRUE)}. 
+(\Sexpr{gene.expression(species.name="ateles fusciceps",gene="FSTL1",complete=TRUE)}). 
 %
@@
 %
-in humans is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
+in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
 three of which are members of a distinct expansion of the growth
@@
 \Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
-is also expressed to a high degree in the human placenta
+is also expressed to a high degree in the \human placenta
 (\cref{tab:top50_human_placenta}).

### 2016-01-05 data/diamond_against_human/gene_to_human_gene.pl
Add diamond against human
Rules to generate the best match for all proteins in other genomes;
primarily useful for describing what unnanoated proteins without 1:1
orthologs are likely to be.

@@ -1,0 +1,70 @@
+#!/usr/bin/perl
+
+use warnings;
+use strict;
+
+use IO::File;
+use Data::Dumper;
+
+my ($prot_to_gene,@alignment_files) = @ARGV;
+
+sub open_file {
+    my $fh = IO::File->new($_[0],'r') or
+        die "Unable to open $_[0] for reading: $!";
+    return $fh;
+}
+
+my $p2g_fh = open_file($prot_to_gene);
+
+my %p2g;
+while (<$p2g_fh>) {
+    next unless /^ENS/;
+    chomp;
+    my ($prot,$gene,$name) = split /\t/;
+    $p2g{$prot} = $gene;
+}
+
+sub print_match {
+    my ($gene,$prot,$human_prot) = @_;
+    print join("\t",$gene,$prot,$p2g{$human_prot},$human_prot)."\n";
+}
+
+print join("\t","gene_id","protein_id","human_gene_id","human_protein")."\n";
+for my $alignment_fh (map {open_file($_)} @alignment_files) {
+    my $previous_id_cont = '';
+    my $previous_gene_id;
+    my $previous_prot_id;
+    my $previous_human_protein;
+    my $previous_bit_score = 0;
+    while (<$alignment_fh>) {
+        chomp;
+        my ($id_cont,$human_protein,$identity_per_len,
+            $len,$mismatches,$gaps,$query_begin,$query_end,
+            $subject_begin,$subject_end,$evalue,$bitscore) = split /\t/;
+        ## we only want the top scoring match
+        next if $id_cont eq $previous_id_cont;
+        $previous_id_cont = $id_cont;
+        my ($prot_id,$gene_id) = split /_/,$id_cont;
+        if (defined $previous_gene_id and
+            $gene_id ne $previous_gene_id) {
+            print_match($previous_gene_id,$previous_prot_id,
+                        $previous_human_protein);
+            $previous_gene_id=undef;
+            $previous_prot_id=undef;
+            $previous_human_protein=undef;
+            $previous_bit_score = 0;
+        }
+        if (not defined $previous_gene_id or
+            $previous_bit_score < $bitscore
+           ) {
+            $previous_gene_id = $gene_id;
+            $previous_prot_id = $prot_id;
+            $previous_human_protein = $human_protein;
+            $previous_bit_score = $bitscore;
+        }
+    }
+    if (defined $previous_gene_id) {
+        print_match($previous_gene_id,$previous_prot_id,
+                    $previous_human_protein);
+    }
+}

### 2016-01-05 common_r_code/plot_tree_matrix.R
use the gene_id for all duplicates in the gene tree

@@ -5,7 +5,9 @@ plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offs
     gene.tree.table[,species:=capfirst(gsub("_"," ",species))]
     gene.tree.table[,short.species:=gsub("^(.)[^ ]* +([^ ]*)","\\1. \\2",species)]
     gene.tree.table[,symbol_or_id:=ifelse(symbol=="NULL",gene_id,paste0(symbol," ",short.species))]
-    gene.tree.table[duplicated(symbol_or_id),symbol_or_id:=paste0(symbol_or_id," dup")]
+    ## use symbol gene_id for duplicates
+    gene.tree.table[duplicated(symbol_or_id)|duplicated(symbol_or_id,fromLast=TRUE),
+                    symbol_or_id:=paste0(symbol," ",gene_id)]
     gene.tree.expression.subset <- gene.tree.expression
     if (!is.na(min.fpkm)) {
         gene.tree.expression.subset <-

### 2016-01-05 data/gtf_types/gene_id_to_type.pl
add GTF types code

@@ -1,0 +1,17 @@
+#!/usr/bin/perl
+
+use warnings;
+use strict;
+
+use Data::Printer;
+
+
+print join("\t",qw(gene_id gene_biotype))."\n";
+while (<>) {
+    next if /^#/;
+    chomp;
+    my @f = split /\t/;
+    next unless $f[2] eq 'gene';
+    my %t = map {/(\S+)\s+"([^"]+)"/ && ($1,$2)} split /;/,$f[8];
+    print join("\t",$t{gene_id},$t{gene_biotype})."\n";
+}

### 2016-01-05 data/diamond_against_human/gene_to_human_gene.pl
include gene name in diamond to human gene output

@@ -21,7 +21,7 @@ while (<$p2g_fh>) {
     next unless /^ENS/;
     chomp;
     my ($prot,$gene,$name) = split /\t/;
-    $p2g{$prot} = $gene;
+    $p2g{$prot} = "$gene\t$name";
 }
 
 sub print_match {
@@ -29,7 +29,7 @@ sub print_match {
     print join("\t",$gene,$prot,$p2g{$human_prot},$human_prot)."\n";
 }
 
-print join("\t","gene_id","protein_id","human_gene_id","human_protein")."\n";
+print join("\t","gene_id","protein_id","human_gene_id","human_gene_symbol","human_protein")."\n";
 for my $alignment_fh (map {open_file($_)} @alignment_files) {
     my $previous_id_cont = '';
     my $previous_gene_id;

### 2016-01-05 data/combine_fpkm.R
include diamond homolog and gene type

@@ -118,6 +118,21 @@ combined.fpkm[name_or_id=="-" | name_or_id=="",name_or_id:=gene_id]
 ### this is wrong for pan paniscus, so fixing it here
 combined.fpkm[human_name=="GH1" & oma_group_name=="GH2",human_name:=NA]
 
+### diamond gene_to_human_gene.txt
+diamond.align <- fread(args[6])
+setkey(diamond.align,"gene_id")
+setnames(diamond.align,"human_gene_symbol","human_alignment_symbol")
+setkey(combined.fpkm,"gene_id")
+combined.fpkm <-
+    diamond.align[,list(gene_id,human_alignment_symbol)][combined.fpkm]
+
+### gtf types
+gtf.types <- fread(args[7])
+setkey(gtf.types,"gene_id")
+setkey(combined.fpkm,"gene_id")
+combined.fpkm <-
+    gtf.types[combined.fpkm]
+setkey(combined.fpkm,"tracking_id")
 
 if (!all(grepl("_per_sample",output.file))) {
     combined.fpkm <-

### 2016-01-05 manuscript/manuscript.Rnw
include tocs to subsubsection
fix up top50 table explanation

@@
   {\end{itemize}}
+\setcounter{tocdepth}{3}
 \oddsidemargin 0.0in 
@@
 \newcommand{\pig}[0]{\textit{S. scrofa}\xspace}
+
+% top 50 gene table explanation. Note that if you modify this, you
+% should probably also modify the human table explanation, which is
+% slightly different.
+\newcommand{\tableexplanation}[2]{The top 50 genes by \ac{FPKM} in #1
+  which are not annotated as Mt\_rRNA, Mt\_tRNA, rRNA or whose \human
+  ortholog is a housekeeping gene. Gene symbol is the gene symbol or
+  Ensembl gene id. Human Symbol is the 1:1 human ortholog or the best
+  human alignment when enclosed in parenthesis. Human Name is the HUGO
+  name corresponding to the human symbol. A complete list of all genes
+  is in \cref{#2}.}
 
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not
-  ribosomal, mitochondrial, or a housekeeping gene in \textit{Homo
-    sapiens}}
+\caption{The top 50 genes by \ac{FPKM} in \human which are not
+  annotated as Mt\_rRNA, Mt\_tRNA, rRNA or a housekeeping gene. Gene
+  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
+  name corresponding to the human symbol. A complete list of all genes
+  is in \cref{tab:all_homo_sapiens_placenta}.}
 \label{tab:top50_human_placenta}
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose \human ortholog is a housekeeping gene in \bonobo }
+\caption{\tableexplanation{\bonobo}{tab:all_pan_paniscus_placenta}}
 \label{tab:top50_bonobo_placenta}

### 2016-01-05 manuscript/manuscript.Rnw
finish up bonobo paragraph+fix top50 table

@@
 top.genes.table <- function(species.name,n=50) {
+    alignment.species <- species.name
+    if (alignment.species=="pan paniscus") {
+        alignment.species <- "pan troglodytes"
+    } else if (alignment.species=="ateles fusciceps") {
+        alignment.species <- "homo sapiens"
+    } else if (grepl("spalax",alignment.species)) {
+        alignment.species <- "mus musculus"
+    }
     temp <- 
@@
                         human_name %in% housekeeping.genes.superset[,gene_short_name]) &
-                               !grepl("((RNA\\d|MT)-|^RP[SL]\\d+$)",human_name) &
-                                   !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[l]?(?:-.+)?$|rRNA$)",gene_short_name)
+                      ! (gene_biotype %in% c("Mt_tRNA","Mt_rRNA","rRNA")) &
+                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?$)",human_name) &
+                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?$)",human_alignment_symbol) &
+                      !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[l]?(?:-.+)?$|rRNA$)",
+                             gene_short_name)
                       ][order(-mean_fpkm)][,list(name_or_id,
+                                                 human_alignment_symbol,
                                                  human_name,mean_fpkm)]
@@
     temp <- temp[!duplicated(name_or_id)][1:n]
+    temp[,used_alignment:=FALSE]
+    temp[is.na(human_name) & !is.na(human_alignment_symbol),
+                    used_alignment:=TRUE]
+    temp[is.na(human_name),human_name:=human_alignment_symbol]
     setkey(temp,"human_name")
@@
         hugo.nomenclature[temp][order(-mean_fpkm),
-                                list(name_or_id,`Approved Symbol`,`Approved Name`,mean_fpkm)]
+                                list(name_or_id,used_alignment,`Approved Symbol`,`Approved Name`,mean_fpkm)]
+    temp[!is.na(name_or_id),
+         name_or_id:=paste0("\\href{http://www.ensembl.org/",
+                             capfirst(gsub(" ","_",alignment.species)),
+                             "/Gene/Summary?g=",name_or_id,"}{",
+                             xtable.sanitize.text(name_or_id),"}")]
     temp[!is.na(`Approved Name`),
@@
          `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Symbol`),"}")]
-    if (species.name=="homo sapiens") {
-        temp[,name_or_id:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",name_or_id,"}{",xtable.sanitize.text(name_or_id),"}")]
-    } else {
-        temp[,name_or_id:=xtable.sanitize.text(name_or_id)]
-    }
+    temp[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
+    temp[,used_alignment:=NULL]
     setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM"))
-    if (species.name=="homo sapiens") {
-        temp <- temp[,-2,with=FALSE]
+    if (species.name=="homo sapiens" || species.name=="ateles fusciceps") {
+        temp[,`Human Symbol`:=NULL]
     }
@@
     fpkm <- combined.fpkm[species==species.name &
-                              gene_short_name==gene,
+                              (gene_short_name==gene | 
+                               gene_id==gene),
                           mean_fpkm]
-
     if (complete) {
+        if (any(grepl("^ENS.+G\\d+$",gene))) {
+            return(paste0(gene,
+                          formatC(fpkm,digits=0,format="f"),
+                          " \\ac{FPKM}"))
+        } else {
             return(paste0("\\textit{",gene,"}=",
@@
                           ))
+        }
     } else {
@@
 %
- which is involved in insulin
-growth factor signaling and angiogenesis during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}. \textit{PRG2} is also expressed at relatively high levels in the
-human placenta (\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")}; Supplemental \cref{tab:all_homo_sapiens_placenta}), although at
-comparatively much lower levels than the \bonobo. A \bonobo gene closely
-related to the chorionic somatomammotropins (ENSPTRG00000034560) is also
-expressed at high levels %
-%
-(\ac{FPKM}=\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560")}); %
-%
-\cref{tab:top50_bonobo_placenta}). Two other Pan genes included within
-this family (ENSPTRG00000030805 which shows similarity to GH2 and
-ENSPTRG00000034560) show expression levels of \ac{FPKM}>89. Six
-pregnancy-specific beta-1-glycoprotein genes were expressed in the
-\bonobo placenta with \ac{FPKM} ranging from 20.885-167.961
-(Supplemental \cref{tab:all_pan_paniscus_placenta}). Placental
-expression of the genes involved in estrogen and progesterone
-production (CYP19A1, HSD3B1 respectively) show lower expression in the
-bonobo as compared to humans %
+which is involved in insulin growth factor signaling and angiogenesis
+during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
+\textit{PRG2} is also expressed at relatively high levels (though
+comparatively much lower) in the human placenta
+(\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")};
+Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
+related to the chorionic somatomammotropins
+are also expressed at relatively high levels (GH1 
+\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
+\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
+GH2
+\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
+\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
+%
+\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
+Placental expression of the genes involved in estrogen and
+progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
+respectively) show lower expression in \bonobo as compared to \human %
 (\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
-\cref{fig:estrogen_evolution_tree};
+\cref{fig:estrogen_evolution_tree}; %
 \Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
 \cref{fig:progesterone_evolution_tree} and
-\cref{tab:top50_human_placenta}) although the expression is fairly similar to that of 
-. In addition, \textit{CGA} is
-expressed in the bonobo placenta
+\cref{tab:top50_human_placenta}) although the expression of
+\textit{HSD3B1} is fairly similar to that of \smonkey %
+(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
+In addition, \textit{CGA} is expressed in the bonobo placenta %
 (\ac{FPKM}=\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
@@
 Two Ensembl-annotated beta subunits of the chorionic gonadotropin
-genes are expressed in \bonobo, one related to CGB and another related
-to CGB7; both are not highly expressed (Table 8). Among the other top
-50 expressed genes, there is high expression of ribosomal proteins
-(e.g., RPL18, RPL23), keratins (e.g., KRT7, KRT19), heat-shock
-proteins (e.g., HSPA5, HSPA1A), serpins (SERPINE1, SERPINE2), insulin
-growth factor 2 (IGF2) and the hormone adrenomedullin (ADM).
+genes are expressed in \bonobo, one related to \textit{CGB} and
+another related to \textit{CGB7}; both are not highly expressed
+(\cref{fig:chorionic_gonadotropin}). Among the other top 50 expressed
+genes, there is high expression of keratins (e.g., \textit{KRT19},
+\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
+\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2}),
+insulin growth factor 2 (\textit{IGF2}) and one of its binding
+proteins \textit{IGFBP3}, and the hormone adrenomedullin
+(\textit{ADM}).
 

### 2016-01-05 manuscript/manuscript.Rnw
add missing armadillo table

@@
 
+% species names
 \newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
@@
 
-\subsection{African Elephant Placenta Transcriptome}
+\subsection{\elephant Placenta Transcriptome}
 
@@
 1987).
+
+\subsection{\armadillo Placenta Transcriptome}
+
+\begin{table}
+\tiny
+<<top.50.armadillo.expressed.genes,echo=FALSE,results="asis">>=
+top.armadillo.genes <- top.genes.table(species.name="dasypus novemcinctus")
+print(xtable(top.armadillo.genes),
+      booktabs=TRUE,
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.armadillo.genes,"Gene symbol")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
+  \armadillo}
+\label{tab:top50_armadillo_placenta}
+\end{table}
+
+The top 50 expressed genes in the term placenta of \armadillo are in
+\cref{tab:top50_armadillo_placenta}.
 

### 2016-01-06 manuscript/manuscript.Rnw
add paragraph on spidermonkey

@@
 
+For the term placenta of \smonkey, assembled contigs from RNA-seq were
+aligned using against \human and the best alignment was chosen as the
+annotation for each assembled contig. In the term placenta of
+\smonkey, the top 50 non-housekeeping genes
+(\cref{tab:top50_spider_monkey_placenta}) include a possible regulator
+of androgen receptor signaling
+\cite{Sampson.ea2012:PAGE4positivityisassociated}, \textit{PAGE4} %
+(\ac{FPKM}=\Sexpr{gene.expression("ateles fusciceps","PAGE4")}), %
+the pregnancy specific glycoprotein \textit{PSG8}
+(\ac{FPKM}=\Sexpr{gene.expression("ateles fusciceps","PSG8")}), %
+a member of the chorionic somatotropins, \textit{GH1} %
+(\ac{FPKM}=\Sexpr{gene.expression("ateles fusciceps","GH1")}), %
+a few HLA genes (\textit{HLA-A}, \textit{HLA-F},
+\cref{tab:top50_spider_monkey_placenta}), and multiple components of
+the annexin complex %
+(\Sexpr{gene.expression("ateles fusciceps","ANXA2",complete=TRUE)}, %
+\Sexpr{gene.expression("ateles fusciceps","S100P",complete=TRUE)}, %
+\Sexpr{gene.expression("ateles fusciceps","S100A11",complete=TRUE)}) %
+which may play a role in angiogenesis, and certainly regulate
+cell-cell interaction in the developing placenta
+\cite{Liu.ea2015:AnnexinA2complexeswith}.
+
 \subsection{\dog Placenta Transcriptome}

### 2016-01-06 manuscript/manuscript.Rnw
fix cgb table link

@@
 another related to \textit{CGB7}; both are not highly expressed
-(\cref{fig:chorionic_gonadotropin}). Among the other top 50 expressed
+(\cref{fig:cgb_evolution_tree}). Among the other top 50 expressed
 genes, there is high expression of keratins (e.g., \textit{KRT19},

### 2016-01-06 manuscript/manuscript.Rnw
move text sanitizing funcs to utility_functions.R
genes.table.sanitize and xtable.sanitize are now in utility functions

@@
 }
-genes.table.sanitize <- function(str){
-    return(str)
-}
-xtable.sanitize.text <- function(str){
-    result <- str
-    result <- gsub("\\\\", "SANITIZE.BACKSLASH", result)
-    result <- gsub("$", "\\$", result, fixed = TRUE)
-    result <- gsub(">", "$>$", result, fixed = TRUE)
-    result <- gsub("<", "$<$", result, fixed = TRUE)
-    result <- gsub("|", "$|$", result, fixed = TRUE)
-    result <- gsub("{", "\\{", result, fixed = TRUE)
-    result <- gsub("}", "\\}", result, fixed = TRUE)
-    result <- gsub("%", "\\%", result, fixed = TRUE)
-    result <- gsub("&", "\\&", result, fixed = TRUE)
-    result <- gsub("_", "\\_", result, fixed = TRUE)
-    result <- gsub("#", "\\#", result, fixed = TRUE)
-    result <- gsub("^", "\\verb|^|", result, fixed = TRUE)
-    result <- gsub("~", "\\~{}", result, fixed = TRUE)
-    result <- gsub("SANITIZE.BACKSLASH", "$\\backslash$",
-                   result, fixed = TRUE)
-    return(result)
-}
 gene.expression <- function(species.name,gene,complete=FALSE) {

### 2016-01-06 common_r_code/utility_functions.R

@@ -101,6 +101,28 @@ english.join <- function(s){
     return(paste(sep="",collapse=" ",s))
 }
 
+genes.table.sanitize <- function(str){
+    return(str)
+}
+xtable.sanitize.text <- function(str){
+    result <- str
+    result <- gsub("\\\\", "SANITIZE.BACKSLASH", result)
+    result <- gsub("$", "\\$", result, fixed = TRUE)
+    result <- gsub(">", "$>$", result, fixed = TRUE)
+    result <- gsub("<", "$<$", result, fixed = TRUE)
+    result <- gsub("|", "$|$", result, fixed = TRUE)
+    result <- gsub("{", "\\{", result, fixed = TRUE)
+    result <- gsub("}", "\\}", result, fixed = TRUE)
+    result <- gsub("%", "\\%", result, fixed = TRUE)
+    result <- gsub("&", "\\&", result, fixed = TRUE)
+    result <- gsub("_", "\\_", result, fixed = TRUE)
+    result <- gsub("#", "\\#", result, fixed = TRUE)
+    result <- gsub("^", "\\verb|^|", result, fixed = TRUE)
+    result <- gsub("~", "\\~{}", result, fixed = TRUE)
+    result <- gsub("SANITIZE.BACKSLASH", "$\\backslash$",
+                   result, fixed = TRUE)
+    return(result)
+}
 
 source("/home/don/projects/xtable/pkg/R/print.xtable.R")
 

### 2016-01-06 manuscript/manuscript.Rnw
add options to fix overfull hboxes
by default include FPKM in gene.expression()

@@
 \usepackage{xspace}
+% this package will stretch the font slightly if necessary.
+\usepackage{microtype}
 \usepackage{zref-xr,zref-user}
@@
 \widowpenalty = 10000
+% set a large emergency stretch to avoid issues with overfull boxes
+\setlength{\emergencystretch}{20em}
 % make the afill smaller
@@
 }
-gene.expression <- function(species.name,gene,complete=FALSE) {
+gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE) {
     fpkm <- combined.fpkm[species==species.name &
@@
                           mean_fpkm]
+    fpkm <- formatC(fpkm,digits=0,format="f")
+    if (units) {
+        fpkm <- paste0(fpkm,
+                       "\\ \\mathrm{FPKM}")
+    }
     if (complete) {
         if (any(grepl("^ENS.+G\\d+$",gene))) {
-            return(paste0(gene,
-                          formatC(fpkm,digits=0,format="f"),
-                          " \\ac{FPKM}"))
+            fpkm <- paste0("\\mathrm{",gene,"}",
+                           fpkm)
         } else {
-            return(paste0("\\textit{",gene,"}=",
-                          formatC(fpkm,digits=0,format="f"),
-                          " \\ac{FPKM}"
-                          ))
+            fpkm <- paste0("\\mathit{",gene,"}$=$",
+                           fpkm)
         }
-    } else {
+    }
+    if (units || complete) {
+        fpkm <- paste0("$",fpkm,"$")
+    }
     return(fpkm)
-    }
 }
@@
 greatest degree of expression %
-(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
+(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
 and the pseudogene \textit{PSG10P} showing the least %
-(\ac{FPKM}=\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:psg_evolution_tree}) . %
+(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:psg_evolution_tree}) . %
 These
@@
 comparatively much lower) in the human placenta
-(\ac{FPKM}=\Sexpr{gene.expression("homo sapiens","PRG2")};
+(\Sexpr{gene.expression("homo sapiens","PRG2")};
 Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
@@
 In addition, \textit{CGA} is expressed in the bonobo placenta %
-(\ac{FPKM}=\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
+(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
 lower levels than \human as well (\cref{tab:top50_human_placenta}).
@@
 \cite{Sampson.ea2012:PAGE4positivityisassociated}, \textit{PAGE4} %
-(\ac{FPKM}=\Sexpr{gene.expression("ateles fusciceps","PAGE4")}), %
+(\Sexpr{gene.expression("ateles fusciceps","PAGE4")}), %
 the pregnancy specific glycoprotein \textit{PSG8}
-(\ac{FPKM}=\Sexpr{gene.expression("ateles fusciceps","PSG8")}), %
+(\Sexpr{gene.expression("ateles fusciceps","PSG8")}), %
 a member of the chorionic somatotropins, \textit{GH1} %
-(\ac{FPKM}=\Sexpr{gene.expression("ateles fusciceps","GH1")}), %
+(\Sexpr{gene.expression("ateles fusciceps","GH1")}), %
 a few HLA genes (\textit{HLA-A}, \textit{HLA-F},
@@
 expression %
-(\ac{FPKM}=%
-\Sexpr{gene.expression("canis familiaris","TMSB10")}) is the thymosin 
-is 
-(\ac{FPKM}=%
-\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
+(\Sexpr{gene.expression("canis familiaris","TMSB10")}) is the thymosin 
+\textit{TMSB10} %
+(\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
 is the hormone relaxin (\textit{RLN2}). Other highly expressed genes

### 2016-01-06 manuscript/manuscript.Rnw
fix up dog paragraph

@@
 \cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
-(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
-In addition, an ortholog of SSLP1 (ENSCAFG00000010638) a gene
+(\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
+In addition, an ortholog of \textit{SSLP1} (ENSCAFG00000010638) a gene
 expressed in the placentas of \cow (%
-\Sexpr{gene.expression("bos tarus","SSLP1")}, %
+\Sexpr{gene.expression("bos taurus","SSLP1",complete=TRUE)}, %
 \cref{tab:top50_cow_placenta}) and \sheep
-(\Sexpr{gene.expression("ovis aries","SSLP1")},
+(\Sexpr{gene.expression("ovis aries","SOLD1",complete=TRUE)},
 \cref{tab:top50_sheep_placenta}) but is not present in primates
@@
 %
-(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
+(\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
 There is no placental expression of the estrogen synthase gene
-(\textit{CYP19A1}) in \dog (\cref{fig:estrogen_evolution_tree}), 
-however expression of \textit{HSD3B1} is relatively high
-(\ac{FPKM}=\Sexpr{gene.expression("canis familiaris","HSD3B1")}).
+(\textit{CYP19A1}) in \dog 
+(\Sexpr{gene.expression("canis familiaris","CYP19A1")}
+\cref{fig:estrogen_evolution_tree}), 
+and expression of \textit{HSD3B7} is also low
+(\Sexpr{gene.expression("canis familiaris","HSD3B7")}, 
+\cref{fig:progesterone_evolution_tree}).
 

### 2016-01-06 manuscript/manuscript.Rnw
discuss cow placenta

@@
 
-\subsection{Cow Placenta Transcriptome}
+\subsection{\cow Placenta Transcriptome}
 
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Bos taurus}}
+\caption{\tableexplanation{\cow}{tab:all_bos_taurus_placenta}}
 \label{tab:top50_cow_placenta}
@@
 
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the dog are shown in Table 5 with a list of all genes in Supplemental
-Table X. The gene with the highest expression (\ac{FPKM}=12488.8) was the
-miRNA gene bta-mir-2887-1. This gene occurs in several copies in the
-cow genome, and it was unclear as to how many or to which of 12
-bta-mir-2887 genes this mapped (Supplemental Table X). A previous
-study indicated that this miRNA gene was expressed in the bovine ovary
-(Hossain et al., 2009). Several members of the ruminant-specific
-pregnancy-associated glycoprotein family show especially high
-expression in the cow placenta (Table 4, Supplemental Table X)
-including PAG2 (\ac{FPKM}=7255.34), PAG10 (2076.42), BT.83780 (1793.72),
-PAG11 (962.19), PAG12 (469.751), BT.88769 (408.584), PAG1 (301.61),
-PAG4 (186.496), PAG6 (157.348), PAG8 (153.225), and PAG16 (103.573).
-In addition, 14 other PAGs were expressed with \ac{FPKM}<40 (Supplemental
-Table X). Other genes with high expression include genes involved with
-regulation of growth (IGF2, PHLDA2), imprinted genes (GNAS, DLK1,
-PHLDA2), and CSH1, a gene that is a member of a ruminant specific
+The top 50 expressed genes as measured by \ac{FPKM} in the term
+placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
+list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
+The gene with the highest expression was \textit{TMSB10} %
+(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
+\textit{TMSB10} was also highly expressed %
+(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
+\dog{}, members of the annexin complex were also highly expressed
+(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
+\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
+at least one of a set of four miRNA with 100\% identity
+(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
+\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
+%
+As these are novel miRNA, there is a possibility that the assignment
+of the short reads to these four miRNA is due to miss-annotation of
+the \cow genome.
+% 
+Several members of the ruminant-specific pregnancy-associated
+glycoprotein family show especially high expression in \cow placenta
+(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_evolution_tree}, and
+Supplemental \cref{tab:all_bos_taurus_placenta}) including
+%
+\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
+%
+28 members of this family have $FPKM \ge 1$, and this family of
+genes has greatly expanded in ruminants (\cow and \sheep{}) although
+the expression levels of this family are higher in \cow than in \sheep
+\cref{fig:pag2_evolution_tree}. Other genes with high expression
+include genes involved with regulation of growth (\textit{IGF2},
+\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
+and \textit{CSH1}, a gene that is a member of a ruminant specific
 expansion of placental prolactins and not directly related to the gene
-of the same name in hominids. Other ruminant-specific placental
-prolactins also show high expression in the cow placenta including
-PRP1 (632.766), PRP5 (396.832), PRP6 (345.296), PRP2 (195.771), in
-addition to 7 other placental prolactins with expression of \ac{FPKM}<51
-(Table 5; Supplemental Table X). SSLP1 (also called SOLD1), a gene
-expressed in the cotyledonary villi of the ruminiant placenta and
-involved in cell proliferation (Ushizawa et al., 2009, 2010), also
-shows evidence of high expression (\ac{FPKM}=1001.45). There is some
-placental expression of the estrogen synthase gene (CYP19A1) in cows
-(75.1346), and comparatively greater expression of HSD3B1 (\ac{FPKM}=
-157.622).
+of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
+ruminant-specific placental prolactins also show high expression in
+the cow placenta including %
+\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
+\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
+\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
+with a total of 14 with $\mathrm{FPKM} \ge 1$.
+%
+\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
+cotyledonary villi of the ruminiant placenta and involved in cell
+proliferation
+\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
+shows evidence of high expression %
+(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
+There is some placental expression of the estrogen synthase gene
+(\textit{CYP19A1}) in cows %
+(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
+comparatively greater expression of \textit{HSD3B1} %
+(\Sexpr{gene.expression("bos taurus","HSD3B1")}).
+
+\begin{figure}
+<<pag2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,min.fpkm=1)
+@ 
+\caption{Expression of genes related to \cow \textit{PAG2} with $\mathrm{FPKM} \ge 1$}
+\label{fig:pag2_evolution_tree}
+\end{figure}
 

### 2016-01-06 manuscript/manuscript.Rnw
fix spalax table references

@@
 \cref{tab:top50_n_galili_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_n_galili_placenta}.
+genes in Supplemental \cref{tab:all_nannospalax_galili_placenta}.
 
@@
 \cref{tab:top50_pig_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_pig_placenta}.
+genes in Supplemental \cref{tab:all_sus_scrofa_placenta}.
 

### 2016-01-06 manuscript/supplemental_material.Rnw
remove empty table
huge gene table now has more information
include annotation
include links
include human alignment where available

change link color to black

--- 2015-11-30 manuscript/supplemental_material.Rnw
@@
 
-<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","palcenta_go_results_mf")))>>=
+<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","palcenta_go_results_mf","hugo_nomenclature","gene_annotation")))>>=
 for (a in paste0("../data/",c("combined_fpkm","gene_trees",
@@
                               "placenta_go_results_bp",
-                              "placenta_go_results_mf"
+                              "placenta_go_results_mf",
+                              "hugo_nomenclature",
+                              "gene_annotation"
                               ))) {
@@
   \label{tab:srr_numbers}
-\end{table}
-
-\begin{table}
-  \begin{tabular}
-  \end{tabular}
 \end{table}
@@
 genes.expression.table <- function(species.name,min.fpkm=50) {
+    alignment.species <- species.name
+    if (alignment.species=="pan paniscus") {
+        alignment.species <- "pan troglodytes"
+    } else if (alignment.species=="ateles fusciceps") {
+        alignment.species <- "homo sapiens"
+    } else if (grepl("spalax",alignment.species)) {
+        alignment.species <- "mus musculus"
+    }
+
     temp <- 
-        combined.fpkm[species==species.name][order(-mean_fpkm)][,list(name_or_id,
+        combined.fpkm[species==species.name
+                      ][order(-mean_fpkm)
+                        ][,list(name_or_id,
+                                human_alignment_symbol,
                                 human_name,mean_fpkm)]
     temp <- temp[!duplicated(name_or_id)][mean_fpkm>=min.fpkm]
-    setnames(temp,c("Gene name","Human Name","FPKM"))
-    if (species.name=="homo sapiens") {
-        temp <- temp[,-2,with=FALSE]
+    temp[,used_alignment:=FALSE]
+    temp[is.na(human_name) & !is.na(human_alignment_symbol),
+         used_alignment:=TRUE]
+    temp[is.na(human_name),human_name:=human_alignment_symbol]
+    setkey(temp,"human_name")
+    temp <- 
+        hugo.nomenclature[temp][order(-mean_fpkm),
+                                list(name_or_id,used_alignment,`Approved Symbol`,`Approved Name`,mean_fpkm)]
+    temp[!is.na(name_or_id),
+         name_or_id:=paste0("\\href{http://www.ensembl.org/",
+                            capfirst(gsub(" ","_",alignment.species)),
+                            "/Gene/Summary?g=",name_or_id,"}{",
+                            xtable.sanitize.text(name_or_id),"}")]
+    temp[!is.na(`Approved Name`),
+         `Approved Name`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Name`),"}")]
+    temp[!is.na(`Approved Symbol`),
+         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Symbol`),"}")]
+    temp[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
+    temp[,used_alignment:=NULL]
+    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM"))
+    if (species.name=="homo sapiens" || species.name=="ateles fusciceps") {
+        temp[,`Human Symbol`:=NULL]
     }
@@
 for (this.species in names(species.names)) {
-    this.caption <- paste0("Expression of all genes with FPKM $\\ge 50$ in \\textit{",
-                           capfirst(this.species),"}")
+    this.caption <- paste0("\\textit{",
+                           capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
     cat(paste0("\\subsection{",this.caption,"}\n\n"))
@@
           caption.placement="top",
+          sanitize.text.function=genes.table.sanitize,
           size="\\tiny"

### 2016-01-06 manuscript/manuscript.Rnw
fix how gene.expression is called cgb.min.expresison

@@
                           mean_fpkm]
+    if (units || complete)
         fpkm <- formatC(fpkm,digits=0,format="f")
@@
     min(sapply(paste0("CGB",c("","1","2","5","7","8")),
-               function(x){gene.expression(species.name="homo sapiens",gene=x)}))
+               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
 cgb.sum.expression <- 
     sum(sapply(paste0("CGB",c("","1","2","5","7","8")),
-               function(x){gene.expression(species.name="homo sapiens",gene=x)}))
+               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
 @

### 2016-01-06 manuscript/manuscript.Rnw
in bos taurus, the gene is HSD3B not HSD3B1.
fix elephant table label

@@
 The top 50 expressed genes in each genome analyzed are shown in
-\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elphant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
+\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
 and all genes in Supplemental
@@
 (\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
-comparatively greater expression of \textit{HSD3B1} %
-(\Sexpr{gene.expression("bos taurus","HSD3B1")}).
+comparatively greater expression of \textit{HSD3B} %
+(\Sexpr{gene.expression("bos taurus","HSD3B")}).
 
@@
   \textit{Loxodonta africana}}
-\label{tab:top50_elphant_placenta}
+\label{tab:top50_elephant_placenta}
 \end{table}

### 2016-01-06 manuscript/manuscript.Rnw
fix TMSB4 error in dog
Fix elephant paragraph

@@
 (\Sexpr{gene.expression("bos taurus","TMSB10")}), and
-\textit{TMSB10} was also highly expressed %
+\textit{TMSB4} was also highly expressed %
 (\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
@@
 
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the elephant are shown in Table 6 with a list of all genes in
-Supplemental Table X. The gene with the highest expression
-(\ac{FPKM}=33439.5) was the natriuretic peptide precursor B (NPPB). The
-gene with the second highest expression (\ac{FPKM}=20233) was the microRNA
-miR-675, which is imbedded in the first exon of the maternally
-imprinted long non-coding RNA gene H19 (Keniry et al., 2012), a gene
-also highly expressed in the elephant placenta (\ac{FPKM}=7164.3). Five
-uncharacterized genes are expressed to a high degree in the elephant
-placenta (Table 6), three of which are a member of an
-elephant-specific expansion of 11 genes closely related to the
-thymus-specific serine protease gene PRSS16. The other 8 PRSS16-like
-genes are also expressed to a lesser degree (FPKM<284.014). Two other
-novel genes with high expression are related to the gene SSLP1 (Table
-6), which is also expressed in the cow and dog placentas (Tables 4 and
-5). Placental expression of the hormones relaxin (RLN1) and prolactin
-(PRL) are exceptionally high in the elephant when compared to other
-mammals (Tables 6 and 8). There is little to no placental expression
-of the estrogen synthase gene, CYP19A1, (FPKM<1) and the enzyme that
-produces progesterone HSD3B1 (\ac{FPKM}=0). Other genes with high
-expression in the elephant placenta include genes involved in
-oxidative stress (TXN), growth (IGF2), and PAEP, a gene usually
-expressed in the endometrium of the uterus in humans (Rutanen et al.,
-1987).
+The top 50 expressed genes as measured by \ac{FPKM} in the term
+placenta of \elephant are shown in \cref{tab:top50_elephant_placenta}
+with a list of all genes in
+\cref{tab:all_loxodonta_africana_placenta}. The gene with the
+highest expression was the natriuretic peptide precursor B
+(\Sexpr{gene.expression("loxodonta africana","NPPB",complete=TRUE)}).
+Two thymosins, %
+\Sexpr{gene.expression("loxodonta africana","TMSB10",complete=TRUE)} %
+and %
+\Sexpr{gene.expression("loxodonta africana","TMSB4Y",complete=TRUE)}
+were also also was highly expressed. The miRNA mir-675
+(\Sexpr{gene.expression("loxodonta
+  africana","ENSLAFG00000027406",complete=TRUE)}). which is imbedded
+in the first exon of the maternally imprinted long non-coding RNA gene
+\textit{H19} \cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene
+which is currently un-annotated in \elephant{}. %
+\DLA{Probably consider fixing the unannotation of H19; it seems pretty
+  obvious to me, though it would require fixing the GTF file for
+  elephant with the right 
+  \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan prediction} for this region}
+Five uncharacterized genes are expressed to a high degree in the
+\elephant placenta (\cref{tab:top50_elephant_placenta}), three of
+which %
+%
+(\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000028493",complete=TRUE)},
+\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000032011",complete=TRUE)},
+\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000031933",complete=TRUE)}) %
+%
+are members of an \elephant{}-specific expansion of 11 genes closely
+related to the thymus-specific serine protease gene \textit{PRSS16}
+\cref{fig:prss16_evolution_tree} The other 8 \textit{PRSS16}-like
+genes are also expressed to a lesser degree %
+(\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000026505",complete=TRUE)}). %
+%
+Two other novel genes with high expression are related to the gene
+\textit{SSLP1} %
+%
+(\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000032659",complete=TRUE)} %
+and %
+%
+\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000029551",complete=TRUE)},
+%
+\cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
+\dog placentas (\cref{tab:top50_dog_placenta,tab:top50_cow_placenta}).
+Placental expression of the hormones relaxin (\textit{RLN}, %
+\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000036090")})
+and prolactin (\textit{PRL}) are exceptionally high in the elephant
+when compared to other mammals
+(\cref{fig:rln2_evolution_tree,fig:prolactin_evolution_tree}). There
+is little to no placental expression of the estrogen synthase gene,
+\textit{CYP19A1} %
+(\Sexpr{gene.expression("loxodonta africana","CYP19A1")}) and the
+enzyme that produces progesterone HSD3B1 %
+(\Sexpr{gene.expression("loxodonta africana","HSD3B")}). Other genes
+with high expression in the elephant placenta include genes involved
+in oxidative stress %
+(\Sexpr{gene.expression("loxodonta africana","TXN",complete=TRUE)}),
+growth %
+(\Sexpr{gene.expression("loxodonta africana","IGF2",complete=TRUE)})
+and %
+\Sexpr{gene.expression("loxodonta africana","PAEP",complete=TRUE)}, %
+a gene usually expressed in the endometrium of the uterus in humans
+\cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.
+
+\begin{figure}
+<<prss16.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of \textit{PRSS16} and related genes}
+\label{fig:prss16_evolution_tree}
+\end{figure}
+
+<<rln2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="RLN2" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of \textit{RLN2} and related genes}
+\label{fig:rln2_evolution_tree}
+\end{figure}
 
@@
 these genes in elephants may indicate an elaboration of this system.
-
-\begin{figure}
-<<prss16.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{Expression of \textit{PRSS16} and related genes}
-\label{fig:prss16_evolution_tree}
-\end{figure}
 

### 2016-01-06 manuscript/manuscript.Rnw
include right spacing in gene.expression output

@@
                                gene_id==gene),
-                          mean_fpkm]
+                          mean_fpkm][1]
     if (units || complete)
@@
         if (any(grepl("^ENS.+G\\d+$",gene))) {
-            fpkm <- paste0("\\mathrm{",gene,"}",
+            fpkm <- paste0("\\mathrm{",gene,"} = ",
                            fpkm)
         } else {
-            fpkm <- paste0("\\mathit{",gene,"}$=$",
+            fpkm <- paste0("\\mathit{",gene,"} = ",
                            fpkm)

### 2016-01-06 manuscript/manuscript.Rnw
rearrange results sections to discuss results under their major type

@@
 
-\subsection{\dog Placenta Transcriptome}
+\subsection{Rodentia}
+
+\subsubsection{Mouse Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
+top.mouse.genes <- top.genes.table(species.name="mus musculus")
+print(xtable(top.mouse.genes),
+      booktabs=TRUE,
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.mouse.genes,"Gene symbol")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
+  \textit{Mus musculus}}
+\label{tab:top50_mouse_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \mouse are shown in
+\cref{tab:top50_mouse_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_mus_musculus_placenta}.
+
+% No idea what mir-6236 is doing
+
+% AC099934.1 probably overlaps with Rn7sl (part of SRP?)
+
+% AC157822.1 overlaps with Rn7s2
+
+% AC122546.1 = no idea
+
+% Gm3788 = psuedogene (Thymosin beta 10)
+
+\DLA{Why are there so many psuedogenes in the mouse? Maybe because of
+  the short read length?}
+
+\subsubsection{\textit{Nannospalax galili} (Blind Mole Rat) Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
+top.galili.genes <- top.genes.table(species.name="nannospalax galili")
+print(xtable(top.galili.genes),
+      booktabs=TRUE,
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.galili.genes,"Gene symbol")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
+  \textit{Nannospalax galili}}
+\label{tab:top50_n_galili_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \textit{Nannospalax galili} are shown in
+\cref{tab:top50_n_galili_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_nannospalax_galili_placenta}.
+
+\subsubsection{\textit{Spalax carmeli} (Blind Mole Rat) Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
+top.carmeli.genes <- top.genes.table(species.name="spalax carmeli")
+print(xtable(top.carmeli.genes),
+      booktabs=TRUE,
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.carmeli.genes,"Gene symbol")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
+  \textit{Spalax carmeli}}
+\label{tab:top50_s_carmeli_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \textit{Spalax carmeli} are shown in
+\cref{tab:top50_s_carmeli_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_spalax_carmeli_placenta}.
+
+\subsection{Laurasiatheria}
+
+\subsubsection{\dog Placenta Transcriptome}
 
@@
 
-\subsection{\cow Placenta Transcriptome}
+\subsubsection{\cow Placenta Transcriptome}
 
@@
 
-\subsection{\elephant Placenta Transcriptome}
+\subsubsection{\pig Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
+top.pig.genes <- top.genes.table(species.name="sus scrofa")
+print(xtable(top.pig.genes),
+      booktabs=TRUE,
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.pig.genes,"Gene symbol")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
+mitochondrial, or whose human ortholog is a housekeeping gene in
+  \pig}
+\label{tab:top50_pig_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \textit{Sus scrofa} are shown in
+\cref{tab:top50_pig_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_sus_scrofa_placenta}.
+
+\subsubsection{\horse Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
+top.horse.genes <- top.genes.table(species.name="equus caballus")
+print(xtable(top.horse.genes),
+      booktabs=TRUE,
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.horse.genes,"Gene symbol")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes in \horse which are
+  not ribosomal, mitochondrial, or whose human ortholog is a
+  housekeeping gene}
+\label{tab:top50_horse_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \textit{Equus Caballus} are shown in
+\cref{tab:top50_horse_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_equus_caballus_placenta}.
+
+\subsubsection{\sheep Placenta Transcriptome}
+
+\begin{table}[h]
+  \tiny
+<<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
+top.sheep.genes <- top.genes.table(species.name="ovis aries")
+print(xtable(top.sheep.genes),
+      booktabs=TRUE,
+      floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+setkey(top.sheep.genes,"Gene symbol")
+@ 
+\caption{Expression in \ac{FPKM} of top 50 genes in \sheep which are
+  not ribosomal, mitochondrial, or whose human ortholog is a
+  housekeeping gene}
+\label{tab:top50_sheep_placenta}
+\end{table}
+
+The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
+in the term placenta of \sheep are shown in
+\cref{tab:top50_sheep_placenta} with a list of all highly expressed
+genes in Supplemental \cref{tab:all_ovis_aries_placenta}.
+
+\subsection{Atlantogenata}
+\subsubsection{\elephant Placenta Transcriptome}
 
@@
 
-\subsection{\armadillo Placenta Transcriptome}
+\subsubsection{\armadillo Placenta Transcriptome}
 
@@
 
-\subsection{Opossum Placenta Transcriptome}
+\subsection{Metatherians}
+
+\subsubsection{\opossum Placenta Transcriptome}
 
@@
 expression of relaxin RLN1 (\ac{FPKM}= 41.3949).
-
-\subsection{Mouse Placenta Transcriptome}
-
-\begin{table}[h]
-  \tiny
-<<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
-top.mouse.genes <- top.genes.table(species.name="mus musculus")
-print(xtable(top.mouse.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.mouse.genes,"Gene symbol")
-@ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Mus musculus}}
-\label{tab:top50_mouse_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \mouse are shown in
-\cref{tab:top50_mouse_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_mus_musculus_placenta}.
-
-% No idea what mir-6236 is doing
-
-% AC099934.1 probably overlaps with Rn7sl (part of SRP?)
-
-% AC157822.1 overlaps with Rn7s2
-
-% AC122546.1 = no idea
-
-% Gm3788 = psuedogene (Thymosin beta 10)
-
-\DLA{Why are there so many psuedogenes in the mouse? Maybe because of
-  the short read length?}
-
-\subsection{\textit{Nannospalax galili} (Blind Mole Rat) Placenta Transcriptome}
-
-\begin{table}[h]
-  \tiny
-<<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
-top.galili.genes <- top.genes.table(species.name="nannospalax galili")
-print(xtable(top.galili.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.galili.genes,"Gene symbol")
-@ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Nannospalax galili}}
-\label{tab:top50_n_galili_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Nannospalax galili} are shown in
-\cref{tab:top50_n_galili_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_nannospalax_galili_placenta}.
-
-\subsection{\textit{Spalax carmeli} (Blind Mole Rat) Placenta Transcriptome}
-
-\begin{table}[h]
-  \tiny
-<<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
-top.carmeli.genes <- top.genes.table(species.name="spalax carmeli")
-print(xtable(top.carmeli.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.carmeli.genes,"Gene symbol")
-@ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Spalax carmeli}}
-\label{tab:top50_s_carmeli_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Spalax carmeli} are shown in
-\cref{tab:top50_s_carmeli_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_spalax_carmeli_placenta}.
-
-\subsection{Wild Boar Placenta Transcriptome}
-
-\begin{table}[h]
-  \tiny
-<<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
-top.pig.genes <- top.genes.table(species.name="sus scrofa")
-print(xtable(top.pig.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.pig.genes,"Gene symbol")
-@ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Sus scrofa}}
-\label{tab:top50_pig_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Sus scrofa} are shown in
-\cref{tab:top50_pig_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_sus_scrofa_placenta}.
-
-\subsection{\horse Placenta Transcriptome}
-
-\begin{table}[h]
-  \tiny
-<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
-top.horse.genes <- top.genes.table(species.name="equus caballus")
-print(xtable(top.horse.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.horse.genes,"Gene symbol")
-@ 
-\caption{Expression in \ac{FPKM} of top 50 genes in \horse which are
-  not ribosomal, mitochondrial, or whose human ortholog is a
-  housekeeping gene}
-\label{tab:top50_horse_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Equus Caballus} are shown in
-\cref{tab:top50_horse_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_equus_caballus_placenta}.
-
-\subsection{\sheep Placenta Transcriptome}
-
-\begin{table}[h]
-  \tiny
-<<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
-top.sheep.genes <- top.genes.table(species.name="ovis aries")
-print(xtable(top.sheep.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.sheep.genes,"Gene symbol")
-@ 
-\caption{Expression in \ac{FPKM} of top 50 genes in \sheep which are
-  not ribosomal, mitochondrial, or whose human ortholog is a
-  housekeeping gene}
-\label{tab:top50_sheep_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \sheep are shown in
-\cref{tab:top50_sheep_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_ovis_aries_placenta}.
 
@@
 
-\subsection{\ac{GSE} Analyses of Individual Transcriptomes}
+\subsubsection{\ac{GSE} Analyses of Individual Transcriptomes}
 

### 2016-01-07 manuscript/manuscript.Rnw
include SEM in FPKM table where available

@@
                                                  human_alignment_symbol,
-                                                 human_name,mean_fpkm)]
+                                                 human_name,mean_fpkm,sd_fpkm,n_fpkm)]
     ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
@@
         hugo.nomenclature[temp][order(-mean_fpkm),
-                                list(name_or_id,used_alignment,`Approved Symbol`,`Approved Name`,mean_fpkm)]
+                                list(name_or_id,used_alignment,`Approved Symbol`,`Approved Name`,
+                                     mean_fpkm,sd_fpkm,n_fpkm)]
     temp[!is.na(name_or_id),
@@
     temp[,used_alignment:=NULL]
-    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM"))
+    temp[,sem_fpkm:=as.numeric(NA)]
+    temp[n_fpkm>1,sem_fpkm:=sd_fpkm/sqrt(n_fpkm)]
+    temp[,sd_fpkm:=NULL]
+    temp[,n_fpkm:=NULL]
+    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM","FPKM SEM"))
+    if (temp[,all(is.na(`FPKM SEM`))]) {
+        temp[,`FPKM SEM`:=NULL]
+    }
     if (species.name=="homo sapiens" || species.name=="ateles fusciceps") {

### 2016-01-07 data/fastq_fpkm/combine_fpkms.R

@@ -104,6 +104,7 @@ gene.fpkms <- rbindlist(gene.fpkms,fill=TRUE)
 
 gene.fpkms[,mean_fpkm := mean(FPKM),tracking_id]
 gene.fpkms[,sd_fpkm := sd(FPKM),tracking_id]
+gene.fpkms[,n_fpkm := length(FPKM),tracking_id]
 # gene.fpkms <- gene.fpkms[!duplicated(tracking_id),]
 
 isoform.fpkms <- list();
@@ -121,6 +122,7 @@ isoform.fpkms <- rbindlist(c(isoform.fpkms),fill=TRUE)
 
 isoform.fpkms[,mean_fpkm := mean(FPKM,na.rm=TRUE),tracking_id]
 isoform.fpkms[,sd_fpkm := sd(FPKM,na.rm=TRUE),tracking_id]
+isoform.fpkms[,n_fpkm := length(na.omit(FPKM)),tracking_id]
 isoform.fpkms <- isoform.fpkms[!duplicated(tracking_id),]
 
 

### 2016-01-07 manuscript/manuscript.Rnw
properly escape `FDR` in the primate specific discussion

@@
 (\cref{fig:fstl1_evolution_tree},
-$\mathrm{FDR}=\Sexpr{pri.all.lin.spec["FSTL1","FDR"]}$), is implicated
+$\mathrm{FDR}=\Sexpr{pri.all.lin.spec["FSTL1",`FDR`]}$), is implicated
 in BMP signaling, and has previously been implicated in inflammation

### 2016-01-07 data/placenta_lineage_specific.R
do not include housekeeping genes in the lineage specific lists

@@ -20,6 +20,10 @@ combined.fpkm.wide <-
 placenta.lineage.specific <-
     combined.fpkm.wide[apply(combined.fpkm.wide[,-1,with=FALSE],1,max) >= 100 & !is.na(human_name),]
 
+### ignore genes which are housekeeping genes
+placenta.lineage.specific <-
+    placenta.lineage.specific[!(human_name %in% housekeeping.genes.superset[,gene_short_name]),]
+
 species <- c("ateles fusciceps", "homo sapiens", "pan paniscus", "mus musculus", 
              "nannospalax galili", "spalax carmeli", "bos taurus", "ovis aries", 
              "sus scrofa", "equus caballus", "canis familiaris", "loxodonta africana", 

### 2016-01-07 manuscript/manuscript.Rnw
discuss the rodent specific lineage

@@
 
-\subsection{Rodentia}
-
-\subsubsection{Mouse Placenta Transcriptome}
+\subsection{Rodentia Placenta Transcriptomes}
+
+\begin{table}
+  \tiny
+<<rodentia.lineage.specific,echo=FALSE,results="asis">>=
+rod.all.lin.spec <- placenta.lineage.specific.table("rod.all")
+setkey(rod.all.lin.spec,"Human Symbol")
+@ 
+\caption{Genes whose expression changed significantly in the Rodentia
+  order as compared to all other analyzed species.}
+\label{tab:rodentia_lineage_specific}
+\end{table}
+
+We examined the expression of three rodentia, \mouse, \carmeli, and
+\galili{}. We lacked sufficient power to detect any significant
+changes on the Rodentia lineage in comparison to all other analyzed
+species (\cref{tab:rodentia_lineage_specific}).
+
+\subsubsection{\mouse Placenta Transcriptome}
 

### 2016-01-07 common_r_code/plot_tree_matrix.R
allow for specifying a subtree

@@ -1,6 +1,7 @@
 library("magrittr")
 library("tidyr")
-plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offset.ratio=0.5,min.fpkm=NA) {
+library("geiger")
+plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offset.ratio=0.5,min.fpkm=NA,subtree=NULL) {
     gene.tree$tip.label <- gsub("^(ENS.+)\\1$","\\1",gene.tree$tip.label)
     gene.tree.table[,species:=capfirst(gsub("_"," ",species))]
     gene.tree.table[,short.species:=gsub("^(.)[^ ]* +([^ ]*)","\\1. \\2",species)]
@@ -20,8 +21,17 @@ plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offs
         drop.tip(gene.tree,
                  gene.tree$tip.label[!(gene.tree$tip.label %in%
                                        gene.tree.expression.subset[,gene_id])])
+    if (!is.null(subtree)) {
+        gene.tree.mrca <- mrca(gene.tree.subset)
+        top.node <- gene.tree.mrca[subtree[1],subtree[2]]
+        gene.tree.subset <-
+            drop.tip(gene.tree.subset,
+                     gene.tree.subset$tip.label[!(gene.tree.subset$tip.label %in%
+                                                  tips(gene.tree.subset,
+                                                       top.node))])
+    }
     gene.tree.expression <-
-        gene.tree.expression[gene_id %in% gene.tree$tip.label,]
+        gene.tree.expression[gene_id %in% gene.tree.subset$tip.label,]
     gene.tree.expression[,mean_fpkm_log:=log10(mean_fpkm+1)]
     gene.tree.expression.matrix <-
         dcast(gene.tree.expression,

### 2016-01-07 manuscript/manuscript.Rnw
write mouse transriptome results paragraph

@@
 gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE) {
+    species.name <- gsub("_"," ",species.name)
     fpkm <- combined.fpkm[species==species.name &
@@
 in the term placenta of \mouse are shown in
-\cref{tab:top50_mouse_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_mus_musculus_placenta}.
-
-% No idea what mir-6236 is doing
-
-% AC099934.1 probably overlaps with Rn7sl (part of SRP?)
-
-% AC157822.1 overlaps with Rn7s2
-
-% AC122546.1 = no idea
-
-% Gm3788 = psuedogene (Thymosin beta 10)
-
-\DLA{Why are there so many psuedogenes in the mouse? Maybe because of
-  the short read length?}
+\cref{tab:top50_mouse_placenta} with a list of all genes with
+$\mathrm{FPKM} \ge 50$ in Supplemental
+\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
+\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
+target \textit{in silico} is the orphan G-coupled receptor
+\textit{Mrgprg},
+\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
+\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
+of which are maternally imprinted (the later of which is a homolog of
+human \textit{H19\_2}),
+\Sexpr{gene.expression("mus_musculus","IGF2",complete=TRUE)}, numerous
+cathepsins and a rodentia specific expansion of prolactin related
+genes. There are also two miRNA which overlap with the of the 7S
+signal recognition particle
+(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
+and
+\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
+and numerous pseudogenes
+(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
+\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
+\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
+\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
+appear to be highly expressed, however, the short read length (34--44)
+in many samples is likely preventing unambiguous assignment of all
+reads to the more lengthy complete transcript, so these shorter
+transcripts should be treated with suspicion, and we will ignore them
+in further discussion. The prolactin-related genes
+(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
+\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
+\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
+\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
+very highly expressed in \mouse in, with %
+\Sexpr{combined.fpkm[species=="mus musculus" &
+  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
+prolactin related genes having $\mathrm{FPKM} \ge 1$
+(\cref{fig:prolactin_evolution_tree}). The cathespins have also
+expanded in rodents, with six showing high expression
+(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
+\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
+\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
+\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
+\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
+\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
+\cref{tab:top50_mouse_placenta,fig:cathespin_evolution_tree}).
+Finally, \textit{Peg10}, an imprinted gene required for proper
+formation of the placenta in mice is also highly expressed
+(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
+reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.
+
+\begin{figure}
+<<cathespin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
+                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"))
+@ 
+\caption{Expression of cathespins related to \textit{CTSL} with $\mathrm{FPKM} \ge 1$}
+\label{fig:cathespin_evolution_tree}
+\end{figure}
 

### 2016-01-07 manuscript/manuscript.Rnw
discuss galili

@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Mus musculus}}
+\caption{\tableexplanation{\mouse}{tab:all_mus_musculus_placenta}}
 \label{tab:top50_mouse_placenta}
@@
 
-\subsubsection{\textit{Nannospalax galili} (Blind Mole Rat) Placenta Transcriptome}
+\subsubsection{\galili Placenta Transcriptome}
 
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Nannospalax galili}}
+\caption{\tableexplanation{\galili}{tab:all_nannospalax_galili_placenta}}
 \label{tab:top50_n_galili_placenta}
@@
 The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Nannospalax galili} are shown in
+in the term placenta of \galili are shown in
 \cref{tab:top50_n_galili_placenta} with a list of all highly expressed
 genes in Supplemental \cref{tab:all_nannospalax_galili_placenta}.
-
-\subsubsection{\textit{Spalax carmeli} (Blind Mole Rat) Placenta Transcriptome}
+Components of the annexin complex are highly expressed
+(\Sexpr{gene.expression("nannospalax_galili","S100a6",complete=TRUE)},
+\Sexpr{gene.expression("nannospalax_galili","S100a11",complete=TRUE)},
+\Sexpr{gene.expression("nannospalax_galili","S100a10",complete=TRUE)},
+and
+\Sexpr{gene.expression("nannospalax_galili","Anxa2",complete=TRUE)}),
+as are members of the thymosin family
+(\Sexpr{gene.expression("nannospalax_galili","Tmsb4x",complete=TRUE)}
+and
+\Sexpr{gene.expression("nannospalax_galili","Tmsb10",complete=TRUE)}),
+various cytoskeleton related proteins
+(\Sexpr{gene.expression("nannospalax_galili","Actc1",complete=TRUE)},
+\Sexpr{gene.expression("nannospalax_galili","Actg2",complete=TRUE)},
+\Sexpr{gene.expression("nannospalax_galili","Acta2",complete=TRUE)},
+\Sexpr{gene.expression("nannospalax_galili","Myl9",complete=TRUE)},
+\Sexpr{gene.expression("nannospalax_galili","Vim",complete=TRUE)}), and
+transmembrane protein of currently unknown function, 
+\Sexpr{gene.expression("nannospalax_galili","Gm10134",complete=TRUE)}.
+
+\subsubsection{\carmeli Placenta Transcriptome}
 

### 2016-01-07 manuscript/manuscript.Rnw
discuss carmeli results

@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Spalax carmeli}}
+\caption{\tableexplanation{\carmeli}{tab:all_spalax_carmeli_placenta}}
 \label{tab:top50_s_carmeli_placenta}
@@
 The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Spalax carmeli} are shown in
-\cref{tab:top50_s_carmeli_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_spalax_carmeli_placenta}.
+in the term placenta of \carmeli are shown in
+\cref{tab:top50_s_carmeli_placenta} with a list of all highly
+expressed genes in Supplemental
+\cref{tab:all_spalax_carmeli_placenta}. Numerous hemoglobin subunits
+ware highly expressed
+(\Sexpr{gene.expression("spalax_carmeli","Hbb-bh1",complete=TRUE)},
+\Sexpr{gene.expression("spalax_carmeli","Hbb-a1",complete=TRUE)}, and
+\Sexpr{gene.expression("spalax_carmeli","Hbb-bs",complete=TRUE)}), as
+well as the relaxin
+\Sexpr{gene.expression("spalax_carmeli","Rln1",complete=TRUE)},
+multiple cathespins
+(\Sexpr{gene.expression("spalax_carmeli","Ctsm",complete=TRUE)},
+\Sexpr{gene.expression("spalax_carmeli","BC051665",complete=TRUE)},
+and \Sexpr{gene.expression("spalax_carmeli","Ctsl",complete=TRUE)},
+\cref{fig:cathespin_evolution_tree}), and many prolactin-related genes
+(\Sexpr{gene.expression("spalax_carmeli","Prl4a1",complete=TRUE)},
+\Sexpr{gene.expression("spalax_carmeli","Prl6a1",complete=TRUE)},
+\Sexpr{gene.expression("spalax_carmeli","Prl3b1",complete=TRUE)}, and
+\Sexpr{gene.expression("spalax_carmeli","Prl7c1",complete=TRUE)},
+\cref{fig:prolactin_evolution_tree}) which shows a similar pattern of
+expansion seen in \mouse.
 

### 2016-01-07 manuscript/manuscript.Rnw
discuss pig and laurasiatheria specific results

@@
 
+\begin{table}
+  \tiny
+<<lau.lineage.specific,echo=FALSE,results="asis">>=
+lau.all.lin.spec <- placenta.lineage.specific.table("lau.all")
+setkey(lau.all.lin.spec,"Human Symbol")
+@ 
+\caption{Genes whose expression changed in the Laurasiatheria
+  superorder as compared to all other analyzed species.}
+\label{tab:laurasiatheria_lineage_specific}
+\end{table}
+
+\begin{table}
+  \tiny
+<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
+eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau")
+setkey(eua.lau.lin.spec,"Human Symbol")
+@ 
+\caption{Genes whose expression changed between the
+  Euasiatherian and Laurasiatherian superorder.}
+\label{tab:laurasiatheria_eurasiatherian_lineage_specific}
+\end{table}
+
+We examined the expression of five Laurasiatherians: \cow, \sheep,
+\pig, \horse, and \dog. We were unable to detect any significant
+differences after correcting for multiple testing between
+Laurasiatherians and all other analyzed species
+(\cref{tab:laurasiatheria_lineage_specific}) or between
+Laurasiatherians and Euasiatherians
+(\cref{tab:laurasiatheria_eurasiatherian_lineage_specific}).
+
 \subsubsection{\dog Placenta Transcriptome}
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \pig}
+\caption{\tableexplanation{\pig}{tab:all_sus_scrofa_placenta}}
 \label{tab:top50_pig_placenta}
@@
 \cref{tab:top50_pig_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_sus_scrofa_placenta}.
+genes in Supplemental \cref{tab:all_sus_scrofa_placenta}. The top two
+expressed transcripts,
+\Sexpr{gene.expression("sus_scrofa","ENSSSCG00000025162",complete=TRUE)}
+and
+\Sexpr{gene.expression("sus_scrofa","ENSSSCG00000021356",complete=TRUE)},
+are miRNA which correspond to the \cow miRNA
+\href{http://www.mirbase.org/cgi-bin/mirna_entry.pl?acc=MI0013080}{bta-mir-2904-1}
+and
+\href{http://www.mirbase.org/cgi-bin/mirna_entry.pl?acc=MIMAT0013845}{bta-mir-2887-1},
+respectively, both of which are also expressed in \cow ovary
+\cite{Hossain.ea2009:Identificationandcharacterizationof}, although
+neither of which have an obvious target gene on the basis of
+\textit{in silico} analysis
+\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}. Other highly
+expressed genes include members of the insulin growth factor family
+(\Sexpr{gene.expression("sus_scrofa","IGF2",complete=TRUE)} and
+\Sexpr{gene.expression("sus_scrofa","IGFBP3",complete=TRUE)}), members
+of the annexin complex
+(\Sexpr{gene.expression("sus_scrofa","S100A1",complete=TRUE)},
+\Sexpr{gene.expression("sus_scrofa","ANXA2",complete=TRUE)},
+\Sexpr{gene.expression("sus_scrofa","S100A12",complete=TRUE)},
+\Sexpr{gene.expression("sus_scrofa","S100A10",complete=TRUE)}, and
+\Sexpr{gene.expression("sus_scrofa","S100A1",complete=TRUE)}), the
+galectin \Sexpr{gene.expression("sus_scrofa","LGALS3",complete=TRUE)}
+and multiple cytoskeleton and extracellular matrix proteins
+(\Sexpr{gene.expression("sus_scrofa","KRT19",complete=TRUE)},
+\Sexpr{gene.expression("sus_scrofa","COL1A1",complete=TRUE)},
+\Sexpr{gene.expression("sus_scrofa","BETA-ACTIN",complete=TRUE)},
+\Sexpr{gene.expression("sus_scrofa","KRT9",complete=TRUE)},
+\Sexpr{gene.expression("sus_scrofa","CLDN4",complete=TRUE)}, and
+\Sexpr{gene.expression("sus_scrofa","KRT9",complete=TRUE)}).
 

### 2016-01-07 manuscript/manuscript.Rnw
use amsmath for the gene names

@@
 \usepackage{pdflscape}
+\usepackage{amsmath}
 \usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
@@
         fpkm <- paste0(fpkm,
-                       "\\ \\mathrm{FPKM}")
+                       "\\ \\text{FPKM}")
     }
@@
         if (any(grepl("^ENS.+G\\d+$",gene))) {
-            fpkm <- paste0("\\mathrm{",gene,"} = ",
+            fpkm <- paste0("\\text{",gene,"} = ",
                            fpkm)
         } else {
-            fpkm <- paste0("\\mathit{",gene,"} = ",
+            fpkm <- paste0("\\text{\\textit{",gene,"}} = ",
                            fpkm)

### 2016-01-07 manuscript/manuscript.Rnw
remove some more ribosomal proteins from the gene list

@@
                       ! (gene_biotype %in% c("Mt_tRNA","Mt_rRNA","rRNA")) &
-                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?$)",human_name) &
-                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?$)",human_alignment_symbol) &
-                      !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[l]?(?:-.+)?$|rRNA$)",
+                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_name) &
+                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_alignment_symbol) &
+                      !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[lL]?(?:-.+)?$|rRNA$)",
                              gene_short_name)

### 2016-01-07 manuscript/manuscript.Rnw
discuss horse placenta transcriptome

@@
 target \textit{in silico} is the orphan G-coupled receptor
-\textit{Mrgprg},
+\textit{Mrgprg}
+\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
 \Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
@@
 \cref{tab:top50_horse_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_equus_caballus_placenta}.
+genes in Supplemental \cref{tab:all_equus_caballus_placenta}. The gene
+with the highest expression was the relaxin,
+\Sexpr{gene.expression("equus_caballus","RLN",complete=TRUE)},
+followed by two miRNA
+(\Sexpr{gene.expression("equus_caballus","ENSECAG00000025328",complete=TRUE)}
+and
+\Sexpr{gene.expression("equus_caballus","ENSECAG00000027056",complete=TRUE)})
+with no known \textit{in silico} targets
+\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}, the thymosin
+\Sexpr{gene.expression("equus_caballus","TMSB4",complete=TRUE)},
+members of the insulin growth factor family
+(\Sexpr{gene.expression("equus_caballus","IGF2",complete=TRUE)} and
+\Sexpr{gene.expression("equus_caballus","IGFBP3",complete=TRUE)}) and
+multiple members of the annexin complex
+(\Sexpr{gene.expression("equus_caballus","S100A6",complete=TRUE)},
+\Sexpr{gene.expression("equus_caballus","S100A1",complete=TRUE)}, and
+\Sexpr{gene.expression("equus_caballus","ANXA2",complete=TRUE)}),
+similar to the pattern of expression seen in \pig and \dog.
 

### 2016-01-07 manuscript/manuscript.Rnw
fix horse table explanation
add missing \begin{figure}

@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes in \horse which are
-  not ribosomal, mitochondrial, or whose human ortholog is a
-  housekeeping gene}
+\caption{\tableexplanation{\horse}{tab:all_equus_caballus_placenta}}
 \label{tab:top50_horse_placenta}
@@
 
+\begin{figure}
 <<rln2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

### 2016-01-07 manuscript/manuscript.Rnw
add HBE1 figure and discuss opossum

@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \armadillo}
+\caption{\tableexplanation{\armadillo}{tab:all_dasypus_novemcinctus_placenta}}
 \label{tab:top50_armadillo_placenta}
@@
 
-The top 50 expressed genes in the term placenta of \armadillo are in
-\cref{tab:top50_armadillo_placenta}.
+The top 50 expressed genes in the term placenta of \armadillo are
+shown in \cref{tab:top50_armadillo_placenta} with a list of all genes
+with $\mathrm{FPKM}\ge 50$ in Supplemental
+\cref{tab:all_dasypus_novemcinctus_placenta}.
 
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Monodelphis domestica}}
+\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica}}
 \label{tab:top50_opossum_placenta}
@@
 
-The top 50 expressed genes as measured by \ac{FPKM} in the term placenta of
-the opossum are shown in Table 7 with a list of all genes in
-Supplemental Table X. The gene with the highest expression
-(\ac{FPKM}=41740.6) was hemoglobin, epsilon (HBE), which is expressed at
-the embryonic stage of development in marsupials, as in other mammals
-(Holland and Gooley, 1997). HBE subunits link up with protein products
-of two hemoglobin, zeta (HBZ) genes, which are also expressed to a
-high degree in the opossum placenta (\ac{FPKM}=2452.19 and \ac{FPKM}=1139.25).
-In addition, there is high expression (\ac{FPKM}=5827.37) of the gene
-hemoglobin, omega (HBO), a hemoglobin found only in marsupials and
-monotremes (Opazo et al., 2008). The hemoglobin gene HBA is also
-expressed in the placenta (\ac{FPKM}=456.978). The gene with the second
-highest expression (\ac{FPKM}=22854.8) is superoxide dismutase 3,
-extracellular (SOD3), which functions as an antioxidant. Four cystatin
-genes (CSTB, CST3, two CSTA-like genes) are expressed to a high
-degree, along with two cathepsin-L related genes (Table 7), which are
-known to interact with cystatins (Turk et al., 2001). The
-uncharacterized gene XM\_001364916.1 also shows high expression in the
-opossum placenta (\ac{FPKM}=13572.8). The gene WFDC2, a member of the whey
-acidic protein (WAP) family is also expressed to a high degree in the
-opossum placenta. Other genes with high expression include genes
-involved in retinol binding (RBP4), growth (IGFBP3), lipid metabolism
-(APOA1, APOA2), and immunosupression (LGALS1). In addition, there was
-little production of CYP19A1 and HSD3B1 (FPKM<2), and moderate
-expression of relaxin RLN1 (\ac{FPKM}= 41.3949).
+The top 50 expressed genes as measured by \ac{FPKM} in the term
+placenta of the \opossum are shown in
+\cref{tab:top50_opossum_placenta} with a list of all genes in
+Supplemental \cref{tab:all_monodelphis_domestica}. The gene with the
+second highest expression is superoxide dismutase 3, extracellular
+(\Sexpr{gene.expression("monodelphis_domestica","SOD3",complete=TRUE)}),
+which functions as an antioxidant. The gene with the second highest
+expression is the hemoglobin subunit epsilon
+(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000011325",complete=TRUE)}),
+which is expressed at the embryonic stage of development in
+marsupials, as in other mammals
+\cite{Holland.ea1997:Characterizationofembryonicglobin}. \textit{HBE}
+subunits link up with protein products of two hemoglobin, zeta
+(\textit{HBZ}) genes, which are also expressed to a high degree in the
+opossum placenta
+(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016635",complete=TRUE)}
+and
+\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016631",complete=TRUE)}).
+In addition, there is high expression of the gene hemoglobin, omega
+(\textit{HBO},
+\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016637",complete=TRUE)}),
+% this is LOC_1000016060 which I figured out by tblastn'ing AAK11486.1
+% against Monodelphis domestica
+a hemoglobin found only in marsupials and monotremes
+\cite{Opazo.ea2008:Genomicevidenceforindependent}. The hemoglobin gene
+\textit{HBA} is also expressed in the placenta
+(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016633",complete=TRUE)}).
+Two cystatin genes
+(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018295",complete=TRUE)}
+and
+\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018296",complete=TRUE)})
+are expressed to a high degree, along with one \textit{CTSL} related
+gene
+(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000002844",complete=TRUE)},
+\cref{fig:cathespin_evolution_tree}), which are known to interact with
+cystatins \cite{Turk.ea2001:Lysosomalcysteineproteasesfacts}. The
+uncharacterized gene \textit{LOC10001195}
+(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
+also shows high expression in the opossum placenta. The gene \textit{WFDC2}, a
+member of the whey acidic protein (WAP) family is also expressed to a
+high degree in the opossum placenta
+(\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).
+Other genes with high expression include members of the annexin
+complex
+(\Sexpr{gene.expression("monodelphis_domestica","S100A6",complete=TRUE)},
+\Sexpr{gene.expression("monodelphis_domestica","ANXA2",complete=TRUE)}).
+In addition, there was little production of CYP19A1 and HSD3B1
+(\Sexpr{gene.expression("monodelphis_domestica","CYP19A1",complete=TRUE)}
+and
+\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000005385",complete=TRUE)},
+respectively).
+
+\begin{figure}
+<<hbe1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="HBE1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of \textit{HBE1} and related genes}
+\label{fig:hbe1_evolution_tree}
+\end{figure}
 

### 2016-01-08 manuscript/manuscript.Rnw
use underscore in elephant descriptions

@@
 with a list of all genes in
-\cref{tab:all_loxodonta_africana_placenta}. The gene with the
-highest expression was the natriuretic peptide precursor B
-(\Sexpr{gene.expression("loxodonta africana","NPPB",complete=TRUE)}).
-Two thymosins, %
-\Sexpr{gene.expression("loxodonta africana","TMSB10",complete=TRUE)} %
-and %
-\Sexpr{gene.expression("loxodonta africana","TMSB4Y",complete=TRUE)}
+\cref{tab:all_loxodonta_africana_placenta}. The gene with the highest
+expression was the natriuretic peptide precursor B
+(\Sexpr{gene.expression("loxodonta_africana","NPPB",complete=TRUE)}).
+Two thymosins,
+\Sexpr{gene.expression("loxodonta_africana","TMSB10",complete=TRUE)}
+and
+\Sexpr{gene.expression("loxodonta_africana","TMSB4Y",complete=TRUE)}
 were also also was highly expressed. The miRNA mir-675
-(\Sexpr{gene.expression("loxodonta
-  africana","ENSLAFG00000027406",complete=TRUE)}). which is imbedded
-in the first exon of the maternally imprinted long non-coding RNA gene
-\textit{H19} \cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene
-which is currently un-annotated in \elephant{}. %
-\DLA{Probably consider fixing the unannotation of H19; it seems pretty
-  obvious to me, though it would require fixing the GTF file for
-  elephant with the right 
-  \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan prediction} for this region}
-Five uncharacterized genes are expressed to a high degree in the
-\elephant placenta (\cref{tab:top50_elephant_placenta}), three of
-which %
-%
-(\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000028493",complete=TRUE)},
-\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000032011",complete=TRUE)},
-\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000031933",complete=TRUE)}) %
-%
+(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027406",complete=TRUE)}).
+which is imbedded in the first exon of the maternally imprinted long
+non-coding RNA gene \textit{H19}
+\cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene which is
+currently un-annotated in \elephant{}. \DLA{Probably consider fixing
+  the unannotation of H19; it seems pretty obvious to me, though it
+  would require fixing the GTF file for elephant with the right
+  \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan
+    prediction} for this region} Five uncharacterized genes are
+expressed to a high degree in the \elephant placenta
+(\cref{tab:top50_elephant_placenta}), three of which
+(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000028493",complete=TRUE)},
+\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032011",complete=TRUE)},
+\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000031933",complete=TRUE)})
 are members of an \elephant{}-specific expansion of 11 genes closely
@@
 \cref{fig:prss16_evolution_tree} The other 8 \textit{PRSS16}-like
-genes are also expressed to a lesser degree %
-(\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000026505",complete=TRUE)}). %
-%
+genes are also expressed to a lesser degree
+(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000026505",complete=TRUE)}).
 Two other novel genes with high expression are related to the gene
-\textit{SSLP1} %
-%
-(\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000032659",complete=TRUE)} %
-and %
-%
-\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000029551",complete=TRUE)},
-%
+\textit{SSLP1}
+(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032659",complete=TRUE)}
+and
+\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000029551",complete=TRUE)},
 \cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
@@
 Placental expression of the hormones relaxin (\textit{RLN}, %
-\Sexpr{gene.expression("loxodonta africana","ENSLAFG00000036090")})
+\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000036090")})
 and prolactin (\textit{PRL}) are exceptionally high in the elephant
@@
 is little to no placental expression of the estrogen synthase gene,
-\textit{CYP19A1} %
-(\Sexpr{gene.expression("loxodonta africana","CYP19A1")}) and the
-enzyme that produces progesterone HSD3B1 %
-(\Sexpr{gene.expression("loxodonta africana","HSD3B")}). Other genes
+\textit{CYP19A1}
+(\Sexpr{gene.expression("loxodonta_africana","CYP19A1")}) and the
+enzyme that produces progesterone HSD3B1
+(\Sexpr{gene.expression("loxodonta_africana","HSD3B")}). Other genes
 with high expression in the elephant placenta include genes involved
-in oxidative stress %
-(\Sexpr{gene.expression("loxodonta africana","TXN",complete=TRUE)}),
-growth %
-(\Sexpr{gene.expression("loxodonta africana","IGF2",complete=TRUE)})
-and %
-\Sexpr{gene.expression("loxodonta africana","PAEP",complete=TRUE)}, %
-a gene usually expressed in the endometrium of the uterus in humans
+in oxidative stress
+(\Sexpr{gene.expression("loxodonta_africana","TXN",complete=TRUE)}),
+growth
+(\Sexpr{gene.expression("loxodonta_africana","IGF2",complete=TRUE)})
+and
+\Sexpr{gene.expression("loxodonta_africana","PAEP",complete=TRUE)}, a
+gene usually expressed in the endometrium of the uterus in humans
 \cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.

### 2016-01-08 manuscript/manuscript.Rnw
discuss armadillo placenta transcriptome

@@
 with $\mathrm{FPKM}\ge 50$ in Supplemental
-\cref{tab:all_dasypus_novemcinctus_placenta}.
+\cref{tab:all_dasypus_novemcinctus_placenta}. The gene with the
+highest expression is
+\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045029",complete=TRUE)}
+which is most similar to the \mouse \textit{Pga5} is expanded in
+\opossum into three members which are expressed in placenta (the other
+two being
+\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033885",complete=TRUE)}
+and
+\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000040431",complete=TRUE)}),
+though the expansion is less than that seen \cow and \sheep
+(\cref{fig:pag2_evolution_tree}). There is also high expression of
+hemoglobin subunit beta,
+\Sexpr{gene.expression("dasypus_novemcinctus","HBB",complete=TRUE)},
+an un-annotated gene with high similarity to \human and \cow
+\textit{OVOS2}
+\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033320",complete=TRUE)}
+(\cref{fig:ovos2_evolution_tree}), the growth factors
+\Sexpr{gene.expression("dasypus_novemcinctus","IGF2",complete=TRUE)}
+and
+\Sexpr{gene.expression("dasypus_novemcinctus","IGFBP4",complete=TRUE)},
+and the thymosin
+\Sexpr{gene.expression("dasypus_novemcinctus","TMSB4X",complete=TRUE)}.
+As has been seen previously, multiple components of the annexin
+complex are also highly expressed, including
+\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045634",complete=TRUE)}
+(\textit{ANXA8L1}),
+\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000017399",complete=TRUE)}
+(\textit{S100B}),
+\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000048504",complete=TRUE)}
+(\textit{S100A10}), and
+\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000004237",complete=TRUE)}
+(\textit{ANXA2}).
+
+\begin{figure}
+<<ovos2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree["ENSDNOG00000033320",tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSDNOG00000033320","ENSMODG00000017988"))
+@ 
+\caption{Expression of \textit{OVOS2} and related genes}
+\label{fig:ovos2_evolution_tree}
+\end{figure}
 

### 2016-01-08 manuscript/manuscript.Rnw
comment out comparative analysis of linage

@@
 
-\subsection{Comparative Analysis of Gene Expression}
-
-We performed a comparative analysis of gene expression by
-reconstructing the evolution of expression values and detecting
-significant shifts in expression along specific lineages. Likelihood
-values for each model for each of the X genes reconstructed, as well
-as likelihood ratio tests, P- and Q-values (after correcting for
-multiple testing) are shown in Supplemental Table X. We identified
-between 4 (Boreoeutheria) and 550 (Mouse) significant shifts (Q <
-0.01) in expression on specific lineages (Table 10). The top 10 genes
-with significant shifts (if applicable) per lineage are shown in Table
-10 and all genes for each branch are shown in Supplemental Tables
-X-XX. Genes that showed shifts on the human lineage were
-overwhelmingly identified as placenta-specific by the
-tissue-specificity index (TSI), with 56 out of 279 genes (20\%) with
-both TSI>0.75 and placenta as one the top two tissues with the highest
-expression. Unsurpisingly, this was not the case with other lineages,
-since we used expression values from a human-derived microarray to
-generate TSI values for each gene. However, this does underscore the
-dynamic nature of expression patterns in the placenta and indeed all
-tissues across mammals.
-
-We then performed DAVID functional annotation clustering analyses on
-each list of genes from each lineage. The top three clusters for each
-lineage are shown in Table 10 and a full list of clusters in
-Supplemental Table X. The top enrichment cluster in the human lineage
-is blood vessel development, followed by extracellular matrix and
-metal ion binding (Table 10). Other significant clusters include those
-related to cell communication/signaling, cytokines, defense response,
-tryptophan metabolism, apoptosis, and skeletal system development
-(Supplemental Table X). The bonobo lineage was enriched in only three
-clusters: apoptosis, Toll-like receptor signalling and apoptosis. The
-mouse displayed enrichment in clusters involved in epigenetic
-modifications such as methyltransferase activity and histone
-modification, as well embryonic morphogeneisis, lipid biosynthesis,
-among others (Supplemental Table X). The top enriched cluster in the
-cow lineage was the response to vitamin A, as well as many clusters
-related to the immune system. The dog showed enrichment in skeletal
-system development, as well as lipid metabolic processes and blood
-circulation. The elephant lineage had only one enriched cluster:
-carbohydrate binding. Genes that showed enrichment on the opossum
-lineage (essentially representing the difference between placental
-mammals and marsupials) were clustered in many categories not seen in
-the other lineages, including clusters related to the mitochondria,
-urea cycle, peroxisome, glycolysis, and transport across the cell
-membrane (Table 10 and Supplemental Table X).
+% Not sure if this is actually worthwhile; I've mentioned some of this
+% already for some of the lineages
+
+% \subsection{Comparative Analysis of Gene Expression}
+% 
+% We performed a comparative analysis of gene expression by
+% reconstructing the evolution of expression values and detecting
+% significant shifts in expression along specific lineages. Likelihood
+% values for each model for each of the X genes reconstructed, as well
+% as likelihood ratio tests, P- and Q-values (after correcting for
+% multiple testing) are shown in Supplemental Table X. We identified
+% between 4 (Boreoeutheria) and 550 (Mouse) significant shifts (Q <
+% 0.01) in expression on specific lineages (Table 10). The top 10 genes
+% with significant shifts (if applicable) per lineage are shown in Table
+% 10 and all genes for each branch are shown in Supplemental Tables
+% X-XX. Genes that showed shifts on the human lineage were
+% overwhelmingly identified as placenta-specific by the
+% tissue-specificity index (TSI), with 56 out of 279 genes (20\%) with
+% both TSI>0.75 and placenta as one the top two tissues with the highest
+% expression. Unsurpisingly, this was not the case with other lineages,
+% since we used expression values from a human-derived microarray to
+% generate TSI values for each gene. However, this does underscore the
+% dynamic nature of expression patterns in the placenta and indeed all
+% tissues across mammals.
+% 
+% We then performed DAVID functional annotation clustering analyses on
+% each list of genes from each lineage. The top three clusters for each
+% lineage are shown in Table 10 and a full list of clusters in
+% Supplemental Table X. The top enrichment cluster in the human lineage
+% is blood vessel development, followed by extracellular matrix and
+% metal ion binding (Table 10). Other significant clusters include those
+% related to cell communication/signaling, cytokines, defense response,
+% tryptophan metabolism, apoptosis, and skeletal system development
+% (Supplemental Table X). The bonobo lineage was enriched in only three
+% clusters: apoptosis, Toll-like receptor signalling and apoptosis. The
+% mouse displayed enrichment in clusters involved in epigenetic
+% modifications such as methyltransferase activity and histone
+% modification, as well embryonic morphogeneisis, lipid biosynthesis,
+% among others (Supplemental Table X). The top enriched cluster in the
+% cow lineage was the response to vitamin A, as well as many clusters
+% related to the immune system. The dog showed enrichment in skeletal
+% system development, as well as lipid metabolic processes and blood
+% circulation. The elephant lineage had only one enriched cluster:
+% carbohydrate binding. Genes that showed enrichment on the opossum
+% lineage (essentially representing the difference between placental
+% mammals and marsupials) were clustered in many categories not seen in
+% the other lineages, including clusters related to the mitochondria,
+% urea cycle, peroxisome, glycolysis, and transport across the cell
+% membrane (Table 10 and Supplemental Table X).
 

### 2016-01-08 manuscript/manuscript.Rnw
discuss sheep placenta

@@
 <<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
-top.sheep.genes <- top.genes.table(species.name="ovis aries")
+top.sheep.genes <- top.genes.table(species.name="ovis aries",n=200)
+top.sheep.genes <-
+    top.sheep.genes[!grepl("\\{(7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\}",
+                           `Gene symbol`) &
+                    !grepl("\\((7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\)",
+                           `Human Symbol`)
+                   ,][1:50,]
 print(xtable(top.sheep.genes),
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes in \sheep which are
-  not ribosomal, mitochondrial, or whose human ortholog is a
-  housekeeping gene}
+\caption{\tableexplanation{\sheep}{tab:all_ovis_aries_placenta}}
 \label{tab:top50_sheep_placenta}
@@
 \cref{tab:top50_sheep_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_ovis_aries_placenta}.
+genes in Supplemental \cref{tab:all_ovis_aries_placenta}. The complete
+list of highly expressed genes includes many snRNAs and rRNAs likely
+due to the RNA extraction and library preparation procedure utilized
+for this sample. Ignoring these, notable highly expressed genes
+include members of the annexin complex
+(\Sexpr{gene.expression("ovis_aries","S100G",complete=TRUE)},
+\Sexpr{gene.expression("ovis_aries","ANXA2",complete=TRUE)}),
+hemoglobin subunits
+(\Sexpr{gene.expression("ovis_aries","ENSOARG00000010944",complete=TRUE)}),
+an ortholog of \textit{SSLP1},
+\Sexpr{gene.expression("ovis_aries","SORD1",complete=TRUE)}, two
+cathespins (\Sexpr{gene.expression("ovis_aries","CTSL",complete=TRUE)}
+and \Sexpr{gene.expression("ovis_aries","CTSB",complete=TRUE)}), and
+placentally expressed transcript 1
+\Sexpr{gene.expression("ovis_aries","PLET1",complete=TRUE)} which may
+play a role in cellular migration
+\cite{Raymond.ea2010:Expressionoforphanprotein}.
 

### 2016-01-08 manuscript/manuscript.Rnw
make legends of gene tree figures more detailed

@@
 
+\newcommand{\genetreeexplanation}[1]{Gene tree and heatmap of
+  expression of genes related to #1. The gene tree was obtained from
+  Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes from species
+  not included in this study or having expression by \ac{FPKM} less
+  than 1 were excluded from the plotted tree. Heatmap cells in grey
+  are from species where there was no expression of the relevant
+  transcript (or a species where that transcript does not apply).}
+
 \begin{document}
@@
 @ 
-\caption{Expression of follistatin-like 1 \textit{FSTLl1} and related
-  genes.}
+\caption{\genetreeexplanation{follistatin-like 1 \textit{FSTLl1}}}
 \label{fig:fstl1_evolution_tree}
@@
 @ 
-\caption{Expression of pregnancy-specific glycoproteins}
+\caption{\genetreeexplanation{pregnancy-specific glycoproteins}}
 \label{fig:psg_evolution_tree}
@@
 @ 
-\caption{Expression of \textit{SERPINE2} genes with $\mathrm{FPKM} \ge 1$}
+\caption{\genetreexplanation{\textit{SERPINE2}}}
 \label{fig:serpine_evolution_tree}
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes in \smonkey which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene}
+\caption{\tableexplanation{\smonkey}{tab:all_ateles_fuciceps_placenta}}
 \label{tab:top50_spider_monkey_placenta}
@@
 cell-cell interaction in the developing placenta
-\cite{Liu.ea2015:AnnexinA2complexeswith}.
+\cite{Liu.ea2015:AnnexinA2complexeswith}. All genes with expression
+greater than 50 \ac{FPKM} are listed in Supplementary
+\cref{tab:all_ateles_fuciceps_placenta}.
 
@@
 @ 
-\caption{Expression in \ac{FPKM} of the top 50 genes in \dog which are
-  not ribosomal, mitochondrial, or whose human ortholog is a
-  housekeeping gene}
+\caption{\tableexplanation{\dog}{tab:all_canis_familiaris_placenta}}
 \label{tab:top50_dog_placenta}
@@
 @ 
-\caption{Expression of genes related to \cow \textit{PAG2} with $\mathrm{FPKM} \ge 1$}
+\caption{\genetreeexplanation{\cow \textit{PAG2}}}
 \label{fig:pag2_evolution_tree}
@@
 @ 
-\caption{Expression in \ac{FPKM} of top 50 genes which are not ribosomal, 
-mitochondrial, or whose human ortholog is a housekeeping gene in
-  \textit{Loxodonta africana}}
+\caption{\tableexplanation{\elephant}{tab:top50_elephant_placenta}}
 \label{tab:top50_elephant_placenta}
@@
 @ 
-\caption{Expression of \textit{PRSS16} and related genes}
+\caption{\genetreeexplanation{\textit{PRSS16}}}
 \label{fig:prss16_evolution_tree}
@@
 @ 
-\caption{Expression of \textit{RLN2} and related genes}
+\caption{\genetreeexplanation{\textit{RLN2}}}
 \label{fig:rln2_evolution_tree}
@@
 @ 
-\caption{Expression of \textit{OVOS2} and related genes}
+\caption{\genetreeexplanation{\cow \textit{OVOS2}}}
 \label{fig:ovos2_evolution_tree}
@@
 @ 
-\caption{Expression of estrogen sythesis genes}
+\caption{\genetreeexplanation{estrogen sythesis genes}}
 \label{fig:estrogen_evolution_tree}
@@
 @ 
-\caption{Expression of progesterone sythesis genes}
+\caption{\genetreeexplanation{progesterone sythesis genes}}
 \label{fig:progesterone_evolution_tree}
@@
 @ 
-\caption{Expression of chorionic gonadotropin}
+\caption{\genetreeexplanation{chorionic gonadotropin}}
 \label{fig:cgb_evolution_tree}
@@
 @ 
-\caption{Expression of chorionic somatomammotropin}
+\caption{\genetreeexplanation{chorionic somatomammotropin}}
 \label{fig:csh1_evolution_tree}
@@
 @ 
-\caption{Expression of prolactin}
+\caption{\genetreeexplanation{prolactin}}
 \label{fig:prolactin_evolution_tree}
@@
 @ 
-\caption{Expression of prolactin receptor}
+\caption{\genetreeexplanation{prolactin receptor}}
 \label{fig:prolactin_receptor_evolution_tree}
@@
 @ 
-\caption{Expression of \textit{TMSB1}}
+\caption{\genetreeexplanation{\textit{TMSB1}}}
 \label{fig:tmsb1_evolution_tree}
@@
 @ 
-\caption{Expression of SSLP1}
+\caption{\genetreeexplanation{\textit{SSLP1}}}
 \label{fig:sslp1_evolution_tree}
@@
 @ 
-\caption{Expression of galectins}
+\caption{\genetreeexplanation{galectins}}
 \label{fig:galectins_lgals13_evolution_tree}

### 2016-01-08 manuscript/manuscript.Rnw
fix up introduction and add gestational period table

@@
 tissues lasts for a proportionally shorter time during gestation
-\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. Marsupials also differ from eutherian mammals in
-that most possess only a choriovitelline placenta, produced from the
-fusion of the chorion and yolk sac, during most of gestation. Although
-some eutherian mammals may possess a choriovitelline placenta at some
-point in development, the major site of nutrient exchange in
-placentals occurs via the chorioallantoic placenta, produced from the
-fusion of the chorion and allantois \cite{Mossman.1987:vertebratefetalmembranes}. It is unclear
+\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. Marsupials also
+differ from eutherian mammals in that most possess only a
+choriovitelline placenta, produced from the fusion of the chorion and
+yolk sac, during most of gestation. Although some eutherian mammals
+may possess a choriovitelline placenta at some point in development,
+the major site of nutrient exchange in placentals occurs via the
+chorioallantoic placenta, produced from the fusion of the chorion and
+allantois \cite{Mossman.1987:vertebratefetalmembranes}. It is unclear
 whether the first placenta was more like that of marsupials or
@@
 Although the morphology of the mammalian placenta has been well
-characterized in a number of species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
-the molecular environment is only beginning to be understood from a
-genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}. Studies of certain model species
-(i.e., humans, mice, cattle) have revealed a stunning divergence in
-the genetics of the placenta, especially in later stages of pregnancy
+characterized in a number of species
+\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
+molecular environment is only beginning to be understood from a
+genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
+Studies of certain model species (i.e., humans, mice, cattle) have
+revealed a stunning divergence in the genetics of the placenta,
+especially in later stages of pregnancy
 \cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
-For example, each lineage has evolved expansive
-placenta-specific gene families with expression patterns that can vary
-in space and time \cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}. These
-include independent expansions of the prolactins in mouse and cow,
-pregnancy-specific glycoproteins (PSGs) in humans and mouse,
+For example, each lineage has evolved expansive placenta-specific gene
+families with expression patterns that can vary in space and time
+\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
+These include independent expansions of the prolactins in mouse and
+cow, pregnancy-specific glycoproteins (PSGs) in humans and mouse,
 pregnancy-associated glycoproteins (PAGs) in cow, and the expansion of
 the growth hormone and chorionic gonadotropin beta genes in hominoid
-primates \cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}. Expression patterns of
-specific genes also can vary widely between species. For example, the
-expression of orthologous genes in the placenta have been found to
-vary substantially between humans and mice \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
-recruitment of SIGLEC6 to placental expression in humans differs from
-both chimpanzees and mice \cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
+primates
+\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
+Expression patterns of specific genes also can vary widely between
+species. For example, the expression of orthologous genes in the
+placenta have been found to vary substantially between humans and mice
+\cite{Su.ea2002:Large-scaleanalysisofhuman}, and the recruitment of
+\textit{SIGLEC6} to placental expression in humans differs from both
+chimpanzees and mice
+\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
 
@@
 the distinct evolutionary interplay between maternal and fetal
-strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}. The maternal strategy is to
-requisition an adequate degree of nutrients to the fetus without
-risking the health of the mother, while the fetal strategy via
-paternal genes is to extract as many resources from the mother as
-possible. Therefore, an interplay between these extremes can set up an
-intricate “arms race”, involving the imprinting of parental alleles,
-production of new genes, and enlistment of existing genes to
-expression at the maternal-fetal interface \cite{Haig1993:Geneticconflictsinhuman}.
-
-One symptom
-of this maternal-fetal conflict is its potential affect on human
-health \cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. 
-In addition to having an invasive
-hemochorial placenta, humans possess a deeply invasive trophoblast
-layer that can penetrate the myometrium of the uterus 
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
- Defects in placental growth are known to be
-associated with the devastating condition of preeclampsia, which can
-cause hypertension, proteinuria, and has the potential to lead to
-death of the mother and/or infant \DLA{(citation)}. Although cases have been
-documented in chimpanzees and gorillas \DLA{(citation)}, preeclampsia has a
-far higher incidence in human pregnancies with as much as 4\% of
-pregnancies being affected in the United States \DLA{(citation)}.
-Chimpanzees and gorillas also have deeply invasive trophoblasts,
-although the progression of invasion in great apes may be slower than
-humans \cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however, large-scale expression
-studies of great ape placentas have not been conducted. The increasing
-affordability of next-generation sequencing technology has enabled the
-large-scale study of specific tissues from a wide array of species to
-address broader phylogenetic questions in the evolution of gene
-expression \cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use Illumina RNA-Seq
-technology to sequence and quantify the mRNA environment of six
-mammalian placentas at term, including that of the \human, \bonobo (the
-sister species to the chimpanzee), \dog, \cow, \elephant, and
-\opossum, representing marsupials as well as three major branches of
-eutherian mammals (Euarchontoglires, Laurasiatheria, Afrotheria).
-These placentas also represent a wide array of placental morphologies
-as well as traits such as number and size of offspring and length of
-gestation \DLA{(Table 1)}. We use this data, along with existing RNA-Seq
-data from mouse \cite{Wang.ea2011:surveyfornovelimprinted}, to address the major differences
-between expression in marsupial and eutherian mammals at term,
-identify the evolution of expression in important placental genes such
-as hormones and galectins, identify novel lineage-specific expression
-patterns, and quantify human specific differences in gene expression
-that may be related to obstetrical disorders such as preeclampsia.
+strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}.
+The maternal strategy is to requisition an adequate degree of
+nutrients to the fetus without risking the health of the mother, while
+the fetal strategy via paternal genes is to extract as many resources
+from the mother as possible. Therefore, an interplay between these
+extremes can set up an intricate “arms race”, involving the imprinting
+of parental alleles, production of new genes, and enlistment of
+existing genes to expression at the maternal-fetal interface
+\cite{Haig1993:Geneticconflictsinhuman}.
+
+One symptom of this maternal-fetal conflict is its potential affect on
+human health
+\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
+addition to having an invasive hemochorial placenta, humans possess a
+deeply invasive trophoblast layer that can penetrate the myometrium of
+the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
+Defects in placental growth are known to be associated with the
+devastating condition of preeclampsia, which can cause hypertension,
+proteinuria, and has the potential to lead to death of the mother
+and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
+Although cases have been documented in chimpanzees
+\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
+gorillas \cite{Thornton.ea1992:Convulsionsinpregnancyin}, preeclampsia
+has a far higher incidence in human pregnancies with as much as 4\% of
+pregnancies being affected in the United States
+\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Chimpanzees
+and gorillas also have deeply invasive trophoblasts, although the
+progression of invasion in great apes may be slower than humans
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however,
+large-scale expression studies of great ape placentas have not been
+conducted. The increasing affordability of next-generation sequencing
+technology has enabled the large-scale study of specific tissues from
+a wide array of species to address broader phylogenetic questions in
+the evolution of gene expression
+\cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use RNA-Seq
+technology to sequence and quantify the mRNA environment of mammalian
+placentas at term, including that of the \human, \bonobo (the sister
+species to the chimpanzee), \smonkey, \mouse, \galili, \carmeli, \dog,
+\cow, \elephant, \armadillo, and \opossum, representing marsupials as
+well as three major branches of eutherian mammals (Euarchontoglires,
+Laurasiatheria, Afrotheria). These placentas also represent a wide
+array of placental morphologies as well as traits such as number and
+size of offspring and length of gestation
+\cref{tab:gestational_periods}. We use this data, along with existing
+RNA-Seq data from \mouse \cite{Wang.ea2011:surveyfornovelimprinted},
+\human \cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
+\horse \cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
+\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
+\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
+differences between expression in marsupial and eutherian mammals at
+term, identify the evolution of expression in important placental
+genes such as hormones and galectins, identify novel lineage-specific
+expression patterns, and quantify human specific differences in gene
+expression that may be related to obstetrical disorders such as
+preeclampsia.
+
+\begin{table}
+  \begin{tabular}{c c c c}
+    \toprule
+    Species                       & Gestation (days) & Litter Size  & Source \\
+    \midrule
+    \textit{Ateles fusciceps}     & 224 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Dasypus novemcinctus} & 135 & 4 & \DLA{Cite Amazing Armadillo: geography book}\\
+    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
+    \textit{Loxodonta africana}   & 644 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
+    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
+    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Pan paniscus}         & 223 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental} \\
+    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
+    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \bottomrule
+  \end{tabular}
+  \caption{Gestational periods in days and average liter sizes for studied species.}
+  \label{tab:gestational_periods}
+\end{table}
 
@@
 cow], \elephant [African savanna elephant], \opossum [gray
-short-tailed opossum], \gallili [?], \carmelli [?]). Information on
+short-tailed opossum], \gallili [blind mole rat from Upper Galilee],
+\carmelli [blind mole rat from northern Israel]). Information on
 specimen ID, place and time of collection, and collector is shown for
@@
 \cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
-\DLA{citation; e-mailed Alec Archibald from the ISGC about what the
-  right one should be}, and \pig
+\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
 \cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
@@
 transcription level in \ac{FPKM}. We considered a gene to be core to
-the transcriptome if an ortholog of that gene was expressed with
-$\mathrm{FPKM} \ge 10$ in all species studied and was not a
-housekeeping gene
-\cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
+the placenta transcriptome if a 1:1 ortholog of that gene was
+expressed with $\mathrm{FPKM} \ge 10$ in all species studied and was
+not a human housekeeping gene according to
+\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
 \Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
-which would be included in the core placenta set if the cutoff was set
-differently.
-%
-\textit{ANXA1} is
-involved in the modulation of the inflammatory response \DLA{It does
-  more stuff; read the huthcinson2011 review and fill this in}
+which would be included in the core placenta set if the \ac{FPKM}
+cutoff was set differently.
+%
+Multiple components of the annexin complex are included in the core
+transcriptome, including \textit{ANXA2}, \textit{ANXA1},
+\textit{S100A11}, \textit{S100A10}. The modulation of  \DLA{It does more stuff; read
+  the huthcinson2011 review and fill this in}
 %

### 2016-01-12 manuscript/manuscript.Rnw
use citet and p{} in the gestational period table
discuss CYP19A1 and HSDB1 expression levels

@@
 \begin{table}
-  \begin{tabular}{c c c c}
+  \begin{tabular}{c c c p{3in}}
     \toprule
@@
     \midrule
-    \textit{Ateles fusciceps}     & 224 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
     \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
@@
     \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
-    \textit{Loxodonta africana}   & 644 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
     \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
@@
     \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Pan paniscus}         & 223 & 1 & \cite{Ernest2003:lifehistorycharacteristicsofplacental} \\
+    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
     \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
@@
 with hormone products or enzymatic products that directly produce
-hormones in multiple mammalian species (\DLA{Table 8}).
+hormones in multiple mammalian species (\cref{fig:hormone_genes}).
 
@@
 production of estrogen and progesterone. High relative expression of
-CYP19A1, the enzyme that synthesizes estrogen, and HSD3B1, the enzyme
-that synthesizes progesterone, in the human, bonobo, and cow,
-generally matches the high levels of placental production of estrogen
-and progesterone in humans, other primates, and ruminants in late
-pregnancy. In addition, HSD3B1 shows high levels in dog, which matches
-the high levels of placental production of progesterone at term in
-that species. \DLA{I think we need citations for this}
+\textit{CYP19A1} (\cref{fig:estrogen_evolution_tree}), the enzyme that
+synthesizes estrogen from androstenedione is seen in \human, \bonobo,
+\cow, and \horse, though not in \sheep, \smonkey. %
+%
+\textit{HSD3B1}, one of two enzymes that synthesizes progesterone from
+pregnalone
+(\cite{Simard.ea2005:Molecularbiologyof3beta-hydroxysteroid},
+\cref{fig:progesterone_evolution_tree}) is also highly expressed in
+\human, \bonobo, \cow, \horse, and \dog, but is not highly expressed
+in \smonkey. The diversity of \textit{CYP19A1} and \textit{HSD3B1}
+expression levels in in primates and ungulates seen here points to the
+need for more complete sampling of placentas at different timepoints
+and in different clades.
 

### 2016-01-12 manuscript/manuscript.Rnw
subset PAG gene tree

@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=1)
+                 gene.tree.table,gene.tree.expression,min.fpkm=1,
+                 subtree=c("ENSSSCG00000006792",
+                           "ENSBTAG00000046085")
+                 )
 @ 

### 2016-01-12 manuscript/manuscript.Rnw
subset PR tree and fix synthesis typo

@@
 @ 
-\caption{\genetreeexplanation{estrogen sythesis genes}}
+\caption{\genetreeexplanation{estrogen synthesis genes}}
 \label{fig:estrogen_evolution_tree}
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{progesterone sythesis genes}}
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSMODG00000005385","ENSLAFG00000027896")
+                 )
+@ 
+\caption{\genetreeexplanation{progesterone synthesis genes}}
 \label{fig:progesterone_evolution_tree}
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSG00000113494","ENSMODG00000020376"))
 @ 

### 2016-01-12 manuscript/manuscript.Rnw
fix fusciceps typo
cite amazing armadillo

@@
     \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Dasypus novemcinctus} & 135 & 4 & \DLA{Cite Amazing Armadillo: geography book}\\
+    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
     \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
@@
 @ 
-\caption{\tableexplanation{\smonkey}{tab:all_ateles_fuciceps_placenta}}
+\caption{\tableexplanation{\smonkey}{tab:all_ateles_fusciceps_placenta}}
 \label{tab:top50_spider_monkey_placenta}
@@
 \cite{Sampson.ea2012:PAGE4positivityisassociated}, \textit{PAGE4} %
-(\Sexpr{gene.expression("ateles fusciceps","PAGE4")}), %
+(\Sexpr{gene.expression("ateles_fusciceps","PAGE4")}), %
 the pregnancy specific glycoprotein \textit{PSG8}
-(\Sexpr{gene.expression("ateles fusciceps","PSG8")}), %
+(\Sexpr{gene.expression("ateles_fusciceps","PSG8")}), %
 a member of the chorionic somatotropins, \textit{GH1} %
-(\Sexpr{gene.expression("ateles fusciceps","GH1")}), %
+(\Sexpr{gene.expression("ateles_fusciceps","GH1")}), %
 a few HLA genes (\textit{HLA-A}, \textit{HLA-F},
 \cref{tab:top50_spider_monkey_placenta}), and multiple components of
-the annexin complex %
-(\Sexpr{gene.expression("ateles fusciceps","ANXA2",complete=TRUE)}, %
-\Sexpr{gene.expression("ateles fusciceps","S100P",complete=TRUE)}, %
-\Sexpr{gene.expression("ateles fusciceps","S100A11",complete=TRUE)}) %
+the annexin complex
+(\Sexpr{gene.expression("ateles_fusciceps","ANXA2",complete=TRUE)}, %
+\Sexpr{gene.expression("ateles_fusciceps","S100P",complete=TRUE)}, %
+\Sexpr{gene.expression("ateles_fusciceps","S100A11",complete=TRUE)}) %
 which may play a role in angiogenesis, and certainly regulate
 cell-cell interaction in the developing placenta
-\cite{Liu.ea2015:AnnexinA2complexeswith}. All genes with expression
+\cite{Liu.ea2015:AnnexinA2complexeswith}.
+\Sexpr{gene.expression("ateles_fusciceps","AL365202.1",complete=TRUE)}
+is also highly expressed, has no known function, and is duplicated
+multiple times in \textit{Macaca mulatta}. All genes with expression
 greater than 50 \ac{FPKM} are listed in Supplementary
-\cref{tab:all_ateles_fuciceps_placenta}.
+\cref{tab:all_ateles_fusciceps_placenta}.
 

### 2016-01-12 manuscript/manuscript.Rnw
set grad.def in plots; include more hormone genes

@@
 
+<<plot.defaults,echo=FALSE,results="hide">>=
+grad.def <- list(low = "white", high = "green", space = "Lab", na.value = "gray50")
+
+@
+
 \section{Abstract}
@@
       + scale_y_discrete(expand=c(0,0))
-      + scale_fill_gradient(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = "grey50",
+      + scale_fill_gradient(low = grad.def[["low"]],
+                            high = grad.def[["high"]],
+                            space = grad.def[["space"]],
+                            na.value = grad.def[["na.value"]],
                             limits=c(0,4),
@@
 hormone.genes <- c("CGA","CGB","CGB1","CGB2","CGB5","CGB8","CSH1","CSH2","CSHL1","CYP19A1",
-                   "GH2","GHRH","HSD3B1","NPPB","NPPC","PRL","RLN1")
+                   "GH2","GHRH","HSD3B1","HSD3B2","HSD17B2","HSD17B1","NPPB","NPPC","PRL","RLN1")
 hormone.expression.nozero <-
@@
 ##         scale_fill_gradientn(colours = colorends, values = rescale(gradientends)),
-         scale_fill_gradient(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = "grey50",
+         scale_fill_gradient(low = grad.def[["low"]],
+                             high = grad.def[["high"]],
+                             space = grad.def[["space"]],
+                             na.value = grad.def[["na.value"]],
                              guide=guide_colorbar(title=expression(log[10](FPKM)))),
@@
                axis.text.x = element_text(angle = 300, hjust = 0, vjust=1,face="italic")),
-         facet_wrap(~type,scale="free_y",ncol=1)
+         facet_grid(type~.,scale="free_y",space="free_y")
          )

### 2016-01-12 manuscript/manuscript.Rnw
text cite yammamoto

@@
 lineage \cite{Wallis2009:ProlactininAfrotheriacharacterization}.
-\cite{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
+\citet{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
 presence of a prolactin in the trophoblast of the elephant placenta

### 2016-01-12 manuscript/manuscript.Rnw
comment out old abstract; clearpage before discussion

@@
 \Sexpr{star.logs[,length(unique(species))]} species of therian (i.e.
-marsupials and “placental”) mammals. We identified significant shifts
-in gene expression along specific evolutionary lineages using a model
-of character evolution. Genes that showed significant elevation in
-expression along the human lineage ($\mathrm{FDR}\le 0.01$) were
-enriched in GO categories such as angiogenesis, tube morphogenesis,
-cell migration, and regulation of inflammatory response. Three of
-these genes (KISS1, ADAM12, GPC3) showed exceptionally high expression
-in humans; these genes are implicated in preeclampsia and may indicate
-that the risk of preeclampsia recently emerged on the human lineage.
+marsupials and “placental”) mammals.
+
+% We identified significant shifts in gene expression along specific
+% evolutionary lineages using a model of character evolution. Genes that
+% showed significant elevation in expression along the human lineage
+% ($\mathrm{FDR}\le 0.01$) were enriched in GO categories such as
+% angiogenesis, tube morphogenesis, cell migration, and regulation of
+% inflammatory response. Three of these genes (KISS1, ADAM12, GPC3)
+% showed exceptionally high expression in humans; these genes are
+% implicated in preeclampsia and may indicate that the risk of
+% preeclampsia recently emerged on the human lineage.
 
@@
 
+\clearpage
+
 \section{Discussion}

### 2016-01-12 manuscript/manuscript.Rnw
discuss CGB and CGA

@@
 \begin{figure}
-<<chorionic.gonadotropin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CGB1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{chorionic gonadotropin}}
-\label{fig:cgb_evolution_tree}
-\end{figure}
-
-\begin{figure}
 <<chorionic.somatomammotropin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
@@
 
-The growth hormone (GH) family has expanded in anthropoid primates,
-resulting in approximately 5 genes in humans (GH1, GH2, CSH1, CSH2,
-CSHL1). All but GH1 show substantial expression in the placenta (Table
-8). There is some evidence that increased expression of these genes
-may be correlated with overall birth weight of offspring in humans
-\DLA{(Mannik et al., 2010)}, and it is possible that the extreme expression
-levels of these genes may be due to high birth weight. Ensembl
-annotations show 4 GH genes present in the chimpanzee genome, and
-three of these genes show high expression in bonobo, although to a
-lesser degree than in human (Table 8). Human placental GH genes are
-involved in increasing blood flow, replacing maternal pituitary
-expression of GH, as well as trophoblast invasion \cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth}, and it is likely that
-they have the same function in the bonobo.
-
 The human gene CGA codes for the alpha subunit of multiple hormones
 including pituitary hormones (luteinizing hormone,
+The \ac{GH} family has expanded in anthropoid primates, resulting in
+approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
+\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
+\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
+between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
+reliably distinguish transcripts between the two). All but
+\textit{GH1} show substantial expression in the \human placenta
+(\cref{fig:csh1_evolution_tree}). There is some evidence that
+increased expression of these genes may be correlated with overall
+birth weight of offspring in \human
+\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
+possible that the extreme expression levels of these genes may be due
+to high birth weight. \human placental \ac{GH} genes are involved in
+increasing blood flow, replacing maternal pituitary expression of
+\ac{GH}, as well as trophoblast invasion
+\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
+and it is likely that they have the same function in \bonobo.
+
+\begin{figure}
+<<chorionic.somatomammotropin.alpha.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CGA" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{chorionic gonadotropin, alpha subunit}}
+\label{fig:cga_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<chorionic.gonadotropin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CGB1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{chorionic gonadotropin, beta subunit}}
+\label{fig:cgb_evolution_tree}
+\end{figure}
+
+The human gene \textit{CGA} codes for the alpha subunit of multiple
+hormones including pituitary hormones (luteinizing hormone,
 follicle-stimulating hormone, thyroid-stimulating hormone), as well as
-numerous chorionic gonadotropins with variable beta subunits \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of the chorionic gonadotropins is to
-prevent the regression of the corpus luteum \cite{Hanson.ea1971:EffectsofHCGand}.
-Among the six placentas we analyzed here, CGA is only expressed in the
-placentas of humans and bonobos (\DLA{Table 8}), which corresponds with the
-emergence of the chorionic gonadotropins in anthropoid primates
-\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly, CGA is expressed at much
-higher levels in both species than the beta subunits with which its
-product dimerizes (\DLA{Table 8}). CGA is one of several genes noted by
-\cite{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone adaptive evolution of the
-human-chimpanzee stem lineage.
-
-Relaxin is a hormone that has multiple effects during pregnancy
+numerous chorionic gonadotropins with variable beta subunits
+\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of
+the chorionic gonadotropins is to prevent the regression of the corpus
+luteum \cite{Hanson.ea1971:EffectsofHCGand}. Among the six placentas
+we analyzed here, \textit{CGA} is only expressed in the placentas of
+\human and \bonobo (\cref{fig:hormone_genes,fig:cga_evolution_tree}),
+which corresponds with the emergence of the chorionic gonadotropins in
+anthropoid primates
+\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly,
+\textit{CGA} is expressed at much higher levels in both species than
+the beta subunits with which its product dimerizes (\cref{fig:hormone_genes,fig:cga_evolution_tree,fig:cgb_evolution_tree}).
+\textit{CGA} is one of several genes noted by
+\citet{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone
+adaptive evolution of the \human{}-\textit{P. troglodytes} stem lineage.
 including increasing maternal blood flow, widening the pubic bone and

### 2016-01-12 manuscript/manuscript.Rnw
discuss PRL in discussion

@@
 expression of the primarily pituitary hormone prolactin (PRL) in the
-placenta. %
-\DLA{This is totally not true. \textit{A. fusciceps} expresses PRL,
-  and so does \textit{S. carmeli}, and \textit{B. taurus}, \textit{M.
-    musculus}, all have expansions of prolatacin.}
-%
-All afrotherians yet analyzed contain only one prolactin
+placenta. All afrotherians yet analyzed contain only one prolactin
 gene, and prolactin has undergone adaptive evolution on the elephant
@@
 previous analysis of the elephant placenta also detected the
-expression of prolactin (Hou et al., 2012). Here we confirm that the
-sole prolactin in the elephant is not only expressed but highly
-expressed in the placenta, and likely functions both in the
-maintenance of the corpus luteum and as a placental lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}. PRL is not expressed in the opossum, even though there
-is evidence that it is expressed in the placenta of another marsupial,
-the tammar wallaby \cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests that
-placental prolactin expression in marsupials may be as diverse as in
-eutherian mammals. Although, PRL is not expressed to any great degree
-in the placenta other mammals, we see very high expression of genes
-which resulted from duplications of PRL in both the cow \DLA{(Table X)}, as
-well as mouse \DLA{(Supplemental Table X)}. \DLA{Describe these}
-
-The human gene CGA codes for the alpha subunit of multiple hormones
-including pituitary hormones (luteinizing hormone,
+expression of prolactin
+\cite{Hou.ea2012:Elephanttranscriptomeprovidesinsights}. Here we
+confirm that the sole prolactin in the elephant is not only expressed
+but highly expressed in the placenta, and likely functions both in the
+maintenance of the corpus luteum and as a placental lactogen
+\cite{Yamamoto.ea2011:PlacentationinAfricanelephant}. \textit{PRL} is
+not expressed in the opossum, even though there is evidence that it is
+expressed in the placenta of another marsupial, the tammar wallaby
+\cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests
+that placental prolactin expression in marsupials may be as diverse as
+in eutherian mammals. Although, \textit{PRL} is not expressed to any
+great degree in the placenta other mammals, we see very high
+expression of genes which resulted from duplications of \textit{PRL}
+in the \mouse, \galili, \carmeli, and \cow
+(\cref{fig:prolactin_evolution_tree}).
+
 The \ac{GH} family has expanded in anthropoid primates, resulting in

### 2016-01-12 manuscript/manuscript.Rnw
move figures to be closer discussion point

@@
 
-In both ruminants and primates, the placenta is responsible for the
-production of estrogen and progesterone. High relative expression of
-\textit{CYP19A1} (\cref{fig:estrogen_evolution_tree}), the enzyme that
-synthesizes estrogen from androstenedione is seen in \human, \bonobo,
-\cow, and \horse, though not in \sheep, \smonkey. %
-%
-\textit{HSD3B1}, one of two enzymes that synthesizes progesterone from
-pregnalone
-(\cite{Simard.ea2005:Molecularbiologyof3beta-hydroxysteroid},
-\cref{fig:progesterone_evolution_tree}) is also highly expressed in
-\human, \bonobo, \cow, \horse, and \dog, but is not highly expressed
-in \smonkey. The diversity of \textit{CYP19A1} and \textit{HSD3B1}
-expression levels in in primates and ungulates seen here points to the
-need for more complete sampling of placentas at different timepoints
-and in different clades.
-
 \begin{figure}
@@
 
+\subsubsection{Estrogen and Progesterone}
+
+In both ruminants and primates, the placenta is responsible for the
+production of estrogen and progesterone. High relative expression of
+\textit{CYP19A1} (\cref{fig:estrogen_evolution_tree}), the enzyme that
+synthesizes estrogen from androstenedione is seen in \human, \bonobo,
+\cow, and \horse, though not in \sheep, \smonkey. %
+%
+\textit{HSD3B1}, one of two enzymes that synthesizes progesterone from
+pregnalone
+(\cite{Simard.ea2005:Molecularbiologyof3beta-hydroxysteroid},
+\cref{fig:progesterone_evolution_tree}) is also highly expressed in
+\human, \bonobo, \cow, \horse, and \dog, but is not highly expressed
+in \smonkey. The diversity of \textit{CYP19A1} and \textit{HSD3B1}
+expression levels in in primates and ungulates seen here points to the
+need for more complete sampling of placentas at different timepoints
+and in different clades.
+
 \begin{figure}
@@
 
-\begin{figure}
-<<prolactin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{prolactin}}
-\label{fig:prolactin_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<prolactin.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRLR" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 subtree=c("ENSG00000113494","ENSMODG00000020376"))
-@ 
-\caption{\genetreeexplanation{prolactin receptor}}
-\label{fig:prolactin_receptor_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<tmsb1.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="TMSB10" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{\textit{TMSB1}}}
-\label{fig:tmsb1_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<sslp1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_id=="ENSCAFG00000010638" & species=="canis familiaris",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{\textit{SSLP1}}}
-\label{fig:sslp1_evolution_tree}
-\end{figure}
+\subsubsection{Prolactin}
 
 The elephant is strikingly the only mammal analyzed that shows high
-expression of the primarily pituitary hormone prolactin (PRL) in the
+expression of the primarily pituitary hormone prolactin (\textit{PRL}) in the
 placenta. All afrotherians yet analyzed contain only one prolactin
@@
 (\cref{fig:prolactin_evolution_tree}).
+
+\begin{figure}
+<<prolactin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{prolactin}}
+\label{fig:prolactin_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<prolactin.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRLR" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSG00000113494","ENSMODG00000020376"))
+@ 
+\caption{\genetreeexplanation{prolactin receptor}}
+\label{fig:prolactin_receptor_evolution_tree}
+\end{figure}
+
+\subsubsection{\ac{GH} family}
 
@@
 
+\begin{figure}
+<<sslp1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_id=="ENSCAFG00000010638" & species=="canis familiaris",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{\textit{SSLP1}}}
+\label{fig:sslp1_evolution_tree}
+\end{figure}
+
 We found evidence of high expression in two novel expansions of genes

### 2016-01-12 manuscript/manuscript.Rnw
discuss RLN

@@
 adaptive evolution of the \human{}-\textit{P. troglodytes} stem lineage.
-including increasing maternal blood flow, widening the pubic bone and
-relaxing uterine musculature in preparation for labor. In both dogs
-and elephants, there is high expression of relaxin in the term
-placenta, with little evidence in other species examined here (Table
-8). Relaxin is known to be produced in the placenta of the pregnant
-dog and is used in the diagnosis of canine pregnancy
-\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof};
-however, the source of relaxin produced by pregnant African elephants
-remains unclear \cite{Meyer.ea2004:Speciesandfetalgender}. The high
-rise in relaxin in elephants has been suggested to be a major signal
-for the initiation of labor
-\cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive} Here we
-show that the placenta of elephants show high expression of relaxin
-and is likely the predominant source of relaxin at term \DLA{(Table 8)}.
-
-Two genes, NPPB and NPPC, code for peptide hormones that are important
+
+\subsubsection{Relaxin}
+
+Relaxin (\textit{RLN}) is a hormone that has multiple effects during
+pregnancy including increasing maternal blood flow, widening the pubic
+bone and relaxing uterine musculature in preparation for labor. In
+\dog, \elephant, \mouse, \carmeli, and \horse there is high expression
+of relaxin in the term placenta, with little evidence in other species
+examined here (\cref{fig:rln2_evolution_tree}). \textit{RLN} is known
+to be produced in the placenta of the pregnant \dog and is used in the
+diagnosis of canine pregnancy
+\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
+the luteal ovary and placenta of \pig
+\cite{Hisaw.ea1948:Relaxininovaryof}, and the placenta of \horse
+\cite{Klonisch.ea1997:Placentallocalizationofrelaxin}. However, the
+source of relaxin produced by pregnant \elephant remains unclear
+\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in relaxin in
+\elephant has been suggested to be a major signal for the initiation
+of labor \cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive}.
+Here we show that the placenta of elephants show high expression of
+relaxin and is likely a major source of \textit{RLN} at term. In
+contrast to the results of
+\cite{Evans.ea1993:mouserelaxingenenucleotide}, we also see expression
+of \textit{RLN} in the placenta of \mouse and \carmeli.
+
 in the inducement of vasodilation, and may be important in increasing

### 2016-01-12 manuscript/manuscript.Rnw
decrease width and use double spacing
capitalize GH
discuss NPPC and NPPB

@@
 \oddsidemargin 0.0in 
-\textwidth 6.5in
+\textwidth 6in
 \raggedbottom
@@
 \maketitle
-\onehalfspacing
+% \onehalfspacing
+\doublespacing
 
@@
 
-\subsubsection{\ac{GH} family}
+\subsubsection{\Ac{GH} family}
 
@@
 
-in the inducement of vasodilation, and may be important in increasing
-maternal blood flow to the fetus (Cameron et al., 1996). There is some
-indication that these proteins are present in the decidua of mice
-(Cameron et al., 1996), but we find very little expression in the
-fetal portion of the mouse placenta (Table 8). Here we find high
-expression of at least one of these genes in dog, cow, and elephant.
-The lack of expression in the hemochorial placentas of
-Euarchontoglires may mean that high expression of these genes are
-essential in increasing maternal blood flow in epitheliochorial and
-endotheliochorial placentas
+\subsubsection{\textit{NPPB} and \textit{NPPC}}
+
+\begin{figure}
+<<nppb.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="NPPB" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{\textit{NPPB}}}
+\label{fig:nppb_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<nppc.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="NPPC" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{\textit{NPPC}}}
+\label{fig:nppc_evolution_tree}
+\end{figure}
+
+Two genes, \textit{NPPB} and \textit{NPPC}, code for peptide hormones
+that are important in the inducement of vasodilation, and may be
+important in increasing maternal blood flow to the fetus
+\cite{Cameron.ea1996:sitesofgeneexpression}. There is some indication
+that these proteins are present in the decidua of mice
+\cite{Cameron.ea1996:sitesofgeneexpression}, but we find very little
+expression in the fetal portion of the mouse placenta
+(\cref{fig:hormone_genes,fig:nppc_evolution_tree,fig:nppb_evolution_tree}).
+However, we do find high expression of at least one of these genes in
+\dog, \cow, and \elephant. The lack of expression in the hemochorial
+placentas of Euarchontoglires may mean that high expression of these
+genes are essential in increasing maternal blood flow in
+epitheliochorial and endotheliochorial placentas.
 

### 2016-01-12 manuscript/manuscript.Rnw
discuss LGALS13

@@
 @ 
-\caption{\genetreeexplanation{galectins}}
+\caption{\genetreeexplanation{\textit{LGALS13}}}
 \label{fig:galectins_lgals13_evolution_tree}
 \end{figure}
-
+%$
 Galectins are recognized as major regulators of pregnancy in both
-fetal and maternal tissues (Than et al., 2012). Indeed, the evolution
-of new lineage-specific galectins has occurred in both primates and
-ruminants, with the genes LGALS13, LGALS14, and LGALS16 recognized as
-being placenta-specific in humans and regulating immune responses by
-inducing apoptosis of maternal T-cells (Than et al., 2009). Indeed,
-LGALS14 has the highest expression of any galectin in the human
-placenta (Table 8), and other primate galectins such as LGALS13 and
-LGALS16 are also expressed in human placenta. However, in the closely
-related bonobo, LGALS14 is not found, and LGALS1 is the most abundant
-galectin. LGALS1 shows generally high expression in all placentas
-analyzed. LGALS1 is also involved in mediating maternal-fetal immune
-tolerance, and is expressed in both maternal and fetal tissue in
-primates (Than et al., 2008). LGALS1 placental expression was thought
-to be confined to eutherian mammals, as there is evidence of adaptive
-evolution at this node, as well as the appearance of estrogen response
-elements in eutherian cis-regulatory regions (Than et al., 2008). It
-was unclear as to whether galectins such as LGALS1 were important in
-the regulation of pregnancy in marsupials, as well as eutherians. Here
-we show substantial expression of galectins in the opossum, especially
-LGALS1 (Table 8). In addition, LGALS2, LGALS3, and LGALS8 are also
+fetal and maternal tissues
+\cite{Than.ea2012:Galectinsguardiansofeutherian}. Indeed, the
+evolution of new lineage-specific galectins has occurred in both
+primates and ruminants, with the genes \textit{LGALS13},
+\textit{LGALS14}, and \textit{LGALS16} recognized as being
+placenta-specific in \human and regulating immune responses by
+inducing apoptosis of maternal T-cells
+\cite{Than.ea2009:primatesubfamilyofgalectins}. Indeed,
+\textit{LGALS14} has the highest expression of any galectin in the
+\human and \smonkey placenta
+(\cref{fig:hormone_genes,fig:galectins_lgals13_evolution_tree}), and
+other primate galectins such as \textit{LGALS13} and \textit{LGALS16}
+are also expressed in \human and \smonkey placenta. However, in the
+closely related \bonobo, \textit{LGALS14} is not found, and
+\textit{LGALS1} is the most abundant galectin. \textit{LGALS1} shows
+generally high expression in all placentas analyzed. \textit{LGALS1}
+is also involved in mediating maternal-fetal immune tolerance, and is
+expressed in both maternal and fetal tissue in primates
+\cite{Than.ea2008:Emergenceofhormonaland}. \textit{LGALS1} placental
+expression was thought to be confined to eutherian mammals, as there
+is evidence of adaptive evolution at this node, as well as the
+appearance of estrogen response elements in eutherian cis-regulatory
+regions \cite{Than.ea2008:Emergenceofhormonaland}. It was unclear as
+to whether galectins such as \textit{LGALS1} were important in the
+regulation of pregnancy in marsupials, as well as eutherians. Here we
+show substantial expression of galectins in the opossum, especially
+\textit{LGALS1} (Table 8). In addition, \textit{LGALS2} is also
 expressed. The presence of galectins at the maternal-fetal interface
-in the opossum implies that it was also present in the ancestral
-placenta of therian mammals. Indeed, adaptive evolution of LGALS1 on
+in \opossum implies that it was also present in the ancestral placenta
+of therian mammals. Indeed, adaptive evolution of \textit{LGALS1} on
 the eutherian stem is likely due to the increased invasive nature of
 eutehrian placentation rather than the origination of maternal-fetal
-tolerance. Although the function of LGALS1 in the marsupial placenta
-needs to be confirmed, it is likely that the mediation of
+tolerance. Although the function of \textit{LGALS1} in the marsupial
+placenta needs to be confirmed, it is likely that the mediation of
 maternal-fetal immune tolerance by galectins is an ancient feature of
@@
 
-The placenta of the opossum Monodelphis domestica is distinctly
-different from that of the term placenta of eutherian mammals in that
-it is derived from both the chorion and yolk sac. Indeed other
-differences between the placentas of marsupials and eutherians include
-lack of invasive extravillous trophoblasts in marsupials, the
-retention of a permeable shell coat for 2/3 of marsupial gestation,
-and the greater degree of imprinting in placenta-expressed genes in
-eutherians (Renfree, 2010; Renfree et al., 2013). In addition, the
-placenta of Monodelphis is also distinct from other marsupials, in
-that it is considered one of the least specialized, and thought be
-closer to the original marsupial condition (Renfree, 2010). In
-Monodelphis, the duration of direct connection between maternal and
-fetal tissues is as short as two days in a gestational period of 13-15
-days. During this period, the placenta is directly engaged in the
-exchange of urea with maternal tissues, owing to the lack of a
-vesicular allantois (Freyer et al., 2007). This differs drastically
+The placenta of \opossum is distinctly different from that of the term
+placenta of eutherian mammals in that it is derived from both the
+chorion and yolk sac. Other differences between the placentas of
+marsupials and eutherians include lack of invasive extravillous
+trophoblasts in marsupials, the retention of a permeable shell coat
+for 2/3 of marsupial gestation, and the greater degree of imprinting
+in placenta-expressed genes in eutherians
+\cite{Renfree2010:ReviewMarsupialsplacentalmammals,Renfree.ea2013:originandevolutionof}.
+In addition, the placenta of \opossum is also distinct from other
+marsupials, in that it is considered one of the least specialized, and
+thought be closer to the original marsupial condition
+\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. In \opossum, the
+duration of direct connection between maternal and fetal tissues is as
+short as two days in a gestational period of 13-15 days
+\cite{Harder.ea1993:Gestationandplacentationin}. During this period,
+the placenta is directly engaged in the exchange of urea with maternal
+tissues, owing to the lack of a vesicular allantois
+\cite{Freyer.ea2007:Placentalfunctionintwo}. This differs drastically
 with the situation in another well-studied marsupial, the tammar
-wallaby (Freyer et al., 2007). Therefore, we must interpret
-differences between the opossum and eutherian placentas with caution,
-as highly expressed genes and functional annotation clusters may
-represent didelphid or Monodelphis specific adaptations.
-
-Genes that show a significant shift on the opossum lineage include
-functional annotation clusters that correspond to active generation of
-energy, transport, and lipid breakdown via the peroxisome (Table 10;
-Supplemental Table X). The choriovitelline placenta of both didelphids
-and the tammar wallaby are known to be involved in glycolysis and
-protein synethsis (Freyer et al., 2003). This implies that the
-choriovitelline placenta in marsupials is an active site of metabolism
-and molecular synthesis, as well as the transport of nutrients from
-the maternal tissues. This is a distinct difference that is not seen
-in significant shifts on other lineages, and may be due to the
-formation of the placenta in the opossum from the yolk sac. In
-addition, the urea cycle is a highly enriched functional annotation
-cluster which directly correlates with the high levels of urea in the
-Monodelphis placenta (Freyer et al., 2007). This is potentially a
-Monodelphis-specific characteristic as discussed above.
+wallaby \cite{Freyer.ea2007:Placentalfunctionintwo}. Therefore, we
+must interpret differences between the opossum and eutherian placentas
+with caution, as highly expressed genes and functional annotation
+clusters may represent didelphid or \opossum specific adaptations.
+
+% Genes that show a significant shift on the \opossum lineage include
+% functional annotation clusters that correspond to active generation of
+% energy, transport, and lipid breakdown via the peroxisome (Table 10;
+% Supplemental Table X). The choriovitelline placenta of both didelphids
+% and the tammar wallaby are known to be involved in glycolysis and
+% protein synethsis (Freyer et al., 2003). This implies that the
+% choriovitelline placenta in marsupials is an active site of metabolism
+% and molecular synthesis, as well as the transport of nutrients from
+% the maternal tissues. This is a distinct difference that is not seen
+% in significant shifts on other lineages, and may be due to the
+% formation of the placenta in the opossum from the yolk sac. In
+% addition, the urea cycle is a highly enriched functional annotation
+% cluster which directly correlates with the high levels of urea in the
+% Monodelphis placenta (Freyer et al., 2007). This is potentially a
+% Monodelphis-specific characteristic as discussed above.
 
@@
 exist in eutherians or are not found in the term eutherian placenta
-(Table X). The existence of embryonic forms of hemoglobin in fetal
-tissues at term, as evidenced by the expression of HBE, HBZ, and HBO
-is a feature of marsupials that is generally due to their very short
-gestation periods. Embryonic hemoglobins in marsupials have a lower
-affinity for oxygen and allow for a lower rate of oxygen uptake from
-maternal tissues (Holland and Gooley, 1997). One novel uncharacterized
-gene that shows high expression (XM\_001364916.1) is not found in
-eutherians; Ensembl gene trees reveal that homologs of this gene exist
-in other amniotes such as birds and turtles, but are not present in
-eutherian mammals. The homolog of this gene in chicken is found to
-expressed in reproductive tract (Froman et al., 2006), but its
-function is unknown. WFDC2 is a member of the whey acidic protein
-(WAP) family and involved in innate immunity. It is expressed in the
-mammary gland of the tammar wallaby during pregnancy and early
-lactation (Sharp et al., 2007), and here in abundance in the opossum
-placenta.
+(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
+of hemoglobin in fetal tissues at term, as evidenced by the expression
+of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
+generally due to their very short gestation periods. Embryonic
+hemoglobins in marsupials have a lower affinity for oxygen and allow
+for a lower rate of oxygen uptake from maternal tissues
+\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
+uncharacterized gene that shows high expression (LOC10001195) is not
+found in eutherians; Ensembl gene trees reveal that homologs of this
+gene exist in other amniotes such as birds and turtles, but are not
+present in eutherian mammals. The homolog of this gene in chicken is
+found to expressed in reproductive tract
+\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
+unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
+family and involved in innate immunity. It is expressed in the mammary
+gland of the tammar wallaby during pregnancy and early lactation
+\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and here in
+relative abundance in the opossum placenta.
 

### 2016-01-12 manuscript/manuscript.Rnw
update authors list and affils

@@
 \begin{document}
-\author[1,2]{Michael R. McGowen}
 \author[1]{Don Armstrong}
+\author[2]{Michael R. McGowen}
 \author[1]{Amy Weckle}
-\author[1]{Jason Caravas}
-\author[3]{Dalen Agnew}
-\author[4]{Kurt Benirschke}
-\author[5]{Sue Savage-Rumbaugh}
-\author[6,7]{Chong Jai Kim}
-\author[8]{Günter P. Wagner}
-\author[6]{Roberto Romero}
-\author[1,6,9,*]{Derek E. Wildman}
+\author[3]{Jason Caravas}
+\author[4]{Dalen Agnew}
+\author[5]{Kurt Benirschke}
+\author[6]{Sue Savage-Rumbaugh}
+\author[7,8]{Chong Jai Kim}
+\author[9]{Günter P. Wagner}
+\author[7]{Roberto Romero}
+\author[1,6,10,*]{Derek E. Wildman}
 \title{The evolution of gene expression in the term placenta of viviparous mammals}
 % These are all wrong, but we'll change them later.
-\affil[1]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
+\affil[1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
 \affil[2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
-\affil[3]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
-\affil[4]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
-\affil[5]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
-\affil[6]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
-\affil[7]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
-\affil[8]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
-\affil[9]{Department of Obstetrics and Gynecology, Wayne State University School of Medicine, Detroit, MI, USA}
+\affil[3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
+\affil[4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
+\affil[5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
+\affil[6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
+\affil[7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
+\affil[8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
+\affil[9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
+\affil[10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
 \affil[*]{To whom correspondence should be addressed}

### 2016-01-12 manuscript/manuscript.Rnw
read vs read length plot is in supplemental

@@
 
-\begin{figure}
-<<reads.vs.read.length,fig.width=8,fig.height=8,echo=FALSE>>=
-print(ggplot(star.logs.wide,
-             aes(x=as.numeric(input.read.length),
-                 y=as.numeric(input.reads),
-                 color=capfirst(species),
-                 shape=capfirst(species)
-                 ))
-      + scale_shape_manual(values=1:nlevels(as.factor(star.logs.wide[,species])),
-                           guide=guide_legend(title="Species")
-                           )
-      + scale_color_discrete(guide=guide_legend(title="Species")
-                             )
-      + geom_point(size=5)
-      + xlab("Trimmed Read Length")
-      + ylab("Read number")
-      + scale_y_log10()
-      + theme(legend.text=element_text(face="italic")
-              )
-      )
-@ 
-\caption{RNA Read number as a function of trimmed read length}
-\label{fig:transcriptome_reads_vs_read_length}
-\end{figure}
-
 \Cref{fig:transcriptome_statistics} shows the RNAseq read and

### 2016-01-12 manuscript/manuscript.Rnw
fix monodelphis placenta ref

@@
 @ 
-\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica}}
+\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica_placenta}}
 \label{tab:top50_opossum_placenta}
@@
 \cref{tab:top50_opossum_placenta} with a list of all genes in
-Supplemental \cref{tab:all_monodelphis_domestica}. The gene with the
+Supplemental \cref{tab:all_monodelphis_domestica_placenta}. The gene with the
 second highest expression is superoxide dismutase 3, extracellular
@@
 %
-Another annexin, \textit{ANXA2} 
+% Another annexin, \textit{ANXA2} 
 %
@@
 \cite{Xiang.ea2013:PromoterhypomethylationofTIMP3} claims
-up-regulation)
+up-regulation) does things
 
@@
 \cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
-uncharacterized gene that shows high expression (LOC10001195) is not
+uncharacterized gene that shows high expression (\textit{LOC10001195}) is not
 found in eutherians; Ensembl gene trees reveal that homologs of this

### 2016-01-12 manuscript/supplemental_material.Rnw
currently transcript mapping is in the main text

@@
 
-\begin{figure}
-<<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
-star.logs.unmapped <- 
-    star.logs[grepl("%",field) & 
-                  grepl("unmapped",field) & 
-                      !grepl("too many mismatches",field),]
-star.logs.unmapped <- 
-    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
-pushViewport(viewport(layout=grid.layout(ncol=1,nrow=2,heights=c(2,1))))
-print(ggplot(star.logs.unmapped,
-             aes(x=capfirst(species),y=as.numeric(gsub("%","",value))))
-      + geom_boxplot()
-      + facet_grid(field~.,scales="free")
-      + ylab("Percent of Reads")
-      + xlab(NULL)
-      + theme_bw()
-      + theme(legend.position="top",
-              plot.margin=unit(c(0,0,0,1.1),"strwidth",data=list("","","","e8")),
-              axis.title.x=element_blank(),
-              axis.text.x = element_blank()),#element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
-      vp=viewport(layout.pos.col=1,layout.pos.row=1)
-      )
-pushViewport(viewport(layout.pos.col=1,layout.pos.row=1))
-grid.text("A",
-          y=unit(1,"npc")-unit(0.1,"cm"),
-          x=unit(0,"npc")+unit(0.1,"cm"),
-          gp=gpar(fontsize=24),
-          just=c("left","top"))
-popViewport()          
-print(ggplot(star.logs[grepl("input reads$",field),],
-             aes(x=capfirst(species),y=as.numeric(value)))
-      + geom_boxplot()
-      + scale_y_log10()
-      + ylab("Reads")
-      + xlab("Species")
-      + theme_bw()
-      + theme(legend.position="top",
-              plot.margin=unit(c(0,1.6,0,0),"strheight",data=list("","TO","","")),
-              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
-      vp=viewport(layout.pos.col=1,layout.pos.row=2))
-pushViewport(viewport(layout.pos.col=1,layout.pos.row=2))
-grid.text("B",
-          y=unit(1,"npc")-unit(0.1,"cm"),
-          x=unit(0,"npc")+unit(0.1,"cm"),
-          gp=gpar(fontsize=24),
-          just=c("left","top"))
-popViewport(2)
-@ 
-\caption{RNAseq read and alignment statistics for all species. A)
-  Boxplot of percentages of unmapped reads in each species; 0\% of
-  reads in all species were unmapped because of mismatches. B) Boxplot
-  of number of RNAseq reads in each species.}
-\label{fig:transcriptome_statistics}
-\end{figure}
+% \begin{figure}
+% <<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
+% star.logs.unmapped <- 
+%     star.logs[grepl("%",field) & 
+%                   grepl("unmapped",field) & 
+%                       !grepl("too many mismatches",field),]
+% star.logs.unmapped <- 
+%     star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
+% pushViewport(viewport(layout=grid.layout(ncol=1,nrow=2,heights=c(2,1))))
+% print(ggplot(star.logs.unmapped,
+%              aes(x=capfirst(species),y=as.numeric(gsub("%","",value))))
+%       + geom_boxplot()
+%       + facet_grid(field~.,scales="free")
+%       + ylab("Percent of Reads")
+%       + xlab(NULL)
+%       + theme_bw()
+%       + theme(legend.position="top",
+%               plot.margin=unit(c(0,0,0,1.1),"strwidth",data=list("","","","e8")),
+%               axis.title.x=element_blank(),
+%               axis.text.x = element_blank()),#element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
+%       vp=viewport(layout.pos.col=1,layout.pos.row=1)
+%       )
+% pushViewport(viewport(layout.pos.col=1,layout.pos.row=1))
+% grid.text("A",
+%           y=unit(1,"npc")-unit(0.1,"cm"),
+%           x=unit(0,"npc")+unit(0.1,"cm"),
+%           gp=gpar(fontsize=24),
+%           just=c("left","top"))
+% popViewport()          
+% print(ggplot(star.logs[grepl("input reads$",field),],
+%              aes(x=capfirst(species),y=as.numeric(value)))
+%       + geom_boxplot()
+%       + scale_y_log10()
+%       + ylab("Reads")
+%       + xlab("Species")
+%       + theme_bw()
+%       + theme(legend.position="top",
+%               plot.margin=unit(c(0,1.6,0,0),"strheight",data=list("","TO","","")),
+%               axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
+%       vp=viewport(layout.pos.col=1,layout.pos.row=2))
+% pushViewport(viewport(layout.pos.col=1,layout.pos.row=2))
+% grid.text("B",
+%           y=unit(1,"npc")-unit(0.1,"cm"),
+%           x=unit(0,"npc")+unit(0.1,"cm"),
+%           gp=gpar(fontsize=24),
+%           just=c("left","top"))
+% popViewport(2)
+% @ 
+% \caption{RNAseq read and alignment statistics for all species. A)
+%   Boxplot of percentages of unmapped reads in each species; 0\% of
+%   reads in all species were unmapped because of mismatches. B) Boxplot
+%   of number of RNAseq reads in each species.}
+% \label{fig:transcriptome_statistics}
+% \end{figure}
 

### 2016-01-12 manuscript/manuscript.Rnw
update conclusion section

@@
 We found evidence of high expression in two novel expansions of genes
-in the elephant genome. One family includes 11 genes that are related
-to the gene PRSS16 gene, a thymus-specific serine protease that may be
-involved regulating the presentation of self-peptides in CD4+
-T-lymphocytes (Fornari et al., 2012). All 11 of these genes are
-expressed in the elephant placenta, three of which have very abundant
-mRNAs (Table 6). It is possible that this elephant-specific gene
-family may be involved in immunity in the elephant placenta and the
-mediation of maternal-fetal immune tolerance; however, future work
-will need to test these hypotheses. Three other genes abundantly
-expressed in the elephant placenta are homologues of SSLP1, a gene
-with predominant expression in the seminal vesicle in mouse (Li et
-al., 2006). Homologues of SSLP1 have been identified as being
-expressed in the placental villi of cow, sheep, and goat. SSLP1 also
-contains an Alu retrotransposon and may be involved in the development
-of villi via cell proliferation (Ushizawa et al., 2009, 2010). We
-verified expression in cow, however its abundant expression was also
-identified in dog. This gene is not found in primates. Expansion of
-these genes in elephants may indicate an elaboration of this system.
+in the \elephant genome. One family includes 11 genes that are related
+to the \textit{PRSS16} gene, a thymus-specific serine protease that
+may be involved regulating the presentation of self-peptides in CD4+
+T-lymphocytes \cite{Fornari.ea2012:TranscriptionprofilingofPrss16}.
+All 11 of these genes are expressed in the elephant placenta, three of
+which have very abundant mRNAs
+(\cref{tab:top50_elephant_placenta,fig:prss16_evolution_tree}). It is
+possible that this elephant-specific gene family may be involved in
+immunity in the elephant placenta and the mediation of maternal-fetal
+immune tolerance; however, future work will need to test these
+hypotheses. Three other genes abundantly expressed in the elephant
+placenta are homologues of \textit{SSLP1}, a gene with predominant
+expression in the seminal vesicle in mouse
+\cite{Li.ea2006:SSLP-1secretedLy-6protein}. Homologues of
+\textit{SSLP1} are also expressed in \cow, \dog, \horse, and \sheep.
+\textit{SSLP1} also contains an Alu retrotransposon and may be
+involved in the development of villi via cell proliferation
+\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}. This
+gene is not found in primates. Expansion of these genes with high
+expression in \elephant may indicate an elaboration of this system.
+
+\subsection{Annexin Complex}
 
@@
 
-The human placenta is distinctive from many other primates in that it
+The \human placenta is distinctive from many other primates in that it
 possesses a deeply invasive trophoblast that penetrates the inner
-myometrium of the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
+myometrium of the uterus
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
 hypothesized that this feature is related to the increased demand for
 resources by the developing fetus to feed the expanded size of the
-human brain (Carter and Pijnenborg, 2011; Robillard et al., 2002).
-Humans have the largest brain in terms of relative size and this
-differs markedly from closely related great apes (Boddy et al., 2012).
-Deep penetration of the myometrium occurs in chimpanzees and gorillas,
-but this penetration may be accelerated in humans (Carter and
-Pijnenborg, 2011).
+human brain
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation,Robillard.ea2002:Evolutionaryadaptationstopre-eclampsiaeclampsia}.
+\human have the largest brain in terms of relative size and this
+differs markedly from closely related great apes
+\cite{Boddy.ea2012:Comparativeanalysisofencephalization}. Deep
+penetration of the myometrium occurs in chimpanzees and gorillas, but
+this penetration may be accelerated in humans
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
 
@@
 preeclampsia is a human disease that occurs in 4\% of pregnancies in
-the United States. Preeclampsia can be attributed to the lack of
-adequate invasion of maternal tissue, and is noted by the onset of
-hypertension and proteinuria (citation). Some cases of preeclampsia
-can develop into eclampsia, a more severe condition resulting in
-seizures during pregnancy and death. Lack of invasion in preeclampsia
-may be attributed to the inefficiency of growth regulators as well as
-the inability to sufficiently suppress the maternal immune response.
-Although there are speculative cases of eclampsia in great apes
-(Baird, 1981; Stout and Lemmon, 1969; Thornton and Onwude, 1992), it
-is exceedingly rare in these species \cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It
-is possible that a comparative evolutionary approach to gene
-expression can identify genes that have divergent expression patterns
-from other mammals as well as association with preeclampsia.
+the United States
+\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Preeclampsia
+can be attributed to the lack of adequate invasion of maternal tissue,
+and is noted by the onset of hypertension and proteinuria
+\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
+preeclampsia can develop into eclampsia, a more severe condition
+resulting in seizures during pregnancy and death. Lack of invasion in
+preeclampsia may be attributed to the inefficiency of growth
+regulators as well as the inability to sufficiently suppress the
+maternal immune response. Although there are speculative cases of
+eclampsia in great apes
+\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
+it is exceedingly rare in these species
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is possible
+that a comparative evolutionary approach to gene expression can
+identify genes that have divergent expression patterns from other
+mammals as well as association with preeclampsia.
 
@@
 and increased blood flow, both of which may correlate with the
-invasion of the trophoblast (Table 8; Supplemental Table X). In
+invasion of the trophoblast (\cref{tab:go_terms_all_species}). In
 addition, clusters such as cytokines and defense response
-(Supplemental Table X) also indicate an a significant change in
-expression of the immune response that may be involved in
+(\cref{tab:go_terms_all_species}) also indicate an a significant
+change in expression of the immune response that may be involved in
 immunosupression at the maternal-fetal boundary to further allow
@@
 
-Many genes that both show expression shifts on the human lineage and
-have very high expression (FPKM>200) are directly related to the
-highly invasive nature of the human placenta. For example, GPC3,
-glypican 3, is an imprinted gene and cell growth regulator, associated
-with Simpson-Golabi-Behmel fetal overgrowth syndrome, and involved in
-intrauterine growth restriction (Monk and Moore, 2004; Pilia et al.,
-1996). GPC3 has the highest expression in humans (Table 2), but is
-also expressed in other mammals to a lesser degree, especially those
-with invasive hemochorial placentas like bonobos (\ac{FPKM}= 199.924),
-spider monkeys (\ac{FPKM}= 96.24), and the mouse (\ac{FPKM}= 80.7545)
-(Supplemental Tables X-XX). ADAM12 increases the availability of
-insulin-like growth factors that lead to the promotion of cell growth
-and placental invasion (Laigaard et al., 2005). (Hou et al., 2009)
-found evidence of adaptive evolution in ADAM12 on the primate lineage,
-and like GPC3 it is expressed in the bonobo (\ac{FPKM}= 46.1553), spider
-monkey (\ac{FPKM}=30.41) and mouse (\ac{FPKM}=2.25391) as well, although to a
-lesser degree (Supplemental Tables X-XX). RHOBTB3 is imprinted and
-associated with trophoblast invasion as well (Bilban et al., 2010;
-Yuen et al., 2011). In addition, one other gene, KISS1, shows
-exceptional abundance in the human placenta (Table 2); however due to
-its absence in elephant Ensmebl gene annotations, it was not part of
-the 1:1 orthologues examined in the comparative study. Here, it is
-only expressed in the term placentas of humans (Table 2) and bonobos
-(\ac{FPKM}=17.6394; Supplemental Table X). KISS1 is thought to be involved
-in the regulation of trophoblast invasion, and shows highest
-expression in the first trimester of pregnancy (Bilban et al., 2004;
-Reynolds et al., 2009).
-
-In addition, several other genes that show expression shifts on the
-human lineage also have major effects on placental function. Genes
-such as TXK and EBI3 are intimately involved in the immune system in
-the placenta. TXK is involved in natural killer cell differentiation
-(Bezman et al., 2012). EBI3 is implicated in natural killer cell
-function an maternal-fetal immunomodulation in the syncytiotrophoblast
-(Devergne et al., 2001). KYNU and KMO are interacting genes that are
-involved in tryptophan metabolism, which may be involved in
-immunosupression and the prevention of fetal rejection (Takikawa,
-2005). THSD7A is a thrombospondin, which inhibits endothelial cell
-migration and tube morphogenesis (Wang et al., 2010).
-
-Many genes that show evidence of a shift on the human lineage show
-evidence of implication in preeclampsia, including ADAM12 (Gack et
-al., 2005; Laigaard et al., 2005) MAOA (Carrasco et al., 2000;
-Bottalico et al., 2004), BCL2 (Ishihara et al., 2002), GPC3 (Gui et
-al., 2012; Ogawa, 2007), TMEM8A, CLIC3, EBI3, FLT1, BCL6, SYDE1, and
-MAN1C1 (Enquobahrie et al., 2008; Winn et al., 2009; Nishizawa, 2011).
-In addition KISS1 is also involved in preeclampsia as well (Sado et
-al., 2011; Qiao et al, 2012). Due to the association of many genes
-that show human-specific shifts as well as high expression, we
-conclude that preeclampsia is possibly a symptom of recent evolution
-of gene expression on the human lineage. Although many genes such as
-GPC3 and ADAM12 show distinctive relatively high expression in bonobo,
-this expression has intensified on the human lineage. Sequencing of
-more placentas of nonhuman primates, including other individuals of
-the genus Pan will help to further characterize this transition.
+% I haven't actually done this analysis, so commented out
+
+% Many genes that both show expression shifts on the human lineage and
+% have very high expression (FPKM>200) are directly related to the
+% highly invasive nature of the human placenta. For example, GPC3,
+% glypican 3, is an imprinted gene and cell growth regulator, associated
+% with Simpson-Golabi-Behmel fetal overgrowth syndrome, and involved in
+% intrauterine growth restriction (Monk and Moore, 2004; Pilia et al.,
+% 1996). GPC3 has the highest expression in humans (Table 2), but is
+% also expressed in other mammals to a lesser degree, especially those
+% with invasive hemochorial placentas like bonobos (\ac{FPKM}= 199.924),
+% spider monkeys (\ac{FPKM}= 96.24), and the mouse (\ac{FPKM}= 80.7545)
+% (Supplemental Tables X-XX). ADAM12 increases the availability of
+% insulin-like growth factors that lead to the promotion of cell growth
+% and placental invasion (Laigaard et al., 2005). (Hou et al., 2009)
+% found evidence of adaptive evolution in ADAM12 on the primate lineage,
+% and like GPC3 it is expressed in the bonobo (\ac{FPKM}= 46.1553), spider
+% monkey (\ac{FPKM}=30.41) and mouse (\ac{FPKM}=2.25391) as well, although to a
+% lesser degree (Supplemental Tables X-XX). RHOBTB3 is imprinted and
+% associated with trophoblast invasion as well (Bilban et al., 2010;
+% Yuen et al., 2011). In addition, one other gene, KISS1, shows
+% exceptional abundance in the human placenta (Table 2); however due to
+% its absence in elephant Ensmebl gene annotations, it was not part of
+% the 1:1 orthologues examined in the comparative study. Here, it is
+% only expressed in the term placentas of humans (Table 2) and bonobos
+% (\ac{FPKM}=17.6394; Supplemental Table X). KISS1 is thought to be involved
+% in the regulation of trophoblast invasion, and shows highest
+% expression in the first trimester of pregnancy (Bilban et al., 2004;
+% Reynolds et al., 2009).
+% 
+% In addition, several other genes that show expression shifts on the
+% human lineage also have major effects on placental function. Genes
+% such as TXK and EBI3 are intimately involved in the immune system in
+% the placenta. TXK is involved in natural killer cell differentiation
+% (Bezman et al., 2012). EBI3 is implicated in natural killer cell
+% function an maternal-fetal immunomodulation in the syncytiotrophoblast
+% (Devergne et al., 2001). KYNU and KMO are interacting genes that are
+% involved in tryptophan metabolism, which may be involved in
+% immunosupression and the prevention of fetal rejection (Takikawa,
+% 2005). THSD7A is a thrombospondin, which inhibits endothelial cell
+% migration and tube morphogenesis (Wang et al., 2010).
+% 
+% Many genes that show evidence of a shift on the human lineage show
+% evidence of implication in preeclampsia, including ADAM12 (Gack et
+% al., 2005; Laigaard et al., 2005) MAOA (Carrasco et al., 2000;
+% Bottalico et al., 2004), BCL2 (Ishihara et al., 2002), GPC3 (Gui et
+% al., 2012; Ogawa, 2007), TMEM8A, CLIC3, EBI3, FLT1, BCL6, SYDE1, and
+% MAN1C1 (Enquobahrie et al., 2008; Winn et al., 2009; Nishizawa, 2011).
+% In addition KISS1 is also involved in preeclampsia as well (Sado et
+% al., 2011; Qiao et al, 2012). Due to the association of many genes
+% that show human-specific shifts as well as high expression, we
+% conclude that preeclampsia is possibly a symptom of recent evolution
+% of gene expression on the human lineage. Although many genes such as
+% GPC3 and ADAM12 show distinctive relatively high expression in bonobo,
+% this expression has intensified on the human lineage. Sequencing of
+% more placentas of nonhuman primates, including other individuals of
+% the genus Pan will help to further characterize this transition.
 
@@
 
-Here we sequenced six placentas spanning the phylogenetic tree of
-mammals. This revealed a dynamic expression pattern of expression of
-lineage-specific expansions of gene families and differential gene
-expression that agreed with the scenario of an “arms race” between
-maternal and fetal tissues that have shaped such disparate gene
-expression profiles. In addition, we highlighted the broad spectrum of
-hormone gene expression in multiple species that suggested that there
-was a lack of conservation in hormone signaling during mammalian
-pregnancy. Major differences between expression in marsupial and
-eutherian mammals at term seemed to be correlated with the differences
-between choriovitelline and chorioamniotic placenta, with the
-marsupial placenta more enriched in genes the contributed to the
-active generation of energy. We also identified that galectin
-expression, especially that of LGALS1, is a common feature of all
-mammalian placentas and was most likely present in the ancestral
-therian. Also, we identified two new expansions of genes that are
-highly expressed in the elephant placenta hat may be involved in
-trophoplast differentiation and immune system function, mirroring
-expansions in other groups such as primates, rodents, and ruminants.
-We also quantified significant lineage-specific shifts in gene
-expression, concentrating on the identification of genes that showed
-gene expression changes on the human lineage. We cautiously propose
-that many of these genes may be related to the invasive nature of the
-trophoblast in humans, although we note that in many of these genes,
-there is in some cases substantial expression in the bonobo as well.
-These genes have the potential to spur further research and may be at
-the root of obstetrical disorders such as preeclampsia. We propose
-that more placentas from diverse lineages, especially those of
-marsupials and primates, to further identify genes and expression
-levels that are specific to marsupials and eutehrians, as well as
-confirm the identity of human-specific expression shifts.
+We sequenced \Sexpr{star.logs[,length(unique(species))]} placentas
+spanning the phylogenetic tree of mammals. This revealed a dynamic
+expression pattern of expression of lineage-specific expansions of
+gene families and differential gene expression that agreed with the
+scenario of an “arms race” between maternal and fetal tissues that
+have shaped such disparate gene expression profiles. In addition, we
+highlighted the broad spectrum of hormone gene expression in multiple
+species that suggested that there is a lack of conservation in hormone
+signaling during mammalian pregnancy. Major differences between
+expression in marsupial and eutherian mammals at term seemed to be
+correlated with the differences between choriovitelline and
+chorioamniotic placenta, with the marsupial placenta more enriched in
+genes the contributed to the active generation of energy. We also
+identified that galectin expression, especially that of
+\textit{LGALS1}, is a common feature of all mammalian placentas and
+was most likely present in the ancestral therian. Also, we identified
+two new expansions of genes that are highly expressed in the \elephant
+placenta hat may be involved in trophoplast differentiation and immune
+system function, mirroring expansions in other groups such as
+primates, rodents, and ruminants. We also quantified significant
+lineage-specific shifts in gene expression, concentrating on the
+identification of genes that showed gene expression changes on the
+human lineage. We cautiously propose that many of these genes may be
+related to the invasive nature of the trophoblast in \human, although
+we note that in many of these genes, there is in some cases
+substantial expression in the \bonobo as well. These genes have the
+potential to spur further research and may be at the root of
+obstetrical disorders such as pre-eclampsia. We propose that more
+placentas from diverse lineages, especially those of marsupials and
+primates, to further identify genes and expression levels that are
+specific to marsupials and eutherians, as well as confirm the identity
+of human-specific expression shifts.
 

### 2016-01-14 manuscript/manuscript.Rnw
add \DEW command for derek
fix broken Igf2 in mouse
cite other PRL papers; include HOXA11 gene tree

@@
 \newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
+\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
 \usepackage[x11names,svgnames]{xcolor}
@@
 human \textit{H19\_2}),
-\Sexpr{gene.expression("mus_musculus","IGF2",complete=TRUE)}, numerous
+\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
 cathepsins and a rodentia specific expansion of prolactin related
@@
 The elephant is strikingly the only mammal analyzed that shows high
-expression of the primarily pituitary hormone prolactin (\textit{PRL}) in the
-placenta. All afrotherians yet analyzed contain only one prolactin
-gene, and prolactin has undergone adaptive evolution on the elephant
-lineage \cite{Wallis2009:ProlactininAfrotheriacharacterization}.
+expression of the primarily pituitary hormone prolactin (\textit{PRL})
+in the placenta, although we see high expression of genes which
+resulted from duplications of \textit{PRL} in the \mouse, \carmeli,
+and \cow (\cref{fig:prolactin_evolution_tree}). All afrotherians yet
+analyzed contain only one prolactin gene, and prolactin has undergone
+adaptive evolution on the \elephant lineage
+\cite{Wallis2009:ProlactininAfrotheriacharacterization}.
 \citet{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
@@
 using immunohistological staining and human derived antibodies. A
-previous analysis of the elephant placenta also detected the
-expression of prolactin
-\cite{Hou.ea2012:Elephanttranscriptomeprovidesinsights}. Here we
-confirm that the sole prolactin in the elephant is not only expressed
-but highly expressed in the placenta, and likely functions both in the
-maintenance of the corpus luteum and as a placental lactogen
-\cite{Yamamoto.ea2011:PlacentationinAfricanelephant}. \textit{PRL} is
-not expressed in the opossum, even though there is evidence that it is
-expressed in the placenta of another marsupial, the tammar wallaby
+previous analysis of \elephant placenta also detected the expression
+of prolactin \cite{Hou.ea2012:Elephanttranscriptomeprovidesinsights}.
+Here we confirm that the sole prolactin in \elephant is not only
+expressed but highly expressed in the placenta, and likely functions
+both in the maintenance of the corpus luteum and as a placental
+lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}.
+\textit{PRL} is not expressed in \opossum, even though there is
+evidence that it is expressed in the placenta of another marsupial,
+\textit{Macropus eugenii} (tammar wallaby)
 \cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests
 that placental prolactin expression in marsupials may be as diverse as
-in eutherian mammals. Although, \textit{PRL} is not expressed to any
-great degree in the placenta other mammals, we see very high
-expression of genes which resulted from duplications of \textit{PRL}
-in the \mouse, \galili, \carmeli, and \cow
-(\cref{fig:prolactin_evolution_tree}).
+in eutherian mammals. One possible explanation of the diversity of
+expression is the presence of endogenous retroviruses such as MER39 in
+primates, MER77 in rodents, and L1-2\_LA in \elephant
+\cite{Emera.ea2012:Convergentevolutionofendometrial} although this
+does not explain the diversity of expression of different isoforms of
+\textit{PRL} in \mouse or their tissue specificity in primates. An
+alternative, that \textit{HOXA11} expression is responsible for
+\textit{PRL} expression
+\cite{Lynch.ea2008:Adaptivechangesintranscription} seems not to be the
+case in placenta, as the rodent with the highest \textit{HOXA11}
+expression, \galili
+(\Sexpr{gene.expression("nannospalax_galili","Hoxa11",complete=TRUE)})
+is the rodent with the lowest expression of \textit{PRL}
+(\Sexpr{gene.expression("nannospalax_galili","Prl",complete=TRUE)},
+\cref{fig:prolactin_evolution_tree}), and \elephant also does not
+express \textit{HOXA11} highly either
+(\Sexpr{gene.expression("loxodonta_africana","HOXA11",complete=TRUE)}).
 
@@
 \label{fig:prolactin_receptor_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<hoxa11.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="HOXA11" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSBTAG00000014738","ENSECAG00000013229")
+                 )
+@ 
+\caption{\genetreeexplanation{\textit{HOXA11}}}
+\label{fig:hoxa11_evolution_tree}
 \end{figure}

### 2016-01-14 manuscript/manuscript.Rnw
cite GH1 gene paper

@@
 between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
-reliably distinguish transcripts between the two). All but
-\textit{GH1} show substantial expression in the \human placenta
-(\cref{fig:csh1_evolution_tree}). There is some evidence that
-increased expression of these genes may be correlated with overall
-birth weight of offspring in \human
+reliably distinguish transcripts between the two). All show
+substantial expression in the \human placenta
+(\cref{fig:csh1_evolution_tree}). The expression of the \ac{GH} family
+in placenta appears to be primate specific, and likely existed at the
+time of the last common ancestor between all extant anthropoids
+\cite{Papper.ea2009:Ancientoriginofplacental}. There is some evidence
+that increased expression of these genes may be correlated with
+overall birth weight of offspring in \human
 \cite{Männik.ea2010:Differentialexpressionprofileof}, and it is

### 2016-01-19 manuscript/manuscript.Rnw
use PSG and PAG acronyms; update introduction

@@
 % showed exceptionally high expression in humans; these genes are
-% implicated in preeclampsia and may indicate that the risk of
-% preeclampsia recently emerged on the human lineage.
+% implicated in pre-eclampsia and may indicate that the risk of
+% pre-eclampsia recently emerged on the human lineage.
 
@@
 specific aspects of placental morphology
-\cite{Capellini.ea2011:Placentationandmaternalinvestment}.
+\cite{Capellini.ea2011:Placentationandmaternalinvestment}
+(\cref{tab:gestational_periods}).
 
@@
 genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
-Studies of certain model species (i.e., humans, mice, cattle) have
+Studies of certain model species (i.e., \human, \mouse, \cow) have
 revealed a stunning divergence in the genetics of the placenta,
@@
 \cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
-These include independent expansions of the prolactins in mouse and
-cow, pregnancy-specific glycoproteins (PSGs) in humans and mouse,
-pregnancy-associated glycoproteins (PAGs) in cow, and the expansion of
-the growth hormone and chorionic gonadotropin beta genes in hominoid
-primates
+These include independent expansions of the prolactins in \mouse,
+\carmeli, \galili, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
+in \cow, genes related to the thymus-specific serine protease
+\textit{PRSS16} in \elephant, and the expansion of the growth hormone
+and chorionic gonadotropin beta genes in hominoid primates
 \cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
@@
 species. For example, the expression of orthologous genes in the
-placenta have been found to vary substantially between humans and mice
-\cite{Su.ea2002:Large-scaleanalysisofhuman}, and the recruitment of
-\textit{SIGLEC6} to placental expression in humans differs from both
-chimpanzees and mice
+placenta have been found to vary substantially between \human and
+\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
+recruitment of \textit{SIGLEC6} to placental expression in \human
+differs from both \textit{P. troglodytes} and \mouse
 \cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
@@
 extremes can set up an intricate “arms race”, involving the imprinting
-of parental alleles, production of new genes, and enlistment of
+of parental alleles, expansion of gene families, and enlistment of
 existing genes to expression at the maternal-fetal interface
@@
 One symptom of this maternal-fetal conflict is its potential affect on
-human health
+\human health
 \cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
-addition to having an invasive hemochorial placenta, humans possess a
+addition to having an invasive hemochorial placenta, \human possess a
 deeply invasive trophoblast layer that can penetrate the myometrium of
@@
 Defects in placental growth are known to be associated with the
-devastating condition of preeclampsia, which can cause hypertension,
+devastating condition of pre-eclampsia, which can cause hypertension,
 proteinuria, and has the potential to lead to death of the mother
 and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
-Although cases have been documented in chimpanzees
+Although cases have been documented in \textit{P. troglodytes}
 \cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
-gorillas \cite{Thornton.ea1992:Convulsionsinpregnancyin}, preeclampsia
-has a far higher incidence in human pregnancies with as much as 4\% of
+\textit{Gorilla gorilla}
+\cite{Thornton.ea1992:Convulsionsinpregnancyin}, pre-eclampsia has a
+far higher incidence in \human pregnancies with as much as 4\% of
 pregnancies being affected in the United States
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Chimpanzees
-and gorillas also have deeply invasive trophoblasts, although the
-progression of invasion in great apes may be slower than humans
+\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. \textit{P.
+  troglodytes} and \textit{G. gorilla} also have deeply invasive
+trophoblasts, although the progression of invasion in great apes may
+be slower than \human
 \cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however,
@@
 placentas at term, including that of the \human, \bonobo (the sister
-species to the chimpanzee), \smonkey, \mouse, \galili, \carmeli, \dog,
-\cow, \elephant, \armadillo, and \opossum, representing marsupials as
-well as three major branches of eutherian mammals (Euarchontoglires,
-Laurasiatheria, Afrotheria). These placentas also represent a wide
-array of placental morphologies as well as traits such as number and
-size of offspring and length of gestation
-\cref{tab:gestational_periods}. We use this data, along with existing
-RNA-Seq data from \mouse \cite{Wang.ea2011:surveyfornovelimprinted},
-\human \cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
-\horse \cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
+species to \textit{P. troglodytes}), \smonkey, \mouse, \galili,
+\carmeli, \dog, \cow, \elephant, \armadillo, and \opossum. These
+species represent three major branches of eutherian mammals
+(Euarchontoglires, Laurasiatheria, Afrotheria) as well as marsupials.
+These placentas also represent a wide array of placental morphologies
+as well as traits such as number and size of offspring and length of
+gestation (\cref{tab:gestational_periods}). We use this data, along with
+existing RNA-Seq data from \mouse
+\cite{Wang.ea2011:surveyfornovelimprinted}, \human
+\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \horse
+\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
 \cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
@@
 genes such as hormones and galectins, identify novel lineage-specific
-expression patterns, and quantify human specific differences in gene
+expression patterns, and quantify \human specific differences in gene
 expression that may be related to obstetrical disorders such as
-preeclampsia.
+pre-eclampsia.
 
@@
 % 
-Eleven pregnancy-specific beta-1-glycoprotein genes are highly-expressed
+Eleven \acp{PSG} are highly-expressed
 (\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
@@
 almost exclusively symptomatic of human pregnancy. For example,
-preeclampsia is a human disease that occurs in 4\% of pregnancies in
+pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
 the United States
@@
 \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
-preeclampsia can develop into eclampsia, a more severe condition
+pre-eclampsia can develop into eclampsia, a more severe condition
 resulting in seizures during pregnancy and death. Lack of invasion in
-preeclampsia may be attributed to the inefficiency of growth
+pre-eclampsia may be attributed to the inefficiency of growth
 regulators as well as the inability to sufficiently suppress the
@@
 identify genes that have divergent expression patterns from other
-mammals as well as association with preeclampsia.
+mammals as well as association with pre-eclampsia.
 
@@
 % Many genes that show evidence of a shift on the human lineage show
-% evidence of implication in preeclampsia, including ADAM12 (Gack et
+% evidence of implication in pre-eclampsia, including ADAM12 (Gack et
 % al., 2005; Laigaard et al., 2005) MAOA (Carrasco et al., 2000;
@@
 % MAN1C1 (Enquobahrie et al., 2008; Winn et al., 2009; Nishizawa, 2011).
-% In addition KISS1 is also involved in preeclampsia as well (Sado et
+% In addition KISS1 is also involved in pre-eclampsia as well (Sado et
 % al., 2011; Qiao et al, 2012). Due to the association of many genes
 % that show human-specific shifts as well as high expression, we
-% conclude that preeclampsia is possibly a symptom of recent evolution
+% conclude that pre-eclampsia is possibly a symptom of recent evolution
 % of gene expression on the human lineage. Although many genes such as

### 2016-01-20 manuscript/manuscript.Rnw
discuss go results for core placenta transcriptome

@@
 
+\begin{table}
+\footnotesize
 <<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=
-placenta.go.results <- rbindlist(list(placenta_go_results_mf[classic.fdr < 0.05,],
-                             placenta_go_results_bp[classic.fdr < 0.05,]))
+placenta.go.results <- rbindlist(list(head(n=20,placenta_go_results_mf[classic.fdr < 0.05,]),
+                             head(n=20,placenta_go_results_bp[classic.fdr < 0.05,])))
 placenta.go.results <- placenta.go.results[,Annotated:=NULL]
+setnames(placenta.go.results,"Significant","# Core")
 setnames(placenta.go.results,"classic","p value")
@@
              digits=2,
-             display=c("d","s","s","d","f","e","e","s"),
-             caption="Top go MF results",
-             label="tab:core_placenta_go_terms"),
+             display=c("d","s","s","d","f","e","e","s")),
       include.rownames=FALSE,
-      floating=TRUE,
       booktabs=TRUE,
-      size="\\tiny"
+      floating=FALSE
       )
 @ 
+\caption{\ac{GO} terms most significantly enriched
+  (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]},
+  \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP)
+  in the core placenta transcriptome as compared to all genes
+  with 1:1 human orthologs in all species. “# Core” is the
+  number of genes in the core transcriptome for each \ac{GO} term;
+  “Expected” is the number of genes expected in the core
+  transcriptome for each term; “p-value” are the odds of this
+  occurring by chance calculated by the Fisher exact test for
+  each term; “FDR” is the false discovery rate as estimated by
+  the Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure}}
+\label{tab:core_placenta_go_terms}
+\end{table}
 
@@
 % http://www.proteinatlas.org/ENSG00000149257-SERPINH1/tissue/placenta
+
+\Cref{tab:core_placenta_go_terms} lists the significant enrichment of
+\ac{GO} terms in the core placenta transcriptome; notable terms
+include those involved in cytoskeletal structure, protein transport,
+and membrane protein structure.
 

### 2016-01-20 manuscript/manuscript.Rnw
discuss core placenta transcriptome
+ make 'the annexin complex' annexin complexes, since there's many
possible annexin complexes.

@@
 \cref{tab:top50_spider_monkey_placenta}), and multiple components of
-the annexin complex
+annexin complexes
 (\Sexpr{gene.expression("ateles_fusciceps","ANXA2",complete=TRUE)}, %
@@
 genes in Supplemental \cref{tab:all_nannospalax_galili_placenta}.
-Components of the annexin complex are highly expressed
+Components of annexin complexes are highly expressed
 (\Sexpr{gene.expression("nannospalax_galili","S100a6",complete=TRUE)},
@@
 (\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
-\dog{}, members of the annexin complex were also highly expressed
+\dog{}, members of annexin complexes were also highly expressed
 (\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
@@
 \Sexpr{gene.expression("sus_scrofa","IGFBP3",complete=TRUE)}), members
-of the annexin complex
+of annexin complexes
 (\Sexpr{gene.expression("sus_scrofa","S100A1",complete=TRUE)},
 \Sexpr{gene.expression("sus_scrofa","ANXA2",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","S100A12",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","S100A10",complete=TRUE)}, and
-\Sexpr{gene.expression("sus_scrofa","S100A1",complete=TRUE)}), the
+\Sexpr{gene.expression("sus_scrofa","S100A12",complete=TRUE)}, and
+\Sexpr{gene.expression("sus_scrofa","S100A10",complete=TRUE)}), the
 galectin \Sexpr{gene.expression("sus_scrofa","LGALS3",complete=TRUE)}
@@
 \Sexpr{gene.expression("equus_caballus","IGFBP3",complete=TRUE)}) and
-multiple members of the annexin complex
+multiple members of annexin complexes
 (\Sexpr{gene.expression("equus_caballus","S100A6",complete=TRUE)},
@@
 for this sample. Ignoring these, notable highly expressed genes
-include members of the annexin complex
+include members of annexin complexes
 (\Sexpr{gene.expression("ovis_aries","S100G",complete=TRUE)},
@@
 %
-Multiple components of the annexin complex are included in the core
+Multiple components of annexin complexes are included in the core
 transcriptome, including \textit{ANXA2}, \textit{ANXA1},
-\textit{S100A11}, \textit{S100A10}. The modulation of  \DLA{It does more stuff; read
-  the huthcinson2011 review and fill this in}
-%
-% Another annexin, \textit{ANXA2} 
-%
-\textit{TIMP3}, which may be down-regulated in pre-eclampsia
-\cite{Ma.ea2014:Down-regulationofTIMP3leads} (though another study
-\cite{Xiang.ea2013:PromoterhypomethylationofTIMP3} claims
-up-regulation) does things
-
-% SPINT2 -- Serine Peptidase inhibitor
-
-% SLC3A2 -- this is probably actually a housekeeping gene; it's
-% involved in amino acid transport
-
-% SQSTM1 -- NFκB signaling (immune supression?)
-
-% KRTCAP2  N-glycosylation (probably housekeeping)
-
-% NRDG1 -- interacts with HSP90
-
-% SDC4 -- proteoglycan (syndecan) cell adhesion
-
-% LAMC1 -- laminin
-
-% LUM -- proteoglycan
-
-% VDAC1 -- mitochondrial gene?
-
-% F11R -- tight junctions
-
-SERPINH1? with TIMP3 
-% http://www.proteinatlas.org/ENSG00000149257-SERPINH1/tissue/placenta
-
+\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
+resolution of the inflammation response through inhibition of
+phospholipase A2 activity and signaling through the formyl peptide
+receptor family, and has previously been hypothesized to be important
+for the maintenance of the anti-inflammatory during pregnancy
+\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
+known to interact with S100A10, S100A6, S100A11, and S100A4
+\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
+its interaction with S100P as well; it is involved in cell-cell
+interaction as well as vesicle trafficking and von Willibrand Factor
+secretion. The expression regulation of another core gene,
+\textit{TIMP3}, has been implicated in pre-eclampsia
+\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
 \Cref{tab:core_placenta_go_terms} lists the significant enrichment of
@@
 include those involved in cytoskeletal structure, protein transport,
-and membrane protein structure.
+membrane protein structure, and inflammation resolution.
+
+% SPINT2 -- Serine Peptidase inhibitor
+
+% SLC3A2 -- this is probably actually a housekeeping gene; it's
+% involved in amino acid transport
+
+% SQSTM1 -- NFκB signaling (immune supression?)
+
+% KRTCAP2  N-glycosylation (probably housekeeping)
+
+% NRDG1 -- interacts with HSP90
+
+% SDC4 -- proteoglycan (syndecan) cell adhesion
+
+% LAMC1 -- laminin
+
+% LUM -- proteoglycan
+
+% VDAC1 -- mitochondrial gene?
+
+% F11R -- tight junctions
+
+% SERPINH1? with TIMP3 
+% http://www.proteinatlas.org/ENSG00000149257-SERPINH1/tissue/placenta
 

### 2016-01-27 manuscript/manuscript.Rnw
prep for genome biology submission

@@
-\documentclass[english,12pt]{article}
+%\documentclass[english,12pt]{article}
+\documentclass[doublespacing]{bmcart_dla}
 \usepackage{fontspec}
-\setmainfont{FreeSerif}
+\setmainfont{FreeSans}
 \setsansfont{FreeSans}
 \setmonofont{FreeMono}
-\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
+%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
+\usepackage[letterpaper]{geometry}
 \usepackage{url}
-\usepackage{fancyhdr}
+% \usepackage{fancyhdr}
 \usepackage{graphicx}
-\usepackage[bf]{caption}
+% \usepackage[bf]{caption}
 \usepackage{rotating}
@@
 \usepackage{amsmath}
-\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
+\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
 \addbibresource{references.bib}
-\usepackage[noblocks,auth-sc]{authblk}
+% \usepackage[noblocks,auth-sc]{authblk}
 \usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
@@
 % \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
-% \AtBeginFigures{\section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
-% \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
+% \AtBeginFigures{\newpage \section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
+% \AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
 \usepackage[nomargin,inline,draft]{fixme}
@@
 \newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
-\usepackage[x11names,svgnames]{xcolor}
-\usepackage{texshade}
-\usepackage{tikz}
+% \usepackage[x11names,svgnames]{xcolor}
+% \usepackage{texshade}
+% \usepackage{tikz}
 \usepackage{nameref}
@@
 \usepackage[capitalize]{cleveref}
-\newenvironment{narrow}[2]{%
-  \begin{list}{}{%
-      \setlength{\topsep}{0pt}%
-      \setlength{\leftmargin}{#1}%
-      \setlength{\rightmargin}{#2}%
-      \setlength{\listparindent}{\parindent}%
-      \setlength{\itemindent}{\parindent}%
-      \setlength{\parsep}{\parskip}}%
-  \item[]}{\end{list}}
-\def\newblock{\hskip}
-\newenvironment{paperquote}{%
-  \begin{quote}%
-     \it
-  }%
-  {\end{quote}}
-\renewcommand{\textfraction}{0.15}
-\renewcommand{\topfraction}{0.85}
-\renewcommand{\bottomfraction}{0.65}
-\renewcommand{\floatpagefraction}{0.60}
+% \newenvironment{narrow}[2]{%
+%   \begin{list}{}{%
+%       \setlength{\topsep}{0pt}%
+%       \setlength{\leftmargin}{#1}%
+%       \setlength{\rightmargin}{#2}%
+%       \setlength{\listparindent}{\parindent}%
+%       \setlength{\itemindent}{\parindent}%
+%       \setlength{\parsep}{\parskip}}%
+%   \item[]}{\end{list}}
+% \def\newblock{\hskip}
+% \newenvironment{paperquote}{%
+%   \begin{quote}%
+%      \it
+%   }%
+%   {\end{quote}}
+% \renewcommand{\textfraction}{0.15}
+% \renewcommand{\topfraction}{0.85}
+% \renewcommand{\bottomfraction}{0.65}
+% \renewcommand{\floatpagefraction}{0.60}
 %\renewcommand{\baselinestretch}{1.8}
-\newenvironment{enumerate*}%
-  {\begin{enumerate}%
-    \setlength{\itemsep}{0pt}%
-    \setlength{\parskip}{0pt}}%
-  {\end{enumerate}}
-\newenvironment{itemize*}%
-  {\begin{itemize}%
-    \setlength{\itemsep}{0pt}%
-    \setlength{\parskip}{0pt}}%
-  {\end{itemize}}
-\setcounter{tocdepth}{3}
-\oddsidemargin 0.0in 
-\textwidth 6in
-\raggedbottom
-\clubpenalty = 10000
-\widowpenalty = 10000
+% \newenvironment{enumerate*}%
+%   {\begin{enumerate}%
+%     \setlength{\itemsep}{0pt}%
+%     \setlength{\parskip}{0pt}}%
+%   {\end{enumerate}}
+% \newenvironment{itemize*}%
+%   {\begin{itemize}%
+%     \setlength{\itemsep}{0pt}%
+%     \setlength{\parskip}{0pt}}%
+%   {\end{itemize}}
+% \setcounter{tocdepth}{3}
+% \oddsidemargin 0.0in 
+% \textwidth 6in
+% \raggedbottom
+% \clubpenalty = 10000
+% \widowpenalty = 10000
 % set a large emergency stretch to avoid issues with overfull boxes
-\setlength{\emergencystretch}{20em}
+% \setlength{\emergencystretch}{20em}
 % make the afill smaller
-\renewcommand\Affilfont{\itshape\small}
+% \renewcommand\Affilfont{\itshape\small}
 \renewcommand\cite[1]{\autocite{#1}}
-\pagestyle{fancy}
+% \pagestyle{fancy}
 
@@
 \begin{document}
-\author[1]{Don Armstrong}
-\author[2]{Michael R. McGowen}
-\author[1]{Amy Weckle}
-\author[3]{Jason Caravas}
-\author[4]{Dalen Agnew}
-\author[5]{Kurt Benirschke}
-\author[6]{Sue Savage-Rumbaugh}
-\author[7,8]{Chong Jai Kim}
-\author[9]{Günter P. Wagner}
-\author[7]{Roberto Romero}
-\author[1,6,10,*]{Derek E. Wildman}
-\title{The evolution of gene expression in the term placenta of viviparous mammals}
-% These are all wrong, but we'll change them later.
-\affil[1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\affil[2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
-\affil[3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
-\affil[4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
-\affil[5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
-\affil[6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
-\affil[7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
-\affil[8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
-\affil[9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
-\affil[10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\affil[*]{To whom correspondence should be addressed}
-\date{\today}
-
-% Acronyms
-\input{acronyms}
-% \twocolumn
-\maketitle
-% \onehalfspacing
-\doublespacing
+\raggedright
 
@@
 
-\section{Abstract}
-
-The placenta is arguably the most anatomically variable organ in
-mammals. Understanding gene expression patterns among placentas from
-multiple species can illuminate the molecular basis of placenta
-function as well as potentially enrich our knowledge of the biology of
-obstetrical syndromes. We profiled the term placenta by using Illumina
+% Acronyms
+\input{acronyms}
+
+\begin{frontmatter}
+\begin{fmbox}
+\dochead{Research}
+\title{The evolution of gene expression in the term placenta of viviparous mammals}%
+%
+\author[addressref={aff1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
+\author[addressref={aff2}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
+\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL} \fnm{Amy} \snm{Weckle}}
+\author[addressref={aff3}]{\inits{J}\fnm{Jason} \snm{Caravas}}
+\author[addressref={aff4}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
+\author[addressref={aff5}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
+\author[addressref={aff6}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
+\author[addressref={aff7,aff8}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
+\author[addressref={aff9}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
+\author[addressref={aff7}]{\inits{R}\fnm{Roberto} \snm{Romero}}
+\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
+% These are all wrong, but we'll change them later.
+\address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
+\address[id=aff2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
+\address[id=aff3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
+\address[id=aff4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
+\address[id=aff5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
+\address[id=aff6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
+\address[id=aff7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
+\address[id=aff8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
+\address[id=aff9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
+\address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
+%\address[id=affcor]{Derek Wildman <wildmand@illinois.edu>}
+
+\begin{artnotes}
+\end{artnotes}
+
+\end{fmbox}
+% \twocolumn
+
+\begin{abstractbox}
+\begin{abstract}
+% \section{Abstract}
+  \parttitle{Background} The placenta is arguably the most
+  anatomically variable organ in mammals. Understanding gene
+  expression patterns among placentas from multiple species can
+  illuminate the molecular basis of placenta function as well as
+  potentially enrich our knowledge of the biology of obstetrical
+  syndromes.
+
+  \parttitle{Results} We profiled the term placenta by using Illumina
   RNA-Seq technology to sequence the transcriptomes of
@@
   marsupials and “placental”) mammals.
+
+  \parttitle{Conclusions}
 
@@
 
-\section{Introduction}
+\end{abstract}
+\end{abstractbox}
+\end{frontmatter}
+
+\section{Background}
 
@@
 \begin{table}
-  \begin{tabular}{c c c p{3in}}
+  \begin{tabular}{c c c c}
     \toprule
@@
 
-Fetal tissue was collected from the placenta of six mammalian species
-(\human [human], \bonobo [bonobo], \dog [domestic dog], \cow [domestic
-cow], \elephant [African savanna elephant], \opossum [gray
-short-tailed opossum], \gallili [blind mole rat from Upper Galilee],
-\carmelli [blind mole rat from northern Israel]). Information on
+Fetal tissue was collected from the placenta of eight mammalian
+species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
+[domestic cow], \elephant [African savanna elephant], \opossum [gray
+short-tailed opossum], \galili [blind mole rat from Upper Galilee],
+\carmeli [blind mole rat from northern Israel]). Information on
 specimen ID, place and time of collection, and collector is shown for
@@
 significantly enriched if the \ac{FDR} as estimated by the Benjamini
-and Hochberg procedure\cite{Benjamini1995:bh_procedure} on from the
+and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
 $p$ values obtained from the Fisher Exact Test were less than or equal
@@
 \cow, \dog, \elephant, \mouse, and \human were determined using
-Ensembl v.80 gene trees via
-Biomart\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the
-biomaRt R package \DLA{Cite the biomaRt package}. A total of
-\Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
+Ensembl v.80 gene trees via Biomart
+\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
+package
+\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
+A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
   1,length(unique(human_name))]} 1:1 orthologous genes were expressed
@@
 \cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
-and all genes in Supplemental
-\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_nannospalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.
+and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_nannospalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.
 
@@
 \begin{table}
-  \tiny
+  \footnotesize
 <<primate.lineage.specific,echo=FALSE,results="asis">>=
@@
 \begin{table}
-  \tiny
+  \footnotesize
 <<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}
-  \tiny
+  \footnotesize
 <<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
@@
 @ 
-\caption{\genetreexplanation{\textit{SERPINE2}}}
+\caption{\genetreeexplanation{\textit{SERPINE2}}}
 \label{fig:serpine_evolution_tree}
@@
 \begin{table}
-  \tiny
+  \footnotesize
 <<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}
-  \tiny
+  \footnotesize
 <<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}
-  \tiny
+  \footnotesize
 <<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}
-  \tiny
+  \footnotesize
 <<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
@@
   in the core placenta transcriptome as compared to all genes
-  with 1:1 human orthologs in all species. “# Core” is the
+  with 1:1 human orthologs in all species. “\# Core” is the
   number of genes in the core transcriptome for each \ac{GO} term;
@@
 
+\begin{backmatter}
+\section*{Competing interests}
+  The authors declare that they have no competing interests.
+
+\section*{Author's contributions}
+Don Armstrong analyzed all of the data and wrote the paper.
+
+Michael McGowan wrote the paper and performed initial analyses.
+
+\section*{Acknowledgements}
+  Text for this section \ldots
+
+\printaddresses
+
+\end{backmatter}
+
 \printbibliography

### 2016-01-27 go_annotation/go_tissue_analysis.R
switch to top 1% instead of top 5%

@@ -33,8 +33,8 @@ do.go.analysis <- function(species.name,go.ontology,top.100=TRUE,exclude.houseke
         analysis.type <- paste0("Top 100",analysis.type)
         gene.list <- all.genes[,factor(as.integer(rank <=100))]
     } else {
-        analysis.type <- paste0("Top 5%",analysis.type)
-        gene.list <- all.genes[,factor(as.integer(rank <= (sum(mean_fpkm>0)*0.05)))]
+        analysis.type <- paste0("Top 1%",analysis.type)
+        gene.list <- all.genes[,factor(as.integer(rank <= (sum(mean_fpkm>1)*0.01)))]
     }
     names(gene.list) <- all.genes[,egid]
     

### 2016-01-28 manuscript/manuscript.Rnw
make math sans
use output.top.genes.table

@@
 \setmonofont{FreeMono}
+\setmathrm{FreeSans}
+\setmathsf{FreeSans}
+\setmathtt{FreeSans}
 %\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
@@
 \usepackage{amsmath}
+\usepackage{sfmath}
 \usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
@@
 @ 
-\caption{\genetreeexplanation{follistatin-like 1 \textit{FSTLl1}}}
+\caption{\genetreeexplanation{follistatin-like 1 \textit{FSTL1}}}
 \label{fig:fstl1_evolution_tree}
@@
 <<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
-top.human.genes <- top.genes.table(species.name="homo sapiens")
-print(xtable(top.human.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.human.genes,"Gene symbol")
+output.top.genes.table(top.genes.table(species.name="homo sapiens"))
 @ 
@@
 top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
-print(xtable(top.bonobo.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+output.top.genes.table(top.bonobo.genes)
 setkey(top.bonobo.genes,"Gene symbol")
@@
 top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
-print(xtable(top.spider.monkey.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
+output.top.genes.table(top.spider.monkey.genes)
 setkey(top.spider.monkey.genes,"Gene symbol")
@@
 \begin{table}
-  \tiny
+  \footnotesize
 <<rodentia.lineage.specific,echo=FALSE,results="asis">>=
@@
 \begin{table}[h]
-  \tiny
+  \footnotesize
 <<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
-top.mouse.genes <- top.genes.table(species.name="mus musculus")
-print(xtable(top.mouse.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.mouse.genes,"Gene symbol")
+output.top.genes.table(top.genes.table(species.name="mus musculus"))
 @ 
@@
 @ 
-\caption{Expression of cathespins related to \textit{CTSL} with $\mathrm{FPKM} \ge 1$}
+\caption{\genetreeexplanation{cathespins related to \textit{CTSL}}}
 \label{fig:cathespin_evolution_tree}
@@
 \begin{table}[h]
-  \tiny
+  \footnotesize
 <<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
-top.galili.genes <- top.genes.table(species.name="nannospalax galili")
-print(xtable(top.galili.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.galili.genes,"Gene symbol")
+output.top.genes.table(top.genes.table(species.name="nannospalax galili"))
 @ 
@@
 \begin{table}[h]
-  \tiny
+  \footnotesize
 <<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
-top.carmeli.genes <- top.genes.table(species.name="spalax carmeli")
-print(xtable(top.carmeli.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.carmeli.genes,"Gene symbol")
+output.top.genes.table(top.genes.table(species.name="spalax carmeli"))
 @ 
@@
 <<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
-top.dog.genes <- top.genes.table(species.name="canis familiaris")
-print(xtable(top.dog.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.dog.genes,"Gene symbol")
+output.top.genes.table(top.genes.table(species.name="canis familiaris"))
 @ 
@@
 <<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
-top.cow.genes <- top.genes.table(species.name="bos taurus")
-print(xtable(top.cow.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.cow.genes,"Gene symbol")
+output.top.genes.table(top.genes.table(species.name="bos taurus"))
 @ 
@@
 \begin{table}[h]
-  \tiny
+  \footnotesize
 <<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
-top.pig.genes <- top.genes.table(species.name="sus scrofa")
-print(xtable(top.pig.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.pig.genes,"Gene symbol")
+output.top.genes.table(top.genes.table(species.name="sus scrofa"))
 @ 
@@
 \begin{table}[h]
-  \tiny
+  \footnotesize
 <<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}[h]
-  \tiny
+  \footnotesize
 <<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
@@
                    ,][1:50,]
-print(xtable(top.sheep.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.sheep.genes,"Gene symbol")
+output.top.genes.table(top.sheep.genes)
 @ 
@@
 \begin{table}
-\tiny
+\footnotesize
 <<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
-top.elephant.genes <- top.genes.table(species.name="loxodonta africana")
-print(xtable(top.elephant.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.elephant.genes,"Gene symbol")
+output.top.genes.table(top.genes.table(species.name="loxodonta africana"))
 @ 
@@
 \begin{table}
-\tiny
+\footnotesize
 <<top.50.armadillo.expressed.genes,echo=FALSE,results="asis">>=
-top.armadillo.genes <- top.genes.table(species.name="dasypus novemcinctus")
-print(xtable(top.armadillo.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.armadillo.genes,"Gene symbol")
+output.top.genes.table(top.genes.table(species.name="dasypus novemcinctus"))
 @ 

### 2016-01-28 manuscript/manuscript.Rnw
add top genes table

@@
 }
+output.top.genes.table <- function(data,output=TRUE) {
+    if (output) {
+        table.display <- c("s","s","s","d")
+        table.align <- c("c","c","p{3in}","r")
+        if (any(colnames(data) %in% "FPKM SEM"))  {
+            table.display <- c(table.display,"d")
+            table.align <- c(table.align,"r")
+        }
+        if (any(colnames(data) %in% "Human Symbol"))  {
+            table.align <- c("c",table.align)
+            table.display <- c("s",table.display)
+        }
+        print(xtable(data,
+                     display=table.display,
+                     align=table.align
+                     ),
+              booktabs=TRUE,
+              floating=FALSE,
+              sanitize.text.function=genes.table.sanitize,
+              include.rownames=FALSE)
+    }
+}
+
 @

### 2016-01-28 manuscript/manuscript.Rnw
round FPKM

@@
     }
+    temp[,`FPKM`:=round(`FPKM`)]
     return(temp)

### 2016-01-28 manuscript/manuscript.Rnw
move some tables and figures to supplemental material

@@
 })
-options(scipen = -1, digits = 2)
+options(scipen = -2, digits = 0)
 library("data.table")
@@
 size of 230 base pairs (bp). Read lengths for species are shown in
-\cref{tab:ref_genome_aligned_species}.
+Supplemental \cref{tab:ref_genome_aligned_species}.
 
@@
 (\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
-Transcriptomes without a suitable reference genome (Reference of ---
-in \cref{tab:ref_genome_aligned_species}) were assembled into a
-reference using Trinity v2014-07-17
-\cite{Haas.ea2013:Denovotranscriptsequence} and annotated using
-Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against either
-the \human or \mouse transcriptome
-(\cref{tab:ref_genome_aligned_species}).
-% 
-All transcriptomes were then aligned to the appropriate reference
-genome using STAR v2.4.2a
+Transcriptomes without a suitable reference genome (\smonkey, \galili,
+\carmeli; italic reference in Supplemental
+\cref{tab:ref_genome_aligned_species}) were assembled 
+using Trinity v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence}
+and annotated using Diamond
+\cite{Buchfink.ea2015:Fastandsensitiveprotein} against either the
+\human or \mouse transcriptome
+(Supplemental \cref{tab:ref_genome_aligned_species}).
+% 
+All transcriptomes were then aligned to the appropriate reference (or
+assembled) genome using STAR v2.4.2a
 \cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
@@
 \cite{Prüfer.ea2012:bonobogenomecomparedwith}.
-
-\begin{table}
-  \begin{tabular}{c c c c}
-    \toprule
-    Species                       & Common Name        & Reference & Read Length \\
-    \midrule
-    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
-    \textit{Bos taurus}           & Cow                & UMD3.1     & 75 \\
-    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 150 \\
-    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 100 \\
-    \textit{Equus caballus}       & Horse              & EquCab2    & 100 \\ 
-    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 75 \\
-    \textit{Homo sapiens}         & Human              & GRCh38     & 75, 150 \\
-    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 75 \\
-    \textit{Mus musculus}         & Mouse              & GRCm38     & 35,44,150 \\
-    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
-    \textit{Ovis aries}           & Sheep              & Oar\_v3.1   & 151 \\
-    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 150 \\
-    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
-    \textit{Sus scrofa}           & Duroc (pig)        & Sscrofa10.2 & 51 \\
-    \bottomrule
-  \end{tabular}
-  \caption{Reference genomes for aligned species}
-  \label{tab:ref_genome_aligned_species}
-\end{table}
 
@@
 
-\begin{figure}
-<<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
-star.logs.unmapped <- 
-    star.logs[grepl("%",field) & 
-                  grepl("unmapped",field) & 
-                      !grepl("too many mismatches",field),]
-star.logs.unmapped <- 
-    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
-pushViewport(viewport(layout=grid.layout(ncol=1,nrow=2,heights=c(2,1))))
-print(ggplot(star.logs.unmapped,
-             aes(x=capfirst(species),y=as.numeric(gsub("%","",value))))
-      + geom_boxplot()
-      + facet_grid(field~.,scales="free")
-      + ylab("Percent of Reads")
-      + xlab(NULL)
-      + theme_bw()
-      + theme(legend.position="top",
-              plot.margin=unit(c(0,0,0,1.1),"strwidth",data=list("","","","e8")),
-              axis.title.x=element_blank(),
-              axis.text.x = element_blank()),#element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
-      vp=viewport(layout.pos.col=1,layout.pos.row=1)
-      )
-pushViewport(viewport(layout.pos.col=1,layout.pos.row=1))
-grid.text("A",
-          y=unit(1,"npc")-unit(0.1,"cm"),
-          x=unit(0,"npc")+unit(0.1,"cm"),
-          gp=gpar(fontsize=24),
-          just=c("left","top"))
-popViewport()          
-print(ggplot(star.logs[grepl("input reads$",field),],
-             aes(x=capfirst(species),y=as.numeric(value)))
-      + geom_boxplot()
-      + scale_y_log10()
-      + ylab("Reads")
-      + xlab("Species")
-      + theme_bw()
-      + theme(legend.position="top",
-              plot.margin=unit(c(0,1.6,0,0),"strheight",data=list("","TO","","")),
-              axis.text.x = element_text(angle = 30, hjust = 1, vjust=1,face="italic")),
-      vp=viewport(layout.pos.col=1,layout.pos.row=2))
-pushViewport(viewport(layout.pos.col=1,layout.pos.row=2))
-grid.text("B",
-          y=unit(1,"npc")-unit(0.1,"cm"),
-          x=unit(0,"npc")+unit(0.1,"cm"),
-          gp=gpar(fontsize=24),
-          just=c("left","top"))
-popViewport(2)
-@ 
-\caption{RNAseq read and alignment statistics for all species. A)
-  Boxplot of percentages of unmapped reads in each species; 0\% of
-  reads in all species were unmapped because of mismatches. B) Boxplot
-  of number of RNAseq reads in each species.}
-\label{fig:transcriptome_statistics}
-\end{figure}
-
-\Cref{fig:transcriptome_statistics} shows the RNAseq read and
+Supplemental \cref{fig:transcriptome_statistics} shows the RNAseq read and
 alignment statistics for all species. For transcriptomes sequenced
@@
 
-\begin{table}
-  \tiny
-<<lau.lineage.specific,echo=FALSE,results="asis">>=
-lau.all.lin.spec <- placenta.lineage.specific.table("lau.all")
-setkey(lau.all.lin.spec,"Human Symbol")
-@ 
-\caption{Genes whose expression changed in the Laurasiatheria
-  superorder as compared to all other analyzed species.}
-\label{tab:laurasiatheria_lineage_specific}
-\end{table}
-
-\begin{table}
-  \tiny
-<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
-eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau")
-setkey(eua.lau.lin.spec,"Human Symbol")
-@ 
-\caption{Genes whose expression changed between the
-  Euasiatherian and Laurasiatherian superorder.}
-\label{tab:laurasiatheria_eurasiatherian_lineage_specific}
-\end{table}
-
 We examined the expression of five Laurasiatherians: \cow, \sheep,
@@
 Laurasiatherians and all other analyzed species
-(\cref{tab:laurasiatheria_lineage_specific}) or between
+(Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
 Laurasiatherians and Euasiatherians
-(\cref{tab:laurasiatheria_eurasiatherian_lineage_specific}).
+(Supplemental \cref{tab:laurasiatheria_eurasiatherian_lineage_specific}).
 
@@
 a hemoglobin found only in marsupials and monotremes
-\cite{Opazo.ea2008:Genomicevidenceforindependent}. The hemoglobin gene
-\textit{HBA} is also expressed in the placenta
+\cite{Opazo.ea2008:Genomicevidenceforindependent} (Supplemental
+\cref{fig:hbe1_evolution_tree}). The hemoglobin gene \textit{HBA} is
+also expressed in the placenta
 (\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016633",complete=TRUE)}).
@@
 (\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
-also shows high expression in the opossum placenta. The gene \textit{WFDC2}, a
-member of the whey acidic protein (WAP) family is also expressed to a
-high degree in the opossum placenta
+also shows high expression in the opossum placenta. The gene
+\textit{WFDC2}, a member of the whey acidic protein (WAP) family is
+also expressed to a high degree in the opossum placenta
 (\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).
@@
 respectively).
-
-\begin{figure}
-<<hbe1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="HBE1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{Expression of \textit{HBE1} and related genes}
-\label{fig:hbe1_evolution_tree}
-\end{figure}
 
@@
 
-\begin{figure}
-<<core.placenta.transcriptome.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
-calculate.distribution.shape <- 
-    function(min.vector) {
-        min.vector <- sort(min.vector,decreasing=TRUE)
-        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
-        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
-        return(temp)
-    }
-
-cptg.nh <- core.placenta.transcriptome.genes.shape[!(human_name %in% housekeeping.genes.superset[,gene_short_name])]
-
-cpt.distribution.shape <- 
-    lapply(c("min_expression",
-             "second_min_expression",
-             "third_min_expression",
-             "fourth_min_expression",
-             "fifth_min_expression"),
-           function(x){
-               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Lowest Expression`:=x]
-           })
-cpt.distribution.shape <- 
-    rbindlist(cpt.distribution.shape)
-
-cpt.distribution.shape <- 
-    cpt.distribution.shape[,`Lowest Expression`:=
-                               gsub("^min_expression$","First",
-                                    `Lowest Expression`)]
-cpt.distribution.shape <- 
-    cpt.distribution.shape[,`Lowest Expression`:=
-                               factor(gsub("^(.)(.+)_min_expression$","\\U\\1\\L\\2",
-                                           `Lowest Expression`,perl=TRUE),
-                                      levels=c("First","Second","Third","Fourth","Fifth"))
-                           ]
-setnames(cpt.distribution.shape,
-         "FPKM",
-         "Expression (FPKM)")
-
-print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 1,],
-             aes(x=`Expression (FPKM)`,color=`Lowest Expression`,y=num_genes))
-      + geom_line(lwd=1)
-      + xlab(expression(FPKM))
-      + ylab("Genes with N-th Lowest Expression")
-      + scale_y_log10()
-      + scale_x_log10()
-      + geom_vline(xintercept=10)
-      + geom_segment(mapping=aes(y=V1,
-                         yend=V1,
-                         xend=10,
-                         x=100,
-                         color=`Lowest Expression`),
-                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
-                                              max(num_genes),`Lowest Expression`],
-                     arrow=arrow(length=unit(0.5,"cm"))
-                   )
-      + geom_text(mapping=aes(x=100,
-                      y=V1,
-                      label=paste0(V1," genes"),override=TRUE),
-                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
-                      max(num_genes),`Lowest Expression`],
-                  show_guide=FALSE,
-                  hjust=0
-                  )
-      + guides(color=guide_legend(title="Lowest\nExpression"))
-      )
-@ 
-\caption{Distribution shape of placenta core genes. The vertical axis shows
-  the number of orthologous gene sets included in the core placenta
-  transcriptome if the least expression cutoff (or second-least, etc.)
-  was set to the horizontal axis. For example, if the cutoff is set to 10
-  (vertical black line), 37 genes would be included if the FPKM of all
-  species must be at least 10, 134 genes would be included if all but
-  one species must have FPKM of at least 10. 655 genes would be
-  included if all but five species must have FPKM of at least 10.}
-\label{fig:core_placenta_transcriptome_shape}
-\end{figure}
-
 \begin{table}
@@
 \caption{\ac{GO} terms most significantly enriched
-  (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]},
+  (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]} MF,
   \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP)
@@
   each term; “FDR” is the false discovery rate as estimated by
-  the Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure}}
+  the Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure}.}
 \label{tab:core_placenta_go_terms}
@@
 \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
-\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
+Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
 which would be included in the core placenta set if the \ac{FPKM}
@@
     go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
-                                         function(x){sum(x <= 0.05)})]
+                                         function(x){sum(as.numeric(as.character(x)) <= 0.05)})]
 go.analysis.wide <- 
@@
     go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
-                         grepl("Top 5% No Housekeeping",type),
+                         grepl("Top 1% No Housekeeping",type),
                      list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
@@
 \caption{GO terms with significant enrichment ($\mathrm{FDR}\le0.05$)
-  in 12 or more of the species analyzed from genes in the top 5\% of
+  in 12 or more of the species analyzed from genes in the top 1\% of
   expression. $n(\mathrm{FDR}\le0.05)$ is the number of species with

### 2016-01-28 manuscript/supplemental_material.Rnw
move some tables and figures to supplemental material

--- 2016-01-06 manuscript/supplemental_material.Rnw
@@
 
+\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
+\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
+\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
+\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
+\newcommand{\horse}[0]{\textit{E. caballus}\xspace}
+\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
+\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
+\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
+\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
+\newcommand{\galili}[0]{\textit{N. galili}\xspace}
+\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
+\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
+\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
+\newcommand{\pig}[0]{\textit{S. scrofa}\xspace}
+
+\newcommand{\genetreeexplanation}[1]{Gene tree and heatmap of
+  expression of genes related to #1. The gene tree was obtained from
+  Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes from species
+  not included in this study or having expression by \ac{FPKM} less
+  than 1 were excluded from the plotted tree. Heatmap cells in grey
+  are from species where there was no expression of the relevant
+  transcript (or a species where that transcript does not apply).}
+
 \begin{document}
@@
 
-\section{Sequence Read Archive}
+\section{Species}
+
+\subsection{Sequence Source}
 
@@
   \label{tab:srr_numbers}
+\end{table}
+
+\subsection{Species Alignment}
+
+\begin{table}
+  \begin{tabular}{c c c c}
+    \toprule
+    Species                       & Common Name        & Reference & Read Length \\
+    \midrule
+    \textit{Ateles fusciceps}     & Spider monkey      & \textit{Homo sapiens} & 100 \\
+    \textit{Bos taurus}           & Cow                & UMD3.1     & 75 \\
+    \textit{Canis familiaris}     & Dog                & CanFam3.1  & 150 \\
+    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 100 \\
+    \textit{Equus caballus}       & Horse              & EquCab2    & 100 \\ 
+    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 75 \\
+    \textit{Homo sapiens}         & Human              & GRCh38     & 75, 150 \\
+    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 75 \\
+    \textit{Mus musculus}         & Mouse              & GRCm38     & 35,44,150 \\
+    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
+    \textit{Ovis aries}           & Sheep              & Oar\_v3.1   & 151 \\
+    \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 150 \\
+    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
+    \textit{Sus scrofa}           & Duroc (pig)        & Sscrofa10.2 & 51 \\
+    \bottomrule
+  \end{tabular}
+  \caption{Reference genomes for aligned species}
+  \label{tab:ref_genome_aligned_species}
 \end{table}
@@
     cat(paste0("\\subsubsection{Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
-                           " the top 5\\% of \\textit{",
+                           " the top 1\\% of \\textit{",
                            capfirst(this.species),"} expressed transcripts}\n\n"))
     this.caption <- paste0("Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
-                           " the top 5\\% of \\textit{",
+                           " the top 1\\% of \\textit{",
                            capfirst(this.species),"} expressed transcripts.")
-    print(xtable(go.results[grepl("Top 5% No Housekeeping",type) & FDR <= 0.05 & species == this.species,
+    print(xtable(go.results[grepl("Top 1% No Housekeeping",type) & FDR <= 0.05 & species == this.species,
                             list(GO.ID,Term,FDR,Ontology)
@@
                  caption=this.caption,
-                 label=paste0("tab:go_terms_enriched_top_five_",gsub(" ","_",this.species)),
+                 label=paste0("tab:go_terms_enriched_top_one_",gsub(" ","_",this.species)),
                  display=c("s","s","s","e","s")
@@
 \section{Gene Expression}
+
+\subsection{Gene Evolution Trees}
+
+\begin{figure}
+<<hbe1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="HBE1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{Expression of \textit{HBE1} and related genes}
+\label{fig:hbe1_evolution_tree}
+\end{figure}
+
+\subsection{Clade specific Expression}
+
+\begin{table}
+  \footnotesize
+<<lau.lineage.specific,echo=FALSE,results="asis">>=
+lau.all.lin.spec <- placenta.lineage.specific.table("lau.all")
+setkey(lau.all.lin.spec,"Human Symbol")
+@ 
+\caption{Genes whose expression changed in the Laurasiatheria
+  superorder as compared to all other analyzed species.}
+\label{tab:laurasiatheria_lineage_specific}
+\end{table}
+
+\begin{table}
+  \footnotesize
+<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
+eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau")
+setkey(eua.lau.lin.spec,"Human Symbol")
+@ 
+\caption{Genes whose expression changed between the
+  Euasiatherian and Laurasiatherian superorder.}
+\label{tab:laurasiatheria_eurasiatherian_lineage_specific}
+\end{table}
+
+\subsection{Species Expression}
+
 <<top.expressed.genes,echo=FALSE,results="asis">>=
@@
 
+\section{Core Placenta Transcriptome}
+
+\begin{figure}
+<<core.placenta.transcriptome.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
+calculate.distribution.shape <- 
+    function(min.vector) {
+        min.vector <- sort(min.vector,decreasing=TRUE)
+        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
+        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
+        return(temp)
+    }
+
+cptg.nh <- core.placenta.transcriptome.genes.shape[!(human_name %in% housekeeping.genes.superset[,gene_short_name])]
+
+cpt.distribution.shape <- 
+    lapply(c("min_expression",
+             "second_min_expression",
+             "third_min_expression",
+             "fourth_min_expression",
+             "fifth_min_expression"),
+           function(x){
+               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Lowest Expression`:=x]
+           })
+cpt.distribution.shape <- 
+    rbindlist(cpt.distribution.shape)
+
+cpt.distribution.shape <- 
+    cpt.distribution.shape[,`Lowest Expression`:=
+                               gsub("^min_expression$","First",
+                                    `Lowest Expression`)]
+cpt.distribution.shape <- 
+    cpt.distribution.shape[,`Lowest Expression`:=
+                               factor(gsub("^(.)(.+)_min_expression$","\\U\\1\\L\\2",
+                                           `Lowest Expression`,perl=TRUE),
+                                      levels=c("First","Second","Third","Fourth","Fifth"))
+                           ]
+setnames(cpt.distribution.shape,
+         "FPKM",
+         "Expression (FPKM)")
+
+print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 1,],
+             aes(x=`Expression (FPKM)`,color=`Lowest Expression`,y=num_genes))
+      + geom_line(lwd=1)
+      + xlab(expression(FPKM))
+      + ylab("Genes with N-th Lowest Expression")
+      + scale_y_log10()
+      + scale_x_log10()
+      + geom_vline(xintercept=10)
+      + geom_segment(mapping=aes(y=V1,
+                         yend=V1,
+                         xend=10,
+                         x=100,
+                         color=`Lowest Expression`),
+                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
+                                              max(num_genes),`Lowest Expression`],
+                     arrow=arrow(length=unit(0.5,"cm"))
+                   )
+      + geom_text(mapping=aes(x=100,
+                      y=V1,
+                      label=paste0(V1," genes"),override=TRUE),
+                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
+                      max(num_genes),`Lowest Expression`],
+                  show_guide=FALSE,
+                  hjust=0
+                  )
+      + guides(color=guide_legend(title="Lowest\nExpression"))
+      )
+@ 
+\caption{Distribution shape of placenta core genes. The vertical axis shows
+  the number of orthologous gene sets included in the core placenta
+  transcriptome if the least expression cutoff (or second-least, etc.)
+  was set to the horizontal axis. For example, if the cutoff is set to 10
+  (vertical black line), 37 genes would be included if the FPKM of all
+  species must be at least 10, 134 genes would be included if all but
+  one species must have FPKM of at least 10. 655 genes would be
+  included if all but five species must have FPKM of at least 10.}
+\label{fig:core_placenta_transcriptome_shape}
+\end{figure}
+
 \printbibliography

### 2016-01-28 manuscript/manuscript.Rnw
move PRLR and HOXA11 tree to supplemnetal

@@
 prolactin related genes having $\mathrm{FPKM} \ge 1$
-(\cref{fig:prolactin_evolution_tree}). The cathespins have also
-expanded in rodents, with six showing high expression
+(\cref{fig:prolactin_evolution_tree}, Supplemental
+\cref{fig:prolactin_receptor_evolution_tree}). The cathespins have
+also expanded in rodents, with six showing high expression
 (\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
@@
 express \textit{HOXA11} highly either
-(\Sexpr{gene.expression("loxodonta_africana","HOXA11",complete=TRUE)}).
+(\Sexpr{gene.expression("loxodonta_africana","HOXA11",complete=TRUE)},
+Supplemental \cref{fig:hoxa11_evolution_tree}).
 
@@
 \label{fig:prolactin_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<prolactin.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRLR" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 subtree=c("ENSG00000113494","ENSMODG00000020376"))
-@ 
-\caption{\genetreeexplanation{prolactin receptor}}
-\label{fig:prolactin_receptor_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<hoxa11.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="HOXA11" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 subtree=c("ENSBTAG00000014738","ENSECAG00000013229")
-                 )
-@ 
-\caption{\genetreeexplanation{\textit{HOXA11}}}
-\label{fig:hoxa11_evolution_tree}
 \end{figure}

### 2016-01-28 manuscript/supplemental_material.Rnw
move PRLR and HOXA11 tree to supplemnetal

@@
 \begin{figure}
+<<prolactin.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRLR" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSG00000113494","ENSMODG00000020376"))
+@ 
+\caption{\genetreeexplanation{prolactin receptor, \textit{PRLR}}}
+\label{fig:prolactin_receptor_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<hoxa11.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="HOXA11" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSBTAG00000014738","ENSECAG00000013229")
+                 )
+@ 
+\caption{\genetreeexplanation{\textit{HOXA11}}}
+\label{fig:hoxa11_evolution_tree}
+\end{figure}
+
+\begin{figure}
 <<hbe1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=

### 2016-01-28 manuscript/manuscript.Rnw
move RLN2, OVOS2, LGAL1, LGALS13 trees to supplmenetal

@@
 5\% of the frequency distribution of \ac{FPKM} (Supplemental
-\cref{tab:go_terms_enriched_top_five_dasypus_novemcinctus,tab:go_terms_enriched_top_five_equus_caballus,tab:go_terms_enriched_top_five_homo_sapiens,tab:go_terms_enriched_top_five_loxodonta_africana,tab:go_terms_enriched_top_five_monodelphis_domestica,tab:go_terms_enriched_top_five_mus_musculus,tab:go_terms_enriched_top_five_ovis_aries,tab:go_terms_enriched_top_five_pan_paniscus,tab:go_terms_enriched_top_five_sus_scrofa,tab:go_terms_enriched_top_five_ateles_fusciceps,tab:go_terms_enriched_top_five_nannospalax_galili,tab:go_terms_enriched_top_five_spalax_carmeli,tab:go_terms_enriched_top_five_bos_taurus,tab:go_terms_enriched_top_five_canis_familiaris});
+\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_nannospalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
 the top 100 genes with the highest \ac{FPKM} (Supplemental
@@
 
-\begin{figure}
-<<rln2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="RLN2" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{\textit{RLN2}}}
-\label{fig:rln2_evolution_tree}
-\end{figure}
-
 \subsubsection{\armadillo Placenta Transcriptome}
@@
 \Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033320",complete=TRUE)}
-(\cref{fig:ovos2_evolution_tree}), the growth factors
+(Supplemental \cref{fig:ovos2_evolution_tree}), the growth factors
 \Sexpr{gene.expression("dasypus_novemcinctus","IGF2",complete=TRUE)}
@@
 
-\begin{figure}
-<<ovos2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree["ENSDNOG00000033320",tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 subtree=c("ENSDNOG00000033320","ENSMODG00000017988"))
-@ 
-\caption{\genetreeexplanation{\cow \textit{OVOS2}}}
-\label{fig:ovos2_evolution_tree}
-\end{figure}
-
 \subsection{Metatherians}
@@
 <<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
-top.opossum.genes <- top.genes.table(species.name="monodelphis domestica")
-print(xtable(top.opossum.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.opossum.genes,"Gene symbol")
+output.top.genes.table(top.genes.table(species.name="monodelphis domestica"))
 @ 
@@
 the individual transcriptomes, we examined \ac{GSE} in 1:1 human
-orthologs which had \ac{FPKM} values in the top 5\% of expressed genes
+orthologs which had \ac{FPKM} values in the top 1\% of expressed genes
 (Supplemental %
-\cref{tab:go_terms_enriched_top_five_dasypus_novemcinctus,tab:go_terms_enriched_top_five_equus_caballus,tab:go_terms_enriched_top_five_homo_sapiens,tab:go_terms_enriched_top_five_loxodonta_africana,tab:go_terms_enriched_top_five_monodelphis_domestica,tab:go_terms_enriched_top_five_mus_musculus,tab:go_terms_enriched_top_five_ovis_aries,tab:go_terms_enriched_top_five_pan_paniscus,tab:go_terms_enriched_top_five_sus_scrofa,tab:go_terms_enriched_top_five_ateles_fusciceps,tab:go_terms_enriched_top_five_nannospalax_galili,tab:go_terms_enriched_top_five_spalax_carmeli,tab:go_terms_enriched_top_five_bos_taurus,tab:go_terms_enriched_top_five_canis_familiaris}) using topGO \cite{topGO:2.20.0}.
+\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_nannospalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris}) using topGO \cite{topGO:2.20.0}.
 %
@@
 of relaxin in the term placenta, with little evidence in other species
-examined here (\cref{fig:rln2_evolution_tree}). \textit{RLN} is known
+examined here (Supplemental \cref{fig:rln2_evolution_tree}). \textit{RLN} is known
 to be produced in the placenta of the pregnant \dog and is used in the
@@
 
-\begin{figure}
-<<nppb.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="NPPB" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{\textit{NPPB}}}
-\label{fig:nppb_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<nppc.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="NPPC" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{\textit{NPPC}}}
-\label{fig:nppc_evolution_tree}
-\end{figure}
-
 Two genes, \textit{NPPB} and \textit{NPPC}, code for peptide hormones
@@
 expression in the fetal portion of the mouse placenta
-(\cref{fig:hormone_genes,fig:nppc_evolution_tree,fig:nppb_evolution_tree}).
+(Supplemental \cref{fig:hormone_genes,fig:nppc_evolution_tree,fig:nppb_evolution_tree}).
 However, we do find high expression of at least one of these genes in
@@
 
-\begin{figure}
-<<galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{Expression of galectins}
-\label{fig:galectins_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{\textit{LGALS13}}}
-\label{fig:galectins_lgals13_evolution_tree}
-\end{figure}
-%$
 Galectins are recognized as major regulators of pregnancy in both
@@
 inducing apoptosis of maternal T-cells
-\cite{Than.ea2009:primatesubfamilyofgalectins}. Indeed,
-\textit{LGALS14} has the highest expression of any galectin in the
-\human and \smonkey placenta
-(\cref{fig:hormone_genes,fig:galectins_lgals13_evolution_tree}), and
-other primate galectins such as \textit{LGALS13} and \textit{LGALS16}
-are also expressed in \human and \smonkey placenta. However, in the
-closely related \bonobo, \textit{LGALS14} is not found, and
-\textit{LGALS1} is the most abundant galectin. \textit{LGALS1} shows
-generally high expression in all placentas analyzed. \textit{LGALS1}
-is also involved in mediating maternal-fetal immune tolerance, and is
-expressed in both maternal and fetal tissue in primates
+\cite{Than.ea2009:primatesubfamilyofgalectins} (Supplemental
+\cref{fig:galectins_lgals13_evolution_tree}). Indeed, \textit{LGALS14}
+has the highest expression of any galectin in the \human and \smonkey
+placenta (\cref{fig:hormone_genes}, Supplemental
+\cref{fig:galectins_lgals13_evolution_tree}), and other primate
+galectins such as \textit{LGALS13} and \textit{LGALS16} are also
+expressed in \human and \smonkey placenta. However, in the closely
+related \bonobo, \textit{LGALS14} is not found, and \textit{LGALS1} is
+the most abundant galectin. \textit{LGALS1} shows generally high
+expression in all placentas analyzed. \textit{LGALS1} is also involved
+in mediating maternal-fetal immune tolerance, and is expressed in both
+maternal and fetal tissue in primates
 \cite{Than.ea2008:Emergenceofhormonaland}. \textit{LGALS1} placental
@@
 show substantial expression of galectins in the opossum, especially
-\textit{LGALS1} (Table 8). In addition, \textit{LGALS2} is also
-expressed. The presence of galectins at the maternal-fetal interface
-in \opossum implies that it was also present in the ancestral placenta
-of therian mammals. Indeed, adaptive evolution of \textit{LGALS1} on
-the eutherian stem is likely due to the increased invasive nature of
-eutehrian placentation rather than the origination of maternal-fetal
-tolerance. Although the function of \textit{LGALS1} in the marsupial
-placenta needs to be confirmed, it is likely that the mediation of
-maternal-fetal immune tolerance by galectins is an ancient feature of
-mammalian viviparity.
+\textit{LGALS1} (Supplemental \cref{fig:galectins_evolution_tree}). In
+addition, \textit{LGALS2} is also expressed. The presence of galectins
+at the maternal-fetal interface in \opossum implies that it was also
+present in the ancestral placenta of therian mammals. Indeed, adaptive
+evolution of \textit{LGALS1} on the eutherian stem is likely due to
+the increased invasive nature of eutehrian placentation rather than
+the origination of maternal-fetal tolerance. Although the function of
+\textit{LGALS1} in the marsupial placenta needs to be confirmed, it is
+likely that the mediation of maternal-fetal immune tolerance by
+galectins is an ancient feature of mammalian viviparity.
 

### 2016-01-28 manuscript/supplemental_material.Rnw
move RLN2, OVOS2, LGAL1, LGALS13 trees to supplmenetal

@@
 
-\subsection{Top 5\% Transcripts Enrichment}
+\subsection{Top 1\% Transcripts Enrichment}
 
@@
 \subsection{Gene Evolution Trees}
+
+\begin{figure}
+<<rln2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="RLN2" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{\textit{RLN2}}}
+\label{fig:rln2_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<ovos2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree["ENSDNOG00000033320",tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSDNOG00000033320","ENSMODG00000017988"))
+@ 
+\caption{\genetreeexplanation{\cow \textit{OVOS2}}}
+\label{fig:ovos2_evolution_tree}
+\end{figure}
 
@@
 
+\begin{figure}
+<<nppb.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="NPPB" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{\textit{NPPB}}}
+\label{fig:nppb_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<nppc.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="NPPC" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{\textit{NPPC}}}
+\label{fig:nppc_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{galectins}}
+\label{fig:galectins_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{\textit{LGALS13}}}
+\label{fig:galectins_lgals13_evolution_tree}
+\end{figure}
+
 \section{Core Placenta Transcriptome}

### 2016-01-29 manuscript/manuscript.Rnw
combine gene tree figures; fix plot to accept vp

@@
 <<plot.defaults,echo=FALSE,results="hide">>=
-grad.def <- list(low = "white", high = "green", space = "Lab", na.value = "gray50")
+grad.def <- list(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = "gray50")
 
@@
 \begin{figure}
-<<psg.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
+<<psg.ctsl.evolution.tree,echo=FALSE,fig.width=16,fig.height=16,warning=FALSE,message=FALSE>>=
+plot.label.size <- 3
+
+pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2)))
 tree.id <-
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1)
-@ 
-\caption{\genetreeexplanation{pregnancy-specific glycoproteins}}
-\label{fig:psg_evolution_tree}
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
+                 vp=viewport(layout.pos.col=1,layout.pos.row=1),
+                 fontsize=plot.label.size
+                 )
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
+                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
+                 vp=viewport(layout.pos.col=2,layout.pos.row=1),
+                 fontsize=plot.label.size
+                 )
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 vp=viewport(layout.pos.col=1,layout.pos.row=2),
+                 fontsize=plot.label.size
+                 )
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,min.fpkm=1,
+                 subtree=c("ENSSSCG00000006792",
+                           "ENSBTAG00000046085"),
+                 vp=viewport(layout.pos.col=2,layout.pos.row=2),
+                 fontsize=plot.label.size
+                 )
+
+popViewport(1)
+grid.text(c("A","B","C","D"),
+          x=unit(c(0,0.5,0,0.5),"npc"),y=unit(c(1,1,0.5,0.5),"npc"),
+          just=c("left","top"),
+          gp=gpar(fontsize="48")
+          )
+@ 
+\caption{\genetreeexplanation{(A) \textit{PSG1}, (B) \textit{CTSL}, (C) \textit{PRSS16}, and (D) \textit{PAG2}}}
+\label{fig:pag2_prss16_ctsl_psg_evolution_tree}
 \end{figure}
@@
 \textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
-\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2}),
-insulin growth factor 2 (\textit{IGF2}) and one of its binding
-proteins \textit{IGFBP3}, and the hormone adrenomedullin
-(\textit{ADM}).
-
-\begin{figure}
-<<serpine.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="SERPINE2" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1)
-@ 
-\caption{\genetreeexplanation{\textit{SERPINE2}}}
-\label{fig:serpine_evolution_tree}
-\end{figure}
+\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
+Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
+2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
+the hormone adrenomedullin (\textit{ADM}).
 
@@
 
-\begin{figure}
-<<cathespin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"))
-@ 
-\caption{\genetreeexplanation{cathespins related to \textit{CTSL}}}
-\label{fig:cathespin_evolution_tree}
-\end{figure}
-
 \subsubsection{\galili Placenta Transcriptome}
@@
 and \Sexpr{gene.expression("spalax_carmeli","Ctsl",complete=TRUE)},
-\cref{fig:cathespin_evolution_tree}), and many prolactin-related genes
+\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), and many prolactin-related genes
 (\Sexpr{gene.expression("spalax_carmeli","Prl4a1",complete=TRUE)},
@@
 glycoprotein family show especially high expression in \cow placenta
-(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_evolution_tree}, and
+(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D, and
 Supplemental \cref{tab:all_bos_taurus_placenta}) including
@@
 the expression levels of this family are higher in \cow than in \sheep
-\cref{fig:pag2_evolution_tree}. Other genes with high expression
+\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D. Other genes with high expression
 include genes involved with regulation of growth (\textit{IGF2},
@@
 (\Sexpr{gene.expression("bos taurus","HSD3B")}).
-
-\begin{figure}
-<<pag2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 subtree=c("ENSSSCG00000006792",
-                           "ENSBTAG00000046085")
-                 )
-@ 
-\caption{\genetreeexplanation{\cow \textit{PAG2}}}
-\label{fig:pag2_evolution_tree}
-\end{figure}
 
@@
 related to the thymus-specific serine protease gene \textit{PRSS16}
-\cref{fig:prss16_evolution_tree} The other 8 \textit{PRSS16}-like
+\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C The other 8 \textit{PRSS16}-like
 genes are also expressed to a lesser degree
@@
 
-\begin{figure}
-<<prss16.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{\textit{PRSS16}}}
-\label{fig:prss16_evolution_tree}
-\end{figure}
-
 \subsubsection{\armadillo Placenta Transcriptome}
@@
 though the expansion is less than that seen \cow and \sheep
-(\cref{fig:pag2_evolution_tree}). There is also high expression of
+(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). There is also high expression of
 hemoglobin subunit beta,
@@
 (\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000002844",complete=TRUE)},
-\cref{fig:cathespin_evolution_tree}), which are known to interact with
+\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), which are known to interact with
 cystatins \cite{Turk.ea2001:Lysosomalcysteineproteasesfacts}. The
@@
 (\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
-also shows high expression in the opossum placenta. The gene
+also shows high expression in \opossum placenta. The gene
 \textit{WFDC2}, a member of the whey acidic protein (WAP) family is
-also expressed to a high degree in the opossum placenta
+also expressed to a high degree in \opossum placenta
 (\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).

### 2016-01-29 common_r_code/plot_tree_matrix.R

@@ -1,7 +1,79 @@
 library("magrittr")
 library("tidyr")
 library("geiger")
-plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offset.ratio=0.5,min.fpkm=NA,subtree=NULL) {
+
+
+don.gheatmap <- function(p, data, offset=0, width=1, low="green", high="red",
+                     color="white", colnames=TRUE, colnames_position="bottom", font.size=4) {
+
+    colnames_position %<>% match.arg(c("bottom", "top"))
+    variable <- value <- lab <- y <- NULL
+    
+    ## if (is.null(width)) {
+    ##     width <- (p$data$x %>% range %>% diff)/30
+    ## }
+
+    ## convert width to width of each cell
+    width <- width * (p$data$x %>% range %>% diff) / ncol(data)
+    
+    isTip <- x <- y <- variable <- value <- from <- to <- NULL
+ 
+    df=p$data
+    df=df[df$isTip,]
+    start <- max(df$x) + offset
+
+    dd <- data[df$label[order(df$y)],]
+    dd$y <- sort(df$y)
+
+    dd$lab <- rownames(dd)
+    ## dd <- melt(dd, id=c("lab", "y"))
+    dd <- gather(dd, variable, value, -c(lab, y))
+    
+    dd$value[is.na(dd$value) | dd$value == ""] <- NA
+
+    V2 <- start + as.numeric(dd$variable) * width
+    mapping <- data.frame(from=dd$variable, to=V2)
+    mapping <- unique(mapping)
+
+    dd$x <- V2
+
+    #if (is.null(color)) {
+        p2 <- p + geom_raster(data=dd, aes(x, y, fill=value), inherit.aes=FALSE)
+##    } else {
+##        p2 <- p + geom_raster(data=dd, aes(x, y, fill=value), color=color, inherit.aes=FALSE)
+##    }
+    if (is(dd$value,"numeric")) {
+        p2 <- p2 + scale_fill_gradient(low=low, high=high, na.value="white")
+    } else {
+        p2 <- p2 + scale_fill_discrete(na.value="white")
+    }
+    
+    if (colnames) {
+        if (colnames_position == "bottom") {
+            y <- min(p$data$y)-0.6
+        } else {
+            y <- max(p$data$y) + 1
+        }
+        p2 <- p2 + geom_text(data=mapping,
+                             aes(x=to, label=from),
+                             angle=90,
+                             hjust=1,
+                             vjust=0.5,
[…]

### 2016-01-29 manuscript/manuscript.Rnw
use phylopic and ggtree for species tree

@@
 species.tree <- 
-    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodona_africana),((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus)))),(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus),((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)))));")
-plot(species.tree,x.lim=18)
-edgelabels("Eutheria",2,frame="none",bg="white",font=1,adj=c(0.75,0),cex=0.9)
-edgelabels("Euarchontoglires",16,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
-edgelabels("Laurasiatheria",7,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
-edgelabels("Atlantogenata",3,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
-edgelabels("Boreoeutheria",6,frame="none",bg="white",font=1,adj=c(0.7,0),cex=0.9)
-edgelabels("Primates",22,frame="none",bg="white",font=1,adj=c(0.5,0),cex=0.9)
-edgelabels("Rodentia",17,frame="none",bg="white",font=1,adj=c(0.7,0),cex=0.9)
+    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
+pp <- ggtree(species.tree)+
+    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
+    scale_x_continuous(limits=c(0,10))
+print(pp %>%
+      ## monodelphis
+      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
+      ## Armadillo
+      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.18)
+      ## elephant
+      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.18)
+      ## dog
+      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.19)
+      ## horse
+      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.18)
+      ## pig
+      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.15)
+      ## sheep
+      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.2)
+      ## cow
+      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.15)
+      ## s carmeli
+      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.15)
+      ## n galili
+      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.15)
+      ## mouse
+      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.2)
+      ## bonobo
+      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=14,width=0.18)
+      ## spider monkey
+      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=12,width=0.25)
+      ## human
+      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
+      )
 @ 

### 2016-01-29 manuscript/manuscript.Rnw
Move more trees to supplemental

@@
 higher expression in the primate lineage
-(\cref{fig:fstl1_evolution_tree},
+(Supplemental \cref{fig:fstl1_evolution_tree},
 $\mathrm{FDR}=\Sexpr{pri.all.lin.spec["FSTL1",`FDR`]}$), is implicated
@@
 lineage by t-test.
-
-\begin{figure}
-<<fstl1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="FSTL1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=5)
-@ 
-\caption{\genetreeexplanation{follistatin-like 1 \textit{FSTL1}}}
-\label{fig:fstl1_evolution_tree}
-\end{figure}
 
@@
 and the pseudogene \textit{PSG10P} showing the least %
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:psg_evolution_tree}) . %
+(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
 These
@@
 
-\begin{table}
-  \footnotesize
-<<rodentia.lineage.specific,echo=FALSE,results="asis">>=
-rod.all.lin.spec <- placenta.lineage.specific.table("rod.all")
-setkey(rod.all.lin.spec,"Human Symbol")
-@ 
-\caption{Genes whose expression changed significantly in the Rodentia
-  order as compared to all other analyzed species.}
-\label{tab:rodentia_lineage_specific}
-\end{table}
-
 We examined the expression of three rodentia, \mouse, \carmeli, and
@@
 changes on the Rodentia lineage in comparison to all other analyzed
-species (\cref{tab:rodentia_lineage_specific}).
+species (Supplemental \cref{tab:rodentia_lineage_specific}).
 
@@
              aes(y=human_name,x=capfirst(Species)))
-      + geom_tile(aes(fill=log10(FPKM)))
+      + geom_raster(aes(fill=log10(FPKM)))
       + scale_x_discrete(expand=c(0,0))
@@
 production of estrogen and progesterone. High relative expression of
-\textit{CYP19A1} (\cref{fig:estrogen_evolution_tree}), the enzyme that
-synthesizes estrogen from androstenedione is seen in \human, \bonobo,
-\cow, and \horse, though not in \sheep, \smonkey. %
+\textit{CYP19A1} (Supplemental \cref{fig:estrogen_evolution_tree}),
+the enzyme that synthesizes estrogen from androstenedione is seen in
+\human, \bonobo, \cow, and \horse, though not in \sheep, \smonkey. %
 %
@@
 (\cite{Simard.ea2005:Molecularbiologyof3beta-hydroxysteroid},
-\cref{fig:progesterone_evolution_tree}) is also highly expressed in
-\human, \bonobo, \cow, \horse, and \dog, but is not highly expressed
-in \smonkey. The diversity of \textit{CYP19A1} and \textit{HSD3B1}
-expression levels in in primates and ungulates seen here points to the
-need for more complete sampling of placentas at different timepoints
-and in different clades.
-
-\begin{figure}
-<<estrogen.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CYP19A1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{estrogen synthesis genes}}
-\label{fig:estrogen_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<progesterone.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="HSD3B1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 subtree=c("ENSMODG00000005385","ENSLAFG00000027896")
-                 )
-@ 
-\caption{\genetreeexplanation{progesterone synthesis genes}}
-\label{fig:progesterone_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<chorionic.somatomammotropin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CSH1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{chorionic somatomammotropin}}
-\label{fig:csh1_evolution_tree}
-\end{figure}
+Supplemental \cref{fig:progesterone_evolution_tree}) is also highly
+expressed in \human, \bonobo, \cow, \horse, and \dog, but is not
+highly expressed in \smonkey. The diversity of \textit{CYP19A1} and
+\textit{HSD3B1} expression levels in in primates and ungulates seen
+here points to the need for more complete sampling of placentas at
+different timepoints and in different clades.
 
@@
 resulted from duplications of \textit{PRL} in the \mouse, \carmeli,
-and \cow (\cref{fig:prolactin_evolution_tree}). All afrotherians yet
+and \cow (Supplemental \cref{fig:prolactin_evolution_tree}). All afrotherians yet
 analyzed contain only one prolactin gene, and prolactin has undergone
@@
 Supplemental \cref{fig:hoxa11_evolution_tree}).
-
-\begin{figure}
-<<prolactin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{prolactin}}
-\label{fig:prolactin_evolution_tree}
-\end{figure}
 
@@
 
-The opossum placenta shows expression of genes that either do not
+The \opossum placenta shows expression of genes that either do not
 exist in eutherians or are not found in the term eutherian placenta
@@
 gland of the tammar wallaby during pregnancy and early lactation
-\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and here in
-relative abundance in the opossum placenta.
+\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and present in
+relative abundance in \opossum placenta.
 

### 2016-01-29 manuscript/supplemental_material.Rnw
move more trees to supplemental

@@
 \begin{figure}
+<<fstl1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="FSTL1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=5)
+@ 
+\caption{\genetreeexplanation{follistatin-like 1 \textit{FSTL1}}}
+\label{fig:fstl1_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<serpine.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="SERPINE2" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1)
+@ 
+\caption{\genetreeexplanation{\textit{SERPINE2}}}
+\label{fig:serpine_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<estrogen.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CYP19A1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{estrogen synthesis genes}}
+\label{fig:estrogen_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<progesterone.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="HSD3B1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSMODG00000005385","ENSLAFG00000027896")
+                 )
+@ 
+\caption{\genetreeexplanation{progesterone synthesis genes}}
+\label{fig:progesterone_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<chorionic.somatomammotropin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CSH1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{chorionic somatomammotropin}}
+\label{fig:csh1_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<prolactin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{prolactin}}
+\label{fig:prolactin_evolution_tree}
+\end{figure}
+
+\begin{figure}
 <<rln2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
@@
 \subsection{Clade specific Expression}
+
+\begin{table}
+  \footnotesize
+<<rodentia.lineage.specific,echo=FALSE,results="asis">>=
+rod.all.lin.spec <- placenta.lineage.specific.table("rod.all")
+setkey(rod.all.lin.spec,"Human Symbol")
+@ 
+\caption{Genes whose expression changed significantly in the Rodentia
+  order as compared to all other analyzed species.}
+\label{tab:rodentia_lineage_specific}
+\end{table}
 

### 2016-01-29 manuscript/supplemental_material.Rnw
include collectors for which we have information

@@
     \toprule
-    Species                       & Experiment Number & Sample Number(s) \\
+    Species                       & Collector & Experiment Number & Sample Number(s) \\
     \midrule
-    \textit{Ateles fusciceps}     &  & \\
-    \textit{Bos taurus}           &  & \\
-    \textit{Canis familiaris}     &  & \\
-    \textit{Dasypus novemcinctus} &  & \\
-    \textit{Equus caballus}       &  & SRR850299,  SRR850300 \\
-    \textit{Loxodonta africana}   &  & \\
-    \textit{Homo sapiens}         & GSE66622 & SRR1850945,
+    \textit{Ateles fusciceps}     &  & & \\
+    \textit{Bos taurus}           & Bob Kreft & & \\
+    \textit{Canis familiaris}     &  & & \\
+    \textit{Dasypus novemcinctus} &  & & \\
+    \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & SRR850299,  SRR850300 \\
+    \textit{Loxodonta africana}   & Kurt Benirschke & & \\
+    \textit{Homo sapiens}         & Karina Richani, \citet{Hughes.ea2015:Evaluatingintra-andinter-individual} & GSE66622 
+                                             & SRR1850945,
                                                SRR1850946, 
@@
                                                SRR1851012 \\
-    \textit{Monodelphis domestica} & & \\
-    \textit{Mus musculus}         &  & SRR243594,
+    \textit{Monodelphis domestica} & K. Smith & & \\
+    \textit{Mus musculus}         & J. Gonzalez, \cite{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
+                                     & SRR243594,
                                        SRR243595,
@@
                                        SRR243593\\
-    \textit{Nannospalax galili}   &  & \\
-    \textit{Ovis aries}           &  & ERR489144, ERR489145\\
+    \textit{Nannospalax galili}   &  & GSE68763 & SRR2016467 \\
+    \textit{Ovis aries}           & \cite{Archibald.ea2010:sheepgenomereferencesequence} & ERR489144, ERR489145\\
     \textit{Pan paniscus}         &  & \\
-    \textit{Spalax carmeli}       &  & \\
-    \textit{Sus scrofa}           &           & SRR543893 \\
+    \textit{Spalax carmeli}       &  & GSE68763 & SRR2016467 \\
+    \textit{Sus scrofa}           & \cite{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 \\
     \bottomrule
@@
 \section{Gene Ontology Analyses}
+
+\begin{table}
+  \footnotesize
+<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
+go.analysis.wide <- 
+    data.table(dcast(data=go.results,
+                     formula=GO.ID+type+Term+Ontology~species,
+                     value.var="FDR"))
+go.analysis.wide <- 
+    go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
+                                         function(x){sum(as.numeric(as.character(x)) <= 0.05)})]
+go.analysis.wide <- 
+    go.analysis.wide[,`max(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
+                                           function(x){if (!any(is.finite(x) & x<=0.05)) {
+                                                           return(NA)
+                                                       } else {
+                                                           max(x[is.finite(x) & x<=0.05],na.rm=TRUE)
+                                                       }})]
+go.analysis.wide.table <- 
+    go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
+                         grepl("Top 1% No Housekeeping",type),
+                     list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
+go.analysis.wide.table <- 
+    go.analysis.wide.table[order(Ontology,-`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
+setnames(go.analysis.wide.table,"GO.ID","GO ID")
+
+print(xtable(go.analysis.wide.table,
+             display=c("d","s","s","s","d","e"),
+             align=  c("c","c","c","c","c","c")
+             ),
+      math.style.exponents=TRUE,
+      include.rownames=FALSE,
+      booktabs=TRUE,
+      floating=FALSE
+      )
+
+@ 
+\caption{GO terms with significant enrichment ($\mathrm{FDR}\le0.05$)
+  in 12 or more of the species analyzed from genes in the top 1\% of
+  expression. $n(\mathrm{FDR}\le0.05)$ is the number of species with
+  significant enrichment ($\mathrm{FDR}\le 0.05$) in the GO term
+  listed. $max(\mathrm{FDR}\le0.05)$ is the maximum FDR value in all
+  species with an FDR of less than $0.05$.}
+\label{tab:go_terms_all_species}
+\end{table}
 
@@
 \label{fig:serpine_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<sslp1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_id=="ENSCAFG00000010638" & species=="canis familiaris",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{\textit{SSLP1}}}
+\label{fig:sslp1_evolution_tree}
 \end{figure}
@@
 
+\begin{table}
+\footnotesize
+<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=
+placenta.go.results <- rbindlist(list(head(n=20,placenta_go_results_mf[classic.fdr < 0.05,]),
+                             head(n=20,placenta_go_results_bp[classic.fdr < 0.05,])))
+placenta.go.results <- placenta.go.results[,Annotated:=NULL]
+setnames(placenta.go.results,"Significant","# Core")
+setnames(placenta.go.results,"classic","p value")
+setnames(placenta.go.results,"classic.fdr","FDR")
+setnames(placenta.go.results,"GO.ID","GO ID")
+setnames(placenta.go.results,"Term","GO Term")
+print(xtable(placenta.go.results,
+             digits=2,
+             display=c("d","s","s","d","f","e","e","s")),
+      include.rownames=FALSE,
+      booktabs=TRUE,
+      floating=FALSE
+      )
+@ 
+\caption{\ac{GO} terms most significantly enriched
+  (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]} MF,
+  \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP)
+  in the core placenta transcriptome as compared to all genes
+  with 1:1 human orthologs in all species. “\# Core” is the
+  number of genes in the core transcriptome for each \ac{GO} term;
+  “Expected” is the number of genes expected in the core
+  transcriptome for each term; “p-value” are the odds of this
+  occurring by chance calculated by the Fisher exact test for
+  each term; “FDR” is the false discovery rate as estimated by
+  the Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure}.}
+\label{tab:core_placenta_go_terms}
+\end{table}
+
 \begin{figure}

### 2016-01-29 common_r_code/plot_tree_matrix.R
use italic fonts in the figures if necessary

@@ -76,7 +76,7 @@ don.gheatmap <- function(p, data, offset=0, width=1, low="green", high="red",
 plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offset.ratio=0.5,min.fpkm=NA,subtree=NULL,vp=NULL,fontsize=3) {
     gene.tree$tip.label <- gsub("^(ENS.+)\\1$","\\1",gene.tree$tip.label)
     gene.tree.table[,species:=capfirst(gsub("_"," ",species))]
-    gene.tree.table[,short.species:=gsub("^(.)[^ ]* +([^ ]*)","\\1. \\2",species)]
+    gene.tree.table[,short.species:=gsub("^(.)[^ ]* +([^ ]{2})[^ ]*","\\1.\\2.",species)]
     gene.tree.table[,symbol_or_id:=ifelse(symbol=="NULL",gene_id,paste0(symbol," ",short.species))]
     ## use symbol gene_id for duplicates
     gene.tree.table[duplicated(symbol_or_id)|duplicated(symbol_or_id,fromLast=TRUE),
@@ -133,12 +133,17 @@ plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offs
 
     text.labels <- p$data[!is.na(p$data$label),]
     text.labels$xpos <- max(p$data$x)+max(p$data$x)*offset.ratio
+    text.labels$face <- "italic"
+    text.labels$face[grepl("^ENS",text.labels$label)] <- "plain"
     print(don.gheatmap(p,data=gene.tree.expression.matrix,
                        width=1.5,
                        offset=max(p$data$x)*offset.ratio,colnames=TRUE)
           + geom_text(aes(x=xpos,
-                          y=y,label=label,
-                          hjust=1),
+                          y=y,
+                          label=label,
+                          hjust=1,
+                          fontface=face
+                          ),
                       size=fontsize,
                       data=text.labels
                       )

### 2016-01-29 manuscript/manuscript.Rnw
move figures to the end
add email addresses

@@
 % Floats at end for journals
-% \usepackage[nolists,figuresfirst,nomarkers,heads,nofighead,notabhead]{endfloat}
-% \AtBeginFigures{\newpage \section*{Figure Legends}\renewcommand{\efloatseparator}{\hbox{}}}
-% \AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
+\usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
+\renewcommand{\efloatseparator}{\mbox{}}
+%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
+%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
 \usepackage[nomargin,inline,draft]{fixme}
@@
 \author[addressref={aff1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
-\author[addressref={aff2}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
+\author[addressref={aff2},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
 \author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL} \fnm{Amy} \snm{Weckle}}
-\author[addressref={aff3}]{\inits{J}\fnm{Jason} \snm{Caravas}}
-\author[addressref={aff4}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
-\author[addressref={aff5}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
-\author[addressref={aff6}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
-\author[addressref={aff7,aff8}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
-\author[addressref={aff9}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
-\author[addressref={aff7}]{\inits{R}\fnm{Roberto} \snm{Romero}}
+\author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}
+\author[addressref={aff4},email={agnewd@dcpah.msu.edu}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
+\author[addressref={aff5},email={kbenirschke@ucsd.edu}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
+\author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
+\author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
+\author[addressref={aff9},email={gunter.wagner@yale.edu}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
+\author[addressref={aff7},email={prbchiefstaff@med.wayne.edu}]{\inits{R}\fnm{Roberto} \snm{Romero}}
 \author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
-% These are all wrong, but we'll change them later.
 \address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
@@
 \address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-%\address[id=affcor]{Derek Wildman <wildmand@illinois.edu>}
 

### 2016-01-29 manuscript/manuscript.Rnw
make pictures slightly smaller
fix plot colors and a few more fixmes

@@
   not included in this study or having expression by \ac{FPKM} less
-  than 1 were excluded from the plotted tree. Heatmap cells in grey
+  than 1 were excluded from the plotted tree. Heatmap cells in white
   are from species where there was no expression of the relevant
@@
 <<plot.defaults,echo=FALSE,results="hide">>=
-grad.def <- list(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = "gray50")
+grad.def <- list(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = NA)
 
@@
 specimen ID, place and time of collection, and collector is shown for
-each sample in \DLA{Supplemental Table 1}. In the case of \bonobo,
+each sample in Supplemental \cref{tab:srr_numbers}. In the case of \bonobo,
 \cow, and \elephant, the placenta was collected at term upon expulsion
@@
 
-\DLA{Add details for new sequences and fix numbers}
+% \DLA{Add details for new sequences and fix numbers}
 
@@
       ## Armadillo
-      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.18)
+      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
       ## elephant
-      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.18)
+      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
       ## dog
-      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.19)
+      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
       ## horse
-      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.18)
+      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
       ## pig
-      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.15)
+      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
       ## sheep
-      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.2)
+      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
       ## cow
-      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.15)
+      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
       ## s carmeli
-      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.15)
+      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
       ## n galili
-      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.15)
+      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
       ## mouse
-      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.2)
+      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
       ## bonobo
-      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=14,width=0.18)
+      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
       ## spider monkey
-      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=12,width=0.25)
+      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
       ## human
@@
 another related to \textit{CGB7}; both are not highly expressed
-(\cref{fig:cgb_evolution_tree}). Among the other top 50 expressed
+(\cref{fig:cga_cgb_evolution_tree}). Among the other top 50 expressed
 genes, there is high expression of keratins (e.g., \textit{KRT19},
@@
 \Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
-\cref{tab:top50_mouse_placenta,fig:cathespin_evolution_tree}).
+\cref{tab:top50_mouse_placenta,fig:pag2_prss16_ctsl_psg_evolution_tree}).
 Finally, \textit{Peg10}, an imprinted gene required for proper
@@
 <<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
-top.horse.genes <- top.genes.table(species.name="equus caballus")
-print(xtable(top.horse.genes),
-      booktabs=TRUE,
-      floating=FALSE,
-      sanitize.text.function=genes.table.sanitize,include.rownames=FALSE)
-setkey(top.horse.genes,"Gene symbol")
+output.top.genes.table(top.genes.table(species.name="equus caballus"))
 @ 
@@
 \cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene which is
-currently un-annotated in \elephant{}. \DLA{Probably consider fixing
-  the unannotation of H19; it seems pretty obvious to me, though it
-  would require fixing the GTF file for elephant with the right
-  \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan
-    prediction} for this region} Five uncharacterized genes are
-expressed to a high degree in the \elephant placenta
-(\cref{tab:top50_elephant_placenta}), three of which
+currently un-annotated in \elephant{}. %
+% \DLA{Probably consider fixing the unannotation of H19; it seems pretty
+%   obvious to me, though it would require fixing the GTF file for
+%   elephant with the right
+%   \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan
+%     prediction} for this region} %
+Five uncharacterized genes are expressed to a high degree in the
+\elephant placenta (\cref{tab:top50_elephant_placenta}), three of
+which
 (\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000028493",complete=TRUE)},
@@
 related to the thymus-specific serine protease gene \textit{PRSS16}
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C The other 8 \textit{PRSS16}-like
-genes are also expressed to a lesser degree
+\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C The other 8
+\textit{PRSS16}-like genes are also expressed to a lesser degree
 (\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000026505",complete=TRUE)}).
@@
 
-\begin{table}
-\footnotesize
-<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=
-placenta.go.results <- rbindlist(list(head(n=20,placenta_go_results_mf[classic.fdr < 0.05,]),
-                             head(n=20,placenta_go_results_bp[classic.fdr < 0.05,])))
-placenta.go.results <- placenta.go.results[,Annotated:=NULL]
-setnames(placenta.go.results,"Significant","# Core")
-setnames(placenta.go.results,"classic","p value")
-setnames(placenta.go.results,"classic.fdr","FDR")
-setnames(placenta.go.results,"GO.ID","GO ID")
-setnames(placenta.go.results,"Term","GO Term")
-print(xtable(placenta.go.results,
-             digits=2,
-             display=c("d","s","s","d","f","e","e","s")),
-      include.rownames=FALSE,
-      booktabs=TRUE,
-      floating=FALSE
-      )
-@ 
-\caption{\ac{GO} terms most significantly enriched
-  (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]} MF,
-  \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP)
-  in the core placenta transcriptome as compared to all genes
-  with 1:1 human orthologs in all species. “\# Core” is the
-  number of genes in the core transcriptome for each \ac{GO} term;
-  “Expected” is the number of genes expected in the core
-  transcriptome for each term; “p-value” are the odds of this
-  occurring by chance calculated by the Fisher exact test for
-  each term; “FDR” is the false discovery rate as estimated by
-  the Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure}.}
-\label{tab:core_placenta_go_terms}
-\end{table}
-
 The
@@
 \cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
-\Cref{tab:core_placenta_go_terms} lists the significant enrichment of
+Supplemental \cref{tab:core_placenta_go_terms} lists the significant enrichment of
 \ac{GO} terms in the core placenta transcriptome; notable terms
@@
 %
-Significant enrichment in genes ($\mathrm{FDR}\le0.05$) in the top 5\%
+Significant enrichment in genes ($\mathrm{FDR}\le0.05$) in the top 1\%
 of \ac{FPKM} values in at least 12 of the
@@
 cell death related pathways among others
-(\cref{tab:go_terms_all_species}). If we instead examine the top 100
+(Supplemental \cref{tab:go_terms_all_species}). If we instead examine the top 100
 genes by \ac{FPKM} in each of the \Sexpr{length(species.names)}
@@
 % metabolism, and lipids.
-
-\begin{table}
-  \footnotesize
-<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
-go.analysis.wide <- 
-    data.table(dcast(data=go.results,
-                     formula=GO.ID+type+Term+Ontology~species,
-                     value.var="FDR"))
-go.analysis.wide <- 
-    go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
-                                         function(x){sum(as.numeric(as.character(x)) <= 0.05)})]
-go.analysis.wide <- 
-    go.analysis.wide[,`max(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
-                                           function(x){if (!any(is.finite(x) & x<=0.05)) {
-                                                           return(NA)
-                                                       } else {
-                                                           max(x[is.finite(x) & x<=0.05],na.rm=TRUE)
-                                                       }})]
-go.analysis.wide.table <- 
-    go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
-                         grepl("Top 1% No Housekeeping",type),
-                     list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
-go.analysis.wide.table <- 
-    go.analysis.wide.table[order(Ontology,-`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
-setnames(go.analysis.wide.table,"GO.ID","GO ID")
-
-print(xtable(go.analysis.wide.table,
-             display=c("d","s","s","s","d","e"),
-             align=  c("c","c","c","c","c","c")
-             ),
-      math.style.exponents=TRUE,
-      include.rownames=FALSE,
-      booktabs=TRUE,
-      floating=FALSE
-      )
-
-@ 
-\caption{GO terms with significant enrichment ($\mathrm{FDR}\le0.05$)
-  in 12 or more of the species analyzed from genes in the top 1\% of
-  expression. $n(\mathrm{FDR}\le0.05)$ is the number of species with
-  significant enrichment ($\mathrm{FDR}\le 0.05$) in the GO term
-  listed. $max(\mathrm{FDR}\le0.05)$ is the maximum FDR value in all
-  species with an FDR of less than $0.05$.}
-\label{tab:go_terms_all_species}
-\end{table}
 
@@
                axis.ticks = element_blank(),
+               panel.background = element_blank(),
                axis.text.x = element_text(angle = 300, hjust = 0, vjust=1,face="italic")),
@@
 
-\subsubsection{\Ac{GH} family}
+\subsubsection{\ac{GH} family}
 
@@
 
+plot.label.size <- 4
+pushViewport(viewport(layout=grid.layout(nrow=1,ncol=2,widths=c(2,3))))
+
 tree.id <-
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{chorionic gonadotropin, alpha subunit}}
-\label{fig:cga_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<chorionic.gonadotropin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 vp=viewport(layout.pos.col=1,layout.pos.row=1),
+                 offset.ratio=0.8,
+                 fontsize=plot.label.size
+                 )
 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{chorionic gonadotropin, beta subunit}}
-\label{fig:cgb_evolution_tree}
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 vp=viewport(layout.pos.col=2,layout.pos.row=1),
+                 subtree=c("ENSMODG00000013768","ENSG00000104818"),
+                 offset.ratio=0.6,
+                 fontsize=plot.label.size)
+popViewport(1)
+grid.text(c("A","B"),
+          x=unit(c(0,1/3),"npc"),y=unit(c(1,1),"npc"),
+          just=c("left","top"),
+          gp=gpar(fontsize="48")
+          )
+@ 
+\caption{\genetreeexplanation{(A) \textit{CGA} (B) \textit{CGB}}}
+\label{fig:cga_cgb_evolution_tree}
 \end{figure}
@@
 we analyzed here, \textit{CGA} is only expressed in the placentas of
-\human and \bonobo (\cref{fig:hormone_genes,fig:cga_evolution_tree}),
+\human and \bonobo (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}),
 which corresponds with the emergence of the chorionic gonadotropins in
@@
 \textit{CGA} is expressed at much higher levels in both species than
-the beta subunits with which its product dimerizes (\cref{fig:hormone_genes,fig:cga_evolution_tree,fig:cgb_evolution_tree}).
+the beta subunits with which its product dimerizes (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}).
 \textit{CGA} is one of several genes noted by
@@
 expression in the fetal portion of the mouse placenta
-(Supplemental \cref{fig:hormone_genes,fig:nppc_evolution_tree,fig:nppb_evolution_tree}).
+(\cref{fig:hormone_genes}, Supplemental \cref{fig:nppc_evolution_tree,fig:nppb_evolution_tree}).
 However, we do find high expression of at least one of these genes in
@@
 
-\begin{figure}
-<<sslp1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_id=="ENSCAFG00000010638" & species=="canis familiaris",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{\textit{SSLP1}}}
-\label{fig:sslp1_evolution_tree}
-\end{figure}
-
 We found evidence of high expression in two novel expansions of genes
@@
 which have very abundant mRNAs
-(\cref{tab:top50_elephant_placenta,fig:prss16_evolution_tree}). It is
+(\cref{tab:top50_elephant_placenta} and \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}C). It is
 possible that this elephant-specific gene family may be involved in

### 2016-01-29 manuscript/supplemental_material.Rnw
fix plot colors and a few more fixmes

@@
   Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes from species
-  not included in this study or having expression by \ac{FPKM} less
+  not included in this study or having expression by FPKM less
   than 1 were excluded from the plotted tree. Heatmap cells in grey

### 2016-01-29 manuscript/supplemental_material.Rnw
remove the GO acronym from supplemental

@@
 @ 
-\caption{\ac{GO} terms most significantly enriched
+\caption{GO terms most significantly enriched
   (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]} MF,
@@
   with 1:1 human orthologs in all species. “\# Core” is the
-  number of genes in the core transcriptome for each \ac{GO} term;
+  number of genes in the core transcriptome for each GO term;
   “Expected” is the number of genes expected in the core

### 2016-01-30 manuscript/supplemental_material.Rnw
fix some of the missing project numbers

@@
 \usepackage{pdflscape}
-\usepackage[backend=biber,natbib=true,hyperref=true,style=reading,autocite=inline]{biblatex}
+\usepackage[backend=biber,natbib=true,hyperref=true,style=nature,citestyle=numeric-comp,autocite=inline]{biblatex}
 \addbibresource{references.bib}
@@
 \begin{table}
-  \begin{tabular}{c c p{3in}}
+  \begin{tabular}{c p{1.5in} c p{2in}}
     \toprule
@@
     \textit{Dasypus novemcinctus} &  & & \\
-    \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & SRR850299,  SRR850300 \\
+    \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & GSM1139275 & SRR850299,  SRR850300 \\
     \textit{Loxodonta africana}   & Kurt Benirschke & & \\
@@
     \textit{Monodelphis domestica} & K. Smith & & \\
-    \textit{Mus musculus}         & J. Gonzalez, \cite{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
+    \textit{Mus musculus}         & J. Gonzalez, \citet{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
                                      & SRR243594,
@@
     \textit{Nannospalax galili}   &  & GSE68763 & SRR2016467 \\
-    \textit{Ovis aries}           & \cite{Archibald.ea2010:sheepgenomereferencesequence} & ERR489144, ERR489145\\
+    \textit{Ovis aries}           & \citet{Archibald.ea2010:sheepgenomereferencesequence} & ERP005642 & ERR489144, ERR489145\\
     \textit{Pan paniscus}         &  & \\
     \textit{Spalax carmeli}       &  & GSE68763 & SRR2016467 \\
-    \textit{Sus scrofa}           & \cite{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 \\
+    \textit{Sus scrofa}           & \citet{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 \\
     \bottomrule

### 2016-02-01 manuscript/manuscript.Rnw
add competing interests and acknowledgements

@@
 
+  Amy Weckle performed all of the RNA extractions and managed the
+  samples.
+
+  Jason Caravas provided samples.
+
+  Dalen Agnew provided samples.
+
+  Kurt Benirschke provided samples.
+
+  Sue Savage-Rumbaugh provided samples.
+
+  Chong J. Kim provided samples.
+
+  Günter P. Wagner provided samples.
+
+  Roberto Romero provided samples.
+
+  Derek E. Wildman conceived the study, wrote the paper, and performed
+  analyses.
+
   \section*{Acknowledgements}
-  Text for this section \ldots
+  The authors would like to specially acknowledge those groups whose
+  publicly available data contributed to the completeness of this
+  work.
 

### 2016-02-03 manuscript/manuscript.Rnw
update the abstract

@@
 
-  \parttitle{Results} We profiled the term placenta by using Illumina
-  RNA-Seq technology to sequence the transcriptomes of
-  \Sexpr{star.logs[,length(unique(species))]} species of therian (i.e.
-  marsupials and “placental”) mammals.
-
-  \parttitle{Conclusions}
+  \parttitle{Results} We measured the expression profiles of the term
+  placenta of \Sexpr{star.logs[,length(unique(species))]} species of
+  therian (i.e. marsupials and “placental”) mammals using RNA-Seq. We
+  identified multiple genes and transcripts which are differentially
+  expressed across lineages including prolactin, relaxin, and
+  galectins, as well as identifying the non-housekeeping genes which
+  are core to placenta function (such as components of annexin
+  complexes) in all \Sexpr{star.logs[,length(unique(species))]}
+  species studied.
+
+  \parttitle{Conclusions} While there are multiple non-housekeeping
+  genes which are highly expressed in all placentas examined and are
+  likely to be involved in the basic functions of the mammalian
+  placenta, we see incredible lineage-specific diversity in the
+  expression and expansion of many genes. These lineage-specific genes
+  are likely to play a role in the anatomical variability of the
+  placenta and may also be implicated in important \human placental
+  disorders such as pre-eclampsia.
 

### 2016-02-05 manuscript/supplemental_material.Rnw
add contributors of samples to supplemental material

@@
     \midrule
-    \textit{Ateles fusciceps}     &  & & \\
+    \textit{Ateles fusciceps}     & Kurt Benirschke & & \\
     \textit{Bos taurus}           & Bob Kreft & & \\
-    \textit{Canis familiaris}     &  & & \\
-    \textit{Dasypus novemcinctus} &  & & \\
+    \textit{Canis familiaris}     & Dalen Agnew & & (Bounty Hubler) \\
+    \textit{Dasypus novemcinctus} & Richard Truman & & \\
     \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & GSM1139275 & SRR850299,  SRR850300 \\
     \textit{Loxodonta africana}   & Kurt Benirschke & & \\
-    \textit{Homo sapiens}         & Karina Richani, \citet{Hughes.ea2015:Evaluatingintra-andinter-individual} & GSE66622 
+    \textit{Homo sapiens}         & Nandor Gabor, \citet{Hughes.ea2015:Evaluatingintra-andinter-individual} & GSE66622 
                                              & SRR1850945,
@@
                                        SRR243593\\
-    \textit{Nannospalax galili}   &  & GSE68763 & SRR2016467 \\
+    \textit{Nannospalax galili}   & Eviator Nevo & GSE68763 & SRR2016467 \\
     \textit{Ovis aries}           & \citet{Archibald.ea2010:sheepgenomereferencesequence} & ERP005642 & ERR489144, ERR489145\\
-    \textit{Pan paniscus}         &  & \\
-    \textit{Spalax carmeli}       &  & GSE68763 & SRR2016467 \\
+    \textit{Pan paniscus}         &  William Mintz Fields & & (Teco)\\
+    \textit{Spalax carmeli}       & Eviator Nevo & GSE68763 & SRR2016467 \\
     \textit{Sus scrofa}           & \citet{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 \\

### 2016-02-10 manuscript/supplemental_material.Rnw
update sources for samples

@@
     \textit{Canis familiaris}     & Dalen Agnew & & (Bounty Hubler) \\
-    \textit{Dasypus novemcinctus} & Richard Truman & & \\
+    \textit{Dasypus novemcinctus} & Richard Truman, Vincent J. Lynch& & \\
     \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & GSM1139275 & SRR850299,  SRR850300 \\
@@
                                                SRR1851012 \\
-    \textit{Monodelphis domestica} & K. Smith & & \\
-    \textit{Mus musculus}         & J. Gonzalez, \citet{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
+    \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & & \\
+    \textit{Mus musculus}         & Juan Gonzalez, \citet{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
                                      & SRR243594,
@@
     \textit{Pan paniscus}         & William Mintz Fields & & (Teco)\\
-    \textit{Spalax carmeli}       & Eviator Nevo & GSE68763 & SRR2016467 \\
+    \textit{Spalax carmeli}       & Anne Dupressoir, Eviator Nevo & GSE68763 & SRR2016467 \\
     \textit{Sus scrofa}           & \citet{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 \\

### 2016-02-12 manuscript/manuscript.Rnw
add eviatar nevo to authors list
add derek changes to introduction

@@
 \author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
+\author[addressref={aff11},email={nevo@research.haifa.ac.il}]{\inits{E}\fnm{Eviatar} \snm{Nevo}}
 \author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
@@
 \address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
+\address[id=aff11]{The Institute of Evolution, University of Haifa, Haifa, Israel}
 
@@
   are likely to play a role in the anatomical variability of the
-  placenta and may also be implicated in important \human placental
+  placenta and may also be implicated in important human placental
   disorders such as pre-eclampsia.
@@
 The placenta arose at some point before the origin of therian mammals,
-a group that includes both marsupials and eutherian or “placental”
-mammals. The term “placental mammals” is misleading, as marsupials
-also possess a placenta, although its direct apposition to maternal
-tissues lasts for a proportionally shorter time during gestation
-\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. Marsupials also
-differ from eutherian mammals in that most possess only a
-choriovitelline placenta, produced from the fusion of the chorion and
-yolk sac, during most of gestation. Although some eutherian mammals
-may possess a choriovitelline placenta at some point in development,
-the major site of nutrient exchange in placentals occurs via the
-chorioallantoic placenta, produced from the fusion of the chorion and
-allantois \cite{Mossman.1987:vertebratefetalmembranes}. It is unclear
-whether the first placenta was more like that of marsupials or
-eutherians; however, using phylogenetic methods, the ancestral
-eutherian was reconstructed as possessing a relatively invasive
-hemochorial placenta, in which the maternal and fetal blood supply is
-separated by one tissue layer, as in humans and mouse
+a group that includes both marsupials and eutherian or “placental
+mammals”. The term “placental mammals” is misleading, as extant
+yinothereians (such as \textit{Ornithorhynchus anatinus}, the
+platypus) possess a rudimentary placenta, as do marsupials, although
+the direct apposition to maternal tissues lasts for a proportionally
+shorter time during gestation
+\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
+Marsupials also differ from eutherian mammals in that most possess
+only a choriovitelline placenta, produced from the fusion of the
+chorion and yolk sac, during most of gestation. Although some
+eutherian mammals may possess a choriovitelline placenta at some point
+in development, the major site of nutrient exchange in placentals
+occurs via the chorioallantoic placenta, produced from the fusion of
+the chorion and allantois
+\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
+the first placenta was more like that of marsupials or eutherians;
+however, using phylogenetic methods, the ancestral eutherian was
+reconstructed as possessing a relatively invasive hemochorial
+placenta, in which the maternal and fetal blood supply is separated by
+one tissue layer, as in humans and mouse
 \cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
+\DLA{To discuss with derek}
 
@@
 genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
-Studies of certain model species (i.e., \human, \mouse, \cow) have
-revealed a stunning divergence in the genetics of the placenta,
-especially in later stages of pregnancy
+Studies of certain model species (i.e., \textit{Homo sapiens},
+\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
+divergence in the genetics of the placenta, especially in later stages
+of pregnancy
 \cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
@@
 These include independent expansions of the prolactins in \mouse,
-\carmeli, \galili, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
+\textit{Spalax carmeli}, \textit{Nanospalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
 in \cow, genes related to the thymus-specific serine protease
-\textit{PRSS16} in \elephant, and the expansion of the growth hormone
+\textit{PRSS16} in \textit{Loxodona africana}, and the expansion of the growth hormone
 and chorionic gonadotropin beta genes in hominoid primates
@@
 recruitment of \textit{SIGLEC6} to placental expression in \human
-differs from both \textit{P. troglodytes} and \mouse
+differs from both \textit{Pan troglodytes} and \mouse
 \cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
@@
 One symptom of this maternal-fetal conflict is its potential affect on
-\human health
+health
 \cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
@@
 technology to sequence and quantify the mRNA environment of mammalian
-placentas at term, including that of the \human, \bonobo (the sister
-species to \textit{P. troglodytes}), \smonkey, \mouse, \galili,
-\carmeli, \dog, \cow, \elephant, \armadillo, and \opossum. These
-species represent three major branches of eutherian mammals
-(Euarchontoglires, Laurasiatheria, Afrotheria) as well as marsupials.
-These placentas also represent a wide array of placental morphologies
-as well as traits such as number and size of offspring and length of
-gestation (\cref{tab:gestational_periods}). We use this data, along with
+placentas at term, including that of the \human, \textit{Pan paniscus}
+(the sister species to \textit{P. troglodytes}), \textit{Ateles
+  fusciceps}, \mouse, \galili, \carmeli, \textit{Canis familiaris},
+\cow, \elephant, \textit{Dasypus novemcinctus}, and
+\textit{Monodelphis domesticus}. These species represent four major
+branches of eutherian mammals (Euarchontoglires, Laurasiatheria,
+Afrotheria, and Xenartha) as well as marsupials. These placentas also represent a
+wide array of placental morphologies as well as traits such as number
+and size of offspring and length of gestation
+(\cref{tab:gestational_periods}). We use this data, along with
 existing RNA-Seq data from \mouse
 \cite{Wang.ea2011:surveyfornovelimprinted}, \human
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \horse
-\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
-\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
+\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \textit{Equus caballus}
+\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis aries}
+\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \textit{Sus scrofa}
 \cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
@@
 
+\DLA{Add QC information}
+
 \subsubsection{Additional Sequences}
@@
 %
-The genome of the closely-related \textit{Pan troglodytes} was used as
-a reference for \bonobo. The separation between the two
-recognized Pan species is generally perceived to be $\approx 1$ Mya
+The genome of the closely-related \textit{P. troglodytes} was used as
+a reference for \bonobo. The separation between the two recognized Pan
+species is generally perceived to be $\approx 1$ million years
 \cite{Prüfer.ea2012:bonobogenomecomparedwith}.
@@
 
-All 1:1 orthologues present among \opossum, \textit{Pan troglodytes},
+All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
 \cow, \dog, \elephant, \mouse, and \human were determined using

### 2016-02-18 common_r_code/plot_tree_matrix.R
allow specifying font size and face

@@ -59,7 +59,7 @@ don.gheatmap <- function(p, data, offset=0, width=1, low="green", high="red",
                              angle=90,
                              hjust=1,
                              vjust=0.5,
-                             y=y, size=font.size, inherit.aes = FALSE)
+                             y=y, size=font.size, fontface="italic",inherit.aes = FALSE)
     }
 
     p2 <- p2 + theme(legend.position="top",
@@ -73,7 +73,7 @@ don.gheatmap <- function(p, data, offset=0, width=1, low="green", high="red",
     return(p2)
 }
 
-plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offset.ratio=0.5,min.fpkm=NA,subtree=NULL,vp=NULL,fontsize=3) {
+plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offset.ratio=0.5,min.fpkm=NA,subtree=NULL,vp=NULL,fontsize=3,width=1.5,axis.text.size=5) {
     gene.tree$tip.label <- gsub("^(ENS.+)\\1$","\\1",gene.tree$tip.label)
     gene.tree.table[,species:=capfirst(gsub("_"," ",species))]
     gene.tree.table[,short.species:=gsub("^(.)[^ ]* +([^ ]{2})[^ ]*","\\1.\\2.",species)]
@@ -135,27 +135,32 @@ plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offs
     text.labels$xpos <- max(p$data$x)+max(p$data$x)*offset.ratio
     text.labels$face <- "italic"
     text.labels$face[grepl("^ENS",text.labels$label)] <- "plain"
-    print(don.gheatmap(p,data=gene.tree.expression.matrix,
-                       width=1.5,
-                       offset=max(p$data$x)*offset.ratio,colnames=TRUE)
-          + geom_text(aes(x=xpos,
-                          y=y,
-                          label=label,
-                          hjust=1,
-                          fontface=face
-                          ),
-                      size=fontsize,
-                      data=text.labels
-                      )
-          + theme(legend.title=element_text(),legend.position="top",
-                  legend.margin=unit(0,"mm"),
-                  plot.margin=unit(c(0,0,0,0),"mm"),
-                  panel.grid=element_line(size=0.5,color="gray50")
-                  )
[…]

### 2016-02-18 manuscript/manuscript.Rnw
describe species abbreviation

@@
   are from species where there was no expression of the relevant
-  transcript (or a species where that transcript does not apply).}
+  transcript (or a species where that transcript does not apply). The
+  species names are abbreviated with the first letter of the genus and
+  first two letters of the species (for example, \textit{Homo sapiens}
+  is \textit{H.sa.}). Genes with duplicate gene names in the same
+  species (or no gene named defined) are listed by the Ensemble gene
+  ID.}
 

### 2016-02-18 manuscript/manuscript.Rnw
add changes suggested by derek

@@
 \cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
-\DLA{To discuss with derek}
+\DLA{To discuss maternal/fetal split with derek}
 
@@
 These include independent expansions of the prolactins in \mouse,
-\textit{Spalax carmeli}, \textit{Nanospalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
+\textit{Spalax carmeli}, \textit{Nannospalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
 in \cow, genes related to the thymus-specific serine protease
@@
                              capfirst(gsub(" ","_",alignment.species)),
-                             "/Gene/Summary?g=",name_or_id,"}{",
-                             xtable.sanitize.text(name_or_id),"}")]
+                             "/Gene/Summary?g=",name_or_id,"}{\\textit{",
+                             xtable.sanitize.text(name_or_id),"}}")]
     temp[!is.na(`Approved Name`),
@@
     temp[!is.na(`Approved Symbol`),
-         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Symbol`),"}")]
+         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{\\textit{",xtable.sanitize.text(`Approved Symbol`),"}}")]
     temp[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
@@
     temp[,n_fpkm:=NULL]
-    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM","FPKM SEM"))
-    if (temp[,all(is.na(`FPKM SEM`))]) {
-        temp[,`FPKM SEM`:=NULL]
+    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM","SEM"))
+    if (temp[,all(is.na(`SEM`))]) {
+        temp[,`SEM`:=NULL]
     }
@@
     if (output) {
-        print(xtable(temp.table[order(`p (t-test)`),],
+        display.table <- temp.table[order(`p (t-test)`),]
+        display.table <-
+            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
+                                                  "}")]
+        print(xtable(display.table,
                      display=c("s","s","s","e","g","f"),
@@
         table.align <- c("c","c","p{3in}","r")
-        if (any(colnames(data) %in% "FPKM SEM"))  {
+        if (any(colnames(data) %in% "SEM"))  {
             table.display <- c(table.display,"d")
@@
 plot.label.size <- 3
-
-pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2)))
+axis.text.size <- 3.5
+
+pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2),clip="off"))
 tree.id <-
@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=1),
-                 fontsize=plot.label.size
+                 vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"),
+                 axis.text.size=axis.text.size,
+                 fontsize=plot.label.size,
                  )
@@
                  subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=1),
-                 fontsize=plot.label.size
+                 vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"),
+                 fontsize=4,
+                 width=3,
+                 axis.text.size=axis.text.size,
+                 offset.ratio=1
                  )
@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=2),
-                 fontsize=plot.label.size
+                 vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"),
+                 fontsize=4,
+                 width=3,
+                 axis.text.size=axis.text.size,
+                 offset.ratio=1.2
                  )
@@
                            "ENSBTAG00000046085"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=2),
-                 fontsize=plot.label.size
+                 vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"),
+                 axis.text.size=axis.text.size,
+                 fontsize=plot.label.size,
+                 offset.ratio=0.6
                  )
@@
 related to the chorionic somatomammotropins
-are also expressed at relatively high levels (GH1 
+are also expressed at relatively high levels (\textit{GH1} 
 \Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
+\textit{GH1} 
 \Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
-GH2
+\textit{GH2}
 \Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
+and \textit{GH2}
 \Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
@@
 (\Sexpr{gene.expression("ateles_fusciceps","PSG8")}), %
-a member of the chorionic somatotropins, \textit{GH1} %
+only one member of the chorionic somatotropins, \textit{GH1} %
 (\Sexpr{gene.expression("ateles_fusciceps","GH1")}), %
@@
 is also highly expressed, has no known function, and is duplicated
-multiple times in \textit{Macaca mulatta}. All genes with expression
-greater than 50 \ac{FPKM} are listed in Supplementary
-\cref{tab:all_ateles_fusciceps_placenta}.
+multiple times in \textit{Macaca mulatta} (Ensembl gene tree
+\href{http://useast.ensembl.org/Multi/GeneTree/Image?gt=ENSGT00580000081348}{ENSGT00580000081348}).
+All genes with expression greater than 50 \ac{FPKM} are listed in
+Supplementary \cref{tab:all_ateles_fusciceps_placenta}.
 
@@
 (\Sexpr{gene.expression("spalax_carmeli","Hbb-bh1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Hbb-a1",complete=TRUE)}, and
+\Sexpr{gene.expression("spalax_carmeli","Hba-a1",complete=TRUE)}, and
 \Sexpr{gene.expression("spalax_carmeli","Hbb-bs",complete=TRUE)}), as
@@
 (Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
-Laurasiatherians and Euasiatherians
-(Supplemental \cref{tab:laurasiatheria_eurasiatherian_lineage_specific}).
+Laurasiatherians and Euarchontoglires
+(Supplemental \cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).
 
@@
 
-\subsection{Atlantogenata}
+\subsection{Atlantogenata (Afrotheria + Xenartha)}
 \subsubsection{\elephant Placenta Transcriptome}
@@
 \Sexpr{gene.expression("loxodonta_africana","PAEP",complete=TRUE)}, a
-gene usually expressed in the endometrium of the uterus in humans
+gene predominantly expressed in the endometrium of the uterus in humans
 \cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.
@@
 which functions as an antioxidant. The gene with the second highest
-expression is the hemoglobin subunit epsilon
+expression is the beta hemoglobin subunit epsilon
 (\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000011325",complete=TRUE)}),
@@
 print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
-             aes(y=human_name,x=capfirst(Species)))
+             aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
       + geom_raster(aes(fill=log10(FPKM)))
@@
               axis.ticks = element_blank(),
+              axis.text.y = element_text(face="italic"),
               axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
@@
                panel.background = element_blank(),
+               axis.text.y = element_text(face="italic"),
                axis.text.x = element_text(angle = 300, hjust = 0, vjust=1,face="italic")),
@@
 
-In both ruminants and primates, the placenta is responsible for the
-production of estrogen and progesterone. High relative expression of
-\textit{CYP19A1} (Supplemental \cref{fig:estrogen_evolution_tree}),
-the enzyme that synthesizes estrogen from androstenedione is seen in
-\human, \bonobo, \cow, and \horse, though not in \sheep, \smonkey. %
+In both ruminants and primates, the placenta is capable of producing
+estrogen and progesterone
+\cite{Strauss.ea1996:Placentalsteroidhormonesynthesis}. High relative
+expression of \textit{CYP19A1} (Supplemental
+\cref{fig:estrogen_evolution_tree}), the enzyme that synthesizes
+estrogen from androstenedione is seen in \human, \bonobo, \cow, and
+\horse, though not in \sheep, \smonkey. %
 %
@@
 is the rodent with the lowest expression of \textit{PRL}
-(\Sexpr{gene.expression("nannospalax_galili","Prl",complete=TRUE)},
-\cref{fig:prolactin_evolution_tree}), and \elephant also does not
+(\cref{fig:prolactin_evolution_tree}), and \elephant also does not
 express \textit{HOXA11} highly either
@@
 
-plot.label.size <- 4
+plot.label.size <- 4.5
 pushViewport(viewport(layout=grid.layout(nrow=1,ncol=2,widths=c(2,3))))
@@
                  vp=viewport(layout.pos.col=1,layout.pos.row=1),
-                 offset.ratio=0.8,
-                 fontsize=plot.label.size
+                 offset.ratio=0.7,
+                 fontsize=plot.label.size,
+                 axis.text.size=3.5
                  )
@@
                  subtree=c("ENSMODG00000013768","ENSG00000104818"),
-                 offset.ratio=0.6,
-                 fontsize=plot.label.size)
+                 offset.ratio=1,
+                 fontsize=plot.label.size,
+                 axis.text.size=3.5
+                 )
 popViewport(1)
 grid.text(c("A","B"),
-          x=unit(c(0,1/3),"npc"),y=unit(c(1,1),"npc"),
+          x=unit(c(0,2/5),"npc"),y=unit(c(1,1),"npc"),
           just=c("left","top"),
@@
 expression in the fetal portion of the mouse placenta
-(\cref{fig:hormone_genes}, Supplemental \cref{fig:nppc_evolution_tree,fig:nppb_evolution_tree}).
-However, we do find high expression of at least one of these genes in
-\dog, \cow, and \elephant. The lack of expression in the hemochorial
-placentas of Euarchontoglires may mean that high expression of these
-genes are essential in increasing maternal blood flow in
-epitheliochorial and endotheliochorial placentas.
+(\cref{fig:hormone_genes}, Supplemental
+\cref{fig:nppc_evolution_tree,fig:nppb_evolution_tree}). While we do
+find high expression of at least one of these genes in \dog, \cow, and
+\elephant, the expression pattern does not follow the placental
+organization (endotheliochorial for \dog and \elphant,
+synepitheliochorial for \cow and \sheep), so it is unlikely that
+expression of these genes are essential for specific placental
+organizations.
 
@@
 
-\subsection{Human Specific Expression Shifts and the Evolution of Preeclampsia}
+\subsection{Human Specific Expression Shifts and the Evolution of Pre-eclampsia}
 
@@
 the United States
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Preeclampsia
+\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
 can be attributed to the lack of adequate invasion of maternal tissue,
@@
 
+  Eviatar Nevo provided samples.
+
   Chong J. Kim provided samples.
 
-  Günter P. Wagner provided samples.
-
-  Roberto Romero provided samples.
+  Günter P. Wagner provided samples and wrote the paper.
+
+  Roberto Romero provided samples and conceived the study.
 

### 2016-02-18 manuscript/supplemental_material.Rnw
fix float; gene tables and caching in supplemental
- use float for table captions on top
- fix species table legend
- fix cache location to fix cache conflicts
- use better gene tree font sizes
- clear pages after tables

@@
 % \AtBeginTables{\section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
+\usepackage{float}
+\floatstyle{plaintop}
+\restylefloat{table}
 \usepackage[nomargin,inline,draft]{fixme}
@@
 opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
-opts_chunk$set(cache=TRUE, autodep=TRUE)
+opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")
 options(device = function(file, width = 8, height = 7, ...) {
@@
   \end{tabular}
-  \caption{Reference genomes for aligned species}
+  \caption{Species names, common names, reference genomes and read 
+    lengths for aligned species. An italic species as a reference 
+    indicates that assembled reads were annotated using the corresponding
+    reference genome as explained in the text.}
   \label{tab:ref_genome_aligned_species}
@@
 
+\clearpage
+
 \subsection{Top 100 Transcripts Enrichment}
@@
 
+\clearpage
+
 \subsection{Top 1\% Transcripts Enrichment}
@@
 @
+
+\clearpage
 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=5)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,min.fpkm=5,
+                 axis.text.size=5,
+                 fontsize=8
+                 )
 @ 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,min.fpkm=1,
+                 axis.text.size=5,
+                 fontsize=4)
 @ 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 axis.text.size=5,
+                 fontsize=6)
 @ 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 axis.text.size=5,
+                 fontsize=7)
 @ 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 subtree=c("ENSMODG00000005385","ENSLAFG00000027896")
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSMODG00000005385","ENSLAFG00000027896"),
+                 axis.text.size=5,
+                 fontsize=6
                  )
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 axis.text.size=5,
+                 fontsize=4)
 @ 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 axis.text.size=5,
+                 fontsize=5)
 @ 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 subtree=c("ENSDNOG00000033320","ENSMODG00000017988"))
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSDNOG00000033320","ENSMODG00000017988"),
+                 axis.text.size=5,
+                 fontsize=5)
 @ 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 subtree=c("ENSG00000113494","ENSMODG00000020376"))
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSG00000113494","ENSMODG00000020376"),
+                 axis.text.size=5,
+                 fontsize=7)
 @ 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 subtree=c("ENSBTAG00000014738","ENSECAG00000013229")
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSBTAG00000014738","ENSECAG00000013229"),
+                 axis.text.size=5,
+                 fontsize=5
                  )
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 axis.text.size=5,
+                 fontsize=5)
 @ 
-\caption{Expression of \textit{HBE1} and related genes}
+\caption{\genetreeexplanation{\textit{HBE1}}}
 \label{fig:hbe1_evolution_tree}
@@
 \subsection{Clade specific Expression}
+
+<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("placenta_lineage_specific")))>>=
+load("../data/placenta_lineage_specific")
+source("/home/don/projects/xtable/pkg/R/print.xtable.R")
+
+placenta.lineage.specific.table <- function(analysis,output=TRUE) {
+    temp.table <- 
+        head(placenta.lineage.specific.analysis[analysis.name==analysis &
+                                                p.value <= 0.05][order(p.value),
+                                                                 list(human_name,p.value,fdr,fc),],
+             n=5)
+    temp.table <- data.table(temp.table)
+    setkey(temp.table,"human_name")
+    temp.table <- 
+        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
+                                            `Approved Name`,p.value,fdr,fc)]
+    setnames(temp.table,
+              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
+              )
+    setkey(temp.table,"Human Symbol")
+    if (output) {
+        display.table <- temp.table[order(`p (t-test)`),]
+        display.table <-
+            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
+                                                  "}")]
+        print(xtable(display.table,
+                     display=c("s","s","s","e","g","f"),
+                     ),
+              booktabs=TRUE,
+              floating=FALSE,
+              math.style.exponents=TRUE,
+              sanitize.text.function=genes.table.sanitize,
+              include.rownames=FALSE)
+
+    }
+    return(invisible(temp.table[order(`p (t-test)`),]))
+}
+output.top.genes.table <- function(data,output=TRUE) {
+    if (output) {
+        table.display <- c("s","s","s","d")
+        table.align <- c("c","c","p{3in}","r")
+        if (any(colnames(data) %in% "SEM"))  {
+            table.display <- c(table.display,"d")
+            table.align <- c(table.align,"r")
+        }
+        if (any(colnames(data) %in% "Human Symbol"))  {
+            table.align <- c("c",table.align)
+            table.display <- c("s",table.display)
+        }
+        print(xtable(data,
+                     display=table.display,
+                     align=table.align
+                     ),
+              booktabs=TRUE,
+              floating=FALSE,
+              sanitize.text.function=genes.table.sanitize,
+              include.rownames=FALSE)
+    }
+}
+
+@
 
@@
 @ 
-\caption{Genes whose expression changed significantly in the Rodentia
+\caption{Genes whose expression changed in the Rodentia
   order as compared to all other analyzed species.}
@@
 \caption{Genes whose expression changed between the
-  Euasiatherian and Laurasiatherian superorder.}
-\label{tab:laurasiatheria_eurasiatherian_lineage_specific}
+  Euarchontoglires and Laurasiatherian superorder.}
+\label{tab:laurasiatheria_euarchontoglires_lineage_specific}
 \end{table}

### 2016-02-19 manuscript/supplemental_material.Rnw
load libraries earlier and clearpage before sections

@@
 \usepackage{cleveref}
+\usepackage{xspace}
 \newenvironment{narrow}[2]{%
@@
 library("org.Hs.eg.db")
-
-@
-
-<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
+library("magrittr")
+library("tidyr")
+
+@
+
+<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
 for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R"))) {
@@
 
+\clearpage
+
 \section{Transcriptome statistics}
@@
 
+\clearpage
+
 \section{Gene Ontology Analyses}
 
-\begin{table}
-  \footnotesize
 <<go.analysis.all.tissue,echo=FALSE,results="asis">>=
@@
              display=c("d","s","s","s","d","e"),
-             align=  c("c","c","c","c","c","c")
+             align=  c("c","c","c","c","c","c"),
+             caption="GO terms with significant enrichment ($\\mathrm{FDR}\\le0.05$)
+  in 12 or more of the species analyzed from genes in the top 1\\% of
+  expression. $n(\\mathrm{FDR}\\le0.05)$ is the number of species with
+  significant enrichment ($\\mathrm{FDR}\\le 0.05$) in the GO term
+  listed. $max(\\mathrm{FDR}\\le0.05)$ is the maximum FDR value in all
+  species with an FDR of less than $0.05$.",
+  label="tab:go_terms_all_species"
   ),
@@
   booktabs=TRUE,
-      floating=FALSE
-      )
-
-@ 
-\caption{GO terms with significant enrichment ($\mathrm{FDR}\le0.05$)
-  in 12 or more of the species analyzed from genes in the top 1\% of
-  expression. $n(\mathrm{FDR}\le0.05)$ is the number of species with
-  significant enrichment ($\mathrm{FDR}\le 0.05$) in the GO term
-  listed. $max(\mathrm{FDR}\le0.05)$ is the maximum FDR value in all
-  species with an FDR of less than $0.05$.}
-\label{tab:go_terms_all_species}
-\end{table}
+  floating=FALSE,
+  tabular.environment="longtable",
+  caption.placement="top",
+  size="\\footnotesize"
+  )
+
+@
 
@@
 
+\begin{figure}
+<<nppb.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="NPPB" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{\textit{NPPB}}}
+\label{fig:nppb_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<nppc.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="NPPC" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{\textit{NPPC}}}
+\label{fig:nppc_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{galectins}}
+\label{fig:galectins_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+@ 
+\caption{\genetreeexplanation{\textit{LGALS13}}}
+\label{fig:galectins_lgals13_evolution_tree}
+\end{figure}
+
+\clearpage
+
 \subsection{Clade specific Expression}
@@
 
-\begin{table}
+\begin{table}[hb]
   \footnotesize
@@
 
+\clearpage
+
 \subsection{Species Expression}
@@
                            capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
-    cat(paste0("\\subsection{",this.caption,"}\n\n"))
+    cat(paste0("\\subsection{\\textit{",capfirst(this.species),"} Placental Gene Expression}\n\n"))
     print(xtable(genes.expression.table(species.name=this.species),
@@
 
-\begin{figure}
-<<nppb.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="NPPB" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{\textit{NPPB}}}
-\label{fig:nppb_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<nppc.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="NPPC" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{\textit{NPPC}}}
-\label{fig:nppc_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{galectins}}
-\label{fig:galectins_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{\textit{LGALS13}}}
-\label{fig:galectins_lgals13_evolution_tree}
-\end{figure}
+\clearpage
 

### 2016-02-19 manuscript/manuscript.Rnw
include magrittr and disable messages from source

@@
 library("scales")
-
-@
-
-<<source.to.latex,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
+library("magrittr")
+library("tidyr")
+
+@
+
+<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
 for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R"))) {
@@
 \begin{fmbox}
-\dochead{Research}
+\dochead{Research}%
 \title{The evolution of gene expression in the term placenta of viviparous mammals}%
@@
 \author[addressref={aff2},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
-\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL} \fnm{Amy} \snm{Weckle}}
+\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL}\fnm{Amy} \snm{Weckle}}
 \author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}

### 2016-02-21 presentation/placenta_viviparous_presentation.Rnw
add presentation

--- 2016-02-19 manuscript/manuscript.Rnw
@@
-%\documentclass[english,12pt]{article}
-\documentclass[doublespacing]{bmcart_dla}
+\documentclass[ignorenonframetext]{beamer}
 \usepackage{fontspec}
-\setmainfont{FreeSans}
+\setmainfont{FreeSerif}
 \setsansfont{FreeSans}
 \setmonofont{FreeMono}
-\setmathrm{FreeSans}
-\setmathsf{FreeSans}
-\setmathtt{FreeSans}
-%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
-\usepackage[letterpaper]{geometry}
 \usepackage{url}
-% \usepackage{fancyhdr}
+\usepackage{fancyhdr}
 \usepackage{graphicx}
-% \usepackage[bf]{caption}
+\usepackage[bf]{caption}
 \usepackage{rotating}
 \usepackage{wrapfig}
+\usepackage{fancybox}
 \usepackage{booktabs}
-\usepackage{float}
+% \usepackage{multirow}
+\usepackage{acronym}
+\usepackage{qrcode}
+\usepackage{cleveref}
+\usepackage[absolute,overlay]{textpos}
+\usepackage[backend=biber,natbib=true,hyperref=true,style=nature]{biblatex}
+\addbibresource{references.bib}
+
+\usepackage{texshade}
+\usepackage{tikz}
+\usetikzlibrary{positioning}
+\usetikzlibrary{arrows.meta}
+\usepackage{xspace}
+\usepackage{nameref}
+\usepackage{zref-xr,zref-user}
+\renewcommand*{\bibfont}{\tiny}
+\usepackage{mathtools,cancel}
+
+\renewcommand{\CancelColor}{\color{red}} %change cancel color to red
+
+\makeatletter
+\let\my@cancelto\cancelto %copy over the original cancelto command
+\newcommand<>{\cancelto}[2]{\alt#3{\my@cancelto{#1}{#2}}{\mathrlap{#2}\phantom{\my@cancelto{#1}{#2}}}}
+% redefine the cancelto command, using \phantom to assure that the
+% result doesn't wiggle up and down with and without the arrow
+\makeatother
+
 \usepackage{multirow}
-\usepackage{setspace}
-\usepackage{wrapfig}
-% \usepackage{txfonts}
-%\usepackage[plain]{fancyref}
-\usepackage{acronym}
 \usepackage{array}
-\usepackage{dcolumn}
-\usepackage{adjustbox}
-\usepackage{longtable}
-\usepackage{pdflscape}
-\usepackage{amsmath}
-\usepackage{sfmath}
-\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
-\addbibresource{references.bib}
-% \usepackage[noblocks,auth-sc]{authblk}
-\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
-%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
-% Floats at end for journals
-\usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
-\renewcommand{\efloatseparator}{\mbox{}}
-%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
-%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
-\usepackage[nomargin,inline,draft]{fixme}
-\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
-\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
-% \usepackage[x11names,svgnames]{xcolor}
-% \usepackage{texshade}
-% \usepackage{tikz}
-\usepackage{nameref}
-\usepackage{xspace}
-% this package will stretch the font slightly if necessary.
-\usepackage{microtype}
-\usepackage{zref-xr,zref-user}
-\zxrsetup{toltxlabel}
-\zexternaldocument*{supplemental_material}
-\usepackage[capitalize]{cleveref}
-% \newenvironment{narrow}[2]{%
-%   \begin{list}{}{%
-%       \setlength{\topsep}{0pt}%
-%       \setlength{\leftmargin}{#1}%
-%       \setlength{\rightmargin}{#2}%
-%       \setlength{\listparindent}{\parindent}%
-%       \setlength{\itemindent}{\parindent}%
-%       \setlength{\parsep}{\parskip}}%
-%   \item[]}{\end{list}}
-% \def\newblock{\hskip}
-% \newenvironment{paperquote}{%
-%   \begin{quote}%
-%      \it
-%   }%
-%   {\end{quote}}
-% \renewcommand{\textfraction}{0.15}
-% \renewcommand{\topfraction}{0.85}
-% \renewcommand{\bottomfraction}{0.65}
-% \renewcommand{\floatpagefraction}{0.60}
-%\renewcommand{\baselinestretch}{1.8}
-% \newenvironment{enumerate*}%
-%   {\begin{enumerate}%
-%     \setlength{\itemsep}{0pt}%
-%     \setlength{\parskip}{0pt}}%
-%   {\end{enumerate}}
-% \newenvironment{itemize*}%
-%   {\begin{itemize}%
-%     \setlength{\itemsep}{0pt}%
-%     \setlength{\parskip}{0pt}}%
-%   {\end{itemize}}
-% \setcounter{tocdepth}{3}
-% \oddsidemargin 0.0in 
-% \textwidth 6in
-% \raggedbottom
-% \clubpenalty = 10000
-% \widowpenalty = 10000
-% set a large emergency stretch to avoid issues with overfull boxes
-% \setlength{\emergencystretch}{20em}
-% make the afill smaller
-% \renewcommand\Affilfont{\itshape\small}
+
+\mode<presentation>{ 
+  \usetheme{CambridgeUS}
+  \usecolortheme{crane}
+%   % http://identitystandards.illinois.edu/graphicstandardsmanual/generalguidelines/colors.html
+  \definecolor{ilboldblue}{HTML}{002058}
+  \definecolor{ilboldorange}{HTML}{E87722}
+  \definecolor{ilblue}{HTML}{606EB2}
+  \definecolor{ilorange}{HTML}{D45D00}
+%   \setbeamercolor{alerted text}{fg=ilboldorange}
+%   \setbeamercolor*{palette primary}{fg=ilboldblue,bg=ilorange!70!white}
+%   \setbeamercolor*{palette secondary}{fg=ilboldblue,bg=ilorange!60!white}
+%   \setbeamercolor*{palette tertiary}{bg=ilblue!60!white,fg=ilboldorange}
+%   \setbeamercolor*{palette quaternary}{fg=ilboldblue,bg=ilorange!80!white}
+%   
+%   \setbeamercolor*{sidebar}{fg=ilorange,bg=ilboldblue}
+%   
+%   \setbeamercolor*{palette sidebar primary}{fg=ilblue!10!white,bg=ilorange}
+%   \setbeamercolor*{palette sidebar secondary}{fg=ilorange}
+%   \setbeamercolor*{palette sidebar tertiary}{fg=ilblue}
+%   \setbeamercolor*{palette sidebar quaternary}{fg=ilorange}
+%   
+%   % \setbeamercolor*{titlelike}{parent=palette primary}
+%   \setbeamercolor{titlelike}{parent=palette primary,fg=ilboldblue,bg=ilorange}
+%   \setbeamercolor{frametitle}{fg=ilboldorange,bg=ilblue!20!white}
+%   \setbeamercolor{frametitle right}{fg=ilboldblue,bg=ilorange}
+%   
+%   \setbeamercolor*{separation line}{}
+%   \setbeamercolor*{fine separation line}{}
+%   \setbeamercovered{transparent}  
+  \logo{\begin{tikzpicture}% Pale figure
+      {\node[opacity=0.734]{\IfFileExists{figures/igb_wordmark.pdf}{\includegraphics[width=2cm,height=1cm,keepaspectratio]{figures/igb_wordmark}}{}%
+        };}%
+    \end{tikzpicture}}
+}
+
+\title[Placenta Gene Expression]{The evolution of gene expression in the term placenta of viviparous mammals}
+\author[Don L Armstrong \textit{et al.}]{Don L Armstrong, Michael R.
+  McGowen, Amy Weckle, Jason Caravas, Dalen Agnew, Kurt Benirschke,
+  Sue Savage-Rumbaugh, Eviatar Nevo, Chong J Kim, Günter P. Wagner,
+  Roberto Romero, and Derek E. Wildman}
+\institute[IGB]{Institute for Genomic Biology, Computing Genomes 
+  for Reproductive Health, University of Illinois, Urbana-Champaign}
+\subject{Placenta Expression}
+
+\date{\today}
 \renewcommand\cite[1]{\autocite{#1}}
@@
 \begin{document}
-\raggedright
 
 <<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
-opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
+opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.8\\textheight",out.extra="keepaspectratio")
 opts_chunk$set(cache=TRUE, autodep=TRUE)
@@
 
-% Acronyms
-\input{acronyms}
-
-\begin{frontmatter}
-\begin{fmbox}
-\dochead{Research}%
-\title{The evolution of gene expression in the term placenta of viviparous mammals}%
-%
-\author[addressref={aff1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
-\author[addressref={aff2},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
-\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL}\fnm{Amy} \snm{Weckle}}
-\author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}
-\author[addressref={aff4},email={agnewd@dcpah.msu.edu}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
-\author[addressref={aff5},email={kbenirschke@ucsd.edu}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
-\author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
-\author[addressref={aff11},email={nevo@research.haifa.ac.il}]{\inits{E}\fnm{Eviatar} \snm{Nevo}}
-\author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
-\author[addressref={aff9},email={gunter.wagner@yale.edu}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
-\author[addressref={aff7},email={prbchiefstaff@med.wayne.edu}]{\inits{R}\fnm{Roberto} \snm{Romero}}
-\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
-\address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[id=aff2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
-\address[id=aff3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[id=aff4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
-\address[id=aff5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
-\address[id=aff6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
-\address[id=aff7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
-\address[id=aff8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[id=aff9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
-\address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[id=aff11]{The Institute of Evolution, University of Haifa, Haifa, Israel}
-
-\begin{artnotes}
-\end{artnotes}
-
-\end{fmbox}
-% \twocolumn
-
-\begin{abstractbox}
-\begin{abstract}
-% \section{Abstract}
-  \parttitle{Background} The placenta is arguably the most
-  anatomically variable organ in mammals. Understanding gene
-  expression patterns among placentas from multiple species can
-  illuminate the molecular basis of placenta function as well as
-  potentially enrich our knowledge of the biology of obstetrical
-  syndromes.
-
-  \parttitle{Results} We measured the expression profiles of the term
-  placenta of \Sexpr{star.logs[,length(unique(species))]} species of
-  therian (i.e. marsupials and “placental”) mammals using RNA-Seq. We
-  identified multiple genes and transcripts which are differentially
-  expressed across lineages including prolactin, relaxin, and
-  galectins, as well as identifying the non-housekeeping genes which
-  are core to placenta function (such as components of annexin
-  complexes) in all \Sexpr{star.logs[,length(unique(species))]}
-  species studied.
-
-  \parttitle{Conclusions} While there are multiple non-housekeeping
-  genes which are highly expressed in all placentas examined and are
-  likely to be involved in the basic functions of the mammalian
-  placenta, we see incredible lineage-specific diversity in the
-  expression and expansion of many genes. These lineage-specific genes
-  are likely to play a role in the anatomical variability of the
-  placenta and may also be implicated in important human placental
-  disorders such as pre-eclampsia.
-
-% We identified significant shifts in gene expression along specific
-% evolutionary lineages using a model of character evolution. Genes that
-% showed significant elevation in expression along the human lineage
-% ($\mathrm{FDR}\le 0.01$) were enriched in GO categories such as
-% angiogenesis, tube morphogenesis, cell migration, and regulation of
-% inflammatory response. Three of these genes (KISS1, ADAM12, GPC3)
-% showed exceptionally high expression in humans; these genes are
-% implicated in pre-eclampsia and may indicate that the risk of
-% pre-eclampsia recently emerged on the human lineage.
-
-\textbf{Keywords:} mammal, gene expression, evolution, RNA-seq, placenta
-
-\end{abstract}
-\end{abstractbox}
-\end{frontmatter}
+\IfFileExists{figures/relevant_xkcd.png}{\frame[plain]{\centering \includegraphics[width=\textwidth,height=0.8\textheight,keepaspectratio]{figures/relevant_xkcd.png}
+
+    \url{https://xkcd.com/1605/}}}
+
+\frame[plain]{\titlepage}
 
@@
 
-The main role of the placenta in viviparous mammals is the transfer of
-nutrients, oxygen, and waste between mother and offspring. Although
-these functions are conserved among mammals, the placenta is perhaps
-the most variable organ in terms of structure and form
-\cite{Mossman.1937:comparativemorphogenesisofthe}. Species differ in
-the shape of attachment (i.e., discoid, diffuse, cotyledonary,
-zonary), degree of invasiveness of fetal tissues into maternal tissues
-(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
-interdigitation of maternal-fetal tissue (i.e., villous, folded,
-trabecular, lamellar, labrynthine)
-\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
-mammals differ drastically in the number and size of their offspring,
-as well as the length of gestation, traits often associated with
-specific aspects of placental morphology
-\cite{Capellini.ea2011:Placentationandmaternalinvestment}
-(\cref{tab:gestational_periods}).
-
-The placenta arose at some point before the origin of therian mammals,
-a group that includes both marsupials and eutherian or “placental
-mammals”. The term “placental mammals” is misleading, as extant
-yinothereians (such as \textit{Ornithorhynchus anatinus}, the
-platypus) possess a rudimentary placenta, as do marsupials, although
-the direct apposition to maternal tissues lasts for a proportionally
-shorter time during gestation
-\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
-Marsupials also differ from eutherian mammals in that most possess
-only a choriovitelline placenta, produced from the fusion of the
-chorion and yolk sac, during most of gestation. Although some
-eutherian mammals may possess a choriovitelline placenta at some point
-in development, the major site of nutrient exchange in placentals
-occurs via the chorioallantoic placenta, produced from the fusion of
-the chorion and allantois
-\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
-the first placenta was more like that of marsupials or eutherians;
-however, using phylogenetic methods, the ancestral eutherian was
-reconstructed as possessing a relatively invasive hemochorial
-placenta, in which the maternal and fetal blood supply is separated by
-one tissue layer, as in humans and mouse
-\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
-\DLA{To discuss maternal/fetal split with derek}
-
-Although the morphology of the mammalian placenta has been well
-characterized in a number of species
-\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
-molecular environment is only beginning to be understood from a
-genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
-Studies of certain model species (i.e., \textit{Homo sapiens},
-\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
-divergence in the genetics of the placenta, especially in later stages
-of pregnancy
-\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
-For example, each lineage has evolved expansive placenta-specific gene
-families with expression patterns that can vary in space and time
-\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
-These include independent expansions of the prolactins in \mouse,
-\textit{Spalax carmeli}, \textit{Nannospalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
-in \cow, genes related to the thymus-specific serine protease
-\textit{PRSS16} in \textit{Loxodona africana}, and the expansion of the growth hormone
-and chorionic gonadotropin beta genes in hominoid primates
-\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
-Expression patterns of specific genes also can vary widely between
-species. For example, the expression of orthologous genes in the
-placenta have been found to vary substantially between \human and
-\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
-recruitment of \textit{SIGLEC6} to placental expression in \human
-differs from both \textit{Pan troglodytes} and \mouse
-\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
-
-One explanation for the divergent paths of the placental interface is
-the distinct evolutionary interplay between maternal and fetal
-strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}.
-The maternal strategy is to requisition an adequate degree of
-nutrients to the fetus without risking the health of the mother, while
-the fetal strategy via paternal genes is to extract as many resources
-from the mother as possible. Therefore, an interplay between these
-extremes can set up an intricate “arms race”, involving the imprinting
-of parental alleles, expansion of gene families, and enlistment of
-existing genes to expression at the maternal-fetal interface
-\cite{Haig1993:Geneticconflictsinhuman}.
-
-One symptom of this maternal-fetal conflict is its potential affect on
-health
-\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
-addition to having an invasive hemochorial placenta, \human possess a
-deeply invasive trophoblast layer that can penetrate the myometrium of
-the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
-Defects in placental growth are known to be associated with the
-devastating condition of pre-eclampsia, which can cause hypertension,
-proteinuria, and has the potential to lead to death of the mother
-and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
-Although cases have been documented in \textit{P. troglodytes}
+\begin{frame}{Function and Evolution of Placenta}
+\begin{itemize}
+\item Placenta functions in the transfer of nutrients, oxygen, and
+  waste between mother and offspring
+\item Placenta is highly variable structure and form in mammals, even though major
+  function is conserved
+\item Placenta likely arose before the origin of therian mammals
+\item Form relatively well characterized, but diversity of molecular
+  environment of the placenta is only recently being understood
+\item Possible role of evolutionary interplay between maternal and
+  fetal strategies
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Impacts on Human Health?}
+\begin{itemize}
+\item Humans possess a deeply invasive trophoblast layer that can
+  penetrate the myometrium of the uterus
+  \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
+\item Defects in placental growth can lead to pre-eclampsia, which can
+  cause hypertension, proteinuria, and mortality of mother and/or
+  infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
+\item Pre-eclampsia has been seen in \textit{Pan troglodytes}
 \cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
 \textit{Gorilla gorilla}
-\cite{Thornton.ea1992:Convulsionsinpregnancyin}, pre-eclampsia has a
-far higher incidence in \human pregnancies with as much as 4\% of
-pregnancies being affected in the United States
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. \textit{P.
-  troglodytes} and \textit{G. gorilla} also have deeply invasive
-trophoblasts, although the progression of invasion in great apes may
-be slower than \human
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however,
-large-scale expression studies of great ape placentas have not been
-conducted. The increasing affordability of next-generation sequencing
-technology has enabled the large-scale study of specific tissues from
-a wide array of species to address broader phylogenetic questions in
-the evolution of gene expression
-\cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use RNA-Seq
-technology to sequence and quantify the mRNA environment of mammalian
-placentas at term, including that of the \human, \textit{Pan paniscus}
-(the sister species to \textit{P. troglodytes}), \textit{Ateles
-  fusciceps}, \mouse, \galili, \carmeli, \textit{Canis familiaris},
-\cow, \elephant, \textit{Dasypus novemcinctus}, and
-\textit{Monodelphis domesticus}. These species represent four major
-branches of eutherian mammals (Euarchontoglires, Laurasiatheria,
-Afrotheria, and Xenartha) as well as marsupials. These placentas also represent a
-wide array of placental morphologies as well as traits such as number
-and size of offspring and length of gestation
-(\cref{tab:gestational_periods}). We use this data, along with
-existing RNA-Seq data from \mouse
-\cite{Wang.ea2011:surveyfornovelimprinted}, \human
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \textit{Equus caballus}
-\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis aries}
-\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \textit{Sus scrofa}
-\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
-differences between expression in marsupial and eutherian mammals at
-term, identify the evolution of expression in important placental
-genes such as hormones and galectins, identify novel lineage-specific
-expression patterns, and quantify \human specific differences in gene
-expression that may be related to obstetrical disorders such as
-pre-eclampsia.
-
-\begin{table}
-  \begin{tabular}{c c c c}
-    \toprule
-    Species                       & Gestation (days) & Litter Size  & Source \\
-    \midrule
-    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
-    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
-    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
-    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
-    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
-    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
-    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \bottomrule
-  \end{tabular}
-  \caption{Gestational periods in days and average liter sizes for studied species.}
-  \label{tab:gestational_periods}
-\end{table}
-
-\section{Methods}
-
-\subsection{Collection of Placental Tissue and RNA Extraction}
-
-Fetal tissue was collected from the placenta of eight mammalian
-species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
-[domestic cow], \elephant [African savanna elephant], \opossum [gray
-short-tailed opossum], \galili [blind mole rat from Upper Galilee],
-\carmeli [blind mole rat from northern Israel]). Information on
-specimen ID, place and time of collection, and collector is shown for
-each sample in Supplemental \cref{tab:srr_numbers}. In the case of \bonobo,
-\cow, and \elephant, the placenta was collected at term upon expulsion
-from the mother following delivery. Placental tissue from \human was
-collected upon delivery via a caesarian section. Near-term placental
-tissue from \dog was collected upon the death of the mother from a car
-accident. Placental tissue from \opossum was sampled at fetal stage
-32, roughly one day prior to birth. Only the fetal portion of the
-placenta was sampled at the maternal-fetal interface (i.e. villous
-portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
-diffuse placenta for \opossum). All tissues were collected and stored
-in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
-could be performed. Tissues were homogenized in Trizol, and total RNA
-was isolated following the manufacturer's instructions. RNA was
-further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
-conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
-following the manufacturer’s protocol to remove any contaminating DNA
-from the sample. Total RNA was then evaluated on a Nanodrop 1000
-(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
-(Santa Clara, CA) to confirm concentration and quality of the
-isolation. RNA quality was assessed using A260/A280 and A260/A230
-absorbance values.
-
-% \DLA{Add details for new sequences and fix numbers}
-
-\subsection{Sequencing}
-
-All RNA-Seq mRNA isolation, library construction, and Illumina
-sequencing was performed by the Applied Genomics Technology Center
-(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
-total RNA and the standard Illumina protocol. Paired-end sequencing
-was performed using the Illumina Genome Analyzer II with an insert
-size of 230 base pairs (bp). Read lengths for species are shown in
-Supplemental \cref{tab:ref_genome_aligned_species}.
-
-\DLA{Add QC information}
-
-\subsubsection{Additional Sequences}
-
-For more complete evolutionary comparisons additional sequences of
-placenta in \mouse \cite{Wang.ea2011:surveyfornovelimprinted}, \human
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
-\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
-\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
-\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
-\ac{SRA} (Supplemental \cref{tab:srr_numbers}).
-
-\subsection{Transcriptome Assembly, Alignment, and Quantification}
-
-The quality of reads was evaluated using FASTQC v0.11.2
-(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
-Transcriptomes without a suitable reference genome (\smonkey, \galili,
-\carmeli; italic reference in Supplemental
-\cref{tab:ref_genome_aligned_species}) were assembled 
-using Trinity v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence}
-and annotated using Diamond
-\cite{Buchfink.ea2015:Fastandsensitiveprotein} against either the
-\human or \mouse transcriptome
-(Supplemental \cref{tab:ref_genome_aligned_species}).
-% 
-All transcriptomes were then aligned to the appropriate reference (or
-assembled) genome using STAR v2.4.2a
-\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
-values in \ac{FPKM}
-\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
-quantified using Cufflinks v2.2.1
-\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
-%
-The genome of the closely-related \textit{P. troglodytes} was used as
-a reference for \bonobo. The separation between the two recognized Pan
-species is generally perceived to be $\approx 1$ million years
-\cite{Prüfer.ea2012:bonobogenomecomparedwith}.
-
+\cite{Thornton.ea1992:Convulsionsinpregnancyin}
+\item Far higher incidence in \textit{Homo sapiens}, with as many as 
+  4\% of pregnancies United States affected
+  \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Overall Hypotheses}
+\begin{itemize}
+\item Gene expression patterns correlate with placenta morphology
+\item Altered expression of genes correlates with invasiveness, and
+  therefore pre-eclampsia
+\item Maternal/Fetal evolutionary strategies will be reflected in gene
+  expression patterns
+\end{itemize}
+\end{frame}
+
+\begin{frame}{Species Studied}
 \begin{figure}
@@
 @ 
-\caption{Species tree for organisms studied}
-\label{fig:species_tree}
-\end{figure}
-
-\subsection{Exploration of Individual Transcriptome Data}
-
-To determine enrichment levels of functional categories of genes in
-highly-expressed transcripts in each transcriptome, we performed
-functional annotation cluster analyses per transcriptome using topGO
-\cite{topGO:2.20.0}. Gene Ontology clusters were considered
-significantly enriched if the \ac{FDR} as estimated by the Benjamini
-and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
-$p$ values obtained from the Fisher Exact Test were less than or equal
-to 0.05. Enrichment tests were considered with two gene lists per
-transcriptome: annotated genes with expression levels of the top
-5\% of the frequency distribution of \ac{FPKM} (Supplemental
-\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_nannospalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
-the top 100 genes with the highest \ac{FPKM} (Supplemental
-\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
-Due to the paucity of annotated gene ontology terms for some species,
-gene lists were compared using the \human Gene Ontology
-using the 1:1 \human ortholog obtained
-
-\subsection{Comparative Analysis of Gene Expression}
-
-All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
-\cow, \dog, \elephant, \mouse, and \human were determined using
-Ensembl v.80 gene trees via Biomart
-\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
-package
-\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
-A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
-  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
-in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
-  sapiens" & mean_fpkm >= 10,length(unique(human_name))]} 1:1
-orthologous genes with \ac{FPKM} $\ge 10$ were retained for future
-analyses of gene expression evolution described below.
-
-% We used the software \ac{OUCH} in the R statistical environment to
-% investigate significant shifts in gene expression levels over
-% time\cite{Butler.2004:phlogeneticcomparativeanalysis}. \ac{OUCH}
-% institutes an \ac{OU} model of evolution, which allows for different
-% optima for quantitative trait values along user-specified branches of
-% a phylogenetic tree. The \ac{OU} model explicitly accounts for change
-% in quantitative trait values by introducing two parameters into the
-% model of stochastic Brownian motion of change in trait value along a
-% phylogenetic tree through time. These parameters are $\theta$, an
-% optimum trait value, and $\alpha$, the strength of selection
-% (interpreted as “pull” toward the optimum). In addition, the \ac{OU}
-% model can also integrate evolutionary change by assuming that each
-% user-specified set of braches has its own separate \ac{OU} parameters
-% \cite{Butler.2004:phlogeneticcomparativeanalysis}.
-% 
-% \DLA{The description of the OU process is terrible; it needs to be
-%   completely redone}
-% 
-% Various \ac{OU} branch models were tested against a model in which no
-% variation in optima ($\theta$) among branches was allowed. The
-% following branches were specified in separate models to have a
-% different optimum than all other branches: human, bonobo, mouse, dog,
-% cow, elephant, opossum, human+bonobo, Euarchontoglires, and
-% Boreoeutheria. Likelihood values were computed for each branch model
-% and compared to an \ac{OU} model with one optimum using likelihood
-% ratio tests with df=1. We filtered out genes with universally low
-% expression levels, and included only genes in which expression in at
-% least one species had an $\mathrm{FPKM}\ge10$ as described above. We
-% corrected for multiple testing using the \ac{FDR} and using the
-% Benjamini-Hochberg correction\cite{Benjamini1995:bh_procedure}. We
-% used the time-calibrated phylogenetic tree of
-% \cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
-% not used in this analyses pruned.
-% 
-% We also calculated the enrichment of functional categories in the
-% lists of genes which displayed a significant shift in expression on
-% each branch using the same methodology as in the per-species gene
-% lists \DLA{reference the tables once this analysis is complete}.
-
-\subsection{Correlation of Tissue Specific Index with Lineage Specific Genes}
-
-To evaluate the extent of tissue specificity of each gene, we
-calculated the \ac{TSI} based on human RNASeq data in 3267 samples in
-126 tissues the Roadmap Epigenomics
-\cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping} and GTEx projects
-\cite{Gtexconsortium.2015:Thegenotypetissueexpression} as well
-additional placenta
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} and uterus
-\cite{Chan.ea2014:Assessmentofmyometrialtranscriptome} samples from
-SRA. The \ac{TSI} measures the extent to which a specific gene has
-predominant expression in only one tissue
-\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}. A gene
-expressed in only one tissue would have a \ac{TSI} of 1, whereas a
-gene expressed in all tissues would have a \ac{TSI} of 0.
-
-% \subsection{Validation of Primate Expression using Spider Monkey Placenta Transcriptome}
-% 
-% To verify expression levels in human and bonobo, we also generated
-% RNA-Seq data for the placenta transcriptome of the black-headed spider
-% monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
-% Ateles is outside the clade, which includes hominids and can be used
-% to polarize expression values of highly expressed genes in Homo and
-% Pan. We sampled placental tissue from an Ateles placenta expelled from
-% the mother shortly after birth (see Supplemental Table X). We used
-% similar procedures discussed above to extract and sequence this
-% sample. [DEREK—I’m unsure how to proceed on describing this so as to
-% not step on Natalie’s toes]
-% 
-% \DLA{Complete discussion of the ateles placenta}
-
-\subsection{Sequence Deposition and Code}
-
-All of the code used to produce every analysis presented in this paper
-is available at
-\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
-sequences used are deposited in SRA, and can be obtained using
-\DLA{add GSE label for the experiment}.
+\end{figure}
+\end{frame}
+
+\begin{frame}{Variation in Gestational Periods and Litter Size}
+\tiny
+\begin{table}
+  \begin{tabular}{c c c c}
+    \toprule
+    Species                       & Gestation (days) & Litter Size  & Source \\
+    \midrule
+    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
+    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
+    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
+    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
+    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
+    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
+    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \bottomrule
+  \end{tabular}
+%  \caption{Gestational periods in days and average liter sizes for studied species.}
+%  \label{tab:gestational_periods}
+\end{table}
+\end{frame}
+
+\begin{frame}{Variations in Placenta Morphology}
+\begin{table}
+\begin{tabular}{c c c}
+  \toprule
+  Species & Interhemal Membrane & Pattern \\
+  \midrule
+  \smonkey & hemochorial & trabecular \\
+  \cow     & epitheliochorial & villous \\
+  \dog     & endotheliochorial & labyrinthine \\
+  \armadillo & hemochorial & trabecular \\
+  \horse & epitheliochorial & villous \\
+  \elephant & endotheliochorial & labyrinthine \\
+  \human & hemochorial & villous \\
+% http://onlinelibrary.wiley.com/doi/10.1002/jez.1402660511/pdf
+  \opossum & epitheliochorial & choriovitelline \\
+  \mouse & hemotrichorial & labyrinthine \\
+  \galili & hemotrichorial & labyrinthine \\
+  \sheep & epitheliochorial & villous \\
+  \bonobo & hemochorial & villous \\
+% http://www.biolreprod.org/content/early/2014/10/17/biolreprod.114.124818.short
+  \carmeli & hemotrichorial & labyrinthine \\
+  \pig & epitheliochorial & diffuse \\
+  \bottomrule
+\end{tabular}
+\end{table}
+\end{frame}
+
+\section{Methods}
+
+\subsection{Tissue/Extraction/Sequencing}
+
+\begin{frame}{Collection of Placental Tissue}
+\begin{itemize}
+\item Collected from 8 mammalian species
+\item Only fetal portion was sampled
+\item Sequenced using GAII with an insert size of 230 bp
+\item Four additional species and additional \mouse and \human samples
+  were obtained from SRA
+\end{itemize}
+\end{frame}
+
+\subsection{Transcriptome Assembly, Alignment, and Quantification}
+
+\begin{frame}{Assembly, Alignment, Annotation and Quantification}
+  \begin{itemize}
+  \item QC Using FASTQC v0.11.2
+  \item Transcriptomes without suitable reference (\smonkey, \galili,
+    \carmeli)
+    \begin{itemize}
+    \item assembled using Trinity
+    \item annotated using Diamond against \human or \mouse
+    \end{itemize}
+  \item Aligned to the reference (or assembled genome) using STAR
+  \item Expression values were quantified using Cufflinks
+  \end{itemize}
+\end{frame}
+
+\subsection{Additional Methods}
+\begin{frame}{Additional Methods}
+  All of the code used to produce every analysis presented in this
+  poster is available at
+  \url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of
+  the sequences used are or will be deposited in SRA.
+\end{frame}
 
@@
 
-Supplemental \cref{fig:transcriptome_statistics} shows the RNAseq read and
-alignment statistics for all species. For transcriptomes sequenced
-with 75 bp %
-%
-(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
-%
-,
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
-with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
-mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
-%
-For transcriptomes sequenced with 150 bp, %
-% The mismatch here is because we're reporting trimmed reads here, not
-% full length reads
-(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
-%
-,
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
-%
-with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
-of reads mapping to the
-respective reference genome (\cref{tab:ref_genome_aligned_species}).
-%
-%
-For transcriptomes sequenced with 35 bp, %
-% The mismatch here is because we're reporting trimmed reads here, not
-% full length reads
-(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
-%
-,
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
-%
-with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
-of reads mapping to the
-respective reference genome (\cref{tab:ref_genome_aligned_species}).
-The top 50 expressed genes in each genome analyzed are shown in
-\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
-and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_nannospalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.
-
-% The percentage of bases that mapped to Ensembl annotated mRNA ranged
-% from ~35\% in elephant to ~85\% in human, with percentages of bases
-% mapping to annotated protein-coding segments of genes ranging from
-% ~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
-% from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
-% from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
-% ranged from ~43\% in elephant to 80\% in bonobo. The number of
-% annotated transcripts to which reads mapped in each respective genomes
-% ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
-% and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
-% greater degree of mapping to annotated protein-coding sequences in
-% humans corresponds to the greater number of Ensembl annotated
-% sequences in the human genome compared to the other five reference
-% genomes (Flicek, 2012).
+\begin{frame}{Transcriptome Mapping Statistics: Reads}
+\begin{figure}
+<<transcriptome.mapping.figure,fig.width=10,fig.height=8,echo=FALSE>>=
+print(ggplot(star.logs[grepl("input reads$",field),list(reads=as.numeric(value),species=capfirst(species))],
+             aes(x=species,y=reads))
+      + coord_flip()
+      + geom_boxplot()
+      + scale_y_log10(breaks=c(1e5,1e6,1e7,1e8,1e9,1e10),labels=trans_format("log10", math_format(10^.x)))
+      + ylab("Reads per Sample")
+      + xlab("Species")
+      + theme_bw()
+      + theme(legend.position="top",
+              axis.text.x = element_text(size=10,face="italic"))
+      )
+@ 
+\end{figure}
+\end{frame}
+
+\begin{frame}{Transcriptome Mapping Statistics: Unmapped}
+\begin{figure}
+<<transcriptome.mapping.figure.unmapped,fig.width=8,fig.height=8,echo=FALSE>>=
+star.logs.unmapped <- 
+    star.logs[grepl("%",field) & 
+                  grepl("unmapped",field) & 
+                      !grepl("too many mismatches",field),]
+star.logs.unmapped <- 
+    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
+print(ggplot(star.logs.unmapped,
+             aes(x=capfirst(species),y=as.numeric(gsub("%","",value))))
+      + geom_boxplot()
+      + facet_grid(field~.)
+      + ylab("Percent of Reads")
+      + xlab("Species")
+      + theme_bw()
+      + coord_flip()
+      + theme(legend.position="top",
+              axis.text.x = element_text(face="italic"))
+      )
+@ 
+\end{figure}
+\end{frame}
 
 <<top.genes.function,echo=FALSE,results="hide">>=
-top.genes.table <- function(species.name,n=50) {
-    alignment.species <- species.name
-    if (alignment.species=="pan paniscus") {
-        alignment.species <- "pan troglodytes"
-    } else if (alignment.species=="ateles fusciceps") {
-        alignment.species <- "homo sapiens"
-    } else if (grepl("spalax",alignment.species)) {
-        alignment.species <- "mus musculus"
-    }
-    temp <- 
-        combined.fpkm[species==species.name &
-                      !(!is.na(human_name) & 
-                        human_name %in% housekeeping.genes.superset[,gene_short_name]) &
-                      ! (gene_biotype %in% c("Mt_tRNA","Mt_rRNA","rRNA")) &
-                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_name) &
-                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_alignment_symbol) &
-                      !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[lL]?(?:-.+)?$|rRNA$)",
-                             gene_short_name)
-                      ][order(-mean_fpkm)][,list(name_or_id,
-                                                 human_alignment_symbol,
-                                                 human_name,mean_fpkm,sd_fpkm,n_fpkm)]
-    ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
-    temp <- temp[!(is.na(name_or_id)|name_or_id==""),]
-    temp <- temp[!duplicated(name_or_id)][1:n]
-    temp[,used_alignment:=FALSE]
-    temp[is.na(human_name) & !is.na(human_alignment_symbol),
-                    used_alignment:=TRUE]
-    temp[is.na(human_name),human_name:=human_alignment_symbol]
-    setkey(temp,"human_name")
-    temp <- 
-        hugo.nomenclature[temp][order(-mean_fpkm),
-                                list(name_or_id,used_alignment,`Approved Symbol`,`Approved Name`,
-                                     mean_fpkm,sd_fpkm,n_fpkm)]
-    temp[!is.na(name_or_id),
-         name_or_id:=paste0("\\href{http://www.ensembl.org/",
-                             capfirst(gsub(" ","_",alignment.species)),
-                             "/Gene/Summary?g=",name_or_id,"}{\\textit{",
-                             xtable.sanitize.text(name_or_id),"}}")]
-    temp[!is.na(`Approved Name`),
-         `Approved Name`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Name`),"}")]
-    temp[!is.na(`Approved Symbol`),
-         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{\\textit{",xtable.sanitize.text(`Approved Symbol`),"}}")]
-    temp[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
-    temp[,used_alignment:=NULL]
-    temp[,sem_fpkm:=as.numeric(NA)]
-    temp[n_fpkm>1,sem_fpkm:=sd_fpkm/sqrt(n_fpkm)]
-    temp[,sd_fpkm:=NULL]
-    temp[,n_fpkm:=NULL]
-    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM","SEM"))
-    if (temp[,all(is.na(`SEM`))]) {
-        temp[,`SEM`:=NULL]
-    }
-    if (species.name=="homo sapiens" || species.name=="ateles fusciceps") {
-        temp[,`Human Symbol`:=NULL]
-    }
-    temp[,`FPKM`:=round(`FPKM`)]
-    return(temp)
-}
+
 gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE) {
@@
                      display=c("s","s","s","e","g","f"),
+                     columns=c("c","c","p{1.5in}","c","c","c")
                      ),
@@
 }
-output.top.genes.table <- function(data,output=TRUE) {
-    if (output) {
-        table.display <- c("s","s","s","d")
-        table.align <- c("c","c","p{3in}","r")
-        if (any(colnames(data) %in% "SEM"))  {
-            table.display <- c(table.display,"d")
-            table.align <- c(table.align,"r")
-        }
-        if (any(colnames(data) %in% "Human Symbol"))  {
-            table.align <- c("c",table.align)
-            table.display <- c("s",table.display)
-        }
-        print(xtable(data,
-                     display=table.display,
-                     align=table.align
-                     ),
-              booktabs=TRUE,
-              floating=FALSE,
-              sanitize.text.function=genes.table.sanitize,
-              include.rownames=FALSE)
-    }
-}
-
-@
+
+@
+
+\subsection{Core Placenta Transcriptome}
+
+\begin{frame}{Core Placenta Transcriptome}
+\begin{figure}
+<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
+print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
+             aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
+      + geom_raster(aes(fill=log10(FPKM)))
+      + scale_x_discrete(expand=c(0,0))
+      + scale_y_discrete(expand=c(0,0))
+      + scale_fill_gradient(low = grad.def[["low"]],
+                            high = grad.def[["high"]],
+                            space = grad.def[["space"]],
+                            na.value = grad.def[["na.value"]],
+                            limits=c(0,4),
+                            guide=guide_colorbar(title=expression(log[10](FPKM)))
+                            )
+      + ylab("Human Ortholog Name")
+      + xlab("Species")
+      + theme_grey()
+      + theme(legend.position="top",
+              axis.ticks = element_blank(),
+              axis.text.y = element_text(face="italic"),
+              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
+      )
+@ 
+\end{figure}
+\end{frame}
+
+\begin{frame}{Core placenta transcriptome}
+\begin{itemize}
+
+\item We considered a gene to be core to the placenta transcriptome if
+  a 1:1 ortholog of that gene was expressed with
+  $\mathrm{FPKM} \ge 10$ in all species studied and was not a human
+  housekeeping gene according to
+  \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
+
+\item Multiple components of annexin complexes are included in the core
+transcriptome, including \textit{ANXA2}, \textit{ANXA1},
+\textit{S100A11}, \textit{S100A10}.
+
+\item ANXA1 is involved in the resolution of the inflammation response
+  through inhibition of phospholipase A2 activity and signaling
+  through the formyl peptide receptor family, and has previously been
+  hypothesized to be important for the maintenance of the
+  anti-inflammatory during pregnancy
+  \cite{Hutchinson.ea2011:Molecularregulatorsofresolution}.
+
+\item ANXA2 is known to interact with S100A10, S100A6, S100A11, and
+  S100A4 \cite{Liu.ea2015:AnnexinA2complexeswith}, and there is
+  evidence for its interaction with S100P as well; it is involved in
+  cell-cell interaction as well as vesicle trafficking and von
+  Willibrand Factor secretion.
+
+\item The expression regulation of another core gene, \textit{TIMP3},
+  has been implicated in pre-eclampsia
+  \cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
+\end{itemize}
+
+\end{frame}
 
@@
 
+\begin{frame}{Primate Placenta Transcriptome}
 \begin{table}
@@
 \end{table}
-
-We examined the expression of three primates, \human, \bonobo, and
-\smonkey. The genes whose expressions are different on the primate
-lineage in comparison to all other analyzed species are listed in
-\cref{tab:primate_lineage_specific}. \textit{FSTL1} has significantly
-higher expression in the primate lineage
-(Supplemental \cref{fig:fstl1_evolution_tree},
-$\mathrm{FDR}=\Sexpr{pri.all.lin.spec["FSTL1",`FDR`]}$), is implicated
-in BMP signaling, and has previously been implicated in inflammation
-and invasiveness
-\cite{Mouillet.ea2015:expressionandpost-transcriptionalregulation}. Of
-non-primates, \horse shows the highest expression
-%
-\Sexpr{gene.expression(species.name="equus caballus",gene="FSTL1",complete=TRUE)}, 
-%
-which is still lower than the lowest primate, \smonkey
-%
-(\Sexpr{gene.expression(species.name="ateles fusciceps",gene="FSTL1",complete=TRUE)}). 
-%
-No other genes were significantly differentially expressed on this
-lineage by t-test.
-
-\subsubsection{\human Placenta Transcriptome}
-\label{sec:human_placenta}
-
-\begin{table}
-  \footnotesize
-<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="homo sapiens"))
-@ 
-\caption{The top 50 genes by \ac{FPKM} in \human which are not
-  annotated as Mt\_rRNA, Mt\_tRNA, rRNA or a housekeeping gene. Gene
-  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
-  name corresponding to the human symbol. A complete list of all genes
-  is in \cref{tab:all_homo_sapiens_placenta}.}
-\label{tab:top50_human_placenta}
-\end{table}
-
-<<cgb.min.expression,results="hide",echo=FALSE,warnings=FALSE>>=
-cgb.min.expression <- 
-    min(sapply(paste0("CGB",c("","1","2","5","7","8")),
-               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
-cgb.sum.expression <- 
-    sum(sapply(paste0("CGB",c("","1","2","5","7","8")),
-               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
-@
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \human are shown in
-\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
-in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
-the highest \ac{FPKM} value
-%
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
-%
-in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
-three of which are members of a distinct expansion of the growth
-hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
-hormone gene family, \ac{GH2} (\ac{FPKM}=%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
-is also expressed to a high degree in the \human placenta
-(\cref{tab:top50_human_placenta}).
-% 
-Eleven \acp{PSG} are highly-expressed
-(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
-greatest degree of expression %
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
-and the pseudogene \textit{PSG10P} showing the least %
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
-These
-genes are members of a hominoid-specific expansion of the \ac{CEA} family
-and known to be expressed by the syncytiotrophoblast
-\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
-%
-Genes involved in the production of estrogens (%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
-% 
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
-% 
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
-% 
-) and progesterone (%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
-\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
-\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
-%
-show exceptionally high expression in the human placenta, as well as %
-%
-\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
-%
-which generates the alpha subnit of multiple human chorionic
-gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
-producing the beta subunits of the chorionic gonadotropins
-(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
-\textit{CGB7}, \textit{CGB8}) have reasonable expression
-($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
-expression is on the order of the expression of expression of
-\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
-%
-%
-Other highly expressed genes in the top 50 include members of the
-hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
-\textit{HBA2}), the maternally imprinted long noncoding RNA gene
-\textit{H19}, genes coding for proteins required for growth and/or
-cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
-\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
-glucose transport (\textit{SLC2A1}) among others
-(\cref{tab:top50_human_placenta}, Supplemental
-\cref{tab:all_homo_sapiens_placenta}).
-
-\begin{figure}
-<<psg.ctsl.evolution.tree,echo=FALSE,fig.width=16,fig.height=16,warning=FALSE,message=FALSE>>=
+\end{frame}
+
+\begin{frame}{PSG1}
+\begin{figure}
+<<psg.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 plot.label.size <- 3
@@
 
-pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2),clip="off"))
 tree.id <-
@@
                  )
-
+@ 
+\end{figure}
+\end{frame}
+
+\begin{frame}{CTSL}
+\begin{figure}
+<<ctsl.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+plot.label.size <- 3
+axis.text.size <- 3.5
 tree.id <-
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"),
-                 fontsize=4,
-                 width=3,
-                 axis.text.size=axis.text.size,
-                 offset.ratio=1.2
-                 )
-
+@ 
+\end{figure}
+\end{frame}
+
+\begin{frame}{PAG2}
+\begin{figure}
+<<pag2.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 tree.id <-
@@
 
-popViewport(1)
-grid.text(c("A","B","C","D"),
-          x=unit(c(0,0.5,0,0.5),"npc"),y=unit(c(1,1,0.5,0.5),"npc"),
-          just=c("left","top"),
-          gp=gpar(fontsize="48")
-          )
-@ 
-\caption{\genetreeexplanation{(A) \textit{PSG1}, (B) \textit{CTSL}, (C) \textit{PRSS16}, and (D) \textit{PAG2}}}
-\label{fig:pag2_prss16_ctsl_psg_evolution_tree}
-\end{figure}
-
-\subsubsection{\bonobo Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
-top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
-output.top.genes.table(top.bonobo.genes)
-setkey(top.bonobo.genes,"Gene symbol")
-@ 
-\caption{\tableexplanation{\bonobo}{tab:all_pan_paniscus_placenta}}
-\label{tab:top50_bonobo_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
-with a list of all genes in Supplemental
-\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
-placenta differs somewhat from that of \human, as evidenced by the
-makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
- in \bonobo is %
-\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
-%
-which is involved in insulin growth factor signaling and angiogenesis
-during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
-\textit{PRG2} is also expressed at relatively high levels (though
-comparatively much lower) in the human placenta
-(\Sexpr{gene.expression("homo sapiens","PRG2")};
-Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
-related to the chorionic somatomammotropins
-are also expressed at relatively high levels (\textit{GH1} 
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
-\textit{GH1} 
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
-\textit{GH2}
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
-and \textit{GH2}
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
-%
-\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
-Placental expression of the genes involved in estrogen and
-progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
-respectively) show lower expression in \bonobo as compared to \human %
-(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
-\cref{fig:estrogen_evolution_tree}; %
-\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
-\cref{fig:progesterone_evolution_tree} and
-\cref{tab:top50_human_placenta}) although the expression of
-\textit{HSD3B1} is fairly similar to that of \smonkey %
-(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
-In addition, \textit{CGA} is expressed in the bonobo placenta %
-(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
-lower levels than \human as well (\cref{tab:top50_human_placenta}).
-%
-Two Ensembl-annotated beta subunits of the chorionic gonadotropin
-genes are expressed in \bonobo, one related to \textit{CGB} and
-another related to \textit{CGB7}; both are not highly expressed
-(\cref{fig:cga_cgb_evolution_tree}). Among the other top 50 expressed
-genes, there is high expression of keratins (e.g., \textit{KRT19},
-\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
-\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
-Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
-2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
-the hormone adrenomedullin (\textit{ADM}).
-
-\subsubsection{\smonkey Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
-top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
-output.top.genes.table(top.spider.monkey.genes)
-setkey(top.spider.monkey.genes,"Gene symbol")
-@ 
-\caption{\tableexplanation{\smonkey}{tab:all_ateles_fusciceps_placenta}}
-\label{tab:top50_spider_monkey_placenta}
-\end{table}
-
-For the term placenta of \smonkey, assembled contigs from RNA-seq were
-aligned using against \human and the best alignment was chosen as the
-annotation for each assembled contig. In the term placenta of
-\smonkey, the top 50 non-housekeeping genes
-(\cref{tab:top50_spider_monkey_placenta}) include a possible regulator
-of androgen receptor signaling
-\cite{Sampson.ea2012:PAGE4positivityisassociated}, \textit{PAGE4} %
-(\Sexpr{gene.expression("ateles_fusciceps","PAGE4")}), %
-the pregnancy specific glycoprotein \textit{PSG8}
-(\Sexpr{gene.expression("ateles_fusciceps","PSG8")}), %
-only one member of the chorionic somatotropins, \textit{GH1} %
-(\Sexpr{gene.expression("ateles_fusciceps","GH1")}), %
-a few HLA genes (\textit{HLA-A}, \textit{HLA-F},
-\cref{tab:top50_spider_monkey_placenta}), and multiple components of
-annexin complexes
-(\Sexpr{gene.expression("ateles_fusciceps","ANXA2",complete=TRUE)}, %
-\Sexpr{gene.expression("ateles_fusciceps","S100P",complete=TRUE)}, %
-\Sexpr{gene.expression("ateles_fusciceps","S100A11",complete=TRUE)}) %
-which may play a role in angiogenesis, and certainly regulate
-cell-cell interaction in the developing placenta
-\cite{Liu.ea2015:AnnexinA2complexeswith}.
-\Sexpr{gene.expression("ateles_fusciceps","AL365202.1",complete=TRUE)}
-is also highly expressed, has no known function, and is duplicated
-multiple times in \textit{Macaca mulatta} (Ensembl gene tree
-\href{http://useast.ensembl.org/Multi/GeneTree/Image?gt=ENSGT00580000081348}{ENSGT00580000081348}).
-All genes with expression greater than 50 \ac{FPKM} are listed in
-Supplementary \cref{tab:all_ateles_fusciceps_placenta}.
-
-\subsection{Rodentia Placenta Transcriptomes}
-
-We examined the expression of three rodentia, \mouse, \carmeli, and
-\galili{}. We lacked sufficient power to detect any significant
-changes on the Rodentia lineage in comparison to all other analyzed
-species (Supplemental \cref{tab:rodentia_lineage_specific}).
-
-\subsubsection{\mouse Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="mus musculus"))
-@ 
-\caption{\tableexplanation{\mouse}{tab:all_mus_musculus_placenta}}
-\label{tab:top50_mouse_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \mouse are shown in
-\cref{tab:top50_mouse_placenta} with a list of all genes with
-$\mathrm{FPKM} \ge 50$ in Supplemental
-\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
-\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
-target \textit{in silico} is the orphan G-coupled receptor
-\textit{Mrgprg}
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
-\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
-\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
-of which are maternally imprinted (the later of which is a homolog of
-human \textit{H19\_2}),
-\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
-cathepsins and a rodentia specific expansion of prolactin related
-genes. There are also two miRNA which overlap with the of the 7S
-signal recognition particle
-(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
-and
-\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
-and numerous pseudogenes
-(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
-appear to be highly expressed, however, the short read length (34--44)
-in many samples is likely preventing unambiguous assignment of all
-reads to the more lengthy complete transcript, so these shorter
-transcripts should be treated with suspicion, and we will ignore them
-in further discussion. The prolactin-related genes
-(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
-\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
-very highly expressed in \mouse in, with %
-\Sexpr{combined.fpkm[species=="mus musculus" &
-  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
-prolactin related genes having $\mathrm{FPKM} \ge 1$
-(\cref{fig:prolactin_evolution_tree}, Supplemental
-\cref{fig:prolactin_receptor_evolution_tree}). The cathespins have
-also expanded in rodents, with six showing high expression
-(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
-\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
-\cref{tab:top50_mouse_placenta,fig:pag2_prss16_ctsl_psg_evolution_tree}).
-Finally, \textit{Peg10}, an imprinted gene required for proper
-formation of the placenta in mice is also highly expressed
-(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
-reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.
-
-\subsubsection{\galili Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="nannospalax galili"))
-@ 
-\caption{\tableexplanation{\galili}{tab:all_nannospalax_galili_placenta}}
-\label{tab:top50_n_galili_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \galili are shown in
-\cref{tab:top50_n_galili_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_nannospalax_galili_placenta}.
-Components of annexin complexes are highly expressed
-(\Sexpr{gene.expression("nannospalax_galili","S100a6",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","S100a11",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","S100a10",complete=TRUE)},
-and
-\Sexpr{gene.expression("nannospalax_galili","Anxa2",complete=TRUE)}),
-as are members of the thymosin family
-(\Sexpr{gene.expression("nannospalax_galili","Tmsb4x",complete=TRUE)}
-and
-\Sexpr{gene.expression("nannospalax_galili","Tmsb10",complete=TRUE)}),
-various cytoskeleton related proteins
-(\Sexpr{gene.expression("nannospalax_galili","Actc1",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","Actg2",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","Acta2",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","Myl9",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","Vim",complete=TRUE)}), and
-transmembrane protein of currently unknown function, 
-\Sexpr{gene.expression("nannospalax_galili","Gm10134",complete=TRUE)}.
-
-\subsubsection{\carmeli Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="spalax carmeli"))
-@ 
-\caption{\tableexplanation{\carmeli}{tab:all_spalax_carmeli_placenta}}
-\label{tab:top50_s_carmeli_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \carmeli are shown in
-\cref{tab:top50_s_carmeli_placenta} with a list of all highly
-expressed genes in Supplemental
-\cref{tab:all_spalax_carmeli_placenta}. Numerous hemoglobin subunits
-ware highly expressed
-(\Sexpr{gene.expression("spalax_carmeli","Hbb-bh1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Hba-a1",complete=TRUE)}, and
-\Sexpr{gene.expression("spalax_carmeli","Hbb-bs",complete=TRUE)}), as
-well as the relaxin
-\Sexpr{gene.expression("spalax_carmeli","Rln1",complete=TRUE)},
-multiple cathespins
-(\Sexpr{gene.expression("spalax_carmeli","Ctsm",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","BC051665",complete=TRUE)},
-and \Sexpr{gene.expression("spalax_carmeli","Ctsl",complete=TRUE)},
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), and many prolactin-related genes
-(\Sexpr{gene.expression("spalax_carmeli","Prl4a1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Prl6a1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Prl3b1",complete=TRUE)}, and
-\Sexpr{gene.expression("spalax_carmeli","Prl7c1",complete=TRUE)},
-\cref{fig:prolactin_evolution_tree}) which shows a similar pattern of
-expansion seen in \mouse.
-
-\subsection{Laurasiatheria}
-
-We examined the expression of five Laurasiatherians: \cow, \sheep,
-\pig, \horse, and \dog. We were unable to detect any significant
-differences after correcting for multiple testing between
-Laurasiatherians and all other analyzed species
-(Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
-Laurasiatherians and Euarchontoglires
-(Supplemental \cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).
-
-\subsubsection{\dog Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="canis familiaris"))
-@ 
-\caption{\tableexplanation{\dog}{tab:all_canis_familiaris_placenta}}
-\label{tab:top50_dog_placenta}
-\end{table}
-
-The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
-in the term placenta of \dog are shown in
-\cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
-\cref{tab:all_canis_familiaris_placenta}. The gene with the highest
-expression %
-(\Sexpr{gene.expression("canis familiaris","TMSB10")}) is the thymosin 
-\textit{TMSB10} %
-(\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
-is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
-among the top 50 (\cref{tab:top50_dog_placenta}) include genes
-involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
-anticoagulation (\textit{TFPI2}, \textit{PLA2G7}), innate immunity
-(\textit{LCN2}, \textit{GM2A}), lipid metabolism (\textit{APOE}),
-inhibition of matrix metalloproteinases (\textit{TIMP2},
-\textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
-\textit{HTRA1}). The gene \textit{LGALS1}, which shows evidence of
-immunosupression during pregnancy
-\cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
-(\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
-In addition, an ortholog of \textit{SSLP1} (ENSCAFG00000010638) a gene
-expressed in the placentas of \cow (%
-\Sexpr{gene.expression("bos taurus","SSLP1",complete=TRUE)}, %
-\cref{tab:top50_cow_placenta}) and \sheep
-(\Sexpr{gene.expression("ovis aries","SOLD1",complete=TRUE)},
-\cref{tab:top50_sheep_placenta}) but is not present in primates
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}
-(\cref{fig:sslp1_evolution_tree}), is also expressed at high levels
-%
-(\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
-There is no placental expression of the estrogen synthase gene
-(\textit{CYP19A1}) in \dog 
-(\Sexpr{gene.expression("canis familiaris","CYP19A1")}
-\cref{fig:estrogen_evolution_tree}), 
-and expression of \textit{HSD3B7} is also low
-(\Sexpr{gene.expression("canis familiaris","HSD3B7")}, 
-\cref{fig:progesterone_evolution_tree}).
-
-\subsubsection{\cow Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="bos taurus"))
-@ 
-\caption{\tableexplanation{\cow}{tab:all_bos_taurus_placenta}}
-\label{tab:top50_cow_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
-list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
-The gene with the highest expression was \textit{TMSB10} %
-(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
-\textit{TMSB4} was also highly expressed %
-(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
-\dog{}, members of annexin complexes were also highly expressed
-(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
-at least one of a set of four miRNA with 100\% identity
-(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
-%
-As these are novel miRNA, there is a possibility that the assignment
-of the short reads to these four miRNA is due to miss-annotation of
-the \cow genome.
-% 
-Several members of the ruminant-specific pregnancy-associated
-glycoprotein family show especially high expression in \cow placenta
-(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D, and
-Supplemental \cref{tab:all_bos_taurus_placenta}) including
-%
-\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
-%
-28 members of this family have $FPKM \ge 1$, and this family of
-genes has greatly expanded in ruminants (\cow and \sheep{}) although
-the expression levels of this family are higher in \cow than in \sheep
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D. Other genes with high expression
-include genes involved with regulation of growth (\textit{IGF2},
-\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
-and \textit{CSH1}, a gene that is a member of a ruminant specific
-expansion of placental prolactins and not directly related to the gene
-of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
-ruminant-specific placental prolactins also show high expression in
-the cow placenta including %
-\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
-with a total of 14 with $\mathrm{FPKM} \ge 1$.
-%
-\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
-cotyledonary villi of the ruminiant placenta and involved in cell
-proliferation
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
-shows evidence of high expression %
-(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
-There is some placental expression of the estrogen synthase gene
-(\textit{CYP19A1}) in cows %
-(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
-comparatively greater expression of \textit{HSD3B} %
-(\Sexpr{gene.expression("bos taurus","HSD3B")}).
-
-\subsubsection{\pig Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="sus scrofa"))
-@ 
-\caption{\tableexplanation{\pig}{tab:all_sus_scrofa_placenta}}
-\label{tab:top50_pig_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Sus scrofa} are shown in
-\cref{tab:top50_pig_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_sus_scrofa_placenta}. The top two
-expressed transcripts,
-\Sexpr{gene.expression("sus_scrofa","ENSSSCG00000025162",complete=TRUE)}
-and
-\Sexpr{gene.expression("sus_scrofa","ENSSSCG00000021356",complete=TRUE)},
-are miRNA which correspond to the \cow miRNA
-\href{http://www.mirbase.org/cgi-bin/mirna_entry.pl?acc=MI0013080}{bta-mir-2904-1}
-and
-\href{http://www.mirbase.org/cgi-bin/mirna_entry.pl?acc=MIMAT0013845}{bta-mir-2887-1},
-respectively, both of which are also expressed in \cow ovary
-\cite{Hossain.ea2009:Identificationandcharacterizationof}, although
-neither of which have an obvious target gene on the basis of
-\textit{in silico} analysis
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}. Other highly
-expressed genes include members of the insulin growth factor family
-(\Sexpr{gene.expression("sus_scrofa","IGF2",complete=TRUE)} and
-\Sexpr{gene.expression("sus_scrofa","IGFBP3",complete=TRUE)}), members
-of annexin complexes
-(\Sexpr{gene.expression("sus_scrofa","S100A1",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","ANXA2",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","S100A12",complete=TRUE)}, and
-\Sexpr{gene.expression("sus_scrofa","S100A10",complete=TRUE)}), the
-galectin \Sexpr{gene.expression("sus_scrofa","LGALS3",complete=TRUE)}
-and multiple cytoskeleton and extracellular matrix proteins
-(\Sexpr{gene.expression("sus_scrofa","KRT19",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","COL1A1",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","BETA-ACTIN",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","KRT9",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","CLDN4",complete=TRUE)}, and
-\Sexpr{gene.expression("sus_scrofa","KRT9",complete=TRUE)}).
-
-\subsubsection{\horse Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="equus caballus"))
-@ 
-\caption{\tableexplanation{\horse}{tab:all_equus_caballus_placenta}}
-\label{tab:top50_horse_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Equus Caballus} are shown in
-\cref{tab:top50_horse_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_equus_caballus_placenta}. The gene
-with the highest expression was the relaxin,
-\Sexpr{gene.expression("equus_caballus","RLN",complete=TRUE)},
-followed by two miRNA
-(\Sexpr{gene.expression("equus_caballus","ENSECAG00000025328",complete=TRUE)}
-and
-\Sexpr{gene.expression("equus_caballus","ENSECAG00000027056",complete=TRUE)})
-with no known \textit{in silico} targets
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}, the thymosin
-\Sexpr{gene.expression("equus_caballus","TMSB4",complete=TRUE)},
-members of the insulin growth factor family
-(\Sexpr{gene.expression("equus_caballus","IGF2",complete=TRUE)} and
-\Sexpr{gene.expression("equus_caballus","IGFBP3",complete=TRUE)}) and
-multiple members of annexin complexes
-(\Sexpr{gene.expression("equus_caballus","S100A6",complete=TRUE)},
-\Sexpr{gene.expression("equus_caballus","S100A1",complete=TRUE)}, and
-\Sexpr{gene.expression("equus_caballus","ANXA2",complete=TRUE)}),
-similar to the pattern of expression seen in \pig and \dog.
-
-\subsubsection{\sheep Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
-top.sheep.genes <- top.genes.table(species.name="ovis aries",n=200)
-top.sheep.genes <-
-    top.sheep.genes[!grepl("\\{(7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\}",
-                           `Gene symbol`) &
-                    !grepl("\\((7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\)",
-                           `Human Symbol`)
-                   ,][1:50,]
-output.top.genes.table(top.sheep.genes)
-@ 
-\caption{\tableexplanation{\sheep}{tab:all_ovis_aries_placenta}}
-\label{tab:top50_sheep_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \sheep are shown in
-\cref{tab:top50_sheep_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_ovis_aries_placenta}. The complete
-list of highly expressed genes includes many snRNAs and rRNAs likely
-due to the RNA extraction and library preparation procedure utilized
-for this sample. Ignoring these, notable highly expressed genes
-include members of annexin complexes
-(\Sexpr{gene.expression("ovis_aries","S100G",complete=TRUE)},
-\Sexpr{gene.expression("ovis_aries","ANXA2",complete=TRUE)}),
-hemoglobin subunits
-(\Sexpr{gene.expression("ovis_aries","ENSOARG00000010944",complete=TRUE)}),
-an ortholog of \textit{SSLP1},
-\Sexpr{gene.expression("ovis_aries","SORD1",complete=TRUE)}, two
-cathespins (\Sexpr{gene.expression("ovis_aries","CTSL",complete=TRUE)}
-and \Sexpr{gene.expression("ovis_aries","CTSB",complete=TRUE)}), and
-placentally expressed transcript 1
-\Sexpr{gene.expression("ovis_aries","PLET1",complete=TRUE)} which may
-play a role in cellular migration
-\cite{Raymond.ea2010:Expressionoforphanprotein}.
-
-\subsection{Atlantogenata (Afrotheria + Xenartha)}
-\subsubsection{\elephant Placenta Transcriptome}
-
-\begin{table}
-\footnotesize
-<<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="loxodonta africana"))
-@ 
-\caption{\tableexplanation{\elephant}{tab:top50_elephant_placenta}}
-\label{tab:top50_elephant_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \elephant are shown in \cref{tab:top50_elephant_placenta}
-with a list of all genes in
-\cref{tab:all_loxodonta_africana_placenta}. The gene with the highest
-expression was the natriuretic peptide precursor B
-(\Sexpr{gene.expression("loxodonta_africana","NPPB",complete=TRUE)}).
-Two thymosins,
-\Sexpr{gene.expression("loxodonta_africana","TMSB10",complete=TRUE)}
-and
-\Sexpr{gene.expression("loxodonta_africana","TMSB4Y",complete=TRUE)}
-were also also was highly expressed. The miRNA mir-675
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027406",complete=TRUE)}).
-which is imbedded in the first exon of the maternally imprinted long
-non-coding RNA gene \textit{H19}
-\cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene which is
-currently un-annotated in \elephant{}. %
-% \DLA{Probably consider fixing the unannotation of H19; it seems pretty
-%   obvious to me, though it would require fixing the GTF file for
-%   elephant with the right
-%   \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan
-%     prediction} for this region} %
-Five uncharacterized genes are expressed to a high degree in the
-\elephant placenta (\cref{tab:top50_elephant_placenta}), three of
-which
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000028493",complete=TRUE)},
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032011",complete=TRUE)},
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000031933",complete=TRUE)})
-are members of an \elephant{}-specific expansion of 11 genes closely
-related to the thymus-specific serine protease gene \textit{PRSS16}
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C The other 8
-\textit{PRSS16}-like genes are also expressed to a lesser degree
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000026505",complete=TRUE)}).
-Two other novel genes with high expression are related to the gene
-\textit{SSLP1}
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032659",complete=TRUE)}
-and
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000029551",complete=TRUE)},
-\cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
-\dog placentas (\cref{tab:top50_dog_placenta,tab:top50_cow_placenta}).
-Placental expression of the hormones relaxin (\textit{RLN}, %
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000036090")})
-and prolactin (\textit{PRL}) are exceptionally high in the elephant
-when compared to other mammals
-(\cref{fig:rln2_evolution_tree,fig:prolactin_evolution_tree}). There
-is little to no placental expression of the estrogen synthase gene,
-\textit{CYP19A1}
-(\Sexpr{gene.expression("loxodonta_africana","CYP19A1")}) and the
-enzyme that produces progesterone HSD3B1
-(\Sexpr{gene.expression("loxodonta_africana","HSD3B")}). Other genes
-with high expression in the elephant placenta include genes involved
-in oxidative stress
-(\Sexpr{gene.expression("loxodonta_africana","TXN",complete=TRUE)}),
-growth
-(\Sexpr{gene.expression("loxodonta_africana","IGF2",complete=TRUE)})
-and
-\Sexpr{gene.expression("loxodonta_africana","PAEP",complete=TRUE)}, a
-gene predominantly expressed in the endometrium of the uterus in humans
-\cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.
-
-\subsubsection{\armadillo Placenta Transcriptome}
-
-\begin{table}
-\footnotesize
-<<top.50.armadillo.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="dasypus novemcinctus"))
-@ 
-\caption{\tableexplanation{\armadillo}{tab:all_dasypus_novemcinctus_placenta}}
-\label{tab:top50_armadillo_placenta}
-\end{table}
-
-The top 50 expressed genes in the term placenta of \armadillo are
-shown in \cref{tab:top50_armadillo_placenta} with a list of all genes
-with $\mathrm{FPKM}\ge 50$ in Supplemental
-\cref{tab:all_dasypus_novemcinctus_placenta}. The gene with the
-highest expression is
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045029",complete=TRUE)}
-which is most similar to the \mouse \textit{Pga5} is expanded in
-\opossum into three members which are expressed in placenta (the other
-two being
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033885",complete=TRUE)}
-and
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000040431",complete=TRUE)}),
-though the expansion is less than that seen \cow and \sheep
-(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). There is also high expression of
-hemoglobin subunit beta,
-\Sexpr{gene.expression("dasypus_novemcinctus","HBB",complete=TRUE)},
-an un-annotated gene with high similarity to \human and \cow
-\textit{OVOS2}
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033320",complete=TRUE)}
-(Supplemental \cref{fig:ovos2_evolution_tree}), the growth factors
-\Sexpr{gene.expression("dasypus_novemcinctus","IGF2",complete=TRUE)}
-and
-\Sexpr{gene.expression("dasypus_novemcinctus","IGFBP4",complete=TRUE)},
-and the thymosin
-\Sexpr{gene.expression("dasypus_novemcinctus","TMSB4X",complete=TRUE)}.
-As has been seen previously, multiple components of the annexin
-complex are also highly expressed, including
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045634",complete=TRUE)}
-(\textit{ANXA8L1}),
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000017399",complete=TRUE)}
-(\textit{S100B}),
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000048504",complete=TRUE)}
-(\textit{S100A10}), and
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000004237",complete=TRUE)}
-(\textit{ANXA2}).
-
-\subsection{Metatherians}
-
-\subsubsection{\opossum Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="monodelphis domestica"))
-@ 
-\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica_placenta}}
-\label{tab:top50_opossum_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the \opossum are shown in
-\cref{tab:top50_opossum_placenta} with a list of all genes in
-Supplemental \cref{tab:all_monodelphis_domestica_placenta}. The gene with the
-second highest expression is superoxide dismutase 3, extracellular
-(\Sexpr{gene.expression("monodelphis_domestica","SOD3",complete=TRUE)}),
-which functions as an antioxidant. The gene with the second highest
-expression is the beta hemoglobin subunit epsilon
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000011325",complete=TRUE)}),
-which is expressed at the embryonic stage of development in
-marsupials, as in other mammals
-\cite{Holland.ea1997:Characterizationofembryonicglobin}. \textit{HBE}
-subunits link up with protein products of two hemoglobin, zeta
-(\textit{HBZ}) genes, which are also expressed to a high degree in the
-opossum placenta
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016635",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016631",complete=TRUE)}).
-In addition, there is high expression of the gene hemoglobin, omega
-(\textit{HBO},
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016637",complete=TRUE)}),
-% this is LOC_1000016060 which I figured out by tblastn'ing AAK11486.1
-% against Monodelphis domestica
-a hemoglobin found only in marsupials and monotremes
-\cite{Opazo.ea2008:Genomicevidenceforindependent} (Supplemental
-\cref{fig:hbe1_evolution_tree}). The hemoglobin gene \textit{HBA} is
-also expressed in the placenta
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016633",complete=TRUE)}).
-Two cystatin genes
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018295",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018296",complete=TRUE)})
-are expressed to a high degree, along with one \textit{CTSL} related
-gene
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000002844",complete=TRUE)},
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), which are known to interact with
-cystatins \cite{Turk.ea2001:Lysosomalcysteineproteasesfacts}. The
-uncharacterized gene \textit{LOC10001195}
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
-also shows high expression in \opossum placenta. The gene
-\textit{WFDC2}, a member of the whey acidic protein (WAP) family is
-also expressed to a high degree in \opossum placenta
-(\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).
-Other genes with high expression include members of the annexin
-complex
-(\Sexpr{gene.expression("monodelphis_domestica","S100A6",complete=TRUE)},
-\Sexpr{gene.expression("monodelphis_domestica","ANXA2",complete=TRUE)}).
-In addition, there was little production of CYP19A1 and HSD3B1
-(\Sexpr{gene.expression("monodelphis_domestica","CYP19A1",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000005385",complete=TRUE)},
-respectively).
-
-\subsection{Core Placenta Transcriptome}
-
-\begin{figure}
-<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
-print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
-             aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
-      + geom_raster(aes(fill=log10(FPKM)))
-      + scale_x_discrete(expand=c(0,0))
-      + scale_y_discrete(expand=c(0,0))
-      + scale_fill_gradient(low = grad.def[["low"]],
-                            high = grad.def[["high"]],
-                            space = grad.def[["space"]],
-                            na.value = grad.def[["na.value"]],
-                            limits=c(0,4),
-                            guide=guide_colorbar(title=expression(log[10](FPKM)))
-                            )
-      + ylab("Human Ortholog Name")
-      + xlab("Species")
-      + theme_grey()
-      + theme(legend.position="top",
-              axis.ticks = element_blank(),
-              axis.text.y = element_text(face="italic"),
-              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
-      )
-@ 
-\caption{Core Placenta Transcriptome.
-  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-  genes expressed (\ac{FPKM}$ \ge 10$) in the placenta of all organisms
-  studied with housekeeping
-  genes \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
-  removed.}
-\label{fig:core_placenta_transcriptome}
-\end{figure}
-
-The
-\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-genes comprising the core placenta transcriptome are shown in
-\cref{fig:core_placenta_transcriptome} ordered by their median
-transcription level in \ac{FPKM}. We considered a gene to be core to
-the placenta transcriptome if a 1:1 ortholog of that gene was
-expressed with $\mathrm{FPKM} \ge 10$ in all species studied and was
-not a human housekeeping gene according to
-\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
-Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
-which would be included in the core placenta set if the \ac{FPKM}
-cutoff was set differently.
-%
-Multiple components of annexin complexes are included in the core
-transcriptome, including \textit{ANXA2}, \textit{ANXA1},
-\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
-resolution of the inflammation response through inhibition of
-phospholipase A2 activity and signaling through the formyl peptide
-receptor family, and has previously been hypothesized to be important
-for the maintenance of the anti-inflammatory during pregnancy
-\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
-known to interact with S100A10, S100A6, S100A11, and S100A4
-\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
-its interaction with S100P as well; it is involved in cell-cell
-interaction as well as vesicle trafficking and von Willibrand Factor
-secretion. The expression regulation of another core gene,
-\textit{TIMP3}, has been implicated in pre-eclampsia
-\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
-Supplemental \cref{tab:core_placenta_go_terms} lists the significant enrichment of
-\ac{GO} terms in the core placenta transcriptome; notable terms
-include those involved in cytoskeletal structure, protein transport,
-membrane protein structure, and inflammation resolution.
-
-% SPINT2 -- Serine Peptidase inhibitor
-
-% SLC3A2 -- this is probably actually a housekeeping gene; it's
-% involved in amino acid transport
-
-% SQSTM1 -- NFκB signaling (immune supression?)
-
-% KRTCAP2  N-glycosylation (probably housekeeping)
-
-% NRDG1 -- interacts with HSP90
-
-% SDC4 -- proteoglycan (syndecan) cell adhesion
-
-% LAMC1 -- laminin
-
-% LUM -- proteoglycan
-
-% VDAC1 -- mitochondrial gene?
-
-% F11R -- tight junctions
-
-% SERPINH1? with TIMP3 
-% http://www.proteinatlas.org/ENSG00000149257-SERPINH1/tissue/placenta
-
-\subsubsection{\ac{GSE} Analyses of Individual Transcriptomes}
-
-In order to identify major functional categories of genes expressed in
-the individual transcriptomes, we examined \ac{GSE} in 1:1 human
-orthologs which had \ac{FPKM} values in the top 1\% of expressed genes
-(Supplemental %
-\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_nannospalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris}) using topGO \cite{topGO:2.20.0}.
-%
-%
-Significant enrichment in genes ($\mathrm{FDR}\le0.05$) in the top 1\%
-of \ac{FPKM} values in at least 12 of the
-\Sexpr{length(species.names)} species included clusters such as
-membrane organization , locomotion, tissue development, extracellular
-matrix organization, regulation of peptidase activity, and various
-cell death related pathways among others
-(Supplemental \cref{tab:go_terms_all_species}). If we instead examine the top 100
-genes by \ac{FPKM} in each of the \Sexpr{length(species.names)}
-species (Supplemental %
-\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris})
-notable enrichments include extracellular matrix, female pregnancy,
-integrin and collagen-related processes, and various cell death
-related processes.
-
-% \textit{Ateles fuciceps} shows enrichment in extracellular
-% matrix and structural molecular activity (Supplemental
-% \cref{tab:go_terms_enriched_ateles_fusciceps}). \textit{Homo sapiens}
-% placentas are enriched in female pregnancy, estrogen biosynthesis,
-% reproductive process, hormone aactivity, oxygen transporter activity,
-% and haptoglobin binding (Supplemental
-% \cref{tab:go_terms_enriched_homo_sapiens}). \textit{Pan paniscus}
-% placentas are enriched in clusters related to stress, apoptosis, and
-% response to abiotic stimuli. \textit{Mus musculus} placentas are
-% enriched in cell junction, vesicle transport, in utero embryonic
-% development, and many cytoskeletal processes. \textit{Nannospalax
-%   galili} placentas are enriched in extracellular matrix organization,
-% integrin binding, and cell death.
-% 
-% Cow shows only four
-% significantly enriched clusters; however many genes highly expressed
-% in the cow are ruminant-specific placenta genes and are not yet
-% annotated to specific GO categories. Dog shows enrichment in genes
-% associated with RNA, as well as homeostasis, glycoproteins, lyosome,
-% and calcium binding. Elephant shows enrichment in genes associated
-% with RNA, coagulation and calcium binding. Opossum shows enrichment in
-% genes involved in apoptosis, oxidative stress, mitochondrion, hormone
-% metabolism, and lipids.
-
-% Not sure if this is actually worthwhile; I've mentioned some of this
-% already for some of the lineages
-
-% \subsection{Comparative Analysis of Gene Expression}
-% 
-% We performed a comparative analysis of gene expression by
-% reconstructing the evolution of expression values and detecting
-% significant shifts in expression along specific lineages. Likelihood
-% values for each model for each of the X genes reconstructed, as well
-% as likelihood ratio tests, P- and Q-values (after correcting for
-% multiple testing) are shown in Supplemental Table X. We identified
-% between 4 (Boreoeutheria) and 550 (Mouse) significant shifts (Q <
-% 0.01) in expression on specific lineages (Table 10). The top 10 genes
-% with significant shifts (if applicable) per lineage are shown in Table
-% 10 and all genes for each branch are shown in Supplemental Tables
-% X-XX. Genes that showed shifts on the human lineage were
-% overwhelmingly identified as placenta-specific by the
-% tissue-specificity index (TSI), with 56 out of 279 genes (20\%) with
-% both TSI>0.75 and placenta as one the top two tissues with the highest
-% expression. Unsurpisingly, this was not the case with other lineages,
-% since we used expression values from a human-derived microarray to
-% generate TSI values for each gene. However, this does underscore the
-% dynamic nature of expression patterns in the placenta and indeed all
-% tissues across mammals.
-% 
-% We then performed DAVID functional annotation clustering analyses on
-% each list of genes from each lineage. The top three clusters for each
-% lineage are shown in Table 10 and a full list of clusters in
-% Supplemental Table X. The top enrichment cluster in the human lineage
-% is blood vessel development, followed by extracellular matrix and
-% metal ion binding (Table 10). Other significant clusters include those
-% related to cell communication/signaling, cytokines, defense response,
-% tryptophan metabolism, apoptosis, and skeletal system development
-% (Supplemental Table X). The bonobo lineage was enriched in only three
-% clusters: apoptosis, Toll-like receptor signalling and apoptosis. The
-% mouse displayed enrichment in clusters involved in epigenetic
-% modifications such as methyltransferase activity and histone
-% modification, as well embryonic morphogeneisis, lipid biosynthesis,
-% among others (Supplemental Table X). The top enriched cluster in the
-% cow lineage was the response to vitamin A, as well as many clusters
-% related to the immune system. The dog showed enrichment in skeletal
-% system development, as well as lipid metabolic processes and blood
-% circulation. The elephant lineage had only one enriched cluster:
-% carbohydrate binding. Genes that showed enrichment on the opossum
-% lineage (essentially representing the difference between placental
-% mammals and marsupials) were clustered in many categories not seen in
-% the other lineages, including clusters related to the mitochondria,
-% urea cycle, peroxisome, glycolysis, and transport across the cell
-% membrane (Table 10 and Supplemental Table X).
-
-\clearpage
-
-\section{Discussion}
+@ 
+\end{figure}
+\end{frame}
 
@@
 
+\begin{frame}{Galectin Expression}
 \begin{figure}
@@
          )
-print(ggplot(combined.long, aes(y=gene, x=species))
+print(ggplot(combined.long[combined.long$type=="Galectin",], aes(y=gene, x=species))
       + common.plot.options
@@
 \end{figure}
-
-\subsubsection{Estrogen and Progesterone}
-
-In both ruminants and primates, the placenta is capable of producing
-estrogen and progesterone
-\cite{Strauss.ea1996:Placentalsteroidhormonesynthesis}. High relative
-expression of \textit{CYP19A1} (Supplemental
-\cref{fig:estrogen_evolution_tree}), the enzyme that synthesizes
-estrogen from androstenedione is seen in \human, \bonobo, \cow, and
-\horse, though not in \sheep, \smonkey. %
+\end{frame}
+
+\begin{frame}{Hormone Expression}
+\begin{figure}
+<<hormone.expression.plot,echo=FALSE>>=
+print(ggplot(combined.long[combined.long$type=="Hormone",], aes(y=gene, x=species))
+      + common.plot.options
+      )
+@ 
+\end{figure}
+\end{frame}
+
+\begin{frame}{IGF Related}
+\begin{figure}
+<<igfrelated.expression.plot,echo=FALSE>>=
+print(ggplot(combined.long[combined.long$type=="IGF Related",], aes(y=gene, x=species))
+      + common.plot.options
+      )
+@ 
+\end{figure}
+\end{frame}
+
+% \begin{frame}{Prolactin}
+% \begin{itemize}
+% \item \elephant is only analyzed mammal with high prolactin expression
+%   (\textit{PRL}) in the placenta
+% \item However, \mouse, \carmeli, and \cow have duplications of
+%   \textit{PRL} with expression}
+% \end{itemize}
 % 
-\textit{HSD3B1}, one of two enzymes that synthesizes progesterone from
-pregnalone
-(\cite{Simard.ea2005:Molecularbiologyof3beta-hydroxysteroid},
-Supplemental \cref{fig:progesterone_evolution_tree}) is also highly
-expressed in \human, \bonobo, \cow, \horse, and \dog, but is not
-highly expressed in \smonkey. The diversity of \textit{CYP19A1} and
-\textit{HSD3B1} expression levels in in primates and ungulates seen
-here points to the need for more complete sampling of placentas at
-different timepoints and in different clades.
-
-\subsubsection{Prolactin}
-
-The elephant is strikingly the only mammal analyzed that shows high
-expression of the primarily pituitary hormone prolactin (\textit{PRL})
-in the placenta, although we see high expression of genes which
-resulted from duplications of \textit{PRL} in the \mouse, \carmeli,
-and \cow (Supplemental \cref{fig:prolactin_evolution_tree}). All afrotherians yet
-analyzed contain only one prolactin gene, and prolactin has undergone
-adaptive evolution on the \elephant lineage
-\cite{Wallis2009:ProlactininAfrotheriacharacterization}.
-\citet{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
-presence of a prolactin in the trophoblast of the elephant placenta
-using immunohistological staining and human derived antibodies. A
-previous analysis of \elephant placenta also detected the expression
-of prolactin \cite{Hou.ea2012:Elephanttranscriptomeprovidesinsights}.
-Here we confirm that the sole prolactin in \elephant is not only
-expressed but highly expressed in the placenta, and likely functions
-both in the maintenance of the corpus luteum and as a placental
-lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}.
-\textit{PRL} is not expressed in \opossum, even though there is
-evidence that it is expressed in the placenta of another marsupial,
-\textit{Macropus eugenii} (tammar wallaby)
-\cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests
-that placental prolactin expression in marsupials may be as diverse as
-in eutherian mammals. One possible explanation of the diversity of
-expression is the presence of endogenous retroviruses such as MER39 in
-primates, MER77 in rodents, and L1-2\_LA in \elephant
-\cite{Emera.ea2012:Convergentevolutionofendometrial} although this
-does not explain the diversity of expression of different isoforms of
-\textit{PRL} in \mouse or their tissue specificity in primates. An
-alternative, that \textit{HOXA11} expression is responsible for
-\textit{PRL} expression
-\cite{Lynch.ea2008:Adaptivechangesintranscription} seems not to be the
-case in placenta, as the rodent with the highest \textit{HOXA11}
-expression, \galili
-(\Sexpr{gene.expression("nannospalax_galili","Hoxa11",complete=TRUE)})
-is the rodent with the lowest expression of \textit{PRL}
-(\cref{fig:prolactin_evolution_tree}), and \elephant also does not
-express \textit{HOXA11} highly either
-(\Sexpr{gene.expression("loxodonta_africana","HOXA11",complete=TRUE)},
-Supplemental \cref{fig:hoxa11_evolution_tree}).
-
-\subsubsection{\ac{GH} family}
-
+% \begin{figure}
+% <<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+% 
+% tree.id <-
+%     genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
+% gene.tree.table <- trees[[tree.id]]$table
+% gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+% gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+% plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+%                  gene.tree.table,gene.tree.expression,
+%                  axis.text.size=5,
+%                  fontsize=5)
+% @ 
+% \end{figure}
+% 
+% \end{frame}
+
+\subsubsection{GH family}
+
+\begin{frame}{Growth Hormone family}
+
+\begin{columns}
+\column{0.4\textwidth}
 The \ac{GH} family has expanded in anthropoid primates, resulting in
@@
 between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
-reliably distinguish transcripts between the two). All show
-substantial expression in the \human placenta
-(\cref{fig:csh1_evolution_tree}). The expression of the \ac{GH} family
-in placenta appears to be primate specific, and likely existed at the
-time of the last common ancestor between all extant anthropoids
-\cite{Papper.ea2009:Ancientoriginofplacental}. There is some evidence
-that increased expression of these genes may be correlated with
-overall birth weight of offspring in \human
-\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
-possible that the extreme expression levels of these genes may be due
-to high birth weight. \human placental \ac{GH} genes are involved in
-increasing blood flow, replacing maternal pituitary expression of
-\ac{GH}, as well as trophoblast invasion
-\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
-and it is likely that they have the same function in \bonobo.
+reliably distinguish transcripts between the two).
+
+\column{0.6\textwidth}
+<<chorionic.somatomammotropin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CSH1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 axis.text.size=5,
+                 fontsize=4)
+@ 
+\end{columns}
+
+\end{frame}
+
+\begin{frame}{CGA}
+
+  The human gene \textit{CGA} codes for the alpha subunit of multiple
+  hormones including pituitary hormones (luteinizing hormone,
+  follicle-stimulating hormone, thyroid-stimulating hormone), as well
+  as numerous chorionic gonadotropins with variable beta subunits
+  \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of
+  the chorionic gonadotropins is to prevent the regression of the
+  corpus luteum \cite{Hanson.ea1971:EffectsofHCGand}. Among the six
+  placentas we analyzed here, \textit{CGA} is only expressed in the
+  placentas of \human and \bonobo
+  (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}), which
+  corresponds with the emergence of the chorionic gonadotropins in
+  anthropoid primates
+  \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly,
+  \textit{CGA} is expressed at much higher levels in both species than
+  the beta subunits with which its product dimerizes \textit{CGA} is
+  one of several genes noted by
+  \citet{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone
+  adaptive evolution of the \human{}-\textit{P. troglodytes} stem
+  lineage.
 
@@
 plot.label.size <- 4.5
-pushViewport(viewport(layout=grid.layout(nrow=1,ncol=2,widths=c(2,3))))
-
 tree.id <-
@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=1),
                  offset.ratio=0.7,
@@
                  )
+@ 
+\end{figure}
+
+\end{frame}
+
+\begin{frame}{CGB}
+\begin{figure}
+<<chorionic.somatomammotropin.beta.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
 
@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=2,layout.pos.row=1),
                  subtree=c("ENSMODG00000013768","ENSG00000104818"),
@@
                  )
-popViewport(1)
-grid.text(c("A","B"),
-          x=unit(c(0,2/5),"npc"),y=unit(c(1,1),"npc"),
-          just=c("left","top"),
-          gp=gpar(fontsize="48")
-          )
-@ 
-\caption{\genetreeexplanation{(A) \textit{CGA} (B) \textit{CGB}}}
-\label{fig:cga_cgb_evolution_tree}
-\end{figure}
-
-The human gene \textit{CGA} codes for the alpha subunit of multiple
-hormones including pituitary hormones (luteinizing hormone,
-follicle-stimulating hormone, thyroid-stimulating hormone), as well as
-numerous chorionic gonadotropins with variable beta subunits
-\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of
-the chorionic gonadotropins is to prevent the regression of the corpus
-luteum \cite{Hanson.ea1971:EffectsofHCGand}. Among the six placentas
-we analyzed here, \textit{CGA} is only expressed in the placentas of
-\human and \bonobo (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}),
-which corresponds with the emergence of the chorionic gonadotropins in
-anthropoid primates
-\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly,
-\textit{CGA} is expressed at much higher levels in both species than
-the beta subunits with which its product dimerizes (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}).
-\textit{CGA} is one of several genes noted by
-\citet{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone
-adaptive evolution of the \human{}-\textit{P. troglodytes} stem lineage.
-
-\subsubsection{Relaxin}
-
+@ 
+\end{figure}
+\end{frame}
+
+\subsection{Relaxin}
+
+\begin{frame}
 Relaxin (\textit{RLN}) is a hormone that has multiple effects during
@@
 
-\subsubsection{\textit{NPPB} and \textit{NPPC}}
-
-Two genes, \textit{NPPB} and \textit{NPPC}, code for peptide hormones
-that are important in the inducement of vasodilation, and may be
-important in increasing maternal blood flow to the fetus
-\cite{Cameron.ea1996:sitesofgeneexpression}. There is some indication
-that these proteins are present in the decidua of mice
-\cite{Cameron.ea1996:sitesofgeneexpression}, but we find very little
-expression in the fetal portion of the mouse placenta
-(\cref{fig:hormone_genes}, Supplemental
-\cref{fig:nppc_evolution_tree,fig:nppb_evolution_tree}). While we do
-find high expression of at least one of these genes in \dog, \cow, and
-\elephant, the expression pattern does not follow the placental
-organization (endotheliochorial for \dog and \elphant,
-synepitheliochorial for \cow and \sheep), so it is unlikely that
-expression of these genes are essential for specific placental
-organizations.
+\end{frame}
 
@@
 
+\begin{frame}{Galectins}
 Galectins are recognized as major regulators of pregnancy in both
@@
 
+\end{frame}
+
 \subsection{The Opossum Placenta and the Evolution of Viviparity}
 
-The placenta of \opossum is distinctly different from that of the term
-placenta of eutherian mammals in that it is derived from both the
-chorion and yolk sac. Other differences between the placentas of
-marsupials and eutherians include lack of invasive extravillous
-trophoblasts in marsupials, the retention of a permeable shell coat
-for 2/3 of marsupial gestation, and the greater degree of imprinting
-in placenta-expressed genes in eutherians
-\cite{Renfree2010:ReviewMarsupialsplacentalmammals,Renfree.ea2013:originandevolutionof}.
-In addition, the placenta of \opossum is also distinct from other
-marsupials, in that it is considered one of the least specialized, and
-thought be closer to the original marsupial condition
-\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. In \opossum, the
-duration of direct connection between maternal and fetal tissues is as
-short as two days in a gestational period of 13-15 days
-\cite{Harder.ea1993:Gestationandplacentationin}. During this period,
-the placenta is directly engaged in the exchange of urea with maternal
-tissues, owing to the lack of a vesicular allantois
-\cite{Freyer.ea2007:Placentalfunctionintwo}. This differs drastically
-with the situation in another well-studied marsupial, the tammar
-wallaby \cite{Freyer.ea2007:Placentalfunctionintwo}. Therefore, we
-must interpret differences between the opossum and eutherian placentas
-with caution, as highly expressed genes and functional annotation
-clusters may represent didelphid or \opossum specific adaptations.
-
-% Genes that show a significant shift on the \opossum lineage include
-% functional annotation clusters that correspond to active generation of
-% energy, transport, and lipid breakdown via the peroxisome (Table 10;
-% Supplemental Table X). The choriovitelline placenta of both didelphids
-% and the tammar wallaby are known to be involved in glycolysis and
-% protein synethsis (Freyer et al., 2003). This implies that the
-% choriovitelline placenta in marsupials is an active site of metabolism
-% and molecular synthesis, as well as the transport of nutrients from
-% the maternal tissues. This is a distinct difference that is not seen
-% in significant shifts on other lineages, and may be due to the
-% formation of the placenta in the opossum from the yolk sac. In
-% addition, the urea cycle is a highly enriched functional annotation
-% cluster which directly correlates with the high levels of urea in the
-% Monodelphis placenta (Freyer et al., 2007). This is potentially a
-% Monodelphis-specific characteristic as discussed above.
-
+\begin{frame}
 The \opossum placenta shows expression of genes that either do not
@@
 
+\end{frame}
+
 \subsection{Novel Gene Expansions in the Elephant Placenta}
 
-We found evidence of high expression in two novel expansions of genes
-in the \elephant genome. One family includes 11 genes that are related
+\begin{frame}{PRSS16}
+
+11 genes that are related
 to the \textit{PRSS16} gene, a thymus-specific serine protease that
@@
 immune tolerance; however, future work will need to test these
-hypotheses. Three other genes abundantly expressed in the elephant
-placenta are homologues of \textit{SSLP1}, a gene with predominant
-expression in the seminal vesicle in mouse
-\cite{Li.ea2006:SSLP-1secretedLy-6protein}. Homologues of
-\textit{SSLP1} are also expressed in \cow, \dog, \horse, and \sheep.
-\textit{SSLP1} also contains an Alu retrotransposon and may be
-involved in the development of villi via cell proliferation
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}. This
-gene is not found in primates. Expansion of these genes with high
-expression in \elephant may indicate an elaboration of this system.
-
-\subsection{Annexin Complex}
-
-\subsection{Human Specific Expression Shifts and the Evolution of Pre-eclampsia}
-
-The \human placenta is distinctive from many other primates in that it
-possesses a deeply invasive trophoblast that penetrates the inner
-myometrium of the uterus
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
-hypothesized that this feature is related to the increased demand for
-resources by the developing fetus to feed the expanded size of the
-human brain
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation,Robillard.ea2002:Evolutionaryadaptationstopre-eclampsiaeclampsia}.
-\human have the largest brain in terms of relative size and this
-differs markedly from closely related great apes
-\cite{Boddy.ea2012:Comparativeanalysisofencephalization}. Deep
-penetration of the myometrium occurs in chimpanzees and gorillas, but
-this penetration may be accelerated in humans
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
-
-The near exclusivity of deep trophoblast invasion in humans has also
-led many to speculate on the origin of various disorders, which are
-almost exclusively symptomatic of human pregnancy. For example,
-pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
-the United States
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
-can be attributed to the lack of adequate invasion of maternal tissue,
-and is noted by the onset of hypertension and proteinuria
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
-pre-eclampsia can develop into eclampsia, a more severe condition
-resulting in seizures during pregnancy and death. Lack of invasion in
-pre-eclampsia may be attributed to the inefficiency of growth
-regulators as well as the inability to sufficiently suppress the
-maternal immune response. Although there are speculative cases of
-eclampsia in great apes
-\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
-it is exceedingly rare in these species
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is possible
-that a comparative evolutionary approach to gene expression can
-identify genes that have divergent expression patterns from other
-mammals as well as association with pre-eclampsia.
-
-Functional annotation clusters significantly enriched on the human
-lineage (blood vessel development, extracellular matrix, tube
-morphogenesis) seem to suggest involvement in remodeling of tissues
-and increased blood flow, both of which may correlate with the
-invasion of the trophoblast (\cref{tab:go_terms_all_species}). In
-addition, clusters such as cytokines and defense response
-(\cref{tab:go_terms_all_species}) also indicate an a significant
-change in expression of the immune response that may be involved in
-immunosupression at the maternal-fetal boundary to further allow
-invasion of fetal tissues.
-
-% I haven't actually done this analysis, so commented out
-
-% Many genes that both show expression shifts on the human lineage and
-% have very high expression (FPKM>200) are directly related to the
-% highly invasive nature of the human placenta. For example, GPC3,
-% glypican 3, is an imprinted gene and cell growth regulator, associated
-% with Simpson-Golabi-Behmel fetal overgrowth syndrome, and involved in
-% intrauterine growth restriction (Monk and Moore, 2004; Pilia et al.,
-% 1996). GPC3 has the highest expression in humans (Table 2), but is
-% also expressed in other mammals to a lesser degree, especially those
-% with invasive hemochorial placentas like bonobos (\ac{FPKM}= 199.924),
-% spider monkeys (\ac{FPKM}= 96.24), and the mouse (\ac{FPKM}= 80.7545)
-% (Supplemental Tables X-XX). ADAM12 increases the availability of
-% insulin-like growth factors that lead to the promotion of cell growth
-% and placental invasion (Laigaard et al., 2005). (Hou et al., 2009)
-% found evidence of adaptive evolution in ADAM12 on the primate lineage,
-% and like GPC3 it is expressed in the bonobo (\ac{FPKM}= 46.1553), spider
-% monkey (\ac{FPKM}=30.41) and mouse (\ac{FPKM}=2.25391) as well, although to a
-% lesser degree (Supplemental Tables X-XX). RHOBTB3 is imprinted and
-% associated with trophoblast invasion as well (Bilban et al., 2010;
-% Yuen et al., 2011). In addition, one other gene, KISS1, shows
-% exceptional abundance in the human placenta (Table 2); however due to
-% its absence in elephant Ensmebl gene annotations, it was not part of
-% the 1:1 orthologues examined in the comparative study. Here, it is
-% only expressed in the term placentas of humans (Table 2) and bonobos
-% (\ac{FPKM}=17.6394; Supplemental Table X). KISS1 is thought to be involved
-% in the regulation of trophoblast invasion, and shows highest
-% expression in the first trimester of pregnancy (Bilban et al., 2004;
-% Reynolds et al., 2009).
-% 
-% In addition, several other genes that show expression shifts on the
-% human lineage also have major effects on placental function. Genes
-% such as TXK and EBI3 are intimately involved in the immune system in
-% the placenta. TXK is involved in natural killer cell differentiation
-% (Bezman et al., 2012). EBI3 is implicated in natural killer cell
-% function an maternal-fetal immunomodulation in the syncytiotrophoblast
-% (Devergne et al., 2001). KYNU and KMO are interacting genes that are
-% involved in tryptophan metabolism, which may be involved in
-% immunosupression and the prevention of fetal rejection (Takikawa,
-% 2005). THSD7A is a thrombospondin, which inhibits endothelial cell
-% migration and tube morphogenesis (Wang et al., 2010).
-% 
-% Many genes that show evidence of a shift on the human lineage show
-% evidence of implication in pre-eclampsia, including ADAM12 (Gack et
-% al., 2005; Laigaard et al., 2005) MAOA (Carrasco et al., 2000;
-% Bottalico et al., 2004), BCL2 (Ishihara et al., 2002), GPC3 (Gui et
-% al., 2012; Ogawa, 2007), TMEM8A, CLIC3, EBI3, FLT1, BCL6, SYDE1, and
-% MAN1C1 (Enquobahrie et al., 2008; Winn et al., 2009; Nishizawa, 2011).
-% In addition KISS1 is also involved in pre-eclampsia as well (Sado et
-% al., 2011; Qiao et al, 2012). Due to the association of many genes
-% that show human-specific shifts as well as high expression, we
-% conclude that pre-eclampsia is possibly a symptom of recent evolution
-% of gene expression on the human lineage. Although many genes such as
-% GPC3 and ADAM12 show distinctive relatively high expression in bonobo,
-% this expression has intensified on the human lineage. Sequencing of
-% more placentas of nonhuman primates, including other individuals of
-% the genus Pan will help to further characterize this transition.
+hypotheses. 
+\begin{figure}
+<<prss16.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"),
+                 fontsize=4,
+                 width=3,
+                 axis.text.size=axis.text.size,
+                 offset.ratio=1.2
+                 )
+
+@ 
+\end{figure}
+\end{frame}
 
@@
 
-We sequenced \Sexpr{star.logs[,length(unique(species))]} placentas
-spanning the phylogenetic tree of mammals. This revealed a dynamic
-expression pattern of expression of lineage-specific expansions of
-gene families and differential gene expression that agreed with the
-scenario of an “arms race” between maternal and fetal tissues that
-have shaped such disparate gene expression profiles. In addition, we
-highlighted the broad spectrum of hormone gene expression in multiple
-species that suggested that there is a lack of conservation in hormone
-signaling during mammalian pregnancy. Major differences between
-expression in marsupial and eutherian mammals at term seemed to be
-correlated with the differences between choriovitelline and
-chorioamniotic placenta, with the marsupial placenta more enriched in
-genes the contributed to the active generation of energy. We also
-identified that galectin expression, especially that of
-\textit{LGALS1}, is a common feature of all mammalian placentas and
-was most likely present in the ancestral therian. Also, we identified
-two new expansions of genes that are highly expressed in the \elephant
-placenta hat may be involved in trophoplast differentiation and immune
-system function, mirroring expansions in other groups such as
-primates, rodents, and ruminants. We also quantified significant
-lineage-specific shifts in gene expression, concentrating on the
-identification of genes that showed gene expression changes on the
-human lineage. We cautiously propose that many of these genes may be
-related to the invasive nature of the trophoblast in \human, although
-we note that in many of these genes, there is in some cases
-substantial expression in the \bonobo as well. These genes have the
-potential to spur further research and may be at the root of
-obstetrical disorders such as pre-eclampsia. We propose that more
-placentas from diverse lineages, especially those of marsupials and
-primates, to further identify genes and expression levels that are
-specific to marsupials and eutherians, as well as confirm the identity
-of human-specific expression shifts.
-
-\begin{backmatter}
-  \section*{Competing interests}
-  The authors declare that they have no competing interests.
-
-  \section*{Author's contributions}
-  Don Armstrong analyzed all of the data and wrote the paper.
-
-  Michael McGowan wrote the paper and performed initial analyses.
-
-  Amy Weckle performed all of the RNA extractions and managed the
-  samples.
-
-  Jason Caravas provided samples.
-
-  Dalen Agnew provided samples.
-
-  Kurt Benirschke provided samples.
-
-  Sue Savage-Rumbaugh provided samples.
-
-  Eviatar Nevo provided samples.
-
-  Chong J. Kim provided samples.
-
-  Günter P. Wagner provided samples and wrote the paper.
-
-  Roberto Romero provided samples and conceived the study.
-
-  Derek E. Wildman conceived the study, wrote the paper, and performed
-  analyses.
-
-  \section*{Acknowledgements}
-  The authors would like to specially acknowledge those groups whose
-  publicly available data contributed to the completeness of this
-  work.
-
-  \printaddresses
-
-\end{backmatter}
-
-\printbibliography
+\begin{frame}{Conclusion}    
+\begin{itemize}
+\item Sequenced placentas of \Sexpr{star.logs[,length(unique(species))]} species
+
+\item lineage-specific expansions of gene families and differential
+  gene expression
+
+\item Identified common expression patterns of
+
+\item Still much more to learn about the changes in placenta
+  expression during placentation
+
+\item Many clades are only represented by a single species; need to
+  have more information to make stronger inferences about evolution of
+  placenta expression
+
+\end{itemize}
+
+\end{frame}
+
+% Acronyms
+\input{acronyms}
+
+\begin{frame}[plain]{References}
+  \begin{center}
+    \mbox{}\vspace{-\baselineskip}
+    \printbibliography[heading=none]
+  \end{center}
+\end{frame}
 

### 2016-02-22 presentation/placenta_viviparous_presentation.Rnw
update presentation

@@
 \usepackage{booktabs}
+\usepackage{pdfpcnotes}
 % \usepackage{multirow}
@@
   \logo{\begin{tikzpicture}% Pale figure
-      {\node[opacity=0.734]{\IfFileExists{figures/igb_wordmark.pdf}{\includegraphics[width=2cm,height=1cm,keepaspectratio]{figures/igb_wordmark}}{}%
+      {\node[opacity=0.5]{\IfFileExists{figures/igb_wordmark.pdf}{\includegraphics[width=2cm,height=1cm,keepaspectratio]{figures/igb_wordmark}}{}%
         };}%
@@
 \begin{frame}{Variation in Gestational Periods and Litter Size}
-\tiny
-\begin{table}
+\footnotesize
+%\begin{table}
   \begin{tabular}{c c c c}
@@
 %  \label{tab:gestational_periods}
-\end{table}
+%\end{table}
 \end{frame}
@@
 \begin{frame}{Variations in Placenta Morphology}
-\begin{table}
+%\begin{table}
+\footnotesize
 \begin{tabular}{c c c}
@@
 \end{tabular}
-\end{table}
+%\end{table}
 \end{frame}
@@
                      display=c("s","s","s","e","g","f"),
-                     columns=c("c","c","p{1.5in}","c","c","c")
+                     align=c("c","c","p{1.5in}","c","c","c")
                      ),
@@
 
-\item We considered a gene to be core to the placenta transcriptome if
-  a 1:1 ortholog of that gene was expressed with
-  $\mathrm{FPKM} \ge 10$ in all species studied and was not a human
-  housekeeping gene according to
+\item Gene is core to the placenta transcriptome if a 1:1 ortholog is
+  expressed with $\mathrm{FPKM} \ge 10$ in all species studied and was
+  not a human housekeeping gene according to
   \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
@@
 
-\item ANXA1 is involved in the resolution of the inflammation response
-  through inhibition of phospholipase A2 activity and signaling
-  through the formyl peptide receptor family, and has previously been
-  hypothesized to be important for the maintenance of the
+\item ANXA1 is involved in resolution of inflammation by inhibiting
+  phospholipase A2 activity and signaling via formyl peptide receptor
+  family
+\item hypothesized to be important for the maintenance of the
   anti-inflammatory during pregnancy
@@
 
-\item ANXA2 is known to interact with S100A10, S100A6, S100A11, and
+\item ANXA2 interacts with S100A10, S100A6, S100A11, and
   S100A4 \cite{Liu.ea2015:AnnexinA2complexeswith}, and there is
-  evidence for its interaction with S100P as well; it is involved in
-  cell-cell interaction as well as vesicle trafficking and von
-  Willibrand Factor secretion.
-
-\item The expression regulation of another core gene, \textit{TIMP3},
-  has been implicated in pre-eclampsia
+  evidence for its interaction with S100P as well
+\item ANXA2 involved in cell-cell interaction as well as vesicle
+  trafficking and von Willibrand Factor secretion.
+
+\item \textit{TIMP3} has been implicated in pre-eclampsia
   \cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
@@
 
-\begin{frame}{PSG1}
+\begin{frame}{Pregnancy-Specific Glycoprotein Expansion}
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item 11 highly expressed \textit{PSG} genes in \human and some
+  expression in \smonkey and \bonobo
+\item \mouse specific expansion and high expression of \textit{PSG} as
+  well
+\item No other species have high expression of \textit{PSG} or
+  \textit{CEACAM}
+\end{itemize}
+\column{0.6\textwidth}
 \begin{figure}
@@
 \end{figure}
-\end{frame}
-
-\begin{frame}{CTSL}
+\end{columns}
+\end{frame}
+
+\begin{frame}{Cathepsin-L expansion}
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item Cathepsins are highly expressed in many placentas
+\item Expanded in \mouse and \galili and possibly also \carmeli to a
+  lesser degree
+\item Intracellular protease which degrades collagen, elastin, and in
+  IGF1R
+\end{itemize}
+\column{0.6\textwidth}
 \begin{figure}
@@
 \end{figure}
-\end{frame}
-
-\begin{frame}{PAG2}
+\end{columns}
+\end{frame}
+
+\begin{frame}{Pregnancy associated glycoprotein}
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item Massive expansion in ruminants (\cow and \sheep)
+\item 28 members of this family have $\mathrm{FPKM} \ge 1$ in \cow
+\item \Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}
+\item \Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}
+\end{itemize}
+\column{0.6\textwidth}
 \begin{figure}
@@
 \end{figure}
+\end{columns}
 \end{frame}
@@
 \begin{frame}{Galectin Expression}
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item Galectins are major regulators of pregnancy in fetal and
+  maternal tissues \cite{Than.ea2012:Galectinsguardiansofeutherian}
+\item \textit{LGALS1} expressed everywhere
+\item Primate expansion of \textit{LGALS13} and \textit{LGALS14}
+\item Likely present in ancestral placenta of therian mammals as it is
+  expressed even in \opossum
+\end{itemize}
+\column{0.6\textwidth}
 \begin{figure}
@@
 
-igfrelated.genes <- unique(grep("^IGF",combined.fpkm[,human_name],value=TRUE))
+igfrelated.genes <- unique(c(grep("^IGF",combined.fpkm[,human_name],value=TRUE),"H19"))
 
@@
          theme_grey(base_size = 9),
-         theme(legend.position = "top",
+         theme(legend.position = "right",
                axis.ticks = element_blank(),
                panel.background = element_blank(),
-               axis.text.y = element_text(face="italic"),
-               axis.text.x = element_text(angle = 300, hjust = 0, vjust=1,face="italic")),
+               axis.text.y = element_text(size=15,face="italic"),
+               axis.text.x = element_text(size=16,angle = 300, hjust = 0, vjust=1,face="italic")),
          facet_grid(type~.,scale="free_y",space="free_y")
@@
 @ 
-\caption{Expression of Hormone Genes, Galectins, and IGF related genes}
-\label{fig:hormone_genes}
-\end{figure}
+\end{figure}
+\end{columns}
 \end{frame}
@@
 \begin{frame}{Hormone Expression}
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item \human specific high expression of \textit{CGA}
+\item \human expansion of \textit{CGB}
+\item Vasodilation-involved proteins (\textit{NPPB} \& \textit{NPPC})
+  highly-expressed in \elephant and \dog, but not expressed elsewhere
+  \pnote{NPPB and NPPC have been seen in decidua of mouse
+    \cite{Cameron.ea1996:sitesofgeneexpression}.}
+\item \textit{PRL} highly expressed in \elephant
+\pnote{more about PRL later}
+\end{itemize}
+\column{0.6\textwidth}
 \begin{figure}
@@
 \begin{frame}{IGF Related}
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item \textit{IGF2} is highly expressed in most clades
+\item \textit{IGF2} promotes the growth and division of cells; highly active during fetal development
+\item \textit{IGF2} binding partners are also highly expressed
+  (\textit{IGFBP3}, \textit{IGFBP5})
+\end{itemize}
+\column{0.6\textwidth}
 \begin{figure}
@@
 
-\begin{frame}{Growth Hormone family}
+\begin{frame}{Growth Hormone (GH) family}
 
@@
 \column{0.4\textwidth}
-The \ac{GH} family has expanded in anthropoid primates, resulting in
+The GH family has expanded in anthropoid primates, resulting in
 approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
@@
                  gene.tree.table,gene.tree.expression,
+                 offset=2,
                  axis.text.size=5,
-                 fontsize=4)
+                 fontsize=10)
 @ 
@@
 
-  The human gene \textit{CGA} codes for the alpha subunit of multiple
-  hormones including pituitary hormones (luteinizing hormone,
-  follicle-stimulating hormone, thyroid-stimulating hormone), as well
-  as numerous chorionic gonadotropins with variable beta subunits
-  \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of
-  the chorionic gonadotropins is to prevent the regression of the
-  corpus luteum \cite{Hanson.ea1971:EffectsofHCGand}. Among the six
-  placentas we analyzed here, \textit{CGA} is only expressed in the
-  placentas of \human and \bonobo
-  (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}), which
-  corresponds with the emergence of the chorionic gonadotropins in
+  \begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item \textit{CGA} codes for the alpha subunit of multiple hormones as
+  well as numerous chorionic gonadotropins with variable beta subunits
+  \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}.
+
+\item Chorionic gonadotropins prevent the regression of the corpus
+  luteum \cite{Hanson.ea1971:EffectsofHCGand}.
+
+\item \textit{CGA} only expressed in the placentas of \human and
+  \bonobo
+
+  \pnote{Expression corresponds with emergence of the chorionic gonadotropins in
     anthropoid primates
-  \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly,
-  \textit{CGA} is expressed at much higher levels in both species than
-  the beta subunits with which its product dimerizes \textit{CGA} is
-  one of several genes noted by
-  \citet{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone
-  adaptive evolution of the \human{}-\textit{P. troglodytes} stem
-  lineage.
-
+    \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}.}
+
+\end{itemize}
+\column{0.6\textwidth}
 \begin{figure}
@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 offset.ratio=0.7,
-                 fontsize=plot.label.size,
-                 axis.text.size=3.5
-                 )
-@ 
-\end{figure}
-
+                 offset.ratio=1,
+                 fontsize=8,
+                 axis.text.size=5
+                 )
+@ 
+\end{figure}
+\end{columns}
 \end{frame}

### 2016-02-22 presentation/placenta_viviparous_presentation.Rnw
fix gene trees for most of the plots

@@
 @ 
-\caption{Genes whose expression changed significantly in the primate
-  order as compared to all other analyzed species.}
-\label{tab:primate_lineage_specific}
 \end{table}
@@
 
-% \begin{frame}{Prolactin}
-% \begin{itemize}
-% \item \elephant is only analyzed mammal with high prolactin expression
-%   (\textit{PRL}) in the placenta
-% \item However, \mouse, \carmeli, and \cow have duplications of
-%   \textit{PRL} with expression}
-% \end{itemize}
-% 
-% \begin{figure}
-% <<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-% 
-% tree.id <-
-%     genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
-% gene.tree.table <- trees[[tree.id]]$table
-% gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-% gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-% plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-%                  gene.tree.table,gene.tree.expression,
-%                  axis.text.size=5,
-%                  fontsize=5)
-% @ 
-% \end{figure}
-% 
-% \end{frame}
+\begin{frame}{Prolactin}
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item \elephant is only analyzed mammal with high prolactin expression
+  (\textit{PRL}) in the placenta
+\item However, \mouse, \carmeli, and \cow have duplications of
+  \textit{PRL} with expression
+\end{itemize}
+\column{0.6\textwidth}
+\begin{figure}
+<<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 axis.text.size=5,
+                 fontsize=4)
+@ 
+\end{figure}
+\end{columns}
+\end{frame}
 
@@
 \begin{frame}{CGB}
+  \begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item \textit{CGB} is highly expressed in \human
+
+\item Some expansion of \textit{CGB} in \bonobo, but not highly expressed
+\item High expression of \textit{LHB} in \smonkey
+\begin{itemize}
+\item May be indicative of expansion
+\item Would have to examine assembled transcripts to tell
+\end{itemize}
+\end{itemize}
+\column{0.6\textwidth}
 \begin{figure}
@@
                  offset.ratio=1,
-                 fontsize=plot.label.size,
+                 fontsize=10,
                  axis.text.size=3.5
@@
 
+\begin{frame}{Relaxin}
+  \begin{columns}
+    \column{0.4\textwidth}
+    \begin{itemize}
+    \item \textit{RLN} increasing maternal blood flow, widening the pubic
+      bone and relaxing uterine musculature in preparation for labor
+    \item We show that the placenta is a source of \textit{RLN} in \dog,
+      \elephant, \mouse, \carmeli, and \horse.
+    \end{itemize}
+    \column{0.6\textwidth}
+    \begin{figure}
+<<rln2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="RLN2" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 axis.text.size=5,
+                 fontsize=5)
+@ 
+\end{figure}
+\end{columns}
+\end{frame}
+
+\subsection{The Evolution of Placental Expression of Galectins}
+
+\begin{frame}{Galectins}
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item Galectins are major regulators of pregnancy in fetal and
+  maternal tissues
+\item \textit{LGALS1} mediates maternal-fetal
+  immune tolerance
+\item substantial expression of galectins in the \opossum
+\item consistent with ancestral placenta may have maternal-fetal
+  immune tolerance mediated by galectins
+\end{itemize}
+\column{0.6\textwidth}
+\begin{figure}
+<<galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 offset=0.8,
+                 axis.text.size=7,
+                 fontsize=7,
+                 subtree=c("ENSMODG00000000457","ENSCAFG00000016577")
+                 )
+@ 
+\end{figure}
+\end{columns}
+\end{frame}
+
 \begin{frame}
-Relaxin (\textit{RLN}) is a hormone that has multiple effects during
-pregnancy including increasing maternal blood flow, widening the pubic
-bone and relaxing uterine musculature in preparation for labor. In
-\dog, \elephant, \mouse, \carmeli, and \horse there is high expression
-of relaxin in the term placenta, with little evidence in other species
-examined here (Supplemental \cref{fig:rln2_evolution_tree}). \textit{RLN} is known
-to be produced in the placenta of the pregnant \dog and is used in the
-diagnosis of canine pregnancy
-\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
-the luteal ovary and placenta of \pig
-\cite{Hisaw.ea1948:Relaxininovaryof}, and the placenta of \horse
-\cite{Klonisch.ea1997:Placentallocalizationofrelaxin}. However, the
-source of relaxin produced by pregnant \elephant remains unclear
-\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in relaxin in
-\elephant has been suggested to be a major signal for the initiation
-of labor \cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive}.
-Here we show that the placenta of elephants show high expression of
-relaxin and is likely a major source of \textit{RLN} at term. In
-contrast to the results of
-\cite{Evans.ea1993:mouserelaxingenenucleotide}, we also see expression
-of \textit{RLN} in the placenta of \mouse and \carmeli.
-
-\end{frame}
-
-\subsection{The Evolution of Placental Expression of Galectins}
-
-\begin{frame}{Galectins}
-Galectins are recognized as major regulators of pregnancy in both
-fetal and maternal tissues
-\cite{Than.ea2012:Galectinsguardiansofeutherian}. Indeed, the
-evolution of new lineage-specific galectins has occurred in both
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item Evolution of new lineage-specific galectins has occurred in both
   primates and ruminants, with the genes \textit{LGALS13},
@@
   inducing apoptosis of maternal T-cells
-\cite{Than.ea2009:primatesubfamilyofgalectins} (Supplemental
-\cref{fig:galectins_lgals13_evolution_tree}). Indeed, \textit{LGALS14}
-has the highest expression of any galectin in the \human and \smonkey
-placenta (\cref{fig:hormone_genes}, Supplemental
-\cref{fig:galectins_lgals13_evolution_tree}), and other primate
-galectins such as \textit{LGALS13} and \textit{LGALS16} are also
-expressed in \human and \smonkey placenta. However, in the closely
-related \bonobo, \textit{LGALS14} is not found, and \textit{LGALS1} is
-the most abundant galectin. \textit{LGALS1} shows generally high
-expression in all placentas analyzed. \textit{LGALS1} is also involved
-in mediating maternal-fetal immune tolerance, and is expressed in both
-maternal and fetal tissue in primates
-\cite{Than.ea2008:Emergenceofhormonaland}. \textit{LGALS1} placental
-expression was thought to be confined to eutherian mammals, as there
-is evidence of adaptive evolution at this node, as well as the
-appearance of estrogen response elements in eutherian cis-regulatory
-regions \cite{Than.ea2008:Emergenceofhormonaland}. It was unclear as
-to whether galectins such as \textit{LGALS1} were important in the
-regulation of pregnancy in marsupials, as well as eutherians. Here we
-show substantial expression of galectins in the opossum, especially
-\textit{LGALS1} (Supplemental \cref{fig:galectins_evolution_tree}). In
-addition, \textit{LGALS2} is also expressed. The presence of galectins
-at the maternal-fetal interface in \opossum implies that it was also
-present in the ancestral placenta of therian mammals. Indeed, adaptive
-evolution of \textit{LGALS1} on the eutherian stem is likely due to
-the increased invasive nature of eutehrian placentation rather than
-the origination of maternal-fetal tolerance. Although the function of
-\textit{LGALS1} in the marsupial placenta needs to be confirmed, it is
-likely that the mediation of maternal-fetal immune tolerance by
-galectins is an ancient feature of mammalian viviparity.
-
+  \cite{Than.ea2009:primatesubfamilyofgalectins}
+\end{itemize}
+
+\column{0.6\textwidth}
+\begin{figure}
+<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 offset=0.8,
+                 axis.text.size=7,
+                 fontsize=10
+                 )
+@ 
+\end{figure}
+\end{columns}
 \end{frame}
@@
 \begin{frame}
-The \opossum placenta shows expression of genes that either do not
-exist in eutherians or are not found in the term eutherian placenta
-(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
-of hemoglobin in fetal tissues at term, as evidenced by the expression
-of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
-generally due to their very short gestation periods. Embryonic
-hemoglobins in marsupials have a lower affinity for oxygen and allow
-for a lower rate of oxygen uptake from maternal tissues
-\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
-uncharacterized gene that shows high expression (\textit{LOC10001195}) is not
-found in eutherians; Ensembl gene trees reveal that homologs of this
-gene exist in other amniotes such as birds and turtles, but are not
-present in eutherian mammals. The homolog of this gene in chicken is
-found to expressed in reproductive tract
-\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
-unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
-family and involved in innate immunity. It is expressed in the mammary
-gland of the tammar wallaby during pregnancy and early lactation
-\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and present in
-relative abundance in \opossum placenta.
-
+\begin{itemize}
+\item The \opossum placenta shows expression of genes that either do
+  not exist in eutherians or are not found in the term eutherian
+  placenta
+
+\item The existence of embryonic forms of hemoglobin in fetal tissues
+  at term, as evidenced by the expression of \textit{HBE} and
+  \textit{HBZ} is a feature of marsupials that is generally due to
+  their very short gestation periods.
+
+\item Embryonic hemoglobins in marsupials have a lower affinity for
+  oxygen and allow for a lower rate of oxygen uptake from maternal
+  tissues \cite{Holland.ea1997:Characterizationofembryonicglobin}.
+
+\item One novel uncharacterized gene that shows high expression
+  (\textit{LOC10001195}) is not found in eutherians; Ensembl gene
+  trees reveal that homologs of this gene exist in other amniotes such
+  as birds and turtles, but are not present in eutherian mammals.
+
+\item \textit{WFDC2} is a member of the whey acidic protein (WAP)
+  family and involved in innate immunity.
+\end{itemize}
 \end{frame}
@@
 \begin{frame}{PRSS16}
-
-11 genes that are related
-to the \textit{PRSS16} gene, a thymus-specific serine protease that
-may be involved regulating the presentation of self-peptides in CD4+
-T-lymphocytes \cite{Fornari.ea2012:TranscriptionprofilingofPrss16}.
-All 11 of these genes are expressed in the elephant placenta, three of
-which have very abundant mRNAs
-(\cref{tab:top50_elephant_placenta} and \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}C). It is
-possible that this elephant-specific gene family may be involved in
-immunity in the elephant placenta and the mediation of maternal-fetal
-immune tolerance; however, future work will need to test these
-hypotheses. 
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item \textit{PRSS16}, a thymus-specific serine protease that
+  regulates the presentation of self-peptides in CD4+ T-lymphocytes
+  \cite{Fornari.ea2012:TranscriptionprofilingofPrss16}.
+\item Possibly involved in the mediation of maternal-fetal immune
+  tolerance in \elephant
+\end{itemize}
+\column{0.6\textwidth}
 \begin{figure}
@@
 
-\item Identified common expression patterns of
+\item Identified common expression patterns of the mammalian
+  expression
 
@@
 
+\item Lots to learn about what is actually driving the changes in
+  expression
+
 \end{itemize}
@@
 
-\begin{frame}[plain]{References}
+\begin{frame}[plain,allowframebreaks]{References}
   \begin{center}

### 2016-03-01 manuscript/manuscript.Rnw
add changes that amy proposed

@@
 from the mother following delivery. Placental tissue from \human was
-collected upon delivery via a caesarian section. Near-term placental
+collected upon delivery via a cesarean section. Near-term placental
 tissue from \dog was collected upon the death of the mother from a car
@@
 
+\DLA{Totally rewrite the conclusion; it's terrible}
+
 We sequenced \Sexpr{star.logs[,length(unique(species))]} placentas
@@
 chorioamniotic placenta, with the marsupial placenta more enriched in
-genes the contributed to the active generation of energy. We also
+genes which contribute to the active generation of energy. We also
 identified that galectin expression, especially that of
@@
 two new expansions of genes that are highly expressed in the \elephant
-placenta hat may be involved in trophoplast differentiation and immune
+placenta that may be involved in trophoplast differentiation and immune
 system function, mirroring expansions in other groups such as
@@
 
-  Jason Caravas provided samples.
+  Jason Caravas performed initial analyses.
 

### 2016-03-07 manuscript/manuscript.Rnw
use Spalax instead of Nannospalax

@@
 \newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
-\newcommand{\galili}[0]{\textit{N. galili}\xspace}
+\newcommand{\galili}[0]{\textit{S. galili}\xspace}
 \newcommand{\sheep}[0]{\textit{O. aries}\xspace}
@@
 These include independent expansions of the prolactins in \mouse,
-\textit{Spalax carmeli}, \textit{Nannospalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
+\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
 in \cow, genes related to the thymus-specific serine protease
@@
     \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Nannospalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
+    \textit{Spalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
     \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
@@
 species.tree <- 
-    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Nannospalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
+    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Spalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
 pp <- ggtree(species.tree)+
@@
 5\% of the frequency distribution of \ac{FPKM} (Supplemental
-\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_nannospalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
+\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
 the top 100 genes with the highest \ac{FPKM} (Supplemental
-\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
+\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
 Due to the paucity of annotated gene ontology terms for some species,
@@
 \cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
-and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_nannospalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.
+and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.
 
@@
 <<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="nannospalax galili"))
-@ 
-\caption{\tableexplanation{\galili}{tab:all_nannospalax_galili_placenta}}
+output.top.genes.table(top.genes.table(species.name="spalax galili"))
+@ 
+\caption{\tableexplanation{\galili}{tab:all_spalax_galili_placenta}}
 \label{tab:top50_n_galili_placenta}
@@
 \cref{tab:top50_n_galili_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_nannospalax_galili_placenta}.
+genes in Supplemental \cref{tab:all_spalax_galili_placenta}.
 Components of annexin complexes are highly expressed
-(\Sexpr{gene.expression("nannospalax_galili","S100a6",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","S100a11",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","S100a10",complete=TRUE)},
+(\Sexpr{gene.expression("spalax_galili","S100a6",complete=TRUE)},
+\Sexpr{gene.expression("spalax_galili","S100a11",complete=TRUE)},
+\Sexpr{gene.expression("spalax_galili","S100a10",complete=TRUE)},
 and
-\Sexpr{gene.expression("nannospalax_galili","Anxa2",complete=TRUE)}),
+\Sexpr{gene.expression("spalax_galili","Anxa2",complete=TRUE)}),
 as are members of the thymosin family
-(\Sexpr{gene.expression("nannospalax_galili","Tmsb4x",complete=TRUE)}
+(\Sexpr{gene.expression("spalax_galili","Tmsb4x",complete=TRUE)}
 and
-\Sexpr{gene.expression("nannospalax_galili","Tmsb10",complete=TRUE)}),
+\Sexpr{gene.expression("spalax_galili","Tmsb10",complete=TRUE)}),
 various cytoskeleton related proteins
-(\Sexpr{gene.expression("nannospalax_galili","Actc1",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","Actg2",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","Acta2",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","Myl9",complete=TRUE)},
-\Sexpr{gene.expression("nannospalax_galili","Vim",complete=TRUE)}), and
+(\Sexpr{gene.expression("spalax_galili","Actc1",complete=TRUE)},
+\Sexpr{gene.expression("spalax_galili","Actg2",complete=TRUE)},
+\Sexpr{gene.expression("spalax_galili","Acta2",complete=TRUE)},
+\Sexpr{gene.expression("spalax_galili","Myl9",complete=TRUE)},
+\Sexpr{gene.expression("spalax_galili","Vim",complete=TRUE)}), and
 transmembrane protein of currently unknown function, 
-\Sexpr{gene.expression("nannospalax_galili","Gm10134",complete=TRUE)}.
+\Sexpr{gene.expression("spalax_galili","Gm10134",complete=TRUE)}.
 
@@
 (Supplemental %
-\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_nannospalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris}) using topGO \cite{topGO:2.20.0}.
+\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris}) using topGO \cite{topGO:2.20.0}.
 %
@@
 species (Supplemental %
-\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_nannospalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris})
+\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris})
 notable enrichments include extracellular matrix, female pregnancy,
@@
 % enriched in cell junction, vesicle transport, in utero embryonic
-% development, and many cytoskeletal processes. \textit{Nannospalax
+% development, and many cytoskeletal processes. \textit{Spalax
 %   galili} placentas are enriched in extracellular matrix organization,
@@
 expression, \galili
-(\Sexpr{gene.expression("nannospalax_galili","Hoxa11",complete=TRUE)})
+(\Sexpr{gene.expression("spalax_galili","Hoxa11",complete=TRUE)})
 is the rodent with the lowest expression of \textit{PRL}

### 2016-03-07 common_r_code/utility_functions.R

@@ -35,10 +35,10 @@ species.names <-
             list(sort.order=4,
                  common.name="mouse",
                  scientific.name="Mus musculus"),
-        "nannospalax galili"=
+        "spalax galili"=
             list(sort.order=5,
-                 name="nannospalax galili",
-                 scientific.name="Nannospalax galili"),
+                 name="spalax galili",
+                 scientific.name="Spalax galili"),
         "spalax carmeli"=
             list(sort.order=6,
                  name="spalax carmeli",

### 2016-03-07 data/fastq_fpkm/combine_fpkms.R

@@ -33,6 +33,9 @@ trinity.diamond.species <- NULL
 trinity.diamond <- list()
 for (file in trinity.diamond.files) {
     species.name <- gsub("_"," ",gsub("_trinity_diamond.txt","",file))
+    if (species.name=="nannospalax galili") {
+        species.name <- "spalax galili"
+    }
     trinity.diamond.species <-
         c(trinity.diamond.species,
           species.name)
@@ -68,11 +71,15 @@ for (file in gene.files) {
         gsub("_"," ",
              gsub("_genes.fpkm_tracking","",
                   gsub("SRR\\d+_","",file)))
+    orig.species.name <- species.name
+    if (species.name=="nannospalax galili") {
+        species.name <- "spalax galili"
+    }
     if (any(species.name %in% trinity.diamond.species)) {
         ## need to merge in the gene names here and then merge down
         ## the calls on the genes to only a single gene
         diamond.file <-
-            paste0(gsub(" ","_",species.name),"_trinity_diamond.txt")
+            paste0(gsub(" ","_",orig.species.name),"_trinity_diamond.txt")
         gene.fpkms[[file]][,tracking_id:=paste0(gsub(" +","_",species.name),
                                 "_",gsub(":.+","",locus))]
         setkey(gene.fpkms[[file]],"tracking_id")
@@ -110,9 +117,14 @@ gene.fpkms[,n_fpkm := length(FPKM),tracking_id]
 isoform.fpkms <- list();
 for (file in isoform.files) {
     isoform.fpkms[[file]] <- fread(file)
-    isoform.fpkms[[file]][,species := gsub("_"," ",
-                                       gsub("_isoforms.fpkm_tracking","",
-                                            gsub("SRR\\d+_","",file)))]
+    species.name <-
+        gsub("_"," ",
+             gsub("_isoforms.fpkm_tracking","",
+                  gsub("SRR\\d+_","",file)))
+    if (species.name=="nannospalax galili") {
+        species.name <- "spalax galili"
+    }
+    isoform.fpkms[[file]][,species := species.name]
     isoform.fpkms[[file]][,file:=file]
     i <- i+1
     setTxtProgressBar(pb,i)
@@ -133,9 +145,14 @@ for (file in star.log.files) {
     star.log$value <- gsub("\\t","",star.log$value)
     star.log$field <- gsub("(^\\s+|\\s+$)","",star.log$field)
     star.logs[[file]] <- data.table(star.log)[!grepl(":$",field),]
[…]

### 2016-03-07 manuscript/supplemental_material.Rnw
use Spalax instead of Nannospalax
fix significantly typo

@@
                                        SRR243593\\
-    \textit{Nannospalax galili}   & Eviator Nevo & GSE68763 & SRR2016467 \\
     \textit{Ovis aries}           & \citet{Archibald.ea2010:sheepgenomereferencesequence} & ERP005642 & ERR489144, ERR489145\\
@@
     \textit{Spalax carmeli}       & Anne Dupressoir, Eviator Nevo & GSE68763 & SRR2016467 \\
+    \textit{Spalax galili}        & Eviator Nevo                  & GSE68763 & SRR2016467 \\
     \textit{Sus scrofa}           & \citet{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 \\
@@
     \textit{Mus musculus}         & Mouse              & GRCm38     & 35,44,150 \\
-    \textit{Nannospalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
     \textit{Ovis aries}           & Sheep              & Oar\_v3.1   & 151 \\
@@
     \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
+    \textit{Spalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
     \textit{Sus scrofa}           & Duroc (pig)        & Sscrofa10.2 & 51 \\
@@
 for (this.species in names(species.names)) {
-    cat(paste0("\\subsubsection{Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
+    cat(paste0("\\subsubsection{Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                            " the top 100 \\textit{",
                            capfirst(this.species),"} expressed transcripts}\n\n"))
-    this.caption <- paste0("Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
+    this.caption <- paste0("Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                            " the top 100 \\textit{",capfirst(this.species),"} expressed transcripts")
@@
 for (this.species in names(species.names)) {
-    cat(paste0("\\subsubsection{Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
+    cat(paste0("\\subsubsection{Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                            " the top 1\\% of \\textit{",
                            capfirst(this.species),"} expressed transcripts}\n\n"))
-    this.caption <- paste0("Signficantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
+    this.caption <- paste0("Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
                            " the top 1\\% of \\textit{",

### 2016-03-07 manuscript/manuscript.Rnw
clean up yinotherian discussion; cite nevo book

@@
 
-The placenta arose at some point before the origin of therian mammals,
-a group that includes both marsupials and eutherian or “placental
-mammals”. The term “placental mammals” is misleading, as extant
-yinothereians (such as \textit{Ornithorhynchus anatinus}, the
+The placenta arose at some point before the origin of extant mammals,
+a group that includes marsupials, monotremes, and eutherian or
+“placental mammals”. The term “placental mammals” is misleading, as
+extant yinotherians (such as \textit{Ornithorhynchus anatinus}, the
 platypus) possess a rudimentary placenta, as do marsupials, although
@@
 short-tailed opossum], \galili [blind mole rat from Upper Galilee],
-\carmeli [blind mole rat from northern Israel]). Information on
-specimen ID, place and time of collection, and collector is shown for
-each sample in Supplemental \cref{tab:srr_numbers}. In the case of \bonobo,
-\cow, and \elephant, the placenta was collected at term upon expulsion
-from the mother following delivery. Placental tissue from \human was
-collected upon delivery via a cesarean section. Near-term placental
-tissue from \dog was collected upon the death of the mother from a car
-accident. Placental tissue from \opossum was sampled at fetal stage
-32, roughly one day prior to birth. Only the fetal portion of the
-placenta was sampled at the maternal-fetal interface (i.e. villous
-portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
-diffuse placenta for \opossum). All tissues were collected and stored
-in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
-could be performed. Tissues were homogenized in Trizol, and total RNA
-was isolated following the manufacturer's instructions. RNA was
-further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
+\carmeli [blind mole rat from northern Israel]
+\cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}). Information
+on specimen ID, place and time of collection, and collector is shown
+for each sample in Supplemental \cref{tab:srr_numbers}. In the case of
+\bonobo, \cow, and \elephant, the placenta was collected at term upon
+expulsion from the mother following delivery. Placental tissue from
+\human was collected upon delivery via a cesarean section. Near-term
+placental tissue from \dog was collected upon the death of the mother
+from a car accident. Placental tissue from \opossum was sampled at
+fetal stage 32, roughly one day prior to birth. Only the fetal portion
+of the placenta was sampled at the maternal-fetal interface (i.e.
+villous portion for \human, \bonobo, \dog, and \elephant; cotyledon
+for \cow; diffuse placenta for \opossum). All tissues were collected
+and stored in RNAlater (Life Technologies, Carlsbad, CA) until RNA
+isolation could be performed. Tissues were homogenized in Trizol, and
+total RNA was isolated following the manufacturer's instructions. RNA
+was further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
 conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),

### 2016-03-08 manuscript/manuscript.Rnw
describe n of samples and include t test explanation

@@
 
-\DLA{Add QC information}
-
 \subsubsection{Additional Sequences}
@@
 For more complete evolutionary comparisons additional sequences of
-placenta in \mouse \cite{Wang.ea2011:surveyfornovelimprinted}, \human
+placenta in \mouse
+(n=\Sexpr{star.logs[grepl("musculus",species),length(unique(file))]})
+\cite{Wang.ea2011:surveyfornovelimprinted}, \human
+(n=\Sexpr{star.logs[grepl("sapiens",species),length(unique(file))]})
 \cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
@@
 \cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
-\ac{SRA} (Supplemental \cref{tab:srr_numbers}).
+\ac{SRA} (Supplemental \cref{tab:srr_numbers}) and analyzed using the
+same methodology as mRNAs sequenced specifically for this study.
 
@@
 to 0.05. Enrichment tests were considered with two gene lists per
-transcriptome: annotated genes with expression levels of the top
-5\% of the frequency distribution of \ac{FPKM} (Supplemental
+transcriptome: annotated genes with expression levels of the top 5\%
+of the frequency distribution of \ac{FPKM} (Supplemental
 \cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
@@
 Due to the paucity of annotated gene ontology terms for some species,
-gene lists were compared using the \human Gene Ontology
-using the 1:1 \human ortholog obtained
+gene lists were compared using the \human Gene Ontology using the 1:1
+\human ortholog. Enrichment was considered out of the set of all 1:1
+orthologs present in each species.
 
@@
 analyses of gene expression evolution described below.
+
+Significant shifts in expression levels between groups of species were
+calculated using Student's t test on 1:1 orthologs using the mean of
+each species to avoid large biases towards samples with $n \gg 1$ and
+considered significant if the \ac{FDR} as estimated by the Benjamini
+and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
+$\le 0.05$.
 

### 2016-03-08 manuscript/common_latex.tex
add common_latex.tex

--- 2016-03-07 manuscript/manuscript.Rnw
@@
-%\documentclass[english,12pt]{article}
-\documentclass[doublespacing]{bmcart_dla}
-\usepackage{fontspec}
-\setmainfont{FreeSans}
-\setsansfont{FreeSans}
-\setmonofont{FreeMono}
-\setmathrm{FreeSans}
-\setmathsf{FreeSans}
-\setmathtt{FreeSans}
-%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
-\usepackage[letterpaper]{geometry}
-\usepackage{url}
-% \usepackage{fancyhdr}
-\usepackage{graphicx}
-% \usepackage[bf]{caption}
-\usepackage{rotating}
-\usepackage{wrapfig}
-\usepackage{booktabs}
-\usepackage{float}
-\usepackage{multirow}
-\usepackage{setspace}
-\usepackage{wrapfig}
-% \usepackage{txfonts}
-%\usepackage[plain]{fancyref}
-\usepackage{acronym}
-\usepackage{array}
-\usepackage{dcolumn}
-\usepackage{adjustbox}
-\usepackage{longtable}
-\usepackage{pdflscape}
-\usepackage{amsmath}
-\usepackage{sfmath}
-\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
-\addbibresource{references.bib}
-% \usepackage[noblocks,auth-sc]{authblk}
-\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
-%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
-% Floats at end for journals
-\usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
-\renewcommand{\efloatseparator}{\mbox{}}
-%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
-%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
-\usepackage[nomargin,inline,draft]{fixme}
-\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
-\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
-% \usepackage[x11names,svgnames]{xcolor}
-% \usepackage{texshade}
-% \usepackage{tikz}
-\usepackage{nameref}
-\usepackage{xspace}
-% this package will stretch the font slightly if necessary.
-\usepackage{microtype}
-\usepackage{zref-xr,zref-user}
-\zxrsetup{toltxlabel}
-\zexternaldocument*{supplemental_material}
-\usepackage[capitalize]{cleveref}
-% \newenvironment{narrow}[2]{%
-%   \begin{list}{}{%
-%       \setlength{\topsep}{0pt}%
-%       \setlength{\leftmargin}{#1}%
-%       \setlength{\rightmargin}{#2}%
-%       \setlength{\listparindent}{\parindent}%
-%       \setlength{\itemindent}{\parindent}%
-%       \setlength{\parsep}{\parskip}}%
-%   \item[]}{\end{list}}
-% \def\newblock{\hskip}
-% \newenvironment{paperquote}{%
-%   \begin{quote}%
-%      \it
-%   }%
-%   {\end{quote}}
-% \renewcommand{\textfraction}{0.15}
-% \renewcommand{\topfraction}{0.85}
-% \renewcommand{\bottomfraction}{0.65}
-% \renewcommand{\floatpagefraction}{0.60}
-%\renewcommand{\baselinestretch}{1.8}
-% \newenvironment{enumerate*}%
-%   {\begin{enumerate}%
-%     \setlength{\itemsep}{0pt}%
-%     \setlength{\parskip}{0pt}}%
-%   {\end{enumerate}}
-% \newenvironment{itemize*}%
-%   {\begin{itemize}%
-%     \setlength{\itemsep}{0pt}%
-%     \setlength{\parskip}{0pt}}%
-%   {\end{itemize}}
-% \setcounter{tocdepth}{3}
-% \oddsidemargin 0.0in 
-% \textwidth 6in
-% \raggedbottom
-% \clubpenalty = 10000
-% \widowpenalty = 10000
-% set a large emergency stretch to avoid issues with overfull boxes
-% \setlength{\emergencystretch}{20em}
-% make the afill smaller
-% \renewcommand\Affilfont{\itshape\small}
-\renewcommand\cite[1]{\autocite{#1}}
-% \pagestyle{fancy}
-
 % species names
@@
 % slightly different.
-\newcommand{\tableexplanation}[2]{The top 50 genes by \ac{FPKM} in #1
-  which are not annotated as Mt\_rRNA, Mt\_tRNA, rRNA or whose \human
+\newcommand{\tableexplanation}[2]{The top 50 genes by FPKM in #1 which
+  are not annotated as Mt\_rRNA, Mt\_tRNA, rRNA or whose \human
   ortholog is a housekeeping gene. Gene symbol is the gene symbol or
@@
   Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes from species
-  not included in this study or having expression by \ac{FPKM} less
-  than 1 were excluded from the plotted tree. Heatmap cells in white
-  are from species where there was no expression of the relevant
-  transcript (or a species where that transcript does not apply). The
-  species names are abbreviated with the first letter of the genus and
-  first two letters of the species (for example, \textit{Homo sapiens}
-  is \textit{H.sa.}). Genes with duplicate gene names in the same
-  species (or no gene named defined) are listed by the Ensemble gene
-  ID.}
+  not included in this study or having expression by FPKM less than 1
+  were excluded from the plotted tree. Heatmap cells in white are from
+  species where there was no expression of the relevant transcript (or
+  a species where that transcript does not apply). The species names
+  are abbreviated with the first letter of the genus and first two
+  letters of the species (for example, \textit{Homo sapiens} is
+  \textit{H.sa.}). Genes with duplicate gene names in the same species
+  (or no gene named defined) are listed by the Ensemble gene ID.}
 
-\begin{document}
-\raggedright
-
-<<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
-opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
-opts_chunk$set(cache=TRUE, autodep=TRUE)
-options(device = function(file, width = 8, height = 7, ...) {
-  cairo_pdf(tempfile(), width = width, height = height, ...)
-})
-options(scipen = -2, digits = 0)
-library("data.table")
-library("ggplot2")
-library("reshape2")
-library("xtable")
-library("grid")
-library("gridExtra")
-library("ape")
-library("ggtree")
-library("topGO")
-library("org.Hs.eg.db")
-library("scales")
-library("magrittr")
-library("tidyr")
-
-@
-
-<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
-for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R"))) {
-    source(a)
-}
-@
-
-<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","placenta_go_results_mf","hugo_nomenclature","gene_annotation")))>>=
-for (a in paste0("../data/",c("combined_fpkm","gene_trees",
-                              "housekeeping_genes_superset",
-                              "placenta_core_transcriptome",
-                              "placenta_go_results_bp",
-                              "placenta_go_results_mf",
-                              "hugo_nomenclature",
-                              "gene_annotation"
-                              ))) {
-    load(a)
-}
-combined.fpkm[,species:=factor(species,levels=names(species.names))]
-setkey(genes.to.tree,"gene_id")
-star.logs[,value:=gsub("%$","",value)]
-star.logs.wide <- 
-    data.table(dcast(star.logs,species+file~field))
-setnames(star.logs.wide,
-         "Average input read length",
-         "input.read.length")
-setnames(star.logs.wide,
-         "Number of input reads",
-         "input.reads")
-setnames(star.logs.wide,
-         "Uniquely mapped reads number",
-         "unique.map.reads")
-star.logs.wide[,`% bad reads`:=(as.numeric(`% of reads mapped to too many loci`)+
-                            as.numeric(`% of reads unmapped: other`)+
-                            as.numeric(`% of reads unmapped: too many mismatches`)+
-                            as.numeric(`% of reads unmapped: too short`))]
-star.logs.wide[,`% good reads`:=100-`% bad reads`]
-setkey(hugo.nomenclature,"Approved Symbol")
-@
-
-<<load.go.results,echo=FALSE,results="hide",cache.extra=tools::md5sum("../go_annotation/go_tissue_analysis")>>=
-load("../go_annotation/go_tissue_analysis")
-
-@
-
-<<plot.defaults,echo=FALSE,results="hide">>=
-grad.def <- list(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = NA)
-
-@
-
-% Acronyms
-\input{acronyms}
-
-\begin{frontmatter}
-\begin{fmbox}
-\dochead{Research}%
-\title{The evolution of gene expression in the term placenta of viviparous mammals}%
-%
-\author[addressref={aff1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
-\author[addressref={aff2},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
-\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL}\fnm{Amy} \snm{Weckle}}
-\author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}
-\author[addressref={aff4},email={agnewd@dcpah.msu.edu}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
-\author[addressref={aff5},email={kbenirschke@ucsd.edu}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
-\author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
-\author[addressref={aff11},email={nevo@research.haifa.ac.il}]{\inits{E}\fnm{Eviatar} \snm{Nevo}}
-\author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
-\author[addressref={aff9},email={gunter.wagner@yale.edu}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
-\author[addressref={aff7},email={prbchiefstaff@med.wayne.edu}]{\inits{R}\fnm{Roberto} \snm{Romero}}
-\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
-\address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[id=aff2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
-\address[id=aff3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[id=aff4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
-\address[id=aff5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
-\address[id=aff6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
-\address[id=aff7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
-\address[id=aff8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[id=aff9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
-\address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[id=aff11]{The Institute of Evolution, University of Haifa, Haifa, Israel}
-
-\begin{artnotes}
-\end{artnotes}
-
-\end{fmbox}
-% \twocolumn
-
-\begin{abstractbox}
-\begin{abstract}
-% \section{Abstract}
-  \parttitle{Background} The placenta is arguably the most
-  anatomically variable organ in mammals. Understanding gene
-  expression patterns among placentas from multiple species can
-  illuminate the molecular basis of placenta function as well as
-  potentially enrich our knowledge of the biology of obstetrical
-  syndromes.
-
-  \parttitle{Results} We measured the expression profiles of the term
-  placenta of \Sexpr{star.logs[,length(unique(species))]} species of
-  therian (i.e. marsupials and “placental”) mammals using RNA-Seq. We
-  identified multiple genes and transcripts which are differentially
-  expressed across lineages including prolactin, relaxin, and
-  galectins, as well as identifying the non-housekeeping genes which
-  are core to placenta function (such as components of annexin
-  complexes) in all \Sexpr{star.logs[,length(unique(species))]}
-  species studied.
-
-  \parttitle{Conclusions} While there are multiple non-housekeeping
-  genes which are highly expressed in all placentas examined and are
-  likely to be involved in the basic functions of the mammalian
-  placenta, we see incredible lineage-specific diversity in the
-  expression and expansion of many genes. These lineage-specific genes
-  are likely to play a role in the anatomical variability of the
-  placenta and may also be implicated in important human placental
-  disorders such as pre-eclampsia.
-
-% We identified significant shifts in gene expression along specific
-% evolutionary lineages using a model of character evolution. Genes that
-% showed significant elevation in expression along the human lineage
-% ($\mathrm{FDR}\le 0.01$) were enriched in GO categories such as
-% angiogenesis, tube morphogenesis, cell migration, and regulation of
-% inflammatory response. Three of these genes (KISS1, ADAM12, GPC3)
-% showed exceptionally high expression in humans; these genes are
-% implicated in pre-eclampsia and may indicate that the risk of
-% pre-eclampsia recently emerged on the human lineage.
-
-\textbf{Keywords:} mammal, gene expression, evolution, RNA-seq, placenta
-
-\end{abstract}
-\end{abstractbox}
-\end{frontmatter}
-
-\section{Background}
-
-The main role of the placenta in viviparous mammals is the transfer of
-nutrients, oxygen, and waste between mother and offspring. Although
-these functions are conserved among mammals, the placenta is perhaps
-the most variable organ in terms of structure and form
-\cite{Mossman.1937:comparativemorphogenesisofthe}. Species differ in
-the shape of attachment (i.e., discoid, diffuse, cotyledonary,
-zonary), degree of invasiveness of fetal tissues into maternal tissues
-(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
-interdigitation of maternal-fetal tissue (i.e., villous, folded,
-trabecular, lamellar, labrynthine)
-\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
-mammals differ drastically in the number and size of their offspring,
-as well as the length of gestation, traits often associated with
-specific aspects of placental morphology
-\cite{Capellini.ea2011:Placentationandmaternalinvestment}
-(\cref{tab:gestational_periods}).
-
-The placenta arose at some point before the origin of therian mammals,
-a group that includes both marsupials and eutherian or “placental
-mammals”. The term “placental mammals” is misleading, as extant
-yinothereians (such as \textit{Ornithorhynchus anatinus}, the
-platypus) possess a rudimentary placenta, as do marsupials, although
-the direct apposition to maternal tissues lasts for a proportionally
-shorter time during gestation
-\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
-Marsupials also differ from eutherian mammals in that most possess
-only a choriovitelline placenta, produced from the fusion of the
-chorion and yolk sac, during most of gestation. Although some
-eutherian mammals may possess a choriovitelline placenta at some point
-in development, the major site of nutrient exchange in placentals
-occurs via the chorioallantoic placenta, produced from the fusion of
-the chorion and allantois
-\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
-the first placenta was more like that of marsupials or eutherians;
-however, using phylogenetic methods, the ancestral eutherian was
-reconstructed as possessing a relatively invasive hemochorial
-placenta, in which the maternal and fetal blood supply is separated by
-one tissue layer, as in humans and mouse
-\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
-\DLA{To discuss maternal/fetal split with derek}
-
-Although the morphology of the mammalian placenta has been well
-characterized in a number of species
-\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
-molecular environment is only beginning to be understood from a
-genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
-Studies of certain model species (i.e., \textit{Homo sapiens},
-\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
-divergence in the genetics of the placenta, especially in later stages
-of pregnancy
-\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
-For example, each lineage has evolved expansive placenta-specific gene
-families with expression patterns that can vary in space and time
-\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
-These include independent expansions of the prolactins in \mouse,
-\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
-in \cow, genes related to the thymus-specific serine protease
-\textit{PRSS16} in \textit{Loxodona africana}, and the expansion of the growth hormone
-and chorionic gonadotropin beta genes in hominoid primates
-\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
-Expression patterns of specific genes also can vary widely between
-species. For example, the expression of orthologous genes in the
-placenta have been found to vary substantially between \human and
-\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
-recruitment of \textit{SIGLEC6} to placental expression in \human
-differs from both \textit{Pan troglodytes} and \mouse
-\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
-
-One explanation for the divergent paths of the placental interface is
-the distinct evolutionary interplay between maternal and fetal
-strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}.
-The maternal strategy is to requisition an adequate degree of
-nutrients to the fetus without risking the health of the mother, while
-the fetal strategy via paternal genes is to extract as many resources
-from the mother as possible. Therefore, an interplay between these
-extremes can set up an intricate “arms race”, involving the imprinting
-of parental alleles, expansion of gene families, and enlistment of
-existing genes to expression at the maternal-fetal interface
-\cite{Haig1993:Geneticconflictsinhuman}.
-
-One symptom of this maternal-fetal conflict is its potential affect on
-health
-\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
-addition to having an invasive hemochorial placenta, \human possess a
-deeply invasive trophoblast layer that can penetrate the myometrium of
-the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
-Defects in placental growth are known to be associated with the
-devastating condition of pre-eclampsia, which can cause hypertension,
-proteinuria, and has the potential to lead to death of the mother
-and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
-Although cases have been documented in \textit{P. troglodytes}
-\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
-\textit{Gorilla gorilla}
-\cite{Thornton.ea1992:Convulsionsinpregnancyin}, pre-eclampsia has a
-far higher incidence in \human pregnancies with as much as 4\% of
-pregnancies being affected in the United States
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. \textit{P.
-  troglodytes} and \textit{G. gorilla} also have deeply invasive
-trophoblasts, although the progression of invasion in great apes may
-be slower than \human
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however,
-large-scale expression studies of great ape placentas have not been
-conducted. The increasing affordability of next-generation sequencing
-technology has enabled the large-scale study of specific tissues from
-a wide array of species to address broader phylogenetic questions in
-the evolution of gene expression
-\cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use RNA-Seq
-technology to sequence and quantify the mRNA environment of mammalian
-placentas at term, including that of the \human, \textit{Pan paniscus}
-(the sister species to \textit{P. troglodytes}), \textit{Ateles
-  fusciceps}, \mouse, \galili, \carmeli, \textit{Canis familiaris},
-\cow, \elephant, \textit{Dasypus novemcinctus}, and
-\textit{Monodelphis domesticus}. These species represent four major
-branches of eutherian mammals (Euarchontoglires, Laurasiatheria,
-Afrotheria, and Xenartha) as well as marsupials. These placentas also represent a
-wide array of placental morphologies as well as traits such as number
-and size of offspring and length of gestation
-(\cref{tab:gestational_periods}). We use this data, along with
-existing RNA-Seq data from \mouse
-\cite{Wang.ea2011:surveyfornovelimprinted}, \human
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \textit{Equus caballus}
-\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis aries}
-\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \textit{Sus scrofa}
-\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
-differences between expression in marsupial and eutherian mammals at
-term, identify the evolution of expression in important placental
-genes such as hormones and galectins, identify novel lineage-specific
-expression patterns, and quantify \human specific differences in gene
-expression that may be related to obstetrical disorders such as
-pre-eclampsia.
-
-\begin{table}
-  \begin{tabular}{c c c c}
-    \toprule
-    Species                       & Gestation (days) & Litter Size  & Source \\
-    \midrule
-    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
-    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
-    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
-    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Spalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
-    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
-    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
-    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \bottomrule
-  \end{tabular}
-  \caption{Gestational periods in days and average liter sizes for studied species.}
-  \label{tab:gestational_periods}
-\end{table}
-
-\section{Methods}
-
-\subsection{Collection of Placental Tissue and RNA Extraction}
-
-Fetal tissue was collected from the placenta of eight mammalian
-species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
-[domestic cow], \elephant [African savanna elephant], \opossum [gray
-short-tailed opossum], \galili [blind mole rat from Upper Galilee],
-\carmeli [blind mole rat from northern Israel]). Information on
-specimen ID, place and time of collection, and collector is shown for
-each sample in Supplemental \cref{tab:srr_numbers}. In the case of \bonobo,
-\cow, and \elephant, the placenta was collected at term upon expulsion
-from the mother following delivery. Placental tissue from \human was
-collected upon delivery via a cesarean section. Near-term placental
-tissue from \dog was collected upon the death of the mother from a car
-accident. Placental tissue from \opossum was sampled at fetal stage
-32, roughly one day prior to birth. Only the fetal portion of the
-placenta was sampled at the maternal-fetal interface (i.e. villous
-portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
-diffuse placenta for \opossum). All tissues were collected and stored
-in RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation
-could be performed. Tissues were homogenized in Trizol, and total RNA
-was isolated following the manufacturer's instructions. RNA was
-further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
-conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
-following the manufacturer’s protocol to remove any contaminating DNA
-from the sample. Total RNA was then evaluated on a Nanodrop 1000
-(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
-(Santa Clara, CA) to confirm concentration and quality of the
-isolation. RNA quality was assessed using A260/A280 and A260/A230
-absorbance values.
-
-% \DLA{Add details for new sequences and fix numbers}
-
-\subsection{Sequencing}
-
-All RNA-Seq mRNA isolation, library construction, and Illumina
-sequencing was performed by the Applied Genomics Technology Center
-(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
-total RNA and the standard Illumina protocol. Paired-end sequencing
-was performed using the Illumina Genome Analyzer II with an insert
-size of 230 base pairs (bp). Read lengths for species are shown in
-Supplemental \cref{tab:ref_genome_aligned_species}.
-
-\DLA{Add QC information}
-
-\subsubsection{Additional Sequences}
-
-For more complete evolutionary comparisons additional sequences of
-placenta in \mouse \cite{Wang.ea2011:surveyfornovelimprinted}, \human
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
-\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
-\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
-\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
-\ac{SRA} (Supplemental \cref{tab:srr_numbers}).
-
-\subsection{Transcriptome Assembly, Alignment, and Quantification}
-
-The quality of reads was evaluated using FASTQC v0.11.2
-(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
-Transcriptomes without a suitable reference genome (\smonkey, \galili,
-\carmeli; italic reference in Supplemental
-\cref{tab:ref_genome_aligned_species}) were assembled 
-using Trinity v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence}
-and annotated using Diamond
-\cite{Buchfink.ea2015:Fastandsensitiveprotein} against either the
-\human or \mouse transcriptome
-(Supplemental \cref{tab:ref_genome_aligned_species}).
-% 
-All transcriptomes were then aligned to the appropriate reference (or
-assembled) genome using STAR v2.4.2a
-\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
-values in \ac{FPKM}
-\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
-quantified using Cufflinks v2.2.1
-\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
-%
-The genome of the closely-related \textit{P. troglodytes} was used as
-a reference for \bonobo. The separation between the two recognized Pan
-species is generally perceived to be $\approx 1$ million years
-\cite{Prüfer.ea2012:bonobogenomecomparedwith}.
-
-\begin{figure}
-<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
-species.tree <- 
-    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Spalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
-pp <- ggtree(species.tree)+
-    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
-    scale_x_continuous(limits=c(0,10))
-print(pp %>%
-      ## monodelphis
-      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
-      ## Armadillo
-      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
-      ## elephant
-      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
-      ## dog
-      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
-      ## horse
-      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
-      ## pig
-      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
-      ## sheep
-      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
-      ## cow
-      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
-      ## s carmeli
-      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
-      ## n galili
-      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
-      ## mouse
-      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
-      ## bonobo
-      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
-      ## spider monkey
-      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
-      ## human
-      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
-      )
-@ 
-\caption{Species tree for organisms studied}
-\label{fig:species_tree}
-\end{figure}
-
-\subsection{Exploration of Individual Transcriptome Data}
-
-To determine enrichment levels of functional categories of genes in
-highly-expressed transcripts in each transcriptome, we performed
-functional annotation cluster analyses per transcriptome using topGO
-\cite{topGO:2.20.0}. Gene Ontology clusters were considered
-significantly enriched if the \ac{FDR} as estimated by the Benjamini
-and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
-$p$ values obtained from the Fisher Exact Test were less than or equal
-to 0.05. Enrichment tests were considered with two gene lists per
-transcriptome: annotated genes with expression levels of the top
-5\% of the frequency distribution of \ac{FPKM} (Supplemental
-\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
-the top 100 genes with the highest \ac{FPKM} (Supplemental
-\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
-Due to the paucity of annotated gene ontology terms for some species,
-gene lists were compared using the \human Gene Ontology
-using the 1:1 \human ortholog obtained
-
-\subsection{Comparative Analysis of Gene Expression}
-
-All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
-\cow, \dog, \elephant, \mouse, and \human were determined using
-Ensembl v.80 gene trees via Biomart
-\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
-package
-\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
-A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
-  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
-in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
-  sapiens" & mean_fpkm >= 10,length(unique(human_name))]} 1:1
-orthologous genes with \ac{FPKM} $\ge 10$ were retained for future
-analyses of gene expression evolution described below.
-
-% We used the software \ac{OUCH} in the R statistical environment to
-% investigate significant shifts in gene expression levels over
-% time\cite{Butler.2004:phlogeneticcomparativeanalysis}. \ac{OUCH}
-% institutes an \ac{OU} model of evolution, which allows for different
-% optima for quantitative trait values along user-specified branches of
-% a phylogenetic tree. The \ac{OU} model explicitly accounts for change
-% in quantitative trait values by introducing two parameters into the
-% model of stochastic Brownian motion of change in trait value along a
-% phylogenetic tree through time. These parameters are $\theta$, an
-% optimum trait value, and $\alpha$, the strength of selection
-% (interpreted as “pull” toward the optimum). In addition, the \ac{OU}
-% model can also integrate evolutionary change by assuming that each
-% user-specified set of braches has its own separate \ac{OU} parameters
-% \cite{Butler.2004:phlogeneticcomparativeanalysis}.
-% 
-% \DLA{The description of the OU process is terrible; it needs to be
-%   completely redone}
-% 
-% Various \ac{OU} branch models were tested against a model in which no
-% variation in optima ($\theta$) among branches was allowed. The
-% following branches were specified in separate models to have a
-% different optimum than all other branches: human, bonobo, mouse, dog,
-% cow, elephant, opossum, human+bonobo, Euarchontoglires, and
-% Boreoeutheria. Likelihood values were computed for each branch model
-% and compared to an \ac{OU} model with one optimum using likelihood
-% ratio tests with df=1. We filtered out genes with universally low
-% expression levels, and included only genes in which expression in at
-% least one species had an $\mathrm{FPKM}\ge10$ as described above. We
-% corrected for multiple testing using the \ac{FDR} and using the
-% Benjamini-Hochberg correction\cite{Benjamini1995:bh_procedure}. We
-% used the time-calibrated phylogenetic tree of
-% \cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
-% not used in this analyses pruned.
-% 
-% We also calculated the enrichment of functional categories in the
-% lists of genes which displayed a significant shift in expression on
-% each branch using the same methodology as in the per-species gene
-% lists \DLA{reference the tables once this analysis is complete}.
-
-\subsection{Correlation of Tissue Specific Index with Lineage Specific Genes}
-
-To evaluate the extent of tissue specificity of each gene, we
-calculated the \ac{TSI} based on human RNASeq data in 3267 samples in
-126 tissues the Roadmap Epigenomics
-\cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping} and GTEx projects
-\cite{Gtexconsortium.2015:Thegenotypetissueexpression} as well
-additional placenta
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} and uterus
-\cite{Chan.ea2014:Assessmentofmyometrialtranscriptome} samples from
-SRA. The \ac{TSI} measures the extent to which a specific gene has
-predominant expression in only one tissue
-\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}. A gene
-expressed in only one tissue would have a \ac{TSI} of 1, whereas a
-gene expressed in all tissues would have a \ac{TSI} of 0.
-
-% \subsection{Validation of Primate Expression using Spider Monkey Placenta Transcriptome}
-% 
-% To verify expression levels in human and bonobo, we also generated
-% RNA-Seq data for the placenta transcriptome of the black-headed spider
-% monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
-% Ateles is outside the clade, which includes hominids and can be used
-% to polarize expression values of highly expressed genes in Homo and
-% Pan. We sampled placental tissue from an Ateles placenta expelled from
-% the mother shortly after birth (see Supplemental Table X). We used
-% similar procedures discussed above to extract and sequence this
-% sample. [DEREK—I’m unsure how to proceed on describing this so as to
-% not step on Natalie’s toes]
-% 
-% \DLA{Complete discussion of the ateles placenta}
-
-\subsection{Sequence Deposition and Code}
-
-All of the code used to produce every analysis presented in this paper
-is available at
-\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
-sequences used are deposited in SRA, and can be obtained using
-\DLA{add GSE label for the experiment}.
-
-\section{Results}
-
-\subsection{Transcriptome Statistics}
-
-Supplemental \cref{fig:transcriptome_statistics} shows the RNAseq read and
-alignment statistics for all species. For transcriptomes sequenced
-with 75 bp %
-%
-(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
-%
-,
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
-with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
-mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
-%
-For transcriptomes sequenced with 150 bp, %
-% The mismatch here is because we're reporting trimmed reads here, not
-% full length reads
-(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
-%
-,
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
-%
-with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
-of reads mapping to the
-respective reference genome (\cref{tab:ref_genome_aligned_species}).
-%
-%
-For transcriptomes sequenced with 35 bp, %
-% The mismatch here is because we're reporting trimmed reads here, not
-% full length reads
-(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
-%
-,
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
-%
-with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
-of reads mapping to the
-respective reference genome (\cref{tab:ref_genome_aligned_species}).
-The top 50 expressed genes in each genome analyzed are shown in
-\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
-and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.
-
-% The percentage of bases that mapped to Ensembl annotated mRNA ranged
-% from ~35\% in elephant to ~85\% in human, with percentages of bases
-% mapping to annotated protein-coding segments of genes ranging from
-% ~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
-% from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
-% from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
-% ranged from ~43\% in elephant to 80\% in bonobo. The number of
-% annotated transcripts to which reads mapped in each respective genomes
-% ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
-% and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
-% greater degree of mapping to annotated protein-coding sequences in
-% humans corresponds to the greater number of Ensembl annotated
-% sequences in the human genome compared to the other five reference
-% genomes (Flicek, 2012).
-
-<<top.genes.function,echo=FALSE,results="hide">>=
-top.genes.table <- function(species.name,n=50) {
-    alignment.species <- species.name
-    if (alignment.species=="pan paniscus") {
-        alignment.species <- "pan troglodytes"
-    } else if (alignment.species=="ateles fusciceps") {
-        alignment.species <- "homo sapiens"
-    } else if (grepl("spalax",alignment.species)) {
-        alignment.species <- "mus musculus"
-    }
-    temp <- 
-        combined.fpkm[species==species.name &
-                      !(!is.na(human_name) & 
-                        human_name %in% housekeeping.genes.superset[,gene_short_name]) &
-                      ! (gene_biotype %in% c("Mt_tRNA","Mt_rRNA","rRNA")) &
-                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_name) &
-                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_alignment_symbol) &
-                      !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[lL]?(?:-.+)?$|rRNA$)",
-                             gene_short_name)
-                      ][order(-mean_fpkm)][,list(name_or_id,
-                                                 human_alignment_symbol,
-                                                 human_name,mean_fpkm,sd_fpkm,n_fpkm)]
-    ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
-    temp <- temp[!(is.na(name_or_id)|name_or_id==""),]
-    temp <- temp[!duplicated(name_or_id)][1:n]
-    temp[,used_alignment:=FALSE]
-    temp[is.na(human_name) & !is.na(human_alignment_symbol),
-                    used_alignment:=TRUE]
-    temp[is.na(human_name),human_name:=human_alignment_symbol]
-    setkey(temp,"human_name")
-    temp <- 
-        hugo.nomenclature[temp][order(-mean_fpkm),
-                                list(name_or_id,used_alignment,`Approved Symbol`,`Approved Name`,
-                                     mean_fpkm,sd_fpkm,n_fpkm)]
-    temp[!is.na(name_or_id),
-         name_or_id:=paste0("\\href{http://www.ensembl.org/",
-                             capfirst(gsub(" ","_",alignment.species)),
-                             "/Gene/Summary?g=",name_or_id,"}{\\textit{",
-                             xtable.sanitize.text(name_or_id),"}}")]
-    temp[!is.na(`Approved Name`),
-         `Approved Name`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Name`),"}")]
-    temp[!is.na(`Approved Symbol`),
-         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{\\textit{",xtable.sanitize.text(`Approved Symbol`),"}}")]
-    temp[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
-    temp[,used_alignment:=NULL]
-    temp[,sem_fpkm:=as.numeric(NA)]
-    temp[n_fpkm>1,sem_fpkm:=sd_fpkm/sqrt(n_fpkm)]
-    temp[,sd_fpkm:=NULL]
-    temp[,n_fpkm:=NULL]
-    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM","SEM"))
-    if (temp[,all(is.na(`SEM`))]) {
-        temp[,`SEM`:=NULL]
-    }
-    if (species.name=="homo sapiens" || species.name=="ateles fusciceps") {
-        temp[,`Human Symbol`:=NULL]
-    }
-    temp[,`FPKM`:=round(`FPKM`)]
-    return(temp)
-}
-gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE) {
-    species.name <- gsub("_"," ",species.name)
-    fpkm <- combined.fpkm[species==species.name &
-                              (gene_short_name==gene | 
-                               gene_id==gene),
-                          mean_fpkm][1]
-    if (units || complete)
-        fpkm <- formatC(fpkm,digits=0,format="f")
-    if (units) {
-        fpkm <- paste0(fpkm,
-                       "\\ \\text{FPKM}")
-    }
-    if (complete) {
-        if (any(grepl("^ENS.+G\\d+$",gene))) {
-            fpkm <- paste0("\\text{",gene,"} = ",
-                           fpkm)
-        } else {
-            fpkm <- paste0("\\text{\\textit{",gene,"}} = ",
-                           fpkm)
-        }
-    }
-    if (units || complete) {
-        fpkm <- paste0("$",fpkm,"$")
-    }
-    return(fpkm)
-}
-
-@
-
-<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("placenta_lineage_specific")))>>=
-load("../data/placenta_lineage_specific")
-source("/home/don/projects/xtable/pkg/R/print.xtable.R")
-
-placenta.lineage.specific.table <- function(analysis,output=TRUE) {
-    temp.table <- 
-        head(placenta.lineage.specific.analysis[analysis.name==analysis &
-                                                p.value <= 0.05][order(p.value),
-                                                                 list(human_name,p.value,fdr,fc),],
-             n=5)
-    temp.table <- data.table(temp.table)
-    setkey(temp.table,"human_name")
-    temp.table <- 
-        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
-                                            `Approved Name`,p.value,fdr,fc)]
-    setnames(temp.table,
-              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
-              )
-    setkey(temp.table,"Human Symbol")
-    if (output) {
-        display.table <- temp.table[order(`p (t-test)`),]
-        display.table <-
-            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
-                                                  "}")]
-        print(xtable(display.table,
-                     display=c("s","s","s","e","g","f"),
-                     ),
-              booktabs=TRUE,
-              floating=FALSE,
-              math.style.exponents=TRUE,
-              sanitize.text.function=genes.table.sanitize,
-              include.rownames=FALSE)
-
-    }
-    return(invisible(temp.table[order(`p (t-test)`),]))
-}
-output.top.genes.table <- function(data,output=TRUE) {
-    if (output) {
-        table.display <- c("s","s","s","d")
-        table.align <- c("c","c","p{3in}","r")
-        if (any(colnames(data) %in% "SEM"))  {
-            table.display <- c(table.display,"d")
-            table.align <- c(table.align,"r")
-        }
-        if (any(colnames(data) %in% "Human Symbol"))  {
-            table.align <- c("c",table.align)
-            table.display <- c("s",table.display)
-        }
-        print(xtable(data,
-                     display=table.display,
-                     align=table.align
-                     ),
-              booktabs=TRUE,
-              floating=FALSE,
-              sanitize.text.function=genes.table.sanitize,
-              include.rownames=FALSE)
-    }
-}
-
-@
-
-\subsection{Primate Placenta Transcriptomes}
-
-\begin{table}
-  \footnotesize
-<<primate.lineage.specific,echo=FALSE,results="asis">>=
-pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
-setkey(pri.all.lin.spec,"Human Symbol")
-@ 
-\caption{Genes whose expression changed significantly in the primate
-  order as compared to all other analyzed species.}
-\label{tab:primate_lineage_specific}
-\end{table}
-
-We examined the expression of three primates, \human, \bonobo, and
-\smonkey. The genes whose expressions are different on the primate
-lineage in comparison to all other analyzed species are listed in
-\cref{tab:primate_lineage_specific}. \textit{FSTL1} has significantly
-higher expression in the primate lineage
-(Supplemental \cref{fig:fstl1_evolution_tree},
-$\mathrm{FDR}=\Sexpr{pri.all.lin.spec["FSTL1",`FDR`]}$), is implicated
-in BMP signaling, and has previously been implicated in inflammation
-and invasiveness
-\cite{Mouillet.ea2015:expressionandpost-transcriptionalregulation}. Of
-non-primates, \horse shows the highest expression
-%
-\Sexpr{gene.expression(species.name="equus caballus",gene="FSTL1",complete=TRUE)}, 
-%
-which is still lower than the lowest primate, \smonkey
-%
-(\Sexpr{gene.expression(species.name="ateles fusciceps",gene="FSTL1",complete=TRUE)}). 
-%
-No other genes were significantly differentially expressed on this
-lineage by t-test.
-
-\subsubsection{\human Placenta Transcriptome}
-\label{sec:human_placenta}
-
-\begin{table}
-  \footnotesize
-<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="homo sapiens"))
-@ 
-\caption{The top 50 genes by \ac{FPKM} in \human which are not
-  annotated as Mt\_rRNA, Mt\_tRNA, rRNA or a housekeeping gene. Gene
-  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
-  name corresponding to the human symbol. A complete list of all genes
-  is in \cref{tab:all_homo_sapiens_placenta}.}
-\label{tab:top50_human_placenta}
-\end{table}
-
-<<cgb.min.expression,results="hide",echo=FALSE,warnings=FALSE>>=
-cgb.min.expression <- 
-    min(sapply(paste0("CGB",c("","1","2","5","7","8")),
-               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
-cgb.sum.expression <- 
-    sum(sapply(paste0("CGB",c("","1","2","5","7","8")),
-               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
-@
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \human are shown in
-\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
-in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
-the highest \ac{FPKM} value
-%
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
-%
-in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
-three of which are members of a distinct expansion of the growth
-hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
-hormone gene family, \ac{GH2} (\ac{FPKM}=%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
-is also expressed to a high degree in the \human placenta
-(\cref{tab:top50_human_placenta}).
-% 
-Eleven \acp{PSG} are highly-expressed
-(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
-greatest degree of expression %
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
-and the pseudogene \textit{PSG10P} showing the least %
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
-These
-genes are members of a hominoid-specific expansion of the \ac{CEA} family
-and known to be expressed by the syncytiotrophoblast
-\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
-%
-Genes involved in the production of estrogens (%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
-% 
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
-% 
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
-% 
-) and progesterone (%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
-\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
-\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
-%
-show exceptionally high expression in the human placenta, as well as %
-%
-\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
-%
-which generates the alpha subnit of multiple human chorionic
-gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
-producing the beta subunits of the chorionic gonadotropins
-(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
-\textit{CGB7}, \textit{CGB8}) have reasonable expression
-($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
-expression is on the order of the expression of expression of
-\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
-%
-%
-Other highly expressed genes in the top 50 include members of the
-hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
-\textit{HBA2}), the maternally imprinted long noncoding RNA gene
-\textit{H19}, genes coding for proteins required for growth and/or
-cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
-\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
-glucose transport (\textit{SLC2A1}) among others
-(\cref{tab:top50_human_placenta}, Supplemental
-\cref{tab:all_homo_sapiens_placenta}).
-
-\begin{figure}
-<<psg.ctsl.evolution.tree,echo=FALSE,fig.width=16,fig.height=16,warning=FALSE,message=FALSE>>=
-plot.label.size <- 3
-axis.text.size <- 3.5
-
-pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2),clip="off"))
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"),
-                 axis.text.size=axis.text.size,
-                 fontsize=plot.label.size,
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"),
-                 fontsize=4,
-                 width=3,
-                 axis.text.size=axis.text.size,
-                 offset.ratio=1
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"),
-                 fontsize=4,
-                 width=3,
-                 axis.text.size=axis.text.size,
-                 offset.ratio=1.2
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 subtree=c("ENSSSCG00000006792",
-                           "ENSBTAG00000046085"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"),
-                 axis.text.size=axis.text.size,
-                 fontsize=plot.label.size,
-                 offset.ratio=0.6
-                 )
-
-popViewport(1)
-grid.text(c("A","B","C","D"),
-          x=unit(c(0,0.5,0,0.5),"npc"),y=unit(c(1,1,0.5,0.5),"npc"),
-          just=c("left","top"),
-          gp=gpar(fontsize="48")
-          )
-@ 
-\caption{\genetreeexplanation{(A) \textit{PSG1}, (B) \textit{CTSL}, (C) \textit{PRSS16}, and (D) \textit{PAG2}}}
-\label{fig:pag2_prss16_ctsl_psg_evolution_tree}
-\end{figure}
-
-\subsubsection{\bonobo Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
-top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
-output.top.genes.table(top.bonobo.genes)
-setkey(top.bonobo.genes,"Gene symbol")
-@ 
-\caption{\tableexplanation{\bonobo}{tab:all_pan_paniscus_placenta}}
-\label{tab:top50_bonobo_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
-with a list of all genes in Supplemental
-\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
-placenta differs somewhat from that of \human, as evidenced by the
-makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
- in \bonobo is %
-\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
-%
-which is involved in insulin growth factor signaling and angiogenesis
-during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
-\textit{PRG2} is also expressed at relatively high levels (though
-comparatively much lower) in the human placenta
-(\Sexpr{gene.expression("homo sapiens","PRG2")};
-Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
-related to the chorionic somatomammotropins
-are also expressed at relatively high levels (\textit{GH1} 
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
-\textit{GH1} 
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
-\textit{GH2}
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
-and \textit{GH2}
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
-%
-\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
-Placental expression of the genes involved in estrogen and
-progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
-respectively) show lower expression in \bonobo as compared to \human %
-(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
-\cref{fig:estrogen_evolution_tree}; %
-\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
-\cref{fig:progesterone_evolution_tree} and
-\cref{tab:top50_human_placenta}) although the expression of
-\textit{HSD3B1} is fairly similar to that of \smonkey %
-(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
-In addition, \textit{CGA} is expressed in the bonobo placenta %
-(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
-lower levels than \human as well (\cref{tab:top50_human_placenta}).
-%
-Two Ensembl-annotated beta subunits of the chorionic gonadotropin
-genes are expressed in \bonobo, one related to \textit{CGB} and
-another related to \textit{CGB7}; both are not highly expressed
-(\cref{fig:cga_cgb_evolution_tree}). Among the other top 50 expressed
-genes, there is high expression of keratins (e.g., \textit{KRT19},
-\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
-\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
-Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
-2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
-the hormone adrenomedullin (\textit{ADM}).
-
-\subsubsection{\smonkey Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
-top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
-output.top.genes.table(top.spider.monkey.genes)
-setkey(top.spider.monkey.genes,"Gene symbol")
-@ 
-\caption{\tableexplanation{\smonkey}{tab:all_ateles_fusciceps_placenta}}
-\label{tab:top50_spider_monkey_placenta}
-\end{table}
-
-For the term placenta of \smonkey, assembled contigs from RNA-seq were
-aligned using against \human and the best alignment was chosen as the
-annotation for each assembled contig. In the term placenta of
-\smonkey, the top 50 non-housekeeping genes
-(\cref{tab:top50_spider_monkey_placenta}) include a possible regulator
-of androgen receptor signaling
-\cite{Sampson.ea2012:PAGE4positivityisassociated}, \textit{PAGE4} %
-(\Sexpr{gene.expression("ateles_fusciceps","PAGE4")}), %
-the pregnancy specific glycoprotein \textit{PSG8}
-(\Sexpr{gene.expression("ateles_fusciceps","PSG8")}), %
-only one member of the chorionic somatotropins, \textit{GH1} %
-(\Sexpr{gene.expression("ateles_fusciceps","GH1")}), %
-a few HLA genes (\textit{HLA-A}, \textit{HLA-F},
-\cref{tab:top50_spider_monkey_placenta}), and multiple components of
-annexin complexes
-(\Sexpr{gene.expression("ateles_fusciceps","ANXA2",complete=TRUE)}, %
-\Sexpr{gene.expression("ateles_fusciceps","S100P",complete=TRUE)}, %
-\Sexpr{gene.expression("ateles_fusciceps","S100A11",complete=TRUE)}) %
-which may play a role in angiogenesis, and certainly regulate
-cell-cell interaction in the developing placenta
-\cite{Liu.ea2015:AnnexinA2complexeswith}.
-\Sexpr{gene.expression("ateles_fusciceps","AL365202.1",complete=TRUE)}
-is also highly expressed, has no known function, and is duplicated
-multiple times in \textit{Macaca mulatta} (Ensembl gene tree
-\href{http://useast.ensembl.org/Multi/GeneTree/Image?gt=ENSGT00580000081348}{ENSGT00580000081348}).
-All genes with expression greater than 50 \ac{FPKM} are listed in
-Supplementary \cref{tab:all_ateles_fusciceps_placenta}.
-
-\subsection{Rodentia Placenta Transcriptomes}
-
-We examined the expression of three rodentia, \mouse, \carmeli, and
-\galili{}. We lacked sufficient power to detect any significant
-changes on the Rodentia lineage in comparison to all other analyzed
-species (Supplemental \cref{tab:rodentia_lineage_specific}).
-
-\subsubsection{\mouse Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="mus musculus"))
-@ 
-\caption{\tableexplanation{\mouse}{tab:all_mus_musculus_placenta}}
-\label{tab:top50_mouse_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \mouse are shown in
-\cref{tab:top50_mouse_placenta} with a list of all genes with
-$\mathrm{FPKM} \ge 50$ in Supplemental
-\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
-\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
-target \textit{in silico} is the orphan G-coupled receptor
-\textit{Mrgprg}
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
-\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
-\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
-of which are maternally imprinted (the later of which is a homolog of
-human \textit{H19\_2}),
-\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
-cathepsins and a rodentia specific expansion of prolactin related
-genes. There are also two miRNA which overlap with the of the 7S
-signal recognition particle
-(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
-and
-\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
-and numerous pseudogenes
-(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
-appear to be highly expressed, however, the short read length (34--44)
-in many samples is likely preventing unambiguous assignment of all
-reads to the more lengthy complete transcript, so these shorter
-transcripts should be treated with suspicion, and we will ignore them
-in further discussion. The prolactin-related genes
-(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
-\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
-very highly expressed in \mouse in, with %
-\Sexpr{combined.fpkm[species=="mus musculus" &
-  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
-prolactin related genes having $\mathrm{FPKM} \ge 1$
-(\cref{fig:prolactin_evolution_tree}, Supplemental
-\cref{fig:prolactin_receptor_evolution_tree}). The cathespins have
-also expanded in rodents, with six showing high expression
-(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
-\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
-\cref{tab:top50_mouse_placenta,fig:pag2_prss16_ctsl_psg_evolution_tree}).
-Finally, \textit{Peg10}, an imprinted gene required for proper
-formation of the placenta in mice is also highly expressed
-(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
-reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.
-
-\subsubsection{\galili Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="spalax galili"))
-@ 
-\caption{\tableexplanation{\galili}{tab:all_spalax_galili_placenta}}
-\label{tab:top50_n_galili_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \galili are shown in
-\cref{tab:top50_n_galili_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_spalax_galili_placenta}.
-Components of annexin complexes are highly expressed
-(\Sexpr{gene.expression("spalax_galili","S100a6",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","S100a11",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","S100a10",complete=TRUE)},
-and
-\Sexpr{gene.expression("spalax_galili","Anxa2",complete=TRUE)}),
-as are members of the thymosin family
-(\Sexpr{gene.expression("spalax_galili","Tmsb4x",complete=TRUE)}
-and
-\Sexpr{gene.expression("spalax_galili","Tmsb10",complete=TRUE)}),
-various cytoskeleton related proteins
-(\Sexpr{gene.expression("spalax_galili","Actc1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Actg2",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Acta2",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Myl9",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Vim",complete=TRUE)}), and
-transmembrane protein of currently unknown function, 
-\Sexpr{gene.expression("spalax_galili","Gm10134",complete=TRUE)}.
-
-\subsubsection{\carmeli Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="spalax carmeli"))
-@ 
-\caption{\tableexplanation{\carmeli}{tab:all_spalax_carmeli_placenta}}
-\label{tab:top50_s_carmeli_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \carmeli are shown in
-\cref{tab:top50_s_carmeli_placenta} with a list of all highly
-expressed genes in Supplemental
-\cref{tab:all_spalax_carmeli_placenta}. Numerous hemoglobin subunits
-ware highly expressed
-(\Sexpr{gene.expression("spalax_carmeli","Hbb-bh1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Hba-a1",complete=TRUE)}, and
-\Sexpr{gene.expression("spalax_carmeli","Hbb-bs",complete=TRUE)}), as
-well as the relaxin
-\Sexpr{gene.expression("spalax_carmeli","Rln1",complete=TRUE)},
-multiple cathespins
-(\Sexpr{gene.expression("spalax_carmeli","Ctsm",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","BC051665",complete=TRUE)},
-and \Sexpr{gene.expression("spalax_carmeli","Ctsl",complete=TRUE)},
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), and many prolactin-related genes
-(\Sexpr{gene.expression("spalax_carmeli","Prl4a1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Prl6a1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Prl3b1",complete=TRUE)}, and
-\Sexpr{gene.expression("spalax_carmeli","Prl7c1",complete=TRUE)},
-\cref{fig:prolactin_evolution_tree}) which shows a similar pattern of
-expansion seen in \mouse.
-
-\subsection{Laurasiatheria}
-
-We examined the expression of five Laurasiatherians: \cow, \sheep,
-\pig, \horse, and \dog. We were unable to detect any significant
-differences after correcting for multiple testing between
-Laurasiatherians and all other analyzed species
-(Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
-Laurasiatherians and Euarchontoglires
-(Supplemental \cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).
-
-\subsubsection{\dog Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="canis familiaris"))
-@ 
-\caption{\tableexplanation{\dog}{tab:all_canis_familiaris_placenta}}
-\label{tab:top50_dog_placenta}
-\end{table}
-
-The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
-in the term placenta of \dog are shown in
-\cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
-\cref{tab:all_canis_familiaris_placenta}. The gene with the highest
-expression %
-(\Sexpr{gene.expression("canis familiaris","TMSB10")}) is the thymosin 
-\textit{TMSB10} %
-(\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
-is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
-among the top 50 (\cref{tab:top50_dog_placenta}) include genes
-involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
-anticoagulation (\textit{TFPI2}, \textit{PLA2G7}), innate immunity
-(\textit{LCN2}, \textit{GM2A}), lipid metabolism (\textit{APOE}),
-inhibition of matrix metalloproteinases (\textit{TIMP2},
-\textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
-\textit{HTRA1}). The gene \textit{LGALS1}, which shows evidence of
-immunosupression during pregnancy
-\cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
-(\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
-In addition, an ortholog of \textit{SSLP1} (ENSCAFG00000010638) a gene
-expressed in the placentas of \cow (%
-\Sexpr{gene.expression("bos taurus","SSLP1",complete=TRUE)}, %
-\cref{tab:top50_cow_placenta}) and \sheep
-(\Sexpr{gene.expression("ovis aries","SOLD1",complete=TRUE)},
-\cref{tab:top50_sheep_placenta}) but is not present in primates
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}
-(\cref{fig:sslp1_evolution_tree}), is also expressed at high levels
-%
-(\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
-There is no placental expression of the estrogen synthase gene
-(\textit{CYP19A1}) in \dog 
-(\Sexpr{gene.expression("canis familiaris","CYP19A1")}
-\cref{fig:estrogen_evolution_tree}), 
-and expression of \textit{HSD3B7} is also low
-(\Sexpr{gene.expression("canis familiaris","HSD3B7")}, 
-\cref{fig:progesterone_evolution_tree}).
-
-\subsubsection{\cow Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="bos taurus"))
-@ 
-\caption{\tableexplanation{\cow}{tab:all_bos_taurus_placenta}}
-\label{tab:top50_cow_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
-list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
-The gene with the highest expression was \textit{TMSB10} %
-(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
-\textit{TMSB4} was also highly expressed %
-(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
-\dog{}, members of annexin complexes were also highly expressed
-(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
-at least one of a set of four miRNA with 100\% identity
-(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
-%
-As these are novel miRNA, there is a possibility that the assignment
-of the short reads to these four miRNA is due to miss-annotation of
-the \cow genome.
-% 
-Several members of the ruminant-specific pregnancy-associated
-glycoprotein family show especially high expression in \cow placenta
-(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D, and
-Supplemental \cref{tab:all_bos_taurus_placenta}) including
-%
-\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
-%
-28 members of this family have $FPKM \ge 1$, and this family of
-genes has greatly expanded in ruminants (\cow and \sheep{}) although
-the expression levels of this family are higher in \cow than in \sheep
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D. Other genes with high expression
-include genes involved with regulation of growth (\textit{IGF2},
-\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
-and \textit{CSH1}, a gene that is a member of a ruminant specific
-expansion of placental prolactins and not directly related to the gene
-of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
-ruminant-specific placental prolactins also show high expression in
-the cow placenta including %
-\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
-with a total of 14 with $\mathrm{FPKM} \ge 1$.
-%
-\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
-cotyledonary villi of the ruminiant placenta and involved in cell
-proliferation
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
-shows evidence of high expression %
-(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
-There is some placental expression of the estrogen synthase gene
-(\textit{CYP19A1}) in cows %
-(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
-comparatively greater expression of \textit{HSD3B} %
-(\Sexpr{gene.expression("bos taurus","HSD3B")}).
-
-\subsubsection{\pig Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="sus scrofa"))
-@ 
-\caption{\tableexplanation{\pig}{tab:all_sus_scrofa_placenta}}
-\label{tab:top50_pig_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Sus scrofa} are shown in
-\cref{tab:top50_pig_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_sus_scrofa_placenta}. The top two
-expressed transcripts,
-\Sexpr{gene.expression("sus_scrofa","ENSSSCG00000025162",complete=TRUE)}
-and
-\Sexpr{gene.expression("sus_scrofa","ENSSSCG00000021356",complete=TRUE)},
-are miRNA which correspond to the \cow miRNA
-\href{http://www.mirbase.org/cgi-bin/mirna_entry.pl?acc=MI0013080}{bta-mir-2904-1}
-and
-\href{http://www.mirbase.org/cgi-bin/mirna_entry.pl?acc=MIMAT0013845}{bta-mir-2887-1},
-respectively, both of which are also expressed in \cow ovary
-\cite{Hossain.ea2009:Identificationandcharacterizationof}, although
-neither of which have an obvious target gene on the basis of
-\textit{in silico} analysis
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}. Other highly
-expressed genes include members of the insulin growth factor family
-(\Sexpr{gene.expression("sus_scrofa","IGF2",complete=TRUE)} and
-\Sexpr{gene.expression("sus_scrofa","IGFBP3",complete=TRUE)}), members
-of annexin complexes
-(\Sexpr{gene.expression("sus_scrofa","S100A1",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","ANXA2",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","S100A12",complete=TRUE)}, and
-\Sexpr{gene.expression("sus_scrofa","S100A10",complete=TRUE)}), the
-galectin \Sexpr{gene.expression("sus_scrofa","LGALS3",complete=TRUE)}
-and multiple cytoskeleton and extracellular matrix proteins
-(\Sexpr{gene.expression("sus_scrofa","KRT19",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","COL1A1",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","BETA-ACTIN",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","KRT9",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","CLDN4",complete=TRUE)}, and
-\Sexpr{gene.expression("sus_scrofa","KRT9",complete=TRUE)}).
-
-\subsubsection{\horse Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="equus caballus"))
-@ 
-\caption{\tableexplanation{\horse}{tab:all_equus_caballus_placenta}}
-\label{tab:top50_horse_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Equus Caballus} are shown in
-\cref{tab:top50_horse_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_equus_caballus_placenta}. The gene
-with the highest expression was the relaxin,
-\Sexpr{gene.expression("equus_caballus","RLN",complete=TRUE)},
-followed by two miRNA
-(\Sexpr{gene.expression("equus_caballus","ENSECAG00000025328",complete=TRUE)}
-and
-\Sexpr{gene.expression("equus_caballus","ENSECAG00000027056",complete=TRUE)})
-with no known \textit{in silico} targets
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}, the thymosin
-\Sexpr{gene.expression("equus_caballus","TMSB4",complete=TRUE)},
-members of the insulin growth factor family
-(\Sexpr{gene.expression("equus_caballus","IGF2",complete=TRUE)} and
-\Sexpr{gene.expression("equus_caballus","IGFBP3",complete=TRUE)}) and
-multiple members of annexin complexes
-(\Sexpr{gene.expression("equus_caballus","S100A6",complete=TRUE)},
-\Sexpr{gene.expression("equus_caballus","S100A1",complete=TRUE)}, and
-\Sexpr{gene.expression("equus_caballus","ANXA2",complete=TRUE)}),
-similar to the pattern of expression seen in \pig and \dog.
-
-\subsubsection{\sheep Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
-top.sheep.genes <- top.genes.table(species.name="ovis aries",n=200)
-top.sheep.genes <-
-    top.sheep.genes[!grepl("\\{(7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\}",
-                           `Gene symbol`) &
-                    !grepl("\\((7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\)",
-                           `Human Symbol`)
-                   ,][1:50,]
-output.top.genes.table(top.sheep.genes)
-@ 
-\caption{\tableexplanation{\sheep}{tab:all_ovis_aries_placenta}}
-\label{tab:top50_sheep_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \sheep are shown in
-\cref{tab:top50_sheep_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_ovis_aries_placenta}. The complete
-list of highly expressed genes includes many snRNAs and rRNAs likely
-due to the RNA extraction and library preparation procedure utilized
-for this sample. Ignoring these, notable highly expressed genes
-include members of annexin complexes
-(\Sexpr{gene.expression("ovis_aries","S100G",complete=TRUE)},
-\Sexpr{gene.expression("ovis_aries","ANXA2",complete=TRUE)}),
-hemoglobin subunits
-(\Sexpr{gene.expression("ovis_aries","ENSOARG00000010944",complete=TRUE)}),
-an ortholog of \textit{SSLP1},
-\Sexpr{gene.expression("ovis_aries","SORD1",complete=TRUE)}, two
-cathespins (\Sexpr{gene.expression("ovis_aries","CTSL",complete=TRUE)}
-and \Sexpr{gene.expression("ovis_aries","CTSB",complete=TRUE)}), and
-placentally expressed transcript 1
-\Sexpr{gene.expression("ovis_aries","PLET1",complete=TRUE)} which may
-play a role in cellular migration
-\cite{Raymond.ea2010:Expressionoforphanprotein}.
-
-\subsection{Atlantogenata (Afrotheria + Xenartha)}
-\subsubsection{\elephant Placenta Transcriptome}
-
-\begin{table}
-\footnotesize
-<<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="loxodonta africana"))
-@ 
-\caption{\tableexplanation{\elephant}{tab:top50_elephant_placenta}}
-\label{tab:top50_elephant_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \elephant are shown in \cref{tab:top50_elephant_placenta}
-with a list of all genes in
-\cref{tab:all_loxodonta_africana_placenta}. The gene with the highest
-expression was the natriuretic peptide precursor B
-(\Sexpr{gene.expression("loxodonta_africana","NPPB",complete=TRUE)}).
-Two thymosins,
-\Sexpr{gene.expression("loxodonta_africana","TMSB10",complete=TRUE)}
-and
-\Sexpr{gene.expression("loxodonta_africana","TMSB4Y",complete=TRUE)}
-were also also was highly expressed. The miRNA mir-675
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027406",complete=TRUE)}).
-which is imbedded in the first exon of the maternally imprinted long
-non-coding RNA gene \textit{H19}
-\cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene which is
-currently un-annotated in \elephant{}. %
-% \DLA{Probably consider fixing the unannotation of H19; it seems pretty
-%   obvious to me, though it would require fixing the GTF file for
-%   elephant with the right
-%   \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan
-%     prediction} for this region} %
-Five uncharacterized genes are expressed to a high degree in the
-\elephant placenta (\cref{tab:top50_elephant_placenta}), three of
-which
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000028493",complete=TRUE)},
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032011",complete=TRUE)},
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000031933",complete=TRUE)})
-are members of an \elephant{}-specific expansion of 11 genes closely
-related to the thymus-specific serine protease gene \textit{PRSS16}
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C The other 8
-\textit{PRSS16}-like genes are also expressed to a lesser degree
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000026505",complete=TRUE)}).
-Two other novel genes with high expression are related to the gene
-\textit{SSLP1}
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032659",complete=TRUE)}
-and
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000029551",complete=TRUE)},
-\cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
-\dog placentas (\cref{tab:top50_dog_placenta,tab:top50_cow_placenta}).
-Placental expression of the hormones relaxin (\textit{RLN}, %
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000036090")})
-and prolactin (\textit{PRL}) are exceptionally high in the elephant
-when compared to other mammals
-(\cref{fig:rln2_evolution_tree,fig:prolactin_evolution_tree}). There
-is little to no placental expression of the estrogen synthase gene,
-\textit{CYP19A1}
-(\Sexpr{gene.expression("loxodonta_africana","CYP19A1")}) and the
-enzyme that produces progesterone HSD3B1
-(\Sexpr{gene.expression("loxodonta_africana","HSD3B")}). Other genes
-with high expression in the elephant placenta include genes involved
-in oxidative stress
-(\Sexpr{gene.expression("loxodonta_africana","TXN",complete=TRUE)}),
-growth
-(\Sexpr{gene.expression("loxodonta_africana","IGF2",complete=TRUE)})
-and
-\Sexpr{gene.expression("loxodonta_africana","PAEP",complete=TRUE)}, a
-gene predominantly expressed in the endometrium of the uterus in humans
-\cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.
-
-\subsubsection{\armadillo Placenta Transcriptome}
-
-\begin{table}
-\footnotesize
-<<top.50.armadillo.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="dasypus novemcinctus"))
-@ 
-\caption{\tableexplanation{\armadillo}{tab:all_dasypus_novemcinctus_placenta}}
-\label{tab:top50_armadillo_placenta}
-\end{table}
-
-The top 50 expressed genes in the term placenta of \armadillo are
-shown in \cref{tab:top50_armadillo_placenta} with a list of all genes
-with $\mathrm{FPKM}\ge 50$ in Supplemental
-\cref{tab:all_dasypus_novemcinctus_placenta}. The gene with the
-highest expression is
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045029",complete=TRUE)}
-which is most similar to the \mouse \textit{Pga5} is expanded in
-\opossum into three members which are expressed in placenta (the other
-two being
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033885",complete=TRUE)}
-and
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000040431",complete=TRUE)}),
-though the expansion is less than that seen \cow and \sheep
-(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). There is also high expression of
-hemoglobin subunit beta,
-\Sexpr{gene.expression("dasypus_novemcinctus","HBB",complete=TRUE)},
-an un-annotated gene with high similarity to \human and \cow
-\textit{OVOS2}
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033320",complete=TRUE)}
-(Supplemental \cref{fig:ovos2_evolution_tree}), the growth factors
-\Sexpr{gene.expression("dasypus_novemcinctus","IGF2",complete=TRUE)}
-and
-\Sexpr{gene.expression("dasypus_novemcinctus","IGFBP4",complete=TRUE)},
-and the thymosin
-\Sexpr{gene.expression("dasypus_novemcinctus","TMSB4X",complete=TRUE)}.
-As has been seen previously, multiple components of the annexin
-complex are also highly expressed, including
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045634",complete=TRUE)}
-(\textit{ANXA8L1}),
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000017399",complete=TRUE)}
-(\textit{S100B}),
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000048504",complete=TRUE)}
-(\textit{S100A10}), and
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000004237",complete=TRUE)}
-(\textit{ANXA2}).
-
-\subsection{Metatherians}
-
-\subsubsection{\opossum Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="monodelphis domestica"))
-@ 
-\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica_placenta}}
-\label{tab:top50_opossum_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the \opossum are shown in
-\cref{tab:top50_opossum_placenta} with a list of all genes in
-Supplemental \cref{tab:all_monodelphis_domestica_placenta}. The gene with the
-second highest expression is superoxide dismutase 3, extracellular
-(\Sexpr{gene.expression("monodelphis_domestica","SOD3",complete=TRUE)}),
-which functions as an antioxidant. The gene with the second highest
-expression is the beta hemoglobin subunit epsilon
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000011325",complete=TRUE)}),
-which is expressed at the embryonic stage of development in
-marsupials, as in other mammals
-\cite{Holland.ea1997:Characterizationofembryonicglobin}. \textit{HBE}
-subunits link up with protein products of two hemoglobin, zeta
-(\textit{HBZ}) genes, which are also expressed to a high degree in the
-opossum placenta
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016635",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016631",complete=TRUE)}).
-In addition, there is high expression of the gene hemoglobin, omega
-(\textit{HBO},
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016637",complete=TRUE)}),
-% this is LOC_1000016060 which I figured out by tblastn'ing AAK11486.1
-% against Monodelphis domestica
-a hemoglobin found only in marsupials and monotremes
-\cite{Opazo.ea2008:Genomicevidenceforindependent} (Supplemental
-\cref{fig:hbe1_evolution_tree}). The hemoglobin gene \textit{HBA} is
-also expressed in the placenta
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016633",complete=TRUE)}).
-Two cystatin genes
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018295",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018296",complete=TRUE)})
-are expressed to a high degree, along with one \textit{CTSL} related
-gene
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000002844",complete=TRUE)},
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), which are known to interact with
-cystatins \cite{Turk.ea2001:Lysosomalcysteineproteasesfacts}. The
-uncharacterized gene \textit{LOC10001195}
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
-also shows high expression in \opossum placenta. The gene
-\textit{WFDC2}, a member of the whey acidic protein (WAP) family is
-also expressed to a high degree in \opossum placenta
-(\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).
-Other genes with high expression include members of the annexin
-complex
-(\Sexpr{gene.expression("monodelphis_domestica","S100A6",complete=TRUE)},
-\Sexpr{gene.expression("monodelphis_domestica","ANXA2",complete=TRUE)}).
-In addition, there was little production of CYP19A1 and HSD3B1
-(\Sexpr{gene.expression("monodelphis_domestica","CYP19A1",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000005385",complete=TRUE)},
-respectively).
-
-\subsection{Core Placenta Transcriptome}
-
-\begin{figure}
-<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
-print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
-             aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
-      + geom_raster(aes(fill=log10(FPKM)))
-      + scale_x_discrete(expand=c(0,0))
-      + scale_y_discrete(expand=c(0,0))
-      + scale_fill_gradient(low = grad.def[["low"]],
-                            high = grad.def[["high"]],
-                            space = grad.def[["space"]],
-                            na.value = grad.def[["na.value"]],
-                            limits=c(0,4),
-                            guide=guide_colorbar(title=expression(log[10](FPKM)))
-                            )
-      + ylab("Human Ortholog Name")
-      + xlab("Species")
-      + theme_grey()
-      + theme(legend.position="top",
-              axis.ticks = element_blank(),
-              axis.text.y = element_text(face="italic"),
-              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
-      )
-@ 
-\caption{Core Placenta Transcriptome.
-  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-  genes expressed (\ac{FPKM}$ \ge 10$) in the placenta of all organisms
-  studied with housekeeping
-  genes \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
-  removed.}
-\label{fig:core_placenta_transcriptome}
-\end{figure}
-
-The
-\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-genes comprising the core placenta transcriptome are shown in
-\cref{fig:core_placenta_transcriptome} ordered by their median
-transcription level in \ac{FPKM}. We considered a gene to be core to
-the placenta transcriptome if a 1:1 ortholog of that gene was
-expressed with $\mathrm{FPKM} \ge 10$ in all species studied and was
-not a human housekeeping gene according to
-\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
-Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
-which would be included in the core placenta set if the \ac{FPKM}
-cutoff was set differently.
-%
-Multiple components of annexin complexes are included in the core
-transcriptome, including \textit{ANXA2}, \textit{ANXA1},
-\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
-resolution of the inflammation response through inhibition of
-phospholipase A2 activity and signaling through the formyl peptide
-receptor family, and has previously been hypothesized to be important
-for the maintenance of the anti-inflammatory during pregnancy
-\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
-known to interact with S100A10, S100A6, S100A11, and S100A4
-\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
-its interaction with S100P as well; it is involved in cell-cell
-interaction as well as vesicle trafficking and von Willibrand Factor
-secretion. The expression regulation of another core gene,
-\textit{TIMP3}, has been implicated in pre-eclampsia
-\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
-Supplemental \cref{tab:core_placenta_go_terms} lists the significant enrichment of
-\ac{GO} terms in the core placenta transcriptome; notable terms
-include those involved in cytoskeletal structure, protein transport,
-membrane protein structure, and inflammation resolution.
-
-% SPINT2 -- Serine Peptidase inhibitor
-
-% SLC3A2 -- this is probably actually a housekeeping gene; it's
-% involved in amino acid transport
-
-% SQSTM1 -- NFκB signaling (immune supression?)
-
-% KRTCAP2  N-glycosylation (probably housekeeping)
-
-% NRDG1 -- interacts with HSP90
-
-% SDC4 -- proteoglycan (syndecan) cell adhesion
-
-% LAMC1 -- laminin
-
-% LUM -- proteoglycan
-
-% VDAC1 -- mitochondrial gene?
-
-% F11R -- tight junctions
-
-% SERPINH1? with TIMP3 
-% http://www.proteinatlas.org/ENSG00000149257-SERPINH1/tissue/placenta
-
-\subsubsection{\ac{GSE} Analyses of Individual Transcriptomes}
-
-In order to identify major functional categories of genes expressed in
-the individual transcriptomes, we examined \ac{GSE} in 1:1 human
-orthologs which had \ac{FPKM} values in the top 1\% of expressed genes
-(Supplemental %
-\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris}) using topGO \cite{topGO:2.20.0}.
-%
-%
-Significant enrichment in genes ($\mathrm{FDR}\le0.05$) in the top 1\%
-of \ac{FPKM} values in at least 12 of the
-\Sexpr{length(species.names)} species included clusters such as
-membrane organization , locomotion, tissue development, extracellular
-matrix organization, regulation of peptidase activity, and various
-cell death related pathways among others
-(Supplemental \cref{tab:go_terms_all_species}). If we instead examine the top 100
-genes by \ac{FPKM} in each of the \Sexpr{length(species.names)}
-species (Supplemental %
-\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris})
-notable enrichments include extracellular matrix, female pregnancy,
-integrin and collagen-related processes, and various cell death
-related processes.
-
-% \textit{Ateles fuciceps} shows enrichment in extracellular
-% matrix and structural molecular activity (Supplemental
-% \cref{tab:go_terms_enriched_ateles_fusciceps}). \textit{Homo sapiens}
-% placentas are enriched in female pregnancy, estrogen biosynthesis,
-% reproductive process, hormone aactivity, oxygen transporter activity,
-% and haptoglobin binding (Supplemental
-% \cref{tab:go_terms_enriched_homo_sapiens}). \textit{Pan paniscus}
-% placentas are enriched in clusters related to stress, apoptosis, and
-% response to abiotic stimuli. \textit{Mus musculus} placentas are
-% enriched in cell junction, vesicle transport, in utero embryonic
-% development, and many cytoskeletal processes. \textit{Spalax
-%   galili} placentas are enriched in extracellular matrix organization,
-% integrin binding, and cell death.
-% 
-% Cow shows only four
-% significantly enriched clusters; however many genes highly expressed
-% in the cow are ruminant-specific placenta genes and are not yet
-% annotated to specific GO categories. Dog shows enrichment in genes
-% associated with RNA, as well as homeostasis, glycoproteins, lyosome,
-% and calcium binding. Elephant shows enrichment in genes associated
-% with RNA, coagulation and calcium binding. Opossum shows enrichment in
-% genes involved in apoptosis, oxidative stress, mitochondrion, hormone
-% metabolism, and lipids.
-
-% Not sure if this is actually worthwhile; I've mentioned some of this
-% already for some of the lineages
-
-% \subsection{Comparative Analysis of Gene Expression}
-% 
-% We performed a comparative analysis of gene expression by
-% reconstructing the evolution of expression values and detecting
-% significant shifts in expression along specific lineages. Likelihood
-% values for each model for each of the X genes reconstructed, as well
-% as likelihood ratio tests, P- and Q-values (after correcting for
-% multiple testing) are shown in Supplemental Table X. We identified
-% between 4 (Boreoeutheria) and 550 (Mouse) significant shifts (Q <
-% 0.01) in expression on specific lineages (Table 10). The top 10 genes
-% with significant shifts (if applicable) per lineage are shown in Table
-% 10 and all genes for each branch are shown in Supplemental Tables
-% X-XX. Genes that showed shifts on the human lineage were
-% overwhelmingly identified as placenta-specific by the
-% tissue-specificity index (TSI), with 56 out of 279 genes (20\%) with
-% both TSI>0.75 and placenta as one the top two tissues with the highest
-% expression. Unsurpisingly, this was not the case with other lineages,
-% since we used expression values from a human-derived microarray to
-% generate TSI values for each gene. However, this does underscore the
-% dynamic nature of expression patterns in the placenta and indeed all
-% tissues across mammals.
-% 
-% We then performed DAVID functional annotation clustering analyses on
-% each list of genes from each lineage. The top three clusters for each
-% lineage are shown in Table 10 and a full list of clusters in
-% Supplemental Table X. The top enrichment cluster in the human lineage
-% is blood vessel development, followed by extracellular matrix and
-% metal ion binding (Table 10). Other significant clusters include those
-% related to cell communication/signaling, cytokines, defense response,
-% tryptophan metabolism, apoptosis, and skeletal system development
-% (Supplemental Table X). The bonobo lineage was enriched in only three
-% clusters: apoptosis, Toll-like receptor signalling and apoptosis. The
-% mouse displayed enrichment in clusters involved in epigenetic
-% modifications such as methyltransferase activity and histone
-% modification, as well embryonic morphogeneisis, lipid biosynthesis,
-% among others (Supplemental Table X). The top enriched cluster in the
-% cow lineage was the response to vitamin A, as well as many clusters
-% related to the immune system. The dog showed enrichment in skeletal
-% system development, as well as lipid metabolic processes and blood
-% circulation. The elephant lineage had only one enriched cluster:
-% carbohydrate binding. Genes that showed enrichment on the opossum
-% lineage (essentially representing the difference between placental
-% mammals and marsupials) were clustered in many categories not seen in
-% the other lineages, including clusters related to the mitochondria,
-% urea cycle, peroxisome, glycolysis, and transport across the cell
-% membrane (Table 10 and Supplemental Table X).
-
-\clearpage
-
-\section{Discussion}
-
-\subsection{Evolution of Placental Hormones}
-
-Reproductive hormones are central to the regulation of pregnancy and
-their source, timing, and quantity can vary widely from species to
-species \cite{Carter2012:Evolutionofplacentalfunction}. Expression and synthesis of hormones is
-perhaps the most well-documented physiological process in the placenta
-in many species. Here we quantified the extreme disparity in genes
-with hormone products or enzymatic products that directly produce
-hormones in multiple mammalian species (\cref{fig:hormone_genes}).
-
-\begin{figure}
-<<hormone.genes.plot,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-gene.expression.nozero <- function(genes) {
-    temp <-
-        data.matrix.dt(dcast(combined.fpkm[human_name %in% genes,],
-                             human_name~species,
-                             value.var="FPKM",
-                             fun.aggregate=function(x){x[1]}))
-    temp <-
-        temp[apply(temp,1,function(x){sum(x,na.rm=TRUE)})>5,]
-    return(temp)
-}
-
-### hormone genes
-hormone.genes <- c("CGA","CGB","CGB1","CGB2","CGB5","CGB8","CSH1","CSH2","CSHL1","CYP19A1",
-                   "GH2","GHRH","HSD3B1","HSD3B2","HSD17B2","HSD17B1","NPPB","NPPC","PRL","RLN1")
-hormone.expression.nozero <-
-    gene.expression.nozero(hormone.genes)
-
-hor.exp.long <- melt(log10(hormone.expression.nozero+1)[rev(1:nrow(hormone.expression.nozero)),])
-colnames(hor.exp.long)[1:3] <- c("gene","species","fpkm")
-hor.exp.long <- hor.exp.long[hor.exp.long$gene!="IGF2",]
-
-### galectin genes
-galectin.genes <- unique(grep("^LGALS",combined.fpkm[,human_name],value=TRUE))
-
-galectin.expression.nozero <-
-    gene.expression.nozero(galectin.genes)
-
-gal.exp.long <- melt(log10(galectin.expression.nozero+1)[rev(1:nrow(galectin.expression.nozero)),])
-colnames(gal.exp.long)[1:3] <- c("gene","species","fpkm")
-
-### IGF related genes
-
-igfrelated.genes <- unique(grep("^IGF",combined.fpkm[,human_name],value=TRUE))
-
-igfrelated.expression.nozero <-
-    gene.expression.nozero(igfrelated.genes)
-
-igf.exp.long <- melt(log10(igfrelated.expression.nozero+1)[rev(1:nrow(igfrelated.expression.nozero)),])
-colnames(igf.exp.long)[1:3] <- c("gene","species","fpkm")
-
-igf.exp.long$type <- "IGF Related"
-gal.exp.long$type <- "Galectin"
-hor.exp.long$type <- "Hormone"
-
-### scalerange <- range(c(igf.exp.long$fpkm,gal.exp.long$fpkm,hor.exp.long$fpkm),na.rm=TRUE)
-### gradientends <- scalerange + rep(c(0,100,200), each=2)
-### colorends <- c("white", "red", "white", "green", "white", "blue")
-### 
-### gal.exp.long$fpkm <- gal.exp.long$fpkm+100
-### hor.exp.long$fpkm <- hor.exp.long$fpkm+200
-
-combined.long <-
-    rbind(igf.exp.long,
-          gal.exp.long,
-          hor.exp.long)
-combined.long$species <-
-    factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(combined.long$species)),
-           levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(names(species.names))))
-
-common.plot.options <- 
-    list(geom_tile(aes(fill = fpkm), colour = "white"),
-##         scale_fill_gradientn(colours = colorends, values = rescale(gradientends)),
-         scale_fill_gradient(low = grad.def[["low"]],
-                             high = grad.def[["high"]],
-                             space = grad.def[["space"]],
-                             na.value = grad.def[["na.value"]],
-                             guide=guide_colorbar(title=expression(log[10](FPKM)))),
-         scale_x_discrete("", expand = c(0, 0)),
-         scale_y_discrete("", expand = c(0, 0)),
-         theme_grey(base_size = 9),
-         theme(legend.position = "top",
-               axis.ticks = element_blank(),
-               panel.background = element_blank(),
-               axis.text.y = element_text(face="italic"),
-               axis.text.x = element_text(angle = 300, hjust = 0, vjust=1,face="italic")),
-         facet_grid(type~.,scale="free_y",space="free_y")
-         )
-print(ggplot(combined.long, aes(y=gene, x=species))
-      + common.plot.options
-      )
-@ 
-\caption{Expression of Hormone Genes, Galectins, and IGF related genes}
-\label{fig:hormone_genes}
-\end{figure}
-
-\subsubsection{Estrogen and Progesterone}
-
-In both ruminants and primates, the placenta is capable of producing
-estrogen and progesterone
-\cite{Strauss.ea1996:Placentalsteroidhormonesynthesis}. High relative
-expression of \textit{CYP19A1} (Supplemental
-\cref{fig:estrogen_evolution_tree}), the enzyme that synthesizes
-estrogen from androstenedione is seen in \human, \bonobo, \cow, and
-\horse, though not in \sheep, \smonkey. %
-%
-\textit{HSD3B1}, one of two enzymes that synthesizes progesterone from
-pregnalone
-(\cite{Simard.ea2005:Molecularbiologyof3beta-hydroxysteroid},
-Supplemental \cref{fig:progesterone_evolution_tree}) is also highly
-expressed in \human, \bonobo, \cow, \horse, and \dog, but is not
-highly expressed in \smonkey. The diversity of \textit{CYP19A1} and
-\textit{HSD3B1} expression levels in in primates and ungulates seen
-here points to the need for more complete sampling of placentas at
-different timepoints and in different clades.
-
-\subsubsection{Prolactin}
-
-The elephant is strikingly the only mammal analyzed that shows high
-expression of the primarily pituitary hormone prolactin (\textit{PRL})
-in the placenta, although we see high expression of genes which
-resulted from duplications of \textit{PRL} in the \mouse, \carmeli,
-and \cow (Supplemental \cref{fig:prolactin_evolution_tree}). All afrotherians yet
-analyzed contain only one prolactin gene, and prolactin has undergone
-adaptive evolution on the \elephant lineage
-\cite{Wallis2009:ProlactininAfrotheriacharacterization}.
-\citet{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
-presence of a prolactin in the trophoblast of the elephant placenta
-using immunohistological staining and human derived antibodies. A
-previous analysis of \elephant placenta also detected the expression
-of prolactin \cite{Hou.ea2012:Elephanttranscriptomeprovidesinsights}.
-Here we confirm that the sole prolactin in \elephant is not only
-expressed but highly expressed in the placenta, and likely functions
-both in the maintenance of the corpus luteum and as a placental
-lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}.
-\textit{PRL} is not expressed in \opossum, even though there is
-evidence that it is expressed in the placenta of another marsupial,
-\textit{Macropus eugenii} (tammar wallaby)
-\cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests
-that placental prolactin expression in marsupials may be as diverse as
-in eutherian mammals. One possible explanation of the diversity of
-expression is the presence of endogenous retroviruses such as MER39 in
-primates, MER77 in rodents, and L1-2\_LA in \elephant
-\cite{Emera.ea2012:Convergentevolutionofendometrial} although this
-does not explain the diversity of expression of different isoforms of
-\textit{PRL} in \mouse or their tissue specificity in primates. An
-alternative, that \textit{HOXA11} expression is responsible for
-\textit{PRL} expression
-\cite{Lynch.ea2008:Adaptivechangesintranscription} seems not to be the
-case in placenta, as the rodent with the highest \textit{HOXA11}
-expression, \galili
-(\Sexpr{gene.expression("spalax_galili","Hoxa11",complete=TRUE)})
-is the rodent with the lowest expression of \textit{PRL}
-(\cref{fig:prolactin_evolution_tree}), and \elephant also does not
-express \textit{HOXA11} highly either
-(\Sexpr{gene.expression("loxodonta_africana","HOXA11",complete=TRUE)},
-Supplemental \cref{fig:hoxa11_evolution_tree}).
-
-\subsubsection{\ac{GH} family}
-
-The \ac{GH} family has expanded in anthropoid primates, resulting in
-approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
-\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
-\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
-between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
-reliably distinguish transcripts between the two). All show
-substantial expression in the \human placenta
-(\cref{fig:csh1_evolution_tree}). The expression of the \ac{GH} family
-in placenta appears to be primate specific, and likely existed at the
-time of the last common ancestor between all extant anthropoids
-\cite{Papper.ea2009:Ancientoriginofplacental}. There is some evidence
-that increased expression of these genes may be correlated with
-overall birth weight of offspring in \human
-\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
-possible that the extreme expression levels of these genes may be due
-to high birth weight. \human placental \ac{GH} genes are involved in
-increasing blood flow, replacing maternal pituitary expression of
-\ac{GH}, as well as trophoblast invasion
-\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
-and it is likely that they have the same function in \bonobo.
-
-\begin{figure}
-<<chorionic.somatomammotropin.alpha.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-plot.label.size <- 4.5
-pushViewport(viewport(layout=grid.layout(nrow=1,ncol=2,widths=c(2,3))))
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CGA" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=1),
-                 offset.ratio=0.7,
-                 fontsize=plot.label.size,
-                 axis.text.size=3.5
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CGB1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=2,layout.pos.row=1),
-                 subtree=c("ENSMODG00000013768","ENSG00000104818"),
-                 offset.ratio=1,
-                 fontsize=plot.label.size,
-                 axis.text.size=3.5
-                 )
-popViewport(1)
-grid.text(c("A","B"),
-          x=unit(c(0,2/5),"npc"),y=unit(c(1,1),"npc"),
-          just=c("left","top"),
-          gp=gpar(fontsize="48")
-          )
-@ 
-\caption{\genetreeexplanation{(A) \textit{CGA} (B) \textit{CGB}}}
-\label{fig:cga_cgb_evolution_tree}
-\end{figure}
-
-The human gene \textit{CGA} codes for the alpha subunit of multiple
-hormones including pituitary hormones (luteinizing hormone,
-follicle-stimulating hormone, thyroid-stimulating hormone), as well as
-numerous chorionic gonadotropins with variable beta subunits
-\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of
-the chorionic gonadotropins is to prevent the regression of the corpus
-luteum \cite{Hanson.ea1971:EffectsofHCGand}. Among the six placentas
-we analyzed here, \textit{CGA} is only expressed in the placentas of
-\human and \bonobo (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}),
-which corresponds with the emergence of the chorionic gonadotropins in
-anthropoid primates
-\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly,
-\textit{CGA} is expressed at much higher levels in both species than
-the beta subunits with which its product dimerizes (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}).
-\textit{CGA} is one of several genes noted by
-\citet{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone
-adaptive evolution of the \human{}-\textit{P. troglodytes} stem lineage.
-
-\subsubsection{Relaxin}
-
-Relaxin (\textit{RLN}) is a hormone that has multiple effects during
-pregnancy including increasing maternal blood flow, widening the pubic
-bone and relaxing uterine musculature in preparation for labor. In
-\dog, \elephant, \mouse, \carmeli, and \horse there is high expression
-of relaxin in the term placenta, with little evidence in other species
-examined here (Supplemental \cref{fig:rln2_evolution_tree}). \textit{RLN} is known
-to be produced in the placenta of the pregnant \dog and is used in the
-diagnosis of canine pregnancy
-\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
-the luteal ovary and placenta of \pig
-\cite{Hisaw.ea1948:Relaxininovaryof}, and the placenta of \horse
-\cite{Klonisch.ea1997:Placentallocalizationofrelaxin}. However, the
-source of relaxin produced by pregnant \elephant remains unclear
-\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in relaxin in
-\elephant has been suggested to be a major signal for the initiation
-of labor \cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive}.
-Here we show that the placenta of elephants show high expression of
-relaxin and is likely a major source of \textit{RLN} at term. In
-contrast to the results of
-\cite{Evans.ea1993:mouserelaxingenenucleotide}, we also see expression
-of \textit{RLN} in the placenta of \mouse and \carmeli.
-
-\subsubsection{\textit{NPPB} and \textit{NPPC}}
-
-Two genes, \textit{NPPB} and \textit{NPPC}, code for peptide hormones
-that are important in the inducement of vasodilation, and may be
-important in increasing maternal blood flow to the fetus
-\cite{Cameron.ea1996:sitesofgeneexpression}. There is some indication
-that these proteins are present in the decidua of mice
-\cite{Cameron.ea1996:sitesofgeneexpression}, but we find very little
-expression in the fetal portion of the mouse placenta
-(\cref{fig:hormone_genes}, Supplemental
-\cref{fig:nppc_evolution_tree,fig:nppb_evolution_tree}). While we do
-find high expression of at least one of these genes in \dog, \cow, and
-\elephant, the expression pattern does not follow the placental
-organization (endotheliochorial for \dog and \elphant,
-synepitheliochorial for \cow and \sheep), so it is unlikely that
-expression of these genes are essential for specific placental
-organizations.
-
-\subsection{The Evolution of Placental Expression of Galectins}
-
-Galectins are recognized as major regulators of pregnancy in both
-fetal and maternal tissues
-\cite{Than.ea2012:Galectinsguardiansofeutherian}. Indeed, the
-evolution of new lineage-specific galectins has occurred in both
-primates and ruminants, with the genes \textit{LGALS13},
-\textit{LGALS14}, and \textit{LGALS16} recognized as being
-placenta-specific in \human and regulating immune responses by
-inducing apoptosis of maternal T-cells
-\cite{Than.ea2009:primatesubfamilyofgalectins} (Supplemental
-\cref{fig:galectins_lgals13_evolution_tree}). Indeed, \textit{LGALS14}
-has the highest expression of any galectin in the \human and \smonkey
-placenta (\cref{fig:hormone_genes}, Supplemental
-\cref{fig:galectins_lgals13_evolution_tree}), and other primate
-galectins such as \textit{LGALS13} and \textit{LGALS16} are also
-expressed in \human and \smonkey placenta. However, in the closely
-related \bonobo, \textit{LGALS14} is not found, and \textit{LGALS1} is
-the most abundant galectin. \textit{LGALS1} shows generally high
-expression in all placentas analyzed. \textit{LGALS1} is also involved
-in mediating maternal-fetal immune tolerance, and is expressed in both
-maternal and fetal tissue in primates
-\cite{Than.ea2008:Emergenceofhormonaland}. \textit{LGALS1} placental
-expression was thought to be confined to eutherian mammals, as there
-is evidence of adaptive evolution at this node, as well as the
-appearance of estrogen response elements in eutherian cis-regulatory
-regions \cite{Than.ea2008:Emergenceofhormonaland}. It was unclear as
-to whether galectins such as \textit{LGALS1} were important in the
-regulation of pregnancy in marsupials, as well as eutherians. Here we
-show substantial expression of galectins in the opossum, especially
-\textit{LGALS1} (Supplemental \cref{fig:galectins_evolution_tree}). In
-addition, \textit{LGALS2} is also expressed. The presence of galectins
-at the maternal-fetal interface in \opossum implies that it was also
-present in the ancestral placenta of therian mammals. Indeed, adaptive
-evolution of \textit{LGALS1} on the eutherian stem is likely due to
-the increased invasive nature of eutehrian placentation rather than
-the origination of maternal-fetal tolerance. Although the function of
-\textit{LGALS1} in the marsupial placenta needs to be confirmed, it is
-likely that the mediation of maternal-fetal immune tolerance by
-galectins is an ancient feature of mammalian viviparity.
-
-\subsection{The Opossum Placenta and the Evolution of Viviparity}
-
-The placenta of \opossum is distinctly different from that of the term
-placenta of eutherian mammals in that it is derived from both the
-chorion and yolk sac. Other differences between the placentas of
-marsupials and eutherians include lack of invasive extravillous
-trophoblasts in marsupials, the retention of a permeable shell coat
-for 2/3 of marsupial gestation, and the greater degree of imprinting
-in placenta-expressed genes in eutherians
-\cite{Renfree2010:ReviewMarsupialsplacentalmammals,Renfree.ea2013:originandevolutionof}.
-In addition, the placenta of \opossum is also distinct from other
-marsupials, in that it is considered one of the least specialized, and
-thought be closer to the original marsupial condition
-\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. In \opossum, the
-duration of direct connection between maternal and fetal tissues is as
-short as two days in a gestational period of 13-15 days
-\cite{Harder.ea1993:Gestationandplacentationin}. During this period,
-the placenta is directly engaged in the exchange of urea with maternal
-tissues, owing to the lack of a vesicular allantois
-\cite{Freyer.ea2007:Placentalfunctionintwo}. This differs drastically
-with the situation in another well-studied marsupial, the tammar
-wallaby \cite{Freyer.ea2007:Placentalfunctionintwo}. Therefore, we
-must interpret differences between the opossum and eutherian placentas
-with caution, as highly expressed genes and functional annotation
-clusters may represent didelphid or \opossum specific adaptations.
-
-% Genes that show a significant shift on the \opossum lineage include
-% functional annotation clusters that correspond to active generation of
-% energy, transport, and lipid breakdown via the peroxisome (Table 10;
-% Supplemental Table X). The choriovitelline placenta of both didelphids
-% and the tammar wallaby are known to be involved in glycolysis and
-% protein synethsis (Freyer et al., 2003). This implies that the
-% choriovitelline placenta in marsupials is an active site of metabolism
-% and molecular synthesis, as well as the transport of nutrients from
-% the maternal tissues. This is a distinct difference that is not seen
-% in significant shifts on other lineages, and may be due to the
-% formation of the placenta in the opossum from the yolk sac. In
-% addition, the urea cycle is a highly enriched functional annotation
-% cluster which directly correlates with the high levels of urea in the
-% Monodelphis placenta (Freyer et al., 2007). This is potentially a
-% Monodelphis-specific characteristic as discussed above.
-
-The \opossum placenta shows expression of genes that either do not
-exist in eutherians or are not found in the term eutherian placenta
-(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
-of hemoglobin in fetal tissues at term, as evidenced by the expression
-of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
-generally due to their very short gestation periods. Embryonic
-hemoglobins in marsupials have a lower affinity for oxygen and allow
-for a lower rate of oxygen uptake from maternal tissues
-\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
-uncharacterized gene that shows high expression (\textit{LOC10001195}) is not
-found in eutherians; Ensembl gene trees reveal that homologs of this
-gene exist in other amniotes such as birds and turtles, but are not
-present in eutherian mammals. The homolog of this gene in chicken is
-found to expressed in reproductive tract
-\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
-unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
-family and involved in innate immunity. It is expressed in the mammary
-gland of the tammar wallaby during pregnancy and early lactation
-\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and present in
-relative abundance in \opossum placenta.
-
-\subsection{Novel Gene Expansions in the Elephant Placenta}
-
-We found evidence of high expression in two novel expansions of genes
-in the \elephant genome. One family includes 11 genes that are related
-to the \textit{PRSS16} gene, a thymus-specific serine protease that
-may be involved regulating the presentation of self-peptides in CD4+
-T-lymphocytes \cite{Fornari.ea2012:TranscriptionprofilingofPrss16}.
-All 11 of these genes are expressed in the elephant placenta, three of
-which have very abundant mRNAs
-(\cref{tab:top50_elephant_placenta} and \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}C). It is
-possible that this elephant-specific gene family may be involved in
-immunity in the elephant placenta and the mediation of maternal-fetal
-immune tolerance; however, future work will need to test these
-hypotheses. Three other genes abundantly expressed in the elephant
-placenta are homologues of \textit{SSLP1}, a gene with predominant
-expression in the seminal vesicle in mouse
-\cite{Li.ea2006:SSLP-1secretedLy-6protein}. Homologues of
-\textit{SSLP1} are also expressed in \cow, \dog, \horse, and \sheep.
-\textit{SSLP1} also contains an Alu retrotransposon and may be
-involved in the development of villi via cell proliferation
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}. This
-gene is not found in primates. Expansion of these genes with high
-expression in \elephant may indicate an elaboration of this system.
-
-\subsection{Annexin Complex}
-
-\subsection{Human Specific Expression Shifts and the Evolution of Pre-eclampsia}
-
-The \human placenta is distinctive from many other primates in that it
-possesses a deeply invasive trophoblast that penetrates the inner
-myometrium of the uterus
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
-hypothesized that this feature is related to the increased demand for
-resources by the developing fetus to feed the expanded size of the
-human brain
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation,Robillard.ea2002:Evolutionaryadaptationstopre-eclampsiaeclampsia}.
-\human have the largest brain in terms of relative size and this
-differs markedly from closely related great apes
-\cite{Boddy.ea2012:Comparativeanalysisofencephalization}. Deep
-penetration of the myometrium occurs in chimpanzees and gorillas, but
-this penetration may be accelerated in humans
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
-
-The near exclusivity of deep trophoblast invasion in humans has also
-led many to speculate on the origin of various disorders, which are
-almost exclusively symptomatic of human pregnancy. For example,
-pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
-the United States
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
-can be attributed to the lack of adequate invasion of maternal tissue,
-and is noted by the onset of hypertension and proteinuria
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
-pre-eclampsia can develop into eclampsia, a more severe condition
-resulting in seizures during pregnancy and death. Lack of invasion in
-pre-eclampsia may be attributed to the inefficiency of growth
-regulators as well as the inability to sufficiently suppress the
-maternal immune response. Although there are speculative cases of
-eclampsia in great apes
-\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
-it is exceedingly rare in these species
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is possible
-that a comparative evolutionary approach to gene expression can
-identify genes that have divergent expression patterns from other
-mammals as well as association with pre-eclampsia.
-
-Functional annotation clusters significantly enriched on the human
-lineage (blood vessel development, extracellular matrix, tube
-morphogenesis) seem to suggest involvement in remodeling of tissues
-and increased blood flow, both of which may correlate with the
-invasion of the trophoblast (\cref{tab:go_terms_all_species}). In
-addition, clusters such as cytokines and defense response
-(\cref{tab:go_terms_all_species}) also indicate an a significant
-change in expression of the immune response that may be involved in
-immunosupression at the maternal-fetal boundary to further allow
-invasion of fetal tissues.
-
-% I haven't actually done this analysis, so commented out
-
-% Many genes that both show expression shifts on the human lineage and
-% have very high expression (FPKM>200) are directly related to the
-% highly invasive nature of the human placenta. For example, GPC3,
-% glypican 3, is an imprinted gene and cell growth regulator, associated
-% with Simpson-Golabi-Behmel fetal overgrowth syndrome, and involved in
-% intrauterine growth restriction (Monk and Moore, 2004; Pilia et al.,
-% 1996). GPC3 has the highest expression in humans (Table 2), but is
-% also expressed in other mammals to a lesser degree, especially those
-% with invasive hemochorial placentas like bonobos (\ac{FPKM}= 199.924),
-% spider monkeys (\ac{FPKM}= 96.24), and the mouse (\ac{FPKM}= 80.7545)
-% (Supplemental Tables X-XX). ADAM12 increases the availability of
-% insulin-like growth factors that lead to the promotion of cell growth
-% and placental invasion (Laigaard et al., 2005). (Hou et al., 2009)
-% found evidence of adaptive evolution in ADAM12 on the primate lineage,
-% and like GPC3 it is expressed in the bonobo (\ac{FPKM}= 46.1553), spider
-% monkey (\ac{FPKM}=30.41) and mouse (\ac{FPKM}=2.25391) as well, although to a
-% lesser degree (Supplemental Tables X-XX). RHOBTB3 is imprinted and
-% associated with trophoblast invasion as well (Bilban et al., 2010;
-% Yuen et al., 2011). In addition, one other gene, KISS1, shows
-% exceptional abundance in the human placenta (Table 2); however due to
-% its absence in elephant Ensmebl gene annotations, it was not part of
-% the 1:1 orthologues examined in the comparative study. Here, it is
-% only expressed in the term placentas of humans (Table 2) and bonobos
-% (\ac{FPKM}=17.6394; Supplemental Table X). KISS1 is thought to be involved
-% in the regulation of trophoblast invasion, and shows highest
-% expression in the first trimester of pregnancy (Bilban et al., 2004;
-% Reynolds et al., 2009).
-% 
-% In addition, several other genes that show expression shifts on the
-% human lineage also have major effects on placental function. Genes
-% such as TXK and EBI3 are intimately involved in the immune system in
-% the placenta. TXK is involved in natural killer cell differentiation
-% (Bezman et al., 2012). EBI3 is implicated in natural killer cell
-% function an maternal-fetal immunomodulation in the syncytiotrophoblast
-% (Devergne et al., 2001). KYNU and KMO are interacting genes that are
-% involved in tryptophan metabolism, which may be involved in
-% immunosupression and the prevention of fetal rejection (Takikawa,
-% 2005). THSD7A is a thrombospondin, which inhibits endothelial cell
-% migration and tube morphogenesis (Wang et al., 2010).
-% 
-% Many genes that show evidence of a shift on the human lineage show
-% evidence of implication in pre-eclampsia, including ADAM12 (Gack et
-% al., 2005; Laigaard et al., 2005) MAOA (Carrasco et al., 2000;
-% Bottalico et al., 2004), BCL2 (Ishihara et al., 2002), GPC3 (Gui et
-% al., 2012; Ogawa, 2007), TMEM8A, CLIC3, EBI3, FLT1, BCL6, SYDE1, and
-% MAN1C1 (Enquobahrie et al., 2008; Winn et al., 2009; Nishizawa, 2011).
-% In addition KISS1 is also involved in pre-eclampsia as well (Sado et
-% al., 2011; Qiao et al, 2012). Due to the association of many genes
-% that show human-specific shifts as well as high expression, we
-% conclude that pre-eclampsia is possibly a symptom of recent evolution
-% of gene expression on the human lineage. Although many genes such as
-% GPC3 and ADAM12 show distinctive relatively high expression in bonobo,
-% this expression has intensified on the human lineage. Sequencing of
-% more placentas of nonhuman primates, including other individuals of
-% the genus Pan will help to further characterize this transition.
-
-\section{Conclusion}
-
-\DLA{Totally rewrite the conclusion; it's terrible}
-
-We sequenced \Sexpr{star.logs[,length(unique(species))]} placentas
-spanning the phylogenetic tree of mammals. This revealed a dynamic
-expression pattern of expression of lineage-specific expansions of
-gene families and differential gene expression that agreed with the
-scenario of an “arms race” between maternal and fetal tissues that
-have shaped such disparate gene expression profiles. In addition, we
-highlighted the broad spectrum of hormone gene expression in multiple
-species that suggested that there is a lack of conservation in hormone
-signaling during mammalian pregnancy. Major differences between
-expression in marsupial and eutherian mammals at term seemed to be
-correlated with the differences between choriovitelline and
-chorioamniotic placenta, with the marsupial placenta more enriched in
-genes which contribute to the active generation of energy. We also
-identified that galectin expression, especially that of
-\textit{LGALS1}, is a common feature of all mammalian placentas and
-was most likely present in the ancestral therian. Also, we identified
-two new expansions of genes that are highly expressed in the \elephant
-placenta that may be involved in trophoplast differentiation and immune
-system function, mirroring expansions in other groups such as
-primates, rodents, and ruminants. We also quantified significant
-lineage-specific shifts in gene expression, concentrating on the
-identification of genes that showed gene expression changes on the
-human lineage. We cautiously propose that many of these genes may be
-related to the invasive nature of the trophoblast in \human, although
-we note that in many of these genes, there is in some cases
-substantial expression in the \bonobo as well. These genes have the
-potential to spur further research and may be at the root of
-obstetrical disorders such as pre-eclampsia. We propose that more
-placentas from diverse lineages, especially those of marsupials and
-primates, to further identify genes and expression levels that are
-specific to marsupials and eutherians, as well as confirm the identity
-of human-specific expression shifts.
-
-\begin{backmatter}
-  \section*{Competing interests}
-  The authors declare that they have no competing interests.
-
-  \section*{Author's contributions}
-  Don Armstrong analyzed all of the data and wrote the paper.
-
-  Michael McGowan wrote the paper and performed initial analyses.
-
-  Amy Weckle performed all of the RNA extractions and managed the
-  samples.
-
-  Jason Caravas performed initial analyses.
-
-  Dalen Agnew provided samples.
-
-  Kurt Benirschke provided samples.
-
-  Sue Savage-Rumbaugh provided samples.
-
-  Eviatar Nevo provided samples.
-
-  Chong J. Kim provided samples.
-
-  Günter P. Wagner provided samples and wrote the paper.
-
-  Roberto Romero provided samples and conceived the study.
-
-  Derek E. Wildman conceived the study, wrote the paper, and performed
-  analyses.
-
-  \section*{Acknowledgements}
-  The authors would like to specially acknowledge those groups whose
-  publicly available data contributed to the completeness of this
-  work.
-
-  \printaddresses
-
-\end{backmatter}
-
-\printbibliography
-
-\end{document}
+%%% Local Variables:
+%%% mode: latex
+%%% TeX-master: "manuscript.Rnw"
+%%% End:

### 2016-03-08 manuscript/manuscript.Rnw
add common_latex.tex

@@
 
-% species names
-\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
-\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
-\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
-\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
-\newcommand{\horse}[0]{\textit{E. caballus}\xspace}
-\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
-\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
-\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
-\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
-\newcommand{\galili}[0]{\textit{S. galili}\xspace}
-\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
-\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
-\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
-\newcommand{\pig}[0]{\textit{S. scrofa}\xspace}
-
-% top 50 gene table explanation. Note that if you modify this, you
-% should probably also modify the human table explanation, which is
-% slightly different.
-\newcommand{\tableexplanation}[2]{The top 50 genes by \ac{FPKM} in #1
-  which are not annotated as Mt\_rRNA, Mt\_tRNA, rRNA or whose \human
-  ortholog is a housekeeping gene. Gene symbol is the gene symbol or
-  Ensembl gene id. Human Symbol is the 1:1 human ortholog or the best
-  human alignment when enclosed in parenthesis. Human Name is the HUGO
-  name corresponding to the human symbol. A complete list of all genes
-  is in \cref{#2}.}
-
-\newcommand{\genetreeexplanation}[1]{Gene tree and heatmap of
-  expression of genes related to #1. The gene tree was obtained from
-  Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes from species
-  not included in this study or having expression by \ac{FPKM} less
-  than 1 were excluded from the plotted tree. Heatmap cells in white
-  are from species where there was no expression of the relevant
-  transcript (or a species where that transcript does not apply). The
-  species names are abbreviated with the first letter of the genus and
-  first two letters of the species (for example, \textit{Homo sapiens}
-  is \textit{H.sa.}). Genes with duplicate gene names in the same
-  species (or no gene named defined) are listed by the Ensemble gene
-  ID.}
+\input{common_latex}
 

### 2016-03-08 manuscript/supplemental_material.Rnw
add common_latex.tex

@@
 
-\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
-\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
-\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
-\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
-\newcommand{\horse}[0]{\textit{E. caballus}\xspace}
-\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
-\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
-\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
-\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
-\newcommand{\galili}[0]{\textit{S. galili}\xspace}
-\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
-\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
-\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
-\newcommand{\pig}[0]{\textit{S. scrofa}\xspace}
-
-\newcommand{\genetreeexplanation}[1]{Gene tree and heatmap of
-  expression of genes related to #1. The gene tree was obtained from
-  Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes from species
-  not included in this study or having expression by FPKM less
-  than 1 were excluded from the plotted tree. Heatmap cells in grey
-  are from species where there was no expression of the relevant
-  transcript (or a species where that transcript does not apply).}
+\input{common_latex}
 

### 2016-03-09 presentation/placenta_viviparous_presentation.Rnw
update placenta viviparous presentation
use minted and morewrites

@@
 \documentclass[ignorenonframetext]{beamer}
+\usepackage{morewrites}
 \usepackage{fontspec}
@@
 \usepackage{mathtools,cancel}
+\usepackage{minted}
+\usepackage{tcolorbox}
+\usepackage{etoolbox}
+\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
+\AfterEndEnvironment{minted}{\end{tcolorbox}}%
 
@@
 \begin{frame}{Species Studied}
-\begin{figure}
+% \begin{figure}
 <<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
@@
 
-\subsection{Primate Placenta Transcriptomes}
-
-\begin{frame}{Primate Placenta Transcriptome}
-\begin{table}
-  \footnotesize
-<<primate.lineage.specific,echo=FALSE,results="asis">>=
-pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
-setkey(pri.all.lin.spec,"Human Symbol")
-@ 
-\end{table}
-\end{frame}
-
-\begin{frame}{Pregnancy-Specific Glycoprotein Expansion}
-\begin{columns}
-\column{0.4\textwidth}
-\begin{itemize}
-\item 11 highly expressed \textit{PSG} genes in \human and some
-  expression in \smonkey and \bonobo
-\item \mouse specific expansion and high expression of \textit{PSG} as
-  well
-\item No other species have high expression of \textit{PSG} or
-  \textit{CEACAM}
-\end{itemize}
-\column{0.6\textwidth}
-\begin{figure}
-<<psg.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-plot.label.size <- 3
-axis.text.size <- 3.5
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"),
-                 axis.text.size=axis.text.size,
-                 fontsize=plot.label.size,
-                 )
-@ 
-\end{figure}
-\end{columns}
-\end{frame}
-
-\begin{frame}{Cathepsin-L expansion}
-\begin{columns}
-\column{0.4\textwidth}
-\begin{itemize}
-\item Cathepsins are highly expressed in many placentas
-\item Expanded in \mouse and \galili and possibly also \carmeli to a
-  lesser degree
-\item Intracellular protease which degrades collagen, elastin, and in
-  IGF1R
-\end{itemize}
-\column{0.6\textwidth}
-\begin{figure}
-<<ctsl.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-plot.label.size <- 3
-axis.text.size <- 3.5
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"),
-                 fontsize=4,
-                 width=3,
-                 axis.text.size=axis.text.size,
-                 offset.ratio=1
-                 )
-
-@ 
-\end{figure}
-\end{columns}
-\end{frame}
-
-\begin{frame}{Pregnancy associated glycoprotein}
-\begin{columns}
-\column{0.4\textwidth}
-\begin{itemize}
-\item Massive expansion in ruminants (\cow and \sheep)
-\item 28 members of this family have $\mathrm{FPKM} \ge 1$ in \cow
-\item \Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}
-\item \Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}
-\end{itemize}
-\column{0.6\textwidth}
-\begin{figure}
-<<pag2.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 subtree=c("ENSSSCG00000006792",
-                           "ENSBTAG00000046085"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"),
-                 axis.text.size=axis.text.size,
-                 fontsize=plot.label.size,
-                 offset.ratio=0.6
-                 )
-
-@ 
-\end{figure}
-\end{columns}
-\end{frame}
-
 \subsection{Evolution of Placental Hormones}
@@
 
+\subsection{Primate Placenta Transcriptomes}
+
+\begin{frame}{Primate Placenta Transcriptome}
+\begin{table}
+  \footnotesize
+<<primate.lineage.specific,echo=FALSE,results="asis">>=
+pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
+setkey(pri.all.lin.spec,"Human Symbol")
+@ 
+\end{table}
+\end{frame}
+
+\begin{frame}{Pregnancy-Specific Glycoprotein Expansion}
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item 11 highly expressed \textit{PSG} genes in \human and some
+  expression in \smonkey and \bonobo
+\item \mouse specific expansion and high expression of \textit{PSG} as
+  well
+\item No other species have high expression of \textit{PSG} or
+  \textit{CEACAM}
+\end{itemize}
+\column{0.6\textwidth}
+\begin{figure}
+<<psg.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+plot.label.size <- 3
+axis.text.size <- 3.5
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
+                 vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"),
+                 axis.text.size=axis.text.size,
+                 fontsize=plot.label.size,
+                 )
+@ 
+\end{figure}
+\end{columns}
+\end{frame}
+
+\begin{frame}{Cathepsin-L expansion}
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item Cathepsins are highly expressed in many placentas
+\item Expanded in \mouse and \galili and possibly also \carmeli to a
+  lesser degree
+\item Intracellular protease which degrades collagen, elastin, and in
+  IGF1R
+\end{itemize}
+\column{0.6\textwidth}
+\begin{figure}
+<<ctsl.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+plot.label.size <- 3
+axis.text.size <- 3.5
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
+                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
+                 vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"),
+                 fontsize=4,
+                 width=3,
+                 axis.text.size=axis.text.size,
+                 offset.ratio=1
+                 )
+
+@ 
+\end{figure}
+\end{columns}
+\end{frame}
+
+\begin{frame}{Pregnancy associated glycoprotein}
+\begin{columns}
+\column{0.4\textwidth}
+\begin{itemize}
+\item Massive expansion in ruminants (\cow and \sheep)
+\item 28 members of this family have $\mathrm{FPKM} \ge 1$ in \cow
+\item \Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}
+\item \Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}
+\end{itemize}
+\column{0.6\textwidth}
+\begin{figure}
+<<pag2.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,min.fpkm=1,
+                 subtree=c("ENSSSCG00000006792",
+                           "ENSBTAG00000046085"),
+                 vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"),
+                 axis.text.size=axis.text.size,
+                 fontsize=plot.label.size,
+                 offset.ratio=0.6
+                 )
+
+@ 
+\end{figure}
+\end{columns}
+\end{frame}
+
 \begin{frame}{Prolactin}

### 2016-03-09 presentation/placenta_viviparous_presentation.Rnw
add section on makefiles

@@
 
+\subsection{Using Make}
+
+\begin{frame}{Makefile rules}
+\centering
+\includegraphics[keepaspectratio,height=0.8\textheight]{figures/default}
+\end{frame}
+
+\begin{frame}[fragile]{Common Options}
+\tiny
+\inputminted[firstline=8, lastline=36]{make}{../fastq/common_makefile}
+\end{frame}
+
+\begin{frame}[fragile]{FASTQC}
+\inputminted[firstline=78, lastline=83]{make}{../fastq/common_makefile}
+\end{frame}
+
+\begin{frame}[fragile]{Trinity}
+\tiny
+\inputminted[firstline=211, lastline=222]{make}{../fastq/common_makefile}
+\end{frame}
+
+\begin{frame}[fragile]{Diamond}
+\tiny
+\inputminted[firstline=249, lastline=268]{make}{../fastq/common_makefile}
+\end{frame}
+
+\begin{frame}[fragile]{STAR Indexes}
+\tiny
+\inputminted[firstline=86, lastline=105]{make}{../fastq/common_makefile}
+\end{frame}
+
+\begin{frame}[fragile]{STAR}
+\tiny
+\inputminted[firstline=106, lastline=118]{make}{../fastq/common_makefile}
+\end{frame}
+
+\begin{frame}[fragile]{Cufflinks}
+\tiny
+\inputminted[firstline=43, lastline=60]{make}{../fastq/common_makefile}
+\end{frame}
+
 \subsection{Additional Methods}

### 2016-03-09 presentation/placenta_viviparous_presentation.Rnw
make figures larger and more centered

@@
 @ 
-\end{figure}
+% \end{figure}
 \end{frame}
@@
 \begin{frame}{Variation in Gestational Periods and Litter Size}
+\centering
 \footnotesize
@@
 %\begin{table}
+\centering
 \footnotesize
@@
 \begin{figure}
-<<transcriptome.mapping.figure,fig.width=10,fig.height=8,echo=FALSE>>=
+<<transcriptome.mapping.figure,fig.width=4,fig.height=4,echo=FALSE>>=
 print(ggplot(star.logs[grepl("input reads$",field),list(reads=as.numeric(value),species=capfirst(species))],
@@
       + theme(legend.position="top",
-              axis.text.x = element_text(size=10,face="italic"))
+              axis.text.x = element_text(size=10),
+              axis.text.y = element_text(size=10,face="italic"))
       )
@@
 \begin{figure}
-<<transcriptome.mapping.figure.unmapped,fig.width=8,fig.height=8,echo=FALSE>>=
+<<transcriptome.mapping.figure.unmapped,fig.width=4,fig.height=4,echo=FALSE>>=
 star.logs.unmapped <- 
@@
       + theme(legend.position="top",
-              axis.text.x = element_text(face="italic"))
+              axis.text.x = element_text(size=10),
+              axis.text.y = element_text(face="italic",size=10))
       )
@@
 \begin{figure}
-<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
+<<core.placenta.transcriptome,fig.width=10,fig.height=6,out.height="0.8\\textheight",echo=FALSE>>=
 print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
-             aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
+             aes(x=human_name,y=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
       + geom_raster(aes(fill=log10(FPKM)))
@@
                             )
-      + ylab("Human Ortholog Name")
-      + xlab("Species")
+      + xlab("Human Ortholog Name")
+      + ylab("Species")
       + theme_grey()
@@
               axis.text.y = element_text(face="italic"),
-              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
+              axis.text.x = element_text(angle=90,hjust=1,face="italic"))
       )
@@
 \begin{figure}
-<<hormone.genes.plot,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+<<hormone.genes.plot,echo=FALSE,warning=FALSE,message=FALSE>>=
 

### 2016-03-09 presentation/placenta_viviparous_presentation.Rnw
remove segments (bz time) from hpcbio pres.

@@
 
-\subsection{Primate Placenta Transcriptomes}
-
-\begin{frame}{Primate Placenta Transcriptome}
-\begin{table}
-  \footnotesize
-<<primate.lineage.specific,echo=FALSE,results="asis">>=
-pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
-setkey(pri.all.lin.spec,"Human Symbol")
-@ 
-\end{table}
-\end{frame}
-
-\begin{frame}{Pregnancy-Specific Glycoprotein Expansion}
-\begin{columns}
-\column{0.4\textwidth}
-\begin{itemize}
-\item 11 highly expressed \textit{PSG} genes in \human and some
-  expression in \smonkey and \bonobo
-\item \mouse specific expansion and high expression of \textit{PSG} as
-  well
-\item No other species have high expression of \textit{PSG} or
-  \textit{CEACAM}
-\end{itemize}
-\column{0.6\textwidth}
-\begin{figure}
-<<psg.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-plot.label.size <- 3
-axis.text.size <- 3.5
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"),
-                 axis.text.size=axis.text.size,
-                 fontsize=plot.label.size,
-                 )
-@ 
-\end{figure}
-\end{columns}
-\end{frame}
-
-\begin{frame}{Cathepsin-L expansion}
-\begin{columns}
-\column{0.4\textwidth}
-\begin{itemize}
-\item Cathepsins are highly expressed in many placentas
-\item Expanded in \mouse and \galili and possibly also \carmeli to a
-  lesser degree
-\item Intracellular protease which degrades collagen, elastin, and in
-  IGF1R
-\end{itemize}
-\column{0.6\textwidth}
-\begin{figure}
-<<ctsl.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-plot.label.size <- 3
-axis.text.size <- 3.5
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"),
-                 fontsize=4,
-                 width=3,
-                 axis.text.size=axis.text.size,
-                 offset.ratio=1
-                 )
-
-@ 
-\end{figure}
-\end{columns}
-\end{frame}
-
-\begin{frame}{Pregnancy associated glycoprotein}
-\begin{columns}
-\column{0.4\textwidth}
-\begin{itemize}
-\item Massive expansion in ruminants (\cow and \sheep)
-\item 28 members of this family have $\mathrm{FPKM} \ge 1$ in \cow
-\item \Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}
-\item \Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}
-\end{itemize}
-\column{0.6\textwidth}
-\begin{figure}
-<<pag2.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 subtree=c("ENSSSCG00000006792",
-                           "ENSBTAG00000046085"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"),
-                 axis.text.size=axis.text.size,
-                 fontsize=plot.label.size,
-                 offset.ratio=0.6
-                 )
-
-@ 
-\end{figure}
-\end{columns}
-\end{frame}
-
-\begin{frame}{Prolactin}
-\begin{columns}
-\column{0.4\textwidth}
-\begin{itemize}
-\item \elephant is only analyzed mammal with high prolactin expression
-  (\textit{PRL}) in the placenta
-\item However, \mouse, \carmeli, and \cow have duplications of
-  \textit{PRL} with expression
-\end{itemize}
-\column{0.6\textwidth}
-\begin{figure}
-<<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
-                 axis.text.size=5,
-                 fontsize=4)
-@ 
-\end{figure}
-\end{columns}
-\end{frame}
-
-\subsubsection{GH family}
-
-\begin{frame}{Growth Hormone (GH) family}
-
-\begin{columns}
-\column{0.4\textwidth}
-The GH family has expanded in anthropoid primates, resulting in
-approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
-\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
-\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
-between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
-reliably distinguish transcripts between the two).
-
-\column{0.6\textwidth}
-<<chorionic.somatomammotropin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CSH1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
-                 offset=2,
-                 axis.text.size=5,
-                 fontsize=10)
-@ 
-\end{columns}
-
-\end{frame}
-
-\begin{frame}{CGA}
-
-  \begin{columns}
-\column{0.4\textwidth}
-\begin{itemize}
-\item \textit{CGA} codes for the alpha subunit of multiple hormones as
-  well as numerous chorionic gonadotropins with variable beta subunits
-  \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}.
-
-\item Chorionic gonadotropins prevent the regression of the corpus
-  luteum \cite{Hanson.ea1971:EffectsofHCGand}.
-
-\item \textit{CGA} only expressed in the placentas of \human and
-  \bonobo
-
-  \pnote{Expression corresponds with emergence of the chorionic gonadotropins in
-    anthropoid primates
-    \cite{Maston.ea2002:Chorionicgonadotropinhasrecent}.}
-
-\end{itemize}
-\column{0.6\textwidth}
-\begin{figure}
-<<chorionic.somatomammotropin.alpha.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-plot.label.size <- 4.5
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CGA" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 offset.ratio=1,
-                 fontsize=8,
-                 axis.text.size=5
-                 )
-@ 
-\end{figure}
-\end{columns}
-\end{frame}
-
 \begin{frame}{CGB}
@@
 \end{columns}
-\end{frame}
-
-\begin{frame}
-\begin{columns}
-\column{0.4\textwidth}
-\begin{itemize}
-\item Evolution of new lineage-specific galectins has occurred in both
-  primates and ruminants, with the genes \textit{LGALS13},
-  \textit{LGALS14}, and \textit{LGALS16} recognized as being
-  placenta-specific in \human and regulating immune responses by
-  inducing apoptosis of maternal T-cells
-  \cite{Than.ea2009:primatesubfamilyofgalectins}
-\end{itemize}
-
-\column{0.6\textwidth}
-\begin{figure}
-<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
-                 offset=0.8,
-                 axis.text.size=7,
-                 fontsize=10
-                 )
-@ 
-\end{figure}
-\end{columns}
-\end{frame}
-
-\subsection{The Opossum Placenta and the Evolution of Viviparity}
-
-\begin{frame}
-\begin{itemize}
-\item The \opossum placenta shows expression of genes that either do
-  not exist in eutherians or are not found in the term eutherian
-  placenta
-
-\item The existence of embryonic forms of hemoglobin in fetal tissues
-  at term, as evidenced by the expression of \textit{HBE} and
-  \textit{HBZ} is a feature of marsupials that is generally due to
-  their very short gestation periods.
-
-\item Embryonic hemoglobins in marsupials have a lower affinity for
-  oxygen and allow for a lower rate of oxygen uptake from maternal
-  tissues \cite{Holland.ea1997:Characterizationofembryonicglobin}.
-
-\item One novel uncharacterized gene that shows high expression
-  (\textit{LOC10001195}) is not found in eutherians; Ensembl gene
-  trees reveal that homologs of this gene exist in other amniotes such
-  as birds and turtles, but are not present in eutherian mammals.
-
-\item \textit{WFDC2} is a member of the whey acidic protein (WAP)
-  family and involved in innate immunity.
-\end{itemize}
 \end{frame}

### 2016-03-10 common_r_code/plot_tree_matrix.R
allow the angle to be 0 for plotting

@@ -4,7 +4,7 @@ library("geiger")
 
 
 don.gheatmap <- function(p, data, offset=0, width=1, low="green", high="red",
-                     color="white", colnames=TRUE, colnames_position="bottom", font.size=4) {
+                     color="white", colnames=TRUE, colnames_position="bottom", font.size=4,font.angle=90) {
 
     colnames_position %<>% match.arg(c("bottom", "top"))
     variable <- value <- lab <- y <- NULL
@@ -56,8 +56,8 @@ don.gheatmap <- function(p, data, offset=0, width=1, low="green", high="red",
         }
         p2 <- p2 + geom_text(data=mapping,
                              aes(x=to, label=from),
-                             angle=90,
-                             hjust=1,
+                             angle=font.angle,
+                             hjust=ifelse(font.angle==0,0.5,1),
                              vjust=0.5,
                              y=y, size=font.size, fontface="italic",inherit.aes = FALSE)
     }
@@ -73,7 +73,7 @@ don.gheatmap <- function(p, data, offset=0, width=1, low="green", high="red",
     return(p2)
 }
 
-plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offset.ratio=0.5,min.fpkm=NA,subtree=NULL,vp=NULL,fontsize=3,width=1.5,axis.text.size=5) {
+plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offset.ratio=0.5,min.fpkm=NA,subtree=NULL,vp=NULL,fontsize=3,width=1.5,axis.text.size=5,axis.text.angle=90) {
     gene.tree$tip.label <- gsub("^(ENS.+)\\1$","\\1",gene.tree$tip.label)
     gene.tree.table[,species:=capfirst(gsub("_"," ",species))]
     gene.tree.table[,short.species:=gsub("^(.)[^ ]* +([^ ]{2})[^ ]*","\\1.\\2.",species)]
@@ -138,7 +138,10 @@ plot.tree.matrix <- function(gene.tree,gene.tree.table,gene.tree.expression,offs
     g1 <-
         (don.gheatmap(p,data=gene.tree.expression.matrix,
                       width=width,
-                      offset=max(p$data$x)*offset.ratio,font.size=axis.text.size,colnames=TRUE)
+                      offset=max(p$data$x)*offset.ratio,
+                      font.size=axis.text.size,
+                      font.angle=axis.text.angle,
+                      colnames=TRUE)
[…]

### 2016-03-10 data/human_mouse_lineage_specific.R
add source code to build human/mouse linage specific
This code uses a nested anova

@@ -1,0 +1,99 @@
+library("data.table")
+library("reshape2")
+library("org.Hs.eg.db")
+library("lme4")
+
+args <- c("combined_fpkm_per_sample","housekeeping_genes_superset","human_mouse_lineage_specific")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+load(args[1])
+load(args[2])
+
+species <- c("ateles fusciceps", "homo sapiens", "pan paniscus", "mus musculus", 
+             "spalax galili", "spalax carmeli", "bos taurus", "ovis aries", 
+             "sus scrofa", "equus caballus", "canis familiaris", "loxodonta africana", 
+             "dasypus novemcinctus", "monodelphis domestica")
+
+
+placenta.lineage.specific <-
+    data.table(human_name=combined.fpkm[FPKM >= 100 & !is.na(human_name),unique(human_name)])
+
+### ignore genes which are housekeeping genes
+placenta.lineage.specific <-
+    placenta.lineage.specific[!(human_name %in% housekeeping.genes.superset[,gene_short_name]),]
+setkey(combined.fpkm,"human_name")
+
+calculate.mann.whitney.test <- function(x,g1,g2){
+    data.subset <- combined.fpkm[x,list(species,FPKM)]
+    data.subset[,this.group:=NA]
+    data.subset[species %in% g1,this.group:=TRUE]
+    data.subset[species %in% g2,this.group:=FALSE]
+    data.subset <- data.subset[!is.na(this.group)]
+    p.value <- NA;
+    try({
+        this.sum <-
+            summary(lmer(FPKM~1+this.group+(1|species),
+                         data.subset))
+        p.value <- pt(this.sum$coefficients["this.groupTRUE",
+                                            "t value"],
+                      df=this.sum$ngrps,lower.tail=FALSE)
+    },silent=TRUE);
+    return(p.value)
+}
+
+
+calculate.fc <- function(x,g1,g2){
+    data.subset <- combined.fpkm[x]
+    fc <- NA
+    try({
+        first <- combined.fpkm[x][species %in% g1,FPKM]
+        second <- combined.fpkm[x][species %in% g2,FPKM]
+        fc <- mean(as.numeric(first))/mean(as.numeric(second))
+        if (fc < 1) {fc <- -1/fc}
+        if (any(!is.finite(fc))) {
+            fc <- max(c(mean(as.numeric(first)),mean(as.numeric(second)),na.rm=TRUE))
+        }},
[…]

### 2016-03-10 data/placenta_lineage_specific.R
fix nanospalax/spalax in linage specific

@@ -25,7 +25,7 @@ placenta.lineage.specific <-
     placenta.lineage.specific[!(human_name %in% housekeeping.genes.superset[,gene_short_name]),]
 
 species <- c("ateles fusciceps", "homo sapiens", "pan paniscus", "mus musculus", 
-             "nannospalax galili", "spalax carmeli", "bos taurus", "ovis aries", 
+             "spalax galili", "spalax carmeli", "bos taurus", "ovis aries",
              "sus scrofa", "equus caballus", "canis familiaris", "loxodonta africana", 
              "dasypus novemcinctus", "monodelphis domestica")
 

### 2016-03-10 manuscript/manuscript.Rnw
add changes Michael requested

@@
 
+  \DLA{Complete rewrite abstract to make the points that we have found
+    stronger and less weak.}
+
 % We identified significant shifts in gene expression along specific
@@
 reconstructed as possessing a relatively invasive hemochorial
-placenta, in which the maternal and fetal blood supply is separated by
-one tissue layer, as in humans and mouse
+placenta, as in humans and mouse
 \cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
-\DLA{To discuss maternal/fetal split with derek}
 
@@
 in \cow, genes related to the thymus-specific serine protease
-\textit{PRSS16} in \textit{Loxodona africana}, and the expansion of the growth hormone
+\textit{PRSS16} in \textit{Loxodona africana} (in this work),
+ and the expansion of the growth hormone
 and chorionic gonadotropin beta genes in hominoid primates
@@
 be slower than \human
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}; however,
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. However,
 large-scale expression studies of great ape placentas have not been
@@
 the evolution of gene expression
-\cite{Brawand.ea2011:evolutionofgeneexpression}. Here we use RNA-Seq
+\cite{Brawand.ea2011:evolutionofgeneexpression}.
+
+Here we use RNA-Seq
 technology to sequence and quantify the mRNA environment of mammalian
@@
 
-\subsection{Correlation of Tissue Specific Index with Lineage Specific Genes}
-
-To evaluate the extent of tissue specificity of each gene, we
-calculated the \ac{TSI} based on human RNASeq data in 3267 samples in
-126 tissues the Roadmap Epigenomics
-\cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping} and GTEx projects
-\cite{Gtexconsortium.2015:Thegenotypetissueexpression} as well
-additional placenta
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} and uterus
-\cite{Chan.ea2014:Assessmentofmyometrialtranscriptome} samples from
-SRA. The \ac{TSI} measures the extent to which a specific gene has
-predominant expression in only one tissue
-\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}. A gene
-expressed in only one tissue would have a \ac{TSI} of 1, whereas a
-gene expressed in all tissues would have a \ac{TSI} of 0.
+% \subsection{Correlation of Tissue Specific Index with Lineage Specific Genes}
+% 
+% To evaluate the extent of tissue specificity of each gene, we
+% calculated the \ac{TSI} based on human RNASeq data in 3267 samples in
+% 126 tissues the Roadmap Epigenomics
+% \cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping} and GTEx projects
+% \cite{Gtexconsortium.2015:Thegenotypetissueexpression} as well
+% additional placenta
+% \cite{Hughes.ea2015:Evaluatingintra-andinter-individual} and uterus
+% \cite{Chan.ea2014:Assessmentofmyometrialtranscriptome} samples from
+% SRA. The \ac{TSI} measures the extent to which a specific gene has
+% predominant expression in only one tissue
+% \cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}. A gene
+% expressed in only one tissue would have a \ac{TSI} of 1, whereas a
+% gene expressed in all tissues would have a \ac{TSI} of 0.
 
@@
 
-<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("placenta_lineage_specific")))>>=
+<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("placenta_lineage_specific","human_mouse_lineage_specific")))>>=
 load("../data/placenta_lineage_specific")
+load("../data/human_mouse_lineage_specific")
 source("/home/don/projects/xtable/pkg/R/print.xtable.R")
 
-placenta.lineage.specific.table <- function(analysis,output=TRUE) {
+placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=5) {
     temp.table <- 
-        head(placenta.lineage.specific.analysis[analysis.name==analysis &
+        head(lineage.table[analysis.name==analysis &
                            p.value <= 0.05][order(p.value),
                                             list(human_name,p.value,fdr,fc),],
-             n=5)
+             n=members)
     temp.table <- data.table(temp.table)
@@
                      display=c("s","s","s","e","g","f"),
+                     align=c("c","c","p{2in}","r","r","r")
                      ),
@@
   \footnotesize
+<<human.lineage.specific,echo=FALSE,results="asis">>=
+hum.all.lin.spec <-
+    placenta.lineage.specific.table("hum.all",
+                                    lineage.table=human.mouse.linage.specific.analysis,
+                                    members=40)
+setkey(hum.all.lin.spec,"Human Symbol")
+@ 
+\caption{Genes whose expression changed significantly in \human
+  as compared to all other analyzed species.}
+\label{tab:human_lineage_specific}
+\end{table}
+
+\begin{table}
+  \footnotesize
 <<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
@@
 ($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
-expression is on the order of the expression of expression of
+expression is on the order of the expression of
 \textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
@@
 \cref{tab:all_homo_sapiens_placenta}).
+
+\DLA{Discuss pre-eclampsia specific genes here to make this point stronger}
 
@@
 
-We examined the expression of three rodentia, \mouse, \carmeli, and
-\galili{}. We lacked sufficient power to detect any significant
-changes on the Rodentia lineage in comparison to all other analyzed
-species (Supplemental \cref{tab:rodentia_lineage_specific}).
+We examined the expression of three rodents, \mouse, \carmeli, and
+\galili{}. We did not detect any significant changes on the Rodentia
+lineage in comparison to all other analyzed species (Supplemental
+\cref{tab:rodentia_lineage_specific}), likely because of a lack power
+to overcome multiple testing.
 
 \subsubsection{\mouse Placenta Transcriptome}
+
+\begin{table}
+<<mouse.lineage.specific,echo=FALSE,results="asis">>=
+mou.all.lin.spec <-
+    placenta.lineage.specific.table("mou.all",
+                                    lineage.table=human.mouse.linage.specific.analysis)
+setkey(mou.all.lin.spec,"Human Symbol")
+@ 
+\caption{Genes whose expression changed significantly in \mouse
+  as compared to all other analyzed species.}
+\label{tab:mouse_lineage_specific}
+\end{table}
 
@@
 
+\DLA{Make the discussion of pre-eclampsia clearer and more defined;
+  possibly include some of the discuss below. Also, this is not
+  necessarily a shift on the human line.}
+
 % I haven't actually done this analysis, so commented out

### 2016-03-10 manuscript/supplemental_material.Rnw
fix species table and add more expression figures

@@
 <<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
-opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
+opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.65\\textheight",out.extra="keepaspectratio")
 opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")
@@
 \begin{table}
-  \begin{tabular}{c p{1.5in} c p{2in}}
+  \raggedright
+  \begin{tabular}{c >{\raggedright}p{1.5in} c >{\raggedright\arraybackslash}p{2in}}
     \toprule
-    Species                       & Collector & Experiment Number & Sample Number(s) \\
+    Species                       & Collector & Experiment & Sample(s) \\
     \midrule
@@
     \textit{Loxodonta africana}   & Kurt Benirschke & & \\
-    \textit{Homo sapiens}         & Nandor Gabor, \citet{Hughes.ea2015:Evaluatingintra-andinter-individual} & GSE66622 
+    \textit{Homo sapiens}         & \citet{Hughes.ea2015:Evaluatingintra-andinter-individual} & GSE66622 
                                              & SRR1850945,
@@
     \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & & \\
-    \textit{Mus musculus}         & Juan Gonzalez, \citet{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
+    \textit{Mus musculus}         & \citet{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
                                      & SRR243594, 
@@
     \textit{Spalax carmeli}       & Anne Dupressoir, Eviator Nevo & GSE68763 & SRR2016467 \\
-    \textit{Spalax galili}        & Eviator Nevo                  & GSE68763 & SRR2016467 \\
+    \textit{Spalax galili}        & Eviator Nevo                  & GSE68763 & SRR2016466 \\
     \textit{Sus scrofa}           & \citet{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 \\
@@
 \end{table}
+
+\clearpage
 
@@
 
-\begin{figure}
+\begin{figure}[hb]
 <<transcriptome.mapping.figure,fig.width=8,fig.height=8,echo=FALSE>>=
@@
 
+\begin{figure}[hb]
+<<adam12.evolution.tree,echo=FALSE,out.height="0.6\\textheight",fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="ADAM12" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,min.fpkm=5,
+                 axis.text.size=5,
+                 fontsize=8
+                 )
+@ 
+\caption{\genetreeexplanation{\textit{ADAM12}}}
+\label{fig:adam12_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<kiss1.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="KISS1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,min.fpkm=5,
+                 axis.text.size=5,
+                 axis.text.angle=0,
+                 fontsize=5
+                 )
+@ 
+\caption{\genetreeexplanation{\textit{KISS1}}}
+\label{fig:kiss1_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<gpc3.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="GPC3" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,min.fpkm=5,
+                 axis.text.size=5,
+                 fontsize=8
+                 )
+@ 
+\caption{\genetreeexplanation{\textit{GPC3}}}
+\label{fig:gpc3_evolution_tree}
+\end{figure}
+
 \begin{figure}
@@
 @
+
+\begin{table}[hb]
+  \footnotesize
+<<human.lineage.specific,echo=FALSE,results="asis">>=
+hum.all.lin.spec <- placenta.lineage.specific.table("hum.all")
+setkey(hum.all.lin.spec,"Human Symbol")
+@ 
+\caption{Genes whose expression changed in \human
+  as compared to all other analyzed species.}
+\label{tab:human_lineage_specific}
+\end{table}
 

### 2016-03-11 manuscript/supplemental_material.Rnw
add the GSE numbers to the supplement and the manuscript

@@
     \midrule
-    \textit{Ateles fusciceps}     & Kurt Benirschke & & \\
-    \textit{Bos taurus}           & Bob Kreft & & \\
-    \textit{Canis familiaris}     & Dalen Agnew & & (Bounty Hubler) \\
-    \textit{Dasypus novemcinctus} & Richard Truman, Vincent J. Lynch& & \\
+    \textit{Ateles fusciceps}     & Kurt Benirschke & GSE79121 & \\
+    \textit{Bos taurus}           & Bob Kreft & GSE79121 & \\
+    \textit{Canis familiaris}     & Dalen Agnew & GSE79121 & (Bounty Hubler) \\
+    \textit{Dasypus novemcinctus} & Richard Truman, Vincent J. Lynch& GSE79121 & \\
     \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & GSM1139275 & SRR850299,  SRR850300 \\
-    \textit{Loxodonta africana}   & Kurt Benirschke & & \\
+    \textit{Loxodonta africana}   & Kurt Benirschke & GSE79121 & \\
     \textit{Homo sapiens}         & \citet{Hughes.ea2015:Evaluatingintra-andinter-individual} & GSE66622 
@@
                                                SRR1851012 \\
-    \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & & \\
+    \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & GSE79121 & \\
     \textit{Mus musculus}         & \citet{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
@@
     \textit{Ovis aries}           & \citet{Archibald.ea2010:sheepgenomereferencesequence} & ERP005642 & ERR489144, ERR489145\\
-    \textit{Pan paniscus}         & William Mintz Fields & & (Teco)\\
+    \textit{Pan paniscus}         & William Mintz Fields & GSE79121 & (Teco) \\
     \textit{Spalax carmeli}       & Anne Dupressoir, Eviator Nevo & GSE68763 & SRR2016467 \\

### 2016-03-17 manuscript/supplemental_material.Rnw
add more ple-eclampsia genes

@@
 
+\subsubsection{Pre-eclampsia genes}
+
 \begin{figure}[hb]
@@
 \label{fig:kiss1_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<pappa.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PAPPA" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,min.fpkm=5,
+                 axis.text.size=5,
+                 axis.text.angle=0,
+                 fontsize=5
+                 )
+@ 
+\caption{\genetreeexplanation{\textit{PAPPA}}}
+\label{fig:pappa_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<tgfb1.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="TGFB1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 subtree=c("ENSMODG00000008630","ENSSSCG00000026296"),
+                 gene.tree.table,gene.tree.expression,min.fpkm=5,
+                 axis.text.size=5,
+                 fontsize=5
+                 )
+@ 
+\caption{\genetreeexplanation{\textit{TGFB1}}}
+\label{fig:tgfb1_evolution_tree}
 \end{figure}

### 2016-03-22 manuscript/supplemental_material.Rnw
add SRX numbers

@@
 
-\DLA{Add in SRR and GSE numbers for everything here once I have them.}
-
 \begin{table}
@@
     \midrule
-    \textit{Ateles fusciceps}     & Kurt Benirschke & GSE79121 & \\
-    \textit{Bos taurus}           & Bob Kreft & GSE79121 & \\
-    \textit{Canis familiaris}     & Dalen Agnew & GSE79121 & (Bounty Hubler) \\
-    \textit{Dasypus novemcinctus} & Richard Truman, Vincent J. Lynch& GSE79121 & \\
+    \textit{Ateles fusciceps}     & Kurt Benirschke & GSE79121 & SRX1629201 \\
+    \textit{Bos taurus}           & Bob Kreft & GSE79121 & SRX1629203 \\
+    \textit{Canis familiaris}     & Dalen Agnew & GSE79121 & SRX1629204 \\
+    \textit{Dasypus novemcinctus} & Richard Truman, Vincent J. Lynch& GSE79121 & SRX1629200 \\
     \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & GSM1139275 & SRR850299,  SRR850300 \\
-    \textit{Loxodonta africana}   & Kurt Benirschke & GSE79121 & \\
+    \textit{Loxodonta africana}   & Kurt Benirschke & GSE79121 & SRX1629205 \\
     \textit{Homo sapiens}         & \citet{Hughes.ea2015:Evaluatingintra-andinter-individual} & GSE66622 
@@
                                                SRR1851012 \\
-    \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & GSE79121 & \\
+    \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & GSE79121 & SRX1629207 \\
     \textit{Mus musculus}         & \citet{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
@@
     \textit{Ovis aries}           & \citet{Archibald.ea2010:sheepgenomereferencesequence} & ERP005642 & ERR489144, ERR489145\\
-    \textit{Pan paniscus}         & William Mintz Fields & GSE79121 & (Teco) \\
+    \textit{Pan paniscus}         & William Mintz Fields & GSE79121 & SRX1629202 \\
     \textit{Spalax carmeli}       & Anne Dupressoir, Eviator Nevo & GSE68763 & SRR2016467 \\

### 2016-03-22 literature_mining/human_lineage_genes.R
add preeclampsia and human lineage to lit search

@@ -1,0 +1,11 @@
+library("data.table")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+load(args[1])
+write.table(human.mouse.lineage.specific.analysis[analysis.name=="hum.all" & fdr <= 0.05 & abs(fc) > 2,human_name],
+            quote=FALSE,
+            sep="",
+            col.names=FALSE,
+            row.names=FALSE,
+            file=args[length(args)])

### 2016-03-22 manuscript/common_latex.tex
add optional argument for FPKM cutoff

@@
 
-\newcommand{\genetreeexplanation}[1]{Gene tree and heatmap of
-  expression of genes related to #1. The gene tree was obtained from
+\newcommand{\genetreeexplanation}[2][1]{Gene tree and heatmap of
+  expression of genes related to #2. The gene tree was obtained from
   Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes from species
-  not included in this study or having expression by FPKM less than 1
+  not included in this study or having expression by FPKM less than #1
   were excluded from the plotted tree. Heatmap cells in white are from

### 2016-03-22 manuscript/manuscript.Rnw
add the GSE numbers to the supplement and the manuscript
add PE specific genes

@@
 \url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
-sequences used are deposited in SRA, and can be obtained using
-\DLA{add GSE label for the experiment}.
+sequences used are deposited in SRA, and can be obtained under the
+series
+\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}.
 
@@
 
+### Pre-eclampsia genes FSTL3 EPAS1 ADAM12 GPC3 TGFB1 PAPPA
+
+pe.expression.nozero <-
+    gene.expression.nozero(c("ADAM12","CRH","KISS1","PAPPA"))
+
+pe.exp.long <- melt(log10(pe.expression.nozero+1)[rev(1:nrow(pe.expression.nozero)),])
+colnames(pe.exp.long)[1:3] <- c("gene","species","fpkm")
+
 igf.exp.long$type <- "IGF Related"
@@
 hor.exp.long$type <- "Hormone"
-
-### scalerange <- range(c(igf.exp.long$fpkm,gal.exp.long$fpkm,hor.exp.long$fpkm),na.rm=TRUE)
-### gradientends <- scalerange + rep(c(0,100,200), each=2)
-### colorends <- c("white", "red", "white", "green", "white", "blue")
-### 
-### gal.exp.long$fpkm <- gal.exp.long$fpkm+100
-### hor.exp.long$fpkm <- hor.exp.long$fpkm+200
+pe.exp.long$type <- "PE"
 
@@
           gal.exp.long,
-          hor.exp.long)
+          hor.exp.long,
+          pe.exp.long
+          )
 combined.long$species <-
@@
     list(geom_tile(aes(fill = fpkm), colour = "white"),
-##         scale_fill_gradientn(colours = colorends, values = rescale(gradientends)),
          scale_fill_gradient(low = grad.def[["low"]],
@@
 @ 
-\caption{Expression of Hormone Genes, Galectins, and IGF related genes}
+\caption{Expression of 1:1 Human orthologs of Hormone Genes, Galectins,
+  genes implicated in pre-eclampsia (PE), and \textit{IGF}-related genes}
 \label{fig:hormone_genes}

### 2016-03-22 manuscript/manuscript.Rnw
fix lineage typo

@@
     placenta.lineage.specific.table("hum.all",
-                                    lineage.table=human.mouse.linage.specific.analysis,
+                                    lineage.table=human.mouse.lineage.specific.analysis,
                                     members=40)

### 2016-03-22 manuscript/supplemental_material.Rnw
use FPKM=1 for most gene trees; add PE genes

@@
 \begin{figure}[hb]
-<<adam12.evolution.tree,echo=FALSE,out.height="0.6\\textheight",fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<crh.evolution.tree,echo=FALSE,out.height="0.6\\textheight",fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="CRH" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+                 gene.tree.table,gene.tree.expression,
+                 min.fpkm=1,
+                 axis.text.size=5,
+                 axis.text.angle=0,
+                 fontsize=8
+                 )
+@ 
+\caption{\genetreeexplanation{\textit{CRH}}}
+\label{fig:crh_evolution_tree}
+\end{figure}
+
+\begin{figure}[hb]
+<<adam12.evolution.tree,echo=FALSE,out.height="0.6\\textheight",fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=5,
-                 axis.text.size=5,
-                 fontsize=8
+                 gene.tree.table,
+                 gene.tree.expression,
+                 subtree=c("ENSMODG00000010538","ENSSSCG00000010746"),
+                 min.fpkm=0.1,
+                 axis.text.size=4,
+                 fontsize=5
                  )
@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=5,
+                 gene.tree.table,gene.tree.expression,
+                 min.fpkm=0.1,
                  axis.text.size=5,
@@
 @ 
-\caption{\genetreeexplanation{\textit{KISS1}}}
+\caption{\genetreeexplanation[0.1]{\textit{KISS1}}}
 \label{fig:kiss1_evolution_tree}
@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=5,
+                 gene.tree.table,gene.tree.expression,
+                 subtree=c("ENSSSCG00000005498","ENSG00000182752"),
+                 min.fpkm=0.1,
                  axis.text.size=5,
@@
 @ 
-\caption{\genetreeexplanation{\textit{PAPPA}}}
+\caption{\genetreeexplanation[0.1]{\textit{PAPPA}}}
 \label{fig:pappa_evolution_tree}
@@
                  subtree=c("ENSMODG00000008630","ENSSSCG00000026296"),
-                 gene.tree.table,gene.tree.expression,min.fpkm=5,
-                 axis.text.size=5,
+                 gene.tree.table,gene.tree.expression,
+                 min.fpkm=1,
+                 axis.text.size=3.5,
                  fontsize=5
@@
 \begin{figure}
-<<gpc3.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<gpc3.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=5,
-                 axis.text.size=5,
-                 fontsize=8
+                 gene.tree.table,
+                 subtree=c("ENSCAFG00000018864","ENSLAFG00000025706"),
+                 gene.tree.expression,
+                 min.fpkm=1,
+                 axis.text.size=3.5,
+                 fontsize=6
                  )
@@
 
+\clearpage
+
+\subsubsection{Primate Specific Evolution}
+
 \begin{figure}
@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=5,
+                 gene.tree.table,
+                 gene.tree.expression,
+                 min.fpkm=1,
                  axis.text.size=5,
@@
 plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=1,
+                 gene.tree.table,
+                 gene.tree.expression,
+                 min.fpkm=1,
                  axis.text.size=5,

### 2016-03-22 manuscript/supplemental_material.Rnw
make per-gene tables narrower

@@
     cat(paste0("\\subsection{\\textit{",capfirst(this.species),"} Placental Gene Expression}\n\n"))
+    xt.align <- c("r","l","l","p{2.5in}","r")
+    if (this.species == "homo sapiens" || this.species == "ateles fusciceps") {
+        xt.align <- xt.align[-2]
+    }
     print(xtable(genes.expression.table(species.name=this.species),
                  caption=this.caption,
+                 align=xt.align,
                  label=paste0("tab:all_",gsub(" ","_",this.species),"_placenta")

### 2016-03-23 data/human_mouse_lineage_specific.R
use parallel in human mouse lineage specific

@@ -1,4 +1,5 @@
 library("data.table")
+library("parallel")
 library("reshape2")
 library("org.Hs.eg.db")
 library("lme4")
@@ -25,7 +26,7 @@ placenta.lineage.specific <-
 setkey(combined.fpkm,"human_name")
 
 calculate.mann.whitney.test <- function(x,g1,g2){
-    data.subset <- combined.fpkm[x,list(species,FPKM)]
+    data.subset <- combined.fpkm[x,list(species,FPKM=sum(FPKM)),by=file][!duplicated(file),]
     data.subset[,this.group:=NA]
     data.subset[species %in% g1,this.group:=TRUE]
     data.subset[species %in% g2,this.group:=FALSE]
@@ -35,6 +36,8 @@ calculate.mann.whitney.test <- function(x,g1,g2){
         this.sum <-
             summary(lmer(FPKM~1+this.group+(1|species),
                          data.subset))
+        ## this will actually underestimate p, but that's OK, because
+        ## we have a pretty large n here anyway
         p.value <- pt(this.sum$coefficients["this.groupTRUE",
                                             "t value"],
                       df=this.sum$ngrps,lower.tail=FALSE)
@@ -49,10 +52,10 @@ calculate.fc <- function(x,g1,g2){
     try({
         first <- combined.fpkm[x][species %in% g1,FPKM]
         second <- combined.fpkm[x][species %in% g2,FPKM]
-        fc <- mean(as.numeric(first))/mean(as.numeric(second))
-        if (fc < 1) {fc <- -1/fc}
+        fc <- log2(mean(as.numeric(first))/mean(as.numeric(second)))
+        ## if (fc < 1) {fc <- -1/fc}
         if (any(!is.finite(fc))) {
-            fc <- max(c(mean(as.numeric(first)),mean(as.numeric(second)),na.rm=TRUE))
+            fc <- log2(max(c(mean(as.numeric(first)),mean(as.numeric(second)),na.rm=TRUE)))
         }},
         silent=TRUE)
     return(fc)
@@ -74,19 +77,21 @@ human.mouse.lineage.specific.analysis <- list()
 for (c.n in names(comparisons)) {
     results <- data.table(placenta.lineage.specific[,human_name])
     setnames(results,"human_name")
-    results[,p.value:=apply(placenta.lineage.specific,1,
-                            calculate.mann.whitney.test,
-                            comparisons[[c.n]]$group1,
-                            comparisons[[c.n]]$group2
-                            )]
[…]

### 2016-03-23 manuscript/manuscript.Rnw
include all of the analysis members in the table

@@
     temp.table <- 
-        head(lineage.table[analysis.name==analysis &
+        lineage.table[analysis.name==analysis &
                       p.value <= 0.05][order(p.value),
-                                            list(human_name,p.value,fdr,fc),],
-             n=members)
+                                       list(human_name,p.value,fdr,fc),]
     temp.table <- data.table(temp.table)
@@
     if (output) {
-        display.table <- temp.table[order(`p (t-test)`),]
+        display.table <- temp.table[order(`p (t-test)`),][1:members,]
+
         display.table <-

### 2016-03-23 manuscript/manuscript.Rnw
discuss pre-eclampsia genes

@@
   \footnotesize
-<<human.lineage.specific,echo=FALSE,results="asis">>=
-hum.all.lin.spec <-
-    placenta.lineage.specific.table("hum.all",
-                                    lineage.table=human.mouse.lineage.specific.analysis,
-                                    members=40)
-setkey(hum.all.lin.spec,"Human Symbol")
-@ 
-\caption{Genes whose expression changed significantly in \human
-  as compared to all other analyzed species.}
-\label{tab:human_lineage_specific}
-\end{table}
-
-\begin{table}
-  \footnotesize
 <<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
@@
 
-\DLA{Discuss pre-eclampsia specific genes here to make this point stronger}
+\paragraph{Pre-eclampsia \human Lineage Specific Genes}
+
+\begin{table}
+  \footnotesize
+<<human.lineage.specific,echo=FALSE,results="asis">>=
+hum.all.lin.spec <-
+    placenta.lineage.specific.table("hum.all",
+                                    lineage.table=human.mouse.lineage.specific.analysis,
+                                    members=40)
+setkey(hum.all.lin.spec,"Human Symbol")
+@ 
+\caption{Genes whose expression changed significantly in \human
+  as compared to all other analyzed species.}
+\label{tab:human_lineage_specific}
+\end{table}
+
+Using a mixed-effects linear model to account for repeated testing of
+\human and \mouse, we identified multiple genes which are
+differentially expressed in the \human lineage in comparison to all
+other lineages (\cref{tab:human_lineage_specific}). These genes
+include multiple genes which are thought to be involved in
+pre-eclampsia. Notably, \textit{CRH}
+(FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
+FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
+\cref{fig:adam12_evolution_tree}), has been shown to be more highly
+expressed in placentas from pre-eclamptic pregnancies
+\cite{Petsas.ea2012:Aberrantexpressionofcorticotropin-releasing,Paiva.ea2011:MeasurementofmRNAtranscripts},
+where it is hypothesized to activate decidual macrophages. Likewise,
+\textit{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
+FC=\Sexpr{hum.all.lin.spec["KISS1",`Fold Change`]}, Supplemental
+\cref{fig:kiss1_evolution_tree}) also shows increased expression in
+pre-eclamptic placentas
+\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}, and is
+involved in modulating trophoblast invasion. Two other genes,
+\textit{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
+FC=\Sexpr{hum.all.lin.spec["PAPPA",`Fold Change`]} Supplemental
+\cref{fig:pappa_evolution_tree}) and \textit{ADAM12}
+(\Sexpr{hum.all.lin.spec["ADAM12",`FDR`]},
+FC=\Sexpr{hum.all.lin.spec["ADAM12",`Fold Change`]}, Supplemental
+\cref{fig:adam12_evolution_tree}) are also differentially expressed
+but are negatively correlated with pre-eclampsia
+\cite{Spencer.ea2008:ADAM12sinmaternalserum}.
 
@@
 
-### Pre-eclampsia genes FSTL3 EPAS1 ADAM12 GPC3 TGFB1 PAPPA
+### Pre-eclampsia genes ADAM12 CRH KISS1 PAPPA
 

### 2016-03-23 manuscript/supplemental_material.Rnw
include too many alignment in star plot

@@
 
-\begin{table}
+\begin{table}[hb]
+  \centering
   \begin{tabular}{c c c c}
@@
     star.logs[grepl("%",field) & 
-                  grepl("unmapped",field) & 
+                  grepl("unmapped|too many loci",field) & 
                       !grepl("too many mismatches",field),]
 star.logs.unmapped <- 
-    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads unmapped: ","",field))]
+    star.logs.unmapped[,field:=capfirst(gsub("\\% of reads (unmapped:|mapped to) ","",field))]
 pushViewport(viewport(layout=grid.layout(ncol=1,nrow=2,heights=c(2,1))))
@@
       + facet_grid(field~.,scales="free")
-      + ylab("Percent of Reads")
+      + ylab("Percentage of Reads Unmapped")
       + xlab(NULL)

### 2016-03-24 data/csv_files/species_fpkm.R
add fpkm files for a database paper and/or supplemental

@@ -1,0 +1,15 @@
+library("data.table")
+args <- c("../combined_fpkm","homo_sapiens","homo_sapiens_mean_fpkm.txt")
+args <- commandArgs(trailingOnly=TRUE)
+this.species <- gsub("_"," ",args[2])
+
+load(args[1])
+combined.fpkm[,`:=`(FPKM=NULL,
+                    FPKM_conf_lo=NULL,
+                    FPKM_conf_hi=NULL,
+                    FPKM_status=NULL
+                    )]
+write.table(file=args[length(args)],
+            combined.fpkm[species==this.species,][order(-mean_fpkm),],
+            sep="\t")
+

### 2016-03-24 manuscript/manuscript.Rnw
include too many alignment in star plot
we're not indicating the place/time of collection

@@
 \cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}). Information
-on specimen ID, place and time of collection, and collector is shown
-for each sample in Supplemental \cref{tab:srr_numbers}. In the case of
-\bonobo, \cow, and \elephant, the placenta was collected at term upon
-expulsion from the mother following delivery. Placental tissue from
-\human was collected upon delivery via a cesarean section. Near-term
-placental tissue from \dog was collected upon the death of the mother
-from a car accident. Placental tissue from \opossum was sampled at
-fetal stage 32, roughly one day prior to birth. Only the fetal portion
-of the placenta was sampled at the maternal-fetal interface (i.e.
-villous portion for \human, \bonobo, \dog, and \elephant; cotyledon
-for \cow; diffuse placenta for \opossum). All tissues were collected
-and stored in RNAlater (Life Technologies, Carlsbad, CA) until RNA
-isolation could be performed. Tissues were homogenized in Trizol, and
-total RNA was isolated following the manufacturer's instructions. RNA
-was further purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in
-conjunction with the RNase-Free DNase Set (Qiagen, Valencia, CA),
-following the manufacturer’s protocol to remove any contaminating DNA
-from the sample. Total RNA was then evaluated on a Nanodrop 1000
-(Thermo Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100
-(Santa Clara, CA) to confirm concentration and quality of the
-isolation. RNA quality was assessed using A260/A280 and A260/A230
-absorbance values.
+on specimen ID, and collector is shown for each sample in Supplemental
+\cref{tab:srr_numbers}. In the case of \bonobo, \cow, and \elephant,
+the placenta was collected at term upon expulsion from the mother
+following delivery. Placental tissue from \human was collected upon
+delivery via a cesarean section. Near-term placental tissue from \dog
+was collected upon the death of the mother from a car accident.
+Placental tissue from \opossum was sampled at fetal stage 32, roughly
+one day prior to birth. Only the fetal portion of the placenta was
+sampled at the maternal-fetal interface (i.e. villous portion for
+\human, \bonobo, \dog, and \elephant; cotyledon for \cow; diffuse
+placenta for \opossum). All tissues were collected and stored in
+RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation could
+be performed. Tissues were homogenized in Trizol, and total RNA was
+isolated following the manufacturer's instructions. RNA was further
+purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction
+with the RNase-Free DNase Set (Qiagen, Valencia, CA), following the
+manufacturer’s protocol to remove any contaminating DNA from the
+sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
+Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
+Clara, CA) to confirm concentration and quality of the isolation. RNA
+quality was assessed using A260/A280 and A260/A230 absorbance values.
 
@@
 
-\subsection{Annexin Complex}
-
-\subsection{Human Specific Expression Shifts and the Evolution of Pre-eclampsia}
+\subsection{Annexin and Fetal Immunotolerance}
+
+\DLA{Finish out this section siting Aarli work and Linke work}
+
+\subsection{\human Specific Expression Shifts and the Evolution of Pre-eclampsia}
 

### 2016-03-24 manuscript/manuscript.Rnw
explain how diamond annotation was chosen

@@
 \carmeli; italic reference in Supplemental
-\cref{tab:ref_genome_aligned_species}) were assembled 
-using Trinity v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence}
-and annotated using Diamond
-\cite{Buchfink.ea2015:Fastandsensitiveprotein} against either the
-\human or \mouse transcriptome
-(Supplemental \cref{tab:ref_genome_aligned_species}).
+\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
+v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
+using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
+either the \human or \mouse transcriptome (Supplemental
+\cref{tab:ref_genome_aligned_species}). For annotation using Diamond,
+the match with the highest bitscore was chosen as the annotation;
+transcripts where the highest scoring match had a bitscore below 40
+were left unannotated.
 % 

### 2016-03-24 manuscript/manuscript.Rnw
remove ouch discussion

@@
 in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
-  sapiens" & mean_fpkm >= 10,length(unique(human_name))]} 1:1
-orthologous genes with \ac{FPKM} $\ge 10$ were retained for future
+  sapiens" & mean_fpkm >= 100,length(unique(human_name))]} 1:1
+orthologous genes with \ac{FPKM} $\ge 100$ were retained for future
 analyses of gene expression evolution described below.
@@
 and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
-$\le 0.05$.
-
-% We used the software \ac{OUCH} in the R statistical environment to
-% investigate significant shifts in gene expression levels over
-% time\cite{Butler.2004:phlogeneticcomparativeanalysis}. \ac{OUCH}
-% institutes an \ac{OU} model of evolution, which allows for different
-% optima for quantitative trait values along user-specified branches of
-% a phylogenetic tree. The \ac{OU} model explicitly accounts for change
-% in quantitative trait values by introducing two parameters into the
-% model of stochastic Brownian motion of change in trait value along a
-% phylogenetic tree through time. These parameters are $\theta$, an
-% optimum trait value, and $\alpha$, the strength of selection
-% (interpreted as “pull” toward the optimum). In addition, the \ac{OU}
-% model can also integrate evolutionary change by assuming that each
-% user-specified set of braches has its own separate \ac{OU} parameters
-% \cite{Butler.2004:phlogeneticcomparativeanalysis}.
-% 
-% \DLA{The description of the OU process is terrible; it needs to be
-%   completely redone}
-% 
-% Various \ac{OU} branch models were tested against a model in which no
-% variation in optima ($\theta$) among branches was allowed. The
-% following branches were specified in separate models to have a
-% different optimum than all other branches: human, bonobo, mouse, dog,
-% cow, elephant, opossum, human+bonobo, Euarchontoglires, and
-% Boreoeutheria. Likelihood values were computed for each branch model
-% and compared to an \ac{OU} model with one optimum using likelihood
-% ratio tests with df=1. We filtered out genes with universally low
-% expression levels, and included only genes in which expression in at
-% least one species had an $\mathrm{FPKM}\ge10$ as described above. We
-% corrected for multiple testing using the \ac{FDR} and using the
-% Benjamini-Hochberg correction\cite{Benjamini1995:bh_procedure}. We
-% used the time-calibrated phylogenetic tree of
-% \cite{Meredith.ea2011:ImpactsofCretaceousTerrestrial} with all species
-% not used in this analyses pruned.
-% 
-% We also calculated the enrichment of functional categories in the
-% lists of genes which displayed a significant shift in expression on
-% each branch using the same methodology as in the per-species gene
-% lists \DLA{reference the tables once this analysis is complete}.
-
-% \subsection{Correlation of Tissue Specific Index with Lineage Specific Genes}
-% 
-% To evaluate the extent of tissue specificity of each gene, we
-% calculated the \ac{TSI} based on human RNASeq data in 3267 samples in
-% 126 tissues the Roadmap Epigenomics
-% \cite{Bernstein.ea2010:NIHRoadmapEpigenomicsMapping} and GTEx projects
-% \cite{Gtexconsortium.2015:Thegenotypetissueexpression} as well
-% additional placenta
-% \cite{Hughes.ea2015:Evaluatingintra-andinter-individual} and uterus
-% \cite{Chan.ea2014:Assessmentofmyometrialtranscriptome} samples from
-% SRA. The \ac{TSI} measures the extent to which a specific gene has
-% predominant expression in only one tissue
-% \cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}. A gene
-% expressed in only one tissue would have a \ac{TSI} of 1, whereas a
-% gene expressed in all tissues would have a \ac{TSI} of 0.
-
-% \subsection{Validation of Primate Expression using Spider Monkey Placenta Transcriptome}
-% 
-% To verify expression levels in human and bonobo, we also generated
-% RNA-Seq data for the placenta transcriptome of the black-headed spider
-% monkey (\textit{Ateles fusciceps}) using similar protocols discussed above.
-% Ateles is outside the clade, which includes hominids and can be used
-% to polarize expression values of highly expressed genes in Homo and
-% Pan. We sampled placental tissue from an Ateles placenta expelled from
-% the mother shortly after birth (see Supplemental Table X). We used
-% similar procedures discussed above to extract and sequence this
-% sample. [DEREK—I’m unsure how to proceed on describing this so as to
-% not step on Natalie’s toes]
-% 
-% \DLA{Complete discussion of the ateles placenta}
-
-\subsection{Sequence Deposition and Code}
+$\le 0.05$. For \human and \mouse lineage specific calculations, a
+type-III anova was calculated using
+lme4~\cite{Bates.ea2015:FittingLinearMixed-EffectsModels} with each
+species having separate random effects.
+
+\subsection{Sequences, FPKMs, and Code}
 
@@
 \url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
-sequences used are deposited in SRA, and can be obtained under the
-series
-\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}.
+sequences used are deposited in SRA, and new sequences for this work
+can be obtained under the series
+\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
+(Supplemental \cref{tab:srr_numbers}). The mean FPKM for gene in each
+species with ortholog annotations are located in the supplemental data
+files \Sexpr{english.join(paste0(gsub(" ","_",
+  star.logs[,unique(species)]),"_mean_fpkm.txt"))}.
 

### 2016-03-24 manuscript/manuscript.Rnw
add mouse lineage specific discussion

@@
 
-\begin{table}
-<<mouse.lineage.specific,echo=FALSE,results="asis">>=
-mou.all.lin.spec <-
-    placenta.lineage.specific.table("mou.all",
-                                    lineage.table=human.mouse.linage.specific.analysis)
-setkey(mou.all.lin.spec,"Human Symbol")
-@ 
-\caption{Genes whose expression changed significantly in \mouse
-  as compared to all other analyzed species.}
-\label{tab:mouse_lineage_specific}
-\end{table}
-
 \begin{table}[h]
@@
 reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.
+
+% \paragraph{\mouse Lineage Specific Genes}
+
+\begin{table}
+  \footnotesize
+<<mouse.lineage.specific,echo=FALSE,results="asis">>=
+mou.all.lin.spec <-
+    placenta.lineage.specific.table("mou.all",
+                                    lineage.table=human.mouse.lineage.specific.analysis,
+                                    members=40
+                                    )
+setkey(mou.all.lin.spec,"Human Symbol")
+@ 
+\caption{\lineagetableexplanationtypeiii{\mouse}{all other analyzed species}}
+\label{tab:mouse_lineage_specific}
+\end{table}
+
+Using a mixed-effects linear model to account for repeated testing of
+\human and \mouse, we identified multiple genes which are
+differentially expressed in the \mouse lineage in comparison to all
+other lineages (\cref{tab:mouse_lineage_specific}). Notable
+mouse-specific increases in expression include include genes like
+\textit{Slc38a4} which are involved in nutrient transport and are
+implicated in fetal growth restriction
+\cite{Sharma.ea2016:Tpbpa-Cre-mediateddeletionofTFAP2C}.
 

### 2016-03-25 manuscript/manuscript.Rnw
fix typos and make it log_2 FC

@@
 species with ortholog annotations are located in the supplemental data
-files \Sexpr{english.join(paste0(gsub(" ","_",
+files \Sexpr{english.join(paste0(gsub(" ","\\_",
   star.logs[,unique(species)]),"_mean_fpkm.txt"))}.
@@
                                                   "}")]
+        setnames(display.table,"Fold Change","$\\mathrm{log}_2$ FC")
         print(xtable(display.table,
@@
 \elephant, the expression pattern does not follow the placental
-organization (endotheliochorial for \dog and \elphant,
+organization (endotheliochorial for \dog and \elephant,
 synepitheliochorial for \cow and \sheep), so it is unlikely that

### 2016-03-25 manuscript/manuscript.Rnw
use the linage table explanation text

@@
 @ 
-\caption{Genes whose expression changed significantly in the primate
-  order as compared to all other analyzed species.}
+\caption{\linagetableexplanation{primates}}
 \label{tab:primate_lineage_specific}
@@
 @ 
-\caption{Genes whose expression changed significantly in \human
-  as compared to all other analyzed species.}
+\caption{\lineagetableexplanationtypeiii{\human}{all other analyzed species}}
 \label{tab:human_lineage_specific}
@@
 \cref{tab:rodentia_lineage_specific}), likely because of a lack power
-to overcome multiple testing.
+to overcome the multiple testing correction. Additional species and/or
+samples in each of these lineages will be necessary to detect lineage
+specific shifts.
 

### 2016-03-25 manuscript/supplemental_material.Rnw
use the linage table explanation text

@@
   \footnotesize
-<<human.lineage.specific,echo=FALSE,results="asis">>=
-hum.all.lin.spec <- placenta.lineage.specific.table("hum.all")
-setkey(hum.all.lin.spec,"Human Symbol")
-@ 
-\caption{Genes whose expression changed in \human
-  as compared to all other analyzed species.}
-\label{tab:human_lineage_specific}
-\end{table}
-
-\begin{table}[hb]
-  \footnotesize
 <<rodentia.lineage.specific,echo=FALSE,results="asis">>=
@@
 @ 
-\caption{Genes whose expression changed in the Rodentia
-  order as compared to all other analyzed species.}
+\caption{\lineagetableexplanation{the rodentia order}{all other species}}
 \label{tab:rodentia_lineage_specific}
@@
 @ 
-\caption{Genes whose expression changed in the Laurasiatheria
-  superorder as compared to all other analyzed species.}
+\caption{\lineagetableexplanation{the laurasiatheria
+    superorder}{all other species}}
 \label{tab:laurasiatheria_lineage_specific}
@@
 @ 
-\caption{Genes whose expression changed between the
-  Euarchontoglires and Laurasiatherian superorder.}
+\caption{\linagetableexplanation{ the euarchontoglires superorder}{ the laurasiatherian superorder}}
 \label{tab:laurasiatheria_euarchontoglires_lineage_specific}

### 2016-03-25 manuscript/common_latex.tex
add lineage table explanation to common_latex

@@
 
+\newcommand{\lineagetableexplanation}[3]%
+[A two-sided t test was used to calculate p values.]{%
+  Genes whose expression changed in #2 as compared to #3. #1 FDR
+  is the False Discovery Rate as estimated by the Benjamini and
+  Hochberg procedure \cite{Benjamini1995:bh_procedure}.
+  $\mathrm{log}_2$ FC is the base-2 logarithm of the fold change from
+  #2 to #3 .}
+
+\newcommand{\lineagetableexplanationtypeiii}[2]{%
+  \lineagetableexplanation[A nested type-III anova was used to
+  calculate p values.]{#1}{#2}}
+
 %%% Local Variables:

### 2016-03-25 data/placenta_lineage_specific.R
calculate log2_fc

@@ -61,9 +61,8 @@ calculate.t.test <- function(x,g1,g2){
 calculate.fc <- function(x,g1,g2){
     fc <- NA
     try({fc <- mean(as.numeric(x[g1]))/mean(as.numeric(x[g2]))
-        if (fc < 1) {fc <- -1/fc}
         if (any(!is.finite(fc))) {
-            fc <- max(c(mean(as.numeric(x[g1])),mean(as.numeric(x[g2])),na.rm=TRUE))
+            fc <- log2(max(c(mean(as.numeric(x[g1])),mean(as.numeric(x[g2])),na.rm=TRUE)))
         }},
         silent=TRUE)
     return(fc)

### 2016-03-25 literature_mining/mouse_lineage_genes.R
add rules to make mouse lineage genes for literature search

@@ -1,0 +1,11 @@
+library("data.table")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+load(args[1])
+write.table(human.mouse.lineage.specific.analysis[analysis.name=="mou.all" & fdr <= 0.05 & abs(fc) > 2,human_name],
+            quote=FALSE,
+            sep="",
+            col.names=FALSE,
+            row.names=FALSE,
+            file=args[length(args)])

### 2016-03-25 manuscript/manuscript.Rnw
add FDR cutoff and ordering by FC for genes

@@
 
-placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=5) {
+placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=5,orderbyfc=FALSE,fdr.cutoff=1) {
     temp.table <- 
@@
     if (output) {
-        display.table <- temp.table[order(`p (t-test)`),][1:members,]
+        display.table <- temp.table[`FDR`<=fdr.cutoff,][order(`p (t-test)`),]
+        if (orderbyfc) {
+            display.table <- temp.table[order(-abs(`Fold Change`)),]
+        }
+        display.table <- display.table[1:members,]
 
@@
                                     lineage.table=human.mouse.lineage.specific.analysis,
-                                    members=40)
+                                    members=40,
+                                    orderbyfc=TRUE,
+                                    fdr.cutoff=0.001
+                                    )
 setkey(hum.all.lin.spec,"Human Symbol")
@@
                                     lineage.table=human.mouse.lineage.specific.analysis,
-                                    members=40
+                                    members=40,
+                                    orderbyfc=TRUE,
+                                    fdr.cutoff=0.001
                                     )
@@
 differentially expressed in the \mouse lineage in comparison to all
-other lineages (\cref{tab:mouse_lineage_specific}). Notable
+other lineages (\cref{tab:mouse_lineage_specific}). \mouse
+\textit{Gkn2} (FDR=\Sexpr{mou.all.lin.spec["GKN2",`FDR`]},
+FC=\Sexpr{mou.all.lin.spec["GKN2",`Fold Change`]}) has been shown to
+have a pattern of expression distinct from the \human pattern, and has
+been suggested to be important for placental development
+\cite{Fahlbusch.ea2015:Trophoblastexpressiondynamicsof}. Notable
 mouse-specific increases in expression include include genes like
-\textit{Slc38a4} which are involved in nutrient transport and are
-implicated in fetal growth restriction
-\cite{Sharma.ea2016:Tpbpa-Cre-mediateddeletionofTFAP2C}.
+\textit{Slc38a4} (FDR=\Sexpr{mou.all.lin.spec["SLC38A4",`FDR`]},
+FC=\Sexpr{mou.all.lin.spec["SLC38A4",`Fold Change`]}) and \textit{Afp}
+(FDR=\Sexpr{mou.all.lin.spec["AFP",`FDR`]},
+FC=\Sexpr{mou.all.lin.spec["AFP",`Fold Change`]}) which are implicated
+in fetal growth restriction
+\cite{Sharma.ea2016:Tpbpa-Cre-mediateddeletionofTFAP2C} and placenta
+abnormalities in humans
+\cite{Waller.ea1996:associationbetweenmaternalserum}, and
+\textit{Cnx2} (FDR=\Sexpr{mou.all.lin.spec["CNX2",`FDR`]},
+FC=\Sexpr{mou.all.lin.spec["CNX2",`Fold Change`]}) which is critical
+for the exchange of metabolites in the placenta labyrinth
+\cite{Gabriel.ea1998:Transplacentaluptakeofglucose}.
 

### 2016-03-25 manuscript/manuscript.Rnw
remove comparative evolution section
fix link to crh figure
ditch opossum shift
Completely rewrite conclusion

@@
 FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
-\cref{fig:adam12_evolution_tree}), has been shown to be more highly
+\cref{fig:crh_evolution_tree}), has been shown to be more highly
 expressed in placentas from pre-eclamptic pregnancies
@@
 
-% \textit{Ateles fuciceps} shows enrichment in extracellular
-% matrix and structural molecular activity (Supplemental
-% \cref{tab:go_terms_enriched_ateles_fusciceps}). \textit{Homo sapiens}
-% placentas are enriched in female pregnancy, estrogen biosynthesis,
-% reproductive process, hormone aactivity, oxygen transporter activity,
-% and haptoglobin binding (Supplemental
-% \cref{tab:go_terms_enriched_homo_sapiens}). \textit{Pan paniscus}
-% placentas are enriched in clusters related to stress, apoptosis, and
-% response to abiotic stimuli. \textit{Mus musculus} placentas are
-% enriched in cell junction, vesicle transport, in utero embryonic
-% development, and many cytoskeletal processes. \textit{Spalax
-%   galili} placentas are enriched in extracellular matrix organization,
-% integrin binding, and cell death.
-% 
-% Cow shows only four
-% significantly enriched clusters; however many genes highly expressed
-% in the cow are ruminant-specific placenta genes and are not yet
-% annotated to specific GO categories. Dog shows enrichment in genes
-% associated with RNA, as well as homeostasis, glycoproteins, lyosome,
-% and calcium binding. Elephant shows enrichment in genes associated
-% with RNA, coagulation and calcium binding. Opossum shows enrichment in
-% genes involved in apoptosis, oxidative stress, mitochondrion, hormone
-% metabolism, and lipids.
-
-% Not sure if this is actually worthwhile; I've mentioned some of this
-% already for some of the lineages
-
-% \subsection{Comparative Analysis of Gene Expression}
-% 
-% We performed a comparative analysis of gene expression by
-% reconstructing the evolution of expression values and detecting
-% significant shifts in expression along specific lineages. Likelihood
-% values for each model for each of the X genes reconstructed, as well
-% as likelihood ratio tests, P- and Q-values (after correcting for
-% multiple testing) are shown in Supplemental Table X. We identified
-% between 4 (Boreoeutheria) and 550 (Mouse) significant shifts (Q <
-% 0.01) in expression on specific lineages (Table 10). The top 10 genes
-% with significant shifts (if applicable) per lineage are shown in Table
-% 10 and all genes for each branch are shown in Supplemental Tables
-% X-XX. Genes that showed shifts on the human lineage were
-% overwhelmingly identified as placenta-specific by the
-% tissue-specificity index (TSI), with 56 out of 279 genes (20\%) with
-% both TSI>0.75 and placenta as one the top two tissues with the highest
-% expression. Unsurpisingly, this was not the case with other lineages,
-% since we used expression values from a human-derived microarray to
-% generate TSI values for each gene. However, this does underscore the
-% dynamic nature of expression patterns in the placenta and indeed all
-% tissues across mammals.
-% 
-% We then performed DAVID functional annotation clustering analyses on
-% each list of genes from each lineage. The top three clusters for each
-% lineage are shown in Table 10 and a full list of clusters in
-% Supplemental Table X. The top enrichment cluster in the human lineage
-% is blood vessel development, followed by extracellular matrix and
-% metal ion binding (Table 10). Other significant clusters include those
-% related to cell communication/signaling, cytokines, defense response,
-% tryptophan metabolism, apoptosis, and skeletal system development
-% (Supplemental Table X). The bonobo lineage was enriched in only three
-% clusters: apoptosis, Toll-like receptor signalling and apoptosis. The
-% mouse displayed enrichment in clusters involved in epigenetic
-% modifications such as methyltransferase activity and histone
-% modification, as well embryonic morphogeneisis, lipid biosynthesis,
-% among others (Supplemental Table X). The top enriched cluster in the
-% cow lineage was the response to vitamin A, as well as many clusters
-% related to the immune system. The dog showed enrichment in skeletal
-% system development, as well as lipid metabolic processes and blood
-% circulation. The elephant lineage had only one enriched cluster:
-% carbohydrate binding. Genes that showed enrichment on the opossum
-% lineage (essentially representing the difference between placental
-% mammals and marsupials) were clustered in many categories not seen in
-% the other lineages, including clusters related to the mitochondria,
-% urea cycle, peroxisome, glycolysis, and transport across the cell
-% membrane (Table 10 and Supplemental Table X).
-
 \clearpage
@@
 
-% Genes that show a significant shift on the \opossum lineage include
-% functional annotation clusters that correspond to active generation of
-% energy, transport, and lipid breakdown via the peroxisome (Table 10;
-% Supplemental Table X). The choriovitelline placenta of both didelphids
-% and the tammar wallaby are known to be involved in glycolysis and
-% protein synethsis (Freyer et al., 2003). This implies that the
-% choriovitelline placenta in marsupials is an active site of metabolism
-% and molecular synthesis, as well as the transport of nutrients from
-% the maternal tissues. This is a distinct difference that is not seen
-% in significant shifts on other lineages, and may be due to the
-% formation of the placenta in the opossum from the yolk sac. In
-% addition, the urea cycle is a highly enriched functional annotation
-% cluster which directly correlates with the high levels of urea in the
-% Monodelphis placenta (Freyer et al., 2007). This is potentially a
-% Monodelphis-specific characteristic as discussed above.
-
 The \opossum placenta shows expression of genes that either do not
@@
 it is exceedingly rare in these species
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is possible
-that a comparative evolutionary approach to gene expression can
-identify genes that have divergent expression patterns from other
-mammals as well as association with pre-eclampsia.
-
-Functional annotation clusters significantly enriched on the human
-lineage (blood vessel development, extracellular matrix, tube
-morphogenesis) seem to suggest involvement in remodeling of tissues
-and increased blood flow, both of which may correlate with the
-invasion of the trophoblast (\cref{tab:go_terms_all_species}). In
-addition, clusters such as cytokines and defense response
-(\cref{tab:go_terms_all_species}) also indicate an a significant
-change in expression of the immune response that may be involved in
-immunosupression at the maternal-fetal boundary to further allow
-invasion of fetal tissues.
-
-\DLA{Make the discussion of pre-eclampsia clearer and more defined;
-  possibly include some of the discuss below. Also, this is not
-  necessarily a shift on the human line.}
-
-% I haven't actually done this analysis, so commented out
-
-% Many genes that both show expression shifts on the human lineage and
-% have very high expression (FPKM>200) are directly related to the
-% highly invasive nature of the human placenta. For example, GPC3,
-% glypican 3, is an imprinted gene and cell growth regulator, associated
-% with Simpson-Golabi-Behmel fetal overgrowth syndrome, and involved in
-% intrauterine growth restriction (Monk and Moore, 2004; Pilia et al.,
-% 1996). GPC3 has the highest expression in humans (Table 2), but is
-% also expressed in other mammals to a lesser degree, especially those
-% with invasive hemochorial placentas like bonobos (\ac{FPKM}= 199.924),
-% spider monkeys (\ac{FPKM}= 96.24), and the mouse (\ac{FPKM}= 80.7545)
-% (Supplemental Tables X-XX). ADAM12 increases the availability of
-% insulin-like growth factors that lead to the promotion of cell growth
-% and placental invasion (Laigaard et al., 2005). (Hou et al., 2009)
-% found evidence of adaptive evolution in ADAM12 on the primate lineage,
-% and like GPC3 it is expressed in the bonobo (\ac{FPKM}= 46.1553), spider
-% monkey (\ac{FPKM}=30.41) and mouse (\ac{FPKM}=2.25391) as well, although to a
-% lesser degree (Supplemental Tables X-XX). RHOBTB3 is imprinted and
-% associated with trophoblast invasion as well (Bilban et al., 2010;
-% Yuen et al., 2011). In addition, one other gene, KISS1, shows
-% exceptional abundance in the human placenta (Table 2); however due to
-% its absence in elephant Ensmebl gene annotations, it was not part of
-% the 1:1 orthologues examined in the comparative study. Here, it is
-% only expressed in the term placentas of humans (Table 2) and bonobos
-% (\ac{FPKM}=17.6394; Supplemental Table X). KISS1 is thought to be involved
-% in the regulation of trophoblast invasion, and shows highest
-% expression in the first trimester of pregnancy (Bilban et al., 2004;
-% Reynolds et al., 2009).
-% 
-% In addition, several other genes that show expression shifts on the
-% human lineage also have major effects on placental function. Genes
-% such as TXK and EBI3 are intimately involved in the immune system in
-% the placenta. TXK is involved in natural killer cell differentiation
-% (Bezman et al., 2012). EBI3 is implicated in natural killer cell
-% function an maternal-fetal immunomodulation in the syncytiotrophoblast
-% (Devergne et al., 2001). KYNU and KMO are interacting genes that are
-% involved in tryptophan metabolism, which may be involved in
-% immunosupression and the prevention of fetal rejection (Takikawa,
-% 2005). THSD7A is a thrombospondin, which inhibits endothelial cell
-% migration and tube morphogenesis (Wang et al., 2010).
-% 
-% Many genes that show evidence of a shift on the human lineage show
-% evidence of implication in pre-eclampsia, including ADAM12 (Gack et
-% al., 2005; Laigaard et al., 2005) MAOA (Carrasco et al., 2000;
-% Bottalico et al., 2004), BCL2 (Ishihara et al., 2002), GPC3 (Gui et
-% al., 2012; Ogawa, 2007), TMEM8A, CLIC3, EBI3, FLT1, BCL6, SYDE1, and
-% MAN1C1 (Enquobahrie et al., 2008; Winn et al., 2009; Nishizawa, 2011).
-% In addition KISS1 is also involved in pre-eclampsia as well (Sado et
-% al., 2011; Qiao et al, 2012). Due to the association of many genes
-% that show human-specific shifts as well as high expression, we
-% conclude that pre-eclampsia is possibly a symptom of recent evolution
-% of gene expression on the human lineage. Although many genes such as
-% GPC3 and ADAM12 show distinctive relatively high expression in bonobo,
-% this expression has intensified on the human lineage. Sequencing of
-% more placentas of nonhuman primates, including other individuals of
-% the genus Pan will help to further characterize this transition.
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
+
+The identification of multiple genes (\textit{CRH}, \textit{ADAM12},
+\textit{KISS1}, \textit{PAPPA}, \textit{IL1RL1}, and \textit{PLAC1};
+\cref{tab:human_lineage_specific}) which are associated with
+pre-eclampsia which have increased expression on the \human lineage
+may explain the larger proportion of pre-eclampsia observed in humans.
+However, there are still genes which are associated with pre-eclampsia
+which are also more highly expressed in the \mouse lineage such as
+\textit{Afp} and \textit{Gjb2}. Furthermore, the onset of
+pre-eclampsia is well before term; all of our placentas are term
+placentas and provide limited information about the transcriptome
+during the onset of pre-eclampsia.
 
@@
 
-\DLA{Totally rewrite the conclusion; it's terrible}
-
-We sequenced \Sexpr{star.logs[,length(unique(species))]} placentas
-spanning the phylogenetic tree of mammals. This revealed a dynamic
-expression pattern of expression of lineage-specific expansions of
-gene families and differential gene expression that agreed with the
-scenario of an “arms race” between maternal and fetal tissues that
-have shaped such disparate gene expression profiles. In addition, we
-highlighted the broad spectrum of hormone gene expression in multiple
-species that suggested that there is a lack of conservation in hormone
-signaling during mammalian pregnancy. Major differences between
-expression in marsupial and eutherian mammals at term seemed to be
-correlated with the differences between choriovitelline and
-chorioamniotic placenta, with the marsupial placenta more enriched in
-genes which contribute to the active generation of energy. We also
-identified that galectin expression, especially that of
-\textit{LGALS1}, is a common feature of all mammalian placentas and
-was most likely present in the ancestral therian. Also, we identified
-two new expansions of genes that are highly expressed in the \elephant
-placenta that may be involved in trophoplast differentiation and immune
-system function, mirroring expansions in other groups such as
-primates, rodents, and ruminants. We also quantified significant
-lineage-specific shifts in gene expression, concentrating on the
-identification of genes that showed gene expression changes on the
-human lineage. We cautiously propose that many of these genes may be
-related to the invasive nature of the trophoblast in \human, although
-we note that in many of these genes, there is in some cases
-substantial expression in the \bonobo as well. These genes have the
-potential to spur further research and may be at the root of
-obstetrical disorders such as pre-eclampsia. We propose that more
-placentas from diverse lineages, especially those of marsupials and
-primates, to further identify genes and expression levels that are
-specific to marsupials and eutherians, as well as confirm the identity
-of human-specific expression shifts.
+We analyzed the sequenced transcriptome of
+\Sexpr{star.logs[,length(unique(file))]} placentas from
+\Sexpr{star.logs[,length(unique(species))]} species spanning the
+phylogenetic tree of mammals. This wealth of data revealed both
+lineage-specific expansions and differential expression of gene
+families as well as a set of core, non-housekeeping genes which are
+critical for the proper organization and functioning of the placenta.
+
+There is diversity in the placental expression of hormones, including
+estrogen, progesterone, prolactin, relaxin, natriuretic peptides, and
+the growth hormone family amount the species and lineages analyzed. We
+also identified genes differentially expressed on the \human lineage
+which are implicated in pre-eclampsia, including \textit{ADAM12},
+\textit{CRH}, and \textit{KISS1}.
+
+The core placenta transcriptome was also identified, which contains
+genes which are involved in maternal-fetal tolerance, such as the
+annexins, and genes responsible for cell-cell and cell-matrix
+interactions, such as \textit{LGALS1}. These genes were almost
+certainly present in the ancestral therian placenta. We hypothesize
+that modifications to their expression or function will compromise the
+development and organization of the placenta.
+
+This study has identified many tantalizing molecular pathways which
+illuminate the vast diversity of placental morphology and may explain
+some placental malfunctions in \human{}. However, the strength of our
+hypotheses are limited by our limited data; more placentas from
+multiple members of diverse clades and different timepoints are still
+needed to come to a more complete understanding of the genomic and
+molecular basis for placenta morphology and function. We look forward
+to collecting and analyzing such a dataset.
 

### 2016-03-25 manuscript/manuscript.Rnw
clean up RLN discussion; remove \paragraph{}

@@
 
-\paragraph{Pre-eclampsia \human Lineage Specific Genes}
+% \paragraph{Pre-eclampsia \human Lineage Specific Genes}
 
@@
 
-\subsubsection{\ac{GH} family}
-
-The \ac{GH} family has expanded in anthropoid primates, resulting in
-approximately 5 genes in \human (\textit{GH1}, \textit{GH2},
-\textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at least three in
-\bonobo (two \textit{GH2} and \textit{GH1}; owing to the similarity
-between the two \textit{GH1} in \textit{Pan troglodytes} we cannot
-reliably distinguish transcripts between the two). All show
+\subsubsection{Growth hormone family}
+
+The growth hormone family of genes has expanded in anthropoid
+primates, resulting in approximately 5 genes in \human (\textit{GH1},
+\textit{GH2}, \textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at
+least three in \bonobo (two \textit{GH2} and \textit{GH1}; owing to
+the similarity between the two \textit{GH1} in \textit{P. troglodytes}
+we cannot reliably distinguish transcripts between the two). All show
 substantial expression in the \human placenta
@@
 to be produced in the placenta of the pregnant \dog and is used in the
-diagnosis of canine pregnancy
+diagnosis of \dog pregnancy
 \cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
@@
 source of relaxin produced by pregnant \elephant remains unclear
-\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in relaxin in
+\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in \textit{RLN} in
 \elephant has been suggested to be a major signal for the initiation
@@
 Here we show that the placenta of elephants show high expression of
-relaxin and is likely a major source of \textit{RLN} at term. In
+\textit{RLN} and is likely a major source of \textit{RLN} at term. In
 contrast to the results of

### 2016-03-25 manuscript/manuscript.Rnw
completely rewrite abstract

@@
   anatomically variable organ in mammals. Understanding gene
-  expression patterns among placentas from multiple species can
-  illuminate the molecular basis of placenta function as well as
-  potentially enrich our knowledge of the biology of obstetrical
-  syndromes.
-
-  \parttitle{Results} We measured the expression profiles of the term
-  placenta of \Sexpr{star.logs[,length(unique(species))]} species of
-  therian (i.e. marsupials and “placental”) mammals using RNA-Seq. We
-  identified multiple genes and transcripts which are differentially
-  expressed across lineages including prolactin, relaxin, and
-  galectins, as well as identifying the non-housekeeping genes which
-  are core to placenta function (such as components of annexin
-  complexes) in all \Sexpr{star.logs[,length(unique(species))]}
-  species studied.
-
-  \parttitle{Conclusions} While there are multiple non-housekeeping
-  genes which are highly expressed in all placentas examined and are
-  likely to be involved in the basic functions of the mammalian
-  placenta, we see incredible lineage-specific diversity in the
-  expression and expansion of many genes. These lineage-specific genes
-  are likely to play a role in the anatomical variability of the
-  placenta and may also be implicated in important human placental
-  disorders such as pre-eclampsia.
-
-  \DLA{Complete rewrite abstract to make the points that we have found
-    stronger and less weak.}
-
-% We identified significant shifts in gene expression along specific
-% evolutionary lineages using a model of character evolution. Genes that
-% showed significant elevation in expression along the human lineage
-% ($\mathrm{FDR}\le 0.01$) were enriched in GO categories such as
-% angiogenesis, tube morphogenesis, cell migration, and regulation of
-% inflammatory response. Three of these genes (KISS1, ADAM12, GPC3)
-% showed exceptionally high expression in humans; these genes are
-% implicated in pre-eclampsia and may indicate that the risk of
-% pre-eclampsia recently emerged on the human lineage.
+  expression patterns among placentas from multiple species
+  illuminates the molecular basis of placenta function as well
+  obstetrical syndromes affecting humans.
+
+  \parttitle{Results} We measured the expression profiles of
+  \Sexpr{star.logs[,length(unique(file))]} term placenta of
+  \Sexpr{star.logs[,length(unique(species))]} species of mammals. We
+  found lineage-specific differential expression of prolactin,
+  relaxin, and galectins as well as increased expression of the
+  pre-eclampsia associated genes \textit{ADAM12}, \textit{CRH}, and
+  \textit{PAPPA} in humans. We also identified genes which are
+  expressed in all placentas, including genes with immune-modulating
+  properties (\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and
+  \textit{S100A10}) and cell-cell interactions (\textit{LAMC1},
+  \textit{LUM}, and \textit{LGALS1}).
+
+  \parttitle{Conclusions} A diverse array of genes are differentially
+  expressed and are likely to play a role in the anatomical
+  variability of the placenta in the placentas of mammalian lineages.
+  However, the expression of a subset of
+  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
+  genes is essential for the proper functioning of the placenta in all
+  species analyzed.
 

### 2016-03-28 manuscript/manuscript.Rnw
fill out discussion of core placenta transcriptome

@@
 
-\subsection{Annexin and Fetal Immunotolerance}
-
-\DLA{Finish out this section siting Aarli work and Linke work}
+\subsection{Core placenta transcriptome}
+
+The transcriptomes from multiple different species enabled us to
+identify the set of expressed genes which are common to all placentas.
+Components of the annexin complexes, including including
+\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
+are expressed in all placentas examined, and argue for the evolution
+importance of annexins in placenta function, where they likely
+establish maternal-fetal
+tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
+We suggest that further research into molecular role of these annexin
+complexes may illuminate how immune tolerance is maintained during
+placentation. The core placenta transcriptome is also enriched in
+genes involved in maintaining the structure of the placenta and
+transporting proteins, as would be expected given the function of the
+placenta.
 
@@
 interactions, such as \textit{LGALS1}. These genes were almost
-certainly present in the ancestral therian placenta. We hypothesize
-that modifications to their expression or function will compromise the
-development and organization of the placenta.
+certainly present in the placenta of the common mammalian ancestor. We
+hypothesize that modifications to their expression or function will
+compromise the development and organization of the placenta.
 

### 2016-03-28 manuscript/manuscript.Rnw
fix placenta table and file expansion

@@
 species with ortholog annotations are located in the supplemental data
-files \Sexpr{english.join(paste0(gsub(" ","\\_",
-  star.logs[,unique(species)]),"_mean_fpkm.txt"))}.
+files \Sexpr{english.join(gsub("_","\\\\_",paste0(gsub(" ","_",
+  star.logs[,unique(species)]),"_mean_fpkm.txt")))}.
 
@@
 @ 
-\caption{\linagetableexplanation{primates}}
+\caption{\lineagetableexplanation{primates}{all other analyzed species}}
 \label{tab:primate_lineage_specific}

### 2016-04-04 manuscript/manuscript.Rnw
add funding support

@@
 
+  \section*{Funding}
+  Partial support for this research came from the \textit{Eunice
+    Kennedy Shriver} National Institute of Child Health and Human
+  Development (NICHD), National Institutes of Health, U.S. Department
+  of Health and Human Services (N01-HD-3342), and the National Science
+  Foundation BCS-0827546.
+
   \printaddresses

### 2016-04-04 manuscript/manuscript.Rnw
it's SOLD1 not SORD1

@@
 an ortholog of \textit{SSLP1},
-\Sexpr{gene.expression("ovis_aries","SORD1",complete=TRUE)}, two
+\Sexpr{gene.expression("ovis_aries","SOLD1",complete=TRUE)}, two
 cathespins (\Sexpr{gene.expression("ovis_aries","CTSL",complete=TRUE)}

### 2016-04-06 data/go_analysis.R
calculate go on permissive core transcriptome

@@ -23,6 +23,7 @@ res.table <-
                         classic=result.fisher,
                         orderBy="classic",
                         topNodes=length(result.fisher@score)))
+res.table[,classic:=gsub("< ","",classic)]
 res.table[,classic.fdr:=p.adjust(method="BH",classic)]
 res.table[,Ontology:=args[3]]
 

### 2016-04-06 data/csv_files/species_fpkm.R
make csv files include all analyses

@@ -1,15 +1,8 @@
 library("data.table")
-args <- c("../combined_fpkm","homo_sapiens","homo_sapiens_mean_fpkm.txt")
+args <- c("../combined_fpkm","all_species_mean_fpkm.txt")
 args <- commandArgs(trailingOnly=TRUE)
-this.species <- gsub("_"," ",args[2])
-
 load(args[1])
-combined.fpkm[,`:=`(FPKM=NULL,
-                    FPKM_conf_lo=NULL,
-                    FPKM_conf_hi=NULL,
-                    FPKM_status=NULL
-                    )]
 write.table(file=args[length(args)],
-            combined.fpkm[species==this.species,][order(-mean_fpkm),],
+            combined.fpkm[order(species,-mean_fpkm),],
             sep="\t")
 

### 2016-04-06 data/csv_files/species_per_sample_fpkm.R

@@ -1,0 +1,8 @@
+library("data.table")
+args <- c("../combined_fpkm_per_sample","all_species_per_sample_mean_fpkm.txt")
+args <- commandArgs(trailingOnly=TRUE)
+load(args[1])
+write.table(file=args[length(args)],
+            combined.fpkm[order(species,-mean_fpkm,file),],
+            sep="\t")
+

### 2016-04-06 data/placenta_core_transcriptome.R
change the naming of the placenta core transcriptome

@@ -38,21 +38,21 @@ combined.fpkm.wide[,median_expression :=
                              1,median)]
 combined.fpkm.wide[,max_expression :=
                        apply(combined.fpkm.wide[,-1,with=FALSE],1,max.na.zero)]
-combined.fpkm.wide[,fifth_min_expression :=
+combined.fpkm.wide[,all_but_four_expression :=
                        apply(combined.fpkm.wide[,-1,with=FALSE],1,function(x){min.but.one(x,4)})]
-combined.fpkm.wide[,fourth_min_expression :=
+combined.fpkm.wide[,all_but_three_expression :=
                        apply(combined.fpkm.wide[,-1,with=FALSE],1,function(x){min.but.one(x,3)})]
-combined.fpkm.wide[,third_min_expression :=
+combined.fpkm.wide[,all_but_two_expression :=
                        apply(combined.fpkm.wide[,-1,with=FALSE],1,function(x){min.but.one(x,2)})]
-combined.fpkm.wide[,second_min_expression :=
+combined.fpkm.wide[,all_but_one_expression :=
                        apply(combined.fpkm.wide[,-1,with=FALSE],1,min.but.one)]
-combined.fpkm.wide[,min_expression :=
+combined.fpkm.wide[,all_expression :=
                        apply(combined.fpkm.wide[,-1,with=FALSE],1,min.na.zero)]
 setkey(combined.fpkm.wide,"human_name")
 
 ## all genes which are not housekeeping genes which have minimum 
 core.placenta.transcriptome.long <- 
-    data.table(melt(combined.fpkm.wide[min_expression >= 10,
+    data.table(melt(combined.fpkm.wide[all_expression >= 10,
                                        ][order(-median_expression)
                                          ][!(human_name %in% 
                                                  housekeeping.genes.superset[,gene_short_name])],
@@ -68,11 +68,11 @@ core.placenta.transcriptome.long[,human_name:=
 core.placenta.transcriptome.genes.shape <-
     combined.fpkm.wide[,list(human_name,
                              median_expression,
-                             min_expression,
-                             second_min_expression,
-                             third_min_expression,
-                             fourth_min_expression,
-                             fifth_min_expression,
+                             all_expression,
+                             all_but_one_expression,
+                             all_but_two_expression,
+                             all_but_three_expression,
[…]

### 2016-04-06 manuscript/manuscript.Rnw
make Michael and equal contributor

@@
 %
-\author[addressref={aff1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
-\author[addressref={aff2},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
+\author[addressref={aff1},email={don@donarmstrong.com},noteref={n1}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
+\author[addressref={aff2},email={1mmcgowen1@gmail.com},noteref={n1}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
 \author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL}\fnm{Amy} \snm{Weckle}}
@@
 \begin{artnotes}
+\end{artnotes}
+
+\begin{artnotes}
+\note[id=n1]{Equal contributor} % note, connected to author
 \end{artnotes}

### 2016-04-06 manuscript/manuscript.Rnw
clean up manuscript and supplemental material

@@
 \usepackage{amsmath}
-\usepackage{sfmath}
+% \usepackage{sfmath}
 \usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
@@
 % \usepackage[noblocks,auth-sc]{authblk}
-\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
+\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
 %\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
@@
 the most variable organ in terms of structure and form
-\cite{Mossman.1937:comparativemorphogenesisofthe}. Species differ in
+\cite{Mossman.1937:comparativemorphogenesisofthe,Moffett.ea2006:Immunologyofplacentationin}. Species differ in
 the shape of attachment (i.e., discoid, diffuse, cotyledonary,
@@
 
-% \DLA{Add details for new sequences and fix numbers}
-
 \subsection{Sequencing}
@@
 \ac{SRA} (Supplemental \cref{tab:srr_numbers}) and analyzed using the
-same methodology as mRNAs sequenced specifically for this study.
+same methodology as RNAs sequenced specifically for this study.
 
@@
 species with ortholog annotations are located in the supplemental data
-files \Sexpr{english.join(gsub("_","\\\\_",paste0(gsub(" ","_",
-  star.logs[,unique(species)]),"_mean_fpkm.txt")))}.
+file all\_species\_mean\_fpkm.txt.xz and the per-sample FPKM values are in
+all\_species\_per\_sample\_fpkm.txt.xz.
 
@@
                           mean_fpkm][1]
+    if (is.na(fpkm)) {
+        stop("FPKM is NA!!!!")
+    }
     if (units || complete)
@@
                                                   "}")]
-        setnames(display.table,"Fold Change","$\\mathrm{log}_2$ FC")
+        setnames(display.table,"Fold Change","$\\text{log}_2$ FC")
         print(xtable(display.table,
@@
 (Supplemental \cref{fig:fstl1_evolution_tree},
-$\mathrm{FDR}=\Sexpr{pri.all.lin.spec["FSTL1",`FDR`]}$), is implicated
+$\text{FDR}=\Sexpr{pri.all.lin.spec["FSTL1",`FDR`]}$), is implicated
 in BMP signaling, and has previously been implicated in inflammation
@@
 hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
-hormone gene family, \ac{GH2} (\ac{FPKM}=%
+hormone gene family, \ac{GH2} (%
 \Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
@@
 \textit{CGB7}, \textit{CGB8}) have reasonable expression
-($\mathrm{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
+($\text{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
 expression is on the order of the expression of
@@
 \cref{tab:top50_mouse_placenta} with a list of all genes with
-$\mathrm{FPKM} \ge 50$ in Supplemental
+$\text{FPKM} \ge 50$ in Supplemental
 \cref{tab:all_mus_musculus_placenta}. The top expressed genes include
@@
   grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
-prolactin related genes having $\mathrm{FPKM} \ge 1$
+prolactin related genes having $\text{FPKM} \ge 1$
 (\cref{fig:prolactin_evolution_tree}, Supplemental
@@
 Several members of the ruminant-specific pregnancy-associated
-glycoprotein family show especially high expression in \cow placenta
+glycoprotein family \cite{Green.ea2000:Pregnancy-associatedbovineandovine} 
+show especially high expression in \cow placenta
 (\cref{tab:top50_cow_placenta}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D, and
@@
 the expression levels of this family are higher in \cow than in \sheep
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D. Other genes with high expression
+(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). Other genes with high expression
 include genes involved with regulation of growth (\textit{IGF2},
@@
 \Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
-with a total of 14 with $\mathrm{FPKM} \ge 1$.
+with a total of 14 with $\text{FPKM} \ge 1$.
 %
@@
 Placental expression of the hormones relaxin (\textit{RLN}, %
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000036090")})
+\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000017411")})
 and prolactin (\textit{PRL}) are exceptionally high in the elephant
@@
 (\Sexpr{gene.expression("loxodonta_africana","CYP19A1")}) and the
-enzyme that produces progesterone HSD3B1
-(\Sexpr{gene.expression("loxodonta_africana","HSD3B")}). Other genes
+enzyme that produces progesterone, \textit{HSD3B1}
+(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027896")}). Other genes
 with high expression in the elephant placenta include genes involved
@@
 shown in \cref{tab:top50_armadillo_placenta} with a list of all genes
-with $\mathrm{FPKM}\ge 50$ in Supplemental
+with $\text{FPKM}\ge 50$ in Supplemental
 \cref{tab:all_dasypus_novemcinctus_placenta}. The gene with the
@@
 <<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
-print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
+print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
+                                                     Species),],
              aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
@@
 the placenta transcriptome if a 1:1 ortholog of that gene was
-expressed with $\mathrm{FPKM} \ge 10$ in all species studied and was
+expressed with $\text{FPKM} \ge 10$ in all species studied and was
 not a human housekeeping gene according to
@@
 \cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
-Supplemental \cref{tab:core_placenta_go_terms} lists the significant enrichment of
-\ac{GO} terms in the core placenta transcriptome; notable terms
-include those involved in cytoskeletal structure, protein transport,
-membrane protein structure, and inflammation resolution.
+Supplemental \cref{tab:core_placenta_go_terms} lists the significant
+enrichment of \ac{GO} terms in the core placenta transcriptome;
+notable terms include those involved in cytoskeletal structure,
+protein transport, membrane protein structure, and inflammation
+resolution. Increasing the permissiveness of the core placenta
+transcriptome by only requiring 10 of 14 species to express 1:1
+orthologs with FPKM $>10$ and results in the same terms also being
+significantly enriched, along with additional terms (Supplemental
+\cref{tab:core_placenta_go_terms}).
 
@@
 %
-Significant enrichment in genes ($\mathrm{FDR}\le0.05$) in the top 1\%
+Significant enrichment in genes ($\text{FDR}\le0.05$) in the top 1\%
 of \ac{FPKM} values in at least 12 of the
@@
 @ 
-\caption{Expression of 1:1 Human orthologs of Hormone Genes, Galectins,
-  genes implicated in pre-eclampsia (PE), and \textit{IGF}-related genes}
+\caption{Expression in $\text{log}_{10}(\text{FPKM})$ of 1:1 \human
+  orthologs of Hormone Genes, Galectins, genes implicated in
+  pre-eclampsia (PE), and \textit{IGF}-related genes}
 \label{fig:hormone_genes}
@@
 expression of the primarily pituitary hormone prolactin (\textit{PRL})
-in the placenta, although we see high expression of genes which
+in the term placenta, although we see high expression of genes which
 resulted from duplications of \textit{PRL} in the \mouse, \carmeli,
-and \cow (Supplemental \cref{fig:prolactin_evolution_tree}). All afrotherians yet
-analyzed contain only one prolactin gene, and prolactin has undergone
-adaptive evolution on the \elephant lineage
+and \cow (Supplemental \cref{fig:prolactin_evolution_tree}). The
+results in \mouse are relatively consistent with the temporal
+expression results of
+\citet{Simmons.ea2008:Spatialandtemporalexpression}. For example, we
+see very high expression of \textit{Prl3b1} and \textit{Prl8a9} in the
+term placentas of \mouse, but low expression of \textit{Prl7a1} and
+\textit{Prl8a2} which are expressed at earlier embryonic stages. %
+All afrotherians yet analyzed contain only one prolactin gene, and
+prolactin has undergone adaptive evolution on the \elephant lineage
 \cite{Wallis2009:ProlactininAfrotheriacharacterization}.
@@
 expression, \galili
-(\Sexpr{gene.expression("spalax_galili","Hoxa11",complete=TRUE)})
-is the rodent with the lowest expression of \textit{PRL}
+(\Sexpr{gene.expression("spalax_galili","Hoxa11",complete=TRUE)}) is
+the rodent with the lowest expression of \textit{PRL}
 (\cref{fig:prolactin_evolution_tree}), and \elephant also does not
@@
 time of the last common ancestor between all extant anthropoids
-\cite{Papper.ea2009:Ancientoriginofplacental}. There is some evidence
-that increased expression of these genes may be correlated with
-overall birth weight of offspring in \human
+\cite{Papper.ea2009:Ancientoriginofplacental,Soares2004:prolactinandgrowthhormone}.
+There is some evidence that increased expression of these genes may be
+correlated with overall birth weight of offspring in \human
 \cite{Männik.ea2010:Differentialexpressionprofileof}, and it is

### 2016-04-06 manuscript/supplemental_material.Rnw
fix linage typo
clean up manuscript and supplemental material

@@
 \usepackage{wrapfig}
+\usepackage{amsmath}
 \usepackage{booktabs}
@@
 \usepackage[noblocks,auth-sc]{authblk}
-\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black]{hyperref}
+\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=blue,citecolor=blue]{hyperref}
 %\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
@@
 library("grid")
+library("scales")
 library("gridExtra")
@@
 
-<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","palcenta_go_results_mf","hugo_nomenclature","gene_annotation")))>>=
+<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","hugo_nomenclature","gene_annotation")))>>=
 for (a in paste0("../data/",c("combined_fpkm","gene_trees",
@@
                               "placenta_core_transcriptome",
-                              "placenta_go_results_bp",
-                              "placenta_go_results_mf",
                               "hugo_nomenclature",
@@
   \raggedright
-  \begin{tabular}{c >{\raggedright}p{1.5in} c >{\raggedright\arraybackslash}p{2in}}
+  \small
+  \begin{tabular}{c >{\raggedright}p{1.5in} c >{\raggedright\arraybackslash}p{2in} c}
     \toprule
-    Species                       & Collector & Experiment & Sample(s) \\
+    Species                       & Collector & Experiment & Sample(s) & Library \\
     \midrule
-    \textit{Ateles fusciceps}     & Kurt Benirschke & GSE79121 & SRX1629201 \\
-    \textit{Bos taurus}           & Bob Kreft & GSE79121 & SRX1629203 \\
-    \textit{Canis familiaris}     & Dalen Agnew & GSE79121 & SRX1629204 \\
-    \textit{Dasypus novemcinctus} & Richard Truman, Vincent J. Lynch& GSE79121 & SRX1629200 \\
-    \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & GSM1139275 & SRR850299,  SRR850300 \\
-    \textit{Loxodonta africana}   & Kurt Benirschke & GSE79121 & SRX1629205 \\
+    \textit{Ateles fusciceps}     & Kurt Benirschke & GSE79121 & SRX1629201 & Total\\
+    \textit{Bos taurus}           & Bob Kreft & GSE79121 & SRX1629203 & Total \\
+    \textit{Canis familiaris}     & Dalen Agnew & GSE79121 & SRX1629204 & Total \\
+    \textit{Dasypus novemcinctus} & Richard Truman, Vincent J. Lynch& GSE79121 & SRX1629200 & Total\\
+    \textit{Equus caballus}       & \citet{Coleman.ea2010:Structuralannotationofequine} & GSM1139275 & SRR850299,  SRR850300 & mRNA \\
+    \textit{Loxodonta africana}   & Kurt Benirschke & GSE79121 & SRX1629205 & Total \\
     \textit{Homo sapiens}         & \citet{Hughes.ea2015:Evaluatingintra-andinter-individual} & GSE66622 
@@
                                                SRR1851011, 
-                                               SRR1851012 \\
-    \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & GSE79121 & SRX1629207 \\
+                                               SRR1851012  & mRNA \\
+    \textit{Monodelphis domestica} & Karen Smith, Gunter Wagner & GSE79121 & SRX1629207 & Total \\
     \textit{Mus musculus}         & \citet{Wang.ea2011:surveyfornovelimprinted} & GSE29690 
@@
                                        SRR243592, 
-                                       SRR243593\\
-    \textit{Ovis aries}           & \citet{Archibald.ea2010:sheepgenomereferencesequence} & ERP005642 & ERR489144, ERR489145\\
-    \textit{Pan paniscus}         & William Mintz Fields & GSE79121 & SRX1629202 \\
-    \textit{Spalax carmeli}       & Anne Dupressoir, Eviator Nevo & GSE68763 & SRR2016467 \\
-    \textit{Spalax galili}        & Eviator Nevo                  & GSE68763 & SRR2016466 \\
-    \textit{Sus scrofa}           & \citet{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 \\
+                                       SRR243593 & mRNA \\
+    \textit{Ovis aries}           & \citet{Archibald.ea2010:sheepgenomereferencesequence} & ERP005642 & ERR489144, ERR489145 & Total \\
+    \textit{Pan paniscus}         & William Mintz Fields & GSE79121 & SRX1629202 & Total\\
+    \textit{Spalax carmeli}       & Anne Dupressoir, Eviator Nevo & GSE68763 & SRR2016467 & Total\\
+    \textit{Spalax galili}        & Eviator Nevo                  & GSE68763 & SRR2016466 & Total\\
+    \textit{Sus scrofa}           & \citet{Ahanda.ea2012:Predictionofaltered3-} &  SRP014902 & SRR543893 & mRNA \\
     \bottomrule
   \end{tabular}
-  \caption{Species, experiment and species numbers used in the analyses presented in this paper}
+  \caption{Species, experiment and species numbers used in the analyses presented in this paper. 
+    The Library column indicates the type of RNA selection performed: Total indicates rRNA depletion; 
+    mRNA indicates poly-A selection.}
   \label{tab:srr_numbers}
@@
 
-\section{Gene Ontology Analyses}
-
-<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
-go.analysis.wide <- 
-    data.table(dcast(data=go.results,
-                     formula=GO.ID+type+Term+Ontology~species,
-                     value.var="FDR"))
-go.analysis.wide <- 
-    go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
-                                         function(x){sum(as.numeric(as.character(x)) <= 0.05)})]
-go.analysis.wide <- 
-    go.analysis.wide[,`max(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
-                                           function(x){if (!any(is.finite(x) & x<=0.05)) {
-                                                           return(NA)
-                                                       } else {
-                                                           max(x[is.finite(x) & x<=0.05],na.rm=TRUE)
-                                                       }})]
-go.analysis.wide.table <- 
-    go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
-                         grepl("Top 1% No Housekeeping",type),
-                     list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
-go.analysis.wide.table <- 
-    go.analysis.wide.table[order(Ontology,-`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
-setnames(go.analysis.wide.table,"GO.ID","GO ID")
-
-print(xtable(go.analysis.wide.table,
-             display=c("d","s","s","s","d","e"),
-             align=  c("c","c","c","c","c","c"),
-             caption="GO terms with significant enrichment ($\\mathrm{FDR}\\le0.05$)
-  in 12 or more of the species analyzed from genes in the top 1\\% of
-  expression. $n(\\mathrm{FDR}\\le0.05)$ is the number of species with
-  significant enrichment ($\\mathrm{FDR}\\le 0.05$) in the GO term
-  listed. $max(\\mathrm{FDR}\\le0.05)$ is the maximum FDR value in all
-  species with an FDR of less than $0.05$.",
-  label="tab:go_terms_all_species"
-  ),
-  math.style.exponents=TRUE,
-  include.rownames=FALSE,
-  booktabs=TRUE,
-  floating=FALSE,
-  tabular.environment="longtable",
-  caption.placement="top",
-  size="\\footnotesize"
-  )
-
-@
-
-\clearpage
-
-\subsection{Top 100 Transcripts Enrichment}
-
-<<go.analysis.all.species,echo=FALSE,results="asis">>=
-for (this.species in names(species.names)) {
-    cat(paste0("\\subsubsection{Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
-                           " the top 100 \\textit{",
-                           capfirst(this.species),"} expressed transcripts}\n\n"))
-    this.caption <- paste0("Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
-                           " the top 100 \\textit{",capfirst(this.species),"} expressed transcripts")
-    this.label <- paste0("tab:go_terms_enriched_",gsub(" ","_",this.species))
-    print(xtable(go.results[grepl("100 No Housekeeping",type) & FDR <= 0.05 & species == this.species,
-                            list(GO.ID,Term,FDR,Ontology)
-                            ][order(FDR)],
-                 caption=this.caption,
-                 label=this.label,
-                 display=c("s","s","s","e","s")
-                 ),
-          size="\\small",
-          booktabs=TRUE,
-          tabular.environment="longtable",
-          caption.placement="top",
-          floating=FALSE)
-}
-@
-
-\clearpage
-
-\subsection{Top 1\% Transcripts Enrichment}
-
-<<go.analysis.top.5p.all.species,echo=FALSE,results="asis">>=
-for (this.species in names(species.names)) {
-    cat(paste0("\\subsubsection{Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
-                           " the top 1\\% of \\textit{",
-                           capfirst(this.species),"} expressed transcripts}\n\n"))
-    this.caption <- paste0("Significantly enriched ($\\mathrm{FDR}\\le0.05$) GO terms in ",
-                           " the top 1\\% of \\textit{",
-                           capfirst(this.species),"} expressed transcripts.")
-    print(xtable(go.results[grepl("Top 1% No Housekeeping",type) & FDR <= 0.05 & species == this.species,
-                            list(GO.ID,Term,FDR,Ontology)
-                            ][order(FDR)],
-                 caption=this.caption,
-                 label=paste0("tab:go_terms_enriched_top_one_",gsub(" ","_",this.species)),
-                 display=c("s","s","s","e","s")
-                 ),
-          size="\\small",
-          booktabs=TRUE, 
-          tabular.environment="longtable",
-          caption.placement="top",
-          floating=FALSE)
-}
-@
-
-\clearpage
+\section{Species Expression}
 
@@
 
-\section{Gene Expression}
-
-\subsection{Gene Evolution Trees}
-
-\subsubsection{Pre-eclampsia genes}
+<<top.expressed.genes,echo=FALSE,results="asis">>=
+for (this.species in names(species.names)) {
+    this.caption <- paste0("\\textit{",
+                           capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
+    cat(paste0("\\subsubsection{\\textit{",capfirst(this.species),"} Placental Gene Expression}\n\n"))
+    xt.align <- c("r","l","l","p{2.5in}","r")
+    if (this.species == "homo sapiens" || this.species == "ateles fusciceps") {
+        xt.align <- xt.align[-2]
+    }
+    print(xtable(genes.expression.table(species.name=this.species),
+                 caption=this.caption,
+                 align=xt.align,
+                 label=paste0("tab:all_",gsub(" ","_",this.species),"_placenta")
+                 ),
+          booktabs=TRUE,
+          tabular.environment="longtable",
+          floating=FALSE,
+          caption.placement="top",
+          sanitize.text.function=genes.table.sanitize,
+          size="\\tiny"
+          )
+}
+@
+
+\clearpage
+
+\section{Gene Ontology Analyses}
+
+<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
+go.analysis.wide <- 
+    data.table(dcast(data=go.results,
+                     formula=GO.ID+type+Term+Ontology~species,
+                     value.var="FDR"))
+go.analysis.wide <- 
+    go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
+                                         function(x){sum(as.numeric(as.character(x)) <= 0.05)})]
+go.analysis.wide <- 
+    go.analysis.wide[,`max(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
+                                           function(x){if (!any(is.finite(x) & x<=0.05)) {
+                                                           return(NA)
+                                                       } else {
+                                                           max(x[is.finite(x) & x<=0.05],na.rm=TRUE)
+                                                       }})]
+go.analysis.wide.table <- 
+    go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
+                         grepl("Top 1% No Housekeeping",type),
+                     list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
+go.analysis.wide.table <- 
+    go.analysis.wide.table[order(Ontology,-`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
+setnames(go.analysis.wide.table,"GO.ID","GO ID")
+
+print(xtable(go.analysis.wide.table,
+             display=c("d","s","s","s","d","e"),
+             align=  c("c","c","c","c","c","c"),
+             caption="GO terms with significant enrichment ($\\text{FDR}\\le0.05$)
+  in 12 or more of the species analyzed from genes in the top 1\\% of
+  expression. $n(\\text{FDR}\\le0.05)$ is the number of species with
+  significant enrichment ($\\text{FDR}\\le 0.05$) in the GO term
+  listed. $max(\\text{FDR}\\le0.05)$ is the maximum FDR value in all
+  species with an FDR of less than $0.05$.",
+  label="tab:go_terms_all_species"
+  ),
+  math.style.exponents=TRUE,
+  include.rownames=FALSE,
+  booktabs=TRUE,
+  floating=FALSE,
+  tabular.environment="longtable",
+  caption.placement="top",
+  size="\\footnotesize"
+  )
+
+@
+
+\clearpage
+
+\subsection{Top 1\% Transcripts Enrichment}
+
+<<go.analysis.top.5p.all.species,echo=FALSE,results="asis">>=
+for (this.species in names(species.names)) {
+    cat(paste0("\\subsubsection{Significantly enriched ($\\text{FDR}\\le0.05$) GO terms in ",
+                           " the top 1\\% of \\textit{",
+                           capfirst(this.species),"} expressed transcripts}\n\n"))
+    this.caption <- paste0("Significantly enriched ($\\text{FDR}\\le0.05$) GO terms in ",
+                           " the top 1\\% of \\textit{",
+                           capfirst(this.species),"} expressed transcripts.")
+    print(xtable(go.results[grepl("Top 1% No Housekeeping",type) & FDR <= 0.05 & species == this.species,
+                            list(GO.ID,Term,FDR,Ontology)
+                            ][order(FDR)],
+                 caption=this.caption,
+                 label=paste0("tab:go_terms_enriched_top_one_",gsub(" ","_",this.species)),
+                 display=c("s","s","s","e","s")
+                 ),
+          size="\\small",
+          booktabs=TRUE, 
+          math.style.exponents=TRUE,
+          include.rownames=FALSE,
+          tabular.environment="longtable",
+          caption.placement="top",
+          floating=FALSE)
+}
+@
+
+\clearpage
+
+\subsection{Top 100 Transcripts Enrichment}
+
+<<go.analysis.all.species,echo=FALSE,results="asis">>=
+for (this.species in names(species.names)) {
+    cat(paste0("\\subsubsection{Significantly enriched ($\\text{FDR}\\le0.05$) GO terms in ",
+                           " the top 100 \\textit{",
+                           capfirst(this.species),"} expressed transcripts}\n\n"))
+    this.caption <- paste0("Significantly enriched ($\\text{FDR}\\le0.05$) GO terms in ",
+                           " the top 100 \\textit{",capfirst(this.species),"} expressed transcripts")
+    this.label <- paste0("tab:go_terms_enriched_",gsub(" ","_",this.species))
+    print(xtable(go.results[grepl("100 No Housekeeping",type) & FDR <= 0.05 & species == this.species,
+                            list(GO.ID,Term,FDR,Ontology)
+                            ][order(FDR)],
+                 caption=this.caption,
+                 label=this.label,
+                 display=c("s","s","s","e","s")
+                 ),
+          size="\\small",
+          booktabs=TRUE,
+          math.style.exponents=TRUE,
+          include.rownames=FALSE,
+          tabular.environment="longtable",
+          caption.placement="top",
+          floating=FALSE)
+}
+@
+
+\clearpage
+
+\section{Gene Evolution Trees}
+
+\subsection{Pre-eclampsia genes}
 
@@
 
-\subsubsection{Primate Specific Evolution}
+\subsection{Primate Specific Evolution}
 
@@
 
+\subsection{Other Gene Trees}
+
 \begin{figure}
@@
 
-\subsection{Clade specific Expression}
+\section{Clade specific Expression}
 
@@
 @ 
-\caption{\linagetableexplanation{ the euarchontoglires superorder}{ the laurasiatherian superorder}}
+\caption{\lineagetableexplanation{ the euarchontoglires superorder}{ the laurasiatherian superorder}}
 \label{tab:laurasiatheria_euarchontoglires_lineage_specific}
@@
 
-\subsection{Species Expression}
-
-<<top.expressed.genes,echo=FALSE,results="asis">>=
-for (this.species in names(species.names)) {
-    this.caption <- paste0("\\textit{",
-                           capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
-    cat(paste0("\\subsection{\\textit{",capfirst(this.species),"} Placental Gene Expression}\n\n"))
-    xt.align <- c("r","l","l","p{2.5in}","r")
-    if (this.species == "homo sapiens" || this.species == "ateles fusciceps") {
-        xt.align <- xt.align[-2]
-    }
-    print(xtable(genes.expression.table(species.name=this.species),
-                 caption=this.caption,
-                 align=xt.align,
-                 label=paste0("tab:all_",gsub(" ","_",this.species),"_placenta")
-                 ),
-          booktabs=TRUE,
-          tabular.environment="longtable",
-          floating=FALSE,
-          caption.placement="top",
-          sanitize.text.function=genes.table.sanitize,
-          size="\\tiny"
-          )
-}
-@
-
-\clearpage
-
 \section{Core Placenta Transcriptome}
@@
 \footnotesize
-<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis">>=
+<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis",cache.extra=tools::md5sum(paste0("../data/placenta_go_results_",c("bp","mf","permissive_bp","permissive_mf")))>>=
+load("../data/placenta_go_results_mf")
+load("../data/placenta_go_results_bp")
+load("../data/placenta_go_results_permissive_mf")
+load("../data/placenta_go_results_permissive_bp")
+
 placenta.go.results <- rbindlist(list(head(n=20,placenta_go_results_mf[classic.fdr < 0.05,]),
@@
 placenta.go.results <- placenta.go.results[,Annotated:=NULL]
-setnames(placenta.go.results,"Significant","# Core")
+setnames(placenta.go.results,"Significant","Core")
 setnames(placenta.go.results,"classic","p value")
@@
 setnames(placenta.go.results,"Term","GO Term")
-print(xtable(placenta.go.results,
+
+placenta.go.results.permissive <-
+    rbindlist(list(placenta_go_results_permissive_mf,
+                   placenta_go_results_permissive_bp))
+setnames(placenta.go.results.permissive,"Significant","Core")
+setnames(placenta.go.results.permissive,"classic","p value")
+setnames(placenta.go.results.permissive,"classic.fdr","Permissive FDR")
+setnames(placenta.go.results.permissive,"GO.ID","GO ID")
+setnames(placenta.go.results.permissive,"Term","GO Term")
+setkey(placenta.go.results,"GO ID")
+setkey(placenta.go.results.permissive,"GO ID")
+
+combined.placenta.go.results <-
+    placenta.go.results.permissive[,list(`GO ID`,`Permissive FDR`)
+                                   ][placenta.go.results][order(Ontology,FDR),]
+combined.placenta.go.results <-
+    combined.placenta.go.results[,`p value`:=as.numeric(`p value`)]
+setcolorder(combined.placenta.go.results,
+            c("Ontology",
+              "GO ID",
+              "GO Term",
+              "Core",
+              "Expected",
+              "p value",
+              "FDR",
+              "Permissive FDR"))
+combined.placenta.go.results <-
+    combined.placenta.go.results[,`p value`:=NULL]
+setnames(combined.placenta.go.results,"Ontology","Ont.")
+setnames(combined.placenta.go.results,"Expected","Exp.")
+print(xtable(combined.placenta.go.results,
              digits=2,
-             display=c("d","s","s","d","f","e","e","s")),
+             display=c("d","s","s","s","d","f","e","e")),
       include.rownames=FALSE,
       booktabs=TRUE,
+      math.style.exponents=TRUE,
       floating=FALSE
@@
   (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]} MF,
-  \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP)
-  in the core placenta transcriptome as compared to all genes
-  with 1:1 human orthologs in all species. “\# Core” is the
-  number of genes in the core transcriptome for each GO term;
-  “Expected” is the number of genes expected in the core
-  transcriptome for each term; “p-value” are the odds of this
-  occurring by chance calculated by the Fisher exact test for
-  each term; “FDR” is the false discovery rate as estimated by
-  the Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure}.}
+  \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP) in
+  the core placenta transcriptome ($\text{FPKM}>10$ in 14
+  of 14 species) as compared to all genes with 1:1 human orthologs in
+  all species. “Ont.” is the GO Ontology, where MF is molecular function 
+  and BP is biological process; “Core” is the number of genes in the core
+  transcriptome for each GO term; “Exp.” is the number of genes
+  expected in the core transcriptome for each term; 
+  “FDR” is the false discovery rate as estimated by the
+  Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure} based on
+  p values calculated by the Fisher Exact Test; 
+  “Permissive FDR” is the FDR resulting from the equivalent GO
+  analysis using the permissive core placenta transcriptome 
+  (1:1 orthologs with expression of at least 10 FPKM at least 10 of 14 species) 
+  showing that the enriched GO terms are relatively insensitive to the precise
+  set of species included.}
 \label{tab:core_placenta_go_terms}
@@
 <<core.placenta.transcriptome.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
+num.species <- star.logs.wide[,length(unique(species))]
 calculate.distribution.shape <- 
@@
 cpt.distribution.shape <- 
-    lapply(c("min_expression",
-             "second_min_expression",
-             "third_min_expression",
-             "fourth_min_expression",
-             "fifth_min_expression"),
+    lapply(c("all_expression",
+             "all_but_one_expression",
+             "all_but_two_expression",
+             "all_but_three_expression",
+             "all_but_four_expression"),
            function(x){
-               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Lowest Expression`:=x]
+               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
            })
+fix.expression.names <- function(x){
+    x <- gsub("^all_expression$",
+              paste0(num.species," of ",num.species," species"),
+              x
+              )
+    nc.levels <- paste0(num.species," of ",num.species," species")
+    ## number coutner
+    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
+    for (a in names(nc)) {
+        x <- gsub(paste0("^all_but_",a,"_expression$"),
+              paste0(num.species-nc[a]," of ",num.species," species"),
+              x
+              )
+        nc.levels <-
+            c(nc.levels,
+              paste0(num.species-nc[a]," of ",num.species," species"))
+    }
+    return(factor(x,levels=nc.levels))
+}
+
 cpt.distribution.shape <- 
@@
 cpt.distribution.shape <- 
-    cpt.distribution.shape[,`Lowest Expression`:=
-                               gsub("^min_expression$","First",
-                                    `Lowest Expression`)]
-cpt.distribution.shape <- 
-    cpt.distribution.shape[,`Lowest Expression`:=
-                               factor(gsub("^(.)(.+)_min_expression$","\\U\\1\\L\\2",
-                                           `Lowest Expression`,perl=TRUE),
-                                      levels=c("First","Second","Third","Fourth","Fifth"))
-                           ]
+    cpt.distribution.shape[,`Minimum Expression in`:=
+                                fix.expression.names(`Minimum Expression in`)]
+
 setnames(cpt.distribution.shape,
@@
 
-print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 1,],
-             aes(x=`Expression (FPKM)`,color=`Lowest Expression`,y=num_genes))
+print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],
+             aes(x=`Expression (FPKM)`,color=`Minimum Expression in`,y=num_genes))
       + geom_line(lwd=1)
-      + xlab(expression(FPKM))
-      + ylab("Genes with N-th Lowest Expression")
-      + scale_y_log10()
-      + scale_x_log10()
+      + xlab("Minimum Expression cutoff (FPKM)")
+      + ylab("Genes with N-th Minimum Expression")
+      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                      labels = trans_format("log10", math_format(10^.x)))
+      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                      labels = trans_format("log10", math_format(10^.x)))
       + geom_vline(xintercept=10)
@@
                          xend=10,
-                         x=100,
-                         color=`Lowest Expression`),
+                         x=1000,
+                         color=`Minimum Expression in`),
                      data=cpt.distribution.shape[`Expression (FPKM)` > 10,
-                                              max(num_genes),`Lowest Expression`],
+                                              max(num_genes),`Minimum Expression in`],
                      arrow=arrow(length=unit(0.5,"cm"))
                    )
-      + geom_text(mapping=aes(x=100,
+      + geom_text(mapping=aes(x=1000,
                       y=V1,
@@
                   data=cpt.distribution.shape[`Expression (FPKM)` > 10,
-                      max(num_genes),`Lowest Expression`],
+                      max(num_genes),`Minimum Expression in`],
                   show_guide=FALSE,
@@
                   )
-      + guides(color=guide_legend(title="Lowest\nExpression"))
-      )
-@ 
-\caption{Distribution shape of placenta core genes. The vertical axis shows
-  the number of orthologous gene sets included in the core placenta
-  transcriptome if the least expression cutoff (or second-least, etc.)
-  was set to the horizontal axis. For example, if the cutoff is set to 10
-  (vertical black line), 37 genes would be included if the FPKM of all
-  species must be at least 10, 134 genes would be included if all but
-  one species must have FPKM of at least 10. 655 genes would be
-  included if all but five species must have FPKM of at least 10.}
+      + guides(color=guide_legend(title="Minimum\nExpression in"))
+      )
+@ 
+\caption{Distribution shape of placenta core genes. The vertical axis
+  shows the number of 1:1 orthologous genes included in the core
+  placenta transcriptome if %
+  \Sexpr{cpt.distribution.shape[,levels(`Minimum Expression in`)[1]]}
+  %
+  (or %
+  \Sexpr{cpt.distribution.shape[,levels(`Minimum Expression in`)[2]]},
+  % 
+  etc.) have a minimum expression in FPKM equal to the horizontal axis.
+  For example, if the cutoff is set to 10 (vertical black line), 
+  % 
+  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
+    & `Minimum Expression in` == '14 of 14 species',max(num_genes)]}
+  %
+  genes would be included if the FPKM of a set of 1:1 orthologs of all 14 species must be at
+  least 10, %
+  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
+    & `Minimum Expression in` == "13 of 14 species",max(num_genes)]}
+  %
+  1:1 orthologs would be included if 13 of 14 species have an
+   with an FPKM of at least 10, and 
+  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
+    & `Minimum Expression in` == '10 of 14 species',max(num_genes)]}
+  genes would be included if 10 of 14 species
+  must have an FPKM of at least 10.}
 \label{fig:core_placenta_transcriptome_shape}
@@
 
+\clearpage
+
 \printbibliography

### 2016-04-07 figures/species_ordering.R
nannospalax is spalax

@@ -4,7 +4,7 @@ species.ordering <-
       "ateles fusciceps",
       "mus musculus",
       "spalax carmeli",
-      "nannospalax galili",
+      "spalax galili",
       "bos taurus",
       "canis familiaris",
       "dasypus novemcinctus",

### 2016-04-07 manuscript/manuscript.Rnw
use sans math, black links, and fix noteref

@@
 \usepackage{amsmath}
-% \usepackage{sfmath}
+\usepackage{sansmath}
 \usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
@@
 % \usepackage[noblocks,auth-sc]{authblk}
-\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=blue,urlcolor=blue,citecolor=blue]{hyperref}
+\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue]{hyperref}
 %\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
@@
 %
-\author[addressref={aff1},email={don@donarmstrong.com},noteref={n1}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
-\author[addressref={aff2},email={1mmcgowen1@gmail.com},noteref={n1}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
+\author[addressref={aff1},noteref={n1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
+\author[addressref={aff2},noteref={n1},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
 \author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL}\fnm{Amy} \snm{Weckle}}

### 2016-04-07 manuscript/manuscript.Rnw
clean up typos and figure references

@@
   \parttitle{Results} We measured the expression profiles of
-  \Sexpr{star.logs[,length(unique(file))]} term placenta of
+  \Sexpr{star.logs[,length(unique(file))]} term placentas of
   \Sexpr{star.logs[,length(unique(species))]} species of mammals. We
@@
 quantified using Cufflinks v2.2.1
-\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions}.
+\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
+\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
 %
@@
 species is generally perceived to be $\approx 1$ million years
-\cite{Prüfer.ea2012:bonobogenomecomparedwith}.
+\cite{Prüfer.ea2012:bonobogenomecomparedwith}. %
+%
 
@@
 @ 
-\caption{Species tree for organisms studied}
+\caption{Species tree for organisms studied showing the relationship
+  between species. Silhouettes of species were obtained from
+  \url{http://phylopic.org}. }
 \label{fig:species_tree}
@@
         lineage.table[analysis.name==analysis &
-                      p.value <= 0.05][order(p.value),
+                      p.value <= 0.05 &
+                      fdr <= fdr.cutoff][order(p.value),
                                        list(human_name,p.value,fdr,fc),]
@@
 three of which are members of a distinct expansion of the growth
-hormone gene family within hominoids (\cref{fig:galectins_evolution_tree}). Another member of the growth
+hormone gene family within hominoids (Supplemental \cref{fig:galectins_evolution_tree}). Another member of the growth
 hormone gene family, \ac{GH2} (%
@@
 %
-28 members of this family have $FPKM \ge 1$, and this family of
+28 members of this family have $\text{FPKM} \ge 1$, and this family of
 genes has greatly expanded in ruminants (\cow and \sheep{}) although
@@
 related to the thymus-specific serine protease gene \textit{PRSS16}
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C The other 8
+(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C). The other 8
 \textit{PRSS16}-like genes are also expressed to a lesser degree
@@
 <<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
-print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
-                                                     Species),],
-             aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
+cptl <-
+    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
+                                            Species),]
+cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
+                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
+                                          capfirst(names(species.names))))]
+print(ggplot(cptl,
+             aes(y=human_name,x=Species))
       + geom_raster(aes(fill=log10(FPKM)))
@@
               axis.text.y = element_text(face="italic"),
-              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
+              axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic"))
       )
@@
   \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-  genes expressed (\ac{FPKM}$ \ge 10$) in the placenta of all organisms
-  studied with housekeeping
-  genes \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
-  removed.}
+  1:1 \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
+  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
+  studied with housekeeping genes
+  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
+  ordered (from top to bottom) by median expression. Supplemental
+  \cref{fig:core_placenta_transcriptome_shape} shows the shape of the
+  distribution of the number of genes in the core placenta
+  transcriptome if different FPKM and species restriction criteria are
+  utilized.}
 \label{fig:core_placenta_transcriptome}
@@
                axis.text.y = element_text(face="italic"),
-               axis.text.x = element_text(angle = 300, hjust = 0, vjust=1,face="italic")),
+               axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic")),
          facet_grid(type~.,scale="free_y",space="free_y")
@@
   orthologs of Hormone Genes, Galectins, genes implicated in
-  pre-eclampsia (PE), and \textit{IGF}-related genes}
+  pre-eclampsia (PE), and \textit{IGF}-related genes. Empty locations
+  indicate no detectable expression and/or no 1:1 \human ortholog in
+  that species.}
 \label{fig:hormone_genes}

### 2016-04-07 manuscript/supplemental_material.Rnw
change the default to 0.6

@@
 <<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
-opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.65\\textheight",out.extra="keepaspectratio")
+opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.6\\textheight",out.extra="keepaspectratio")
 opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")

### 2016-04-07 manuscript/supplemental_material.Rnw
fix species ordering and figure sizes

@@
                  axis.text.angle=0,
-                 fontsize=8
+                 fontsize=6.5
                  )
@@
 
-\begin{figure}
-<<fstl1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+\begin{figure}[hbf]
+<<fstl1.evolution.tree,echo=FALSE,out.height="0.6\\textheight",fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
                  min.fpkm=1,
-                 axis.text.size=5,
-                 fontsize=8
+                 axis.text.size=3.5,
+                 fontsize=5
                  )
@@
 
+\clearpage
+
 \subsection{Other Gene Trees}
 
-\begin{figure}
-<<serpine.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+\begin{figure}[hbf]
+<<serpine.evolution.tree,echo=FALSE,out.height="0.6\\textheight",fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
                  min.fpkm=1,
-                 axis.text.size=5,
-                 fontsize=4)
+                 axis.text.size=3.5,
+                 fontsize=3)
 @ 
@@
 \begin{figure}
-<<sslp1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<sslp1.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
                  gene.tree.table,gene.tree.expression,
-                 axis.text.size=5,
-                 fontsize=6)
+                 axis.text.size=3.5,
+                 fontsize=5)
 @ 
@@
 \begin{figure}
-<<estrogen.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<estrogen.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
                  gene.tree.table,gene.tree.expression,
-                 axis.text.size=5,
-                 fontsize=7)
+                 axis.text.size=3.5,
+                 fontsize=4.5)
 @ 
@@
 \begin{figure}
-<<progesterone.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<progesterone.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
                  subtree=c("ENSMODG00000005385","ENSLAFG00000027896"),
-                 axis.text.size=5,
-                 fontsize=6
+                 axis.text.size=3.5,
+                 fontsize=4.5
                  )
@@
 \begin{figure}
-<<chorionic.somatomammotropin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<chorionic.somatomammotropin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
                  gene.tree.table,gene.tree.expression,
-                 axis.text.size=5,
+                 axis.text.size=3.5,
                  fontsize=4)
@@
 \begin{figure}
-<<prolactin.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
                  gene.tree.table,gene.tree.expression,
-                 axis.text.size=5,
-                 fontsize=5)
+                 axis.text.size=3.5,
+                 fontsize=3.5)
 @ 
@@
 \begin{figure}
-<<rln2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<rln2.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
 \begin{figure}
-<<ovos2.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<ovos2.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
                  subtree=c("ENSDNOG00000033320","ENSMODG00000017988"),
-                 axis.text.size=5,
-                 fontsize=5)
+                 axis.text.size=3.5,
+                 fontsize=4)
 @ 
@@
 \begin{figure}
-<<prolactin.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<prolactin.receptor.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
                  subtree=c("ENSG00000113494","ENSMODG00000020376"),
-                 axis.text.size=5,
-                 fontsize=7)
+                 axis.text.size=3.5,
+                 fontsize=4)
 @ 
@@
 \begin{figure}
-<<hoxa11.receptor.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<hoxa11.receptor.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
                  subtree=c("ENSBTAG00000014738","ENSECAG00000013229"),
-                 axis.text.size=5,
-                 fontsize=5
+                 axis.text.size=3.5,
+                 fontsize=4
                  )
@@
 \begin{figure}
-<<hbe1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<hbe1.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
                  gene.tree.table,gene.tree.expression,
-                 axis.text.size=5,
-                 fontsize=5)
+                 subtree=c("ENSBTAG00000037815",
+                           "ENSMUSG00000078621"),
+                 axis.text.size=3.5,
+                 fontsize=4)
 @ 
@@
 \begin{figure}
-<<nppb.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<nppb.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 axis.text.size=3.5,
+                 fontsize=6
+                 )
 @ 
@@
 \begin{figure}
-<<nppc.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<nppc.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 axis.text.size=3.5,
+                 fontsize=3
+                 )
 @ 
@@
 \begin{figure}
-<<galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<galectins.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
-@ 
-\caption{\genetreeexplanation{galectins}}
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 axis.text.size=3.5,
+                 fontsize=3.5
+                 )
+@ 
+\caption{\genetreeexplanation{\textit{LGALS1} and \textit{LGALS2}}}
 \label{fig:galectins_evolution_tree}
@@
 \begin{figure}
-<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 
@@
 gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 axis.text.size=3.5,
+                 fontsize=4.5
+                 )
 @ 

### 2016-04-17 manuscript/supplemental_material.Rnw
call clearpage first

@@
                            capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
-    cat(paste0("\\subsubsection{\\textit{",capfirst(this.species),"} Placental Gene Expression}\n\n"))
+    cat(paste0("\\clearpage\\subsubsection{\\textit{",capfirst(this.species),"} Placental Gene Expression}\n\n"))
     xt.align <- c("r","l","l","p{2.5in}","r")

### 2016-04-18 manuscript/manuscript.Rnw
input.read.length was not numeric; fixes dog read length

@@
 star.logs.wide[,`% good reads`:=100-`% bad reads`]
+star.logs.wide[,input.reads:=as.numeric(input.reads)]
 setkey(hugo.nomenclature,"Approved Symbol")

### 2016-04-18 manuscript/manuscript.Rnw
add Derek suggested changes

@@
   variability of the placenta in the placentas of mammalian lineages.
-  However, the expression of a subset of
+  However, using the strictest filtering strategy the expression of only
   \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
@@
   species analyzed.
+
+\DLA{Include something about pre-eclampsia}
 
@@
 the shape of attachment (i.e., discoid, diffuse, cotyledonary,
-zonary), degree of invasiveness of fetal tissues into maternal tissues
+zonary), degree of intimacy between fetal tissues and maternal tissues
 (i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
@@
 “placental mammals”. The term “placental mammals” is misleading, as
-extant yinotherians (such as \textit{Ornithorhynchus anatinus}, the
+extant monotremes (such as \textit{Ornithorhynchus anatinus}, the
 platypus) possess a rudimentary placenta, as do marsupials, although
@@
  and the expansion of the growth hormone
-and chorionic gonadotropin beta genes in hominoid primates
+and chorionic gonadotropin beta genes in anthropoid primates
 \cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
@@
     \toprule
-    Species                       & Gestation (days) & Litter Size  & Source \\
+    Species                       & Gestation (days) & \multicolumn{1}{p{1in}}{\centering Average Litter Size}  & Source \\
     \midrule
@@
 three of which are members of a distinct expansion of the growth
-hormone gene family within hominoids (Supplemental \cref{fig:galectins_evolution_tree}). Another member of the growth
+hormone gene family within catarrhines (Supplemental \cref{fig:galectins_evolution_tree}). Another member of the growth
 hormone gene family, \ac{GH2} (%
@@
 Other highly expressed genes in the top 50 include members of the
-hemoglobin family (e.g. \textit{HBA1}, \textit{HBB}, \textit{HBG2},
-\textit{HBA2}), the maternally imprinted long noncoding RNA gene
+hemoglobin family (e.g. \textit{HBA1}, \textit{HBA2}, \textit{HBB},
+and \textit{HBG2}), the maternally imprinted long noncoding RNA gene
 \textit{H19}, genes coding for proteins required for growth and/or
@@
 
-We examined the expression of five Laurasiatherians: \cow, \sheep,
+We examined the expression of five laurasiatherians: \cow, \sheep,
 \pig, \horse, and \dog. We were unable to detect any significant
 differences after correcting for multiple testing between
-Laurasiatherians and all other analyzed species
+laurasiatherians and all other analyzed species
 (Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
-Laurasiatherians and Euarchontoglires
+laurasiatherians and Euarchontoglires
 (Supplemental \cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).
@@
 The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Equus Caballus} are shown in
+in the term placenta of \textit{Equus caballus} are shown in
 \cref{tab:top50_horse_placenta} with a list of all highly expressed
@@
 receptor family, and has previously been hypothesized to be important
-for the maintenance of the anti-inflammatory during pregnancy
+for the maintenance of the anti-inflammatory state during pregnancy
 \cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
@@
 different timepoints and in different clades.
+
+\DLA{discuss progesterone withdrawal and the lou muglin review}
 
@@
 The \opossum placenta shows expression of genes that either do not
-exist in eutherians or are not found in the term eutherian placenta
+exist in eutherians or are not expressed in the term eutherian placenta
 (\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
@@
 
+\DLA{Alu was primate specific? Check to see whether this is real.}
+
 \subsection{Core placenta transcriptome}
@@
 \human have the largest brain in terms of relative size and this
-differs markedly from closely related great apes
+differs markedly from closely related apes
 \cite{Boddy.ea2012:Comparativeanalysisofencephalization}. Deep
@@
 can be attributed to the lack of adequate invasion of maternal tissue,
-and is noted by the onset of hypertension and proteinuria
+and is noted by the onset of hypertension and proteinuria %
+%
+\DLA{Discuss the new ACOG requirements for diagnosis which does not
+  require proteinuria.}
+%
 \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
@@
 There is diversity in the placental expression of hormones, including
-estrogen, progesterone, prolactin, relaxin, natriuretic peptides, and
-the growth hormone family amount the species and lineages analyzed. We
-also identified genes differentially expressed on the \human lineage
-which are implicated in pre-eclampsia, including \textit{ADAM12},
-\textit{CRH}, and \textit{KISS1}.
+estrogen, progesterone, prolactin, relaxin, the growth hormone family, and
+natriuretic peptides among the species and lineages analyzed.
+We also identified genes differentially expressed on the \human
+lineage which are implicated in pre-eclampsia, including
+\textit{ADAM12}, \textit{CRH}, and \textit{KISS1}.
 

### 2016-04-18 manuscript/manuscript.Rnw
add in common names

@@
     \toprule
-    Species                       & Gestation (days) & \multicolumn{1}{p{1in}}{\centering Average Litter Size}  & Source \\
+    Species                       & Common Name & Gestation (days) & \multicolumn{1}{p{1in}}{\centering Average Litter Size}  & Source \\
     \midrule
-    \textit{Ateles fusciceps}     & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Bos taurus}           & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Canis familiaris}     & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Dasypus novemcinctus} & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
-    \textit{Equus caballus}       & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
-    \textit{Loxodonta africana}   & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Homo sapiens}         & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Monodelphis domestica}& 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
-    \textit{Mus musculus}         & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Spalax galili}   & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
-    \textit{Ovis aries}           & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Pan paniscus}         & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
-    \textit{Spalax carmeli}       & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
-    \textit{Sus scrofa}           & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Ateles fusciceps}     & spider monkey & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Bos taurus}           & domestic cow & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Canis familiaris}     & domestic dog & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Dasypus novemcinctus} & nine-banded armadillo & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
+    \textit{Equus caballus}       & domestic horse & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
+    \textit{Loxodonta africana}   & African avanna elephant & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Homo sapiens}         & human & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Monodelphis domestica}& gray short-tailed opossum & 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
+    \textit{Mus musculus}         & mouse & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Spalax galili}   & blind mole rat from Upper Galilee & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
+    \textit{Ovis aries}           & domestic sheep & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Pan paniscus}         & bonobo & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
+    \textit{Spalax carmeli}       & blind mole rat from northern Israel & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
+    \textit{Sus scrofa}           & domestic pig & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
     \bottomrule
@@
 
-Fetal tissue was collected from the placenta of eight mammalian
-species (\human [human], \bonobo [bonobo], \dog [domestic dog], \cow
+Fetal tissue was collected from the placenta of nine mammalian
+species (\human [human], \bonobo [bonobo], \smonkey [spider monkey], \dog [domestic dog], \cow
 [domestic cow], \elephant [African savanna elephant], \opossum [gray

### 2016-04-18 manuscript/supplemental_material.Rnw
clearpage after each species, not before

--- 2016-04-07 manuscript/supplemental_material.Rnw
@@
           )
-}
-@
-
-\clearpage
+    cat("\\clearpage\n")
+}
+@
 

### 2016-04-18 manuscript/supplemental_material.Rnw
make the species names in sup match the text

@@
     \textit{Canis familiaris}     & Dog                & CanFam3.1  & 150 \\
-    \textit{Dasypus novemcinctus} & 9-Banded Armadillo & DASNOV3    & 100 \\
+    \textit{Dasypus novemcinctus} & Nine-Banded Armadillo & DASNOV3    & 100 \\
     \textit{Equus caballus}       & Horse              & EquCab2    & 100 \\ 
-    \textit{Loxodonta africana}   & Elephant           & loxAfr3    & 75 \\
+    \textit{Loxodonta africana}   & African savanna elephant           & loxAfr3    & 75 \\
     \textit{Homo sapiens}         & Human              & GRCh38     & 75, 150 \\
-    \textit{Monodelphis domestica} & Opossum           & BROAD05    & 75 \\
+    \textit{Monodelphis domestica} & Gray short-tailed opossum           & BROAD05    & 75 \\
     \textit{Mus musculus}         & Mouse              & GRCm38     & 35,44,150 \\
@@
     \textit{Pan paniscus}         & Bonobo             & CHIMP2.1.4 & 150 \\
-    \textit{Spalax carmeli}       & Blind mole rat     & \textit{Mus musculus} & 100 \\
-    \textit{Spalax galili}   & Blind mole rat     & \textit{Mus musculus} & 100 \\
-    \textit{Sus scrofa}           & Duroc (pig)        & Sscrofa10.2 & 51 \\
+    \textit{Spalax carmeli}       & Blind mole rat from Upper Galilee    & \textit{Mus musculus} & 100 \\
+    \textit{Spalax galili}   & Blind mole rat from Northern Israel    & \textit{Mus musculus} & 100 \\
+    \textit{Sus scrofa}           & Domestic pig (Duroc)        & Sscrofa10.2 & 51 \\
     \bottomrule

### 2016-04-18 manuscript/manuscript.Rnw
fix the commas and input.read.length issue

@@
 star.logs.wide[,input.reads:=as.numeric(input.reads)]
+star.logs.wide[,input.read.length:=as.numeric(input.read.length)]
 setkey(hugo.nomenclature,"Approved Symbol")
@@
 %
-(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])})
-%
-,
+(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])}),%
+%
 the number of trimmed reads ranged from %
@@
 %
-For transcriptomes sequenced with 150 bp, %
+For transcriptomes sequenced with 150 bp %
 % The mismatch here is because we're reporting trimmed reads here, not
 % full length reads
-(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])})
-%
-,
+(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])}),
+%
 the number of trimmed reads ranged from %
@@
 %
-For transcriptomes sequenced with 35 bp, %
+For transcriptomes sequenced with 35 bp %
 % The mismatch here is because we're reporting trimmed reads here, not
 % full length reads
-(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])})
-%
-,
+(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])}),
+%
 the number of trimmed reads ranged from %

### 2016-04-18 manuscript/manuscript.Rnw
captialize common names

@@
     \midrule
-    \textit{Ateles fusciceps}     & spider monkey & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Bos taurus}           & domestic cow & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Canis familiaris}     & domestic dog & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Dasypus novemcinctus} & nine-banded armadillo & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
-    \textit{Equus caballus}       & domestic horse & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
-    \textit{Loxodonta africana}   & African avanna elephant & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Homo sapiens}         & human & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Monodelphis domestica}& gray short-tailed opossum & 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
-    \textit{Mus musculus}         & mouse & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Spalax galili}   & blind mole rat from Upper Galilee & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
-    \textit{Ovis aries}           & domestic sheep & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Pan paniscus}         & bonobo & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
-    \textit{Spalax carmeli}       & blind mole rat from northern Israel & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
-    \textit{Sus scrofa}           & domestic pig & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Ateles fusciceps}     & Spider monkey & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Bos taurus}           & Cow & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Canis familiaris}     & Dog & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Dasypus novemcinctus} & Nine-banded Armadillo & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
+    \textit{Equus caballus}       & Horse & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
+    \textit{Loxodonta africana}   & African savanna elephant & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Homo sapiens}         & Human & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Monodelphis domestica}& Gray short-tailed opossum & 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
+    \textit{Mus musculus}         & Mouse & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Spalax galili}   & Blind mole rat from Upper Galilee & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
+    \textit{Ovis aries}           & Domestic sheep & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Pan paniscus}         & Bonobo & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
+    \textit{Spalax carmeli}       & Blind mole rat from northern Israel & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
+    \textit{Sus scrofa}           & Domestic pig (Duroc) & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
     \bottomrule

### 2016-04-18 manuscript/manuscript.Rnw
center and ragged common name
the second mention of horse is just \horse
document why we're using a ram

@@
 \begin{table}
-  \begin{tabular}{c c c c}
+  \begin{tabular}{c >{\raggedright \centering}p{1in} c c c}
     \toprule
@@
       %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
-      ## sheep
+      ## sheep -- this is a male, the sample is female, but there aren't any females in phylopic
       %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
@@
 The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Equus caballus} are shown in
+in the term placenta of \horse are shown in
 \cref{tab:top50_horse_placenta} with a list of all highly expressed

### 2016-04-18 manuscript/manuscript.Rnw
discuss muglia review of P4

@@
 expressed in \human, \bonobo, \cow, \horse, and \dog, but is not
-highly expressed in \smonkey. The diversity of \textit{CYP19A1} and
-\textit{HSD3B1} expression levels in in primates and ungulates seen
-here points to the need for more complete sampling of placentas at
-different timepoints and in different clades.
-
-\DLA{discuss progesterone withdrawal and the lou muglin review}
+highly expressed in \smonkey, \sheep, \pig, \elephant, \armadillo, or
+\opossum. This is consistent with the previously observed local
+progesterone withdrawal in \sheep, and the leuteo-placental shift of
+progesterone expression in boreoeutheria in contrast to afrotherians
+which do not show such a shift
+\cite{Swaggart.ea2015:Genomicsofpretermbirth}. The diversity of
+\textit{CYP19A1} and \textit{HSD3B1} expression levels in in primates
+and ungulates seen here points to the need for more complete sampling
+of placentas at different timepoints and in different clades.
 

### 2016-04-18 manuscript/manuscript.Rnw
resolve the alu retrotransposon issue

@@
   variability of the placenta in the placentas of mammalian lineages.
-  However, using the strictest filtering strategy the expression of only
+  Furthermore, multiple genes which are associated with pre-eclampsia
+  (\textit{CRH}, \textit{ADAM12}, \textit{KISS1}, \textit{PAPPA},
+  \textit{IL1RL1}, and \textit{PLAC1}) show increased expression on
+  the human lineage. Still, using the strictest filtering strategy the
+  expression we find that a core set of only
   \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
@@
   species analyzed.
-
-\DLA{Include something about pre-eclampsia}
 
@@
 \textit{SSLP1} are also expressed in \cow, \dog, \horse, and \sheep.
-\textit{SSLP1} also contains an Alu retrotransposon and may be
+\textit{SSLP1} also contains an Alu-like retrotransposon and may be
 involved in the development of villi via cell proliferation
@@
 
-\DLA{Alu was primate specific? Check to see whether this is real.}
+% Bovine has their own alu-like sequence: PMID 7835712
 

### 2016-04-18 manuscript/manuscript.Rnw
discuss ACOG2013

@@
 can be attributed to the lack of adequate invasion of maternal tissue,
-and is noted by the onset of hypertension and proteinuria %
-%
-\DLA{Discuss the new ACOG requirements for diagnosis which does not
-  require proteinuria.}
-%
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
+and is noted by the onset of hypertension and proteinuria (or in the
+absence of proteinuria, thrombocytopenia, impaired liver function,
+renal insufficiency, pulmonary edema, or cerebral or visual
+disturbances) %
+\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current,ACOG2013:hypertensionpregnancy}. Some cases of
 pre-eclampsia can develop into eclampsia, a more severe condition

### 2016-04-19 manuscript/manuscript.Rnw
fix grammar error in abstract
switch to SINE repeat from Alu repeat

@@
   \textit{IL1RL1}, and \textit{PLAC1}) show increased expression on
-  the human lineage. Still, using the strictest filtering strategy the
-  expression we find that a core set of only
+  the human lineage. Still, using the strictest filtering strategy
+  we find a core set of only
   \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
@@
 \textit{SSLP1} are also expressed in \cow, \dog, \horse, and \sheep.
-\textit{SSLP1} also contains an Alu-like retrotransposon and may be
+\textit{SSLP1} also contains an SINE repeat and may be
 involved in the development of villi via cell proliferation

### 2016-04-25 poster/placenta_viviparous_poster.Rnw
add start of poster

--- 2016-04-18 manuscript/manuscript.Rnw
@@
-%\documentclass[english,12pt]{article}
-\documentclass[doublespacing]{bmcart_dla}
+\documentclass[a0paper,landscape]{tikzposter}
 \usepackage{fontspec}
-\setmainfont{FreeSans}
+\setmainfont{FreeSerif}
 \setsansfont{FreeSans}
 \setmonofont{FreeMono}
-\setmathrm{FreeSans}
-\setmathsf{FreeSans}
-\setmathtt{FreeSans}
-%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
-\usepackage[letterpaper]{geometry}
 \usepackage{url}
-% \usepackage{fancyhdr}
+\usepackage{fancyhdr}
 \usepackage{graphicx}
-% \usepackage[bf]{caption}
+\usepackage[bf]{caption}
 \usepackage{rotating}
 \usepackage{wrapfig}
+\usepackage{fancybox}
 \usepackage{booktabs}
-\usepackage{float}
+% \usepackage{multirow}
+\usepackage{acronym}
+\usepackage{qrcode}
+\usepackage[backend=biber,natbib=true,hyperref=true,style=nature]{biblatex}
+\addbibresource{references.bib}
+% \usepackage[nomargin,inline,draft]{fixme}
+% \newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
+\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
+\usepackage{texshade}
+\usepackage{tikz}
+\usepackage{nameref}
+\usepackage{zref-xr,zref-user}
+\renewcommand*{\bibfont}{\small}
+
+% The textpos package is necessary to position textblocks at arbitary 
+% places on the page.  Use showboxes option to show outlines of textboxes.
+% \usepackage[absolute]{textpos}
+\usepackage[absolute,showboxes]{textpos}
+
 \usepackage{multirow}
-\usepackage{setspace}
-\usepackage{wrapfig}
-% \usepackage{txfonts}
-%\usepackage[plain]{fancyref}
-\usepackage{acronym}
 \usepackage{array}
-\usepackage{dcolumn}
-\usepackage{adjustbox}
-\usepackage{longtable}
-\usepackage{pdflscape}
-\usepackage{amsmath}
-\usepackage{sansmath}
-\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
-\addbibresource{references.bib}
-% \usepackage[noblocks,auth-sc]{authblk}
-\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue]{hyperref}
-%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
-% Floats at end for journals
-\usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
-\renewcommand{\efloatseparator}{\mbox{}}
-%\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
-%\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
-\usepackage[nomargin,inline,draft]{fixme}
-\newcommand{\DLA}[1]{\textcolor{red}{\fxnote{DLA: #1}}}
-\newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
-% \usepackage[x11names,svgnames]{xcolor}
-% \usepackage{texshade}
-% \usepackage{tikz}
-\usepackage{nameref}
-\usepackage{xspace}
-% this package will stretch the font slightly if necessary.
-\usepackage{microtype}
-\usepackage{zref-xr,zref-user}
-\zxrsetup{toltxlabel}
-\zexternaldocument*{supplemental_material}
-\usepackage[capitalize]{cleveref}
-% \newenvironment{narrow}[2]{%
-%   \begin{list}{}{%
-%       \setlength{\topsep}{0pt}%
-%       \setlength{\leftmargin}{#1}%
-%       \setlength{\rightmargin}{#2}%
-%       \setlength{\listparindent}{\parindent}%
-%       \setlength{\itemindent}{\parindent}%
-%       \setlength{\parsep}{\parskip}}%
-%   \item[]}{\end{list}}
-% \def\newblock{\hskip}
-% \newenvironment{paperquote}{%
-%   \begin{quote}%
-%      \it
-%   }%
-%   {\end{quote}}
-% \renewcommand{\textfraction}{0.15}
-% \renewcommand{\topfraction}{0.85}
-% \renewcommand{\bottomfraction}{0.65}
-% \renewcommand{\floatpagefraction}{0.60}
-%\renewcommand{\baselinestretch}{1.8}
-% \newenvironment{enumerate*}%
-%   {\begin{enumerate}%
-%     \setlength{\itemsep}{0pt}%
-%     \setlength{\parskip}{0pt}}%
-%   {\end{enumerate}}
-% \newenvironment{itemize*}%
-%   {\begin{itemize}%
-%     \setlength{\itemsep}{0pt}%
-%     \setlength{\parskip}{0pt}}%
-%   {\end{itemize}}
-% \setcounter{tocdepth}{3}
-% \oddsidemargin 0.0in 
-% \textwidth 6in
-% \raggedbottom
-% \clubpenalty = 10000
-% \widowpenalty = 10000
-% set a large emergency stretch to avoid issues with overfull boxes
-% \setlength{\emergencystretch}{20em}
-% make the afill smaller
-% \renewcommand\Affilfont{\itshape\small}
-\renewcommand\cite[1]{\autocite{#1}}
-% \pagestyle{fancy}
 
-\input{common_latex}
+\title{The Evolution of Gene Expression in the Term Placenta of
+  Viviparous Mammals}
+
+\author{Don L. Armstrong, Michael R. McGowen, Amy Weckle, Jason
+  Caravas, Dalen Agnew, Sue Savage-Rumbaugh, Eviatar Nevo, Chong J.
+  Kim, Günter P. Wagner, Roberto Romero, and Derek Wildman}
+\institute{Institute for Genomic Biology, Computing Genomes for
+  Reproductive Health, University of Illinois, Urbana-Champaign}
+\usetheme{Default}
+
+% use this
+% http://www.identitystandards.illinois.edu/graphicstandardsmanual/othermedia/posterpresentation.html
+\definecolorstyle{UofI}{
+  \definecolor{boldBlue}{HTML}{002058}
+  \definecolor{boldOrange}{HTML}{E87722}
+  \definecolor{professionalBlue}{HTML}{606EB2}
+  \definecolor{professionalOrange}{HTML}{D45D00}
+  }{
+    \colorlet{backgroundcolor}{professionalBlue}
+    \colorlet{framecolor}{black}
+    \colorlet{titlefgcolor}{boldBlue}
+    \colorlet{titlebgcolor}{white}
+    \colorlet{blockbodybgcolor}{white}
+    \colorlet{blocktitlebgcolor}{professionalOrange}
+    \colorlet{blocktitlefgcolor}{boldBlue}
+    }
+\usecolorstyle{UofI}
+\defineblockstyle{UofIMinimal}{
+    titlewidthscale=1, bodywidthscale=1, titleleft,
+    titleoffsetx=0pt, titleoffsety=0pt, bodyoffsetx=0pt, bodyoffsety=0pt,
+    bodyverticalshift=0pt, roundedcorners=0, linewidth=0.2cm,
+    titleinnersep=1cm, bodyinnersep=1cm
+}{
+    \begin{scope}[line width=\blocklinewidth, rounded corners=\blockroundedcorners]
+       \ifBlockHasTitle %
+           \draw[draw=none,fill=blockbodybgcolor]
+               (blockbody.south west) rectangle (blocktitle.north east);
+           \draw[color=blocktitlebgcolor, loosely dashed]
+               (blocktitle.south west) -- (blocktitle.south east);%
+       \else
+             \draw[draw=none,fill=blockbodybgcolor]
+                 (blockbody.south west) rectangle (blockbody.north east);
+        \fi
+    \end{scope}
+}
+
+\useblockstyle{UofIMinimal}
+\definebackgroundstyle{UofIVertical}{
+    \draw[line width=0pt, bottom color=backgroundcolor, 
+    top color=white] (bottomleft) rectangle (topright);
+}
+\usebackgroundstyle{UofIVertical}
+\definetitlestyle{UofITitle}{
+    width=\paperwidth, roundedcorners=0, linewidth=0pt, innersep=12mm,
+    titletotopverticalspace=5mm, titletoblockverticalspace=10mm,
+    titlegraphictotitledistance=10pt
+}{}
+\usetitlestyle{UofITitle}
+
+\makeatletter
+\renewcommand\TP@maketitle{%
+   \begin{minipage}{0.8\textwidth}
+        \color{titlefgcolor}
+        {\bfseries \Huge \sc \@title \par}
+        \vspace*{1em}
+        {\huge \@author \par}
+        \vspace*{1em}
+        {\LARGE \@institute}
+    \end{minipage}%
+    \hfill
+    \begin{minipage}{0.2\linewidth}
+      \includegraphics[width=0.8\linewidth,height=6cm,keepaspectratio]{figures/igb_wordmark}
+    \end{minipage}
+}
+\makeatother
 
 \begin{document}
-\raggedright
 
@@
 
-% Acronyms
 \input{acronyms}
 
-\begin{frontmatter}
-\begin{fmbox}
-\dochead{Research}%
-\title{The evolution of gene expression in the term placenta of viviparous mammals}%
-%
-\author[addressref={aff1},noteref={n1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
-\author[addressref={aff2},noteref={n1},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
-\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL}\fnm{Amy} \snm{Weckle}}
-\author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}
-\author[addressref={aff4},email={agnewd@dcpah.msu.edu}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
-\author[addressref={aff5},email={kbenirschke@ucsd.edu}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
-\author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
-\author[addressref={aff11},email={nevo@research.haifa.ac.il}]{\inits{E}\fnm{Eviatar} \snm{Nevo}}
-\author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
-\author[addressref={aff9},email={gunter.wagner@yale.edu}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
-\author[addressref={aff7},email={prbchiefstaff@med.wayne.edu}]{\inits{R}\fnm{Roberto} \snm{Romero}}
-\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
-\address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[id=aff2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
-\address[id=aff3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[id=aff4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
-\address[id=aff5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
-\address[id=aff6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
-\address[id=aff7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
-\address[id=aff8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[id=aff9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
-\address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[id=aff11]{The Institute of Evolution, University of Haifa, Haifa, Israel}
+\maketitle
 
-\begin{artnotes}
-\end{artnotes}
+\begin{columns}
 
-\begin{artnotes}
-\note[id=n1]{Equal contributor} % note, connected to author
-\end{artnotes}
+  \column{0.25}
 
-\end{fmbox}
-% \twocolumn
+  \block{Introduction}{
 
-\begin{abstractbox}
-\begin{abstract}
-% \section{Abstract}
-  \parttitle{Background} The placenta is arguably the most
-  anatomically variable organ in mammals. Understanding gene
-  expression patterns among placentas from multiple species
-  illuminates the molecular basis of placenta function as well
-  obstetrical syndromes affecting humans.
+%     Our approach identified MIR1914, MIR126, SNORD12, and SNORD5 as
+%     uterus specific. CGB, LGALS13, and members of the PSG family were
+%     identified as placenta specific. The accuracy of the classifier
+%     was verified using RNAseq transcriptome data not in the training
+%     set. We show the applicability of the classifier to simulated EVs
+%     by down sampling reads to $10^5$ reads and biasing the reads to
+%     account for preferential association of RNA motifs with EV
+%     packaging machinery. When coupled with single-EV sequencing, our
+%     classifier enables clinicians to monitor disease-specific markers
+%     over the course of treatment without biopsies, increasing the
+%     ability of clinicians to adapt treatment modalities to the
+%     individual patient. 
+  }
 
-  \parttitle{Results} We measured the expression profiles of
-  \Sexpr{star.logs[,length(unique(file))]} term placentas of
-  \Sexpr{star.logs[,length(unique(species))]} species of mammals. We
-  found lineage-specific differential expression of prolactin,
-  relaxin, and galectins as well as increased expression of the
-  pre-eclampsia associated genes \textit{ADAM12}, \textit{CRH}, and
-  \textit{PAPPA} in humans. We also identified genes which are
-  expressed in all placentas, including genes with immune-modulating
-  properties (\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and
-  \textit{S100A10}) and cell-cell interactions (\textit{LAMC1},
-  \textit{LUM}, and \textit{LGALS1}).
+  \block{Methods}{     
+}
 
-  \parttitle{Conclusions} A diverse array of genes are differentially
-  expressed and are likely to play a role in the anatomical
-  variability of the placenta in the placentas of mammalian lineages.
-  Furthermore, multiple genes which are associated with pre-eclampsia
-  (\textit{CRH}, \textit{ADAM12}, \textit{KISS1}, \textit{PAPPA},
-  \textit{IL1RL1}, and \textit{PLAC1}) show increased expression on
-  the human lineage. Still, using the strictest filtering strategy the
-  expression we find that a core set of only
-  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-  genes is essential for the proper functioning of the placenta in all
-  species analyzed.
-
-\textbf{Keywords:} mammal, gene expression, evolution, RNA-seq, placenta
-
-\end{abstract}
-\end{abstractbox}
-\end{frontmatter}
-
-\section{Background}
-
-The main role of the placenta in viviparous mammals is the transfer of
-nutrients, oxygen, and waste between mother and offspring. Although
-these functions are conserved among mammals, the placenta is perhaps
-the most variable organ in terms of structure and form
-\cite{Mossman.1937:comparativemorphogenesisofthe,Moffett.ea2006:Immunologyofplacentationin}. Species differ in
-the shape of attachment (i.e., discoid, diffuse, cotyledonary,
-zonary), degree of intimacy between fetal tissues and maternal tissues
-(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
-interdigitation of maternal-fetal tissue (i.e., villous, folded,
-trabecular, lamellar, labrynthine)
-\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
-mammals differ drastically in the number and size of their offspring,
-as well as the length of gestation, traits often associated with
-specific aspects of placental morphology
-\cite{Capellini.ea2011:Placentationandmaternalinvestment}
-(\cref{tab:gestational_periods}).
-
-The placenta arose at some point before the origin of extant mammals,
-a group that includes marsupials, monotremes, and eutherian or
-“placental mammals”. The term “placental mammals” is misleading, as
-extant monotremes (such as \textit{Ornithorhynchus anatinus}, the
-platypus) possess a rudimentary placenta, as do marsupials, although
-the direct apposition to maternal tissues lasts for a proportionally
-shorter time during gestation
-\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
-Marsupials also differ from eutherian mammals in that most possess
-only a choriovitelline placenta, produced from the fusion of the
-chorion and yolk sac, during most of gestation. Although some
-eutherian mammals may possess a choriovitelline placenta at some point
-in development, the major site of nutrient exchange in placentals
-occurs via the chorioallantoic placenta, produced from the fusion of
-the chorion and allantois
-\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
-the first placenta was more like that of marsupials or eutherians;
-however, using phylogenetic methods, the ancestral eutherian was
-reconstructed as possessing a relatively invasive hemochorial
-placenta, as in humans and mouse
-\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
-
-Although the morphology of the mammalian placenta has been well
-characterized in a number of species
-\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
-molecular environment is only beginning to be understood from a
-genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
-Studies of certain model species (i.e., \textit{Homo sapiens},
-\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
-divergence in the genetics of the placenta, especially in later stages
-of pregnancy
-\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
-For example, each lineage has evolved expansive placenta-specific gene
-families with expression patterns that can vary in space and time
-\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
-These include independent expansions of the prolactins in \mouse,
-\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
-in \cow, genes related to the thymus-specific serine protease
-\textit{PRSS16} in \textit{Loxodona africana} (in this work),
- and the expansion of the growth hormone
-and chorionic gonadotropin beta genes in anthropoid primates
-\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
-Expression patterns of specific genes also can vary widely between
-species. For example, the expression of orthologous genes in the
-placenta have been found to vary substantially between \human and
-\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
-recruitment of \textit{SIGLEC6} to placental expression in \human
-differs from both \textit{Pan troglodytes} and \mouse
-\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
-
-One explanation for the divergent paths of the placental interface is
-the distinct evolutionary interplay between maternal and fetal
-strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}.
-The maternal strategy is to requisition an adequate degree of
-nutrients to the fetus without risking the health of the mother, while
-the fetal strategy via paternal genes is to extract as many resources
-from the mother as possible. Therefore, an interplay between these
-extremes can set up an intricate “arms race”, involving the imprinting
-of parental alleles, expansion of gene families, and enlistment of
-existing genes to expression at the maternal-fetal interface
-\cite{Haig1993:Geneticconflictsinhuman}.
-
-One symptom of this maternal-fetal conflict is its potential affect on
-health
-\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
-addition to having an invasive hemochorial placenta, \human possess a
-deeply invasive trophoblast layer that can penetrate the myometrium of
-the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
-Defects in placental growth are known to be associated with the
-devastating condition of pre-eclampsia, which can cause hypertension,
-proteinuria, and has the potential to lead to death of the mother
-and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
-Although cases have been documented in \textit{P. troglodytes}
-\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
-\textit{Gorilla gorilla}
-\cite{Thornton.ea1992:Convulsionsinpregnancyin}, pre-eclampsia has a
-far higher incidence in \human pregnancies with as much as 4\% of
-pregnancies being affected in the United States
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. \textit{P.
-  troglodytes} and \textit{G. gorilla} also have deeply invasive
-trophoblasts, although the progression of invasion in great apes may
-be slower than \human
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. However,
-large-scale expression studies of great ape placentas have not been
-conducted. The increasing affordability of next-generation sequencing
-technology has enabled the large-scale study of specific tissues from
-a wide array of species to address broader phylogenetic questions in
-the evolution of gene expression
-\cite{Brawand.ea2011:evolutionofgeneexpression}.
-
-Here we use RNA-Seq
-technology to sequence and quantify the mRNA environment of mammalian
-placentas at term, including that of the \human, \textit{Pan paniscus}
-(the sister species to \textit{P. troglodytes}), \textit{Ateles
-  fusciceps}, \mouse, \galili, \carmeli, \textit{Canis familiaris},
-\cow, \elephant, \textit{Dasypus novemcinctus}, and
-\textit{Monodelphis domesticus}. These species represent four major
-branches of eutherian mammals (Euarchontoglires, Laurasiatheria,
-Afrotheria, and Xenartha) as well as marsupials. These placentas also represent a
-wide array of placental morphologies as well as traits such as number
-and size of offspring and length of gestation
-(\cref{tab:gestational_periods}). We use this data, along with
-existing RNA-Seq data from \mouse
-\cite{Wang.ea2011:surveyfornovelimprinted}, \human
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \textit{Equus caballus}
-\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis aries}
-\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \textit{Sus scrofa}
-\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
-differences between expression in marsupial and eutherian mammals at
-term, identify the evolution of expression in important placental
-genes such as hormones and galectins, identify novel lineage-specific
-expression patterns, and quantify \human specific differences in gene
-expression that may be related to obstetrical disorders such as
-pre-eclampsia.
-
-\begin{table}
-  \begin{tabular}{c >{\raggedright \centering}p{1in} c c c}
-    \toprule
-    Species                       & Common Name & Gestation (days) & \multicolumn{1}{p{1in}}{\centering Average Litter Size}  & Source \\
-    \midrule
-    \textit{Ateles fusciceps}     & Spider monkey & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Bos taurus}           & Cow & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Canis familiaris}     & Dog & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Dasypus novemcinctus} & Nine-banded Armadillo & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
-    \textit{Equus caballus}       & Horse & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
-    \textit{Loxodonta africana}   & African savanna elephant & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Homo sapiens}         & Human & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Monodelphis domestica}& Gray short-tailed opossum & 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
-    \textit{Mus musculus}         & Mouse & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Spalax galili}   & Blind mole rat from Upper Galilee & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
-    \textit{Ovis aries}           & Domestic sheep & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Pan paniscus}         & Bonobo & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
-    \textit{Spalax carmeli}       & Blind mole rat from northern Israel & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
-    \textit{Sus scrofa}           & Domestic pig (Duroc) & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \bottomrule
-  \end{tabular}
-  \caption{Gestational periods in days and average liter sizes for studied species.}
-  \label{tab:gestational_periods}
-\end{table}
-
-\section{Methods}
-
-\subsection{Collection of Placental Tissue and RNA Extraction}
-
-Fetal tissue was collected from the placenta of nine mammalian
-species (\human [human], \bonobo [bonobo], \smonkey [spider monkey], \dog [domestic dog], \cow
-[domestic cow], \elephant [African savanna elephant], \opossum [gray
-short-tailed opossum], \galili [blind mole rat from Upper Galilee],
-\carmeli [blind mole rat from northern Israel]
-\cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}). Information
-on specimen ID, and collector is shown for each sample in Supplemental
-\cref{tab:srr_numbers}. In the case of \bonobo, \cow, and \elephant,
-the placenta was collected at term upon expulsion from the mother
-following delivery. Placental tissue from \human was collected upon
-delivery via a cesarean section. Near-term placental tissue from \dog
-was collected upon the death of the mother from a car accident.
-Placental tissue from \opossum was sampled at fetal stage 32, roughly
-one day prior to birth. Only the fetal portion of the placenta was
-sampled at the maternal-fetal interface (i.e. villous portion for
-\human, \bonobo, \dog, and \elephant; cotyledon for \cow; diffuse
-placenta for \opossum). All tissues were collected and stored in
-RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation could
-be performed. Tissues were homogenized in Trizol, and total RNA was
-isolated following the manufacturer's instructions. RNA was further
-purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction
-with the RNase-Free DNase Set (Qiagen, Valencia, CA), following the
-manufacturer’s protocol to remove any contaminating DNA from the
-sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
-Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
-Clara, CA) to confirm concentration and quality of the isolation. RNA
-quality was assessed using A260/A280 and A260/A230 absorbance values.
-
-\subsection{Sequencing}
-
-All RNA-Seq mRNA isolation, library construction, and Illumina
-sequencing was performed by the Applied Genomics Technology Center
-(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
-total RNA and the standard Illumina protocol. Paired-end sequencing
-was performed using the Illumina Genome Analyzer II with an insert
-size of 230 base pairs (bp). Read lengths for species are shown in
-Supplemental \cref{tab:ref_genome_aligned_species}.
-
-\subsubsection{Additional Sequences}
-
-For more complete evolutionary comparisons additional sequences of
-placenta in \mouse
-(n=\Sexpr{star.logs[grepl("musculus",species),length(unique(file))]})
-\cite{Wang.ea2011:surveyfornovelimprinted}, \human
-(n=\Sexpr{star.logs[grepl("sapiens",species),length(unique(file))]})
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
-\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
-\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
-\cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
-\ac{SRA} (Supplemental \cref{tab:srr_numbers}) and analyzed using the
-same methodology as RNAs sequenced specifically for this study.
-
-\subsection{Transcriptome Assembly, Alignment, and Quantification}
-
-The quality of reads was evaluated using FASTQC v0.11.2
-(\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
-Transcriptomes without a suitable reference genome (\smonkey, \galili,
-\carmeli; italic reference in Supplemental
-\cref{tab:ref_genome_aligned_species}) were assembled using Trinity
-v2014-07-17 \cite{Haas.ea2013:Denovotranscriptsequence} and annotated
-using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
-either the \human or \mouse transcriptome (Supplemental
-\cref{tab:ref_genome_aligned_species}). For annotation using Diamond,
-the match with the highest bitscore was chosen as the annotation;
-transcripts where the highest scoring match had a bitscore below 40
-were left unannotated.
-% 
-All transcriptomes were then aligned to the appropriate reference (or
-assembled) genome using STAR v2.4.2a
-\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
-values in \ac{FPKM}
-\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
-quantified using Cufflinks v2.2.1
-\cite{Trapnell.ea2009:TopHatdiscoveringsplicejunctions} (Supplemental
-\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}).
-%
-The genome of the closely-related \textit{P. troglodytes} was used as
-a reference for \bonobo. The separation between the two recognized Pan
-species is generally perceived to be $\approx 1$ million years
-\cite{Prüfer.ea2012:bonobogenomecomparedwith}. %
-%
+\column{0.25}
 
@@
 
-\subsection{Exploration of Individual Transcriptome Data}
+\column{0.25}
 
-To determine enrichment levels of functional categories of genes in
-highly-expressed transcripts in each transcriptome, we performed
-functional annotation cluster analyses per transcriptome using topGO
-\cite{topGO:2.20.0}. Gene Ontology clusters were considered
-significantly enriched if the \ac{FDR} as estimated by the Benjamini
-and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
-$p$ values obtained from the Fisher Exact Test were less than or equal
-to 0.05. Enrichment tests were considered with two gene lists per
-transcriptome: annotated genes with expression levels of the top 5\%
-of the frequency distribution of \ac{FPKM} (Supplemental
-\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
-the top 100 genes with the highest \ac{FPKM} (Supplemental
-\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
-Due to the paucity of annotated gene ontology terms for some species,
-gene lists were compared using the \human Gene Ontology using the 1:1
-\human ortholog. Enrichment was considered out of the set of all 1:1
-orthologs present in each species.
+\column{0.25}
 
-\subsection{Comparative Analysis of Gene Expression}
-
-All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
-\cow, \dog, \elephant, \mouse, and \human were determined using
-Ensembl v.80 gene trees via Biomart
-\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
-package
-\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
-A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
-  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
-in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
-  sapiens" & mean_fpkm >= 100,length(unique(human_name))]} 1:1
-orthologous genes with \ac{FPKM} $\ge 100$ were retained for future
-analyses of gene expression evolution described below.
-
-Significant shifts in expression levels between groups of species were
-calculated using Student's t test on 1:1 orthologs using the mean of
-each species to avoid large biases towards samples with $n \gg 1$ and
-considered significant if the \ac{FDR} as estimated by the Benjamini
-and Hochberg procedure \cite{Benjamini1995:bh_procedure} was
-$\le 0.05$. For \human and \mouse lineage specific calculations, a
-type-III anova was calculated using
-lme4~\cite{Bates.ea2015:FittingLinearMixed-EffectsModels} with each
-species having separate random effects.
-
-\subsection{Sequences, FPKMs, and Code}
-
-All of the code used to produce every analysis presented in this paper
-is available at
-\url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
-sequences used are deposited in SRA, and new sequences for this work
-can be obtained under the series
-\href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
-(Supplemental \cref{tab:srr_numbers}). The mean FPKM for gene in each
-species with ortholog annotations are located in the supplemental data
-file all\_species\_mean\_fpkm.txt.xz and the per-sample FPKM values are in
-all\_species\_per\_sample\_fpkm.txt.xz.
-
-\section{Results}
-
-\subsection{Transcriptome Statistics}
-
-Supplemental \cref{fig:transcriptome_statistics} shows the RNAseq read and
-alignment statistics for all species. For transcriptomes sequenced
-with 75 bp %
-%
-(\Sexpr{species(star.logs.wide[input.read.length > 70 & input.read.length < 80,unique(species)])}),%
-%
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,max(as.numeric(input.reads))]}
-with an average of \Sexpr{star.logs.wide[input.read.length > 70 & input.read.length < 80,mean(`% good reads`)]}\% of reads 
-mapping to the respective reference genome (\cref{tab:ref_genome_aligned_species}). 
-%
-For transcriptomes sequenced with 150 bp %
-% The mismatch here is because we're reporting trimmed reads here, not
-% full length reads
-(\Sexpr{species(star.logs.wide[input.read.length > 120 & input.read.length < 160,unique(species)])}),
-%
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length  > 120 & input.read.length < 160,max(as.numeric(input.reads))]}
-%
-with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
-of reads mapping to the
-respective reference genome (\cref{tab:ref_genome_aligned_species}).
-%
-%
-For transcriptomes sequenced with 35 bp %
-% The mismatch here is because we're reporting trimmed reads here, not
-% full length reads
-(\Sexpr{species(star.logs.wide[input.read.length < 40 ,unique(species)])}),
-%
-the number of trimmed reads ranged from %
-\Sexpr{star.logs.wide[input.read.length < 40 ,min(as.numeric(input.reads))]} 
-%
-to %
-\Sexpr{star.logs.wide[input.read.length  < 40  ,max(as.numeric(input.reads))]}
-%
-with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
-of reads mapping to the
-respective reference genome (\cref{tab:ref_genome_aligned_species}).
-The top 50 expressed genes in each genome analyzed are shown in
-\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
-and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.
-
-% The percentage of bases that mapped to Ensembl annotated mRNA ranged
-% from ~35\% in elephant to ~85\% in human, with percentages of bases
-% mapping to annotated protein-coding segments of genes ranging from
-% ~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
-% from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
-% from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
-% ranged from ~43\% in elephant to 80\% in bonobo. The number of
-% annotated transcripts to which reads mapped in each respective genomes
-% ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
-% and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
-% greater degree of mapping to annotated protein-coding sequences in
-% humans corresponds to the greater number of Ensembl annotated
-% sequences in the human genome compared to the other five reference
-% genomes (Flicek, 2012).
-
-<<top.genes.function,echo=FALSE,results="hide">>=
-top.genes.table <- function(species.name,n=50) {
-    alignment.species <- species.name
-    if (alignment.species=="pan paniscus") {
-        alignment.species <- "pan troglodytes"
-    } else if (alignment.species=="ateles fusciceps") {
-        alignment.species <- "homo sapiens"
-    } else if (grepl("spalax",alignment.species)) {
-        alignment.species <- "mus musculus"
-    }
-    temp <- 
-        combined.fpkm[species==species.name &
-                      !(!is.na(human_name) & 
-                        human_name %in% housekeeping.genes.superset[,gene_short_name]) &
-                      ! (gene_biotype %in% c("Mt_tRNA","Mt_rRNA","rRNA")) &
-                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_name) &
-                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_alignment_symbol) &
-                      !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[lL]?(?:-.+)?$|rRNA$)",
-                             gene_short_name)
-                      ][order(-mean_fpkm)][,list(name_or_id,
-                                                 human_alignment_symbol,
-                                                 human_name,mean_fpkm,sd_fpkm,n_fpkm)]
-    ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
-    temp <- temp[!(is.na(name_or_id)|name_or_id==""),]
-    temp <- temp[!duplicated(name_or_id)][1:n]
-    temp[,used_alignment:=FALSE]
-    temp[is.na(human_name) & !is.na(human_alignment_symbol),
-                    used_alignment:=TRUE]
-    temp[is.na(human_name),human_name:=human_alignment_symbol]
-    setkey(temp,"human_name")
-    temp <- 
-        hugo.nomenclature[temp][order(-mean_fpkm),
-                                list(name_or_id,used_alignment,`Approved Symbol`,`Approved Name`,
-                                     mean_fpkm,sd_fpkm,n_fpkm)]
-    temp[!is.na(name_or_id),
-         name_or_id:=paste0("\\href{http://www.ensembl.org/",
-                             capfirst(gsub(" ","_",alignment.species)),
-                             "/Gene/Summary?g=",name_or_id,"}{\\textit{",
-                             xtable.sanitize.text(name_or_id),"}}")]
-    temp[!is.na(`Approved Name`),
-         `Approved Name`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Name`),"}")]
-    temp[!is.na(`Approved Symbol`),
-         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{\\textit{",xtable.sanitize.text(`Approved Symbol`),"}}")]
-    temp[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
-    temp[,used_alignment:=NULL]
-    temp[,sem_fpkm:=as.numeric(NA)]
-    temp[n_fpkm>1,sem_fpkm:=sd_fpkm/sqrt(n_fpkm)]
-    temp[,sd_fpkm:=NULL]
-    temp[,n_fpkm:=NULL]
-    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM","SEM"))
-    if (temp[,all(is.na(`SEM`))]) {
-        temp[,`SEM`:=NULL]
-    }
-    if (species.name=="homo sapiens" || species.name=="ateles fusciceps") {
-        temp[,`Human Symbol`:=NULL]
-    }
-    temp[,`FPKM`:=round(`FPKM`)]
-    return(temp)
-}
-gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE) {
-    species.name <- gsub("_"," ",species.name)
-    fpkm <- combined.fpkm[species==species.name &
-                              (gene_short_name==gene | 
-                               gene_id==gene),
-                          mean_fpkm][1]
-    if (is.na(fpkm)) {
-        stop("FPKM is NA!!!!")
-    }
-    if (units || complete)
-        fpkm <- formatC(fpkm,digits=0,format="f")
-    if (units) {
-        fpkm <- paste0(fpkm,
-                       "\\ \\text{FPKM}")
-    }
-    if (complete) {
-        if (any(grepl("^ENS.+G\\d+$",gene))) {
-            fpkm <- paste0("\\text{",gene,"} = ",
-                           fpkm)
-        } else {
-            fpkm <- paste0("\\text{\\textit{",gene,"}} = ",
-                           fpkm)
-        }
-    }
-    if (units || complete) {
-        fpkm <- paste0("$",fpkm,"$")
-    }
-    return(fpkm)
+\block{Conclusions}{
 }
 
-@
+\block{References}{
 
-<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("placenta_lineage_specific","human_mouse_lineage_specific")))>>=
-load("../data/placenta_lineage_specific")
-load("../data/human_mouse_lineage_specific")
-source("/home/don/projects/xtable/pkg/R/print.xtable.R")
-
-placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=5,orderbyfc=FALSE,fdr.cutoff=1) {
-    temp.table <- 
-        lineage.table[analysis.name==analysis &
-                      p.value <= 0.05 &
-                      fdr <= fdr.cutoff][order(p.value),
-                                       list(human_name,p.value,fdr,fc),]
-    temp.table <- data.table(temp.table)
-    setkey(temp.table,"human_name")
-    temp.table <- 
-        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
-                                            `Approved Name`,p.value,fdr,fc)]
-    setnames(temp.table,
-              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
-              )
-    setkey(temp.table,"Human Symbol")
-    if (output) {
-        display.table <- temp.table[`FDR`<=fdr.cutoff,][order(`p (t-test)`),]
-        if (orderbyfc) {
-            display.table <- temp.table[order(-abs(`Fold Change`)),]
-        }
-        display.table <- display.table[1:members,]
-
-        display.table <-
-            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
-                                                  "}")]
-        setnames(display.table,"Fold Change","$\\text{log}_2$ FC")
-        print(xtable(display.table,
-                     display=c("s","s","s","e","g","f"),
-                     align=c("c","c","p{2in}","r","r","r")
-                     ),
-              booktabs=TRUE,
-              floating=FALSE,
-              math.style.exponents=TRUE,
-              sanitize.text.function=genes.table.sanitize,
-              include.rownames=FALSE)
-
-    }
-    return(invisible(temp.table[order(`p (t-test)`),]))
-}
-output.top.genes.table <- function(data,output=TRUE) {
-    if (output) {
-        table.display <- c("s","s","s","d")
-        table.align <- c("c","c","p{3in}","r")
-        if (any(colnames(data) %in% "SEM"))  {
-            table.display <- c(table.display,"d")
-            table.align <- c(table.align,"r")
-        }
-        if (any(colnames(data) %in% "Human Symbol"))  {
-            table.align <- c("c",table.align)
-            table.display <- c("s",table.display)
-        }
-        print(xtable(data,
-                     display=table.display,
-                     align=table.align
-                     ),
-              booktabs=TRUE,
-              floating=FALSE,
-              sanitize.text.function=genes.table.sanitize,
-              include.rownames=FALSE)
-    }
+  \begin{center}
+    \mbox{}\vspace{-\baselineskip}
+    \printbibliography[heading=none]
+  \end{center}
 }
 
-@
-
-\subsection{Primate Placenta Transcriptomes}
-
-\begin{table}
-  \footnotesize
-<<primate.lineage.specific,echo=FALSE,results="asis">>=
-pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
-setkey(pri.all.lin.spec,"Human Symbol")
-@ 
-\caption{\lineagetableexplanation{primates}{all other analyzed species}}
-\label{tab:primate_lineage_specific}
-\end{table}
-
-We examined the expression of three primates, \human, \bonobo, and
-\smonkey. The genes whose expressions are different on the primate
-lineage in comparison to all other analyzed species are listed in
-\cref{tab:primate_lineage_specific}. \textit{FSTL1} has significantly
-higher expression in the primate lineage
-(Supplemental \cref{fig:fstl1_evolution_tree},
-$\text{FDR}=\Sexpr{pri.all.lin.spec["FSTL1",`FDR`]}$), is implicated
-in BMP signaling, and has previously been implicated in inflammation
-and invasiveness
-\cite{Mouillet.ea2015:expressionandpost-transcriptionalregulation}. Of
-non-primates, \horse shows the highest expression
+\block{Code and Datasets}{
+The code and underlying data for this poster can be found at
+\begin{center}
+\qrcode[padding]{http://dla2.us/p/pv2016}
+\url{http://dla2.us/p/pv2016}
+\end{center}
 % 
-\Sexpr{gene.expression(species.name="equus caballus",gene="FSTL1",complete=TRUE)}, 
-%
-which is still lower than the lowest primate, \smonkey
-%
-(\Sexpr{gene.expression(species.name="ateles fusciceps",gene="FSTL1",complete=TRUE)}). 
-%
-No other genes were significantly differentially expressed on this
-lineage by t-test.
-
-\subsubsection{\human Placenta Transcriptome}
-\label{sec:human_placenta}
-
-\begin{table}
-  \footnotesize
-<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="homo sapiens"))
-@ 
-\caption{The top 50 genes by \ac{FPKM} in \human which are not
-  annotated as Mt\_rRNA, Mt\_tRNA, rRNA or a housekeeping gene. Gene
-  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
-  name corresponding to the human symbol. A complete list of all genes
-  is in \cref{tab:all_homo_sapiens_placenta}.}
-\label{tab:top50_human_placenta}
-\end{table}
-
-<<cgb.min.expression,results="hide",echo=FALSE,warnings=FALSE>>=
-cgb.min.expression <- 
-    min(sapply(paste0("CGB",c("","1","2","5","7","8")),
-               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
-cgb.sum.expression <- 
-    sum(sapply(paste0("CGB",c("","1","2","5","7","8")),
-               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
-@
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \human are shown in
-\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
-in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
-the highest \ac{FPKM} value
-%
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
-%
-in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
-three of which are members of a distinct expansion of the growth
-hormone gene family within catarrhines (Supplemental \cref{fig:galectins_evolution_tree}). Another member of the growth
-hormone gene family, \ac{GH2} (%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
-is also expressed to a high degree in the \human placenta
-(\cref{tab:top50_human_placenta}).
-% 
-Eleven \acp{PSG} are highly-expressed
-(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
-greatest degree of expression %
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
-and the pseudogene \textit{PSG10P} showing the least %
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
-These
-genes are members of a hominoid-specific expansion of the \ac{CEA} family
-and known to be expressed by the syncytiotrophoblast
-\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
-%
-Genes involved in the production of estrogens (%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
-% 
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
-% 
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
-% 
-) and progesterone (%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
-\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
-\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
-%
-show exceptionally high expression in the human placenta, as well as %
-%
-\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
-%
-which generates the alpha subnit of multiple human chorionic
-gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
-producing the beta subunits of the chorionic gonadotropins
-(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
-\textit{CGB7}, \textit{CGB8}) have reasonable expression
-($\text{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
-expression is on the order of the expression of
-\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
-%
-%
-Other highly expressed genes in the top 50 include members of the
-hemoglobin family (e.g. \textit{HBA1}, \textit{HBA2}, \textit{HBB},
-and \textit{HBG2}), the maternally imprinted long noncoding RNA gene
-\textit{H19}, genes coding for proteins required for growth and/or
-cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
-\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
-glucose transport (\textit{SLC2A1}) among others
-(\cref{tab:top50_human_placenta}, Supplemental
-\cref{tab:all_homo_sapiens_placenta}).
-
-% \paragraph{Pre-eclampsia \human Lineage Specific Genes}
-
-\begin{table}
-  \footnotesize
-<<human.lineage.specific,echo=FALSE,results="asis">>=
-hum.all.lin.spec <-
-    placenta.lineage.specific.table("hum.all",
-                                    lineage.table=human.mouse.lineage.specific.analysis,
-                                    members=40,
-                                    orderbyfc=TRUE,
-                                    fdr.cutoff=0.001
-                                    )
-setkey(hum.all.lin.spec,"Human Symbol")
-@ 
-\caption{\lineagetableexplanationtypeiii{\human}{all other analyzed species}}
-\label{tab:human_lineage_specific}
-\end{table}
-
-Using a mixed-effects linear model to account for repeated testing of
-\human and \mouse, we identified multiple genes which are
-differentially expressed in the \human lineage in comparison to all
-other lineages (\cref{tab:human_lineage_specific}). These genes
-include multiple genes which are thought to be involved in
-pre-eclampsia. Notably, \textit{CRH}
-(FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
-FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
-\cref{fig:crh_evolution_tree}), has been shown to be more highly
-expressed in placentas from pre-eclamptic pregnancies
-\cite{Petsas.ea2012:Aberrantexpressionofcorticotropin-releasing,Paiva.ea2011:MeasurementofmRNAtranscripts},
-where it is hypothesized to activate decidual macrophages. Likewise,
-\textit{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
-FC=\Sexpr{hum.all.lin.spec["KISS1",`Fold Change`]}, Supplemental
-\cref{fig:kiss1_evolution_tree}) also shows increased expression in
-pre-eclamptic placentas
-\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}, and is
-involved in modulating trophoblast invasion. Two other genes,
-\textit{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
-FC=\Sexpr{hum.all.lin.spec["PAPPA",`Fold Change`]} Supplemental
-\cref{fig:pappa_evolution_tree}) and \textit{ADAM12}
-(\Sexpr{hum.all.lin.spec["ADAM12",`FDR`]},
-FC=\Sexpr{hum.all.lin.spec["ADAM12",`Fold Change`]}, Supplemental
-\cref{fig:adam12_evolution_tree}) are also differentially expressed
-but are negatively correlated with pre-eclampsia
-\cite{Spencer.ea2008:ADAM12sinmaternalserum}.
-
-\begin{figure}
-<<psg.ctsl.evolution.tree,echo=FALSE,fig.width=16,fig.height=16,warning=FALSE,message=FALSE>>=
-plot.label.size <- 3
-axis.text.size <- 3.5
-
-pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2),clip="off"))
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"),
-                 axis.text.size=axis.text.size,
-                 fontsize=plot.label.size,
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"),
-                 fontsize=4,
-                 width=3,
-                 axis.text.size=axis.text.size,
-                 offset.ratio=1
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"),
-                 fontsize=4,
-                 width=3,
-                 axis.text.size=axis.text.size,
-                 offset.ratio=1.2
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 subtree=c("ENSSSCG00000006792",
-                           "ENSBTAG00000046085"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"),
-                 axis.text.size=axis.text.size,
-                 fontsize=plot.label.size,
-                 offset.ratio=0.6
-                 )
-
-popViewport(1)
-grid.text(c("A","B","C","D"),
-          x=unit(c(0,0.5,0,0.5),"npc"),y=unit(c(1,1,0.5,0.5),"npc"),
-          just=c("left","top"),
-          gp=gpar(fontsize="48")
-          )
-@ 
-\caption{\genetreeexplanation{(A) \textit{PSG1}, (B) \textit{CTSL}, (C) \textit{PRSS16}, and (D) \textit{PAG2}}}
-\label{fig:pag2_prss16_ctsl_psg_evolution_tree}
-\end{figure}
-
-\subsubsection{\bonobo Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
-top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
-output.top.genes.table(top.bonobo.genes)
-setkey(top.bonobo.genes,"Gene symbol")
-@ 
-\caption{\tableexplanation{\bonobo}{tab:all_pan_paniscus_placenta}}
-\label{tab:top50_bonobo_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
-with a list of all genes in Supplemental
-\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
-placenta differs somewhat from that of \human, as evidenced by the
-makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
- in \bonobo is %
-\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
-%
-which is involved in insulin growth factor signaling and angiogenesis
-during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
-\textit{PRG2} is also expressed at relatively high levels (though
-comparatively much lower) in the human placenta
-(\Sexpr{gene.expression("homo sapiens","PRG2")};
-Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
-related to the chorionic somatomammotropins
-are also expressed at relatively high levels (\textit{GH1} 
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
-\textit{GH1} 
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
-\textit{GH2}
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
-and \textit{GH2}
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
-%
-\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
-Placental expression of the genes involved in estrogen and
-progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
-respectively) show lower expression in \bonobo as compared to \human %
-(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
-\cref{fig:estrogen_evolution_tree}; %
-\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
-\cref{fig:progesterone_evolution_tree} and
-\cref{tab:top50_human_placenta}) although the expression of
-\textit{HSD3B1} is fairly similar to that of \smonkey %
-(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
-In addition, \textit{CGA} is expressed in the bonobo placenta %
-(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
-lower levels than \human as well (\cref{tab:top50_human_placenta}).
-%
-Two Ensembl-annotated beta subunits of the chorionic gonadotropin
-genes are expressed in \bonobo, one related to \textit{CGB} and
-another related to \textit{CGB7}; both are not highly expressed
-(\cref{fig:cga_cgb_evolution_tree}). Among the other top 50 expressed
-genes, there is high expression of keratins (e.g., \textit{KRT19},
-\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
-\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
-Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
-2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
-the hormone adrenomedullin (\textit{ADM}).
-
-\subsubsection{\smonkey Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
-top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
-output.top.genes.table(top.spider.monkey.genes)
-setkey(top.spider.monkey.genes,"Gene symbol")
-@ 
-\caption{\tableexplanation{\smonkey}{tab:all_ateles_fusciceps_placenta}}
-\label{tab:top50_spider_monkey_placenta}
-\end{table}
-
-For the term placenta of \smonkey, assembled contigs from RNA-seq were
-aligned using against \human and the best alignment was chosen as the
-annotation for each assembled contig. In the term placenta of
-\smonkey, the top 50 non-housekeeping genes
-(\cref{tab:top50_spider_monkey_placenta}) include a possible regulator
-of androgen receptor signaling
-\cite{Sampson.ea2012:PAGE4positivityisassociated}, \textit{PAGE4} %
-(\Sexpr{gene.expression("ateles_fusciceps","PAGE4")}), %
-the pregnancy specific glycoprotein \textit{PSG8}
-(\Sexpr{gene.expression("ateles_fusciceps","PSG8")}), %
-only one member of the chorionic somatotropins, \textit{GH1} %
-(\Sexpr{gene.expression("ateles_fusciceps","GH1")}), %
-a few HLA genes (\textit{HLA-A}, \textit{HLA-F},
-\cref{tab:top50_spider_monkey_placenta}), and multiple components of
-annexin complexes
-(\Sexpr{gene.expression("ateles_fusciceps","ANXA2",complete=TRUE)}, %
-\Sexpr{gene.expression("ateles_fusciceps","S100P",complete=TRUE)}, %
-\Sexpr{gene.expression("ateles_fusciceps","S100A11",complete=TRUE)}) %
-which may play a role in angiogenesis, and certainly regulate
-cell-cell interaction in the developing placenta
-\cite{Liu.ea2015:AnnexinA2complexeswith}.
-\Sexpr{gene.expression("ateles_fusciceps","AL365202.1",complete=TRUE)}
-is also highly expressed, has no known function, and is duplicated
-multiple times in \textit{Macaca mulatta} (Ensembl gene tree
-\href{http://useast.ensembl.org/Multi/GeneTree/Image?gt=ENSGT00580000081348}{ENSGT00580000081348}).
-All genes with expression greater than 50 \ac{FPKM} are listed in
-Supplementary \cref{tab:all_ateles_fusciceps_placenta}.
-
-\subsection{Rodentia Placenta Transcriptomes}
-
-We examined the expression of three rodents, \mouse, \carmeli, and
-\galili{}. We did not detect any significant changes on the Rodentia
-lineage in comparison to all other analyzed species (Supplemental
-\cref{tab:rodentia_lineage_specific}), likely because of a lack power
-to overcome the multiple testing correction. Additional species and/or
-samples in each of these lineages will be necessary to detect lineage
-specific shifts.
-
-\subsubsection{\mouse Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="mus musculus"))
-@ 
-\caption{\tableexplanation{\mouse}{tab:all_mus_musculus_placenta}}
-\label{tab:top50_mouse_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \mouse are shown in
-\cref{tab:top50_mouse_placenta} with a list of all genes with
-$\text{FPKM} \ge 50$ in Supplemental
-\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
-\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
-target \textit{in silico} is the orphan G-coupled receptor
-\textit{Mrgprg}
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
-\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
-\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
-of which are maternally imprinted (the later of which is a homolog of
-human \textit{H19\_2}),
-\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
-cathepsins and a rodentia specific expansion of prolactin related
-genes. There are also two miRNA which overlap with the of the 7S
-signal recognition particle
-(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
-and
-\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
-and numerous pseudogenes
-(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
-appear to be highly expressed, however, the short read length (34--44)
-in many samples is likely preventing unambiguous assignment of all
-reads to the more lengthy complete transcript, so these shorter
-transcripts should be treated with suspicion, and we will ignore them
-in further discussion. The prolactin-related genes
-(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
-\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
-very highly expressed in \mouse in, with %
-\Sexpr{combined.fpkm[species=="mus musculus" &
-  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
-prolactin related genes having $\text{FPKM} \ge 1$
-(\cref{fig:prolactin_evolution_tree}, Supplemental
-\cref{fig:prolactin_receptor_evolution_tree}). The cathespins have
-also expanded in rodents, with six showing high expression
-(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
-\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
-\cref{tab:top50_mouse_placenta,fig:pag2_prss16_ctsl_psg_evolution_tree}).
-Finally, \textit{Peg10}, an imprinted gene required for proper
-formation of the placenta in mice is also highly expressed
-(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
-reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.
-
-% \paragraph{\mouse Lineage Specific Genes}
-
-\begin{table}
-  \footnotesize
-<<mouse.lineage.specific,echo=FALSE,results="asis">>=
-mou.all.lin.spec <-
-    placenta.lineage.specific.table("mou.all",
-                                    lineage.table=human.mouse.lineage.specific.analysis,
-                                    members=40,
-                                    orderbyfc=TRUE,
-                                    fdr.cutoff=0.001
-                                    )
-setkey(mou.all.lin.spec,"Human Symbol")
-@ 
-\caption{\lineagetableexplanationtypeiii{\mouse}{all other analyzed species}}
-\label{tab:mouse_lineage_specific}
-\end{table}
-
-Using a mixed-effects linear model to account for repeated testing of
-\human and \mouse, we identified multiple genes which are
-differentially expressed in the \mouse lineage in comparison to all
-other lineages (\cref{tab:mouse_lineage_specific}). \mouse
-\textit{Gkn2} (FDR=\Sexpr{mou.all.lin.spec["GKN2",`FDR`]},
-FC=\Sexpr{mou.all.lin.spec["GKN2",`Fold Change`]}) has been shown to
-have a pattern of expression distinct from the \human pattern, and has
-been suggested to be important for placental development
-\cite{Fahlbusch.ea2015:Trophoblastexpressiondynamicsof}. Notable
-mouse-specific increases in expression include include genes like
-\textit{Slc38a4} (FDR=\Sexpr{mou.all.lin.spec["SLC38A4",`FDR`]},
-FC=\Sexpr{mou.all.lin.spec["SLC38A4",`Fold Change`]}) and \textit{Afp}
-(FDR=\Sexpr{mou.all.lin.spec["AFP",`FDR`]},
-FC=\Sexpr{mou.all.lin.spec["AFP",`Fold Change`]}) which are implicated
-in fetal growth restriction
-\cite{Sharma.ea2016:Tpbpa-Cre-mediateddeletionofTFAP2C} and placenta
-abnormalities in humans
-\cite{Waller.ea1996:associationbetweenmaternalserum}, and
-\textit{Cnx2} (FDR=\Sexpr{mou.all.lin.spec["CNX2",`FDR`]},
-FC=\Sexpr{mou.all.lin.spec["CNX2",`Fold Change`]}) which is critical
-for the exchange of metabolites in the placenta labyrinth
-\cite{Gabriel.ea1998:Transplacentaluptakeofglucose}.
-
-\subsubsection{\galili Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="spalax galili"))
-@ 
-\caption{\tableexplanation{\galili}{tab:all_spalax_galili_placenta}}
-\label{tab:top50_n_galili_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \galili are shown in
-\cref{tab:top50_n_galili_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_spalax_galili_placenta}.
-Components of annexin complexes are highly expressed
-(\Sexpr{gene.expression("spalax_galili","S100a6",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","S100a11",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","S100a10",complete=TRUE)},
-and
-\Sexpr{gene.expression("spalax_galili","Anxa2",complete=TRUE)}),
-as are members of the thymosin family
-(\Sexpr{gene.expression("spalax_galili","Tmsb4x",complete=TRUE)}
-and
-\Sexpr{gene.expression("spalax_galili","Tmsb10",complete=TRUE)}),
-various cytoskeleton related proteins
-(\Sexpr{gene.expression("spalax_galili","Actc1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Actg2",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Acta2",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Myl9",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Vim",complete=TRUE)}), and
-transmembrane protein of currently unknown function, 
-\Sexpr{gene.expression("spalax_galili","Gm10134",complete=TRUE)}.
-
-\subsubsection{\carmeli Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="spalax carmeli"))
-@ 
-\caption{\tableexplanation{\carmeli}{tab:all_spalax_carmeli_placenta}}
-\label{tab:top50_s_carmeli_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \carmeli are shown in
-\cref{tab:top50_s_carmeli_placenta} with a list of all highly
-expressed genes in Supplemental
-\cref{tab:all_spalax_carmeli_placenta}. Numerous hemoglobin subunits
-ware highly expressed
-(\Sexpr{gene.expression("spalax_carmeli","Hbb-bh1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Hba-a1",complete=TRUE)}, and
-\Sexpr{gene.expression("spalax_carmeli","Hbb-bs",complete=TRUE)}), as
-well as the relaxin
-\Sexpr{gene.expression("spalax_carmeli","Rln1",complete=TRUE)},
-multiple cathespins
-(\Sexpr{gene.expression("spalax_carmeli","Ctsm",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","BC051665",complete=TRUE)},
-and \Sexpr{gene.expression("spalax_carmeli","Ctsl",complete=TRUE)},
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), and many prolactin-related genes
-(\Sexpr{gene.expression("spalax_carmeli","Prl4a1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Prl6a1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Prl3b1",complete=TRUE)}, and
-\Sexpr{gene.expression("spalax_carmeli","Prl7c1",complete=TRUE)},
-\cref{fig:prolactin_evolution_tree}) which shows a similar pattern of
-expansion seen in \mouse.
-
-\subsection{Laurasiatheria}
-
-We examined the expression of five laurasiatherians: \cow, \sheep,
-\pig, \horse, and \dog. We were unable to detect any significant
-differences after correcting for multiple testing between
-laurasiatherians and all other analyzed species
-(Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
-laurasiatherians and Euarchontoglires
-(Supplemental \cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).
-
-\subsubsection{\dog Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="canis familiaris"))
-@ 
-\caption{\tableexplanation{\dog}{tab:all_canis_familiaris_placenta}}
-\label{tab:top50_dog_placenta}
-\end{table}
-
-The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
-in the term placenta of \dog are shown in
-\cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
-\cref{tab:all_canis_familiaris_placenta}. The gene with the highest
-expression %
-(\Sexpr{gene.expression("canis familiaris","TMSB10")}) is the thymosin 
-\textit{TMSB10} %
-(\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
-is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
-among the top 50 (\cref{tab:top50_dog_placenta}) include genes
-involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
-anticoagulation (\textit{TFPI2}, \textit{PLA2G7}), innate immunity
-(\textit{LCN2}, \textit{GM2A}), lipid metabolism (\textit{APOE}),
-inhibition of matrix metalloproteinases (\textit{TIMP2},
-\textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
-\textit{HTRA1}). The gene \textit{LGALS1}, which shows evidence of
-immunosupression during pregnancy
-\cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
-(\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
-In addition, an ortholog of \textit{SSLP1} (ENSCAFG00000010638) a gene
-expressed in the placentas of \cow (%
-\Sexpr{gene.expression("bos taurus","SSLP1",complete=TRUE)}, %
-\cref{tab:top50_cow_placenta}) and \sheep
-(\Sexpr{gene.expression("ovis aries","SOLD1",complete=TRUE)},
-\cref{tab:top50_sheep_placenta}) but is not present in primates
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}
-(\cref{fig:sslp1_evolution_tree}), is also expressed at high levels
-%
-(\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
-There is no placental expression of the estrogen synthase gene
-(\textit{CYP19A1}) in \dog 
-(\Sexpr{gene.expression("canis familiaris","CYP19A1")}
-\cref{fig:estrogen_evolution_tree}), 
-and expression of \textit{HSD3B7} is also low
-(\Sexpr{gene.expression("canis familiaris","HSD3B7")}, 
-\cref{fig:progesterone_evolution_tree}).
-
-\subsubsection{\cow Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="bos taurus"))
-@ 
-\caption{\tableexplanation{\cow}{tab:all_bos_taurus_placenta}}
-\label{tab:top50_cow_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
-list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
-The gene with the highest expression was \textit{TMSB10} %
-(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
-\textit{TMSB4} was also highly expressed %
-(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
-\dog{}, members of annexin complexes were also highly expressed
-(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
-at least one of a set of four miRNA with 100\% identity
-(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
-%
-As these are novel miRNA, there is a possibility that the assignment
-of the short reads to these four miRNA is due to miss-annotation of
-the \cow genome.
-% 
-Several members of the ruminant-specific pregnancy-associated
-glycoprotein family \cite{Green.ea2000:Pregnancy-associatedbovineandovine} 
-show especially high expression in \cow placenta
-(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D, and
-Supplemental \cref{tab:all_bos_taurus_placenta}) including
-%
-\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
-%
-28 members of this family have $\text{FPKM} \ge 1$, and this family of
-genes has greatly expanded in ruminants (\cow and \sheep{}) although
-the expression levels of this family are higher in \cow than in \sheep
-(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). Other genes with high expression
-include genes involved with regulation of growth (\textit{IGF2},
-\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
-and \textit{CSH1}, a gene that is a member of a ruminant specific
-expansion of placental prolactins and not directly related to the gene
-of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
-ruminant-specific placental prolactins also show high expression in
-the cow placenta including %
-\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
-with a total of 14 with $\text{FPKM} \ge 1$.
-%
-\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
-cotyledonary villi of the ruminiant placenta and involved in cell
-proliferation
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
-shows evidence of high expression %
-(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
-There is some placental expression of the estrogen synthase gene
-(\textit{CYP19A1}) in cows %
-(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
-comparatively greater expression of \textit{HSD3B} %
-(\Sexpr{gene.expression("bos taurus","HSD3B")}).
-
-\subsubsection{\pig Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="sus scrofa"))
-@ 
-\caption{\tableexplanation{\pig}{tab:all_sus_scrofa_placenta}}
-\label{tab:top50_pig_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Sus scrofa} are shown in
-\cref{tab:top50_pig_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_sus_scrofa_placenta}. The top two
-expressed transcripts,
-\Sexpr{gene.expression("sus_scrofa","ENSSSCG00000025162",complete=TRUE)}
-and
-\Sexpr{gene.expression("sus_scrofa","ENSSSCG00000021356",complete=TRUE)},
-are miRNA which correspond to the \cow miRNA
-\href{http://www.mirbase.org/cgi-bin/mirna_entry.pl?acc=MI0013080}{bta-mir-2904-1}
-and
-\href{http://www.mirbase.org/cgi-bin/mirna_entry.pl?acc=MIMAT0013845}{bta-mir-2887-1},
-respectively, both of which are also expressed in \cow ovary
-\cite{Hossain.ea2009:Identificationandcharacterizationof}, although
-neither of which have an obvious target gene on the basis of
-\textit{in silico} analysis
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}. Other highly
-expressed genes include members of the insulin growth factor family
-(\Sexpr{gene.expression("sus_scrofa","IGF2",complete=TRUE)} and
-\Sexpr{gene.expression("sus_scrofa","IGFBP3",complete=TRUE)}), members
-of annexin complexes
-(\Sexpr{gene.expression("sus_scrofa","S100A1",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","ANXA2",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","S100A12",complete=TRUE)}, and
-\Sexpr{gene.expression("sus_scrofa","S100A10",complete=TRUE)}), the
-galectin \Sexpr{gene.expression("sus_scrofa","LGALS3",complete=TRUE)}
-and multiple cytoskeleton and extracellular matrix proteins
-(\Sexpr{gene.expression("sus_scrofa","KRT19",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","COL1A1",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","BETA-ACTIN",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","KRT9",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","CLDN4",complete=TRUE)}, and
-\Sexpr{gene.expression("sus_scrofa","KRT9",complete=TRUE)}).
-
-\subsubsection{\horse Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="equus caballus"))
-@ 
-\caption{\tableexplanation{\horse}{tab:all_equus_caballus_placenta}}
-\label{tab:top50_horse_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \horse are shown in
-\cref{tab:top50_horse_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_equus_caballus_placenta}. The gene
-with the highest expression was the relaxin,
-\Sexpr{gene.expression("equus_caballus","RLN",complete=TRUE)},
-followed by two miRNA
-(\Sexpr{gene.expression("equus_caballus","ENSECAG00000025328",complete=TRUE)}
-and
-\Sexpr{gene.expression("equus_caballus","ENSECAG00000027056",complete=TRUE)})
-with no known \textit{in silico} targets
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}, the thymosin
-\Sexpr{gene.expression("equus_caballus","TMSB4",complete=TRUE)},
-members of the insulin growth factor family
-(\Sexpr{gene.expression("equus_caballus","IGF2",complete=TRUE)} and
-\Sexpr{gene.expression("equus_caballus","IGFBP3",complete=TRUE)}) and
-multiple members of annexin complexes
-(\Sexpr{gene.expression("equus_caballus","S100A6",complete=TRUE)},
-\Sexpr{gene.expression("equus_caballus","S100A1",complete=TRUE)}, and
-\Sexpr{gene.expression("equus_caballus","ANXA2",complete=TRUE)}),
-similar to the pattern of expression seen in \pig and \dog.
-
-\subsubsection{\sheep Placenta Transcriptome}
-
-\begin{table}[h]
-  \footnotesize
-<<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
-top.sheep.genes <- top.genes.table(species.name="ovis aries",n=200)
-top.sheep.genes <-
-    top.sheep.genes[!grepl("\\{(7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\}",
-                           `Gene symbol`) &
-                    !grepl("\\((7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\)",
-                           `Human Symbol`)
-                   ,][1:50,]
-output.top.genes.table(top.sheep.genes)
-@ 
-\caption{\tableexplanation{\sheep}{tab:all_ovis_aries_placenta}}
-\label{tab:top50_sheep_placenta}
-\end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \sheep are shown in
-\cref{tab:top50_sheep_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_ovis_aries_placenta}. The complete
-list of highly expressed genes includes many snRNAs and rRNAs likely
-due to the RNA extraction and library preparation procedure utilized
-for this sample. Ignoring these, notable highly expressed genes
-include members of annexin complexes
-(\Sexpr{gene.expression("ovis_aries","S100G",complete=TRUE)},
-\Sexpr{gene.expression("ovis_aries","ANXA2",complete=TRUE)}),
-hemoglobin subunits
-(\Sexpr{gene.expression("ovis_aries","ENSOARG00000010944",complete=TRUE)}),
-an ortholog of \textit{SSLP1},
-\Sexpr{gene.expression("ovis_aries","SOLD1",complete=TRUE)}, two
-cathespins (\Sexpr{gene.expression("ovis_aries","CTSL",complete=TRUE)}
-and \Sexpr{gene.expression("ovis_aries","CTSB",complete=TRUE)}), and
-placentally expressed transcript 1
-\Sexpr{gene.expression("ovis_aries","PLET1",complete=TRUE)} which may
-play a role in cellular migration
-\cite{Raymond.ea2010:Expressionoforphanprotein}.
-
-\subsection{Atlantogenata (Afrotheria + Xenartha)}
-\subsubsection{\elephant Placenta Transcriptome}
-
-\begin{table}
-\footnotesize
-<<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="loxodonta africana"))
-@ 
-\caption{\tableexplanation{\elephant}{tab:top50_elephant_placenta}}
-\label{tab:top50_elephant_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \elephant are shown in \cref{tab:top50_elephant_placenta}
-with a list of all genes in
-\cref{tab:all_loxodonta_africana_placenta}. The gene with the highest
-expression was the natriuretic peptide precursor B
-(\Sexpr{gene.expression("loxodonta_africana","NPPB",complete=TRUE)}).
-Two thymosins,
-\Sexpr{gene.expression("loxodonta_africana","TMSB10",complete=TRUE)}
-and
-\Sexpr{gene.expression("loxodonta_africana","TMSB4Y",complete=TRUE)}
-were also also was highly expressed. The miRNA mir-675
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027406",complete=TRUE)}).
-which is imbedded in the first exon of the maternally imprinted long
-non-coding RNA gene \textit{H19}
-\cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene which is
-currently un-annotated in \elephant{}. %
-% \DLA{Probably consider fixing the unannotation of H19; it seems pretty
-%   obvious to me, though it would require fixing the GTF file for
-%   elephant with the right
-%   \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan
-%     prediction} for this region} %
-Five uncharacterized genes are expressed to a high degree in the
-\elephant placenta (\cref{tab:top50_elephant_placenta}), three of
-which
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000028493",complete=TRUE)},
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032011",complete=TRUE)},
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000031933",complete=TRUE)})
-are members of an \elephant{}-specific expansion of 11 genes closely
-related to the thymus-specific serine protease gene \textit{PRSS16}
-(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C). The other 8
-\textit{PRSS16}-like genes are also expressed to a lesser degree
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000026505",complete=TRUE)}).
-Two other novel genes with high expression are related to the gene
-\textit{SSLP1}
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032659",complete=TRUE)}
-and
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000029551",complete=TRUE)},
-\cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
-\dog placentas (\cref{tab:top50_dog_placenta,tab:top50_cow_placenta}).
-Placental expression of the hormones relaxin (\textit{RLN}, %
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000017411")})
-and prolactin (\textit{PRL}) are exceptionally high in the elephant
-when compared to other mammals
-(\cref{fig:rln2_evolution_tree,fig:prolactin_evolution_tree}). There
-is little to no placental expression of the estrogen synthase gene,
-\textit{CYP19A1}
-(\Sexpr{gene.expression("loxodonta_africana","CYP19A1")}) and the
-enzyme that produces progesterone, \textit{HSD3B1}
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027896")}). Other genes
-with high expression in the elephant placenta include genes involved
-in oxidative stress
-(\Sexpr{gene.expression("loxodonta_africana","TXN",complete=TRUE)}),
-growth
-(\Sexpr{gene.expression("loxodonta_africana","IGF2",complete=TRUE)})
-and
-\Sexpr{gene.expression("loxodonta_africana","PAEP",complete=TRUE)}, a
-gene predominantly expressed in the endometrium of the uterus in humans
-\cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.
-
-\subsubsection{\armadillo Placenta Transcriptome}
-
-\begin{table}
-\footnotesize
-<<top.50.armadillo.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="dasypus novemcinctus"))
-@ 
-\caption{\tableexplanation{\armadillo}{tab:all_dasypus_novemcinctus_placenta}}
-\label{tab:top50_armadillo_placenta}
-\end{table}
-
-The top 50 expressed genes in the term placenta of \armadillo are
-shown in \cref{tab:top50_armadillo_placenta} with a list of all genes
-with $\text{FPKM}\ge 50$ in Supplemental
-\cref{tab:all_dasypus_novemcinctus_placenta}. The gene with the
-highest expression is
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045029",complete=TRUE)}
-which is most similar to the \mouse \textit{Pga5} is expanded in
-\opossum into three members which are expressed in placenta (the other
-two being
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033885",complete=TRUE)}
-and
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000040431",complete=TRUE)}),
-though the expansion is less than that seen \cow and \sheep
-(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). There is also high expression of
-hemoglobin subunit beta,
-\Sexpr{gene.expression("dasypus_novemcinctus","HBB",complete=TRUE)},
-an un-annotated gene with high similarity to \human and \cow
-\textit{OVOS2}
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033320",complete=TRUE)}
-(Supplemental \cref{fig:ovos2_evolution_tree}), the growth factors
-\Sexpr{gene.expression("dasypus_novemcinctus","IGF2",complete=TRUE)}
-and
-\Sexpr{gene.expression("dasypus_novemcinctus","IGFBP4",complete=TRUE)},
-and the thymosin
-\Sexpr{gene.expression("dasypus_novemcinctus","TMSB4X",complete=TRUE)}.
-As has been seen previously, multiple components of the annexin
-complex are also highly expressed, including
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045634",complete=TRUE)}
-(\textit{ANXA8L1}),
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000017399",complete=TRUE)}
-(\textit{S100B}),
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000048504",complete=TRUE)}
-(\textit{S100A10}), and
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000004237",complete=TRUE)}
-(\textit{ANXA2}).
-
-\subsection{Metatherians}
-
-\subsubsection{\opossum Placenta Transcriptome}
-
-\begin{table}
-  \footnotesize
-<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="monodelphis domestica"))
-@ 
-\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica_placenta}}
-\label{tab:top50_opossum_placenta}
-\end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the \opossum are shown in
-\cref{tab:top50_opossum_placenta} with a list of all genes in
-Supplemental \cref{tab:all_monodelphis_domestica_placenta}. The gene with the
-second highest expression is superoxide dismutase 3, extracellular
-(\Sexpr{gene.expression("monodelphis_domestica","SOD3",complete=TRUE)}),
-which functions as an antioxidant. The gene with the second highest
-expression is the beta hemoglobin subunit epsilon
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000011325",complete=TRUE)}),
-which is expressed at the embryonic stage of development in
-marsupials, as in other mammals
-\cite{Holland.ea1997:Characterizationofembryonicglobin}. \textit{HBE}
-subunits link up with protein products of two hemoglobin, zeta
-(\textit{HBZ}) genes, which are also expressed to a high degree in the
-opossum placenta
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016635",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016631",complete=TRUE)}).
-In addition, there is high expression of the gene hemoglobin, omega
-(\textit{HBO},
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016637",complete=TRUE)}),
-% this is LOC_1000016060 which I figured out by tblastn'ing AAK11486.1
-% against Monodelphis domestica
-a hemoglobin found only in marsupials and monotremes
-\cite{Opazo.ea2008:Genomicevidenceforindependent} (Supplemental
-\cref{fig:hbe1_evolution_tree}). The hemoglobin gene \textit{HBA} is
-also expressed in the placenta
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016633",complete=TRUE)}).
-Two cystatin genes
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018295",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018296",complete=TRUE)})
-are expressed to a high degree, along with one \textit{CTSL} related
-gene
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000002844",complete=TRUE)},
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), which are known to interact with
-cystatins \cite{Turk.ea2001:Lysosomalcysteineproteasesfacts}. The
-uncharacterized gene \textit{LOC10001195}
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
-also shows high expression in \opossum placenta. The gene
-\textit{WFDC2}, a member of the whey acidic protein (WAP) family is
-also expressed to a high degree in \opossum placenta
-(\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).
-Other genes with high expression include members of the annexin
-complex
-(\Sexpr{gene.expression("monodelphis_domestica","S100A6",complete=TRUE)},
-\Sexpr{gene.expression("monodelphis_domestica","ANXA2",complete=TRUE)}).
-In addition, there was little production of CYP19A1 and HSD3B1
-(\Sexpr{gene.expression("monodelphis_domestica","CYP19A1",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000005385",complete=TRUE)},
-respectively).
-
-\subsection{Core Placenta Transcriptome}
-
-\begin{figure}
-<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
-cptl <-
-    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
-                                            Species),]
-cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
-                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
-                                          capfirst(names(species.names))))]
-print(ggplot(cptl,
-             aes(y=human_name,x=Species))
-      + geom_raster(aes(fill=log10(FPKM)))
-      + scale_x_discrete(expand=c(0,0))
-      + scale_y_discrete(expand=c(0,0))
-      + scale_fill_gradient(low = grad.def[["low"]],
-                            high = grad.def[["high"]],
-                            space = grad.def[["space"]],
-                            na.value = grad.def[["na.value"]],
-                            limits=c(0,4),
-                            guide=guide_colorbar(title=expression(log[10](FPKM)))
-                            )
-      + ylab("Human Ortholog Name")
-      + xlab("Species")
-      + theme_grey()
-      + theme(legend.position="top",
-              axis.ticks = element_blank(),
-              axis.text.y = element_text(face="italic"),
-              axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic"))
-      )
-@ 
-\caption{Core Placenta Transcriptome.
-  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-  1:1 \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
-  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
-  studied with housekeeping genes
-  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
-  ordered (from top to bottom) by median expression. Supplemental
-  \cref{fig:core_placenta_transcriptome_shape} shows the shape of the
-  distribution of the number of genes in the core placenta
-  transcriptome if different FPKM and species restriction criteria are
-  utilized.}
-\label{fig:core_placenta_transcriptome}
-\end{figure}
-
-The
-\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-genes comprising the core placenta transcriptome are shown in
-\cref{fig:core_placenta_transcriptome} ordered by their median
-transcription level in \ac{FPKM}. We considered a gene to be core to
-the placenta transcriptome if a 1:1 ortholog of that gene was
-expressed with $\text{FPKM} \ge 10$ in all species studied and was
-not a human housekeeping gene according to
-\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
-Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
-which would be included in the core placenta set if the \ac{FPKM}
-cutoff was set differently.
-%
-Multiple components of annexin complexes are included in the core
-transcriptome, including \textit{ANXA2}, \textit{ANXA1},
-\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
-resolution of the inflammation response through inhibition of
-phospholipase A2 activity and signaling through the formyl peptide
-receptor family, and has previously been hypothesized to be important
-for the maintenance of the anti-inflammatory state during pregnancy
-\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
-known to interact with S100A10, S100A6, S100A11, and S100A4
-\cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
-its interaction with S100P as well; it is involved in cell-cell
-interaction as well as vesicle trafficking and von Willibrand Factor
-secretion. The expression regulation of another core gene,
-\textit{TIMP3}, has been implicated in pre-eclampsia
-\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
-Supplemental \cref{tab:core_placenta_go_terms} lists the significant
-enrichment of \ac{GO} terms in the core placenta transcriptome;
-notable terms include those involved in cytoskeletal structure,
-protein transport, membrane protein structure, and inflammation
-resolution. Increasing the permissiveness of the core placenta
-transcriptome by only requiring 10 of 14 species to express 1:1
-orthologs with FPKM $>10$ and results in the same terms also being
-significantly enriched, along with additional terms (Supplemental
-\cref{tab:core_placenta_go_terms}).
-
-% SPINT2 -- Serine Peptidase inhibitor
-
-% SLC3A2 -- this is probably actually a housekeeping gene; it's
-% involved in amino acid transport
-
-% SQSTM1 -- NFκB signaling (immune supression?)
-
-% KRTCAP2  N-glycosylation (probably housekeeping)
-
-% NRDG1 -- interacts with HSP90
-
-% SDC4 -- proteoglycan (syndecan) cell adhesion
-
-% LAMC1 -- laminin
-
-% LUM -- proteoglycan
-
-% VDAC1 -- mitochondrial gene?
-
-% F11R -- tight junctions
-
-% SERPINH1? with TIMP3 
-% http://www.proteinatlas.org/ENSG00000149257-SERPINH1/tissue/placenta
-
-\subsubsection{\ac{GSE} Analyses of Individual Transcriptomes}
-
-In order to identify major functional categories of genes expressed in
-the individual transcriptomes, we examined \ac{GSE} in 1:1 human
-orthologs which had \ac{FPKM} values in the top 1\% of expressed genes
-(Supplemental %
-\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris}) using topGO \cite{topGO:2.20.0}.
-%
-%
-Significant enrichment in genes ($\text{FDR}\le0.05$) in the top 1\%
-of \ac{FPKM} values in at least 12 of the
-\Sexpr{length(species.names)} species included clusters such as
-membrane organization , locomotion, tissue development, extracellular
-matrix organization, regulation of peptidase activity, and various
-cell death related pathways among others
-(Supplemental \cref{tab:go_terms_all_species}). If we instead examine the top 100
-genes by \ac{FPKM} in each of the \Sexpr{length(species.names)}
-species (Supplemental %
-\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris})
-notable enrichments include extracellular matrix, female pregnancy,
-integrin and collagen-related processes, and various cell death
-related processes.
-
-\clearpage
-
-\section{Discussion}
-
-\subsection{Evolution of Placental Hormones}
-
-Reproductive hormones are central to the regulation of pregnancy and
-their source, timing, and quantity can vary widely from species to
-species \cite{Carter2012:Evolutionofplacentalfunction}. Expression and synthesis of hormones is
-perhaps the most well-documented physiological process in the placenta
-in many species. Here we quantified the extreme disparity in genes
-with hormone products or enzymatic products that directly produce
-hormones in multiple mammalian species (\cref{fig:hormone_genes}).
-
-\begin{figure}
-<<hormone.genes.plot,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-gene.expression.nozero <- function(genes) {
-    temp <-
-        data.matrix.dt(dcast(combined.fpkm[human_name %in% genes,],
-                             human_name~species,
-                             value.var="FPKM",
-                             fun.aggregate=function(x){x[1]}))
-    temp <-
-        temp[apply(temp,1,function(x){sum(x,na.rm=TRUE)})>5,]
-    return(temp)
+}
+\block{}{
+  \includegraphics[width=\linewidth,height=6cm,keepaspectratio]{figures/uofi_mark}
 }
 
-### hormone genes
-hormone.genes <- c("CGA","CGB","CGB1","CGB2","CGB5","CGB8","CSH1","CSH2","CSHL1","CYP19A1",
-                   "GH2","GHRH","HSD3B1","HSD3B2","HSD17B2","HSD17B1","NPPB","NPPC","PRL","RLN1")
-hormone.expression.nozero <-
-    gene.expression.nozero(hormone.genes)
-
-hor.exp.long <- melt(log10(hormone.expression.nozero+1)[rev(1:nrow(hormone.expression.nozero)),])
-colnames(hor.exp.long)[1:3] <- c("gene","species","fpkm")
-hor.exp.long <- hor.exp.long[hor.exp.long$gene!="IGF2",]
-
-### galectin genes
-galectin.genes <- unique(grep("^LGALS",combined.fpkm[,human_name],value=TRUE))
-
-galectin.expression.nozero <-
-    gene.expression.nozero(galectin.genes)
-
-gal.exp.long <- melt(log10(galectin.expression.nozero+1)[rev(1:nrow(galectin.expression.nozero)),])
-colnames(gal.exp.long)[1:3] <- c("gene","species","fpkm")
-
-### IGF related genes
-
-igfrelated.genes <- unique(grep("^IGF",combined.fpkm[,human_name],value=TRUE))
-
-igfrelated.expression.nozero <-
-    gene.expression.nozero(igfrelated.genes)
-
-igf.exp.long <- melt(log10(igfrelated.expression.nozero+1)[rev(1:nrow(igfrelated.expression.nozero)),])
-colnames(igf.exp.long)[1:3] <- c("gene","species","fpkm")
-
-### Pre-eclampsia genes ADAM12 CRH KISS1 PAPPA
-
-pe.expression.nozero <-
-    gene.expression.nozero(c("ADAM12","CRH","KISS1","PAPPA"))
-
-pe.exp.long <- melt(log10(pe.expression.nozero+1)[rev(1:nrow(pe.expression.nozero)),])
-colnames(pe.exp.long)[1:3] <- c("gene","species","fpkm")
-
-igf.exp.long$type <- "IGF Related"
-gal.exp.long$type <- "Galectin"
-hor.exp.long$type <- "Hormone"
-pe.exp.long$type <- "PE"
-
-combined.long <-
-    rbind(igf.exp.long,
-          gal.exp.long,
-          hor.exp.long,
-          pe.exp.long
-          )
-combined.long$species <-
-    factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(combined.long$species)),
-           levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(names(species.names))))
-
-common.plot.options <- 
-    list(geom_tile(aes(fill = fpkm), colour = "white"),
-         scale_fill_gradient(low = grad.def[["low"]],
-                             high = grad.def[["high"]],
-                             space = grad.def[["space"]],
-                             na.value = grad.def[["na.value"]],
-                             guide=guide_colorbar(title=expression(log[10](FPKM)))),
-         scale_x_discrete("", expand = c(0, 0)),
-         scale_y_discrete("", expand = c(0, 0)),
-         theme_grey(base_size = 9),
-         theme(legend.position = "top",
-               axis.ticks = element_blank(),
-               panel.background = element_blank(),
-               axis.text.y = element_text(face="italic"),
-               axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic")),
-         facet_grid(type~.,scale="free_y",space="free_y")
-         )
-print(ggplot(combined.long, aes(y=gene, x=species))
-      + common.plot.options
-      )
-@ 
-\caption{Expression in $\text{log}_{10}(\text{FPKM})$ of 1:1 \human
-  orthologs of Hormone Genes, Galectins, genes implicated in
-  pre-eclampsia (PE), and \textit{IGF}-related genes. Empty locations
-  indicate no detectable expression and/or no 1:1 \human ortholog in
-  that species.}
-\label{fig:hormone_genes}
-\end{figure}
-
-\subsubsection{Estrogen and Progesterone}
-
-In both ruminants and primates, the placenta is capable of producing
-estrogen and progesterone
-\cite{Strauss.ea1996:Placentalsteroidhormonesynthesis}. High relative
-expression of \textit{CYP19A1} (Supplemental
-\cref{fig:estrogen_evolution_tree}), the enzyme that synthesizes
-estrogen from androstenedione is seen in \human, \bonobo, \cow, and
-\horse, though not in \sheep, \smonkey. %
-%
-\textit{HSD3B1}, one of two enzymes that synthesizes progesterone from
-pregnalone
-(\cite{Simard.ea2005:Molecularbiologyof3beta-hydroxysteroid},
-Supplemental \cref{fig:progesterone_evolution_tree}) is also highly
-expressed in \human, \bonobo, \cow, \horse, and \dog, but is not
-highly expressed in \smonkey, \sheep, \pig, \elephant, \armadillo, or
-\opossum. This is consistent with the previously observed local
-progesterone withdrawal in \sheep, and the leuteo-placental shift of
-progesterone expression in boreoeutheria in contrast to afrotherians
-which do not show such a shift
-\cite{Swaggart.ea2015:Genomicsofpretermbirth}. The diversity of
-\textit{CYP19A1} and \textit{HSD3B1} expression levels in in primates
-and ungulates seen here points to the need for more complete sampling
-of placentas at different timepoints and in different clades.
-
-\subsubsection{Prolactin}
-
-The elephant is strikingly the only mammal analyzed that shows high
-expression of the primarily pituitary hormone prolactin (\textit{PRL})
-in the term placenta, although we see high expression of genes which
-resulted from duplications of \textit{PRL} in the \mouse, \carmeli,
-and \cow (Supplemental \cref{fig:prolactin_evolution_tree}). The
-results in \mouse are relatively consistent with the temporal
-expression results of
-\citet{Simmons.ea2008:Spatialandtemporalexpression}. For example, we
-see very high expression of \textit{Prl3b1} and \textit{Prl8a9} in the
-term placentas of \mouse, but low expression of \textit{Prl7a1} and
-\textit{Prl8a2} which are expressed at earlier embryonic stages. %
-All afrotherians yet analyzed contain only one prolactin gene, and
-prolactin has undergone adaptive evolution on the \elephant lineage
-\cite{Wallis2009:ProlactininAfrotheriacharacterization}.
-\citet{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
-presence of a prolactin in the trophoblast of the elephant placenta
-using immunohistological staining and human derived antibodies. A
-previous analysis of \elephant placenta also detected the expression
-of prolactin \cite{Hou.ea2012:Elephanttranscriptomeprovidesinsights}.
-Here we confirm that the sole prolactin in \elephant is not only
-expressed but highly expressed in the placenta, and likely functions
-both in the maintenance of the corpus luteum and as a placental
-lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}.
-\textit{PRL} is not expressed in \opossum, even though there is
-evidence that it is expressed in the placenta of another marsupial,
-\textit{Macropus eugenii} (tammar wallaby)
-\cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests
-that placental prolactin expression in marsupials may be as diverse as
-in eutherian mammals. One possible explanation of the diversity of
-expression is the presence of endogenous retroviruses such as MER39 in
-primates, MER77 in rodents, and L1-2\_LA in \elephant
-\cite{Emera.ea2012:Convergentevolutionofendometrial} although this
-does not explain the diversity of expression of different isoforms of
-\textit{PRL} in \mouse or their tissue specificity in primates. An
-alternative, that \textit{HOXA11} expression is responsible for
-\textit{PRL} expression
-\cite{Lynch.ea2008:Adaptivechangesintranscription} seems not to be the
-case in placenta, as the rodent with the highest \textit{HOXA11}
-expression, \galili
-(\Sexpr{gene.expression("spalax_galili","Hoxa11",complete=TRUE)}) is
-the rodent with the lowest expression of \textit{PRL}
-(\cref{fig:prolactin_evolution_tree}), and \elephant also does not
-express \textit{HOXA11} highly either
-(\Sexpr{gene.expression("loxodonta_africana","HOXA11",complete=TRUE)},
-Supplemental \cref{fig:hoxa11_evolution_tree}).
-
-\subsubsection{Growth hormone family}
-
-The growth hormone family of genes has expanded in anthropoid
-primates, resulting in approximately 5 genes in \human (\textit{GH1},
-\textit{GH2}, \textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at
-least three in \bonobo (two \textit{GH2} and \textit{GH1}; owing to
-the similarity between the two \textit{GH1} in \textit{P. troglodytes}
-we cannot reliably distinguish transcripts between the two). All show
-substantial expression in the \human placenta
-(\cref{fig:csh1_evolution_tree}). The expression of the \ac{GH} family
-in placenta appears to be primate specific, and likely existed at the
-time of the last common ancestor between all extant anthropoids
-\cite{Papper.ea2009:Ancientoriginofplacental,Soares2004:prolactinandgrowthhormone}.
-There is some evidence that increased expression of these genes may be
-correlated with overall birth weight of offspring in \human
-\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
-possible that the extreme expression levels of these genes may be due
-to high birth weight. \human placental \ac{GH} genes are involved in
-increasing blood flow, replacing maternal pituitary expression of
-\ac{GH}, as well as trophoblast invasion
-\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
-and it is likely that they have the same function in \bonobo.
-
-\begin{figure}
-<<chorionic.somatomammotropin.alpha.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-plot.label.size <- 4.5
-pushViewport(viewport(layout=grid.layout(nrow=1,ncol=2,widths=c(2,3))))
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CGA" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=1),
-                 offset.ratio=0.7,
-                 fontsize=plot.label.size,
-                 axis.text.size=3.5
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CGB1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=2,layout.pos.row=1),
-                 subtree=c("ENSMODG00000013768","ENSG00000104818"),
-                 offset.ratio=1,
-                 fontsize=plot.label.size,
-                 axis.text.size=3.5
-                 )
-popViewport(1)
-grid.text(c("A","B"),
-          x=unit(c(0,2/5),"npc"),y=unit(c(1,1),"npc"),
-          just=c("left","top"),
-          gp=gpar(fontsize="48")
-          )
-@ 
-\caption{\genetreeexplanation{(A) \textit{CGA} (B) \textit{CGB}}}
-\label{fig:cga_cgb_evolution_tree}
-\end{figure}
-
-The human gene \textit{CGA} codes for the alpha subunit of multiple
-hormones including pituitary hormones (luteinizing hormone,
-follicle-stimulating hormone, thyroid-stimulating hormone), as well as
-numerous chorionic gonadotropins with variable beta subunits
-\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of
-the chorionic gonadotropins is to prevent the regression of the corpus
-luteum \cite{Hanson.ea1971:EffectsofHCGand}. Among the six placentas
-we analyzed here, \textit{CGA} is only expressed in the placentas of
-\human and \bonobo (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}),
-which corresponds with the emergence of the chorionic gonadotropins in
-anthropoid primates
-\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly,
-\textit{CGA} is expressed at much higher levels in both species than
-the beta subunits with which its product dimerizes (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}).
-\textit{CGA} is one of several genes noted by
-\citet{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone
-adaptive evolution of the \human{}-\textit{P. troglodytes} stem lineage.
-
-\subsubsection{Relaxin}
-
-Relaxin (\textit{RLN}) is a hormone that has multiple effects during
-pregnancy including increasing maternal blood flow, widening the pubic
-bone and relaxing uterine musculature in preparation for labor. In
-\dog, \elephant, \mouse, \carmeli, and \horse there is high expression
-of relaxin in the term placenta, with little evidence in other species
-examined here (Supplemental \cref{fig:rln2_evolution_tree}). \textit{RLN} is known
-to be produced in the placenta of the pregnant \dog and is used in the
-diagnosis of \dog pregnancy
-\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
-the luteal ovary and placenta of \pig
-\cite{Hisaw.ea1948:Relaxininovaryof}, and the placenta of \horse
-\cite{Klonisch.ea1997:Placentallocalizationofrelaxin}. However, the
-source of relaxin produced by pregnant \elephant remains unclear
-\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in \textit{RLN} in
-\elephant has been suggested to be a major signal for the initiation
-of labor \cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive}.
-Here we show that the placenta of elephants show high expression of
-\textit{RLN} and is likely a major source of \textit{RLN} at term. In
-contrast to the results of
-\cite{Evans.ea1993:mouserelaxingenenucleotide}, we also see expression
-of \textit{RLN} in the placenta of \mouse and \carmeli.
-
-\subsubsection{\textit{NPPB} and \textit{NPPC}}
-
-Two genes, \textit{NPPB} and \textit{NPPC}, code for peptide hormones
-that are important in the inducement of vasodilation, and may be
-important in increasing maternal blood flow to the fetus
-\cite{Cameron.ea1996:sitesofgeneexpression}. There is some indication
-that these proteins are present in the decidua of mice
-\cite{Cameron.ea1996:sitesofgeneexpression}, but we find very little
-expression in the fetal portion of the mouse placenta
-(\cref{fig:hormone_genes}, Supplemental
-\cref{fig:nppc_evolution_tree,fig:nppb_evolution_tree}). While we do
-find high expression of at least one of these genes in \dog, \cow, and
-\elephant, the expression pattern does not follow the placental
-organization (endotheliochorial for \dog and \elephant,
-synepitheliochorial for \cow and \sheep), so it is unlikely that
-expression of these genes are essential for specific placental
-organizations.
-
-\subsection{The Evolution of Placental Expression of Galectins}
-
-Galectins are recognized as major regulators of pregnancy in both
-fetal and maternal tissues
-\cite{Than.ea2012:Galectinsguardiansofeutherian}. Indeed, the
-evolution of new lineage-specific galectins has occurred in both
-primates and ruminants, with the genes \textit{LGALS13},
-\textit{LGALS14}, and \textit{LGALS16} recognized as being
-placenta-specific in \human and regulating immune responses by
-inducing apoptosis of maternal T-cells
-\cite{Than.ea2009:primatesubfamilyofgalectins} (Supplemental
-\cref{fig:galectins_lgals13_evolution_tree}). Indeed, \textit{LGALS14}
-has the highest expression of any galectin in the \human and \smonkey
-placenta (\cref{fig:hormone_genes}, Supplemental
-\cref{fig:galectins_lgals13_evolution_tree}), and other primate
-galectins such as \textit{LGALS13} and \textit{LGALS16} are also
-expressed in \human and \smonkey placenta. However, in the closely
-related \bonobo, \textit{LGALS14} is not found, and \textit{LGALS1} is
-the most abundant galectin. \textit{LGALS1} shows generally high
-expression in all placentas analyzed. \textit{LGALS1} is also involved
-in mediating maternal-fetal immune tolerance, and is expressed in both
-maternal and fetal tissue in primates
-\cite{Than.ea2008:Emergenceofhormonaland}. \textit{LGALS1} placental
-expression was thought to be confined to eutherian mammals, as there
-is evidence of adaptive evolution at this node, as well as the
-appearance of estrogen response elements in eutherian cis-regulatory
-regions \cite{Than.ea2008:Emergenceofhormonaland}. It was unclear as
-to whether galectins such as \textit{LGALS1} were important in the
-regulation of pregnancy in marsupials, as well as eutherians. Here we
-show substantial expression of galectins in the opossum, especially
-\textit{LGALS1} (Supplemental \cref{fig:galectins_evolution_tree}). In
-addition, \textit{LGALS2} is also expressed. The presence of galectins
-at the maternal-fetal interface in \opossum implies that it was also
-present in the ancestral placenta of therian mammals. Indeed, adaptive
-evolution of \textit{LGALS1} on the eutherian stem is likely due to
-the increased invasive nature of eutehrian placentation rather than
-the origination of maternal-fetal tolerance. Although the function of
-\textit{LGALS1} in the marsupial placenta needs to be confirmed, it is
-likely that the mediation of maternal-fetal immune tolerance by
-galectins is an ancient feature of mammalian viviparity.
-
-\subsection{The Opossum Placenta and the Evolution of Viviparity}
-
-The placenta of \opossum is distinctly different from that of the term
-placenta of eutherian mammals in that it is derived from both the
-chorion and yolk sac. Other differences between the placentas of
-marsupials and eutherians include lack of invasive extravillous
-trophoblasts in marsupials, the retention of a permeable shell coat
-for 2/3 of marsupial gestation, and the greater degree of imprinting
-in placenta-expressed genes in eutherians
-\cite{Renfree2010:ReviewMarsupialsplacentalmammals,Renfree.ea2013:originandevolutionof}.
-In addition, the placenta of \opossum is also distinct from other
-marsupials, in that it is considered one of the least specialized, and
-thought be closer to the original marsupial condition
-\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. In \opossum, the
-duration of direct connection between maternal and fetal tissues is as
-short as two days in a gestational period of 13-15 days
-\cite{Harder.ea1993:Gestationandplacentationin}. During this period,
-the placenta is directly engaged in the exchange of urea with maternal
-tissues, owing to the lack of a vesicular allantois
-\cite{Freyer.ea2007:Placentalfunctionintwo}. This differs drastically
-with the situation in another well-studied marsupial, the tammar
-wallaby \cite{Freyer.ea2007:Placentalfunctionintwo}. Therefore, we
-must interpret differences between the opossum and eutherian placentas
-with caution, as highly expressed genes and functional annotation
-clusters may represent didelphid or \opossum specific adaptations.
-
-The \opossum placenta shows expression of genes that either do not
-exist in eutherians or are not expressed in the term eutherian placenta
-(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
-of hemoglobin in fetal tissues at term, as evidenced by the expression
-of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
-generally due to their very short gestation periods. Embryonic
-hemoglobins in marsupials have a lower affinity for oxygen and allow
-for a lower rate of oxygen uptake from maternal tissues
-\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
-uncharacterized gene that shows high expression (\textit{LOC10001195}) is not
-found in eutherians; Ensembl gene trees reveal that homologs of this
-gene exist in other amniotes such as birds and turtles, but are not
-present in eutherian mammals. The homolog of this gene in chicken is
-found to expressed in reproductive tract
-\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
-unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
-family and involved in innate immunity. It is expressed in the mammary
-gland of the tammar wallaby during pregnancy and early lactation
-\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and present in
-relative abundance in \opossum placenta.
-
-\subsection{Novel Gene Expansions in the Elephant Placenta}
-
-We found evidence of high expression in two novel expansions of genes
-in the \elephant genome. One family includes 11 genes that are related
-to the \textit{PRSS16} gene, a thymus-specific serine protease that
-may be involved regulating the presentation of self-peptides in CD4+
-T-lymphocytes \cite{Fornari.ea2012:TranscriptionprofilingofPrss16}.
-All 11 of these genes are expressed in the elephant placenta, three of
-which have very abundant mRNAs
-(\cref{tab:top50_elephant_placenta} and \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}C). It is
-possible that this elephant-specific gene family may be involved in
-immunity in the elephant placenta and the mediation of maternal-fetal
-immune tolerance; however, future work will need to test these
-hypotheses. Three other genes abundantly expressed in the elephant
-placenta are homologues of \textit{SSLP1}, a gene with predominant
-expression in the seminal vesicle in mouse
-\cite{Li.ea2006:SSLP-1secretedLy-6protein}. Homologues of
-\textit{SSLP1} are also expressed in \cow, \dog, \horse, and \sheep.
-\textit{SSLP1} also contains an Alu-like retrotransposon and may be
-involved in the development of villi via cell proliferation
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}. This
-gene is not found in primates. Expansion of these genes with high
-expression in \elephant may indicate an elaboration of this system.
-
-% Bovine has their own alu-like sequence: PMID 7835712
-
-\subsection{Core placenta transcriptome}
-
-The transcriptomes from multiple different species enabled us to
-identify the set of expressed genes which are common to all placentas.
-Components of the annexin complexes, including including
-\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
-are expressed in all placentas examined, and argue for the evolution
-importance of annexins in placenta function, where they likely
-establish maternal-fetal
-tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
-We suggest that further research into molecular role of these annexin
-complexes may illuminate how immune tolerance is maintained during
-placentation. The core placenta transcriptome is also enriched in
-genes involved in maintaining the structure of the placenta and
-transporting proteins, as would be expected given the function of the
-placenta.
-
-\subsection{\human Specific Expression Shifts and the Evolution of Pre-eclampsia}
-
-The \human placenta is distinctive from many other primates in that it
-possesses a deeply invasive trophoblast that penetrates the inner
-myometrium of the uterus
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
-hypothesized that this feature is related to the increased demand for
-resources by the developing fetus to feed the expanded size of the
-human brain
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation,Robillard.ea2002:Evolutionaryadaptationstopre-eclampsiaeclampsia}.
-\human have the largest brain in terms of relative size and this
-differs markedly from closely related apes
-\cite{Boddy.ea2012:Comparativeanalysisofencephalization}. Deep
-penetration of the myometrium occurs in chimpanzees and gorillas, but
-this penetration may be accelerated in humans
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
-
-The near exclusivity of deep trophoblast invasion in humans has also
-led many to speculate on the origin of various disorders, which are
-almost exclusively symptomatic of human pregnancy. For example,
-pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
-the United States
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
-can be attributed to the lack of adequate invasion of maternal tissue,
-and is noted by the onset of hypertension and proteinuria %
-%
-\DLA{Discuss the new ACOG requirements for diagnosis which does not
-  require proteinuria.}
-%
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
-pre-eclampsia can develop into eclampsia, a more severe condition
-resulting in seizures during pregnancy and death. Lack of invasion in
-pre-eclampsia may be attributed to the inefficiency of growth
-regulators as well as the inability to sufficiently suppress the
-maternal immune response. Although there are speculative cases of
-eclampsia in great apes
-\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
-it is exceedingly rare in these species
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
-
-The identification of multiple genes (\textit{CRH}, \textit{ADAM12},
-\textit{KISS1}, \textit{PAPPA}, \textit{IL1RL1}, and \textit{PLAC1};
-\cref{tab:human_lineage_specific}) which are associated with
-pre-eclampsia which have increased expression on the \human lineage
-may explain the larger proportion of pre-eclampsia observed in humans.
-However, there are still genes which are associated with pre-eclampsia
-which are also more highly expressed in the \mouse lineage such as
-\textit{Afp} and \textit{Gjb2}. Furthermore, the onset of
-pre-eclampsia is well before term; all of our placentas are term
-placentas and provide limited information about the transcriptome
-during the onset of pre-eclampsia.
-
-\section{Conclusion}
-
-We analyzed the sequenced transcriptome of
-\Sexpr{star.logs[,length(unique(file))]} placentas from
-\Sexpr{star.logs[,length(unique(species))]} species spanning the
-phylogenetic tree of mammals. This wealth of data revealed both
-lineage-specific expansions and differential expression of gene
-families as well as a set of core, non-housekeeping genes which are
-critical for the proper organization and functioning of the placenta.
-
-There is diversity in the placental expression of hormones, including
-estrogen, progesterone, prolactin, relaxin, the growth hormone family, and
-natriuretic peptides among the species and lineages analyzed.
-We also identified genes differentially expressed on the \human
-lineage which are implicated in pre-eclampsia, including
-\textit{ADAM12}, \textit{CRH}, and \textit{KISS1}.
-
-The core placenta transcriptome was also identified, which contains
-genes which are involved in maternal-fetal tolerance, such as the
-annexins, and genes responsible for cell-cell and cell-matrix
-interactions, such as \textit{LGALS1}. These genes were almost
-certainly present in the placenta of the common mammalian ancestor. We
-hypothesize that modifications to their expression or function will
-compromise the development and organization of the placenta.
-
-This study has identified many tantalizing molecular pathways which
-illuminate the vast diversity of placental morphology and may explain
-some placental malfunctions in \human{}. However, the strength of our
-hypotheses are limited by our limited data; more placentas from
-multiple members of diverse clades and different timepoints are still
-needed to come to a more complete understanding of the genomic and
-molecular basis for placenta morphology and function. We look forward
-to collecting and analyzing such a dataset.
-
-\begin{backmatter}
-  \section*{Competing interests}
-  The authors declare that they have no competing interests.
-
-  \section*{Author's contributions}
-  Don Armstrong analyzed all of the data and wrote the paper.
-
-  Michael McGowan wrote the paper and performed initial analyses.
-
-  Amy Weckle performed all of the RNA extractions and managed the
-  samples.
-
-  Jason Caravas performed initial analyses.
-
-  Dalen Agnew provided samples.
-
-  Kurt Benirschke provided samples.
-
-  Sue Savage-Rumbaugh provided samples.
-
-  Eviatar Nevo provided samples.
-
-  Chong J. Kim provided samples.
-
-  Günter P. Wagner provided samples and wrote the paper.
-
-  Roberto Romero provided samples and conceived the study.
-
-  Derek E. Wildman conceived the study, wrote the paper, and performed
-  analyses.
-
-  \section*{Acknowledgements}
-  The authors would like to specially acknowledge those groups whose
-  publicly available data contributed to the completeness of this
-  work.
-
-  \section*{Funding}
-  Partial support for this research came from the \textit{Eunice
-    Kennedy Shriver} National Institute of Child Health and Human
-  Development (NICHD), National Institutes of Health, U.S. Department
-  of Health and Human Services (N01-HD-3342), and the National Science
-  Foundation BCS-0827546.
-
-  \printaddresses
-
-\end{backmatter}
-
-\printbibliography
+\end{columns}
 

### 2016-04-27 poster/placenta_viviparous_poster.Rnw
add conclusions and switch to KISS1 instead of CGA

@@
 \usepackage{fancybox}
+\usepackage{xspace}
 \usepackage{booktabs}
@@
 \usepackage{qrcode}
-\usepackage[backend=biber,natbib=true,hyperref=true,style=nature]{biblatex}
+\usepackage[backend=biber,maxnames=1,natbib=true,hyperref=true,style=nature]{biblatex}
 \addbibresource{references.bib}
@@
 \usetheme{Default}
+
+% species names
+\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
+\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
+\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
+\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
+\newcommand{\horse}[0]{\textit{E. caballus}\xspace}
+\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
+\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
+\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
+\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
+\newcommand{\galili}[0]{\textit{N. galili}\xspace}
+\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
+\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
+\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
+\newcommand{\pig}[0]{\textit{S. scrofa}\xspace}
 
@@
     \hfill
-    \begin{minipage}{0.2\linewidth}
-      \includegraphics[width=0.8\linewidth,height=6cm,keepaspectratio]{figures/igb_wordmark}
+    \begin{minipage}{0.13\linewidth}
+      \begin{flushright}
+        \includegraphics[width=\linewidth,height=6cm,keepaspectratio]{figures/igb_wordmark}
+      \end{flushright}
     \end{minipage}
@@
 <<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
-opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
+opts_chunk$set(dev="cairo_pdf",out.width="0.9\\colwidth",out.extra="keepaspectratio")
 opts_chunk$set(cache=TRUE, autodep=TRUE)
@@
 
-  \column{0.25}
+  \column{0.2}
 
   \block{Introduction}{
+\begin{itemize}
+\item Function and Evolution of Placenta
+    \begin{itemize}
+    \item Major function is the transfer of nutrients, oxygen, and
+      waste between mother and offspring
+    \item Highly variable structure and form in mammals, even though
+      major function is conserved
+    \item Likely arose before the origin of therian mammals
+    \item Form relatively well characterized, but diversity of
+      molecular environment understudied
+    \item Role in the evolutionary interplay between maternal and fetal
+      strategies
+    \end{itemize}
+
+  \item Impacts on Human Health
+    \begin{itemize}
+    \item Humans possess a deeply invasive trophoblast layer that can
+      penetrate the myometrium of the uterus
+      \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
+    \item Defects in placental growth can lead to pre-eclampsia, which can
+      cause hypertension, proteinuria, and mortality of mother and/or
+      infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
+    \item Pre-eclampsia has been seen in \textit{Pan troglodytes}
+      \cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
+      \textit{Gorilla gorilla}
+      \cite{Thornton.ea1992:Convulsionsinpregnancyin}
+    \item Far higher incidence in \textit{Homo sapiens}, with as many as 
+      4\% of pregnancies United States affected
+      \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
+    \end{itemize}
+
+  \item Hypotheses
+    \begin{itemize}
+    \item Gene expression patterns correlate with placenta morphology
+    \item Altered expression of genes correlates with invasiveness, and
+      therefore pre-eclampsia
+    \item Maternal/Fetal evolutionary strategies will be reflected in gene
+      expression patterns
+    \end{itemize}
+  \end{itemize}
 
@@
   \block{Methods}{
+    \small
+    \begin{itemize}
+    \item Fetal placenta tissue from 9 mammalian species collected;
+      total RNA isolated and sequenced using Illumina GA II with an
+      average read size of 150nt
+    \item Additional mammalian placenta sequences obtained from SRA
+    \item QCed using FASTQC, aligned using STAR, and quantified using
+      Cufflinks. Species missing suitable reference (Spider monkey,
+      \textit{S. galili}, \textit{S. carmeli}) were assembled using
+      Trinity and annotated using Diamond against the Human or Mouse
+      transcriptome.
+    \item Ensembl build 80 was the source of annotation.
+    \item Enrichment in human and mouse were calculated using a
+      type-III anova, all other enrichments used Student's t test.
+    \end{itemize}
   }
 
-\column{0.25}
+\column{0.2}
 
-\begin{figure}
-<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
+<<top.genes.function,echo=FALSE,results="hide">>=
+top.genes.table <- function(species.name,n=50) {
+    alignment.species <- species.name
+    if (alignment.species=="pan paniscus") {
+        alignment.species <- "pan troglodytes"
+    } else if (alignment.species=="ateles fusciceps") {
+        alignment.species <- "homo sapiens"
+    } else if (grepl("spalax",alignment.species)) {
+        alignment.species <- "mus musculus"
+    }
+    temp <- 
+        combined.fpkm[species==species.name &
+                      !(!is.na(human_name) & 
+                        human_name %in% housekeeping.genes.superset[,gene_short_name]) &
+                      ! (gene_biotype %in% c("Mt_tRNA","Mt_rRNA","rRNA")) &
+                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_name) &
+                      !grepl("((RNA\\d|MT)-|^RP[SL]P?\\d+[ABCD]?[L]?$)",human_alignment_symbol) &
+                      !grepl("(^[Mm][Tt]-|^R[pP][lsLS]\\d+[abcdABCD]?[lL]?(?:-.+)?$|rRNA$)",
+                             gene_short_name)
+                      ][order(-mean_fpkm)][,list(name_or_id,
+                                                 human_alignment_symbol,
+                                                 human_name,mean_fpkm,sd_fpkm,n_fpkm)]
+    ## remove any genes which have a null name_or_id; they are probably from unassembled contigs
+    temp <- temp[!(is.na(name_or_id)|name_or_id==""),]
+    temp <- temp[!duplicated(name_or_id)][1:n]
+    temp[,used_alignment:=FALSE]
+    temp[is.na(human_name) & !is.na(human_alignment_symbol),
+                    used_alignment:=TRUE]
+    temp[is.na(human_name),human_name:=human_alignment_symbol]
+    setkey(temp,"human_name")
+    temp <- 
+        hugo.nomenclature[temp][order(-mean_fpkm),
+                                list(name_or_id,used_alignment,`Approved Symbol`,`Approved Name`,
+                                     mean_fpkm,sd_fpkm,n_fpkm)]
+    temp[!is.na(name_or_id),
+         name_or_id:=paste0("\\href{http://www.ensembl.org/",
+                             capfirst(gsub(" ","_",alignment.species)),
+                             "/Gene/Summary?g=",name_or_id,"}{\\textit{",
+                             xtable.sanitize.text(name_or_id),"}}")]
+    temp[!is.na(`Approved Name`),
+         `Approved Name`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{",xtable.sanitize.text(`Approved Name`),"}")]
+    temp[!is.na(`Approved Symbol`),
+         `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",`Approved Symbol`,"}{\\textit{",xtable.sanitize.text(`Approved Symbol`),"}}")]
+    temp[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
+    temp[,used_alignment:=NULL]
+    temp[,sem_fpkm:=as.numeric(NA)]
+    temp[n_fpkm>1,sem_fpkm:=sd_fpkm/sqrt(n_fpkm)]
+    temp[,sd_fpkm:=NULL]
+    temp[,n_fpkm:=NULL]
+    setnames(temp,c("Gene symbol","Human Symbol","Human Name","FPKM","SEM"))
+    if (temp[,all(is.na(`SEM`))]) {
+        temp[,`SEM`:=NULL]
+    }
+    if (species.name=="homo sapiens" || species.name=="ateles fusciceps") {
+        temp[,`Human Symbol`:=NULL]
+    }
+    temp[,`FPKM`:=round(`FPKM`)]
+    return(temp)
+}
+gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE) {
+    species.name <- gsub("_"," ",species.name)
+    fpkm <- combined.fpkm[species==species.name &
+                              (gene_short_name==gene | 
+                               gene_id==gene),
+                          mean_fpkm][1]
+    if (is.na(fpkm)) {
+        stop("FPKM is NA!!!!")
+    }
+    if (units || complete)
+        fpkm <- formatC(fpkm,digits=0,format="f")
+    if (units) {
+        fpkm <- paste0(fpkm,
+                       "\\ \\text{FPKM}")
+    }
+    if (complete) {
+        if (any(grepl("^ENS.+G\\d+$",gene))) {
+            fpkm <- paste0("\\text{",gene,"} = ",
+                           fpkm)
+        } else {
+            fpkm <- paste0("\\text{\\textit{",gene,"}} = ",
+                           fpkm)
+        }
+    }
+    if (units || complete) {
+        fpkm <- paste0("$",fpkm,"$")
+    }
+    return(fpkm)
+}
+
+@
+
+<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("placenta_lineage_specific","human_mouse_lineage_specific")))>>=
+load("../data/placenta_lineage_specific")
+load("../data/human_mouse_lineage_specific")
+source("/home/don/projects/xtable/pkg/R/print.xtable.R")
+
+placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=5,orderbyfc=FALSE,fdr.cutoff=1) {
+    temp.table <- 
+        lineage.table[analysis.name==analysis &
+                      p.value <= 0.05 &
+                      fdr <= fdr.cutoff][order(p.value),
+                                       list(human_name,p.value,fdr,fc),]
+    temp.table <- data.table(temp.table)
+    setkey(temp.table,"human_name")
+    temp.table <- 
+        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
+                                            `Approved Name`,p.value,fdr,fc)]
+    setnames(temp.table,
+              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
+              )
+    setkey(temp.table,"Human Symbol")
+    if (output) {
+        display.table <- temp.table[`FDR`<=fdr.cutoff,][order(`p (t-test)`),]
+        if (orderbyfc) {
+            display.table <- temp.table[order(-abs(`Fold Change`)),]
+        }
+        display.table <- display.table[1:members,]
+        display.table[,`p (t-test)`:=NULL]
+        display.table <-
+            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
+                                                  "}")]
+        setnames(display.table,"Fold Change","$\\mathrm{log}_2$ FC")
+        print(xtable(display.table,
+                     display=c("s","s","s","g","f"),
+                     align=c("c","c","p{3.3in}","r","r")
+                     ),
+              booktabs=TRUE,
+              floating=FALSE,
+              math.style.exponents=TRUE,
+              sanitize.text.function=genes.table.sanitize,
+              include.rownames=FALSE)
+
+    }
+    return(invisible(temp.table[order(`p (t-test)`),]))
+}
+output.top.genes.table <- function(data,output=TRUE) {
+    if (output) {
+        table.display <- c("s","s","s","d")
+        table.align <- c("c","c","p{3in}","r")
+        if (any(colnames(data) %in% "SEM"))  {
+            table.display <- c(table.display,"d")
+            table.align <- c(table.align,"r")
+        }
+        if (any(colnames(data) %in% "Human Symbol"))  {
+            table.align <- c("c",table.align)
+            table.display <- c("s",table.display)
+        }
+        print(xtable(data,
+                     display=table.display,
+                     align=table.align
+                     ),
+              booktabs=TRUE,
+              floating=FALSE,
+              sanitize.text.function=genes.table.sanitize,
+              include.rownames=FALSE)
+    }
+}
+
+@
+
+\block{Phylogeny of Species}{
+\centering
+<<species.tree,fig.width=10,fig.height=8,echo=FALSE>>=
 species.tree <- 
@@
 @ 
-\caption{Species tree for organisms studied showing the relationship
+\captionof{figure}{Species tree for organisms studied showing the relationship
   between species. Silhouettes of species were obtained from
@@
 \label{fig:species_tree}
-\end{figure}
+}
 
-\column{0.25}
+% \block{Variation in Gestational Period and Litter Size}{
+%   \small
+%   \begin{tabular}{>{\raggedright \centering}p{2in} >{\raggedright \centering}p{2.5in} c c }
+%     \toprule
+%     Species                       & Common Name & \multicolumn{1}{p{1in}}{\centering Gestation (days)} & \multicolumn{1}{p{1in}}{\centering Average Litter Size} \\
+%     \midrule
+%     \textit{Ateles fusciceps}     & Spider monkey & 224 & 1 \\
+%     \textit{Bos taurus}           & Cow & 278 & 1 \\
+%     \textit{Canis familiaris}     & Dog & 61 & 4 \\
+%     \textit{Dasypus novemcinctus} & Nine-banded Armadillo & 150 & 4 \\
+%     \textit{Equus caballus}       & Horse & 336 & 1 \\
+%     \textit{Loxodonta africana}   & African savanna elephant & 644 & 1 \\
+%     \textit{Homo sapiens}         & Human & 270 & 1.1 \\
+%     \textit{Monodelphis domestica}& Gray short-tailed opossum & 15 & 7.1 \\
+%     \textit{Mus musculus}         & Mouse & 20 & 6.1 \\
+%     \textit{Spalax galili}   & Blind mole rat from Upper Galilee & 28-42 & 4 \\
+%     \textit{Ovis aries}           & Domestic sheep & 151 & 1.35 \\
+%     \textit{Pan paniscus}         & Bonobo & 223 & 1 \\
+%     \textit{Spalax carmeli}       & Blind mole rat from northern Israel & 28-42 & 4 \\
+%     \textit{Sus scrofa}           & Domestic pig (Duroc) & 114 & 5.8 \\
+%     \bottomrule
+%   \end{tabular}
+%   \captionof{table}{Gestational periods in days and average liter sizes for studied species.}
+% }
 
-\column{0.25}
+% \block{Variations in Placenta Morphology}{
+% \begin{tabular}{c c c}
+%   \toprule
+%   Species & Interhemal Membrane & Pattern \\
+%   \midrule
+%   \smonkey & hemochorial & trabecular \\
+%   \cow     & epitheliochorial & villous \\
+%   \dog     & endotheliochorial & labyrinthine \\
+%   \armadillo & hemochorial & trabecular \\
+%   \horse & epitheliochorial & villous \\
+%   \elephant & endotheliochorial & labyrinthine \\
+%   \human & hemochorial & villous \\
+% % http://onlinelibrary.wiley.com/doi/10.1002/jez.1402660511/pdf
+%   \opossum & epitheliochorial & choriovitelline \\
+%   \mouse & hemotrichorial & labyrinthine \\
+%   \galili & hemotrichorial & labyrinthine \\
+%   \sheep & epitheliochorial & villous \\
+%   \bonobo & hemochorial & villous \\
+% % http://www.biolreprod.org/content/early/2014/10/17/biolreprod.114.124818.short
+%   \carmeli & hemotrichorial & labyrinthine \\
+%   \pig & epitheliochorial & diffuse \\
+%   \bottomrule
+% \end{tabular}
+% }
+
+\block{Human-specific Genes}{
+{\small
+<<human.lineage.specific,echo=FALSE,results="asis">>=
+hum.all.lin.spec <-
+    placenta.lineage.specific.table("hum.all",
+                                    lineage.table=human.mouse.lineage.specific.analysis,
+                                    members=10,
+                                    orderbyfc=TRUE,
+                                    fdr.cutoff=0.001
+                                    )
+setkey(hum.all.lin.spec,"Human Symbol")
+@
+}
+\captionof{table}{1:1 Human orthologous genes whose expression
+  changed on the human lineage as opposed to all other analyzed
+  species.}
+}
+
+\block{Primate-specific Genes}{
+{\small
+<<primate.lineage.specific,echo=FALSE,results="asis">>=
+pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
+setkey(pri.all.lin.spec,"Human Symbol")
+@ 
+}
+}
+
+\column{0.2}
+
+\block{Core Placenta Transcriptome}{
+<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
+cptl <-
+    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
+                                            Species),]
+cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
+                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
+                                          capfirst(names(species.names))))]
+print(ggplot(cptl,
+             aes(y=human_name,x=Species))
+      + geom_raster(aes(fill=log10(FPKM)))
+      + scale_x_discrete(expand=c(0,0))
+      + scale_y_discrete(expand=c(0,0))
+      + scale_fill_gradient(low = grad.def[["low"]],
+                            high = grad.def[["high"]],
+                            space = grad.def[["space"]],
+                            na.value = grad.def[["na.value"]],
+                            limits=c(0,4),
+                            guide=guide_colorbar(title=expression(log[10](FPKM)))
+                            )
+      + ylab("Human Ortholog Name")
+      + xlab("Species")
+      + theme_grey()
+      + theme(legend.position="top",
+              axis.ticks = element_blank(),
+              axis.text.y = element_text(face="italic"),
+              axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic"))
+      )
+@ 
+\captionof{figure}{Core Placenta Transcriptome.
+  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
+  1:1 human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
+  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
+  studied with housekeeping genes
+  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
+  ordered (from top to bottom) by median expression.}
+\label{fig:core_placenta_transcriptome}
+}
+
+\block{\textit{PRSS16} Expression}{
+<<prss16.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"),
+                 fontsize=4,
+                 width=3,
+                 axis.text.size=3.5,
+                 offset.ratio=1.2
+                 )
+
+@ 
+\captionof{figure}{\textit{PRSS16} is highly expressed and massively
+  expanded in \elephant, but not in \armadillo. It is currently
+  unknown whether this high expression and expansion is a shared
+  feature afrotherians.}
+}
+
+\column{0.2}
+
+\block{\textit{CGA} Expression}{
+
+<<kiss1.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
+
+plot.label.size <- 4.5
+tree.id <-
+    genes.to.tree[combined.fpkm[gene_short_name=="KISS1" & species=="homo sapiens",gene_id],tree]
+gene.tree.table <- trees[[tree.id]]$table
+gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
+plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
+                 offset.ratio=1,
+                 fontsize=8,
+                 axis.text.size=5
+                 )
+@ 
+\captionof{figure}{\textit{KISS1}, which is implicated in
+  pre-eclampsia, is significantly more expressed on the human lineage
+  in comparison to all other lineages eamined. }
+}
+
+\block{Hormones, Galectins, \textit{IGF} and Pre-eclampsia}{
+<<hormone.genes.plot,echo=FALSE,fig.width=8,fig.height=10,warning=FALSE,message=FALSE>>=
+
+gene.expression.nozero <- function(genes) {
+    temp <-
+        data.matrix.dt(dcast(combined.fpkm[human_name %in% genes,],
+                             human_name~species,
+                             value.var="FPKM",
+                             fun.aggregate=function(x){x[1]}))
+    temp <-
+        temp[apply(temp,1,function(x){sum(x,na.rm=TRUE)})>5,]
+    return(temp)
+}
+
+### hormone genes
+hormone.genes <- c("CGA","CGB","CGB1","CGB2","CGB5","CGB8","CSH1","CSH2","CSHL1","CYP19A1",
+                   "GH2","GHRH","HSD3B1","HSD3B2","HSD17B2","HSD17B1","NPPB","NPPC","PRL","RLN1")
+hormone.expression.nozero <-
+    gene.expression.nozero(hormone.genes)
+
+hor.exp.long <- melt(log10(hormone.expression.nozero+1)[rev(1:nrow(hormone.expression.nozero)),])
+colnames(hor.exp.long)[1:3] <- c("gene","species","fpkm")
+hor.exp.long <- hor.exp.long[hor.exp.long$gene!="IGF2",]
+
+### galectin genes
+galectin.genes <- unique(grep("^LGALS",combined.fpkm[,human_name],value=TRUE))
+
+galectin.expression.nozero <-
+    gene.expression.nozero(galectin.genes)
+
+gal.exp.long <- melt(log10(galectin.expression.nozero+1)[rev(1:nrow(galectin.expression.nozero)),])
+colnames(gal.exp.long)[1:3] <- c("gene","species","fpkm")
+
+### IGF related genes
+
+igfrelated.genes <- unique(grep("^IGF",combined.fpkm[,human_name],value=TRUE))
+
+igfrelated.expression.nozero <-
+    gene.expression.nozero(igfrelated.genes)
+
+igf.exp.long <- melt(log10(igfrelated.expression.nozero+1)[rev(1:nrow(igfrelated.expression.nozero)),])
+colnames(igf.exp.long)[1:3] <- c("gene","species","fpkm")
+
+### Pre-eclampsia genes ADAM12 CRH KISS1 PAPPA
+
+pe.expression.nozero <-
+    gene.expression.nozero(c("ADAM12","CRH","KISS1","PAPPA"))
+
+pe.exp.long <- melt(log10(pe.expression.nozero+1)[rev(1:nrow(pe.expression.nozero)),])
+colnames(pe.exp.long)[1:3] <- c("gene","species","fpkm")
+
+igf.exp.long$type <- "IGF Related"
+gal.exp.long$type <- "Galectin"
+hor.exp.long$type <- "Hormone"
+pe.exp.long$type <- "PE"
+
+combined.long <-
+    rbind(igf.exp.long,
+          gal.exp.long,
+          hor.exp.long,
+          pe.exp.long
+          )
+combined.long$species <-
+    factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(combined.long$species)),
+           levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(names(species.names))))
+
+common.plot.options <- 
+    list(geom_raster(aes(fill = fpkm)),
+         scale_fill_gradient(low = grad.def[["low"]],
+                             high = grad.def[["high"]],
+                             space = grad.def[["space"]],
+                             na.value = grad.def[["na.value"]],
+                             guide=guide_colorbar(title=expression(log[10](FPKM)))),
+         ylab("Human Ortholog Name"),
+         xlab("Species"),
+         theme_grey(),
+         theme(legend.position = "top",
+               axis.ticks = element_blank(),
+               panel.background = element_blank(),
+               axis.text.y = element_text(face="italic"),
+               axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic")),
+         facet_grid(type~.,scale="free_y",space="free_y")
+         )
+print(ggplot(combined.long, aes(y=gene, x=species))
+      + common.plot.options
+      )
+@ 
+\captionof{figure}{Expression in $\mathrm{log}_{10}(\mathrm{FPKM})$ of 1:1 \human
+  orthologs of Hormone Genes, Galectins, genes implicated in
+  pre-eclampsia (PE), and \textit{IGF}-related genes. Empty locations
+  indicate no detectable expression and/or no 1:1 \human\ ortholog in
+  that species.}
+}
+
+\column{0.2}
 
 \block{Conclusions}{
+\begin{itemize}
+\item Expression resource of
+  \Sexpr{star.logs[,length(unique(species))]} species representing the
+  breadth of mammalia for future research.
+
+\item Lineage-specific expansions of gene families and differential
+  gene expression
+
+\item Components of annexin complexes, including \textit{ANXA2},
+  \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10} are expressed
+  in all placentas examined, and argue for the evolution importance of
+  annexins in placenta function, where they likely establish
+  maternal-fetal
+  tolerance\cite{Liu.ea2015:AnnexinA2complexeswith}.
+
+\item We identified of multiple genes (\textit{CRH},
+  \textit{ADAM12}, \textit{KISS1}, \textit{PAPPA}, \textit{IL1RL1},
+  and \textit{PLAC1}) which are associated with pre-eclampsia which
+  have increased expression on the \human lineage and may explain
+  larger proportion of pre-eclampsia observed in humans.
+
+\item Placenta samples from additional species and time points are
+  needed to more completely understand the complex molecular milieu
+  of the developing placenta.
+
+\end{itemize}
 }
@@
 
-\block{Code and Datasets}{
+\block{}{
 The code and underlying data for this poster can be found at
@@
 \end{center}
-% 
-}
-\block{}{
+
   \includegraphics[width=\linewidth,height=6cm,keepaspectratio]{figures/uofi_mark}

### 2016-04-28 poster/placenta_viviparous_poster.Rnw
KISS1 is not CGA
update the hypotheses to match the conclusions better

@@
     \item Major function is the transfer of nutrients, oxygen, and
-      waste between mother and offspring
+      waste between mother and offspring.
     \item Highly variable structure and form in mammals, even though
-      major function is conserved
-    \item Likely arose before the origin of therian mammals
+      major function is conserved.
+    \item Likely arose before the origin of therian mammals.
     \item Form relatively well characterized, but diversity of
-      molecular environment understudied
+      molecular environment understudied.
     \item Role in the evolutionary interplay between maternal and fetal
-      strategies
+      strategies.
     \end{itemize}
@@
       \textit{Gorilla gorilla}
-      \cite{Thornton.ea1992:Convulsionsinpregnancyin}
+      \cite{Thornton.ea1992:Convulsionsinpregnancyin}.
     \item Far higher incidence in \textit{Homo sapiens}, with as many as 
@@
     \begin{itemize}
-    \item Gene expression patterns correlate with placenta morphology
-    \item Altered expression of genes correlates with invasiveness, and
-      therefore pre-eclampsia
-    \item Maternal/Fetal evolutionary strategies will be reflected in gene
-      expression patterns
+    \item Mechanisms of maintaining maternal-fetal tolerance will be
+      conserved.
+    \item Altered expression of genes correlates with invasiveness,
+      and therefore pre-eclampsia
+    \item Maternal/Fetal evolutionary strategies will be reflected in
+      gene expression patterns
     \end{itemize}
@@
 
-\block{\textit{CGA} Expression}{
+\block{\textit{KISS1} Expression}{
 

### 2016-04-29 poster/placenta_viviparous_poster.Rnw
update poster plot margins

@@
               axis.ticks = element_blank(),
+              plot.margin = unit(c(0,0,0,0),"mm"),
               axis.text.y = element_text(face="italic"),
@@
 @ 
-\captionof{figure}{Core Placenta Transcriptome.
-  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-  1:1 human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
-  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
-  studied with housekeeping genes
-  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
-  ordered (from top to bottom) by median expression.}
+\captionof{figure}{\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
+  1:1 human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in
+  placentas of all \Sexpr{star.logs[,length(unique(species))]} species
+  studied ordered by median expression without housekeeping genes
+  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.}
 \label{fig:core_placenta_transcriptome}
@@
                axis.ticks = element_blank(),
+               plot.margin = unit(c(0,0,0,0),"mm"),
                panel.background = element_blank(),

### 2016-05-16 manuscript/manuscript.Rnw
don't include figures
fix double include typo
start working on reformatting for genes & development

@@
-%\documentclass[english,12pt]{article}
-\documentclass[doublespacing]{bmcart_dla}
+\documentclass[english,12pt]{article}
 \usepackage{fontspec}
-\setmainfont{FreeSans}
+\setmainfont{FreeSerif}
 \setsansfont{FreeSans}
 \setmonofont{FreeMono}
-\setmathrm{FreeSans}
+\setmathrm{FreeSerif}
 \setmathsf{FreeSans}
 \setmathtt{FreeSans}
-%\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
-\usepackage[letterpaper]{geometry}
+\usepackage[letterpaper,left=0.75in,right=0.75in,top=0.75in,bottom=0.75in,headheight=14pt]{geometry}
 \usepackage{url}
-% \usepackage{fancyhdr}
+\usepackage{fancyhdr}
 \usepackage{graphicx}
@@
 \usepackage{sansmath}
-\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
+\usepackage[backend=biber,natbib=true,hyperref=true,citestyle=authoryear-comp,firstinits=true,sorting=nyt,bibstyle=authoryear,autocite=inline,maxcitenames=1,mincitenames=1,maxbibnames=10,minbibnames=10]{biblatex}
+\DefineBibliographyStrings{english}{andothers = {\emph{et\addabbrvspace al\adddot}}}
 \addbibresource{references.bib}
-% \usepackage[noblocks,auth-sc]{authblk}
+\usepackage[noblocks,auth-sc]{authblk}
 \usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue]{hyperref}
@@
 %\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
+\renewcommand{\includegraphics}[2][]{}
 \usepackage[nomargin,inline,draft]{fixme}
@@
 \usepackage[capitalize]{cleveref}
-% \newenvironment{narrow}[2]{%
-%   \begin{list}{}{%
-%       \setlength{\topsep}{0pt}%
-%       \setlength{\leftmargin}{#1}%
-%       \setlength{\rightmargin}{#2}%
-%       \setlength{\listparindent}{\parindent}%
-%       \setlength{\itemindent}{\parindent}%
-%       \setlength{\parsep}{\parskip}}%
-%   \item[]}{\end{list}}
-% \def\newblock{\hskip}
-% \newenvironment{paperquote}{%
-%   \begin{quote}%
-%      \it
-%   }%
-%   {\end{quote}}
-% \renewcommand{\textfraction}{0.15}
-% \renewcommand{\topfraction}{0.85}
-% \renewcommand{\bottomfraction}{0.65}
-% \renewcommand{\floatpagefraction}{0.60}
-%\renewcommand{\baselinestretch}{1.8}
-% \newenvironment{enumerate*}%
-%   {\begin{enumerate}%
-%     \setlength{\itemsep}{0pt}%
-%     \setlength{\parskip}{0pt}}%
-%   {\end{enumerate}}
-% \newenvironment{itemize*}%
-%   {\begin{itemize}%
-%     \setlength{\itemsep}{0pt}%
-%     \setlength{\parskip}{0pt}}%
-%   {\end{itemize}}
-% \setcounter{tocdepth}{3}
-% \oddsidemargin 0.0in 
-% \textwidth 6in
-% \raggedbottom
-% \clubpenalty = 10000
-% \widowpenalty = 10000
+\raggedbottom
+\clubpenalty = 10000
+\widowpenalty = 10000
 % set a large emergency stretch to avoid issues with overfull boxes
@@
 % make the afill smaller
-% \renewcommand\Affilfont{\itshape\small}
+\renewcommand\Affilfont{\itshape\small}
 \renewcommand\cite[1]{\autocite{#1}}
@@
 \begin{document}
-\raggedright
 
@@
 
-\begin{frontmatter}
-\begin{fmbox}
-\dochead{Research}%
 \title{The evolution of gene expression in the term placenta of viviparous mammals}%
 %
-\author[addressref={aff1},noteref={n1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
-\author[addressref={aff2},noteref={n1},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
-\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL}\fnm{Amy} \snm{Weckle}}
-\author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}
-\author[addressref={aff4},email={agnewd@dcpah.msu.edu}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
-\author[addressref={aff5},email={kbenirschke@ucsd.edu}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
-\author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
-\author[addressref={aff11},email={nevo@research.haifa.ac.il}]{\inits{E}\fnm{Eviatar} \snm{Nevo}}
-\author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
-\author[addressref={aff9},email={gunter.wagner@yale.edu}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
-\author[addressref={aff7},email={prbchiefstaff@med.wayne.edu}]{\inits{R}\fnm{Roberto} \snm{Romero}}
-\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
-\address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[id=aff2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
-\address[id=aff3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[id=aff4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
-\address[id=aff5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
-\address[id=aff6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
-\address[id=aff7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
-\address[id=aff8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[id=aff9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
-\address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[id=aff11]{The Institute of Evolution, University of Haifa, Haifa, Israel}
-
-\begin{artnotes}
-\end{artnotes}
-
-\begin{artnotes}
-\note[id=n1]{Equal contributor} % note, connected to author
-\end{artnotes}
-
-\end{fmbox}
-% \twocolumn
-
-\begin{abstractbox}
-\begin{abstract}
-% \section{Abstract}
-  \parttitle{Background} The placenta is arguably the most
+%\author[addressref={aff1},noteref={n1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
+\author[1,†]{Don L. Armstrong}
+%\author[addressref={aff2},noteref={n1},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
+\author[2,†]{Michael R. McGowen}
+%\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL}\fnm{Amy} \snm{Weckle}}
+\author[1]{Amy Weckle}
+%\author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}
+\author[3]{Jason Caravas}
+%\author[addressref={aff4},email={agnewd@dcpah.msu.edu}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
+\author[4]{Dalen Agnew}
+%\author[addressref={aff5},email={kbenirschke@ucsd.edu}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
+\author[5]{Kurt Benirschke}
+%\author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
+\author[6]{Sue Savage-Rumbaugh}
+%\author[addressref={aff11},email={nevo@research.haifa.ac.il}]{\inits{E}\fnm{Eviatar} \snm{Nevo}}
+\author[7]{Eviatar Nevo}
+%\author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
+\author[8,9]{Chong J. Kim}
+%\author[addressref={aff9},email={gunter.wagner@yale.edu}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
+\author[10]{Gunter Wagner}
+% \author[addressref={aff7},email={prbchiefstaff@med.wayne.edu}]{\inits{R}\fnm{Roberto} \snm{Romero}}
+\author[8]{Roberto Romero}
+%\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
+\author[1,11,*]{Derek E. Wildman}
+\affil[1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
+\affil[2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
+\affil[3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
+\affil[4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
+\affil[5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
+\affil[6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
+\affil[7]{The Institute of Evolution, University of Haifa, Haifa, Israel}
+\affil[8]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
+\affil[9]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
+\affil[10]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
+\affil[11]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
+\affil[†]{Equal contributors}
+\affil[*]{To whom correspondence should be addressed: wildman@illinois.edu}
+
+\maketitle
+
+ \section{Abstract}
+  The placenta is arguably the most
   anatomically variable organ in mammals. Understanding gene
@@
 
-  \parttitle{Results} We measured the expression profiles of
+  We measured the expression profiles of
   \Sexpr{star.logs[,length(unique(file))]} term placentas of
@@
 
-  \parttitle{Conclusions} A diverse array of genes are differentially
+  A diverse array of genes are differentially
   expressed and are likely to play a role in the anatomical
@@
 \textbf{Keywords:} mammal, gene expression, evolution, RNA-seq, placenta
-
-\end{abstract}
-\end{abstractbox}
-\end{frontmatter}
 
@@
         table.display <- c("s","s","s","d")
-        table.align <- c("c","c","p{3in}","r")
+        table.align <- c("c","c","p{4in}","r")
         if (any(colnames(data) %in% "SEM"))  {
@@
 \begin{table}
-  \footnotesize
+  \small
 <<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}
-  \footnotesize
+  \small
 <<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}
-  \footnotesize
+  \small
 <<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}[h]
-  \footnotesize
+  \small
 <<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}[h]
-  \footnotesize
+  \small
 <<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}[h]
-  \footnotesize
+  \small
 <<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}
-  \footnotesize
+  \small
 <<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}
-  \footnotesize
+  \small
 <<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}[h]
-  \footnotesize
+  \small
 <<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}[h]
-  \footnotesize
+  \small
 <<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}[h]
-  \footnotesize
+  \small
 <<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
@@
 \begin{table}
-  \footnotesize
+  \small
 <<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
@@
 identify the set of expressed genes which are common to all placentas.
-Components of the annexin complexes, including including
+Components of the annexin complexes, including
 \textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}

### 2016-06-08 manuscript/manuscript.Rnw
Housekeeping genes shape in different species

--- 2016-04-19 manuscript/manuscript.Rnw
@@
 
+In contrast to the shape of the number of genes included in the core
+placenta transcriptome, the set of \human housekeeping genes expressed
+all placentas is flatter (Supplemental
+\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
+of the total fraction of genes. However, there is still a large
+difference between the set of housekeeping genes which are expressed
+in \human and those which are expressed in all placentas.
+
 % SPINT2 -- Serine Peptidase inhibitor

### 2016-06-08 manuscript/supplemental_material.Rnw
Housekeeping genes shape in different species

@@
 
+\begin{figure}
+<<housekeeping.placenta.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
+## housekeeping set as apposed to non-housekeeping set
+cptg.h <- core.placenta.transcriptome.genes.shape[(human_name %in% housekeeping.genes.superset[,gene_short_name])]
+
+cpt.h.distribution.shape <- 
+    lapply(c("all_expression",
+             "all_but_one_expression",
+             "all_but_two_expression",
+             "all_but_three_expression",
+             "all_but_four_expression"),
+           function(x){
+               calculate.distribution.shape(cptg.h[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
+           })
+fix.expression.names <- function(x){
+    x <- gsub("^all_expression$",
+              paste0(num.species," of ",num.species," species"),
+              x
+              )
+    nc.levels <- paste0(num.species," of ",num.species," species")
+    ## number coutner
+    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
+    for (a in names(nc)) {
+        x <- gsub(paste0("^all_but_",a,"_expression$"),
+              paste0(num.species-nc[a]," of ",num.species," species"),
+              x
+              )
+        nc.levels <-
+            c(nc.levels,
+              paste0(num.species-nc[a]," of ",num.species," species"))
+    }
+    return(factor(x,levels=nc.levels))
+}
+
+cpt.h.distribution.shape <- 
+    rbindlist(cpt.h.distribution.shape)
+
+cpt.h.distribution.shape <- 
+    cpt.h.distribution.shape[,`Minimum Expression in`:=
+                                fix.expression.names(`Minimum Expression in`)]
+
+setnames(cpt.h.distribution.shape,
+         "FPKM",
+         "Expression (FPKM)")
+
+print(ggplot(cpt.h.distribution.shape[`Expression (FPKM)` > 0.1,],
+             aes(x=`Expression (FPKM)`,color=`Minimum Expression in`,y=num_genes))
+      + geom_line(lwd=1)
+      + xlab("Minimum Expression cutoff (FPKM)")
+      + ylab("Genes with N-th Minimum Expression")
+      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                      labels = trans_format("log10", math_format(10^.x)))
+      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                      labels = trans_format("log10", math_format(10^.x)))
+      + geom_vline(xintercept=10)
+      + geom_segment(mapping=aes(y=V1,
+                         yend=V1,
+                         xend=10,
+                         x=250,
+                         color=`Minimum Expression in`),
+                     data=cpt.h.distribution.shape[`Expression (FPKM)` > 10,
+                                              max(num_genes),`Minimum Expression in`],
+                     arrow=arrow(length=unit(0.5,"cm"))
+                   )
+      + geom_text(mapping=aes(x=250,
+                      y=V1,
+                      label=paste0(V1," genes"),override=TRUE),
+                  data=cpt.h.distribution.shape[`Expression (FPKM)` > 10,
+                      max(num_genes),`Minimum Expression in`],
+                  show_guide=FALSE,
+                  hjust=0
+                  )
+      + guides(color=guide_legend(title="Minimum\nExpression in"))
+      )
+@ 
+\caption{Distribution shape of housekeeping genes in placentas out of
+  \Sexpr{pts[,sum(expressed.housekeeping)]} \human housekeeping 
+  genes which are expressed in any placenta. The vertical axis
+  shows the number of 1:1 orthologous \human housekeeping genes
+  expressed in 
+  placenta transcriptomes if %
+  \Sexpr{cpt.h.distribution.shape[,levels(`Minimum Expression in`)[1]]}
+  %
+  (or %
+  \Sexpr{cpt.h.distribution.shape[,levels(`Minimum Expression in`)[2]]},
+  % 
+  etc.) have a minimum expression in FPKM equal to the horizontal axis.
+  For example, if the cutoff is set to 10 (vertical black line), 
+  % 
+  \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
+    & `Minimum Expression in` == '14 of 14 species',max(num_genes)]}
+  %
+  genes would be included if the FPKM of a set of 1:1 orthologs of all 14 species must be at
+  least 10, %
+  \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
+    & `Minimum Expression in` == "13 of 14 species",max(num_genes)]}
+  %
+  1:1 orthologs would be included if 13 of 14 species have an
+   with an FPKM of at least 10, and 
+  \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
+    & `Minimum Expression in` == '10 of 14 species',max(num_genes)]}
+  genes would be included if 10 of 14 species
+  must have an FPKM of at least 10.}
+\label{fig:core_housekeeping_transcriptome_shape}
+\end{figure}
+
+\clearpage
+
 \printbibliography

### 2016-06-08 data/placenta_core_transcriptome.R

@@ -86,7 +86,21 @@ pts <- pts[,placenta.ts.permissive:=all.genes %in%
                                                         ! is.na(human_name),
                                                         unique(human_name)]
            ]
+pts[,placenta.ts.all:=all.genes %in%
+         core.placenta.transcriptome.genes.shape[all_expression > 10 &
+                                                 ! is.na(human_name),
+                                                 unique(human_name)]]
+pts[,placenta.ts.one:=all.genes %in%
+         core.placenta.transcriptome.genes.shape[all_expression > 1 &
+                                                 ! is.na(human_name),
+                                                 unique(human_name)]]
+expressed.housekeeping.genes <-
+    core.placenta.transcriptome.genes.shape[max_expression > 10 &
+                                            human_name %in%
+                                            housekeeping.genes.superset[,gene_short_name]
+                                            ][,unique(human_name)]
 pts <- pts[,housekeeping:=all.genes %in% housekeeping.genes.superset[,gene_short_name]]
+pts <- pts[,expressed.housekeeping:=all.genes %in% expressed.housekeeping.genes]
 pts <- pts[,egid:=as.vector(unlist(sapply(mget(all.genes,org.Hs.egSYMBOL2EG,ifnotfound=NA),
                                           function(x){x[1]})))]
 ## this list contains all possible genes with a 0 if it is not in the

### 2016-08-12 manuscript/manuscript.Rnw
increase alpha in phylopic

@@
 \begin{figure}
-<<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
+<<species.tree,fig.width=8,fig.height=6,out.width="\\columnwidth",echo=FALSE>>=
 species.tree <- 
@@
       ## monodelphis
-      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.5,node=1,width=0.2)
+      phylopic("91324e57-b3f1-42e0-abe3-43e5bc8aa4c6", color="#86B875",alpha=0.3,node=1,width=0.2)
       ## Armadillo
-      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.5,node=2,width=0.15)
+      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.3,node=2,width=0.15)
       ## elephant
-      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.5,node=3,width=0.15)
+      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.3,node=3,width=0.15)
       ## dog
-      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.5,node=4,width=0.15)
+      %>% phylopic("4d83a0cd-cf06-4a32-9a5a-0a6b644158c1",color="#FFAACC",alpha=0.3,node=4,width=0.15)
       ## horse
-      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.5,node=5,width=0.15)
+      %>% phylopic("071ee517-a0f1-4d19-aa29-812b9f86cb53",color="#22BBBB",alpha=0.3,node=5,width=0.15)
       ## pig
-      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.5,node=6,width=0.12)
+      %>% phylopic("008d6d88-d1be-470a-8c70-73625c3fb4fb",color="#33AA11",alpha=0.3,node=6,width=0.12)
       ## sheep -- this is a male, the sample is female, but there aren't any females in phylopic
-      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.5,node=7,width=0.18)
+      %>% phylopic("cbea1964-89bf-4ce1-99b0-58fe0f9170b3",color="#004411",alpha=0.3,node=7,width=0.18)
       ## cow
-      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.5,node=8,width=0.12)
+      %>% phylopic("415714b4-859c-4d1c-9ce0-9e1081613df7",color="#881111",alpha=0.3,node=8,width=0.12)
       ## s carmeli
-      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.5,node=9,width=0.1)
+      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#881188",alpha=0.3,node=9,width=0.1)
       ## n galili
-      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.5,node=10,width=0.1)
+      %>% phylopic("287c58ad-009b-4878-9a9e-59fd7e0fe66f",color="#888888",alpha=0.3,node=10,width=0.1)
       ## mouse
-      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.5,node=11,width=0.18)
+      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.3,node=11,width=0.18)
       ## bonobo
-      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.5,node=12,width=0.18)
+      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.3,node=12,width=0.18)
       ## spider monkey
-      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.5,node=14,width=0.25)
+      %>% phylopic("aceb287d-84cf-46f1-868c-4797c4ac54a8",color="#FF11FF",alpha=0.3,node=14,width=0.25)
       ## human
-      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.5,node=13,width=0.18)
+      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.3,node=13,width=0.18)
       )

### 2016-08-12 manuscript/manuscript.Rnw
use mtime instead of md5sum

@@
 
-<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
+<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=file.info(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))[,"mtime"]>>=
 for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R"))) {
@@
 
-<<load.data,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","placenta_go_results_mf","hugo_nomenclature","gene_annotation")))>>=
+<<load.data,echo=FALSE,results="hide",cache.extra=file.info(paste0("../data/",c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome","placenta_go_results_bp","placenta_go_results_mf","hugo_nomenclature","gene_annotation")))[,"mtime"]>>=
 for (a in paste0("../data/",c("combined_fpkm","gene_trees",

### 2016-08-12 manuscript/manuscript.Rnw
remove top50 tables

@@
 
-\begin{table}
-  \footnotesize
-<<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="homo sapiens"))
-@ 
-\caption{The top 50 genes by \ac{FPKM} in \human which are not
-  annotated as Mt\_rRNA, Mt\_tRNA, rRNA or a housekeeping gene. Gene
-  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
-  name corresponding to the human symbol. A complete list of all genes
-  is in \cref{tab:all_homo_sapiens_placenta}.}
-\label{tab:top50_human_placenta}
-\end{table}
+% \begin{table}
+%   \footnotesize
+% <<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
+% output.top.genes.table(top.genes.table(species.name="homo sapiens"))
+% @ 
+% \caption{The top 50 genes by \ac{FPKM} in \human which are not
+%   annotated as Mt\_rRNA, Mt\_tRNA, rRNA or a housekeeping gene. Gene
+%   symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
+%   name corresponding to the human symbol. A complete list of all genes
+%   is in \cref{tab:all_homo_sapiens_placenta}.}
+% \label{tab:top50_human_placenta}
+% \end{table}
 
@@
 
-\begin{table}
-  \footnotesize
-<<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
-top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
-output.top.genes.table(top.bonobo.genes)
-setkey(top.bonobo.genes,"Gene symbol")
-@ 
-\caption{\tableexplanation{\bonobo}{tab:all_pan_paniscus_placenta}}
-\label{tab:top50_bonobo_placenta}
-\end{table}
+% \begin{table}
+%   \footnotesize
+% <<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
+% top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
+% output.top.genes.table(top.bonobo.genes)
+% setkey(top.bonobo.genes,"Gene symbol")
+% @ 
+% \caption{\tableexplanation{\bonobo}{tab:all_pan_paniscus_placenta}}
+% \label{tab:top50_bonobo_placenta}
+% \end{table}
 
@@
 
-\begin{table}[h]
-  \footnotesize
-<<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="mus musculus"))
-@ 
-\caption{\tableexplanation{\mouse}{tab:all_mus_musculus_placenta}}
-\label{tab:top50_mouse_placenta}
-\end{table}
+% \begin{table}[h]
+%   \footnotesize
+% <<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
+% output.top.genes.table(top.genes.table(species.name="mus musculus"))
+% @ 
+% \caption{\tableexplanation{\mouse}{tab:all_mus_musculus_placenta}}
+% \label{tab:top50_mouse_placenta}
+% \end{table}
 
@@
 
-\begin{table}[h]
-  \footnotesize
-<<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="spalax galili"))
-@ 
-\caption{\tableexplanation{\galili}{tab:all_spalax_galili_placenta}}
-\label{tab:top50_n_galili_placenta}
-\end{table}
+% \begin{table}[h]
+%   \footnotesize
+% <<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
+% output.top.genes.table(top.genes.table(species.name="spalax galili"))
+% @ 
+% \caption{\tableexplanation{\galili}{tab:all_spalax_galili_placenta}}
+% \label{tab:top50_n_galili_placenta}
+% \end{table}
 
@@
 
-\begin{table}[h]
-  \footnotesize
-<<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="spalax carmeli"))
-@ 
-\caption{\tableexplanation{\carmeli}{tab:all_spalax_carmeli_placenta}}
-\label{tab:top50_s_carmeli_placenta}
-\end{table}
+% \begin{table}[h]
+%   \footnotesize
+% <<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
+% output.top.genes.table(top.genes.table(species.name="spalax carmeli"))
+% @ 
+% \caption{\tableexplanation{\carmeli}{tab:all_spalax_carmeli_placenta}}
+% \label{tab:top50_s_carmeli_placenta}
+% \end{table}
 
@@
 
-\begin{table}
-  \footnotesize
-<<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="canis familiaris"))
-@ 
-\caption{\tableexplanation{\dog}{tab:all_canis_familiaris_placenta}}
-\label{tab:top50_dog_placenta}
-\end{table}
+% \begin{table}
+%   \footnotesize
+% <<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
+% output.top.genes.table(top.genes.table(species.name="canis familiaris"))
+% @ 
+% \caption{\tableexplanation{\dog}{tab:all_canis_familiaris_placenta}}
+% \label{tab:top50_dog_placenta}
+% \end{table}
 
@@
 
-\begin{table}
-  \footnotesize
-<<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="bos taurus"))
-@ 
-\caption{\tableexplanation{\cow}{tab:all_bos_taurus_placenta}}
-\label{tab:top50_cow_placenta}
-\end{table}
+% \begin{table}
+%   \footnotesize
+% <<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
+% output.top.genes.table(top.genes.table(species.name="bos taurus"))
+% @ 
+% \caption{\tableexplanation{\cow}{tab:all_bos_taurus_placenta}}
+% \label{tab:top50_cow_placenta}
+% \end{table}
 
@@
 
-\begin{table}[h]
-  \footnotesize
-<<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="sus scrofa"))
-@ 
-\caption{\tableexplanation{\pig}{tab:all_sus_scrofa_placenta}}
-\label{tab:top50_pig_placenta}
-\end{table}
+% \begin{table}[h]
+%   \footnotesize
+% <<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
+% output.top.genes.table(top.genes.table(species.name="sus scrofa"))
+% @ 
+% \caption{\tableexplanation{\pig}{tab:all_sus_scrofa_placenta}}
+% \label{tab:top50_pig_placenta}
+% \end{table}
 
@@
 
-\begin{table}[h]
-  \footnotesize
-<<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="equus caballus"))
-@ 
-\caption{\tableexplanation{\horse}{tab:all_equus_caballus_placenta}}
-\label{tab:top50_horse_placenta}
-\end{table}
+% \begin{table}[h]
+%   \footnotesize
+% <<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
+% output.top.genes.table(top.genes.table(species.name="equus caballus"))
+% @ 
+% \caption{\tableexplanation{\horse}{tab:all_equus_caballus_placenta}}
+% \label{tab:top50_horse_placenta}
+% \end{table}
 
@@
 
-\begin{table}[h]
-  \footnotesize
-<<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
-top.sheep.genes <- top.genes.table(species.name="ovis aries",n=200)
-top.sheep.genes <-
-    top.sheep.genes[!grepl("\\{(7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\}",
-                           `Gene symbol`) &
-                    !grepl("\\((7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\)",
-                           `Human Symbol`)
-                   ,][1:50,]
-output.top.genes.table(top.sheep.genes)
-@ 
-\caption{\tableexplanation{\sheep}{tab:all_ovis_aries_placenta}}
-\label{tab:top50_sheep_placenta}
-\end{table}
+% \begin{table}[h]
+%   \footnotesize
+% <<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
+% top.sheep.genes <- top.genes.table(species.name="ovis aries",n=200)
+% top.sheep.genes <-
+%     top.sheep.genes[!grepl("\\{(7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\}",
+%                            `Gene symbol`) &
+%                     !grepl("\\((7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\)",
+%                            `Human Symbol`)
+%                    ,][1:50,]
+% output.top.genes.table(top.sheep.genes)
+% @ 
+% \caption{\tableexplanation{\sheep}{tab:all_ovis_aries_placenta}}
+% \label{tab:top50_sheep_placenta}
+% \end{table}
 
@@
 
-\begin{table}
-\footnotesize
-<<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="loxodonta africana"))
-@ 
-\caption{\tableexplanation{\elephant}{tab:top50_elephant_placenta}}
-\label{tab:top50_elephant_placenta}
-\end{table}
+% \begin{table}
+% \footnotesize
+% <<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
+% output.top.genes.table(top.genes.table(species.name="loxodonta africana"))
+% @ 
+% \caption{\tableexplanation{\elephant}{tab:top50_elephant_placenta}}
+% \label{tab:top50_elephant_placenta}
+% \end{table}
 
@@
 
-\begin{table}
-\footnotesize
-<<top.50.armadillo.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="dasypus novemcinctus"))
-@ 
-\caption{\tableexplanation{\armadillo}{tab:all_dasypus_novemcinctus_placenta}}
-\label{tab:top50_armadillo_placenta}
-\end{table}
+% \begin{table}
+% \footnotesize
+% <<top.50.armadillo.expressed.genes,echo=FALSE,results="asis">>=
+% output.top.genes.table(top.genes.table(species.name="dasypus novemcinctus"))
+% @ 
+% \caption{\tableexplanation{\armadillo}{tab:all_dasypus_novemcinctus_placenta}}
+% \label{tab:top50_armadillo_placenta}
+% \end{table}
 
@@
 
-\begin{table}
-  \footnotesize
-<<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
-output.top.genes.table(top.genes.table(species.name="monodelphis domestica"))
-@ 
-\caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica_placenta}}
-\label{tab:top50_opossum_placenta}
-\end{table}
+% \begin{table}
+%   \footnotesize
+% <<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
+% output.top.genes.table(top.genes.table(species.name="monodelphis domestica"))
+% @ 
+% \caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica_placenta}}
+% \label{tab:top50_opossum_placenta}
+% \end{table}
 
@@
 
-  \printaddresses
-
-\end{backmatter}
+%  \printaddresses
+
+%\end{backmatter}
 

### 2016-08-12 manuscript/manuscript.Rnw
add main mentioned genes table

@@
 }
+@
+
+<<gene.expression,results="hide",echo=FALSE>>=
+mentioned.genes <- load_cache(label="save_mentioned_genes",object="saved_mentioned_genes",list())
 gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE) {
     species.name <- gsub("_"," ",species.name)
+    mentioned.genes[[paste0(species.name,"_",gene)]] <<-
+        list("species.name"=species.name,
+             "gene"=gene)
     fpkm <- combined.fpkm[species==species.name &
@@
                                                   "}")]
+        # setnames(display.table,"Human Symbol","\\multicolumn{1}{p{0.5in}}{Human Symbol}")6
         setnames(display.table,"Fold Change","$\\text{log}_2$ FC")
@@
                      display=c("s","s","s","e","g","f"),
-                     align=c("c","c","p{2in}","r","r","r")
+                     align=c("c","p{0.5in}","p{1in}","r","r","r")
                      ),
@@
 
+<<mentioned.genes.table.functions,echo=FALSE,results="asis">>=
+mentioned.genes.table <- list()
+if (length(mentioned.genes)==0) {
+    mentioned.genes.table[["No genes"]] <-
+        data.table(species="No genes",
+                   name_or_id="No genes",
+                   human_alignment_symbol="NOGENE",
+                   human_name="NONAME",
+                   mean_fpkm = 0,
+                   sd_fpkm = 1,
+                   n_fpkm = 2)
+}
+for (entry in names(mentioned.genes)) {
+    mentioned.genes.table[[entry]] <-
+        combined.fpkm[species==mentioned.genes[[entry]][["species.name"]] &
+                      (gene_short_name==mentioned.genes[[entry]][["gene"]] | 
+                       gene_id==mentioned.genes[[entry]][["gene"]]),
+                      list(species,
+                           name_or_id,
+                           human_alignment_symbol,
+                           human_name,
+                           mean_fpkm,
+                           sd_fpkm,
+                           n_fpkm)]
+}
+mentioned.genes.table <-
+    rbindlist(mentioned.genes.table)
+mentioned.genes.table[,used_alignment:=FALSE]
+mentioned.genes.table[is.na(human_name) & !is.na(human_alignment_symbol),
+                      used_alignment:=TRUE]
+mentioned.genes.table[is.na(human_name),human_name:=human_alignment_symbol]
+setkey(mentioned.genes.table,"human_name")
+mentioned.genes.table <- 
+    hugo.nomenclature[mentioned.genes.table][,
+                                             list(species,
+                                                  name_or_id,
+                                                  used_alignment,
+                                                  `Approved Symbol`,
+                                                  `Approved Name`,
+                                                  mean_fpkm,sd_fpkm,n_fpkm)]
+mentioned.genes.table[,alignment.species:=species]
+mentioned.genes.table[alignment.species=="pan paniscus",alignment.species:="pan troglodytes"]
+mentioned.genes.table[alignment.species=="ateles fusciceps",alignment.species:="homo sapiens"]
+mentioned.genes.table[grepl("spalax",alignment.species),alignment.species:="mus musculus"]
+mentioned.genes.table[!is.na(name_or_id),
+                      name_or_id:=paste0("\\href{http://www.ensembl.org/",
+                                         capfirst(gsub(" ","_",alignment.species)),
+                                         "/Gene/Summary?g=",name_or_id,"}{\\textit{",
+                                         xtable.sanitize.text(name_or_id),"}}")]
+mentioned.genes.table[!is.na(`Approved Name`),
+                      `Approved Name`:=paste0("\\href{http://www.ensembl.org/",
+                                              "Homo_sapiens/Gene/Summary?g=",
+                                              `Approved Symbol`,"}{",
+                                              xtable.sanitize.text(`Approved Name`),
+                                              "}")]
+mentioned.genes.table[!is.na(`Approved Symbol`),
+                      `Approved Symbol`:=paste0("\\href{http://www.ensembl.org/",
+                                                "Homo_sapiens/Gene/Summary?g=",
+                                                `Approved Symbol`,
+                                                "}{\\textit{",
+                                                xtable.sanitize.text(`Approved Symbol`),
+                                                "}}")]
+mentioned.genes.table[used_alignment==TRUE,`Approved Symbol`:=paste0("(",`Approved Symbol`,")")]
+mentioned.genes.table[,sem_fpkm:=as.numeric(NA)]
+mentioned.genes.table[n_fpkm>1,sem_fpkm:=sd_fpkm/sqrt(n_fpkm)]
+mentioned.genes.table[,used_alignment:=NULL]
+mentioned.genes.table[,alignment.species:=NULL]
+mentioned.genes.table[,sd_fpkm:=NULL]
+mentioned.genes.table[,n_fpkm:=NULL]
+mentioned.genes.table[,`Approved Name`:=NULL]
+### order by species and then by FPKM
+setorder(mentioned.genes.table,`Approved Symbol`,species,-mean_fpkm)
+### italicize species names
+mentioned.genes.table[,species:=paste0("\\textit{",gsub("^(.)","\\U\\1",species,perl=TRUE),"}")]
+setnames(mentioned.genes.table,c("Species","Gene Symbol",
+                                 "Human Symbol",
+                                 ## "Human Name",
+                                 "FPKM","SEM"))
+mentioned.genes.table[,`FPKM`:=round(`FPKM`)]
+
+output.mentioned.genes.table <- function(subset.table=quote(TRUE),
+                                         human.symbol=TRUE) {
+    mgt.subset <- mentioned.genes.table[eval(subset.table)]
+    mgt.display <- c("s","s","s","s","d","d")
+    mgt.align <- c("c","c",
+                   ">{\\raggedright \\centering}p{0.5in}",
+                   ">{\\raggedright \\centering}p{0.5in}",
+                   ##">{\\raggedright}p{4in}",
+                   "r","r")
+    if (!human.symbol) {
+        mgt.subset[,`Human Symbol`:=NULL]
+        mgt.display <- mgt.display[-4]
+        mgt.align <- mgt.align[-4]
+    }
+
+    print(xtable(mgt.subset,
+                 display=mgt.display,
+                 align=mgt.align
+                 ),
+          math.style.exponents=TRUE,
+          booktabs=TRUE,
+          floating=FALSE,
+          sanitize.text.function=genes.table.sanitize,
+          include.rownames=FALSE)
+}
+@
+
 \subsection{Primate Placenta Transcriptomes}
+
+\begin{table}
+\footnotesize
+<<primate.genes,echo=FALSE,results="asis">>=
+output.mentioned.genes.table(subset.table=quote(grepl("Homo|Ateles|Pan",Species)),
+                             human.symbol=FALSE
+                             )
+@ 
+\caption{Notable genes ordered by species and \ac{FPKM}. Gene
+  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
+  name corresponding to the human symbol. A complete table of 
+  genes with \ac{FPKM} $\ge 50$ is located in the supplemental material }
+\label{tab:notable_genes}
+\end{table}
 
@@
 
+<<save_mentioned_genes,echo=FALSE,results="hide">>=
+saved_mentioned_genes <- mentioned.genes
+@
+
 \end{document}

### 2016-08-12 manuscript/manuscript.Rnw
change manuscript to use placenta's TeX style

@@
 
-\begin{frontmatter}
-\begin{fmbox}
-\dochead{Research}%
-\title{The evolution of gene expression in the term placenta of viviparous mammals}%
-%
-\author[addressref={aff1},noteref={n1},email={don@donarmstrong.com}]{\inits{DL}\fnm{Don L} \snm{Armstrong}}
-\author[addressref={aff2},noteref={n1},email={1mmcgowen1@gmail.com}]{\inits{MR}\fnm{Michael R.} \snm{McGowen}}
-\author[addressref={aff1},email={aweckle@illinois.edu}]{\inits{AL}\fnm{Amy} \snm{Weckle}}
-\author[addressref={aff3},email={jacaravas@gmail.com}]{\inits{J}\fnm{Jason} \snm{Caravas}}
-\author[addressref={aff4},email={agnewd@dcpah.msu.edu}]{\inits{D}\fnm{Dalen} \snm{Agnew}}
-\author[addressref={aff5},email={kbenirschke@ucsd.edu}]{\inits{K}\fnm{Kurt} \snm{Benirschke}}
-\author[addressref={aff6},email={suerumbaugh@gmail.com}]{\inits{S}\fnm{Sue} \snm{Savage-Rumbaugh}}
-\author[addressref={aff11},email={nevo@research.haifa.ac.il}]{\inits{E}\fnm{Eviatar} \snm{Nevo}}
-\author[addressref={aff7,aff8},email={cjkim@med.wayne.edu}]{\inits{CJ}\fnm{Chong J} \snm{Kim}}
-\author[addressref={aff9},email={gunter.wagner@yale.edu}]{\inits{GP}\fnm{Günter P.} \snm{Wagner}}
-\author[addressref={aff7},email={prbchiefstaff@med.wayne.edu}]{\inits{R}\fnm{Roberto} \snm{Romero}}
-\author[addressref={aff1,aff6,aff10},corref={aff1},email={wildmand@illinois.edu}]{\inits{DE}\fnm{Derek E.} \snm{Wildman}}
-\address[id=aff1]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[id=aff2]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
-\address[id=aff3]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[id=aff4]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
-\address[id=aff5]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
-\address[id=aff6]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
-\address[id=aff7]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
-\address[id=aff8]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[id=aff9]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
-\address[id=aff10]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[id=aff11]{The Institute of Evolution, University of Haifa, Haifa, Israel}
-
-\begin{artnotes}
-\end{artnotes}
-
-\begin{artnotes}
-\note[id=n1]{Equal contributor} % note, connected to author
-\end{artnotes}
-
-\end{fmbox}
-% \twocolumn
-
-\begin{abstractbox}
+\title{The core transcriptome of the mammal placenta}%
+%
+\author[affigb]{Don L. Armstrong\fnref{n1}}
+\ead{don@donarmstrong.com}
+\author[afflon]{Michael R. McGowen\fnref{n1}}
+\ead{1mmcgowen1@gmail.com}
+\author[affigb]{Amy Weckle}
+\ead{aweckle@illinois.edu}
+\author[affwayne]{Jason Caravas}
+\ead{jacaravas@gmail.com}
+\author[affmich]{Dalen Agnew}
+\ead{agnewd@dcpah.msu.edu}
+\author[affuscd]{Kurt Benirschke}
+\ead{kbenirschke@ucsd.edu}
+\author[affapetrust]{Sue Savage-Rumbaugh}
+\ead{suerumbaugh@gmail.com}
+\author[affhaifa]{Eviatar Nevo}
+\ead{nevo@research.haifa.ac.il}
+\author[affnih,affwaynepath]{Chong J Kim}
+\ead{cjkim@med.wayne.edu}
+\author[affyale]{Günter P. Wagner}
+\ead{gunter.wagner@yale.edu}
+\author[affnih]{Roberto Romero}
+\ead{prbchiefstaff@med.wayne.edu}
+\author[affigb,affapetrust,affmip]{Derek E. Wildman\corref{cor1}}
+\ead{wildmand@illinois.edu}
+\cortext[cor1]{Corresponding author}
+\address[affigb]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
+\address[afflon]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
+\address[affwayne]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
+\address[affmich]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
+\address[affuscd]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
+\address[affapetrust]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
+\address[affnih]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
+\address[affwaynepath]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
+\address[affyale]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
+\address[affmip]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
+\address[affhaifa]{The Institute of Evolution, University of Haifa, Haifa, Israel}
+
 \begin{abstract}
-% \section{Abstract}
-  \parttitle{Background} The placenta is arguably the most
-  anatomically variable organ in mammals. Understanding gene
-  expression patterns among placentas from multiple species
-  illuminates the molecular basis of placenta function as well
-  obstetrical syndromes affecting humans.
-
-  \parttitle{Results} We measured the expression profiles of
+  \textbf{Background} The placenta is arguably the most
+  anatomically variable organ in mammals.
+
+  \textbf{Results} We measured the expression profiles of
   \Sexpr{star.logs[,length(unique(file))]} term placentas of

### 2016-08-12 manuscript/manuscript.Rnw
add keywords
switch to colwidth

@@
 <<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
-opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
+opts_chunk$set(dev="cairo_pdf",out.width="\\colwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
 opts_chunk$set(cache=TRUE, autodep=TRUE)
@@
 
-\textbf{Keywords:} mammal, gene expression, evolution, RNA-seq, placenta
-
 \end{abstract}
-\end{abstractbox}
-\end{frontmatter}
+
+\begin{keyword}
+mammal \sep  gene expression \sep evolution \sep RNA-seq \sep placenta
+\end{keyword}
+
+\maketitle
 

### 2016-08-12 manuscript/manuscript.Rnw
add possible linebreak after ENS
add rodent table

@@
                                          "/Gene/Summary?g=",name_or_id,"}{\\textit{",
-                                         xtable.sanitize.text(name_or_id),"}}")]
+                                         gsub("(ENS[^0-9]+)","\\1\\\\linebreak[0]",
+                                              xtable.sanitize.text(name_or_id)),"}}")]
 mentioned.genes.table[!is.na(`Approved Name`),
@@
 ### italicize species names
-mentioned.genes.table[,species:=paste0("\\textit{",gsub("^(.)","\\U\\1",species,perl=TRUE),"}")]
+mentioned.genes.table[,species:=paste0("\\textit{",gsub("^(.)\\S+","\\U\\1.",species,perl=TRUE),"}")]
 setnames(mentioned.genes.table,c("Species","Gene Symbol",
@@
         mgt.align <- mgt.align[-4]
+        mgt.align[3] <- ">{\\raggedright \\centering}p{1.2in}"
     }
@@
 
-\begin{table}
+\begin{table}[htb]
 \footnotesize
 <<primate.genes,echo=FALSE,results="asis">>=
-output.mentioned.genes.table(subset.table=quote(grepl("Homo|Ateles|Pan",Species)),
-                             human.symbol=FALSE
+output.mentioned.genes.table(subset.table=quote(grepl("sapiens|fusciceps|paniscus",Species)),
+                             human.symbol=TRUE
                              )
@@
   genes with \ac{FPKM} $\ge 50$ is located in the supplemental material }
-\label{tab:notable_genes}
+\label{tab:notable_primate_genes}
 \end{table}
 
-\begin{table}
+\begin{table}[htb]
   \footnotesize
@@
 
-\begin{figure}
-<<psg.ctsl.evolution.tree,echo=FALSE,fig.width=16,fig.height=16,warning=FALSE,message=FALSE>>=
+\begin{figure*}
+<<psg.ctsl.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
 plot.label.size <- 3
@@
 \subsection{Rodentia Placenta Transcriptomes}
+
+\begin{table}
+  \footnotesize
+<<rodentia.genes,echo=FALSE,results="asis">>=
+output.mentioned.genes.table(subset.table=quote(grepl("musculus|galili|carmeli",Species)),
+                             human.symbol=TRUE
+                             )
+@ 
+\caption{Notable genes ordered by species and \ac{FPKM}. Gene
+  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
+  name corresponding to the human symbol. A complete table of 
+  genes with \ac{FPKM} $\ge 50$ is located in the supplemental material }
+\label{tab:notable_non_primate_genes}
+\end{table}
 

### 2016-08-12 manuscript/manuscript.Rnw
remove ateles top 50
remove human name from diff table

@@
                                                   "}")]
+        display.table[,`Human Name`:=NULL]
         # setnames(display.table,"Human Symbol","\\multicolumn{1}{p{0.5in}}{Human Symbol}")6
@@
         print(xtable(display.table,
-                     display=c("s","s","s","e","g","f"),
-                     align=c("c","p{0.5in}","p{1in}","r","r","r")
+                     display=c("s","s","e","g","f"),
+                     align=c("c","p{0.5in}","r","r","r")
                      ),
@@
 
-\begin{table}
-  \footnotesize
-<<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
-top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
-output.top.genes.table(top.spider.monkey.genes)
-setkey(top.spider.monkey.genes,"Gene symbol")
-@ 
-\caption{\tableexplanation{\smonkey}{tab:all_ateles_fusciceps_placenta}}
-\label{tab:top50_spider_monkey_placenta}
-\end{table}
+% \begin{table}
+%   \footnotesize
+% <<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
+% top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
+% output.top.genes.table(top.spider.monkey.genes)
+% setkey(top.spider.monkey.genes,"Gene symbol")
+% @ 
+% \caption{\tableexplanation{\smonkey}{tab:all_ateles_fusciceps_placenta}}
+% \label{tab:top50_spider_monkey_placenta}
+% \end{table}
 

### 2016-08-12 manuscript/manuscript.Rnw
add changes for elsarticle

@@
 %\documentclass[english,12pt]{article}
-\documentclass[doublespacing]{bmcart_dla}
+\documentclass[5p]{elsarticle_dla}
 \usepackage{fontspec}
-\setmainfont{FreeSans}
+\setmainfont{FreeSerif}
 \setsansfont{FreeSans}
 \setmonofont{FreeMono}
-\setmathrm{FreeSans}
+\setmathrm{FreeSerif}
 \setmathsf{FreeSans}
-\setmathtt{FreeSans}
+\setmathtt{FreeSerif}
 %\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
 \usepackage[letterpaper]{geometry}
-\usepackage{url}
-% \usepackage{fancyhdr}
+%\usepackage{url}
 \usepackage{graphicx}
-% \usepackage[bf]{caption}
-\usepackage{rotating}
-\usepackage{wrapfig}
+%\usepackage{rotating}
+%\usepackage{wrapfig}
 \usepackage{booktabs}
-\usepackage{float}
+%\usepackage{float}
 \usepackage{multirow}
-\usepackage{setspace}
-\usepackage{wrapfig}
-% \usepackage{txfonts}
-%\usepackage[plain]{fancyref}
+%\usepackage{setspace}
 \usepackage{acronym}
 \usepackage{array}
-\usepackage{dcolumn}
-\usepackage{adjustbox}
 \usepackage{longtable}
-\usepackage{pdflscape}
 \usepackage{amsmath}
-\usepackage{sansmath}
 \usepackage[backend=biber,natbib=true,hyperref=true,citestyle=numeric-comp,style=nature,autocite=inline]{biblatex}
 \addbibresource{references.bib}
-% \usepackage[noblocks,auth-sc]{authblk}
 \usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue]{hyperref}
-%\usepackage[hyperfigures,bookmarks,colorlinks,citecolor=black,filecolor=black,linkcolor=black,urlcolor=black]{hyperref}
 % Floats at end for journals
-\usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
-\renewcommand{\efloatseparator}{\mbox{}}
+% \usepackage[nolists,figuresfirst,nomarkers,heads]{endfloat}
+% \renewcommand{\efloatseparator}{\mbox{}}
 %\AtBeginFigures{\newpage \renewcommand{\efloatseparator}{\mbox{}}\section*{Figure Legends}}
 %\AtBeginTables{\newpage \section*{Tables}\renewcommand{\efloatseparator}{\mbox{}}}
-\renewcommand{\includegraphics}[2][]{}
+% \renewcommand{\includegraphics}[2][]{}
 \usepackage[nomargin,inline,draft]{fixme}
@@
 % this package will stretch the font slightly if necessary.
-\usepackage{microtype}
+% \usepackage{microtype}
 \usepackage{zref-xr,zref-user}
@@
 % \renewcommand\Affilfont{\itshape\small}
-\renewcommand\cite[1]{\autocite{#1}}
+%\renewcommand\cite[1]{\autocite{#1}}
+%\renewcommand\citet[1]{\textcite{#1}}
 % \pagestyle{fancy}
@@
 \begin{document}
-\raggedright
+% \raggedright
 
@@
 
-  \parttitle{Conclusions} A diverse array of genes are differentially
+  \textbf{Conclusions} A diverse array of genes are differentially
   expressed and are likely to play a role in the anatomical
@@
 \begin{table}
-  \begin{tabular}{c >{\raggedright \centering}p{1in} c c c}
+  \small
+  \begin{tabular}{>{\raggedright }p{0.7in}
+                  >{\raggedright }p{0.7in}
+                  c c 
+                  p{0.65in}}
     \toprule
-    Species                       & Common Name & Gestation (days) & \multicolumn{1}{p{1in}}{\centering Average Litter Size}  & Source \\
+    {\centering Species  }                     & {\centering Common Name }& 
+       \multicolumn{1}{p{0.43in}}{\raggedright \centering Gestation (days)} & 
+       \multicolumn{1}{p{0.3in}}{\raggedright \centering Litter Size}  & Source \\
     \midrule
@@
 \label{fig:pag2_prss16_ctsl_psg_evolution_tree}
-\end{figure}
+\end{figure*}
 
@@
 
-\section{Conclusion}
+\subsection{Conclusion}
 
@@
 
-\begin{backmatter}
+%\begin{backmatter}
   \section*{Competing interests}

### 2016-08-12 common_r_code/plot_ancestral_state_tree.R
add start of ancestral state tree

@@ -1,0 +1,57 @@
+library("data.table")
+library("phytools")
+
+plot.ancestral.state.tree <- function(gene_selector,genes.to.tree,combined.fpkm,trees,small.distance=0.0001) {
+    capfirst <- function(s, strict = FALSE) {
+        paste(toupper(substring(s, 1, 1)),
+        {s <- substring(s, 2); if(strict) tolower(s) else s},
+        sep = "")
+    }
+    tree.id <-
+        genes.to.tree[combined.fpkm[eval(gene_selector),gene_id],
+                      unique(tree)]
+    if (length(tree.id)!=1)
+        stop(paste0("Number of possible trees is not equal to 1: ",length(tree.id)))
+    tree.data <- trees[[tree.id]]$tree.dat
+    tree.data <- gsub(":0([\\),])",paste0(":",small.distance,"\\1"),tree.data)
+    gene.tree <- read.tree(text=tree.data)
+    gene.tree$tip.label <- gsub("^(ENS.+)\\1$","\\1",gene.tree$tip.label)
+    
+    ## this is the table which corresponds to the gene tree
+    gene.tree.table <- trees[[tree.id]]$table
+    gene.tree.table[,species:=capfirst(gsub("_"," ",species))]
+    gene.tree.table[,short.species:=gsub("^(.)[^ ]* +([^ ]{2})[^ ]*","\\1.\\2.",species)]
+    gene.tree.table[,symbol_or_id:=ifelse(symbol=="NULL",gene_id,paste0(symbol," ",short.species))]
+    ## use symbol gene_id for duplicates
+    gene.tree.table[duplicated(symbol_or_id)|duplicated(symbol_or_id,fromLast=TRUE),
+                    symbol_or_id:=paste0(symbol," ",gene_id)]
+    setkey(gene.tree.table,"protein_id")
+    
+    ## convert the protein id into gene_id
+    gene.tree$tip.label <-
+        gene.tree.table[gene.tree$tip.label,gene_id]
+
+    ## pull expression values for this tree
+    gene.tree.expression <-
+        combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
+    
+    gene.tree.subset <-
+        drop.tip(gene.tree,
+                 gene.tree$tip.label[!(gene.tree$tip.label %in%
+                                       gene.tree.expression[,gene_id])])
+    
+    gene.tree.expression <-
+        gene.tree.expression[gene_id %in% gene.tree.subset$tip.label,]
[…]

### 2016-08-12 ancestral_states/ancestral_state.R
use the mean per-species fpkm

@@ -3,7 +3,7 @@ library("data.table")
 library("phytools")
 
 ### arguments for debugging
-args <- c("combined_fpkm_per_sample","gene_trees","CYP19A1","combined_ancestral_states")
+args <- c("combined_fpkm","gene_trees","CYP19A1","combined_ancestral_states")
 
 args <- commandArgs(trailingOnly=TRUE)
 
@@ -26,6 +26,9 @@ calculate.ancestral.state <- function(gene,small.distance=0.001) {
     tree.id <-
         genes.to.tree[combined.fpkm[gene_short_name==gene & species=="homo sapiens",gene_id],
                       unique(tree)]
+    if (is.null(tree.id))
+        stop(paste0("Unable to find tree for",gene))
+    message(paste("gene is",gene,"tree id is",tree.id))
     tree.data <- trees[[tree.id]]$tree.dat
     tree.data <- gsub(":0([\\),])",paste0(":",small.distance,"\\1"),tree.data)
     gene.tree <- read.tree(text=tree.data)

### 2016-08-16 manuscript/manuscript.Rnw
load phytools; it's columnwidth

@@
 <<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
-opts_chunk$set(dev="cairo_pdf",out.width="\\colwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
+opts_chunk$set(dev="cairo_pdf",out.width="\\columnwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
 opts_chunk$set(cache=TRUE, autodep=TRUE)
@@
 library("tidyr")
-
+library("phytools")
 @

### 2016-08-16 manuscript/manuscript.Rnw
plot ancestral trees

@@
 
+<<load_plot_ancestral_trees,message=FALSE,echo=FALSE,results="hide",cache.extra=file.info(paste0("../common_r_code/",c("plot_ancestral_state_tree.R")))[,"mtime"]>>=
+for (a in paste0("../common_r_code/",c("plot_ancestral_state_tree.R"))) {
+    source(a)
+}
+@
+
 <<plot.defaults,echo=FALSE,results="hide">>=
-grad.def <- list(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = NA)
+#grad.def <- list(low = "#132B43", high = "#56B1F7", space = "Lab", na.value = NA)
+grad.def <- list(low = "gray90", high = "blue", space = "Lab", na.value = NA)
 
@@
 
-\begin{figure*}
-<<psg.ctsl.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-plot.label.size <- 3
-axis.text.size <- 3.5
-
+\begin{figure*}[htb]
+<<psg.ctsl.evolution.tree,echo=FALSE,out.width="\\textwidth",fig.width=8,fig.height=8>>=
 pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2),clip="off"))
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PSG1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"),
-                 axis.text.size=axis.text.size,
-                 fontsize=plot.label.size,
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CTSL" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(gene.tree,gene.tree.table,gene.tree.expression,min.fpkm=1,
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PSG1" & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                subtree=c("ENSG00000221826","ENSG00000124469"),
+                                trees=trees
+                                )+
+      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.2)+
+      theme(legend.position="none"),
+      vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"))
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CTSL" & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
                                 subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"),
-                 fontsize=4,
-                 width=3,
-                 axis.text.size=axis.text.size,
-                 offset.ratio=1
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRSS16" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"),
-                 fontsize=4,
-                 width=3,
-                 axis.text.size=axis.text.size,
-                 offset.ratio=1.2
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PAG2" & species=="bos taurus",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,min.fpkm=1,
-                 subtree=c("ENSSSCG00000006792",
+                                trees=trees)+
+      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,1)+
+      theme(legend.position="none"),
+      vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"))
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PRSS16" &
+                                                    species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                subtree=c(   "ENSG00000112812",
+                                          "ENSLAFG00000032211"),
+                                combined.fpkm=combined.fpkm,
+                                trees=trees)+
+      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,1.2)+
+      theme(legend.position="none"),
+      vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"))
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PAG2" &
+                                                    species=="bos taurus"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                subtree=c("ENSSSCG00000013097",
                                           "ENSBTAG00000046085"),
-                 vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"),
-                 axis.text.size=axis.text.size,
-                 fontsize=plot.label.size,
-                 offset.ratio=0.6
-                 )
-
+                                trees=trees) +
+      geom_tiplab(size=2,color="black",fontface="italic") +
+      xlim(0,0.8),
+      vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"))
 popViewport(1)
+
 grid.text(c("A","B","C","D"),
@@
           just=c("left","top"),
-          gp=gpar(fontsize="48")
+          gp=gpar(fontsize="24")
           )
@@
 
-\begin{figure}
+\begin{figure}[htb]
 <<hormone.genes.plot,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
@@
 
-\begin{figure}
-<<chorionic.somatomammotropin.alpha.evolution.tree,echo=FALSE,fig.width=12,fig.height=12,warning=FALSE,message=FALSE>>=
-
-plot.label.size <- 4.5
+\begin{figure}[htb]
+<<chorionic.somatomammotropin.alpha.evolution.tree,echo=FALSE,fig.width=4,fig.height=4,warning=FALSE,message=FALSE>>=
+
 pushViewport(viewport(layout=grid.layout(nrow=1,ncol=2,widths=c(2,3))))
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CGA" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=1,layout.pos.row=1),
-                 offset.ratio=0.7,
-                 fontsize=plot.label.size,
-                 axis.text.size=3.5
-                 )
-
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CGB1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 vp=viewport(layout.pos.col=2,layout.pos.row=1),
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CGA" &
+                                                    species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.8)+
+      theme(legend.position="none"),
+      vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"))
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CGB1" &
+                                                    species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
                                 subtree=c("ENSMODG00000013768","ENSG00000104818"),
-                 offset.ratio=1,
-                 fontsize=plot.label.size,
-                 axis.text.size=3.5
-                 )
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,1),
+      vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"))
 popViewport(1)

### 2016-08-16 common_r_code/plot_ancestral_state_tree.R
remove subtrees

@@ -1,5 +1,7 @@
 library("data.table")
 library("phytools")
+library("magrittr")
+library("geiger")
 ##' Plot the Ancestral State Tree
 ##'
 ##' See examples in manuscript.Rnw for more details
@@ -12,7 +14,7 @@ library("phytools")
 ##' @param ... Additional options passed to contMap
 ##' @return whatever contMap() returns
 ##' @author Don Armstrong
-plot.ancestral.state.tree <- function(gene_selector,genes.to.tree,combined.fpkm,trees,small.distance=0.0001,...) {
+plot.ancestral.state.tree <- function(gene_selector,genes.to.tree,combined.fpkm,trees,subtree=NULL,small.distance=0.0001) {
     capfirst <- function(s, strict = FALSE) {
         paste(toupper(substring(s, 1, 1)),
         {s <- substring(s, 2); if(strict) tolower(s) else s},
@@ -45,13 +47,31 @@ plot.ancestral.state.tree <- function(gene_selector,genes.to.tree,combined.fpkm,
     ## pull expression values for this tree
     gene.tree.expression <-
         combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-    
-    gene.tree.expression[,duplicate_gene_id:=gene_id]
 
     gene.tree.subset <-
         drop.tip(gene.tree,
                  gene.tree$tip.label[!(gene.tree$tip.label %in%
                                        gene.tree.expression[,gene_id])])
+    if (!is.null(subtree)) {
+        gene.tree.mrca <- mrca(gene.tree.subset)
+        top.node <- gene.tree.mrca[subtree[1],subtree[2]]
+        gene.tree.subset <-
+            drop.tip(gene.tree.subset,
+                     gene.tree.subset$tip.label[!(gene.tree.subset$tip.label %in%
+                                                  tips(gene.tree.subset,
+                                                       top.node))])
+    }
+    gene.tree.expression <-
+        gene.tree.expression[gene_id %in% gene.tree.subset$tip.label,]
+
+    gene.tree.expression[,duplicate_gene_id:=gene_id]
+
+    gene.tree.subset <-
+        drop.tip(gene.tree.subset,
+                 gene.tree.subset$tip.label[!(gene.tree.subset$tip.label %in%
+                                              gene.tree.expression[,gene_id])])
+
+
     
     for (duplicate.gene_id in unique(gene.tree.expression[duplicated(gene_id),gene_id])) {
[…]

### 2016-08-16 common_r_code/plot_ancestral_state_tree.R
fix extraneous 6

@@ -134,7 +134,7 @@ plot.ancestral.state.tree <- function(gene_selector,genes.to.tree,combined.fpkm,
     return(ggtree(f.tr,aes(color=log10(fpkm+1)),size=1)+
            # geom_tiplab(color="black")+
            scale_color_gradient(low="grey90",high="blue",
-                                limits=c(0,4),6
+                                limits=c(0,5),
                                 guide=guide_colorbar(title=expression(log[10](FPKM))))+
            theme(legend.position="bottom"))
 }

### 2016-08-16 manuscript/supplemental_material.Rnw
increase headheight
switch to plot ancestral state trees

@@
 \setmonofont{FreeMono}
-\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
+\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=15pt]{geometry}
 \usepackage{url}
@@
 
-<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
-for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R"))) {
+<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R","plot_ancestral_state_tree.R")))>>=
+for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R",
+                                       "utility_functions.R",
+                                       "plot_ancestral_state_tree.R"))) {
     source(a)
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CRH" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
-                 min.fpkm=1,
-                 axis.text.size=5,
-                 axis.text.angle=0,
-                 fontsize=6.5
-                 )
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CRH" &
+                                                    species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.37))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="ADAM12" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,
-                 gene.tree.expression,
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="ADAM12" &
+                                                    species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
                                 subtree=c("ENSMODG00000010538","ENSSSCG00000010746"),
-                 min.fpkm=0.1,
-                 axis.text.size=4,
-                 fontsize=5
-                 )
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.25))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="KISS1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
-                 min.fpkm=0.1,
-                 axis.text.size=5,
-                 axis.text.angle=0,
-                 fontsize=5
-                 )
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="KISS1"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.7))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PAPPA" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PAPPA"
+                                                    & species=="homo sapiens"),
                                 subtree=c("ENSSSCG00000005498","ENSG00000182752"),
-                 min.fpkm=0.1,
-                 axis.text.size=5,
-                 axis.text.angle=0,
-                 fontsize=5
-                 )
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.15))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="TGFB1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="TGFB1"
+                                                    & species=="homo sapiens"),
                                 subtree=c("ENSMODG00000008630","ENSSSCG00000026296"),
-                 gene.tree.table,gene.tree.expression,
-                 min.fpkm=1,
-                 axis.text.size=3.5,
-                 fontsize=5
-                 )
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.3))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="GPC3" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="GPC3"
+                                                    & species=="homo sapiens"),
                                 subtree=c("ENSCAFG00000018864","ENSLAFG00000025706"),
-                 gene.tree.expression,
-                 min.fpkm=1,
-                 axis.text.size=3.5,
-                 fontsize=6
-                 )
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.13))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="FSTL1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,
-                 gene.tree.expression,
-                 min.fpkm=1,
-                 axis.text.size=3.5,
-                 fontsize=5
-                 )
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="FSTL1"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=5,color="black",fontface="italic")+xlim(0,3))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="SERPINE2" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,
-                 gene.tree.expression,
-                 min.fpkm=1,
-                 axis.text.size=3.5,
-                 fontsize=3)
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="SERPINE2"
+                                                    & species=="homo sapiens"),
+                                subtree=c("ENSBTAG00000008717",
+                                          "ENSMODG00000004913"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=4,color="black",fontface="italic")+xlim(0,0.8))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_id=="ENSCAFG00000010638" & species=="canis familiaris",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
-                 axis.text.size=3.5,
-                 fontsize=5)
+print(plot.ancestral.state.tree(gene_selector=quote(gene_id=="ENSCAFG00000010638" &
+                                                    species=="canis familiaris"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,1.2))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CYP19A1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
-                 axis.text.size=3.5,
-                 fontsize=4.5)
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CYP19A1"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.37))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="HSD3B1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="HSD3B1"
+                                                    & species=="homo sapiens"),
                                 subtree=c("ENSMODG00000005385","ENSLAFG00000027896"),
-                 axis.text.size=3.5,
-                 fontsize=4.5
-                 )
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.6))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="CSH1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
-                 axis.text.size=3.5,
-                 fontsize=4)
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CSH1"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.9))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRL" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
-                 axis.text.size=3.5,
-                 fontsize=3.5)
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PRL"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=3,color="black",fontface="italic")+xlim(0,1))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="RLN2" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
-                 axis.text.size=3.5,
-                 fontsize=3.5)
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="RLN2"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.8))
 @ 
@@
 
-tree.id <-
-    genes.to.tree["ENSDNOG00000033320",tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
+print(plot.ancestral.state.tree(gene_selector=quote(gene_id=="ENSDNOG00000033320"),
                                 subtree=c("ENSDNOG00000033320","ENSMODG00000017988"),
-                 axis.text.size=3.5,
-                 fontsize=4)
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=5,color="black",fontface="italic")+xlim(0,1.3))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="PRLR" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PRLR"
+                                                    & species=="homo sapiens"),
                                 subtree=c("ENSG00000113494","ENSMODG00000020376"),
-                 axis.text.size=3.5,
-                 fontsize=4)
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.3))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="HOXA11" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="HOXA11"
+                                                    & species=="homo sapiens"),
                                 subtree=c("ENSBTAG00000014738","ENSECAG00000013229"),
-                 axis.text.size=3.5,
-                 fontsize=4
-                 )
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.4))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="HBE1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),
-                 gene.tree.table,gene.tree.expression,
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="HBE1"
+                                                    & species=="homo sapiens"),
                                 subtree=c("ENSBTAG00000037815",
                                           "ENSMUSG00000078621"),
-                 axis.text.size=3.5,
-                 fontsize=4)
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=4,color="black",fontface="italic")+xlim(0,0.6))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="NPPB" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 axis.text.size=3.5,
-                 fontsize=6
-                 )
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="NPPB"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.4))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="NPPC" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 axis.text.size=3.5,
-                 fontsize=3
-                 )
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="NPPC"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.32))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="LGALS1" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 axis.text.size=3.5,
-                 fontsize=3.5
-                 )
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="LGALS1"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,1))
 @ 
@@
 
-tree.id <-
-    genes.to.tree[combined.fpkm[gene_short_name=="LGALS13" & species=="homo sapiens",gene_id],tree]
-gene.tree.table <- trees[[tree.id]]$table
-gene.tree.expression <- combined.fpkm[gene_id %in% gene.tree.table[,gene_id]]
-gene.tree <- read.tree(text=trees[[tree.id]]$tree.dat)
-plot.tree.matrix(read.tree(text=trees[[tree.id]]$tree.dat),gene.tree.table,gene.tree.expression,
-                 axis.text.size=3.5,
-                 fontsize=4.5
-                 )
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="LGALS13"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,1.5))
 @ 
@@
                       max(num_genes),`Minimum Expression in`],
-                  show_guide=FALSE,
+                  show.legend=FALSE,
                   hjust=0

### 2016-09-01 common_r_code/plot_ancestral_state_tree.R
allow specifying line width

@@ -14,7 +14,7 @@ library("geiger")
 ##' @param ... Additional options passed to contMap
 ##' @return whatever contMap() returns
 ##' @author Don Armstrong
-plot.ancestral.state.tree <- function(gene_selector,genes.to.tree,combined.fpkm,trees,subtree=NULL,small.distance=0.0001) {
+plot.ancestral.state.tree <- function(gene_selector,genes.to.tree,combined.fpkm,trees,subtree=NULL,small.distance=0.0001,tree.line.width=1) {
     capfirst <- function(s, strict = FALSE) {
         paste(toupper(substring(s, 1, 1)),
         {s <- substring(s, 2); if(strict) tolower(s) else s},
@@ -131,7 +131,7 @@ plot.ancestral.state.tree <- function(gene_selector,genes.to.tree,combined.fpkm,
         as.numeric(gene.tree.expression.vector[f.tr$label[f.tr$isTip]])
     f.tr$fpkm[!f.tr$isTip] <-
         theanc[as.character(f.tr$node[!f.tr$isTip])]
-    return(ggtree(f.tr,aes(color=log10(fpkm+1)),size=1)+
+    return(ggtree(f.tr,aes(color=log10(fpkm+1)),size=tree.line.width)+
            # geom_tiplab(color="black")+
            scale_color_gradient(low="grey90",high="blue",
                                 limits=c(0,5),

### 2016-09-01 presentation/placenta_viviparous_presentation.Rnw
add core placenta shape plot

--- 2016-03-09 presentation/placenta_viviparous_presentation.Rnw
@@
 \documentclass[ignorenonframetext]{beamer}
-\usepackage{morewrites}
 \usepackage{fontspec}
@@
 \usepackage{mathtools,cancel}
-\usepackage{minted}
-\usepackage{tcolorbox}
-\usepackage{etoolbox}
-\BeforeBeginEnvironment{minted}{\begin{tcolorbox}}%
-\AfterEndEnvironment{minted}{\end{tcolorbox}}%
 
@@
 
-<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R")))>>=
-for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R"))) {
+<<source.to.latex,echo=FALSE,message=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R","utility_functions.R","plot_ancestral_state_tree.R")))>>=
+for (a in paste0("../common_r_code/",c("to_latex.R","plot_tree_matrix.R",
+                                       "utility_functions.R",
+                                       "plot_ancestral_state_tree.R"
+                                       ))) {
     source(a)
@@
 \begin{frame}{Species Studied}
-% \begin{figure}
+\begin{figure}
 <<species.tree,fig.width=8,fig.height=6,echo=FALSE>>=
@@
 
-\subsection{Using Make}
-
-\begin{frame}{Makefile rules}
-\centering
-\includegraphics[keepaspectratio,height=0.8\textheight]{figures/default}
-\end{frame}
-
-\begin{frame}[fragile]{Common Options}
-\tiny
-\inputminted[firstline=8, lastline=36]{make}{../fastq/common_makefile}
-\end{frame}
-
-\begin{frame}[fragile]{FASTQC}
-\inputminted[firstline=78, lastline=83]{make}{../fastq/common_makefile}
-\end{frame}
-
-\begin{frame}[fragile]{Trinity}
-\tiny
-\inputminted[firstline=211, lastline=222]{make}{../fastq/common_makefile}
-\end{frame}
-
-\begin{frame}[fragile]{Diamond}
-\tiny
-\inputminted[firstline=249, lastline=268]{make}{../fastq/common_makefile}
-\end{frame}
-
-\begin{frame}[fragile]{STAR Indexes}
-\tiny
-\inputminted[firstline=86, lastline=105]{make}{../fastq/common_makefile}
-\end{frame}
-
-\begin{frame}[fragile]{STAR}
-\tiny
-\inputminted[firstline=106, lastline=118]{make}{../fastq/common_makefile}
-\end{frame}
-
-\begin{frame}[fragile]{Cufflinks}
-\tiny
-\inputminted[firstline=43, lastline=60]{make}{../fastq/common_makefile}
-\end{frame}
-
 \subsection{Additional Methods}
@@
 
+\begin{frame}{Core Placenta Transcriptome Shape}
+\begin{figure}
+<<core.placenta.transcriptome.distribution,fig.width=6,fig.height=4,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
+num.species <- star.logs.wide[,length(unique(species))]
+calculate.distribution.shape <- 
+    function(min.vector) {
+        min.vector <- sort(min.vector,decreasing=TRUE)
+        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
+        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
+        return(temp)
+    }
+
+cptg.nh <- core.placenta.transcriptome.genes.shape[!(human_name %in% housekeeping.genes.superset[,gene_short_name])]
+
+cpt.distribution.shape <- 
+    lapply(c("all_expression",
+             "all_but_one_expression",
+             "all_but_two_expression",
+             "all_but_three_expression",
+             "all_but_four_expression"),
+           function(x){
+               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
+           })
+fix.expression.names <- function(x){
+    x <- gsub("^all_expression$",
+              paste0(num.species," of ",num.species," species"),
+              x
+              )
+    nc.levels <- paste0(num.species," of ",num.species," species")
+    ## number coutner
+    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
+    for (a in names(nc)) {
+        x <- gsub(paste0("^all_but_",a,"_expression$"),
+              paste0(num.species-nc[a]," of ",num.species," species"),
+              x
+              )
+        nc.levels <-
+            c(nc.levels,
+              paste0(num.species-nc[a]," of ",num.species," species"))
+    }
+    return(factor(x,levels=nc.levels))
+}
+
+cpt.distribution.shape <- 
+    rbindlist(cpt.distribution.shape)
+
+cpt.distribution.shape <- 
+    cpt.distribution.shape[,`Minimum Expression in`:=
+                                fix.expression.names(`Minimum Expression in`)]
+
+setnames(cpt.distribution.shape,
+         "FPKM",
+         "Expression (FPKM)")
+
+print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],
+             aes(x=`Expression (FPKM)`,color=`Minimum Expression in`,y=num_genes))
+      + geom_line(lwd=1)
+      + xlab("Minimum Expression cutoff (FPKM)")
+      + ylab("Genes with N-th Minimum Expression")
+      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                      labels = trans_format("log10", math_format(10^.x)))
+      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                      labels = trans_format("log10", math_format(10^.x)))
+      + geom_vline(xintercept=10)
+      + geom_segment(mapping=aes(y=V1,
+                         yend=V1,
+                         xend=10,
+                         x=1000,
+                         color=`Minimum Expression in`),
+                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
+                                              max(num_genes),`Minimum Expression in`],
+                     arrow=arrow(length=unit(0.5,"cm"))
+                   )
+      + geom_text(mapping=aes(x=1000,
+                      y=V1,
+                      label=paste0(V1," genes"),override=TRUE),
+                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
+                      max(num_genes),`Minimum Expression in`],
+                  show.legend=FALSE,
+                  hjust=0
+                  )
+      + guides(color=guide_legend(title="Minimum\nExpression in"))
+      )
+@ 
+\end{figure}
+\end{frame}
+
 \begin{frame}{Core placenta transcriptome}

### 2016-09-01 presentation/placenta_viviparous_presentation.Rnw
add egfr gene tree

@@
 
+\begin{frame}{EGFR}
+  \begin{figure}
+<<egfr_expression,echo=FALSE,fig.width=5,fig.height=5>>=
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="EGFR" &
+                                                    species=="homo sapiens"),
+                                subtree=c("ENSG00000146648","ENSMODG00000009033"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees,
+                                tree.line.width=2
+                                )+
+      geom_tiplab(size=5,color="black",fontface="italic")+xlim(0,0.4))
+
+@ 
+  \end{figure}
+\end{frame}
+
 \begin{frame}{Hormone Expression}

### 2016-09-01 manuscript/supplemental_material.Rnw
add placenta classification table

@@
 
+\clearpage
+
 \section{Species}
+
+\subsection{Placental Classification}
+
+\begin{table}[hb]
+<<placental_classification,echo=FALSE,results="asis">>=
+placental.classification <- fread("../data/placental_classification.txt")
+print(xtable(placental.classification,
+             display=c("s","s","s","s","s"),
+             align=c("c","c","c","c","c")),
+      include.rownames=FALSE,
+      floating=FALSE,
+      booktabs=TRUE
+      )
+
+@ 
+\caption{Classification of Placenta shapes}
+\label{tab:placental_classification}
+\end{table}
 
@@
   \small
-  \begin{tabular}{c >{\raggedright}p{1.5in} c >{\raggedright\arraybackslash}p{2in} c}
+  \begin{tabular}{c >{\raggedright}p{1.2in} c >{\raggedright\arraybackslash}p{1.8in} c}
     \toprule

### 2016-09-01 manuscript/supplemental_material.Rnw
add short captions

@@
 @ 
-\caption{RNAseq read and alignment statistics for all species. A)
+\caption[RNAseq statistics]{RNAseq read and alignment statistics for all species. A)
   Boxplot of percentages of unmapped reads in each species; 0\% of
@@
              align=  c("c","c","c","c","c","c"),
-             caption="GO terms with significant enrichment ($\\text{FDR}\\le0.05$)
+             caption=c("GO terms with significant enrichment ($\\text{FDR}\\le0.05$)
   in 12 or more of the species analyzed from genes in the top 1\\% of
@@
   listed. $max(\\text{FDR}\\le0.05)$ is the maximum FDR value in all
-  species with an FDR of less than $0.05$.",
+  species with an FDR of less than $0.05$.","GO terms in 12 or more species"),
   label="tab:go_terms_all_species"
@@
                             ][order(FDR)],
-                 caption=this.caption,
+                 caption=c(this.caption,paste0("GO terms from \\textit{",
+                                               capfirst(this.species),"} top 1\\%")),
                  label=paste0("tab:go_terms_enriched_top_one_",gsub(" ","_",this.species)),
@@
                             ][order(FDR)],
-                 caption=this.caption,
+                 caption=c(this.caption,paste0("GO terms from \\textit{",
+                                               capfirst(this.species),"} top 100")),
                  label=this.label,

### 2016-09-08 manuscript/extended_discussion.Rnw
copy manuscript to extended discussion

--- 2016-08-16 manuscript/manuscript.Rnw
@@
 
-\title{The core transcriptome of the mammal placenta}%
+\title{The evolution of the transcriptome of mammalian placentas}%
 %
@@
 respective reference genome (\cref{tab:ref_genome_aligned_species}).
-The top 50 expressed genes in each genome analyzed are shown in
-\cref{tab:top50_human_placenta,tab:top50_bonobo_placenta,tab:top50_spider_monkey_placenta,tab:top50_dog_placenta,tab:top50_cow_placenta,tab:top50_elephant_placenta,tab:top50_opossum_placenta,tab:top50_mouse_placenta,tab:top50_n_galili_placenta,tab:top50_s_carmeli_placenta,tab:top50_pig_placenta,tab:top50_horse_placenta,tab:top50_sheep_placenta}
-and all genes in Supplemental \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.
-
-% The percentage of bases that mapped to Ensembl annotated mRNA ranged
-% from ~35\% in elephant to ~85\% in human, with percentages of bases
-% mapping to annotated protein-coding segments of genes ranging from
-% ~34\% (elephant) to ~54\% (human), untranslated regions (UTRs) ranging
-% from ~0.6\% (elephant) to ~31\% (human), and known introns ranging
-% from ~5.5\% (dog) to 11\% (bonobo). Median percent genomic coverage
-% ranged from ~43\% in elephant to 80\% in bonobo. The number of
-% annotated transcripts to which reads mapped in each respective genomes
-% ranged from 4119 (bonobo) to 6506 (human) for transcripts with \ac{FPKM}>10
-% and 11689 (dog) to 30733 (human) for transcripts with \ac{FPKM}>1. The
-% greater degree of mapping to annotated protein-coding sequences in
-% humans corresponds to the greater number of Ensembl annotated
-% sequences in the human genome compared to the other five reference
-% genomes (Flicek, 2012).
+The expression of all genes with \ac{FPKM} $\ge 50$ is shown in Supplemental
+\cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.
 
@@
 mentioned.genes <- load_cache(label="save_mentioned_genes",object="saved_mentioned_genes",list())
-gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE) {
+gene.expression <- function(species.name,gene,complete=FALSE,units=TRUE,mention.only=FALSE) {
     species.name <- gsub("_"," ",species.name)
@@
              "gene"=gene)
+    if (mention.only) {
+        return("")
+    }
     fpkm <- combined.fpkm[species==species.name &
@@
         if (any(grepl("^ENS.+G\\d+$",gene))) {
-            fpkm <- paste0("\\text{",gene,"} = ",
+            fpkm <- paste0(gsub("(ENS)(.+)(G)0*(\\d+)","\\1\\\\linebreak[0]\\2\\\\linebreak[0]\\3...\\4",gene)," = ",
                            fpkm)
         } else {
-            fpkm <- paste0("\\text{\\textit{",gene,"}} = ",
+            fpkm <- paste0("\\textit{",gene,"} = ",
                            fpkm)
@@
     }
-    if (units || complete) {
-        fpkm <- paste0("$",fpkm,"$")
-    }
+    ## if (units || complete) {
+    ##     fpkm <- paste0("$",fpkm,"$")
+    ## }
     return(fpkm)
@@
 
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \human are shown in
-\cref{tab:top50_human_placenta} with all genes with \ac{FPKM} $\ge 50$
-in Supplemental \cref{tab:all_homo_sapiens_placenta}. The gene with by far
-the highest \ac{FPKM} value
+The gene with by far the highest \ac{FPKM} value
 %
@@
 is also expressed to a high degree in the \human placenta
-(\cref{tab:top50_human_placenta}).
+(Supplemental \cref{tab:all_homo_sapiens_placenta}).
 % 
@@
 which generates the alpha subnit of multiple human chorionic
-gonadotropin hormones (\cref{tab:top50_human_placenta}). The genes
+gonadotropin hormones (Supplemental \cref{tab:all_homo_sapiens_placenta}). The genes
 producing the beta subunits of the chorionic gonadotropins
@@
 glucose transport (\textit{SLC2A1}) among others
-(\cref{tab:top50_human_placenta}, Supplemental
+(Supplemental \cref{tab:all_homo_sapiens_placenta}, Supplemental
 \cref{tab:all_homo_sapiens_placenta}).
@@
                                 trees=trees)+
-      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,1)+
+      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.8)+
       theme(legend.position="none"),
@@
                                 trees=trees)+
-      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,1.2)+
-      theme(legend.position="none"),
+      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,1.2),
       vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"))
@@
       geom_tiplab(size=2,color="black",fontface="italic") +
-      xlim(0,0.8),
+      xlim(0,0.8)+
+      theme(legend.position="none"),
       vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"))
@@
 The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \bonobo are shown in \cref{tab:top50_bonobo_placenta}
+placenta of \bonobo are shown in Supplemental \cref{tab:all_pan_paniscus_placenta}
 with a list of all genes in Supplemental
@@
 placenta differs somewhat from that of \human, as evidenced by the
-makeup of the top 50 genes (\cref{tab:top50_bonobo_placenta}). The highest expression
+makeup of the top 50 genes (Supplemental \cref{tab:all_pan_paniscus_placenta}). The highest expression
  in \bonobo is %
@@
 \cref{fig:progesterone_evolution_tree} and
-\cref{tab:top50_human_placenta}) although the expression of
+Supplemental \cref{tab:all_homo_sapiens_placenta}) although the expression of
 \textit{HSD3B1} is fairly similar to that of \smonkey %
@@
 (\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
-lower levels than \human as well (\cref{tab:top50_human_placenta}).
+lower levels than \human as well (Supplemental \cref{tab:all_homo_sapiens_placenta}).
 %
@@
 \smonkey, the top 50 non-housekeeping genes
-(\cref{tab:top50_spider_monkey_placenta}) include a possible regulator
+(Supplemental \cref{tab:all_spider_monkey_placenta}) include a possible regulator
 of androgen receptor signaling
@@
 a few HLA genes (\textit{HLA-A}, \textit{HLA-F},
-\cref{tab:top50_spider_monkey_placenta}), and multiple components of
+Supplemental \cref{tab:all_spider_monkey_placenta}), and multiple components of
 annexin complexes
@@
 in the term placenta of \mouse are shown in
-\cref{tab:top50_mouse_placenta} with a list of all genes with
+Supplemental \cref{tab:all_mus_musculus_placenta} with a list of all genes with
 $\text{FPKM} \ge 50$ in Supplemental
@@
 \Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
-\cref{tab:top50_mouse_placenta,fig:pag2_prss16_ctsl_psg_evolution_tree}).
+Supplemental \cref{tab:all_mus_musculus_placenta,fig:pag2_prss16_ctsl_psg_evolution_tree}).
 Finally, \textit{Peg10}, an imprinted gene required for proper
@@
 in the term placenta of \galili are shown in
-\cref{tab:top50_n_galili_placenta} with a list of all highly expressed
+Supplemental \cref{tab:all_n_galili_placenta} with a list of all highly expressed
 genes in Supplemental \cref{tab:all_spalax_galili_placenta}.
@@
 in the term placenta of \carmeli are shown in
-\cref{tab:top50_s_carmeli_placenta} with a list of all highly
+Supplemental \cref{tab:all_s_carmeli_placenta} with a list of all highly
 expressed genes in Supplemental
@@
 in the term placenta of \dog are shown in
-\cref{tab:top50_dog_placenta} with a list of all genes in Supplemental
+Supplemental \cref{tab:all_canis_familiaris_placenta} with a list of all genes in Supplemental
 \cref{tab:all_canis_familiaris_placenta}. The gene with the highest
@@
 is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
-among the top 50 (\cref{tab:top50_dog_placenta}) include genes
+among the top 50 (Supplemental \cref{tab:all_canis_familiaris_placenta}) include genes
 involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
@@
 \Sexpr{gene.expression("bos taurus","SSLP1",complete=TRUE)}, %
-\cref{tab:top50_cow_placenta}) and \sheep
+Supplemental \cref{tab:all_bos_taurus_placenta}) and \sheep
 (\Sexpr{gene.expression("ovis aries","SOLD1",complete=TRUE)},
-\cref{tab:top50_sheep_placenta}) but is not present in primates
+Supplemental \cref{tab:all_sheep_placenta}) but is not present in primates
 \cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}
@@
 The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the \cow are shown in \cref{tab:top50_cow_placenta} with a
+placenta of the \cow are shown in Supplemental \cref{tab:all_bos_taurus_placenta} with a
 list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
@@
 show especially high expression in \cow placenta
-(\cref{tab:top50_cow_placenta}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D, and
+(Supplemental \cref{tab:all_bos_taurus_placenta}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D, and
 Supplemental \cref{tab:all_bos_taurus_placenta}) including
@@
 in the term placenta of \textit{Sus scrofa} are shown in
-\cref{tab:top50_pig_placenta} with a list of all highly expressed
+Supplemental \cref{tab:all_pig_placenta} with a list of all highly expressed
 genes in Supplemental \cref{tab:all_sus_scrofa_placenta}. The top two
@@
 in the term placenta of \horse are shown in
-\cref{tab:top50_horse_placenta} with a list of all highly expressed
+Supplemental \cref{tab:all_equus_caballus_placenta} with a list of all highly expressed
 genes in Supplemental \cref{tab:all_equus_caballus_placenta}. The gene
@@
 in the term placenta of \sheep are shown in
-\cref{tab:top50_sheep_placenta} with a list of all highly expressed
+Supplemental \cref{tab:all_sheep_placenta} with a list of all highly expressed
 genes in Supplemental \cref{tab:all_ovis_aries_placenta}. The complete
@@
 The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \elephant are shown in \cref{tab:top50_elephant_placenta}
+placenta of \elephant are shown in Supplemental \cref{tab:all_loxodonta_africana_placenta}
 with a list of all genes in
@@
 Five uncharacterized genes are expressed to a high degree in the
-\elephant placenta (\cref{tab:top50_elephant_placenta}), three of
+\elephant placenta (Supplemental \cref{tab:all_loxodonta_africana_placenta}), three of
 which
@@
 \cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
-\dog placentas (\cref{tab:top50_dog_placenta,tab:top50_cow_placenta}).
+\dog placentas (Supplemental \cref{tab:all_canis_familiaris_placenta,tab:all_bos_taurus_placenta}).
 Placental expression of the hormones relaxin (\textit{RLN}, %
@@
 The top 50 expressed genes in the term placenta of \armadillo are
-shown in \cref{tab:top50_armadillo_placenta} with a list of all genes
+shown in Supplemental \cref{tab:all_dasypus_novemcinctus_placenta} with a list of all genes
 with $\text{FPKM}\ge 50$ in Supplemental
@@
 placenta of the \opossum are shown in
-\cref{tab:top50_opossum_placenta} with a list of all genes in
+Supplemental \cref{tab:all_opossum_placenta} with a list of all genes in
 Supplemental \cref{tab:all_monodelphis_domestica_placenta}. The gene with the
@@
                                 )+
-      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.8)+
+      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.4)+
       theme(legend.position="none"),
@@
                                 )+
-      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,1),
+      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.5),
       vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"))
@@
 exist in eutherians or are not expressed in the term eutherian placenta
-(\cref{tab:top50_opossum_placenta}). The existence of embryonic forms
+(Supplemental \cref{tab:all_opossum_placenta}). The existence of embryonic forms
 of hemoglobin in fetal tissues at term, as evidenced by the expression
@@
 which have very abundant mRNAs
-(\cref{tab:top50_elephant_placenta} and \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}C). It is
+(Supplemental \cref{tab:all_loxodonta_africana_placenta} and \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}C). It is
 possible that this elephant-specific gene family may be involved in

### 2016-09-08 manuscript/common_latex.tex
we're only showing the top 40 genes in the type III anova
update the gene tree explanation

@@
 
-\newcommand{\genetreeexplanation}[2][1]{Gene tree and heatmap of
-  expression of genes related to #2. The gene tree was obtained from
-  Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes from species
-  not included in this study or having expression by FPKM less than #1
-  were excluded from the plotted tree. Heatmap cells in white are from
-  species where there was no expression of the relevant transcript (or
-  a species where that transcript does not apply). The species names
-  are abbreviated with the first letter of the genus and first two
-  letters of the species (for example, \textit{Homo sapiens} is
-  \textit{H.sa.}). Genes with duplicate gene names in the same species
-  (or no gene named defined) are listed by the Ensemble gene ID.}
+\newcommand{\genetreeexplanation}[2][1]{Gene tree showing expression
+  and ancestral expression of genes related to #2. The gene tree was
+  obtained from Ensembl~\cite{Cunningham.ea2015:Ensembl2015} and genes
+  from species not included in this study were excluded from the
+  plotted tree. The coloring of the internal nodes of the plots
+  indicates the ancestral expression as estimated by fastAnc using
+  phylotools. Genes with duplicate gene names in the same species (or
+  no gene named defined) are listed by the Ensemble gene ID with
+  leading zeros elided.}
 
@@
   \lineagetableexplanation[A nested type-III anova was used to
-  calculate p values.]{#1}{#2}}
+  calculate p values, and the top 40 genes by $\text{log}_2$
+  Fold-Change which changed significantly (FDR $\le 0.05$) are
+  shown. ]{#1}{#2}}
 

### 2016-09-08 manuscript/manuscript.Rnw
remove commented out latex code
change title
genes are now in supplemental

@@
 \usepackage[capitalize]{cleveref}
-% \newenvironment{narrow}[2]{%
-%   \begin{list}{}{%
-%       \setlength{\topsep}{0pt}%
-%       \setlength{\leftmargin}{#1}%
-%       \setlength{\rightmargin}{#2}%
-%       \setlength{\listparindent}{\parindent}%
-%       \setlength{\itemindent}{\parindent}%
-%       \setlength{\parsep}{\parskip}}%
-%   \item[]}{\end{list}}
-% \def\newblock{\hskip}
-% \newenvironment{paperquote}{%
-%   \begin{quote}%
-%      \it
-%   }%
-%   {\end{quote}}
-% \renewcommand{\textfraction}{0.15}
-% \renewcommand{\topfraction}{0.85}
-% \renewcommand{\bottomfraction}{0.65}
-% \renewcommand{\floatpagefraction}{0.60}
-%\renewcommand{\baselinestretch}{1.8}
-% \newenvironment{enumerate*}%
-%   {\begin{enumerate}%
-%     \setlength{\itemsep}{0pt}%
-%     \setlength{\parskip}{0pt}}%
-%   {\end{enumerate}}
-% \newenvironment{itemize*}%
-%   {\begin{itemize}%
-%     \setlength{\itemsep}{0pt}%
-%     \setlength{\parskip}{0pt}}%
-%   {\end{itemize}}
-% \setcounter{tocdepth}{3}
-% \oddsidemargin 0.0in 
-% \textwidth 6in
-% \raggedbottom
-% \clubpenalty = 10000
-% \widowpenalty = 10000
-% set a large emergency stretch to avoid issues with overfull boxes
-% \setlength{\emergencystretch}{20em}
-% make the afill smaller
-% \renewcommand\Affilfont{\itshape\small}
-%\renewcommand\cite[1]{\autocite{#1}}
-%\renewcommand\citet[1]{\textcite{#1}}
-% \pagestyle{fancy}
 

### 2016-09-09 manuscript/manuscript.Rnw
gut the manuscript for rewriting

@@
 \begin{abstract}
-  \textbf{Background} The placenta is arguably the most
+  \textbf{Introduction} The placenta is arguably the most
   anatomically variable organ in mammals.
+
+  \textbf{Methods}
 
@@
   \Sexpr{star.logs[,length(unique(file))]} term placentas of
-  \Sexpr{star.logs[,length(unique(species))]} species of mammals. We
-  found lineage-specific differential expression of prolactin,
-  relaxin, and galectins as well as increased expression of the
-  pre-eclampsia associated genes \textit{ADAM12}, \textit{CRH}, and
-  \textit{PAPPA} in humans. We also identified genes which are
-  expressed in all placentas, including genes with immune-modulating
-  properties (\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and
-  \textit{S100A10}) and cell-cell interactions (\textit{LAMC1},
-  \textit{LUM}, and \textit{LGALS1}).
-
-  \textbf{Conclusions} A diverse array of genes are differentially
-  expressed and are likely to play a role in the anatomical
-  variability of the placenta in the placentas of mammalian lineages.
-  Furthermore, multiple genes which are associated with pre-eclampsia
-  (\textit{CRH}, \textit{ADAM12}, \textit{KISS1}, \textit{PAPPA},
-  \textit{IL1RL1}, and \textit{PLAC1}) show increased expression on
-  the human lineage. Still, using the strictest filtering strategy
+  \Sexpr{star.logs[,length(unique(species))]} species of mammals 
+  also identified genes which are expressed in all placentas,
+  including genes with immune-modulating properties (\textit{ANXA2},
+  \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}) and
+  cell-cell interactions (\textit{LAMC1}, \textit{LUM}, and
+  \textit{LGALS1}).
+
+  \textbf{Discussion}
+
+  Still, using the strictest filtering strategy
   we find a core set of only
@@
 
-\section{Background}
-
-The main role of the placenta in viviparous mammals is the transfer of
-nutrients, oxygen, and waste between mother and offspring. Although
-these functions are conserved among mammals, the placenta is perhaps
-the most variable organ in terms of structure and form
-\cite{Mossman.1937:comparativemorphogenesisofthe,Moffett.ea2006:Immunologyofplacentationin}. Species differ in
-the shape of attachment (i.e., discoid, diffuse, cotyledonary,
-zonary), degree of intimacy between fetal tissues and maternal tissues
-(i.e. epitheliochorial, endotheliochorial, hemochorial), and degree of
-interdigitation of maternal-fetal tissue (i.e., villous, folded,
-trabecular, lamellar, labrynthine)
-\cite{Mossman.1987:vertebratefetalmembranes}. In addition, therian
-mammals differ drastically in the number and size of their offspring,
-as well as the length of gestation, traits often associated with
-specific aspects of placental morphology
-\cite{Capellini.ea2011:Placentationandmaternalinvestment}
-(\cref{tab:gestational_periods}).
-
-The placenta arose at some point before the origin of extant mammals,
-a group that includes marsupials, monotremes, and eutherian or
-“placental mammals”. The term “placental mammals” is misleading, as
-extant monotremes (such as \textit{Ornithorhynchus anatinus}, the
-platypus) possess a rudimentary placenta, as do marsupials, although
-the direct apposition to maternal tissues lasts for a proportionally
-shorter time during gestation
-\cite{Selwood.ea2006:Trophoblastandhypoblastin,Renfree2010:ReviewMarsupialsplacentalmammals}.
-Marsupials also differ from eutherian mammals in that most possess
-only a choriovitelline placenta, produced from the fusion of the
-chorion and yolk sac, during most of gestation. Although some
-eutherian mammals may possess a choriovitelline placenta at some point
-in development, the major site of nutrient exchange in placentals
-occurs via the chorioallantoic placenta, produced from the fusion of
-the chorion and allantois
-\cite{Mossman.1987:vertebratefetalmembranes}. It is unclear whether
-the first placenta was more like that of marsupials or eutherians;
-however, using phylogenetic methods, the ancestral eutherian was
-reconstructed as possessing a relatively invasive hemochorial
-placenta, as in humans and mouse
-\cite{Elliot.ea2009:Phylogeneticevidenceforearly,Wildman.ea2006:Evolutionofmammalianplacenta}.
-
-Although the morphology of the mammalian placenta has been well
-characterized in a number of species
-\cite{Mossman.1987:vertebratefetalmembranes}, the diversity of the
-molecular environment is only beginning to be understood from a
-genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
-Studies of certain model species (i.e., \textit{Homo sapiens},
-\textit{Mus musculus}, \textit{Bos taurus}) have revealed a stunning
-divergence in the genetics of the placenta, especially in later stages
-of pregnancy
-\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
-For example, each lineage has evolved expansive placenta-specific gene
-families with expression patterns that can vary in space and time
-\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
-These include independent expansions of the prolactins in \mouse,
-\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG} in \human and \mouse, \acp{PAG}
-in \cow, genes related to the thymus-specific serine protease
-\textit{PRSS16} in \textit{Loxodona africana} (in this work),
- and the expansion of the growth hormone
-and chorionic gonadotropin beta genes in anthropoid primates
-\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
-Expression patterns of specific genes also can vary widely between
-species. For example, the expression of orthologous genes in the
-placenta have been found to vary substantially between \human and
-\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
-recruitment of \textit{SIGLEC6} to placental expression in \human
-differs from both \textit{Pan troglodytes} and \mouse
-\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
-
-One explanation for the divergent paths of the placental interface is
-the distinct evolutionary interplay between maternal and fetal
-strategies during pregnancy \cite{Haig1993:Geneticconflictsinhuman}.
-The maternal strategy is to requisition an adequate degree of
-nutrients to the fetus without risking the health of the mother, while
-the fetal strategy via paternal genes is to extract as many resources
-from the mother as possible. Therefore, an interplay between these
-extremes can set up an intricate “arms race”, involving the imprinting
-of parental alleles, expansion of gene families, and enlistment of
-existing genes to expression at the maternal-fetal interface
-\cite{Haig1993:Geneticconflictsinhuman}.
-
-One symptom of this maternal-fetal conflict is its potential affect on
-health
-\cite{Hollegaard.ea2013:Parent-offspringconflictandpersistence}. In
-addition to having an invasive hemochorial placenta, \human possess a
-deeply invasive trophoblast layer that can penetrate the myometrium of
-the uterus \cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
-Defects in placental growth are known to be associated with the
-devastating condition of pre-eclampsia, which can cause hypertension,
-proteinuria, and has the potential to lead to death of the mother
-and/or infant \cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}.
-Although cases have been documented in \textit{P. troglodytes}
-\cite{Stout.ea1969:Glomerularcapillaryendothelialswelling}, and
-\textit{Gorilla gorilla}
-\cite{Thornton.ea1992:Convulsionsinpregnancyin}, pre-eclampsia has a
-far higher incidence in \human pregnancies with as much as 4\% of
-pregnancies being affected in the United States
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. \textit{P.
-  troglodytes} and \textit{G. gorilla} also have deeply invasive
-trophoblasts, although the progression of invasion in great apes may
-be slower than \human
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. However,
-large-scale expression studies of great ape placentas have not been
-conducted. The increasing affordability of next-generation sequencing
-technology has enabled the large-scale study of specific tissues from
-a wide array of species to address broader phylogenetic questions in
-the evolution of gene expression
-\cite{Brawand.ea2011:evolutionofgeneexpression}.
-
-Here we use RNA-Seq
-technology to sequence and quantify the mRNA environment of mammalian
-placentas at term, including that of the \human, \textit{Pan paniscus}
-(the sister species to \textit{P. troglodytes}), \textit{Ateles
-  fusciceps}, \mouse, \galili, \carmeli, \textit{Canis familiaris},
-\cow, \elephant, \textit{Dasypus novemcinctus}, and
-\textit{Monodelphis domesticus}. These species represent four major
-branches of eutherian mammals (Euarchontoglires, Laurasiatheria,
-Afrotheria, and Xenartha) as well as marsupials. These placentas also represent a
-wide array of placental morphologies as well as traits such as number
-and size of offspring and length of gestation
-(\cref{tab:gestational_periods}). We use this data, along with
-existing RNA-Seq data from \mouse
-\cite{Wang.ea2011:surveyfornovelimprinted}, \human
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual}, \textit{Equus caballus}
-\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis aries}
-\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \textit{Sus scrofa}
-\cite{Ahanda.ea2012:Predictionofaltered3-} to address the major
-differences between expression in marsupial and eutherian mammals at
-term, identify the evolution of expression in important placental
-genes such as hormones and galectins, identify novel lineage-specific
-expression patterns, and quantify \human specific differences in gene
-expression that may be related to obstetrical disorders such as
-pre-eclampsia.
+\section{Introduction}
 
@@
 \end{figure}
-
-\subsection{Exploration of Individual Transcriptome Data}
-
-To determine enrichment levels of functional categories of genes in
-highly-expressed transcripts in each transcriptome, we performed
-functional annotation cluster analyses per transcriptome using topGO
-\cite{topGO:2.20.0}. Gene Ontology clusters were considered
-significantly enriched if the \ac{FDR} as estimated by the Benjamini
-and Hochberg procedure \cite{Benjamini1995:bh_procedure} on from the
-$p$ values obtained from the Fisher Exact Test were less than or equal
-to 0.05. Enrichment tests were considered with two gene lists per
-transcriptome: annotated genes with expression levels of the top 5\%
-of the frequency distribution of \ac{FPKM} (Supplemental
-\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris});
-the top 100 genes with the highest \ac{FPKM} (Supplemental
-\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris}).
-Due to the paucity of annotated gene ontology terms for some species,
-gene lists were compared using the \human Gene Ontology using the 1:1
-\human ortholog. Enrichment was considered out of the set of all 1:1
-orthologs present in each species.
 
@@
 
-\subsection{Primate Placenta Transcriptomes}
-
-\begin{table}[htb]
-\footnotesize
-<<primate.genes,echo=FALSE,results="asis">>=
-output.mentioned.genes.table(subset.table=quote(grepl("sapiens|fusciceps|paniscus",Species)),
-                             human.symbol=TRUE
-                             )
-@ 
-\caption{Notable genes ordered by species and \ac{FPKM}. Gene
-  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
-  name corresponding to the human symbol. A complete table of 
-  genes with \ac{FPKM} $\ge 50$ is located in the supplemental material }
-\label{tab:notable_primate_genes}
-\end{table}
-
-\begin{table}[htb]
-  \footnotesize
-<<primate.lineage.specific,echo=FALSE,results="asis">>=
-pri.all.lin.spec <- placenta.lineage.specific.table("pri.all")
-setkey(pri.all.lin.spec,"Human Symbol")
-@ 
-\caption{\lineagetableexplanation{primates}{all other analyzed species}}
-\label{tab:primate_lineage_specific}
-\end{table}
-
-We examined the expression of three primates, \human, \bonobo, and
-\smonkey. The genes whose expressions are different on the primate
-lineage in comparison to all other analyzed species are listed in
-\cref{tab:primate_lineage_specific}. \textit{FSTL1} has significantly
-higher expression in the primate lineage
-(Supplemental \cref{fig:fstl1_evolution_tree},
-$\text{FDR}=\Sexpr{pri.all.lin.spec["FSTL1",`FDR`]}$), is implicated
-in BMP signaling, and has previously been implicated in inflammation
-and invasiveness
-\cite{Mouillet.ea2015:expressionandpost-transcriptionalregulation}. Of
-non-primates, \horse shows the highest expression
-%
-\Sexpr{gene.expression(species.name="equus caballus",gene="FSTL1",complete=TRUE)}, 
-%
-which is still lower than the lowest primate, \smonkey
-%
-(\Sexpr{gene.expression(species.name="ateles fusciceps",gene="FSTL1",complete=TRUE)}). 
-%
-No other genes were significantly differentially expressed on this
-lineage by t-test.
-
-\subsubsection{\human Placenta Transcriptome}
-\label{sec:human_placenta}
-
-% \begin{table}
-%   \footnotesize
-% <<top.50.human.expressed.genes,echo=FALSE,results="asis">>=
-% output.top.genes.table(top.genes.table(species.name="homo sapiens"))
-% @ 
-% \caption{The top 50 genes by \ac{FPKM} in \human which are not
-%   annotated as Mt\_rRNA, Mt\_tRNA, rRNA or a housekeeping gene. Gene
-%   symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
-%   name corresponding to the human symbol. A complete list of all genes
-%   is in \cref{tab:all_homo_sapiens_placenta}.}
-% \label{tab:top50_human_placenta}
-% \end{table}
-
-<<cgb.min.expression,results="hide",echo=FALSE,warnings=FALSE>>=
-cgb.min.expression <- 
-    min(sapply(paste0("CGB",c("","1","2","5","7","8")),
-               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
-cgb.sum.expression <- 
-    sum(sapply(paste0("CGB",c("","1","2","5","7","8")),
-               function(x){gene.expression(species.name="homo sapiens",gene=x,units=FALSE)}))
-@
-
-The gene with by far the highest \ac{FPKM} value
-%
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="CSH1")}) %
-%
-in \human is \textit{CSH1}, followed by \textit{CSH2}, and \textit{CSHL1}, all
-three of which are members of a distinct expansion of the growth
-hormone gene family within catarrhines (Supplemental \cref{fig:galectins_evolution_tree}). Another member of the growth
-hormone gene family, \ac{GH2} (%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="GH2")}), %
-is also expressed to a high degree in the \human placenta
-(Supplemental \cref{tab:all_homo_sapiens_placenta}).
-% 
-Eleven \acp{PSG} are highly-expressed
-(\ac{FPKM}$\ge 100$) in the term placenta of \human, with \textit{PSG1} showing the
-greatest degree of expression %
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG1")}) % 
-and the pseudogene \textit{PSG10P} showing the least %
-(\Sexpr{gene.expression(species.name="homo sapiens",gene="PSG10P")}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~A) . %
-These
-genes are members of a hominoid-specific expansion of the \ac{CEA} family
-and known to be expressed by the syncytiotrophoblast
-\cite{Rawn.ea2008:evolutionregulationandfunction,Lei.ea1992:Cloningandexpressionof}. %
-%
-Genes involved in the production of estrogens (%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP19A1",complete=TRUE)},
-% 
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD17B1",complete=TRUE)}, and
-% 
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)}%
-% 
-) and progesterone (%
-\Sexpr{gene.expression(species.name="homo sapiens",gene="HSD3B1",complete=TRUE)},
-\Sexpr{gene.expression(species.name="homo sapiens",gene="CYP11A1",complete=TRUE)}, and
-\Sexpr{gene.expression(species.name="homo sapiens",gene="ADM",complete=TRUE)})
-%
-show exceptionally high expression in the human placenta, as well as %
-%
-\textit{CGA} (FPKM=\Sexpr{gene.expression(species.name="homo sapiens",gene="CGA")}, %
-%
-which generates the alpha subnit of multiple human chorionic
-gonadotropin hormones (Supplemental \cref{tab:all_homo_sapiens_placenta}). The genes
-producing the beta subunits of the chorionic gonadotropins
-(\textit{CGB}, \textit{CGB1}, \textit{CGB2}, \textit{CGB5},
-\textit{CGB7}, \textit{CGB8}) have reasonable expression
-($\text{FPKM} \ge \Sexpr{cgb.min.expression}$), and the sum of their
-expression is on the order of the expression of
-\textit{CGA} ($\sum \mathit{CGBs}=\Sexpr{cgb.sum.expression}$ FPKM) .
-%
-%
-Other highly expressed genes in the top 50 include members of the
-hemoglobin family (e.g. \textit{HBA1}, \textit{HBA2}, \textit{HBB},
-and \textit{HBG2}), the maternally imprinted long noncoding RNA gene
-\textit{H19}, genes coding for proteins required for growth and/or
-cell proliferation (\textit{ADAM12}, \textit{KISS1}, \textit{IGF2},
-\textit{GPC3}, \textit{CSF3R}), anticoagulation (\textit{TFPI2}), and
-glucose transport (\textit{SLC2A1}) among others
-(Supplemental \cref{tab:all_homo_sapiens_placenta}, Supplemental
-\cref{tab:all_homo_sapiens_placenta}).
-
-% \paragraph{Pre-eclampsia \human Lineage Specific Genes}
+\subsection{Pre-eclampsia \human Lineage Specific Genes}
 
@@
 
-\subsubsection{\bonobo Placenta Transcriptome}
-
-% \begin{table}
-%   \footnotesize
-% <<top.50.bonobo.expressed.genes,echo=FALSE,results="asis">>=
-% top.bonobo.genes <- top.genes.table(species.name="pan paniscus")
-% output.top.genes.table(top.bonobo.genes)
-% setkey(top.bonobo.genes,"Gene symbol")
-% @ 
-% \caption{\tableexplanation{\bonobo}{tab:all_pan_paniscus_placenta}}
-% \label{tab:top50_bonobo_placenta}
-% \end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \bonobo are shown in Supplemental \cref{tab:all_pan_paniscus_placenta}
-with a list of all genes in Supplemental
-\cref{tab:all_pan_paniscus_placenta}. Expression in the \bonobo
-placenta differs somewhat from that of \human, as evidenced by the
-makeup of the top 50 genes (Supplemental \cref{tab:all_pan_paniscus_placenta}). The highest expression
- in \bonobo is %
-\Sexpr{gene.expression("pan paniscus","PRG2",complete=TRUE)}, %
-%
-which is involved in insulin growth factor signaling and angiogenesis
-during pregnancy \cite{Weyer.ea2011:Placentalregulationofpeptide}.
-\textit{PRG2} is also expressed at relatively high levels (though
-comparatively much lower) in the human placenta
-(\Sexpr{gene.expression("homo sapiens","PRG2")};
-Supplemental \cref{tab:all_homo_sapiens_placenta}). Four \bonobo genes
-related to the chorionic somatomammotropins
-are also expressed at relatively high levels (\textit{GH1} 
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034560",complete=TRUE)}, 
-\textit{GH1} 
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034212",complete=TRUE)}, 
-\textit{GH2}
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000030805",complete=TRUE)}, 
-and \textit{GH2}
-\Sexpr{gene.expression("pan paniscus","ENSPTRG00000034209",complete=TRUE)}, 
-%
-\cref{tab:all_pan_paniscus_placenta}, \cref{fig:csh1_evolution_tree}).
-Placental expression of the genes involved in estrogen and
-progesterone production (\textit{CYP19A1}, \textit{HSD3B1}
-respectively) show lower expression in \bonobo as compared to \human %
-(\Sexpr{gene.expression("pan paniscus","CYP19A1",complete=TRUE)},
-\cref{fig:estrogen_evolution_tree}; %
-\Sexpr{gene.expression("pan paniscus","HSD3B1",complete=TRUE)},
-\cref{fig:progesterone_evolution_tree} and
-Supplemental \cref{tab:all_homo_sapiens_placenta}) although the expression of
-\textit{HSD3B1} is fairly similar to that of \smonkey %
-(\Sexpr{gene.expression("ateles fusciceps","HSD3B1",complete=TRUE)}. %
-In addition, \textit{CGA} is expressed in the bonobo placenta %
-(\Sexpr{gene.expression("pan paniscus","CGA")}), but at much
-lower levels than \human as well (Supplemental \cref{tab:all_homo_sapiens_placenta}).
-%
-Two Ensembl-annotated beta subunits of the chorionic gonadotropin
-genes are expressed in \bonobo, one related to \textit{CGB} and
-another related to \textit{CGB7}; both are not highly expressed
-(\cref{fig:cga_cgb_evolution_tree}). Among the other top 50 expressed
-genes, there is high expression of keratins (e.g., \textit{KRT19},
-\textit{KRT18}), heat-shock proteins (e.g., \textit{HSP90B1},
-\textit{HSPA1A}), serpins (\textit{SERPINE1}, \textit{SERPINE2},
-Supplemental \cref{fig:serpine_evolution_tree}), insulin growth factor
-2 (\textit{IGF2}) and one of its binding proteins \textit{IGFBP3}, and
-the hormone adrenomedullin (\textit{ADM}).
-
-\subsubsection{\smonkey Transcriptome}
-
-% \begin{table}
-%   \footnotesize
-% <<top.50.spider.monkey.expressed.genes,echo=FALSE,results="asis">>=
-% top.spider.monkey.genes <- top.genes.table(species.name="ateles fusciceps")
-% output.top.genes.table(top.spider.monkey.genes)
-% setkey(top.spider.monkey.genes,"Gene symbol")
-% @ 
-% \caption{\tableexplanation{\smonkey}{tab:all_ateles_fusciceps_placenta}}
-% \label{tab:top50_spider_monkey_placenta}
-% \end{table}
-
-For the term placenta of \smonkey, assembled contigs from RNA-seq were
-aligned using against \human and the best alignment was chosen as the
-annotation for each assembled contig. In the term placenta of
-\smonkey, the top 50 non-housekeeping genes
-(Supplemental \cref{tab:all_spider_monkey_placenta}) include a possible regulator
-of androgen receptor signaling
-\cite{Sampson.ea2012:PAGE4positivityisassociated}, \textit{PAGE4} %
-(\Sexpr{gene.expression("ateles_fusciceps","PAGE4")}), %
-the pregnancy specific glycoprotein \textit{PSG8}
-(\Sexpr{gene.expression("ateles_fusciceps","PSG8")}), %
-only one member of the chorionic somatotropins, \textit{GH1} %
-(\Sexpr{gene.expression("ateles_fusciceps","GH1")}), %
-a few HLA genes (\textit{HLA-A}, \textit{HLA-F},
-Supplemental \cref{tab:all_spider_monkey_placenta}), and multiple components of
-annexin complexes
-(\Sexpr{gene.expression("ateles_fusciceps","ANXA2",complete=TRUE)}, %
-\Sexpr{gene.expression("ateles_fusciceps","S100P",complete=TRUE)}, %
-\Sexpr{gene.expression("ateles_fusciceps","S100A11",complete=TRUE)}) %
-which may play a role in angiogenesis, and certainly regulate
-cell-cell interaction in the developing placenta
-\cite{Liu.ea2015:AnnexinA2complexeswith}.
-\Sexpr{gene.expression("ateles_fusciceps","AL365202.1",complete=TRUE)}
-is also highly expressed, has no known function, and is duplicated
-multiple times in \textit{Macaca mulatta} (Ensembl gene tree
-\href{http://useast.ensembl.org/Multi/GeneTree/Image?gt=ENSGT00580000081348}{ENSGT00580000081348}).
-All genes with expression greater than 50 \ac{FPKM} are listed in
-Supplementary \cref{tab:all_ateles_fusciceps_placenta}.
-
-\subsection{Rodentia Placenta Transcriptomes}
-
-\begin{table}
-  \footnotesize
-<<rodentia.genes,echo=FALSE,results="asis">>=
-output.mentioned.genes.table(subset.table=quote(grepl("musculus|galili|carmeli",Species)),
-                             human.symbol=TRUE
-                             )
-@ 
-\caption{Notable genes ordered by species and \ac{FPKM}. Gene
-  symbol is the gene symbol or Ensembl gene id. Human Name is the HUGO
-  name corresponding to the human symbol. A complete table of 
-  genes with \ac{FPKM} $\ge 50$ is located in the supplemental material }
-\label{tab:notable_non_primate_genes}
-\end{table}
-
-We examined the expression of three rodents, \mouse, \carmeli, and
-\galili{}. We did not detect any significant changes on the Rodentia
-lineage in comparison to all other analyzed species (Supplemental
-\cref{tab:rodentia_lineage_specific}), likely because of a lack power
-to overcome the multiple testing correction. Additional species and/or
-samples in each of these lineages will be necessary to detect lineage
-specific shifts.
-
-\subsubsection{\mouse Placenta Transcriptome}
-
-% \begin{table}[h]
-%   \footnotesize
-% <<top.50.mouse.expressed.genes,echo=FALSE,results="asis">>=
-% output.top.genes.table(top.genes.table(species.name="mus musculus"))
-% @ 
-% \caption{\tableexplanation{\mouse}{tab:all_mus_musculus_placenta}}
-% \label{tab:top50_mouse_placenta}
-% \end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \mouse are shown in
-Supplemental \cref{tab:all_mus_musculus_placenta} with a list of all genes with
-$\text{FPKM} \ge 50$ in Supplemental
-\cref{tab:all_mus_musculus_placenta}. The top expressed genes include
-\Sexpr{gene.expression("mus_musculus","Mir6236",complete=TRUE)}, whose
-target \textit{in silico} is the orphan G-coupled receptor
-\textit{Mrgprg}
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget},
-\Sexpr{gene.expression("mus_musculus","H19",complete=TRUE)} and
-\Sexpr{gene.expression("mus_musculus","Gm27786",complete=TRUE)}, both
-of which are maternally imprinted (the later of which is a homolog of
-human \textit{H19\_2}),
-\Sexpr{gene.expression("mus_musculus","Igf2",complete=TRUE)}, numerous
-cathepsins and a rodentia specific expansion of prolactin related
-genes. There are also two miRNA which overlap with the of the 7S
-signal recognition particle
-(\Sexpr{gene.expression("mus_musculus","AC157822.1",complete=TRUE)}
-and
-\Sexpr{gene.expression("mus_musculus","AC099934.1",complete=TRUE)})
-and numerous pseudogenes
-(\Sexpr{gene.expression("mus_musculus","Gm11966",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm10736",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm10180",complete=TRUE)}
-\Sexpr{gene.expression("mus_musculus","Gm29216",complete=TRUE)}) which
-appear to be highly expressed, however, the short read length (34--44)
-in many samples is likely preventing unambiguous assignment of all
-reads to the more lengthy complete transcript, so these shorter
-transcripts should be treated with suspicion, and we will ignore them
-in further discussion. The prolactin-related genes
-(\Sexpr{gene.expression("mus_musculus","Prl3b1",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Prl2b1",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Prl8a9",complete=TRUE)}, and
-\Sexpr{gene.expression("mus_musculus","Prl8a6",complete=TRUE)}) are
-very highly expressed in \mouse in, with %
-\Sexpr{combined.fpkm[species=="mus musculus" &
-  grepl("Prl(\\d.\\d)?",gene_short_name), sum(mean_fpkm >= 1)]}
-prolactin related genes having $\text{FPKM} \ge 1$
-(\cref{fig:prolactin_evolution_tree}, Supplemental
-\cref{fig:prolactin_receptor_evolution_tree}). The cathespins have
-also expanded in rodents, with six showing high expression
-(\Sexpr{gene.expression("mus_musculus","Ctsj",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Cts6",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Ctsr",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Ctsq",complete=TRUE)},
-\Sexpr{gene.expression("mus_musculus","Cts3",complete=TRUE)}, and
-\Sexpr{gene.expression("mus_musculus","Ctsm",complete=TRUE)},
-Supplemental \cref{tab:all_mus_musculus_placenta,fig:pag2_prss16_ctsl_psg_evolution_tree}).
-Finally, \textit{Peg10}, an imprinted gene required for proper
-formation of the placenta in mice is also highly expressed
-(\Sexpr{gene.expression("mus_musculus","Peg10")}), as previously
-reported \cite{Ono.ea2006:DeletionofPeg10imprinted}.
-
-% \paragraph{\mouse Lineage Specific Genes}
-
-\begin{table}
-  \footnotesize
-<<mouse.lineage.specific,echo=FALSE,results="asis">>=
-mou.all.lin.spec <-
-    placenta.lineage.specific.table("mou.all",
-                                    lineage.table=human.mouse.lineage.specific.analysis,
-                                    members=40,
-                                    orderbyfc=TRUE,
-                                    fdr.cutoff=0.001
-                                    )
-setkey(mou.all.lin.spec,"Human Symbol")
-@ 
-\caption{\lineagetableexplanationtypeiii{\mouse}{all other analyzed species}}
-\label{tab:mouse_lineage_specific}
-\end{table}
-
-Using a mixed-effects linear model to account for repeated testing of
-\human and \mouse, we identified multiple genes which are
-differentially expressed in the \mouse lineage in comparison to all
-other lineages (\cref{tab:mouse_lineage_specific}). \mouse
-\textit{Gkn2} (FDR=\Sexpr{mou.all.lin.spec["GKN2",`FDR`]},
-FC=\Sexpr{mou.all.lin.spec["GKN2",`Fold Change`]}) has been shown to
-have a pattern of expression distinct from the \human pattern, and has
-been suggested to be important for placental development
-\cite{Fahlbusch.ea2015:Trophoblastexpressiondynamicsof}. Notable
-mouse-specific increases in expression include include genes like
-\textit{Slc38a4} (FDR=\Sexpr{mou.all.lin.spec["SLC38A4",`FDR`]},
-FC=\Sexpr{mou.all.lin.spec["SLC38A4",`Fold Change`]}) and \textit{Afp}
-(FDR=\Sexpr{mou.all.lin.spec["AFP",`FDR`]},
-FC=\Sexpr{mou.all.lin.spec["AFP",`Fold Change`]}) which are implicated
-in fetal growth restriction
-\cite{Sharma.ea2016:Tpbpa-Cre-mediateddeletionofTFAP2C} and placenta
-abnormalities in humans
-\cite{Waller.ea1996:associationbetweenmaternalserum}, and
-\textit{Cnx2} (FDR=\Sexpr{mou.all.lin.spec["CNX2",`FDR`]},
-FC=\Sexpr{mou.all.lin.spec["CNX2",`Fold Change`]}) which is critical
-for the exchange of metabolites in the placenta labyrinth
-\cite{Gabriel.ea1998:Transplacentaluptakeofglucose}.
-
-\subsubsection{\galili Placenta Transcriptome}
-
-% \begin{table}[h]
-%   \footnotesize
-% <<top.50.n.galili.expressed.genes,echo=FALSE,results="asis">>=
-% output.top.genes.table(top.genes.table(species.name="spalax galili"))
-% @ 
-% \caption{\tableexplanation{\galili}{tab:all_spalax_galili_placenta}}
-% \label{tab:top50_n_galili_placenta}
-% \end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \galili are shown in
-Supplemental \cref{tab:all_n_galili_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_spalax_galili_placenta}.
-Components of annexin complexes are highly expressed
-(\Sexpr{gene.expression("spalax_galili","S100a6",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","S100a11",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","S100a10",complete=TRUE)},
-and
-\Sexpr{gene.expression("spalax_galili","Anxa2",complete=TRUE)}),
-as are members of the thymosin family
-(\Sexpr{gene.expression("spalax_galili","Tmsb4x",complete=TRUE)}
-and
-\Sexpr{gene.expression("spalax_galili","Tmsb10",complete=TRUE)}),
-various cytoskeleton related proteins
-(\Sexpr{gene.expression("spalax_galili","Actc1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Actg2",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Acta2",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Myl9",complete=TRUE)},
-\Sexpr{gene.expression("spalax_galili","Vim",complete=TRUE)}), and
-transmembrane protein of currently unknown function, 
-\Sexpr{gene.expression("spalax_galili","Gm10134",complete=TRUE)}.
-
-\subsubsection{\carmeli Placenta Transcriptome}
-
-% \begin{table}[h]
-%   \footnotesize
-% <<top.50.s.carmeli.expressed.genes,echo=FALSE,results="asis">>=
-% output.top.genes.table(top.genes.table(species.name="spalax carmeli"))
-% @ 
-% \caption{\tableexplanation{\carmeli}{tab:all_spalax_carmeli_placenta}}
-% \label{tab:top50_s_carmeli_placenta}
-% \end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \carmeli are shown in
-Supplemental \cref{tab:all_s_carmeli_placenta} with a list of all highly
-expressed genes in Supplemental
-\cref{tab:all_spalax_carmeli_placenta}. Numerous hemoglobin subunits
-ware highly expressed
-(\Sexpr{gene.expression("spalax_carmeli","Hbb-bh1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Hba-a1",complete=TRUE)}, and
-\Sexpr{gene.expression("spalax_carmeli","Hbb-bs",complete=TRUE)}), as
-well as the relaxin
-\Sexpr{gene.expression("spalax_carmeli","Rln1",complete=TRUE)},
-multiple cathespins
-(\Sexpr{gene.expression("spalax_carmeli","Ctsm",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","BC051665",complete=TRUE)},
-and \Sexpr{gene.expression("spalax_carmeli","Ctsl",complete=TRUE)},
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), and many prolactin-related genes
-(\Sexpr{gene.expression("spalax_carmeli","Prl4a1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Prl6a1",complete=TRUE)},
-\Sexpr{gene.expression("spalax_carmeli","Prl3b1",complete=TRUE)}, and
-\Sexpr{gene.expression("spalax_carmeli","Prl7c1",complete=TRUE)},
-\cref{fig:prolactin_evolution_tree}) which shows a similar pattern of
-expansion seen in \mouse.
-
-\subsection{Laurasiatheria}
-
-We examined the expression of five laurasiatherians: \cow, \sheep,
-\pig, \horse, and \dog. We were unable to detect any significant
-differences after correcting for multiple testing between
-laurasiatherians and all other analyzed species
-(Supplemental \cref{tab:laurasiatheria_lineage_specific}) or between
-laurasiatherians and Euarchontoglires
-(Supplemental \cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).
-
-\subsubsection{\dog Placenta Transcriptome}
-
-% \begin{table}
-%   \footnotesize
-% <<top.50.dog.expressed.genes,echo=FALSE,results="asis">>=
-% output.top.genes.table(top.genes.table(species.name="canis familiaris"))
-% @ 
-% \caption{\tableexplanation{\dog}{tab:all_canis_familiaris_placenta}}
-% \label{tab:top50_dog_placenta}
-% \end{table}
-
-The top 50 non-housekeeping expressed genes as measured by \ac{FPKM}
-in the term placenta of \dog are shown in
-Supplemental \cref{tab:all_canis_familiaris_placenta} with a list of all genes in Supplemental
-\cref{tab:all_canis_familiaris_placenta}. The gene with the highest
-expression %
-(\Sexpr{gene.expression("canis familiaris","TMSB10")}) is the thymosin 
-\textit{TMSB10} %
-(\Sexpr{gene.expression("canis familiaris","ENSCAFG00000002115")}) %
-is the hormone relaxin (\textit{RLN2}). Other highly expressed genes
-among the top 50 (Supplemental \cref{tab:all_canis_familiaris_placenta}) include genes
-involved in the cytoskeleton (\textit{TMSB10}, \textit{TMSB4X}),
-anticoagulation (\textit{TFPI2}, \textit{PLA2G7}), innate immunity
-(\textit{LCN2}, \textit{GM2A}), lipid metabolism (\textit{APOE}),
-inhibition of matrix metalloproteinases (\textit{TIMP2},
-\textit{TIMP3}), and growth factors (\textit{PGF}, \textit{EGFR},
-\textit{HTRA1}). The gene \textit{LGALS1}, which shows evidence of
-immunosupression during pregnancy
-\cite{Than.ea2008:Emergenceofhormonaland}, is highly expressed %
-(\Sexpr{gene.expression("canis familiaris","LGALS1")}). %
-In addition, an ortholog of \textit{SSLP1} (ENSCAFG00000010638) a gene
-expressed in the placentas of \cow (%
-\Sexpr{gene.expression("bos taurus","SSLP1",complete=TRUE)}, %
-Supplemental \cref{tab:all_bos_taurus_placenta}) and \sheep
-(\Sexpr{gene.expression("ovis aries","SOLD1",complete=TRUE)},
-Supplemental \cref{tab:all_sheep_placenta}) but is not present in primates
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}
-(\cref{fig:sslp1_evolution_tree}), is also expressed at high levels
-%
-(\Sexpr{gene.expression("canis familiaris","ENSCAFG00000010638")}). %
-There is no placental expression of the estrogen synthase gene
-(\textit{CYP19A1}) in \dog 
-(\Sexpr{gene.expression("canis familiaris","CYP19A1")}
-\cref{fig:estrogen_evolution_tree}), 
-and expression of \textit{HSD3B7} is also low
-(\Sexpr{gene.expression("canis familiaris","HSD3B7")}, 
-\cref{fig:progesterone_evolution_tree}).
-
-\subsubsection{\cow Placenta Transcriptome}
-
-% \begin{table}
-%   \footnotesize
-% <<top.50.cow.expressed.genes,echo=FALSE,results="asis">>=
-% output.top.genes.table(top.genes.table(species.name="bos taurus"))
-% @ 
-% \caption{\tableexplanation{\cow}{tab:all_bos_taurus_placenta}}
-% \label{tab:top50_cow_placenta}
-% \end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the \cow are shown in Supplemental \cref{tab:all_bos_taurus_placenta} with a
-list of all genes in Supplemental \cref{tab:all_bos_taurus_placenta}.
-The gene with the highest expression was \textit{TMSB10} %
-(\Sexpr{gene.expression("bos taurus","TMSB10")}), and
-\textit{TMSB4} was also highly expressed %
-(\Sexpr{gene.expression("bos taurus","TMSB4")}). As in
-\dog{}, members of annexin complexes were also highly expressed
-(\Sexpr{gene.expression("bos taurus","S100G",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ANXA2",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","S100A2",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","S100A11",complete=TRUE)}), as well as %
-at least one of a set of four miRNA with 100\% identity
-(\Sexpr{gene.expression("bos taurus","ENSBTAG00000047046",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000047154",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000045710",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000046351",complete=TRUE)}). 
-%
-As these are novel miRNA, there is a possibility that the assignment
-of the short reads to these four miRNA is due to miss-annotation of
-the \cow genome.
-% 
-Several members of the ruminant-specific pregnancy-associated
-glycoprotein family \cite{Green.ea2000:Pregnancy-associatedbovineandovine} 
-show especially high expression in \cow placenta
-(Supplemental \cref{tab:all_bos_taurus_placenta}, \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D, and
-Supplemental \cref{tab:all_bos_taurus_placenta}) including
-%
-\Sexpr{gene.expression("bos taurus","PAG2",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PAG10",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000022759",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000013214",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000019226",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","ENSBTAG00000024611",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PAG12",complete=TRUE)}. %
-%
-28 members of this family have $\text{FPKM} \ge 1$, and this family of
-genes has greatly expanded in ruminants (\cow and \sheep{}) although
-the expression levels of this family are higher in \cow than in \sheep
-(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). Other genes with high expression
-include genes involved with regulation of growth (\textit{IGF2},
-\textit{PHLDA2}), imprinted genes (\textit{DLK1}, \textit{PHLDA2}),
-and \textit{CSH1}, a gene that is a member of a ruminant specific
-expansion of placental prolactins and not directly related to the gene
-of the same name in hominids \ref{fig:prolactin_evolution_tree}. Other
-ruminant-specific placental prolactins also show high expression in
-the cow placenta including %
-\Sexpr{gene.expression("bos taurus","PRP1",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP5",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP6",complete=TRUE)}, %
-\Sexpr{gene.expression("bos taurus","PRP8",complete=TRUE)}, and %
-\Sexpr{gene.expression("bos taurus","PRP7",complete=TRUE)}, %
-with a total of 14 with $\text{FPKM} \ge 1$.
-%
-\textit{SSLP1} (\textit{SOLD1} in \sheep{}), a gene expressed in the
-cotyledonary villi of the ruminiant placenta and involved in cell
-proliferation
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}, also
-shows evidence of high expression %
-(\Sexpr{gene.expression("bos taurus","SSLP1")}). %
-There is some placental expression of the estrogen synthase gene
-(\textit{CYP19A1}) in cows %
-(\Sexpr{gene.expression("bos taurus","CYP19A1")}), and
-comparatively greater expression of \textit{HSD3B} %
-(\Sexpr{gene.expression("bos taurus","HSD3B")}).
-
-\subsubsection{\pig Placenta Transcriptome}
-
-% \begin{table}[h]
-%   \footnotesize
-% <<top.50.pig.expressed.genes,echo=FALSE,results="asis">>=
-% output.top.genes.table(top.genes.table(species.name="sus scrofa"))
-% @ 
-% \caption{\tableexplanation{\pig}{tab:all_sus_scrofa_placenta}}
-% \label{tab:top50_pig_placenta}
-% \end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \textit{Sus scrofa} are shown in
-Supplemental \cref{tab:all_pig_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_sus_scrofa_placenta}. The top two
-expressed transcripts,
-\Sexpr{gene.expression("sus_scrofa","ENSSSCG00000025162",complete=TRUE)}
-and
-\Sexpr{gene.expression("sus_scrofa","ENSSSCG00000021356",complete=TRUE)},
-are miRNA which correspond to the \cow miRNA
-\href{http://www.mirbase.org/cgi-bin/mirna_entry.pl?acc=MI0013080}{bta-mir-2904-1}
-and
-\href{http://www.mirbase.org/cgi-bin/mirna_entry.pl?acc=MIMAT0013845}{bta-mir-2887-1},
-respectively, both of which are also expressed in \cow ovary
-\cite{Hossain.ea2009:Identificationandcharacterizationof}, although
-neither of which have an obvious target gene on the basis of
-\textit{in silico} analysis
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}. Other highly
-expressed genes include members of the insulin growth factor family
-(\Sexpr{gene.expression("sus_scrofa","IGF2",complete=TRUE)} and
-\Sexpr{gene.expression("sus_scrofa","IGFBP3",complete=TRUE)}), members
-of annexin complexes
-(\Sexpr{gene.expression("sus_scrofa","S100A1",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","ANXA2",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","S100A12",complete=TRUE)}, and
-\Sexpr{gene.expression("sus_scrofa","S100A10",complete=TRUE)}), the
-galectin \Sexpr{gene.expression("sus_scrofa","LGALS3",complete=TRUE)}
-and multiple cytoskeleton and extracellular matrix proteins
-(\Sexpr{gene.expression("sus_scrofa","KRT19",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","COL1A1",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","BETA-ACTIN",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","KRT9",complete=TRUE)},
-\Sexpr{gene.expression("sus_scrofa","CLDN4",complete=TRUE)}, and
-\Sexpr{gene.expression("sus_scrofa","KRT9",complete=TRUE)}).
-
-\subsubsection{\horse Placenta Transcriptome}
-
-% \begin{table}[h]
-%   \footnotesize
-% <<top.50.horse.expressed.genes,echo=FALSE,results="asis">>=
-% output.top.genes.table(top.genes.table(species.name="equus caballus"))
-% @ 
-% \caption{\tableexplanation{\horse}{tab:all_equus_caballus_placenta}}
-% \label{tab:top50_horse_placenta}
-% \end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \horse are shown in
-Supplemental \cref{tab:all_equus_caballus_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_equus_caballus_placenta}. The gene
-with the highest expression was the relaxin,
-\Sexpr{gene.expression("equus_caballus","RLN",complete=TRUE)},
-followed by two miRNA
-(\Sexpr{gene.expression("equus_caballus","ENSECAG00000025328",complete=TRUE)}
-and
-\Sexpr{gene.expression("equus_caballus","ENSECAG00000027056",complete=TRUE)})
-with no known \textit{in silico} targets
-\cite{Agarwal.ea2015:PredictingeffectivemicroRNAtarget}, the thymosin
-\Sexpr{gene.expression("equus_caballus","TMSB4",complete=TRUE)},
-members of the insulin growth factor family
-(\Sexpr{gene.expression("equus_caballus","IGF2",complete=TRUE)} and
-\Sexpr{gene.expression("equus_caballus","IGFBP3",complete=TRUE)}) and
-multiple members of annexin complexes
-(\Sexpr{gene.expression("equus_caballus","S100A6",complete=TRUE)},
-\Sexpr{gene.expression("equus_caballus","S100A1",complete=TRUE)}, and
-\Sexpr{gene.expression("equus_caballus","ANXA2",complete=TRUE)}),
-similar to the pattern of expression seen in \pig and \dog.
-
-\subsubsection{\sheep Placenta Transcriptome}
-
-% \begin{table}[h]
-%   \footnotesize
-% <<top.50.sheep.expressed.genes,echo=FALSE,results="asis">>=
-% top.sheep.genes <- top.genes.table(species.name="ovis aries",n=200)
-% top.sheep.genes <-
-%     top.sheep.genes[!grepl("\\{(7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\}",
-%                            `Gene symbol`) &
-%                     !grepl("\\((7SK|.+_([SM]RP|nuc)|SNOR[AD]\\d+|U\\d+|HIST.*)\\)",
-%                            `Human Symbol`)
-%                    ,][1:50,]
-% output.top.genes.table(top.sheep.genes)
-% @ 
-% \caption{\tableexplanation{\sheep}{tab:all_ovis_aries_placenta}}
-% \label{tab:top50_sheep_placenta}
-% \end{table}
-
-The 50 non-housekeeping genes with highest expression (by \ac{FPKM})
-in the term placenta of \sheep are shown in
-Supplemental \cref{tab:all_sheep_placenta} with a list of all highly expressed
-genes in Supplemental \cref{tab:all_ovis_aries_placenta}. The complete
-list of highly expressed genes includes many snRNAs and rRNAs likely
-due to the RNA extraction and library preparation procedure utilized
-for this sample. Ignoring these, notable highly expressed genes
-include members of annexin complexes
-(\Sexpr{gene.expression("ovis_aries","S100G",complete=TRUE)},
-\Sexpr{gene.expression("ovis_aries","ANXA2",complete=TRUE)}),
-hemoglobin subunits
-(\Sexpr{gene.expression("ovis_aries","ENSOARG00000010944",complete=TRUE)}),
-an ortholog of \textit{SSLP1},
-\Sexpr{gene.expression("ovis_aries","SOLD1",complete=TRUE)}, two
-cathespins (\Sexpr{gene.expression("ovis_aries","CTSL",complete=TRUE)}
-and \Sexpr{gene.expression("ovis_aries","CTSB",complete=TRUE)}), and
-placentally expressed transcript 1
-\Sexpr{gene.expression("ovis_aries","PLET1",complete=TRUE)} which may
-play a role in cellular migration
-\cite{Raymond.ea2010:Expressionoforphanprotein}.
-
-\subsection{Atlantogenata (Afrotheria + Xenartha)}
-\subsubsection{\elephant Placenta Transcriptome}
-
-% \begin{table}
-% \footnotesize
-% <<top.50.elephant.expressed.genes,echo=FALSE,results="asis">>=
-% output.top.genes.table(top.genes.table(species.name="loxodonta africana"))
-% @ 
-% \caption{\tableexplanation{\elephant}{tab:top50_elephant_placenta}}
-% \label{tab:top50_elephant_placenta}
-% \end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of \elephant are shown in Supplemental \cref{tab:all_loxodonta_africana_placenta}
-with a list of all genes in
-\cref{tab:all_loxodonta_africana_placenta}. The gene with the highest
-expression was the natriuretic peptide precursor B
-(\Sexpr{gene.expression("loxodonta_africana","NPPB",complete=TRUE)}).
-Two thymosins,
-\Sexpr{gene.expression("loxodonta_africana","TMSB10",complete=TRUE)}
-and
-\Sexpr{gene.expression("loxodonta_africana","TMSB4Y",complete=TRUE)}
-were also also was highly expressed. The miRNA mir-675
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027406",complete=TRUE)}).
-which is imbedded in the first exon of the maternally imprinted long
-non-coding RNA gene \textit{H19}
-\cite{Keniry.ea2012:H19lincRNAisdevelopmental}, a gene which is
-currently un-annotated in \elephant{}. %
-% \DLA{Probably consider fixing the unannotation of H19; it seems pretty
-%   obvious to me, though it would require fixing the GTF file for
-%   elephant with the right
-%   \href{http://useast.ensembl.org/Loxodonta_africana/Transcript/Summary?db=core;pt=GENSCAN00000034833;r=scaffold_145:855987-899506}{genscan
-%     prediction} for this region} %
-Five uncharacterized genes are expressed to a high degree in the
-\elephant placenta (Supplemental \cref{tab:all_loxodonta_africana_placenta}), three of
-which
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000028493",complete=TRUE)},
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032011",complete=TRUE)},
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000031933",complete=TRUE)})
-are members of an \elephant{}-specific expansion of 11 genes closely
-related to the thymus-specific serine protease gene \textit{PRSS16}
-(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~C). The other 8
-\textit{PRSS16}-like genes are also expressed to a lesser degree
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000026505",complete=TRUE)}).
-Two other novel genes with high expression are related to the gene
-\textit{SSLP1}
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000032659",complete=TRUE)}
-and
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000029551",complete=TRUE)},
-\cref{fig:sslp1_evolution_tree}), which is also expressed in \cow and
-\dog placentas (Supplemental \cref{tab:all_canis_familiaris_placenta,tab:all_bos_taurus_placenta}).
-Placental expression of the hormones relaxin (\textit{RLN}, %
-\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000017411")})
-and prolactin (\textit{PRL}) are exceptionally high in the elephant
-when compared to other mammals
-(\cref{fig:rln2_evolution_tree,fig:prolactin_evolution_tree}). There
-is little to no placental expression of the estrogen synthase gene,
-\textit{CYP19A1}
-(\Sexpr{gene.expression("loxodonta_africana","CYP19A1")}) and the
-enzyme that produces progesterone, \textit{HSD3B1}
-(\Sexpr{gene.expression("loxodonta_africana","ENSLAFG00000027896")}). Other genes
-with high expression in the elephant placenta include genes involved
-in oxidative stress
-(\Sexpr{gene.expression("loxodonta_africana","TXN",complete=TRUE)}),
-growth
-(\Sexpr{gene.expression("loxodonta_africana","IGF2",complete=TRUE)})
-and
-\Sexpr{gene.expression("loxodonta_africana","PAEP",complete=TRUE)}, a
-gene predominantly expressed in the endometrium of the uterus in humans
-\cite{Rutanen.ea1987:Progesterone-associatedproteinsPP12and}.
-
-\subsubsection{\armadillo Placenta Transcriptome}
-
-% \begin{table}
-% \footnotesize
-% <<top.50.armadillo.expressed.genes,echo=FALSE,results="asis">>=
-% output.top.genes.table(top.genes.table(species.name="dasypus novemcinctus"))
-% @ 
-% \caption{\tableexplanation{\armadillo}{tab:all_dasypus_novemcinctus_placenta}}
-% \label{tab:top50_armadillo_placenta}
-% \end{table}
-
-The top 50 expressed genes in the term placenta of \armadillo are
-shown in Supplemental \cref{tab:all_dasypus_novemcinctus_placenta} with a list of all genes
-with $\text{FPKM}\ge 50$ in Supplemental
-\cref{tab:all_dasypus_novemcinctus_placenta}. The gene with the
-highest expression is
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045029",complete=TRUE)}
-which is most similar to the \mouse \textit{Pga5} is expanded in
-\opossum into three members which are expressed in placenta (the other
-two being
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033885",complete=TRUE)}
-and
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000040431",complete=TRUE)}),
-though the expansion is less than that seen \cow and \sheep
-(\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~D). There is also high expression of
-hemoglobin subunit beta,
-\Sexpr{gene.expression("dasypus_novemcinctus","HBB",complete=TRUE)},
-an un-annotated gene with high similarity to \human and \cow
-\textit{OVOS2}
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000033320",complete=TRUE)}
-(Supplemental \cref{fig:ovos2_evolution_tree}), the growth factors
-\Sexpr{gene.expression("dasypus_novemcinctus","IGF2",complete=TRUE)}
-and
-\Sexpr{gene.expression("dasypus_novemcinctus","IGFBP4",complete=TRUE)},
-and the thymosin
-\Sexpr{gene.expression("dasypus_novemcinctus","TMSB4X",complete=TRUE)}.
-As has been seen previously, multiple components of the annexin
-complex are also highly expressed, including
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000045634",complete=TRUE)}
-(\textit{ANXA8L1}),
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000017399",complete=TRUE)}
-(\textit{S100B}),
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000048504",complete=TRUE)}
-(\textit{S100A10}), and
-\Sexpr{gene.expression("dasypus_novemcinctus","ENSDNOG00000004237",complete=TRUE)}
-(\textit{ANXA2}).
-
-\subsection{Metatherians}
-
-\subsubsection{\opossum Placenta Transcriptome}
-
-% \begin{table}
-%   \footnotesize
-% <<top.50.opossum.expressed.genes,echo=FALSE,results="asis">>=
-% output.top.genes.table(top.genes.table(species.name="monodelphis domestica"))
-% @ 
-% \caption{\tableexplanation{\opossum}{tab:all_monodelphis_domestica_placenta}}
-% \label{tab:top50_opossum_placenta}
-% \end{table}
-
-The top 50 expressed genes as measured by \ac{FPKM} in the term
-placenta of the \opossum are shown in
-Supplemental \cref{tab:all_opossum_placenta} with a list of all genes in
-Supplemental \cref{tab:all_monodelphis_domestica_placenta}. The gene with the
-second highest expression is superoxide dismutase 3, extracellular
-(\Sexpr{gene.expression("monodelphis_domestica","SOD3",complete=TRUE)}),
-which functions as an antioxidant. The gene with the second highest
-expression is the beta hemoglobin subunit epsilon
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000011325",complete=TRUE)}),
-which is expressed at the embryonic stage of development in
-marsupials, as in other mammals
-\cite{Holland.ea1997:Characterizationofembryonicglobin}. \textit{HBE}
-subunits link up with protein products of two hemoglobin, zeta
-(\textit{HBZ}) genes, which are also expressed to a high degree in the
-opossum placenta
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016635",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016631",complete=TRUE)}).
-In addition, there is high expression of the gene hemoglobin, omega
-(\textit{HBO},
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016637",complete=TRUE)}),
-% this is LOC_1000016060 which I figured out by tblastn'ing AAK11486.1
-% against Monodelphis domestica
-a hemoglobin found only in marsupials and monotremes
-\cite{Opazo.ea2008:Genomicevidenceforindependent} (Supplemental
-\cref{fig:hbe1_evolution_tree}). The hemoglobin gene \textit{HBA} is
-also expressed in the placenta
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000016633",complete=TRUE)}).
-Two cystatin genes
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018295",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000018296",complete=TRUE)})
-are expressed to a high degree, along with one \textit{CTSL} related
-gene
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000002844",complete=TRUE)},
-\cref{fig:pag2_prss16_ctsl_psg_evolution_tree}~B), which are known to interact with
-cystatins \cite{Turk.ea2001:Lysosomalcysteineproteasesfacts}. The
-uncharacterized gene \textit{LOC10001195}
-(\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000006547",complete=TRUE)})
-also shows high expression in \opossum placenta. The gene
-\textit{WFDC2}, a member of the whey acidic protein (WAP) family is
-also expressed to a high degree in \opossum placenta
-(\Sexpr{gene.expression("monodelphis_domestica","WFDC2",complete=TRUE)}).
-Other genes with high expression include members of the annexin
-complex
-(\Sexpr{gene.expression("monodelphis_domestica","S100A6",complete=TRUE)},
-\Sexpr{gene.expression("monodelphis_domestica","ANXA2",complete=TRUE)}).
-In addition, there was little production of CYP19A1 and HSD3B1
-(\Sexpr{gene.expression("monodelphis_domestica","CYP19A1",complete=TRUE)}
-and
-\Sexpr{gene.expression("monodelphis_domestica","ENSMODG00000005385",complete=TRUE)},
-respectively).
-
 \subsection{Core Placenta Transcriptome}
@@
 
-\subsubsection{\ac{GSE} Analyses of Individual Transcriptomes}
+\subsection{\ac{GSE} Analyses of Individual Transcriptomes}
 
@@
 \subsection{Evolution of Placental Hormones}
-
-Reproductive hormones are central to the regulation of pregnancy and
-their source, timing, and quantity can vary widely from species to
-species \cite{Carter2012:Evolutionofplacentalfunction}. Expression and synthesis of hormones is
-perhaps the most well-documented physiological process in the placenta
-in many species. Here we quantified the extreme disparity in genes
-with hormone products or enzymatic products that directly produce
-hormones in multiple mammalian species (\cref{fig:hormone_genes}).
-
-\begin{figure}[htb]
-<<hormone.genes.plot,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-gene.expression.nozero <- function(genes) {
-    temp <-
-        data.matrix.dt(dcast(combined.fpkm[human_name %in% genes,],
-                             human_name~species,
-                             value.var="FPKM",
-                             fun.aggregate=function(x){x[1]}))
-    temp <-
-        temp[apply(temp,1,function(x){sum(x,na.rm=TRUE)})>5,]
-    return(temp)
-}
-
-### hormone genes
-hormone.genes <- c("CGA","CGB","CGB1","CGB2","CGB5","CGB8","CSH1","CSH2","CSHL1","CYP19A1",
-                   "GH2","GHRH","HSD3B1","HSD3B2","HSD17B2","HSD17B1","NPPB","NPPC","PRL","RLN1")
-hormone.expression.nozero <-
-    gene.expression.nozero(hormone.genes)
-
-hor.exp.long <- melt(log10(hormone.expression.nozero+1)[rev(1:nrow(hormone.expression.nozero)),])
-colnames(hor.exp.long)[1:3] <- c("gene","species","fpkm")
-hor.exp.long <- hor.exp.long[hor.exp.long$gene!="IGF2",]
-
-### galectin genes
-galectin.genes <- unique(grep("^LGALS",combined.fpkm[,human_name],value=TRUE))
-
-galectin.expression.nozero <-
-    gene.expression.nozero(galectin.genes)
-
-gal.exp.long <- melt(log10(galectin.expression.nozero+1)[rev(1:nrow(galectin.expression.nozero)),])
-colnames(gal.exp.long)[1:3] <- c("gene","species","fpkm")
-
-### IGF related genes
-
-igfrelated.genes <- unique(grep("^IGF",combined.fpkm[,human_name],value=TRUE))
-
-igfrelated.expression.nozero <-
-    gene.expression.nozero(igfrelated.genes)
-
-igf.exp.long <- melt(log10(igfrelated.expression.nozero+1)[rev(1:nrow(igfrelated.expression.nozero)),])
-colnames(igf.exp.long)[1:3] <- c("gene","species","fpkm")
-
-### Pre-eclampsia genes ADAM12 CRH KISS1 PAPPA
-
-pe.expression.nozero <-
-    gene.expression.nozero(c("ADAM12","CRH","KISS1","PAPPA"))
-
-pe.exp.long <- melt(log10(pe.expression.nozero+1)[rev(1:nrow(pe.expression.nozero)),])
-colnames(pe.exp.long)[1:3] <- c("gene","species","fpkm")
-
-igf.exp.long$type <- "IGF Related"
-gal.exp.long$type <- "Galectin"
-hor.exp.long$type <- "Hormone"
-pe.exp.long$type <- "PE"
-
-combined.long <-
-    rbind(igf.exp.long,
-          gal.exp.long,
-          hor.exp.long,
-          pe.exp.long
-          )
-combined.long$species <-
-    factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(combined.long$species)),
-           levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(names(species.names))))
-
-common.plot.options <- 
-    list(geom_tile(aes(fill = fpkm), colour = "white"),
-         scale_fill_gradient(low = grad.def[["low"]],
-                             high = grad.def[["high"]],
-                             space = grad.def[["space"]],
-                             na.value = grad.def[["na.value"]],
-                             guide=guide_colorbar(title=expression(log[10](FPKM)))),
-         scale_x_discrete("", expand = c(0, 0)),
-         scale_y_discrete("", expand = c(0, 0)),
-         theme_grey(base_size = 9),
-         theme(legend.position = "top",
-               axis.ticks = element_blank(),
-               panel.background = element_blank(),
-               axis.text.y = element_text(face="italic"),
-               axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic")),
-         facet_grid(type~.,scale="free_y",space="free_y")
-         )
-print(ggplot(combined.long, aes(y=gene, x=species))
-      + common.plot.options
-      )
-@ 
-\caption{Expression in $\text{log}_{10}(\text{FPKM})$ of 1:1 \human
-  orthologs of Hormone Genes, Galectins, genes implicated in
-  pre-eclampsia (PE), and \textit{IGF}-related genes. Empty locations
-  indicate no detectable expression and/or no 1:1 \human ortholog in
-  that species.}
-\label{fig:hormone_genes}
-\end{figure}
 
@@
 
-\subsection{The Opossum Placenta and the Evolution of Viviparity}
-
-The placenta of \opossum is distinctly different from that of the term
-placenta of eutherian mammals in that it is derived from both the
-chorion and yolk sac. Other differences between the placentas of
-marsupials and eutherians include lack of invasive extravillous
-trophoblasts in marsupials, the retention of a permeable shell coat
-for 2/3 of marsupial gestation, and the greater degree of imprinting
-in placenta-expressed genes in eutherians
-\cite{Renfree2010:ReviewMarsupialsplacentalmammals,Renfree.ea2013:originandevolutionof}.
-In addition, the placenta of \opossum is also distinct from other
-marsupials, in that it is considered one of the least specialized, and
-thought be closer to the original marsupial condition
-\cite{Renfree2010:ReviewMarsupialsplacentalmammals}. In \opossum, the
-duration of direct connection between maternal and fetal tissues is as
-short as two days in a gestational period of 13-15 days
-\cite{Harder.ea1993:Gestationandplacentationin}. During this period,
-the placenta is directly engaged in the exchange of urea with maternal
-tissues, owing to the lack of a vesicular allantois
-\cite{Freyer.ea2007:Placentalfunctionintwo}. This differs drastically
-with the situation in another well-studied marsupial, the tammar
-wallaby \cite{Freyer.ea2007:Placentalfunctionintwo}. Therefore, we
-must interpret differences between the opossum and eutherian placentas
-with caution, as highly expressed genes and functional annotation
-clusters may represent didelphid or \opossum specific adaptations.
-
-The \opossum placenta shows expression of genes that either do not
-exist in eutherians or are not expressed in the term eutherian placenta
-(Supplemental \cref{tab:all_opossum_placenta}). The existence of embryonic forms
-of hemoglobin in fetal tissues at term, as evidenced by the expression
-of \textit{HBE} and \textit{HBZ} is a feature of marsupials that is
-generally due to their very short gestation periods. Embryonic
-hemoglobins in marsupials have a lower affinity for oxygen and allow
-for a lower rate of oxygen uptake from maternal tissues
-\cite{Holland.ea1997:Characterizationofembryonicglobin}. One novel
-uncharacterized gene that shows high expression (\textit{LOC10001195}) is not
-found in eutherians; Ensembl gene trees reveal that homologs of this
-gene exist in other amniotes such as birds and turtles, but are not
-present in eutherian mammals. The homolog of this gene in chicken is
-found to expressed in reproductive tract
-\cite{Froman.ea2006:expressedsequencetaganalysis}, but its function is
-unknown. \textit{WFDC2} is a member of the whey acidic protein (WAP)
-family and involved in innate immunity. It is expressed in the mammary
-gland of the tammar wallaby during pregnancy and early lactation
-\cite{Sharp.ea2007:Molecularevolutionofmonotreme}, and present in
-relative abundance in \opossum placenta.
-
-\subsection{Novel Gene Expansions in the Elephant Placenta}
-
-We found evidence of high expression in two novel expansions of genes
-in the \elephant genome. One family includes 11 genes that are related
-to the \textit{PRSS16} gene, a thymus-specific serine protease that
-may be involved regulating the presentation of self-peptides in CD4+
-T-lymphocytes \cite{Fornari.ea2012:TranscriptionprofilingofPrss16}.
-All 11 of these genes are expressed in the elephant placenta, three of
-which have very abundant mRNAs
-(Supplemental \cref{tab:all_loxodonta_africana_placenta} and \cref{fig:pag2_prss16_ctsl_psg_evolution_tree}C). It is
-possible that this elephant-specific gene family may be involved in
-immunity in the elephant placenta and the mediation of maternal-fetal
-immune tolerance; however, future work will need to test these
-hypotheses. Three other genes abundantly expressed in the elephant
-placenta are homologues of \textit{SSLP1}, a gene with predominant
-expression in the seminal vesicle in mouse
-\cite{Li.ea2006:SSLP-1secretedLy-6protein}. Homologues of
-\textit{SSLP1} are also expressed in \cow, \dog, \horse, and \sheep.
-\textit{SSLP1} also contains an SINE repeat and may be
-involved in the development of villi via cell proliferation
-\cite{Ushizawa.ea2009:Characterizationandexpressionanalysis}. This
-gene is not found in primates. Expansion of these genes with high
-expression in \elephant may indicate an elaboration of this system.
-
-% Bovine has their own alu-like sequence: PMID 7835712
-
 \subsection{Core placenta transcriptome}

### 2016-09-09 manuscript/manuscript.Rnw
remove discussion of placental hormones

@@
 
-\subsection{Evolution of Placental Hormones}
-
-\subsubsection{Estrogen and Progesterone}
-
-In both ruminants and primates, the placenta is capable of producing
-estrogen and progesterone
-\cite{Strauss.ea1996:Placentalsteroidhormonesynthesis}. High relative
-expression of \textit{CYP19A1} (Supplemental
-\cref{fig:estrogen_evolution_tree}), the enzyme that synthesizes
-estrogen from androstenedione is seen in \human, \bonobo, \cow, and
-\horse, though not in \sheep, \smonkey. %
-%
-\textit{HSD3B1}, one of two enzymes that synthesizes progesterone from
-pregnalone
-(\cite{Simard.ea2005:Molecularbiologyof3beta-hydroxysteroid},
-Supplemental \cref{fig:progesterone_evolution_tree}) is also highly
-expressed in \human, \bonobo, \cow, \horse, and \dog, but is not
-highly expressed in \smonkey, \sheep, \pig, \elephant, \armadillo, or
-\opossum. This is consistent with the previously observed local
-progesterone withdrawal in \sheep, and the leuteo-placental shift of
-progesterone expression in boreoeutheria in contrast to afrotherians
-which do not show such a shift
-\cite{Swaggart.ea2015:Genomicsofpretermbirth}. The diversity of
-\textit{CYP19A1} and \textit{HSD3B1} expression levels in in primates
-and ungulates seen here points to the need for more complete sampling
-of placentas at different timepoints and in different clades.
-
-\subsubsection{Prolactin}
-
-The elephant is strikingly the only mammal analyzed that shows high
-expression of the primarily pituitary hormone prolactin (\textit{PRL})
-in the term placenta, although we see high expression of genes which
-resulted from duplications of \textit{PRL} in the \mouse, \carmeli,
-and \cow (Supplemental \cref{fig:prolactin_evolution_tree}). The
-results in \mouse are relatively consistent with the temporal
-expression results of
-\citet{Simmons.ea2008:Spatialandtemporalexpression}. For example, we
-see very high expression of \textit{Prl3b1} and \textit{Prl8a9} in the
-term placentas of \mouse, but low expression of \textit{Prl7a1} and
-\textit{Prl8a2} which are expressed at earlier embryonic stages. %
-All afrotherians yet analyzed contain only one prolactin gene, and
-prolactin has undergone adaptive evolution on the \elephant lineage
-\cite{Wallis2009:ProlactininAfrotheriacharacterization}.
-\citet{Yamamoto.ea2011:PlacentationinAfricanelephant} noted the
-presence of a prolactin in the trophoblast of the elephant placenta
-using immunohistological staining and human derived antibodies. A
-previous analysis of \elephant placenta also detected the expression
-of prolactin \cite{Hou.ea2012:Elephanttranscriptomeprovidesinsights}.
-Here we confirm that the sole prolactin in \elephant is not only
-expressed but highly expressed in the placenta, and likely functions
-both in the maintenance of the corpus luteum and as a placental
-lactogen \cite{Yamamoto.ea2011:PlacentationinAfricanelephant}.
-\textit{PRL} is not expressed in \opossum, even though there is
-evidence that it is expressed in the placenta of another marsupial,
-\textit{Macropus eugenii} (tammar wallaby)
-\cite{Menzies.ea2011:Placentalexpressionofpituitary}. This suggests
-that placental prolactin expression in marsupials may be as diverse as
-in eutherian mammals. One possible explanation of the diversity of
-expression is the presence of endogenous retroviruses such as MER39 in
-primates, MER77 in rodents, and L1-2\_LA in \elephant
-\cite{Emera.ea2012:Convergentevolutionofendometrial} although this
-does not explain the diversity of expression of different isoforms of
-\textit{PRL} in \mouse or their tissue specificity in primates. An
-alternative, that \textit{HOXA11} expression is responsible for
-\textit{PRL} expression
-\cite{Lynch.ea2008:Adaptivechangesintranscription} seems not to be the
-case in placenta, as the rodent with the highest \textit{HOXA11}
-expression, \galili
-(\Sexpr{gene.expression("spalax_galili","Hoxa11",complete=TRUE)}) is
-the rodent with the lowest expression of \textit{PRL}
-(\cref{fig:prolactin_evolution_tree}), and \elephant also does not
-express \textit{HOXA11} highly either
-(\Sexpr{gene.expression("loxodonta_africana","HOXA11",complete=TRUE)},
-Supplemental \cref{fig:hoxa11_evolution_tree}).
-
-\subsubsection{Growth hormone family}
-
-The growth hormone family of genes has expanded in anthropoid
-primates, resulting in approximately 5 genes in \human (\textit{GH1},
-\textit{GH2}, \textit{CSH1}, \textit{CSH2}, \textit{CSHL1}) and at
-least three in \bonobo (two \textit{GH2} and \textit{GH1}; owing to
-the similarity between the two \textit{GH1} in \textit{P. troglodytes}
-we cannot reliably distinguish transcripts between the two). All show
-substantial expression in the \human placenta
-(\cref{fig:csh1_evolution_tree}). The expression of the \ac{GH} family
-in placenta appears to be primate specific, and likely existed at the
-time of the last common ancestor between all extant anthropoids
-\cite{Papper.ea2009:Ancientoriginofplacental,Soares2004:prolactinandgrowthhormone}.
-There is some evidence that increased expression of these genes may be
-correlated with overall birth weight of offspring in \human
-\cite{Männik.ea2010:Differentialexpressionprofileof}, and it is
-possible that the extreme expression levels of these genes may be due
-to high birth weight. \human placental \ac{GH} genes are involved in
-increasing blood flow, replacing maternal pituitary expression of
-\ac{GH}, as well as trophoblast invasion
-\cite{Alsat.ea1998:Physiologicalroleofhuman,Lacroix.ea2005:Stimulationofhumantrophoblast,Schiessl.ea2007:Roleofplacentalgrowth},
-and it is likely that they have the same function in \bonobo.
-
-\begin{figure}[htb]
-<<chorionic.somatomammotropin.alpha.evolution.tree,echo=FALSE,fig.width=4,fig.height=4,warning=FALSE,message=FALSE>>=
-
-pushViewport(viewport(layout=grid.layout(nrow=1,ncol=2,widths=c(2,3))))
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CGA" &
-                                                    species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.4)+
-      theme(legend.position="none"),
-      vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"))
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CGB1" &
-                                                    species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                subtree=c("ENSMODG00000013768","ENSG00000104818"),
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.5),
-      vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"))
-popViewport(1)
-grid.text(c("A","B"),
-          x=unit(c(0,2/5),"npc"),y=unit(c(1,1),"npc"),
-          just=c("left","top"),
-          gp=gpar(fontsize="24")
-          )
-@ 
-\caption{\genetreeexplanation{(A) \textit{CGA} (B) \textit{CGB}}}
-\label{fig:cga_cgb_evolution_tree}
-\end{figure}
-
-The human gene \textit{CGA} codes for the alpha subunit of multiple
-hormones including pituitary hormones (luteinizing hormone,
-follicle-stimulating hormone, thyroid-stimulating hormone), as well as
-numerous chorionic gonadotropins with variable beta subunits
-\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. The function of
-the chorionic gonadotropins is to prevent the regression of the corpus
-luteum \cite{Hanson.ea1971:EffectsofHCGand}. Among the six placentas
-we analyzed here, \textit{CGA} is only expressed in the placentas of
-\human and \bonobo (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}),
-which corresponds with the emergence of the chorionic gonadotropins in
-anthropoid primates
-\cite{Maston.ea2002:Chorionicgonadotropinhasrecent}. However, oddly,
-\textit{CGA} is expressed at much higher levels in both species than
-the beta subunits with which its product dimerizes (\cref{fig:hormone_genes,fig:cga_cgb_evolution_tree}).
-\textit{CGA} is one of several genes noted by
-\citet{Uddin.ea2008:Distinctgenomicsignaturesof} to have undergone
-adaptive evolution of the \human{}-\textit{P. troglodytes} stem lineage.
-
-\subsubsection{Relaxin}
-
-Relaxin (\textit{RLN}) is a hormone that has multiple effects during
-pregnancy including increasing maternal blood flow, widening the pubic
-bone and relaxing uterine musculature in preparation for labor. In
-\dog, \elephant, \mouse, \carmeli, and \horse there is high expression
-of relaxin in the term placenta, with little evidence in other species
-examined here (Supplemental \cref{fig:rln2_evolution_tree}). \textit{RLN} is known
-to be produced in the placenta of the pregnant \dog and is used in the
-diagnosis of \dog pregnancy
-\cite{Steinetz.ea1989:Serumrelaxinandprogesterone,Tsutsui.ea1991:Determinationofsourceof},
-the luteal ovary and placenta of \pig
-\cite{Hisaw.ea1948:Relaxininovaryof}, and the placenta of \horse
-\cite{Klonisch.ea1997:Placentallocalizationofrelaxin}. However, the
-source of relaxin produced by pregnant \elephant remains unclear
-\cite{Meyer.ea2004:Speciesandfetalgender}. The high rise in \textit{RLN} in
-\elephant has been suggested to be a major signal for the initiation
-of labor \cite{Niemuller.ea1998:Plasmaconcentrationsofimmunoreactive}.
-Here we show that the placenta of elephants show high expression of
-\textit{RLN} and is likely a major source of \textit{RLN} at term. In
-contrast to the results of
-\cite{Evans.ea1993:mouserelaxingenenucleotide}, we also see expression
-of \textit{RLN} in the placenta of \mouse and \carmeli.
-
-\subsubsection{\textit{NPPB} and \textit{NPPC}}
-
-Two genes, \textit{NPPB} and \textit{NPPC}, code for peptide hormones
-that are important in the inducement of vasodilation, and may be
-important in increasing maternal blood flow to the fetus
-\cite{Cameron.ea1996:sitesofgeneexpression}. There is some indication
-that these proteins are present in the decidua of mice
-\cite{Cameron.ea1996:sitesofgeneexpression}, but we find very little
-expression in the fetal portion of the mouse placenta
-(\cref{fig:hormone_genes}, Supplemental
-\cref{fig:nppc_evolution_tree,fig:nppb_evolution_tree}). While we do
-find high expression of at least one of these genes in \dog, \cow, and
-\elephant, the expression pattern does not follow the placental
-organization (endotheliochorial for \dog and \elephant,
-synepitheliochorial for \cow and \sheep), so it is unlikely that
-expression of these genes are essential for specific placental
-organizations.
-
-\subsection{The Evolution of Placental Expression of Galectins}
-
-Galectins are recognized as major regulators of pregnancy in both
-fetal and maternal tissues
-\cite{Than.ea2012:Galectinsguardiansofeutherian}. Indeed, the
-evolution of new lineage-specific galectins has occurred in both
-primates and ruminants, with the genes \textit{LGALS13},
-\textit{LGALS14}, and \textit{LGALS16} recognized as being
-placenta-specific in \human and regulating immune responses by
-inducing apoptosis of maternal T-cells
-\cite{Than.ea2009:primatesubfamilyofgalectins} (Supplemental
-\cref{fig:galectins_lgals13_evolution_tree}). Indeed, \textit{LGALS14}
-has the highest expression of any galectin in the \human and \smonkey
-placenta (\cref{fig:hormone_genes}, Supplemental
-\cref{fig:galectins_lgals13_evolution_tree}), and other primate
-galectins such as \textit{LGALS13} and \textit{LGALS16} are also
-expressed in \human and \smonkey placenta. However, in the closely
-related \bonobo, \textit{LGALS14} is not found, and \textit{LGALS1} is
-the most abundant galectin. \textit{LGALS1} shows generally high
-expression in all placentas analyzed. \textit{LGALS1} is also involved
-in mediating maternal-fetal immune tolerance, and is expressed in both
-maternal and fetal tissue in primates
-\cite{Than.ea2008:Emergenceofhormonaland}. \textit{LGALS1} placental
-expression was thought to be confined to eutherian mammals, as there
-is evidence of adaptive evolution at this node, as well as the
-appearance of estrogen response elements in eutherian cis-regulatory
-regions \cite{Than.ea2008:Emergenceofhormonaland}. It was unclear as
-to whether galectins such as \textit{LGALS1} were important in the
-regulation of pregnancy in marsupials, as well as eutherians. Here we
-show substantial expression of galectins in the opossum, especially
-\textit{LGALS1} (Supplemental \cref{fig:galectins_evolution_tree}). In
-addition, \textit{LGALS2} is also expressed. The presence of galectins
-at the maternal-fetal interface in \opossum implies that it was also
-present in the ancestral placenta of therian mammals. Indeed, adaptive
-evolution of \textit{LGALS1} on the eutherian stem is likely due to
-the increased invasive nature of eutehrian placentation rather than
-the origination of maternal-fetal tolerance. Although the function of
-\textit{LGALS1} in the marsupial placenta needs to be confirmed, it is
-likely that the mediation of maternal-fetal immune tolerance by
-galectins is an ancient feature of mammalian viviparity.
-
 \subsection{Core placenta transcriptome}

### 2016-09-10 data/go_analysis.R
add rules to calculate placenta class specific analyses

@@ -9,8 +9,7 @@ args <- commandArgs(trailingOnly=TRUE)
 
 load(args[1])
 
-data <- eval(parse(text=args[2]))
-
+data <- as.factor(eval(parse(text=args[2])))
 godata <- new("topGOdata",
               ontology=args[3],
               allGenes=data,

### 2016-09-10 data/placenta_classification_specific.R

@@ -1,0 +1,73 @@
+library("data.table")
+library("reshape2")
+library("org.Hs.eg.db")
+
+args <- c("combined_fpkm","housekeeping_genes_superset",
+          "placental_classification.txt","placenta_classification_specific")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+load(args[1])
+load(args[2])
+placenta_classification <- fread(args[3])
+
+combined.fpkm.wide <- 
+    data.table(dcast(combined.fpkm,
+                     human_name~species,
+                     fun.aggregate=sum,
+                     value.var="mean_fpkm"))
+
+### ignore genes which aren't expressed at least 10 FPKM in some
+### tissue and have FPKM >= 1 in more than 1 species
+placenta.class.specific <-
+    combined.fpkm.wide[apply(combined.fpkm.wide[,-1,with=FALSE],1,max) >= 10 &
+                       apply(combined.fpkm.wide[,-1,with=FALSE],1,function(x){sum(x >= 1)}) > 1 &
+                       !is.na(human_name),]
+
+### ignore genes which are housekeeping genes
+placenta.class.specific <-
+    placenta.class.specific[!(human_name %in% housekeeping.genes.superset[,gene_short_name]),]
+
+placenta.class.specific <-
+    data.table(melt(placenta.class.specific,id.vars="human_name",
+                    variable.name="species",
+                    value.name="mean_fpkm"))
+setkey(placenta.class.specific,
+       "species")
+setkey(placenta_classification,"species")
+placenta_classification <-
+    placenta.class.specific[placenta_classification]      
+
+calculate.aov <- function(gene,factor="barrier") {
+    summary(aov(glm(as.formula(paste0("mean_fpkm~",factor,"+0")),
+                    data=placenta_classification[human_name==gene,]))
+            )[[1]][factor,"Pr(>F)"]
+}
+
+placenta.classification.p <-
+    data.table(data.frame(shape=sapply(placenta_classification[,unique(human_name)],
+                                       calculate.aov,factor="shape"),
+                          barrier=sapply(placenta_classification[,unique(human_name)],
+                                         calculate.aov,factor="barrier"),
+                          interdigitation=sapply(placenta_classification[,unique(human_name)],
+                                                 calculate.aov,factor="interdigitation")),
+               keep.rownames=TRUE
+               )
+
[…]

### 2016-09-10 manuscript/manuscript.Rnw
update methods

@@
 
-  \textbf{Methods}
-
-  \textbf{Results} We measured the expression profiles of
-  \Sexpr{star.logs[,length(unique(file))]} term placentas of
-  \Sexpr{star.logs[,length(unique(species))]} species of mammals 
-  also identified genes which are expressed in all placentas,
+  \textbf{Method} Using RNA-seq, we measured the expression profiles
+  of \Sexpr{star.logs[,length(unique(file))]} term placentas of
+  \Sexpr{star.logs[,length(unique(species))]} species of mammals.
+
+  \textbf{Results} We identified a set of
+  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
+  core genes which are expressed (FPKM $\ge 10$) in all placentas.
+
   including genes with immune-modulating properties (\textit{ANXA2},

### 2016-09-12 manuscript/manuscript.Rnw
load placental_classification

@@
 grad.def <- list(low = "gray90", high = "blue", space = "Lab", na.value = NA)
+
+@
+
+<<placental_classification,echo=FALSE,results="hide",cache.extra=file.info(dir("../data","placenta.*classification.*",full.names=TRUE))[,"mtime"]>>=
+placenta_classification <- fread("../data/placental_classification.txt")
+load("../data/placenta_classification_p")
+for (mf.bp in c("mf","bp")) {
+    for (class in c("interdigitation","barrier","shape")) {
+        load(paste0("../data/placenta_classification_go_",mf.bp,"_",class))
+    }
+}
 

### 2016-09-12 manuscript/manuscript.Rnw
rewrite introduction

@@
 \section{Introduction}
+
+The placenta is arguably one of the most variable organs in different
+species of mammals in terms of the shape of attachment
+(\Sexpr{english.join(placenta_classification[,levels(as.factor(shape))])}),
+degree of intimacy between fetal tissues and maternal tissues
+(\Sexpr{english.join(placenta_classification[,levels(as.factor(intimacy))])}),
+and degree of interdigitation of maternal and fetal tissues
+(\Sexpr{english.join(placenta_classification[,levels(as.factor(interdigitation))])})
+while its primary functions (transfer of nutrients and oxygen to the
+fetus, removal of wastes and carbon dioxide) are strongly conserved.
 

### 2016-09-12 manuscript/supplemental_material.Rnw
add placenta classification tables

@@
 
+<<placental_classification,echo=FALSE,results="hide",cache.extra=file.info(dir("../data","placenta.*classification.*",full.names=TRUE))[,"mtime"]>>=
+placenta_classification <- fread("../data/placental_classification.txt")
+load("../data/placenta_classification_p")
+for (mf.bp in c("mf","bp")) {
+    for (class in c("interdigitation","barrier","shape")) {
+        load(paste0("../data/placenta_classification_go_",mf.bp,"_",class))
+    }
+}
+
+@
+
 \begin{table}[hb]
-<<placental_classification,echo=FALSE,results="asis">>=
-placental.classification <- fread("../data/placental_classification.txt")
+<<placental_classification_table,echo=FALSE,results="asis">>=
 print(xtable(placental.classification,
@@
 
+\section{Placenta Classification Analyses}
+
+<<placenta_class_table,echo=FALSE,results="hide">>=
+placenta_class_table <- function(class="shape") {
+    subtable <- placenta.classification.p[,c("rn",class,paste0(class,".fdr")),with=FALSE]
+    setnames(subtable,c("Gene","p","FDR"))
+    subtable <- head(subtable[FDR <= 0.05,][order(p),],n=50)
+    print(xtable(subtable,
+                 align=c("c","c","r","r"),
+                 display=c("s","s","g","g")
+                 ),
+          booktabs=TRUE,
+          floating=FALSE,
+          include.rownames=FALSE,
+          math.style.exponents=TRUE,
+          size="\\tiny")
+}
+
+@
+
+\subsection{Attachment Shape}
+
+\begin{table}
+<<attachment_shape_significant,echo=FALSE,results="asis">>=
+placenta_class_table("shape")
+@ 
+\caption{Top 50 genes which are associated with the shape of placenta attachment}
+\label{tab:shape_genes}
+\end{table}
+
+\subsection{Maternal-Fetal Intimacy}
+
+<<intimacy_significant,echo=FALSE,results="asis">>=
+placenta_class_table("barrier")
+@ 
+\caption{Top 50 genes which are associated with the degree of placenta intimacy}
+\label{tab:barrier_genes}
+\end{table}
+
+\subsection{Maternal-Fetal Interdigitation}
+
+<<interdigitation_significant,echo=FALSE,results="asis">>=
+placenta_class_table("interdigitation")
+@ 
+\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation}
+\label{tab:interdigitation_genes}
+\end{table}
+
 \section{Gene Ontology Analyses}

### 2016-09-19 manuscript/manuscript.Rnw
move classification table to supplemental material

@@
 
-\begin{table}
-  \small
-  \begin{tabular}{>{\raggedright }p{0.7in}
-                  >{\raggedright }p{0.7in}
-                  c c 
-                  p{0.65in}}
-    \toprule
-    {\centering Species  }                     & {\centering Common Name }& 
-       \multicolumn{1}{p{0.43in}}{\raggedright \centering Gestation (days)} & 
-       \multicolumn{1}{p{0.3in}}{\raggedright \centering Litter Size}  & Source \\
-    \midrule
-    \textit{Ateles fusciceps}     & Spider monkey & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Bos taurus}           & Cow & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Canis familiaris}     & Dog & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Dasypus novemcinctus} & Nine-banded Armadillo & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
-    \textit{Equus caballus}       & Horse & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
-    \textit{Loxodonta africana}   & African savanna elephant & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Homo sapiens}         & Human & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Monodelphis domestica}& Gray short-tailed opossum & 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
-    \textit{Mus musculus}         & Mouse & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Spalax galili}   & Blind mole rat from Upper Galilee & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
-    \textit{Ovis aries}           & Domestic sheep & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Pan paniscus}         & Bonobo & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
-    \textit{Spalax carmeli}       & Blind mole rat from northern Israel & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
-    \textit{Sus scrofa}           & Domestic pig (Duroc) & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \bottomrule
-  \end{tabular}
-  \caption{Gestational periods in days and average liter sizes for studied species.}
-  \label{tab:gestational_periods}
-\end{table}
+Even though the placenta's primary function is conserved among
 

### 2016-09-19 manuscript/supplemental_material.Rnw
move classification table to supplemental material

@@
 
+\begin{table}
+  \small
+  \begin{tabular}{>{\raggedright }p{0.7in}
+                  >{\raggedright }p{0.7in}
+                  c c 
+                  p{0.65in}}
+    \toprule
+    {\centering Species  }                     & {\centering Common Name }& 
+       \multicolumn{1}{p{0.43in}}{\raggedright \centering Gestation (days)} & 
+       \multicolumn{1}{p{0.3in}}{\raggedright \centering Litter Size}  & Source \\
+    \midrule
+    \textit{Ateles fusciceps}     & Spider monkey & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Bos taurus}           & Cow & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Canis familiaris}     & Dog & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Dasypus novemcinctus} & Nine-banded Armadillo & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
+    \textit{Equus caballus}       & Horse & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
+    \textit{Loxodonta africana}   & African savanna elephant & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
+    \textit{Homo sapiens}         & Human & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Monodelphis domestica}& Gray short-tailed opossum & 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
+    \textit{Mus musculus}         & Mouse & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Spalax galili}   & Blind mole rat from Upper Galilee & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
+    \textit{Ovis aries}           & Domestic sheep & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \textit{Pan paniscus}         & Bonobo & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
+    \textit{Spalax carmeli}       & Blind mole rat from northern Israel & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
+    \textit{Sus scrofa}           & Domestic pig (Duroc) & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
+    \bottomrule
+  \end{tabular}
+  \caption{Gestational periods in days and average liter sizes for studied species.}
+  \label{tab:gestational_periods}
+\end{table}
+
 \subsection{Sequence Source}

### 2016-09-19 manuscript/manuscript.Rnw
load rsqlite

@@
 library("ape")
+library("EBImage")
 library("ggtree")
+library("RSQLite")
 library("topGO")

### 2016-09-19 manuscript/supplemental_material.Rnw
load rsqlite

@@
 library("ggtree")
+library("RSQLite")
 library("topGO")

### 2016-09-19 manuscript/manuscript.Rnw
rewrite starting introduction paragraph

@@
 
-The placenta is arguably one of the most variable organs in different
-species of mammals in terms of the shape of attachment
-(\Sexpr{english.join(placenta_classification[,levels(as.factor(shape))])}),
-degree of intimacy between fetal tissues and maternal tissues
-(\Sexpr{english.join(placenta_classification[,levels(as.factor(intimacy))])}),
-and degree of interdigitation of maternal and fetal tissues
-(\Sexpr{english.join(placenta_classification[,levels(as.factor(interdigitation))])})
-while its primary functions (transfer of nutrients and oxygen to the
-fetus, removal of wastes and carbon dioxide) are strongly conserved.
-
 Even though the placenta's primary function is conserved among
+mammals, it varies widely in terms of the shape of attachment, degree
+of intimacy between fetal tissues and maternal tissues and degree of
+interdigitation of maternal and fetal tissues in different species of
+mammals
+\cite{Mossman.1987:vertebratefetalmembranes,Moffett.ea2006:Immunologyofplacentationin}
+(Supplemental \cref{tab:placental_classification}). Mammals also
+differ drastically in the number and size of their offspring, as well
+as the length of gestation; these traits are often with specific
+aspects of placental morphology
+\cite{Capellini.ea2011:Placentationandmaternalinvestment}
+(Supplemental \cref{tab:gestational_periods}).
 

### 2016-09-19 data/placenta_classification.R
build placenta classification

@@ -1,0 +1,32 @@
+library("data.table")
+
+
+args <- c("../data/placental_classification.txt","placental_classification")
+args <- commandArgs(trailingOnly=TRUE)
+
+placenta_classification <- fread(args[1])
+
+placenta_classification[,species:=factor(species)]
+placenta_classification[,shape:=factor(shape)]
+placenta_classification[,interdigitiation:=
+                             factor(interdigitation,
+                                    levels=c("villous",
+                                             "folded",
+                                             "trabecular",
+                                             "lamellar",
+                                             "labrynthine"
+                                             ),
+                                    ordered=TRUE
+                                    )]
+
+placenta_classification[,barrier:=
+                             factor(barrier,
+                                    levels=c("epitheliochorial",
+                                             "endotheliochorial",
+                                             "hemochorial"
+                                             ),
+                                    ordered=TRUE
+                                    )]
+
+save(placenta_classification,
+     file=args[length(args)])

### 2016-09-19 manuscript/manuscript.Rnw
add questions/previous studies to introduction

@@
 \cite{Capellini.ea2011:Placentationandmaternalinvestment}
-(Supplemental \cref{tab:gestational_periods}).
+(Supplemental \cref{tab:gestational_periods}). Although the morphology
+of the mammalian placenta has been well characterized in a number of
+species \cite{Mossman.1987:vertebratefetalmembranes}, the diversity of
+the molecular environment is only beginning to be understood from a
+genetic perspective \cite{Carter2012:Evolutionofplacentalfunction}.
+
+These observations lead to two related questions: 1) what are the
+genes which are responsible for the \emph{core functionality of the
+  placenta} and 2) which genes are associated with \emph{changes in
+  placenta morphology}?
+
+Previous studies of some species (i.e., \textit{Homo sapiens},
+\textit{Mus musculus}, \textit{Bos taurus}) have provided some
+preliminary answers to these questions. They revealed a stunning
+divergence in the genetics of the placenta, especially in later stages
+of pregnancy
+\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
+For example, each lineage has evolved expansive placenta-specific gene
+families with expression patterns that can vary in space and time
+\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
+These include independent expansions of the prolactins in \mouse,
+\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG}
+in \human and \mouse, \acp{PAG} in \cow, genes related to the
+thymus-specific serine protease \textit{PRSS16} in \textit{Loxodona
+  africana} (in this work), and the expansion of the growth hormone
+and chorionic gonadotropin beta genes in anthropoid primates
+\cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
+Expression patterns of specific genes also can vary widely between
+species. For example, the expression of orthologous genes in the
+placenta have been found to vary substantially between \human and
+\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
+recruitment of \textit{SIGLEC6} to placental expression in \human
+differs from both \textit{Pan troglodytes} and \mouse
+\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
+
+In order to address our two questions in a more systematic way, we
+sequenced and quantified the mRNA environment of mammalian placentas
+at term, including that of the \human, \textit{Pan paniscus} (the
+sister species to \textit{P. troglodytes}), \textit{Ateles fusciceps},
+\mouse, \galili, \carmeli, \textit{Canis familiaris}, \cow, \elephant,
+\textit{Dasypus novemcinctus}, and \textit{Monodelphis domesticus}.
+These species represent four major branches of eutherian mammals
+(Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
+as marsupials. These placentas also represent a wide array of
+placental morphologies as well as traits such as number and size of
+offspring and length of gestation (\cref{tab:gestational_periods}). We
+use this data, along with existing RNA-Seq data from \mouse
+\cite{Wang.ea2011:surveyfornovelimprinted}, \human
+\cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
+\textit{Equus caballus}
+\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis
+  aries} \cite{Archibald.ea2010:sheepgenomereferencesequence}, and
+\textit{Sus scrofa} \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
+identify the non-housekeeping genes which are expressed in the
+placentas of all species analyzed and therefore represent the
+\emph{core functionality of the placenta}; 2) identify the genes whose
+expression level changes significantly in each of the morphology
+types.
 

### 2016-09-20 manuscript/manuscript.Rnw
remove messages from tree

@@
 \begin{figure}
-<<species.tree,fig.width=8,fig.height=6,out.width="\\columnwidth",echo=FALSE>>=
+<<species.tree,fig.width=8,fig.height=6,out.width="\\columnwidth",echo=FALSE,message=FALSE,warning=FALSE>>=
 species.tree <- 

### 2016-09-20 manuscript/manuscript.Rnw
switch to bar plot

@@
                                           capfirst(names(species.names))))]
-print(ggplot(cptl,
-             aes(y=human_name,x=Species))
-      + geom_raster(aes(fill=log10(FPKM)))
-      + scale_x_discrete(expand=c(0,0))
-      + scale_y_discrete(expand=c(0,0))
-      + scale_fill_gradient(low = grad.def[["low"]],
-                            high = grad.def[["high"]],
-                            space = grad.def[["space"]],
-                            na.value = grad.def[["na.value"]],
-                            limits=c(0,4),
-                            guide=guide_colorbar(title=expression(log[10](FPKM)))
-                            )
-      + ylab("Human Ortholog Name")
-      + xlab("Species")
-      + theme_grey()
-      + theme(legend.position="top",
-              axis.ticks = element_blank(),
-              axis.text.y = element_text(face="italic"),
-              axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic"))
-      )
+
+print(ggplot(cptl,aes(x=human_name,y=FPKM))+
+      geom_boxplot(outlier.shape=NA)+
+      scale_y_log10()+
+      geom_point(aes(color=Species,shape=Species))+
+      scale_shape_manual(guide=guide_legend(nrow=2,
+                                            keywidth=unit(3,"mm"),
+                                            keyheight=unit(3,"mm"),
+                                            label.theme=element_text(face="italic",
+                                                                     angle=0,
+                                                                     size=9)),
+                         values=0:18)+
+      coord_flip()+
+      theme(legend.position="bottom"))
+
+# print(ggplot(cptl,
+#              aes(y=human_name,x=Species))
+#       + geom_raster(aes(fill=log10(FPKM)))
+#       + scale_x_discrete(expand=c(0,0))
+#       + scale_y_discrete(expand=c(0,0))
+#       + scale_fill_gradient(low = grad.def[["low"]],
+#                             high = grad.def[["high"]],
+#                             space = grad.def[["space"]],
+#                             na.value = grad.def[["na.value"]],
+#                             limits=c(0,4),
+#                             guide=guide_colorbar(title=expression(log[10](FPKM)))
+#                             )
+#       + ylab("Human Ortholog Name")
+#       + xlab("Species")
+#       + theme_grey()
+#       + theme(legend.position="top",
+#               axis.ticks = element_blank(),
+#               axis.text.y = element_text(face="italic"),
+#               axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic")) 
+#      )
 @ 

### 2016-09-20 manuscript/manuscript.Rnw
start discussing shape

@@
 In order to address our two questions in a more systematic way, we
-sequenced and quantified the mRNA environment of mammalian placentas
-at term, including that of the \human, \textit{Pan paniscus} (the
-sister species to \textit{P. troglodytes}), \textit{Ateles fusciceps},
+sequenced and quantified the total RNA of mammalian placentas at term,
+including that of the \human, \textit{Pan paniscus} (the sister
+species to \textit{P. troglodytes}), \textit{Ateles fusciceps},
 \mouse, \galili, \carmeli, \textit{Canis familiaris}, \cow, \elephant,
@@
 \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
-Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
+\cref{fig:core_placenta_transcriptome_shape} shows the number of genes
 which would be included in the core placenta set if the \ac{FPKM}
-cutoff was set differently.
+cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
+\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>1 & 
+`Minimum Expression in` == '14 of 14 species',max(num_genes)]} %
+genes in the core transcriptome.
 %
@@
 
-\subsection{\ac{GSE} Analyses of Individual Transcriptomes}
-
-In order to identify major functional categories of genes expressed in
-the individual transcriptomes, we examined \ac{GSE} in 1:1 human
-orthologs which had \ac{FPKM} values in the top 1\% of expressed genes
-(Supplemental %
-\cref{tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_monodelphis_domestica,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_canis_familiaris}) using topGO \cite{topGO:2.20.0}.
-%
-%
-Significant enrichment in genes ($\text{FDR}\le0.05$) in the top 1\%
-of \ac{FPKM} values in at least 12 of the
-\Sexpr{length(species.names)} species included clusters such as
-membrane organization , locomotion, tissue development, extracellular
-matrix organization, regulation of peptidase activity, and various
-cell death related pathways among others
-(Supplemental \cref{tab:go_terms_all_species}). If we instead examine the top 100
-genes by \ac{FPKM} in each of the \Sexpr{length(species.names)}
-species (Supplemental %
-\cref{tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_monodelphis_domestica,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_canis_familiaris})
-notable enrichments include extracellular matrix, female pregnancy,
-integrin and collagen-related processes, and various cell death
-related processes.
+\subsection{Placenta Morphology and Genetics}
+
+\subsubsection{Placenta Attachment Shape}
+
+\subsubsection{Degree of Intimacy}
+
+\subsubsection{Interdigitation}
 

### 2016-09-20 data/placenta_classification_specific.R
load placenta classifciation

@@ -9,7 +9,7 @@ args <- commandArgs(trailingOnly=TRUE)
 
 load(args[1])
 load(args[2])
-placenta_classification <- fread(args[3])
+load(args[3])
 
 combined.fpkm.wide <- 
     data.table(dcast(combined.fpkm,

### 2016-09-23 manuscript/supplemental_material.Rnw
do gsea for morphology associations

@@
 
+\clearpage
+
 \begin{table}
@@
 
-\section{Gene Ontology Analyses}
+\section{GSEA}
+
+\subsection{Pathway Analysis}
+\begin{table}
+<<pathway_analysis,echo=FALSE,results="asis">>=
+pathway <- list()
+for (database in c("kegg","reactome","biocarta","netpath")) {
+    load(paste0("../data/placenta_gsea_",database))
+    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
+    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
+    pathway[[database]] <-
+        gsea.table
+}
+pathway.analysis <-
+    rbindlist(pathway)[fdr < 0.1,]
+rm(pathway)
+pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
+                                gsub("_+"," ",
+                                     Pathway))]
+pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]
+
+setnames(pathway.analysis,"p_values","p")
+setorder(pathway.analysis,"p")
+print(xtable(pathway.analysis,
+             display=c("s","s","f","f","s"),
+             align=c("c","p{2in}","r","r","c")),
+      math.style.exponents=TRUE,
+      include.rownames=FALSE,
+      booktabs=TRUE,
+      floating=FALSE)      
+@ 
+\caption[Core Placenta Pathway Enrichment]{Pathways with significant and suggestive enrichment
+  ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
+  biocarta, reactome, and kegg pathway databases}
+\label{tab:core_placenta_pathway_analysis}
+\end{table}
+
+\begin{table}
+<<pathway_analysis_class,echo=FALSE,results="asis">>=
+pathway <- list()
+for (class in c("shape","interdigitation","barrier")) {
+    for (database in c("kegg","reactome","biocarta","netpath")) {
+        load(paste0("../data/placenta_gsea_",class,"_",database))
+        eval(parse(text=paste0("gsea.table <- placenta_gsea_",class,"_",database)))
+        gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
+        gsea.table[,Class:=gsub("^(.)","\\U\\1",class,perl=TRUE)]
+        pathway[[paste0(class,"_",database)]] <-
+            gsea.table
+    }
+}
+placenta_class_gsea <-
+    rbindlist(pathway)[fdr < 0.1,]
+placenta_class_gsea[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
+                                   gsub("_+"," ",
+                                        Pathway))]
+placenta_class_gsea[,Pathway:=gsub(".+\\%","",Pathway)]
+setnames(placenta_class_gsea,"p_values","p")
+setorder(placenta_class_gsea,"p")
+print(xtable(placenta_class_gsea,
+             display=c("s","s","f","f","s"),
+             align=c("c","p{2in}","r","r","c")),
+      math.style.exponents=TRUE,
+      include.rownames=FALSE,
+      booktabs=TRUE,
+      floating=FALSE)      
+@ 
+\caption[Placenta Classification Pathway Enrichment]{Pathways with
+  significant and suggestive enrichment ($\text{FDR}\le 0.1$) in genes
+  associated with placenta morphology using the biocarta, reactome,
+  and kegg pathway databases.}
+\label{tab:core_placenta_pathway_analysis}
+\end{table}
+
+\subsection{GO Ontology}
 

### 2016-09-23 data/placenta_classification_specific.R

@@ -96,12 +96,11 @@ placenta.classification.p[,fdr.overall:=p.adjust(p,method="BH")]
 placenta.classification.significance <- list()
 for (class in c("shape","barrier","interdigitation")) {
     placenta.classification.significance[[class]] <-
-        ifelse(placenta.classification.p[!is.na(egid),
-                                         paste0(class,".fdr"),
-                                         with=FALSE] <= 0.05,
+        ifelse(placenta.classification.p[!is.na(egid) & type=="aov" & factor==class,
+                                         fdr] <= 0.05,
                1,0)
     names(placenta.classification.significance[[class]]) <-
-        placenta.classification.p[!is.na(egid),egid]
+        placenta.classification.p[!is.na(egid) & type=="aov" & factor==class,egid]
 }
 
 save(placenta.classification.p,

### 2016-09-23 manuscript/supplemental_material.Rnw
load common placenta classification table

@@
 <<placental_classification,echo=FALSE,results="hide",cache.extra=file.info(dir("../data","placenta.*classification.*",full.names=TRUE))[,"mtime"]>>=
-placenta_classification <- fread("../data/placental_classification.txt")
+load("../data/placenta_classification")
 load("../data/placenta_classification_p")
@@
 <<placental_classification_table,echo=FALSE,results="asis">>=
-print(xtable(placental.classification,
+print(xtable(placenta_classification,
              display=c("s","s","s","s","s"),

### 2016-09-23 data/placenta_classification.R
fix labyrinthine typo

@@ -8,13 +8,13 @@ placenta_classification <- fread(args[1])
 
 placenta_classification[,species:=factor(species)]
 placenta_classification[,shape:=factor(shape)]
-placenta_classification[,interdigitiation:=
+placenta_classification[,interdigitation:=
                              factor(interdigitation,
                                     levels=c("villous",
                                              "folded",
                                              "trabecular",
                                              "lamellar",
-                                             "labrynthine"
+                                             "labyrinthine"
                                              ),
                                     ordered=TRUE
                                     )]

### 2016-09-28 manuscript/supplemental_material.Rnw
add pathway plots for EGF and actin cytoskeleton

@@
 
+<<placenta_pathway_egid,echo=FALSE,results="hide">>=
+load("../data/placenta_core_transcriptome_complete")
+plac_path_egid <- function(column) {
+    expr <- placenta.core.transcriptome.complete[!is.na(egid),eval(column)]
+    names(expr) <- placenta.core.transcriptome.complete[!is.na(egid),egid]
+    expr
+}
+
+plac_path_genes <- function(genes,column) {
+    result <- as.numeric(NULL)
+    result[!is.na(genes)] <-
+        placenta.core.transcriptome.complete[na.omit(genes),eval(column)]
+    names(result) <- names(genes)
+    result
+}
+
+@
+
+<<plot_pathway,echo=FALSE,result="hide">>=
+library("KEGGgraph")
+library("plotrix")
+makeAttr <- function(graph, default, valNodeList) {
+  tmp <- nodes(graph)
+  x <- rep(default, length(tmp)); names(x) <- tmp
+
+  if(!missing(valNodeList)) {
+    stopifnot(is.list(valNodeList))
+    allnodes <- unlist(valNodeList)
+    stopifnot(all(allnodes %in% tmp))
+    for(i in seq(valNodeList)) {
+      x[valNodeList[[i]]] <- names(valNodeList)[i]
+    }
+  }
+  return(x)
+}
+graph_attrs <- function(graph,nodecolorcolumn,cutpoint=1) {
+    outs <- sapply(edges(graph), length) > 0
+    ins <- sapply(inEdges(graph), length) > 0
+    ios <- outs | ins
+    ## translate the KEGG IDs into Gene Symbol
+    ioGeneID <- translateKEGGID2GeneID(names(ios))
+    nodesNames <- sapply(mget(ioGeneID, org.Hs.egSYMBOL, ifnotfound=NA), "[[",1)
+    names(nodesNames) <- names(ios)
+    nodevalue <- plac_path_genes(nodesNames,nodecolorcolumn)
+    nodevalue.max <- max(pretty(nodevalue))
+    nodevalue.na <- is.na(nodevalue)
+    nodevalue[is.na(nodevalue)] <- 0
+    nodecolors <- apply(colorRamp(c("gray","blue"))(nodevalue/nodevalue.max),
+                        1,
+                        function(x){rgb(x[1],x[2],x[3],maxColorValue=255)})
+    nodecolors[nodevalue.na] <- "#FFFFFF"
+    names(nodecolors) <- names(nodevalue)
+    subgraph <- graph #subGraph(names(nodecolors)[nodevalue >= 0.1],
+                       #  graph)
+    for (node in names(nodevalue)[nodevalue < cutpoint]) {
+         subgraph <- removeNode(node,subgraph)
+    }
+    nAttrs <- list();
+    # nAttrs$fillcolor <- makeAttr(graph, "lightgrey", list(orange=names(ios)[ios]))
+    nAttrs$label <- nodesNames[nodes(subgraph)]
+    nAttrs$fillcolor <- nodecolors[nodes(subgraph)]
+    nAttrs$fontsize <- makeAttr(subgraph,14)
+    nAttrs$width <- makeAttr(subgraph,2)
+    nAttrs$shape <- makeAttr(subgraph,"box")
+    list(nAttrs=nAttrs,
+         graph=subgraph,
+         max=nodevalue.max
+         )
+}
+
+plot_pathway <- function(pathway="hsa04012",pathcolor=quote(median_expression),cutpoint=1) {
+
+    the.graph <- KEGGgraph::parseKGML2Graph(KEGGgraph::getKGMLurl(pathway))
+
+    graph.attrs <- graph_attrs(the.graph,nodecolorcolumn=pathcolor,cutpoint=cutpoint)
+    plot(graph.attrs$graph,"dot",
+         nodeAttrs=graph.attrs$nAttrs,
+         attrs=list(graph=list(rankdir="LR")))
+    color.legend(100,100,200,200,c("0",as.character(graph.attrs$max)),
+                 apply(colorRamp(c("gray","blue"))(seq(0,1,length.out=10)),
+                       1,
+                       function(x){rgb(x[1],x[2],x[3],maxColorValue=255)}))
+}
+
+@
+
+\begin{figure}
+<<egf_pathway,echo=FALSE,message=FALSE,fig.width=6,fig.height=6,fig.out.width="\\textwidth">>=
+plot_pathway("hsa04012")
+@
+\caption{EGF KEGG pathway}
+\label{fig:egf_kegg_pathway}
+\end{figure}
+
+<<actin_pathway,echo=FALSE,results="hide",message=FALSE>>=
+library("pathview")
+pathview(gene.data=plac_path_egid(quote(as.numeric(median_expression))),
+         pathway.id="hsa04810",
+         both.dirs=list(gene=FALSE,cpd=FALSE),
+         limit=list(gene=200,cpd=29),
+         high=list(gene="blue"),
+         net.signature=FALSE)
+@
+
+\begin{figure}
+  \includegraphics{hsa04810.pathview.png}
+  \caption{KEGG Actin Cytoskeleton Pathway}
+  \label{fig:kegg_actin}
+\end{figure}
+
 <<placenta_class_table,echo=FALSE,results="hide">>=
 placenta_class_table <- function(class="shape") {
-    subtable <- placenta.classification.p[,c("rn",class,paste0(class,".fdr")),with=FALSE]
+    subtable <- placenta.classification.p[factor==class & type == "aov",list(gene,p,fdr)]
     setnames(subtable,c("Gene","p","FDR"))

### 2016-09-28 manuscript/manuscript.Rnw
add core placenta plot to manuscript

@@
 %\usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=14pt]{geometry}
-\usepackage[letterpaper]{geometry}
+%\usepackage[letterpaper]{geometry}
 %\usepackage{url}
@@
 
+<<core.placenta.transcriptome.distribution,results="hide",echo=FALSE,message=FALSE>>=
+num.species <- star.logs.wide[,length(unique(species))]
+calculate.distribution.shape <- 
+    function(min.vector) {
+        min.vector <- sort(min.vector,decreasing=TRUE)
+        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
+        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
+        return(temp)
+    }
+
+cptg.nh <- core.placenta.transcriptome.genes.shape[!(human_name %in% housekeeping.genes.superset[,gene_short_name])]
+
+cpt.distribution.shape <- 
+    lapply(c("all_expression",
+             "all_but_one_expression",
+             "all_but_two_expression",
+             "all_but_three_expression",
+             "all_but_four_expression"),
+           function(x){
+               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
+           })
+fix.expression.names <- function(x){
+    x <- gsub("^all_expression$",
+              paste0(num.species," of ",num.species," species"),
+              x
+              )
+    nc.levels <- paste0(num.species," of ",num.species," species")
+    ## number coutner
+    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
+    for (a in names(nc)) {
+        x <- gsub(paste0("^all_but_",a,"_expression$"),
+              paste0(num.species-nc[a]," of ",num.species," species"),
+              x
+              )
+        nc.levels <-
+            c(nc.levels,
+              paste0(num.species-nc[a]," of ",num.species," species"))
+    }
+    return(factor(x,levels=nc.levels))
+}
+
+cpt.distribution.shape <- 
+    rbindlist(cpt.distribution.shape)
+
+cpt.distribution.shape <- 
+    cpt.distribution.shape[,`Minimum Expression in`:=
+                                fix.expression.names(`Minimum Expression in`)]
+
+setnames(cpt.distribution.shape,
+         "FPKM",
+         "Expression (FPKM)")
+@
+
 The
@@
 \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
-\cref{fig:core_placenta_transcriptome_shape} shows the number of genes
+Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
 which would be included in the core placenta set if the \ac{FPKM}

### 2016-09-28 manuscript/supplemental_material.Rnw
add core placenta plot to manuscript

@@
           math.style.exponents=TRUE,
-          size="\\tiny")
+          size="\\scriptsize")
 }
@@
 
-\begin{table}
+\begin{table}[h]
 <<attachment_shape_significant,echo=FALSE,results="asis">>=
@@
 \end{table}
+\clearpage
 
@@
 
+\begin{table}[h]
 <<intimacy_significant,echo=FALSE,results="asis">>=
@@
 
+\clearpage
+
 \subsection{Maternal-Fetal Interdigitation}
 
+\begin{table}[h]
 <<interdigitation_significant,echo=FALSE,results="asis">>=
@@
 
+\clearpage
+
 \section{GSEA}
@@
 \begin{table}
-<<pathway_analysis,echo=FALSE,results="asis">>=
+<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
 pathway <- list()
@@
 print(xtable(pathway.analysis,
-             display=c("s","s","f","f","s"),
-             align=c("c","p{2in}","r","r","c")),
+             display=c("s","s","g","g","s"),
+             align=c("c","p{3.5in}","r","r","c")),
       math.style.exponents=TRUE,
@@
 \begin{table}
-<<pathway_analysis_class,echo=FALSE,results="asis">>=
+<<pathway_analysis_class,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
 pathway <- list()
@@
 print(xtable(placenta_class_gsea,
-             display=c("s","s","f","f","s"),
-             align=c("c","p{2in}","r","r","c")),
+             display=c("s","s","g","g","s","s"),
+             align=c("c","p{3in}","r","r","c","c")),
       math.style.exponents=TRUE,

### 2016-09-29 manuscript/manuscript.Rnw
stop loading EBImage
 + EBImage breaks channel and makes an odd error message

@@
 library("ape")
-library("EBImage")
 library("ggtree")
@@
 pp <- ggtree(species.tree)+
-    geom_text(aes(label=label),hjust=0,size=5,fontface="italic")+
+    geom_text(aes(label=gsub("_"," ",label)),hjust=0,size=5,fontface="italic")+
     scale_x_continuous(limits=c(0,10))

### 2016-09-29 manuscript/manuscript.Rnw
fix line break issues

@@
 \addbibresource{references.bib}
-\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue]{hyperref}
+\usepackage[hyperfigures,bookmarks,colorlinks,linkcolor=black,urlcolor=black,citecolor=blue,breaklinks=true]{hyperref}
 % Floats at end for journals
@@
 \usepackage[capitalize]{cleveref}
-
+\def\UrlBreaks{\do\.\do\@\do\\\do\/\do\!\do\_\do\|\do\;\do\>\do\]%
+\do\)\do\,\do\?\do\&\do\'\do+\do\=\do\#%
+\do\/%
+}%
 \input{common_latex}
+\tolerance=5000
+\emergencystretch=0.1pt
 
@@
 
+\begin{sloppypar}
 The quality of reads was evaluated using FASTQC v0.11.2
@@
 %
+\end{sloppypar}
 

### 2016-09-29 manuscript/manuscript.Rnw
start working on discussion

@@
 sequencing was performed by the Applied Genomics Technology Center
-(AGTC) at Wayne State University School of Medicine using 2.5-5 µg of
+(AGTC) at Wayne State University School of Me\-di\-cine using 2.5-5 µg of
 total RNA and the standard Illumina protocol. Paired-end sequencing
@@
 `Minimum Expression in` == '14 of 14 species',max(num_genes)]} %
-genes in the core transcriptome.
-%
+genes in the core transcriptome. %
+In contrast to the shape of the number of genes included in the core
+placenta transcriptome, the set of \human housekeeping genes expressed
+all placentas is flatter (Supplemental
+\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
+of the total fraction of genes. However, there is still a large
+difference between the set of housekeeping genes which are expressed
+in \human and those which are expressed in all placentas.
+
+\subsubsection{Annexin Complexes}
+
 Multiple components of annexin complexes are included in the core
@@
 
-In contrast to the shape of the number of genes included in the core
-placenta transcriptome, the set of \human housekeeping genes expressed
-all placentas is flatter (Supplemental
-\cref{fig:core_housekeeping_transcriptome_shape}), and includes more
-of the total fraction of genes. However, there is still a large
-difference between the set of housekeeping genes which are expressed
-in \human and those which are expressed in all placentas.
+\subsubsection{Invasion: EGF and Auxin Y in the Core Placenta}
 
@@
 
+\subsubsection{Annexin Complexes}
 The transcriptomes from multiple different species enabled us to
@@
 placenta.
+
+\subsubsection{Invasion: EGF and Auxin Y in the Core Placenta}
 

### 2016-10-04 manuscript/manuscript.Rnw
update manuscript for gsea
explain GSEA and ANOVA for morphology

@@
 species having separate random effects.
+
+Significant shifts in expression levels which correlate with placenta
+morphology (shape, interdigitation, and intimacy) were identified
+using ANOVA on generalized linear models with the $\log$ \ac{FPKM} as
+the response variable and the morphology as the predictor variable.
+Placenta morphology with a defined order (interdigitation, intimacy)
+were additionally analyzed using proportional odds logistic regression
+using the MASS package in R~\cite{Venables2002:MASS}.
+
+\subsection{Gene Set and Pathway Enrichment Analysis}
+
+Gene sets and pathways in were obtained from MSigdb v5.1 hallmark gene
+set \cite{Liberzon.ea2011:MolecularsignaturesdatabaseMSigDB} and Gene
+Ontology which represent a superset of available pathway databases,
+including Reactome, Kegg, Netpath, and GO. We calculated the
+over-representation of gene lists in these gene sets using Fisher's
+exact test and corrected for multiple testing by estimating the
+\ac{FDR} using the Benjamini and Hochberg procedure
+\cite{Benjamini1995:bh_procedure}.
 

### 2016-10-04 data/placenta_classification.R
use intimacy instead of barrier

@@ -19,8 +19,8 @@ placenta_classification[,interdigitation:=
                                     ordered=TRUE
                                     )]
 
-placenta_classification[,barrier:=
-                             factor(barrier,
+placenta_classification[,intimacy:=
+                             factor(intimacy,
                                     levels=c("epitheliochorial",
                                              "endotheliochorial",
                                              "hemochorial"

### 2016-10-04 data/placenta_classification_specific.R

@@ -4,7 +4,7 @@ library("org.Hs.eg.db")
 library("parallel")
 
 args <- c("combined_fpkm","housekeeping_genes_superset",
-          "placental_classification.txt","placenta_classification_specific")
+          "placenta_classification","placenta_classification_specific")
 
 args <- commandArgs(trailingOnly=TRUE)
 
@@ -51,7 +51,7 @@ placenta_classification <-
     placenta.class.specific[placenta_classification]      
 
 
-calculate.aov <- function(gene,factor="barrier") {
+calculate.aov <- function(gene,factor="intimacy") {
     temp.glm <- glm(as.formula(paste0("mean_fpkm~",factor,"+0")),
                     data=placenta_classification[human_name==gene,])
     temp.glm.coef <- summary(temp.glm)$coefficients
@@ -78,10 +78,11 @@ calculate.aov <- function(gene,factor="barrier") {
 }
 
 placenta.classification.p <-
-    rbindlist(lapply(c("shape","barrier","interdigitation"),
+    rbindlist(lapply(c("shape","intimacy","interdigitation"),
                      function(x){
                          rbindlist(mclapply(placenta_classification[,unique(human_name)],
-                                            calculate.aov,factor=x,mc.cores=num.cores))}))
+                                            calculate.aov,factor=x,mc.cores=num.cores))})
+              )
 
 ## throw out analyses with infinite statistic
 placenta.classification.p <-
@@ -94,7 +95,7 @@ placenta.classification.p[,fdr:=p.adjust(p,method="BH"),by=list(factor,type)]
 placenta.classification.p[,fdr.overall:=p.adjust(p,method="BH")]
 
 placenta.classification.significance <- list()
-for (class in c("shape","barrier","interdigitation")) {
+for (class in c("shape","intimacy","interdigitation")) {
     placenta.classification.significance[[class]] <-
         ifelse(placenta.classification.p[!is.na(egid) & type=="aov" & factor==class,
                                          fdr] <= 0.05,

### 2016-10-04 data/placenta_classification_specific.R
calculate using polr

@@ -77,12 +77,38 @@ calculate.aov <- function(gene,factor="intimacy") {
                               "p"=temp.glm.coef[,"Pr(>|t|)"])))
 }
 
+calculate.polr <- function(gene,factor="intimacy") {
+    temp.coef <-
+        coef(summary(polr(as.formula(paste0(factor,"~mean_fpkm")),
+                          placenta_classification[human_name==gene],
+                          Hess=TRUE)))
+    data.table("gene"=gene,
+               "egid"=name.to.ensembl[gene,egid],
+               "gene_id"=name.to.ensembl[gene,gene_id],
+               "factor"=factor,
+               "type"="polr",
+               "level"=factor,
+               "coefficient"=temp.coef["mean_fpkm","Value"],
+               "statistic"=temp.glm.coef["mean_fpkm","t value"],
+               ## approximate p using the Z distribution
+               "p"=pnorm(temp.glm.coef["mean_fpkm","t value"],
+                         lower.tail=FALSE)*2)
+}
+    
+
 placenta.classification.p <-
     rbindlist(lapply(c("shape","intimacy","interdigitation"),
                      function(x){
                          rbindlist(mclapply(placenta_classification[,unique(human_name)],
                                             calculate.aov,factor=x,mc.cores=num.cores))})
               )
+placenta.classification.p <-
+    rbindlist(c(list(orig=placenta.classification.p),
+                lapply(c("intimacy","interdigitation"),
+                       function(x){
+                           rbindlist(mclapply(placenta_classification[,unique(human_name)],
+                                              calculate.aov,factor=x,mc.cores=num.cores))})
+                ))
 
 ## throw out analyses with infinite statistic
 placenta.classification.p <-

### 2016-10-04 manuscript/supplemental_material.Rnw
plot too short vs read length

@@
 \begin{figure}
-<<reads.vs.read.length,fig.width=8,fig.height=8,echo=FALSE>>=
+<<reads.vs.read.length,fig.width=8,fig.height=6,echo=FALSE>>=
+grid::pushViewport(grid::viewport(layout=grid::grid.layout(ncol=2,nrow=1,widths=c(1,1.4))))
 print(ggplot(star.logs.wide,
@@
       + xlab("Trimmed Read Length")
-      + ylab("Read number")
+      + ylab("Number of input reads")
       + scale_y_log10()
-      + theme(legend.text=element_text(face="italic")
-              )
-      )
+      + theme(legend.position="none"
+              ),
+      vp=grid::viewport(layout.pos.row=1,layout.pos.col=1)
+      )
+print(ggplot(star.logs.wide,
+             aes(x=as.numeric(input.read.length),
+                 y=as.numeric(`% of reads unmapped: too short`)#,
+                 # color=capfirst(species),
+                 # shape=capfirst(species)
+                 ))
+      + scale_shape_manual(values=1:nlevels(as.factor(star.logs.wide[,species])),
+                           guide=guide_legend(title="Species")
+                           )
+      + scale_color_discrete(guide=guide_legend(title="Species")
+                             )
+      + geom_point(aes(color=capfirst(species),shape=capfirst(species)),size=5)
+      + stat_smooth(method="lm")
+      + xlab("Trimmed Read Length")
+      + ylab("Unmapped Too Short %")
+      + theme(legend.text=element_text(face="italic"),
+              legend.position="right"
+              ),
+            vp=grid::viewport(layout.pos.row=1,layout.pos.col=2)
+      )
+grid::popViewport()
 @ 

### 2016-10-04 manuscript/manuscript.Rnw
discuss EGF and auxin y invasion

@@
 
+The core placenta transcriptome is significantly enriched in genes
+which are involved in multiple biological pathways, including multiple
+pathways related to translation (ribosome, translation, signal
+recognition particle targeting to the rough endoplasamic reticulum,
+etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
+translation, the next most abundant set of significantly enriched
+molecular pathways are related to the migration of the trophoblast
+into the endometrium, including the EGFR1 pathway and the Actin Y
+pathway. Two other pathways, the mPR/Progesterone pathway and Rho cell
+motility pathways, feed into the Actin Y pathway as well. Supplemental
+\cref{fig:egf_kegg_pathway} shows the EGF pathway in humans with the
+median expression in FPKM of genes in the pathway shown. Notably, many
+of the genes on the GRB2->MYC pathway are strongly expressed; their
+activation leads to an increase in migration and invasion in response
+to EGF binding to EGFR1. The Actin cytoskeleton pathway is depicted in
+Supplemental \cref{fig:kegg_actin}, which shows that \textit{FN1} and
+\textit{ITG} are highly expressed, which activate Rac, and potentially
+lead to the formation of focal adhesions and lamellipodia formation
+through their inhibition of GSN and HSPC300.
+
 % SPINT2 -- Serine Peptidase inhibitor

### 2016-10-04 manuscript/extended_discussion.Rnw
fix cache location and remove extra table

@@
 opts_chunk$set(dev="cairo_pdf",out.width="\\columnwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
-opts_chunk$set(cache=TRUE, autodep=TRUE)
+opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="extended_discussion_cache/")
 options(device = function(file, width = 8, height = 7, ...) {
@@
 pre-eclampsia.
-
-\begin{table}
-  \small
-  \begin{tabular}{>{\raggedright }p{0.7in}
-                  >{\raggedright }p{0.7in}
-                  c c 
-                  p{0.65in}}
-    \toprule
-    {\centering Species  }                     & {\centering Common Name }& 
-       \multicolumn{1}{p{0.43in}}{\raggedright \centering Gestation (days)} & 
-       \multicolumn{1}{p{0.3in}}{\raggedright \centering Litter Size}  & Source \\
-    \midrule
-    \textit{Ateles fusciceps}     & Spider monkey & 224 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Bos taurus}           & Cow & 278 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Canis familiaris}     & Dog & 61 & 4 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Dasypus novemcinctus} & Nine-banded Armadillo & 150 & 4 & \citet{Smith1984:amazingarmadillo} \\
-    \textit{Equus caballus}       & Horse & 336 & 1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation}\\
-    \textit{Loxodonta africana}   & African savanna elephant & 644 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental}\\
-    \textit{Homo sapiens}         & Human & 270 & 1.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Monodelphis domestica}& Gray short-tailed opossum & 15 & 7.1 & \citet{Harder.ea1993:Gestationandplacentationin} \\
-    \textit{Mus musculus}         & Mouse & 20 & 6.1 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Spalax galili}   & Blind mole rat from Upper Galilee & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals} \\
-    \textit{Ovis aries}           & Domestic sheep & 151 & 1.35 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \textit{Pan paniscus}         & Bonobo & 223 & 1 & \citet{Ernest2003:lifehistorycharacteristicsofplacental} \\
-    \textit{Spalax carmeli}       & Blind mole rat from northern Israel & 28-42 & 4 & \citet{Nevo1997:mosiacevolutionofsubterranianmammals}\\
-    \textit{Sus scrofa}           & Domestic pig (Duroc) & 114 & 5.8 & \citet{Kiltie1982:intraspecificvariationmammaliangestation} \\
-    \bottomrule
-  \end{tabular}
-  \caption{Gestational periods in days and average liter sizes for studied species.}
-  \label{tab:gestational_periods}
-\end{table}
 

### 2016-10-04 manuscript/extended_discussion.Rnw
use different figure path; remove author list

@@
-%\documentclass[english,12pt]{article}
-\documentclass[5p]{elsarticle_dla}
+\documentclass[english,12pt]{article}
+%\documentclass[5p]{elsarticle_dla}
 \usepackage{fontspec}
@@
 opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="extended_discussion_cache/")
+opts_chunk$set(fig.path="extended_discussion_figure/")
 options(device = function(file, width = 8, height = 7, ...) {
@@
 \title{The evolution of the transcriptome of mammalian placentas}%
-%
-\author[affigb]{Don L. Armstrong\fnref{n1}}
-\ead{don@donarmstrong.com}
-\author[afflon]{Michael R. McGowen\fnref{n1}}
-\ead{1mmcgowen1@gmail.com}
-\author[affigb]{Amy Weckle}
-\ead{aweckle@illinois.edu}
-\author[affwayne]{Jason Caravas}
-\ead{jacaravas@gmail.com}
-\author[affmich]{Dalen Agnew}
-\ead{agnewd@dcpah.msu.edu}
-\author[affuscd]{Kurt Benirschke}
-\ead{kbenirschke@ucsd.edu}
-\author[affapetrust]{Sue Savage-Rumbaugh}
-\ead{suerumbaugh@gmail.com}
-\author[affhaifa]{Eviatar Nevo}
-\ead{nevo@research.haifa.ac.il}
-\author[affnih,affwaynepath]{Chong J Kim}
-\ead{cjkim@med.wayne.edu}
-\author[affyale]{Günter P. Wagner}
-\ead{gunter.wagner@yale.edu}
-\author[affnih]{Roberto Romero}
-\ead{prbchiefstaff@med.wayne.edu}
-\author[affigb,affapetrust,affmip]{Derek E. Wildman\corref{cor1}}
-\ead{wildmand@illinois.edu}
-\cortext[cor1]{Corresponding author}
-\address[affigb]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[afflon]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
-\address[affwayne]{Center for Molecular Medicine and Genetics, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[affmich]{Diagnostic Center for Population and Animal Health, College of Veterinary Medicine, Michigan State University, East Lansing, MI, USA}
-\address[affuscd]{Department of Pathology, University of California San Diego, La Jolla, CA, USA}
-\address[affapetrust]{Great Ape Trust/Bonobo Hope Sanctuary, Des Moines, IA, USA}
-\address[affnih]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
-\address[affwaynepath]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
-\address[affyale]{Department of Ecology and Evolutionary Biology and Yale Systems Biology Institute, Yale University, New Haven, CT, USA}
-\address[affmip]{Department of Molecular and Integrative Physiology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
-\address[affhaifa]{The Institute of Evolution, University of Haifa, Haifa, Israel}
-
-\begin{abstract}
-  \textbf{Background} The placenta is arguably the most
-  anatomically variable organ in mammals.
-
-  \textbf{Results} We measured the expression profiles of
-  \Sexpr{star.logs[,length(unique(file))]} term placentas of
-  \Sexpr{star.logs[,length(unique(species))]} species of mammals. We
-  found lineage-specific differential expression of prolactin,
-  relaxin, and galectins as well as increased expression of the
-  pre-eclampsia associated genes \textit{ADAM12}, \textit{CRH}, and
-  \textit{PAPPA} in humans. We also identified genes which are
-  expressed in all placentas, including genes with immune-modulating
-  properties (\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and
-  \textit{S100A10}) and cell-cell interactions (\textit{LAMC1},
-  \textit{LUM}, and \textit{LGALS1}).
-
-  \textbf{Conclusions} A diverse array of genes are differentially
-  expressed and are likely to play a role in the anatomical
-  variability of the placenta in the placentas of mammalian lineages.
-  Furthermore, multiple genes which are associated with pre-eclampsia
-  (\textit{CRH}, \textit{ADAM12}, \textit{KISS1}, \textit{PAPPA},
-  \textit{IL1RL1}, and \textit{PLAC1}) show increased expression on
-  the human lineage. Still, using the strictest filtering strategy
-  we find a core set of only
-  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-  genes is essential for the proper functioning of the placenta in all
-  species analyzed.
-
-\end{abstract}
-
-\begin{keyword}
-mammal \sep  gene expression \sep evolution \sep RNA-seq \sep placenta
-\end{keyword}
-
-\maketitle
 

### 2016-10-04 manuscript/supplemental_material.Rnw
use different figure path; remove author list

@@
 opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")
+opts_chunk$set(fig.path="supplemental_figure/")
 options(device = function(file, width = 8, height = 7, ...) {

### 2016-10-05 manuscript/manuscript.Rnw
change the position of the table

@@
 
-\begin{table}
+\begin{table}[tbh]
   \footnotesize
@@
 
-\begin{figure*}[htb]
-<<psg.ctsl.evolution.tree,echo=FALSE,out.width="\\textwidth",fig.width=8,fig.height=8>>=
-pushViewport(viewport(layout=grid.layout(nrow=2,ncol=2),clip="off"))
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PSG1" & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                subtree=c("ENSG00000221826","ENSG00000124469"),
-                                trees=trees
-                                )+
-      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.2)+
-      theme(legend.position="none"),
-      vp=viewport(layout.pos.col=1,layout.pos.row=1,clip="off"))
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CTSL" & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                subtree=c("ENSMODG00000002844","ENSMUSG00000055679"),
-                                trees=trees)+
-      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,0.8)+
-      theme(legend.position="none"),
-      vp=viewport(layout.pos.col=2,layout.pos.row=1,clip="off"))
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PRSS16" &
-                                                    species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                subtree=c(   "ENSG00000112812",
-                                          "ENSLAFG00000032211"),
-                                combined.fpkm=combined.fpkm,
-                                trees=trees)+
-      geom_tiplab(size=2,color="black",fontface="italic")+xlim(0,1.2),
-      vp=viewport(layout.pos.col=1,layout.pos.row=2,clip="off"))
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PAG2" &
-                                                    species=="bos taurus"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                subtree=c("ENSSSCG00000013097",
-                                          "ENSBTAG00000046085"),
-                                trees=trees) +
-      geom_tiplab(size=2,color="black",fontface="italic") +
-      xlim(0,0.8)+
-      theme(legend.position="none"),
-      vp=viewport(layout.pos.col=2,layout.pos.row=2,clip="off"))
-popViewport(1)
-
-grid.text(c("A","B","C","D"),
-          x=unit(c(0,0.5,0,0.5),"npc"),y=unit(c(1,1,0.5,0.5),"npc"),
-          just=c("left","top"),
-          gp=gpar(fontsize="24")
-          )
-@ 
-\caption{\genetreeexplanation{(A) \textit{PSG1}, (B) \textit{CTSL}, (C) \textit{PRSS16}, and (D) \textit{PAG2}}}
-\label{fig:pag2_prss16_ctsl_psg_evolution_tree}
-\end{figure*}
-
 \subsection{Core Placenta Transcriptome}
 
-\begin{figure}
+\begin{figure}[tbh]
 <<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=

### 2016-10-19 manuscript/manuscript.Rnw
it's intimacy, not barrier

@@
 for (mf.bp in c("mf","bp")) {
-    for (class in c("interdigitation","barrier","shape")) {
+    for (class in c("interdigitation","intimacy","shape")) {
         load(paste0("../data/placenta_classification_go_",mf.bp,"_",class))

### 2016-11-11 manuscript/manuscript.Rnw
make core placenta plot larger

@@
 \begin{figure}[tbh]
-<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
+<<core.placenta.transcriptome,fig.width=5,fig.height=8,out.height="0.8\\textheight",echo=FALSE>>=
 cptl <-
@@
       geom_boxplot(outlier.shape=NA)+
-      scale_y_log10()+
+      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
       geom_point(aes(color=Species,shape=Species))+
-      scale_shape_manual(guide=guide_legend(nrow=2,
+      scale_shape_manual("Species",
+                         guide=guide_legend(#nrow=1,
                                             keywidth=unit(3,"mm"),
@@
                                                                      angle=0,
-                                                                     size=9)),
+                                                                     size=9),
+                                            title.position="top"
+                                            ),
                          values=0:18)+
+      # scale_color_discrete(NULL)+
       coord_flip()+
-      theme(legend.position="bottom"))
-
-# print(ggplot(cptl,
-#              aes(y=human_name,x=Species))
-#       + geom_raster(aes(fill=log10(FPKM)))
-#       + scale_x_discrete(expand=c(0,0))
-#       + scale_y_discrete(expand=c(0,0))
-#       + scale_fill_gradient(low = grad.def[["low"]],
-#                             high = grad.def[["high"]],
-#                             space = grad.def[["space"]],
-#                             na.value = grad.def[["na.value"]],
-#                             limits=c(0,4),
-#                             guide=guide_colorbar(title=expression(log[10](FPKM)))
-#                             )
-#       + ylab("Human Ortholog Name")
-#       + xlab("Species")
-#       + theme_grey()
-#       + theme(legend.position="top",
-#               axis.ticks = element_blank(),
-#               axis.text.y = element_text(face="italic"),
-#               axis.text.x = element_text(angle = 90, hjust = 1, vjust=0,face="italic")) 
-#      )
+      xlab("Human Ortholog Name")+
+      ylab("FPKM")+
+      theme(legend.position=c(0.7,0.15),
+            axis.text.y=element_text(face="italic"),
+            legend.margin=unit(0,"mm"),
+            legend.justification="left",
+            plot.margin=unit(c(0,0,0,0),"mm")
+            ))
 @ 

### 2016-11-11 manuscript/manuscript.Rnw
start discussing morphology of placenta attachment et al

@@
 median expression in FPKM of genes in the pathway shown. Notably, many
-of the genes on the GRB2->MYC pathway are strongly expressed; their
+of the genes on the GRB2→MYC pathway are strongly expressed; their
 activation leads to an increase in migration and invasion in response
@@
 
+We correlated different aspects of placenta morphology, including
+shape, intimacy, and interdigitation with the mRNA transcription
+profiles of all 1:1 human orthologs with \ac{FPKM} greater than 10 in
+one species, and \ac{FPKM} greater than 1 at least two species.
+
 \subsubsection{Placenta Attachment Shape}
 
+Supplemental \cref{tab:shape_genes} shows the top 50 genes which are
+associated with the shape of placenta attachment, as well as the shape
+in which the difference is most significant.
+
+\textit{EREG} -- ligand of EGFR, known to be expressed in some
+placentas but apparently not present in the term placentas of \human.
+
+\textit{UNC} -- totally unknown
+
 \subsubsection{Degree of Intimacy}
+
+Genes in multiple immune system-related pathways are altered in increasing degrees of intimacy
 

### 2016-11-11 data/placenta_classification_specific.R
replace 0s with small non-zero

@@ -51,8 +51,14 @@ setkey(placenta_classification,"species")
 placenta_classification <-
     placenta.class.specific[placenta_classification]      
 
+set.seed(30621)
+### replace 0s with very small non-zero to avoid trouble with
+### underestimating the variance
+placenta_classification[mean_fpkm==0,
+                        mean_fpkm:=rnorm(sum(mean_fpkm==0,na.rm=TRUE))/100]
 
 calculate.aov <- function(gene,factor="intimacy") {
+    temp.data <- placenta_classification[human_name==gene,]
     temp.glm <- glm(as.formula(paste0("mean_fpkm~",factor,"+0")),
                     data=placenta_classification[human_name==gene,])
     temp.glm.coef <- summary(temp.glm)$coefficients

### 2016-11-11 data/placenta_classification_specific.R
pnorm needs the abs of the t value

@@ -102,7 +102,7 @@ calculate.polr <- function(gene,factor="intimacy") {
                "coefficient"=temp.coef["mean_fpkm","Value"],
                "statistic"=temp.coef["mean_fpkm","t value"],
                ## approximate p using the Z distribution
-               "p"=pnorm(as.numeric(temp.coef["mean_fpkm","t value"]),
+               "p"=pnorm(abs(as.numeric(temp.coef["mean_fpkm","t value"])),
                          lower.tail=FALSE)*2)
 }
     

### 2016-11-11 data/placenta_classification_specific.R
use polr analysis

@@ -134,11 +134,20 @@ placenta.classification.p[,fdr.overall:=p.adjust(p,method="BH")]
 placenta.classification.significance <- list()
 for (class in c("shape","intimacy","interdigitation")) {
     placenta.classification.significance[[class]] <-
-        ifelse(placenta.classification.p[!is.na(egid) & type=="aov" & factor==class,
-                                         fdr] <= 0.05,
-               1,0)
+        factor(ifelse(placenta.classification.p[!is.na(egid) & type=="aov" & factor==class,
+                                                fdr] <= 0.05,
+                      1,0),levels=c("0","1"))
     names(placenta.classification.significance[[class]]) <-
         placenta.classification.p[!is.na(egid) & type=="aov" & factor==class,egid]
+}
+
+for (class in c("intimacy","interdigitation")) {
+    placenta.classification.significance[[paste0(class,".polr")]] <-
+        factor(ifelse(placenta.classification.p[!is.na(egid) & type=="polr" & factor==class,
+                                                fdr] <= 0.05,
+                      1,0),levels=c("0","1"))
+    names(placenta.classification.significance[[paste0(class,".polr")]]) <-
+        placenta.classification.p[!is.na(egid) & type=="polr" & factor==class,egid]
 }
 
 save(placenta.classification.p,

### 2016-11-11 manuscript/supplemental_material.Rnw
fix placenta class analysis name
switch from barrier to intimacy

@@
 \usepackage{multirow}
+\usepackage{multicol}
 \usepackage{setspace}
@@
 \usepackage{xspace}
-\newenvironment{narrow}[2]{%
-  \begin{list}{}{%
-      \setlength{\topsep}{0pt}%
-      \setlength{\leftmargin}{#1}%
-      \setlength{\rightmargin}{#2}%
-      \setlength{\listparindent}{\parindent}%
-      \setlength{\itemindent}{\parindent}%
-      \setlength{\parsep}{\parskip}}%
-  \item[]}{\end{list}}
+% \usepackage[below]{placeins}
+% \makeatletter
+% \AtBeginDocument{%
+%   \expandafter\renewcommand\expandafter\subsection\expandafter
+%     {\expandafter\@fb@secFB\subsection}%
+%   \newcommand\@fb@secFB{\FloatBarrier
+%     \gdef\@fb@afterHHook{\@fb@topbarrier \gdef\@fb@afterHHook{}}}%
+%   \g@addto@macro\@afterheading{\@fb@afterHHook}%
+%   \gdef\@fb@afterHHook{}%
+% }
+% \makeatother
 \def\newblock{\hskip}
@@
 <<placenta_class_table,echo=FALSE,results="hide">>=
-placenta_class_table <- function(class="shape") {
-    subtable <- placenta.classification.p[factor==class & type == "aov",list(gene,p,fdr)]
+placenta_class_table <- function(class="shape",analysis.type="aov") {
+    subtable <- placenta.classification.p[factor==class & type == analysis.type,
+                                          list(gene,p,fdr)]
     setnames(subtable,c("Gene","p","FDR"))
-    subtable <- head(subtable[FDR <= 0.05,][order(p),],n=50)
+    subtable <- head(subtable[order(p),],n=50)
+    top.type <- function(gene.name) {
+        placenta.classification.p[gene==gene.name & factor==class
+                                  & type =="glm",][order(-abs(coefficient))][1,level]
+    }
+    subtable[,`level_type`:=sapply(Gene,top.type)]
+    setnames(subtable,"level_type",
+             paste0("Most Significant ",
+                    gsub("^(.)","\\U\\1",class,perl=TRUE)))
     print(xtable(subtable,
-                 align=c("c","c","r","r"),
-                 display=c("s","s","g","g")
+                 align=c("c","c","r","r","c"),
+                 display=c("s","s","g","g","s")
                  ),
@@
 
+\clearpage
+
 \subsection{Attachment Shape}
 
-\begin{table}[h]
+\begin{table}
 <<attachment_shape_significant,echo=FALSE,results="asis">>=
@@
 \end{table}
-\clearpage
 
@@
 <<intimacy_significant,echo=FALSE,results="asis">>=
-placenta_class_table("barrier")
+placenta_class_table("intimacy")
 @ 
 \caption{Top 50 genes which are associated with the degree of placenta intimacy}
-\label{tab:barrier_genes}
+\label{tab:intimacy_genes}
 \end{table}
 
-\clearpage
+\begin{table}[h]
+<<intimacy_significant_polr,echo=FALSE,results="asis">>=
+placenta_class_table("intimacy",analysis.type="polr")
+@ 
+\caption{Top 50 genes which are associated with the degree of placenta intimacy; ordered analysis}
+\label{tab:intimacy_genes_polr}
+\end{table}
 
@@
 \label{tab:interdigitation_genes}
+\end{table}
+
+\begin{table}[h]
+<<interdigitation_significant_polr,echo=FALSE,results="asis">>=
+placenta_class_table("interdigitation",analysis.type="polr")
+@ 
+\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation; ordered analysis}
+\label{tab:interdigitation_genes_polr}
 \end{table}
@@
 \begin{table}
+  \small
 <<pathway_analysis_class,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
 pathway <- list()
-for (class in c("shape","interdigitation","barrier")) {
+for (class in c("shape","interdigitation","intimacy")) {
     for (database in c("kegg","reactome","biocarta","netpath")) {
@@
 placenta_class_gsea <-
-    rbindlist(pathway)[fdr < 0.1,]
+    rbindlist(pathway)[fdr < 0.2,]
 placenta_class_gsea[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
@@
 \caption[Placenta Classification Pathway Enrichment]{Pathways with
-  significant and suggestive enrichment ($\text{FDR}\le 0.1$) in genes
+  significant and suggestive enrichment ($\text{FDR}\le 0.2$) in genes
   associated with placenta morphology using the biocarta, reactome,
   and kegg pathway databases.}
-\label{tab:core_placenta_pathway_analysis}
+\label{tab:placenta_pathway_class_analysis}
 \end{table}

### 2016-11-12 manuscript/manuscript.Rnw
discuss placenta attachment shape

@@
 associated with the shape of placenta attachment, as well as the shape
-in which the difference is most significant.
-
-\textit{EREG} -- ligand of EGFR, known to be expressed in some
-placentas but apparently not present in the term placentas of \human.
-
-\textit{UNC} -- totally unknown
+in which the difference is most significant. The most significant
+genes which were associated with diffuse placenta attachment include
+\textit{EREG}, a ligand of EGFR, \textit{UNC93A}, a homolog of
+\textit{UNC93B1} which is involved in the balancing between Tlr7 and
+Tlr9 immune response in mice\DLA{cite TLR7/TLR9 paper}, and
+\textit{WNT5A}, which inhibits the invasion of the trophoblast cell
+line HTR8/SVneo \textit{in vitro} \DLA{Cite this paper after reading
+  the comments on it (PMID: 26865089)}.
 

### 2016-11-12 manuscript/manuscript.Rnw
discuss placenta intimacy and interdigitation

@@
 
-\subsubsection{Degree of Intimacy}
-
-Genes in multiple immune system-related pathways are altered in increasing degrees of intimacy
-
-\subsubsection{Interdigitation}
+\subsubsection{Degree of Intimacy and Interdigitation}
+
+For the degree of intimacy and interdigitation, we saw multiple genes
+which were associated with different levels of interdigitiation and
+intimacy (Supplemental
+\cref{tab:intimacy_genes,tab:interdigitation_genes}) using an
+unordered factor approach, but when we used an ordered factor
+analysis, none of these genes were retained (Supplemental
+\cref{tab:intimacy_genes_polr,tab:interdigitation_genes_polr}). We
+suspect this is due to the small number of intermediate
+interdigitation and intimacy placentas that are available. Future
+analyses of additional placentas with intermediate interdigitation and
+intimacy should better illuminate the pathways which are responsible
+for these placental morphologies.
 

### 2016-11-12 literature_mining/placenta_class_specific.R
add rules for literature mining for placenta genes

@@ -1,0 +1,11 @@
+library("data.table")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+load(args[1])
+write.table(placenta.classification.p[fdr<0.05 & ((type=="aov" & factor=="shape")| type=="polr"),unique(gene)],
+            quote=FALSE,
+            sep="",
+            col.names=FALSE,
+            row.names=FALSE,
+            file=args[length(args)])

### 2016-11-12 literature_mining/placenta_class_terms.R

@@ -1,0 +1,13 @@
+library("data.table")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+load(args[1])
+write.table(c(placenta_classification[,levels(shape)],
+              placenta_classification[,levels(interdigitation)],
+              placenta_classification[,levels(intimacy)]),
+            quote=FALSE,
+            sep="",
+            col.names=FALSE,
+            row.names=FALSE,
+            file=args[length(args)])

### 2016-11-14 manuscript/manuscript.Rnw
add Priya
move core placenta distribution shape to paper

@@
 \ead{aweckle@illinois.edu}
+\author[affigb]{Priyadarshini Pantham}
+\ead{pantham@illinois.edu}
 \author[affwayne]{Jason Caravas}
@@
 
-<<core.placenta.transcriptome.distribution,results="hide",echo=FALSE,message=FALSE>>=
+\begin{figure}
+<<core.placenta.transcriptome.distribution,fig.width=5,fig.height=5,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
 num.species <- star.logs.wide[,length(unique(species))]
@@
            function(x){
-               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
+               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Min. Exp. in`:=x]
            })
@@
     x <- gsub("^all_expression$",
-              paste0(num.species," of ",num.species," species"),
+              paste0(num.species," of ",num.species),
               x
               )
-    nc.levels <- paste0(num.species," of ",num.species," species")
+    nc.levels <- paste0(num.species," of ",num.species)
     ## number coutner
@@
         x <- gsub(paste0("^all_but_",a,"_expression$"),
-              paste0(num.species-nc[a]," of ",num.species," species"),
+              paste0(num.species-nc[a]," of ",num.species),
               x
@@
             c(nc.levels,
-              paste0(num.species-nc[a]," of ",num.species," species"))
+              paste0(num.species-nc[a]," of ",num.species))
     }
@@
 cpt.distribution.shape <- 
-    cpt.distribution.shape[,`Minimum Expression in`:=
-                                fix.expression.names(`Minimum Expression in`)]
+    cpt.distribution.shape[,`Min. Exp. in`:=
+                                fix.expression.names(`Min. Exp. in`)]
 
@@
          "Expression (FPKM)")
-@
+
+print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],
+             aes(x=`Expression (FPKM)`,color=`Min. Exp. in`,y=num_genes))
+      + geom_line(lwd=1)
+      + xlab("Minimum Expression cutoff (FPKM)")
+      + ylab("Number of Genes Meeting FPKM Cutoff in N species")
+      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                      labels = trans_format("log10", math_format(10^.x)))
+      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
+                      labels = trans_format("log10", math_format(10^.x)))
+      + geom_vline(xintercept=10,linetype="dotted")
+      + geom_segment(mapping=aes(y=V1,
+                         yend=V1,
+                         xend=10,
+                         x=300,
+                         color=`Min. Exp. in`),
+                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
+                                              max(num_genes),`Min. Exp. in`],
+                     arrow=arrow(length=unit(0.5,"cm")),
+                     show.legend=FALSE
+                     )
+      + geom_text(mapping=aes(x=300,
+                      y=V1,
+                      label=paste0(V1," genes"),override=TRUE),
+                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
+                      max(num_genes),`Min. Exp. in`],
+                  show.legend=FALSE,
+                  hjust=0
+                  )
+      + guides(color=guide_legend(title=paste0("Min. Expression in\nn of ",num.species," Species",
+                                  keywidth=unit(3,"mm"),
+                                  keyheight=unit(3,"mm")
+                                  ))
+      + theme_bw()
+      + theme(legend.position=c(1,1),
+              legend.margin=unit(0,"mm"),
+              legend.justification=c(1,1),
+              plot.margin=unit(c(0,0,0,0),"mm")
+              )
+      )
+@ 
+\caption{Distribution shape of placenta core genes. The vertical axis
+  shows the number of 1:1 orthologous genes included in the core
+  placenta transcriptome if %
+  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[1]]}
+  %
+  (or %
+  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[2]]},
+  % 
+  etc.) have a minimum expression in FPKM equal to the horizontal axis.
+  For example, if the cutoff is set to 10 (vertical black line), 
+  % 
+  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
+    & `Min. Exp. in` == '14 of 14',max(num_genes)]}
+  %
+  genes would be included if the FPKM of a set of 1:1 orthologs of all 14 species must be at
+  least 10, %
+  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
+    & `Min. Exp. in` == '13 of 14',max(num_genes)]}
+  %
+  1:1 orthologs would be included if 13 of 14 species have an
+   with an FPKM of at least 10, and 
+  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
+    & `Min. Exp. in` == '10 of 14',max(num_genes)]}
+  genes would be included if 10 of 14 species
+  must have an FPKM of at least 10.}
+\label{fig:core_placenta_transcriptome_shape}
+\end{figure}
 
@@
 \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
-Supplemental \cref{fig:core_placenta_transcriptome_shape} shows the number of genes
+\Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
 which would be included in the core placenta set if the \ac{FPKM}
@@
 \Sexpr{cpt.distribution.shape[`Expression (FPKM)`>1 & 
-`Minimum Expression in` == '14 of 14 species',max(num_genes)]} %
+`Min. Exp. in` == '14 of 14',max(num_genes)]} %
 genes in the core transcriptome. %
@@
 
+  Priyadarshini Pantham wrote the paper and performed gene enrichment
+  analyses.
+
   Jason Caravas performed initial analyses.

### 2016-11-14 manuscript/supplemental_material.Rnw
move core placenta distribution shape to paper

@@
 
-\begin{figure}
-<<core.placenta.transcriptome.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
-num.species <- star.logs.wide[,length(unique(species))]
-calculate.distribution.shape <- 
-    function(min.vector) {
-        min.vector <- sort(min.vector,decreasing=TRUE)
-        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
-        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
-        return(temp)
-    }
-
-cptg.nh <- core.placenta.transcriptome.genes.shape[!(human_name %in% housekeeping.genes.superset[,gene_short_name])]
-
-cpt.distribution.shape <- 
-    lapply(c("all_expression",
-             "all_but_one_expression",
-             "all_but_two_expression",
-             "all_but_three_expression",
-             "all_but_four_expression"),
-           function(x){
-               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
-           })
-fix.expression.names <- function(x){
-    x <- gsub("^all_expression$",
-              paste0(num.species," of ",num.species," species"),
-              x
-              )
-    nc.levels <- paste0(num.species," of ",num.species," species")
-    ## number coutner
-    nc <- c("one"=1,"two"=2,"three"=3,"four"=4)
-    for (a in names(nc)) {
-        x <- gsub(paste0("^all_but_",a,"_expression$"),
-              paste0(num.species-nc[a]," of ",num.species," species"),
-              x
-              )
-        nc.levels <-
-            c(nc.levels,
-              paste0(num.species-nc[a]," of ",num.species," species"))
-    }
-    return(factor(x,levels=nc.levels))
-}
-
-cpt.distribution.shape <- 
-    rbindlist(cpt.distribution.shape)
-
-cpt.distribution.shape <- 
-    cpt.distribution.shape[,`Minimum Expression in`:=
-                                fix.expression.names(`Minimum Expression in`)]
-
-setnames(cpt.distribution.shape,
-         "FPKM",
-         "Expression (FPKM)")
-
-print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],
-             aes(x=`Expression (FPKM)`,color=`Minimum Expression in`,y=num_genes))
-      + geom_line(lwd=1)
-      + xlab("Minimum Expression cutoff (FPKM)")
-      + ylab("Genes with N-th Minimum Expression")
-      + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
-                      labels = trans_format("log10", math_format(10^.x)))
-      + scale_x_log10(breaks = trans_breaks("log10", function(x) 10^x),
-                      labels = trans_format("log10", math_format(10^.x)))
-      + geom_vline(xintercept=10)
-      + geom_segment(mapping=aes(y=V1,
-                         yend=V1,
-                         xend=10,
-                         x=1000,
-                         color=`Minimum Expression in`),
-                     data=cpt.distribution.shape[`Expression (FPKM)` > 10,
-                                              max(num_genes),`Minimum Expression in`],
-                     arrow=arrow(length=unit(0.5,"cm"))
-                   )
-      + geom_text(mapping=aes(x=1000,
-                      y=V1,
-                      label=paste0(V1," genes"),override=TRUE),
-                  data=cpt.distribution.shape[`Expression (FPKM)` > 10,
-                      max(num_genes),`Minimum Expression in`],
-                  show.legend=FALSE,
-                  hjust=0
-                  )
-      + guides(color=guide_legend(title="Minimum\nExpression in"))
-      )
-@ 
-\caption{Distribution shape of placenta core genes. The vertical axis
-  shows the number of 1:1 orthologous genes included in the core
-  placenta transcriptome if %
-  \Sexpr{cpt.distribution.shape[,levels(`Minimum Expression in`)[1]]}
-  %
-  (or %
-  \Sexpr{cpt.distribution.shape[,levels(`Minimum Expression in`)[2]]},
-  % 
-  etc.) have a minimum expression in FPKM equal to the horizontal axis.
-  For example, if the cutoff is set to 10 (vertical black line), 
-  % 
-  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
-    & `Minimum Expression in` == '14 of 14 species',max(num_genes)]}
-  %
-  genes would be included if the FPKM of a set of 1:1 orthologs of all 14 species must be at
-  least 10, %
-  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
-    & `Minimum Expression in` == "13 of 14 species",max(num_genes)]}
-  %
-  1:1 orthologs would be included if 13 of 14 species have an
-   with an FPKM of at least 10, and 
-  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
-    & `Minimum Expression in` == '10 of 14 species',max(num_genes)]}
-  genes would be included if 10 of 14 species
-  must have an FPKM of at least 10.}
-\label{fig:core_placenta_transcriptome_shape}
-\end{figure}
-
 \clearpage

### 2016-11-14 manuscript/acronyms.tex
add more genes
add Gene Set Enrichement acronym
discuss start of OUCH analysis
add discussion of tissue specificity index
add sra archive
add GH acronym
use PSG and PAG acronyms; update introduction
discuss go results for core placenta transcriptome
prep for genome biology submission
change capitalization of GH
switch to acro from acronym

@@
-\acrodef{EV}[EV]{Extracellular Vesicle}
-\acrodef{SVM}{Support Vector Machine}
-\acrodef{ROC}{Receiver Operating Characteristic}
-\acrodef{KNN}{K Nearest Neighbors}
-\acrodef{SRA}{Sequence Read Archive}
-\acrodef{PRG2}[\textit{PRG2}]{Proteoglycan 2, bone marrow}
-\acrodef{FPKM}{Fragments Per Kilobase of exon per Million reads}
-\acrodef{CSH1}[\textit{CSH1}]{chorionic somatomammotropin hormone 1}
-\acrodef{CSH2}[\textit{CSH2}]{chorionic somatomammotropin hormone 2}
-\acrodef{CSHL1}[\textit{CSHL1}]{chorionic somatomammotropin hormone-like 1}
-\acrodef{GH2}[\textit{GH2}]{growth hormone 2}
+\DeclareAcronym{EV}{short=EV,long=Extracellular Vesicle}
+\DeclareAcronym{SVM}{long=Support Vector Machine}
+\DeclareAcronym{ROC}{long=Receiver Operating Characteristic}
+\DeclareAcronym{KNN}{long=K Nearest Neighbors}
+\DeclareAcronym{SRA}{long=Sequence Read Archive}
+\DeclareAcronym{PRG2}{short=\textit{PRG2},long=Proteoglycan 2, bone marrow}
+\DeclareAcronym{FPKM}{long=Fragments Per Kilobase of exon per Million reads}
+\DeclareAcronym{CSH1}{short=\textit{CSH1},long=chorionic somatomammotropin hormone 1}
+\DeclareAcronym{CSH2}{short=\textit{CSH2},long=chorionic somatomammotropin hormone 2}
+\DeclareAcronym{CSHL1}{short=\textit{CSHL1},long=chorionic somatomammotropin hormone-like 1}
+\DeclareAcronym{GH2}{short=\textit{GH2},long=growth hormone 2}
+\DeclareAcronym{CEA}{long=carcinoembryonic antigen}
+\DeclareAcronym{HSD3B1}{short=\textit{HSD3B1},long=3-β Hydroxysteroid dehydrogenase 1}
+\DeclareAcronym{GSE}{long=Gene Set Enrichment}
+\DeclareAcronym{OU}{long=Ornstein-Uhlenbeck}
+\DeclareAcronym{OUCH}{long=\ac{OU} Comparative Hypotheses}
+\DeclareAcronym{TSI}{long=Tissue Specificity Index}
+\DeclareAcronym{SRA}{long=Sequence Read Archive}
+\DeclareAcronym{GH}{long=Growth Hormone}
+\DeclareAcronym{PSG}{long=Pregnancy-Specific Glycoprotein}
+\DeclareAcronym{PAG}{long=Pregnancy-Associated Glycoprotein}
+\DeclareAcronym{GO}{long=Gene Ontology}
+\DeclareAcronym{FDR}{long=False Discovery Rate}
 

### 2016-11-14 manuscript/manuscript.Rnw
switch to acro from acronym
use bw/gene expression in core plot

@@
 %\usepackage{setspace}
-\usepackage{acronym}
+\usepackage{acro}
 \usepackage{array}
@@
       xlab("Human Ortholog Name")+
-      ylab("FPKM")+
+      ylab("Gene Expression (FPKM)")+
+      theme_bw() +
       theme(legend.position=c(0.7,0.15),

### 2016-11-14 manuscript/manuscript.Rnw
it's actin y
remove discussion of brain size

@@
 
-\subsubsection{Invasion: EGF and Auxin Y in the Core Placenta}
+Another annexin which is highly expressed in the core placenta is
+\textit{ANXA5}, a gene which was originally identified as
+
+\subsubsection{Invasion: EGF and Actin Y in the Core Placenta}
 
@@
 
-\clearpage
-
 \section{Discussion}
@@
 
-\subsubsection{Invasion: EGF and Auxin Y in the Core Placenta}
+\subsubsection{Invasion: EGF and Actin Y in the Core Placenta}
 
@@
 \cite{Carter.ea2011:Evolutionofinvasiveplacentation,Robillard.ea2002:Evolutionaryadaptationstopre-eclampsiaeclampsia}.
-\human have the largest brain in terms of relative size and this
-differs markedly from closely related apes
-\cite{Boddy.ea2012:Comparativeanalysisofencephalization}. Deep
-penetration of the myometrium occurs in chimpanzees and gorillas, but
-this penetration may be accelerated in humans
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
 

### 2016-11-14 manuscript/manuscript.Rnw
discuss Tlr9/7 response in mice

@@
   \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-  core genes which are expressed (FPKM $\ge 10$) in all placentas.
+  core genes which are expressed (FPKM $\ge 10$) in all placentas,
 
   including genes with immune-modulating properties (\textit{ANXA2},
-  \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}) and
-  cell-cell interactions (\textit{LAMC1}, \textit{LUM}, and
-  \textit{LGALS1}).
+  \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}), cell-cell
+  interactions (\textit{LAMC1}, \textit{LUM}, and \textit{LGALS1}),
+  and trophoblast fusion (\textit{ANXA5} and \textit{ANXA1}).
 
@@
 \textit{UNC93B1} which is involved in the balancing between Tlr7 and
-Tlr9 immune response in mice\DLA{cite TLR7/TLR9 paper}, and
+Tlr9 immune response in
+mice~\cite{Brinkmann.ea2007:interactionbetweenERmembrane}, and
 \textit{WNT5A}, which inhibits the invasion of the trophoblast cell
@@
 disturbances) %
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current,ACOG2013:hypertensionpregnancy}. Some cases of
+\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
 pre-eclampsia can develop into eclampsia, a more severe condition
@@
 regulators as well as the inability to sufficiently suppress the
-maternal immune response. Although there are speculative cases of
-eclampsia in great apes
+maternal immune response. %
+Although there are speculative cases of eclampsia in great apes
 \cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},

### 2016-11-15 data/combine_fpkm.R
update human_alignment for homo and smonkey

@@ -134,6 +134,9 @@ combined.fpkm <-
     gtf.types[combined.fpkm]
 setkey(combined.fpkm,"tracking_id")
 
+combined.fpkm[species=="homo sapiens", human_alignment_symbol:=human_name]
+combined.fpkm[species=="ateles fusciceps", human_alignment_symbol:=human_name]
+
 if (!all(grepl("_per_sample",output.file))) {
     combined.fpkm <-
         combined.fpkm[!duplicated(tracking_id)]

### 2016-11-15 data/homology_table.R
fix homology table

@@ -16,11 +16,13 @@ homology.table <-
 ### ditch rows which are all zero
 
 gene.sum <-
-    apply(homology.table,1,function(x){sum(as.numeric(x[2:9]),na.rm=TRUE)})
+    apply(homology.table,1,function(x){sum(as.numeric(x[-1]),na.rm=TRUE)})
 homology.table <- homology.table[gene.sum>0 & !is.na(homology.table$human_name),]
-rownames(homology.table) <-
+rownames.homology.table <-
     homology.table[,1]
 homology.table <- data.matrix(homology.table[,-1])
+rownames(homology.table) <-
+    rownames.homology.table
 
 save(homology.table,
      file=args[length(args)])

### 2016-11-15 manuscript/manuscript.Rnw
discuss annexin in the context of cell fusion

@@
 
-Another annexin which is highly expressed in the core placenta is
-\textit{ANXA5}, a gene which was originally identified as
+Another annexin which is highly expressed in all term placentas is
+\textit{ANXA5}, which was originally identified as a an anti-coagulant
+protein, but is also involved in syncitium formation, membrane repair,
+and reduction of spontaneous
+abortion~\cite{Ueki.ea2012:Lossofmaternalannexin}. Other notable
+syncytiotrophoblast fusion proteins, Syncytin-1 (\textit{ERVW-1}) and -2
+(\textit{ERVFRD-1}) were not present in the core placenta, as they (or
+their homologs) are not highly expressed in \dog, \bonobo, or \sheep,
+nor was \textit{MFSD2A}, the Syncytin-2 receptor (\ac{FPKM} $\ge 10$
+only in \human, \mouse, and \dog).
 
@@
 placenta.
+
+\subsubsection{Formation of the syncytiotrophoblast}
+
+While Syncitin-1 and -2 are implicated in the formation of the
+syncytiotrophoblast in
+\human~\cite{Rawn.ea2008:evolutionregulationandfunction} and homologs
+of \textit{ERVFRD-1} and \textit{ERVW-1} are widely expressed in most
+of the placentas analyzed here, they are not expressed in \bonobo,
+\sheep, and \dog, even though these mammals form
+syncytiotrophoblasts~\cite{Mess.ea2006:Evolutionarytransformationsoffetal}.
+Because the formation of syncytiotrophoblast is the stem state of
+Eutheria, Afrotheria, Euarchontoglires, and
+Laurasiatheria~\cite{Mess.ea2006:Evolutionarytransformationsoffetal},
+the gene(s) responsible for cytotrophoblast fusion to form
+syncytiotrophoblast is likely to be highly expressed in the core
+placenta. \textit{ANXA5} is known to be expressed at multiple stages
+of placentation, and is capable of repairing membrane defects and
+promoting the fusion of cells~\DLA{Cite more papers here which define
+  this.} in the presence of Ca$^{2+}$. While there are likely other
+genes which are responsible for the positioning of cytotrophoblast
+cells in close proximity to start the fusion process, \textit{ANXA5}
+is likely to be essential for the completion of fusion.
+
+\DLA{Possibility of membrane repair due to maternal immune response as
+  well during this process, as well as repair of membrane due to high
+  secretion and adsoroption of nutrients/waste in the placenta.}
+
+% Cows have BERV-K1 and -K2, but it's not annotated in the ENSEMBL
+% build of the cow genome that we have
 

### 2016-11-15 manuscript/manuscript.Rnw
tweak distribution plot

@@
                   )
-      + guides(color=guide_legend(title=paste0("Min. Expression in\nn of ",num.species," Species",
+      + guides(color=guide_legend(title=paste0("Min. Expression in\nn of ",num.species," Species"),
                                   keywidth=unit(3,"mm"),
-                                  keyheight=unit(3,"mm")
+                                  keyheight=unit(3,"mm"),
+                                  title.hjust=0.5,
+                                  label.position="right"
                                   ))
       + theme_bw()
-      + theme(legend.position=c(1,1),
+      + theme(legend.position=c(0.98,0.98),
               legend.margin=unit(0,"mm"),
@@
 cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
-\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>1 & 
+\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & 
 `Min. Exp. in` == '14 of 14',max(num_genes)]} %
@@
 
-\subsubsection{Annexin Complexes}
+\subsubsection{Immune Regulation}
 The transcriptomes from multiple different species enabled us to

### 2016-11-17 figures/heatmap_gene_by_functional_grouping/heatmap_gene_by_functional_grouping_plot.R
remove outdated figures

@@ -1,96 +1,0 @@
-args <- c("../../data/combined_fpkm",
-          "go_functional_grouping"
-          )
-args <- commandArgs(trailingOnly=TRUE)
-
-library(data.table)
-library(reshape2)
-library(ggplot2)
-library(grid)
-
-
-### load("../../data/combined_fpkm")
-load(args[1])
-### load("go_functional_grouping")
-load(args[2])
-
-pdf(file=args[length(args)],width=10,height=20)
-data.matrix.dt <- function(x){
-    temp <- data.matrix(as.data.frame(x)[,-1])
-    rownames(temp) <- as.data.frame(x)[,1]
-    temp
-}
-
-gene.expression.nozero <- function(genes,expression,min.exp=5,log=FALSE) {
-    agg.fun <- function(x){x[1]}
-    if (log) {
-        agg.fun <- function(x){log2(x[1]+1)}
-    }
-    temp <-
-        data.matrix.dt(dcast(expression[human_name %in% genes,],
-                             human_name~species,
-                             value.var="FPKM",
-                             fun.aggregate=agg.fun))
-    return(temp[apply(temp,1,
-                      function(x){
-                          sum(x,na.rm=TRUE)
-                      })> agg.fun(min.exp),])
-}
-
-
-
-go.terms <- c("^blood vessel development$",
-              "^extracellular matrix$",
-              "^tube morphogenesis$",
-              "apoptosis",
-              "lumen$")
-go.genes <- list()
-for (i  in 1:length(go.terms)) {
-    go.term <- go.terms[i]
-    go.genes[[go.term]] <-
-        go.functional.groups[grepl(go.term,name_1006),hgnc_symbol]
-}
-# pushViewport(viewport(layout=grid.layout(ncol=1,nrow=length(go.terms),
-#                            heights=sapply(go.genes,length))))
-for (i  in 1:length(go.terms)) {
-    go.term <- go.terms[i]
-    go.expression.nozero <-
-        gene.expression.nozero(genes=go.genes[[go.term]],
-                               expression=combined.fpkm,
-                               log=TRUE
-                               )
-    go.exp.nz.long <-
-        melt(go.expression.nozero,
-             value.name="fpkm"
-             )
-    colnames(go.exp.nz.long)[1:2] <- c("gene","species")
-
-    ## reorder by expression level
-    go.exp.nz.long$gene <-
-        reorder(go.exp.nz.long$gene,
[…]

### 2016-11-17 figures/hormone_genes/hormone_genes.R

@@ -1,131 +1,0 @@
-library(data.table)
-library(reshape2)
-library(ggplot2)
-library(grid)
-library(scales)
-
-args <- commandArgs(trailingOnly=TRUE)
-
-### ../../data/combined_fpkm
-load(args[1])
-### ../../data/go_gene_association
-load(args[2])
-### ../species_ordering.R
-source(args[3])
-
-
-plot_type <- "onepage"
-if (any(grepl("multipage",args[length(args)]))) {
-    plot_type <- "multipage"
-}
-pdf(args[length(args)],width=8,height=8)
-
-hormone.genes <-
-    go.gene.association[go_name=="hormone activity" & go_namespace=="molecular_function",gene_name]
-data.matrix.dt <- function(x){
-    temp <- data.matrix(as.data.frame(x)[,-1])
-    rownames(temp) <- as.data.frame(x)[,1]
-    temp
-}
-
-gene.expression.nozero <- function(genes) {
-    temp <-
-        data.matrix.dt(dcast(combined.fpkm[human_name %in% genes,],
-                             human_name~species,
-                             value.var="FPKM",
-                             fun.aggregate=function(x){x[1]}))
-    temp <-
-        temp[apply(temp,1,function(x){sum(x,na.rm=TRUE)})>5,]
-    return(temp)
-}
-
-### hormone genes
-hormone.genes <- c("CGA","CGB","CGB1","CGB2","CGB5","CGB8","CSH1","CSH2","CSHL1","CYP19A1",
-                   "GH2","GHRH","HSD3B1","NPPB","NPPC","PRL","RLN1")
-hormone.expression.nozero <-
-    gene.expression.nozero(hormone.genes)
-
-
-
-hor.exp.long <- melt(log2(hormone.expression.nozero+1)[rev(1:nrow(hormone.expression.nozero)),])
-colnames(hor.exp.long)[1:3] <- c("gene","species","fpkm")
-hor.exp.long <- hor.exp.long[hor.exp.long$gene!="IGF2",]
-
-### galectin genes
-galectin.genes <- unique(grep("^LGALS",combined.fpkm[,human_name],value=TRUE))
-
-galectin.expression.nozero <-
-    gene.expression.nozero(galectin.genes)
-
-gal.exp.long <- melt(log2(galectin.expression.nozero+1)[rev(1:nrow(galectin.expression.nozero)),])
[…]

### 2016-11-17 go_annotation/go_placenta_classification_analysis.R
add go placenta classification analysis

@@ -1,0 +1,109 @@
+library("data.table")
+library(doMC)
+ncore = multicore:::detectCores()
+registerDoMC(cores = ncore)
+library("topGO")
+library("org.Hs.eg.db")
+
+args <- c("placenta_classification","housekeeping_genes_superset","go_bos_taurus_mf")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+load(args[1])
+load(args[2])
+
+
+
+do.go.analysis <- function(species.name,go.ontology,top.100=TRUE,exclude.housekeeping=FALSE) {
+
+    all.genes <- combined.fpkm[species==species.name & !is.na(human_name),]
+    all.genes[,egid:=as.vector(unlist(sapply(mget(all.genes[,human_name],
+                   org.Hs.egSYMBOL2EG,ifnotfound=NA),function(x){x[1]})))]
+    all.genes <- all.genes[!is.na(egid),]
+
+    analysis.type <- ""
+    if (exclude.housekeeping) {
+        all.genes <- all.genes[!(human_name %in% housekeeping.genes.superset[,gene_short_name]),]
+        analysis.type <- " No Housekeeping"
+    }
+    
+    all.genes[,rank:=rank(-mean_fpkm)]
+
+    if (top.100) {
+        analysis.type <- paste0("Top 100",analysis.type)
+        gene.list <- all.genes[,factor(as.integer(rank <=100))]
+    } else {
+        analysis.type <- paste0("Top 1%",analysis.type)
+        gene.list <- all.genes[,factor(as.integer(rank <= (sum(mean_fpkm>1)*0.01)))]
+    }
+    names(gene.list) <- all.genes[,egid]
+    
+    godata <- new("topGOdata",
+                  ontology=go.ontology,
+                  allGenes=gene.list,
+                  annot=annFUN.org,
+                  mapping="org.Hs.eg.db")
+    test.stat <- new("classicCount", testStatistic = GOFisherTest, name = "Fisher test")
+    result.fisher <- getSigGroups(godata,test.stat)
+    res.table <- 
+        data.table(GenTable(godata,
+                            classic=result.fisher,
+                            orderBy="classic",
+                            topNodes=length(result.fisher@score)))
+    res.table[,classic:=as.numeric(gsub("< ","",classic))]
+    setnames(res.table,"classic","p value")
+    res.table[,FDR:=p.adjust(method="BH",`p value`)]
+    res.table[,Ontology:=go.ontology]
[…]

### 2016-11-17 manuscript/manuscript.Rnw
change title again
show the housekeeping genes which are core

@@
 
-\title{The evolution of the transcriptome of mammalian placentas}%
+\title{The core transcriptome of mammalian placentas and the
+  divergence of expression with placental shape}%
 %
@@
     core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
-                                            Species),]
+                                            Species) & housekeeping==FALSE,]
 cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),

### 2016-11-17 manuscript/supplemental_material.Rnw
switch to acro from acronym
show the housekeeping genes which are core

@@
 % \usepackage{txfonts}
-\usepackage{acronym}
+\usepackage{acro}
 \usepackage{array}
@@
 
+\begin{figure}[tbh]
+<<core.placenta.transcriptome.housekeeping,fig.width=5,fig.height=10,out.height="0.8\\textheight",echo=FALSE>>=
+cptlh <-
+    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
+                                            Species) & housekeeping==TRUE,]
+cptlh <- cptlh[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
+                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
+                                          capfirst(names(species.names))))]
+
+print(ggplot(cptlh,aes(x=human_name,y=FPKM))+
+      geom_boxplot(outlier.shape=NA)+
+      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
+      geom_point(aes(color=Species,shape=Species))+
+      scale_shape_manual("Species",
+                         guide=guide_legend(#nrow=1,
+                                            keywidth=unit(3,"mm"),
+                                            keyheight=unit(3,"mm"),
+                                            label.theme=element_text(face="italic",
+                                                                     angle=0,
+                                                                     size=9),
+                                            title.position="top"
+                                            ),
+                         values=0:18)+
+      # scale_color_discrete(NULL)+
+      coord_flip()+
+      xlab("Human Ortholog Name")+
+      ylab("Gene Expression (FPKM)")+
+      theme_bw() +
+      theme(legend.position=c(0.7,0.15),
+            axis.text.y=element_text(face="italic",size=5),
+            legend.margin=unit(0,"mm"),
+            legend.justification="left",
+            plot.margin=unit(c(0,0,0,0),"mm")
+            ))
+@ 
+\caption{Housekeeping genes excluded from the Core Placenta Transcriptome.
+  \Sexpr{cptlh[,length(unique(human_name))]}
+  1:1 \human ortholog genes expressed ($\mathrm{FPKM} \ge 10$) in the
+  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
+  studied which are housekeeping genes~%
+  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
+  ordered (from top to bottom) by median expression.}
+\label{fig:core_placenta_transcriptome_housekeeping}
+\end{figure}
+
+\clearpage
+
 \begin{figure}

### 2016-11-17 data/placenta_core_transcriptome.R

@@ -54,12 +54,15 @@ setkey(combined.fpkm.wide,"human_name")
 core.placenta.transcriptome.long <- 
     data.table(melt(combined.fpkm.wide[all_expression >= 10,
                                        ][order(-median_expression)
-                                         ][!(human_name %in% 
-                                                 housekeeping.genes.superset[,gene_short_name])],
+                                         ],
                     id.vars="human_name",
                     variable.name="Species",
                     value.name="FPKM"
                     ))
+core.placenta.transcriptome.long[,housekeeping:=FALSE]
+core.placenta.transcriptome.long[human_name %in% 
+                                 housekeeping.genes.superset[,gene_short_name],
+                                 housekeeping:=TRUE]
 core.placenta.transcriptome.long[,human_name:=
                                      reorder(human_name,
                                              combined.fpkm.wide[core.placenta.transcriptome.long[,human_name],median_expression])]

### 2016-11-17 manuscript/supplemental_material.Rnw
display the whole human-lineage specific table

@@
 load("../data/placenta_lineage_specific")
+load("../data/human_mouse_lineage_specific")
 source("/home/don/projects/xtable/pkg/R/print.xtable.R")
 
-placenta.lineage.specific.table <- function(analysis,output=TRUE) {
+placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=40,orderbyfc=FALSE,fdr.cutoff=1,caption=NULL,label=NULL) {
     temp.table <- 
-        head(placenta.lineage.specific.analysis[analysis.name==analysis &
-                                                p.value <= 0.05][order(p.value),
-                                                                 list(human_name,p.value,fdr,fc),],
-             n=5)
+        lineage.table[analysis.name==analysis &
+                      p.value <= 0.05 &
+                      fdr <= fdr.cutoff][order(p.value),
+                                       list(human_name,p.value,fdr,fc),]
     temp.table <- data.table(temp.table)
@@
     if (output) {
-        display.table <- temp.table[order(`p (t-test)`),]
+        display.table <- temp.table[`FDR`<=fdr.cutoff,][order(`p (t-test)`),]
+        if (orderbyfc) {
+            display.table <- temp.table[order(-abs(`Fold Change`)),]
+        }
+        if (!is.na(members)) {
+            members <- min(c(members,nrow(display.table)))
+            display.table <- display.table[1:members,]
+        }
+
         display.table <-
@@
                                                   "}")]
+        display.table[,`Human Name`:=NULL]
+        # setnames(display.table,"Human Symbol","\\multicolumn{1}{p{0.5in}}{Human Symbol}")6
+        setnames(display.table,"Fold Change","$\\text{log}_2$ FC")
         print(xtable(display.table,
-                     display=c("s","s","s","e","g","f"),
+                     display=c("s","s","e","g","f"),
+                     caption=caption,
+                     label=label,
+                     align=c("c","p{0.7in}","r","r","r")
                      ),
@@
               floating=FALSE,
+              tabular.environment=ifelse(is.na(members),"longtable","tabular"),
               math.style.exponents=TRUE,
               sanitize.text.function=genes.table.sanitize,
+              size="\\small",
               include.rownames=FALSE)
@@
 
+<<human.lineage.specific,echo=FALSE,results="asis">>=
+hum.all.lin.spec <-
+    placenta.lineage.specific.table("hum.all",
+                                    lineage.table=human.mouse.lineage.specific.analysis,
+                                    members=NA,
+                                    caption="\\lineagetableexplanationtypeiii{\\human}{all other species}",
+                                    label="tab:huamn_linage_specific_all"
+                                    )
+setkey(hum.all.lin.spec,"Human Symbol")
+@
+
 \begin{table}[hb]
@@
 @ 
-\caption{\lineagetableexplanation{the rodentia order}{all other species}}
+\caption{\lineagetableexplanationtypeiii{the rodentia order}{all other species}}
 \label{tab:rodentia_lineage_specific}
@@
 cptg.h <- core.placenta.transcriptome.genes.shape[(human_name %in% housekeeping.genes.superset[,gene_short_name])]
+num.species <- star.logs.wide[,length(unique(species))]
+
+calculate.distribution.shape <- 
+    function(min.vector) {
+        min.vector <- sort(min.vector,decreasing=TRUE)
+        temp <- data.table(FPKM=min.vector,num_genes=1:length(min.vector))
+        temp <- temp[order(-num_genes)][!duplicated(FPKM)]
+        return(temp)
+    }
 

### 2016-11-17 manuscript/manuscript.Rnw
add FPKM 1 text to the figure

@@
                   )
+      + geom_segment(mapping=aes(yend=V1,
+                         y=100,
+                         xend=1,
+                         x=1,
+                         color=`Min. Exp. in`),
+                     data=cpt.distribution.shape[`Expression (FPKM)` > 1 &
+                                                 `Min. Exp. in` == '14 of 14',
+                                                 max(num_genes),`Min. Exp. in`],
+                     arrow=arrow(length=unit(0.5,"cm")),
+                     show.legend=FALSE
+                     )
+      + geom_text(mapping=aes(x=1,
+                      y=100,
+                      label=paste0(V1," genes"),
+                      override=TRUE),
+                  angle=90,
+                  data=cpt.distribution.shape[`Expression (FPKM)` > 1 &
+                                              `Min. Exp. in` == '14 of 14',
+                                              max(num_genes),`Min. Exp. in`],
+                  show.legend=FALSE,
+                  hjust=1
+                  )
       + guides(color=guide_legend(title=paste0("Min. Expression in\nn of ",num.species," Species"),

### 2016-11-17 manuscript/manuscript.Rnw
tweak discussion of annexins and other genes

@@
 into the endometrium, including the EGFR1 pathway and the Actin Y
-pathway. Two other pathways, the mPR/Progesterone pathway and Rho cell
-motility pathways, feed into the Actin Y pathway as well. Supplemental
-\cref{fig:egf_kegg_pathway} shows the EGF pathway in humans with the
-median expression in FPKM of genes in the pathway shown. Notably, many
-of the genes on the GRB2→MYC pathway are strongly expressed; their
-activation leads to an increase in migration and invasion in response
-to EGF binding to EGFR1. The Actin cytoskeleton pathway is depicted in
-Supplemental \cref{fig:kegg_actin}, which shows that \textit{FN1} and
-\textit{ITG} are highly expressed, which activate Rac, and potentially
-lead to the formation of focal adhesions and lamellipodia formation
-through their inhibition of GSN and HSPC300.
+pathway. Two other significant pathways, the mPR/Progesterone pathway
+and Rho cell motility pathways, feed into the Actin Y pathway as well.
+Supplemental \cref{fig:egf_kegg_pathway} shows the EGF pathway in
+humans with the median expression in FPKM of genes in the pathway
+shown. Notably, many of the genes on the GRB2→MYC pathway are strongly
+expressed; their activation leads to an increase in migration and
+invasion in response to EGF binding to
+EGFR1~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}. The Actin
+cytoskeleton pathway is depicted in Supplemental
+\cref{fig:kegg_actin}, which shows that \textit{FN1} and \textit{ITG}
+are highly expressed, which activate Rac, and potentially lead to the
+formation of focal adhesions and lamellipodia formation through their
+inhibition of GSN and HSPC300. Other compoenents of cell-cell
+adhesion, including \textit{F11R} which forms tight junctions in
+epithelia, proteglycans \textit{SDC4} and \textit{LUM}, and the
+laminin \textit{LAMC1}.
 
@@
 mice~\cite{Brinkmann.ea2007:interactionbetweenERmembrane}, and
-\textit{WNT5A}, which inhibits the invasion of the trophoblast cell
-line HTR8/SVneo \textit{in vitro} \DLA{Cite this paper after reading
-  the comments on it (PMID: 26865089)}.
+\textit{WNT5A}, which is a member of the Wnt ligand family whose
+singaling pathway has been implicated in motility in human trophoblast
+cells~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}.
 
@@
 
-\subsubsection{Immune Regulation}
+The core placenta transcriptome is enriched in multiple genes which
+are critical for the function of the placenta, including vesicular
+transport, immune regulation, cell fusion and membrane repair, and
+invasion.
+
+\subsubsection{Immune Regulation and Annexins}
 The transcriptomes from multiple different species enabled us to
@@
 of placentation, and is capable of repairing membrane defects and
-promoting the fusion of cells~\DLA{Cite more papers here which define
-  this.} in the presence of Ca$^{2+}$. While there are likely other
-genes which are responsible for the positioning of cytotrophoblast
-cells in close proximity to start the fusion process, \textit{ANXA5}
-is likely to be essential for the completion of fusion.
-
-\DLA{Possibility of membrane repair due to maternal immune response as
-  well during this process, as well as repair of membrane due to high
-  secretion and adsoroption of nutrients/waste in the placenta.}
+promoting the fusion of
+cells~\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing} in
+the presence of Ca$^{2+}$. While there are likely other genes which
+are responsible for the positioning of cytotrophoblast cells in close
+proximity to start the fusion process, \textit{ANXA5} is likely to be
+essential for the completion of fusion and the maintenance of membrane
+repair during the massive flux of nutrients and waste in the placenta.
+
+% \DLA{Possibility of membrane repair due to maternal immune response as
+%   well during this process, as well as repair of membrane due to high
+%   secretion and adsoroption of nutrients/waste in the placenta.}
 
@@
 
-\subsubsection{Invasion: EGF and Actin Y in the Core Placenta}
+\subsubsection{Invasion: EGF and CD9 in the Core Placenta}
+
+The high level of expression of components of the EGFR signaling
+pathway in the core placenta transcriptome (including \textit{GRB2}, \textit{EGFR}, \textit{MAPK3}, \textit{MAP2K2}, \textit{SHC1}, \textit{AKT3}, \textit{AKT2}, \textit{ERBB3}
+
+% Role of CD9 in invasion; commonality between placentas
 

### 2016-11-17 manuscript/manuscript.Rnw
move human specific up
discuss EGFR signalling

@@
 values in \ac{FPKM}
-\cite{Trapnell.ea2010:Transcriptassemblyandquantification} were
+% \cite{Trapnell.ea2010:Transcriptassemblyandquantification}
+were
 quantified using Cufflinks v2.2.1
@@
 \cow, \dog, \elephant, \mouse, and \human were determined using
-Ensembl v.80 gene trees via Biomart
-\cite{Kasprzyk2011:BioMartdrivingparadigmchange} using the biomaRt R
-package
-\cite{Durinck.ea2009:Mappingidentifiersforintegration,Durinck.ea2005:BioMartandBioconductorpowerful}.
+Ensembl v.80 gene trees via Biomart using the biomaRt R package
+\cite{Durinck.ea2009:Mappingidentifiersforintegration}. %
 A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
@@
 formation of focal adhesions and lamellipodia formation through their
-inhibition of GSN and HSPC300. Other compoenents of cell-cell
+inhibition of GSN and HSPC300. Other components of cell-cell
 adhesion, including \textit{F11R} which forms tight junctions in
@@
 
+\subsection{\human Specific Expression Shifts and the Evolution of Pre-eclampsia}
+
+The \human placenta is distinctive from many other primates in that it
+possesses a deeply invasive trophoblast that penetrates the inner
+myometrium of the uterus
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
+hypothesized that this feature is related to the increased demand for
+resources by the developing fetus to feed the expanded size of the
+human brain
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation,Robillard.ea2002:Evolutionaryadaptationstopre-eclampsiaeclampsia}.
+
+The near exclusivity of deep trophoblast invasion in humans has also
+led many to speculate on the origin of various disorders, which are
+almost exclusively symptomatic of human pregnancy. For example,
+pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
+the United States
+\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
+can be attributed to the lack of adequate invasion of maternal tissue,
+and is noted by the onset of hypertension and proteinuria (or in the
+absence of proteinuria, thrombocytopenia, impaired liver function,
+renal insufficiency, pulmonary edema, or cerebral or visual
+disturbances) %
+\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
+pre-eclampsia can develop into eclampsia, a more severe condition
+resulting in seizures during pregnancy and death. Lack of invasion in
+pre-eclampsia may be attributed to the inefficiency of growth
+regulators as well as the inability to sufficiently suppress the
+maternal immune response. %
+Although there are speculative cases of eclampsia in great apes
+\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
+it is exceedingly rare in these species
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
+
+The identification of multiple genes (\textit{CRH}, \textit{ADAM12},
+\textit{KISS1}, \textit{PAPPA}, \textit{IL1RL1}, and \textit{PLAC1};
+\cref{tab:human_lineage_specific}) which are associated with
+pre-eclampsia which have increased expression on the \human lineage
+may explain the larger proportion of pre-eclampsia observed in humans.
+However, there are still genes which are associated with pre-eclampsia
+which are also more highly expressed in the \mouse lineage such as
+\textit{Afp} and \textit{Gjb2}. Furthermore, the onset of
+pre-eclampsia is well before term; all of our placentas are term
+placentas and provide limited information about the transcriptome
+during the onset of pre-eclampsia.
+
 \subsection{Core placenta transcriptome}
@@
 complexes may illuminate how immune tolerance is maintained during
-placentation. The core placenta transcriptome is also enriched in
-genes involved in maintaining the structure of the placenta and
-transporting proteins, as would be expected given the function of the
-placenta.
+placentation.
 
@@
 
-The high level of expression of components of the EGFR signaling
-pathway in the core placenta transcriptome (including \textit{GRB2}, \textit{EGFR}, \textit{MAPK3}, \textit{MAP2K2}, \textit{SHC1}, \textit{AKT3}, \textit{AKT2}, \textit{ERBB3}
-
-% Role of CD9 in invasion; commonality between placentas
-
-\subsection{\human Specific Expression Shifts and the Evolution of Pre-eclampsia}
-
-The \human placenta is distinctive from many other primates in that it
-possesses a deeply invasive trophoblast that penetrates the inner
-myometrium of the uterus
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
-hypothesized that this feature is related to the increased demand for
-resources by the developing fetus to feed the expanded size of the
-human brain
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation,Robillard.ea2002:Evolutionaryadaptationstopre-eclampsiaeclampsia}.
-
-The near exclusivity of deep trophoblast invasion in humans has also
-led many to speculate on the origin of various disorders, which are
-almost exclusively symptomatic of human pregnancy. For example,
-pre-eclampsia is a human disease that occurs in 4\% of pregnancies in
-the United States
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Pre-eclampsia
-can be attributed to the lack of adequate invasion of maternal tissue,
-and is noted by the onset of hypertension and proteinuria (or in the
-absence of proteinuria, thrombocytopenia, impaired liver function,
-renal insufficiency, pulmonary edema, or cerebral or visual
-disturbances) %
-\cite{Chaiworapongsa.ea2014:Pre-eclampsiapart1current}. Some cases of
-pre-eclampsia can develop into eclampsia, a more severe condition
-resulting in seizures during pregnancy and death. Lack of invasion in
-pre-eclampsia may be attributed to the inefficiency of growth
-regulators as well as the inability to sufficiently suppress the
-maternal immune response. %
-Although there are speculative cases of eclampsia in great apes
-\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
-it is exceedingly rare in these species
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
-
-The identification of multiple genes (\textit{CRH}, \textit{ADAM12},
-\textit{KISS1}, \textit{PAPPA}, \textit{IL1RL1}, and \textit{PLAC1};
-\cref{tab:human_lineage_specific}) which are associated with
-pre-eclampsia which have increased expression on the \human lineage
-may explain the larger proportion of pre-eclampsia observed in humans.
-However, there are still genes which are associated with pre-eclampsia
-which are also more highly expressed in the \mouse lineage such as
-\textit{Afp} and \textit{Gjb2}. Furthermore, the onset of
-pre-eclampsia is well before term; all of our placentas are term
-placentas and provide limited information about the transcriptome
-during the onset of pre-eclampsia.
+The high expression of components of the EGFR signaling pathway in the
+core placenta transcriptome (including \textit{GRB2}, \textit{RALB})
+as well as annexins which are involved in the location of EGF receptor
+localization (\textit{ANXA1} and \textit{ANXA2}) in the core placenta
+is consistent with previous findingsshowing that alterations to the
+EGF signaling system change invasion of trophoblast cells into the
+spiral arteries~\cite{Armant.ea2015:Reducedexpressionofepidermal}.
+
+% TAGLN2 -- likely to be produced by the placenta as well; possibly
+% due to the smooth muscle transition?
 

### 2016-11-17 manuscript/manuscript.Rnw
discuss CD9 in the core placenta

@@
 spiral arteries~\cite{Armant.ea2015:Reducedexpressionofepidermal}.
+Another component of the core placenta transcriptome, \textit{CD9},
+has been shown to suppress \human extravillous trophoblast
+invasion~\cite{Matsumoto.ea2016:CD9suppresseshumanextravillous}, and
+we hypothesize that it is responsible for maintaining the balance
+between invasion and attachment in the other placentas examined here
+as well.
 

### 2016-11-18 manuscript/manuscript.Rnw
refer to transcriptome statistics

@@
 All transcriptomes were then aligned to the appropriate reference (or
-assembled) genome using STAR v2.4.2a
-\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} and expression
-values in \ac{FPKM}
+assembled) genome using STAR v2.4.2a~%
+\cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} (Supplemental
+\cref{fig:transcriptome_statistics,fig:transcriptome_reads_vs_read_length})
+and expression values in \ac{FPKM}
 % \cite{Trapnell.ea2010:Transcriptassemblyandquantification}
@@
 of reads mapping to the
-respective reference genome (\cref{tab:ref_genome_aligned_species}).
+respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
 %
@@
 %
-with an average of \Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
-of reads mapping to the
-respective reference genome (\cref{tab:ref_genome_aligned_species}).
-The expression of all genes with \ac{FPKM} $\ge 50$ is shown in Supplemental
+with an average of %
+\Sexpr{star.logs.wide[input.read.length < 40, mean(`% good reads`)]}\%
+%
+of reads mapping to the respective reference genome (Supplemental
+\cref{tab:ref_genome_aligned_species}). Read length was not a
+predictor of unmapped read percentage (Supplemental
+\cref{fig:transcriptome_reads_vs_read_length}). The expression of
+all genes with \ac{FPKM} $\ge 50$ is shown in Supplemental
 \cref{tab:all_ateles_fusciceps_placenta,tab:all_homo_sapiens_placenta,tab:all_pan_paniscus_placenta,tab:all_mus_musculus_placenta,tab:all_spalax_galili_placenta,tab:all_spalax_carmeli_placenta,tab:all_bos_taurus_placenta,tab:all_ovis_aries_placenta,tab:all_sus_scrofa_placenta,tab:all_equus_caballus_placenta,tab:all_canis_familiaris_placenta,tab:all_loxodonta_africana_placenta,tab:all_dasypus_novemcinctus_placenta,tab:all_monodelphis_domestica_placenta}.

### 2016-11-18 manuscript/manuscript.Rnw
mention per-species go analyses

@@
 @
+
+% \subsubsection{Per-species Gene Set and Pathway Enrichment}
+
+We performed \ac{gsea} on both the top 1\% of transcripts
+(Supplemental \cref{tab:go_terms_enriched_top_one_ateles_fusciceps,
+  tab:go_terms_enriched_top_one_homo_sapiens,
+  tab:go_terms_enriched_top_one_pan_paniscus,
+  tab:go_terms_enriched_top_one_mus_musculus,
+  tab:go_terms_enriched_top_one_spalax_galili,
+  tab:go_terms_enriched_top_one_spalax_carmeli,
+  tab:go_terms_enriched_top_one_bos_taurus,
+  tab:go_terms_enriched_top_one_ovis_aries,
+  tab:go_terms_enriched_top_one_sus_scrofa,
+  tab:go_terms_enriched_top_one_equus_caballus,
+  tab:go_terms_enriched_top_one_canis_familiaris,
+  tab:go_terms_enriched_top_one_loxodonta_africana,
+  tab:go_terms_enriched_top_one_dasypus_novemcinctus,
+  tab:go_terms_enriched_top_one_monodelphis_domestica}) and the top
+100 transcripts (Supplemental
+\cref{tab:go_terms_enriched_ateles_fusciceps,
+  tab:go_terms_enriched_homo_sapiens,
+  tab:go_terms_enriched_pan_paniscus,
+  tab:go_terms_enriched_mus_musculus,
+  tab:go_terms_enriched_spalax_galili,
+  tab:go_terms_enriched_spalax_carmeli,
+  tab:go_terms_enriched_bos_taurus, tab:go_terms_enriched_ovis_aries,
+  tab:go_terms_enriched_sus_scrofa,
+  tab:go_terms_enriched_equus_caballus,
+  tab:go_terms_enriched_canis_familiaris,
+  tab:go_terms_enriched_loxodonta_africana,
+  tab:go_terms_enriched_dasypus_novemcinctus,
+  tab:go_terms_enriched_monodelphis_domestica}); common processes
+include protein transport and viral transcription/expression
+(Supplemental \cref{tab:go_terms_all_species}).
 

### 2016-11-18 manuscript/supplemental_material.Rnw
move per-species go analyses up to match the paper order
remove lineage table

@@
 
-\section{Placenta Classification Analyses}
-
-<<placenta_pathway_egid,echo=FALSE,results="hide">>=
-load("../data/placenta_core_transcriptome_complete")
-plac_path_egid <- function(column) {
-    expr <- placenta.core.transcriptome.complete[!is.na(egid),eval(column)]
-    names(expr) <- placenta.core.transcriptome.complete[!is.na(egid),egid]
-    expr
-}
-
-plac_path_genes <- function(genes,column) {
-    result <- as.numeric(NULL)
-    result[!is.na(genes)] <-
-        placenta.core.transcriptome.complete[na.omit(genes),eval(column)]
-    names(result) <- names(genes)
-    result
-}
-
-@
-
-<<plot_pathway,echo=FALSE,result="hide">>=
-library("KEGGgraph")
-library("plotrix")
-makeAttr <- function(graph, default, valNodeList) {
-  tmp <- nodes(graph)
-  x <- rep(default, length(tmp)); names(x) <- tmp
-
-  if(!missing(valNodeList)) {
-    stopifnot(is.list(valNodeList))
-    allnodes <- unlist(valNodeList)
-    stopifnot(all(allnodes %in% tmp))
-    for(i in seq(valNodeList)) {
-      x[valNodeList[[i]]] <- names(valNodeList)[i]
-    }
-  }
-  return(x)
-}
-graph_attrs <- function(graph,nodecolorcolumn,cutpoint=1) {
-    outs <- sapply(edges(graph), length) > 0
-    ins <- sapply(inEdges(graph), length) > 0
-    ios <- outs | ins
-    ## translate the KEGG IDs into Gene Symbol
-    ioGeneID <- translateKEGGID2GeneID(names(ios))
-    nodesNames <- sapply(mget(ioGeneID, org.Hs.egSYMBOL, ifnotfound=NA), "[[",1)
-    names(nodesNames) <- names(ios)
-    nodevalue <- plac_path_genes(nodesNames,nodecolorcolumn)
-    nodevalue.max <- max(pretty(nodevalue))
-    nodevalue.na <- is.na(nodevalue)
-    nodevalue[is.na(nodevalue)] <- 0
-    nodecolors <- apply(colorRamp(c("gray","blue"))(nodevalue/nodevalue.max),
-                        1,
-                        function(x){rgb(x[1],x[2],x[3],maxColorValue=255)})
-    nodecolors[nodevalue.na] <- "#FFFFFF"
-    names(nodecolors) <- names(nodevalue)
-    subgraph <- graph #subGraph(names(nodecolors)[nodevalue >= 0.1],
-                       #  graph)
-    for (node in names(nodevalue)[nodevalue < cutpoint]) {
-         subgraph <- removeNode(node,subgraph)
-    }
-    nAttrs <- list();
-    # nAttrs$fillcolor <- makeAttr(graph, "lightgrey", list(orange=names(ios)[ios]))
-    nAttrs$label <- nodesNames[nodes(subgraph)]
-    nAttrs$fillcolor <- nodecolors[nodes(subgraph)]
-    nAttrs$fontsize <- makeAttr(subgraph,14)
-    nAttrs$width <- makeAttr(subgraph,2)
-    nAttrs$shape <- makeAttr(subgraph,"box")
-    list(nAttrs=nAttrs,
-         graph=subgraph,
-         max=nodevalue.max
-         )
-}
-
-plot_pathway <- function(pathway="hsa04012",pathcolor=quote(median_expression),cutpoint=1) {
-
-    the.graph <- KEGGgraph::parseKGML2Graph(KEGGgraph::getKGMLurl(pathway))
-
-    graph.attrs <- graph_attrs(the.graph,nodecolorcolumn=pathcolor,cutpoint=cutpoint)
-    plot(graph.attrs$graph,"dot",
-         nodeAttrs=graph.attrs$nAttrs,
-         attrs=list(graph=list(rankdir="LR")))
-    color.legend(100,100,200,200,c("0",as.character(graph.attrs$max)),
-                 apply(colorRamp(c("gray","blue"))(seq(0,1,length.out=10)),
-                       1,
-                       function(x){rgb(x[1],x[2],x[3],maxColorValue=255)}))
-}
-
-@
-
-\begin{figure}
-<<egf_pathway,echo=FALSE,message=FALSE,fig.width=6,fig.height=6,fig.out.width="\\textwidth">>=
-plot_pathway("hsa04012")
-@
-\caption{EGF KEGG pathway}
-\label{fig:egf_kegg_pathway}
-\end{figure}
-
-<<actin_pathway,echo=FALSE,results="hide",message=FALSE>>=
-library("pathview")
-pathview(gene.data=plac_path_egid(quote(as.numeric(median_expression))),
-         pathway.id="hsa04810",
-         both.dirs=list(gene=FALSE,cpd=FALSE),
-         limit=list(gene=200,cpd=29),
-         high=list(gene="blue"),
-         net.signature=FALSE)
-@
-
-\begin{figure}
-  \includegraphics{hsa04810.pathview.png}
-  \caption{KEGG Actin Cytoskeleton Pathway}
-  \label{fig:kegg_actin}
-\end{figure}
-
-<<placenta_class_table,echo=FALSE,results="hide">>=
-placenta_class_table <- function(class="shape",analysis.type="aov") {
-    subtable <- placenta.classification.p[factor==class & type == analysis.type,
-                                          list(gene,p,fdr)]
-    setnames(subtable,c("Gene","p","FDR"))
-    subtable <- head(subtable[order(p),],n=50)
-    top.type <- function(gene.name) {
-        placenta.classification.p[gene==gene.name & factor==class
-                                  & type =="glm",][order(-abs(coefficient))][1,level]
-    }
-    subtable[,`level_type`:=sapply(Gene,top.type)]
-    setnames(subtable,"level_type",
-             paste0("Most Significant ",
-                    gsub("^(.)","\\U\\1",class,perl=TRUE)))
-    print(xtable(subtable,
-                 align=c("c","c","r","r","c"),
-                 display=c("s","s","g","g","s")
-                 ),
-          booktabs=TRUE,
-          floating=FALSE,
-          include.rownames=FALSE,
-          math.style.exponents=TRUE,
-          size="\\scriptsize")
-}
-
-@
-
-\clearpage
-
-\subsection{Attachment Shape}
-
-\begin{table}
-<<attachment_shape_significant,echo=FALSE,results="asis">>=
-placenta_class_table("shape")
-@ 
-\caption{Top 50 genes which are associated with the shape of placenta attachment}
-\label{tab:shape_genes}
-\end{table}
-
-\subsection{Maternal-Fetal Intimacy}
-
-\begin{table}[h]
-<<intimacy_significant,echo=FALSE,results="asis">>=
-placenta_class_table("intimacy")
-@ 
-\caption{Top 50 genes which are associated with the degree of placenta intimacy}
-\label{tab:intimacy_genes}
-\end{table}
-
-\begin{table}[h]
-<<intimacy_significant_polr,echo=FALSE,results="asis">>=
-placenta_class_table("intimacy",analysis.type="polr")
-@ 
-\caption{Top 50 genes which are associated with the degree of placenta intimacy; ordered analysis}
-\label{tab:intimacy_genes_polr}
-\end{table}
-
-\subsection{Maternal-Fetal Interdigitation}
-
-\begin{table}[h]
-<<interdigitation_significant,echo=FALSE,results="asis">>=
-placenta_class_table("interdigitation")
-@ 
-\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation}
-\label{tab:interdigitation_genes}
-\end{table}
-
-\begin{table}[h]
-<<interdigitation_significant_polr,echo=FALSE,results="asis">>=
-placenta_class_table("interdigitation",analysis.type="polr")
-@ 
-\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation; ordered analysis}
-\label{tab:interdigitation_genes_polr}
-\end{table}
-
-\clearpage
-
-\section{GSEA}
-
-\subsection{Pathway Analysis}
-\begin{table}
-<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
-pathway <- list()
-for (database in c("kegg","reactome","biocarta","netpath")) {
-    load(paste0("../data/placenta_gsea_",database))
-    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
-    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
-    pathway[[database]] <-
-        gsea.table
-}
-pathway.analysis <-
-    rbindlist(pathway)[fdr < 0.1,]
-rm(pathway)
-pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
-                                gsub("_+"," ",
-                                     Pathway))]
-pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]
-
-setnames(pathway.analysis,"p_values","p")
-setorder(pathway.analysis,"p")
-print(xtable(pathway.analysis,
-             display=c("s","s","g","g","s"),
-             align=c("c","p{3.5in}","r","r","c")),
-      math.style.exponents=TRUE,
-      include.rownames=FALSE,
-      booktabs=TRUE,
-      floating=FALSE)      
-@ 
-\caption[Core Placenta Pathway Enrichment]{Pathways with significant and suggestive enrichment
-  ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
-  biocarta, reactome, and kegg pathway databases}
-\label{tab:core_placenta_pathway_analysis}
-\end{table}
-
-\begin{table}
-  \small
-<<pathway_analysis_class,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
-pathway <- list()
-for (class in c("shape","interdigitation","intimacy")) {
-    for (database in c("kegg","reactome","biocarta","netpath")) {
-        load(paste0("../data/placenta_gsea_",class,"_",database))
-        eval(parse(text=paste0("gsea.table <- placenta_gsea_",class,"_",database)))
-        gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
-        gsea.table[,Class:=gsub("^(.)","\\U\\1",class,perl=TRUE)]
-        pathway[[paste0(class,"_",database)]] <-
-            gsea.table
-    }
-}
-placenta_class_gsea <-
-    rbindlist(pathway)[fdr < 0.2,]
-placenta_class_gsea[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
-                                   gsub("_+"," ",
-                                        Pathway))]
-placenta_class_gsea[,Pathway:=gsub(".+\\%","",Pathway)]
-setnames(placenta_class_gsea,"p_values","p")
-setorder(placenta_class_gsea,"p")
-print(xtable(placenta_class_gsea,
-             display=c("s","s","g","g","s","s"),
-             align=c("c","p{3in}","r","r","c","c")),
-      math.style.exponents=TRUE,
-      include.rownames=FALSE,
-      booktabs=TRUE,
-      floating=FALSE)      
-@ 
-\caption[Placenta Classification Pathway Enrichment]{Pathways with
-  significant and suggestive enrichment ($\text{FDR}\le 0.2$) in genes
-  associated with placenta morphology using the biocarta, reactome,
-  and kegg pathway databases.}
-\label{tab:placenta_pathway_class_analysis}
-\end{table}
-
-\clearpage
+\section{Per-Species GSEA}
 
@@
 
+\section{Placenta Classification Analyses}
+
+<<placenta_pathway_egid,echo=FALSE,results="hide">>=
+load("../data/placenta_core_transcriptome_complete")
+plac_path_egid <- function(column) {
+    expr <- placenta.core.transcriptome.complete[!is.na(egid),eval(column)]
+    names(expr) <- placenta.core.transcriptome.complete[!is.na(egid),egid]
+    expr
+}
+
+plac_path_genes <- function(genes,column) {
+    result <- as.numeric(NULL)
+    result[!is.na(genes)] <-
+        placenta.core.transcriptome.complete[na.omit(genes),eval(column)]
+    names(result) <- names(genes)
+    result
+}
+
+@
+
+<<plot_pathway,echo=FALSE,result="hide">>=
+library("KEGGgraph")
+library("plotrix")
+makeAttr <- function(graph, default, valNodeList) {
+  tmp <- nodes(graph)
+  x <- rep(default, length(tmp)); names(x) <- tmp
+
+  if(!missing(valNodeList)) {
+    stopifnot(is.list(valNodeList))
+    allnodes <- unlist(valNodeList)
+    stopifnot(all(allnodes %in% tmp))
+    for(i in seq(valNodeList)) {
+      x[valNodeList[[i]]] <- names(valNodeList)[i]
+    }
+  }
+  return(x)
+}
+graph_attrs <- function(graph,nodecolorcolumn,cutpoint=1) {
+    outs <- sapply(edges(graph), length) > 0
+    ins <- sapply(inEdges(graph), length) > 0
+    ios <- outs | ins
+    ## translate the KEGG IDs into Gene Symbol
+    ioGeneID <- translateKEGGID2GeneID(names(ios))
+    nodesNames <- sapply(mget(ioGeneID, org.Hs.egSYMBOL, ifnotfound=NA), "[[",1)
+    names(nodesNames) <- names(ios)
+    nodevalue <- plac_path_genes(nodesNames,nodecolorcolumn)
+    nodevalue.max <- max(pretty(nodevalue))
+    nodevalue.na <- is.na(nodevalue)
+    nodevalue[is.na(nodevalue)] <- 0
+    nodecolors <- apply(colorRamp(c("gray","blue"))(nodevalue/nodevalue.max),
+                        1,
+                        function(x){rgb(x[1],x[2],x[3],maxColorValue=255)})
+    nodecolors[nodevalue.na] <- "#FFFFFF"
+    names(nodecolors) <- names(nodevalue)
+    subgraph <- graph #subGraph(names(nodecolors)[nodevalue >= 0.1],
+                       #  graph)
+    for (node in names(nodevalue)[nodevalue < cutpoint]) {
+         subgraph <- removeNode(node,subgraph)
+    }
+    nAttrs <- list();
+    # nAttrs$fillcolor <- makeAttr(graph, "lightgrey", list(orange=names(ios)[ios]))
+    nAttrs$label <- nodesNames[nodes(subgraph)]
+    nAttrs$fillcolor <- nodecolors[nodes(subgraph)]
+    nAttrs$fontsize <- makeAttr(subgraph,14)
+    nAttrs$width <- makeAttr(subgraph,2)
+    nAttrs$shape <- makeAttr(subgraph,"box")
+    list(nAttrs=nAttrs,
+         graph=subgraph,
+         max=nodevalue.max
+         )
+}
+
+plot_pathway <- function(pathway="hsa04012",pathcolor=quote(median_expression),cutpoint=1) {
+
+    the.graph <- KEGGgraph::parseKGML2Graph(KEGGgraph::getKGMLurl(pathway))
+
+    graph.attrs <- graph_attrs(the.graph,nodecolorcolumn=pathcolor,cutpoint=cutpoint)
+    plot(graph.attrs$graph,"dot",
+         nodeAttrs=graph.attrs$nAttrs,
+         attrs=list(graph=list(rankdir="LR")))
+    color.legend(100,100,200,200,c("0",as.character(graph.attrs$max)),
+                 apply(colorRamp(c("gray","blue"))(seq(0,1,length.out=10)),
+                       1,
+                       function(x){rgb(x[1],x[2],x[3],maxColorValue=255)}))
+}
+
+@
+
+\begin{figure}
+<<egf_pathway,echo=FALSE,message=FALSE,fig.width=6,fig.height=6,fig.out.width="\\textwidth">>=
+plot_pathway("hsa04012")
+@
+\caption{EGF KEGG pathway}
+\label{fig:egf_kegg_pathway}
+\end{figure}
+
+<<actin_pathway,echo=FALSE,results="hide",message=FALSE>>=
+library("pathview")
+pathview(gene.data=plac_path_egid(quote(as.numeric(median_expression))),
+         pathway.id="hsa04810",
+         both.dirs=list(gene=FALSE,cpd=FALSE),
+         limit=list(gene=200,cpd=29),
+         high=list(gene="blue"),
+         net.signature=FALSE)
+@
+
+\begin{figure}
+  \includegraphics{hsa04810.pathview.png}
+  \caption{KEGG Actin Cytoskeleton Pathway}
+  \label{fig:kegg_actin}
+\end{figure}
+
+<<placenta_class_table,echo=FALSE,results="hide">>=
+placenta_class_table <- function(class="shape",analysis.type="aov") {
+    subtable <- placenta.classification.p[factor==class & type == analysis.type,
+                                          list(gene,p,fdr)]
+    setnames(subtable,c("Gene","p","FDR"))
+    subtable <- head(subtable[order(p),],n=50)
+    top.type <- function(gene.name) {
+        placenta.classification.p[gene==gene.name & factor==class
+                                  & type =="glm",][order(-abs(coefficient))][1,level]
+    }
+    subtable[,`level_type`:=sapply(Gene,top.type)]
+    setnames(subtable,"level_type",
+             paste0("Most Significant ",
+                    gsub("^(.)","\\U\\1",class,perl=TRUE)))
+    print(xtable(subtable,
+                 align=c("c","c","r","r","c"),
+                 display=c("s","s","g","g","s")
+                 ),
+          booktabs=TRUE,
+          floating=FALSE,
+          include.rownames=FALSE,
+          math.style.exponents=TRUE,
+          size="\\scriptsize")
+}
+
+@
+
+\clearpage
+
+\subsection{Attachment Shape}
+
+\begin{table}
+<<attachment_shape_significant,echo=FALSE,results="asis">>=
+placenta_class_table("shape")
+@ 
+\caption{Top 50 genes which are associated with the shape of placenta attachment}
+\label{tab:shape_genes}
+\end{table}
+
+\subsection{Maternal-Fetal Intimacy}
+
+\begin{table}[h]
+<<intimacy_significant,echo=FALSE,results="asis">>=
+placenta_class_table("intimacy")
+@ 
+\caption{Top 50 genes which are associated with the degree of placenta intimacy}
+\label{tab:intimacy_genes}
+\end{table}
+
+\begin{table}[h]
+<<intimacy_significant_polr,echo=FALSE,results="asis">>=
+placenta_class_table("intimacy",analysis.type="polr")
+@ 
+\caption{Top 50 genes which are associated with the degree of placenta intimacy; ordered analysis}
+\label{tab:intimacy_genes_polr}
+\end{table}
+
+\subsection{Maternal-Fetal Interdigitation}
+
+\begin{table}[h]
+<<interdigitation_significant,echo=FALSE,results="asis">>=
+placenta_class_table("interdigitation")
+@ 
+\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation}
+\label{tab:interdigitation_genes}
+\end{table}
+
+\begin{table}[h]
+<<interdigitation_significant_polr,echo=FALSE,results="asis">>=
+placenta_class_table("interdigitation",analysis.type="polr")
+@ 
+\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation; ordered analysis}
+\label{tab:interdigitation_genes_polr}
+\end{table}
+
+\clearpage
+
+\section{GSEA}
+
+\subsection{Pathway Analysis}
+\begin{table}
+<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
+pathway <- list()
+for (database in c("kegg","reactome","biocarta","netpath")) {
+    load(paste0("../data/placenta_gsea_",database))
+    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
+    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
+    pathway[[database]] <-
+        gsea.table
+}
+pathway.analysis <-
+    rbindlist(pathway)[fdr < 0.1,]
+rm(pathway)
+pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
+                                gsub("_+"," ",
+                                     Pathway))]
+pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]
+
+setnames(pathway.analysis,"p_values","p")
+setorder(pathway.analysis,"p")
+print(xtable(pathway.analysis,
+             display=c("s","s","g","g","s"),
+             align=c("c","p{3.5in}","r","r","c")),
+      math.style.exponents=TRUE,
+      include.rownames=FALSE,
+      booktabs=TRUE,
+      floating=FALSE)      
+@ 
+\caption[Core Placenta Pathway Enrichment]{Pathways with significant and suggestive enrichment
+  ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
+  biocarta, reactome, and kegg pathway databases}
+\label{tab:core_placenta_pathway_analysis}
+\end{table}
+
+\begin{table}
+  \small
+<<pathway_analysis_class,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
+pathway <- list()
+for (class in c("shape","interdigitation","intimacy")) {
+    for (database in c("kegg","reactome","biocarta","netpath")) {
+        load(paste0("../data/placenta_gsea_",class,"_",database))
+        eval(parse(text=paste0("gsea.table <- placenta_gsea_",class,"_",database)))
+        gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
+        gsea.table[,Class:=gsub("^(.)","\\U\\1",class,perl=TRUE)]
+        pathway[[paste0(class,"_",database)]] <-
+            gsea.table
+    }
+}
+placenta_class_gsea <-
+    rbindlist(pathway)[fdr < 0.2,]
+placenta_class_gsea[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
+                                   gsub("_+"," ",
+                                        Pathway))]
+placenta_class_gsea[,Pathway:=gsub(".+\\%","",Pathway)]
+setnames(placenta_class_gsea,"p_values","p")
+setorder(placenta_class_gsea,"p")
+print(xtable(placenta_class_gsea,
+             display=c("s","s","g","g","s","s"),
+             align=c("c","p{3in}","r","r","c","c")),
+      math.style.exponents=TRUE,
+      include.rownames=FALSE,
+      booktabs=TRUE,
+      floating=FALSE)      
+@ 
+\caption[Placenta Classification Pathway Enrichment]{Pathways with
+  significant and suggestive enrichment ($\text{FDR}\le 0.2$) in genes
+  associated with placenta morphology using the biocarta, reactome,
+  and kegg pathway databases.}
+\label{tab:placenta_pathway_class_analysis}
+\end{table}
+
+\clearpage
+
 \section{Gene Evolution Trees}
@@
 @
-
-\begin{table}[hb]
-  \footnotesize
-<<rodentia.lineage.specific,echo=FALSE,results="asis">>=
-rod.all.lin.spec <- placenta.lineage.specific.table("rod.all")
-setkey(rod.all.lin.spec,"Human Symbol")
-@ 
-\caption{\lineagetableexplanationtypeiii{the rodentia order}{all other species}}
-\label{tab:rodentia_lineage_specific}
-\end{table}
-
-\begin{table}[hb]
-  \footnotesize
-<<lau.lineage.specific,echo=FALSE,results="asis">>=
-lau.all.lin.spec <- placenta.lineage.specific.table("lau.all")
-setkey(lau.all.lin.spec,"Human Symbol")
-@ 
-\caption{\lineagetableexplanation{the laurasiatheria
-    superorder}{all other species}}
-\label{tab:laurasiatheria_lineage_specific}
-\end{table}
-
-\begin{table}[hb]
-  \footnotesize
-<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
-eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau")
-setkey(eua.lau.lin.spec,"Human Symbol")
-@ 
-\caption{\lineagetableexplanation{ the euarchontoglires superorder}{ the laurasiatherian superorder}}
-\label{tab:laurasiatheria_euarchontoglires_lineage_specific}
-\end{table}
 

### 2016-11-18 manuscript/supplemental_material.Rnw
make species list subsections

@@
                            capfirst(this.species),"} Placental Gene Expression of all genes with FPKM $\\ge 50$")
-    cat(paste0("\\subsubsection{\\textit{",capfirst(this.species),"} Placental Gene Expression}\n\n"))
+    cat(paste0("\\subsection{\\textit{",capfirst(this.species),"} Placental Gene Expression}\n\n"))
     xt.align <- c("r","l","l","p{2.5in}","r")

### 2016-11-18 manuscript/supplemental_material.Rnw
move GO analysis
remove extra figures from supplemental
 - Remove extra gene trees
 - move analyses up

@@
 
-\subsection{GO Ontology}
-
-<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
-go.analysis.wide <- 
-    data.table(dcast(data=go.results,
-                     formula=GO.ID+type+Term+Ontology~species,
-                     value.var="FDR"))
-go.analysis.wide <- 
-    go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
-                                         function(x){sum(as.numeric(as.character(x)) <= 0.05)})]
-go.analysis.wide <- 
-    go.analysis.wide[,`max(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
-                                           function(x){if (!any(is.finite(x) & x<=0.05)) {
-                                                           return(NA)
-                                                       } else {
-                                                           max(x[is.finite(x) & x<=0.05],na.rm=TRUE)
-                                                       }})]
-go.analysis.wide.table <- 
-    go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
-                         grepl("Top 1% No Housekeeping",type),
-                     list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
-go.analysis.wide.table <- 
-    go.analysis.wide.table[order(Ontology,-`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
-setnames(go.analysis.wide.table,"GO.ID","GO ID")
-
-print(xtable(go.analysis.wide.table,
-             display=c("d","s","s","s","d","e"),
-             align=  c("c","c","c","c","c","c"),
-             caption=c("GO terms with significant enrichment ($\\text{FDR}\\le0.05$)
-  in 12 or more of the species analyzed from genes in the top 1\\% of
-  expression. $n(\\text{FDR}\\le0.05)$ is the number of species with
-  significant enrichment ($\\text{FDR}\\le 0.05$) in the GO term
-  listed. $max(\\text{FDR}\\le0.05)$ is the maximum FDR value in all
-  species with an FDR of less than $0.05$.","GO terms in 12 or more species"),
-  label="tab:go_terms_all_species"
-  ),
-  math.style.exponents=TRUE,
-  include.rownames=FALSE,
-  booktabs=TRUE,
-  floating=FALSE,
-  tabular.environment="longtable",
-  caption.placement="top",
-  size="\\footnotesize"
-  )
-
-@
-
-\clearpage
-
 \subsection{Top 1\% Transcripts Enrichment}
@@
 
-\section{Placenta Classification Analyses}
-
-<<placenta_pathway_egid,echo=FALSE,results="hide">>=
-load("../data/placenta_core_transcriptome_complete")
-plac_path_egid <- function(column) {
-    expr <- placenta.core.transcriptome.complete[!is.na(egid),eval(column)]
-    names(expr) <- placenta.core.transcriptome.complete[!is.na(egid),egid]
-    expr
-}
-
-plac_path_genes <- function(genes,column) {
-    result <- as.numeric(NULL)
-    result[!is.na(genes)] <-
-        placenta.core.transcriptome.complete[na.omit(genes),eval(column)]
-    names(result) <- names(genes)
-    result
-}
-
-@
-
-<<plot_pathway,echo=FALSE,result="hide">>=
-library("KEGGgraph")
-library("plotrix")
-makeAttr <- function(graph, default, valNodeList) {
-  tmp <- nodes(graph)
-  x <- rep(default, length(tmp)); names(x) <- tmp
-
-  if(!missing(valNodeList)) {
-    stopifnot(is.list(valNodeList))
-    allnodes <- unlist(valNodeList)
-    stopifnot(all(allnodes %in% tmp))
-    for(i in seq(valNodeList)) {
-      x[valNodeList[[i]]] <- names(valNodeList)[i]
-    }
-  }
-  return(x)
-}
-graph_attrs <- function(graph,nodecolorcolumn,cutpoint=1) {
-    outs <- sapply(edges(graph), length) > 0
-    ins <- sapply(inEdges(graph), length) > 0
-    ios <- outs | ins
-    ## translate the KEGG IDs into Gene Symbol
-    ioGeneID <- translateKEGGID2GeneID(names(ios))
-    nodesNames <- sapply(mget(ioGeneID, org.Hs.egSYMBOL, ifnotfound=NA), "[[",1)
-    names(nodesNames) <- names(ios)
-    nodevalue <- plac_path_genes(nodesNames,nodecolorcolumn)
-    nodevalue.max <- max(pretty(nodevalue))
-    nodevalue.na <- is.na(nodevalue)
-    nodevalue[is.na(nodevalue)] <- 0
-    nodecolors <- apply(colorRamp(c("gray","blue"))(nodevalue/nodevalue.max),
-                        1,
-                        function(x){rgb(x[1],x[2],x[3],maxColorValue=255)})
-    nodecolors[nodevalue.na] <- "#FFFFFF"
-    names(nodecolors) <- names(nodevalue)
-    subgraph <- graph #subGraph(names(nodecolors)[nodevalue >= 0.1],
-                       #  graph)
-    for (node in names(nodevalue)[nodevalue < cutpoint]) {
-         subgraph <- removeNode(node,subgraph)
-    }
-    nAttrs <- list();
-    # nAttrs$fillcolor <- makeAttr(graph, "lightgrey", list(orange=names(ios)[ios]))
-    nAttrs$label <- nodesNames[nodes(subgraph)]
-    nAttrs$fillcolor <- nodecolors[nodes(subgraph)]
-    nAttrs$fontsize <- makeAttr(subgraph,14)
-    nAttrs$width <- makeAttr(subgraph,2)
-    nAttrs$shape <- makeAttr(subgraph,"box")
-    list(nAttrs=nAttrs,
-         graph=subgraph,
-         max=nodevalue.max
-         )
-}
-
-plot_pathway <- function(pathway="hsa04012",pathcolor=quote(median_expression),cutpoint=1) {
-
-    the.graph <- KEGGgraph::parseKGML2Graph(KEGGgraph::getKGMLurl(pathway))
-
-    graph.attrs <- graph_attrs(the.graph,nodecolorcolumn=pathcolor,cutpoint=cutpoint)
-    plot(graph.attrs$graph,"dot",
-         nodeAttrs=graph.attrs$nAttrs,
-         attrs=list(graph=list(rankdir="LR")))
-    color.legend(100,100,200,200,c("0",as.character(graph.attrs$max)),
-                 apply(colorRamp(c("gray","blue"))(seq(0,1,length.out=10)),
-                       1,
-                       function(x){rgb(x[1],x[2],x[3],maxColorValue=255)}))
-}
-
-@
-
-\begin{figure}
-<<egf_pathway,echo=FALSE,message=FALSE,fig.width=6,fig.height=6,fig.out.width="\\textwidth">>=
-plot_pathway("hsa04012")
-@
-\caption{EGF KEGG pathway}
-\label{fig:egf_kegg_pathway}
-\end{figure}
-
-<<actin_pathway,echo=FALSE,results="hide",message=FALSE>>=
-library("pathview")
-pathview(gene.data=plac_path_egid(quote(as.numeric(median_expression))),
-         pathway.id="hsa04810",
-         both.dirs=list(gene=FALSE,cpd=FALSE),
-         limit=list(gene=200,cpd=29),
-         high=list(gene="blue"),
-         net.signature=FALSE)
-@
-
-\begin{figure}
-  \includegraphics{hsa04810.pathview.png}
-  \caption{KEGG Actin Cytoskeleton Pathway}
-  \label{fig:kegg_actin}
-\end{figure}
-
-<<placenta_class_table,echo=FALSE,results="hide">>=
-placenta_class_table <- function(class="shape",analysis.type="aov") {
-    subtable <- placenta.classification.p[factor==class & type == analysis.type,
-                                          list(gene,p,fdr)]
-    setnames(subtable,c("Gene","p","FDR"))
-    subtable <- head(subtable[order(p),],n=50)
-    top.type <- function(gene.name) {
-        placenta.classification.p[gene==gene.name & factor==class
-                                  & type =="glm",][order(-abs(coefficient))][1,level]
-    }
-    subtable[,`level_type`:=sapply(Gene,top.type)]
-    setnames(subtable,"level_type",
-             paste0("Most Significant ",
-                    gsub("^(.)","\\U\\1",class,perl=TRUE)))
-    print(xtable(subtable,
-                 align=c("c","c","r","r","c"),
-                 display=c("s","s","g","g","s")
+\subsection{Common Species Terms}
+
+<<go.analysis.all.tissue,echo=FALSE,results="asis">>=
+go.analysis.wide <- 
+    data.table(dcast(data=go.results,
+                     formula=GO.ID+type+Term+Ontology~species,
+                     value.var="FDR"))
+go.analysis.wide <- 
+    go.analysis.wide[,`n(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
+                                         function(x){sum(as.numeric(as.character(x)) <= 0.05)})]
+go.analysis.wide <- 
+    go.analysis.wide[,`max(FDR ≤ 0.05)`:=apply(go.analysis.wide[,-(1:4),with=FALSE],1,
+                                           function(x){if (!any(is.finite(x) & x<=0.05)) {
+                                                           return(NA)
+                                                       } else {
+                                                           max(x[is.finite(x) & x<=0.05],na.rm=TRUE)
+                                                       }})]
+go.analysis.wide.table <- 
+    go.analysis.wide[!is.na(`n(FDR ≤ 0.05)`) & `n(FDR ≤ 0.05)` >= 12 &
+                         grepl("Top 1% No Housekeeping",type),
+                     list(GO.ID,Term,Ontology,`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
+go.analysis.wide.table <- 
+    go.analysis.wide.table[order(Ontology,-`n(FDR ≤ 0.05)`,`max(FDR ≤ 0.05)`)]
+setnames(go.analysis.wide.table,"GO.ID","GO ID")
+
+print(xtable(go.analysis.wide.table,
+             display=c("d","s","s","s","d","e"),
+             align=  c("c","c","c","c","c","c"),
+             caption=c("GO terms with significant enrichment ($\\text{FDR}\\le0.05$)
+  in 12 or more of the species analyzed from genes in the top 1\\% of
+  expression. $n(\\text{FDR}\\le0.05)$ is the number of species with
+  significant enrichment ($\\text{FDR}\\le 0.05$) in the GO term
+  listed. $max(\\text{FDR}\\le0.05)$ is the maximum FDR value in all
+  species with an FDR of less than $0.05$.","GO terms in 12 or more species"),
+  label="tab:go_terms_all_species"
   ),
-          booktabs=TRUE,
-          floating=FALSE,
-          include.rownames=FALSE,
-          math.style.exponents=TRUE,
-          size="\\scriptsize")
-}
-
-@
-
-\clearpage
-
-\subsection{Attachment Shape}
-
-\begin{table}
-<<attachment_shape_significant,echo=FALSE,results="asis">>=
-placenta_class_table("shape")
-@ 
-\caption{Top 50 genes which are associated with the shape of placenta attachment}
-\label{tab:shape_genes}
-\end{table}
-
-\subsection{Maternal-Fetal Intimacy}
-
-\begin{table}[h]
-<<intimacy_significant,echo=FALSE,results="asis">>=
-placenta_class_table("intimacy")
-@ 
-\caption{Top 50 genes which are associated with the degree of placenta intimacy}
-\label{tab:intimacy_genes}
-\end{table}
-
-\begin{table}[h]
-<<intimacy_significant_polr,echo=FALSE,results="asis">>=
-placenta_class_table("intimacy",analysis.type="polr")
-@ 
-\caption{Top 50 genes which are associated with the degree of placenta intimacy; ordered analysis}
-\label{tab:intimacy_genes_polr}
-\end{table}
-
-\subsection{Maternal-Fetal Interdigitation}
-
-\begin{table}[h]
-<<interdigitation_significant,echo=FALSE,results="asis">>=
-placenta_class_table("interdigitation")
-@ 
-\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation}
-\label{tab:interdigitation_genes}
-\end{table}
-
-\begin{table}[h]
-<<interdigitation_significant_polr,echo=FALSE,results="asis">>=
-placenta_class_table("interdigitation",analysis.type="polr")
-@ 
-\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation; ordered analysis}
-\label{tab:interdigitation_genes_polr}
-\end{table}
-
-\clearpage
-
-\section{GSEA}
-
-\subsection{Pathway Analysis}
-\begin{table}
-<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
-pathway <- list()
-for (database in c("kegg","reactome","biocarta","netpath")) {
-    load(paste0("../data/placenta_gsea_",database))
-    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
-    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
-    pathway[[database]] <-
-        gsea.table
-}
-pathway.analysis <-
-    rbindlist(pathway)[fdr < 0.1,]
-rm(pathway)
-pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
-                                gsub("_+"," ",
-                                     Pathway))]
-pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]
-
-setnames(pathway.analysis,"p_values","p")
-setorder(pathway.analysis,"p")
-print(xtable(pathway.analysis,
-             display=c("s","s","g","g","s"),
-             align=c("c","p{3.5in}","r","r","c")),
   math.style.exponents=TRUE,
@@
   booktabs=TRUE,
-      floating=FALSE)      
-@ 
-\caption[Core Placenta Pathway Enrichment]{Pathways with significant and suggestive enrichment
-  ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
-  biocarta, reactome, and kegg pathway databases}
-\label{tab:core_placenta_pathway_analysis}
-\end{table}
-
-\begin{table}
-  \small
-<<pathway_analysis_class,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
-pathway <- list()
-for (class in c("shape","interdigitation","intimacy")) {
-    for (database in c("kegg","reactome","biocarta","netpath")) {
-        load(paste0("../data/placenta_gsea_",class,"_",database))
-        eval(parse(text=paste0("gsea.table <- placenta_gsea_",class,"_",database)))
-        gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
-        gsea.table[,Class:=gsub("^(.)","\\U\\1",class,perl=TRUE)]
-        pathway[[paste0(class,"_",database)]] <-
-            gsea.table
-    }
-}
-placenta_class_gsea <-
-    rbindlist(pathway)[fdr < 0.2,]
-placenta_class_gsea[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
-                                   gsub("_+"," ",
-                                        Pathway))]
-placenta_class_gsea[,Pathway:=gsub(".+\\%","",Pathway)]
-setnames(placenta_class_gsea,"p_values","p")
-setorder(placenta_class_gsea,"p")
-print(xtable(placenta_class_gsea,
-             display=c("s","s","g","g","s","s"),
-             align=c("c","p{3in}","r","r","c","c")),
-      math.style.exponents=TRUE,
-      include.rownames=FALSE,
-      booktabs=TRUE,
-      floating=FALSE)      
-@ 
-\caption[Placenta Classification Pathway Enrichment]{Pathways with
-  significant and suggestive enrichment ($\text{FDR}\le 0.2$) in genes
-  associated with placenta morphology using the biocarta, reactome,
-  and kegg pathway databases.}
-\label{tab:placenta_pathway_class_analysis}
-\end{table}
+  floating=FALSE,
+  tabular.environment="longtable",
+  caption.placement="top",
+  size="\\footnotesize"
+  )
+
+@
 
@@
 \label{fig:crh_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<kiss1.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="KISS1"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.7))
+@ 
+\caption{\genetreeexplanation[0.1]{\textit{KISS1}}}
+\label{fig:kiss1_evolution_tree}
+\end{figure}
+
+\begin{figure}
+<<pappa.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PAPPA"
+                                                    & species=="homo sapiens"),
+                                subtree=c("ENSSSCG00000005498","ENSG00000182752"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.15))
+@ 
+\caption{\genetreeexplanation[0.1]{\textit{PAPPA}}}
+\label{fig:pappa_evolution_tree}
 \end{figure}
@@
 \label{fig:adam12_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<kiss1.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="KISS1"
-                                                    & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.7))
-@ 
-\caption{\genetreeexplanation[0.1]{\textit{KISS1}}}
-\label{fig:kiss1_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<pappa.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PAPPA"
-                                                    & species=="homo sapiens"),
-                                subtree=c("ENSSSCG00000005498","ENSG00000182752"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.15))
-@ 
-\caption{\genetreeexplanation[0.1]{\textit{PAPPA}}}
-\label{fig:pappa_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<tgfb1.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="TGFB1"
-                                                    & species=="homo sapiens"),
-                                subtree=c("ENSMODG00000008630","ENSSSCG00000026296"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.3))
-@ 
-\caption{\genetreeexplanation{\textit{TGFB1}}}
-\label{fig:tgfb1_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<gpc3.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="GPC3"
-                                                    & species=="homo sapiens"),
-                                subtree=c("ENSCAFG00000018864","ENSLAFG00000025706"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.13))
-@ 
-\caption{\genetreeexplanation{\textit{GPC3}}}
-\label{fig:gpc3_evolution_tree}
-\end{figure}
-
-\clearpage
-
-\subsection{Primate Specific Evolution}
-
-\begin{figure}[hbf]
-<<fstl1.evolution.tree,echo=FALSE,out.height="0.6\\textheight",fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="FSTL1"
-                                                    & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=5,color="black",fontface="italic")+xlim(0,3))
-@ 
-\caption{\genetreeexplanation{follistatin-like 1 \textit{FSTL1}}}
-\label{fig:fstl1_evolution_tree}
-\end{figure}
-
-\clearpage
-
-\subsection{Other Gene Trees}
-
-\begin{figure}[hbf]
-<<serpine.evolution.tree,echo=FALSE,out.height="0.6\\textheight",fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="SERPINE2"
-                                                    & species=="homo sapiens"),
-                                subtree=c("ENSBTAG00000008717",
-                                          "ENSMODG00000004913"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=4,color="black",fontface="italic")+xlim(0,0.8))
-@ 
-\caption{\genetreeexplanation{\textit{SERPINE2}}}
-\label{fig:serpine_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<sslp1.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_id=="ENSCAFG00000010638" &
-                                                    species=="canis familiaris"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,1.2))
-@ 
-\caption{\genetreeexplanation{\textit{SSLP1}}}
-\label{fig:sslp1_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<estrogen.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CYP19A1"
-                                                    & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.37))
-@ 
-\caption{\genetreeexplanation{estrogen synthesis genes}}
-\label{fig:estrogen_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<progesterone.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="HSD3B1"
-                                                    & species=="homo sapiens"),
-                                subtree=c("ENSMODG00000005385","ENSLAFG00000027896"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.6))
-@ 
-\caption{\genetreeexplanation{progesterone synthesis genes}}
-\label{fig:progesterone_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<chorionic.somatomammotropin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CSH1"
-                                                    & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.9))
-@ 
-\caption{\genetreeexplanation{chorionic somatomammotropin}}
-\label{fig:csh1_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<prolactin.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PRL"
-                                                    & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=3,color="black",fontface="italic")+xlim(0,1))
-@ 
-\caption{\genetreeexplanation{prolactin}}
-\label{fig:prolactin_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<rln2.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="RLN2"
-                                                    & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.8))
-@ 
-\caption{\genetreeexplanation{\textit{RLN2}}}
-\label{fig:rln2_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<ovos2.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_id=="ENSDNOG00000033320"),
-                                subtree=c("ENSDNOG00000033320","ENSMODG00000017988"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=5,color="black",fontface="italic")+xlim(0,1.3))
-@ 
-\caption{\genetreeexplanation{\cow \textit{OVOS2}}}
-\label{fig:ovos2_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<prolactin.receptor.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="PRLR"
-                                                    & species=="homo sapiens"),
-                                subtree=c("ENSG00000113494","ENSMODG00000020376"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.3))
-@ 
-\caption{\genetreeexplanation{prolactin receptor, \textit{PRLR}}}
-\label{fig:prolactin_receptor_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<hoxa11.receptor.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="HOXA11"
-                                                    & species=="homo sapiens"),
-                                subtree=c("ENSBTAG00000014738","ENSECAG00000013229"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.4))
-@ 
-\caption{\genetreeexplanation{\textit{HOXA11}}}
-\label{fig:hoxa11_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<hbe1.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="HBE1"
-                                                    & species=="homo sapiens"),
-                                subtree=c("ENSBTAG00000037815",
-                                          "ENSMUSG00000078621"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=4,color="black",fontface="italic")+xlim(0,0.6))
-@ 
-\caption{\genetreeexplanation{\textit{HBE1}}}
-\label{fig:hbe1_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<nppb.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="NPPB"
-                                                    & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.4))
-@ 
-\caption{\genetreeexplanation{\textit{NPPB}}}
-\label{fig:nppb_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<nppc.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="NPPC"
-                                                    & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,0.32))
-@ 
-\caption{\genetreeexplanation{\textit{NPPC}}}
-\label{fig:nppc_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<galectins.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="LGALS1"
-                                                    & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,1))
-@ 
-\caption{\genetreeexplanation{\textit{LGALS1} and \textit{LGALS2}}}
-\label{fig:galectins_evolution_tree}
-\end{figure}
-
-\begin{figure}
-<<lgals13.galectins.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
-
-print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="LGALS13"
-                                                    & species=="homo sapiens"),
-                                genes.to.tree=genes.to.tree,
-                                combined.fpkm=combined.fpkm,
-                                trees=trees
-                                )+
-      geom_tiplab(size=6,color="black",fontface="italic")+xlim(0,1.5))
-@ 
-\caption{\genetreeexplanation{\textit{LGALS13}}}
-\label{fig:galectins_lgals13_evolution_tree}
 \end{figure}
@@
                                     caption="\\lineagetableexplanationtypeiii{\\human}{all other species}",
-                                    label="tab:huamn_linage_specific_all"
+                                    label="tab:human_linage_specific_all"
                                     )
@@
 
+\section{Placenta Classification Analyses}
+
+<<placenta_pathway_egid,echo=FALSE,results="hide">>=
+load("../data/placenta_core_transcriptome_complete")
+plac_path_egid <- function(column) {
+    expr <- placenta.core.transcriptome.complete[!is.na(egid),eval(column)]
+    names(expr) <- placenta.core.transcriptome.complete[!is.na(egid),egid]
+    expr
+}
+
+plac_path_genes <- function(genes,column) {
+    result <- as.numeric(NULL)
+    result[!is.na(genes)] <-
+        placenta.core.transcriptome.complete[na.omit(genes),eval(column)]
+    names(result) <- names(genes)
+    result
+}
+
+@
+
+<<plot_pathway,echo=FALSE,result="hide">>=
+library("KEGGgraph")
+library("plotrix")
+makeAttr <- function(graph, default, valNodeList) {
+  tmp <- nodes(graph)
+  x <- rep(default, length(tmp)); names(x) <- tmp
+
+  if(!missing(valNodeList)) {
+    stopifnot(is.list(valNodeList))
+    allnodes <- unlist(valNodeList)
+    stopifnot(all(allnodes %in% tmp))
+    for(i in seq(valNodeList)) {
+      x[valNodeList[[i]]] <- names(valNodeList)[i]
+    }
+  }
+  return(x)
+}
+graph_attrs <- function(graph,nodecolorcolumn,cutpoint=1) {
+    outs <- sapply(edges(graph), length) > 0
+    ins <- sapply(inEdges(graph), length) > 0
+    ios <- outs | ins
+    ## translate the KEGG IDs into Gene Symbol
+    ioGeneID <- translateKEGGID2GeneID(names(ios))
+    nodesNames <- sapply(mget(ioGeneID, org.Hs.egSYMBOL, ifnotfound=NA), "[[",1)
+    names(nodesNames) <- names(ios)
+    nodevalue <- plac_path_genes(nodesNames,nodecolorcolumn)
+    nodevalue.max <- max(pretty(nodevalue))
+    nodevalue.na <- is.na(nodevalue)
+    nodevalue[is.na(nodevalue)] <- 0
+    nodecolors <- apply(colorRamp(c("gray","blue"))(nodevalue/nodevalue.max),
+                        1,
+                        function(x){rgb(x[1],x[2],x[3],maxColorValue=255)})
+    nodecolors[nodevalue.na] <- "#FFFFFF"
+    names(nodecolors) <- names(nodevalue)
+    subgraph <- graph #subGraph(names(nodecolors)[nodevalue >= 0.1],
+                       #  graph)
+    for (node in names(nodevalue)[nodevalue < cutpoint]) {
+         subgraph <- removeNode(node,subgraph)
+    }
+    nAttrs <- list();
+    # nAttrs$fillcolor <- makeAttr(graph, "lightgrey", list(orange=names(ios)[ios]))
+    nAttrs$label <- nodesNames[nodes(subgraph)]
+    nAttrs$fillcolor <- nodecolors[nodes(subgraph)]
+    nAttrs$fontsize <- makeAttr(subgraph,14)
+    nAttrs$width <- makeAttr(subgraph,2)
+    nAttrs$shape <- makeAttr(subgraph,"box")
+    list(nAttrs=nAttrs,
+         graph=subgraph,
+         max=nodevalue.max
+         )
+}
+
+plot_pathway <- function(pathway="hsa04012",pathcolor=quote(median_expression),cutpoint=1) {
+
+    the.graph <- KEGGgraph::parseKGML2Graph(KEGGgraph::getKGMLurl(pathway))
+
+    graph.attrs <- graph_attrs(the.graph,nodecolorcolumn=pathcolor,cutpoint=cutpoint)
+    plot(graph.attrs$graph,"dot",
+         nodeAttrs=graph.attrs$nAttrs,
+         attrs=list(graph=list(rankdir="LR")))
+    color.legend(100,100,200,200,c("0",as.character(graph.attrs$max)),
+                 apply(colorRamp(c("gray","blue"))(seq(0,1,length.out=10)),
+                       1,
+                       function(x){rgb(x[1],x[2],x[3],maxColorValue=255)}))
+}
+
+@
+
+\begin{figure}
+<<egf_pathway,echo=FALSE,message=FALSE,fig.width=6,fig.height=6,fig.out.width="\\textwidth">>=
+plot_pathway("hsa04012")
+@
+\caption{EGF KEGG pathway}
+\label{fig:egf_kegg_pathway}
+\end{figure}
+
+<<actin_pathway,echo=FALSE,results="hide",message=FALSE>>=
+library("pathview")
+pathview(gene.data=plac_path_egid(quote(as.numeric(median_expression))),
+         pathway.id="hsa04810",
+         both.dirs=list(gene=FALSE,cpd=FALSE),
+         limit=list(gene=200,cpd=29),
+         high=list(gene="blue"),
+         net.signature=FALSE)
+@
+
+\begin{figure}
+  \includegraphics{hsa04810.pathview.png}
+  \caption{KEGG Actin Cytoskeleton Pathway}
+  \label{fig:kegg_actin}
+\end{figure}
+
+<<placenta_class_table,echo=FALSE,results="hide">>=
+placenta_class_table <- function(class="shape",analysis.type="aov") {
+    subtable <- placenta.classification.p[factor==class & type == analysis.type,
+                                          list(gene,p,fdr)]
+    setnames(subtable,c("Gene","p","FDR"))
+    subtable <- head(subtable[order(p),],n=50)
+    top.type <- function(gene.name) {
+        placenta.classification.p[gene==gene.name & factor==class
+                                  & type =="glm",][order(-abs(coefficient))][1,level]
+    }
+    subtable[,`level_type`:=sapply(Gene,top.type)]
+    setnames(subtable,"level_type",
+             paste0("Most Significant ",
+                    gsub("^(.)","\\U\\1",class,perl=TRUE)))
+    print(xtable(subtable,
+                 align=c("c","c","r","r","c"),
+                 display=c("s","s","g","g","s")
+                 ),
+          booktabs=TRUE,
+          floating=FALSE,
+          include.rownames=FALSE,
+          math.style.exponents=TRUE,
+          size="\\scriptsize")
+}
+
+@
+
+\clearpage
+
+\subsection{Attachment Shape}
+
+\begin{table}
+<<attachment_shape_significant,echo=FALSE,results="asis">>=
+placenta_class_table("shape")
+@ 
+\caption{Top 50 genes which are associated with the shape of placenta attachment}
+\label{tab:shape_genes}
+\end{table}
+
+\subsection{Maternal-Fetal Intimacy}
+
+\begin{table}[h]
+<<intimacy_significant,echo=FALSE,results="asis">>=
+placenta_class_table("intimacy")
+@ 
+\caption{Top 50 genes which are associated with the degree of placenta intimacy}
+\label{tab:intimacy_genes}
+\end{table}
+
+\begin{table}[h]
+<<intimacy_significant_polr,echo=FALSE,results="asis">>=
+placenta_class_table("intimacy",analysis.type="polr")
+@ 
+\caption{Top 50 genes which are associated with the degree of placenta intimacy; ordered analysis}
+\label{tab:intimacy_genes_polr}
+\end{table}
+
+\subsection{Maternal-Fetal Interdigitation}
+
+\begin{table}[h]
+<<interdigitation_significant,echo=FALSE,results="asis">>=
+placenta_class_table("interdigitation")
+@ 
+\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation}
+\label{tab:interdigitation_genes}
+\end{table}
+
+\begin{table}[h]
+<<interdigitation_significant_polr,echo=FALSE,results="asis">>=
+placenta_class_table("interdigitation",analysis.type="polr")
+@ 
+\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation; ordered analysis}
+\label{tab:interdigitation_genes_polr}
+\end{table}
+
+\clearpage
+
+\section{GSEA}
+
+\subsection{Pathway Analysis}
+\begin{table}
+<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
+pathway <- list()
+for (database in c("kegg","reactome","biocarta","netpath")) {
+    load(paste0("../data/placenta_gsea_",database))
+    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
+    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
+    pathway[[database]] <-
+        gsea.table
+}
+pathway.analysis <-
+    rbindlist(pathway)[fdr < 0.1,]
+rm(pathway)
+pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
+                                gsub("_+"," ",
+                                     Pathway))]
+pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]
+
+setnames(pathway.analysis,"p_values","p")
+setorder(pathway.analysis,"p")
+print(xtable(pathway.analysis,
+             display=c("s","s","g","g","s"),
+             align=c("c","p{3.5in}","r","r","c")),
+      math.style.exponents=TRUE,
+      include.rownames=FALSE,
+      booktabs=TRUE,
+      floating=FALSE)      
+@ 
+\caption[Core Placenta Pathway Enrichment]{Pathways with significant and suggestive enrichment
+  ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
+  biocarta, reactome, and kegg pathway databases}
+\label{tab:core_placenta_pathway_analysis}
+\end{table}
+
+\begin{table}
+  \small
+<<pathway_analysis_class,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
+pathway <- list()
+for (class in c("shape","interdigitation","intimacy")) {
+    for (database in c("kegg","reactome","biocarta","netpath")) {
+        load(paste0("../data/placenta_gsea_",class,"_",database))
+        eval(parse(text=paste0("gsea.table <- placenta_gsea_",class,"_",database)))
+        gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
+        gsea.table[,Class:=gsub("^(.)","\\U\\1",class,perl=TRUE)]
+        pathway[[paste0(class,"_",database)]] <-
+            gsea.table
+    }
+}
+placenta_class_gsea <-
+    rbindlist(pathway)[fdr < 0.2,]
+placenta_class_gsea[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
+                                   gsub("_+"," ",
+                                        Pathway))]
+placenta_class_gsea[,Pathway:=gsub(".+\\%","",Pathway)]
+setnames(placenta_class_gsea,"p_values","p")
+setorder(placenta_class_gsea,"p")
+print(xtable(placenta_class_gsea,
+             display=c("s","s","g","g","s","s"),
+             align=c("c","p{3in}","r","r","c","c")),
+      math.style.exponents=TRUE,
+      include.rownames=FALSE,
+      booktabs=TRUE,
+      floating=FALSE)      
+@ 
+\caption[Placenta Classification Pathway Enrichment]{Pathways with
+  significant and suggestive enrichment ($\text{FDR}\le 0.2$) in genes
+  associated with placenta morphology using the biocarta, reactome,
+  and kegg pathway databases.}
+\label{tab:placenta_pathway_class_analysis}
+\end{table}
+
+\clearpage
+
 \printbibliography

### 2016-11-18 manuscript/manuscript.Rnw
fix go analyses ref

--- 2016-11-18 manuscript/manuscript.Rnw
@@
 
+% \subsubsection{Per-species Gene Set and Pathway Enrichment}
+
+We performed \ac{gsea} on both the top 1\% of transcripts
+(Supplemental \cref{tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_canis_familiaris,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_monodelphis_domestica}) and the top
+100 transcripts (Supplemental
+\cref{tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_canis_familiaris,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_monodelphis_domestica}); common processes
+include protein transport and viral transcription/expression
+(Supplemental \cref{tab:go_terms_all_species}).
+
 \subsection{Pre-eclampsia \human Lineage Specific Genes}
@@
 differentially expressed in the \human lineage in comparison to all
-other lineages (\cref{tab:human_lineage_specific}). These genes
-include multiple genes which are thought to be involved in
-pre-eclampsia. Notably, \textit{CRH}
-(FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
+other lineages (\cref{tab:human_lineage_specific}, Supplemental
+\cref{tab:human_lineage_specific_all}). These genes include multiple
+genes which are thought to be involved in pre-eclampsia. Notably,
+\textit{CRH} (FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
 FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
@@
 the placenta transcriptome if a 1:1 ortholog of that gene was
-expressed with $\text{FPKM} \ge 10$ in all species studied and was
-not a human housekeeping gene according to
-\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}.
+expressed with $\text{FPKM} \ge 10$ in all species studied and was not
+a human housekeeping gene according to
+\citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
+meet this criteria but are housekeeping genes are depicted in
+Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
 \Cref{fig:core_placenta_transcriptome_shape} shows the number of genes
 which would be included in the core placenta set if the \ac{FPKM}
-cutoff was set differently; for example, choosing an \ac{FPKM} cutoff of 1 yields %
-\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & 
-`Min. Exp. in` == '14 of 14',max(num_genes)]} %
+cutoff was set differently; for example, choosing an \ac{FPKM} cutoff
+of 1 yields %
+\Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
+  == '14 of 14',max(num_genes)]} %
 genes in the core transcriptome. %

### 2016-11-18 manuscript/extended_supplemental_material.Rnw
duplicate supplemental material for cutting
ignore more filex, change extended supplement path

--- 2016-11-11 manuscript/supplemental_material.Rnw
@@
 opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.6\\textheight",out.extra="keepaspectratio")
-opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")
-opts_chunk$set(fig.path="supplemental_figure/")
+opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="extended_supplemental_cache/")
+opts_chunk$set(fig.path="extended_supplemental_figure/")
 options(device = function(file, width = 8, height = 7, ...) {

### 2016-11-18 manuscript/manuscript.Rnw
move the discussion of the go analysis for core
remove extra citations to get to 45

@@
 mammals
-\cite{Mossman.1987:vertebratefetalmembranes,Moffett.ea2006:Immunologyofplacentationin}
+\cite{Mossman.1987:vertebratefetalmembranes} %,Moffett.ea2006:Immunologyofplacentationin}
 (Supplemental \cref{tab:placental_classification}). Mammals also
@@
 of pregnancy
-\cite{Carter2012:Evolutionofplacentalfunction,Hou.ea2012:Elephanttranscriptomeprovidesinsights,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
+\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
 For example, each lineage has evolved expansive placenta-specific gene
 families with expression patterns that can vary in space and time
-\cite{Green.ea2000:Pregnancy-associatedbovineandovine,Simmons.ea2008:Spatialandtemporalexpression}.
+\cite{Green.ea2000:Pregnancy-associatedbovineandovine}.
 These include independent expansions of the prolactins in \mouse,
@@
 
+Supplemental \cref{tab:core_placenta_go_terms} lists the significant
+enrichment of \ac{GO} terms in the core placenta transcriptome;
+notable terms include those involved in cytoskeletal structure,
+protein transport, membrane protein structure, and inflammation
+resolution. Increasing the permissiveness of the core placenta
+transcriptome by only requiring 10 of 14 species to express 1:1
+orthologs with FPKM $>10$ and results in the same terms also being
+significantly enriched, along with additional terms (Supplemental
+\cref{tab:core_placenta_go_terms}). Supplemental
+\cref{tab:core_placenta_pathway_analysis} shows the pathways with
+significant and suggestive enrichment in the core placenta
+transcriptome.
+
 \subsubsection{Annexin Complexes}
@@
 \cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
-Supplemental \cref{tab:core_placenta_go_terms} lists the significant
-enrichment of \ac{GO} terms in the core placenta transcriptome;
-notable terms include those involved in cytoskeletal structure,
-protein transport, membrane protein structure, and inflammation
-resolution. Increasing the permissiveness of the core placenta
-transcriptome by only requiring 10 of 14 species to express 1:1
-orthologs with FPKM $>10$ and results in the same terms also being
-significantly enriched, along with additional terms (Supplemental
-\cref{tab:core_placenta_go_terms}).
 
@@
 maternal immune response. %
-Although there are speculative cases of eclampsia in great apes
-\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
+Although there are speculative cases of eclampsia in great apes, 
+%\cite{Baird1981:Eclampsiainlowlandgorilla,Stout.ea1969:Glomerularcapillaryendothelialswelling,Thornton.ea1992:Convulsionsinpregnancyin},
 it is exceedingly rare in these species
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
+\citeC{arter.ea2011:Evolutionofinvasiveplacentation}.
 

### 2016-11-18 manuscript/manuscript.Rnw
rewrite abstract

@@
   of \Sexpr{star.logs[,length(unique(file))]} term placentas of
-  \Sexpr{star.logs[,length(unique(species))]} species of mammals.
+  \Sexpr{star.logs[,length(unique(species))]} species of mammals, and
+  compared the evolution of expression across clades.
 
   \textbf{Results} We identified a set of
-  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
-  core genes which are expressed (FPKM $\ge 10$) in all placentas,
-
+  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
+  core genes which are expressed (FPKM $\ge 10$) in all placentas, %
+  %
   including genes with immune-modulating properties (\textit{ANXA2},
@@
   interactions (\textit{LAMC1}, \textit{LUM}, and \textit{LGALS1}),
-  and trophoblast fusion (\textit{ANXA5} and \textit{ANXA1}).
-
-  \textbf{Discussion}
-
-  Still, using the strictest filtering strategy
-  we find a core set of only
-  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
+  invasion (\textit{GRB2} and \textit{RALB5}) and trophoblast fusion
+  (\textit{ANXA5} and \textit{ANXA1}). %
+  We also identified multiple pre-eclampsia associated genes which are
+  differentially expressed in \textit{Homo sapiens} when compared to
+  the other %
+  \Sexpr{star.logs[,length(unique(species))-1]} species. %
+
+  \textbf{Discussion} %
+  Using the strictest filtering strategy we find a core set of only
+  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
   genes is essential for the proper functioning of the placenta in all
-  species analyzed.
+  species analyzed. We also found significantly higher expression of
+  the pre-eclampsia associated genes \textit{KISS1}, \textit{CRH},
+  \textit{PAPPA}, and \textit{ADAM12} on the \human lineage.
 

### 2016-11-18 manuscript/supplemental_material.Rnw
move core placenta pathway analysis up
fix human lineage table label

@@
                                     caption="\\lineagetableexplanationtypeiii{\\human}{all other species}",
-                                    label="tab:human_linage_specific_all"
+                                    label="tab:human_lineage_specific_all"
                                     )
@@
 
+\begin{table}
+<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
+pathway <- list()
+for (database in c("kegg","reactome","biocarta","netpath")) {
+    load(paste0("../data/placenta_gsea_",database))
+    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
+    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
+    pathway[[database]] <-
+        gsea.table
+}
+pathway.analysis <-
+    rbindlist(pathway)[fdr < 0.1,]
+rm(pathway)
+pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
+                                gsub("_+"," ",
+                                     Pathway))]
+pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]
+
+setnames(pathway.analysis,"p_values","p")
+setorder(pathway.analysis,"p")
+print(xtable(pathway.analysis,
+             display=c("s","s","g","g","s"),
+             align=c("c","p{3.5in}","r","r","c")),
+      math.style.exponents=TRUE,
+      include.rownames=FALSE,
+      booktabs=TRUE,
+      floating=FALSE)      
+@ 
+\caption[Core Placenta Pathway Enrichment]{Pathways with significant and suggestive enrichment
+  ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
+  biocarta, reactome, and kegg pathway databases}
+\label{tab:core_placenta_pathway_analysis}
+\end{table}
+
 \clearpage
@@
 
-\subsection{Pathway Analysis}
-\begin{table}
-<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
-pathway <- list()
-for (database in c("kegg","reactome","biocarta","netpath")) {
-    load(paste0("../data/placenta_gsea_",database))
-    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
-    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
-    pathway[[database]] <-
-        gsea.table
-}
-pathway.analysis <-
-    rbindlist(pathway)[fdr < 0.1,]
-rm(pathway)
-pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
-                                gsub("_+"," ",
-                                     Pathway))]
-pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]
-
-setnames(pathway.analysis,"p_values","p")
-setorder(pathway.analysis,"p")
-print(xtable(pathway.analysis,
-             display=c("s","s","g","g","s"),
-             align=c("c","p{3.5in}","r","r","c")),
-      math.style.exponents=TRUE,
-      include.rownames=FALSE,
-      booktabs=TRUE,
-      floating=FALSE)      
-@ 
-\caption[Core Placenta Pathway Enrichment]{Pathways with significant and suggestive enrichment
-  ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
-  biocarta, reactome, and kegg pathway databases}
-\label{tab:core_placenta_pathway_analysis}
-\end{table}
-
 \begin{table}

### 2016-11-18 manuscript/manuscript.Rnw
switch to single-column style
reduce length of manuscript some more

@@
 %\documentclass[english,12pt]{article}
-\documentclass[5p]{elsarticle_dla}
+%\documentclass[5p]{elsarticle_dla}
+\documentclass{elsarticle_dla}
 \usepackage{fontspec}
@@
 placenta have been found to vary substantially between \human and
-\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}, and the
-recruitment of \textit{SIGLEC6} to placental expression in \human
-differs from both \textit{Pan troglodytes} and \mouse
-\cite{Linden.ea2007:Human-specificexpressionofSiglec-6}.
-
-In order to address our two questions in a more systematic way, we
+\mouse \cite{Su.ea2002:Large-scaleanalysisofhuman}.
+
+To address our questions, we 
 sequenced and quantified the total RNA of mammalian placentas at term,
-including that of the \human, \textit{Pan paniscus} (the sister
+including \human, \textit{Pan paniscus} (the sister
 species to \textit{P. troglodytes}), \textit{Ateles fusciceps},
@@
 placentas of all species analyzed and therefore represent the
-\emph{core functionality of the placenta}; 2) identify the genes whose
+\emph{core functionality of the placenta}; 2) identify genes whose
 expression level changes significantly in each of the morphology
@@
 \begin{sloppypar}
-The quality of reads was evaluated using FASTQC v0.11.2
+Reads quality was evaluated using FASTQC v0.11.2
 (\url{http://www.bioinformatics.babraham.ac.uk/projects/fastqc/}).
@@
 either the \human or \mouse transcriptome (Supplemental
-\cref{tab:ref_genome_aligned_species}). For annotation using Diamond,
-the match with the highest bitscore was chosen as the annotation;
-transcripts where the highest scoring match had a bitscore below 40
-were left unannotated.
-% 
-All transcriptomes were then aligned to the appropriate reference (or
+\cref{tab:ref_genome_aligned_species}), and the
+the match with the highest bitscore was chosen as the annotation.
+Transcripts where the highest scoring match had a bitscore below 40
+were left un-annotated.
+% 
+All transcriptomes were aligned to the reference (or
 assembled) genome using STAR v2.4.2a~%
@@
 %
-The genome of the closely-related \textit{P. troglodytes} was used as
-a reference for \bonobo. The separation between the two recognized Pan
-species is generally perceived to be $\approx 1$ million years
-\cite{Prüfer.ea2012:bonobogenomecomparedwith}. %
-%
+The genome of the closely-related \textit{P. troglodytes} (separation
+$\approx 1$ million years
+\cite{Prüfer.ea2012:bonobogenomecomparedwith}) was used as a reference
+for \bonobo %
+.
 \end{sloppypar}
@@
 
-All 1:1 orthologs present among \opossum, \textit{Pan troglodytes},
-\cow, \dog, \elephant, \mouse, and \human were determined using
+1:1 orthologs in were determined using
 Ensembl v.80 gene trees via Biomart using the biomaRt R package
@@
   1,length(unique(human_name))]} 1:1 orthologous genes were expressed
-in at least one species. Of these, \Sexpr{combined.fpkm[species!="homo
-  sapiens" & mean_fpkm >= 100,length(unique(human_name))]} 1:1
-orthologous genes with \ac{FPKM} $\ge 100$ were retained for future
-analyses of gene expression evolution described below.
+in at least one species.
 
@@
 
-Gene sets and pathways in were obtained from MSigdb v5.1 hallmark gene
-set \cite{Liberzon.ea2011:MolecularsignaturesdatabaseMSigDB} and Gene
-Ontology which represent a superset of available pathway databases,
-including Reactome, Kegg, Netpath, and GO. We calculated the
-over-representation of gene lists in these gene sets using Fisher's
-exact test and corrected for multiple testing by estimating the
-\ac{FDR} using the Benjamini and Hochberg procedure
+Gene sets and pathways in were obtained from MSigdb v5.1
+\cite{Liberzon.ea2011:MolecularsignaturesdatabaseMSigDB} and Gene
+Ontology to form a superset of available pathway databases. We
+calculated the over-representation of gene lists in these gene sets
+using Fisher's exact test and corrected for multiple testing by
+estimating the \ac{FDR} using the Benjamini and Hochberg procedure
 \cite{Benjamini1995:bh_procedure}.
@@
 \href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
-(Supplemental \cref{tab:srr_numbers}). The mean FPKM for gene in each
+(Supplemental \cref{tab:srr_numbers}). The mean \ac{FPKM} for gene in each
 species with ortholog annotations are located in the supplemental data
-file all\_species\_mean\_fpkm.txt.xz and the per-sample FPKM values are in
+file all\_species\_mean\_fpkm.txt.xz and the per-sample \ac{FPKM} values are in
 all\_species\_per\_sample\_fpkm.txt.xz.
@@
 \Sexpr{star.logs[,length(unique(species))]} species spanning the
-phylogenetic tree of mammals. This wealth of data revealed both
-lineage-specific expansions and differential expression of gene
+phylogenetic tree of mammals. This revealed both
+lineage-specific differential expression of gene
 families as well as a set of core, non-housekeeping genes which are
@@
 
-There is diversity in the placental expression of hormones, including
-estrogen, progesterone, prolactin, relaxin, the growth hormone family, and
-natriuretic peptides among the species and lineages analyzed.
+The core placenta transcriptome contains genes which are involved in
+maternal-fetal tolerance, such as the annexins, and genes responsible
+for cell-cell and cell-matrix interactions, such as \textit{LGALS1}.
+These genes were almost certainly present in the placenta of the
+common mammalian ancestor. We hypothesize that modifications to their
+expression or function will compromise the development and
+organization of the placenta.
+
 We also identified genes differentially expressed on the \human
@@
 \textit{ADAM12}, \textit{CRH}, and \textit{KISS1}.
-
-The core placenta transcriptome was also identified, which contains
-genes which are involved in maternal-fetal tolerance, such as the
-annexins, and genes responsible for cell-cell and cell-matrix
-interactions, such as \textit{LGALS1}. These genes were almost
-certainly present in the placenta of the common mammalian ancestor. We
-hypothesize that modifications to their expression or function will
-compromise the development and organization of the placenta.
 

### 2016-11-28 manuscript/supplemental_material.Rnw
fix placenta classification plot

@@
 <<placental_classification_table,echo=FALSE,results="asis">>=
-print(xtable(placenta_classification,
+plac.class <- copy(placenta_classification[,list(species,shape,interdigitation,intimacy)])
+plac.class[,`species`:=paste0("\\textit{",gsub("^(.)","\\U\\1",`species`,perl=TRUE),"}")]
+plac.class[,`interdigitation`:=gsub("^(.)","\\U\\1",`interdigitation`,perl=TRUE)]
+plac.class[,`shape`:=gsub("^(.)","\\U\\1",`shape`,perl=TRUE)]
+plac.class[,`intimacy`:=gsub("^(.)","\\U\\1",`intimacy`,perl=TRUE)]
+setnames(plac.class,colnames(plac.class),gsub("^(.)","\\U\\1",colnames(plac.class)))
+
+print(xtable(plac.class,
              display=c("s","s","s","s","s"),
@@
       floating=FALSE,
+      sanitize.text.function=genes.table.sanitize,
       booktabs=TRUE

### 2016-11-28 manuscript/manuscript.Rnw
fix carter typo; turn of acro macros
LGALS1 is immunomodulating too

@@
 %\usepackage{setspace}
-\usepackage{acro}
+\usepackage[macros=false]{acro}
 \usepackage{array}
@@
   of \Sexpr{star.logs[,length(unique(file))]} term placentas of
-  \Sexpr{star.logs[,length(unique(species))]} species of mammals, and
-  compared the evolution of expression across clades.
+  \Sexpr{star.logs[,length(unique(species))]} species of mammals
+  representing all major eutherian superordinal clades and marsupials,
+  and compared the evolution of expression across clades.
 
@@
   including genes with immune-modulating properties (\textit{ANXA2},
-  \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}), cell-cell
+  \textit{ANXA1}, \textit{S100A11}, \textit{S100A10}, and \textit{LGALS1}), cell-cell
   interactions (\textit{LAMC1}, \textit{LUM}, and \textit{LGALS1}),
-  invasion (\textit{GRB2} and \textit{RALB5}) and trophoblast fusion
+  invasion (\textit{GRB2} and \textit{RALB5}) and syncytialization
   (\textit{ANXA5} and \textit{ANXA1}). %
@@
 it is exceedingly rare in these species
-\citeC{arter.ea2011:Evolutionofinvasiveplacentation}.
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation}.
 

### 2016-11-28 manuscript/manuscript.Rnw
association was missing in intro

@@
 differ drastically in the number and size of their offspring, as well
-as the length of gestation; these traits are often with specific
-aspects of placental morphology
+as the length of gestation; these traits are often associated with
+specific aspects of placental morphology
 \cite{Capellini.ea2011:Placentationandmaternalinvestment}

### 2016-11-28 manuscript/manuscript.Rnw
add metatheria to species tree

@@
     geom_text(aes(label=gsub("_"," ",label)),hjust=0,size=5,fontface="italic")+
+    geom_text(data=data.frame(x=1,y=1),label="Metatheria",hjust=0,vjust=0,size=5,fontface="italic")+
     scale_x_continuous(limits=c(0,10))

### 2016-11-28 manuscript/manuscript.Rnw
switch phylopic for human
they were sequenced with 76bp
add missing discussion of 100bp sequenced reads

@@
       ## human
-      %>% phylopic("a9f4ebd5-53d7-4de9-9e66-85ff6c2d513e",color="#CCCC44",alpha=0.3,node=13,width=0.18)
+      %>% phylopic("acf1cbec-f6ef-4a82-8ab5-be2f963b93f5",color="#CCCC44",alpha=0.3,node=13,width=0.3)
       )
@@
 alignment statistics for all species. For transcriptomes sequenced
-with 75 bp %
+with 76 bp %
 %
@@
 with an average of \Sexpr{star.logs.wide[input.read.length > 120 & input.read.length < 160,mean(`% good reads`)]}\% %
+of reads mapping to the
+respective reference genome (Supplemental \cref{tab:ref_genome_aligned_species}).
+%
+%
+For transcriptomes sequenced with 100 bp %
+% The mismatch here is because we're reporting trimmed reads here, not
+% full length reads
+(\Sexpr{species(star.logs.wide[input.read.length > 80 & input.read.length < 120,unique(species)])}),
+%
+the number of trimmed reads ranged from %
+\Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,min(as.numeric(input.reads))]} 
+%
+to %
+\Sexpr{star.logs.wide[input.read.length  > 80 & input.read.length < 120,max(as.numeric(input.reads))]}
+%
+with an average of \Sexpr{star.logs.wide[input.read.length > 80 & input.read.length < 120,mean(`% good reads`)]}\% %
 of reads mapping to the

### 2016-11-28 manuscript/manuscript.Rnw
add missing significant
the top 1% and top 100 are by FPKM

@@
 \cref{tab:ref_genome_aligned_species}). Read length was not a
-predictor of unmapped read percentage (Supplemental
+significant predictor of unmapped read percentage (Supplemental
 \cref{fig:transcriptome_reads_vs_read_length}). The expression of
@@
 
-We performed \ac{gsea} on both the top 1\% of transcripts
+We performed \ac{gsea} on both the top 1\% of transcripts by \ac{FPKM}
 (Supplemental \cref{tab:go_terms_enriched_top_one_ateles_fusciceps,tab:go_terms_enriched_top_one_homo_sapiens,tab:go_terms_enriched_top_one_pan_paniscus,tab:go_terms_enriched_top_one_mus_musculus,tab:go_terms_enriched_top_one_spalax_galili,tab:go_terms_enriched_top_one_spalax_carmeli,tab:go_terms_enriched_top_one_bos_taurus,tab:go_terms_enriched_top_one_ovis_aries,tab:go_terms_enriched_top_one_sus_scrofa,tab:go_terms_enriched_top_one_equus_caballus,tab:go_terms_enriched_top_one_canis_familiaris,tab:go_terms_enriched_top_one_loxodonta_africana,tab:go_terms_enriched_top_one_dasypus_novemcinctus,tab:go_terms_enriched_top_one_monodelphis_domestica}) and the top
-100 transcripts (Supplemental
+100 transcripts by \ac{FPKM} (Supplemental
 \cref{tab:go_terms_enriched_ateles_fusciceps,tab:go_terms_enriched_homo_sapiens,tab:go_terms_enriched_pan_paniscus,tab:go_terms_enriched_mus_musculus,tab:go_terms_enriched_spalax_galili,tab:go_terms_enriched_spalax_carmeli,tab:go_terms_enriched_bos_taurus,tab:go_terms_enriched_ovis_aries,tab:go_terms_enriched_sus_scrofa,tab:go_terms_enriched_equus_caballus,tab:go_terms_enriched_canis_familiaris,tab:go_terms_enriched_loxodonta_africana,tab:go_terms_enriched_dasypus_novemcinctus,tab:go_terms_enriched_monodelphis_domestica}); common processes

### 2016-11-28 manuscript/manuscript.Rnw
explain what FC means and how it is calculated
describe how order was defined

@@
 lme4~\cite{Bates.ea2015:FittingLinearMixed-EffectsModels} with each
-species having separate random effects.
+species having separate random effects. Shifts in expression levels
+are reported with $\log_2$ Fold-Change (FC), rounded to the nearest
+integer.
 
@@
 the response variable and the morphology as the predictor variable.
-Placenta morphology with a defined order (interdigitation, intimacy)
-were additionally analyzed using proportional odds logistic regression
-using the MASS package in R~\cite{Venables2002:MASS}.
+Placenta morphology with a definable order (interdigitation and
+intimacy) were additionally analyzed using proportional odds logistic
+regression using the MASS package in R~\cite{Venables2002:MASS}.
 

### 2016-11-28 manuscript/manuscript.Rnw
the dog fetus died too
use the long names everywhere

@@
 
-Previous studies of some species (i.e., \textit{Homo sapiens},
-\textit{Mus musculus}, \textit{Bos taurus}) have provided some
-preliminary answers to these questions. They revealed a stunning
-divergence in the genetics of the placenta, especially in later stages
-of pregnancy
-\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman}.
-For example, each lineage has evolved expansive placenta-specific gene
+Previous studies of some species (i.e., \human, \mouse, \cow, and
+\elephant) have provided some preliminary answers to these questions.
+They revealed a stunning divergence in the genetics of the placenta,
+especially in later stages of pregnancy
+\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman,Hou.ea2012:Elephanttranscriptomeprovidesinsights}}.
+For example, each superorder has evolved expansive placenta-specific gene
 families with expression patterns that can vary in space and time
-\cite{Green.ea2000:Pregnancy-associatedbovineandovine}.
-These include independent expansions of the prolactins in \mouse,
-\textit{Spalax carmeli}, \textit{Spalax galili}, and \cow, \acp{PSG}
-in \human and \mouse, \acp{PAG} in \cow, genes related to the
-thymus-specific serine protease \textit{PRSS16} in \textit{Loxodona
-  africana} (in this work), and the expansion of the growth hormone
-and chorionic gonadotropin beta genes in anthropoid primates
+\cite{Green.ea2000:Pregnancy-associatedbovineandovine}. These include
+independent expansions of the prolactins in \mouse, \carmeli, \galili,
+and \cow, \acp{PSG} in \human and \mouse, \acp{PAG} in \cow, genes
+related to the thymus-specific serine protease \textit{PRSS16} in
+\elephant (in this work), and the expansion of the growth hormone and
+chorionic gonadotropin beta genes in anthropoid primates
 \cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
@@
 sequenced and quantified the total RNA of mammalian placentas at term,
-including \human, \textit{Pan paniscus} (the sister
-species to \textit{P. troglodytes}), \textit{Ateles fusciceps},
-\mouse, \galili, \carmeli, \textit{Canis familiaris}, \cow, \elephant,
-\textit{Dasypus novemcinctus}, and \textit{Monodelphis domesticus}.
+including \human, \bonobo (the sister
+species to \textit{Pan troglodytes}), \smonkey,
+\mouse, \galili, \carmeli, \dog, \cow, \elephant,
+\armadillo, and \opossum.
 These species represent four major branches of eutherian mammals
@@
 \cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
-\textit{Equus caballus}
-\cite{Coleman.ea2010:Structuralannotationofequine}, \textit{Ovis
-  aries} \cite{Archibald.ea2010:sheepgenomereferencesequence}, and
-\textit{Sus scrofa} \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
+\horse
+\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
+\cite{Archibald.ea2010:sheepgenomereferencesequence}, and
+\pig \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
 identify the non-housekeeping genes which are expressed in the
@@
 delivery via a cesarean section. Near-term placental tissue from \dog
-was collected upon the death of the mother from a car accident.
+was collected upon the death of the mother and fetus from a car accident.
 Placental tissue from \opossum was sampled at fetal stage 32, roughly
@@
 For more complete evolutionary comparisons additional sequences of
-placenta in \mouse
+placenta in \mouse [mouse]
 (n=\Sexpr{star.logs[grepl("musculus",species),length(unique(file))]})
@@
 (n=\Sexpr{star.logs[grepl("sapiens",species),length(unique(file))]})
-\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse
-\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep
-\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig
+\cite{Hughes.ea2015:Evaluatingintra-andinter-individual} \horse [horse]
+\cite{Coleman.ea2010:Structuralannotationofequine}, \sheep [sheep] 
+\cite{Archibald.ea2010:sheepgenomereferencesequence}, and \pig [pig]
 \cite{Ahanda.ea2012:Predictionofaltered3-} were obtained from the
@@
 using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
-either the \human or \mouse transcriptome (Supplemental
+either the \human (for \smonkey) or \mouse (for both Spalacines) transcriptome (Supplemental
 \cref{tab:ref_genome_aligned_species}), and the

### 2016-11-28 manuscript/manuscript.Rnw
add priya's comments and changes
fix space error

@@
 \newcommand{\DEW}[1]{\textcolor{purple}{\fxnote{DEW: #1}}}
+\newcommand{\PP}[1]{\textcolor{orange}{\fxnote{PP: #1}}}
 % \usepackage[x11names,svgnames]{xcolor}
@@
 
+\PP{Introduce the characteristics earlier in the introduction}
+
 \section{Discussion}
@@
 
+\PP{Add refs for fusion}
+
 % \DLA{Possibility of membrane repair due to maternal immune response as
@@
 \subsubsection{Invasion: EGF and CD9 in the Core Placenta}
+
+\DEW{Write out FDR for enrichment described}
 
@@
 localization (\textit{ANXA1} and \textit{ANXA2}) in the core placenta
-is consistent with previous findingsshowing that alterations to the
+is consistent with previous findings showing that alterations to the
 EGF signaling system change invasion of trophoblast cells into the
@@
 
+\PP{EGF interacts with S100 from the core placenta as well}
+
 % TAGLN2 -- likely to be produced by the placenta as well; possibly

### 2016-11-28 manuscript/manuscript.Rnw
fix syncytin typo
cite hutchinson for maternal immune response

@@
 regulators as well as the inability to sufficiently suppress the
-maternal immune response. %
+maternal immune response~\cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. %
 Although there are speculative cases of eclampsia in great apes, 
@@
 
+\DEW{Why?}
+
 \subsubsection{Formation of the syncytiotrophoblast}
 
-While Syncitin-1 and -2 are implicated in the formation of the
+While Syncytin-1 and -2 are implicated in the formation of the
 syncytiotrophoblast in

### 2016-11-28 manuscript/manuscript.Rnw
fix inclusion of housekeeping genes
add derek comments

@@
 
+  \DEW{The discussion and results seem redundant. Given the title
+    shouldn't something about shape be mentioned?}
+
 \end{abstract}
@@
 \end{table}
+
+\DEW{Add gene names}
 
@@
 The
-\Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
+\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
 genes comprising the core placenta transcriptome are shown in
@@
 
-Supplemental \cref{tab:shape_genes} shows the top 50 genes which are
-associated with the shape of placenta attachment, as well as the shape
-in which the difference is most significant. The most significant
-genes which were associated with diffuse placenta attachment include
-\textit{EREG}, a ligand of EGFR, \textit{UNC93A}, a homolog of
-\textit{UNC93B1} which is involved in the balancing between Tlr7 and
-Tlr9 immune response in
+Supplemental \cref{tab:shape_genes} shows the 50 genes which are most
+significantly associated with the shape of placenta attachment, as
+well as the shape in which the difference is most significant. The
+most significant genes which were associated with diffuse placenta
+attachment include \textit{EREG}, a ligand of EGFR, \textit{UNC93A}, a
+homolog of \textit{UNC93B1} which is involved in the balancing between
+Tlr7 and Tlr9 immune response in
 mice~\cite{Brinkmann.ea2007:interactionbetweenERmembrane}, and
@@
 
+\DEW{Need to cite NWM grants, IGB, etc.}
+
 %  \printaddresses

### 2016-11-29 data/placenta_core_transcriptome.R
add all but metatherian

@@ -48,11 +48,15 @@ combined.fpkm.wide[,all_but_one_expression :=
                        apply(combined.fpkm.wide[,-1,with=FALSE],1,min.but.one)]
 combined.fpkm.wide[,all_expression :=
                        apply(combined.fpkm.wide[,-1,with=FALSE],1,min.na.zero)]
+combined.fpkm.wide[,all_but_metatherian_expression :=
+                        apply(combined.fpkm.wide[,!grepl("^(monodelphis domestica|human_name|.*_expression)$",
+                                                         colnames(combined.fpkm.wide)),
+                                                 with=FALSE],1,min.na.zero)]
 setkey(combined.fpkm.wide,"human_name")
 
 ## all genes which are not housekeeping genes which have minimum 
 core.placenta.transcriptome.long <- 
-    data.table(melt(combined.fpkm.wide[all_expression >= 10,
+    data.table(melt(combined.fpkm.wide[all_but_metatherian_expression >= 10,
                                        ][order(-median_expression)
                                          ],
                     id.vars="human_name",
@@ -71,6 +75,7 @@ core.placenta.transcriptome.long[,human_name:=
 core.placenta.transcriptome.genes.shape <-
     combined.fpkm.wide[,list(human_name,
                              median_expression,
+                             all_but_metatherian_expression,
                              all_expression,
                              all_but_one_expression,
                              all_but_two_expression,
@@ -85,16 +90,17 @@ placenta.transcriptome.genes <-
 pts <- data.table(all.genes=all.genes)
 pts <- pts[,placenta.ts:=all.genes %in% placenta.transcriptome.genes]
 pts <- pts[,placenta.ts.permissive:=all.genes %in%
-                core.placenta.transcriptome.genes.shape[all_but_four_expression > 10 &
+                core.placenta.transcriptome.genes.shape[(all_but_four_expression > 10  |
+                                                         all_but_metatherian_expression > 10 ) &
                                                         ! is.na(human_name),
                                                         unique(human_name)]
            ]
 pts[,placenta.ts.all:=all.genes %in%
-         core.placenta.transcriptome.genes.shape[all_expression > 10 &
+         core.placenta.transcriptome.genes.shape[all_but_metatherian_expression > 10 &
                                                  ! is.na(human_name),
                                                  unique(human_name)]]
 pts[,placenta.ts.one:=all.genes %in%
-         core.placenta.transcriptome.genes.shape[all_expression > 1 &
[…]

### 2016-11-29 data/placenta_core_transcriptome_complete.R

@@ -48,9 +48,14 @@ combined.fpkm.wide[,all_but_one_expression :=
                        apply(combined.fpkm.wide[,-1,with=FALSE],1,min.but.one)]
 combined.fpkm.wide[,all_expression :=
                         apply(combined.fpkm.wide[,-1,with=FALSE],1,min.na.zero)]
+combined.fpkm.wide[,all_but_metatherian_expression :=
+                        apply(combined.fpkm.wide[,!grepl("^(monodelphis domestica|human_name|.*_expression)$",
+                                                         colnames(combined.fpkm.wide)),
+                                                 with=FALSE],1,min.na.zero)]
 placenta.core.transcriptome.complete <-
     combined.fpkm.wide[,list(human_name,
                              median_expression,
+                             all_but_metatherian_expression,
                              all_expression,
                              all_but_one_expression,
                              all_but_two_expression,

### 2016-11-29 manuscript/acronyms.tex
add gsea acronym
use long names for genes where possible

@@
 \DeclareAcronym{FDR}{long=False Discovery Rate}
+\DeclareAcronym{gsea}{short=GSEA,long=gene set enrichment analysis}
+\DeclareAcronym{CRH}{short=\textit{CRH},long={corticotropin-releasing hormone}}
+\DeclareAcronym{KISS1}{short=\textit{KISS1},long={KiSS-1 metastatis-supressor}}
+\DeclareAcronym{PAPPA}{short=\textit{PAPPA},long={pappalysin 1}}
+\DeclareAcronym{ADAM12}{short=\textit{ADAM12},long={ADAM Metallopeptidase Domain 12}}
+\DeclareAcronym{ANXA1}{short=\textit{ANXA1},long={annexin A1}}
+\DeclareAcronym{ANXA5}{short=\textit{ANXA5},long={annexin A5}}
+\DeclareAcronym{S100A11}{short=\textit{S100A11},long={S100 calcium binding protein A11}}
+\DeclareAcronym{S100A10}{short=\textit{S100A10},long={S100 calcium binding protein A10}}
+\DeclareAcronym{LGALS1}{short=\textit{LGALS1},long={galectin 1}}
+\DeclareAcronym{LAMC1}{short=\textit{LAMC1},long={laminin subunit gamma 1}}
+\DeclareAcronym{LUM}{short=\textit{LUM},long={lumican}}
+\DeclareAcronym{GRB2}{short=\textit{GRB2},long={growth factor receptor bound protein 2}}
+\DeclareAcronym{RALB}{short=\textit{RALB},long={RAS like proto-oncogene B}}
+\DeclareAcronym{ERVFRD-1}{short=\textit{ERVFRD-1},long={endogenous retrovirus group FRD member 1}}
+\DeclareAcronym{ERVW-1}{short=\textit{ERVW-1},long={endogenous retrovirus group W member 1}}
+\DeclareAcronym{MFSD2A}{short=\textit{MFSD2A},long={major facilitator superfamily domain containing 2A}}
+\DeclareAcronym{PRSS16}{short=\textit{PRSS16},long={protease, serine 16}}
+\DeclareAcronym{TIMP3}{short=\textit{TIMP3},long={TIMP metallopeptidase inhibitor 3}}
+\DeclareAcronym{F11R}{short=\textit{F11R},long={F11 receptor}}
+\DeclareAcronym{SDC4}{short=\textit{SDC4},long={syndecan 4}}
+\DeclareAcronym{FN1}{short=\textit{FN1},long={fibronectin 1}}
+\DeclareAcronym{ITGB1}{short=\textit{ITGB1},long={integrin subunit beta 1}}
+\DeclareAcronym{ITGB1BP1}{short=\textit{ITGB1BP1},long={integrin subunit beta 1 binding protein 1}}
+\DeclareAcronym{EREG}{short=\textit{EREG},long={epiregulin}}
+\DeclareAcronym{UNC93A}{short=\textit{UNC93A},long={unc-93 homolog A (\textit{C. elegans})}}
+\DeclareAcronym{UNC93B1}{short=\textit{UNC93B1},long={unc-93 homolog B1 (\textit{C. elegans)}}}
+\DeclareAcronym{WNT5A}{short=\textit{WNT5A},long={Wnt family member 5A}}
+\DeclareAcronym{IL1RL1}{short=\textit{IL1RL1},long={interleukin 1 receptor like 1}}
+\DeclareAcronym{PLAC1}{short=\textit{PLAC1},long={placenta specific 1}}
+\DeclareAcronym{Afp}{short=\textit{Afp},long={alpha fetoprotein}}
+\DeclareAcronym{Gjb2}{short=\textit{Gjb2},long={gap junction protein beta 2}}
+\DeclareAcronym{CD9}{short=\textit{CD9},long={CD9 molecule}}
 

### 2016-11-29 manuscript/manuscript.Rnw
use long names for genes where possible

@@
   %
-  including genes with immune-modulating properties (\textit{ANXA2},
-  \textit{ANXA1}, \textit{S100A11}, \textit{S100A10}, and \textit{LGALS1}), cell-cell
-  interactions (\textit{LAMC1}, \textit{LUM}, and \textit{LGALS1}),
-  invasion (\textit{GRB2} and \textit{RALB5}) and syncytialization
-  (\textit{ANXA5} and \textit{ANXA1}). %
+  including genes with immune-modulating properties (\ac{ANXA2},
+  \ac{ANXA1}, \ac{S100A11}, \ac{S100A10}, and \ac{LGALS1}), cell-cell
+  interactions (\ac{LAMC1}, \ac{LUM}, and \ac{LGALS1}),
+  invasion (\ac{GRB2} and \ac{RALB}) and syncytialization
+  (\ac{ANXA5} and \ac{ANXA1}). %
   We also identified multiple pre-eclampsia associated genes which are
@@
   species analyzed. We also found significantly higher expression of
-  the pre-eclampsia associated genes \textit{KISS1}, \textit{CRH},
-  \textit{PAPPA}, and \textit{ADAM12} on the \human lineage.
+  the pre-eclampsia associated genes \ac{KISS1}, \ac{CRH},
+  \ac{PAPPA}, and \ac{ADAM12} on the \human lineage.
 
@@
 \maketitle
+
+% Reset all acronyms
+\acresetall
 
@@
 and \cow, \acp{PSG} in \human and \mouse, \acp{PAG} in \cow, genes
-related to the thymus-specific serine protease \textit{PRSS16} in
+related to the thymus-specific serine protease \ac{PRSS16} in
 \elephant (in this work), and the expansion of the growth hormone and
@@
 
-\DEW{Add gene names}
+% \DEW{Add gene names}
 
@@
 genes which are thought to be involved in pre-eclampsia. Notably,
-\textit{CRH} (FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
+\ac{CRH} (FDR=\Sexpr{hum.all.lin.spec["CRH",`FDR`]},
 FC=\Sexpr{hum.all.lin.spec["CRH",`Fold Change`]}, Supplemental
@@
 where it is hypothesized to activate decidual macrophages. Likewise,
-\textit{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
+\ac{KISS1} (\Sexpr{hum.all.lin.spec["KISS1",`FDR`]},
 FC=\Sexpr{hum.all.lin.spec["KISS1",`Fold Change`]}, Supplemental
@@
 involved in modulating trophoblast invasion. Two other genes,
-\textit{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
+\ac{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
 FC=\Sexpr{hum.all.lin.spec["PAPPA",`Fold Change`]} Supplemental
-\cref{fig:pappa_evolution_tree}) and \textit{ADAM12}
+\cref{fig:pappa_evolution_tree}) and \ac{ADAM12}
 (\Sexpr{hum.all.lin.spec["ADAM12",`FDR`]},
@@
 Multiple components of annexin complexes are included in the core
-transcriptome, including \textit{ANXA2}, \textit{ANXA1},
-\textit{S100A11}, \textit{S100A10}. ANXA1 is involved in the
+transcriptome, including \ac{ANXA2}, \ac{ANXA1},
+\ac{S100A11}, \ac{S100A10}. ANXA1 is involved in the
 resolution of the inflammation response through inhibition of
@@
 secretion. The expression regulation of another core gene,
-\textit{TIMP3}, has been implicated in pre-eclampsia
+\ac{TIMP3}, has been implicated in pre-eclampsia
 \cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
@@
 Another annexin which is highly expressed in all term placentas is
-\textit{ANXA5}, which was originally identified as a an anti-coagulant
+\ac{ANXA5}, which was originally identified as a an anti-coagulant
 protein, but is also involved in syncitium formation, membrane repair,
@@
 abortion~\cite{Ueki.ea2012:Lossofmaternalannexin}. Other notable
-syncytiotrophoblast fusion proteins, Syncytin-1 (\textit{ERVW-1}) and -2
-(\textit{ERVFRD-1}) were not present in the core placenta, as they (or
+syncytiotrophoblast fusion proteins, Syncytin-1 (\ac{ERVW-1}) and -2
+(\ac{ERVFRD-1}) were not present in the core placenta, as they (or
 their homologs) are not highly expressed in \dog, \bonobo, or \sheep,
-nor was \textit{MFSD2A}, the Syncytin-2 receptor (\ac{FPKM} $\ge 10$
+nor was \ac{MFSD2A}, the Syncytin-2 receptor (\ac{FPKM} $\ge 10$
 only in \human, \mouse, and \dog).
@@
 cytoskeleton pathway is depicted in Supplemental
-\cref{fig:kegg_actin}, which shows that \textit{FN1} and \textit{ITG}
-are highly expressed, which activate Rac, and potentially lead to the
-formation of focal adhesions and lamellipodia formation through their
-inhibition of GSN and HSPC300. Other components of cell-cell
-adhesion, including \textit{F11R} which forms tight junctions in
-epithelia, proteglycans \textit{SDC4} and \textit{LUM}, and the
-laminin \textit{LAMC1}.
+\cref{fig:kegg_actin}, which shows that \ac{ITGB1} and its binding
+protein \ac{ITGB1BP1} as well as \ac{FN1} are highly expressed, which
+activates Rac. This potentially leads to the formation of focal adhesions
+and lamellipodia formation through inhibition of GSN and
+HSPC300. Other components of cell-cell adhesion, including \ac{F11R}
+which forms tight junctions in epithelia, the proteglycans \ac{SDC4} and
+\ac{LUM}, and \ac{LAMC1}.
 
@@
 most significant genes which were associated with diffuse placenta
-attachment include \textit{EREG}, a ligand of EGFR, \textit{UNC93A}, a
-homolog of \textit{UNC93B1} which is involved in the balancing between
+attachment include \ac{EREG}, a ligand of EGFR, \ac{UNC93A}, a
+homolog of \ac{UNC93B1} which is involved in the balancing between
 Tlr7 and Tlr9 immune response in
 mice~\cite{Brinkmann.ea2007:interactionbetweenERmembrane}, and
-\textit{WNT5A}, which is a member of the Wnt ligand family whose
-singaling pathway has been implicated in motility in human trophoblast
+\ac{WNT5A}, which is a member of the Wnt ligand family whose
+signaling pathway has been implicated in motility in human trophoblast
 cells~\cite{Gupta.ea2016:CellSignalingPathwaysInvolved}.
@@
 
-The identification of multiple genes (\textit{CRH}, \textit{ADAM12},
-\textit{KISS1}, \textit{PAPPA}, \textit{IL1RL1}, and \textit{PLAC1};
+The identification of multiple genes (\ac{CRH}, \ac{ADAM12},
+\ac{KISS1}, \ac{PAPPA}, \ac{IL1RL1}, and \ac{PLAC1};
 \cref{tab:human_lineage_specific}) which are associated with
@@
 which are also more highly expressed in the \mouse lineage such as
-\textit{Afp} and \textit{Gjb2}. Furthermore, the onset of
+\ac{Afp} and \ac{Gjb2}. Furthermore, the onset of
 pre-eclampsia is well before term; all of our placentas are term
@@
 Components of the annexin complexes, including
-\textit{ANXA2}, \textit{ANXA1}, \textit{S100A11}, and \textit{S100A10}
+\ac{ANXA2}, \ac{ANXA1}, \ac{S100A11}, and \ac{S100A10}
 are expressed in all placentas examined, and argue for the evolution
@@
 \human~\cite{Rawn.ea2008:evolutionregulationandfunction} and homologs
-of \textit{ERVFRD-1} and \textit{ERVW-1} are widely expressed in most
+of \ac{ERVFRD-1} and \ac{ERVW-1} are widely expressed in most
 of the placentas analyzed here, they are not expressed in \bonobo,
@@
 syncytiotrophoblast is likely to be highly expressed in the core
-placenta. \textit{ANXA5} is known to be expressed at multiple stages
+placenta. \ac{ANXA5} is known to be expressed at multiple stages
 of placentation, and is capable of repairing membrane defects and
@@
 are responsible for the positioning of cytotrophoblast cells in close
-proximity to start the fusion process, \textit{ANXA5} is likely to be
+proximity to start the fusion process, we hypothesize that \ac{ANXA5} is
 essential for the completion of fusion and the maintenance of membrane
@@
 The high expression of components of the EGFR signaling pathway in the
-core placenta transcriptome (including \textit{GRB2}, \textit{RALB})
+core placenta transcriptome (including \ac{GRB2}, \ac{RALB})
 as well as annexins which are involved in the location of EGF receptor
-localization (\textit{ANXA1} and \textit{ANXA2}) in the core placenta
+localization (\ac{ANXA1} and \ac{ANXA2}) in the core placenta
 is consistent with previous findings showing that alterations to the
@@
 spiral arteries~\cite{Armant.ea2015:Reducedexpressionofepidermal}.
-Another component of the core placenta transcriptome, \textit{CD9},
+Another component of the core placenta transcriptome, \ac{CD9},
 has been shown to suppress \human extravillous trophoblast
@@
 maternal-fetal tolerance, such as the annexins, and genes responsible
-for cell-cell and cell-matrix interactions, such as \textit{LGALS1}.
+for cell-cell and cell-matrix interactions, such as \ac{LGALS1}.
 These genes were almost certainly present in the placenta of the
@@
 lineage which are implicated in pre-eclampsia, including
-\textit{ADAM12}, \textit{CRH}, and \textit{KISS1}.
+\ac{ADAM12}, \ac{CRH}, and \ac{KISS1}.
 

### 2016-11-29 manuscript/gene_names.sh

@@ -1,0 +1,6 @@
+#!/bin/bash
+
+for gene in "$@"; do
+    name="$(echo "$gene" | gene_to_aliases|awk -F"$(echo -ne "\t")" '{print $2}')"
+    echo "\\DeclareAcronym{$gene}{short=\\textit{$gene},long={$name}}";
+done;

### 2016-11-29 manuscript/manuscript.Rnw
make the species tree 90mm wide

@@
 \begin{figure}
-<<species.tree,fig.width=8,fig.height=6,out.width="\\columnwidth",echo=FALSE,message=FALSE,warning=FALSE>>=
+<<species.tree,fig.width=8,fig.height=6,out.width="90mm",echo=FALSE,message=FALSE,warning=FALSE>>=
 species.tree <- 
@@
       ## bonobo
-      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.3,node=12,width=0.18)
+      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.3,node=12,width=0.15)
       ## spider monkey
@@
       ## human
-      %>% phylopic("acf1cbec-f6ef-4a82-8ab5-be2f963b93f5",color="#CCCC44",alpha=0.3,node=13,width=0.3)
+      %>% phylopic("acf1cbec-f6ef-4a82-8ab5-be2f963b93f5",color="#AAAA11",alpha=0.3,node=13,width=0.4)
       )

### 2016-11-29 manuscript/manuscript.Rnw
use the eutherian only core placenta

@@
 \begin{figure}[tbh]
-<<core.placenta.transcriptome,fig.width=5,fig.height=8,out.height="0.8\\textheight",echo=FALSE>>=
+<<core.placenta.transcriptome,fig.width=9/2.5,fig.height=8,out.width="90mm",out.height="0.8\\textheight",echo=FALSE>>=
 cptl <-
@@
 
-print(ggplot(cptl,aes(x=human_name,y=FPKM))+
+print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
       geom_boxplot(outlier.shape=NA)+
@@
                                                                      angle=0,
-                                                                     size=9),
+                                                                     size=7),
                                             title.position="top"
@@
       theme_bw() +
-      theme(legend.position=c(0.7,0.15),
-            axis.text.y=element_text(face="italic"),
+      theme(text=element_text(size=9),
+            legend.position=c(0.99,0.41),
+            axis.text.y=element_text(face="italic",size=8),
             legend.margin=unit(0,"mm"),
-            legend.justification="left",
+            legend.justification=c(1,1),
             plot.margin=unit(c(0,0,0,0),"mm")
@@
 \caption{Core Placenta Transcriptome.
-  \Sexpr{core.placenta.transcriptome.long[,length(unique(human_name))]}
+  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
   1:1 \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
-  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
+  placenta of all 13 Eutherians
   studied with housekeeping genes
@@
 \begin{figure}
-<<core.placenta.transcriptome.distribution,fig.width=5,fig.height=5,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
+<<core.placenta.transcriptome.distribution,fig.width=9/2.5,fig.height=9/2.5,out.width="90mm",out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
 num.species <- star.logs.wide[,length(unique(species))]
@@
              "all_but_three_expression",
-             "all_but_four_expression"),
+             "all_but_four_expression",
+             "all_but_metatherian_expression"
+             ),
            function(x){
@@
 fix.expression.names <- function(x){
-    x <- gsub("^all_expression$",
-              paste0(num.species," of ",num.species),
+    x <- gsub("^all_but_metatherian_expression$",
+              paste0("All ",num.species-1," Eutherians"),
               x
               )
-    nc.levels <- paste0(num.species," of ",num.species)
+    x <- gsub("^all_expression$",
+              paste0(num.species," of ",num.species," Species"),
+              x
+              )
+    nc.levels <- c(paste0("All ",num.species-1," Eutherians"),
+                   paste0(num.species," of ",num.species," Species")
+                   )
     ## number coutner
@@
         x <- gsub(paste0("^all_but_",a,"_expression$"),
-              paste0(num.species-nc[a]," of ",num.species),
+              paste0(num.species-nc[a]," of ",num.species," Species"),
               x
@@
             c(nc.levels,
-              paste0(num.species-nc[a]," of ",num.species))
+              paste0(num.species-nc[a]," of ",num.species," Species"))
     }
@@
                                 fix.expression.names(`Min. Exp. in`)]
-
 setnames(cpt.distribution.shape,
@@
       + xlab("Minimum Expression cutoff (FPKM)")
-      + ylab("Number of Genes Meeting FPKM Cutoff in N species")
+      + ylab("Number of Genes Meeting FPKM Cutoff")
       + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
@@
       + geom_vline(xintercept=10,linetype="dotted")
-      + geom_segment(mapping=aes(y=V1,
+      + geom_segment(mapping=aes(y=V1-ifelse(V1<80,20,0),
                          yend=V1,
                          xend=10,
-                         x=300,
+                         x=sapply(V1,
+                                  function(x){
+                                      cpt.distribution.shape[num_genes > x,
+                                                             max(`Expression (FPKM)`)*4]
+                                  }),
                          color=`Min. Exp. in`),
@@
                      )
-      + geom_text(mapping=aes(x=300,
-                      y=V1,
+      + geom_text(mapping=aes(x=sapply(V1,
+                                       function(x){
+                                           cpt.distribution.shape[num_genes > x,
+                                                                  max(`Expression (FPKM)`)*4.1]
+                                       }),
+                      y=V1-ifelse(V1<80,20,0),
                       label=paste0(V1," genes"),override=TRUE),
@@
                   show.legend=FALSE,
-                  hjust=0
+                  hjust=0,
+                  size=8*(5/14)
                   )
@@
                      data=cpt.distribution.shape[`Expression (FPKM)` > 1 &
-                                                 `Min. Exp. in` == '14 of 14',
+                                                 `Min. Exp. in` == 'All 13 Eutherians',
                                                  max(num_genes),`Min. Exp. in`],
@@
                   data=cpt.distribution.shape[`Expression (FPKM)` > 1 &
-                                              `Min. Exp. in` == '14 of 14',
+                                              `Min. Exp. in` == 'All 13 Eutherians',
                                               max(num_genes),`Min. Exp. in`],
                   show.legend=FALSE,
-                  hjust=1
+                  hjust=1,
+                  size=8*(5/14)
                   )
-      + guides(color=guide_legend(title=paste0("Min. Expression in\nn of ",num.species," Species"),
+      + guides(color=guide_legend(title=paste0("Minimum\nExpression in"),
                                   keywidth=unit(3,"mm"),
@@
                                   title.hjust=0.5,
-                                  label.position="right"
+                                  label.position="right",
+                                  label.theme=element_text(size=7,angle=0)
                                   ))
       + theme_bw()
-      + theme(legend.position=c(0.98,0.98),
+      + theme(text=element_text(size=9),
+              legend.position=c(0.99,0.99),
               legend.margin=unit(0,"mm"),
@@
   \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
-    & `Min. Exp. in` == '14 of 14',max(num_genes)]}
-  %
-  genes would be included if the FPKM of a set of 1:1 orthologs of all 14 species must be at
-  least 10, %
+    & `Min. Exp. in` == 'All 13 Eutherians',max(num_genes)]}
+  %
+  genes would be included if the expression of a set of 1:1 orthologs of all 13 Eutherians must be at
+  least 10 FPKM, %
   \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
-    & `Min. Exp. in` == '13 of 14',max(num_genes)]}
+    & `Min. Exp. in` == '13 of 14 Species',max(num_genes)]}
   %
   1:1 orthologs would be included if 13 of 14 species have an
-   with an FPKM of at least 10, and 
+   expression of at least 10 FPKM, and 
   \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
-    & `Min. Exp. in` == '10 of 14',max(num_genes)]}
+    & `Min. Exp. in` == '10 of 14 Species',max(num_genes)]}
   genes would be included if 10 of 14 species
@@
 \Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
-  == '14 of 14',max(num_genes)]} %
+  == '14 of 14 Species',max(num_genes)]} %
 genes in the core transcriptome. %

### 2016-11-29 manuscript/manuscript.Rnw
cite the NWM grant

@@
   Development (NICHD), National Institutes of Health, U.S. Department
-  of Health and Human Services (N01-HD-3342), and the National Science
-  Foundation BCS-0827546.
+  of Health and Human Services (N01-HD-3342), the National Science
+  Foundation (BCS-0827546 and BCS-0751508), and a seed grant from the
+  Institute for Genomic Biology at the University of Illinois,
+  Urbana-Champaign.
 
 \DEW{Need to cite NWM grants, IGB, etc.}
+
+  \DLA{Unclear what the IGB grant is; Derek: which grant did you want
+    me to cite? Mentioned seed grant.}
 

### 2016-11-29 manuscript/manuscript.Rnw
fix wrong oxford comma

@@
 mammals, it varies widely in terms of the shape of attachment, degree
-of intimacy between fetal tissues and maternal tissues and degree of
+of intimacy between fetal tissues and maternal tissues, and degree of
 interdigitation of maternal and fetal tissues in different species of

### 2016-11-29 manuscript/manuscript.Rnw
use bf instead of emph for keywords

@@
 These observations lead to two related questions: 1) what are the
-genes which are responsible for the \emph{core functionality of the
-  placenta} and 2) which genes are associated with \emph{changes in
+genes which are responsible for the \textbf{core functionality of the
+  placenta} and 2) which genes are associated with \textbf{changes in
   placenta morphology}?

### 2016-11-29 manuscript/manuscript.Rnw
fix invasion and pre-eclampsia references

@@
 \cref{fig:kiss1_evolution_tree}) also shows increased expression in
-pre-eclamptic placentas
-\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}, and is
-involved in modulating trophoblast invasion. Two other genes,
+pre-eclamptic placentas, and is
+involved in modulating trophoblast invasion~%
+\cite{Vazquez-Alaniz.ea2011:Comparativeexpressionprofilesfor}. Two other genes,
 \ac{PAPPA} (\Sexpr{hum.all.lin.spec["PAPPA",`FDR`]},
@@
 \cref{fig:adam12_evolution_tree}) are also differentially expressed
-but are negatively correlated with pre-eclampsia
+but are negatively correlated with pre-eclampsia~%
 \cite{Spencer.ea2008:ADAM12sinmaternalserum}.

### 2016-11-30 manuscript/manuscript.Rnw
mention classification when they are discussed
add another digit to output

@@
 })
-options(scipen = -2, digits = 0)
+options(scipen = -2, digits = 1)
 library("data.table")
@@
 one species, and \ac{FPKM} greater than 1 at least two species.
+Supplemental \cref{tab:placental_classification} lists the shape,
+intimacy, and interdigitiation of each species analyzed.
 
@@
 \PP{Introduce the characteristics earlier in the introduction}
+
+\DLA{They're in the first paragraph of the introduction, but I've also
+  referred to them now in the first subsection here.}
 

### 2016-11-30 manuscript/manuscript.Rnw
include the FDR for the gene enrichments

@@
     }
+}
+
+@
+
+<<placenta_go_results,echo=FALSE,results="hide">>=
+placgo <- function(goid,group="mf") {
+    if (group=="mf") {
+        temp <- placenta_go_results_mf
+    } else {
+        temp <- placenta_go_results_bp
+    }
+    paste0("$\\mathrm{FDR}=",
+           number_to_latex(temp[GO.ID==goid,`FDR`]),
+           "$"
+           )
+}
+@
+
+<<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=
+pathway <- list()
+for (database in c("kegg","reactome","biocarta","netpath")) {
+    load(paste0("../data/placenta_gsea_",database))
+    eval(parse(text=paste0("gsea.table <- placenta_gsea_",database)))
+    gsea.table[,Database:=gsub("^(.)","\\U\\1",database,perl=TRUE)]
+    pathway[[database]] <-
+        gsea.table
+}
+pathway.analysis <-
+    rbindlist(pathway)[fdr < 0.1,]
+rm(pathway)
+pathway.analysis[,Pathway:=gsub("^(KEGG|BIOCARTA|REACTOME) ","",
+                                gsub("_+"," ",
+                                     Pathway))]
+pathway.analysis[,Pathway:=gsub(".+\\%","",Pathway)]
+
+setnames(pathway.analysis,"p_values","p")
+setorder(pathway.analysis,"p")
+
+placpath <- function(pathway,database) {
+    paste0("$\\mathrm{FDR}=",
+           number_to_latex(pathway.analysis[grepl(pathway,Pathway) & Database==Database,
+                                            fdr]),
+           "$")
 }
@@
 \ac{ANXA5}, which was originally identified as a an anti-coagulant
-protein, but is also involved in syncitium formation, membrane repair,
+protein, but is also involved in syncytium formation, membrane repair,
 and reduction of spontaneous
@@
 their homologs) are not highly expressed in \dog, \bonobo, or \sheep,
-nor was \ac{MFSD2A}, the Syncytin-2 receptor (\ac{FPKM} $\ge 10$
+nor was the Syncytin-2 receptor, \ac{MFSD2A} (\ac{FPKM} $\ge 10$
 only in \human, \mouse, and \dog).
@@
 which are involved in multiple biological pathways, including multiple
-pathways related to translation (ribosome, translation, signal
-recognition particle targeting to the rough endoplasamic reticulum,
+pathways related to translation (ribosome %
+%
+(\Sexpr{placpath("^RIBOSOME","Kegg")})
+, translation %
+(\Sexpr{placpath("^TRANSLATION","Reactome")})
+%
+, signal
+recognition particle targeting to the rough endoplasmic reticulum %
+(\Sexpr{placpath("^SRP DEPENDENT","Reactome")})
+%
+,
 etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
@@
 molecular pathways are related to the migration of the trophoblast
-into the endometrium, including the EGFR1 pathway and the Actin Y
-pathway. Two other significant pathways, the mPR/Progesterone pathway
+into the endometrium, including the EGFR1 pathway %
+(\Sexpr{placpath("^EGFR1","Netpath")})
+%
+and the Actin Y pathway %
+(\Sexpr{placpath("^ACTINY","Biocarta")})
+%
+. Two other significant pathways, the mPR/Progesterone pathway
 and Rho cell motility pathways, feed into the Actin Y pathway as well.
@@
 
-\DEW{Write out FDR for enrichment described}
-
-The high expression of components of the EGFR signaling pathway in the
-core placenta transcriptome (including \ac{GRB2}, \ac{RALB})
-as well as annexins which are involved in the location of EGF receptor
-localization (\ac{ANXA1} and \ac{ANXA2}) in the core placenta
-is consistent with previous findings showing that alterations to the
-EGF signaling system change invasion of trophoblast cells into the
-spiral arteries~\cite{Armant.ea2015:Reducedexpressionofepidermal}.
-Another component of the core placenta transcriptome, \ac{CD9},
-has been shown to suppress \human extravillous trophoblast
+% \DEW{Write out FDR for enrichment described}
+
+The high expression of components of the EGFR signaling pathway %
+(Enrichment \Sexpr{placpath("^EGFR1","Netpath")}) %
+in the
+core placenta transcriptome (including \ac{GRB2}, \ac{RALB}) as well
+as annexins which are involved in the spatio-temporal regulation of
+EGF receptor signal transduction (\ac{ANXA1}, \ac{ANXA2}, and
+\ac{S100A11} \cite{Poeter.ea2013:DisruptionofannexinA1S100A11}) in the
+core placenta is consistent with previous findings showing that
+alterations to the EGF signaling system change invasion of trophoblast
+cells into the spiral
+arteries~\cite{Armant.ea2015:Reducedexpressionofepidermal}. Another
+component of the core placenta transcriptome, \ac{CD9}, has been shown
+to suppress \human extravillous trophoblast
 invasion~\cite{Matsumoto.ea2016:CD9suppresseshumanextravillous}, and
@@
 
-\PP{EGF interacts with S100 from the core placenta as well}
+% \PP{EGF interacts with S100 from the core placenta as well}
 

### 2016-11-30 manuscript/manuscript.Rnw
clarify that the metatherians are excluded

@@
 identify the non-housekeeping genes which are expressed in the
-placentas of all species analyzed and therefore represent the
+placentas of all eutherian species analyzed and therefore represent the
 \emph{core functionality of the placenta}; 2) identify genes whose
@@
 the placenta transcriptome if a 1:1 ortholog of that gene was
-expressed with $\text{FPKM} \ge 10$ in all species studied and was not
-a human housekeeping gene according to
+expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
+(excluding the metatherian, \opossum) the and was not a human housekeeping
+gene according to
 \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which

### 2016-11-30 manuscript/manuscript.Rnw
refer to how the lineage spec. analyses were done

@@
 \subsection{Comparative Analysis of Gene Expression}
+\label{sec:comp-analys-gene}
 
@@
 
-Using a mixed-effects linear model to account for repeated testing of
-\human and \mouse, we identified multiple genes which are
+Using a type-III ANOVA to account for repeated testing of
+\human and \mouse (see \cref{sec:comp-analys-gene} for details), we identified multiple genes which are
 differentially expressed in the \human lineage in comparison to all

### 2016-11-30 manuscript/manuscript.Rnw
fix proteoglycan typo
discuss interdigitation using ancestral state

@@
 HSPC300. Other components of cell-cell adhesion, including \ac{F11R}
-which forms tight junctions in epithelia, the proteglycans \ac{SDC4} and
+which forms tight junctions in epithelia, the proteoglycans \ac{SDC4} and
 \ac{LUM}, and \ac{LAMC1}.
@@
 \cref{tab:intimacy_genes_polr,tab:interdigitation_genes_polr}). We
-suspect this is due to the small number of intermediate
-interdigitation and intimacy placentas that are available. Future
-analyses of additional placentas with intermediate interdigitation and
-intimacy should better illuminate the pathways which are responsible
-for these placental morphologies.
+suspect that this is due to the fact that the ancestral state of all
+eutherians is a labyrinthine, hemochorial
+placenta~\cite{Wildman.ea2006:Evolutionofmammalianplacenta}. This
+implies that the transcriptome of intermediately invasive or
+interdigitated placentas need not be intermediate between the
+transcriptomes of less intimate (folded/lamellar) or interdigitated
+placentas (epitheliochorial) and most intimate (hemochorial) or
+interdigitated placentas (labyrinthine) because these morphologies
+have evolved independently. An alternative is that the limited number
+of species with placentas of intermediate interdigitation and intimacy
+in this study limits the ability to detect genes which whose
+expression correlates with degree of intimacy and interdigitation.
+Future analyses of additional placentas with intermediate
+interdigitation and intimacy should better illuminate the pathways
+which are responsible for these placental morphologies.
 

### 2016-11-30 manuscript/manuscript.Rnw
discuss fusion proteins more completely

@@
 
-While Syncytin-1 and -2 are implicated in the formation of the
-syncytiotrophoblast in
+While Syncytin-1 and -2 (\ac{ERVW-1} and \ac{ERVFRD-1}, respectively) are implicated
+in the formation of the syncytiotrophoblast in
 \human~\cite{Rawn.ea2008:evolutionregulationandfunction} and homologs
-of \ac{ERVFRD-1} and \ac{ERVW-1} are widely expressed in most
-of the placentas analyzed here, they are not expressed in \bonobo,
-\sheep, and \dog, even though these mammals form
+of \ac{ERVFRD-1} and \ac{ERVW-1} are widely expressed in most of the
+placentas analyzed here, they are not expressed in \bonobo, \sheep,
+and \dog, even though these mammals form
 syncytiotrophoblasts~\cite{Mess.ea2006:Evolutionarytransformationsoffetal}.
@@
 Laurasiatheria~\cite{Mess.ea2006:Evolutionarytransformationsoffetal},
+and syncytiotrophoblast must be continuously replenished
 the gene(s) responsible for cytotrophoblast fusion to form
-syncytiotrophoblast is likely to be highly expressed in the core
-placenta. \ac{ANXA5} is known to be expressed at multiple stages
-of placentation, and is capable of repairing membrane defects and
+syncytiotrophoblast should be highly expressed in the core
+placenta. \ac{ANXA5} is known to be expressed at multiple stages of
+placentation, and is capable of repairing membrane defects and
 promoting the fusion of
 cells~\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing} in
-the presence of Ca$^{2+}$. While there are likely other genes which
+the presence of Ca$^{2+}$. While there may be other genes which
 are responsible for the positioning of cytotrophoblast cells in close
-proximity to start the fusion process, we hypothesize that \ac{ANXA5} is
-essential for the completion of fusion and the maintenance of membrane
-repair during the massive flux of nutrients and waste in the placenta.
+proximity to start the fusion process, \ac{ANXA5} is likely to
+be essential for the completion of fusion and the maintenance of
+membrane repair during the massive flux of nutrients and waste in the
+placenta \cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing,Ueki.ea2012:Lossofmaternalannexin}.
 
 \PP{Add refs for fusion}
+
+\DLA{Added the Carmeille.ea2015 and Ueki.ea2012 I have other
+  references
+  (\cite{Ota.ea2013:ContributionoffetalANXA5,Bouter.ea2015:ReviewAnnexin-A5andcell})
+  but I'd exceed the limit for placenta (45)}
 

### 2016-11-30 manuscript/supplemental_material.Rnw
strip out U from plac names; use perl to UC them

--- 2016-11-18 manuscript/supplemental_material.Rnw
@@
 <<placental_classification_table,echo=FALSE,results="asis">>=
-print(xtable(placenta_classification,
+plac.class <- copy(placenta_classification[,list(species,shape,interdigitation,intimacy)])
+plac.class[,`species`:=paste0("\\textit{",gsub("^(.)","\\U\\1",`species`,perl=TRUE),"}")]
+plac.class[,`interdigitation`:=gsub("^(.)","\\U\\1",`interdigitation`,perl=TRUE)]
+plac.class[,`shape`:=gsub("^(.)","\\U\\1",`shape`,perl=TRUE)]
+plac.class[,`intimacy`:=gsub("^(.)","\\U\\1",`intimacy`,perl=TRUE)]
+setnames(plac.class,colnames(plac.class),gsub("^(.)","\\U\\1",colnames(plac.class),perl=TRUE))
+
+print(xtable(plac.class,
              display=c("s","s","s","s","s"),

### 2016-11-30 manuscript/supplemental_material.Rnw
make the figure more squat

@@
 \begin{figure}[tbh]
-<<core.placenta.transcriptome.housekeeping,fig.width=5,fig.height=10,out.height="0.8\\textheight",echo=FALSE>>=
+<<core.placenta.transcriptome.housekeeping,fig.width=6,fig.height=8,out.height="0.8\\textheight",echo=FALSE>>=
 cptlh <-
@@
 
-print(ggplot(cptlh,aes(x=human_name,y=FPKM))+
+print(ggplot(cptlh[FPKM > 0.9,],aes(x=human_name,y=FPKM))+
       geom_boxplot(outlier.shape=NA)+

### 2016-11-30 manuscript/supplemental_material.Rnw
use the housekeeping shape code from the manuscript

@@
 \begin{figure}
-<<housekeeping.placenta.distribution,fig.width=8,fig.height=8,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
+<<housekeeping.placenta.distribution,fig.width=5,fig.height=5,out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
 ## housekeeping set as apposed to non-housekeeping set
@@
 fix.expression.names <- function(x){
+    x <- gsub("^all_but_metatherian_expression$",
+              paste0("All ",num.species-1," Eutherians"),
+              x
+              )
     x <- gsub("^all_expression$",
-              paste0(num.species," of ",num.species," species"),
+              paste0(num.species," of ",num.species," Species"),
               x
               )
-    nc.levels <- paste0(num.species," of ",num.species," species")
+    nc.levels <- c(paste0("All ",num.species-1," Eutherians"),
+                   paste0(num.species," of ",num.species," Species")
+                   )
     ## number coutner
@@
         x <- gsub(paste0("^all_but_",a,"_expression$"),
-              paste0(num.species-nc[a]," of ",num.species," species"),
+              paste0(num.species-nc[a]," of ",num.species," Species"),
               x
@@
             c(nc.levels,
-              paste0(num.species-nc[a]," of ",num.species," species"))
+              paste0(num.species-nc[a]," of ",num.species," Species"))
     }
@@
 cpt.h.distribution.shape <- 
-    cpt.h.distribution.shape[,`Minimum Expression in`:=
+    cpt.h.distribution.shape[,`Min. Exp. in`:=
                                 fix.expression.names(`Minimum Expression in`)]
@@
 print(ggplot(cpt.h.distribution.shape[`Expression (FPKM)` > 0.1,],
-             aes(x=`Expression (FPKM)`,color=`Minimum Expression in`,y=num_genes))
+             aes(x=`Expression (FPKM)`,color=`Min. Exp. in`,y=num_genes))
       + geom_line(lwd=1)
       + xlab("Minimum Expression cutoff (FPKM)")
-      + ylab("Genes with N-th Minimum Expression")
+      + ylab("Number of Genes Meeting FPKM Cutoff")
       + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
@@
                       labels = trans_format("log10", math_format(10^.x)))
-      + geom_vline(xintercept=10)
-      + geom_segment(mapping=aes(y=V1,
+      + geom_vline(xintercept=10,linetype="dotted")
+      + geom_segment(mapping=aes(y=V1-ifelse(V1<145,40,0),
                          yend=V1,
                          xend=10,
-                         x=250,
-                         color=`Minimum Expression in`),
+                         x=sapply(V1,
+                                  function(x){
+                                      cpt.h.distribution.shape[num_genes > x,
+                                                             max(`Expression (FPKM)`)*2.5]
+                                  }),
+                         color=`Min. Exp. in`),
                      data=cpt.h.distribution.shape[`Expression (FPKM)` > 10,
-                                              max(num_genes),`Minimum Expression in`],
-                     arrow=arrow(length=unit(0.5,"cm"))
-                   )
-      + geom_text(mapping=aes(x=250,
-                      y=V1,
+                                              max(num_genes),`Min. Exp. in`],
+                     arrow=arrow(length=unit(0.5,"cm")),
+                     show.legend=FALSE
+                     )
+      + geom_text(mapping=aes(x=sapply(V1,
+                                       function(x){
+                                           cpt.h.distribution.shape[num_genes > x,
+                                                                  max(`Expression (FPKM)`)*2.6]
+                                       }),
+                      y=V1-ifelse(V1<145,40,0),
                       label=paste0(V1," genes"),override=TRUE),
                   data=cpt.h.distribution.shape[`Expression (FPKM)` > 10,
-                      max(num_genes),`Minimum Expression in`],
+                      max(num_genes),`Min. Exp. in`],
                   show.legend=FALSE,
-                  hjust=0
-                  )
-      + guides(color=guide_legend(title="Minimum\nExpression in"))
+                  hjust=0,
+                  size=8*(5/14)
+                  )
+      + geom_segment(mapping=aes(yend=V1,
+                         y=100,
+                         xend=1,
+                         x=1,
+                         color=`Min. Exp. in`),
+                     data=cpt.h.distribution.shape[`Expression (FPKM)` > 1 &
+                                                 `Min. Exp. in` == 'All 13 Eutherians',
+                                                 max(num_genes),`Min. Exp. in`],
+                     arrow=arrow(length=unit(0.5,"cm")),
+                     show.legend=FALSE
+                     )
+      + geom_text(mapping=aes(x=1,
+                      y=100,
+                      label=paste0(V1," genes"),
+                      override=TRUE),
+                  angle=90,
+                  data=cpt.h.distribution.shape[`Expression (FPKM)` > 1 &
+                                              `Min. Exp. in` == 'All 13 Eutherians',
+                                              max(num_genes),`Min. Exp. in`],
+                  show.legend=FALSE,
+                  hjust=1,
+                  size=8*(5/14)
+                  )
+      + guides(color=guide_legend(title=paste0("Minimum\nExpression in"),
+                                  keywidth=unit(3,"mm"),
+                                  keyheight=unit(3,"mm"),
+                                  title.hjust=0.5,
+                                  label.position="right",
+                                  label.theme=element_text(size=7,angle=0)
+                                  ))
+      + theme_bw()
+      + theme(text=element_text(size=9),
+              legend.position=c(0.99,0.99),
+              legend.margin=unit(0,"mm"),
+              legend.justification=c(1,1),
+              plot.margin=unit(c(0,0,0,0),"mm")
+              )
       )
@@
   placenta transcriptomes if %
-  \Sexpr{cpt.h.distribution.shape[,levels(`Minimum Expression in`)[1]]}
+  \Sexpr{cpt.h.distribution.shape[,levels(`Min. Exp. in`)[1]]}
   %
   (or %
-  \Sexpr{cpt.h.distribution.shape[,levels(`Minimum Expression in`)[2]]},
+  \Sexpr{cpt.h.distribution.shape[,levels(`Min. Exp. in`)[2]]},
   % 
@@
   \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
-    & `Minimum Expression in` == '14 of 14 species',max(num_genes)]}
+    & `Min. Exp. in` == '14 of 14 Species',max(num_genes)]}
   %
-  genes would be included if the FPKM of a set of 1:1 orthologs of all 14 species must be at
-  least 10, %
+  genes would be included if the expression of a set of 1:1 orthologs of all 14 species must be at
+  least 10 FPKM, %
   \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
-    & `Minimum Expression in` == "13 of 14 species",max(num_genes)]}
+    & `Min. Exp. in` == "13 of 14 Species",max(num_genes)]}
   %
   1:1 orthologs would be included if 13 of 14 species have an
-   with an FPKM of at least 10, and 
+   expression of at least 10 FPKM, and 
   \Sexpr{cpt.h.distribution.shape[`Expression (FPKM)` > 10 
-    & `Minimum Expression in` == '10 of 14 species',max(num_genes)]}
+    & `Min. Exp. in` == '10 of 14 Species',max(num_genes)]}
   genes would be included if 10 of 14 species
-  must have an FPKM of at least 10.}
+  must have an expression of at least 10 FPKM.}
 \label{fig:core_housekeeping_transcriptome_shape}

### 2016-11-30 common_r_code/utility_functions.R
add number_to_latex

@@ -86,6 +86,16 @@ species <- function(s,join=TRUE) {
     }
 }
 
+number_to_latex <- function(x) {
+    if (is.numeric(x)) {
+        x <- knitr:::format_sci(x, 'latex')
+        i <- grep('[^-0-9.,]', x)
+        x[i] <- sprintf('\\ensuremath{%s}', x[i])
+        x <- sprintf('\\text{%s}', x)
+    }
+    x
+}
+
 english.join <- function(s){
     ret <- c()
     ## everything but the last gets a , appended if there are more than two

### 2016-11-30 manuscript/manuscript.Rnw
use shape instead of attachment

@@
 
+\listoffixmes
+
 \section{Introduction}
@@
 Even though the placenta's primary function is conserved among
-mammals, it varies widely in terms of the shape of attachment, degree
+mammals, it varies widely in terms of shape, degree
 of intimacy between fetal tissues and maternal tissues, and degree of
@@
 using Diamond \cite{Buchfink.ea2015:Fastandsensitiveprotein} against
-either the \human (for \smonkey) or \mouse (for both Spalacines) transcriptome (Supplemental
+either the \human (\smonkey) or \mouse (\galili and \carmeli) transcriptome (Supplemental
 \cref{tab:ref_genome_aligned_species}), and the
@@
 Supplemental \cref{tab:placental_classification} lists the shape,
-intimacy, and interdigitiation of each species analyzed.
-
-\subsubsection{Placenta Attachment Shape}
+intimacy, and interdigitation of each species analyzed.
+
+\subsubsection{Placenta Shape}
 
 Supplemental \cref{tab:shape_genes} shows the 50 genes which are most
-significantly associated with the shape of placenta attachment, as
+significantly associated with the shape of the placenta, as
 well as the shape in which the difference is most significant. The
 most significant genes which were associated with diffuse placenta
-attachment include \ac{EREG}, a ligand of EGFR, \ac{UNC93A}, a
+shape include \ac{EREG}, a ligand of EGFR, \ac{UNC93A}, a
 homolog of \ac{UNC93B1} which is involved in the balancing between
@@
 For the degree of intimacy and interdigitation, we saw multiple genes
-which were associated with different levels of interdigitiation and
+which were associated with different levels of interdigitation and
 intimacy (Supplemental

### 2016-11-30 manuscript/manuscript.Rnw
clean up the abstract

@@
 \begin{abstract}
-  \textbf{Introduction} The placenta is arguably the most
-  anatomically variable organ in mammals.
+  \textbf{Introduction} The placenta is arguably the most anatomically
+  variable organ in mammals even though its primary function is
+  conserved.
 
@@
   \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
-  core genes which are expressed (FPKM $\ge 10$) in all placentas, %
+  core genes which are expressed (FPKM $\ge 10$) in all eutherian placentas, %
   %
@@
   \Sexpr{star.logs[,length(unique(species))-1]} species. %
+  %
+  Multiple genes are significantly associated with placenta
+  morphology, including \ac{EREG} and \ac{WNT5A} which are both
+  associated with placental shape.
 
   \textbf{Discussion} %
-  Using the strictest filtering strategy we find a core set of only
   \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
-  genes is essential for the proper functioning of the placenta in all
-  species analyzed. We also found significantly higher expression of
-  the pre-eclampsia associated genes \ac{KISS1}, \ac{CRH},
-  \ac{PAPPA}, and \ac{ADAM12} on the \human lineage.
-
-  \DEW{The discussion and results seem redundant. Given the title
-    shouldn't something about shape be mentioned?}
+  genes are important for the core functions of the placenta in all
+  eutherian species analyzed. The molecular functions and pathways
+  enriched in the core placenta align with the evolutionarily
+  conserved functionality of the placenta.
+
+ % \DEW{The discussion and results seem redundant. Given the title
+ %   shouldn't something about shape be mentioned?}
 

### 2016-11-30 manuscript/manuscript.Rnw
clarify why annexins should be studied

@@
 complexes may illuminate how immune tolerance is maintained during
-placentation.
-
-\DEW{Why?}
+placentation because these appear to the the major class of immune
+modulatory transcripts which are expressed in all placentas analyzed.
+
+\DEW{Why?} \DLA{Clarified why I think this is important to study.}
 

### 2016-11-30 manuscript/acronyms.tex
use short/single/long options in acro

--- 2016-11-14 manuscript/acronyms.tex
@@
 \DeclareAcronym{SRA}{long=Sequence Read Archive}
-\DeclareAcronym{PRG2}{short=\textit{PRG2},long=Proteoglycan 2, bone marrow}
-\DeclareAcronym{FPKM}{long=Fragments Per Kilobase of exon per Million reads}
-\DeclareAcronym{CSH1}{short=\textit{CSH1},long=chorionic somatomammotropin hormone 1}
-\DeclareAcronym{CSH2}{short=\textit{CSH2},long=chorionic somatomammotropin hormone 2}
-\DeclareAcronym{CSHL1}{short=\textit{CSHL1},long=chorionic somatomammotropin hormone-like 1}
-\DeclareAcronym{GH2}{short=\textit{GH2},long=growth hormone 2}
 \DeclareAcronym{CEA}{long=carcinoembryonic antigen}
-\DeclareAcronym{HSD3B1}{short=\textit{HSD3B1},long=3-β Hydroxysteroid dehydrogenase 1}
 \DeclareAcronym{GSE}{long=Gene Set Enrichment}
@@
 \DeclareAcronym{FDR}{long=False Discovery Rate}
+\DeclareAcronym{gsea}{short=GSEA,long=gene set enrichment analysis}
+\DeclareAcronym{FPKM}{long=Fragments Per Kilobase of exon per Million reads}
+\DeclareAcronym{HSD3B1}{short=\textit{HSD3B1},long={hydroxy-δ-5-steroid dehydrogenase, 3 β- and steroid δ-isomerase 1},first-style=reversed,single={\textit{HSD3B1} (hydroxy-δ-5-steroid dehydrogenase, 3 β- and steroid δ-isomerase 1)}}
+\DeclareAcronym{CSH1}{short=\textit{CSH1},long={chorionic somatomammotropin hormone 1},first-style=reversed,single={\textit{CSH1} (chorionic somatomammotropin hormone 1)}}
+\DeclareAcronym{CSH2}{short=\textit{CSH2},long={chorionic somatomammotropin hormone 2},first-style=reversed,single={\textit{CSH2} (chorionic somatomammotropin hormone 2)}}
+\DeclareAcronym{CSHL1}{short=\textit{CSHL1},long={chorionic somatomammotropin hormone like 1},first-style=reversed,single={\textit{CSHL1} (chorionic somatomammotropin hormone like 1)}}
+\DeclareAcronym{GH2}{short=\textit{GH2},long={growth hormone 2},first-style=reversed,single={\textit{GH2} (growth hormone 2)}}
+\DeclareAcronym{PRG2}{short=\textit{PRG2},long={proteoglycan 2, pro eosinophil major basic protein},first-style=reversed,single={\textit{PRG2} (proteoglycan 2, pro eosinophil major basic protein)}}
+\DeclareAcronym{CRH}{short=\textit{CRH},long={corticotropin releasing hormone},first-style=reversed,single={\textit{CRH} (corticotropin releasing hormone)}}
+\DeclareAcronym{KISS1}{short=\textit{KISS1},long={KiSS-1 metastasis-suppressor},first-style=reversed,single={\textit{KISS1} (KiSS-1 metastasis-suppressor)}}
+\DeclareAcronym{PAPPA}{short=\textit{PAPPA},long={pappalysin 1},first-style=reversed,single={\textit{PAPPA} (pappalysin 1)}}
+\DeclareAcronym{ADAM12}{short=\textit{ADAM12},long={ADAM metallopeptidase domain 12},first-style=reversed,single={\textit{ADAM12} (ADAM metallopeptidase domain 12)}}
+\DeclareAcronym{ANXA1}{short=\textit{ANXA1},long={annexin A1},first-style=reversed,single={\textit{ANXA1} (annexin A1)}}
+\DeclareAcronym{ANXA2}{short=\textit{ANXA2},long={annexin A2},first-style=reversed,single={\textit{ANXA2} (annexin A2)}}
+\DeclareAcronym{ANXA5}{short=\textit{ANXA5},long={annexin A5},first-style=reversed,single={\textit{ANXA5} (annexin A5)}}
+\DeclareAcronym{S100A11}{short=\textit{S100A11},long={S100 calcium binding protein A11},first-style=reversed,single={\textit{S100A11} (S100 calcium binding protein A11)}}
+\DeclareAcronym{S100A10}{short=\textit{S100A10},long={S100 calcium binding protein A10},first-style=reversed,single={\textit{S100A10} (S100 calcium binding protein A10)}}
+\DeclareAcronym{LGALS1}{short=\textit{LGALS1},long={galectin 1},first-style=reversed,single={\textit{LGALS1} (galectin 1)}}
+\DeclareAcronym{LAMC1}{short=\textit{LAMC1},long={laminin subunit γ 1},first-style=reversed,single={\textit{LAMC1} (laminin subunit γ 1)}}
+\DeclareAcronym{LUM}{short=\textit{LUM},long={lumican},first-style=reversed,single={\textit{LUM} (lumican)}}
+\DeclareAcronym{GRB2}{short=\textit{GRB2},long={growth factor receptor bound protein 2},first-style=reversed,single={\textit{GRB2} (growth factor receptor bound protein 2)}}
+\DeclareAcronym{RALB}{short=\textit{RALB},long={RAS like proto-oncogene B},first-style=reversed,single={\textit{RALB} (RAS like proto-oncogene B)}}
+\DeclareAcronym{ERVFRD-1}{short=\textit{ERVFRD-1},long={endogenous retrovirus group FRD member 1},first-style=reversed,single={\textit{ERVFRD-1} (endogenous retrovirus group FRD member 1)}}
+\DeclareAcronym{ERVW-1}{short=\textit{ERVW-1},long={endogenous retrovirus group W member 1},first-style=reversed,single={\textit{ERVW-1} (endogenous retrovirus group W member 1)}}
+\DeclareAcronym{MFSD2A}{short=\textit{MFSD2A},long={major facilitator superfamily domain containing 2A},first-style=reversed,single={\textit{MFSD2A} (major facilitator superfamily domain containing 2A)}}
+\DeclareAcronym{PRSS16}{short=\textit{PRSS16},long={protease, serine 16},first-style=reversed,single={\textit{PRSS16} (protease, serine 16)}}
+\DeclareAcronym{TIMP3}{short=\textit{TIMP3},long={TIMP metallopeptidase inhibitor 3},first-style=reversed,single={\textit{TIMP3} (TIMP metallopeptidase inhibitor 3)}}
+\DeclareAcronym{F11R}{short=\textit{F11R},long={F11 receptor},first-style=reversed,single={\textit{F11R} (F11 receptor)}}
+\DeclareAcronym{SDC4}{short=\textit{SDC4},long={syndecan 4},first-style=reversed,single={\textit{SDC4} (syndecan 4)}}
+\DeclareAcronym{FN1}{short=\textit{FN1},long={fibronectin 1},first-style=reversed,single={\textit{FN1} (fibronectin 1)}}
+\DeclareAcronym{ITGB1}{short=\textit{ITGB1},long={integrin subunit β 1},first-style=reversed,single={\textit{ITGB1} (integrin subunit β 1)}}
+\DeclareAcronym{ITGB1BP1}{short=\textit{ITGB1BP1},long={integrin subunit β 1 binding protein 1},first-style=reversed,single={\textit{ITGB1BP1} (integrin subunit β 1 binding protein 1)}}
+\DeclareAcronym{EREG}{short=\textit{EREG},long={epiregulin},first-style=reversed,single={\textit{EREG} (epiregulin)}}
+\DeclareAcronym{UNC93A}{short=\textit{UNC93A},long={unc-93 homolog A (C. elegans)},first-style=reversed,single={\textit{UNC93A} (unc-93 homolog A (C. elegans))}}
+\DeclareAcronym{UNC93B1}{short=\textit{UNC93B1},long={unc-93 homolog B1 (C. elegans)},first-style=reversed,single={\textit{UNC93B1} (unc-93 homolog B1 (C. elegans))}}
+\DeclareAcronym{WNT5A}{short=\textit{WNT5A},long={Wnt family member 5A},first-style=reversed,single={\textit{WNT5A} (Wnt family member 5A)}}
+\DeclareAcronym{IL1RL1}{short=\textit{IL1RL1},long={interleukin 1 receptor like 1},first-style=reversed,single={\textit{IL1RL1} (interleukin 1 receptor like 1)}}
+\DeclareAcronym{PLAC1}{short=\textit{PLAC1},long={placenta specific 1},first-style=reversed,single={\textit{PLAC1} (placenta specific 1)}}
+\DeclareAcronym{Afp}{short=\textit{Afp},long={α-fetoprotein},first-style=reversed,single={\textit{Afp} (α-fetoprotein)}}
+\DeclareAcronym{Gjb2}{short=\textit{Gjb2},long={gap junction protein β 2},first-style=reversed,single={\textit{Gjb2} (gap junction protein β 2)}}
+\DeclareAcronym{CD9}{short=\textit{CD9},long={CD9 molecule},first-style=reversed,single={\textit{CD9} (CD9 molecule)}}
 

### 2016-11-30 manuscript/gene_names.sh

@@ -2,5 +2,5 @@
 
 for gene in "$@"; do
     name="$(echo "$gene" | gene_to_aliases|awk -F"$(echo -ne "\t")" '{print $2}')"
-    echo "\\DeclareAcronym{$gene}{short=\\textit{$gene},long={$name}}";
+    echo "\\DeclareAcronym{$gene}{short=\\textit{$gene},long={$name},first-style=reversed,single={\\textit{$gene} ($name)}}";
 done;

### 2016-11-30 manuscript/acronyms.tex
fix acronyms

@@
-\DeclareAcronym{EV}{short=EV,long=Extracellular Vesicle}
-\DeclareAcronym{SVM}{long=Support Vector Machine}
-\DeclareAcronym{ROC}{long=Receiver Operating Characteristic}
-\DeclareAcronym{KNN}{long=K Nearest Neighbors}
-\DeclareAcronym{SRA}{long=Sequence Read Archive}
-\DeclareAcronym{CEA}{long=carcinoembryonic antigen}
-\DeclareAcronym{GSE}{long=Gene Set Enrichment}
-\DeclareAcronym{OU}{long=Ornstein-Uhlenbeck}
-\DeclareAcronym{OUCH}{long=\ac{OU} Comparative Hypotheses}
-\DeclareAcronym{TSI}{long=Tissue Specificity Index}
-\DeclareAcronym{SRA}{long=Sequence Read Archive}
-\DeclareAcronym{GH}{long=Growth Hormone}
-\DeclareAcronym{PSG}{long=Pregnancy-Specific Glycoprotein}
-\DeclareAcronym{PAG}{long=Pregnancy-Associated Glycoprotein}
-\DeclareAcronym{GO}{long=Gene Ontology}
-\DeclareAcronym{FDR}{long=False Discovery Rate}
-\DeclareAcronym{gsea}{short=GSEA,long=gene set enrichment analysis}
-\DeclareAcronym{FPKM}{long=Fragments Per Kilobase of exon per Million reads}
+\DeclareAcronym{EV}{short=EV,long={Extracellular Vesicle}}
+\DeclareAcronym{SVM}{short=SVM,long={Support Vector Machine}}
+\DeclareAcronym{ROC}{short=ROC,long={Receiver Operating Characteristic}}
+\DeclareAcronym{KNN}{short=KNN,long={K Nearest Neighbors}}
+\DeclareAcronym{SRA}{short=SRA,long={Sequence Read Archive}}
+\DeclareAcronym{CEA}{short=CEA,long={carcinoembryonic antigen}}
+\DeclareAcronym{GSE}{short=GSE,long={Gene Set Enrichment}}
+%\DeclareAcronym{OU}{short=OU,long=Ornstein-Uhlenbeck}
+%\DeclareAcronym{OUCH}{short=OUCH,long=\ac*{OU} Comparative Hypotheses}
+\DeclareAcronym{TSI}{short=TSI,long=Tissue Specificity Index}
+\DeclareAcronym{GH}{short=GH,long=Growth Hormone}
+\DeclareAcronym{PSG}{short=PSG,long={Pregnancy-Specific Glycoprotein}}
+\DeclareAcronym{PAG}{short=PAG,long={Pregnancy-Associated Glycoprotein}}
+\DeclareAcronym{GO}{short=GO,long={Gene Ontology}}
+\DeclareAcronym{FDR}{short=FDR,long={False Discovery Rate}}
+\DeclareAcronym{gsea}{short=GSEA,long={gene set enrichment analysis}}
+\DeclareAcronym{FPKM}{short=FPKM,long={Fragments Per Kilobase of exon per Million reads}}
 \DeclareAcronym{HSD3B1}{short=\textit{HSD3B1},long={hydroxy-δ-5-steroid dehydrogenase, 3 β- and steroid δ-isomerase 1},first-style=reversed,single={\textit{HSD3B1} (hydroxy-δ-5-steroid dehydrogenase, 3 β- and steroid δ-isomerase 1)}}

### 2016-12-01 presentation/placenta_viviparous_presentation.Rnw
fix axis italics after coord flip

@@
       + theme(legend.position="top",
-              axis.text.x = element_text(face="italic"))
+              axis.text.y = element_text(face="italic"))
       )

### 2016-12-01 presentation/placenta_viviparous_presentation.Rnw
use the new shapes and core without metatherians

@@
 \begin{figure}
-<<core.placenta.transcriptome,fig.width=8,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
-print(ggplot(core.placenta.transcriptome.long[!grepl("(min|median|max)_expression",Species),],
-             aes(y=human_name,x=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species))))
-      + geom_raster(aes(fill=log10(FPKM)))
-      + scale_x_discrete(expand=c(0,0))
-      + scale_y_discrete(expand=c(0,0))
-      + scale_fill_gradient(low = grad.def[["low"]],
-                            high = grad.def[["high"]],
-                            space = grad.def[["space"]],
-                            na.value = grad.def[["na.value"]],
-                            limits=c(0,4),
-                            guide=guide_colorbar(title=expression(log[10](FPKM)))
-                            )
-      + ylab("Human Ortholog Name")
-      + xlab("Species")
-      + theme_grey()
-      + theme(legend.position="top",
-              axis.ticks = element_blank(),
-              axis.text.y = element_text(face="italic"),
-              axis.text.x = element_text(angle = 340, hjust = 0.1, vjust=1,face="italic"))
-      )
+<<core.placenta.transcriptome,fig.width=7,fig.height=7*9/16,out.height="0.8\\textheight",echo=FALSE>>=
+cptl <-
+    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
+                                            Species) & housekeeping==FALSE,]
+cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
+                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
+                                          capfirst(names(species.names))))]
+cptl[,human_name:=factor(human_name,rev(levels(as.factor(human_name))))]
+
+print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
+      geom_boxplot(outlier.shape=NA)+
+      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
+      geom_point(aes(color=Species,shape=Species))+
+      scale_shape_manual("Species",
+                         guide=guide_legend(ncol=3,
+                                            keywidth=unit(3,"mm"),
+                                            keyheight=unit(3,"mm"),
+                                            label.theme=element_text(face="italic",
+                                                                     angle=0,
+                                                                     size=7),
+                                            title.position="top"
+                                            ),
+                         values=0:18)+
+      # scale_color_discrete(NULL)+
+      # coord_flip()+
+      xlab("Human Ortholog Name")+
+      ylab("Gene Expression (FPKM)")+
+      theme_bw() +
+      theme(text=element_text(size=9),
+            legend.position=c(0.99,0.99),
+            axis.text.x=element_text(face="italic",size=7,angle=90,hjust=1,vjust=0.5),
+            legend.margin=unit(0,"mm"),
+            legend.justification=c(1,1),
+            plot.margin=unit(c(0,0,0,0),"mm")
+            ))
 @ 
@@
              "all_but_three_expression",
-             "all_but_four_expression"),
+             "all_but_four_expression",
+             "all_but_metatherian_expression"
+             ),
            function(x){
-               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Minimum Expression in`:=x]
+               calculate.distribution.shape(cptg.nh[,x,with=FALSE][[1]])[,`Min. Exp. in`:=x]
            })
 fix.expression.names <- function(x){
+    x <- gsub("^all_but_metatherian_expression$",
+              paste0("All ",num.species-1," Eutherians"),
+              x
+              )
     x <- gsub("^all_expression$",
-              paste0(num.species," of ",num.species," species"),
+              paste0(num.species," of ",num.species," Species"),
               x
               )
-    nc.levels <- paste0(num.species," of ",num.species," species")
+    nc.levels <- c(paste0("All ",num.species-1," Eutherians"),
+                   paste0(num.species," of ",num.species," Species")
+                   )
     ## number coutner
@@
 cpt.distribution.shape <- 
-    cpt.distribution.shape[,`Minimum Expression in`:=
-                                fix.expression.names(`Minimum Expression in`)]
-
+    cpt.distribution.shape[,`Min. Exp. in`:=
+                                fix.expression.names(`Min. Exp. in`)]
 setnames(cpt.distribution.shape,
@@
 print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],
-             aes(x=`Expression (FPKM)`,color=`Minimum Expression in`,y=num_genes))
+             aes(x=`Expression (FPKM)`,color=`Min. Exp. in`,y=num_genes))
       + geom_line(lwd=1)
       + xlab("Minimum Expression cutoff (FPKM)")
-      + ylab("Genes with N-th Minimum Expression")
+      + ylab("Number of Genes Meeting FPKM Cutoff")
       + scale_y_log10(breaks = trans_breaks("log10", function(x) 10^x),
@@
                       labels = trans_format("log10", math_format(10^.x)))
-      + geom_vline(xintercept=10)
-      + geom_segment(mapping=aes(y=V1,
+      + geom_vline(xintercept=10,linetype="dotted")
+      + geom_segment(mapping=aes(y=V1-ifelse(V1<80,20,0),
                          yend=V1,
                          xend=10,
-                         x=1000,
-                         color=`Minimum Expression in`),
+                         x=sapply(V1,
+                                  function(x){
+                                      cpt.distribution.shape[num_genes > x,
+                                                             max(`Expression (FPKM)`)*4]
+                                  }),
+                         color=`Min. Exp. in`),
                      data=cpt.distribution.shape[`Expression (FPKM)` > 10,
-                                              max(num_genes),`Minimum Expression in`],
-                     arrow=arrow(length=unit(0.5,"cm"))
-                   )
-      + geom_text(mapping=aes(x=1000,
-                      y=V1,
+                                              max(num_genes),`Min. Exp. in`],
+                     arrow=arrow(length=unit(0.5,"cm")),
+                     show.legend=FALSE
+                     )
+      + geom_text(mapping=aes(x=sapply(V1,
+                                       function(x){
+                                           cpt.distribution.shape[num_genes > x,
+                                                                  max(`Expression (FPKM)`)*4.1]
+                                       }),
+                      y=V1-ifelse(V1<80,20,0),
                       label=paste0(V1," genes"),override=TRUE),
                   data=cpt.distribution.shape[`Expression (FPKM)` > 10,
-                      max(num_genes),`Minimum Expression in`],
+                      max(num_genes),`Min. Exp. in`],
                   show.legend=FALSE,
-                  hjust=0
-                  )
-      + guides(color=guide_legend(title="Minimum\nExpression in"))
+                  hjust=0,
+                  size=8*(5/14)
+                  )
+      + geom_segment(mapping=aes(yend=V1,
+                         y=100,
+                         xend=1,
+                         x=1,
+                         color=`Min. Exp. in`),
+                     data=cpt.distribution.shape[`Expression (FPKM)` > 1 &
+                                                 `Min. Exp. in` == 'All 13 Eutherians',
+                                                 max(num_genes),`Min. Exp. in`],
+                     arrow=arrow(length=unit(0.5,"cm")),
+                     show.legend=FALSE
+                     )
+      + geom_text(mapping=aes(x=1,
+                      y=100,
+                      label=paste0(V1," genes"),
+                      override=TRUE),
+                  angle=90,
+                  data=cpt.distribution.shape[`Expression (FPKM)` > 1 &
+                                              `Min. Exp. in` == 'All 13 Eutherians',
+                                              max(num_genes),`Min. Exp. in`],
+                  show.legend=FALSE,
+                  hjust=1,
+                  size=8*(5/14)
+                  )
+      + guides(color=guide_legend(title=paste0("Minimum\nExpression in"),
+                                  keywidth=unit(3,"mm"),
+                                  keyheight=unit(3,"mm"),
+                                  title.hjust=0.5,
+                                  label.position="right",
+                                  label.theme=element_text(size=7,angle=0)
+                                  ))
+      + theme_bw()
+      + theme(text=element_text(size=9),
+              legend.position=c(0.99,0.99),
+              legend.margin=unit(0,"mm"),
+              legend.justification=c(1,1),
+              plot.margin=unit(c(0,0,0,0),"mm")
+              )
       )

### 2016-12-05 manuscript/extended_supplemental_material.Rnw
fix acronyms
include CGB2 tree in extended supplement

@@
 % \usepackage{txfonts}
-\usepackage{acro}
+\usepackage[macros=false]{acro}
 \usepackage{array}
@@
 \begin{figure}
+<<cgb2.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=
+
+print(plot.ancestral.state.tree(gene_selector=quote(gene_short_name=="CGB2"
+                                                    & species=="homo sapiens"),
+                                genes.to.tree=genes.to.tree,
+                                combined.fpkm=combined.fpkm,
+                                trees=trees
+                                )+
+      geom_tiplab(size=6,color="black",fontface="italic"))
+@ 
+\caption{\genetreeexplanation{\textit{CGB2}}}
+\label{fig:cgb2_evolution_tree}
+\end{figure}
+
+\begin{figure}
 <<hoxa11.receptor.evolution.tree,echo=FALSE,fig.width=8,fig.height=8,warning=FALSE,message=FALSE>>=

### 2016-12-05 manuscript/manuscript.Rnw
fix acronyms
switch from homologs to similar to
 - We aren't claiming that the Syncytins are separated by a speciation
   event (especially as they are endogenous retroviruses)

@@
 %\usepackage{setspace}
-\usepackage[macros=false]{acro}
+\usepackage[macros=false,single=true,single-form=long]{acro}
 \usepackage{array}
@@
 (\ac{ERVFRD-1}) were not present in the core placenta, as they (or
-their homologs) are not highly expressed in \dog, \bonobo, or \sheep,
+genes similar to them) are not highly expressed in \dog, \bonobo, or \sheep,
 nor was the Syncytin-2 receptor, \ac{MFSD2A} (\ac{FPKM} $\ge 10$
@@
 shape include \ac{EREG}, a ligand of EGFR, \ac{UNC93A}, a
-homolog of \ac{UNC93B1} which is involved in the balancing between
+gene similar to \ac{UNC93B1} which is involved in the balancing between
 Tlr7 and Tlr9 immune response in
@@
 in the formation of the syncytiotrophoblast in
-\human~\cite{Rawn.ea2008:evolutionregulationandfunction} and homologs
-of \ac{ERVFRD-1} and \ac{ERVW-1} are widely expressed in most of the
+\human~\cite{Rawn.ea2008:evolutionregulationandfunction} and genes which are similar to 
+\ac{ERVFRD-1} and \ac{ERVW-1} are widely expressed in most of the
 placentas analyzed here, they are not expressed in \bonobo, \sheep,

### 2016-12-06 manuscript/manuscript.Rnw
add a second type III explanation for all genes
calculate core shape earlier

@@
 @ 
-\caption{\lineagetableexplanationtypeiii{\human}{all other analyzed species}}
+\caption{\lineagetableexplanationtypeiiishort{\human}{all other analyzed species}}
 \label{tab:human_lineage_specific}
@@
 
-\begin{figure}[tbh]
-<<core.placenta.transcriptome,fig.width=9/2.5,fig.height=8,out.width="90mm",out.height="0.8\\textheight",echo=FALSE>>=
-cptl <-
-    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
-                                            Species) & housekeeping==FALSE,]
-cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
-                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
-                                          capfirst(names(species.names))))]
-
-print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
-      geom_boxplot(outlier.shape=NA)+
-      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
-      geom_point(aes(color=Species,shape=Species))+
-      scale_shape_manual("Species",
-                         guide=guide_legend(#nrow=1,
-                                            keywidth=unit(3,"mm"),
-                                            keyheight=unit(3,"mm"),
-                                            label.theme=element_text(face="italic",
-                                                                     angle=0,
-                                                                     size=7),
-                                            title.position="top"
-                                            ),
-                         values=0:18)+
-      # scale_color_discrete(NULL)+
-      coord_flip()+
-      xlab("Human Ortholog Name")+
-      ylab("Gene Expression (FPKM)")+
-      theme_bw() +
-      theme(text=element_text(size=9),
-            legend.position=c(0.99,0.41),
-            axis.text.y=element_text(face="italic",size=8),
-            legend.margin=unit(0,"mm"),
-            legend.justification=c(1,1),
-            plot.margin=unit(c(0,0,0,0),"mm")
-            ))
-@ 
-\caption{Core Placenta Transcriptome.
-  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
-  1:1 \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
-  placenta of all 13 Eutherians
-  studied with housekeeping genes
-  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
-  ordered (from top to bottom) by median expression. Supplemental
-  \cref{fig:core_placenta_transcriptome_shape} shows the shape of the
-  distribution of the number of genes in the core placenta
-  transcriptome if different FPKM and species restriction criteria are
-  utilized.}
-\label{fig:core_placenta_transcriptome}
-\end{figure}
-
-\begin{figure}
-<<core.placenta.transcriptome.distribution,fig.width=9/2.5,fig.height=9/2.5,out.width="90mm",out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
+<<core.placenta.transcriptome.distribution.calc,echo=FALSE,message=FALSE,results="hide">>=
 num.species <- star.logs.wide[,length(unique(species))]
@@
 
+@
+
+\begin{figure}[tbh]
+<<core.placenta.transcriptome,fig.width=9/2.5,fig.height=8,out.width="90mm",out.height="0.8\\textheight",echo=FALSE>>=
+cptl <-
+    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
+                                            Species) & housekeeping==FALSE,]
+cptl <- cptl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
+                              levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
+                                          capfirst(names(species.names))))]
+
+print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
+      geom_boxplot(outlier.shape=NA)+
+      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
+      geom_point(aes(color=Species,shape=Species))+
+      scale_shape_manual("Species",
+                         guide=guide_legend(#nrow=1,
+                                            keywidth=unit(3,"mm"),
+                                            keyheight=unit(3,"mm"),
+                                            label.theme=element_text(face="italic",
+                                                                     angle=0,
+                                                                     size=7),
+                                            title.position="top"
+                                            ),
+                         values=0:18)+
+      # scale_color_discrete(NULL)+
+      coord_flip()+
+      xlab("Human Ortholog Name")+
+      ylab("Gene Expression (FPKM)")+
+      theme_bw() +
+      theme(text=element_text(size=9),
+            legend.position=c(0.99,0.41),
+            axis.text.y=element_text(face="italic",size=8),
+            legend.margin=unit(0,"mm"),
+            legend.justification=c(1,1),
+            plot.margin=unit(c(0,0,0,0),"mm")
+            ))
+@ 
+\caption{Core Placenta Transcriptome.
+  \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
+  1:1 \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
+  placenta of all 13 eutherians studied with housekeeping genes
+  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
+  ordered (from top to bottom) by median expression. If we also require
+  expression (\ac{FPKM}$ \ge 10$) in the marsupial, \opossum, there are 
+  \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 &
+    `Min. Exp. in` == '14 of 14 Species',max(num_genes)]} genes
+  which meet this criteria.
+  \Cref{fig:core_placenta_transcriptome_shape} shows the shape of the
+  distribution of the number of genes in the core placenta
+  transcriptome if different \ac{FPKM} and species restriction criteria are
+  utilized.}
+\label{fig:core_placenta_transcriptome}
+\end{figure}
+
+\begin{figure}
+<<core.placenta.transcriptome.distribution,fig.width=9/2.5,fig.height=9/2.5,out.width="90mm",out.height="0.8\\textheight",echo=FALSE,message=FALSE>>=
   print(ggplot(cpt.distribution.shape[`Expression (FPKM)` > 0.1,],

### 2016-12-06 manuscript/manuscript.Rnw
include 13 of 14 species listing
clarify most recent common ancestor listing
clarify pre-eclampsia onset

@@
   (or %
-  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[2]]},
+  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[2]]}, or
+  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[3]]},
   % 
@@
 which are also more highly expressed in the \mouse lineage such as
-\ac{Afp} and \ac{Gjb2}. Furthermore, the onset of
-pre-eclampsia is well before term; all of our placentas are term
-placentas and provide limited information about the transcriptome
-during the onset of pre-eclampsia.
+\ac{Afp} and \ac{Gjb2}. Furthermore, the biological processes which
+ultimately result in pre-eclampsia occur well before term; all of our
+placentas are term placentas and provide limited information about the
+transcriptome during the onset of pre-eclampsia.
 
@@
 maternal-fetal tolerance, such as the annexins, and genes responsible
-for cell-cell and cell-matrix interactions, such as \ac{LGALS1}.
-These genes were almost certainly present in the placenta of the
-common mammalian ancestor. We hypothesize that modifications to their
-expression or function will compromise the development and
-organization of the placenta.
+for cell-cell and cell-matrix interactions, such as \ac{LGALS1}. These
+genes were almost certainly present in the placenta of the most recent
+common mammalian ancestor included in this study. We hypothesize that
+modifications to their expression or function will compromise the
+development and organization of the placenta.
 

### 2016-12-06 manuscript/supplemental_material.Rnw
fix acronyms
add missing \clearpage commands
use short captions for supplemental figures

@@
 
+\clearpage
+
 \begin{figure}
@@
               floating=FALSE,
+              caption.placement="top",
               tabular.environment=ifelse(is.na(members),"longtable","tabular"),
@@
                                     members=NA,
-                                    caption="\\lineagetableexplanationtypeiii{\\human}{all other species}",
+                                    caption=c("\\lineagetableexplanationtypeiii{\\human}{all other species}","\\human lineage specific genes")
                                     label="tab:human_lineage_specific_all"
@@
 @ 
-\caption{GO terms most significantly enriched
+\caption[Core placenta GO Enrichment]{GO terms most significantly enriched
   (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]} MF,
@@
 @ 
-\caption{Top 50 genes which are associated with the shape of placenta attachment}
+\caption[Genes associated with attachment]{Top 50 genes which are associated with the shape of placenta attachment}
 \label{tab:shape_genes}
 \end{table}
+
+\clearpage
 
@@
 @ 
-\caption{Top 50 genes which are associated with the degree of placenta intimacy}
+\caption[Genes associated with intimacy]{Top 50 genes which are associated with the degree of placenta intimacy}
 \label{tab:intimacy_genes}
@@
 @ 
-\caption{Top 50 genes which are associated with the degree of placenta intimacy; ordered analysis}
+\caption[Intimacy genes: ordered analysis]{Top 50 genes which are associated with the degree of placenta intimacy; ordered analysis}
 \label{tab:intimacy_genes_polr}
@@
 
+\clearpage
 \subsection{Maternal-Fetal Interdigitation}
@@
 @ 
-\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation}
+\caption[Genes associated with interdigitation]{Top 50 genes which are associated with the degree of maternal-fetal interdigitation}
 \label{tab:interdigitation_genes}
@@
 @ 
-\caption{Top 50 genes which are associated with the degree of maternal-fetal interdigitation; ordered analysis}
+\caption[Interdigitation genes: ordered analysis]{Top 50 genes which are associated with the degree of maternal-fetal interdigitation; ordered analysis}
 \label{tab:interdigitation_genes_polr}

### 2016-12-06 manuscript/manuscript.Rnw
fix the Laurasiatheria collision

@@
 pp <- ggtree(species.tree)+
-    geom_text(aes(label=gsub("_"," ",label)),hjust=0,size=5,fontface="italic")+
-    geom_text(data=data.frame(x=1,y=1),label="Metatheria",hjust=0,vjust=0,size=5,fontface="italic")+
-    scale_x_continuous(limits=c(0,10))
+    geom_text(aes(label=gsub("_"," ",label)),vjust=0.65,hjust=0,size=4,fontface="italic")+
+    geom_text(data=data.frame(x=1,y=1),label="Metatheria",hjust=0,vjust=0,size=4,fontface="italic")+
+    scale_x_continuous(limits=c(0,10))+
+    theme(legend.margin=unit(0,"mm"),
+          plot.margin=unit(c(0,0,0,0),"mm")
+          ) + ggplot2::xlim(0,9)
+
 print(pp %>%

### 2016-12-06 manuscript/supplemental_material.Rnw
add short captions for supplemental figures

@@
 @ 
-\caption{\genetreeexplanation{\textit{CRH}}}
+\caption[Gene tree of \textit{CRH} Expression]{\genetreeexplanation{\textit{CRH}}}
 \label{fig:crh_evolution_tree}
@@
 @ 
-\caption{\genetreeexplanation[0.1]{\textit{KISS1}}}
+\caption[Gene tree of \textit{KISS1} Expression]{\genetreeexplanation[0.1]{\textit{KISS1}}}
 \label{fig:kiss1_evolution_tree}
@@
 @ 
-\caption{\genetreeexplanation[0.1]{\textit{PAPPA}}}
+\caption[Gene tree of \textit{PAPPA} Expression]{\genetreeexplanation[0.1]{\textit{PAPPA}}}
 \label{fig:pappa_evolution_tree}
@@
 @ 
-\caption{\genetreeexplanation{\textit{ADAM12}}}
+\caption[Gene tree of \textit{ADAM12} Expression]{\genetreeexplanation{\textit{ADAM12}}}
 \label{fig:adam12_evolution_tree}
@@
                                     members=NA,
-                                    caption=c("\\lineagetableexplanationtypeiii{\\human}{all other species}","\\human lineage specific genes")
+                                    caption=c("\\lineagetableexplanationtypeiii{\\human}{all other species}","\\human lineage specific genes"),
                                     label="tab:human_lineage_specific_all"
@@
 @ 
-\caption{Housekeeping genes excluded from the Core Placenta Transcriptome.
+\caption[Housekeeping genes excluded from the Core Placenta Transcriptome]{
+  Housekeeping genes excluded from the Core Placenta Transcriptome.
   \Sexpr{cptlh[,length(unique(human_name))]}
@@
 @ 
-\caption{Distribution shape of housekeeping genes in placentas out of
+\caption[Distribution shape of housekeeping genes in placentas]{
+  Distribution shape of housekeeping genes in placentas out of
   \Sexpr{pts[,sum(expressed.housekeeping)]} \human housekeeping 

### 2016-12-06 manuscript/manuscript.Rnw
resolve outstanding corrections

@@
 
- % \DEW{The discussion and results seem redundant. Given the title
- %   shouldn't something about shape be mentioned?}
-
 \end{abstract}
@@
 \end{table}
-
-% \DEW{Add gene names}
 
@@
 
-\PP{Introduce the characteristics earlier in the introduction}
-
-\DLA{They're in the first paragraph of the introduction, but I've also
-  referred to them now in the first subsection here.}
-
 \section{Discussion}
@@
 modulatory transcripts which are expressed in all placentas analyzed.
-
-\DEW{Why?} \DLA{Clarified why I think this is important to study.}
 
@@
 
-\PP{Add refs for fusion}
-
-\DLA{Added the Carmeille.ea2015 and Ueki.ea2012 I have other
-  references
-  (\cite{Ota.ea2013:ContributionoffetalANXA5,Bouter.ea2015:ReviewAnnexin-A5andcell})
-  but I'd exceed the limit for placenta (45)}
-
-% \DLA{Possibility of membrane repair due to maternal immune response as
-%   well during this process, as well as repair of membrane due to high
-%   secretion and adsoroption of nutrients/waste in the placenta.}
-
 % Cows have BERV-K1 and -K2, but it's not annotated in the ENSEMBL
@@
 \subsubsection{Invasion: EGF and CD9 in the Core Placenta}
-
-% \DEW{Write out FDR for enrichment described}
 
@@
 
-% \PP{EGF interacts with S100 from the core placenta as well}
-
 % TAGLN2 -- likely to be produced by the placenta as well; possibly
@@
 
-\DEW{Need to cite NWM grants, IGB, etc.}
-
-  \DLA{Unclear what the IGB grant is; Derek: which grant did you want
-    me to cite? Mentioned seed grant.}
-
 %  \printaddresses

### 2016-12-06 manuscript/manuscript.Rnw
use short gene names in the intro

@@
   %
-  including genes with immune-modulating properties (\ac{ANXA2},
-  \ac{ANXA1}, \ac{S100A11}, \ac{S100A10}, and \ac{LGALS1}), cell-cell
-  interactions (\ac{LAMC1}, \ac{LUM}, and \ac{LGALS1}),
-  invasion (\ac{GRB2} and \ac{RALB}) and syncytialization
-  (\ac{ANXA5} and \ac{ANXA1}). %
+  including genes with immune-modulating properties (\acs{ANXA2},
+  \acs{ANXA1}, \acs{S100A11}, \acs{S100A10}, and \acs{LGALS1}), cell-cell
+  interactions (\acs{LAMC1}, \acs{LUM}, and \acs{LGALS1}),
+  invasion (\acs{GRB2} and \acs{RALB}) and syncytialization
+  (\acs{ANXA5} and \acs{ANXA1}). %
   We also identified multiple pre-eclampsia associated genes which are
@@
   Multiple genes are significantly associated with placenta
-  morphology, including \ac{EREG} and \ac{WNT5A} which are both
+  morphology, including \acs{EREG} and \acs{WNT5A} which are both
   associated with placental shape.
@@
 \ac{TIMP3}, has been implicated in pre-eclampsia
-\cite{Ma.ea2014:Down-regulationofTIMP3leads,Xiang.ea2013:PromoterhypomethylationofTIMP3}.
+\cite{Ma.ea2014:Down-regulationofTIMP3leads}.
 

### 2016-12-06 manuscript/manuscript.Rnw
don't bother to mention increased brain size

@@
 myometrium of the uterus
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. It is
-hypothesized that this feature is related to the increased demand for
-resources by the developing fetus to feed the expanded size of the
-human brain
-\cite{Carter.ea2011:Evolutionofinvasiveplacentation,Robillard.ea2002:Evolutionaryadaptationstopre-eclampsiaeclampsia}.
-
+\cite{Carter.ea2011:Evolutionofinvasiveplacentation}. %
 The near exclusivity of deep trophoblast invasion in humans has also

### 2016-12-12 manuscript/manuscript.Rnw
update affiliation and email for cjkim (now at Asan)
add equal contributor line which went missing
use RNA-Seq everywhere

@@
 \ead{nevo@research.haifa.ac.il}
-\author[affnih,affwaynepath]{Chong J Kim}
-\ead{cjkim@med.wayne.edu}
+\author[affnih,affasanpath]{Chong J Kim}
+\ead{ckim@amc.seoul.kr}
 \author[affyale]{Günter P. Wagner}
@@
 \cortext[cor1]{Corresponding author}
+\fntext[n1]{Equal contributor}
 \address[affigb]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
@@
 \address[affnih]{Perinatology Research Branch, Eunice Kennedy Shriver National Institute of Child Health and Human Development (NICHD), National Institutes of Health (NIH), Detroit, MI, USA}
+\address[affasan]{Department of Pathology, University of Ulsan College of Medicine, Asan Medical Center, Seoul, Korea}
 \address[affwaynepath]{Department of Pathology, Wayne State University School of Medicine, Detroit, MI, USA}
@@
 
-  \textbf{Method} Using RNA-seq, we measured the expression profiles
+  \textbf{Method} Using RNA-Seq, we measured the expression profiles
   of \Sexpr{star.logs[,length(unique(file))]} term placentas of
@@
 \begin{keyword}
-mammal \sep  gene expression \sep evolution \sep RNA-seq \sep placenta
+mammal \sep  gene expression \sep evolution \sep RNA-Seq \sep placenta
 \end{keyword}
@@
 
-Supplemental \cref{fig:transcriptome_statistics} shows the RNAseq read and
+Supplemental \cref{fig:transcriptome_statistics} shows the RNA-Seq read and
 alignment statistics for all species. For transcriptomes sequenced

### 2016-12-12 manuscript/manuscript.Rnw
increase spacing in species tree labels

@@
 pp <- ggtree(species.tree)+
-    geom_text(aes(label=gsub("_"," ",label)),vjust=0.65,hjust=0,size=4,fontface="italic")+
-    geom_text(data=data.frame(x=1,y=1),label="Metatheria",hjust=0,vjust=0,size=4,fontface="italic")+
+    geom_text(aes(label=gsub("_"," ",label)),nudge_x=0.05,vjust=0.65,hjust=0,size=4,fontface="italic")+
+    geom_text(data=data.frame(x=1,y=1),label="Metatheria",nudge_y=0.05,hjust=0,vjust=0,size=4,fontface="italic")+
     scale_x_continuous(limits=c(0,10))+
@@
       ## Armadillo
-      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.3,node=2,width=0.15)
+      %>% phylopic("5d59b5ce-c1dd-40f6-b295-8d2629b9775e", color="#44EEBB",alpha=0.3,node=2,width=0.13)
       ## elephant
-      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.3,node=3,width=0.15)
+      %>% phylopic("80db1004-bc9f-4318-84e7-cdd9639a1f3e",color="#BB1111",alpha=0.3,node=3,width=0.13)
       ## dog
@@
       ## mouse
-      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.3,node=11,width=0.18)
+      %>% phylopic("6b2b98f6-f879-445f-9ac2-2c2563157025",color="#BB11CC",alpha=0.3,node=11,width=0.15)
       ## bonobo
-      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.3,node=12,width=0.15)
+      %>% phylopic("f06b4dd0-094e-4f37-9105-25fdb4eb1b02",color="#1111FF",alpha=0.3,node=12,width=0.13)
       ## spider monkey

### 2016-12-12 manuscript/manuscript.Rnw
use siunitx for genes table

@@
 %\usepackage{url}
+\usepackage{siunitx}
 \usepackage{graphicx}
@@
         # setnames(display.table,"Human Symbol","\\multicolumn{1}{p{0.5in}}{Human Symbol}")6
-        setnames(display.table,"Fold Change","$\\text{log}_2$ FC")
+        setnames(display.table,"Fold Change","{$\\text{log}_2$ FC}")
+        setnames(display.table,"FDR","{FDR}")
+        setnames(display.table,"p (t-test)","{p (t-test)}")
         print(xtable(display.table,
-                     display=c("s","s","e","g","f"),
-                     align=c("c","p{0.5in}","r","r","r")
+                     display=c("s","s","e","e","f"),
+                     align=c("c","p{0.5in}","S[table-format=1.2e2]","S[table-format=1.2e2]","S[table-format=2.2]")
                      ),
@@
               floating=FALSE,
-              math.style.exponents=TRUE,
+              math.style.exponents=FALSE,
               sanitize.text.function=genes.table.sanitize,

### 2016-12-17 manuscript/manuscript.Rnw
add suggestions from michael

@@
 related to the thymus-specific serine protease \ac{PRSS16} in
-\elephant (in this work), and the expansion of the growth hormone and
+\elephant, and the expansion of the growth hormone and
 chorionic gonadotropin beta genes in anthropoid primates
@@
 offspring and length of gestation (\cref{tab:gestational_periods}). We
-use this data, along with existing RNA-Seq data from \mouse
+use these data, along with existing RNA-Seq data from \mouse
 \cite{Wang.ea2011:surveyfornovelimprinted}, \human
@@
 
-\subsection{Pre-eclampsia \human Lineage Specific Genes}
+\subsection{\human Lineage Specific Genes and Their Association with Pre-eclampsia}
 
@@
             legend.position=c(0.99,0.41),
-            axis.text.y=element_text(face="italic",size=8),
+            axis.text.y=element_text(face="italic",size=7),
             legend.margin=unit(0,"mm"),
@@
   placenta transcriptome if %
-  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[1]]}
+  \Sexpr{cpt.distribution.shape[,tolower(levels(`Min. Exp. in`))[1]]}
   %
   (or %
-  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[2]]}, or
-  \Sexpr{cpt.distribution.shape[,levels(`Min. Exp. in`)[3]]},
+  \Sexpr{cpt.distribution.shape[,tolower(levels(`Min. Exp. in`))[2]]}, or
+  \Sexpr{cpt.distribution.shape[,tolower(levels(`Min. Exp. in`))[3]]},
   % 
@@
 expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
-(excluding the metatherian, \opossum) the and was not a human housekeeping
+(excluding the metatherian, \opossum) and was not a human housekeeping
 gene according to
@@
 
-\subsection{Placenta Morphology and Genetics}
-
-We correlated different aspects of placenta morphology, including
-shape, intimacy, and interdigitation with the mRNA transcription
-profiles of all 1:1 human orthologs with \ac{FPKM} greater than 10 in
-one species, and \ac{FPKM} greater than 1 at least two species.
-Supplemental \cref{tab:placental_classification} lists the shape,
-intimacy, and interdigitation of each species analyzed.
-
-\subsubsection{Placenta Shape}
+\subsection{Placenta Morphology: Shape}
 
@@
 
-\subsubsection{Degree of Intimacy and Interdigitation}
+\subsection{Placenta Morphology: Degree of Intimacy and Interdigitation}
 
@@
 intimacy (Supplemental
-\cref{tab:intimacy_genes,tab:interdigitation_genes}) using an
-unordered factor approach, but when we used an ordered factor
-analysis, none of these genes were retained (Supplemental
+\cref{tab:intimacy_genes,tab:interdigitation_genes}) using ANOVA, but
+when we used an ordered factor analysis (proportional odds logistic
+regression), none of these genes were retained (Supplemental
 \cref{tab:intimacy_genes_polr,tab:interdigitation_genes_polr}). We
@@
 transport, immune regulation, cell fusion and membrane repair, and
-invasion.
+invasion (Supplemental
+\cref{tab:core_placenta_go_terms,tab:core_placenta_pathway_analysis}).
 

### 2016-12-17 manuscript/acronyms.tex
add links to hugo for genes

--- 2016-11-29 manuscript/acronyms.tex
@@
-\DeclareAcronym{EV}{short=EV,long=Extracellular Vesicle}
-\DeclareAcronym{SVM}{long=Support Vector Machine}
-\DeclareAcronym{ROC}{long=Receiver Operating Characteristic}
-\DeclareAcronym{KNN}{long=K Nearest Neighbors}
-\DeclareAcronym{SRA}{long=Sequence Read Archive}
-\DeclareAcronym{PRG2}{short=\textit{PRG2},long=Proteoglycan 2, bone marrow}
-\DeclareAcronym{FPKM}{long=Fragments Per Kilobase of exon per Million reads}
-\DeclareAcronym{CSH1}{short=\textit{CSH1},long=chorionic somatomammotropin hormone 1}
-\DeclareAcronym{CSH2}{short=\textit{CSH2},long=chorionic somatomammotropin hormone 2}
-\DeclareAcronym{CSHL1}{short=\textit{CSHL1},long=chorionic somatomammotropin hormone-like 1}
-\DeclareAcronym{GH2}{short=\textit{GH2},long=growth hormone 2}
-\DeclareAcronym{CEA}{long=carcinoembryonic antigen}
-\DeclareAcronym{HSD3B1}{short=\textit{HSD3B1},long=3-β Hydroxysteroid dehydrogenase 1}
-\DeclareAcronym{GSE}{long=Gene Set Enrichment}
-\DeclareAcronym{OU}{long=Ornstein-Uhlenbeck}
-\DeclareAcronym{OUCH}{long=\ac{OU} Comparative Hypotheses}
-\DeclareAcronym{TSI}{long=Tissue Specificity Index}
-\DeclareAcronym{SRA}{long=Sequence Read Archive}
-\DeclareAcronym{GH}{long=Growth Hormone}
-\DeclareAcronym{PSG}{long=Pregnancy-Specific Glycoprotein}
-\DeclareAcronym{PAG}{long=Pregnancy-Associated Glycoprotein}
-\DeclareAcronym{GO}{long=Gene Ontology}
-\DeclareAcronym{FDR}{long=False Discovery Rate}
-\DeclareAcronym{gsea}{short=GSEA,long=gene set enrichment analysis}
-\DeclareAcronym{CRH}{short=\textit{CRH},long={corticotropin-releasing hormone}}
-\DeclareAcronym{KISS1}{short=\textit{KISS1},long={KiSS-1 metastatis-supressor}}
-\DeclareAcronym{PAPPA}{short=\textit{PAPPA},long={pappalysin 1}}
-\DeclareAcronym{ADAM12}{short=\textit{ADAM12},long={ADAM Metallopeptidase Domain 12}}
-\DeclareAcronym{ANXA1}{short=\textit{ANXA1},long={annexin A1}}
-\DeclareAcronym{ANXA5}{short=\textit{ANXA5},long={annexin A5}}
-\DeclareAcronym{S100A11}{short=\textit{S100A11},long={S100 calcium binding protein A11}}
-\DeclareAcronym{S100A10}{short=\textit{S100A10},long={S100 calcium binding protein A10}}
-\DeclareAcronym{LGALS1}{short=\textit{LGALS1},long={galectin 1}}
-\DeclareAcronym{LAMC1}{short=\textit{LAMC1},long={laminin subunit gamma 1}}
-\DeclareAcronym{LUM}{short=\textit{LUM},long={lumican}}
-\DeclareAcronym{GRB2}{short=\textit{GRB2},long={growth factor receptor bound protein 2}}
-\DeclareAcronym{RALB}{short=\textit{RALB},long={RAS like proto-oncogene B}}
-\DeclareAcronym{ERVFRD-1}{short=\textit{ERVFRD-1},long={endogenous retrovirus group FRD member 1}}
-\DeclareAcronym{ERVW-1}{short=\textit{ERVW-1},long={endogenous retrovirus group W member 1}}
-\DeclareAcronym{MFSD2A}{short=\textit{MFSD2A},long={major facilitator superfamily domain containing 2A}}
-\DeclareAcronym{PRSS16}{short=\textit{PRSS16},long={protease, serine 16}}
-\DeclareAcronym{TIMP3}{short=\textit{TIMP3},long={TIMP metallopeptidase inhibitor 3}}
-\DeclareAcronym{F11R}{short=\textit{F11R},long={F11 receptor}}
-\DeclareAcronym{SDC4}{short=\textit{SDC4},long={syndecan 4}}
-\DeclareAcronym{FN1}{short=\textit{FN1},long={fibronectin 1}}
-\DeclareAcronym{ITGB1}{short=\textit{ITGB1},long={integrin subunit beta 1}}
-\DeclareAcronym{ITGB1BP1}{short=\textit{ITGB1BP1},long={integrin subunit beta 1 binding protein 1}}
-\DeclareAcronym{EREG}{short=\textit{EREG},long={epiregulin}}
-\DeclareAcronym{UNC93A}{short=\textit{UNC93A},long={unc-93 homolog A (\textit{C. elegans})}}
-\DeclareAcronym{UNC93B1}{short=\textit{UNC93B1},long={unc-93 homolog B1 (\textit{C. elegans)}}}
-\DeclareAcronym{WNT5A}{short=\textit{WNT5A},long={Wnt family member 5A}}
-\DeclareAcronym{IL1RL1}{short=\textit{IL1RL1},long={interleukin 1 receptor like 1}}
-\DeclareAcronym{PLAC1}{short=\textit{PLAC1},long={placenta specific 1}}
-\DeclareAcronym{Afp}{short=\textit{Afp},long={alpha fetoprotein}}
-\DeclareAcronym{Gjb2}{short=\textit{Gjb2},long={gap junction protein beta 2}}
-\DeclareAcronym{CD9}{short=\textit{CD9},long={CD9 molecule}}
+\DeclareAcronym{EV}{short=EV,long={Extracellular Vesicle}}
+\DeclareAcronym{SVM}{short=SVM,long={Support Vector Machine}}
+\DeclareAcronym{ROC}{short=ROC,long={Receiver Operating Characteristic}}
+\DeclareAcronym{KNN}{short=KNN,long={K Nearest Neighbors}}
+\DeclareAcronym{SRA}{short=SRA,long={Sequence Read Archive}}
+\DeclareAcronym{CEA}{short=CEA,long={carcinoembryonic antigen}}
+\DeclareAcronym{GSE}{short=GSE,long={Gene Set Enrichment}}
+%\DeclareAcronym{OU}{short=OU,long=Ornstein-Uhlenbeck}
+%\DeclareAcronym{OUCH}{short=OUCH,long=\ac*{OU} Comparative Hypotheses}
+\DeclareAcronym{TSI}{short=TSI,long=Tissue Specificity Index}
+\DeclareAcronym{GH}{short=GH,long=Growth Hormone}
+\DeclareAcronym{PSG}{short=PSG,long={Pregnancy-Specific Glycoprotein}}
+\DeclareAcronym{PAG}{short=PAG,long={Pregnancy-Associated Glycoprotein}}
+\DeclareAcronym{GO}{short=GO,long={Gene Ontology}}
+\DeclareAcronym{FDR}{short=FDR,long={False Discovery Rate}}
+\DeclareAcronym{gsea}{short=GSEA,long={gene set enrichment analysis}}
+\DeclareAcronym{FPKM}{short=FPKM,long={Fragments Per Kilobase of exon per Million reads}}
+\DeclareAcronym{HSD3B1}{short=\textit{HSD3B1},long={hydroxy-δ-5-steroid dehydrogenase, 3 β- and steroid δ-isomerase 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=HSD3B1}{\textit{HSD3B1}} (hydroxy-δ-5-steroid dehydrogenase, 3 β- and steroid δ-isomerase 1)}}
+\DeclareAcronym{CSH1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CSH1}{\textit{CSH1}},long={chorionic somatomammotropin hormone 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CSH1}{\textit{CSH1}} (chorionic somatomammotropin hormone 1)}}
+\DeclareAcronym{CSH2}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CSH2}{\textit{CSH2}},long={chorionic somatomammotropin hormone 2},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CSH2}{\textit{CSH2}} (chorionic somatomammotropin hormone 2)}}
+\DeclareAcronym{CSHL1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CSHL1}{\textit{CSHL1}},long={chorionic somatomammotropin hormone like 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CSHL1}{\textit{CSHL1}} (chorionic somatomammotropin hormone like 1)}}
+\DeclareAcronym{GH2}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=GH2}{\textit{GH2}},long={growth hormone 2},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=GH2}{\textit{GH2}} (growth hormone 2)}}
+\DeclareAcronym{PRG2}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=PRG2}{\textit{PRG2}},long={proteoglycan 2, pro eosinophil major basic protein},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=PRG2}{\textit{PRG2}} (proteoglycan 2, pro eosinophil major basic protein)}}
+\DeclareAcronym{CRH}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CRH}{\textit{CRH}},long={corticotropin releasing hormone},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CRH}{\textit{CRH}} (corticotropin releasing hormone)}}
+\DeclareAcronym{KISS1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=KISS1}{\textit{KISS1}},long={KiSS-1 metastasis-suppressor},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=KISS1}{\textit{KISS1}} (KiSS-1 metastasis-suppressor)}}
+\DeclareAcronym{PAPPA}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=PAPPA}{\textit{PAPPA}},long={pappalysin 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=PAPPA}{\textit{PAPPA}} (pappalysin 1)}}
+\DeclareAcronym{ADAM12}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ADAM12}{\textit{ADAM12}},long={ADAM metallopeptidase domain 12},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ADAM12}{\textit{ADAM12}} (ADAM metallopeptidase domain 12)}}
+\DeclareAcronym{ANXA1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ANXA1}{\textit{ANXA1}},long={annexin A1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ANXA1}{\textit{ANXA1}} (annexin A1)}}
+\DeclareAcronym{ANXA2}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ANXA2}{\textit{ANXA2}},long={annexin A2},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ANXA2}{\textit{ANXA2}} (annexin A2)}}
+\DeclareAcronym{ANXA5}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ANXA5}{\textit{ANXA5}},long={annexin A5},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ANXA5}{\textit{ANXA5}} (annexin A5)}}
+\DeclareAcronym{S100A11}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=S100A11}{\textit{S100A11}},long={S100 calcium binding protein A11 (calgizzarin)},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=S100A11}{\textit{S100A11}} (S100 calcium binding protein A11; calgizzarin)}}
+\DeclareAcronym{S100A10}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=S100A10}{\textit{S100A10}},long={S100 calcium binding protein A10 (calpectin-I light polypeptide)},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=S100A10}{\textit{S100A10}} (S100 calcium binding protein A10; calpectin-I light polypeptide)}}
+\DeclareAcronym{LGALS1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=LGALS1}{\textit{LGALS1}},long={galectin 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=LGALS1}{\textit{LGALS1}} (galectin 1)}}
+\DeclareAcronym{LAMC1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=LAMC1}{\textit{LAMC1}},long={laminin subunit γ 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=LAMC1}{\textit{LAMC1}} (laminin subunit γ 1)}}
+\DeclareAcronym{LUM}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=LUM}{\textit{LUM}},long={lumican},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=LUM}{\textit{LUM}} (lumican)}}
+\DeclareAcronym{GRB2}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=GRB2}{\textit{GRB2}},long={growth factor receptor bound protein 2},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=GRB2}{\textit{GRB2}} (growth factor receptor bound protein 2)}}
+\DeclareAcronym{RALB}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=RALB}{\textit{RALB}},long={RAS like proto-oncogene B},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=RALB}{\textit{RALB}} (RAS like proto-oncogene B)}}
+\DeclareAcronym{ERVFRD-1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ERVFRD-1}{\textit{ERVFRD-1}},long={endogenous retrovirus group FRD member 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ERVFRD-1}{\textit{ERVFRD-1}} (endogenous retrovirus group FRD member 1)}}
+\DeclareAcronym{ERVW-1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ERVW-1}{\textit{ERVW-1}},long={endogenous retrovirus group W member 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ERVW-1}{\textit{ERVW-1}} (endogenous retrovirus group W member 1)}}
+\DeclareAcronym{MFSD2A}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=MFSD2A}{\textit{MFSD2A}},long={major facilitator superfamily domain containing 2A},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=MFSD2A}{\textit{MFSD2A}} (major facilitator superfamily domain containing 2A)}}
+\DeclareAcronym{PRSS16}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=PRSS16}{\textit{PRSS16}},long={protease, serine 16},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=PRSS16}{\textit{PRSS16}} (protease, serine 16)}}
+\DeclareAcronym{TIMP3}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=TIMP3}{\textit{TIMP3}},long={TIMP metallopeptidase inhibitor 3},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=TIMP3}{\textit{TIMP3}} (TIMP metallopeptidase inhibitor 3)}}
+\DeclareAcronym{F11R}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=F11R}{\textit{F11R}},long={F11 receptor},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=F11R}{\textit{F11R}} (F11 receptor)}}
+\DeclareAcronym{SDC4}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=SDC4}{\textit{SDC4}},long={syndecan 4},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=SDC4}{\textit{SDC4}} (syndecan 4)}}
+\DeclareAcronym{FN1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=FN1}{\textit{FN1}},long={fibronectin 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=FN1}{\textit{FN1}} (fibronectin 1)}}
+\DeclareAcronym{ITGB1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ITGB1}{\textit{ITGB1}},long={integrin subunit β 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ITGB1}{\textit{ITGB1}} (integrin subunit β 1)}}
+\DeclareAcronym{ITGB1BP1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ITGB1BP1}{\textit{ITGB1BP1}},long={integrin subunit β 1 binding protein 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=ITGB1BP1}{\textit{ITGB1BP1}} (integrin subunit β 1 binding protein 1)}}
+\DeclareAcronym{EREG}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=EREG}{\textit{EREG}},long={epiregulin},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=EREG}{\textit{EREG}} (epiregulin)}}
+\DeclareAcronym{UNC93A}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93A}{\textit{UNC93A}},long={unc-93 homolog A (C.
+elegans)},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93A}{\textit{UNC93A}} (unc-93 homolog A (C.
+elegans))}}
+\DeclareAcronym{UNC93B1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93B1}{\textit{UNC93B1}},long={unc-93 homolog B1 (C.
+elegans)},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93B1}{\textit{UNC93B1}} (unc-93 homolog B1 (C.
+elegans))}}
+\DeclareAcronym{WNT5A}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=WNT5A}{\textit{WNT5A}},long={Wnt family member 5A},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=WNT5A}{\textit{WNT5A}} (Wnt family member 5A)}}
+\DeclareAcronym{IL1RL1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=IL1RL1}{\textit{IL1RL1}},long={interleukin 1 receptor like 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=IL1RL1}{\textit{IL1RL1}} (interleukin 1 receptor like 1)}}
+\DeclareAcronym{PLAC1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=PLAC1}{\textit{PLAC1}},long={placenta specific 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=PLAC1}{\textit{PLAC1}} (placenta specific 1)}}
+\DeclareAcronym{Afp}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=Afp}{\textit{Afp}},long={α-fetoprotein},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=Afp}{\textit{Afp}} (α-fetoprotein)}}
+\DeclareAcronym{Gjb2}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=Gjb2}{\textit{Gjb2}},long={gap junction protein β 2},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=Gjb2}{\textit{Gjb2}} (gap junction protein β 2)}}
+\DeclareAcronym{CD9}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CD9}{\textit{CD9}},long={CD9 molecule},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CD9}{\textit{CD9}} (CD9 molecule)}}
 

### 2017-02-01 manuscript/manuscript.Rnw
fix Chong J. Kim's affiliation
It's the Carl R. Woese IGB
remove fixmes and clearpage
change are expressed to is expressed

@@
 \ead{nevo@research.haifa.ac.il}
-\author[affnih,affasanpath]{Chong J Kim}
+\author[affnih,affasan]{Chong J. Kim}
 \ead{ckim@amc.seoul.kr}
@@
 \fntext[n1]{Equal contributor}
-\address[affigb]{Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
+\address[affigb]{Carl R. Woese Institute for Genomic Biology, University of Illinois at Urbana-Champaign, Urbana, IL, USA}
 \address[afflon]{School of Biological and Chemical Sciences, Queen Mary, University of London, London, UK}
@@
   \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
-  core genes which are expressed (FPKM $\ge 10$) in all eutherian placentas, %
+  core genes which is expressed (FPKM $\ge 10$) in all eutherian placentas, %
   %
@@
 
-\listoffixmes
+\clearpage 
+% \listoffixmes
 

### 2017-02-06 manuscript/manuscript.Rnw
clarify that the gene shape is for all genes
clarify superorder and species tree
 - taxonomic orders are now approximately
   at the same level for the major taxonomic
   orders
 - indicate that branches are not to scale

@@
 \cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman,Hou.ea2012:Elephanttranscriptomeprovidesinsights}}.
-For example, each superorder has evolved expansive placenta-specific gene
+For example, each superorder (Euarchontoglires, Laurasiatheria,
+Afrotheria, and Xenartha, \cref{fig:species_tree})
+has evolved expansive placenta-specific gene
 families with expression patterns that can vary in space and time
@@
 species.tree <- 
-    read.tree(text="(Monodelphis_domestica,((Dasypus_novemcinctus,Loxodonta_africana)Atlantogenata,((Canis_familiaris,(Equus_caballus,(Sus_scrofa,(Ovis_aries,Bos_taurus))))Laurasiatheria,(((Spalax_carmeli,Spalax_galili),Mus_Musculus)Rodentia,((Pan_paniscus,Homo_sapiens),Ateles_fuciceps)Primates)Euarchontoglires)Boreoeutheria)Eutheria);")
+    read.tree(text="(Monodelphis_domestica:7,((Dasypus_novemcinctus:5,Loxodonta_africana:5)Atlantogenata:1,((Canis_familiaris:4,(Equus_caballus:3,(Sus_scrofa:2,(Ovis_aries:1,Bos_taurus:1):1):1):1)Laurasiatheria:1,(((Spalax_carmeli:1,Spalax_galili:1):1,Mus_Musculus:2)Rodentia:2,((Pan_paniscus:1,Homo_sapiens:1):1,Ateles_fuciceps:2)Primates:2)Euarchontoglires:1)Boreoeutheria:1)Eutheria:1);")
 pp <- ggtree(species.tree)+
@@
     geom_text(data=data.frame(x=1,y=1),label="Metatheria",nudge_y=0.05,hjust=0,vjust=0,size=4,fontface="italic")+
+    geom_text(data=data.frame(x=3.3,y=c(2,3)),label=c("Xenartha","Afrotheria"),nudge_y=0.05,hjust=0,vjust=c(0,0),size=4,fontface="italic")+
     scale_x_continuous(limits=c(0,10))+
-    theme(legend.margin=unit(0,"mm"),
+    theme(legend.spacing=unit(0,"mm"),
           plot.margin=unit(c(0,0,0,0),"mm")
           ) + ggplot2::xlim(0,9)
-
 print(pp %>%
@@
 \caption{Species tree for organisms studied showing the relationship
-  between species. Silhouettes of species were obtained from
-  \url{http://phylopic.org}. }
+  between species. Branch lengths are not to scale. Silhouettes of
+  species were obtained from \url{http://phylopic.org}. }
 \label{fig:species_tree}
@@
 @ 
-\caption{Distribution shape of placenta core genes. The vertical axis
+\caption{Distribution shape of placenta core genes for all possible
+  number of genes. The vertical axis
   shows the number of 1:1 orthologous genes included in the core
-  placenta transcriptome if %
+  placenta transcriptome (for all possible number of genes) if %
   \Sexpr{cpt.distribution.shape[,tolower(levels(`Min. Exp. in`))[1]]}

### 2017-02-06 manuscript/manuscript.Rnw
switch from 1:1 to single-copy ortholog

@@
 
-1:1 orthologs in were determined using
+Single-copy (1-to-1) orthologs in were determined using
 Ensembl v.80 gene trees via Biomart using the biomaRt R package
@@
 A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
-  1,length(unique(human_name))]} 1:1 orthologous genes were expressed
+  1,length(unique(human_name))]} single-copy orthologous genes were expressed
 in at least one species.
@@
 Significant shifts in expression levels between groups of species were
-calculated using Student's t test on 1:1 orthologs using the mean of
+calculated using Student's t test on single-copy orthologs using the mean of
 each species to avoid large biases towards samples with $n \gg 1$ and
@@
   \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
-  1:1 \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
+  single-copy \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
   placenta of all 13 eutherians studied with housekeeping genes
@@
   number of genes. The vertical axis
-  shows the number of 1:1 orthologous genes included in the core
+  shows the number of single-copy orthologous genes included in the core
   placenta transcriptome (for all possible number of genes) if %
@@
   %
-  genes would be included if the expression of a set of 1:1 orthologs of all 13 Eutherians must be at
+  genes would be included if the expression of a set of single-copy orthologs of all 13 Eutherians must be at
   least 10 FPKM, %
@@
   %
-  1:1 orthologs would be included if 13 of 14 species have an
+  single-copy orthologs would be included if 13 of 14 species have an
    expression of at least 10 FPKM, and 
@@
 transcription level in \ac{FPKM}. We considered a gene to be core to
-the placenta transcriptome if a 1:1 ortholog of that gene was
+the placenta transcriptome if a single-copy ortholog of that gene was
 expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
@@
 resolution. Increasing the permissiveness of the core placenta
-transcriptome by only requiring 10 of 14 species to express 1:1
+transcriptome by only requiring 10 of 14 species to express single-copy
 orthologs with FPKM $>10$ and results in the same terms also being

### 2017-02-06 manuscript/manuscript.Rnw
correct typos in annex evolutionary importance ¶

@@
 \ac{ANXA2}, \ac{ANXA1}, \ac{S100A11}, and \ac{S100A10}
-are expressed in all placentas examined, and argue for the evolution
+are expressed in all placentas examined, and argue for the evolutionary
 importance of annexins in placenta function, where they likely
@@
 tolerance\cite{Liu.ea2015:AnnexinA2complexeswith,Hutchinson.ea2011:Molecularregulatorsofresolution,Aarli.ea1997:Suppressiveeffectonlymphoproliferation}.
-We suggest that further research into molecular role of these annexin
+We suggest that further research into the molecular role of these annexin
 complexes may illuminate how immune tolerance is maintained during
-placentation because these appear to the the major class of immune
+placentation, because annexins appear to be the major class of immune
 modulatory transcripts which are expressed in all placentas analyzed.

### 2017-02-13 manuscript/supplemental_material.Rnw
use TSI to show that housekeeping genes are OK

@@
 
+<<tissue_specificity_index_vs_housekeeping,echo=FALSE,message=FALSE>>=
+tissue.specificity.index <- function(expression){
+    expression <- expression[!is.na(expression)]
+    expression <- expression[is.finite(expression)]
+    if (length(expression)==0) {
+        return(0)
+    }
+    if (max(expression)==0) {
+        return(0)
+    }
+    return(sum(1-expression/max(expression))/
+               (length(expression)-1))
+}
+tissue.specific.data <- fread("zcat ../data/rna_tissue.csv.zip")
+tissue.specific.markers <-
+    tissue.specific.data[,list(`Gene name`,
+                               "Tissue Specificity Index"=
+                                   tissue.specificity.index(as.numeric(Value))
+                               ),
+                         by=Gene][!duplicated(Gene),]
+rm(tissue.specific.data)
+
+setkey(tissue.specific.markers,"Gene name")
+core.placenta.genes <-
+    core.placenta.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
+                                            Species),
+                                     list("Median FPKM"=median(FPKM),
+                                          "Gene Type"=ifelse(housekeeping,
+                                                             "Housekeeping","Not Housekeeping"),
+                                          housekeeping=housekeeping[1]
+                                          ),
+                                     by=human_name]
+setkey(core.placenta.genes,"human_name")
+core.placenta.genes.tsi <-
+    tissue.specific.markers[core.placenta.genes[!duplicated(human_name)]]
+rm(tissue.specific.markers,core.placenta.genes)
+@
+
+\begin{figure}
+<<tissue_specificity_index_vs_median_fpkm,echo=FALSE,message=FALSE>>=
+grid::pushViewport(grid::viewport(layout=grid::grid.layout(ncol=2,nrow=1,widths=c(1.5,1))))
+print(ggplot(core.placenta.genes.tsi,
+             aes(x=`Median FPKM`,
+                 y=`Tissue Specificity Index`,
+                 color=ifelse(housekeeping,"Housekeeping","Not\nHousekeeping")))
+      + geom_point()
+      + stat_ellipse()
+      + ylab("Tissue Specificity Index")
+      + xlab("Median FPKM")
+      + scale_x_log10(labels=trans_format("log10", math_format(10^.x)))
+      + scale_color_discrete("")
+      + theme(legend.position="top"),
+      vp=grid::viewport(layout.pos.row=1,layout.pos.col=1)
+      )
+print(ggplot(core.placenta.genes.tsi,
+             aes(x=ifelse(housekeeping,"Housekeeping","Not Housekeeping"),
+                 y=`Tissue Specificity Index`,
+                 color=ifelse(housekeeping,"Housekeeping","Not\nHousekeeping")))
+      + geom_boxplot(outlier.color=NA)
+      + geom_jitter()
+      + ylab("Tissue Specificity Index")
+      + xlab("Gene Type")
+      + scale_color_discrete("")
+      + theme(legend.position="top"),
+      vp=grid::viewport(layout.pos.row=1,layout.pos.col=2)
+      )
+grid::popViewport()
+grid::grid.text(c("A","B"),
+                y=unit(1,"npc")-unit(0.1,"cm"),
+                x=unit(c(0,1.5/2.5),"npc")+unit(0.1,"cm"),
+                gp=gpar(fontsize=24),
+                just=c("left","top"))
+
+@ 
+\caption[Tissue specificity index with housekeeping]{Tissue
+  specificity index and housekeeping genes. In general, the set of
+  housekeeping genes identified by
+  \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} has lower
+  tissue specificity
+  index~\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}
+  than genes which are not tissue specific (calculated based on the
+  Human Protein Atlas version 16.1 RNA-seq data
+  \cite{Uhlén.ea2015:ProteomicsTissue-basedmapof}). Some of the genes
+  with higher tissue specificity (\textit{NPC} and \textit{CISD1} have
+  outlier expression in a single tissue, even though they are
+  moderately expressed in every tissue.)}
+\label{fig:tissue_specificity_index_with_housekeeping}
+\end{figure}
+
+\clearpage
+
+<<tissue_specificity_index_of_placenta_core_genes,echo=FALSE,results="asis">>=
+print(xtable(core.placenta.genes.tsi[order(-`Median FPKM`),
+                                     list(`Gene name`,
+                                          `Tissue Specificity Index`,
+                                          `Median FPKM`,
+                                          `Gene Type`
+                                          )],
+             digits=2,
+             display=c("s","s","f","f","s"),
+             caption=c("Tissue specificity index~\\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles} (calculated based on the Human Protein Atlas version 16.1 RNA-seq data \\cite{Uhlén.ea2015:ProteomicsTissue-basedmapof}) of Placenta Core Genes ordered by median expression with housekeeping status~\\cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} indicated.",
+                       "Tissue specificity of Placenta Core Genes"
+                       ),
+             label="tab:tissue_specificity_index_of_plaenta_core_genes"
+             ),
+      include.rownames=FALSE,
+      booktabs=TRUE,
+      math.style.exponents=TRUE,
+      size="\\footnotesize",
+      caption.placement="top",
+      tabular.environment="longtable",
+      floating=FALSE
+      )
+@
+
 \begin{figure}

### 2017-02-13 manuscript/manuscript.Rnw
discuss TSI in manuscript; move enrichment table

@@
 \cref{fig:core_housekeeping_transcriptome_shape}), and includes more
-of the total fraction of genes. However, there is still a large
-difference between the set of housekeeping genes which are expressed
-in \human and those which are expressed in all placentas.
+of the total fraction of genes. An alternative approach to defining
+housekeeping genes is to use the tissue specificity
+index~\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}.
+Supplemental \cref{tab:tissue_specificity_index_of_plaenta_core_genes}
+lists the tissue specificity index for all housekeeping and
+non-housekeeping genes; housekeeping genes expressed in all placentas
+have a lesser median tissue specificity index than non-housekeeping
+genes (Supplemental
+\cref{fig:tissue_specificity_index_with_housekeeping}).
 

### 2017-02-13 manuscript/manuscript.Rnw
Describe information on specimen ID, etc

@@
 \carmeli [blind mole rat from northern Israel]
-\cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}). Information
-on specimen ID, and collector is shown for each sample in Supplemental
+\cite{Nevo2001:adaptiveradiationofblindsubterraneanmole};
+n=1 for all collected species). Information
+on specimen ID, number of specimens, and collector is shown for each sample in Supplemental
 \cref{tab:srr_numbers}. In the case of \bonobo, \cow, and \elephant,
@@
 
-\subsubsection{Additional Sequences}
-
-For more complete evolutionary comparisons additional sequences of
+\subsubsection{Additional Placenta Transcriptomes}
+
+For more complete evolutionary comparisons additional transcriptomes of
 placenta in \mouse [mouse]

### 2017-02-15 manuscript/manuscript.Rnw
Indicate genes assoc. with adverse pregnancy
Show which genes are associated with adverse pregnancy outcomes using
literature search from pubmed and link to those results in Table 1.

@@
 
+To identify genes which have evidence of association with adverse
+pregnancy outcomes and other placenta-relevant terms of interest, we
+used the “pubmed_to_grid” component of
+Function2Gene~\cite{Armstrong2008:function2gene} which conducts a term
+by gene search of PubMed results.
+
 \subsection{Sequences, FPKMs, and Code}
@@
 
+literature.mining <-
+    data.table(read.table("../literature_mining/human_lineage_vs_terms.txt",
+                          sep="\t",skip=0,stringsAsFactors=FALSE,
+                          check.names=FALSE,header=TRUE),
+               keep.rownames=TRUE)
+setnames(literature.mining,"rn","human_name")
+setkey(literature.mining,"human_name")
+literature.mining[,`Adverse Pregnancy Outcomes`:=`preeclampsia`+
+                       IUGR+`placenta previa`+`gestational diabetes`+
+                       `premature birth`,
+                  by=human_name]
+literature.mining.terms <-
+    paste0("%28%22preeclampsia%22[All%20Fields]%20OR%20",
+           "%22pre-eclampsia%22[All%20Fields]%20OR%20",
+           "%22eclampsia%22[All%20Fields]%20OR%20",
+           "%22IUGR%22[All%20Fields]%20OR%20",
+           "%22iugr%22[All%20Fields]%20OR%20",
+           "%22intrauterine%20growth%20restriction%22[All%20Fields]%20OR%20",
+           "%22placenta%20previa%22[All%20Fields]%20OR%20",
+           "%22gestational%20diabetes%22[All%20Fields]%20OR%20",
+           "%22premature%20birth%22[All%20Fields]%20OR%20",
+           "%22premature%20labor%22[All%20Fields]%29")
 placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=5,orderbyfc=FALSE,fdr.cutoff=1) {
@@
     temp.table <-
+        literature.mining[,list(human_name,`Adverse Pregnancy Outcomes`)
+                          ][temp.table]
+    temp.table[,
+               `Adv. Preg. Publications`:="0"
+               ]
+    temp.table[`Adverse Pregnancy Outcomes` >= 2,
+               `Adv. Preg. Publications`:="≥ 2"
+               ]
+    temp.table[`Adverse Pregnancy Outcomes` >= 5,
+               `Adv. Preg. Publications`:="≥ 5"
+               ]
+    temp.table[`Adverse Pregnancy Outcomes` >= 20,
+               `Adv. Preg. Publications`:="≥ 20"
+               ]
+    temp.table[`Adverse Pregnancy Outcomes` == 1,
+               `Adv. Preg. Publications`:="1"
+               ]
+    setkey(temp.table,"human_name")
+    temp.table <- 
         hugo.nomenclature[temp.table][,list(`Approved Symbol`,
-                                            `Approved Name`,p.value,fdr,fc)]
+                                            `Approved Name`,p.value,fdr,fc,
+                                            `Adv. Preg. Publications`
+                                            )]
     setnames(temp.table,
-              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
+             c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change",
+               "Adv. Preg. Pub."
+               )
               )
@@
         display.table <-
-            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
-                                                  "}")]
+            display.table[,`Adv. Preg. Pub.`:=
+                               paste0("\\href{https://www.ncbi.nlm.nih.gov/pubmed/?term=%22",
+                                      `Human Symbol`,
+                                      "%22[All%20Fields]%20AND%20",
+                                      literature.mining.terms,"}{",
+                                      `Adv. Preg. Pub.`,"}"
+                                      )]
+        display.table <-
+            display.table[,`Human Symbol`:=
+                               paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",
+                                      `Human Symbol`,"}{",
+                                      "\\textit{",xtable.sanitize.text(`Human Symbol`),
+                                      "}}")]
         display.table[,`Human Name`:=NULL]
-        # setnames(display.table,"Human Symbol","\\multicolumn{1}{p{0.5in}}{Human Symbol}")6
+        # setnames(display.table,"Human Symbol","\\multicolumn{1}{p{0.5in}}{Human Symbol}")
         setnames(display.table,"Fold Change","{$\\text{log}_2$ FC}")
@@
         print(xtable(display.table,
-                     display=c("s","s","e","e","f"),
-                     align=c("c","p{0.5in}","S[table-format=1.2e2]","S[table-format=1.2e2]","S[table-format=2.2]")
+                     display=c("s","s","e","e","f","s"),
+                     align=c("c","p{0.5in}","S[table-format=1.2e2]","S[table-format=1.2e2]","S[table-format=2.2]","p{0.5in}")
                      ),
@@
 @ 
-\caption{\lineagetableexplanationtypeiiishort{\human}{all other analyzed species}}
+\caption{\lineagetableexplanationtypeiiishort{\human}{all other
+    analyzed species} Adv. Preg. Pub. is the approximate number of
+  publications indexed by PubMed mentioning this gene and an adverse
+  pregnancy outcome (pre-eclampsia, premature labor, gestational
+  diabetes, intrauterine growth restriction, or premature birth) and
+  links to the most current PubMed search results.}
 \label{tab:human_lineage_specific}

### 2017-02-15 manuscript/manuscript.Rnw
bold core functionality
discuss orthologies by species type

@@
 placentas of all eutherian species analyzed and therefore represent the
-\emph{core functionality of the placenta}; 2) identify genes whose
+\textbf{core functionality of the placenta}; 2) identify genes whose
 expression level changes significantly in each of the morphology
@@
 
-Single-copy (1-to-1) orthologs in were determined using
-Ensembl v.80 gene trees via Biomart using the biomaRt R package
+Single-copy (1-to-1) orthologs in were determined using Ensembl v.80
+gene trees via Biomart using the biomaRt R package
 \cite{Durinck.ea2009:Mappingidentifiersforintegration}. %
 A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
-  1,length(unique(human_name))]} single-copy orthologous genes were expressed
-in at least one species.
+  1,length(unique(human_name))]} single-copy orthologous genes were
+expressed in at least one species. We also obtained orthology from the
+perspective of each species to all other species using biomaRt package
+and Ensembl v.80 database to determine the percentage of highly
+expressed genes which had an orthology of one-to-one, one-to-many,
+many-to-many, or novel/unknown.
 
@@
 significant predictor of unmapped read percentage (Supplemental
-\cref{fig:transcriptome_reads_vs_read_length}). The expression of
+\cref{fig:transcriptome_reads_vs_read_length}). %
+%
+Between 20\% and 40\% of all highly expressed genes in each species
+were single-copy orthologs with all other analyzed species, with the
+remainder being mostly one-to-many or many-to-many in comparison to at
+least one other species (Supplemental
+\cref{tab:orthologs_in_each_species}).
+%
+The expression of
 all genes with \ac{FPKM} $\ge 50$ is shown in Supplemental

### 2017-02-15 manuscript/supplemental_material.Rnw
discuss TSI in manuscript; move enrichment table
discuss orthologies by species type

@@
 
+\subsection{Orthology types in each Species}
+
+\begin{table}[h]
+  \footnotesize
+<<orthologs_in_each_species,echo=FALSE,results="asis">>=
+load("../data/orthologs_by_species_ranking")
+orthologs.by.species[,percentage:=ratio*100]
+orthologs.by.species[,orthology_type:=
+                          factor(orthology_type,
+                                 levels=c("one2one",
+                                          "one2many",
+                                          "many2many",
+                                          "novel/unknown"))]
+orthology.wide.table <-
+    data.table(dcast(orthologs.by.species,
+                     type+orthology_type~species,
+                     value.var="percentage"))
+setnames(orthology.wide.table,
+         "type","Cutoff")
+orthology.wide.table[,Cutoff:=xtable.sanitize.text(Cutoff)]
+orthology.wide.table[,orthology_type:=xtable.sanitize.text(orthology_type)]
+setnames(orthology.wide.table,
+         "orthology_type","Orthology Type")
+setnames(orthology.wide.table,
+         colnames(orthology.wide.table)[-(1:2)],
+         paste0("\\textit{",
+                gsub("^(.)","\\U\\1",
+                     colnames(orthology.wide.table)[-(1:2)],
+                     perl=TRUE),
+                "}"))
+print(xtable(orthology.wide.table,
+             digits=0
+             ),
+      rotate.colnames=TRUE,
+      booktabs=TRUE,
+      floating=FALSE,
+      include.rownames=FALSE,
+      sanitize.text.function=genes.table.sanitize
+      )
+@ 
+\caption{Percentage of orthology in the top 100 and top 1\% of
+  placenta transcripts as identified by Ensembl. If a gene has the
+  same orthology in comparison to all sequenced species, it is
+  reported as that orthology; if it had an orthology of many2many in
+  any species, it is reported as many2many; if it had an orthology of
+  one2many in any species, but no many2many, it is reported as
+  one2many. Novel/unknown genes had no reported orthologies to any
+  sequenced species.}
+\label{tab:orthologs_in_each_species}
+\end{table}
+
+\clearpage
+
 <<genes.expression,echo=FALSE,results="hide">>=
@@
 \section{Core Placenta Transcriptome}
-
-\begin{table}
-\footnotesize
-<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis",cache.extra=tools::md5sum(paste0("../data/placenta_go_results_",c("bp","mf","permissive_bp","permissive_mf")))>>=
-load("../data/placenta_go_results_mf")
-load("../data/placenta_go_results_bp")
-load("../data/placenta_go_results_permissive_mf")
-load("../data/placenta_go_results_permissive_bp")
-
-placenta.go.results <- rbindlist(list(head(n=20,placenta_go_results_mf[classic.fdr < 0.05,]),
-                             head(n=20,placenta_go_results_bp[classic.fdr < 0.05,])))
-placenta.go.results <- placenta.go.results[,Annotated:=NULL]
-setnames(placenta.go.results,"Significant","Core")
-setnames(placenta.go.results,"classic","p value")
-setnames(placenta.go.results,"classic.fdr","FDR")
-setnames(placenta.go.results,"GO.ID","GO ID")
-setnames(placenta.go.results,"Term","GO Term")
-
-placenta.go.results.permissive <-
-    rbindlist(list(placenta_go_results_permissive_mf,
-                   placenta_go_results_permissive_bp))
-setnames(placenta.go.results.permissive,"Significant","Core")
-setnames(placenta.go.results.permissive,"classic","p value")
-setnames(placenta.go.results.permissive,"classic.fdr","Permissive FDR")
-setnames(placenta.go.results.permissive,"GO.ID","GO ID")
-setnames(placenta.go.results.permissive,"Term","GO Term")
-setkey(placenta.go.results,"GO ID")
-setkey(placenta.go.results.permissive,"GO ID")
-
-combined.placenta.go.results <-
-    placenta.go.results.permissive[,list(`GO ID`,`Permissive FDR`)
-                                   ][placenta.go.results][order(Ontology,FDR),]
-combined.placenta.go.results <-
-    combined.placenta.go.results[,`p value`:=as.numeric(`p value`)]
-setcolorder(combined.placenta.go.results,
-            c("Ontology",
-              "GO ID",
-              "GO Term",
-              "Core",
-              "Expected",
-              "p value",
-              "FDR",
-              "Permissive FDR"))
-combined.placenta.go.results <-
-    combined.placenta.go.results[,`p value`:=NULL]
-setnames(combined.placenta.go.results,"Ontology","Ont.")
-setnames(combined.placenta.go.results,"Expected","Exp.")
-print(xtable(combined.placenta.go.results,
-             digits=2,
-             display=c("d","s","s","s","d","f","e","e")),
-      include.rownames=FALSE,
-      booktabs=TRUE,
-      math.style.exponents=TRUE,
-      floating=FALSE
-      )
-@ 
-\caption[Core placenta GO Enrichment]{GO terms most significantly enriched
-  (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]} MF,
-  \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP) in
-  the core placenta transcriptome ($\text{FPKM}>10$ in 14
-  of 14 species) as compared to all genes with 1:1 human orthologs in
-  all species. “Ont.” is the GO Ontology, where MF is molecular function 
-  and BP is biological process; “Core” is the number of genes in the core
-  transcriptome for each GO term; “Exp.” is the number of genes
-  expected in the core transcriptome for each term; 
-  “FDR” is the false discovery rate as estimated by the
-  Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure} based on
-  p values calculated by the Fisher Exact Test; 
-  “Permissive FDR” is the FDR resulting from the equivalent GO
-  analysis using the permissive core placenta transcriptome 
-  (1:1 orthologs with expression of at least 10 FPKM at least 10 of 14 species) 
-  showing that the enriched GO terms are relatively insensitive to the precise
-  set of species included.}
-\label{tab:core_placenta_go_terms}
-\end{table}
-
-\clearpage
 
@@
 \begin{table}
+\footnotesize
+<<go.analysis,warning=FALSE,message=FALSE,echo=FALSE,results="asis",cache.extra=tools::md5sum(paste0("../data/placenta_go_results_",c("bp","mf","permissive_bp","permissive_mf")))>>=
+load("../data/placenta_go_results_mf")
+load("../data/placenta_go_results_bp")
+load("../data/placenta_go_results_permissive_mf")
+load("../data/placenta_go_results_permissive_bp")
+
+placenta.go.results <- rbindlist(list(head(n=20,placenta_go_results_mf[classic.fdr < 0.05,]),
+                             head(n=20,placenta_go_results_bp[classic.fdr < 0.05,])))
+placenta.go.results <- placenta.go.results[,Annotated:=NULL]
+setnames(placenta.go.results,"Significant","Core")
+setnames(placenta.go.results,"classic","p value")
+setnames(placenta.go.results,"classic.fdr","FDR")
+setnames(placenta.go.results,"GO.ID","GO ID")
+setnames(placenta.go.results,"Term","GO Term")
+
+placenta.go.results.permissive <-
+    rbindlist(list(placenta_go_results_permissive_mf,
+                   placenta_go_results_permissive_bp))
+setnames(placenta.go.results.permissive,"Significant","Core")
+setnames(placenta.go.results.permissive,"classic","p value")
+setnames(placenta.go.results.permissive,"classic.fdr","Permissive FDR")
+setnames(placenta.go.results.permissive,"GO.ID","GO ID")
+setnames(placenta.go.results.permissive,"Term","GO Term")
+setkey(placenta.go.results,"GO ID")
+setkey(placenta.go.results.permissive,"GO ID")
+
+combined.placenta.go.results <-
+    placenta.go.results.permissive[,list(`GO ID`,`Permissive FDR`)
+                                   ][placenta.go.results][order(Ontology,FDR),]
+combined.placenta.go.results <-
+    combined.placenta.go.results[,`p value`:=as.numeric(`p value`)]
+setcolorder(combined.placenta.go.results,
+            c("Ontology",
+              "GO ID",
+              "GO Term",
+              "Core",
+              "Expected",
+              "p value",
+              "FDR",
+              "Permissive FDR"))
+combined.placenta.go.results <-
+    combined.placenta.go.results[,`p value`:=NULL]
+setnames(combined.placenta.go.results,"Ontology","Ont.")
+setnames(combined.placenta.go.results,"Expected","Exp.")
+print(xtable(combined.placenta.go.results,
+             digits=2,
+             display=c("d","s","s","s","d","f","e","e")),
+      include.rownames=FALSE,
+      booktabs=TRUE,
+      math.style.exponents=TRUE,
+      floating=FALSE
+      )
+@ 
+\caption[Core placenta GO Enrichment]{GO terms most significantly enriched
+  (\Sexpr{placenta.go.results[,sum(Ontology=="MF")]} MF,
+  \Sexpr{placenta.go.results[,sum(Ontology=="BP")]} BP) in
+  the core placenta transcriptome ($\text{FPKM}>10$ in 14
+  of 14 species) as compared to all genes with 1:1 human orthologs in
+  all species. “Ont.” is the GO Ontology, where MF is molecular function 
+  and BP is biological process; “Core” is the number of genes in the core
+  transcriptome for each GO term; “Exp.” is the number of genes
+  expected in the core transcriptome for each term; 
+  “FDR” is the false discovery rate as estimated by the
+  Benjamini and Hochberg procedure \cite{Benjamini1995:bh_procedure} based on
+  p values calculated by the Fisher Exact Test; 
+  “Permissive FDR” is the FDR resulting from the equivalent GO
+  analysis using the permissive core placenta transcriptome 
+  (1:1 orthologs with expression of at least 10 FPKM at least 10 of 14 species) 
+  showing that the enriched GO terms are relatively insensitive to the precise
+  set of species included.}
+\label{tab:core_placenta_go_terms}
+\end{table}
+
+\clearpage
+
+\begin{table}
 <<pathway_analysis,echo=FALSE,results="asis",cache.extra=file.info(dir("../data","placenta_gsea.*",full.names=TRUE))[,"mtime"]>>=

### 2017-02-15 data/ensembl_all_ortholog_types.R

@@ -1,0 +1,34 @@
+library("data.table")
+
+
+load("ensembl_all_orthologs")
+
+resolve_orthology <- function(orthology_type) {
+    orthology_type <- gsub("ortholog_","",
+                           orthology_type)
+    if (all(orthology_type==orthology_type[1])) {
+        return(orthology_type[1])
+    }
+    if (any(orthology_type=="many2many")) {
+        return("many2many")
+    }
+    if (any(orthology_type=="one2many")) {
+        return("one2many")
+    }
+    return(NA)
+}
+
+
+ortholog.types <-
+    orthologs[,list(species=primary[1],
+                    orthology_type=resolve_orthology(homolog_orthology_type)),
+              by=ensembl_gene_id]
+setnames(ortholog.types,
+         "ensembl_gene_id",
+         "gene_id")
+setkey(ortholog.types,
+       "gene_id")
+
+save(file="ensembl_all_ortholog_types",
+     ortholog.types
+     )

### 2017-02-15 data/ensembl_all_orthologs.R

@@ -1,0 +1,62 @@
+library(biomaRt)
+library(data.table)
+
+args <- commandArgs(trailingOnly=TRUE)
+
+species <-
+    c("hsapiens",
+      "dnovemcinctus",
+      "ptroglodytes",
+      "btaurus",
+      "cfamiliaris",
+      "lafricana",
+      "mmusculus",
+      "mdomestica",
+      "ecaballus",
+      "sscrofa",
+      "oaries"
+      )
+
+orthologs <- list()
+
+species.permute <- expand.grid(species,species)
+
+for (i in 1:nrow(species.permute)) {
+    main.species <- species.permute[i,"Var1"]
+    secondary.species <- species.permute[i,"Var2"]
+    if (secondary.species == main.species) {
+        next;
+    }
+    mart <- useMart("ENSEMBL_MART_ENSEMBL",
+                    dataset=paste0(main.species,"_gene_ensembl"),
+                    host="may2015.archive.ensembl.org")
+
+    orth.name <- paste0(main.species,":",
+                        secondary.species)
+    orths <-
+        data.table(getBM(c("ensembl_gene_id",
+                           paste0(secondary.species,
+                                  paste0("_homolog_",
+                                         c("ensembl_gene",
+                                           "orthology_type"))
+                                  )),
+                         values =TRUE,
+                         mart = mart,
+                         bmHeader=FALSE))
+    setnames(orths,
+             paste0(secondary.species,
+                                  paste0("_homolog_",
+                                         c("ensembl_gene",
+                                           "orthology_type"))
+                    ),
+             c("homolog_ensembl_gene_id","homolog_orthology_type"))
+    orths <- orths[homolog_ensembl_gene_id!="",]
+    orths[,primary:=main.species]
+    orths[,secondary:=secondary.species]
+    orthologs[[orth.name]] <- orths
+}
+orthologs <-
+    rbindlist(orthologs)
+
+save(orthologs,
+     file=args[length(args)])

### 2017-02-15 data/orthologs_by_species_ranking.R

@@ -1,0 +1,61 @@
+library("data.table")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+## combined_fpkm
+load(args[1])
+
+## ensembl_all_ortholog_types
+load(args[2])
+
+## cuts -- top 100, top 1%
+
+table_orthologs <- function(this.species,top.100=TRUE) {
+    top.results <-
+        combined.fpkm[species==this.species &
+                      !is.na(gene_id),][order(-FPKM)]
+    if (top.100) {
+        top.results <-
+            top.results[1:100,]
+    } else {
+        top.results <-
+            top.results[1:ceiling(nrow(top.results)*0.01),]
+    }
+    setkey(top.results,"gene_id")
+    species.results <-
+        merge(top.results,
+              ortholog.types,
+              all.x=TRUE,
+              all.y=FALSE,
+              by.x="gene_id",
+              by.y="gene_id")[,table(orthology_type,useNA="ifany")]
+    species.results <-
+        data.table(species.results)
+    species.results[is.na(orthology_type),orthology_type:="novel/unknown"]
+    species.results[,species:=this.species]
+    if (top.100) {
+        species.results[,type:="Top 100"]
+    } else {
+        species.results[,type:="Top 1%"]
+    }
+    species.results
+}
+
+
+orthologs.by.species <- list()
+
+for (specie in combined.fpkm[,as.character(unique(species))]) {
+    orthologs.by.species[[specie]] <-
+        table_orthologs(specie)
+    orthologs.by.species[[paste0(specie,"2")]] <-
+        table_orthologs(specie,top.100=FALSE)
+
+}
+
+orthologs.by.species <-
+    rbindlist(orthologs.by.species)
+orthologs.by.species[,ratio:=N/sum(N),by=list(species,type)]
+
+save(orthologs.by.species,
+     file=args[length(args)])
+

### 2017-02-16 manuscript/common_latex.tex
use the long names everywhere
add a second type III explanation for all genes
switch from 1:1 to single-copy ortholog
Include differential expression in non-humans
Include four additional tables in the supplemental material which
shows the differential expression of single-copy orthologs in each of
the species.

@@
 % species names
-\newcommand{\smonkey}[0]{\textit{A. fusciceps}\xspace}
-\newcommand{\cow}[0]{\textit{B. taurus}\xspace}
-\newcommand{\dog}[0]{\textit{C. familiaris}\xspace}
-\newcommand{\armadillo}[0]{\textit{D. novemcinctus}\xspace}
-\newcommand{\horse}[0]{\textit{E. caballus}\xspace}
-\newcommand{\elephant}[0]{\textit{L. africana}\xspace}
-\newcommand{\human}[0]{\textit{H. sapiens}\xspace}
-\newcommand{\opossum}[0]{\textit{M. domestica}\xspace}
-\newcommand{\mouse}[0]{\textit{M. musculus}\xspace}
-\newcommand{\galili}[0]{\textit{S. galili}\xspace}
-\newcommand{\sheep}[0]{\textit{O. aries}\xspace}
-\newcommand{\bonobo}[0]{\textit{P. paniscus}\xspace}
-\newcommand{\carmeli}[0]{\textit{S. carmeli}\xspace}
-\newcommand{\pig}[0]{\textit{S. scrofa}\xspace}
+\newcommand{\smonkey}[0]{\textit{Ateles fusciceps}\xspace}
+\newcommand{\cow}[0]{\textit{Bos taurus}\xspace}
+\newcommand{\dog}[0]{\textit{Canis familiaris}\xspace}
+\newcommand{\armadillo}[0]{\textit{Dasypus novemcinctus}\xspace}
+\newcommand{\horse}[0]{\textit{Equus caballus}\xspace}
+\newcommand{\elephant}[0]{\textit{Loxodonta africana}\xspace}
+\newcommand{\human}[0]{\textit{Homo sapiens}\xspace}
+\newcommand{\opossum}[0]{\textit{Monodelphis domestica}\xspace}
+\newcommand{\mouse}[0]{\textit{Mus musculus}\xspace}
+\newcommand{\galili}[0]{\textit{Spalax galili}\xspace}
+\newcommand{\sheep}[0]{\textit{Ovis aries}\xspace}
+\newcommand{\bonobo}[0]{\textit{Pan paniscus}\xspace}
+\newcommand{\carmeli}[0]{\textit{Spalax carmeli}\xspace}
+\newcommand{\pig}[0]{\textit{Sus scrofa}\xspace}
 
@@
   ortholog is a housekeeping gene. Gene symbol is the gene symbol or
-  Ensembl gene id. Human Symbol is the 1:1 human ortholog or the best
+  Ensembl gene id. Human Symbol is the single-copy \human ortholog or the best
   human alignment when enclosed in parenthesis. Human Name is the HUGO
@@
   $\mathrm{log}_2$ FC is the base-2 logarithm of the fold change from
-  #2 to #3 .}
+  #2 to #3.}
 
 \newcommand{\lineagetableexplanationtypeiii}[2]{%
+  \lineagetableexplanation[A nested type-III anova was used to
+  calculate p values.]{#1}{#2}}
+
+\newcommand{\lineagetableexplanationtypeiiishort}[2]{%
   \lineagetableexplanation[A nested type-III anova was used to

### 2017-02-16 manuscript/manuscript.Rnw
Include differential expression in non-humans
Include four additional tables in the supplemental material which
shows the differential expression of single-copy orthologs in each of
the species.

@@
 
-<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(paste0("../data/",c("placenta_lineage_specific","human_mouse_lineage_specific")))>>=
+<<load.placenta.lineage.specific,echo=FALSE,results="hide",cache.extra=tools::md5sum(c(paste0("../data/",c("placenta_lineage_specific","human_mouse_lineage_specific")),"../literature_mining/human_lineage_vs_terms.txt"))>>=
 load("../data/placenta_lineage_specific")
@@
 Using a type-III ANOVA to account for repeated testing of
-\human and \mouse (see \cref{sec:comp-analys-gene} for details), we identified multiple genes which are
+\human (see \cref{sec:comp-analys-gene} for details), we identified multiple genes which are
 differentially expressed in the \human lineage in comparison to all
@@
 \cite{Spencer.ea2008:ADAM12sinmaternalserum}.
+
+\subsection{Other Lineage and Clade Specific Genes}
+
+We also compared differences in the \mouse lineage in comparison to
+all other lineages using a type-III ANOVA as in the \human analysis,
+and found few genes which are highly differentially expressed and are
+also associated with adverse pregnancy outcomes. Comparisons between
+rodentia (\mouse, \galili, and \carmeli) and all other species using
+the mean of each species expression did not indicate any significant
+differences (Supplemental \cref{tab:rodentia_lineage_specific}) nor
+did the equivalent comparisons between laurasiatheria and all other
+species (Supplemental \cref{tab:laurasiatheria_lineage_specific}) and
+laurasiatheria and euarchontoglires (Supplemental
+\cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).
+Additional placenta samples of each species and the collection of the
+placentas of additional laurasiatherians may resolve the lack of
+significance due to insufficient power.
 

### 2017-02-16 manuscript/supplemental_material.Rnw
Include differential expression in non-humans
Include four additional tables in the supplemental material which
shows the differential expression of single-copy orthologs in each of
the species.

@@
 
-placenta.lineage.specific.table <- function(analysis,output=TRUE,lineage.table=placenta.lineage.specific.analysis,members=40,orderbyfc=FALSE,fdr.cutoff=1,caption=NULL,label=NULL) {
+placenta.lineage.specific.table <- function(analysis,output=TRUE,
+                                            lineage.table=placenta.lineage.specific.analysis,
+                                            lit.mining=NULL,
+                                            lit.mining.terms=NULL,
+                                            members=40,
+                                            orderbyfc=FALSE,
+                                            fdr.cutoff=1,caption=NULL,
+                                            label=NULL) {
     temp.table <- 
@@
     setkey(temp.table,"human_name")
+    if (!is.null(lit.mining)) {
         temp.table <-
-        hugo.nomenclature[temp.table][,list(`Approved Symbol`,
-                                            `Approved Name`,p.value,fdr,fc)]
+            lit.mining[,list(human_name,`Adverse Pregnancy Outcomes`)
+                       ][temp.table]
+        temp.table[,
+                   `Adv. Preg. Pub.`:="0"
+                   ]
+        temp.table[`Adverse Pregnancy Outcomes` >= 2,
+               `Adv. Preg. Pub.`:="≥ 2"
+               ]
+        temp.table[`Adverse Pregnancy Outcomes` >= 5,
+                   `Adv. Preg. Pub.`:="≥ 5"
+                   ]
+        temp.table[`Adverse Pregnancy Outcomes` >= 20,
+                   `Adv. Preg. Pub.`:="≥ 20"
+               ]
+        temp.table[`Adverse Pregnancy Outcomes` == 1,
+                   `Adv. Preg. Pub.`:="1"
+                   ]
+        temp.table[,`Adverse Pregnancy Outcomes`:=NULL]
+    }
+    keep.cols <-
+        c("Approved Symbol","Approved Name","p.value","fdr","fc")
+    if (!is.null(lit.mining)) {
+        keep.cols <- c(keep.cols,"Adv. Preg. Pub.")
+    }
+    temp.table <- 
+        hugo.nomenclature[temp.table][,keep.cols,with=FALSE]
     setnames(temp.table,
+             c("Approved Symbol","Approved Name","p.value","fdr","fc"),
              c("Human Symbol","Human Name","p (t-test)","FDR","Fold Change")
@@
         display.table <- temp.table[`FDR`<=fdr.cutoff,][order(`p (t-test)`),]
+        col.disp <-
+            c("s","s","e","g","f")
+        col.align <-
+            c("c",">{\\centering\\arraybackslash}p{1in}","r","r","r")
+        if (!is.null(lit.mining)) {
+            col.disp <- c(col.disp,"s")
+            col.align <- c(col.align,">{\\centering\\arraybackslash}p{0.75in}")
+        }
         if (orderbyfc) {
@@
         }
-
+        if (!is.null(lit.mining)) {
             display.table <-
-            display.table[,`Human Symbol`:=paste0("\\textit{",xtable.sanitize.text(`Human Symbol`),
-                                                  "}")]
+                display.table[,`Adv. Preg. Pub.`:=
+                                   paste0("\\href{https://www.ncbi.nlm.nih.gov/pubmed/?term=%22",
+                                          `Human Symbol`,
+                                          "%22[All%20Fields]%20AND%20",
+                                          literature.mining.terms,"}{",
+                                          `Adv. Preg. Pub.`,"}"
+                                          )]
+        }
+
+        display.table <-
+            display.table[,`Human Symbol`:=
+                               paste0("\\href{http://www.ensembl.org/Homo_sapiens/Gene/Summary?g=",
+                                      `Human Symbol`,"}{",
+                                      "\\textit{",xtable.sanitize.text(`Human Symbol`),
+                                      "}}")]
         display.table[,`Human Name`:=NULL]
@@
         print(xtable(display.table,
-                     display=c("s","s","e","g","f"),
+                     display=col.disp,
                      caption=caption,
                      label=label,
-                     align=c("c","p{0.7in}","r","r","r")
+                     align=col.align
                      ),
@@
 
-<<human.lineage.specific,echo=FALSE,results="asis">>=
+\subsection{\human specific expression}
+
+<<human.lineage.specific,echo=FALSE,results="asis",cache.extra=file.info("../literature_mining/human_lineage_vs_terms.txt")[,"mtime"]>>=
+literature.mining <-
+    data.table(read.table("../literature_mining/human_lineage_vs_terms.txt",
+                          sep="\t",skip=0,stringsAsFactors=FALSE,
+                          check.names=FALSE,header=TRUE),
+               keep.rownames=TRUE)
+setnames(literature.mining,"rn","human_name")
+setkey(literature.mining,"human_name")
+literature.mining[,`Adverse Pregnancy Outcomes`:=`preeclampsia`+
+                       IUGR+`placenta previa`+`gestational diabetes`+
+                       `premature birth`,
+                  by=human_name]
+literature.mining.terms <-
+    paste0("%28%22preeclampsia%22[All%20Fields]%20OR%20",
+           "%22pre-eclampsia%22[All%20Fields]%20OR%20",
+           "%22eclampsia%22[All%20Fields]%20OR%20",
+           "%22IUGR%22[All%20Fields]%20OR%20",
+           "%22iugr%22[All%20Fields]%20OR%20",
+           "%22intrauterine%20growth%20restriction%22[All%20Fields]%20OR%20",
+           "%22placenta%20previa%22[All%20Fields]%20OR%20",
+           "%22gestational%20diabetes%22[All%20Fields]%20OR%20",
+           "%22premature%20birth%22[All%20Fields]%20OR%20",
+           "%22premature%20labor%22[All%20Fields]%29")
+
 hum.all.lin.spec <-
@@
                                     lineage.table=human.mouse.lineage.specific.analysis,
+                                    lit.mining=literature.mining,
+                                    lit.mining.terms=literature.mining.terms,
                                     members=NA,
+                                    fdr.cutoff=0.05,
+                                    orderbyfc=TRUE,
                                     caption=c("\\lineagetableexplanationtypeiii{\\human}{all other species}","\\human lineage specific genes"),
@@
 @
+
+\clearpage
+
+\subsection{\mouse specific expression}
+
+<<mouse.lineage.specific,echo=FALSE,results="asis",cache.extra=file.info("../literature_mining/mouse_lineage_vs_terms.txt")[,"mtime"]>>=
+mouse.literature.mining <-
+    data.table(read.table("../literature_mining/mouse_lineage_vs_terms.txt",
+                          sep="\t",skip=0,stringsAsFactors=FALSE,
+                          check.names=FALSE,header=TRUE),
+               keep.rownames=TRUE)
+setnames(mouse.literature.mining,"rn","human_name")
+setkey(mouse.literature.mining,"human_name")
+mouse.literature.mining[,`Adverse Pregnancy Outcomes`:=`preeclampsia`+
+                             IUGR+`placenta previa`+`gestational diabetes`+
+                             `premature birth`,
+                        by=human_name]
+mou.all.lin.spec <-
+    placenta.lineage.specific.table("mou.all",
+                                    lineage.table=human.mouse.lineage.specific.analysis,
+                                    lit.mining=mouse.literature.mining,
+                                    lit.mining.terms=literature.mining.terms,
+                                    fdr.cutoff=0.05,
+                                    members=NA,
+                                    orderbyfc=TRUE,
+                                    caption=c("\\lineagetableexplanationtypeiii{\\mouse}{all other species}","\\mouse lineage specific genes"),
+                                    label="tab:mouse_lineage_specific_all"
+                                    )
+setkey(mou.all.lin.spec,"Human Symbol")
+@
+
+\clearpage
+
+\subsection{Rodentia, Laurasiatheria, and Euarchontoglires Comparisons}
+
+\begin{table}[hb]
+  \footnotesize
+<<rodentia.lineage.specific,echo=FALSE,results="asis">>=
+rod.all.lin.spec <- placenta.lineage.specific.table("rod.all",fdr.cutoff=0.5)
+setkey(rod.all.lin.spec,"Human Symbol")
+@ 
+\caption{\lineagetableexplanation{the rodentia order}{all other species}}
+\label{tab:rodentia_lineage_specific}
+\end{table}
+
+\begin{table}[hb]
+  \footnotesize
+<<lau.lineage.specific,echo=FALSE,results="asis">>=
+lau.all.lin.spec <- placenta.lineage.specific.table("lau.all",fdr.cutoff=0.5)
+setkey(lau.all.lin.spec,"Human Symbol")
+@ 
+\caption{\lineagetableexplanation{the laurasiatheria
+    superorder}{all other species}}
+\label{tab:laurasiatheria_lineage_specific}
+\end{table}
+
+\begin{table}[hb]
+  \footnotesize
+<<eua.lau.lineage.specific,echo=FALSE,results="asis">>=
+eua.lau.lin.spec <- placenta.lineage.specific.table("eua.lau",fdr.cutoff=1)
+setkey(eua.lau.lin.spec,"Human Symbol")
+@ 
+\caption{\lineagetableexplanation{ the euarchontoglires superorder}{ the laurasiatherian superorder}}
+\label{tab:laurasiatheria_euarchontoglires_lineage_specific}
+\end{table}
 

### 2017-02-16 data/placenta_lineage_specific.R

@@ -60,7 +60,7 @@ calculate.t.test <- function(x,g1,g2){
 
 calculate.fc <- function(x,g1,g2){
     fc <- NA
-    try({fc <- mean(as.numeric(x[g1]))/mean(as.numeric(x[g2]))
+    try({fc <- log2(mean(as.numeric(x[g1]))/mean(as.numeric(x[g2])))
         if (any(!is.finite(fc))) {
             fc <- log2(max(c(mean(as.numeric(x[g1])),mean(as.numeric(x[g2])),na.rm=TRUE)))
         }},

### 2017-02-20 data/ensembl_gene_trees.R
add ensembl gene trees rule

@@ -1,0 +1,54 @@
+library(biomaRt)
+library(data.table)
+
+args <- commandArgs(trailingOnly=TRUE)
+
+species <-
+    c("hsapiens",
+      "dnovemcinctus",
+      "ptroglodytes",
+      "btaurus",
+      "cfamiliaris",
+      "lafricana",
+      "mmusculus",
+      "mdomestica",
+      "ecaballus",
+      "sscrofa",
+      "oaries"
+      )
+
+orthologs <- list()
+
+for (i in 1:length(species)) {
+    main.species <- species[i]
+    mart <- useMart("ENSEMBL_MART_ENSEMBL",
+                    dataset=paste0(main.species,"_gene_ensembl"),
+                    host="may2015.archive.ensembl.org")
+
+    orths <-
+        data.table(getBM(c("ensembl_gene_id",
+                           paste0(secondary.species,
+                                  paste0("_homolog_",
+                                         c("ensembl_gene",
+                                           "orthology_type"))
+                                  )),
+                         values =TRUE,
+                         mart = mart,
+                         bmHeader=FALSE))
+    setnames(orths,
+             paste0(secondary.species,
+                                  paste0("_homolog_",
+                                         c("ensembl_gene",
+                                           "orthology_type"))
+                    ),
+             c("homolog_ensembl_gene_id","homolog_orthology_type"))
+    orths <- orths[homolog_ensembl_gene_id!="",]
+    orths[,primary:=main.species]
+    orths[,secondary:=secondary.species]
+    orthologs[[orth.name]] <- orths
+}
+orthologs <-
+    rbindlist(orthologs)
+
+save(orthologs,
+     file=args[length(args)])

### 2017-02-20 data/placenta_core_tree_transcriptome.R
calculate placenta core tree transcriptome

@@ -1,0 +1,184 @@
+library("data.table")
+library("reshape2")
+library("org.Hs.eg.db")
+library("Biostrings")
+
+args <- c("combined_fpkm","gene_trees","housekeeping_genes_superset","placenta_core_transcriptome")
+
+args <- commandArgs(trailingOnly=TRUE)
+
+load(args[1])
+load(args[2])
+load(args[3])
+
+combined.fpkm <- genes.to.tree[combined.fpkm]
+
+collapse.gene.names <- function(x,min.collapse=2) {
+    ## longest common substring
+    if (is.null(x) || length(x)==0) {
+        return(as.character(NA))
+    }
+    x <- sort(unique(x))
+    str_collapse <- function(y,len) {
+        if (len == 1 || length(y) < 2) {
+            return(y)
+        }
+        y.tree <-
+            gsub(paste0("^(.{",len,"}).*$"),"\\1",y[1])
+        y.rem <-
+            gsub(paste0("^.{",len,"}"),"",y)
+        y.rem.prefix <-
+            sum(combn(y.rem,2,function(x){lcprefix(x[1],x[2])}) >= 2)
+        if (length(y.rem) > 3 &&
+            y.rem.prefix >= 2
+            ) {
+            y.rem <- 
+                collapse.gene.names(y.rem,min.collapse=1)
+        }
+        paste0(y.tree,
+               "{",paste(collapse=",",
+                         y.rem),"}")
+    }
+    i <- 1
+    ret <- NULL
+    while (i <= length(x)) {
+        col.pmin <-
+            pmin(sapply(x,lcprefix,x[i]))
+        collapseable <-
+            which(col.pmin > min.collapse)
+        if (length(collapseable) == 0) {
+            ret <- c(ret,x[i])
+            i <- i+1
+        } else {
+            ret <- c(ret,
+                     str_collapse(x[collapseable],
+                                  min(col.pmin[collapseable]))
+                     )
+            i <- max(collapseable)+1
+        }
+    }
+    return(paste0(collapse=",",ret))
+}
+
+combined.fpkm[!is.na(tree),
+              tree_genes:=collapse.gene.names(na.omit(human_name)),
+              by=tree]
+
+combined.fpkm.wide <-
+    data.table(dcast(combined.fpkm[!is.na(tree),],
+                     tree+tree_genes~species,
+                     fun.aggregate=mean,
[…]

### 2017-02-20 manuscript/supplemental_material.Rnw
switch to legend.spacing

@@
             axis.text.y=element_text(face="italic",size=5),
-            legend.margin=unit(0,"mm"),
+            legend.spacing=unit(0,"mm"),
             legend.justification="left",
@@
               legend.position=c(0.99,0.99),
-              legend.margin=unit(0,"mm"),
+              legend.spacing=unit(0,"mm"),
               legend.justification=c(1,1),

### 2017-02-20 manuscript/supplemental_material.Rnw
add plots of placenta core tree transcriptome

@@
 
+\subsection{Tree-based Core Placenta Transcriptome}
+
+<<load_core_placenta_tree_transcriptome,echo=FALSE,results="hide",cache.extra=file.info("../data/placenta_core_tree_transcriptome")[,"mtime"]>>=
+load("../data/placenta_core_tree_transcriptome")
+
+@
+
+\begin{figure}[tbh]
+<<core_placenta_tree_transcriptome,fig.width=7,fig.height=8,out.height="0.8\\textheight",echo=FALSE>>=
+cpttl <-
+    core.placenta.tree.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
+                                                 Species) & housekeeping==FALSE,]
+cpttl <- cpttl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
+                                levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
+                                            capfirst(names(species.names))))]
+
+print(ggplot(cpttl[FPKM > 0.1,],aes(x=tree_genes,y=FPKM))+
+      geom_boxplot(outlier.shape=NA)+
+      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
+      geom_point(aes(color=Species,shape=Species))+
+      scale_shape_manual("Species",
+                         guide=guide_legend(#nrow=1,
+                                            keywidth=unit(3,"mm"),
+                                            keyheight=unit(3,"mm"),
+                                            label.theme=element_text(face="italic",
+                                                                     angle=0,
+                                                                     size=9),
+                                            title.position="top"
+                                            ),
+                         values=0:18)+
+      # scale_color_discrete(NULL)+
+      coord_flip()+
+      xlab("Genes in Tree")+
+      ylab("Gene Expression (FPKM)")+
+      theme_bw() +
+      theme(legend.position=c(0.7,0.15),
+            axis.text.y=element_text(face="italic",size=5),
+            legend.spacing=unit(0,"mm"),
+            legend.justification="left",
+            plot.margin=unit(c(0,0,0,0),"mm")
+            ))
+@ 
+\caption[Tree-based Core Placenta Transcriptome]{
+  Tree-based Core Placenta Transcriptome.
+  \Sexpr{cpttl[,length(unique(tree_genes))]}
+  gene trees with mean expression ($\mathrm{FPKM} \ge 10$) in the
+  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
+  studied which are not housekeeping~%
+  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
+  ordered (from top to bottom) by median expression.}
+\label{fig:core_placenta_tree_transcriptome}
+\end{figure}
+
+\begin{figure}
+<<core_placenta_tree_transcriptome_housekeeping,fig.width=9,fig.height=18,out.height="0.8\\textheight",echo=FALSE>>=
+cpttlh <-
+    core.placenta.tree.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
+                                                 Species) & housekeeping==TRUE,]
+cpttlh <- cpttlh[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
+                                levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",
+                                            capfirst(names(species.names))))]
+
+print(ggplot(cpttlh[FPKM > 0.1,],aes(x=tree_genes,y=FPKM))+
+      geom_boxplot(outlier.shape=NA)+
+      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
+      geom_point(aes(color=Species,shape=Species))+
+      scale_shape_manual("Species",
+                         guide=guide_legend(#nrow=1,
+                                            keywidth=unit(3,"mm"),
+                                            keyheight=unit(3,"mm"),
+                                            label.theme=element_text(face="italic",
+                                                                     angle=0,
+                                                                     size=9),
+                                            title.position="top"
+                                            ),
+                         values=0:18)+
+      # scale_color_discrete(NULL)+
+      coord_flip()+
+      xlab("Genes in Tree")+
+      ylab("Gene Expression (FPKM)")+
+      theme_bw() +
+      theme(legend.position=c(0.7,0.15),
+            axis.text.y=element_text(face="italic",size=5),
+            legend.spacing=unit(0,"mm"),
+            legend.justification="left",
+            plot.margin=unit(c(0,0,0,0),"mm")
+            ))
+@ 
+\caption[Tree-based Core Placenta Transcriptome: Housekeeping Genes]{
+  Tree-based Core Placenta Transcriptome: Housekeeping Genes.
+  \Sexpr{cpttlh[,length(unique(tree_genes))]}
+  gene trees with mean expression ($\mathrm{FPKM} \ge 10$) in the
+  placenta of all \Sexpr{star.logs[,length(unique(species))]} species
+  studied which are housekeeping~%
+  \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}
+  ordered (from top to bottom) by median expression.}
+\label{fig:core_placenta_tree_transcriptome_housekeeping}
+\end{figure}
+
+\clearpage
+
 \section{Placenta Classification Analyses}

### 2017-02-20 manuscript/manuscript.Rnw
discuss core placenta tree transcriptome

@@
 
+<<load_core_placenta_tree_transcriptome,echo=FALSE,results="hide",cache.extra=file.info("../data/placenta_core_tree_transcriptome")[,"mtime"]>>=
+load("../data/placenta_core_tree_transcriptome")
+cpttl <-
+    core.placenta.tree.transcriptome.long[!grepl("(min|median|max|all)_expression|all_but_",
+                                                 Species) & housekeeping==FALSE,]
+@
+
 The
@@
 
+If we use gene trees to define the core placenta transcriptome instead
+of single-copy orthologs (where the mean expression for genes in the
+tree which are present must be $\text{FPKM} \ge 10$), we find
+\Sexpr{cpttl[,length(unique(tree_genes))]} unique gene trees which
+include nearly all of the genes present in the core placenta
+transcriptome (Supplemental
+\cref{fig:core_placenta_tree_transcriptome,fig:core_placenta_tree_transcriptome_housekeeping}).
+
 \subsubsection{Annexin Complexes}

### 2017-02-20 manuscript/manuscript.Rnw
Define the precise numerical values for total fraction

@@
 \cref{fig:core_housekeeping_transcriptome_shape}), and includes more
-of the total fraction of genes. An alternative approach to defining
+of the total fraction of genes (%
+%
+\Sexpr{core.placenta.transcriptome.long[housekeeping==TRUE,length(unique(human_name))]/
+  core.placenta.transcriptome.long[,length(unique(human_name))]} %
+%
+%
+vs %
+\Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]/
+  core.placenta.transcriptome.long[,length(unique(human_name))]}%
+%
+). An alternative approach to defining
 housekeeping genes is to use the tissue specificity

### 2017-02-20 manuscript/manuscript.Rnw
add limitations section
Add limitations section to address a reviewer critique. Additionally
fix a time point typo.

@@
 
+\subsection{Limitations}
+
+This study includes only term placentas, and our conclusions are
+limited to the set of genes which continue to be expressed at term. We
+suspect that many genes which are responsible for establishing shape,
+intimacy, and interdigitation are expressed and function at earlier
+time points where placental morphology is being established.
+Furthermore, while these samples have been dissected to only include
+the fetal interface, they likely include substantial cell and tissue
+heterogeneity. We are therfore not yet able to assign specific genes
+in our core placental transcriptome to specific cell types and
+tissues. We believe that samples of placentas from earlier in
+gestation and from multiple locations within the placenta and at
+multiple will address these limitations.
+
 % TAGLN2 -- likely to be produced by the placenta as well; possibly
@@
 hypotheses are limited by our limited data; more placentas from
-multiple members of diverse clades and different timepoints are still
+multiple members of diverse clades and different time points are still
 needed to come to a more complete understanding of the genomic and

### 2017-03-08 data/combine_fpkm.R
Switch to using OMA to define ortholog pairs
 - Use ensembl as a backup
 - Test that the orthologs are correctly assigned
 - Completely codify the rules in source and check them

@@ -1,6 +1,6 @@
-library(data.table)
-library(reshape2)
-library(parallel)
+library("data.table")
+library("reshape2")
+library("parallel")
 
 args <- commandArgs(trailingOnly=TRUE)
 
@@ -10,59 +10,98 @@ if (is.null(getOption("mc.cores"))) {
 
 output.file <- args[1]
 
-homology.file <- args[2]
+## fastq_fpkm/combined_fpkms
+load(args[3])
 
-load(args[4])
-
-
-### from ftp://ftp.informatics.jax.org/pub/reports/HOM_AllOrganism.rpt
-homology.mouse.human <- fread(homology.file)
-setnames(homology.mouse.human,"HomoloGene ID","homologene.id")
-setnames(homology.mouse.human,"Common Organism Name","organism.name")
-homology.table <-
-    data.table(dcast(homology.mouse.human,
-                     homologene.id~organism.name,
-                     fun.aggregate=function(x){x[1]},value.var="Symbol"))
-setnames(homology.table,
-         c("mouse, laboratory",
-           "dog, domestic",
-           "rhesus macaque",
-           "western clawed frog"
-           ),
-         c("mouse",
-           "dog",
-           "macaque",
-           "frog"
-           ))
-human.homolog <- function(gene,species="mouse") {
-    return(homology.table[eval(parse(text=paste0(species,"==\"",gene,"\""))),
-                          human][1])
-}
-
-oma.groups <- fread(args[3])
+## oma-groups_long.txt
+oma.groups <- fread(args[2])
 oma.groups[,species:=gsub("\\d+","",oma_entry)]
+oma.groups <- oma.groups[grepl("G0",ensembl),]
 setkey(oma.groups,"ensembl")
 oma.groups.human <- oma.groups[species=="HUMAN"]
-setkey(oma.groups.human,"group_num")
+setkey(oma.groups.human,"fingerprint")
 
-oma.homolog <- function(ensembl.id){
-    gene.fpkms[tracking_id %in% oma.groups.human[group_num==oma.groups[ensembl.id,group_num],
-                                                    ensembl],
-                  gene_short_name][1]
+
+### ensembl_all_orthologs
+load(args[4])
+orthologs.subset <-
+    orthologs[secondary=="hsapiens",]
+setkey(orthologs.subset,"ensembl_gene_id")
+
+human_gene_names <-
[…]

### 2017-03-08 data/ensembl_orthologs_long.R

@@ -3,7 +3,10 @@ library(data.table)
 
 args <- commandArgs(trailingOnly=TRUE)
 
-human <- useMart("ensembl",dataset="hsapiens_gene_ensembl")
+mart <- useMart("ENSEMBL_MART_ENSEMBL",
+                dataset="hsapiens_gene_ensembl",
+                host="may2015.archive.ensembl.org")
+
 
 ensembl.orthologs.long <- list()
 
@@ -18,7 +21,7 @@ for (species in c("dnovemcinctus",
         data.table(getBM(c("ensembl_gene_id",
                            paste0(species,"_",
                                   "homolog_ensembl_gene")),
-                         values =TRUE, mart = human, bmHeader=FALSE))
+                         values =TRUE, mart = mart, bmHeader=FALSE))
     setnames(ensembl.orthologs.long[[species]],
              paste0(species,"_",
                     "homolog_ensembl_gene"),
@@ -27,12 +30,6 @@ for (species in c("dnovemcinctus",
     ensembl.orthologs.long[[species]] <-
         ensembl.orthologs.long[[species]][homolog_ensembl_gene!="",]
 }
-
-hgnc.symbols <-
-    data.table(getBM(c("ensembl_gene_id",
-                       "hgnc_symbol"),
-                     mart=human))
-
 
 ensembl.orthologs.long <- rbindlist(ensembl.orthologs.long)
 

### 2017-03-08 data/housekeeping_genes_superset.R
fix the name of the housekeeping genes files

@@ -1,6 +1,6 @@
 library("data.table")
 
-args <- c("housekeeping_genes_superset.txt"
+args <- c("housekeeping_genes_superset.txt",
           "housekeeping_genes_superset"
           )
 
@@ -8,7 +8,7 @@ args <- commandArgs(trailingOnly=TRUE)
 
 
 
-housekeeping_genes_superset <-
+housekeeping.genes.superset <-
     data.table(read.table(args[1]))
 
 save(housekeeping.genes.superset,file=args[length(args)])

### 2017-03-08 manuscript/manuscript.Rnw
use ortholog pairs and describe OMA analysis

@@
 
-Single-copy (1-to-1) orthologs in were determined using Ensembl v.80
-gene trees via Biomart using the biomaRt R package
-\cite{Durinck.ea2009:Mappingidentifiersforintegration}. %
-A total of \Sexpr{combined.fpkm[species!="homo sapiens" & mean_fpkm >=
-  1,length(unique(human_name))]} single-copy orthologous genes were
-expressed in at least one species. We also obtained orthology from the
-perspective of each species to all other species using biomaRt package
-and Ensembl v.80 database to determine the percentage of highly
-expressed genes which had an orthology of one-to-one, one-to-many,
-many-to-many, or novel/unknown.
+Orthologous pairs between \human and all other species were identified
+using the \ac{OMA} database
+\cite{Altenhoff.ea2015:OMAorthologydatabasein}. In cases where a gene
+did not have a human ortholog pair identified by \ac{OMA}, Ensembl v.80
+one2one orthologs obtained from Biomart using the biomaRt R package
+\cite{Durinck.ea2009:Mappingidentifiersforintegration} %
+were used to select the appropriate orthologous pair. A total of
+\Sexpr{combined.fpkm[,list(n=length(gene_id),
+  max.fpkm=max(mean_fpkm)),by=human_name][n==14 & max.fpkm >
+  10,length(human_name)]} ortholog pair sets in all 14 species were
+expressed (\ac{FPKM} $\ge 10$) in at least one species. We also
+obtained orthology from the perspective of each species to all other
+species using biomaRt package and Ensembl v.80 database to determine
+the percentage of highly expressed genes which had an orthology of
+one-to-one, one-to-many, many-to-many, or novel/unknown. For future
+analyses by other investigators, we have also provided the human gene
+which most closely aligns to each expressed annotated
+protein-generating sequence of all species analyzed using DIAMOND
+\cite{Buchfink.ea2015:Fastandsensitiveprotein}.
 
 Significant shifts in expression levels between groups of species were
-calculated using Student's t test on single-copy orthologs using the mean of
+calculated using Student's t test on ortholog pairs using the mean of
 each species to avoid large biases towards samples with $n \gg 1$ and
@@
 Between 20\% and 40\% of all highly expressed genes in each species
-were single-copy orthologs with all other analyzed species, with the
+were ortholog pairs with all other analyzed species, with the
 remainder being mostly one-to-many or many-to-many in comparison to at
@@
   \Sexpr{core.placenta.transcriptome.long[housekeeping==FALSE,length(unique(human_name))]}
-  single-copy \human ortholog genes expressed (\ac{FPKM}$ \ge 10$) in the
+  ortholog pairs of \human genes expressed (\ac{FPKM}$ \ge 10$) in the
   placenta of all 13 eutherians studied with housekeeping genes
@@
       + theme(text=element_text(size=9),
-              legend.position=c(0.99,0.99),
-              legend.margin=unit(0,"mm"),
-              legend.justification=c(1,1),
+              legend.position=c(0.01,0.01),
+              legend.margin=margin(0,0,0,0,unit="mm"),
+              legend.spacing=unit(0,"mm"),
+              legend.justification=c(0,0),
               plot.margin=unit(c(0,0,0,0),"mm")
@@
   number of genes. The vertical axis
-  shows the number of single-copy orthologous genes included in the core
+  shows the number of ortholog pairs included in the core
   placenta transcriptome (for all possible number of genes) if %
@@
   %
-  genes would be included if the expression of a set of single-copy orthologs of all 13 Eutherians must be at
-  least 10 FPKM, %
+  genes would be included if the expression of a set of ortholog pairs of all 13 Eutherians must be at
+9  least 10 FPKM, %
   \Sexpr{cpt.distribution.shape[`Expression (FPKM)` > 10 
@@
   %
-  single-copy orthologs would be included if 13 of 14 species have an
+  ortholog pairs would be included if 13 of 14 species have an
    expression of at least 10 FPKM, and 
@@
 transcription level in \ac{FPKM}. We considered a gene to be core to
-the placenta transcriptome if a single-copy ortholog of that gene was
+the placenta transcriptome if an ortholog pair of that gene was
 expressed with $\text{FPKM} \ge 10$ in all eutherian species studied
@@
 resolution. Increasing the permissiveness of the core placenta
-transcriptome by only requiring 10 of 14 species to express single-copy
-orthologs with FPKM $>10$ and results in the same terms also being
+transcriptome by only requiring 10 of 14 species to express an
+ortholog pair with FPKM $>10$ and results in the same terms also being
 significantly enriched, along with additional terms (Supplemental
@@
 If we use gene trees to define the core placenta transcriptome instead
-of single-copy orthologs (where the mean expression for genes in the
-tree which are present must be $\text{FPKM} \ge 10$), we find
+of ortholog pairs (where the mean expression for genes in the tree
+which are present must be $\text{FPKM} \ge 10$), we find
 \Sexpr{cpttl[,length(unique(tree_genes))]} unique gene trees which

### 2017-03-08 manuscript/manuscript.Rnw
fix pubmed_to_grid latex escape

@@
 pregnancy outcomes and other placenta-relevant terms of interest, we
-used the “pubmed_to_grid” component of
+used the “pubmed\_to\_grid” component of
 Function2Gene~\cite{Armstrong2008:function2gene} which conducts a term

### 2017-03-08 manuscript/manuscript.Rnw
use adjustbox centering

@@
 \usepackage{longtable}
+\usepackage[export]{adjustbox}[2011/08/13]
 \usepackage{amsmath}
@@
 <<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
-opts_chunk$set(dev="cairo_pdf",out.width="\\columnwidth",out.height="0.7\\textheight",out.extra="keepaspectratio")
+opts_chunk$set(dev="cairo_pdf",out.width="\\columnwidth",out.height="0.7\\textheight",out.extra="keepaspectratio,center")
 opts_chunk$set(cache=TRUE, autodep=TRUE)

### 2017-03-08 manuscript/manuscript.Rnw
fix } typo in reference

@@
 especially in later stages of pregnancy
-\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman,Hou.ea2012:Elephanttranscriptomeprovidesinsights}}.
+\cite{Carter2012:Evolutionofplacentalfunction,Knox.ea2008:Genomicevolutionofplacenta,Su.ea2002:Large-scaleanalysisofhuman,Hou.ea2012:Elephanttranscriptomeprovidesinsights}.
 For example, each superorder (Euarchontoglires, Laurasiatheria,

### 2017-03-08 manuscript/manuscript.Rnw
indicate when the pubmed literature search was run

@@
 \caption{\lineagetableexplanationtypeiiishort{\human}{all other
-    analyzed species} Adv. Preg. Pub. is the approximate number of
-  publications indexed by PubMed mentioning this gene and an adverse
-  pregnancy outcome (pre-eclampsia, premature labor, gestational
-  diabetes, intrauterine growth restriction, or premature birth) and
-  links to the most current PubMed search results.}
+    analyzed species} The human symbol links to a brief explanation of
+  the gene function in Ensembl. Adv. Preg. Pub. is the approximate
+  number of publications indexed by PubMed mentioning this gene and an
+  adverse pregnancy outcome (pre-eclampsia, premature labor,
+  gestational diabetes, intrauterine growth restriction, or premature
+  birth) as of February 2017, and links to the most current PubMed
+  search results.}
 \label{tab:human_lineage_specific}

### 2017-03-08 manuscript/manuscript.Rnw
make the core placenta plot two-column

@@
 \begin{figure}[tbh]
-<<core.placenta.transcriptome,fig.width=9/2.5,fig.height=8,out.width="90mm",out.height="0.8\\textheight",echo=FALSE>>=
+<<core.placenta.transcriptome,fig.width=19/2.5,fig.height=6,out.width="190mm",echo=FALSE>>=
 cptl <-
@@
                                           capfirst(names(species.names))))]
-
+cptl <- cptl[,median_FPKM:=median(FPKM),by=human_name]
+cptl <- cptl[,facet:="high"]
+cptl <- cptl[median_FPKM < median(median_FPKM),facet:="low"]
 print(ggplot(cptl[FPKM > 0.1,],aes(x=human_name,y=FPKM))+
       geom_boxplot(outlier.shape=NA)+
-      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
+      scale_y_log10(breaks=c(10^1,10^2,10^3,10^4),
+                    labels=trans_format("log10", math_format(10^.x)))+
       geom_point(aes(color=Species,shape=Species))+
-      scale_shape_manual("Species",
+      scale_shape_manual(NULL,##"Species",
                          guide=guide_legend(#nrow=1,
@@
                          values=0:18)+
-      # scale_color_discrete(NULL)+
+      scale_color_discrete(NULL)+
       coord_flip()+
@@
       theme_bw() +
-      theme(text=element_text(size=9),
-            legend.position=c(0.99,0.41),
-            axis.text.y=element_text(face="italic",size=7),
-            legend.margin=unit(0,"mm"),
-            legend.justification=c(1,1),
-            plot.margin=unit(c(0,0,0,0),"mm")
-            ))
+      expand_limits(y=c(10^4)) + 
+      theme(text=element_text(size=7.5),
+            legend.position=c(0.99,0.01),
+            axis.text.y=element_text(face="italic",size=7.5),
+            legend.margin=margin(0,0,0,0,unit="mm"),
+            legend.spacing=unit(0,"mm"),
+            legend.justification=c(1,0),
+            plot.margin=unit(c(0,0,0,0),"mm"),
+            strip.background = element_blank(),
+            strip.text.x = element_blank()
+            )
+      + facet_wrap(~facet,scales="free")
+      )
 @ 

### 2017-03-08 data/placenta_core_transcriptome.R
we only want those genes which have NA, use max

@@ -10,9 +10,9 @@ load(args[1])
 load(args[2])
 
 combined.fpkm.wide <- 
-    data.table(dcast(combined.fpkm,
+    data.table(dcast(combined.fpkm[!is.na(human_name),],
                      human_name~species,
-                     fun.aggregate=sum,
+                     fun.aggregate=max,
                      value.var="mean_fpkm"))
 
 min.na.zero <- function(x){

### 2017-03-08 manuscript/manuscript.Rnw
move the legend in the CPT shape plot
don't use reactome; discuss changes in GSEA due to core change

@@
 pathway <- list()
-for (database in c("kegg","reactome","biocarta","netpath")) {
+for (database in c("kegg","biocarta","netpath")) {
     load(paste0("../data/placenta_gsea_",database))
@@
       + geom_vline(xintercept=10,linetype="dotted")
-      + geom_segment(mapping=aes(y=V1-ifelse(V1<80,20,0),
+      + geom_segment(mapping=aes(y=V1-ifelse(V1<100,20,0),
                          yend=V1,
@@
                                       cpt.distribution.shape[num_genes > x,
-                                                             max(`Expression (FPKM)`)*4]
+                                                             max(`Expression (FPKM)`)*2]
                                   }),
@@
                                            cpt.distribution.shape[num_genes > x,
-                                                                  max(`Expression (FPKM)`)*4.1]
+                                                                  max(`Expression (FPKM)`)*2.1]
                                        }),
-                      y=V1-ifelse(V1<80,20,0),
-                      label=paste0(V1," genes"),override=TRUE),
+                      y=V1-ifelse(V1<100,20,0),
+                      label=paste0(V1," genes")),inherit.aes=FALSE,
                   data=cpt.distribution.shape[`Expression (FPKM)` > 10,
@@
       + geom_segment(mapping=aes(yend=V1,
-                         y=100,
+                         y=10^2.5,
                          xend=1,
@@
       + geom_text(mapping=aes(x=1,
-                      y=100,
-                      label=paste0(V1," genes"),
-                      override=TRUE),
+                      y=10^2.5,
+                      label=paste0(V1," genes")),
+                      inherit.aes=FALSE,
                   angle=90,
@@
 The core placenta transcriptome is significantly enriched in genes
-which are involved in multiple biological pathways, including multiple
-pathways related to translation (ribosome %
-%
-(\Sexpr{placpath("^RIBOSOME","Kegg")})
-, translation %
-(\Sexpr{placpath("^TRANSLATION","Reactome")})
-%
-, signal
-recognition particle targeting to the rough endoplasmic reticulum %
-(\Sexpr{placpath("^SRP DEPENDENT","Reactome")})
-%
-,
-etc.) (Supplemental \cref{tab:core_placenta_pathway_analysis}). Beyond
-translation, the next most abundant set of significantly enriched
-molecular pathways are related to the migration of the trophoblast
-into the endometrium, including the EGFR1 pathway %
-(\Sexpr{placpath("^EGFR1","Netpath")})
+which are involved in multiple biological pathways (Supplemental
+\cref{tab:core_placenta_pathway_analysis}). The most abundant set
+of significantly enriched molecular pathways are related to the
+migration of the trophoblast into the endometrium, including the EGFR1
+pathway %
+(\Sexpr{placpath("^EGFR1","Netpath")}), Focal Adhesion pathway
+(\Sexpr{placpath("^EGFR1","Kegg")})
 %
 and the Actin Y pathway %
-(\Sexpr{placpath("^ACTINY","Biocarta")})
+(\Sexpr{placpath("^ACTINY","Biocarta")})%
 %

### 2017-03-08 manuscript/supplemental_material.Rnw
move the legend in the supplemental CPTL shape plot
don't use reactome; discuss changes in GSEA due to core change

@@
       + geom_vline(xintercept=10,linetype="dotted")
-      + geom_segment(mapping=aes(y=V1-ifelse(V1<145,40,0),
+      + geom_segment(mapping=aes(y=V1-ifelse(V1<160,40,0),
                          yend=V1,
@@
                                        }),
-                      y=V1-ifelse(V1<145,40,0),
-                      label=paste0(V1," genes"),override=TRUE),
+                      y=V1-ifelse(V1<160,40,0),
+                      label=paste0(V1," genes")),
+                  inherit.aes=FALSE,
                   data=cpt.h.distribution.shape[`Expression (FPKM)` > 10,
@@
       + geom_text(mapping=aes(x=1,
-                      y=100,
-                      label=paste0(V1," genes"),
-                      override=TRUE),
+                      y=10^2.5,
+                      label=paste0(V1," genes")),
+                  inherit.aes=FALSE,
                   angle=90,
@@
       + theme(text=element_text(size=9),
-              legend.position=c(0.99,0.99),
+              legend.position=c(0.01,0.01),
               legend.spacing=unit(0,"mm"),
-              legend.justification=c(1,1),
+              legend.justification=c(0,0),
               plot.margin=unit(c(0,0,0,0),"mm")
@@
   ($\text{FDR}\le 0.1$) in the core placenta transcriptome using the
-  biocarta, reactome, and kegg pathway databases}
+  biocarta, netpath, and kegg pathway databases}
 \label{tab:core_placenta_pathway_analysis}

### 2017-03-08 manuscript/manuscript.Rnw
document that APOB and AFP are associated in mice

@@
 and found few genes which are highly differentially expressed and are
-also associated with adverse pregnancy outcomes. Comparisons between
-rodentia (\mouse, \galili, and \carmeli) and all other species using
-the mean of each species expression did not indicate any significant
-differences (Supplemental \cref{tab:rodentia_lineage_specific}) nor
-did the equivalent comparisons between laurasiatheria and all other
-species (Supplemental \cref{tab:laurasiatheria_lineage_specific}) and
-laurasiatheria and euarchontoglires (Supplemental
+also associated with adverse pregnancy outcomes, including \ac{APOB}
+and \ac{AFP} (Supplemental \cref{tab:mouse_lineage_specific_all}).
+Comparisons between rodentia (\mouse, \galili, and \carmeli) and all
+other species using the mean of each species expression did not
+indicate any significant differences (Supplemental
+\cref{tab:rodentia_lineage_specific}) nor did the equivalent
+comparisons between laurasiatheria and all other species (Supplemental
+\cref{tab:laurasiatheria_lineage_specific}) and laurasiatheria and
+euarchontoglires (Supplemental
 \cref{tab:laurasiatheria_euarchontoglires_lineage_specific}).

### 2017-03-08 manuscript/manuscript.Rnw
discuss that houses are different

@@
 %
-). An alternative approach to defining
-housekeeping genes is to use the tissue specificity
+). An alternative approach to identify housekeeping genes is to use
+the tissue specificity
 index~\cite{Yanai.ea2005:Genome-widemidrangetranscriptionprofiles}.
-Supplemental \cref{tab:tissue_specificity_index_of_plaenta_core_genes}
-lists the tissue specificity index for all housekeeping and
-non-housekeeping genes; housekeeping genes expressed in all placentas
-have a lesser median tissue specificity index than non-housekeeping
-genes (Supplemental
-\cref{fig:tissue_specificity_index_with_housekeeping}).
+Supplemental
+\cref{tab:tissue_specificity_index_of_placenta_core_genes} lists the
+tissue specificity index for all housekeeping and non-housekeeping
+genes; housekeeping genes expressed in all placentas have a lesser
+median tissue specificity index than non-housekeeping genes
+(Supplemental \cref{fig:tissue_specificity_index_with_housekeeping}).
+We also note that housekeeping genes have been defined in \human, and
+may not be housekeeping genes in other species; generating tissue
+expression atlases and examining the evolution of expression in
+non-model organisms is a promising area of future research.
 

### 2017-03-08 manuscript/manuscript.Rnw
describe how the fetal villous portion was separated

@@
 
-Fetal tissue was collected from the placenta of nine mammalian
-species (\human [human], \bonobo [bonobo], \smonkey [spider monkey], \dog [domestic dog], \cow
-[domestic cow], \elephant [African savanna elephant], \opossum [gray
-short-tailed opossum], \galili [blind mole rat from Upper Galilee],
-\carmeli [blind mole rat from northern Israel]
-\cite{Nevo2001:adaptiveradiationofblindsubterraneanmole};
-n=1 for all collected species). Information
-on specimen ID, number of specimens, and collector is shown for each sample in Supplemental
+Fetal tissue was collected from the placenta of nine mammalian species
+(\human [human], \bonobo [bonobo], \smonkey [spider monkey], \dog
+[domestic dog], \cow [domestic cow], \elephant [African savanna
+elephant], \opossum [gray short-tailed opossum], \galili [blind mole
+rat from Upper Galilee], \carmeli [blind mole rat from northern
+Israel] \cite{Nevo2001:adaptiveradiationofblindsubterraneanmole}; n=1
+for all collected species). Information on specimen ID, number of
+specimens, and collector is shown for each sample in Supplemental
 \cref{tab:srr_numbers}. In the case of \bonobo, \cow, and \elephant,
@@
 delivery via a cesarean section. Near-term placental tissue from \dog
-was collected upon the death of the mother and fetus from a car accident.
-Placental tissue from \opossum was sampled at fetal stage 32, roughly
-one day prior to birth. Only the fetal portion of the placenta was
-sampled at the maternal-fetal interface (i.e. villous portion for
-\human, \bonobo, \dog, and \elephant; cotyledon for \cow; diffuse
-placenta for \opossum). All tissues were collected and stored in
-RNAlater (Life Technologies, Carlsbad, CA) until RNA isolation could
-be performed. Tissues were homogenized in Trizol, and total RNA was
-isolated following the manufacturer's instructions. RNA was further
-purified using RNeasy Midi Kit (Qiagen, Valencia, CA), in conjunction
-with the RNase-Free DNase Set (Qiagen, Valencia, CA), following the
-manufacturer’s protocol to remove any contaminating DNA from the
-sample. Total RNA was then evaluated on a Nanodrop 1000 (Thermo
-Scientific, Wilmington, DE) and an Agilent Bioanalyzer 2100 (Santa
-Clara, CA) to confirm concentration and quality of the isolation. RNA
-quality was assessed using A260/A280 and A260/A230 absorbance values.
+was collected upon the death of the mother and fetus from a car
+accident. Placental tissue from \opossum was sampled at fetal stage
+32, roughly one day prior to birth. Only the fetal villous portion of
+the placenta was sampled at the maternal-fetal interface (i.e. villous
+portion for \human, \bonobo, \dog, and \elephant; cotyledon for \cow;
+diffuse placenta for \opossum). The placenta membranes were separated
+from the fetal villous portion using forceps and a scapel, and the
+resulting fetal villous portion was washed twice in PBS. All tissues
+were collected and stored in RNAlater (Life Technologies, Carlsbad,
+CA) until RNA isolation could be performed. Tissues were homogenized
+in Trizol, and total RNA was isolated following the manufacturer's
+instructions. RNA was further purified using RNeasy Midi Kit (Qiagen,
+Valencia, CA), in conjunction with the RNase-Free DNase Set (Qiagen,
+Valencia, CA), following the manufacturer’s protocol to remove any
+contaminating DNA from the sample. Total RNA was then evaluated on a
+Nanodrop 1000 (Thermo Scientific, Wilmington, DE) and an Agilent
+Bioanalyzer 2100 (Santa Clara, CA) to confirm concentration and
+quality of the isolation. RNA quality was assessed using A260/A280 and
+A260/A230 absorbance values.
 

### 2017-03-08 manuscript/acronyms.tex
add OMA as an acronym
italicize C. elegans

@@
 \DeclareAcronym{FPKM}{short=FPKM,long={Fragments Per Kilobase of exon per Million reads}}
+\DeclareAcronym{OMA}{short=OMA,long={Orthologous MAtrix project}}
 \DeclareAcronym{HSD3B1}{short=\textit{HSD3B1},long={hydroxy-δ-5-steroid dehydrogenase, 3 β- and steroid δ-isomerase 1},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=HSD3B1}{\textit{HSD3B1}} (hydroxy-δ-5-steroid dehydrogenase, 3 β- and steroid δ-isomerase 1)}}
@@
 \DeclareAcronym{EREG}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=EREG}{\textit{EREG}},long={epiregulin},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=EREG}{\textit{EREG}} (epiregulin)}}
-\DeclareAcronym{UNC93A}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93A}{\textit{UNC93A}},long={unc-93 homolog A (C.
-elegans)},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93A}{\textit{UNC93A}} (unc-93 homolog A (C.
-elegans))}}
-\DeclareAcronym{UNC93B1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93B1}{\textit{UNC93B1}},long={unc-93 homolog B1 (C.
-elegans)},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93B1}{\textit{UNC93B1}} (unc-93 homolog B1 (C.
-elegans))}}
+\DeclareAcronym{UNC93A}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93A}{\textit{UNC93A}},long={unc-93 homolog A (\textit{C.
+elegans})},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93A}{\textit{UNC93A}} (unc-93 homolog A (\textit{C.
+elegans}))}}
+\DeclareAcronym{UNC93B1}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93B1}{\textit{UNC93B1}},long={unc-93 homolog B1 (\textit{C.
+elegans})},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=UNC93B1}{\textit{UNC93B1}} (unc-93 homolog B1 (\textit{C.
+elegans}))}}
 \DeclareAcronym{WNT5A}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=WNT5A}{\textit{WNT5A}},long={Wnt family member 5A},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=WNT5A}{\textit{WNT5A}} (Wnt family member 5A)}}

### 2017-03-08 manuscript/manuscript.Rnw
fix a/an typo

@@
 Another annexin which is highly expressed in all term placentas is
-\ac{ANXA5}, which was originally identified as a an anti-coagulant
+\ac{ANXA5}, which was originally identified as an anti-coagulant
 protein, but is also involved in syncytium formation, membrane repair,

### 2017-03-08 manuscript/supplemental_material.Rnw
fix placenta typo in supp table label

@@
                        ),
-             label="tab:tissue_specificity_index_of_plaenta_core_genes"
+             label="tab:tissue_specificity_index_of_placenta_core_genes"
              ),

### 2017-03-08 manuscript/supplemental_material.Rnw
switch away from kegg pathview to KGML2Graph

@@
 \setmonofont{FreeMono}
+\setmathrm{FreeSerif}
+\setmathsf{FreeSans}
+\setmathtt{FreeSerif}
 \usepackage[letterpaper,left=1in,right=1in,top=1in,bottom=1in,headheight=15pt]{geometry}
 \usepackage{url}
+\usepackage{siunitx}
 \usepackage{fancyhdr}
@@
 \usepackage{booktabs}
-\usepackage{float}
+%\usepackage{float}
 \usepackage{multirow}
 \usepackage{multicol}
-\usepackage{setspace}
-\usepackage{wrapfig}
-% \usepackage{txfonts}
 \usepackage[macros=false]{acro}
@@
 \usepackage{dcolumn}
-\usepackage{adjustbox}
 \usepackage{longtable}
+\usepackage[export]{adjustbox}[2011/08/13]
 \usepackage{pdflscape}
@@
 <<load.libraries,echo=FALSE,results="hide",warning=FALSE,message=FALSE,error=FALSE,cache=FALSE>>=
-opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.6\\textheight",out.extra="keepaspectratio")
+opts_chunk$set(dev="cairo_pdf",out.width="\\textwidth",out.height="0.6\\textheight",out.extra="keepaspectratio,center")
 opts_chunk$set(cache=TRUE, autodep=TRUE,cache.path="supplemental_cache/")
@@
 <<lau.lineage.specific,echo=FALSE,results="asis">>=
-lau.all.lin.spec <- placenta.lineage.specific.table("lau.all",fdr.cutoff=0.5)
+lau.all.lin.spec <- placenta.lineage.specific.table("lau.all",fdr.cutoff=1)
 setkey(lau.all.lin.spec,"Human Symbol")
@@
                                           capfirst(names(species.names))))]
-
+cptlh <- cptlh[,median_FPKM:=median(FPKM),by=human_name]
+cptlh <- cptlh[,facet:="high"]
+cptlh <- cptlh[median_FPKM < median(median_FPKM),facet:="low"]
 print(ggplot(cptlh[FPKM > 0.9,],aes(x=human_name,y=FPKM))+
       geom_boxplot(outlier.shape=NA)+
-      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
+      scale_y_log10(breaks=c(10^1,10^2,10^3,10^4),
+                    labels=trans_format("log10", math_format(10^.x)))+
       geom_point(aes(color=Species,shape=Species))+
-      scale_shape_manual("Species",
+      scale_shape_manual(NULL,##"Species",
                          guide=guide_legend(#nrow=1,
@@
       theme_bw() +
-      theme(legend.position=c(0.7,0.15),
-            axis.text.y=element_text(face="italic",size=5),
+      expand_limits(y=c(10^4)) + 
+      theme(text=element_text(size=7.5),
+            legend.position=c(0.99,0.01),
+            axis.text.y=element_text(face="italic",size=7.5),
+            legend.margin=margin(0,0,0,0,unit="mm"),
             legend.spacing=unit(0,"mm"),
-            legend.justification="left",
-            plot.margin=unit(c(0,0,0,0),"mm")
-            ))
+            legend.justification=c(1,0),
+            plot.margin=unit(c(0,0,0,0),"mm"),
+            strip.background = element_blank(),
+            strip.text.x = element_blank()
+            )
+      + facet_wrap(~facet,scales="free")
+      )
 @ 
@@
       + geom_vline(xintercept=10,linetype="dotted")
-      + geom_segment(mapping=aes(y=V1-ifelse(V1<160,40,0),
+      + geom_segment(mapping=aes(y=V1-ifelse(V1<170,40,0),
                          yend=V1,
@@
                                        }),
-                      y=V1-ifelse(V1<160,40,0),
+                      y=V1-ifelse(V1<170,40,0),
                       label=paste0(V1," genes")),
@@
 \begin{figure}[tbh]
-<<core_placenta_tree_transcriptome,fig.width=7,fig.height=8,out.height="0.8\\textheight",echo=FALSE>>=
+<<core_placenta_tree_transcriptome,fig.width=12,fig.height=8,out.width="\\textwidth",out.height="0.8\\textheight",echo=FALSE>>=
 cpttl <-
@@
                                             capfirst(names(species.names))))]
-
-print(ggplot(cpttl[FPKM > 0.1,],aes(x=tree_genes,y=FPKM))+
+cpttl <- cpttl[,median_FPKM:=median(FPKM,na.rm=TRUE),by=tree_genes]
+cpttl <- cpttl[,facet:="high"]
+cpttl <- cpttl[median_FPKM < median(median_FPKM),facet:="low"]
+print(ggplot(cpttl[FPKM > 0.1,],aes(x=reorder(tree_genes,median_FPKM),y=FPKM))+
       geom_boxplot(outlier.shape=NA)+
-      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
+      scale_y_log10(breaks=c(10^1,10^2,10^3,10^4),
+                    labels=trans_format("log10", math_format(10^.x)))+
       geom_point(aes(color=Species,shape=Species))+
-      scale_shape_manual("Species",
+      scale_shape_manual(NULL,##"Species",
                          guide=guide_legend(#nrow=1,
@@
 \begin{figure}
-<<core_placenta_tree_transcriptome_housekeeping,fig.width=9,fig.height=18,out.height="0.8\\textheight",echo=FALSE>>=
+<<core_placenta_tree_transcriptome_housekeeping,fig.width=12,fig.height=12,out.height="0.8\\textheight",echo=FALSE>>=
 cpttlh <-
@@
 
-print(ggplot(cpttlh[FPKM > 0.1,],aes(x=tree_genes,y=FPKM))+
+cpttlh <- cpttlh[,median_FPKM:=median(FPKM,na.rm=TRUE),by=tree_genes]
+cpttlh <- cpttlh[,facet:="high"]
+cpttlh <- cpttlh[median_FPKM < median(median_FPKM),facet:="low"]
+print(ggplot(cpttlh[FPKM > 0.1,],aes(x=reorder(tree_genes,median_FPKM),y=FPKM))+
       geom_boxplot(outlier.shape=NA)+
-      scale_y_log10(labels=trans_format("log10", math_format(10^.x)))+
+      scale_y_log10(breaks=c(10^1,10^2,10^3,10^4),
+                    labels=trans_format("log10", math_format(10^.x)))+
       geom_point(aes(color=Species,shape=Species))+
-      scale_shape_manual("Species",
+      scale_shape_manual(NULL,##"Species",
                          guide=guide_legend(#nrow=1,
@@
 @
-\caption{EGF KEGG pathway}
+\caption{EGF KEGG pathway.}
 \label{fig:egf_kegg_pathway}
@@
 
-<<actin_pathway,echo=FALSE,results="hide",message=FALSE>>=
-library("pathview")
-pathview(gene.data=plac_path_egid(quote(as.numeric(median_expression))),
-         pathway.id="hsa04810",
-         both.dirs=list(gene=FALSE,cpd=FALSE),
-         limit=list(gene=200,cpd=29),
-         high=list(gene="blue"),
-         net.signature=FALSE)
-@
-
 \begin{figure}
-  \includegraphics{hsa04810.pathview.png}
-  \caption{KEGG Actin Cytoskeleton Pathway}
+<<actin_pathway,echo=FALSE,results="hide",message=FALSE,warning=FALSE>>=
+plot_pathway("hsa04810",pathcolor=quote(log10(median_expression+1)),cutpoint=0.1)
+#library("pathview")
+# pathview(gene.data=plac_path_egid(quote(as.numeric(median_expression))),
+#          pathway.id="hsa04810",
+#          both.dirs=list(gene=FALSE,cpd=FALSE),
+#          limit=list(gene=200,cpd=29),
+#          high=list(gene="blue"),
+#          net.signature=FALSE)
+@
+
+%  \includegraphics{hsa04810.pathview.png}
+  \caption{KEGG Actin Cytoskeleton Pathway. Expression in $\log_{10}$ FPKM.}
   \label{fig:kegg_actin}

### 2017-03-09 manuscript/manuscript.Rnw
fix minor issues; add acknowledgements

@@
 Even though the placenta's primary function is conserved among
-mammals, it varies widely in terms of shape, degree
+mammals, the placenta varies widely in terms of shape, degree
 of intimacy between fetal tissues and maternal tissues, and degree of
@@
 
-To address our questions, we 
+To address our questions we
 sequenced and quantified the total RNA of mammalian placentas at term,
-including \human, \bonobo (the sister
-species to \textit{Pan troglodytes}), \smonkey,
-\mouse, \galili, \carmeli, \dog, \cow, \elephant,
-\armadillo, and \opossum.
+including \human [human], \bonobo [bonobo], (the sister
+species to \textit{Pan troglodytes}), \smonkey [spider monkey],
+\mouse [mouse], \galili [blind mole rat from Upper Galilee], \carmeli [blind mole rat from northern Israel], \dog [domestic dog], \cow [domestic cow], \elephant [African savanna elephant],
+\armadillo [nine-banded armadillo], and \opossum [gray short-tailed opossum].
 These species represent four major branches of eutherian mammals
 (Euarchontoglires, Laurasiatheria, Afrotheria, and Xenartha) as well
-as marsupials. These placentas also represent a wide array of
+as marsupials (\cref{fig:species_tree}). These placentas also represent a wide array of
 placental morphologies as well as traits such as number and size of
-offspring and length of gestation (\cref{tab:gestational_periods}). We
+offspring and length of gestation (Supplemental \cref{tab:gestational_periods}). We
 use these data, along with existing RNA-Seq data from \mouse 
@@
 \cite{Hughes.ea2015:Evaluatingintra-andinter-individual},
-\horse
+\horse [horse]
 \cite{Coleman.ea2010:Structuralannotationofequine}, \sheep 
 \cite{Archibald.ea2010:sheepgenomereferencesequence}, and
-\pig \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
+\pig [domestic pig] \cite{Ahanda.ea2012:Predictionofaltered3-} to: 1)
 identify the non-housekeeping genes which are expressed in the
@@
 \textbf{core functionality of the placenta}; 2) identify genes whose
-expression level changes significantly in each of the morphology
+expression level changes significantly in each of the placenta morphology
 types.
@@
 diffuse placenta for \opossum). The placenta membranes were separated
-from the fetal villous portion using forceps and a scapel, and the
+from the fetal villous portion by dissection, and the
 resulting fetal villous portion was washed twice in PBS. All tissues
@@
 which most closely aligns to each expressed annotated
-protein-generating sequence of all species analyzed using DIAMOND
+protein-generating sequence of all species analyzed using Diamond
 \cite{Buchfink.ea2015:Fastandsensitiveprotein}.
@@
 species having separate random effects. Shifts in expression levels
-are reported with $\log_2$ Fold-Change (FC), rounded to the nearest
-integer.
+are reported in $\log_2$ Fold-Change (FC).
 
@@
 All of the code used to produce every analysis presented in this paper
-is available at
+and the manuscript itself is available at
 \url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
@@
 %
-Between 20\% and 40\% of all highly expressed genes in each species
+Between 20\% and 37\% of all highly expressed genes in each species
 were ortholog pairs with all other analyzed species, with the
@@
 genes were almost certainly present in the placenta of the most recent
-common mammalian ancestor included in this study. We hypothesize that
-modifications to their expression or function will compromise the
+common mammalian ancestor of the species included in this study. We hypothesize that
+modifications to the expression or function of genes in the core placenta transcriptome
+will compromise the
 development and organization of the placenta.
@@
   publicly available data contributed to the completeness of this
-  work.
+  work, as well as the suggestions of the anonymous reviewers which
+  improved this paper and the advice of Monica Uddin in preparing this
+  manuscript.
 

### 2017-03-17 manuscript/manuscript.Rnw
use 13 eutherians instead of 14 of 14
include precise numbers for gene trees
 - Calculate permissive gene tree as well (FPKM ≥ 5)
 - Make sure to add back in the human name for humans even if there
   isn't a human species

@@
 
-<<load_core_placenta_tree_transcriptome,echo=FALSE,results="hide",cache.extra=file.info("../data/placenta_core_tree_transcriptome")[,"mtime"]>>=
+<<load_core_placenta_tree_transcriptome,echo=FALSE,results="hide",cache.extra=file.info(paste0("../data/placenta_core_tree_transcriptome",c("","_permissive")))[,"mtime"]>>=
 load("../data/placenta_core_tree_transcriptome")
+load("../data/placenta_core_tree_transcriptome_permissive")
 cpttl <-
@@
 \Sexpr{cpt.distribution.shape[`Expression (FPKM)`>=1 & `Min. Exp. in`
-  == '14 of 14 Species',max(num_genes)]} %
+  == 'All 13 Eutherians',max(num_genes)]} %
 genes in the core transcriptome. %
@@
 \Sexpr{cpttl[,length(unique(tree_genes))]} unique gene trees which
-include nearly all of the genes present in the core placenta
-transcriptome (Supplemental
+include most
+(\Sexpr{length(intersect(core.placenta.tree.human.genes.in.tree,
+  core.placenta.transcriptome.long[,as.character(unique(human_name))]))/
+  core.placenta.transcriptome.long[,length(unique(human_name))]*100}\%)
+of the genes present in the core placenta transcriptome (Supplemental
 \cref{fig:core_placenta_tree_transcriptome,fig:core_placenta_tree_transcriptome_housekeeping}).
+Relaxing the mean gene tree expression requirement to
+$\text{FPKM} \ge 5$ includes nearly all of the genes
+(\Sexpr{length(intersect(core.placenta.tree.human.genes.in.permissive.tree,
+  core.placenta.transcriptome.long[,as.character(unique(human_name))]))/
+  core.placenta.transcriptome.long[,length(unique(human_name))]*100}\%).
 

### 2017-03-17 manuscript/supplemental_material.Rnw
include precise numbers for gene trees
 - Calculate permissive gene tree as well (FPKM ≥ 5)
 - Make sure to add back in the human name for humans even if there
   isn't a human species

@@
                                                  Species) & housekeeping==FALSE,]
-cpttl <- cpttl[,Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
+cpttl <- cpttl[!is.na(tree_genes),Species:=factor(gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",capfirst(Species)),
                                 levels=gsub("^([^ ])[^ ]* +([^ ]*)","\\1. \\2",

### 2017-03-17 data/combine_fpkm.R

@@ -123,6 +123,8 @@ if (!is.null(non.1.1.orthologs)) {
                human_name:=NA]
 }
 
+gene.fpkms[species=="homo sapiens",human_name:=gene_short_name]
+
 ### this shouldn't be required, but I've triggered some bug in the
 ### data.table code
 combined.fpkm <- data.table(data.frame(gene.fpkms))

### 2017-03-17 data/placenta_core_tree_transcriptome_permissive.R

@@ -113,7 +113,7 @@ setkey(combined.fpkm.wide,"tree_genes")
 
 ## all genes which are not housekeeping genes which have minimum 
 core.placenta.tree.transcriptome.long <- 
-    data.table(melt(combined.fpkm.wide[all_but_metatherian_expression >= 10,
+    data.table(melt(combined.fpkm.wide[all_but_metatherian_expression >= 5,
                                        ][order(-median_expression)
                                          ],
                     id.vars=c("tree","tree_genes"),
@@ -138,7 +138,7 @@ core.placenta.tree.transcriptome.long[,tree_genes:=
                                                                       ][core.placenta.tree.transcriptome.long[,as.character(tree_genes)],
                                                                             median_expression])]
 
-core.placenta.tree.human.genes.in.tree <-
+core.placenta.tree.human.genes.in.permissive.tree <-
     combined.fpkm[tree %in% as.character(core.placenta.tree.transcriptome.long[,unique(tree)]),
                   unique(human_name)]
 
@@ -183,10 +183,7 @@ pts <- pts[,housekeeping:=all.trees %in% housekeeping.genes.superset[,gene_short
 pts.tree <- pts[,expressed.housekeeping:=all.trees %in% expressed.housekeeping.genes]
 
 
-save(core.placenta.tree.transcriptome.long,
-     core.placenta.tree.transcriptome.genes.shape,
-     core.placenta.tree.human.genes.in.tree,
-     pts.tree,
+save(core.placenta.tree.human.genes.in.permissive.tree,
      file=args[length(args)])
 
 

### 2017-03-17 common_r_code/simple_grange.R
add simple grange and stat ellipse

@@ -1,0 +1,7 @@
+library("GenomicRanges")
+simple.grange <- function(chr,start,end) {
+    return(GRanges(seqnames=Rle(paste0("chr",gsub("^chr","",as.character(chr))),
+                       rep(1,length(chr))),
+                   ranges=IRanges(start=start,
+                       end=end)))
+}

### 2017-03-17 common_r_code/stat-ellipse.R

@@ -1,0 +1,37 @@
+require(proto)
+
+StatEllipse <- proto(ggplot2:::Stat,
+	{
+		required_aes <- c("x", "y")
+		default_geom <- function(.) GeomPath
+		objname <- "ellipse"
+
+		calculate_groups <- function(., data, scales, ...){
+			.super$calculate_groups(., data, scales,...)
+		}
+		calculate <- function(., data, scales, level = 0.75, segments = 51,...){
+      dfn <- 2
+      dfd <- length(data$x) - 1
+      if (dfd < 3){
+      	ellipse <- rbind(c(NA,NA))	
+      } else {
+          require(MASS)
+          v <- cov.trob(cbind(data$x, data$y))
+          shape <- v$cov
+          center <- v$center
+          radius <- sqrt(dfn * qf(level, dfn, dfd))
+          angles <- (0:segments) * 2 * pi/segments
+          unit.circle <- cbind(cos(angles), sin(angles))
+          ellipse <- t(center + radius * t(unit.circle %*% chol(shape)))
+      }
+    
+      ellipse <- as.data.frame(ellipse)
+      colnames(ellipse) <- c("x","y")
+      return(ellipse)
+		}
+	}
+)
+
+stat_ellipse <- function(mapping=NULL, data=NULL, geom="path", position="identity", ...) {
+  StatEllipse$new(mapping=mapping, data=data, geom=geom, position=position, ...)
+}

### 2017-03-17 manuscript/acronyms.tex
add apob and afp acronyms; remove psg and pag acronyms

@@
 \DeclareAcronym{GH}{short=GH,long=Growth Hormone}
-\DeclareAcronym{PSG}{short=PSG,long={Pregnancy-Specific Glycoprotein}}
-\DeclareAcronym{PAG}{short=PAG,long={Pregnancy-Associated Glycoprotein}}
+%\DeclareAcronym{PSG}{short=PSG,long=Pregnancy-Specific Glycoprotein}
+%\DeclareAcronym{PAG}{short=PAG,long=Pregnancy-Associated Glycoprotein}
 \DeclareAcronym{GO}{short=GO,long={Gene Ontology}}
@@
 \DeclareAcronym{CD9}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CD9}{\textit{CD9}},long={CD9 molecule},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=CD9}{\textit{CD9}} (CD9 molecule)}}
+\DeclareAcronym{AFP}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=AFP}{\textit{AFP}},long={alpha fetoprotein},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=AFP}{\textit{AFP}} (alpha fetoprotein)}}
+\DeclareAcronym{APOB}{short=\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=APOB}{\textit{APOB}},long={apolipoprotein B},first-style=reversed,single={\href{http://www.genenames.org/cgi-bin/gene_symbol_report?match=APOB}{\textit{APOB}} (apolipoprotein B)}}
 

### 2017-03-17 manuscript/manuscript.Rnw
line break on -
move where acronyms are defined
add derek's suggested changes; more completely discuss ERV capture

@@
 \do\)\do\,\do\?\do\&\do\'\do+\do\=\do\#%
-\do\/%
+\do\/\do\-%
 }%
@@
 \emergencystretch=0.1pt
+
+% Acronyms
+\input{acronyms}
 
@@
 @
-
-% Acronyms
-\input{acronyms}
 
@@
 These observations lead to two related questions: 1) what are the
-genes which are responsible for the \textbf{core functionality of the
+genes which are responsible for the \textbf{core functions of the
   placenta} and 2) which genes are associated with \textbf{changes in
@@
 For example, each superorder (Euarchontoglires, Laurasiatheria,
-Afrotheria, and Xenartha, \cref{fig:species_tree})
-has evolved expansive placenta-specific gene
-families with expression patterns that can vary in space and time
+Afrotheria, and Xenartha, \cref{fig:species_tree}) has evolved
+expansive placenta-specific gene families with expression patterns
+that can vary in space and time
 \cite{Green.ea2000:Pregnancy-associatedbovineandovine}. These include
 independent expansions of the prolactins in \mouse, \carmeli, \galili,
-and \cow, \acp{PSG} in \human and \mouse, \acp{PAG} in \cow, genes
-related to the thymus-specific serine protease \ac{PRSS16} in
-\elephant, and the expansion of the growth hormone and
-chorionic gonadotropin beta genes in anthropoid primates
+and \cow, Pregnancy-Specific Glycoproteins in \human and \mouse,
+Pregnancy-Associated Glycoproteins in \cow, genes related to the
+thymus-specific serine protease \ac{PRSS16} in \elephant, and the
+expansion of the growth hormone and chorionic gonadotropin beta gene
+families in anthropoid primates
 \cite{Haig2008:Placentalgrowthhormone-relatedproteins,Rawn.ea2008:evolutionregulationandfunction}.
@@
 accident. Placental tissue from \opossum was sampled at fetal stage
-32, roughly one day prior to birth. Only the fetal villous portion of
+32, roughly one day prior to birth. Only the fetal portion of
 the placenta was sampled at the maternal-fetal interface (i.e. villous
@@
 diffuse placenta for \opossum). The placenta membranes were separated
-from the fetal villous portion by dissection, and the
-resulting fetal villous portion was washed twice in PBS. All tissues
+from the fetal portion by dissection, and the
+resulting fetal portion was washed twice in PBS. All tissues
 were collected and stored in RNAlater (Life Technologies, Carlsbad,
@@
 \cite{Dobin.ea2013:STARultrafastuniversalRNA-seq} (Supplemental
-\cref{fig:transcriptome_statistics,fig:transcriptome_reads_vs_read_length})
+\cref{fig:transcriptome_statistics,fig:transcriptome_reads_vs_read_length,tab:ref_genome_aligned_species})
 and expression values in \ac{FPKM}
@@
 obtained orthology from the perspective of each species to all other
-species using biomaRt package and Ensembl v.80 database to determine
-the percentage of highly expressed genes which had an orthology of
+species using the biomaRt package and the Ensembl v.80 database to determine
+the percentage of highly expressed genes that had an orthology of
 one-to-one, one-to-many, many-to-many, or novel/unknown. For future
@@
 \url{https://github.com/uiuc-cgm/placenta-viviparous.git}. All of the
-sequences used are deposited in SRA, and new sequences for this work
-can be obtained under the series
+sequences and assembled transcriptomes used are deposited in SRA, and
+new sequences for this work can be obtained under the series
 \href{http://www.ncbi.nlm.nih.gov/geo/query/acc.cgi?acc=GSE79121}{GSE79121}
-(Supplemental \cref{tab:srr_numbers}). The mean \ac{FPKM} for gene in each
-species with ortholog annotations are located in the supplemental data
-file all\_species\_mean\_fpkm.txt.xz and the per-sample \ac{FPKM} values are in
-all\_species\_per\_sample\_fpkm.txt.xz.
+(Supplemental \cref{tab:srr_numbers}). The mean \ac{FPKM} for gene in
+each species with ortholog annotations are located in the supplemental
+data file all\_species\_mean\_fpkm.txt.xz and the per-sample \ac{FPKM}
+values are in all\_species\_per\_sample\_fpkm.txt.xz.
 
@@
 Multiple components of annexin complexes are included in the core
-transcriptome, including \ac{ANXA2}, \ac{ANXA1},
-\ac{S100A11}, \ac{S100A10}. ANXA1 is involved in the
+transcriptome, including \ac{ANXA1}, \ac{ANXA2},
+\ac{S100A10}, \ac{S100A11}. ANXA1 is involved in the
 resolution of the inflammation response through inhibition of
@@
 \cite{Hutchinson.ea2011:Molecularregulatorsofresolution}. ANXA2 is
-known to interact with S100A10, S100A6, S100A11, and S100A4
+known to interact with S100A4, S100A6, S100A10, and S100A11
 \cite{Liu.ea2015:AnnexinA2complexeswith}, and there is evidence for
@@
 
-While Syncytin-1 and -2 (\ac{ERVW-1} and \ac{ERVFRD-1}, respectively) are implicated
-in the formation of the syncytiotrophoblast in
-\human~\cite{Rawn.ea2008:evolutionregulationandfunction} and genes which are similar to 
-\ac{ERVFRD-1} and \ac{ERVW-1} are widely expressed in most of the
-placentas analyzed here, they are not expressed in \bonobo, \sheep,
-and \dog, even though these mammals form
+The evolutionarily convergent capture of endogenous retroviruses to
+form expressed fusogenic proteins such as Syncytin-1 and -2
+(\ac{ERVW-1} and \ac{ERVFRD-1}, respectively) has been implicated in
+syncytiotrophoblast formation in \human and many other
+species~\cite{Rawn.ea2008:evolutionregulationandfunction}. While genes
+which are similar to \ac{ERVFRD-1} and \ac{ERVW-1} are expressed in
+most of the placentas analyzed here, there are no highly expressed
+genes with sequence similarity in \bonobo and \dog, even though these
+mammals form
 syncytiotrophoblasts~\cite{Mess.ea2006:Evolutionarytransformationsoffetal}.
@@
 Laurasiatheria~\cite{Mess.ea2006:Evolutionarytransformationsoffetal},
-and syncytiotrophoblast must be continuously replenished
-the gene(s) responsible for cytotrophoblast fusion to form
-syncytiotrophoblast should be highly expressed in the core
-placenta. \ac{ANXA5} is known to be expressed at multiple stages of
+and the syncytiotrophoblast must be continuously replenished, the
+gene(s) responsible for cytotrophoblast fusion to form the
+syncytiotrophoblast should be highly expressed in the core placenta.
+\ac{ANXA5} is known to be expressed at multiple stages of
 placentation, and is capable of repairing membrane defects and
@@
 cells~\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing} in
-the presence of Ca$^{2+}$. While there may be other genes which
-are responsible for the positioning of cytotrophoblast cells in close
-proximity to start the fusion process, \ac{ANXA5} is likely to
-be essential for the completion of fusion and the maintenance of
-membrane repair during the massive flux of nutrients and waste in the
-placenta \cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing,Ueki.ea2012:Lossofmaternalannexin}.
+the presence of Ca$^{2+}$. We hypothesize that \ac{ANXA5} is essential
+for the completion of fusion and the maintenance of membrane repair
+during the flux of nutrients and waste in the placenta
+\cite{Carmeille.ea2015:Annexin-A5promotesmembraneresealing,Ueki.ea2012:Lossofmaternalannexin}.
 
@@
 Furthermore, while these samples have been dissected to only include
-the fetal interface, they likely include substantial cell and tissue
+the fetal side, they likely include substantial cell and tissue
 heterogeneity. We are therfore not yet able to assign specific genes
@@
 The core placenta transcriptome contains genes which are involved in
-maternal-fetal tolerance, such as the annexins, and genes responsible
-for cell-cell and cell-matrix interactions, such as \ac{LGALS1}. These
+maternal-fetal tolerance and genes responsible
+for cell-cell and cell-matrix interactions, such as the annexins and \ac{LGALS1}. These
 genes were almost certainly present in the placenta of the most recent

### 2017-03-19 manuscript/manuscript.Rnw
add missing 'and' and left to right to manuscript
clarify what the housekeeping core placenta figure contains

@@
   \cite{Eisenberg.ea2013:Humanhousekeepinggenesrevisited} removed
-  ordered (from top to bottom) by median expression. If we also require
+  ordered (from top to bottom, left to right) by median expression. If we also require
   expression (\ac{FPKM}$ \ge 10$) in the marsupial, \opossum, there are 
@@
 \citet{Eisenberg.ea2013:Humanhousekeepinggenesrevisited}. Genes which
-meet this criteria but are housekeeping genes are depicted in
+meet the FPKM criterion and are housekeeping genes are depicted in
 Supplemental \cref{fig:core_placenta_transcriptome_housekeeping}.
@@
 transcriptome, including \ac{ANXA1}, \ac{ANXA2},
-\ac{S100A10}, \ac{S100A11}. ANXA1 is involved in the
+\ac{S100A10}, and \ac{S100A11}. ANXA1 is involved in the
 resolution of the inflammation response through inhibition of
